### 2022-02-03 article/article.tex
add current writings

\documentclass[10pt]{article}
\renewcommand{\baselinestretch}{1.8}

\usepackage[b4paper,left=0.8in,right=0.8in,top=0.8in,bottom=0.8in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{showkeys}
\usepackage{amsmath}

\usepackage{multicol}
\setlength\columnsep{24pt}

\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

\algdef{S}[FUNCTION]{Function}
   [3]{{\texttt{\textsl{#1}}} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \texttt{\textproc{#1}}}

\algrenewcommand\Call[2]{\texttt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

\newcommand\keywordfont{\sffamily\bfseries}
\algrenewcommand\algorithmicend{{\keywordfont end}}
\algrenewcommand\algorithmicfor{{\keywordfont for}}
\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
\algrenewcommand\algorithmicdo{{\keywordfont do}}
\algrenewcommand\algorithmicuntil{{\keywordfont until}}
\algrenewcommand\algorithmicfunction{{\keywordfont function}}
\algrenewcommand\algorithmicif{{\keywordfont if}}
\algrenewcommand\algorithmicthen{{\keywordfont then}}
\algrenewcommand\algorithmicelse{{\keywordfont else}}
\algrenewcommand\algorithmicreturn{{\keywordfont return}}

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

\title{Wait-free Queues with Polygarithmic Step Complexity}
\date{\today}

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\begin{document}
\maketitle

%\begin{abstract}
%This is the paper's abstract \ldots
%\end{abstract}

\section{Previous Work}
\begin{table}[hbt]
  \begin{center}
  \begin{tabular}{l|c|c|c}
    & Type & Progress Property  & Complexity \\ \hline
    MichaelS96~\cite{DBLP:conf/podc/MichaelS96} & Singly Linked List & Lock-free & $\Omega(p)$ \\ \hline
    MoirNSS05~\cite{DBLP:conf/spaa/MoirNSS05} & Linked List with Elemination & & \\ \hline
    HoffmanSS07~\cite{DBLP:conf/opodis/HoffmanSS07} & Linked List with Baskets & & $\Omega(p)$ \\ \hline
    Ladan-MozesS08 ~\cite{DBLP:journals/dc/Ladan-MozesS08} & Doubly linked list & Lock-free & $\Omega(p)$ \\ \hline
    MilmanKLLP18~\cite{DBLP:conf/spaa/MilmanKLLP18} & Linked List with batching same type ops & Lock-free & \\ \hline
    GidenstamST10~\cite{DBLP:conf/opodis/GidenstamST10} & Linked list of arrays & Lock-free & $\Omega(p+len(array))$ \\ \hline
    KoganP11~\cite{DBLP:conf/ppopp/KoganP11} &  Linkedlist with head and tail pointers & Wait-free & $\textsc{O}(p\times\text{bakery-max})$
  \end{tabular}
  \caption{Linearizable ME-MD Queus}
  \end{center}
\end{table}

\section{ Universal Construction using Tournament Tree with Big CAS Objects }
\paragraph{}
Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions. He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses. Here, we first introduce a universal construction using $\textsc{O}(\log p)$ CAS operations~\cite{DBLP:conf/fsttcs/JayantiP05}. We use the universal construction as a stepping stone towards our queue algorithm, so we will not explain it in too much detail.
\paragraph{}
We introduce our universal construction in Algorithm~\ref{alg1} with a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes. Nodes are CAS objects, and each leaf is assigned to one process exclusively. Each CAS object stores a sequence of operations. When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root. The sequence of operations stored in Node $n$'s CAS object shows the order of the operations propagated up to $n$.
The history of operations stored at the root is the linearization ordering. The operation $op$ is linearized when it is appended to the root.

\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}

Leaf $l_i$ stores the sequence of the operations invoked by $P_i$. 
The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$. In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
After a process adding a new operations to its leaf, it has to propagate new operations up to the root. \textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$). In each step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again. If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) finishes.

\begin{figure}

\begin{center}

\begin{subfigure}[b]{.49\textwidth}

  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{Operations after $||$ are new.}

\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}

  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{New operations are added to the parent node.}

\end{subfigure}

\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
\end{center}
\end{figure}

\begin{algorithm}
\caption{Universal Construction Idea}\label{alg1}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
\State \texttt{l\textsubscript{i}.\Call{append}{op}}
\State \Call{Propagate}{parent of l\textsubscript{i}}
\State \texttt{Run} \textsf{the sequence stored in root}
\State \texttt{\Return op}\textsf{'s response from line 4}
\EndFunction{Do}

\Statex

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n==root}} \Return
\ElsIf{!\Call{Refresh}{n}}
\State \Call{Refresh}{$n$} \EndIf
\State \Call{Propagate}{parent of n}
\EndFunction{Propagate}

\columnbreak

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{old=} \Call{Read}{n}
\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
\State \texttt{new= old$\cdot$new}
\State \Return n.\Call{CAS}{old, new}
\EndFunction{Refresh}

\end{multicols}
\end{algorithmic}
\end{algorithm}

$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}). However, to compute the result of an operation, we only use the total ordering that is stored at the root. Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable size CAS objects and still achieve polygarithmic step complexity.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
\end{center}
\end{figure}

\section{Block Tree}

%\subsection{Idea in a nutshell}
We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes. First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node. Second, each operation is linearized when its element is appended to the root.

\subsection{Sequence of propagated operations}
\paragraph{}
The basic idea behind the universal construction is to create a linearization of operations invoked by processes. If we design a fast shared data structure in which processes can append elements to a sequence, we can use it to implement practical, fast implementaion of an object $O$ by using the sequence data structure to keep track of the sequence of operations  on $O$. In the following sections, we introduce our solution called Block~Tree.

%\begin{lemma}
%    Consider an instance of \textsc{Propagate(}\textnormal{n}\textsc{)}. When it terminates all operations that were in $n$ before \textsc{Propagate(}\textnormal{n}\textsc{)} will be in the root. \texttt{(not suitable place)}
%\end{lemma}

\subsection{Sequence of Sets of Concurrent Operations \label{subsec::block}}
\paragraph{}
 In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root. However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).

 \begin{figure}[h]
\begin{center}
\begin{tikzpicture}

\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves. Since we linearize operations when they are added to the root, ordering the blocks in the root is important.}
\end{center}
\end{figure}

The definition of linearizability allows concurrent operations to be reordered arbitrarily. Thus, a group of concurrent operations can be appended to our root sequence as one block without specifying the order among the operations.

%\begin{lemma}
%Operations in the same block can be linearized in any order.
%\end{lemma}

\subsection{Using arrays instead of big CAS Objects}
\paragraph{}
 We used unbounded CAS objects storing sequences as big words in the universal construction. One can represent sequences as arrays to overcome this implementation problem. Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.

\subsection{Augmenting Tree to make \textsc{Refresh}() Step faster}
\paragraph{}
Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied. This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations. Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set. Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)

\begin{figure}

\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]

\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
 [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
 [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:block} Showing concurrent operation sets with blocks. Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively. Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child. Blocks between two lines $||$ are propagated together to the parent. For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent. The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.}
\end{figure}

Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$. For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) also points to elements from blocks (0,5) and (1,2). 
%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.

\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
\end{definition}

\begin{definition}\{Subblock\}
  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$. Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
\end{definition}
%{\TODO: describe how blocks work with a good example how exactly it is done}

\paragraph{}
We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
%\begin{lemma}
%\label{lem:block_size}
%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
%\end{lemma}
%\begin{proof}
%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
%\end{proof}

\paragraph{}
In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes. So in a complete propagation, at most $2p$ blocks are merged into one block. \texttt{(maybe useful for analysis)}
%\begin{lemma}
%A block in the root cannot contain more than one operation from a process.
%\end{lemma}
%\begin{proof}
%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
%\end{proof}

\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
%\begin{definition}
%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
%\end{definition}
\paragraph{}
\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence. We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element. To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.

Furthermore, in each block, we store the prefix sum of left and right elements. Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).

\begin{figure}
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]

\end{tikzpicture}
\end{center}
\caption{\label{fig:prefix} Using Prefix sums in blocks. When we want to find block b elements in its children, we can use binary search. The number below each block shows the count of elements in the previous blocks.}
\end{figure}

\begin{figure}[hbt]
\centering
  \includegraphics[width=5in]{pics/pointers}
  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
\end{figure}

\paragraph{}
Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$. From lemma $\ref{lem:block_size}$ we know a block size is at most $p$. So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.

\subsection{Block Tree Algorithm}
\paragraph{}
Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements. It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.

\paragraph{Design of a block tree}

Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
Each internal node of the tree contains an array of blocks of elements.
Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks. Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order. Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively. Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).

In order to compute elements of a block faster we store prefix-sum blocks(block i has tuple(right-sum=$\#$right ops in previous block, left-sum=$\#$left ops in previous blocks)[See Figure \ref{fig:prefix}]. Here is the algorithm to get elements of a block.

\paragraph{Specification}
A block tree is a shared data structure that stores a sequence of elements.  It has two methods \texttt{Append(e)} and \texttt{Get(i)}. \texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence. \texttt{Get(i)} returns \texttt{i}th element stored in the sequence.

\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.

\paragraph{Membership}
Element \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
\end{itemize}

\paragraph{Order of elements inside node}
Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
\begin{itemize}
  \item The block containing d is before the block containing e.
  \item e and d are in the same block and d is in the left child and e is in the right child.
  \item d is before e in the same child's order.
\end{itemize}

\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/tree}
  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
\end{figure}

\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children. \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block. Other attributes are assigned values followed by definition.
\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
\end{figure}

%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}

\paragraph{Double Refresh}
Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
\begin{proof}
\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block. If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence. So suppose two \texttt{Refresh}es fail. Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}. So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}. After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
\begin{proof}
Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks. So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end. Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
\end{proof}

\paragraph{Total Order}
Sequence represented by the Block Tree is the sequence of the blocks stored in the root.

\paragraph{Linearization Points}
\texttt{Get(i)} is linearized when it terminates. \texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.

\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
\begin{proof}
If there are more than $p$ subblocks, then there is more than one block from process pl. \texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9). So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
\end{proof}

\paragraph{Computing \texttt{Get(n, b, i)}}
To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$. Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.

\paragraph{How \texttt{Refresh(n)} works.}
\begin{enumerate}
  \item Read n's counter and head
  \item Create block b
  \item CAS b into n
  \item If previous succeed:
  \begin{enumerate}
  \item Update sup of b's ending subblocks
  \item Increment children's counters
  \end{enumerate}
\end{enumerate}

%\paragraph{Time-Space Relation} There are at most $p^2$ blocks with same \texttt{time} attribute.
%\begin{proof}
%If a process dies at line 28, the children counter won't increase, and each process aggregates most block, and since there are at most p concurrent processes, then there will be at most $p^2$ blocks with same \texttt{time} attribute.
%\end{proof}
%
%\paragraph{Time-Propagate() Relation} In a \texttt{CreateBlock()} procedure, the blocks that are seen have at most 2 different \texttt{time} attribute values.
%
%\paragraph{Time-Super Relation} sup[i+1]-sup[i] is $\leq p$
%
%\paragraph{\texttt{b}'s Superblock position} SuperBlock of \texttt{b} is within range $\pm p$ of the \texttt{\texttt{sup[b.time]}}.
%\begin{proof}
%
%\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=6in]{pics/super}
  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
\end{figure}

\paragraph{Computing superblock}
\begin{enumerate}
  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10. For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
  \end{proof} 
  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
   Follows from the order of lines 37, 26, 27.
 \end{proof}
%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
  \item \texttt{super[i+1]-super[i]}$\leq p$
\begin{proof}
 In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27. Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set. If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27. So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
 \end{proof}
  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
\begin{proof}
\texttt{super[i]} is the index of the superblock of a block containing block b. It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.  \texttt{super(b)} is the real superblock of b. \texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}. If \texttt{b.time} is \texttt{t} we have:
$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
  \end{proof} 
\end{enumerate}

%The block tree contains a tournament tree in which leaves are assigned to processes, and each process appends elements to its leaf and tries to propagate it up to the root. The final sequence is stored at the root. \texttt{Append} after appending to leaf calls \texttt{Propagate} which iterates over the path to root. In each \texttt{Propagate(n)} invocation on node n it tries ti \texttt{Refresh} the sequence stored at node n with the new elements in n's children. After adding the element to the root \texttt{Index(op)} finds the order of the element in the root.

%\textsf{I finally thought maybe a sequence of elements is better for telling specification.}

\begin{algorithm}
\caption{Block Tree \label{alg2}}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex \textbf{Structure}

\Statex $\blacktriangleright$ \texttt{\textsl{element} e}
\begin{itemize}
%\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}.ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks}\textsf{: blocks stored in \texttt{n}}
\item \texttt{\textsl{int} head= 1}\textsf{: index of first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{}
\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is \texttt{i}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf node} l\textsubscript{i} extends \textsl{node}}
\begin{itemize}
\item \texttt{\textsl{operation[]} ops}\textsf{: elements that are invoked to \texttt{Append()} to the blok tree by process i}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, sum\textsubscript{left}}

  \textsf{~~\#operations from the left subblocks of b, prefix sum of num\textsubscript{left}}

  \item \texttt{\textsl{int} num\textsubscript{right}, sum\textsubscript{right}}

  \textsf{~\#operations from the right subblocks of b, prefix sum of num\textsubscript{right}}

  \item \texttt{\textsl{int} sum}

  \textsf{\# operation in b}

  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}

  \textsf{~~index of \texttt{b}'s last subblock}
  \item \texttt{\textsl{int} time}

\end{itemize}

%\Statex \textbf{Examples}
%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

\Statex

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.head}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.time= op.loc}
\State \texttt{b.sum= 1}
\State \texttt{l\textsubscript{pid}.ops[op.loc]= op}
\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
\State \texttt{op.head+= 1}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.loc, b} 
\EndFunction{Append}

\columnbreak

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\If{\texttt{n {\keywordfont is} root}} \Return
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

\Statex

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{i= n.head}
\State \texttt{c= n.counter}
\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, c}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \Return{\texttt{true}}
\Else 
\State \Return{ \texttt{false}}
\EndIf
\State \texttt{\Call{CAS}{n.head, i, i+1}}
\EndFunction{Index}

\Statex

\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int c} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.time= c}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{j= n.dir.head}
\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
\State \texttt{end= n.dir.blocks[j]}
\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].time}
\State \texttt{b.end\textsubscript{dir}= j}
\State \texttt{b.num\textsubscript{dir}= end.sum - start.sum}
\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
\EndFor
\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
\EndFunction{CreateBlock}

\end{multicols}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Block Tree Continued}
\begin{algorithmic}[1]
\setcounter{ALG@line}{46}

\Function{element}{Get}{\textsl{int} i} 
\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.head}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
\EndFunction{Get}

\Statex

\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}

\Statex \Comment{\textmd{Searches~the value \texttt{i} of the given prefix sum \texttt{type} on the node \texttt{n} in the domain from \texttt{start} to \texttt{end} blocks. \texttt{f} is one of \texttt{sum, sum\textsubscript{left}, sum\textsubscript{right}}. Does binary search on \textsl{field} \texttt{f} of the blocks stored in \texttt{n.blocks} and returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
\State \Return \texttt{result block's index}
\EndFunction{BSearch}

\Statex

\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}
\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[i]}
\Else
\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} 
\EndIf
\EndIf
\EndFunction{Get}

\Statex
\Statex $\leadsto$ \textsf{Precondition:~\texttt{i}th operation in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns rank in the  
root of the \texttt{i}th operation in node \texttt{n}.}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].time]-p, sup[n.blocks[b].time]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
\EndIf
\State \Return\Call{Index}{n.parent, i, superBlock}
\EndIf
\EndFunction{Index}

\Statex

%\Function{list}{GetElements}{block b} 
%\ForEach{block {\keywordfont in} \Call{GetSubBlocks}{b}}
%\State result.append(\Call{GetElements}{})
%\EndFor
%\State \Return result
%\EndFunction{GetElements}
%
%\Statex
%
%\Function{list}{GetSubBlocks}{b}
%\State n:=b.node
%\State b[-1]=b's previous block
%\ForEach{direction \{left, right\}}
%\State init=\Call{BSearch}{n.direction, b[-1].direction}
%\State end=\Call{BSearch}{n.direction, b.direction}
%\State result.append([init:end])
%\EndFor
%\State \Return result
%\EndFunction{GetSubBlocks}

\end{algorithmic}
\end{algorithm}

\section{Implementing Queue using Block Tree}
\subsection{Idea in a nutshell}
With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log ^2p+\log n)$ steps. A queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence. If the queue is empty it returns \texttt{null}. Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.  So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
%
%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{}
Next, we describe how to use block tree to implement queues. The block tree, maintains the history of all operations, not only the current state of the queue. Now consider the following history of operations. What should each \texttt{Dequeue()} return?

\begin{table}[hbt]
\centering
  \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
    \hline \texttt{ENQ(5)}& \texttt{ENQ(2)}& \texttt{DEQ()}& \texttt{ENQ(3)}&\texttt{DEQ()}& \texttt{DEQ()}& \texttt{DEQ()}& \texttt{ENQ(4)}& \texttt{ENQ(6)}& \texttt{DEQ()}\\ \hline
  \end{tabular}
  \caption{An example histoy of operations on the queue}
\end{table}

\begin{definition}
A non-null dequeue is one that returns a non-null value.
\end{definition}

\paragraph{}
In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order. Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null. If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}. So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.

%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}

%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}

\paragraph{}
In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism. So now the problem is to find the result of each Dequeue. From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations. In the next example, operations in a cell are concurrent. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

\begin{table}[hbt]
\centering
  \begin{tabular}{c|c|c|c}
    \hline \texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
  \end{tabular}
  \caption{An example history of operation blocks on the queue}
\end{table}

\paragraph{}
Now, we claimed that by knowing the current size of the queue and the number of non-null dequeue operations before the current dequeue, we could compute the index of the resulting \texttt{Enqueue()}. We apply this approach to blocks; if we store the size of the queue after each block of operations happens and the number of non-null dequeues dequeues till a block, we can compute each dequeue's index of result in \textsc{O}$(1)$ steps.

\begin{table}[hbt]
\centering
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
                size & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
  \caption{Augmented history of operation blocks on the queue}
\end{table}

%\begin{definition}
%  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
%  
%  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
%  
%\end{definition}

Size and the number of non-null dequeues for $b$th block could be computed this way:\\
\texttt{size[b]= max(size[b-1] +enqueues[b] -dequeues[b], 0)}\\
\texttt{non-null dequeues[b]= non-null dequeues[b-1] +dequeues[b] -size[b-1] -enqueues[b]}

Given \texttt{DEQ} is in block \texttt{b}, \texttt{response(DEQ)} would be:\\
\texttt{(size[b-1]- index of DEQ in the block's dequeus >=0) ? ENQ[non-null dequeus[b-1]+ index of DEQ in the block's dequeus] : null;}

\section{Main Algorithm}

\paragraph{Specification}
A Queue is a shared data structure that stores a sequence of elements. It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}. \texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence. \texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.

%Now that we know how to augment Block Tree root history to compute the response, we will present the complete algorithm. \texttt{Enqueue()} and \texttt{Dequeue()} are the main functions. \texttt{Enqueue(e)} adds \texttt{e} to the queue and \texttt{Dequeue()} returns the first element in the queue and removes it from the queue. Block Tree functionalities are very similar. \texttt{Append(e)} does what \texttt{Enqueue()} does and if we know the current index of the head of the queue we can call \texttt{Get()} to find its value.

\begin{figure}[hbt]
\centering
  \includegraphics[width=5in]{pics/queue}
  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
\end{figure}

\begin{algorithm}
\caption{Queue \label{algQ}}
\begin{algorithmic}[1]
\begin{multicols}{2}

%\Statex \textbf{Structure}

\Statex $\blacktriangleright$ \texttt{\textsl{Operation}}
\begin{itemize}
%\item \texttt{\textsl{string} name}
%\item \texttt{\textsl{string} type$\in$\{DEQ,ENQ\}}
\item \texttt{\textsl{Object} element}\textsf{: argument of an enqueue, if null this is a dequeue}
%\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in the leaf's \texttt{ops}}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{Node}}
\begin{itemize}
\item \texttt{\textsl{*Node} left, right, parent}
\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0}
\item \texttt{\textsl{int} head= 1}\textsf{: index of the first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{}
\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is field \texttt{i}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{Leaf} extends \textsl{Node}}
\Statex \Comment{\textsf{\texttt{l\textsubscript{i}} is the Leaf for process \texttt{i} and \texttt{i} is in range \texttt{0} to \texttt{p-1}}}
\begin{itemize}
\item \texttt{\textsl{Operation[]} ops}\textsf{: invoked operations}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{Root} extends \textsl{Node}}
\begin{itemize}
\item \texttt{override \textsl{Root Block[]} blocks}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{Block}}

\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}

  \textsf{~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}

  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}

  \textsf{~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}

  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}

  \textsf{~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}

  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}

  \textsf{~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}

  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}

  \textsf{\# enqueue, dequeue operations in the block}

  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}

  \textsf{sum of \# enqueue, dequeue operations in blocks up to this one}

  \item \texttt{\textsl{int} num, sum}

  \textsf{total \# operations in block, prefix sum of \texttt{num}}

  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}

  \textsf{~~index of this block's last subblock in the left and right child}
  \item \texttt{\textsl{int} group}

\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{Root Block} extends \textsl{Block}}
\begin{itemize}
  \item \texttt{\textsl{int} size}

  \textsf{size of queue after this block's operations finish}

  \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}

  \textsf{count of non-null dequeus up to this block}

\end{itemize}

%\Statex \textbf{Examples}
%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

\Statex

\Function{void}{Enqueue}{\textsl{Object} e}
\State \texttt{op= NEW(\textsl{operation})}
%\State \texttt{op.type= ENQ}
\State \texttt{op.element= e}
%\State \texttt{op.pid= this.pid}
\State \Call{Append}{op}
\EndFunction{Enqueue}

\Statex

\Function{Object}{Dequeue()}{}
\State \texttt{op= NEW(\textsl{operation})}
%\State \texttt{op.type= DEQ}
%\State \texttt{op.pid= this.pid}
\State  \Call{Append}{op} \Comment{$i$th DEQ in $b$th block}
\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, op.loc, 1}
\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned}
\If{\texttt{res }\textbf{is }\texttt{null}}
\State \Return \texttt{null}
\Else
\State \Return \Call{Get}{res}
\EndIf
\EndFunction{Dequeue}

\Statex

\Function{int}{ComputeHead}{int i, int b}\Comment{Computes head of the queue when \texttt{i}th dequeue in \texttt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
\If{\texttt{root.blocks[b-1].size + root.blocks[b].num\textsubscript{enq} - i $<$ 0}}
\State \Return \texttt{-1}
\Else{ \Return \texttt{root.blocks[b-1].sum\textsubscript{non-null deq} + i}}
\EndIf
\EndFunction{ComputeHead}

\Statex

\Function{void}{Append}{\textsl{operation} op} 
\State \texttt{pid= id of process performing \textsc{Append}}
\State \texttt{op.loc= l\textsubscript{pid}.head}
\State \texttt{op.head+= 1}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.group= op.loc}
\If{\texttt{op.element==null}} 
\texttt{b.sum\textsubscript{deq}=1} \Else \texttt{~b.sum\textsubscript{enq}=1} \EndIf
%\State \texttt{b.sum\textsubscript{op.type}= 1}
\State \texttt{l\textsubscript{pid}.ops[op.loc]= op} \label{addOP}
\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
\State \Call{Propagate}{l\textsubscript{pid}.parent} 
\EndFunction{Append}

\end{multicols}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Queue Continued}
\begin{algorithmic}[1]
\setcounter{ALG@line}{33}
\begin{multicols}{2}

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\If{\texttt{n.parent \textbf{is} null}} 
\State \Call{Propagate}{n.parent}
\EndIf
\EndFunction{Propagate}

\Statex

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{h= n.head}
%\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
\State \texttt{t= n.counter}
\State \texttt{<new, t\textsubscript{left}, t\textsubscript{right}>= \Call{CreateBlock}{n, h, t}}
\If{\texttt{new.num==0}} \Return{\texttt{true}}
\ElsIf{\Call{CAS}{n.blocks[h], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[t\textsubscript{dir}], null, h+1}}
\State \texttt{\Call{CAS}{n.dir.counter, t\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{\texttt{true}}
\Else
\State \texttt{\Call{CAS}{n.head, h, h+1}}
\State \Return{ \texttt{false}}
\EndIf
\EndFunction{Refresh}

\Statex

\Function{element}{Get}{\textsl{int} i}
\Comment{Returns $i$th Enqueue.}
\State \texttt{res= \Call{BSearch}{root, sum\textsubscript{enq}, i, 0, root.head}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
\EndFunction{Get}

\Statex

\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}

\Statex \Comment{\textmd{Does binary search for~the value \texttt{i} of the given prefix sum \texttt{feild}. \texttt{f} is one of \texttt{sum, sum\textsubscript{left}, sum\textsubscript{right}}. Returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
%\State \Return \texttt{result block's index}
\EndFunction{BSearch}

\pagebreak

\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i, int t} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]} with time field=\texttt{t}. Returns the created block as well as values read from each child counter feild.}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.group= t}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{lastIndex= n.dir.head} \label{lastLine}
\State \texttt{prevIndex= n.blocks[i-1].end\textsubscript{dir}} \label{prevLine}
\State \texttt{lastBlock= n.dir.blocks[lastIndex]}
\State \texttt{prevBlock= n.dir.blocks[prevIndex]}
\State \texttt{t\textsubscript{dir}= n.dir.counter}
\State \texttt{b.end\textsubscript{dir}= lastIndex}
\State \texttt{b.num\textsubscript{enq-dir}= lastBlock.sum\textsubscript{enq} - prevBlock.sum\textsubscript{enq}}
\State \texttt{b.num\textsubscript{deq-dir}= lastBlock.sum\textsubscript{deq} - prevBlock.sum\textsubscript{deq}}
\State \texttt{b.sum\textsubscript{enq-dir}= n.blocks[i-1].sum\textsubscript{enq-dir} + b.num\textsubscript{enq-dir}}
\State \texttt{b.sum\textsubscript{deq-dir}= n.blocks[i-1].sum\textsubscript{deq-dir} + b.num\textsubscript{deq-dir}}
\EndFor
\State \texttt{b.num\textsubscript{enq}= b.num\textsubscript{enq-left} + b.num\textsubscript{enq-right}}
\State \texttt{b.num\textsubscript{deq}= b.num\textsubscript{deq-left} + b.num\textsubscript{deq-right}}
\State \texttt{b.num= b.num\textsubscript{enq} + b.num\textsubscript{deq}}
\State \texttt{b.sum= n.blocks[i-1].sum + b.num}

\If{\texttt{n.parent} \textbf{is} \texttt{null}}
\State \texttt{b.size= max(root.blocks[i-1].size + b.num\textsubscript{enq} - b.num\textsubscript{deq}, 0)}
\State \texttt{b.sum\textsubscript{non-null deq}= root.blocks[i-1].sum\textsubscript{non-null deq} + max( b.num\textsubscript{deq} - root.blocks[i-1].size - b.num\textsubscript{enq}, 0)}
\EndIf

\State \Return \texttt{b, t\textsubscript{left}, t\textsubscript{rightt}}
\EndFunction{CreateBlock}

\end{multicols}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Queue Continued}
\begin{algorithmic}[1]
\setcounter{ALG@line}{83}

\Statex $\leadsto$ \textsf{Precondition:~\texttt{n.blocks[b]} contains $\geq$\texttt{i} enqueues.}
\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
\Else
\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\textsubscript{enq}} 
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{enq-left}}
\State\texttt{subBlock=\Call{BSearch}{n.right, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{right}+1, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\textsubscript{enq}} 
\EndIf
\EndIf
\EndFunction{Get}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \texttt{b}th block of node \texttt{n} has propagated up to the root and \texttt{i}th dequeue resides in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks[b-1]+i, b}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq-right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq} + n.blocks[superBlock].sum\textsubscript{deq-left}}
\EndIf
\State \Return\Call{Index}{n.parent, superBlock, i}
\EndIf
\EndFunction{Index}

\end{algorithmic}
\end{algorithm}

\subsection{Pseudocode description}

\paragraph{Tournament Tree}
In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree. Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}. Each process adds $op$ to its leaf. In each internal node an ordering of operations in its subtree is stored. All processes agree on the total ordering of all operations stored in the root. This ordering will be the linearization of the operations.
\paragraph{Implicit Storing Blocks}
For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks. A Block is a constant size object that implicitly represents a sequence of operations. In each node there is an array of Blocks.

\begin{definition}[Block]
A block is an object that stores some statistics described in Algorithm Queue.   
\end{definition}

\begin{definition}[\label{def::subblock}Subblock]
Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
\end{definition}
Block $b$ contains subblocks in the left and right children. WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$. See Figure \ref{fig::createBlock} .

\begin{definition}[Membership of an operation in a block]
Operation \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
 \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
 \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
\end{itemize}

\end{definition}

We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s. In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}. Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}. There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step. It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases. Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}. This helps us to know which blocks have aggregated together to a block, not precisely though. We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.

In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}. In order to make \texttt{BSearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.

To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}. \texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.

\texttt{Enqueue(e)} just \texttt{append}s an \texttt{operation} with \texttt{element e} to the \texttt{root}. \texttt{Dequeue()} appends an \texttt{operation} to the root and computes its ordering and the \texttt{enqueue operation} containing the head before it calling \texttt{ComputeHead()} and then \texttt{gets} and returns the \texttt{operation}'s element.

\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index. As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times. \texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.

We only want to know the \texttt{element} of \texttt{enqueue} operations and compute ordering for \texttt{dequeue} operations. That's the reason here \texttt{Get()} searches between enqueues only and \texttt{Index()} returns ordering of a dequeue among dequeues. \texttt{Get(n, b ,i)} decides the requested element is in which child of n and continues to search recursively. \texttt{index(n, i, b)} calculates the ordering of the given operation in \texttt{n}'s parent each step and finally returns the result among total ordering.

\subsection{Complexity Analysis}

\texttt{Enqueue()} operations do a constant number of steps and an \texttt{Append()} which calls \texttt{Propagate()}. Let $p$ be the number of processes and $m$ be the count of operations invoked. In a \texttt{Propagate()} step there are a constant number of steps taken at each level and the height of the tree is $\log(p)$. So \texttt{Enqueue()} takes $O(\log p)$ steps.

\texttt{Get()} searches among all blocks at first and then iterates over a path from the root to a leafof the tree and in each step searches in a domain of size $p$ , so it takes $\log^2 p+\log m$. \texttt{Index()} is a path from a leaf to the root which each step calls a $\log p$ step search so it takes $\log^2p$. \texttt{Dequeue()} calls \texttt{Append()}, it also calls \texttt{Index()} and \texttt{Get()} that take $\log^2 p$, $\log^2p +\log n$.

\subsection{Linearizability Proof}

\begin{definition}
If \texttt{n.blocks[i]==b}  we call \texttt{i} the \emph{index} of block \texttt{b} in node \texttt{n}. Block \texttt{b} is before block \texttt{b$^\prime$} in node \texttt{n} if and only if \texttt{b}'s index is smaller than \texttt{b$^\prime$}'s. Block \texttt{b} is propagated to node \texttt{n} or set \texttt{S} if \texttt{b} is in \texttt{n.blocks} or \texttt{S} or is a subblock of a block in \texttt{n.blocks} or \texttt{S}.
\end{definition}

%\begin{definition}
%  Block \texttt{b} in node \texttt{n} is new if \texttt{b} is not a subblock of any block in \texttt{n.parent.blocks[]}.
%\end{definition}

\begin{definition}
  Block \texttt{b} in node \texttt{n} is in \emph{Established(n, t)} if \texttt{n.head} is greater than \texttt{b}'s index at time $t$.
\end{definition}

\begin{lemma}[headProgress] \label{lem::headProgress}
\texttt{n.head} is non-decreasing over time.
\end{lemma}
\begin{proof}
  Simply because \texttt{n.head} is only incremented.
\end{proof}

\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line 52(\texttt{h=n.head}) might be 1 bit behind the first empty block.
\end{lemma}
\begin{proof}
 Because at the end of every \texttt{Refresh()} with block size greater than 0 (Lines 53,56) \texttt{n.head} is incremented. Maybe some process goes to sleep before incrementing the head, but after sleeping if \texttt{h} does not increase then \texttt{CAS} in Line 52 is going to be failed and nothing is going to be appended to \texttt{n.blocks}.
\end{proof}

%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
%  If block \texttt{b} in node \texttt{n} is established, then all the blocks in \texttt{n.blocks[]} before \texttt{b} are established. If block \texttt{b} in node \texttt{n} is not new then all the blocks in \texttt{n.blocks[]} before \texttt{b} are not new.
%\end{lemma}
%\begin{proof}
%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
%\end{proof}

\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
\end{lemma}
\begin{proof}
Because blocks are only appended(not modified) with CAS to \texttt{n.blocks[n.head]} and \texttt{n.head} is non-decreasing.
\end{proof}

\begin{lemma}[createBlock] \label{lem::createBlock}
  Suppose \texttt{CreateBlock(n, h, x)} is invoked at time $t$. The blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ that are not propagated to $Established(n, t)$, are subblock of the block returned by \texttt{CreateBlock(n, h, x)}.
\end{lemma}
\begin{proof}
 We prove the claim for the left child. Blocks in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock. Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Suppose \texttt{Refresh(n)}'s \texttt{CAS(n.blocks[h], null, new)} returns \texttt{true}. Let $t$ be the time \texttt{Refresh(n)} is invoked, blocks propgatated to $Established(n.left, t)$ and $Established(n.right, t)$ are propagated to in $Established(n, t)$ after \texttt{CAS(n.blocks[h], null, new)}.
\end{lemma}
\begin{proof}
By Lemma~\ref{lem::createBlock} \texttt{new} contains \texttt{n}'s childrens' established blocks before Line 43 which is  appended to \texttt{n.blocks} by \texttt{CAS} in Line 48.
\end{proof}

\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
    If instance $r$ of \texttt{Refresh(n)} returns \texttt{false}, then there is another successful instance $r^\prime$ of \texttt{Refresh(n)} that has performed a successful \texttt{CAS(n.blocks[h], null, new)}(Line 49) after Line 43(\texttt{h= n.head}) of $r$.
\end{lemma}
\begin{proof}
  If there is no other concurrent successful \texttt{Refresh(n)} then \texttt{Refresh(n)} would succeed in Line 48. So there is another \texttt{Refresh(n)}, that has to \texttt{CAS}ed successfully its block in \texttt{n.blocks[h]} after Line 43 of \texttt{Refresh(n)}. Otherwise the other \texttt{Refresh(n)} should have read \texttt{h$^\prime$>h} instead of \texttt{h} for \texttt{n.head}(Line 52).
\end{proof}

\begin{lemma}[Double Refresh] \label{doubleRefresh}
  Consider two consecutive instances $r_1$,$r_2$ of \texttt{Refresh(n)} by the same process (Lines 35,36). Let be the time before $r_1$ invoked. After $r_2$'s \texttt{CAS} all the blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ are in $Established(n, t)$.
\end{lemma}
\begin{proof}

If Line 35 (first Refresh which we call $R_1$) returns \texttt{true}, the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \texttt{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}. $R_1^\prime$ may or may not have propagated some subblocks of \texttt{new} to \texttt{n}. It is obvious that the \texttt{new} constructed by the second Refresh in Line 36 contains the blocks in \texttt{new} by $R_1$ which $R_1^\prime$ did not contain, since \texttt{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \texttt{n.blocks[h]} was not \texttt{null} before $R_2$'s \texttt{CAS}. Furthermore, \texttt{n.blocks[h]} was \texttt{null} before reading \texttt{h} by $R_1$. So there is a successful \texttt{Refresh()} after the read of \texttt{h} in $R_1$ and before the \texttt{CAS} of $R_2$. This \texttt{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
\end{proof}

\begin{figure}[hbt]
  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
  \caption{$R_2^\prime$'s \texttt{CAS} is executed after \texttt{h1=n.head}.}
\end{figure}

 Block \texttt{new} is created of new established subblocks of children of \texttt{n}(Lemma \ref{lem::createBlock}, Line 46). If \texttt{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \texttt{n}.

\begin{lemma}[Double Refresh] \label{doublyRefresh}
All operations in \texttt{n}'s children's blocks before line \texttt{35} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{37}.
\end{lemma}
\begin{proof}
Suppose block \texttt{b} with index \texttt{i} is in the the left child of \texttt{n} before the line 35. By Lemma~\ref{head} it follows that \texttt{n.left.head} is greater than \texttt{i}. \texttt{Refresh()} calls \texttt{CreateBlock()} and creates a block from blocks between \texttt{n.blocks[n.head].end\textsubscript{left}} and \texttt{n.left.head} in the left child, which contains \texttt{b} as well. First it tries to append it in \texttt{n.blocks.head} and if it was succsuful it continues recursivley. If not it tries again, and if the second call of \texttt{Refresh()} in Line 36 fails. It means there is another \texttt{Refresh} which has reah its \texttt{i} after the Line 35, so it contains \texttt{b} as well.
\end{proof}
\texttt{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block. If a \texttt{Refresh()} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence. So suppose two \texttt{Refresh}es fail. Since the first \texttt{Refresh()} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first \texttt{Refresh()}, was successful before the second \texttt{Refresh()}. So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh()} that assuredly has read block before the mentioned line \texttt{35}. After all it means if any of the \texttt{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.

\begin{lemma}[Append]\label{append}
 When \text{\texttt{Append(op)}} is finished, \texttt{op} appears exactly once in a block of \texttt{root.blocks}.
\end{lemma}
\begin{proof}
\texttt{Append(op)} adds \texttt{op} to \texttt{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \texttt{Propagate()} recursively propagates \texttt{op} up to the root.
By lemma \ref{doublyRefresh} we know that operation \texttt{op} propagates from child to parent at each level.
\end{proof}

\begin{lemma}[Block Size Upper Bound]\label{blockSize}
Each block in a node contains at most one operation from each processs.  
\end{lemma}
\begin{proof}
  Note that \texttt{prevIndex} and \texttt{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}. After a block has propagated to the parent the blocks between \texttt{prevIndex} and \texttt{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}). The number of new operations which have not propagated yet to the parent cannot be more than $p$. If so by the law of pigeonholes there is a process which has appended two concurrent operations.
\end{proof}

\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
Each block in a node  has at most $p$ subblocks.
\end{lemma}
\begin{proof}
  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
\end{proof}

\begin{definition} [Ordering of operations inside a node] \label{ordering}
$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.

\begin{itemize}
  \item $E(n,i)$ is the sequence of enqueue operations that are member of \texttt{n.blocks[i]} ordered by process id.
  \item 
$D(n,i)$ is the sequence of dequeue operations that are member of \texttt{n.blocks[i]} ordered by process id.
\item $D(n)=D(n,1).D(n,2).D(n,3)...$
\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}

\begin{theorem}
The queue implementation is linearizable.
\end{theorem}
\begin{proof}
  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
  \begin{enumerate}
    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}. The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$. \\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered. It remains to prove  Dequeue $d$ returns the correct response according to the linearization order. By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly. If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
  \end{enumerate} 
\end{proof}

\begin{lemma}[Get] \label{get}
\texttt{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
\end{lemma}
\begin{proof}
It is obvious that \texttt{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$. To find the \texttt{i}th enqueue in block \texttt{b} of an internal node \texttt{n} in line 87 it is decided that it resides in the left child or the right child. This decision is made by Definition of $E(n,b)$. After that Lines 88, 92 search the proper subblocks of \texttt{b}. From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \texttt{prevBlock} and \texttt{lastBlock} block of the \texttt{CreateBlock()}.
\end{proof}

\begin{lemma}[Index]
 Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
\end{lemma}
\begin{proof}
  \texttt{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \texttt{n} in \texttt{n.parent} and then computes the order in $D(n.parent, superblock)$. Then calls \texttt{index()} on \texttt{n.parent} recursively. It is easy to see why the second is correct. Correctness of computing superblock comes from Lemma \ref{superBlock}.
\end{proof}

\begin{lemma}[Computing SuperBlock]\label{superBlock}
  If \texttt{Index(n,b,i)} performs line 101, then \texttt{superblock} contains \texttt{i}th Dequeue in \texttt{b}th block of node \texttt{n}.
\end{lemma}
\begin{proof}
\begin{enumerate}
 \item Value read for \texttt{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
  Values \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} are set before incrementing in lines 26,27.

 \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\\$\blacktriangleright$
  Follows from the order of lines 60, 48, 49.

% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
 \item \texttt{super[i+1]-super[i]}$\leq p$
\\$\blacktriangleright$
 In a Refresh with successful CAS in line 46, \texttt{super} and \texttt{counter} are set for each child in lines 48,49. Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set. If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 49. So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.

 \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
\\$\blacktriangleright$
\texttt{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}. It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing. \texttt{super(b)} is the real superblock of b. \texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}. If \texttt{b.time} is \texttt{t} we have:
$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$

\end{enumerate}
\end{proof}

\begin{lemma}[Computing Queue's Head] \label{computeHead}
  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q. If Q is empty \texttt{ComputeHead(i,b)} returns -1, otherwise
 returns index in $E(root,b)$ of X.
  \end{lemma}

\begin{lemma}[Validity of \texttt{head}]\label{head}
No two blocks are written in the same index in \texttt{n.blocks}.
\end{lemma}
\begin{proof}
  \texttt{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \texttt{head} read. If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \texttt{head}. If a process dies before line 51, another process will increment \texttt{head} in line 54.
\end{proof}

\begin{lemma}[Validity of \texttt{super} and \texttt{counter}]\label{superCounter}
If \texttt{super[i] $\neq$ null}, then \texttt{super[i]} in node \texttt{n} is the superblock of a block with \texttt{time=i}.
\end{lemma}
\begin{proof}
After a successful CAS in line 46 \texttt{super} and \texttt{counter} are modified in both children. \texttt{super[i]} is supposed to be the superblock of a block with \texttt{time}=i and \texttt{counter} is the timer in each node. \texttt{super[i]} and \texttt{counter=i} are expected to update after a bunch of blocks with \texttt{time=i} have been aggregated together into a block in the parent. If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \texttt{time}. We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity. If a process dies right after line 48, then \texttt{counter} will remain the same and \texttt{super[i]} is correct. Furthermore we are sure that when \texttt{super[i]} is read it will not be \texttt{null}.
\end{proof}

\begin{lemma}[Search Ranges]\label{search}
  Preconditions of all invocation of \texttt{BSearch} are satisfied.
\end{lemma}
\begin{proof}

Line 83: \texttt{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.

Line 88: This search tries to find the ith enqueue, knowing that it is in the left child. Search is done over the left subblocks. The start and end of the range are followed by definition. Line 92 is the same.

Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \texttt{super[i]} is at most $p$, since at most $p$ processes could die.

\end{proof}

\paragraph{Overall}

The main difference between the main algorithm and the block tree is that we separate enqueues and dequeues, compute the number of non-null dequeues and the Queue's size in each block in the root. We emphasize that these changes work correctly; every other claim in Block trees applies here.

\section{Make first level search fast}
\begin{algorithm}
\caption{Modified Queue \label{algQx}}
\begin{algorithmic}[1]
\begin{multicols}{2}

%\Statex \textbf{Structure}

%\Statex $\blacktriangleright$ \texttt{extends Queue Algorithm}

%\Statex $\blacktriangleright$\texttt{\textsl{RBTNode} RBTRoot}  \Statex
%  \textsf{pointer to an RBTNode, that is root of the RBTree}

\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[RBTNode]} rootBlocks}
  \Statex \textsf{A persistant redblack tree supporting \texttt{append(n),get(i),split(j)}}. \texttt{append(n)} returns \texttt{true} in case successful.

\Statex $\blacktriangleright$ \texttt{\textsl{RBTNode}}
\begin{itemize}

  \item \texttt{\textsl{RootBlock} block}

  \item \texttt{\textsl{int} age}

  \textsf{number of finished operations in the block}

  \item \texttt{\textsl{int} step}

  \textsf{If this node is the root of the RBTree, then step shows the number of appended nodes to the RBTree}

%  \item \texttt{some RBTRee attributes like left,right children}

\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf} extends Node}
\begin{itemize}
  \item \texttt{\textsl{int[]} response}

  \textsf{leaf.response[i] stores response of leaf.ops[i]}

  \item \texttt{\textsl{int} maxOld}

  \textsf{Index of the youngest old block in the root that this process has seen yet.}

\end{itemize}

%\Statex $\blacktriangleright$ \texttt{\textsl{RBT(RBTnode n)}}
%  \Statex \textsf{Gives us RBTree  persistentoperations if n is root of a RBTRee, for example RBT(n).append(x) appends x if to the redblack tree with n as root. A persistent append is an append which only contains the changed nodes.}

\Statex

\Function{boolean}{Refresh}{\textsl{node} n} \Comment if n is root
\State \texttt{new=CreateBlock()} \Comment TODO
\If{\texttt{new.num==0}} \Return{\texttt{true}}
\ElsIf{\Call{RBTAppend}{new}}
\Statex..\EndIf
\EndFunction{Refresh}

\Statex

\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{search in the current RBTRee}
\Else
..
\EndIf
\EndFunction{Index}

\Statex

\Function{Object}{Dequeue()}{}
\State ..
\State \texttt{N\textsubscript{i}= RBTnode n that n.block contains the invocation of Dequeue}
\State \texttt{N\textsubscript{r}= RBTnode n that n.block contains the response of Dequeue}
\State \texttt{N\textsubscript{i}.age= N\textsubscript{i}.age+1} \Comment{this is a shared counter}
\State \texttt{N\textsubscript{r}.age= N\textsubscript{r}.age+1}
\If{\texttt{N\textsubscript{i} or N\textsubscript{r} become old}} \texttt{update maxOld}
\EndIf
\EndFunction{Dequeue}

\pagebreak
%\Function{void}{Update(Root Block B)}{}
%\If{\texttt{B.todo==0}}
%\If{\texttt{B.left.old-subtree} and \texttt{B.right.old-subtree}}
%\State \texttt{B.old-subtree= True}
%\If{\texttt{B.parent!=root}}
%\State \texttt{Update(B.parent)}
%\EndIf\EndIf
%\EndIf
%\EndFunction{Update}
%\Statex

\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{rootBlocks}}}
\State \texttt{root= rootBlocks.root}
\If{\texttt{root.step\%$p^2$==0}}
\State \texttt{Help()}
\State \texttt{CollectGarbage()}
\EndIf
\State \texttt{new= RBTnode(b,0,null)}
%\State \texttt{r= pointer to root of the RBT(h).append(n)}
%\State \texttt{*r.step= h.step+1}
%\State \texttt{RBTRoot.CAS(h,r)}
\State \Return \texttt{rootBlocks.append(new)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\Comment{Helps pending operations}
\State{last= l.head-1}
\For{\texttt{leaf l in leaves}}
\If{\texttt{l.blocks[last] is not null}}
\If{op \textbf{is} DEQ}
\State\texttt{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
\State \texttt{write the response to l.responses[last]}
\EndIf
\EndIf
\EndFor
\EndFunction{Help}
\Statex

\Function{void}{CollectGrabage}{}\Comment{Collects the old root blocks.}
\State \texttt{l=FindYoungestOld(Root.Blocks.root)}
\State \texttt{t1,t2= RBT.split(l)}
\State \texttt{RBTRoot.CAS(t2.root)}
\EndFunction{CollectGrabage}

\Statex

\Function{Block}{FindYoungestOld}{b}
\State\Return  \texttt{read all maxOld values among leaves and decide the largest one} \Comment{There is no need to do a sasDK;Lnapgshot.}
\EndFunction{findYoungestOld}

\Statex

\Function{response}{FallBack}{op i}

\If{operation i in leaf l cannot find its desired RootBlock}

\State \Return \texttt{l.response[i]}
\EndIf

\EndFunction{FallBack}

\end{multicols}
\end{algorithmic}
\end{algorithm}

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2022-02-03 jayanti/Jayanti.tex
add current writings

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\definecolor{light-gray}{gray}{0.95}
\usepackage{listings}
\lstset{
    numbers=left,
    breaklines=true,
    backgroundcolor=\color{light-gray},
    tabsize=2,
    basicstyle=\ttfamily,
}
\usepackage{setspace}
\renewcommand{\baselinestretch}{1.3} 
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{fancyhdr}
\renewcommand{\baselinestretch}{1.34} 
\usepackage{geometry}
\usepackage{enumitem}
\usepackage[]{algorithm2e}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt plus 1pt}
\setlength{\headheight}{13.6pt}
\newcommand\question[1]{\vspace{1.2em}\hrule\textbf{ #1}\vspace{.5em}\hrule}
\renewcommand\part[1]{\vspace{.10in}\textbf{#1)} \enspace}
\pagestyle{fancyplain}
\fancyhf{}
\renewcommand{\headrulewidth}{1.4pt}
\lhead{\textbf{\ANDREWID }}
\cfoot{\textbf{Project } - \thepage   }
\rhead{\rule[-1ex]{0pt}{3.5ex} Hossein Naderi }
\newcommand\ANDREWID{Linearizability correctness of Jayanti MESD queue} 
\newcommand\HWNUM{1}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\begin{document}

\begin{definition}[v-visioner]
  A propagate procedure reading timestamp v (Line 17) in one of its refresh procedures is called v-visioner.
\end{definition}

\begin{theorem}
Proposed queue is linearizable. linearization point of each enqueue(p,v) is right after first v-visioner propagation is completed (Line 10). And dequeue(p) operation is linearized just after reading root's value (Line 11).
\end{theorem}
\begin{proof}
We proposed an ordering for the linearization of operations. We claim that if operations are completed in the order, we mentioned then the results of operations are the same, which means the queue is consistent. It is trivial to see that if $op_1$ is finished before $op_2$ the linearization point of $op_1$ is before $op_2$ (real-time ordering  property of linearization). We break the rest of the proof according to the possible conflicting cases.

\begin{itemize}
\item Enqueue(p,v) \& Dequeue(q) have conflicts
\begin{itemize}
\item Enqueue is linearized before the Dequeue\\
This means the value v is propagated before reading the root. So whether or not if v is the smallest value, then we can consider its operation is completed because the dequeue is going to see the result of top of the tree after the propagation of v.
\item Dequeue is linearized before the Enqueue\\
This means the dequeue has read the root element, so whether or not finishes or nor the dequee is going to return the soot element, so it has taken effect before enqueue operation.
\end{itemize}
\item Two Enqueue operations Conflicting \\
In this case, one of the enqueue operations' inserted value has propagated before the other one and has linearized before. After the linearization point of the earlier operation, the root value is updated, so we can interpret it has finished before the other.
\end{itemize}
\end{proof}

\begin{definition}[invariant]
The root element in the tree is the value with the smallest timestamp in the tree.
\end{definition}
I think the invariant needed is for the correctness of the algorithm and not the linearizability in order to show the dequeue is returning the smallest timestamp. I don't think it holds for the whole tree because maybe after propagating some element, another small element is still propagating.

\end{document}

### 2022-02-03 jayanti/Universal Construction using Jayanti with CAS.tex
add current writings

\documentclass[10pt]{article}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\definecolor{light-gray}{gray}{0.95}
\usepackage{listings}
\lstset{
    numbers=left,
    breaklines=true,
    backgroundcolor=\color{light-gray},
    tabsize=2,
    basicstyle=\ttfamily,
}
\usepackage{setspace}
\renewcommand{\baselinestretch}{1.3} 
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{fancyhdr}
\renewcommand{\baselinestretch}{1.34} 
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{multicol}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt plus 1pt}
\setlength{\headheight}{13.6pt}
\newcommand\question[1]{\vspace{1.2em}\hrule\textbf{ #1}\vspace{.5em}\hrule}
\renewcommand\part[1]{\vspace{.10in}\textbf{#1)} \enspace}
\pagestyle{fancyplain}
\fancyhf{}
\renewcommand{\headrulewidth}{1.4pt}
\lhead{\textbf{\ANDREWID }}
\cfoot{\textbf{Third Attempt } - \thepage   }
\rhead{\rule[-1ex]{0pt}{3.5ex} Hossein Naderi }
\newcommand\ANDREWID{Universal Construction using Jayanti Approach} 
\newcommand\HWNUM{1}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\begin{document}

%\textit{In this paper we're trying to use Tournament trees in order to create a faster Universal Construction using CAS objects.}

%
%\question{What we need to satisfy in our design}
%We have given a sequential object with some type of queries. We need to make a middle ware which gets  queries by different processes and in a nonblocking way finishes operations. Each process add an operation to be done on the sequential object some times. We're going to create a consensus order on all operations among processes. Each operation is done on the sequential object when it is added to final list.
%
%Since our problem is a lot like maintaining a queue, my first idea is to take advantage of Jayanti FIFO queue. Since it can handle concurrent enqueue queries and single dequeues. Whenever a process wants to do some operation it enqueues it and increases a shared counter value. An Idle process is always looking at counter value and it it is greater than 0 dequeues from queue and does the job.
%
%
%
%\begin{minipage}{.5\textwidth}
%Shared Objects:
%\begin{itemize}
%  \item \textit{shared} Q: a MESD FIFO Queue
%\item \textit{shared} counter
%\item O: given sequential object
%\end{itemize}
%\begin{algorithmic}[1]
%\Function{$do_{client}$}{operation op} 
%\State $enqueue(op)$
%\State $inc(counter)$
%\EndFunction
%\end{algorithmic}
%
%\end{minipage}
%\begin{minipage}{0.5\textwidth}
%\begin{algorithmic}[1]
%\Function{$do_{master}$}{} 
%\Loop
%\If{$counter>0$}{ $dec(counter)$}
%\State $op=dequeue(T)$
%\State $dec(counter)$
%\State $\delta(O,op)$
%\EndIf
%\EndLoop
%\EndFunction
%\end{algorithmic}
%
%    \end{minipage}
%
%\pagebreak
%\question{Correctness and Analysis}
%Our goal is to maintain a list that keeps linearization ordering of operations among all processes. We have to show its linearizable and prove progress property. It is non blocking since counter and queue are wait free so client operations do not lock each other and since there is only one dequeue it is deadlock free too.
%
%But there are problems with crashes: What happens if in an $d0_{client}$ an op is enqueued but the counter is not incremented? Since $do_{master}$ is only running by one process and it's not terminating then we don't have to set linearization point to it. $do_{client}$ is after line 3 when counter is 0 and line 2 when counter is $>0$. Nonconcurent operations are trivial so Let's talk about concurrent ones.
%
%Cilent operations took $O(log\; n)$ and the Master is constantly dequeuing the queue.
%\begin{proof}
%\begin{itemize}
%  \item $do_{client} 1$ is linearized before $do_{client} 2$\\ \textit{TODO}
%
%\end{itemize}
%\end{proof}
%
%\question{Negative points of algorithm}
%The main problem with this design is its single point failure. When master dies then no operations will be done on the given sequential object.
%
%There is another inherent problem with universal constructions that I don't get. I mean since our final object is sequential and only one operation is running at a time so the only improvement here is that there is a consensus among processors on which one to do it's operation. So if we create a consensus object that gives permissions to processes to do their operation it will be enough. It's not wait free since a process may halt but I guess it's nonblocking. It's very similar to task scheduling in operating system.
%
%\pagebreak
%\question{Our Meeting Idea}
%Instead of maintaining a single object shared among processes, let's give a sequential object to each process and only share history among them. Our design is a tournament tree which its leaves are assigned to processes and they are a queue of jobs.
\question{Design \& Notation}
A tournament tree in which each process is assigned to one of its leaves. Each node of the tree stores a string of form $"<operation, process>^*"$ and has a regex parser for reading its words. Each process has a local copy of the history of the given sequential object and a local copy o.

We show set subtraction with notation "-" in the algorithm. And "." for concatenation on two strings. 
In the rest of the paper, when we say node $n$ we mean $n.value$. Unless we explicitly call tree related attributes like $n.children$ or $n.parent$.

\begin{algorithm}
\caption{}\label{alg}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex Shared Objects
\Statex \hspace{\algorithmicindent}tree T
\Statex Local Objects
\Statex \hspace{\algorithmicindent}Sequential Object o
\Statex \hspace{\algorithmicindent}List snapshot
\Statex

\Function{Do}{operation op} 
\State l= p's assigned leaf in tree
\State l.$<op, p>$
\State \Call{Propagate}{l}
\State history= \Call{Read}{root}
\State changeset= histroy-snapshot
\State snapshot= history
\ForAll{$<op^\prime, p^\prime>$: changeset}
\If{p$^\prime$=p}
\State res=o.op$^\prime$
\Else
\State o.op$^\prime$
\EndIf
\EndFor
\State \Return res
\EndFunction

\Statex

\Function{Propagate}{node n}
\If{n==root} \Return
\ElsIf{!\Call{Refresh}{n.parent}}
\State \Call{Refresh}{n.parent} \EndIf
\State \Call{Propagate}{n.parent}
\EndFunction

\Statex

\Function{Refresh}{node n}
\State old= \Call{Read}{n}
\ForAll{child: n.children}
\State posted.(child-n)
\Statex \Comment ops that child contains but n doesn't
\EndFor
\State new+= old.posted
\State res=\Call{CAS$_n$}{old, new}
\State \Return res
\Statex\Comment{\texttt{true}: Success, \texttt{false}: Failure}
\EndFunction
\end{multicols}
\end{algorithmic}
\end{algorithm}

\question{Correctness}
We're going to show our design satisfies correctness and progress.
\textsc{Do(\textnormal{op})} finishes in finite steps, so our design is wait-free. It remains to show Algorithm \ref{alg} is linearizable.

%\begin{lemma}
%  For every failed \textsc{Refresh(n)} instance $refresh_i$, there is at exactly one successful concurrent $refresh$ that has changed $n$ before $refresh_i$.
%\end{lemma}
%\begin{proof}
%   It is trivial to see that there are some successful \textsc{Refresh(n)} operations which have read same value as $refresh_i$. Because CAS operations which only concats strings and properties of \textsc{CAS$_n$()} then there is exactly one.
%\end{proof}

\begin{lemma}
  If $op$ is in node $n$ and then a process calls a successful \textsc{Refresh(}\textnormal{n.parent}\textsc{)} or 2 consecutive \textsc{Refresh(}\textnormal{n.parent}\textsc{)}s, $op$ will be in $n.parent$ after those procedures terminates.
\end{lemma}
\begin{proof}
In case of one successful \textsc{Refresh()}, the \textsc{CAS}$_n$ has added $op$ to $n.parent$.
Consider 2 consecutive failed $refresh_1$, $refresh_2$ on node $n$. For each failed \textsc{Refresh()} $r$ there is at least a \textsc{Refresh()} $w$ which has succeeded to do \textsc{CAS} with the same old vale read as $r$.
From Lemma 1 let the successful \textsc{Refresh()}es corresponding to $refresh_1$ and $refresh_2$ be $refresh_3$ and $refresh_4$ respectively.
If $refresh_3$ has read $op$ then $op$ will be in $n.parent$ after $refresh_3$. Otherwise $refresh_4$'s \textsc{Read} cannot be before \textsc{CAS} of $refresh_3$.
And since the \textsc{CAS} of $refresh_3$ is after concatenating $op$ to $n$ then $refresh_4$ has read $op$ and stores it in $n.parent$.
\end{proof}

\begin{lemma}
  Consider an instance of \textsc{Propagate(}\textnormal{n}\textsc{)}. When it terminates all operations that were in $n$ before \textsc{Propagate(}\textnormal{n}\textsc{)} will be in the root.
\end{lemma}
\begin{proof}
  We prove this by induction on depth of node n. From Lemma 1, all operations in $n$ will be in $n.parent$ after one recursive call of \textsc{Propagate(}\textnormal{n}\textsc{)}. By induction hypothesis to \textsc{Propagate(}\textnormal{n.parent}\textsc{)}  which is then called recursively, \textsc{Propagate(}\textnormal{n}\textsc{)} satisfies the claim.
\end{proof}

\begin{theorem}
Algorithm \ref{alg} is linearizable.  
\end{theorem}
\begin{proof}
   Lemma 2 shows us that if \textsc{Propagate} procedure of a \textsc{Do(\textnormal{op})} is terminated then $op$ will be in root. let the linearization point of a finished \textsc{Do(\textnormal{op})} be when $op$ is  concatenated to the root by some \textsc{Propagate} procedure. If several operations are propagated at the same time then they are linearized in the order of adding to root. Since every process runs operation on  $o$ in order of the $history$ stored in the root, then results are consistent with linearization order.

  If $do_1$ is finished before $do_2$ begins, then $do_1$ is linearized before $do_2$.
  From Lemma 2 we know that after line 4 of Algorithm 1 $do_1$ operations are propagated to the root. So $do_1$ is linearized before $do_2$.
\end{proof}

\question{Analysis \& Conclusion}
  Each \textsc{Do(\textnormal{op})} has a \textsc{Propagate} which uses at most $2\log(p)$ \textsc{CAS} operations. The history is repeated by each process locally. So after $n$ \textsc{Do(\textnormal{op})} operations among $p$ processes there will be $\Omega(\log p + np)$ steps per \textsc{Do(\textnormal{op})}. However the number of shared memory accesses is $\Omega(\log p)$. Since CAS operations work with large words the algorithm is not practical, but it shows that  without restricting size of shared words the lower bound on universal construction shared memory accesses is lower than $\Omega(\log p)$.

\end{document}

### 2022-02-03 jayanti/describing-alogorithm.tex
add current writings

\documentclass[10pt,table]{article}
\usepackage[dvipsnames]{xcolor}
%\usepackage{hyperref}
%\usepackage{amsmath}
\usepackage{tikz}
\usepackage{qtree}
\usepackage{tikz-qtree}
%\usepackage{pgfplots}
\usepackage{listings}
%\definecolor{light-gray}{gray}{0.95}
\usepackage{listings}
\usepackage{setspace}
\renewcommand{\baselinestretch}{1.3} 
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{amsmath,amssymb,amsthm,amsfonts,amssymb}
\usepackage{fancyhdr}
\usepackage[b4paper]{geometry}
\usepackage{enumitem}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multicol}
\setlength\columnsep{30pt}
\usepackage{setspace}

\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt plus 1pt}
\setlength{\headheight}{13.6pt}
\newcommand\question[1]{\vspace{1.2em}\hrule\textbf{ #1}\vspace{.5em}\hrule}
\renewcommand\part[1]{\vspace{.10in}\textbf{#1)} \enspace}
\pagestyle{fancyplain}
\fancyhf{}
\renewcommand{\headrulewidth}{1.4pt}
\lhead{\textbf{\ANDREWID }}
\cfoot{\textbf{First Draft } - \thepage   }
\rhead{\rule[-1ex]{0pt}{3.5ex} Hossein Naderi }
\newcommand\ANDREWID{Trying to explain the algorithm in simple language} 
\newcommand\HWNUM{1}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

\begin{document}

\question{Previous Work}

\begin{table}[hbt]
  \begin{center}
  \begin{tabular}{l|c|c|c|c|c}
    & Type & Progress Property & Conditions & Amortized Time Complexity & Space \\
    \hline
    & & & & & \\
  \end{tabular}
  \caption{}
  \end{center}
\end{table}

\question{Goal}
The model consists of $p$ processes. And the problem is to implement linearizable shared queue Q supporting $<ENQ, DEQ>$ operations between them.

%
%\begin{figure}[h]
%\begin{center}
%\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
%
%\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
%%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
%
%%Curve Lines [id:da24404451870478872] 
%\draw    (102,43) .. controls (143.58,47.95) and (138.12,104.85) .. (199.13,107.92) ;
%\draw [shift={(201,108)}, rotate = 181.82] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%%Curve Lines [id:da9272135872377072] 
%\draw    (102,121) .. controls (141.6,91.3) and (162.58,119.43) .. (224.12,117.08) ;
%\draw [shift={(226,117)}, rotate = 537.27] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%%Curve Lines [id:da5453903537732101] 
%\draw    (101,182) .. controls (158.42,196.85) and (149.19,136.23) .. (203.34,122.4) ;
%\draw [shift={(205,122)}, rotate = 526.9300000000001] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%%Shape: Grid [id:dp5927814714859243] 
%\draw  [draw opacity=0] (238,105) -- (350,105) -- (350,126) -- (238,126) -- cycle ; \draw   (238,105) -- (238,126)(258,105) -- (258,126)(278,105) -- (278,126)(298,105) -- (298,126)(318,105) -- (318,126)(338,105) -- (338,126) ; \draw   (238,105) -- (350,105)(238,125) -- (350,125) ; \draw    ;
%%Curve Lines [id:da33961815970017184] 
%\draw    (359,115) .. controls (398.6,85.3) and (399,117.35) .. (457.22,115.08) ;
%\draw [shift={(459,115)}, rotate = 537.14] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%%Curve Lines [id:da9041409242658852] 
%\draw    (359,115) .. controls (400.58,104.11) and (400.02,164.77) .. (458.22,160.16) ;
%\draw [shift={(460,160)}, rotate = 534.29] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%%Curve Lines [id:da2855643108836211] 
%\draw    (359,115) .. controls (379.79,51.64) and (412.34,55.91) .. (458.59,71.52) ;
%\draw [shift={(460,72)}, rotate = 198.8] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) .. controls (6.95,-1.4) and (3.31,-0.3) .. (0,0) .. controls (3.31,0.3) and (6.95,1.4) .. (10.93,3.29)   ;
%
%\end{tikzpicture}
%\caption{Multiple queries on the shared Queue}
%\end{center}
%\end{figure}

%We show operation $o$ requested by process $p$ by $o_p$ and the shared queue by $Q$ and local queues by $q$.

If $op$ finishes before $op^\prime$ starts, it has to take effect on the $Q$ before another. For concurrent operations from processes on $Q$, the implementation can decide which to put before the other. So we want to design an algorithm that gives processes responses and does their operation on the $Q$. Since the linearization is a complete ordering among given operations, our problem is to agree on a linearization of the operations.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}
\Tree [.$n_0$ [.$n_1$ [.$n_3$ $p_1$ $p_2$ ] [.$n_4$ $p_3$ $p_4$ ] ]
          [.$n_2$ [.$n_5$ $p_5$ $p_6$ ] [.$n_6$ $p_7$ $p_8$ ] ] ]
\end{tikzpicture}
\caption{The total ordering of all requests operation propagated up to the root is stored in the root. A request it done if it has reached to the root and has been added to the history. In each node we store which requests has been propagated up to it.}
\end{center}
\end{figure}

A usual way \cite{JayantiP05} to agree on something is to use tournament trees. Consider a tournament tree where each process is assigned to one of its leaves. Each process adds an operation to its leaf and tries to propagate it up to the root. The first one that reaches the root is the first operation. Thus the tournament tree is like a competition between processes to determine the total ordering. Propagating an operation from a node means trying to move it one level higher than its parent. Sometimes, two operations from two nodes are propagated up to the parent concurrently; we have to choose one to be before the other arbitrarily.

Now two problems arise:
\begin{itemize}
  \item Since the tournament tree is shared among all the processes, then at some point, there may be multiple processes trying to propagate their operations at the same node. So we need a lock-free procedure for propagating operations from children of a node to it.
  \item One way to implement a shared object with the tournament tree is to store in each node the ordering of operations propagated up to that node. Merging of the two children of the root gives us the total ordering at some point. But keeping all these data is not memory-efficient. For example, all dequeues straight after a queue that gets empty can return null without knowing the whole history before.
\end{itemize} 
\pagebreak
\question{First Try}

In the first attempt, we store each subtree ordering of operation in its root. At each propagation step, we append children's new operations to the parent ordering.  The merging step is heavy, and this way is not memory-efficient. Each ordering merge step is $O(p)$, and there will be $\log(p)$ propagate steps, and it will take $O(\#operations)$ to compute the operation response.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}

\Tree [.{\begin{tabular}{l|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ & ... \\ \hline\end{tabular}}  [.{\begin{tabular}{l|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}}
      {\begin{tabular}{l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline\end{tabular}} ] [.{\begin{tabular}{l|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{First Try: We show operation $i$ from process $j$ with $op_j^i$. In each node, we store the ordering of all the operations propagated up to it.}
\end{center}
\end{figure}

\begin{algorithm}[h]
\caption{First Try Algorithm}
\begin{algorithmic}[1]

\Procedure{Do}{node n, operation op}
\State n.leaf.append(op)
\State \Call{Propagate}{n}
\State \Return \Call{Compute}{op}
\EndProcedure
\Statex
\Procedure{Propagate}{node n}
\If{r$\neq$root}
\State \Call{MergeChilrenOrderingInto}{n.parent}
\State \Call{Propagate}{r.parent}
\EndIf
\EndProcedure
\Statex
\Procedure{MergeChilrenOrderingInto}{node n}
\State new=n.children.new-operations
\State n.ordering.append(new)
\EndProcedure

\end{algorithmic}
\end{algorithm}

\question{Second Try}

On the first try, we store the ordering of all the operations in the subtree in its root. But it's not necessary for concurrent operations. If two operations are propagated to a node simultaneously, they will be propagated up to the root together so we can store their ordering in the root. This improvement does not change the time complexity of merge steps.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}

\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
      {\begin{tabular}{l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ... \\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ... \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{Second Try: In each internal node, we store the set of all the operations propagated together up to it, and one can arbitrarily linearize the total of concurrent operations.}
\end{center}
\end{figure}

\pagebreak
\question{Third Try}

We call sets in the ordering from Algorithm 2 "blocks". Here we are proposing that if you store some constant statistics in each set, you can compute the set. If you know how many operations from the left child and right child are in block b, then with knowing the count of operations before block b in node n, we can find out which operation from each child of n has propagated to block b. This leads us to the main algorithm. Using this approach, we can make merge steps $O(\log p)$. As we said before, if we know some data like what is the size of the queue, it helps us to compute results faster. In this way, we propose our algorithm to compute the dequeue operation in $O(\log^2p)$.

%\begin{figure}[h]
%\begin{center}
%\begin{tikzpicture}
%  
%\Tree [.{\begin{tabular}{l|c|c|c|c|c|c|c|c|c}  \hline $root$ & $\pi$ & $\chi$ & $\varphi$ & $\Delta$ & $\theta$ & $\Upsilon$ & $\xi$ & $\eta$ & ... \\ \hline\end{tabular}}  [.{\begin{tabular}{l|c|c|c|c}  \hline $n_1$ & $B_{11}$ & $B_{12}$ & $B_{13}$ & ... \\ \hline\end{tabular}}
%      {\begin{tabular}{l|c|c|c}  \hline $p_1$ & $\varphi$ & $\Upsilon$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_2$ & $\Delta$ & $\theta$ & ... \\ \hline\end{tabular}} ] [.{\begin{tabular}{l|c|c|c}  \hline $n_2$ & $B_{21}$ & $B_{22}$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_3$ & $\chi$ & $\eta$ & ... \\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_4$ & $\pi$ & $\xi$ & ... \\ \hline\end{tabular}} ] ]
%
%\end{tikzpicture}
%\caption{Third Try: In each node we store the summary of operation propagated together(some constant size statistics called blocks). Pros: Quick merges and low memory use.}
%\end{center}
%\end{figure}

%If each propagation step is not heavy in our design, then there are at most $\log(p)$ steps. We are trying to implement steps in the nodes $\log(p)$ too. In the remainder, we focus on two parts of the algorithm; the first one deals with the lock-free property of the merge procedure on nodes of the tournament tree, and the second one is about why our design works fast and correctly as a shared queue.

\question{Merge Step}
In each merge step on node n, we read n.children new operations and try to append them to the n's ordering. Here we propose a wait-free approach using two CAS operations. First, we create a block of newly added operations to n's children(exist in n.children but not in n). After that, we try to append them to the last block; if it wasn't successful, we try again. If one of the $CAS$ operations was successful, then the block is propagated up to the root; otherwise, it means after the first unsuccessful CAS, another operation has come, and it reads current operations.

\texttt{This new algorithm needs proof}

\begin{algorithm}[h]
\caption{Merge Step}
\begin{algorithmic}[1]

\Procedure{MergeChilrenOrderingInto}{node n}

\State block=\Call{CreateBlock}{n.left,n.right}
\State last=n.last \Comment{Index of the first empty cell of n's array of blocks.}
\If{\Call{CAS}{n.last, last, last+1}}
n[last]=block
\ElsIf{\State\Call{CAS}{n.last, last+1, last+2}}
n[last]=block
\EndIf
\EndProcedure

\end{algorithmic}
\end{algorithm}

\question{Data structure details}
Here we are talking about details of the information stored in blocks and the root.

Tree Structure:
\begin{itemize}
  \item Leaf $l$ of the tournamnet tree is the list of the operations of process $p$.
  \item Interval node $n$ stores an array of blocks($n.blocks$) and index of the first empty cell of the array($n.last$).
  \item Root like interval nodes stores blocks with two additional data: size(size of the queue after the block), req($\#$returning deques in each block).
\end{itemize}

In each block we store:
\begin{itemize}
  \item Accumulative $\#$left enqs,$\#$right enqs,$\#$left deqs,$\#$right deqs
  \item pointers to last block merged from left child and righ child
\end{itemize}

How to find the ith enq among all operations? Find the bock containing ith operation in the root using binary search. Decide the operation is in which child and continue recursively.

\texttt{how to draw lists as nodes of the tree and draw edges from cells?}

\begin{center}
\begin{tikzpicture}

\Tree [.1 2 3 ]

\end{tikzpicture}
\end{center}

And these coloumns in the root:\\

{\begin{tabular}{c|c|c|c|c|c|c|c|c|}
    \hline enq & 5,6 & 5,8 & 6,3 & 3,7 & 10,2 & 1,2 & 2,3 & ... \\
    \hline deq & 3,11 & 3,6 & 5,10 & 7,9 & 3,1 & 2,0 & 9,8 & ... \\
    \hline rdeq & 11 & 9 & 13 & 10 & 4 & 2 & 14 & ... \\ 
    \hline size & 0 & 4 & 0 & 0 & 8 & 9 & 0 &... \\ \hline\end{tabular}}

\question{Complete Algorithm with the Logic related to the Queue}

\begin{algorithm}
\caption{Main Algorithm}\label{alg}
\begin{algorithmic}[1]
\onehalfspacing

\Function{Do}{operation op}
\State add p to this.ops
\State \Call{Propagate}{this.ops}
\State \Comment{When is op added to the root?}
\If{op is a deq}
\State before-size: size of the block before block containing op
\State e: \#enqs in the block containing op
\State d: \#deqs in the block containing op before op
\If{before-size + e - d $<$ 1}
\State \Return{null}
\Else{}
\State d: \#rdeqs before the op in all the ordering
\State \Return{enq(d+1)} \#d+1th enq value in all the enqs
\EndIf
\EndIf
\EndFunction
\Statex

\Function{Propagate}{node n}
\State b=\Call{Create-Block}{n}
\If{!\Call{tryAppend}{b, n}}
\Call{tryAppend}{b, n}
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction
\Statex

\Function{Create-Block}{n}
\Statex \Comment constructs block of the new operation in children of n. if n is the root, it has extra fields: size, rdeqs.
\EndFunction
\Statex

\Function{tryAppend}{b, n}
\Statex \Comment tries to append b to the last of the n's list.
\EndFunction
\Statex

\Function{Index}{operation op, level $\in$ nodes, type $\in\{$block, operation$\}$}
\Statex \Comment returns index of op in the given level, e.g  Index(op, root, block) return ordering of the block containing op in the root blocks.
\EndFunction
\Statex

\Function{Access}{i, level $\in$ nodes, type $\in\{$enq, deq$\}$}
\Statex \Comment returns i-th operation of given type in the given node subtree.
\EndFunction
\Statex

\Function{Prefix-Sum}{i, level $\in$ nodes, type$\in\{$enq, deq, rdeq$\}$}
\Statex \Comment computes how many of the given type operations are before the ith operation in the given level. For rdeq it will only get root level.
\EndFunction

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Block Tree}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Statex $\blacktriangleright$ leaf $l_i$: list of operations
\Statex $\blacktriangleright$ internal node $n$ of BT: list of blocks and index $last$
\Statex $\blacktriangleright$ block $b$: 4 statistics of concurrent operations aggregated together to a block in a \textsc{Refresh} consisting [$left$:$ \#$left ops, $right$: $\#$right ops, $left-sum$: prefix sum left ops, $right-sum$: prefix sum right ops]
\Statex $\blacktriangleright$ index $last$: index of last block of node $n$
\Statex

\Function{void}{Append}{operation op, pid i} 
\State $l_i$.append($op$)
\State \Call{Propagate}{parent of $l_i$}
\EndFunction{Append}

\Statex

\Function{void}{Propagate}{node $n$}
\If{$n$==root} \Return
\Else
\State $new$=\Call{CreateBlock}{n}
\If{\Call{CAS}{$n$.last, now, now+1}}
last block of $n$=now
\ElsIf{\Call{CAS}{$n$.last, now, now+1}}
last block of $n$=now
\EndIf
\EndIf
\State \Call{Propagate}{parent of $n$}
\EndFunction{Propagate}

\Statex

\Function{boolean}{CreateBlock}{node n}
\State $current$= last block of $n$
\State $start$= \Call{BSearch}{crrent.left-sum}
\State $left$=0
\ForEach{$block$}{ $start$: first null block of left child of $n$}
\State $left$+=$block$.left
\EndFor
\State $left-sum$=current.left-sum+left
\State do lines 14 to 19 for right
\State \Return [left,right,left-sum,right-sum]
\EndFunction{CreateBlock}

\columnbreak

\Function{element}{GetIndex}{block b, index i} 
\State n:=b.node
\If{i$\leq$b.left}
\State sb=\Call{BSearch}{left child of $n$, i-b.right.sum}
\State \Call{GetIndex}{sb, b.left-sum} 
\Else ~sb=\Call{ BSearch}{right child of $n$, i-b.left.sum}
\State \Call{GetIndex}{sb, b.right-sum} \EndIf
\EndFunction{GetIndex}

\Statex

\Function{list}{GetElements}{block b} 
\ForEach{block \textbf{in} \Call{GetSubBlocks}{b}}
\State result.append(\Call{GetElements}{})
\EndFor
\State \Return result
\EndFunction{GetElements}

\Statex

\Function{list}{GetSubBlocks}{b}
\State n:=b.node
\State b[-1]=b's previous block
\ForEach{direction \{left, right\}}
\State init=\Call{BSearch}{n.direction, b[-1].direction}
\State end=\Call{BSearch}{n.direction, b.direction}
\State result.append([init:end])
\EndFor
\State \Return result
\EndFunction{GetSubBlocks}

\end{multicols}
\end{algorithmic}
\end{algorithm}

\bibliographystyle{plain}
\bibliography{refrences}

\end{document}

### 2022-02-03 proposal/proposal.tex
add current writings

\documentclass{article}
\usepackage{hyperref}
\usepackage{graphicx}

\renewcommand{\baselinestretch}{1.5}

\begin{document}

\title{Thesis Proposal}
\author{Hossein Naderi}
\maketitle

\begin{enumerate}
  \item Importance of implementing shared data structures
\begin{itemize}
  \item how it helps developers
  \item why lock free data structures are better than using locks?
\end{itemize} 
  \item  What properties matter for a distributed data structure?
\begin{itemize}
  \item Correctness condition: Linearizability definition
  \item Wait-freeness
\end{itemize} 
  \item Previous works on distributed queues: more detailed
\begin{itemize}
  \item why we are studying queus
  \item $p$ factor in time complexity of all previous algorithms
  \item types of current queues in the literature
  \item previous impractical universal constructions that use long words
\end{itemize}
  \item Our contribution
\begin{itemize}
  \item polylogarithmic algorithm with respect to $p$ and $Q$ size of the queue
\end{itemize}
  \item Explaining ideas
\begin{itemize}
  \item using tournament tree to agree on the linearization order
  \item double refresh
  \item how to compute a dequeue response
  \item how to be sure the number of blocks in the root is related to the size of the queue
  \item memory management as a goal
\end{itemize}
\end{enumerate}

\pagebreak

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a linked list simultaneously. One solution is to use locks; whenever a process wants to query on a data structure, it locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow parallelism since only one process can make progress in one moment. Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.

\paragraph{}
The question that may arise is, ``What properties matter for a lock-free data structure?''.
Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A safety condition tells us the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.

The standard safety condition is called linearizability, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure 1 is an example of execution on a linearizable queue. The arrow shows time, and rectangles show the time between invocation and termination of an operation. Since ENQ(A) and ENQ(B) are concurrent ENQ(B) may have taken effect before ENQ(A). Execution in figure 2 is not consistent since A has been enqueued before B, so it has to be dequeued first.

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{good}
  \caption{An example of a linearizable execution. Both could take effect first since ENQ(A), and ENQ(B) are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[width=3in]{bad}
  \caption{An example of an execution that is not linearizable. Since ENQ(A) has completed before ENQ(B) is invoked the DEQ() should return A.}
\end{figure}

An algorithm is \textit{wait-free} if each operation terminates after a finite number of steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free. From now on, we are going to focus on wait freedom.

%Since executions on a shared data structure are different from sequential ones, the correctness condition also differs. The strongest correctness condition is linearizability, defined as "...". Lock-free is good, but a wait-free algorithm guarantees us every operation terminates after finite steps.

\paragraph{Previous Work}

In the following paragraphs, we look at previous work on distributed queues.
Michael \& Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. It is included in the standard Java Concurrency Package. Their idea is to store the queue in a singly~linked~list to compute the head and tail of the queue from the head and tail of the linked list. To insert a node to the linked list, they use compare and swap operations. If $p$ processes try to enqueue simultaneously, only one can succeed and the others have to retry. This makes the worst-case number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can $\Omega(p)$ take steps.

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea. The elimination mechanism benefits the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing operations such as pushes and pops to exchange values. Their algorithm makes it possible for failed aged operations to complement. Empirical evaluation of their work is better than MS-queue, but the worst case is still the same, in case there are $p$ concurrent enqueus their algorithm is not better than MS-queue.

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Since the operations in a basket are concurrent, we can order them in any way. Operations in a basket try to find their position in queue one by one by using CAS operations. The operations in different baskets can be executed in parallel. However, like the previous algorithm, if still there are $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues. They use a doubly linked list and do fewer compare and swap operations than MS-queue. But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one. The amortized worst-case complexity is still $\Omega(p)$ CASes.

Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining. Their queue is linearizable but not lock-free. After adding an operation acquiring the lock, they maintain history in publication records and compute all active operations responses. They believe their algorithm in real-world assumptions works well.

Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. Global head and tail point to arrays and content of arrays are marked if dequeued and written by CAS operations. Their data structure is lock-free but it updates lazily. Threads have a cache view on the queue and update it if it gets old.

Kogan and Perank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on MS-queue and use Herlihy's helping technique to achieve wait-freeness. Their step complexity is $\Omega(p)$ because of the helping mechanism.

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching. Their idea of batching allows a sequence of operations to be submitted as a batch for later execution. It supports a new notion introduced by the authors called Extended Medium Futures Linearizability.

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.

%FIFO queues have a wide range of use in OS and applications. The current state-of-the-art queues are implemented using linked lists; that's why they have factor p in their time complexity.

%\href{https://docs.google.com/spreadsheets/d/1cL1tgXXdljkh462sMwkTVMHH_k0MBTmslIVM3xA5VS4/edit#gid=0}{Table of previous works}

%There is a connection between queues and universal constructions. We can implement a universal construction using a queue. We can store operations in the shared queue and compute operations' responses using the queue's content. Some impractical universal constructions are using big words as "...".

\paragraph{Our contribution}
In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList. Our idea is similar to the Jayanti MESD queue.

\bibliography{main} 
\bibliographystyle{ieeetr}

\end{document}

### 2022-02-04 highlevel/highlevel.tex
add highlevel description

\documentclass[12pt]{article}
\usepackage{graphicx}
\begin{document}

\section*{High-level description of the algorithm}

In the following sections, we discuss how the abstract version algorithm works and explain its requirements for implementation. The general idea is to have processes agree on the linearization and then respond fast from the history.

\subsection*{Tree specs}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/tree}
  \caption{An example of a linearizable execution. Both could take effect first since ENQ(A), and ENQ(B) are concurrent.}
\end{figure}
A block is an ordering on some operations. Note that it does not necessarily contain the operations. Each node contains an ordering of blocks appended to it. Process $p_i$ adds $op$ to a block in its node and propagates it to the root, which stores the total ordering.

\paragraph{Propagate(node n):}
\begin{enumerate}
  \item Create a block from $n$'s children's blocks which are not already in $n$
  \item Append the block to $n$
  \item If 2 was not successful, do 1-2 one more time
\end{enumerate}

\paragraph{Requirements}
Creating a block takes less than $O(\log p)$ steps. If some processes try to append their block into node $n$ simultaneously, at least one of them succeeds.

\paragraph{Linearization}
Operations are linearized when they are added to the root. The operations in a block are linearized as the block's ordering. Methods $Get(i)$ and $Index(op)$ give us information about the linearization.

\paragraph{Get(i):} gets the $i$th operation in the total ordering
\begin{enumerate}
  \item $B$= find the block containing $i$th in the root 
\marginpar{\begin{flushleft}\scriptsize{$\log \#ops$ steps}\end{flushleft}}
  \item Get(B, i-pre(B), root) 
\marginpar{\begin{flushleft}\scriptsize{pre(B) is \#operations before B}\end{flushleft}}
\end{enumerate}

\paragraph{Get(B,i,n):} gets the $i$th operation in block B in node n
\begin{enumerate}
  \item $SB$= find the subblock of B containing $i$th op
\marginpar{\begin{flushleft}\scriptsize{$\log p$ steps}\end{flushleft}}
  \item Get(SB, i-pre(SB), B's node)
\end{enumerate}

\paragraph{Requirements} Each level of Get() takes $\log p$ steps, and there are $\log p$ levels. So Get may take $O(\log \#ops + \log^2 p)$. In the following, we discuss that Get is going to be called on the current values of the Queue, so it is going to be $\log Q$.

\paragraph{Index(op):} Where is the position of propagated operation op? It will be done by recursively calling Index(n,b,i) in the path from a leaf up to the root.

\paragraph{Index(n,B,i):}
\begin{enumerate}
  \item $SB$= find the block in n's parent that contains ith op in block B of node n
\marginpar{\begin{flushleft}\scriptsize{$\log p$ steps}\end{flushleft}}
  \item Index(n,B,i+pre(SB))
\end{enumerate}

\paragraph{Requirements} find the superblock of a block in $\log p$ steps

\section*{Transform to a Queue} We are going to augment the root blocks to find out which index is the head of Queue while a Dequeue is linearized.

\paragraph{Enqueue(v):} Add ENQ(v) to the tree
\paragraph{Dequeue():}
\begin{enumerate}
  \item Add DEQ() to the tree
  \item Index the DEQ() order
  \item Compute the index of the ENQ() which is the response of DEQ()
\marginpar{\begin{flushleft}\scriptsize{might be null}\end{flushleft}}
\item Get(i) and return the argument
\end{enumerate}

\paragraph{Requirements}
Augment root blocks so that 3 is fast. Get(i) takes $\log Q+ \log^2 p$.

\end{document}

### 2022-02-08 proposal/proposal.tex
add pics for related work

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 \usepackage{hyperref}
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{good}
+  \center\includegraphics[width=4in]{pics/good}
   \caption{An example of a linearizable execution. Both could take effect first since ENQ(A), and ENQ(B) are concurrent.}
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=3in]{bad}
+  \center\includegraphics[width=3in]{pics/bad}
   \caption{An example of an execution that is not linearizable. Since ENQ(A) has completed before ENQ(B) is invoked the DEQ() should return A.}
@@
 It is included in the standard Java Concurrency Package.
-Their idea is to store the queue in a singly~linked~list to compute the head and tail of the queue from the head and tail of the linked list.
-To insert a node to the linked list, they use compare and swap operations.
+Their idea is to store the queue elements in a singly~linked~list.
+Head points to the first not dequeued node in the linked list and tail points to last element in the queue.
+To insert a node to the linked list, they use atomic primitive operations like LL/SC or CAS.
 If $p$ processes try to enqueue simultaneously, only one can succeed and the others have to retry.
@@
 
+\begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/msq}
+  \caption{MS-queue structure, Enquque and Dequeue operations}
+\end{figure}
+
 Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea.
@@
 An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing operations such as pushes and pops to exchange values.
+Elimination works by allowing opposing operations such as consecutive enqueues and dequeues in the elimination array to exchange values.
 Their algorithm makes it possible for failed aged operations to complement.
-Empirical evaluation of their work is better than MS-queue, but the worst case is still the same, in case there are $p$ concurrent enqueus their algorithm is not better than MS-queue.
+Empirical evaluation of their work is better than MS-queue, but the worst case is still the same, in case there are $p$ concurrent enqueues their algorithm is not better than MS-queue.
 
@@
 Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
+Each basket has a time interval in which all its nodes’ enqueue operations overlap.
 Since the operations in a basket are concurrent, we can order them in any way.
-Operations in a basket try to find their position in queue one by one by using CAS operations.
-The operations in different baskets can be executed in parallel.
-However, like the previous algorithm, if still there are $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
+Operations in a basket try to find their order in the basket one by one by using CAS operations.
+However, like the previous algorithms, if still there are $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/baskets}
+  \caption{Baskets queue idea}
+\end{figure}
 

### 2022-02-22 proposal/proposal.tex
minor edit on propsal and added our contribution

@@
 \documentclass[12pt]{article}
+\usepackage[a4paper]{geometry}
 \usepackage{hyperref}
 \usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{tikz-qtree}
 
@@
 
-\begin{enumerate}
-  \item Importance of implementing shared data structures
-\begin{itemize}
-  \item how it helps developers
-  \item why lock free data structures are better than using locks?
-\end{itemize} 
-  \item  What properties matter for a distributed data structure?
-\begin{itemize}
-  \item Correctness condition: Linearizability definition
-  \item Wait-freeness
-\end{itemize} 
-  \item Previous works on distributed queues: more detailed
-\begin{itemize}
-  \item why we are studying queus
-  \item $p$ factor in time complexity of all previous algorithms
-  \item types of current queues in the literature
-  \item previous impractical universal constructions that use long words
-\end{itemize}
-  \item Our contribution
-\begin{itemize}
-  \item polylogarithmic algorithm with respect to $p$ and $Q$ size of the queue
-\end{itemize}
-  \item Explaining ideas
-\begin{itemize}
-  \item using tournament tree to agree on the linearization order
-  \item double refresh
-  \item how to compute a dequeue response
-  \item how to be sure the number of blocks in the root is related to the size of the queue
-  \item memory management as a goal
-\end{itemize}
-\end{enumerate}
+%\begin{enumerate}
+%  \item Importance of implementing shared data structures
+%\begin{itemize}
+%  \item how it helps developers
+%  \item why lock free data structures are better than using locks?
+%\end{itemize} 
+%  \item  What properties matter for a distributed data structure?
+%\begin{itemize}
+%  \item Correctness condition: Linearizability definition
+%  \item Wait-freeness
+%\end{itemize} 
+%  \item Previous works on distributed queues: more detailed
+%\begin{itemize}
+%  \item why we are studying queues
+%  \item $p$ factor in time complexity of all previous algorithms
+%  \item types of current queues in the literature
+%  \item previous impractical universal constructions that use long words
+%\end{itemize}
+%  \item Our contribution
+%\begin{itemize}
+%  \item polylogarithmic algorithm with respect to $p$ and $Q$ size of the queue
+%\end{itemize}
+%  \item Explaining ideas
+%\begin{itemize}
+%  \item using tournament tree to agree on the linearization order
+%  \item double refresh
+%  \item how to compute a dequeue response
+%  \item how to be sure the number of blocks in the root is related to the size of the queue
+%  \item memory management as a goal
+%\end{itemize}
+%\end{enumerate}
 
@@
   \center\includegraphics[width=4in]{pics/msq}
-  \caption{MS-queue structure, Enquque and Dequeue operations}
+  \caption{MS-queue structure, Enqueue and Dequeue operations}
 \end{figure}
@@
 An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing operations such as consecutive enqueues and dequeues in the elimination array to exchange values.
+Elimination works by allowing opposing operations such as consecutive enqueues and dequeues when the queue is initially empty in the elimination array to exchange values.
 Their algorithm makes it possible for failed aged operations to complement.
@@
 Operations in a basket try to find their order in the basket one by one by using CAS operations.
-However, like the previous algorithms, if still there are $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
+However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
 
@@
 \paragraph{Our contribution}
-In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList. Our idea is similar to the Jayanti MESD queue.
+In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList.
+Our idea is similar to the Jayanti MESD queue.
+But we do not use CAS operations with big words.
+
+\paragraph{}
+Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
+He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
+We introduce a universal construction using $\textsc{O}(\log p)$ CAS per operation~\cite{DBLP:conf/fsttcs/JayantiP05} called Block tree.
+We use the block tree as a stepping stone towards our queue algorithm.
+\paragraph{}
+A block tree is a Tournament tree shared among $p$ processes.
+Each process has a leaf, and it appends its operations to its leaf.
+After that, the process tries to propagate up to the tree's root.
+An ordering is stored in each node of the tree of operations propagated to the node.
+All processes agree on the sequence stored in the root (Linearization ordering).
+
+\begin{center}
+\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
+          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
+\end{center}
+
+The goal is to make each propagate step ensure that the operations are propagated up to the root and fast.
+
+Our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using CAS operations.
+The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
+Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$) which has appended the talked operations.
+
+To make CAS fast, we do not append operations explicitly, but only remember their number. This allows us to CAS fixed-size objects in each \textsc{Refresh}($n$).
+
+\begin{figure}
+\begin{center}
+\begin{subfigure}[b]{.49\textwidth}
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}}
+  \caption{Operations after $||$ are new.}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{.49\textwidth}
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}}
+  \caption{New operations are added to the parent node.}
+\end{subfigure}
+\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
+\end{center}
+\end{figure}
+
+We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization.
+Having methods Append(op), Get(i), Index(op), we can have the linearization shared among the processes.
+To have a queue, we will implement Enqueue and Dequeue using our block tree.
+An Enqueue can be done using just an Append operation in the block tree.
+Dequeue() consists of an Append, Index, compute the corresponding response and return it using Get().
+
+Our future work is to manage the memory of the tree, not to have infinite orderings.
 

### 2022-02-25 proposal/proposal.tex
added remainig parts to the proposal

@@
 
+\section{Introduction}
+
 \paragraph{}
@@
 It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a linked list simultaneously.
-One solution is to use locks; whenever a process wants to query on a data structure, it locks it, and others cannot use it until the lock is released.
+One solution is to use locks; whenever a process wants to do an update query on a data structure, it locks it, and others cannot use it until the lock is released.
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
@@
 
-\paragraph{Previous Work}
+\section{Previous Work}
 
-In the following paragraphs, we look at previous work on distributed queues.
+In the following paragraphs, we look at previous lock-free queues.
 Michael \& Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
@@
 Their idea is to store the queue elements in a singly~linked~list.
-Head points to the first not dequeued node in the linked list and tail points to last element in the queue.
+Head points to the first not dequeued node in the linked list and tail points to the last element in the queue.
 To insert a node to the linked list, they use atomic primitive operations like LL/SC or CAS.
-If $p$ processes try to enqueue simultaneously, only one can succeed and the others have to retry.
+If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
 This makes the worst-case number of steps to be $\Omega(p)$ per enqueue.
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/msq}
+  \center\includegraphics[width=4in]{pics/msqueue}
   \caption{MS-queue structure, Enqueue and Dequeue operations}
@@
 Their algorithm makes it possible for failed aged operations to complement.
-Empirical evaluation of their work is better than MS-queue, but the worst case is still the same, in case there are $p$ concurrent enqueues their algorithm is not better than MS-queue.
+Empirical evaluation of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
 
@@
 Global head and tail point to arrays and content of arrays are marked if dequeued and written by CAS operations.
-Their data structure is lock-free but it updates lazily.
+Their data structure is lock-free, but it updates lazily.
 Threads have a cache view on the queue and update it if it gets old.
@@
 
-Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
+%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
+In the worst-case step complexity of all the papers discussed above, there is a $p$ term which comes from the case all $p$ processes try to do an enqueue at the same time.
+We are focusing on seeing if we can do this in sublinear in terms of $p$ or not.
 
@@
 
-\paragraph{Our contribution}
+\section{Our contribution}
 In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList.
-Our idea is similar to the Jayanti MESD queue.
-But we do not use CAS operations with big words.
-
-\paragraph{}
+Our idea is similar to the Jayanti MESD queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use CAS operations with big words.
+The new work about our idea is that even if there are $p$ processes trying to enqueue simultaneously, each enqueue takes $\log p^2 +Q$.
 Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
 He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
-We introduce a universal construction using $\textsc{O}(\log p)$ CAS per operation~\cite{DBLP:conf/fsttcs/JayantiP05} called Block tree.
+We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ CAS per operation.
 We use the block tree as a stepping stone towards our queue algorithm.
 \paragraph{}
-A block tree is a Tournament tree shared among $p$ processes.
+A block tree is a tournament tree shared among $p$ processes.
 Each process has a leaf, and it appends its operations to its leaf.
@@
 All processes agree on the sequence stored in the root (Linearization ordering).
-
+\begin{figure}
 \begin{center}
@@
 \end{center}
+\caption{Each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root which the final ordering is stored.}  
+\end{figure}
 
-The goal is to make each propagate step ensure that the operations are propagated up to the root and fast.
+The goal is to make each propagate step ensure that the operations are propagated up to the root and fast. And then use the linearization ordering to answer the dequeue operations.
 
-Our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using CAS operations.
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using CAS operations.
 The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
 Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$) which has appended the talked operations.
-
-To make CAS fast, we do not append operations explicitly, but only remember their number. This allows us to CAS fixed-size objects in each \textsc{Refresh}($n$).
+To make CAS fast, we do not append operations explicitly, but only remember their number.
+This allows us to CAS fixed-size objects in each \textsc{Refresh}($n$).
+To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure.
 
@@
 \end{subfigure}
-\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
+\caption{\label{fig:ucexample} Propagate Step, operations in children after || are new.}
 \end{center}
@@
 We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization.
-Having methods Append(op), Get(i), Index(op), we can have the linearization shared among the processes.
-To have a queue, we will implement Enqueue and Dequeue using our block tree.
+Get(i) finds the block containing $i$th in the root and then finds its subblock recursively to reach a leaf.
+Index() is somewhat similar but more complicated from bottom to top; the main challenge is that each level of the tree should take polylogarithmic steps with respect to $p$.
+Having methods Get(i), Index(op), we can have the linearization shared among the processes.
+We will implement Enqueue and Dequeue using our block tree to have a queue.
 An Enqueue can be done using just an Append operation in the block tree.
-Dequeue() consists of an Append, Index, compute the corresponding response and return it using Get().
+Dequeue() consists of an Append, Index, compute the corresponding response, and return it using Get().
 
-Our future work is to manage the memory of the tree, not to have infinite orderings.
+However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing $i$th operation in the root level.
+We make it fast by storing prefix sums but still, it is proportional to the number of all operations.
+We come with a garbage collection only for the root level to reach $\log Q$.
+The main idea is to make use of the fast split of persistent red-black trees.
+
+\section{Next steps}
+The current algorithm only works as a queue, but other data structures like Stacks share similarities.
+It might be possible to come up with a generalization for all these types.
+We do not care about the space, and we use infinite-sized arrays that are not practical.
+Our future work is to manage the memory of the tree and make it practical.
 

### 2022-03-01 proposal/proposal.tex
add planning section

@@
 \title{Thesis Proposal}
-\author{Hossein Naderi}
+\author{Hossein Naderi \and Eric Ruppert}
 \maketitle
@@
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow parallelism since only one process can make progress in one moment.
+Moreover, locks do not allow complete parallelism since only one process holding the lock can make progress in the moment.
 Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.
 
-\paragraph{}
 The question that may arise is, ``What properties matter for a lock-free data structure?''.
@@
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A safety condition tells us the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.
+A safety condition tells us that the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.
 
-The standard safety condition is called linearizability, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure 1 is an example of execution on a linearizable queue.
+The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
+Figure \ref{fig::goodexample} is an example of execution on a linearizable empty queue.
 The arrow shows time, and rectangles show the time between invocation and termination of an operation.
-Since ENQ(A) and ENQ(B) are concurrent ENQ(B) may have taken effect before ENQ(A).
-Execution in figure 2 is not consistent since A has been enqueued before B, so it has to be dequeued first.
+Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not have taken effect before \texttt{Enqueue(A)}.
+Execution in figure \ref{fig::badexample} is not consistent since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
 
 \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/good}
-  \caption{An example of a linearizable execution. Both could take effect first since ENQ(A), and ENQ(B) are concurrent.}
+  \center\includegraphics[scale=0.5]{pics/good}
+  \caption{\label{fig::goodexample}An example of a linearizable execution. Both could take effect first since \texttt{Enqueue(A)}, and \texttt{Enqueue(B)} are concurrent.}
 \end{figure}
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=3in]{pics/bad}
-  \caption{An example of an execution that is not linearizable. Since ENQ(A) has completed before ENQ(B) is invoked the DEQ() should return A.}
+  \center\includegraphics[scale=0.5]{pics/bad}
+  \caption{\label{fig::badexample}An example of an execution that is not linearizable.
+Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A}.}
 \end{figure}
@@
 A wait-free algorithm is also lock-free.
-From now on, we are going to focus on wait freedom.
+%From now on, we are going to focus on wait freedom.
 
@@
 Head points to the first not dequeued node in the linked list and tail points to the last element in the queue.
-To insert a node to the linked list, they use atomic primitive operations like LL/SC or CAS.
+To insert a node to the linked list, they use atomic primitive operations like LL/SC or \texttt{CAS}.
 If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the worst-case number of steps to be $\Omega(p)$ per enqueue.
-Similarly, dequeue can $\Omega(p)$ take steps.
+This makes the worst-\texttt{CAS}e number of steps to be $\Omega(p)$ per enqueue.
+Similarly, dequeue can take $\Omega(p)$ steps.
 
 \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/msqueue}
+  \center\includegraphics[scale=0.5]{pics/msqueue}
   \caption{MS-queue structure, Enqueue and Dequeue operations}
@@
 An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing operations such as consecutive enqueues and dequeues when the queue is initially empty in the elimination array to exchange values.
+Elimination works by allowing opposing operations such as consecutive enqueues and dequeues  in the elimination array to exchange values when the queue is empty.
 Their algorithm makes it possible for failed aged operations to complement.
-Empirical evaluation of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+Empirical evaluation of their work is better than MS-queue, but the worst \texttt{CAS}e is still the same; in \texttt{CAS}e there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
 
@@
 Since the operations in a basket are concurrent, we can order them in any way.
-Operations in a basket try to find their order in the basket one by one by using CAS operations.
+Operations in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
 However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/baskets}
+  \center\includegraphics[scale=0.3]{pics/baskets}
   \caption{Baskets queue idea}
@@
 They use a doubly linked list and do fewer compare and swap operations than MS-queue.
-But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one.
-The amortized worst-case complexity is still $\Omega(p)$ CASes.
+But as before, the worst \texttt{CAS}e is when there are $p$ concurrent enqueues which have to be enqueued one by one.
+The amortized worst-\texttt{CAS}e complexity is still $\Omega(p)$ \texttt{CAS}es.
 
@@
 Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-Global head and tail point to arrays and content of arrays are marked if dequeued and written by CAS operations.
+Global head and tail point to arrays and content of arrays are marked if dequeued and written by \texttt{CAS} operations.
 Their data structure is lock-free, but it updates lazily.
@@
 %Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
-In the worst-case step complexity of all the papers discussed above, there is a $p$ term which comes from the case all $p$ processes try to do an enqueue at the same time.
-We are focusing on seeing if we can do this in sublinear in terms of $p$ or not.
+In the worst-\texttt{CAS}e step complexity of all the papers discussed above, there is a $p$ term which comes from the \texttt{CAS}e all $p$ processes try to do an enqueue at the same time.
+We are focusing to see if we can do this in sublinear steps in terms of $p$ or not.
 
@@
 \section{Our contribution}
-In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList.
-Our idea is similar to the Jayanti MESD queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use CAS operations with big words.
-The new work about our idea is that even if there are $p$ processes trying to enqueue simultaneously, each enqueue takes $\log p^2 +Q$.
-Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
-He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
-We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ CAS per operation.
-We use the block tree as a stepping stone towards our queue algorithm.
-\paragraph{}
-A block tree is a tournament tree shared among $p$ processes.
+In this work, we design a queue with $\log^2 p +Q$ steps per operation, outperforming queues built by the LinkedList approach.
+Our idea is similar to Jayanti 's Multi Enqueuer-Single Dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words.
+The new work about our idea is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $\log^2 p +Q$.
+Jayanti proved an $\Omega(\log p)$ lower bound on the worst-\texttt{CAS}e shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
+He also introduced~ a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
+We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ \texttt{CAS} per operation called \textit{Block tree}.
+Then we use the block tree as a stepping stone towards our queue algorithm.
+A block tree is a tournament tree shared among $p$ processes (See Figure~\ref{fig::blocktree}).
 Each process has a leaf, and it appends its operations to its leaf.
@@
 \end{center}
-\caption{Each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root which the final ordering is stored.}  
+\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
+Each node tries to propagate its operations up to the root which the final ordering is stored.}
 \end{figure}
 
-The goal is to make each propagate step ensure that the operations are propagated up to the root and fast. And then use the linearization ordering to answer the dequeue operations.
+The goal here is to ensure in each propagate step, the new operations are propagated up to the parent in $\log p$ steps (See Figure \ref{fig::propagstep}).
+And then use the linearization ordering to answer the dequeue operations.
 
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using CAS operations.
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using \texttt{CAS} operations.
 The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
-Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$) which has appended the talked operations.
-To make CAS fast, we do not append operations explicitly, but only remember their number.
-This allows us to CAS fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure.
+Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
+Instead of storing operations explicitly in the nodes, we only keep track of the number of them.
+This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
+To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.
 
@@
 \end{tikzpicture}}
-  \caption{Operations after $||$ are new.}
+  \caption{\label{fig::propagstep}Operations after $||$ are new.}
 \end{subfigure}
@@
 We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization.
-Get(i) finds the block containing $i$th in the root and then finds its subblock recursively to reach a leaf.
+Get(i) finds the block containing the $i$th in the root and then finds its sub-block recursively to reach a leaf.
 Index() is somewhat similar but more complicated from bottom to top; the main challenge is that each level of the tree should take polylogarithmic steps with respect to $p$.
 Having methods Get(i), Index(op), we can have the linearization shared among the processes.
-We will implement Enqueue and Dequeue using our block tree to have a queue.
-An Enqueue can be done using just an Append operation in the block tree.
-Dequeue() consists of an Append, Index, compute the corresponding response, and return it using Get().
+Now we can implement Enqueue and Dequeue using our block tree.
+An \texttt{Enqueue(e)} can be done using just an \texttt{Append()} operation with input argument \texttt{e} in the block tree.
+\texttt{Dequeue()} consists of an \texttt{Append()}, \texttt{Index()}, compute the corresponding response, and return it using \texttt{Get()}.
 
-However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing $i$th operation in the root level.
-We make it fast by storing prefix sums but still, it is proportional to the number of all operations.
+\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the containing sub-block or super-block of the given operation.
+We use prefix sum for binary search and some extra constant information that we do not discuss here.
+
+However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing the $i$th operation at the root level.
 We come with a garbage collection only for the root level to reach $\log Q$.
 The main idea is to make use of the fast split of persistent red-black trees.
+So our algorithm works in $O(\log^2 p +Q)$ steps.
 
@@
 
+\section{Planning}
+The plan is to finish the writing of the thesis by Aug 2021. The thesis will contain:
+\begin{itemize}
+  \item More detailed discussion of previous work
+  \item Description of the algorithm, layer by layer
+  \item Pseudocode of the algorithm in two parts; regarding memory management
+  \item Proof of linearizability and time analysis
+\end{itemize}
+
+We expect to submit a paper of our work to \textit{36th Int-| Symposium on Distributed Computing
+October 25-27, 2022, Augusta, Georgia, USA} by May 13.
+
 \bibliography{main} 

### 2022-03-02 proposal/proposal.tex
fix typo

@@
 \title{Thesis Proposal}
-\author{Hossein Naderi \and Eric Ruppert}
+\author{Hossein Naderi \\supervised by Eric Ruppert}
 \maketitle
@@
 If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the worst-\texttt{CAS}e number of steps to be $\Omega(p)$ per enqueue.
+This makes the worst-case number of steps to be $\Omega(p)$ per enqueue.
 Similarly, dequeue can take $\Omega(p)$ steps.
@@
 Their algorithm makes it possible for failed aged operations to complement.
-Empirical evaluation of their work is better than MS-queue, but the worst \texttt{CAS}e is still the same; in \texttt{CAS}e there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+Empirical evaluation of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
 
@@
 They use a doubly linked list and do fewer compare and swap operations than MS-queue.
-But as before, the worst \texttt{CAS}e is when there are $p$ concurrent enqueues which have to be enqueued one by one.
-The amortized worst-\texttt{CAS}e complexity is still $\Omega(p)$ \texttt{CAS}es.
+But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one.
+The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
 
@@
 %Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
-In the worst-\texttt{CAS}e step complexity of all the papers discussed above, there is a $p$ term which comes from the \texttt{CAS}e all $p$ processes try to do an enqueue at the same time.
+In the worst-case step complexity of all the papers discussed above, there is a $p$ term which comes from the case all $p$ processes try to do an enqueue at the same time.
 We are focusing to see if we can do this in sublinear steps in terms of $p$ or not.
@@
 The new work about our idea is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $\log^2 p +Q$.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-\texttt{CAS}e shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
+Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
 He also introduced~ a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.

### 2022-03-09 proposal/proposal.tex
apply eric's feedback

@@
 This is why distributed algorithms have become important.
-It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a linked list simultaneously.
-One solution is to use locks; whenever a process wants to do an update query on a data structure, it locks it, and others cannot use it until the lock is released.
+It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a sequential linked list simultaneously, processes $p,q$ read the same tail node, $p$ changes the pointer of tail to its new node and after that $q$ does the same.
+In this run, $p$'s update is overwritten.
+One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released.
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only one process holding the lock can make progress in the moment.
+Moreover, locks do not allow complete parallelism since only one process holding the lock can make progress.
 Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.
@@
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A safety condition tells us that the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.
+A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
 
@@
 Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not have taken effect before \texttt{Enqueue(A)}.
-Execution in figure \ref{fig::badexample} is not consistent since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
+The execution in Figure \ref{fig::badexample} is not consistent since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
 
@@
   \center\includegraphics[scale=0.5]{pics/good}
-  \caption{\label{fig::goodexample}An example of a linearizable execution. Both could take effect first since \texttt{Enqueue(A)}, and \texttt{Enqueue(B)} are concurrent.}
+  \caption{\label{fig::goodexample}An example of a linearizable execution. Either could take effect first since \texttt{Enqueue(A)}, and \texttt{Enqueue(B)} are concurrent.}
 \end{figure}
@@
 
+There are various progress properties; the strongest is wait-freedom, and the more common is loc-freedom.
 An algorithm is \textit{wait-free} if each operation terminates after a finite number of steps.
@@
 
-\section{Previous Work}
-
+\section{Related Work}
+\subsection{List based queues}
 In the following paragraphs, we look at previous lock-free queues.
-Michael \& Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
+Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
 It is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly~linked~list.
-Head points to the first not dequeued node in the linked list and tail points to the last element in the queue.
-To insert a node to the linked list, they use atomic primitive operations like LL/SC or \texttt{CAS}.
+Their idea is to store the queue elements in a singly-linked list (See Figure~\ref{fig::msq}).
+Head points to the first node in the linked list that has not been dequeued, and tail points to the last element in the queue.
+To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
 If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
@@
   \center\includegraphics[scale=0.5]{pics/msqueue}
-  \caption{MS-queue structure, Enqueue and Dequeue operations}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.}
 \end{figure}
@@
 Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea.
-The elimination mechanism benefits the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
+The elimination mechanism has the dual purpose of allowing operations to be complete in parallel and reducing contention for the queue.
 An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing operations such as consecutive enqueues and dequeues  in the elimination array to exchange values when the queue is empty.
+Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
 Their algorithm makes it possible for failed aged operations to complement.
-Empirical evaluation of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+The empirical evaluation showed the throughput of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
 
@@
 Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes’ enqueue operations overlap.
+Each basket has a time interval in which all its nodes' enqueue operations overlap.
 Since the operations in a basket are concurrent, we can order them in any way.
-Operations in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
-However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.
+Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
+However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$ per operation.
 
@@
   \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption{Baskets queue idea}
+  \caption{Baskets queue idea.}
 \end{figure}
@@
 Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues.
-They use a doubly linked list and do fewer compare and swap operations than MS-queue.
+They use a doubly-linked list and do fewer \texttt{CAS} operations than MS-queue.
 But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one.
@@
 After adding an operation acquiring the lock, they maintain history in publication records and compute all active operations responses.
-They believe their algorithm in real-world assumptions works well.
+They present experiments that show their algorithm performs well in some situations.
 
 Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-Global head and tail point to arrays and content of arrays are marked if dequeued and written by \texttt{CAS} operations.
-Their data structure is lock-free, but it updates lazily.
-Threads have a cache view on the queue and update it if it gets old.
+Global head and tail pointers point to arrays containing the first and last element in the queue and the elements when they are written by \texttt{CAS} operations and dequeued.
+Their data structure is lock-free.
+An enqueue or a dequeue searches for the array head or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
+Still, if the head array is empty and $p$ processes try to enqueue simultaneously, step complexity remains $\Omega(p)$.
 
-Kogan and Perank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on MS-queue and use Herlihy's helping technique to achieve wait-freeness.
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.5]{pics/sundell}
+    \caption{\label{fig::sundell}Global pointer point to arrays, dequeued elements are red and current elements are in green.}
+\end{figure}
+
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on MS-queue and use Herlihy's helping technique to achieve wait-freedom.
 Their step complexity is $\Omega(p)$ because of the helping mechanism.
 
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
+%todo: comparison
+%Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
 Their idea of batching allows a sequence of operations to be submitted as a batch for later execution.
@@
 %Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
-In the worst-case step complexity of all the papers discussed above, there is a $p$ term which comes from the case all $p$ processes try to do an enqueue at the same time.
-We are focusing to see if we can do this in sublinear steps in terms of $p$ or not.
+In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously.
+Morrison and Afek introduce this execution as \textit{CAS retry problem} ~\cite{DBLP:conf/ppopp/MorrisonA13}.
+We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.
+
+\subsection{Universal constructions}
+Herlihy introduced the consensus hierarchy and discussed the possibility of implementation of shared objects from other objects~\cite{10.1145/114005.102808}.
+A universal construction is an object which if it implements any other objects.
+We can implement a concurrent queue using a universal construction.
+Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
+He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
+His universal construction can be used to create any data structure, but his implementation is not practical because of using unreasonably sized \texttt{CAS} operations.
+
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of log $p$ using $O(\log n$)bit \texttt{LL/SC} objects, which $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.
 
@@
 
-\section{Our contribution}
-In this work, we design a queue with $\log^2 p +Q$ steps per operation, outperforming queues built by the LinkedList approach.
-Our idea is similar to Jayanti 's Multi Enqueuer-Single Dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words.
-The new work about our idea is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $\log^2 p +Q$.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also introduced~ a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ \texttt{CAS} per operation called \textit{Block tree}.
+\section{Our Contribution}
+In this work, we design a queue with $\log^2 p +\log q$ steps per operation, which $q$ is the size of the queue, outperforming queues built by the LinkedList approach.
+Our idea is similar to Jayanti's Multi Enqueuer-Single Dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
+The new feature of our implementation is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $O(\log^2 p +\log q)$ steps.
+We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
 Then we use the block tree as a stepping stone towards our queue algorithm.
-A block tree is a tournament tree shared among $p$ processes (See Figure~\ref{fig::blocktree}).
+A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}).
 Each process has a leaf, and it appends its operations to its leaf.
-After that, the process tries to propagate up to the tree's root.
-An ordering is stored in each node of the tree of operations propagated to the node.
-All processes agree on the sequence stored in the root (Linearization ordering).
+After that, the process tries to propagate its new operation up to the tree's root.
+An ordering of operations propagated up to a node is stored in that node.
+All processes agree on the sequence stored in the root and this is used as the linearization ordering.
 \begin{figure}
@@
 \caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root which the final ordering is stored.}
+Each node tries to propagate its operations up to the root, which stores the final ordering.}
 \end{figure}
 
-The goal here is to ensure in each propagate step, the new operations are propagated up to the parent in $\log p$ steps (See Figure \ref{fig::propagstep}).
-And then use the linearization ordering to answer the dequeue operations.
+The goal here is to ensure in each propagate step; the new operations are propagated up to the parent in $\log p$ steps (See~Figure~\ref{fig::propagstep}).
+Then a dequeue operation uses the linearization ordering to compute its answer.
 
 In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using \texttt{CAS} operations.
-The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
-Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
+The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in \texttt{n}'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
+%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
 Instead of storing operations explicitly in the nodes, we only keep track of the number of them.
@@
 \end{subfigure}
-\caption{\label{fig:ucexample} Propagate Step, operations in children after || are new.}
+\caption{\label{fig:ucexample} Successful \texttt{Refresh}, operations in children after $||$ are new.}
 \end{center}
@@
 
-We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization.
-Get(i) finds the block containing the $i$th in the root and then finds its sub-block recursively to reach a leaf.
-Index() is somewhat similar but more complicated from bottom to top; the main challenge is that each level of the tree should take polylogarithmic steps with respect to $p$.
-Having methods Get(i), Index(op), we can have the linearization shared among the processes.
+We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
+\texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf.
+\texttt{Index()} is  similar but more complicated, finding superblocks from a leaf to the root.
+The main challenge in each level of the \texttt{Get(i)}, \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
+After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)}, \texttt{Index(op)}.
 Now we can implement Enqueue and Dequeue using our block tree.
-An \texttt{Enqueue(e)} can be done using just an \texttt{Append()} operation with input argument \texttt{e} in the block tree.
-\texttt{Dequeue()} consists of an \texttt{Append()}, \texttt{Index()}, compute the corresponding response, and return it using \texttt{Get()}.
+An \texttt{Enqueue(e)} is appending an operation with input argument \texttt{e} in the block tree.
+To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree.
+Then finds the rank of the \texttt{DEQ} using \texttt{Index()}, after knowing the rank of \texttt{DEQ} it computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
+Finally returns the argument of the answer \texttt{ENQ} using \texttt{Get(i)}.
 
-\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the containing sub-block or super-block of the given operation.
-We use prefix sum for binary search and some extra constant information that we do not discuss here.
+\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation.
+We use prefix sum for binary search and a constant amount of information that we do not discuss here.
 
-However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing the $i$th operation at the root level.
-We come with a garbage collection only for the root level to reach $\log Q$.
-The main idea is to make use of the fast split of persistent red-black trees.
-So our algorithm works in $O(\log^2 p +Q)$ steps.
+The algorithm works as a queue, but \texttt{Get(i)}  may take a long time since it has to find the block containing the $i$th operation at the root level.
+We came up with a garbage collection only for the root level to reduce the time to find this block to $O(\log q)$.
+The main idea is to remove a block after its operations are finished.
+We can search in Persistent red-black trees with size $n$ in $O(\log n)$ steps and split the tree by a key in $O(\log n)$ steps.
+We will store the blocks in the root in a persistent red-black tree to be able only to maintain the remaining information of the queue.
+So our algorithm works in $O(\log^2 p +\log q)$ steps.
 
 \section{Next steps}
-The current algorithm only works as a queue, but other data structures like Stacks share similarities.
+The current algorithm only works as a queue, but other data structures like stacks share similarities.
 It might be possible to come up with a generalization for all these types.
-We do not care about the space, and we use infinite-sized arrays that are not practical.
+We have not yet considered space usage, and we use infinite-sized arrays that are not practical.
 Our future work is to manage the memory of the tree and make it practical.
@@
 \section{Planning}
-The plan is to finish the writing of the thesis by Aug 2021. The thesis will contain:
+The plan is to finish the writing of the thesis by August 2021. The thesis will contain:
 \begin{itemize}
@@
 
-We expect to submit a paper of our work to \textit{36th Int-| Symposium on Distributed Computing
+We expect to submit a paper of our work to \textit{36th International Symposium on Distributed Computing
 October 25-27, 2022, Augusta, Georgia, USA} by May 13.

### 2022-03-17 proposal/proposal.tex
apply eric's feedback

@@
 This is why distributed algorithms have become important.
-It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a sequential linked list simultaneously, processes $p,q$ read the same tail node, $p$ changes the pointer of tail to its new node and after that $q$ does the same.
+It is not hard to see why multiple processes cannot update sequential data structures designed for one process.
+For example, consider two processes trying to insert some values into a sequential linked list simultaneously.
+Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same.
 In this run, $p$'s update is overwritten.
@@
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only one process holding the lock can make progress.
-Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.
+Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
+Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
+Also there is a common problem in previous shared queues we're going to resolve which make their worst case slow.
 
@@
 The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} is an example of execution on a linearizable empty queue.
-The arrow shows time, and rectangles show the time between invocation and termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not have taken effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not consistent since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
+Figure \ref{fig::goodexample} is an example of execution on a linearizable  queue that is initially empty.
+The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
+Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
+The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
 
-\begin{figure}[hbt]
+\begin{figure}[t]
   \center\includegraphics[scale=0.5]{pics/good}
-  \caption{\label{fig::goodexample}An example of a linearizable execution. Either could take effect first since \texttt{Enqueue(A)}, and \texttt{Enqueue(B)} are concurrent.}
+  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
 \end{figure}
 
-\begin{figure}[hbt]
+\begin{figure}[t]
   \center\includegraphics[scale=0.5]{pics/bad}
@@
 
-There are various progress properties; the strongest is wait-freedom, and the more common is loc-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of steps.
+There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
+An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
 We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free.
+A wait-free algorithm is also lock-free but not vice versa since a lock-free algorithm may terminate after an exponential number of steps.
 %From now on, we are going to focus on wait freedom.
@@
 Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-It is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (See Figure~\ref{fig::msq}).
-Head points to the first node in the linked list that has not been dequeued, and tail points to the last element in the queue.
+A version of it is included in the standard Java Concurrency Package.
+Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
+Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue.
 To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
 If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the worst-case number of steps to be $\Omega(p)$ per enqueue.
+This makes the amortized number of steps to be $\Omega(p)$ per enqueue.
 Similarly, dequeue can take $\Omega(p)$ steps.
 
-\begin{figure}[hbt]
+\begin{figure}[t]
   \center\includegraphics[scale=0.5]{pics/msqueue}
-  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrow shows new pointer and gray dashed arrow shows the old arrow.}
 \end{figure}
 
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea.
-The elimination mechanism has the dual purpose of allowing operations to be complete in parallel and reducing contention for the queue.
+Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination technique.
+The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
 An Elimination Queue consists of an MS-queue augmented with an elimination array.
 Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for failed aged operations to complement.
+Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
 The empirical evaluation showed the throughput of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
@@
 Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
-However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$ per operation.
+However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
 
-\begin{figure}[hbt]
+\begin{figure}[t]
   \center\includegraphics[scale=0.3]{pics/baskets}
@@
 Their queue is linearizable but not lock-free.
-After adding an operation acquiring the lock, they maintain history in publication records and compute all active operations responses.
+Their main idea is that with knowledge of all the history of operations, it might be possible to answer queries faster than doing them one by one.
+In our work we also maintain the whole history.
 They present experiments that show their algorithm performs well in some situations.
@@
 Their data structure is lock-free.
-An enqueue or a dequeue searches for the array head or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
-Still, if the head array is empty and $p$ processes try to enqueue simultaneously, step complexity remains $\Omega(p)$.
+An enqueue or a dequeue searches for the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
+Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.
 
-\begin{figure}[hbt]
+\begin{figure}[t]
   \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption{\label{fig::sundell}Global pointer point to arrays, dequeued elements are red and current elements are in green.}
+    \caption{\label{fig::sundell}Global pointer point to arrays. Head and Tail elements are blue dequeued elements are red and current elements of the queue are in green.}
 \end{figure}
@@
 In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously.
-Morrison and Afek introduce this execution as \textit{CAS retry problem} ~\cite{DBLP:conf/ppopp/MorrisonA13}.
+Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
+It is not limited to list-based queues and array-based queues share CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} .
 We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.
@@
 \subsection{Universal constructions}
-Herlihy introduced the consensus hierarchy and discussed the possibility of implementation of shared objects from other objects~\cite{10.1145/114005.102808}.
-A universal construction is an object which if it implements any other objects.
+Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
+A \textit{universal construction} is an object that can implement a shared object from any given sequential object.
 We can implement a concurrent queue using a universal construction.
@@
 He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-His universal construction can be used to create any data structure, but his implementation is not practical because of using unreasonably sized \texttt{CAS} operations.
+His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.
 
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of log $p$ using $O(\log n$)bit \texttt{LL/SC} objects, which $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of log $p$ using $O(\log n$)bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
 Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.
@@
 \section{Our Contribution}
-In this work, we design a queue with $\log^2 p +\log q$ steps per operation, which $q$ is the size of the queue, outperforming queues built by the LinkedList approach.
-Our idea is similar to Jayanti's Multi Enqueuer-Single Dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
-The new feature of our implementation is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $O(\log^2 p +\log q)$ steps.
+In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, outperforming queues built by the LinkedList approach.
+Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
 We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
@@
 All processes agree on the sequence stored in the root and this is used as the linearization ordering.
-\begin{figure}
+
+\begin{figure}[t]
 \begin{center}
@@
 \caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root, which stores the final ordering.}
+Each node tries to propagate its operations up to the root, which stores the final ordering of all operations.}
 \end{figure}
 
-The goal here is to ensure in each propagate step; the new operations are propagated up to the parent in $\log p$ steps (See~Figure~\ref{fig::propagstep}).
-Then a dequeue operation uses the linearization ordering to compute its answer.
+The goal here is to ensure in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
+Then, a dequeue operation uses the linearization ordering to compute its answer.
 
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using \texttt{CAS} operations.
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a\texttt{CAS} operation.
 The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in \texttt{n}'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
@@
 This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.
+To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.
 
-\begin{figure}
+\begin{figure}[t]
 \begin{center}
@@
 \end{tikzpicture}}
-  \caption{\label{fig::propagstep}Operations after $||$ are new.}
+  \caption{Operations after $||$ are new.}
 \end{subfigure}
@@
 \end{subfigure}
-\caption{\label{fig:ucexample} Successful \texttt{Refresh}, operations in children after $||$ are new.}
+\caption{\label{fig::propagstep} Successful \texttt{Refresh}, operations in children after $||$ are new.}
 \end{center}
@@
 \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf.
-\texttt{Index()} is  similar but more complicated, finding superblocks from a leaf to the root.
-The main challenge in each level of the \texttt{Get(i)}, \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
-After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)}, \texttt{Index(op)}.
+\texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root.
+The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
+After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
 Now we can implement Enqueue and Dequeue using our block tree.
-An \texttt{Enqueue(e)} is appending an operation with input argument \texttt{e} in the block tree.
+An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
 To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree.
-Then finds the rank of the \texttt{DEQ} using \texttt{Index()}, after knowing the rank of \texttt{DEQ} it computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
-Finally returns the argument of the answer \texttt{ENQ} using \texttt{Get(i)}.
+Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, using the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
+Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
 
 \texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation.
-We use prefix sum for binary search and a constant amount of information that we do not discuss here.
+Each block stores a constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
 
@@
 We came up with a garbage collection only for the root level to reduce the time to find this block to $O(\log q)$.
-The main idea is to remove a block after its operations are finished.
-We can search in Persistent red-black trees with size $n$ in $O(\log n)$ steps and split the tree by a key in $O(\log n)$ steps.
+The main idea is to remove a block after its operations are no longer needed.
 We will store the blocks in the root in a persistent red-black tree to be able only to maintain the remaining information of the queue.
-So our algorithm works in $O(\log^2 p +\log q)$ steps.
+It is possible to search in persistent red-black trees with size $n$ in $O(\log n)$ steps and split the tree by a key in $O(\log n)$ steps.
+Finally, the algorithm works in $O(\log^2 p +\log q)$ steps.
 
@@
   \item Description of the algorithm, layer by layer
-  \item Pseudocode of the algorithm in two parts; regarding memory management
+  \item Pseudocode of the algorithm
   \item Proof of linearizability and time analysis

### 2022-03-18 proposal/proposal.tex
finalize proposal

@@
 Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
-Also there is a common problem in previous shared queues we're going to resolve which make their worst case slow.
+Also we are going to resolve a common problem in previous shared queues which makes their worst case slow.
 
-The question that may arise is, ``What properties matter for a lock-free data structure?''.
-Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa since a lock-free algorithm may terminate after an exponential number of steps.
+A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every often but never terminates.
 %From now on, we are going to focus on wait freedom.
@@
 \section{Related Work}
-\subsection{List based queues}
+\subsection{List-based Queues}
 In the following paragraphs, we look at previous lock-free queues.
@@
   \center\includegraphics[scale=0.5]{pics/msqueue}
-  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrow shows new pointer and gray dashed arrow shows the old arrow.}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrows show new pointers and gray dashed arrows show the old pointers.}
 \end{figure}
@@
   \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption{Baskets queue idea.}
+  \caption{Baskets queue idea. There is a mechanism in the algorithm that orders the operations in a basket by the order of which \texttt{CAS}es first successfully.}
 \end{figure}
@@
   \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption{\label{fig::sundell}Global pointer point to arrays. Head and Tail elements are blue dequeued elements are red and current elements of the queue are in green.}
+    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
 \end{figure}
@@
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-It is not limited to list-based queues and array-based queues share CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} .
+It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} .
 We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.
@@
 Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
-A \textit{universal construction} is an object that can implement a shared object from any given sequential object.
+A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
 We can implement a concurrent queue using a universal construction.
@@
 
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of log $p$ using $O(\log n$)bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
 Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.
@@
 \section{Our Contribution}
-In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, outperforming queues built by the LinkedList approach.
-Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
+In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, avoiding the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays.
 We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
@@
 All processes agree on the sequence stored in the root and this is used as the linearization ordering.
+Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
 
@@
 To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree.
-Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, using the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
+Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
 Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

### 2022-03-23 proposal/proposal.tex
apply franck feedback on proposal

@@
 \title{Thesis Proposal}
-\author{Hossein Naderi \\supervised by Eric Ruppert}
+\author{Hossein Naderibeni \\supervised by Eric Ruppert}
 \maketitle
@@
 Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
-Also we are going to resolve a common problem in previous shared queues which makes their worst case slow.
+%We are going to resolve a common problem in previous shared queues which makes their worst case slow.
 
@@
 The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} is an example of execution on a linearizable  queue that is initially empty.
+Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty.
 The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
@@
   \caption{\label{fig::badexample}An example of an execution that is not linearizable.
-Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A}.}
+Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
 \end{figure}
@@
 We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every often but never terminates.
+A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every so often but never terminates.
 %From now on, we are going to focus on wait freedom.
@@
 Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
 
@@
   \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption{Baskets queue idea. There is a mechanism in the algorithm that orders the operations in a basket by the order of which \texttt{CAS}es first successfully.}
+  \caption{Baskets queue idea.
+There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}.
+To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again.
+The successful process will be the next one in the basket and so on.}
 \end{figure}
@@
 
-Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-Global head and tail pointers point to arrays containing the first and last element in the queue and the elements when they are written by \texttt{CAS} operations and dequeued.
+Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
+Global head and tail pointers point to arrays containing the first and last elements in the queue.
+Global pointers are up to date, but head and tail pointers may be behind in time.
+An enqueue or a dequeue searches in the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
 Their data structure is lock-free.
-An enqueue or a dequeue searches for the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
 Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.
@@
 
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on MS-queue and use Herlihy's helping technique to achieve wait-freedom.
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique to achieve wait-freedom.
 Their step complexity is $\Omega(p)$ because of the helping mechanism.
@@
 
-The goal here is to ensure in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
+The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
 Then, a dequeue operation uses the linearization ordering to compute its answer.
 
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a\texttt{CAS} operation.
-The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in \texttt{n}'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation.
+The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
 %Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
@@
 This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.
+To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.
 
@@
 An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
-To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree.
+To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
 Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.

### 2022-03-23 article/psuedocode.tex
minimalize datastructure structure

--- 2022-02-03 article/article.tex
@@
 
-\title{Wait-free Queues with Polygarithmic Step Complexity}
-\date{\today}
-
 \usepackage{amsmath,amssymb,amsthm}
@@
 \begin{document}
-\maketitle
-
-%\begin{abstract}
-%This is the paper's abstract \ldots
-%\end{abstract}
-
-\section{Previous Work}
-\begin{table}[hbt]
-  \begin{center}
-  \begin{tabular}{l|c|c|c}
-    & Type & Progress Property  & Complexity \\ \hline
-    MichaelS96~\cite{DBLP:conf/podc/MichaelS96} & Singly Linked List & Lock-free & $\Omega(p)$ \\ \hline
-    MoirNSS05~\cite{DBLP:conf/spaa/MoirNSS05} & Linked List with Elemination & & \\ \hline
-    HoffmanSS07~\cite{DBLP:conf/opodis/HoffmanSS07} & Linked List with Baskets & & $\Omega(p)$ \\ \hline
-    Ladan-MozesS08 ~\cite{DBLP:journals/dc/Ladan-MozesS08} & Doubly linked list & Lock-free & $\Omega(p)$ \\ \hline
-    MilmanKLLP18~\cite{DBLP:conf/spaa/MilmanKLLP18} & Linked List with batching same type ops & Lock-free & \\ \hline
-    GidenstamST10~\cite{DBLP:conf/opodis/GidenstamST10} & Linked list of arrays & Lock-free & $\Omega(p+len(array))$ \\ \hline
-    KoganP11~\cite{DBLP:conf/ppopp/KoganP11} &  Linkedlist with head and tail pointers & Wait-free & $\textsc{O}(p\times\text{bakery-max})$
-  \end{tabular}
-  \caption{Linearizable ME-MD Queus}
-  \end{center}
-\end{table}
-
-\section{ Universal Construction using Tournament Tree with Big CAS Objects }
-\paragraph{}
-Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
-He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
-Here, we first introduce a universal construction using $\textsc{O}(\log p)$ CAS operations~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use the universal construction as a stepping stone towards our queue algorithm, so we will not explain it in too much detail.
-\paragraph{}
-We introduce our universal construction in Algorithm~\ref{alg1} with a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes.
-Nodes are CAS objects, and each leaf is assigned to one process exclusively.
-Each CAS object stores a sequence of operations.
-When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root.
-The sequence of operations stored in Node $n$'s CAS object shows the order of the operations propagated up to $n$.
-The history of operations stored at the root is the linearization ordering. The operation $op$ is linearized when it is appended to the root.
-
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-
-Leaf $l_i$ stores the sequence of the operations invoked by $P_i$. 
-The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$.
-In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
-After a process adding a new operations to its leaf, it has to propagate new operations up to the root.
-\textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$).
-In each step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again.
-If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
-Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) finishes.
-
-\begin{figure}
-
-\begin{center}
-
-\begin{subfigure}[b]{.49\textwidth}
-
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}
-}
-  \caption{Operations after $||$ are new.}
-
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.49\textwidth}
-
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}
-}
-  \caption{New operations are added to the parent node.}
-
-\end{subfigure}
-
-\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
-\end{center}
-\end{figure}
-
-\begin{algorithm}
-\caption{Universal Construction Idea}\label{alg1}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
-\State \texttt{l\textsubscript{i}.\Call{append}{op}}
-\State \Call{Propagate}{parent of l\textsubscript{i}}
-\State \texttt{Run} \textsf{the sequence stored in root}
-\State \texttt{\Return op}\textsf{'s response from line 4}
-\EndFunction{Do}
-
-\Statex
-
-\Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{n==root}} \Return
-\ElsIf{!\Call{Refresh}{n}}
-\State \Call{Refresh}{$n$} \EndIf
-\State \Call{Propagate}{parent of n}
-\EndFunction{Propagate}
-
-\columnbreak
-
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{old=} \Call{Read}{n}
-\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
-\State \texttt{new= old$\cdot$new}
-\State \Return n.\Call{CAS}{old, new}
-\EndFunction{Refresh}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
-The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}).
-However, to compute the result of an operation, we only use the total ordering that is stored at the root.
-Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable size CAS objects and still achieve polygarithmic step complexity.
-
-\begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]
-
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
-      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]
-
-\end{tikzpicture}
-\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
-\end{center}
-\end{figure}
-
-\section{Block Tree}
-
-%\subsection{Idea in a nutshell}
-We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes.
-First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node.
-Second, each operation is linearized when its element is appended to the root.
-
-\subsection{Sequence of propagated operations}
-\paragraph{}
-The basic idea behind the universal construction is to create a linearization of operations invoked by processes.
-If we design a fast shared data structure in which processes can append elements to a sequence, we can use it to implement practical, fast implementaion of an object $O$ by using the sequence data structure to keep track of the sequence of operations  on $O$.
-In the following sections, we introduce our solution called Block~Tree.
-
-%\begin{lemma}
-%    Consider an instance of \textsc{Propagate(}\textnormal{n}\textsc{)}.
-When it terminates all operations that were in $n$ before \textsc{Propagate(}\textnormal{n}\textsc{)} will be in the root.
-\texttt{(not suitable place)}
-%\end{lemma}
-
-\subsection{Sequence of Sets of Concurrent Operations \label{subsec::block}}
-\paragraph{}
- In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root.
-However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
-
- \begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}
-
-\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-
-\end{tikzpicture}
-\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves.
-Since we linearize operations when they are added to the root, ordering the blocks in the root is important.}
-\end{center}
-\end{figure}
-
-The definition of linearizability allows concurrent operations to be reordered arbitrarily.
-Thus, a group of concurrent operations can be appended to our root sequence as one block without specifying the order among the operations.
-
-%\begin{lemma}
-%Operations in the same block can be linearized in any order.
-%\end{lemma}
-
-\subsection{Using arrays instead of big CAS Objects}
-\paragraph{}
- We used unbounded CAS objects storing sequences as big words in the universal construction.
-One can represent sequences as arrays to overcome this implementation problem.
-Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.
-
-\subsection{Augmenting Tree to make \textsc{Refresh}() Step faster}
-\paragraph{}
-Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied.
-This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations.
-Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set.
-Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)
-
-\begin{figure}
-
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
-	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]
-
-\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:block} Showing concurrent operation sets with blocks.
-Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
-Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
-Blocks between two lines $||$ are propagated together to the parent.
-For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent.
-The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.}
-\end{figure}
-
-Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
-For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
-Block (5,3) also points to elements from blocks (0,5) and (1,2).
-%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.
-
-\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
-\end{definition}
-
-\begin{definition}\{Subblock\}
-  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$.
-Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
-\end{definition}
-%{\TODO: describe how blocks work with a good example how exactly it is done}
-
-\paragraph{}
-We choose to linearize operations in a block from the left child before those from the right child as a convention.
-Operations within a block of the root can be ordered in any way that is convenient.
-In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
-So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
-All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
-%\begin{lemma}
-%\label{lem:block_size}
-%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
-%\end{lemma}
-%\begin{proof}
-%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
-%\end{proof}
-
-\paragraph{}
-In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes.
-So in a complete propagation, at most $2p$ blocks are merged into one block.
-\texttt{(maybe useful for analysis)}
-%\begin{lemma}
-%A block in the root cannot contain more than one operation from a process.
-%\end{lemma}
-%\begin{proof}
-%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
-%\end{proof}
-
-\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
-%\begin{definition}
-%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
-%\end{definition}
-\paragraph{}
-\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence.
-We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
-To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.
-
-Furthermore, in each block, we store the prefix sum of left and right elements.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]
-
-\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:prefix} Using Prefix sums in blocks.
-When we want to find block b elements in its children, we can use binary search.
-The number below each block shows the count of elements in the previous blocks.}
-\end{figure}
-
-\begin{figure}[hbt]
-\centering
-  \includegraphics[width=5in]{pics/pointers}
-  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
-\end{figure}
-
-\paragraph{}
-Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
-From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
-So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
-
-\subsection{Block Tree Algorithm}
-\paragraph{}
-Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements.
-It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.
-
-\paragraph{Design of a block tree}
-
-Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
-Each internal node of the tree contains an array of blocks of elements.
-Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks.
-Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order.
-Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively.
-Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
-Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
-The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
-After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).
-
-In order to compute elements of a block faster we store prefix-sum blocks(block i has tuple(right-sum=$\#$right ops in previous block, left-sum=$\#$left ops in previous blocks)[See Figure \ref{fig:prefix}].
-Here is the algorithm to get elements of a block.
-
-\paragraph{Specification}
-A block tree is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Append(e)} and \texttt{Get(i)}.
-\texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence.
-\texttt{Get(i)} returns \texttt{i}th element stored in the sequence.
-
-\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.
-
-\paragraph{Membership}
-Element \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
-  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
-\end{itemize}
-
-\paragraph{Order of elements inside node}
-Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
-\begin{itemize}
-  \item The block containing d is before the block containing e.
-  \item e and d are in the same block and d is in the left child and e is in the right child.
-  \item d is before e in the same child's order.
-\end{itemize}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
-\end{figure}
-
-\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
-\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
-Other attributes are assigned values followed by definition.
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
-\end{figure}
-
-%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}
-
-\paragraph{Double Refresh}
-Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
-\begin{proof}
-\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block.
-If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
-So suppose two \texttt{Refresh}es fail.
-Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}.
-So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}.
-After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-\end{proof}
-\begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
-\end{figure}
-
-\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
-\begin{proof}
-Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks.
-So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end.
-Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
-\end{proof}
-
-\paragraph{Total Order}
-Sequence represented by the Block Tree is the sequence of the blocks stored in the root.
-
-\paragraph{Linearization Points}
-\texttt{Get(i)} is linearized when it terminates.
-\texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.
-
-\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
-\begin{proof}
-If there are more than $p$ subblocks, then there is more than one block from process pl.
-\texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9).
-So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
-\end{proof}
-
-\paragraph{Computing \texttt{Get(n, b, i)}}
-To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$.
-Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.
-
-\paragraph{How \texttt{Refresh(n)} works.}
-\begin{enumerate}
-  \item Read n's counter and head
-  \item Create block b
-  \item CAS b into n
-  \item If previous succeed:
-  \begin{enumerate}
-  \item Update sup of b's ending subblocks
-  \item Increment children's counters
-  \end{enumerate}
-\end{enumerate}
-
-%\paragraph{Time-Space Relation} There are at most $p^2$ blocks with same \texttt{time} attribute.
-%\begin{proof}
-%If a process dies at line 28, the children counter won't increase, and each process aggregates most block, and since there are at most p concurrent processes, then there will be at most $p^2$ blocks with same \texttt{time} attribute.
-%\end{proof}
-%
-%\paragraph{Time-Propagate() Relation} In a \texttt{CreateBlock()} procedure, the blocks that are seen have at most 2 different \texttt{time} attribute values.
-%
-%\paragraph{Time-Super Relation} sup[i+1]-sup[i] is $\leq p$
-%
-%\paragraph{\texttt{b}'s Superblock position} SuperBlock of \texttt{b} is within range $\pm p$ of the \texttt{\texttt{sup[b.time]}}.
-%\begin{proof}
-%
-%\end{proof}
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/super}
-  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
-\end{figure}
-
-\paragraph{Computing superblock}
-\begin{enumerate}
-  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
-    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10.
-For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
-  \end{proof} 
-  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
-   Follows from the order of lines 37, 26, 27.
- \end{proof}
-%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-  \item \texttt{super[i+1]-super[i]}$\leq p$
-\begin{proof}
- In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27.
-Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
-If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27.
-So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
- \end{proof}
-  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
-\begin{proof}
-\texttt{super[i]} is the index of the superblock of a block containing block b.
-It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
-\texttt{super(b)} is the real superblock of b.
-\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
-If \texttt{b.time} is \texttt{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-  \end{proof} 
-\end{enumerate}
-
-%The block tree contains a tournament tree in which leaves are assigned to processes, and each process appends elements to its leaf and tries to propagate it up to the root.
-The final sequence is stored at the root.
-\texttt{Append} after appending to leaf calls \texttt{Propagate} which iterates over the path to root.
-In each \texttt{Propagate(n)} invocation on node n it tries ti \texttt{Refresh} the sequence stored at node n with the new elements in n's children.
-After adding the element to the root \texttt{Index(op)} finds the order of the element in the root.
-
-%\textsf{I finally thought maybe a sequence of elements is better for telling specification.}
-
-\begin{algorithm}
-\caption{Block Tree \label{alg2}}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-\Statex \textbf{Structure}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{element} e}
-\begin{itemize}
-%\item \texttt{\textsl{string} name}
-\item \texttt{\textsl{int} pid}
-\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}.ops}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
-\begin{itemize}
-\item \texttt{\textsl{*node} left, right, parent}
-\item \texttt{\textsl{block[]} blocks}\textsf{: blocks stored in \texttt{n}}
-\item \texttt{\textsl{int} head= 1}\textsf{: index of first empty cell of \texttt{blocks}}
-\item \texttt{\textsl{int} counter= 0}\textsf{}
-\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is \texttt{i}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{leaf node} l\textsubscript{i} extends \textsl{node}}
-\begin{itemize}
-\item \texttt{\textsl{operation[]} ops}\textsf{: elements that are invoked to \texttt{Append()} to the blok tree by process i}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
-\begin{itemize}
-  \item \texttt{\textsl{int} num\textsubscript{left}, sum\textsubscript{left}}
-
-  \textsf{~~\#operations from the left subblocks of b, prefix sum of num\textsubscript{left}}
-
-  \item \texttt{\textsl{int} num\textsubscript{right}, sum\textsubscript{right}}
-
-  \textsf{~\#operations from the right subblocks of b, prefix sum of num\textsubscript{right}}
-
-  \item \texttt{\textsl{int} sum}
-
-  \textsf{\# operation in b}
-
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
-
-  \textsf{~~index of \texttt{b}'s last subblock}
-  \item \texttt{\textsl{int} time}
-
-\end{itemize}
-
-%\Statex \textbf{Examples}
-%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
-%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
-%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
-
-\Statex
-
-\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
-\State \texttt{op.loc= l\textsubscript{pid}.head}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.time= op.loc}
-\State \texttt{b.sum= 1}
-\State \texttt{l\textsubscript{pid}.ops[op.loc]= op}
-\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
-\State \texttt{op.head+= 1}
-\State \Call{Propagate}{l\textsubscript{pid}.parent}
-\State \Return \Call{Index}{l\textsubscript{pid}, op.loc, b} 
-\EndFunction{Append}
-
-\columnbreak
-
-\Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{not} \Call{Refresh}{n}}
-\State \Call{Refresh}{n}
-\EndIf
-\If{\texttt{n {\keywordfont is} root}} \Return
-\EndIf
-\State \Call{Propagate}{n.parent}
-\EndFunction{Propagate}
-
-\Statex
-
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{i= n.head}
-\State \texttt{c= n.counter}
-\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, c}}
-\If{\Call{CAS}{n.blocks[i+1], null, new}}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
-\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
-\EndFor
-\State \Return{\texttt{true}}
-\Else 
-\State \Return{ \texttt{false}}
-\EndIf
-\State \texttt{\Call{CAS}{n.head, i, i+1}}
-\EndFunction{Index}
-
-\Statex
-
-\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int c} 
-\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.time= c}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{j= n.dir.head}
-\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
-\State \texttt{end= n.dir.blocks[j]}
-\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].time}
-\State \texttt{b.end\textsubscript{dir}= j}
-\State \texttt{b.num\textsubscript{dir}= end.sum - start.sum}
-\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
-\EndFor
-\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
-\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
-\EndFunction{CreateBlock}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Block Tree Continued}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{46}
-
-\Function{element}{Get}{\textsl{int} i} 
-\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.head}}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
-\EndFunction{Get}
-
-\Statex
-
-\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
-\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}
-
-\Statex \Comment{\textmd{Searches~the value \texttt{i} of the given prefix sum \texttt{type} on the node \texttt{n} in the domain from \texttt{start} to \texttt{end} blocks.
-\texttt{f} is one of \texttt{sum, sum\textsubscript{left}, sum\textsubscript{right}}.
-Does binary search on \textsl{field} \texttt{f} of the blocks stored in \texttt{n.blocks} and returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
-\State \Return \texttt{result block's index}
-\EndFunction{BSearch}
-
-\Statex
-
-\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
-\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}
-\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[i]}
-\Else
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
-\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
-\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
-\Else
-\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
-\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
-\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} 
-\EndIf
-\EndIf
-\EndFunction{Get}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition:~\texttt{i}th operation in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
-\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns rank in the  
-root of the \texttt{i}th operation in node \texttt{n}.}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
-\Else
-\State \texttt{dir= (n.parent.left==n)? left: right}
-\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].time]-p, sup[n.blocks[b].time]+p}}
-\If{\texttt{dir {\keywordfont is} left}}
-\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
-\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
-\EndIf
-\State \Return\Call{Index}{n.parent, i, superBlock}
-\EndIf
-\EndFunction{Index}
-
-\Statex
-
-%\Function{list}{GetElements}{block b} 
-%\ForEach{block {\keywordfont in} \Call{GetSubBlocks}{b}}
-%\State result.append(\Call{GetElements}{})
-%\EndFor
-%\State \Return result
-%\EndFunction{GetElements}
-%
-%\Statex
-%
-%\Function{list}{GetSubBlocks}{b}
-%\State n:=b.node
-%\State b[-1]=b's previous block
-%\ForEach{direction \{left, right\}}
-%\State init=\Call{BSearch}{n.direction, b[-1].direction}
-%\State end=\Call{BSearch}{n.direction, b.direction}
-%\State result.append([init:end])
-%\EndFor
-%\State \Return result
-%\EndFunction{GetSubBlocks}
-
-\end{algorithmic}
-\end{algorithm}
-
-\section{Implementing Queue using Block Tree}
-\subsection{Idea in a nutshell}
-With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log ^2p+\log n)$ steps.
-A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
-If the queue is empty it returns \texttt{null}.
-Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.
-So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
-%
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
-\paragraph{}
-Next, we describe how to use block tree to implement queues.
-The block tree, maintains the history of all operations, not only the current state of the queue.
-Now consider the following history of operations.
-What should each \texttt{Dequeue()} return?
-
-\begin{table}[hbt]
-\centering
-  \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
-    \hline \texttt{ENQ(5)}& \texttt{ENQ(2)}& \texttt{DEQ()}& \texttt{ENQ(3)}&\texttt{DEQ()}& \texttt{DEQ()}& \texttt{DEQ()}& \texttt{ENQ(4)}& \texttt{ENQ(6)}& \texttt{DEQ()}\\ \hline
-  \end{tabular}
-  \caption{An example histoy of operations on the queue}
-\end{table}
-
-\begin{definition}
-A non-null dequeue is one that returns a non-null value.
-\end{definition}
-
-\paragraph{}
-In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order.
-Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null.
-If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}.
-So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.
-
-%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
-
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}
-
-\paragraph{}
-In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism.
-So now the problem is to find the result of each Dequeue.
-From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
-In the next example, operations in a cell are concurrent.
-\texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
-We will next describe how these values can be computed efficiently.
-
-\begin{table}[hbt]
-\centering
-  \begin{tabular}{c|c|c|c}
-    \hline \texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
-  \end{tabular}
-  \caption{An example history of operation blocks on the queue}
-\end{table}
-
-\paragraph{}
-Now, we claimed that by knowing the current size of the queue and the number of non-null dequeue operations before the current dequeue, we could compute the index of the resulting \texttt{Enqueue()}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations happens and the number of non-null dequeues dequeues till a block, we can compute each dequeue's index of result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbt]
-\centering
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
-    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
-        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
-            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
-                size & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-  \caption{Augmented history of operation blocks on the queue}
-\end{table}
-
-%\begin{definition}
-%  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
-%  
-%  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
-%  
-%\end{definition}
-
-Size and the number of non-null dequeues for $b$th block could be computed this way:\\
-\texttt{size[b]= max(size[b-1] +enqueues[b] -dequeues[b], 0)}\\
-\texttt{non-null dequeues[b]= non-null dequeues[b-1] +dequeues[b] -size[b-1] -enqueues[b]}
-
-Given \texttt{DEQ} is in block \texttt{b}, \texttt{response(DEQ)} would be:\\
-\texttt{(size[b-1]- index of DEQ in the block's dequeus >=0) ? ENQ[non-null dequeus[b-1]+ index of DEQ in the block's dequeus] : null;}
-
-\section{Main Algorithm}
-
-\paragraph{Specification}
-A Queue is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
-\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
-\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
-
-%Now that we know how to augment Block Tree root history to compute the response, we will present the complete algorithm.
-\texttt{Enqueue()} and \texttt{Dequeue()} are the main functions.
-\texttt{Enqueue(e)} adds \texttt{e} to the queue and \texttt{Dequeue()} returns the first element in the queue and removes it from the queue.
-Block Tree functionalities are very similar.
-\texttt{Append(e)} does what \texttt{Enqueue()} does and if we know the current index of the head of the queue we can call \texttt{Get()} to find its value.
-
-\begin{figure}[hbt]
-\centering
-  \includegraphics[width=5in]{pics/queue}
-  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
-\end{figure}
 
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{Operation}}
+\Statex $\diamondsuit$ \texttt{\textsl{Local}}
 \begin{itemize}
-%\item \texttt{\textsl{string} name}
-%\item \texttt{\textsl{string} type$\in$\{DEQ,ENQ\}}
-\item \texttt{\textsl{Object} element}\textsf{: argument of an enqueue, if null this is a dequeue}
-%\item \texttt{\textsl{int} pid}
-\item \texttt{\textsl{int} loc\textsf{: location in the leaf's \texttt{ops}}}
+\item \texttt{\textsl{*Node} leaf\textsf{: pointer the the process's leaf in the tree}}
 \end{itemize}
+
+\Statex
+
+\Statex $\diamondsuit$ \texttt{\textsl{Shared}}
+\begin{itemize}
+\item \texttt{\textsl{Tree}} to complete
+\end{itemize}
+
+\Statex
+\Statex $\diamondsuit$ \texttt{\textsl{Structures}}
 
@@
 \item \texttt{\textsl{*Node} left, right, parent}
-\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0}
+\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \texttt{en} pointers to the first block of the coresponding children.
+\texttt{blocks[i]} returns the $i$th block stored.}
 \item \texttt{\textsl{int} head= 1}\textsf{: index of the first empty cell of \texttt{blocks}}
@@
 \item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is field \texttt{i}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Leaf} extends \textsl{Node}}
-\Statex \Comment{\textsf{\texttt{l\textsubscript{i}} is the Leaf for process \texttt{i} and \texttt{i} is in range \texttt{0} to \texttt{p-1}}}
-\begin{itemize}
-\item \texttt{\textsl{Operation[]} ops}\textsf{: invoked operations}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Root} extends \textsl{Node}}
-\begin{itemize}
-\item \texttt{override \textsl{Root Block[]} blocks}
 \end{itemize}
@@
   \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}
+  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
+  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
+  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
+  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
+  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
+  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
+  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
+  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
+  \textsf{: \# enqueue, dequeue operations in the block}
+  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
+  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
+  \item \texttt{\textsl{int} num, sum}
+  \textsf{: total \# operations in block, prefix sum of \texttt{num}}
+  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
+  \textsf{:~~index of the last subblock in the left and right child}
+  \item \texttt{\textsl{int} group}
+  \textsf{: id of the group of blocks including this propagated together, more precisely the value of the node's counter when propagating this block.}
+\end{itemize}
 
-  \textsf{~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
+\pagebreak
 
-  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
-
-  \textsf{~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
-
-  \textsf{~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
-
-  \textsf{~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
-
-  \textsf{\# enqueue, dequeue operations in the block}
-
-  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
-
-  \textsf{sum of \# enqueue, dequeue operations in blocks up to this one}
-
-  \item \texttt{\textsl{int} num, sum}
-
-  \textsf{total \# operations in block, prefix sum of \texttt{num}}
-
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
-
-  \textsf{~~index of this block's last subblock in the left and right child}
-  \item \texttt{\textsl{int} group}
-
+\Statex $\blacktriangleright$ \texttt{\textsl{Leaf Block} extends \textsl{Block}}
+\begin{itemize}
+  \item \texttt{\textsl{Object} element}
+  \textsf{Each block in a leaf also represents an operation. The \texttt{element} shows the operations argument if it is an enqueue, and if it is a dequeue the value is \texttt{null}.}
 \end{itemize}
@@
   \item \texttt{\textsl{int} size}
-
   \textsf{size of queue after this block's operations finish}
-
   \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
-
   \textsf{count of non-null dequeus up to this block}
-
 \end{itemize}
-
-%\Statex \textbf{Examples}
-%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
-%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
-%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
 
@@
 \Function{void}{Enqueue}{\textsl{Object} e}
-\State \texttt{op= NEW(\textsl{operation})}
-%\State \texttt{op.type= ENQ}
-\State \texttt{op.element= e}
-%\State \texttt{op.pid= this.pid}
-\State \Call{Append}{op}
+\State \texttt{block b= \Call{new}{\textsl{block}}}
+\State \texttt{b.element= e}
+\State \texttt{b.sum\textsubscript{enq}=1}
+\State \Call{Append}{b}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{}
-\State \texttt{op= NEW(\textsl{operation})}
-%\State \texttt{op.type= DEQ}
-%\State \texttt{op.pid= this.pid}
-\State  \Call{Append}{op} \Comment{$i$th DEQ in $b$th block}
+\State \texttt{block b= \Call{new}{\textsl{block}}}
+\State \texttt{b.element= null}
+\State \texttt{b.sum\textsubscript{deq}=1}
+\State \Call{Append}{b}
 \State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, op.loc, 1}
 \State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned}
-\If{\texttt{res }\textbf{is }\texttt{null}}
-\State \Return \texttt{null}
-\Else
 \State \Return \Call{Get}{res}
-\EndIf
 \EndFunction{Dequeue}
@@
 
-\Function{void}{Append}{\textsl{operation} op} 
-\State \texttt{pid= id of process performing \textsc{Append}}
-\State \texttt{op.loc= l\textsubscript{pid}.head}
-\State \texttt{op.head+= 1}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.group= op.loc}
-\If{\texttt{op.element==null}} 
-\texttt{b.sum\textsubscript{deq}=1} \Else \texttt{~b.sum\textsubscript{enq}=1} \EndIf
-%\State \texttt{b.sum\textsubscript{op.type}= 1}
-\State \texttt{l\textsubscript{pid}.ops[op.loc]= op} \label{addOP}
-\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
-\State \Call{Propagate}{l\textsubscript{pid}.parent} 
+\Function{void}{Append}{\textsl{block} b} 
+\State \texttt{b.group= this.leaf.head}
+\State \texttt{l\textsubscript{pid}.blocks[this.leaf.head]= b}
+\State \texttt{this.leaf.head+=1}
+\State \Call{Propagate}{this.leaf.parent} 
 \EndFunction{Append}
@@
 
+%##########################################
+
 \begin{algorithm}
-\caption{Queue Continued}
+%\caption{Queue Continued}
 \begin{algorithmic}[1]
@@
 \EndIf
-\If{\texttt{n.parent \textbf{is} null}} 
+\If{\texttt{n.parent} \textbf{is not} \texttt{null}} 
 \State \Call{Propagate}{n.parent}
@@
 %\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
-\State \texttt{t= n.counter}
-\State \texttt{<new, t\textsubscript{left}, t\textsubscript{right}>= \Call{CreateBlock}{n, h, t}}
+\State \texttt{c= n.counter}
+\State \texttt{<new, c\textsubscript{left}, c\textsubscript{right}>= \Call{CreateBlock}{n, h}}
+\State \texttt{new.group= c}
 \If{\texttt{new.num==0}} \Return{\texttt{true}}
@@
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{\Call{CAS}{n.dir.super[t\textsubscript{dir}], null, h+1}}
-\State \texttt{\Call{CAS}{n.dir.counter, t\textsubscript{dir}, c\textsubscript{dir}+1}}
+\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, h+1}}
+\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
 \EndFor
@@
 \Comment{Returns $i$th Enqueue.}
+\If{\texttt{i }\textbf{is }\texttt{null}}
+\State \Return \texttt{null}
+\EndIf
 \State \texttt{res= \Call{BSearch}{root, sum\textsubscript{enq}, i, 0, root.head}}
@@
 \Statex \Comment{\textmd{Does binary search for~the value \texttt{i} of the given prefix sum \texttt{feild}.
-\texttt{f} is one of \texttt{sum, sum\textsubscript{left}, sum\textsubscript{right}}.
 Returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
@@
 
-\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i, int t} 
-\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]} with time field=\texttt{t}. Returns the created block as well as values read from each child counter feild.}
+\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
+\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
 \State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.group= t}
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
@@
 \State \texttt{prevBlock= n.dir.blocks[prevIndex]}
-\State \texttt{t\textsubscript{dir}= n.dir.counter}
+\State \texttt{c\textsubscript{dir}= n.dir.counter}
 \State \texttt{b.end\textsubscript{dir}= lastIndex}
@@
 
-\State \Return \texttt{b, t\textsubscript{left}, t\textsubscript{rightt}}
+\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{right}}
 \EndFunction{CreateBlock}
@@
 
-\subsection{Pseudocode description}
+%##########################################
 
-\paragraph{Tournament Tree}
-In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree.
-Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}.
-Each process adds $op$ to its leaf.
-In each internal node an ordering of operations in its subtree is stored.
-All processes agree on the total ordering of all operations stored in the root.
-This ordering will be the linearization of the operations.
-\paragraph{Implicit Storing Blocks}
-For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks.
-A Block is a constant size object that implicitly represents a sequence of operations.
-In each node there is an array of Blocks.
-
-\begin{definition}[Block]
-A block is an object that stores some statistics described in Algorithm Queue.   
-\end{definition}
-
-\begin{definition}[\label{def::subblock}Subblock]
-Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-\end{definition}
-Block $b$ contains subblocks in the left and right children.
-WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$.
-See Figure \ref{fig::createBlock} .
-
-\begin{definition}[Membership of an operation in a block]
-Operation \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
- \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
- \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-\end{itemize}
-
-\end{definition}
-
-We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s.
-In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}.
-Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
-There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step.
-It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases.
-Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}.
-This helps us to know which blocks have aggregated together to a block, not precisely though.
-We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.
-
-In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}.
-In order to make \texttt{BSearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
-
-To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}.
-\texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.
-
-\texttt{Enqueue(e)} just \texttt{append}s an \texttt{operation} with \texttt{element e} to the \texttt{root}.
-\texttt{Dequeue()} appends an \texttt{operation} to the root and computes its ordering and the \texttt{enqueue operation} containing the head before it calling \texttt{ComputeHead()} and then \texttt{gets} and returns the \texttt{operation}'s element.
-
-\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index.
-As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times.
-\texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.
-
-We only want to know the \texttt{element} of \texttt{enqueue} operations and compute ordering for \texttt{dequeue} operations.
-That's the reason here \texttt{Get()} searches between enqueues only and \texttt{Index()} returns ordering of a dequeue among dequeues.
-\texttt{Get(n, b ,i)} decides the requested element is in which child of n and continues to search recursively.
-\texttt{index(n, i, b)} calculates the ordering of the given operation in \texttt{n}'s parent each step and finally returns the result among total ordering.
-
-\subsection{Complexity Analysis}
-
-\texttt{Enqueue()} operations do a constant number of steps and an \texttt{Append()} which calls \texttt{Propagate()}.
-Let $p$ be the number of processes and $m$ be the count of operations invoked.
-In a \texttt{Propagate()} step there are a constant number of steps taken at each level and the height of the tree is $\log(p)$.
-So \texttt{Enqueue()} takes $O(\log p)$ steps.
-
-\texttt{Get()} searches among all blocks at first and then iterates over a path from the root to a leafof the tree and in each step searches in a domain of size $p$ , so it takes $\log^2 p+\log m$.
-\texttt{Index()} is a path from a leaf to the root which each step calls a $\log p$ step search so it takes $\log^2p$.
-\texttt{Dequeue()} calls \texttt{Append()}, it also calls \texttt{Index()} and \texttt{Get()} that take $\log^2 p$, $\log^2p +\log n$.
-
-\subsection{Linearizability Proof}
-
-\begin{definition}
-If \texttt{n.blocks[i]==b}  we call \texttt{i} the \emph{index} of block \texttt{b} in node \texttt{n}.
-Block \texttt{b} is before block \texttt{b$^\prime$} in node \texttt{n} if and only if \texttt{b}'s index is smaller than \texttt{b$^\prime$}'s.
-Block \texttt{b} is propagated to node \texttt{n} or set \texttt{S} if \texttt{b} is in \texttt{n.blocks} or \texttt{S} or is a subblock of a block in \texttt{n.blocks} or \texttt{S}.
-\end{definition}
-
-%\begin{definition}
-%  Block \texttt{b} in node \texttt{n} is new if \texttt{b} is not a subblock of any block in \texttt{n.parent.blocks[]}.
-%\end{definition}
-
-\begin{definition}
-  Block \texttt{b} in node \texttt{n} is in \emph{Established(n, t)} if \texttt{n.head} is greater than \texttt{b}'s index at time $t$.
-\end{definition}
-
-\begin{lemma}[headProgress] \label{lem::headProgress}
-\texttt{n.head} is non-decreasing over time.
-\end{lemma}
-\begin{proof}
-  Simply because \texttt{n.head} is only incremented.
-\end{proof}
-
-\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line 52(\texttt{h=n.head}) might be 1 bit behind the first empty block.
-\end{lemma}
-\begin{proof}
- Because at the end of every \texttt{Refresh()} with block size greater than 0 (Lines 53,56) \texttt{n.head} is incremented.
-Maybe some process goes to sleep before incrementing the head, but after sleeping if \texttt{h} does not increase then \texttt{CAS} in Line 52 is going to be failed and nothing is going to be appended to \texttt{n.blocks}.
-\end{proof}
-
-%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
-%  If block \texttt{b} in node \texttt{n} is established, then all the blocks in \texttt{n.blocks[]} before \texttt{b} are established.
-If block \texttt{b} in node \texttt{n} is not new then all the blocks in \texttt{n.blocks[]} before \texttt{b} are not new.
-%\end{lemma}
-%\begin{proof}
-%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
-%\end{proof}
-
-\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
-  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
-\end{lemma}
-\begin{proof}
-Because blocks are only appended(not modified) with CAS to \texttt{n.blocks[n.head]} and \texttt{n.head} is non-decreasing.
-\end{proof}
-
-\begin{lemma}[createBlock] \label{lem::createBlock}
-  Suppose \texttt{CreateBlock(n, h, x)} is invoked at time $t$.
-The blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ that are not propagated to $Established(n, t)$, are subblock of the block returned by \texttt{CreateBlock(n, h, x)}.
-\end{lemma}
-\begin{proof}
- We prove the claim for the left child.
-Blocks in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
-Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
-See Figure~\ref{fig::createBlock}.
-The right child is the same.
-\end{proof}
-
-\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Suppose \texttt{Refresh(n)}'s \texttt{CAS(n.blocks[h], null, new)} returns \texttt{true}.
-Let $t$ be the time \texttt{Refresh(n)} is invoked, blocks propgatated to $Established(n.left, t)$ and $Established(n.right, t)$ are propagated to in $Established(n, t)$ after \texttt{CAS(n.blocks[h], null, new)}.
-\end{lemma}
-\begin{proof}
-By Lemma~\ref{lem::createBlock} \texttt{new} contains \texttt{n}'s childrens' established blocks before Line 43 which is  appended to \texttt{n.blocks} by \texttt{CAS} in Line 48.
-\end{proof}
-
-\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-    If instance $r$ of \texttt{Refresh(n)} returns \texttt{false}, then there is another successful instance $r^\prime$ of \texttt{Refresh(n)} that has performed a successful \texttt{CAS(n.blocks[h], null, new)}(Line 49) after Line 43(\texttt{h= n.head}) of $r$.
-\end{lemma}
-\begin{proof}
-  If there is no other concurrent successful \texttt{Refresh(n)} then \texttt{Refresh(n)} would succeed in Line 48.
-So there is another \texttt{Refresh(n)}, that has to \texttt{CAS}ed successfully its block in \texttt{n.blocks[h]} after Line 43 of \texttt{Refresh(n)}.
-Otherwise the other \texttt{Refresh(n)} should have read \texttt{h$^\prime$>h} instead of \texttt{h} for \texttt{n.head}(Line 52).
-\end{proof}
-
-\begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive instances $r_1$,$r_2$ of \texttt{Refresh(n)} by the same process (Lines 35,36).
-Let be the time before $r_1$ invoked.
-After $r_2$'s \texttt{CAS} all the blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ are in $Established(n, t)$.
-\end{lemma}
-\begin{proof}
-
-If Line 35 (first Refresh which we call $R_1$) returns \texttt{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \texttt{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
-$R_1^\prime$ may or may not have propagated some subblocks of \texttt{new} to \texttt{n}.
-It is obvious that the \texttt{new} constructed by the second Refresh in Line 36 contains the blocks in \texttt{new} by $R_1$ which $R_1^\prime$ did not contain, since \texttt{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
-If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
-If not, it is deduced that \texttt{n.blocks[h]} was not \texttt{null} before $R_2$'s \texttt{CAS}.
-Furthermore, \texttt{n.blocks[h]} was \texttt{null} before reading \texttt{h} by $R_1$.
-So there is a successful \texttt{Refresh()} after the read of \texttt{h} in $R_1$ and before the \texttt{CAS} of $R_2$.
-This \texttt{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
-See Figure~\ref{}.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
-  \caption{$R_2^\prime$'s \texttt{CAS} is executed after \texttt{h1=n.head}.}
-\end{figure}
-
- Block \texttt{new} is created of new established subblocks of children of \texttt{n}(Lemma \ref{lem::createBlock}, Line 46).
-If \texttt{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \texttt{n}.
-
-\begin{lemma}[Double Refresh] \label{doublyRefresh}
-All operations in \texttt{n}'s children's blocks before line \texttt{35} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{37}.
-\end{lemma}
-\begin{proof}
-Suppose block \texttt{b} with index \texttt{i} is in the the left child of \texttt{n} before the line 35.
-By Lemma~\ref{head} it follows that \texttt{n.left.head} is greater than \texttt{i}.
-\texttt{Refresh()} calls \texttt{CreateBlock()} and creates a block from blocks between \texttt{n.blocks[n.head].end\textsubscript{left}} and \texttt{n.left.head} in the left child, which contains \texttt{b} as well.
-First it tries to append it in \texttt{n.blocks.head} and if it was succsuful it continues recursivley.
-If not it tries again, and if the second call of \texttt{Refresh()} in Line 36 fails.
-It means there is another \texttt{Refresh} which has reah its \texttt{i} after the Line 35, so it contains \texttt{b} as well.
-\end{proof}
-\texttt{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
-If a \texttt{Refresh()} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
-So suppose two \texttt{Refresh}es fail.
-Since the first \texttt{Refresh()} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first \texttt{Refresh()}, was successful before the second \texttt{Refresh()}.
-So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh()} that assuredly has read block before the mentioned line \texttt{35}.
-After all it means if any of the \texttt{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-
-\begin{lemma}[Append]\label{append}
- When \text{\texttt{Append(op)}} is finished, \texttt{op} appears exactly once in a block of \texttt{root.blocks}.
-\end{lemma}
-\begin{proof}
-\texttt{Append(op)} adds \texttt{op} to \texttt{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \texttt{Propagate()} recursively propagates \texttt{op} up to the root.
-By lemma \ref{doublyRefresh} we know that operation \texttt{op} propagates from child to parent at each level.
-\end{proof}
-
-\begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block in a node contains at most one operation from each processs.  
-\end{lemma}
-\begin{proof}
-  Note that \texttt{prevIndex} and \texttt{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
-After a block has propagated to the parent the blocks between \texttt{prevIndex} and \texttt{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
-The number of new operations which have not propagated yet to the parent cannot be more than $p$.
-If so by the law of pigeonholes there is a process which has appended two concurrent operations.
-\end{proof}
-
-\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
-Each block in a node  has at most $p$ subblocks.
-\end{lemma}
-\begin{proof}
-  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
-\end{proof}
-
-\begin{definition} [Ordering of operations inside a node] \label{ordering}
-$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
-
-\begin{itemize}
-  \item $E(n,i)$ is the sequence of enqueue operations that are member of \texttt{n.blocks[i]} ordered by process id.
-  \item 
-$D(n,i)$ is the sequence of dequeue operations that are member of \texttt{n.blocks[i]} ordered by process id.
-\item $D(n)=D(n,1).D(n,2).D(n,3)...$
-\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
-  \begin{enumerate}
-    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
-The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
-    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
-\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
-It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
-By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
-If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
-  \end{enumerate} 
-\end{proof}
-
-\begin{lemma}[Get] \label{get}
-\texttt{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
-\end{lemma}
-\begin{proof}
-It is obvious that \texttt{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
-To find the \texttt{i}th enqueue in block \texttt{b} of an internal node \texttt{n} in line 87 it is decided that it resides in the left child or the right child.
-This decision is made by Definition of $E(n,b)$.
-After that Lines 88, 92 search the proper subblocks of \texttt{b}.
-From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \texttt{prevBlock} and \texttt{lastBlock} block of the \texttt{CreateBlock()}.
-\end{proof}
-
-\begin{lemma}[Index]
- Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
-\end{lemma}
-\begin{proof}
-  \texttt{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \texttt{n} in \texttt{n.parent} and then computes the order in $D(n.parent, superblock)$.
-Then calls \texttt{index()} on \texttt{n.parent} recursively.
-It is easy to see why the second is correct.
-Correctness of computing superblock comes from Lemma \ref{superBlock}.
-\end{proof}
-
-\begin{lemma}[Computing SuperBlock]\label{superBlock}
-  If \texttt{Index(n,b,i)} performs line 101, then \texttt{superblock} contains \texttt{i}th Dequeue in \texttt{b}th block of node \texttt{n}.
-\end{lemma}
-\begin{proof}
-\begin{enumerate}
- \item Value read for \texttt{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
-  Values \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} are set before incrementing in lines 26,27.
-
- \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\\$\blacktriangleright$
-  Follows from the order of lines 60, 48, 49.
-
-% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
- \item \texttt{super[i+1]-super[i]}$\leq p$
-\\$\blacktriangleright$
- In a Refresh with successful CAS in line 46, \texttt{super} and \texttt{counter} are set for each child in lines 48,49.
-Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
-If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 49.
-So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
-
- \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
-\\$\blacktriangleright$
-\texttt{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
-It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
-\texttt{super(b)} is the real superblock of b.
-\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
-If \texttt{b.time} is \texttt{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-
-\end{enumerate}
-\end{proof}
-
-\begin{lemma}[Computing Queue's Head] \label{computeHead}
-  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q.
-If Q is empty \texttt{ComputeHead(i,b)} returns -1, otherwise
- returns index in $E(root,b)$ of X.
-  \end{lemma}
-
-\begin{lemma}[Validity of \texttt{head}]\label{head}
-No two blocks are written in the same index in \texttt{n.blocks}.
-\end{lemma}
-\begin{proof}
-  \texttt{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \texttt{head} read.
-If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \texttt{head}.
-If a process dies before line 51, another process will increment \texttt{head} in line 54.
-\end{proof}
-
-\begin{lemma}[Validity of \texttt{super} and \texttt{counter}]\label{superCounter}
-If \texttt{super[i] $\neq$ null}, then \texttt{super[i]} in node \texttt{n} is the superblock of a block with \texttt{time=i}.
-\end{lemma}
-\begin{proof}
-After a successful CAS in line 46 \texttt{super} and \texttt{counter} are modified in both children.
-\texttt{super[i]} is supposed to be the superblock of a block with \texttt{time}=i and \texttt{counter} is the timer in each node.
-\texttt{super[i]} and \texttt{counter=i} are expected to update after a bunch of blocks with \texttt{time=i} have been aggregated together into a block in the parent.
-If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \texttt{time}.
-We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
-If a process dies right after line 48, then \texttt{counter} will remain the same and \texttt{super[i]} is correct.
-Furthermore we are sure that when \texttt{super[i]} is read it will not be \texttt{null}.
-\end{proof}
-
-\begin{lemma}[Search Ranges]\label{search}
-  Preconditions of all invocation of \texttt{BSearch} are satisfied.
-\end{lemma}
-\begin{proof}
-
-Line 83: \texttt{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-
-Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-
-Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \texttt{super[i]} is at most $p$, since at most $p$ processes could die.
-
-\end{proof}
-
-\paragraph{Overall}
-
-The main difference between the main algorithm and the block tree is that we separate enqueues and dequeues, compute the number of non-null dequeues and the Queue's size in each block in the root.
-We emphasize that these changes work correctly; every other claim in Block trees applies here.
-
-\section{Make first level search fast}
 \begin{algorithm}
-\caption{Modified Queue \label{algQx}}
+%\caption{Modified Queue \label{algQx}}
 \begin{algorithmic}[1]
 \begin{multicols}{2}
-
-%\Statex \textbf{Structure}
-
-%\Statex $\blacktriangleright$ \texttt{extends Queue Algorithm}
-
-%\Statex $\blacktriangleright$\texttt{\textsl{RBTNode} RBTRoot}  \Statex
-%  \textsf{pointer to an RBTNode, that is root of the RBTree}
 
@@
 
-%  \item \texttt{some RBTRee attributes like left,right children}
-
 \end{itemize}
@@
 \end{itemize}
-
-%\Statex $\blacktriangleright$ \texttt{\textsl{RBT(RBTnode n)}}
-%  \Statex \textsf{Gives us RBTree  persistentoperations if n is root of a RBTRee, for example RBT(n).append(x) appends x if to the redblack tree with n as root.
-A persistent append is an append which only contains the changed nodes.}
 
@@
 \pagebreak
-%\Function{void}{Update(Root Block B)}{}
-%\If{\texttt{B.todo==0}}
-%\If{\texttt{B.left.old-subtree} and \texttt{B.right.old-subtree}}
-%\State \texttt{B.old-subtree= True}
-%\If{\texttt{B.parent!=root}}
-%\State \texttt{Update(B.parent)}
-%\EndIf\EndIf
-%\EndIf
-%\EndFunction{Update}
-%\Statex
 
@@
 \State \texttt{new= RBTnode(b,0,null)}
-%\State \texttt{r= pointer to root of the RBT(h).append(n)}
-%\State \texttt{*r.step= h.step+1}
-%\State \texttt{RBTRoot.CAS(h,r)}
 \State \Return \texttt{rootBlocks.append(new)}
@@
 
-\bibliographystyle{abbrv}
-\bibliography{main}
-
 \end{document}

### 2022-03-24 article/psuedocode.tex
fix garbage collection mistakes

@@
 \begin{itemize}
-\item \texttt{\textsl{Tree}} to complete
+\item \texttt{\textsl{Tree}} \textit{to complete, how?}
 \end{itemize}
@@
 \item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is field \texttt{i}}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \texttt{\textsl{leaf} extends Node}
+\begin{itemize}
+  \item \texttt{\textsl{int[]} response}
+
+  \textsf{leaf.response[i] stores response of leaf.ops[i]}
+
+  \item \texttt{\textsl{int} maxOld}
+
+  \textsf{Index of the youngest old block in the root that this process has seen yet.}
+
 \end{itemize}
@@
   \textsf{: id of the group of blocks including this propagated together, more precisely the value of the node's counter when propagating this block.}
+  \item \texttt{\textsl{int} order}
+  \textsf{: the index of the block in the node containing it}
 \end{itemize}
@@
   \item \texttt{\textsl{int} size}
-  \textsf{size of queue after this block's operations finish}
+  \textsf{: size of queue after this block's operations finish}
   \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
-  \textsf{count of non-null dequeus up to this block}
+  \textsf{: count of non-null dequeus up to this block}
+  \item \texttt{\textsl{int} age}
+  \textsf{: number of finished operations in the block}
 \end{itemize}
@@
 \State \Call{Append}{b}
-\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, op.loc, 1}
-\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned}
+\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, b.order, 1}
+\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned} \label{deqRest}
 \State \Return \Call{Get}{res}
+\State \texttt{b\textsubscript{i}= b}\Comment{block in the root contains the invocation of dequeue}
+\State \texttt{b\textsubscript{r}= root.blocks.get(sum\textsubscript{enq}==i)}\Comment{block in the root contains the invocation of dequeue}
+\State \texttt{b\textsubscript{i}.age= b\textsubscript{i}.age+1}
+\State \texttt{b\textsubscript{r}.age= b\textsubscript{r}.age+1}
+\If{\texttt{b\textsubscript{i} or b\textsubscript{r} become old}} \texttt{update maxOld}
+\EndIf
 \EndFunction{Dequeue}
@@
 \If{\texttt{new.num==0}} \Return{\texttt{true}}
+\ElsIf{\texttt{n} \textbf{is} \texttt{root}}
+\If{\texttt{root.blocks.append(new)}}\State\texttt{goto \ref{okcas}}\EndIf
 \ElsIf{\Call{CAS}{n.blocks[h], null, new}}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
+\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}} \label{okcas}
 \State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, h+1}}
@@
 \EndIf
-\State \texttt{res= \Call{BSearch}{root, sum\textsubscript{enq}, i, 0, root.head}}
+\State \texttt{res= root.blocks.get(sum\textsubscript{enq}==i).order}
 \State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
@@
 \State \texttt{block b= \Call{new}{\textsl{block}}}
+\State \texttt{b.order= i}
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
@@
 \begin{algorithm}
-\caption{Queue Continued}
 \begin{algorithmic}[1]
@@
 \Else
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
+\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in the left child of \texttt{n}}
 \State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
@@
 \Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks[b-1]+i, b}
+\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks.get(order==b-1).sum\textsubscript{deq}+i, b}
 \Else
@@
 \begin{algorithm}
-%\caption{Modified Queue \label{algQx}}
 \begin{algorithmic}[1]
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[RBTNode]} rootBlocks}
-  \Statex \textsf{A persistant redblack tree supporting \texttt{append(n),get(i),split(j)}}. \texttt{append(n)} returns \texttt{true} in case successful.
+\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[rootBlock]}}
+  \Statex \textsf{A persistant red-black tree supporting \texttt{append(b, key),get(key=i),split(j)}}. \texttt{append(b, key)} returns \texttt{true} in case successful.
 
-\Statex $\blacktriangleright$ \texttt{\textsl{RBTNode}}
-\begin{itemize}
-
-  \item \texttt{\textsl{RootBlock} block}
-
-  \item \texttt{\textsl{int} age}
-
-  \textsf{number of finished operations in the block}
-
-  \item \texttt{\textsl{int} step}
-
-  \textsf{If this node is the root of the RBTree, then step shows the number of appended nodes to the RBTree}
-
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{leaf} extends Node}
-\begin{itemize}
-  \item \texttt{\textsl{int[]} response}
-
-  \textsf{leaf.response[i] stores response of leaf.ops[i]}
-
-  \item \texttt{\textsl{int} maxOld}
-
-  \textsf{Index of the youngest old block in the root that this process has seen yet.}
-
-\end{itemize}
-
-\Statex
-
-\Function{boolean}{Refresh}{\textsl{node} n} \Comment if n is root
-\State \texttt{new=CreateBlock()} \Comment TODO
-\If{\texttt{new.num==0}} \Return{\texttt{true}}
-\ElsIf{\Call{RBTAppend}{new}}
-\Statex..\EndIf
-\EndFunction{Refresh}
-
-\Statex
-
-\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{search in the current RBTRee}
-\Else
-..
-\EndIf
-\EndFunction{Index}
-
-\Statex
-
-\Function{Object}{Dequeue()}{}
-\State ..
-\State \texttt{N\textsubscript{i}= RBTnode n that n.block contains the invocation of Dequeue}
-\State \texttt{N\textsubscript{r}= RBTnode n that n.block contains the response of Dequeue}
-\State \texttt{N\textsubscript{i}.age= N\textsubscript{i}.age+1} \Comment{this is a shared counter}
-\State \texttt{N\textsubscript{r}.age= N\textsubscript{r}.age+1}
-\If{\texttt{N\textsubscript{i} or N\textsubscript{r} become old}} \texttt{update maxOld}
-\EndIf
-\EndFunction{Dequeue}
-
-\pagebreak
-
-\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{rootBlocks}}}
-\State \texttt{root= rootBlocks.root}
-\If{\texttt{root.step\%$p^2$==0}}
+\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{root.blocks}}}
+\State \texttt{step= root.head}
+\If{\texttt{step\%$p^2$==0}}
 \State \texttt{Help()}
@@
 \EndIf
-\State \texttt{new= RBTnode(b,0,null)}
-\State \Return \texttt{rootBlocks.append(new)}
+\State \texttt{b.age= 0}
+\State \Return \texttt{root.blocks.append(b, b.order)}
 \EndFunction{RBTAppend}
@@
 \Function{void}{Help}{}\Comment{Helps pending operations}
-\State{last= l.head-1}
-\For{\texttt{leaf l in leaves}}
-\If{\texttt{l.blocks[last] is not null}}
-\If{op \textbf{is} DEQ}
-\State\texttt{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
-\State \texttt{write the response to l.responses[last]}
+\For{\texttt{leaf l} \textbf{in leaves}}\Comment{\textit{how to iterate over them?}}
+\State{\texttt{last= l.head-1}}
+\If{\texttt{l.blocks[last]} \textbf{is not} \texttt{null}}
+\If{\texttt{l.blocks[last].element==null}} \Comment{operation is dequeue}
+\State \texttt{goto \ref{deqRest} with} \Comment{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
+\State \texttt{l.responses[last]= response}
 \EndIf

### 2022-03-25 article/psuedocode.tex
complete garbage collection methods

@@
 \item \texttt{\textsl{*Node} left, right, parent}
-\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \texttt{en} pointers to the first block of the coresponding children.
-\texttt{blocks[i]} returns the $i$th block stored.}
+\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \texttt{en} pointers to the first block of the corresponding children.
+\texttt{blocks[i]} returns the $i$th block stored.
+In the root node it is implemented with a persistent red-black tree and it is a big array in other nodes.}
 \item \texttt{\textsl{int} head= 1}\textsf{: index of the first empty cell of \texttt{blocks}}
@@
 \State \texttt{b\textsubscript{r}.age= b\textsubscript{r}.age+1}
-\If{\texttt{b\textsubscript{i} or b\textsubscript{r} become old}} \texttt{update maxOld}
+\If{\texttt{b\textsubscript{i}==b\textsubscript{i}.num or b\textsubscript{r}==b.num}}\Comment{become old}
+\State \texttt{this.leaf.maxOld= Max(b\textsubscript{i}*(b\textsubscript{i}-b\textsubscript{i}.num), b\textsubscript{r}*(b\textsubscript{r}-b\textsubscript{r}.num), this.leaf.maxold)}
 \EndIf
@@
 \If{\texttt{n.parent} \textbf{is} \texttt{null}}
+\State \texttt{Cast(b, RootBlock)} \Comment{cast block to a root block}
 \State \texttt{b.size= max(root.blocks[i-1].size + b.num\textsubscript{enq} - b.num\textsubscript{deq}, 0)}
@@
 \Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
-\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
+\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.blocks[b].element}
 \Else
@@
 \If{\texttt{l.blocks[last].element==null}} \Comment{operation is dequeue}
-\State \texttt{goto \ref{deqRest} with} \Comment{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
+\State \texttt{goto \ref{deqRest} with these values <>} \Comment{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate(). \textit{TODO}}
 \State \texttt{l.responses[last]= response}
@@
 \Function{Block}{FindYoungestOld}{b}
-\State\Return  \texttt{read all maxOld values among leaves and decide the largest one} \Comment{There is no need to do a sasDK;Lnapgshot.}
+\For{\texttt{leaf l} \textbf{in leaves}}
+\State\texttt{max= Max(l.maxOld, max)}
+\EndFor
+\State\Return\texttt{max} \Comment{This snapshot suffies.}
 \EndFunction{findYoungestOld}
@@
 
-\If{operation i in leaf l cannot find its desired RootBlock}
+\If{a dequeue cannot find the root block}
 
-\State \Return \texttt{l.response[i]}
+\State \Return \texttt{this.leaf.response(block.order)}
 \EndIf

### 2022-03-25 proposal/proposal.tex
apply committee feedback

@@
 
-\title{Thesis Proposal}
+\title{Wait-free Queues with Polylogarithmic\\ Step Complexity}
 \author{Hossein Naderibeni \\supervised by Eric Ruppert}
@@
 Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
-Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
-%We are going to resolve a common problem in previous shared queues which makes their worst case slow.
+Current approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
+In this work, we are going to introduce a lock-free queue with better performance in the worst case than the previous lock-free queues.
 
@@
 We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every so often but never terminates.
+A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm there might be an operation that takes infinitely many steps but never terminates.
 %From now on, we are going to focus on wait freedom.
@@
   \center\includegraphics[scale=0.5]{pics/msqueue}
-  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrows show new pointers and gray dashed arrows show the old pointers.}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
+In the first diagram the first element has been dequeued.
+Red arrows show new pointers and gray dashed arrows show the old pointers.}
 \end{figure}

### 2022-03-28 proposal/proposal.tex
add first paragraph

@@
 
-\paragraph{}
+In this work, we are going to introduce a novel lock-free queue.
+Linearizable lock-free data structures are the current standard requirements for designing distributed algorithms.
+All lock-free queues in the literature yet have a common problem in their worst case.
+Our algorithm solves this problem, performing better than them.
+
 Shared data structures have become an essential field in distributed algorithms research.
@@
 Current approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
-In this work, we are going to introduce a lock-free queue with better performance in the worst case than the previous lock-free queues.
 

### 2022-03-28 article/psuedocode.tex
fix meetining 13 notes

--- 2022-03-24 article/psuedocode.tex
@@
 
-\algnewcommand\algorithmicforeach{\textbf{for each}}
+\algnewcommand\algorithmicforeach{\bf{for each}}
 \algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
@@
 \algdef{S}[FUNCTION]{Function}
-   [3]{{\texttt{\textsl{#1}}} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}
+   [3]{{\tt{\sl{#1}}} \textproc{\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
 
 \algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \texttt{\textproc{#1}}}
+   [1]{\algorithmicend\ \tt{\textproc{#1}}}
 
-\algrenewcommand\Call[2]{\texttt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+\algrenewcommand\Call[2]{\tt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
 
@@
 
+\newcommand{\sub}[1]{\textsubscript{#1}}
+\renewcommand{\tt}[1]{\texttt{#1}}
+\renewcommand{\sl}[1]{\textsl{#1}}
+\renewcommand{\it}[1]{\textit{#1}}
+\renewcommand{\bf}[1]{\textbf{#1}}
+\newcommand{\cmt}[1]{\Comment{#1}}
+
 \usepackage{amsmath,amssymb,amsthm}
@@
 
-%\Statex \textbf{Structure}
+%\Statex \bf{Structure}
 
-\Statex $\diamondsuit$ \texttt{\textsl{Local}}
+\Statex $\diamondsuit$ \tt{\sl{Local}}
 \begin{itemize}
-\item \texttt{\textsl{*Node} leaf\textsf{: pointer the the process's leaf in the tree}}
+\item \tt{\sl{*Node} leaf\textsf{: pointer the the process's leaf in the tree}}
 \end{itemize}
@@
 
-\Statex $\diamondsuit$ \texttt{\textsl{Shared}}
+\Statex $\diamondsuit$ \tt{\sl{Shared}}
 \begin{itemize}
-\item \texttt{\textsl{Tree}} \textit{to complete, how?}
+\item \tt{\sl{Tree}} \textsf{: A binary tree of \tt{Node}s is shared among the processes.
+It can be implemented with a 1 index based array of size $p$.
+Such that the root is index 1, the left child and the right child of a node with index i are indices 2i, 2i+1 in the array.}
 \end{itemize}
@@
 \Statex
-\Statex $\diamondsuit$ \texttt{\textsl{Structures}}
+\Statex $\diamondsuit$ \tt{\sl{Structures}}
 
-\Statex $\blacktriangleright$ \texttt{\textsl{Node}}
+\Statex $\blacktriangleright$ \tt{\sl{Node}}
 \begin{itemize}
-\item \texttt{\textsl{*Node} left, right, parent}
-\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \texttt{en} pointers to the first block of the coresponding children.
-\texttt{blocks[i]} returns the $i$th block stored.}
-\item \texttt{\textsl{int} head= 1}\textsf{: index of the first empty cell of \texttt{blocks}}
-\item \texttt{\textsl{int} counter= 0}\textsf{}
-\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is field \texttt{i}}
+\item \tt{\sl{*Node} left, right, parent}
+\item \tt{\sl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \tt{en} pointers to the first block of the corresponding children.
+\tt{blocks[i]} returns the $i$th block stored.
+In the root node it is implemented with a persistent red-black tree and it is a big array in the other nodes.}
+\item \tt{\sl{int} head= 1}\textsf{: index of the first empty cell of \tt{blocks}}
+\item \tt{\sl{int} counter= 0}\textsf{}
+\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \tt{time} is field \tt{i}}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \texttt{\textsl{leaf} extends Node}
+\Statex $\blacktriangleright$ \tt{\sl{leaf} extends Node}
 \begin{itemize}
-  \item \texttt{\textsl{int[]} response}
+  \item \tt{\sl{int[]} response}
 
@@
 
-  \item \texttt{\textsl{int} maxOld}
+  \item \tt{\sl{int} \tt{last\sub{done}}}
 
-  \textsf{Index of the youngest old block in the root that this process has seen yet.}
+  \textsf{Each process stores the index of the most recent block that the process has finished its last operation.
+An enqueue operation is finished if it has appended its element to the root and a dequeue operation is finished when it computes its response.}
 
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{Block}}
 
 \begin{itemize}
-  \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}
-  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
-  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
-  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
-  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
-  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
-  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
-  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
-  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
+  \item \tt{\sl{int} num\sub{enq-left}, sum\sub{enq-left}}
+  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \tt{num\sub{enq-left}}}
+  \item \tt{\sl{int} num\sub{deq-left}, sum\sub{deq-left}}
+  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \tt{num\sub{deq-left}}}
+  \item \tt{\sl{int} num\sub{enq-right}, \tt{sum\sub{enq-right}}}
+  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \tt{num\sub{enq-right}}}
+  \item \tt{\sl{int} num\sub{deq-right}, \tt{sum\sub{deq-right}}}
+  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \tt{num\sub{deq-right}}}
+  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
   \textsf{: \# enqueue, dequeue operations in the block}
-  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
+  \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
   \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
-  \item \texttt{\textsl{int} num, sum}
-  \textsf{: total \# operations in block, prefix sum of \texttt{num}}
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
+  \item \tt{\sl{int} num, sum}
+  \textsf{: total \# operations in block, prefix sum of \tt{num}}
+  \item \tt{\sl{int} end\sub{left}, end\sub{right}}
   \textsf{:~~index of the last subblock in the left and right child}
-  \item \texttt{\textsl{int} group}
-  \textsf{: id of the group of blocks including this propagated together, more precisely the value of the node's counter when propagating this block.}
-  \item \texttt{\textsl{int} order}
+  \item \tt{\sl{int} group}
+  \textsf{: id of the group of blocks including this propagated together, more precisely the value read from the node \tt{n}'s counter when propagating this block to the node \tt{n}.}
+  \item \tt{\sl{int} order}
   \textsf{: the index of the block in the node containing it}
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{Leaf Block} extends \textsl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{Leaf Block} extends \sl{Block}}
 \begin{itemize}
-  \item \texttt{\textsl{Object} element}
-  \textsf{Each block in a leaf also represents an operation. The \texttt{element} shows the operations argument if it is an enqueue, and if it is a dequeue the value is \texttt{null}.}
+  \item \tt{\sl{Object} element}
+  \textsf{Each block in a leaf represents an operation. The \tt{element} shows the operation's argument if it is an enqueue, and if it is a dequeue this value is \tt{null}.}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \texttt{\textsl{Root Block} extends \textsl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{Root Block} extends \sl{Block}}
 \begin{itemize}
-  \item \texttt{\textsl{int} size}
+  \item \tt{\sl{int} size}
   \textsf{: size of queue after this block's operations finish}
-  \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
+  \item \tt{\sl{int} sum\sub{non-null deq}}
   \textsf{: count of non-null dequeus up to this block}
-  \item \texttt{\textsl{int} age}
+  \item \tt{\sl{int} num\sub{done}}
   \textsf{: number of finished operations in the block}
@@
 
-\Function{void}{Enqueue}{\textsl{Object} e}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.element= e}
-\State \texttt{b.sum\textsubscript{enq}=1}
+\Function{void}{Enqueue}{\sl{Object} e}
+\State \tt{block b= \Call{new}{\sl{block}}}
+\State \tt{b.element= e}
+\State \tt{b.num\sub{enq}=1}
+\State \tt{b.sum\sub{enq}= this.blocks[leaf.head].sum\sub{enq}+1}
 \State \Call{Append}{b}
@@
 \Function{Object}{Dequeue()}{}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.element= null}
-\State \texttt{b.sum\textsubscript{deq}=1}
+\State \tt{block b= \Call{new}{\sl{block}}}
+\State \tt{b.element= null}
+\State \tt{b.num\sub{deq}=1}
+\State \tt{b.sum\sub{deq}= this.blocks[leaf.head].sum\sub{deq}+1}
 \State \Call{Append}{b}
-\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, b.order, 1}
-\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned} \label{deqRest}
-\State \Return \Call{Get}{res}
-\State \texttt{b\textsubscript{i}= b}\Comment{block in the root contains the invocation of dequeue}
-\State \texttt{b\textsubscript{r}= root.blocks.get(sum\textsubscript{enq}==i)}\Comment{block in the root contains the invocation of dequeue}
-\State \texttt{b\textsubscript{i}.age= b\textsubscript{i}.age+1}
-\State \texttt{b\textsubscript{r}.age= b\textsubscript{r}.age+1}
-\If{\texttt{b\textsubscript{i} or b\textsubscript{r} become old}} \texttt{update maxOld}
+\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf.
+\tt{b\sub{i}} is the block in the root containing it.}
+\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1.} \label{deqRest}
+\If{\tt{index\sub{response}!=-1}}
+\State \tt{output= null}
+\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
+\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
+\State \tt{this.leaf.last\sub{done}= b\sub{r}}
 \EndIf
+\Else
+\State \tt{output= \Call{Get}{res}}
+\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response})}\cmt{block in the root contains response enqueue.}
+\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
+\State \tt{b\sub{r}.num\sub{done}= b\sub{r}.num\sub{done}+1}
+\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
+\State \tt{this.leaf.last\sub{done}= b\sub{r}}
+\ElsIf {\tt{b\sub{i}.num\sub{done}==b\sub{i}.num}}
+\State \tt{this.leaf.last\sub{done}= b\sub{i}}
+\EndIf
+\EndIf
+\State \Return{\tt{output}}
 \EndFunction{Dequeue}
-
-\Statex
-
-\Function{int}{ComputeHead}{int i, int b}\Comment{Computes head of the queue when \texttt{i}th dequeue in \texttt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
-\If{\texttt{root.blocks[b-1].size + root.blocks[b].num\textsubscript{enq} - i $<$ 0}}
-\State \Return \texttt{-1}
-\Else{ \Return \texttt{root.blocks[b-1].sum\textsubscript{non-null deq} + i}}
-\EndIf
-\EndFunction{ComputeHead}
-
-\Statex
-
-\Function{void}{Append}{\textsl{block} b} 
-\State \texttt{b.group= this.leaf.head}
-\State \texttt{l\textsubscript{pid}.blocks[this.leaf.head]= b}
-\State \texttt{this.leaf.head+=1}
-\State \Call{Propagate}{this.leaf.parent} 
-\EndFunction{Append}
 
@@
 \begin{algorithm}
-%\caption{Queue Continued}
 \begin{algorithmic}[1]
@@
 
-\Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{not} \Call{Refresh}{n}}
+\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes head of the queue when \tt{i}th dequeue in \tt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
+\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
+\State \Return \tt{-1}
+\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
+\EndIf
+\EndFunction{ComputeDeqRes}
+
+\Statex
+
+\Function{void}{Append}{\sl{block} b} 
+\State \tt{b.group= this.leaf.head}
+\State \tt{l\sub{pid}.blocks[this.leaf.head]= b}
+\State \tt{this.leaf.head+=1}
+\State \Call{Propagate}{this.leaf.parent} 
+\EndFunction{Append}
+
+\Statex
+
+\Function{void}{Propagate}{\sl{node} n}
+\If{\tt{not} \Call{Refresh}{n}}
 \State \Call{Refresh}{n}
 \EndIf
-\If{\texttt{n.parent} \textbf{is not} \texttt{null}} 
+\If{\tt{n.parent} \bf{is not} \tt{null}} 
 \State \Call{Propagate}{n.parent}
@@
 
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{h= n.head}
-%\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
-\State \texttt{c= n.counter}
-\State \texttt{<new, c\textsubscript{left}, c\textsubscript{right}>= \Call{CreateBlock}{n, h}}
-\State \texttt{new.group= c}
-\If{\texttt{new.num==0}} \Return{\texttt{true}}
-\ElsIf{\texttt{n} \textbf{is} \texttt{root}}
-\If{\texttt{root.blocks.append(new)}}\State\texttt{goto \ref{okcas}}\EndIf
-\ElsIf{\Call{CAS}{n.blocks[h], null, new}}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}} \label{okcas}
-\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, h+1}}
-\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
+\Function{boolean}{Refresh}{\sl{node} n}
+\State \tt{h= n.head}
+%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
+\State \tt{c= n.counter}
+\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}}
+\State \tt{new.group= c}
+\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
+\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{space in he first of the new line?}}
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
+\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}}
+\State \tt{\Call{CAS}{n.dir.counter, c\sub{dir}, c\sub{dir}+1}}
 \EndFor
-\State \texttt{\Call{CAS}{n.head, h, h+1}}
-\State \Return{\texttt{true}}
+\State \tt{\Call{CAS}{n.head, h, h+1}}
+\State \Return{\tt{true}}
 \Else
-\State \texttt{\Call{CAS}{n.head, h, h+1}}
-\State \Return{ \texttt{false}}
+\State \tt{\Call{CAS}{n.head, h, h+1}}
+\State \Return{ \tt{false}}
 \EndIf
@@
 
-\Function{element}{Get}{\textsl{int} i}
-\Comment{Returns $i$th Enqueue.}
-\If{\texttt{i }\textbf{is }\texttt{null}}
-\State \Return \texttt{null}
+\Function{element}{Get}{\sl{int} i}
+\cmt{Returns $i$th Enqueue.}
+\If{\tt{i }\bf{is }\tt{null}}
+\State \Return \tt{null}
 \EndIf
-\State \texttt{res= root.blocks.get(sum\textsubscript{enq}==i).order}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
+\State \tt{res= root.blocks.get(enq, i).order}
+\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
 \EndFunction{Get}
@@
 
-\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
-\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}
+\Statex $\leadsto$ \textsf{Precondition: \tt{n.blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
+\Function{int}{BSearch}{\sl{node} n, \sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
 
-\Statex \Comment{\textmd{Does binary search for~the value \texttt{i} of the given prefix sum \texttt{feild}.
-Returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
-%\State \Return \texttt{result block's index}
+\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{feild}.
+Returns the index of the leftmost block in \tt{n.blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
+%\State \Return \tt{result block's index}
 \EndFunction{BSearch}
 
-\pagebreak
+\Statex
 
-\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
-\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.order= i}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{lastIndex= n.dir.head} \label{lastLine}
-\State \texttt{prevIndex= n.blocks[i-1].end\textsubscript{dir}} \label{prevLine}
-\State \texttt{lastBlock= n.dir.blocks[lastIndex]}
-\State \texttt{prevBlock= n.dir.blocks[prevIndex]}
-\State \texttt{c\textsubscript{dir}= n.dir.counter}
-\State \texttt{b.end\textsubscript{dir}= lastIndex}
-\State \texttt{b.num\textsubscript{enq-dir}= lastBlock.sum\textsubscript{enq} - prevBlock.sum\textsubscript{enq}}
-\State \texttt{b.num\textsubscript{deq-dir}= lastBlock.sum\textsubscript{deq} - prevBlock.sum\textsubscript{deq}}
-\State \texttt{b.sum\textsubscript{enq-dir}= n.blocks[i-1].sum\textsubscript{enq-dir} + b.num\textsubscript{enq-dir}}
-\State \texttt{b.sum\textsubscript{deq-dir}= n.blocks[i-1].sum\textsubscript{deq-dir} + b.num\textsubscript{deq-dir}}
+\Function{<Block, int, int>}{CreateBlock}{\sl{node} n, \sl{int} i} 
+\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
+\State \tt{block b= \Call{new}{\sl{block}}}
+\If{\tt{n} \bf{is} \tt{root}}
+\State \tt{block b= \Call{new}{\sl{root block}}}
+\EndIf
+\State \tt{b.order= i}
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
+\State \tt{lastIndex= n.dir.head} \label{lastLine}
+\State \tt{prevIndex= n.blocks[i-1].end\sub{dir}} \label{prevLine}
+\State \tt{lastBlock= n.dir.blocks[lastIndex]}
+\State \tt{prevBlock= n.dir.blocks[prevIndex]}
+\State \tt{c\sub{dir}= n.dir.counter}
+\State \tt{b.end\sub{dir}= lastIndex}
+\State \tt{b.num\sub{enq-dir}= lastBlock.sum\sub{enq} - prevBlock.sum\sub{enq}}
+\State \tt{b.num\sub{deq-dir}= lastBlock.sum\sub{deq} - prevBlock.sum\sub{deq}}
+\State \tt{b.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
+\State \tt{b.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
 \EndFor
-\State \texttt{b.num\textsubscript{enq}= b.num\textsubscript{enq-left} + b.num\textsubscript{enq-right}}
-\State \texttt{b.num\textsubscript{deq}= b.num\textsubscript{deq-left} + b.num\textsubscript{deq-right}}
-\State \texttt{b.num= b.num\textsubscript{enq} + b.num\textsubscript{deq}}
-\State \texttt{b.sum= n.blocks[i-1].sum + b.num}
+\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
+\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
+\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
+\State \tt{b.sum= n.blocks[i-1].sum + b.num}
 
-\If{\texttt{n.parent} \textbf{is} \texttt{null}}
-\State \texttt{b.size= max(root.blocks[i-1].size + b.num\textsubscript{enq} - b.num\textsubscript{deq}, 0)}
-\State \texttt{b.sum\textsubscript{non-null deq}= root.blocks[i-1].sum\textsubscript{non-null deq} + max( b.num\textsubscript{deq} - root.blocks[i-1].size - b.num\textsubscript{enq}, 0)}
+\If{\tt{n.parent} \bf{is} \tt{null}}
+\State \tt{b.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
+\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
 \EndIf
 
-\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{right}}
+\State \Return \tt{b, c\sub{left}, c\sub{right}}
 \EndFunction{CreateBlock}
@@
 
-\Statex $\leadsto$ \textsf{Precondition:~\texttt{n.blocks[b]} contains $\geq$\texttt{i} enqueues.}
-\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
-\Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
-\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
+\Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains $\geq$\tt{i} enqueues.}
+\Function{element}{Get}{\sl{node} n, \sl{int} b, \sl{int} i} 
+\cmt{\textmd{Returns the \tt{i}th Enqueue in \tt{b}th block of node \tt{n}}}
+\If{\tt{n {\keywordfont is} leaf}} \Return \tt{n.blocks[b].element}
 \Else
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in the left child of \texttt{n}}
-\State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
-\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\textsubscript{enq}} 
+\If{\tt{i $\leq$ n.blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of \tt{n}}
+\State \tt{subBlock= \Call{BSearch}{n.left, sum\sub{enq}, i, n.blocks[b-1].end\sub{left}+1, n.blocks[b].end\sub{left}}}
+\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\sub{enq}} 
 \Else
-\State \texttt{i= i-n.blocks[b].num\textsubscript{enq-left}}
-\State\texttt{subBlock=\Call{BSearch}{n.right, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{right}+1, n.blocks[b].end\textsubscript{right}}}
-\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\textsubscript{enq}} 
+\State \tt{i= i-n.blocks[b].num\sub{enq-left}}
+\State\tt{subBlock=\Call{BSearch}{n.right, sum\sub{enq}, i, n.blocks[b-1].end\sub{right}+1, n.blocks[b].end\sub{right}}}
+\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\sub{enq}} 
 \EndIf
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \texttt{b}th block of node \texttt{n} has propagated up to the root and \texttt{i}th dequeue resides in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
-\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks.get(order==b-1).sum\textsubscript{deq}+i, b}
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of node \tt{n} has propagated up to the root and \tt{i}th dequeue resides in node \tt{n} is in block \tt{b} of node \tt{n}.}
+\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root among dequeus, of ith dequeue in bth block of node n.}
+\If{\tt{n {\keywordfont is} root}} \Return \tt{root.blocks.get(order==b-1).sum\sub{deq}+i, b}
 \Else
-\State \texttt{dir= (n.parent.left==n)? left: right}
-\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
-\If{\texttt{dir {\keywordfont is} left}}
-\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq-right}}
-\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq} + n.blocks[superBlock].sum\textsubscript{deq-left}}
+\State \tt{dir= (n.parent.left==n)? left: right}
+\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
+\If{\tt{dir {\keywordfont is} left}}
+\State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq-right}}
+\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}}
 \EndIf
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[rootBlock]}}
-  \Statex \textsf{A persistant red-black tree supporting \texttt{append(b, key),get(key=i),split(j)}}. \texttt{append(b, key)} returns \texttt{true} in case successful.
+\Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
+  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}. \tt{append(b, key)} returns \tt{true} in case successful.
 
-\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{root.blocks}}}
-\State \texttt{step= root.head}
-\If{\texttt{step\%$p^2$==0}}
-\State \texttt{Help()}
-\State \texttt{CollectGarbage()}
+\Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
+\State \tt{step= root.head}
+\If{\tt{step\%$p^2$==0}}
+\State \tt{Help()}
+\State \tt{CollectGarbage()}
 \EndIf
-\State \texttt{b.age= 0}
-\State \Return \texttt{root.blocks.append(b, b.order)}
+\State \tt{b.age= 0}
+\State \Return \tt{root.blocks.append(b, b.order)}
 \EndFunction{RBTAppend}
@@
 
-\Function{void}{Help}{}\Comment{Helps pending operations}
-\For{\texttt{leaf l} \textbf{in leaves}}\Comment{\textit{how to iterate over them?}}
-\State{\texttt{last= l.head-1}}
-\If{\texttt{l.blocks[last]} \textbf{is not} \texttt{null}}
-\If{\texttt{l.blocks[last].element==null}} \Comment{operation is dequeue}
-\State \texttt{goto \ref{deqRest} with} \Comment{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
-\State \texttt{l.responses[last]= response}
+\Function{void}{Help}{}\cmt{Helps pending operations}
+\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{how to iterate over them?}}
+\State{\tt{last= l.head-1}}
+\If{\tt{l.blocks[last]} \bf{is not} \tt{null}}
+\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
+\State \tt{goto \ref{deqRest} with these values <>} \cmt{run \tt{Dequeue()}  for \tt{l.ops[last]} after Propagate(). \it{TODO}}
+\State \tt{l.responses[last]= response}
 \EndIf
@@
 
-\Function{void}{CollectGrabage}{}\Comment{Collects the old root blocks.}
-\State \texttt{l=FindYoungestOld(Root.Blocks.root)}
-\State \texttt{t1,t2= RBT.split(l)}
-\State \texttt{RBTRoot.CAS(t2.root)}
-\EndFunction{CollectGrabage}
+\Function{void}{CollectGarbage}{}\cmt{Collects the old root blocks.}
+\State \tt{l=FindYoungestOld(Root.Blocks.root)}
+\State \tt{t1,t2= RBT.split(l)}
+\State \tt{RBTRoot.CAS(t2.root)}
+\EndFunction{CollectGarbage}
 
@@
 \Function{Block}{FindYoungestOld}{b}
-\State\Return  \texttt{read all maxOld values among leaves and decide the largest one} \Comment{There is no need to do a sasDK;Lnapgshot.}
+\For{\tt{leaf l} \bf{in leaves}}
+\State\tt{max= Max(l.maxOld, max)}
+\EndFor
+\State\Return\tt{max} \cmt{This snapshot suffies.}
 \EndFunction{findYoungestOld}
@@
 
-\Function{response}{FallBack}{op i}
+\Function{response}{FallBack}{op i} \cmt{\it{really necessary?}}
 
-\If{operation i in leaf l cannot find its desired RootBlock}
+\If{a dequeue cannot find the root block}
 
-\State \Return \texttt{l.response[i]}
+\State \Return \tt{this.leaf.response(block.order)}
 \EndIf

### 2022-04-01 article/psuedocode.tex
fix fallback, add comments

@@
   \textsf{: id of the group of blocks including this propagated together, more precisely the value read from the node \tt{n}'s counter when propagating this block to the node \tt{n}.}
-  \item \tt{\sl{int} order}
-  \textsf{: the index of the block in the node containing it}
 \end{itemize}
@@
   \textsf{: count of non-null dequeus up to this block}
-  \item \tt{\sl{int} num\sub{done}}
+  \item \tt{\sl{int} num\sub{finished}}
   \textsf{: number of finished operations in the block}
+    \item \tt{\sl{int} order}
+  \textsf{: the index of the block in the node containing it. Useful in the root since in the PBRT we do not keep indices in another way.}
 \end{itemize}
-
-\Statex
-
-\Function{void}{Enqueue}{\sl{Object} e}
-\State \tt{block b= \Call{new}{\sl{block}}}
-\State \tt{b.element= e}
-\State \tt{b.num\sub{enq}=1}
-\State \tt{b.sum\sub{enq}= this.blocks[leaf.head].sum\sub{enq}+1}
-\State \Call{Append}{b}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Function{Object}{Dequeue()}{}
-\State \tt{block b= \Call{new}{\sl{block}}}
-\State \tt{b.element= null}
-\State \tt{b.num\sub{deq}=1}
-\State \tt{b.sum\sub{deq}= this.blocks[leaf.head].sum\sub{deq}+1}
-\State \Call{Append}{b}
-\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf.
-\tt{b\sub{i}} is the block in the root containing it.}
-\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1.} \label{deqRest}
-\If{\tt{index\sub{response}!=-1}}
-\State \tt{output= null}
-\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
-\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
-\State \tt{this.leaf.last\sub{done}= b\sub{r}}
-\EndIf
-\Else
-\State \tt{output= \Call{Get}{res}}
-\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response})}\cmt{block in the root contains response enqueue.}
-\State \tt{b\sub{i}.num\sub{done}= b\sub{i}.num\sub{done}+1}
-\State \tt{b\sub{r}.num\sub{done}= b\sub{r}.num\sub{done}+1}
-\If{\tt{b\sub{r}.num\sub{done}==b\sub{r}.num}}\cmt{become old}
-\State \tt{this.leaf.last\sub{done}= b\sub{r}}
-\ElsIf {\tt{b\sub{i}.num\sub{done}==b\sub{i}.num}}
-\State \tt{this.leaf.last\sub{done}= b\sub{i}}
-\EndIf
-\EndIf
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
 
@@
 
-\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes head of the queue when \tt{i}th dequeue in \tt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
-\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
-\State \Return \tt{-1}
-\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
+\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
+\State \tt{block b= \Call{new}{\sl{leaf block}}}
+\State \tt{b.element= e}
+\State \tt{b.num\sub{enq}=1}
+\State \tt{b.sum\sub{enq}= this.leaf.blocks[this.leaf.head].sum\sub{enq}+1}
+\State \Call{Append}{b}
+\EndFunction{Enqueue}
+
+\Statex
+\Function{Object}{Dequeue()}{}
+\State \tt{block b= \Call{new}{\sl{leaf block}}} \cmt{Creates a null element block, appends it to the tree, computes its order among operations, then computes its response index if it exists and returns the response's element.}
+\State \tt{b.element= null}
+\State \tt{b.num\sub{deq}=1}
+\State \tt{b.sum\sub{deq}= this.leaf.blocks[this.leaf.head].sum\sub{deq}+1} \cmt{\tt{this} is the current running process}
+\State \Call{Append}{b}
+\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf.
+\tt{b\sub{i}} is the index of the block in the root containing it.
+Since only one invocation of \tt{dequeue} is running by \tt{this} process at one time we are allowed to used \tt{this.leaf.head} safely.}
+\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
+\If{\tt{index\sub{response}!=-1}}
+\State \tt{output= null}
+\State \tt{b\sub{deq}=root.blocks[b\sub{i}]}
+\State \tt{b\sub{deq}.num\sub{finished}.inc()} \cmt{shared counter}
+\If{\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}}\cmt{all the operations in the block containing the dequeue are finished.}
+\State \tt{this.leaf.last\sub{done}= b\sub{i}}
 \EndIf
-\EndFunction{ComputeDeqRes}
+\Else
+\State \tt{output= \Call{Get}{index\sub{response}}}
+\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response}).order}\cmt{index of the block in the root contains response enqueue.}
+\State \tt{b\sub{enq}=root.blocks[b\sub{r}]}
+\State \tt{b\sub{enq}.num\sub{finished}.inc()}
+\State \tt{b\sub{deq}.num\sub{finished}.inc()}
+\If{\tt{b\sub{enq}.num\sub{finished}==b\sub{enq}.num}}\cmt{become done}
+\State \tt{this.leaf.last\sub{done}= b\sub{r}}
+\ElsIf {\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}} \cmt{\tt{b\sub{deq}} comes after \tt{b\sub{enq}}.}
+\State \tt{this.leaf.last\sub{done}= b\sub{i}} \cmt{\tt{this.leaf.last\sub{done}} is an increasing value.}
+\EndIf
+\EndIf
+\State \Return{\tt{output}}
+\EndFunction{Dequeue}
 
@@
 \Function{void}{Append}{\sl{block} b} 
-\State \tt{b.group= this.leaf.head}
-\State \tt{l\sub{pid}.blocks[this.leaf.head]= b}
-\State \tt{this.leaf.head+=1}
+\State \tt{b.group= this.leaf.head} \label{appendStart} \cmt{Only this block is propagated from the leaf by itself.}
+\State \tt{this.leaf.blocks[this.leaf.head]= b} 
+\State \tt{this.leaf.head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
 \State \Call{Propagate}{this.leaf.parent} 
@@
 \If{\tt{not} \Call{Refresh}{n}}
-\State \Call{Refresh}{n}
+\State \Call{Refresh}{n} \cmt{Lemma Double Refresh}
 \EndIf
-\If{\tt{n.parent} \bf{is not} \tt{null}} 
+\If{\tt{n} \bf{is not} \tt{root}} \cmt{To check anode is the root we can check its index if the tree is implemented by an array or check if \tt{n.parent} is not \tt{null}.} 
 \State \Call{Propagate}{n.parent}
@@
 
+\Function{element}{Get}{\sl{int} i}
+\cmt{Returns $i$th Enqueue.}
+\State \tt{res= root.blocks.get(enq, i).order}
+\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
+\EndFunction{Get}
+
+\Statex
+
+\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns the index of the the head of the queue or -1 if queue is empty.}
+\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
+\State \Return \tt{-1}
+\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
+\EndIf
+\EndFunction{ComputeDeqRes}
+
+\end{multicols}
+\end{algorithmic}
+\end{algorithm}
+
+%##########################################
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{33}
+\begin{multicols}{2}
+
 \Function{boolean}{Refresh}{\sl{node} n}
@@
 \State \tt{c= n.counter}
-\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}}
+\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}} \cmt{\tt{c\sub{left}, c\sub{right}} are the values read from \tt{n}'s children's counters.}
 \State \tt{new.group= c}
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{space in he first of the new line?}}
+\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{how to put space in he first of the new line?}}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}}
+\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}} \cmt{Superblock's Lemma}
 \State \tt{\Call{CAS}{n.dir.counter, c\sub{dir}, c\sub{dir}+1}}
@@
 \Else
-\State \tt{\Call{CAS}{n.head, h, h+1}}
+\State \tt{\Call{CAS}{n.head, h, h+1}} \cmt{Even if another process wins, help to increase the head. It might fell sleep before increasing.}
 \State \Return{ \tt{false}}
@@
 \EndFunction{Refresh}
-
-\Statex
-
-\Function{element}{Get}{\sl{int} i}
-\cmt{Returns $i$th Enqueue.}
-\If{\tt{i }\bf{is }\tt{null}}
-\State \Return \tt{null}
-\EndIf
-\State \tt{res= root.blocks.get(enq, i).order}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
-\EndFunction{Get}
 
@@
 
-\Statex
+\pagebreak
 
 \Function{<Block, int, int>}{CreateBlock}{\sl{node} n, \sl{int} i} 
-\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}. Returns the created block as well as values read from each child counter feild.}
+\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}.
+Returns the created block as well as values read from each child counter field.
+The values are used for incrementing children's counters if the block was appended to \tt{n.blocks} successfully.
+\it{Does it need help?
+I think no but in that case we do not have to pass these values to the calling line.}}
 \State \tt{block b= \Call{new}{\sl{block}}}
 \If{\tt{n} \bf{is} \tt{root}}
-\State \tt{block b= \Call{new}{\sl{root block}}}
+\State \tt{b= \Call{new}{\sl{root block}}}
 \EndIf
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{lastIndex= n.dir.head} \label{lastLine}
-\State \tt{prevIndex= n.blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{lastBlock= n.dir.blocks[lastIndex]}
-\State \tt{prevBlock= n.dir.blocks[prevIndex]}
+\State \tt{index\sub{last}= n.dir.head} \label{lastLine}
+\State \tt{index\sub{prev}= n.blocks[i-1].end\sub{dir}} \label{prevLine}
+\State \tt{block\sub{last}= n.dir.blocks[index\sub{last}]}
+\State \tt{block\sub{prev}= n.dir.blocks[index\sub{prev}]} \cmt{\tt{n.dir.blocks[index\sub{prev}..index\sub{last}]} are merged to one block.}
 \State \tt{c\sub{dir}= n.dir.counter}
-\State \tt{b.end\sub{dir}= lastIndex}
-\State \tt{b.num\sub{enq-dir}= lastBlock.sum\sub{enq} - prevBlock.sum\sub{enq}}
-\State \tt{b.num\sub{deq-dir}= lastBlock.sum\sub{deq} - prevBlock.sum\sub{deq}}
+\State \tt{b.end\sub{dir}= index\sub{last}}
+\State \tt{b.num\sub{enq-dir}= block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
+\State \tt{b.num\sub{deq-dir}= block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
 \State \tt{b.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
@@
 
-\If{\tt{n.parent} \bf{is} \tt{null}}
+\If{\tt{n} \bf{is} \tt{root}}
 \State \tt{b.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of node \tt{n} has propagated up to the root and \tt{i}th dequeue resides in node \tt{n} is in block \tt{b} of node \tt{n}.}
-\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root among dequeus, of ith dequeue in bth block of node n.}
-\If{\tt{n {\keywordfont is} root}} \Return \tt{root.blocks.get(order==b-1).sum\sub{deq}+i, b}
+\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
+\If{\tt{n {\keywordfont is} root}} \Return \tt{i, b}
 \Else
-\State \tt{dir= (n.parent.left==n)? left: right}
-\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
+\State \tt{dir= (n.parent.left==n)? left: right} \cmt{check \tt{n} is a left or a right child}
+\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}} \cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
 \If{\tt{dir {\keywordfont is} left}}
 \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq-right}}
-\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}}
+\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from \tt{n}'s right child}
 \EndIf
@@
 \Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
-  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}. \tt{append(b, key)} returns \tt{true} in case successful.
+  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+  \tt{append(b, key)} returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.
 
@@
 \State \tt{step= root.head}
-\If{\tt{step\%$p^2$==0}}
+\If{\tt{step\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root. Used in lemma's using the size of the PBRT.}
 \State \tt{Help()}
@@
 \EndIf
-\State \tt{b.age= 0}
+\State \tt{b.num\sub{finished}= 0}
 \State \Return \tt{root.blocks.append(b, b.order)}
@@
 \Function{void}{Help}{}\cmt{Helps pending operations}
-\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{how to iterate over them?}}
+\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{if the tree is implemented with an array we can iterate over the second half of the array.}}
 \State{\tt{last= l.head-1}}
-\If{\tt{l.blocks[last]} \bf{is not} \tt{null}}
+\cmt{\tt{l.blocks[last]} can not be \tt{null} because of lines \ref{appendStart}-\ref{appendEnd}.}
 \If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
@@
 \State \tt{l.responses[last]= response}
-\EndIf
 \EndIf
@@
 
-\Function{void}{CollectGarbage}{}\cmt{Collects the old root blocks.}
-\State \tt{l=FindYoungestOld(Root.Blocks.root)}
-\State \tt{t1,t2= RBT.split(l)}
+\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
+\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
+\State \tt{t1,t2= RBT.split(order, s)}
 \State \tt{RBTRoot.CAS(t2.root)}
@@
 
-\Function{Block}{FindYoungestOld}{b}
+\Function{Block}{FindMostRecentDone}{b}
 \For{\tt{leaf l} \bf{in leaves}}
@@
 
-\If{a dequeue cannot find the root block}
+\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
 

### 2022-04-12 article/psuedocode.tex
apply meeting 4 feedbacks till createblock

@@
 \begin{algorithm}
-\caption{Queue \label{algQ}}
+\caption{Fields description}
 \begin{algorithmic}[1]
+\setcounter{ALG@line}{100}
 \begin{multicols}{2}
@@
 
-\Statex $\diamondsuit$ \tt{\sl{Local}}
+\Statex $\diamondsuit$ \tt{\sl{Shared}}
 \begin{itemize}
-\item \tt{\sl{*Node} leaf\textsf{: pointer the the process's leaf in the tree}}
+\item \tt{\sl{Node*} tree[]} \textsf{: A binary tree of \tt{Node}s such that \tt{Tree[0]} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.}
 \end{itemize}
@@
 
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
+\Statex $\diamondsuit$ \tt{\sl{Local}}
 \begin{itemize}
-\item \tt{\sl{Tree}} \textsf{: A binary tree of \tt{Node}s is shared among the processes.
-It can be implemented with a 1 index based array of size $p$.
-Such that the root is index 1, the left child and the right child of a node with index i are indices 2i, 2i+1 in the array.}
+\item \tt{\sl{int} leaf\textsf{: \tt{Tree[leaf]} is the process's leaf in the tree.}}
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{*Node} left, right, parent}
-\item \tt{\sl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0 and \tt{en} pointers to the first block of the corresponding children.
-\tt{blocks[i]} returns the $i$th block stored.
-In the root node it is implemented with a persistent red-black tree and it is a big array in the other nodes.}
-\item \tt{\sl{int} head= 1}\textsf{: index of the first empty cell of \tt{blocks}}
-\item \tt{\sl{int} counter= 0}\textsf{}
-\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \tt{time} is field \tt{i}}
+%\item \tt{\sl{*Node} left(), right(), parent()} \textsf{: computed as mentioned in the \tt{Tree[]}}
+\item \tt{\sl{BlockList} blocks}\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+Initially  empty, when \tt{blocks.tryAppend(b)} returns true \tt{b} is appended to the end of the list and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end} pointers to the first block of the corresponding children.}
+
+\item \tt{\sl{int} counter= 0}\textsf{} \textsf{\# groups have been propagated from the node. It may be behind its real value.}
+\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} that its \tt{group} field is \tt{i}.}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{leaf} extends Node}
+\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
 \begin{itemize}
-  \item \tt{\sl{int[]} response}
 
-  \textsf{leaf.response[i] stores response of leaf.ops[i]}
+\item \tt{\sl{PBRT} blocks}
+
+  \textsf{Implemented with a persistent red-black tree.}
+
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
+\begin{itemize}
+
+\item \tt{\sl{Block[]} blocks}
+
+  \textsf{Implemented with an array and using CAS for appending to the head.}
+
+\item \tt{\sl{int} head= 1}\textsf{: \#\tt{block}s in the \tt{blocks}.}
+
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
+\begin{itemize}
 
@@
 
-  \textsf{Each process stores the index of the most recent block that the process has finished its last operation.
-An enqueue operation is finished if it has appended its element to the root and a dequeue operation is finished when it computes its response.}
+  \textsf{Each process stores the index of the most recent block in the root that the process has finished the last operation of the block.
+\tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response.}
 
 \end{itemize}
+\pagebreak
 
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{Block b}} \Comment{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
 
 \begin{itemize}
-  \item \tt{\sl{int} num\sub{enq-left}, sum\sub{enq-left}}
-  \textsf{:~~\#enqueues from subblocks in left child, prefix sum of \tt{num\sub{enq-left}}}
-  \item \tt{\sl{int} num\sub{deq-left}, sum\sub{deq-left}}
-  \textsf{:~~\#dequeues from subblocks in left child, prefix sum of \tt{num\sub{deq-left}}}
-  \item \tt{\sl{int} num\sub{enq-right}, \tt{sum\sub{enq-right}}}
-  \textsf{: ~\#enqueues from subblocks in right child, prefix sum of \tt{num\sub{enq-right}}}
-  \item \tt{\sl{int} num\sub{deq-right}, \tt{sum\sub{deq-right}}}
-  \textsf{: ~\#dequeues from subblocks in right child, prefix sum of \tt{num\sub{deq-right}}}
-  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in the block}
-  \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: sum of \# enqueue, dequeue operations in blocks up to this one}
-  \item \tt{\sl{int} num, sum}
-  \textsf{: total \# operations in block, prefix sum of \tt{num}}
-  \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~index of the last subblock in the left and right child}
+%  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
+%  \textsf{: \# enqueue, dequeue operations in the block}
+%  \item \tt{\sl{int} num, sum}
+%  \textsf{: total \# operations in block, prefix sum of \tt{num}}
+
   \item \tt{\sl{int} group}
-  \textsf{: id of the group of blocks including this propagated together, more precisely the value read from the node \tt{n}'s counter when propagating this block to the node \tt{n}.}
+  \textsf{: the value read from the node \tt{n.counter} when appending this block to \tt{n}.}
 \end{itemize}
 
-\pagebreak
-
-\Statex $\blacktriangleright$ \tt{\sl{Leaf Block} extends \sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
 \begin{itemize}
@@
   \textsf{Each block in a leaf represents an operation. The \tt{element} shows the operation's argument if it is an enqueue, and if it is a dequeue this value is \tt{null}.}
+
+    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+  \textsf{: sum of \# enqueue, dequeue operations in the leaf's blocks up to this one}
+
+    \item \tt{\sl{Object} response}
+
+  \textsf{response stores the response of the operation in the \tt{LeafBlock}.}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{Root Block} extends \sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
+\begin{itemize}
+  \item \tt{\sl{int} sum\sub{enq-left}}
+  \textsf{:~~\#enqueues from the subblocks in the left child + \tt{b[-1].sum\sub{enq-left}}}
+  \item \tt{\sl{int} sum\sub{deq-left}}
+  \textsf{:~~\#dequeues from the subblocks in the left child + \tt{b[-1].sum\sub{deq-left}}}
+  \item \tt{\sl{int} sum\sub{enq-right}}
+  \textsf{:~~\#enqueues from the subblocks in the right child + \tt{b[-1].sum\sub{enq-right}}}
+  \item \tt{\sl{int} sum\sub{enq-right}}
+  \textsf{:~~\#dequeues from the subblocks in the right child + \tt{b[-1].sum\sub{deq-right}}}
+    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
+  \textsf{:~~index of the last subblock in the left and right child}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
 \begin{itemize}
@@
   \textsf{: count of non-null dequeus up to this block}
-  \item \tt{\sl{int} num\sub{finished}}
+  \item \tt{\sl{counter} num\sub{finished}}
   \textsf{: number of finished operations in the block}
@@
 
+\Statex $\blacktriangleright$ \tt{Conventions}
+\begin{itemize}
+  \item \tt{i} : index of ith operation in the tree
+  \item \tt{j} : index of jth operation in a node
+  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
+  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
+\end{itemize}
 \end{multicols}
@@
 \begin{algorithm}
+\caption{\tt{\sl{Queue}}}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{33}
+\setcounter{ALG@line}{200}
 \begin{multicols}{2}
@@
 \Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
-\State \tt{block b= \Call{new}{\sl{leaf block}}}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
 \State \tt{b.element= e}
-\State \tt{b.num\sub{enq}=1}
-\State \tt{b.sum\sub{enq}= this.leaf.blocks[this.leaf.head].sum\sub{enq}+1}
-\State \Call{Append}{b}
+%\State \tt{b.num\sub{enq}=1}
+\State \tt{newBlock.sum\sub{enq}= tree[leaf].blocks[tree[leaf].head].sum\sub{enq}+1}
+\State \tt{newBlock.sum\sub{deq}= tree[leaf].blocks[tree[leaf].head].sum\sub{deq}}
+\State \tt{tree[leaf].}\Call{Append}{newBlock}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{}
-\State \tt{block b= \Call{new}{\sl{leaf block}}} \cmt{Creates a null element block, appends it to the tree, computes its order among operations, then computes its response index if it exists and returns the response's element.}
-\State \tt{b.element= null}
-\State \tt{b.num\sub{deq}=1}
-\State \tt{b.sum\sub{deq}= this.leaf.blocks[this.leaf.head].sum\sub{deq}+1} \cmt{\tt{this} is the current running process}
-\State \Call{Append}{b}
-\State \tt{<i, b\sub{i}>=} \Call{Index}{this.leaf, this.leaf.head, 1} \cmt{\tt{i} is the order in the root among all dequeues, of the dequeue in the last block in the process's leaf.
-\tt{b\sub{i}} is the index of the block in the root containing it.
-Since only one invocation of \tt{dequeue} is running by \tt{this} process at one time we are allowed to used \tt{this.leaf.head} safely.}
-\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
-\If{\tt{index\sub{response}!=-1}}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a null element block, appends it to the tree, computes its order among operations, then computes its response index if it exists and returns the response's element.}
+\State \tt{newBlock.element= null}
+\State \tt{newBlock.sum\sub{enq}= tree[leaf].blocks[tree[leaf].head].sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq}= tree[leaf].blocks[tree[leaf].head].sum\sub{deq}+1}
+\State \tt{tree[leaf].}\Call{Append}{newBlock}
+\State \tt{<i, b\sub{i}>=} \tt{tree[leaf].}\Call{Index}{tree[leaf].head, 1} \cmt{\tt{i} is the rank among all dequeues of the dequeue in the root and \tt{b\sub{i}} is the index of the block in the root containing it.}
+\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b\sub{i}} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
+\If{\tt{index\sub{response}==-1}}
 \State \tt{output= null}
-\State \tt{b\sub{deq}=root.blocks[b\sub{i}]}
-\State \tt{b\sub{deq}.num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}}\cmt{all the operations in the block containing the dequeue are finished.}
-\State \tt{this.leaf.last\sub{done}= b\sub{i}}
+\State \tt{tree[0].blocks[b\sub{i}].num\sub{finished}.inc()} \cmt{shared counter}
+\If{\tt{tree[0].blocks[b\sub{i}].num\sub{finished}==tree[0].blocks[b\sub{i}].num}}\cmt{all the operations in the block containing the dequeue are finished.}
+\State \tt{tree[leaf].last\sub{done}= b\sub{i}}
 \EndIf
@@
 \State \tt{output= \Call{Get}{index\sub{response}}}
-\State \tt{b\sub{r}= root.blocks.get(enq, index\sub{response}).order}\cmt{index of the block in the root contains response enqueue.}
-\State \tt{b\sub{enq}=root.blocks[b\sub{r}]}
-\State \tt{b\sub{enq}.num\sub{finished}.inc()}
-\State \tt{b\sub{deq}.num\sub{finished}.inc()}
-\If{\tt{b\sub{enq}.num\sub{finished}==b\sub{enq}.num}}\cmt{become done}
-\State \tt{this.leaf.last\sub{done}= b\sub{r}}
-\ElsIf {\tt{b\sub{deq}.num\sub{finished}==b\sub{deq}.num}} \cmt{\tt{b\sub{deq}} comes after \tt{b\sub{enq}}.}
-\State \tt{this.leaf.last\sub{done}= b\sub{i}} \cmt{\tt{this.leaf.last\sub{done}} is an increasing value.}
+\State \tt{b\sub{r}= tree[0].blocks.get(enq, index\sub{response}).order}\cmt{index of the block in the root that contains response enqueue.}
+\State \tt{root.blocks[b\sub{r}].num\sub{finished}.inc()}
+\State \tt{root.blocks[b\sub{r}].num\sub{finished}.inc()}
+\If{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}\cmt{become done}
+\State \tt{tree[leaf].last\sub{done}= b\sub{r}}
+\ElsIf {\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}} \cmt{\tt{root.blocks[b\sub{r}]} comes after \tt{root.blocks[b\sub{i}]}.}
+\State \tt{this.leaf.last\sub{done}= b\sub{i}}
 \EndIf
@@
 \EndFunction{Dequeue}
-
-\Statex
-
-\Function{void}{Append}{\sl{block} b} 
-\State \tt{b.group= this.leaf.head} \label{appendStart} \cmt{Only this block is propagated from the leaf by itself.}
-\State \tt{this.leaf.blocks[this.leaf.head]= b} 
-\State \tt{this.leaf.head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
-\State \Call{Propagate}{this.leaf.parent} 
-\EndFunction{Append}
-
-\Statex
-
-\Function{void}{Propagate}{\sl{node} n}
-\If{\tt{not} \Call{Refresh}{n}}
-\State \Call{Refresh}{n} \cmt{Lemma Double Refresh}
-\EndIf
-\If{\tt{n} \bf{is not} \tt{root}} \cmt{To check anode is the root we can check its index if the tree is implemented by an array or check if \tt{n.parent} is not \tt{null}.} 
-\State \Call{Propagate}{n.parent}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Function{element}{Get}{\sl{int} i}
-\cmt{Returns $i$th Enqueue.}
-\State \tt{res= root.blocks.get(enq, i).order}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
-\EndFunction{Get}
 
@@
 \begin{algorithm}
+\caption{\tt{\sl{Node}}}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{33}
+\setcounter{ALG@line}{300}
 \begin{multicols}{2}
 
-\Function{boolean}{Refresh}{\sl{node} n}
-\State \tt{h= n.head}
+\Function{void}{Propagate}{}
+\If{\tt{not} this.\Call{Refresh}{}}
+\State this.\Call{Refresh}{this} \cmt{Lemma Double Refresh}
+\EndIf
+\If{\tt{this} \bf{is not} \tt{root}} \cmt{To check a node is the root we can check its index is 0.} 
+\State \tt{this.parent().}\Call{Propagate()}{}
+\EndIf
+\EndFunction{Propagate}
+
+\Statex
+
+\Function{boolean}{Refresh}{}
+\State \tt{h= head}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{c= n.counter}
+\State \tt{c= counter}
 \State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}} \cmt{\tt{c\sub{left}, c\sub{right}} are the values read from \tt{n}'s children's counters.}
@@
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{(\tt{n} \bf{is} \tt{root} \bf{and} \tt{root.blocks.append(new)}) \bf{or} \\ (\tt{n} \bf{is not} \tt{root} \bf{and} \Call{CAS}{n.blocks[h], null, new})} \cmt{\it{how to put space in he first of the new line?}}
+\ElsIf{\tt{root.blocks.tryAppend(new)}} \cmt{\it{how to put space in he first of the new line?}}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{n.dir.super[c\sub{dir}], null, h+1}} \cmt{Superblock's Lemma}
-\State \tt{\Call{CAS}{n.dir.counter, c\sub{dir}, c\sub{dir}+1}}
+\State \tt{\Call{CAS}{dir.super[c\sub{dir}], null, h+1}} \cmt{Superblock's Lemma}
+\State \tt{\Call{CAS}{dir.counter, c\sub{dir}, c\sub{dir}+1}}
 \EndFor
-\State \tt{\Call{CAS}{n.head, h, h+1}}
+\State \tt{\Call{CAS}{head, h, h+1}}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{n.head, h, h+1}} \cmt{Even if another process wins, help to increase the head. It might fell sleep before increasing.}
+\State \tt{\Call{CAS}{head, h, h+1}} \cmt{Even if another process wins, help to increase the head. It might fell sleep before increasing.}
 \State \Return{ \tt{false}}
@@
 
+%##########################################
+
 \begin{algorithm}
+\caption{Node}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{83}
+\setcounter{ALG@line}{400}
 
@@
 \begin{algorithm}
+\caption{Leaf}
 \begin{algorithmic}[1]
+\setcounter{ALG@line}{500}
+
+\Function{void}{Append}{\sl{block} b} 
+\State \tt{head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
+\State \tt{b.group= head} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
+\State \tt{blocks[head]= b} 
+\State \tt{this.parent().}\Call{Propagate()}{} 
+\EndFunction{Append}
+
+\end{algorithmic}
+\end{algorithm}
+
+%##########################################
+
+\begin{algorithm}
+\caption{Root}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{600}
+
+\Function{element}{Get}{\sl{int} i}
+\cmt{Returns $i$th Enqueue.}
+\State \tt{res= root.blocks.get(enq, i).order}
+\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
+\EndFunction{Get}
+
+\end{algorithmic}
+\end{algorithm}
+
+%##########################################
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{700}
 \begin{multicols}{2}

### 2022-04-19 article/psuedocode.tex
apply meeting 4-2 notes

@@
 \begin{itemize}
-\item \tt{\sl{int} leaf\textsf{: \tt{Tree[leaf]} is the process's leaf in the tree.}}
+\item \tt{\sl{int} leaf\textsf{: \tt{tree[leaf]} is the process's leaf in the tree.}}
 \end{itemize}
@@
 
-\item \tt{\sl{int} counter= 0}\textsf{} \textsf{\# groups have been propagated from the node. It may be behind its real value.}
+\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks have been propagated from the node. It may be behind its real value.}
 \item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} that its \tt{group} field is \tt{i}.}
@@
 
-\Statex $\blacktriangleright$ \tt{\sl{Block b}} \Comment{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
+\Statex $\blacktriangleright$ \tt{\sl{Block b}} \cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
 
@@
   \item \tt{\sl{int} group}
-  \textsf{: the value read from the node \tt{n.counter} when appending this block to \tt{n}.}
+  \textsf{: the value read from the node \tt{n.num\sub{propagated}} when appending this block to \tt{n}.}
 \end{itemize}
@@
   \item \tt{\sl{Object} element}
-  \textsf{Each block in a leaf represents an operation. The \tt{element} shows the operation's argument if it is an enqueue, and if it is a dequeue this value is \tt{null}.}
+  \textsf{: Each block in a leaf represents an operation. The \tt{element} shows the operation's argument if it is an enqueue, and if it is a dequeue this value is \tt{null}.}
 
     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: sum of \# enqueue, dequeue operations in the leaf's blocks up to this one}
+  \textsf{: number of enqueue, dequeue operations in the leaf's containing this block till this block}
 
@@
     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~index of the last subblock in the left and right child}
+  \textsf{:~~index of the last subblock of the block in the left and right child}
 \end{itemize}
@@
   \item \tt{\sl{int} size}
-  \textsf{: size of queue after this block's operations finish}
-  \item \tt{\sl{int} sum\sub{non-null deq}}
-  \textsf{: count of non-null dequeus up to this block}
+  \textsf{: size of the queue after this block's operations finish}
+%  \item \tt{\sl{int} sum\sub{non-null deq}}
+%  \textsf{: count of non-null dequeus up to this block}
   \item \tt{\sl{counter} num\sub{finished}}
@@
     \item \tt{\sl{int} order}
-  \textsf{: the index of the block in the node containing it. Useful in the root since in the PBRT we do not keep indices in another way.}
+  \textsf{: the index of the block in the node containing the block. Useful in the root since in the PBRT we do not keep indices as key.}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{Conventions}
-\begin{itemize}
-  \item \tt{i} : index of ith operation in the tree
-  \item \tt{j} : index of jth operation in a node
-  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
-  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
-\end{itemize}
+%\Statex $\blacktriangleright$ \tt{Conventions}
+%\begin{itemize}
+%  \item \tt{i} : index of ith operation in the tree
+%  \item \tt{j} : index of jth operation in a node
+%  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
+%  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
+%\end{itemize}
+
 \end{multicols}
@@
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{b.element= e}
+\State \tt{newBlock.element= e}
 %\State \tt{b.num\sub{enq}=1}
@@
 \Function{Object}{Dequeue()}{}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a null element block, appends it to the tree, computes its order among operations, then computes its response index if it exists and returns the response's element.}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes its response; if it exists returns the response's element.}
 \State \tt{newBlock.element= null}
@@
 \State \tt{tree[leaf].}\Call{Append}{newBlock}
-\State \tt{<i, b\sub{i}>=} \tt{tree[leaf].}\Call{Index}{tree[leaf].head, 1} \cmt{\tt{i} is the rank among all dequeues of the dequeue in the root and \tt{b\sub{i}} is the index of the block in the root containing it.}
-\State \tt{index\sub{response}=} \Call{ComputeDeqRes}{i, b\sub{i}} \cmt{\tt{index\sub{response}} is the index of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
-\If{\tt{index\sub{response}==-1}}
+\State \tt{<r\sub{deq}, b\sub{r\sub{deq}}>=} \tt{tree[leaf].}\Call{Index}{tree[leaf].head, 1} \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{r\sub{deq}}} is the index of the block in the root containing the dequeue.}
+\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{r\sub{deq}}} \cmt{\tt{r\sub{deq}} is the rank of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
+\If{\tt{r\sub{enq}==-1}}
 \State \tt{output= null}
-\State \tt{tree[0].blocks[b\sub{i}].num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{tree[0].blocks[b\sub{i}].num\sub{finished}==tree[0].blocks[b\sub{i}].num}}\cmt{all the operations in the block containing the dequeue are finished.}
-\State \tt{tree[leaf].last\sub{done}= b\sub{i}}
+\State \tt{tree[0].blocks[b\sub{r\sub{deq}}].num\sub{finished}.inc()} \cmt{shared counter}
+\If{\tt{tree[0].blocks[b\sub{r\sub{deq}}].num\sub{finished}==tree[0].blocks[b\sub{r\sub{deq}}].num}}\cmt{all the operations in the block containing the dequeue are finished.}
+\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{deq}}}
 \EndIf
+\pagebreak
 \Else
 \State \tt{output= \Call{Get}{index\sub{response}}}
-\State \tt{b\sub{r}= tree[0].blocks.get(enq, index\sub{response}).order}\cmt{index of the block in the root that contains response enqueue.}
-\State \tt{root.blocks[b\sub{r}].num\sub{finished}.inc()}
-\State \tt{root.blocks[b\sub{r}].num\sub{finished}.inc()}
-\If{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}\cmt{become done}
-\State \tt{tree[leaf].last\sub{done}= b\sub{r}}
-\ElsIf {\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}} \cmt{\tt{root.blocks[b\sub{r}]} comes after \tt{root.blocks[b\sub{i}]}.}
-\State \tt{this.leaf.last\sub{done}= b\sub{i}}
+\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
+\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
+\If{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}
+\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{enq}}}
+\ElsIf {\tt{root.blocks[b\sub{r\sub{deq}}].num\sub{finished}==root.blocks[b\sub{r\sub{deq}}].num}} \cmt{\tt{root.blocks[b\sub{r}]} comes after \tt{root.blocks[b\sub{i}]}.}
+\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{deq}}}
 \EndIf
@@
 
-\Function{int}{ComputeDeqRes}{int i, int b}\cmt{Computes the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns the index of the the head of the queue or -1 if queue is empty.}
+\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
+Returns <-1,0> if the queue is empty.}
 \If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
-\State \Return \tt{-1}
-\Else{ \Return \tt{root.blocks[b-1].sum\sub{non-null deq} + i}}
+\State \Return \tt{-1, 0}
+\Else
+\State \tt{res= root.blocks[b-1].sum\sub{deq}- root.blocks[b-1].size + i}
+\State \Return \tt{<res, tree[0].blocks.get(enq, res).order>}
 \EndIf
-\EndFunction{ComputeDeqRes}
+\EndFunction{FindResponse}
 
@@
 
-\Function{void}{Propagate}{}
-\If{\tt{not} this.\Call{Refresh}{}}
-\State this.\Call{Refresh}{this} \cmt{Lemma Double Refresh}
+\Function{void}{Propagate()}{}
+\If{\bf{not} \tt{this.}\Call{Refresh()}{}}
+\State \tt{this.}\Call{Refresh()}{} \cmt{Lemma Double Refresh}
 \EndIf
@@
 
-\Function{boolean}{Refresh}{}
+\Function{boolean}{Refresh()}{}
 \State \tt{h= head}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{c= counter}
-\State \tt{<new, c\sub{left}, c\sub{right}>= \Call{CreateBlock}{n, h}} \cmt{\tt{c\sub{left}, c\sub{right}} are the values read from \tt{n}'s children's counters.}
-\State \tt{new.group= c}
+\State \tt{<new, n\sub{left}, n\sub{right}>= this.\Call{CreateBlock}{h}} \cmt{\tt{n\sub{left}, n\sub{right}} are the values read from the children's \tt{num\sub{propagated}}s.}
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{\tt{root.blocks.tryAppend(new)}} \cmt{\it{how to put space in he first of the new line?}}
+\ElsIf{\tt{root.blocks.tryAppend(new)}}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[c\sub{dir}], null, h+1}} \cmt{Superblock's Lemma}
-\State \tt{\Call{CAS}{dir.counter, c\sub{dir}, c\sub{dir}+1}}
+\State \tt{\Call{CAS}{dir.super[n\sub{dir}], null, h+1}}
+\State \tt{\Call{CAS}{dir.num\sub{propagated}, n\sub{dir}, n\sub{dir}+1}}
 \EndFor
@@
 
-\Statex $\leadsto$ \textsf{Precondition: \tt{n.blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
-\Function{int}{BSearch}{\sl{node} n, \sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
+\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
 
 \Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{feild}.
-Returns the index of the leftmost block in \tt{n.blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
+Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
 %\State \Return \tt{result block's index}
@@
 
-\Function{<Block, int, int>}{CreateBlock}{\sl{node} n, \sl{int} i} 
-\Statex\cmt{Creates a block to insert into \tt{n.blocks[i]}.
-Returns the created block as well as values read from each child counter field.
-The values are used for incrementing children's counters if the block was appended to \tt{n.blocks} successfully.
-\it{Does it need help?
-I think no but in that case we do not have to pass these values to the calling line.}}
-\State \tt{block b= \Call{new}{\sl{block}}}
-\If{\tt{n} \bf{is} \tt{root}}
-\State \tt{b= \Call{new}{\sl{root block}}}
-\EndIf
-\State \tt{b.order= i}
+\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
+\Statex\cmt{Creates a block to be inserted into \tt{this.blocks[i]}.
+Returns the created block as well as values read from each childnum\sub{propagated} field.
+The values are used for incrementing children's num\sub{propagated}s if the block was appended to \tt{this.blocks} successfully.}
+\State \tt{block newBlock= \Call{new}{\sl{block}}}
+\State \tt{newBlock.group= num\sub{propagated}}
+\State \tt{newBlock.order= i}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \State \tt{block\sub{last}= n.dir.blocks[index\sub{last}]}
-\State \tt{block\sub{prev}= n.dir.blocks[index\sub{prev}]} \cmt{\tt{n.dir.blocks[index\sub{prev}..index\sub{last}]} are merged to one block.}
-\State \tt{c\sub{dir}= n.dir.counter}
-\State \tt{b.end\sub{dir}= index\sub{last}}
-\State \tt{b.num\sub{enq-dir}= block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
-\State \tt{b.num\sub{deq-dir}= block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
-\State \tt{b.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
-\State \tt{b.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
+\State \tt{block\sub{prev}= n.dir.blocks[index\sub{prev}]}
+\State \cmt{\tt{newBlock} includes \tt{n.dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
+\State \tt{n\sub{dir}= n.dir.num\sub{propagated}}
+\State \tt{newBlock.end\sub{dir}= index\sub{last}}
+\State \tt{newBlock.num\sub{enq-dir}= block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
+\State \tt{newBlock.num\sub{deq-dir}= block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
+\State \tt{newBlock.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
+\State \tt{newBlock.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
 \EndFor
-\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
-\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
-\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
-\State \tt{b.sum= n.blocks[i-1].sum + b.num}
+%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
+%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
+%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
+%\State \tt{b.sum= n.blocks[i-1].sum + b.num}
 
 \If{\tt{n} \bf{is} \tt{root}}
-\State \tt{b.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
-\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
+\State \tt{newBlock.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
+%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
 \EndIf
@@
 
-\Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains $\geq$\tt{i} enqueues.}
-\Function{element}{Get}{\sl{node} n, \sl{int} b, \sl{int} i} 
-\cmt{\textmd{Returns the \tt{i}th Enqueue in \tt{b}th block of node \tt{n}}}
-\If{\tt{n {\keywordfont is} leaf}} \Return \tt{n.blocks[b].element}
+\Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains \tt{i}th enqueue in the node.}
+\Function{element}{Get}{\sl{int} b, \sl{int} i} 
+\If{\tt{this {\keywordfont is} leaf}} \Return \tt{this.blocks[b].element}
 \Else
-\If{\tt{i $\leq$ n.blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of \tt{n}}
-\State \tt{subBlock= \Call{BSearch}{n.left, sum\sub{enq}, i, n.blocks[b-1].end\sub{left}+1, n.blocks[b].end\sub{left}}}
-\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\sub{enq}} 
+\If{\tt{i $\leq$ this.blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
+\State \tt{subBlock= this.left\Call{BSearch}{sum\sub{enq}, i, this.blocks[b-1].end\sub{left}+1, this.blocks[b].end\sub{left}}}
+\State \Return\tt{this.left.}\Call{Get}{i-this.left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
 \Else
-\State \tt{i= i-n.blocks[b].num\sub{enq-left}}
-\State\tt{subBlock=\Call{BSearch}{n.right, sum\sub{enq}, i, n.blocks[b-1].end\sub{right}+1, n.blocks[b].end\sub{right}}}
-\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\sub{enq}} 
+\State \tt{i= i-this.blocks[b].num\sub{enq-left}}
+\State\tt{subBlock= this.right\Call{BSearch}{sum\sub{enq}, i, this.blocks[b-1].end\sub{right}+1, this.blocks[b].end\sub{right}}}
+\State \Return\tt{this.right.}\Call{Get}{i-this.right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
 \EndIf
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of node \tt{n} has propagated up to the root and \tt{i}th dequeue resides in node \tt{n} is in block \tt{b} of node \tt{n}.}
-\Function{<int, int>}{Index}{\sl{node} n, \sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
-\If{\tt{n {\keywordfont is} root}} \Return \tt{i, b}
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{this.blocks[b]}.}
+\Function{<int, int>}{Index}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
+\If{\tt{this {\keywordfont is} root}} \Return \tt{<i, b>}
 \Else
-\State \tt{dir= (n.parent.left==n)? left: right} \cmt{check \tt{n} is a left or a right child}
-\State \tt{superBlock= \Call{BSearch}{n.parent, n.sum\sub{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}} \cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
+\State \tt{dir= (this.parent().left==n)? left: right} \cmt{check \tt{n} is a left or a right child}
+\State \tt{superBlock= \Call{BSearch}{this.parent(), this.sum\sub{deq-dir}, i, super[this.blocks[b].group]-p, super[this.blocks[b].group]+p}} \cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
 \If{\tt{dir {\keywordfont is} left}}
-\State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq-right}}
-\Else \State \tt{i+= n.parent.blocks[superBlock-1].sum\sub{deq} + n.blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from \tt{n}'s right child}
+\State \tt{i+= this.parent().blocks[superBlock-1].sum\sub{deq-right}}
+\Else \State \tt{i+= this.parent.blocks[superBlock-1].sum\sub{deq} + this.blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from the right child}
 \EndIf
-\State \Return\Call{Index}{n.parent, superBlock, i}
+\State \Return\Call{Index}{this.parent(), superBlock, i}
 \EndIf

### 2022-04-21 article/psuedocode.tex
apply april 20 feedbacks

@@
 \begin{itemize}
-\item \tt{\sl{Node*} tree[]} \textsf{: A binary tree of \tt{Node}s such that \tt{Tree[0]} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.}
+\item \tt{\sl{Node*} tree[]} \textsf{: A binary tree of \tt{Node}s such that \tt{Tree[0]} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.} \label{treeRules}
 \end{itemize}
@@
 \begin{itemize}
-%\item \tt{\sl{*Node} left(), right(), parent()} \textsf{: computed as mentioned in the \tt{Tree[]}}
+\item \tt{\sl{*Node} left(), right(), parent()} \textsf{: initialized  when creating the tree as mentioned in the \tt{tree[]} description \ref{treeRules}.}
 \item \tt{\sl{BlockList} blocks}\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
@@
 
+\begin{footnotesize}
+
+\it{Conventions}
+\begin{itemize}
+  \item \tt{b\sub{x}}: the block containing \tt{x}
+  \item \tt{r\sub{x}}: rank of \tt{x} in the current scope
+\end{itemize}
+
+\it{Unwritten rules}
+\begin{itemize}
+ \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b>0 and x $\in$ \{enq, deq\}})
+ \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b>0})
+  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}})
+\end{itemize}
+\end{footnotesize}
+
 %##########################################
@@
 \State \tt{tree[leaf].}\Call{Append}{newBlock}
-\State \tt{<r\sub{deq}, b\sub{r\sub{deq}}>=} \tt{tree[leaf].}\Call{Index}{tree[leaf].head, 1} \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{r\sub{deq}}} is the index of the block in the root containing the dequeue.}
-\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{r\sub{deq}}} \cmt{\tt{r\sub{deq}} is the rank of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
+\State \tt{<r\sub{deq}, b\sub{r\sub{deq}}>=} \tt{tree[leaf].}\Call{Index\sub{Deq}}{tree[leaf].head, 1}
+\Statex \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{r\sub{deq}}} is the index of the block in the root containing the dequeue.}
+\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{r\sub{deq}}} 
+\Statex \cmt{\tt{r\sub{enq}} is the rank of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
 \If{\tt{r\sub{enq}==-1}}
@@
 \Else
-\State \tt{output= \Call{Get}{index\sub{response}}}
+\State \tt{output= \Call{Get\sub{enq}}{r\sub{enq}, b\sub{r\sub{enq}}}}
 \State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
@@
 \Else
-\State \tt{res= root.blocks[b-1].sum\sub{deq}- root.blocks[b-1].size + i}
-\State \Return \tt{<res, tree[0].blocks.get(enq, res).order>}
+\State \tt{response= root.blocks[b-1].sum\sub{deq}- root.blocks[b-1].size + i}
+\State \Return \tt{<res, tree[0].blocks.get(enq, response).order>}
 \EndIf
@@
 \If{\tt{this} \bf{is not} \tt{root}} \cmt{To check a node is the root we can check its index is 0.} 
-\State \tt{this.parent().}\Call{Propagate()}{}
+\State \tt{this.parent.}\Call{Propagate()}{}
 \EndIf
@@
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, n\sub{left}, n\sub{right}>= this.\Call{CreateBlock}{h}} \cmt{\tt{n\sub{left}, n\sub{right}} are the values read from the children's \tt{num\sub{propagated}}s.}
+\State \tt{<new, np\sub{left}, np\sub{right}>= this.\Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}}s.}
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[n\sub{dir}], null, h+1}}
-\State \tt{\Call{CAS}{dir.num\sub{propagated}, n\sub{dir}, n\sub{dir}+1}}
+\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}}
+\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
 \EndFor
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= n.dir.head} \label{lastLine}
-\State \tt{index\sub{prev}= n.blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{block\sub{last}= n.dir.blocks[index\sub{last}]}
-\State \tt{block\sub{prev}= n.dir.blocks[index\sub{prev}]}
-\State \cmt{\tt{newBlock} includes \tt{n.dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{n\sub{dir}= n.dir.num\sub{propagated}}
+\State \tt{index\sub{last}= dir.head} \label{lastLine}
+\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
+\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
+\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
+\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
+\State \tt{n\sub{dir}= dir.num\sub{propagated}}
 \State \tt{newBlock.end\sub{dir}= index\sub{last}}
-\State \tt{newBlock.num\sub{enq-dir}= block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
-\State \tt{newBlock.num\sub{deq-dir}= block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
-\State \tt{newBlock.sum\sub{enq-dir}= n.blocks[i-1].sum\sub{enq-dir} + b.num\sub{enq-dir}}
-\State \tt{newBlock.sum\sub{deq-dir}= n.blocks[i-1].sum\sub{deq-dir} + b.num\sub{deq-dir}}
+\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
 \EndFor
@@
 
-\If{\tt{n} \bf{is} \tt{root}}
+\If{\tt{this} \bf{is} \tt{root}}
 \State \tt{newBlock.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
@@
 
-\State \Return \tt{b, c\sub{left}, c\sub{right}}
+\State \Return \tt{b, np\sub{left}, np\sub{right}}
 \EndFunction{CreateBlock}
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains \tt{i}th enqueue in the node.}
-\Function{element}{Get}{\sl{int} b, \sl{int} i} 
-\If{\tt{this {\keywordfont is} leaf}} \Return \tt{this.blocks[b].element}
+\Function{element}{Get\sub{Enq}}{\sl{int} b, \sl{int} i} 
+\If{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
+\State \tt{subBlock= left\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}}
+\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
 \Else
-\If{\tt{i $\leq$ this.blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
-\State \tt{subBlock= this.left\Call{BSearch}{sum\sub{enq}, i, this.blocks[b-1].end\sub{left}+1, this.blocks[b].end\sub{left}}}
-\State \Return\tt{this.left.}\Call{Get}{i-this.left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
-\Else
-\State \tt{i= i-this.blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= this.right\Call{BSearch}{sum\sub{enq}, i, this.blocks[b-1].end\sub{right}+1, this.blocks[b].end\sub{right}}}
-\State \Return\tt{this.right.}\Call{Get}{i-this.right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
+\State \tt{i= i-blocks[b].num\sub{enq-left}}
+\State\tt{subBlock= right\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}}
+\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
 \EndIf
-\EndIf
-\EndFunction{Get}
+\EndFunction{Get\sub{Enq}}
 
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{this.blocks[b]}.}
-\Function{<int, int>}{Index}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
-\If{\tt{this {\keywordfont is} root}} \Return \tt{<i, b>}
-\Else
-\State \tt{dir= (this.parent().left==n)? left: right} \cmt{check \tt{n} is a left or a right child}
-\State \tt{superBlock= \Call{BSearch}{this.parent(), this.sum\sub{deq-dir}, i, super[this.blocks[b].group]-p, super[this.blocks[b].group]+p}} \cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{blocks[b]}.}
+\Function{<int, int>}{Index\sub{Deq}}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
+\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
+\State \tt{superBlock= \Call{BSearch}{parent, sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
+\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
 \If{\tt{dir {\keywordfont is} left}}
-\State \tt{i+= this.parent().blocks[superBlock-1].sum\sub{deq-right}}
-\Else \State \tt{i+= this.parent.blocks[superBlock-1].sum\sub{deq} + this.blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from the right child}
+\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
+\Else \State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq} + blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from the right child}
 \EndIf
-\State \Return\Call{Index}{this.parent(), superBlock, i}
-\EndIf
+\State \Return\Call{Index\sub{Deq}}{this.parent, superBlock, i}
+
 \EndFunction{Index}
@@
 
-\Function{void}{Append}{\sl{block} b} 
+\Function{void}{Append}{\sl{block} blk} 
 \State \tt{head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
-\State \tt{b.group= head} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
-\State \tt{blocks[head]= b} 
-\State \tt{this.parent().}\Call{Propagate()}{} 
+\State \tt{blk.group= head} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
+\State \tt{blocks[head]= blk} 
+\State \tt{parent.}\Call{Propagate()}{} 
 \EndFunction{Append}
+
+\Statex
+
+\Function{element}{Get\sub{Enq}}{\sl{int} b, \sl{int} i} 
+\State\Return \tt{blocks[b].element}
+\EndFunction{Get\sub{Enq}}
 
@@
 
-\Function{element}{Get}{\sl{int} i}
-\cmt{Returns $i$th Enqueue.}
-\State \tt{res= root.blocks.get(enq, i).order}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\sub{enq}}}
-\EndFunction{Get}
+\Function{<int, int>}{Index\sub{Deq}}{\sl{int} b, \sl{int} i} 
+\State\Return \tt{<i, b>}
+\EndFunction{Index\sub{Deq}}
 

### 2022-04-22 article/psuedocode.tex
apply apr 22 morning notes

@@
 \begin{itemize}
-\item \tt{\sl{Node*} tree[]} \textsf{: A binary tree of \tt{Node}s such that \tt{Tree[0]} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.} \label{treeRules}
+\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is a pointer to the root node.}
+%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{int} leaf\textsf{: \tt{tree[leaf]} is the process's leaf in the tree.}}
+\item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{*Node} left(), right(), parent()} \textsf{: initialized  when creating the tree as mentioned in the \tt{tree[]} description \ref{treeRules}.}
-\item \tt{\sl{BlockList} blocks}\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b)} returns true \tt{b} is appended to the end of the list and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end} pointers to the first block of the corresponding children.}
-
-\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks have been propagated from the node. It may be behind its real value.}
-\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} that its \tt{group} field is \tt{i}.}
+\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
+\item \tt{\sl{BlockList} blocks}
+\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
+Since it is incremented after propagating, it may be behind by 1.}
+\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} whose \tt{group} field is \tt{i}.}
 \end{itemize}
@@
 
-  \textsf{Implemented with a persistent red-black tree.}
+  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
 
@@
 
-  \textsf{Implemented with an array and using CAS for appending to the head.}
+  \textsf{\tt{BlockList} is implemented with an array.}
 
-\item \tt{\sl{int} head= 1}\textsf{: \#\tt{block}s in the \tt{blocks}.}
+\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
 
@@
 
-  \textsf{Each process stores the index of the most recent block in the root that the process has finished the last operation of the block.
-\tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response.}
+  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
+A block is finished if all of its operations are finished.
+\tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response.
+\it{put the definitions before the pseudocode}}
 
@@
 
-\Statex $\blacktriangleright$ \tt{\sl{Block b}} \cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
+\Statex $\blacktriangleright$ \tt{\sl{Block}} 
+%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
+\cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
+\it{put the definitions before the pseudocode}}
 
@@
   \item \tt{\sl{int} group}
-  \textsf{: the value read from the node \tt{n.num\sub{propagated}} when appending this block to \tt{n}.}
+  \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
 \end{itemize}
@@
   \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents an operation. The \tt{element} shows the operation's argument if it is an enqueue, and if it is a dequeue this value is \tt{null}.}
+  \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}
+
+    \item \tt{\sl{Object} response}
+  \textsf{: stores the response of the operation in the \tt{LeafBlock}.}
 
     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: number of enqueue, dequeue operations in the leaf's containing this block till this block}
+  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
 
-    \item \tt{\sl{Object} response}
-
-  \textsf{response stores the response of the operation in the \tt{LeafBlock}.}
 \end{itemize}
@@
 \begin{itemize}
-  \item \tt{\sl{int} sum\sub{enq-left}}
-  \textsf{:~~\#enqueues from the subblocks in the left child + \tt{b[-1].sum\sub{enq-left}}}
-  \item \tt{\sl{int} sum\sub{deq-left}}
-  \textsf{:~~\#dequeues from the subblocks in the left child + \tt{b[-1].sum\sub{deq-left}}}
-  \item \tt{\sl{int} sum\sub{enq-right}}
-  \textsf{:~~\#enqueues from the subblocks in the right child + \tt{b[-1].sum\sub{enq-right}}}
-  \item \tt{\sl{int} sum\sub{enq-right}}
-  \textsf{:~~\#dequeues from the subblocks in the right child + \tt{b[-1].sum\sub{deq-right}}}
     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
   \textsf{:~~index of the last subblock of the block in the left and right child}
+  \item \tt{\sl{int} sum\sub{enq-left}}
+  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{deq-left}}
+  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{enq-right}}
+  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
+  \item \tt{\sl{int} sum\sub{deq-right}}
+  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
 \end{itemize}
@@
 \begin{itemize}
-  \item \tt{\sl{int} size}
-  \textsf{: size of the queue after this block's operations finish}
+  \item \tt{\sl{int} length}
+  \textsf{: length of the queue after performing all operations in the prefix for this block}
 %  \item \tt{\sl{int} sum\sub{non-null deq}}
@@
     \item \tt{\sl{int} order}
-  \textsf{: the index of the block in the node containing the block. Useful in the root since in the PBRT we do not keep indices as key.}
+  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
 \end{itemize}
@@
 
-\it{Conventions}
+\it{Variable naming:}
 \begin{itemize}
-  \item \tt{b\sub{x}}: the block containing \tt{x}
-  \item \tt{r\sub{x}}: rank of \tt{x} in the current scope
+  \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
+  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
 \end{itemize}
 
-\it{Unwritten rules}
+\it{Abbreviations:}
 \begin{itemize}
- \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b>0 and x $\in$ \{enq, deq\}})
- \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b>0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}})
+ \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
+ \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
+  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
 \end{itemize}
@@
 %\State \tt{b.num\sub{enq}=1}
-\State \tt{newBlock.sum\sub{enq}= tree[leaf].blocks[tree[leaf].head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= tree[leaf].blocks[tree[leaf].head].sum\sub{deq}}
-\State \tt{tree[leaf].}\Call{Append}{newBlock}
+\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
+\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
+\State \tt{leaf.}\Call{Append}{newBlock}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes its response; if it exists returns the response's element.}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
 \State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= tree[leaf].blocks[tree[leaf].head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= tree[leaf].blocks[tree[leaf].head].sum\sub{deq}+1}
-\State \tt{tree[leaf].}\Call{Append}{newBlock}
-\State \tt{<r\sub{deq}, b\sub{r\sub{deq}}>=} \tt{tree[leaf].}\Call{Index\sub{Deq}}{tree[leaf].head, 1}
-\Statex \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{r\sub{deq}}} is the index of the block in the root containing the dequeue.}
-\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{r\sub{deq}}} 
-\Statex \cmt{\tt{r\sub{enq}} is the rank of the enqueue which is the response to the dequeue or -1 if the response is \tt{null}.} \label{deqRest}
+\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
+\State \tt{leaf.}\Call{Append}{newBlock}
+\State \tt{<r\sub{deq}, b\sub{deq}>=} \tt{leaf.}\Call{IndexDeq}{leaf.size, 1}
+\Statex \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{deq}} is the index of the block in the root containing the dequeue.}
+\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
+\Statex \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue and \tt{b\sub{deq}} is the index of the block of it in the blocklist.
+If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
 \If{\tt{r\sub{enq}==-1}}
 \State \tt{output= null}
-\State \tt{tree[0].blocks[b\sub{r\sub{deq}}].num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{tree[0].blocks[b\sub{r\sub{deq}}].num\sub{finished}==tree[0].blocks[b\sub{r\sub{deq}}].num}}\cmt{all the operations in the block containing the dequeue are finished.}
-\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{deq}}}
+\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
+\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
+%\cmt{all the operations in the block containing the dequeue are finished.}
+\State \tt{leaf.last\sub{done}= b\sub{deq}}
 \EndIf
@@
 \Else
-\State \tt{output= \Call{Get\sub{enq}}{r\sub{enq}, b\sub{r\sub{enq}}}}
+\State \tt{output= \Call{GetEnq}{r\sub{enq}, b\sub{r\sub{enq}}}} \cmt{getting the reponse's \tt{element}.}
 \State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
 \State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
-\If{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}
-\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{enq}}}
-\ElsIf {\tt{root.blocks[b\sub{r\sub{deq}}].num\sub{finished}==root.blocks[b\sub{r\sub{deq}}].num}} \cmt{\tt{root.blocks[b\sub{r}]} comes after \tt{root.blocks[b\sub{i}]}.}
-\State \tt{tree[leaf].last\sub{done}= b\sub{r\sub{deq}}}
+\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
+\State \tt{leaf.last\sub{done}= b\sub{deq}}
+\ElsIf{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}
+\State \tt{leaf.last\sub{done}= b\sub{r\sub{enq}}}
 \EndIf
@@
 \Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
-Returns <-1,0> if the queue is empty.}
-\If{\tt{root.blocks[b-1].size + root.blocks[b].num\sub{enq} - i $<$ 0}}
-\State \Return \tt{-1, 0}
+Returns \tt{<-1,-->} if the queue is empty.}
+\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
+\State \Return \tt{<-1,-->}
 \Else
-\State \tt{response= root.blocks[b-1].sum\sub{deq}- root.blocks[b-1].size + i}
-\State \Return \tt{<res, tree[0].blocks.get(enq, response).order>}
+\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}.
+$r$th non-null dequeue returns the element of th $r$th enqueue.
+We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues.
+Note that the $i$th dequeue in the given block is not a non-null dequeue.}
+\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
+\State \Return \tt{<r\sub{enq}, root.blocks.get(enq, r\sub{enq}).order>}
 \EndIf
@@
 \Function{boolean}{Refresh()}{}
-\State \tt{h= head}
+\State \tt{h= size}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
@@
 \EndFor
-\State \tt{\Call{CAS}{head, h, h+1}}
+\State \tt{\Call{CAS}{size, h, h+1}}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{head, h, h+1}} \cmt{Even if another process wins, help to increase the head. It might fell sleep before increasing.}
+\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. It might fell sleep before increasing.}
 \State \Return{ \tt{false}}
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.head} \label{lastLine}
+\State \tt{index\sub{last}= dir.size} \label{lastLine}
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.size= max(root.blocks[i-1].size + b.num\sub{enq} - b.num\sub{deq}, 0)}
-%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].size - b.num\sub{enq}, 0)}
+\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
+%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
 \EndIf
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains \tt{i}th enqueue in the node.}
-\Function{element}{Get\sub{Enq}}{\sl{int} b, \sl{int} i} 
+\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
 \If{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
@@
 \EndIf
-\EndFunction{Get\sub{Enq}}
+\EndFunction{GetEnq}
 
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{blocks[b]}.}
-\Function{<int, int>}{Index\sub{Deq}}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
+\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
@@
 \EndIf
-\State \Return\Call{Index\sub{Deq}}{this.parent, superBlock, i}
+\State \Return\Call{IndexDeq}{this.parent, superBlock, i}
 
@@
 \Function{void}{Append}{\sl{block} blk} 
-\State \tt{head+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
-\State \tt{blk.group= head} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
-\State \tt{blocks[head]= blk} 
+\State \tt{size+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
+\State \tt{blk.group= size} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
+\State \tt{blocks[size]= blk} 
 \State \tt{parent.}\Call{Propagate()}{} 
@@
 
-\Function{element}{Get\sub{Enq}}{\sl{int} b, \sl{int} i} 
+\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
 \State\Return \tt{blocks[b].element}
-\EndFunction{Get\sub{Enq}}
+\EndFunction{GetEnq}
 
@@
 
-\Function{<int, int>}{Index\sub{Deq}}{\sl{int} b, \sl{int} i} 
+\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} 
 \State\Return \tt{<i, b>}
-\EndFunction{Index\sub{Deq}}
+\EndFunction{IndexDeq}
+
+\end{algorithmic}
+\end{algorithm}
+
+%##########################################
+
+\begin{algorithm}
+\caption{BlockList}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{700}
+
+\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+Initially  empty, when \tt{blocks.tryAppend(b)} returns true \tt{b} is appended to the end of the list and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
+
+\Statex
+\Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
+\Function{boolean}{TryAppend}{\sl{block} blk} 
+\State \tt{TODO}
+\EndFunction{TryAppend}
+
+\Statex
+\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
+\Function{boolean}{TryAppend}{\sl{block} blk} 
+\State \tt{TODO}
+\EndFunction{TryAppend}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
+\setcounter{ALG@line}{800}
 \begin{multicols}{2}
@@
 \Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
-\State \tt{step= root.head}
+\State \tt{step= root.size}
 \If{\tt{step\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root. Used in lemma's using the size of the PBRT.}
@@
 \For{\tt{leaf l} \bf{in leaves}}\cmt{\it{if the tree is implemented with an array we can iterate over the second half of the array.}}
-\State{\tt{last= l.head-1}}
+\State{\tt{last= l.size-1}}
 \cmt{\tt{l.blocks[last]} can not be \tt{null} because of lines \ref{appendStart}-\ref{appendEnd}.}

### 2022-04-26 article/psuedocode.tex
apply meeting 4-5 feedbacks

@@
 \item \tt{\sl{BlockList} blocks}
+\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
 \item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
 Since it is incremented after propagating, it may be behind by 1.}
-\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} whose \tt{group} field is \tt{i}.}
+\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
 \end{itemize}
@@
   \textsf{\tt{BlockList} is implemented with an array.}
-
-\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
 
@@
 \State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<r\sub{deq}, b\sub{deq}>=} \tt{leaf.}\Call{IndexDeq}{leaf.size, 1}
-\Statex \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{deq}} is the index of the block in the root containing the dequeue.}
-\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
-\Statex \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue and \tt{b\sub{deq}} is the index of the block of it in the blocklist.
-If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
-\If{\tt{r\sub{enq}==-1}}
-\State \tt{output= null}
-\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
-%\cmt{all the operations in the block containing the dequeue are finished.}
-\State \tt{leaf.last\sub{done}= b\sub{deq}}
-\EndIf
-\pagebreak
-\Else
-\State \tt{output= \Call{GetEnq}{r\sub{enq}, b\sub{r\sub{enq}}}} \cmt{getting the reponse's \tt{element}.}
-\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
-\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
-\State \tt{leaf.last\sub{done}= b\sub{deq}}
-\ElsIf{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}
-\State \tt{leaf.last\sub{done}= b\sub{r\sub{enq}}}
-\EndIf
-\EndIf
-\State \Return{\tt{output}}
+\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
 \EndFunction{Dequeue}
@@
 \State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
-\State \Return \tt{<r\sub{enq}, root.blocks.get(enq, r\sub{enq}).order>}
+\State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
 \EndIf
@@
 \Function{void}{Propagate()}{}
-\If{\bf{not} \tt{this.}\Call{Refresh()}{}}
-\State \tt{this.}\Call{Refresh()}{} \cmt{Lemma Double Refresh}
+\If{\bf{not} \Call{Refresh()}{}}
+\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
 \EndIf
-\If{\tt{this} \bf{is not} \tt{root}} \cmt{To check a node is the root we can check its index is 0.} 
-\State \tt{this.parent.}\Call{Propagate()}{}
+\If{\tt{this} \bf{is not} \tt{root}}
+\State \tt{parent.}\Call{Propagate()}{}
 \EndIf
@@
 \Function{boolean}{Refresh()}{}
-\State \tt{h= size}
+\State \tt{s= size}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, np\sub{left}, np\sub{right}>= this.\Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}}s.}
+\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{\tt{root.blocks.tryAppend(new)}}
+\ElsIf{\tt{blocks.tryAppend(new, s)}}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}}
+\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
 \State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
@@
 \Else
-\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. It might fell sleep before increasing.}
+\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
 \State \Return{ \tt{false}}
@@
 
-\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{feild}.
+\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
 Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
@@
 \Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\Statex\cmt{Creates a block to be inserted into \tt{this.blocks[i]}.
-Returns the created block as well as values read from each childnum\sub{propagated} field.
-The values are used for incrementing children's num\sub{propagated}s if the block was appended to \tt{this.blocks} successfully.}
+\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}.
+Returns the created block as well as values read from each child's num\sub{propagated} field.
+These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
 \State \tt{block newBlock= \Call{new}{\sl{block}}}
@@
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
+\State \tt{newBlock.end\sub{dir}= index\sub{last}}
 \State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
@@
 \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{n\sub{dir}= dir.num\sub{propagated}}
-\State \tt{newBlock.end\sub{dir}= index\sub{last}}
+\State \tt{this\sub{dir}= dir.num\sub{propagated}}
 \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
@@
 
-\State \Return \tt{b, np\sub{left}, np\sub{right}}
+\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
 \EndFunction{CreateBlock}
@@
 
-\Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains \tt{i}th enqueue in the node.}
+\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
 \Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
-\If{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
-\State \tt{subBlock= left\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}}
+\If{\tt{this} \bf{is} \tt{leaf}}
+\State\Return \tt{blocks[b].element}
+\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
+\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
 \State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}}
+\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
 \State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{blocks[b]}.}
-\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
+\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
+\If{\tt{this} \bf{is} \tt{root}}
+\State\Return \tt{<b, i>}
+\Else
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
-\State \tt{superBlock= \Call{BSearch}{parent, sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
+\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
-\If{\tt{dir {\keywordfont is} left}}
-\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
-\Else \State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq} + blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from the right child}
+\If{\tt{dir {\keywordfont is} right}}
+%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
+\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
 \EndIf
-\State \Return\Call{IndexDeq}{this.parent, superBlock, i}
-
+\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
+\EndIf
 \EndFunction{Index}
@@
 
-\Function{void}{Append}{\sl{block} blk} 
-\State \tt{size+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
-\State \tt{blk.group= size} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
+\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
+\State \tt{size+=1} \label{appendEnd} 
+\State \tt{blk.group= size} \label{appendStart} 
 \State \tt{blocks[size]= blk} 
@@
 
-\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
-\State\Return \tt{blocks[b].element}
-\EndFunction{GetEnq}
-
-\end{algorithmic}
-\end{algorithm}
-
-%##########################################
-
-\begin{algorithm}
-\caption{Root}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{600}
-
-\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} 
-\State\Return \tt{<i, b>}
-\EndFunction{IndexDeq}
+\Function{Object}{HelpDequeue()}{}
+\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
+\State \tt{<r\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
+ \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
+If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
+\If{\tt{r\sub{enq}==-1}}
+\State \tt{output= null}
+\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
+\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
+\State \tt{last\sub{done}= b\sub{deq}}
+\EndIf
+\pagebreak
+\Else
+\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
+\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
+\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
+\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
+\State \tt{last\sub{done}= b\sub{deq}}
+\ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
+\State \tt{last\sub{done}= b\sub{enq}}
+\EndIf
+\EndIf
+\State \Return{\tt{output}}
+\EndFunction{Dequeue}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
+\setcounter{ALG@line}{600}
 
 \Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b)} returns true \tt{b} is appended to the end of the list and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
+Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
 
@@
 \Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
-\Function{boolean}{TryAppend}{\sl{block} blk} 
-\State \tt{TODO}
+  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+\Statex \tt{root: }\sf{pointer to the root of the PBRT}
+\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
+\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
+\State \tt{Help()}
+\State \tt{CollectGarbage()}
+\EndIf
+\State \tt{blk.num\sub{finished}= 0}
+\State \tt{*oldRoot= \&root.blocks.root}
+\State \tt{*newRoot= root.blocks.Append(blk).root}
+\State \Return \tt{CAS(root, oldRoot, newRoot)}
 \EndFunction{TryAppend}
@@
 \Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Function{boolean}{TryAppend}{\sl{block} blk} 
-\State \tt{TODO}
+\Statex \tt{blocks[]: }\sf{array of blocks}
+\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
+\State \Return{\tt{CAS(blocks[n], null, blk)}}
 \EndFunction{TryAppend}
@@
 
-\Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
-  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-  \tt{append(b, key)} returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.
-
-\Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
-\State \tt{step= root.size}
-\If{\tt{step\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root. Used in lemma's using the size of the PBRT.}
-\State \tt{Help()}
-\State \tt{CollectGarbage()}
-\EndIf
-\State \tt{b.num\sub{finished}= 0}
-\State \Return \tt{root.blocks.append(b, b.order)}
-\EndFunction{RBTAppend}
-\Statex
-
 \Function{void}{Help}{}\cmt{Helps pending operations}
-\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{if the tree is implemented with an array we can iterate over the second half of the array.}}
+\For{\tt{leaf l} \bf{in leaves}}
 \State{\tt{last= l.size-1}}
-\cmt{\tt{l.blocks[last]} can not be \tt{null} because of lines \ref{appendStart}-\ref{appendEnd}.}
+\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
 \If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
-\State \tt{goto \ref{deqRest} with these values <>} \cmt{run \tt{Dequeue()}  for \tt{l.ops[last]} after Propagate(). \it{TODO}}
-\State \tt{l.responses[last]= response}
+\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
 \EndIf
@@
 \State\Return\tt{max} \cmt{This snapshot suffies.}
-\EndFunction{findYoungestOld}
+\EndFunction{FindMostRecentDone}
 
@@
 
-\Function{response}{FallBack}{op i} \cmt{\it{really necessary?}}
+\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
 

### 2022-04-28 article/psuedocode.tex
add old lemmas

@@
 \newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
+\renewcommand{\tt}[1]{\tt{#1}}
 \renewcommand{\sl}[1]{\textsl{#1}}
@@
 \renewcommand{\bf}[1]{\textbf{#1}}
+\renewcommand{\nf}[1]{\normalfont{#1}}
 \newcommand{\cmt}[1]{\Comment{#1}}
@@
 \begin{document}
+
+\section{Pseudocode}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+                     %
+                     %
+                     %
+                     %
+                     %
+                     %
+                     %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+\section{Proof of Linearizability}
+
+\begin{definition}
+If \tt{n.blocks[i]==b}  we call \tt{i} the \emph{index} of block \tt{b} in node \tt{n}.
+Block \tt{b} is before block \tt{b$^\prime$} in node \tt{n} if and only if \tt{b}'s index is smaller than \tt{b$^\prime$}'s.
+Block \tt{b} is propagated to node \tt{n} or set \tt{S} if \tt{b} is in \tt{n.blocks} or \tt{S} or is a subblock of a block in \tt{n.blocks} or \tt{S}.
+\end{definition}
+
+%\begin{definition}
+%  Block \tt{b} in node \tt{n} is new if \tt{b} is not a subblock of any block in \tt{n.parent.blocks[]}.
+%\end{definition}
+
+\begin{definition}
+  Block \tt{b} in node \tt{n} is in the set \emph{Established(n, t)} if \tt{n.size} is greater than \tt{b}'s index at time $t$.
+\end{definition}
+
+\begin{lemma}[sizeProgress] \label{lem::sizeProgress}
+{\nf{\tt{n.size}}} is non-decreasing over time.
+\end{lemma}
+%\begin{proof}
+%  Simply because \tt{n.head} is only incremented.
+%\end{proof}
+
+\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}({\nf{\tt{h=n.size}}}) might be 1 bit behind the first empty block in the node.
+\end{lemma}
+%\begin{proof}
+% Because at the end of every \tt{Refresh()} with block size greater than 0 (Lines 53,56) \tt{n.head} is incremented.
+Maybe some process goes to sleep before incrementing the head, but after sleeping if \tt{h} does not increase then \tt{CAS} in Line 52 is going to be failed and nothing is going to be appended to \tt{n.blocks}.
+%\end{proof}
+
+%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
+%  If block \tt{b} in node \tt{n} is established, then all the blocks in \tt{n.blocks[]} before \tt{b} are established.
+If block \tt{b} in node \tt{n} is not new then all the blocks in \tt{n.blocks[]} before \tt{b} are not new.
+%\end{lemma}
+%\begin{proof}
+%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
+%\end{proof}
+
+\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
+  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
+\end{lemma}
+%\begin{proof}
+%Because blocks are only appended(not modified) with CAS to \tt{n.blocks[n.head]} and \tt{n.head} is non-decreasing.
+%\end{proof}
+
+\begin{lemma}[createBlock] \label{lem::createBlock}
+  Suppose \tt{n.CreateBlock( h, x)} is invoked at time $t$.
+The blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ that are not propagated to $Established(n, t)$, are subblock of the block returned by \tt{CreateBlock(n, h, x)}.
+\end{lemma}
+%\begin{proof}
+% We prove the claim for the left child.
+Blocks in \tt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
+Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
+See Figure~\ref{fig::createBlock}.
+The right child is the same.
+%\end{proof}
+
+\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
+  Suppose \tt{Refresh(n)}'s \tt{CAS(n.blocks[h], null, new)} returns \tt{true}.
+Let $t$ be the time \tt{Refresh(n)} is invoked, blocks propgatated to $Established(n.left, t)$ and $Established(n.right, t)$ are propagated to in $Established(n, t)$ after \tt{CAS(n.blocks[h], null, new)}.
+\end{lemma}
+%\begin{proof}
+%By Lemma~\ref{lem::createBlock} \tt{new} contains \tt{n}'s childrens' established blocks before Line 43 which is  appended to \tt{n.blocks} by \tt{CAS} in Line 48.
+%\end{proof}
+
+\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
+    If instance $r$ of \tt{Refresh(n)} returns \tt{false}, then there is another successful instance $r^\prime$ of \tt{Refresh(n)} that has performed a successful \tt{CAS(n.blocks[h], null, new)}(Line 49) after Line 43(\tt{h= n.head}) of $r$.
+\end{lemma}
+%\begin{proof}
+%  If there is no other concurrent successful \tt{Refresh(n)} then \tt{Refresh(n)} would succeed in Line 48.
+So there is another \tt{Refresh(n)}, that has to \tt{CAS}ed successfully its block in \tt{n.blocks[h]} after Line 43 of \tt{Refresh(n)}.
+Otherwise the other \tt{Refresh(n)} should have read \tt{h$^\prime$>h} instead of \tt{h} for \tt{n.head}(Line 52).
+%\end{proof}
+
+\begin{lemma}[Double Refresh] \label{doubleRefresh}
+  Consider two consecutive instances $r_1$,$r_2$ of \tt{Refresh(n)} by the same process (Lines 35,36).
+Let be the time before $r_1$ invoked.
+After $r_2$'s \tt{CAS} all the blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ are in $Established(n, t)$.
+\end{lemma}
+%\begin{proof}
+%  
+%
+%If Line 35 (first Refresh which we call $R_1$) returns \tt{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
+If not, then there is another successful instance of \tt{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
+$R_1^\prime$ may or may not have propagated some subblocks of \tt{new} to \tt{n}.
+It is obvious that the \tt{new} constructed by the second Refresh in Line 36 contains the blocks in \tt{new} by $R_1$ which $R_1^\prime$ did not contain, since \tt{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
+If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
+If not, it is deduced that \tt{n.blocks[h]} was not \tt{null} before $R_2$'s \tt{CAS}.
+Furthermore, \tt{n.blocks[h]} was \tt{null} before reading \tt{h} by $R_1$.
+So there is a successful \tt{Refresh()} after the read of \tt{h} in $R_1$ and before the \tt{CAS} of $R_2$.
+This \tt{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
+See Figure~\ref{}.
+%\end{proof}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
+  \caption{$R_2^\prime$'s \tt{CAS} is executed after \tt{h1=n.head}.}
+\end{figure}
+
+ Block \tt{new} is created of new established subblocks of children of \tt{n}(Lemma \ref{lem::createBlock}, Line 46).
+If \tt{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \tt{n}.
+
+\begin{lemma}[Double Refresh] \label{doublyRefresh}
+All operations in \tt{n}'s children's blocks before line \tt{35} are guaranteed to be in \tt{n}'s blocks after Line~\tt{37}.
+\end{lemma}
+%\begin{proof}
+%Suppose block \tt{b} with index \tt{i} is in the the left child of \tt{n} before the line 35.
+By Lemma~\ref{head} it follows that \tt{n.left.head} is greater than \tt{i}.
+\tt{Refresh()} calls \tt{CreateBlock()} and creates a block from blocks between \tt{n.blocks[n.head].end\textsubscript{left}} and \tt{n.left.head} in the left child, which contains \tt{b} as well.
+First it tries to append it in \tt{n.blocks.head} and if it was succsuful it continues recursivley.
+If not it tries again, and if the second call of \tt{Refresh()} in Line 36 fails.
+It means there is another \tt{Refresh} which has reah its \tt{i} after the Line 35, so it contains \tt{b} as well.
+%\end{proof}
+\tt{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
+If a \tt{Refresh()} procedure returns true it means it has appended the block created by \tt{CreateBlock()} into the parent node's sequence.
+So suppose two \tt{Refresh}es fail.
+Since the first \tt{Refresh()} was not successful, it means another CAS operation by a \tt{Refresh}, concurrent to the first \tt{Refresh()}, was successful before the second \tt{Refresh()}.
+So it means the second failed \tt{Refresh} is concurrent with a successful \tt{Refresh()} that assuredly has read block before the mentioned line \tt{35}.
+After all it means if any of the \tt{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
+
+\begin{lemma}[Append]\label{append}
+ When \text{\tt{Append(op)}} is finished, \tt{op} appears exactly once in a block of \tt{root.blocks}.
+\end{lemma}
+%\begin{proof}
+%\tt{Append(op)} adds \tt{op} to \tt{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \tt{Propagate()} recursively propagates \tt{op} up to the root.
+%By lemma \ref{doublyRefresh} we know that operation \tt{op} propagates from child to parent at each level.
+%\end{proof}
+
+\begin{lemma}[Block Size Upper Bound]\label{blockSize}
+Each block in a node contains at most one operation from each processs.  
+\end{lemma}
+%\begin{proof}
+%  Note that \tt{prevIndex} and \tt{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
+After a block has propagated to the parent the blocks between \tt{prevIndex} and \tt{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
+The number of new operations which have not propagated yet to the parent cannot be more than $p$.
+If so by the law of pigeonholes there is a process which has appended two concurrent operations.
+%\end{proof}
+
+\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
+Each block in a node  has at most $p$ subblocks.
+\end{lemma}
+%\begin{proof}
+%  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
+%\end{proof}
+
+\begin{definition} [Ordering of operations inside a node] \label{ordering}
+$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
+
+\begin{itemize}
+  \item $E(n,i)$ is the sequence of enqueue operations that are member of \tt{n.blocks[i]} ordered by process id.
+  \item 
+$D(n,i)$ is the sequence of dequeue operations that are member of \tt{n.blocks[i]} ordered by process id.
+\item $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+\end{itemize}
+\end{definition}
+
+\begin{theorem}
+The queue implementation is linearizable.
+\end{theorem}
+%\begin{proof}
+%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
+%  \begin{enumerate}
+%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
+The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
+%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
+\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
+It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
+By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
+If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
+%  \end{enumerate} 
+%\end{proof}
+
+\begin{lemma}[Get] \label{get}
+\tt{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
+\end{lemma}
+%\begin{proof}
+%It is obvious that \tt{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
+To find the \tt{i}th enqueue in block \tt{b} of an internal node \tt{n} in line 87 it is decided that it resides in the left child or the right child.
+This decision is made by Definition of $E(n,b)$.
+After that Lines 88, 92 search the proper subblocks of \tt{b}.
+From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \tt{prevBlock} and \tt{lastBlock} block of the \tt{CreateBlock()}.
+%\end{proof}
+
+\begin{lemma}[Index]
+ Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
+\end{lemma}
+%\begin{proof}
+%  \tt{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \tt{n} in \tt{n.parent} and then computes the order in $D(n.parent, superblock)$.
+Then calls \tt{index()} on \tt{n.parent} recursively.
+It is easy to see why the second is correct.
+Correctness of computing superblock comes from Lemma \ref{superBlock}.
+%\end{proof}
+
+\begin{lemma}[Computing SuperBlock]\label{superBlock}
+  If \tt{Index(n,b,i)} performs line 101, then \tt{superblock} contains \tt{i}th Dequeue in \tt{b}th block of node \tt{n}.
+\end{lemma}
+%\begin{proof}
+%\begin{enumerate}
+% \item Value read for \tt{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
+%  Values \tt{c\textsubscript{dir}} read in lines 23, \tt{super} are set before incrementing in lines 26,27.
+%
+% \item \tt{super[]} preserves order from child to parent; if in a child block \tt{b} is before \tt{c} then \tt{b.time} $\leq$ \tt{c.time} and \tt{super[b.time]} $\leq$ \tt{super[c.time]}\\$\blacktriangleright$
+%  Follows from the order of lines 60, 48, 49.
+%
+%% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
+%% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
+% \item \tt{super[i+1]-super[i]}$\leq p$
+%\\$\blacktriangleright$
+% In a Refresh with successful CAS in line 46, \tt{super} and \tt{counter} are set for each child in lines 48,49.
+Assume the current value of the counter in node \tt{n} is \tt{i+1} and still \tt{super[i+1]} is not set.
+If an instance of successful \tt{Refresh(n)} finishes \tt{super[i+1]} is set a new value and a block is added after \tt{n.parent[sup[i]]}.
+There could be at most $p$ successful unfinished concurrent instances of \tt{Refresh()} that have not reached line 49.
+So the distance between \tt{super[i+1]} and \tt{super[i]} is less than $p$.
+%
+% \item Superblock of \tt{b} is within range $\pm 2p$ of the \tt{super[b.time]}.
+%\\$\blacktriangleright$
+%\tt{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
+It is trivial to see that \tt{n.super} and \tt{n.b.counter} are increasing.
+\tt{super(b)} is the real superblock of b.
+\tt{super(t]} is the index of the superblock of the last block with time \tt{t}.
+If \tt{b.time} is \tt{t} we have:
+%$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+%
+%\end{enumerate}
+%\end{proof}
+
+\begin{lemma}[Computing Queue's Head] \label{computeHead}
+  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q. If Q is empty \tt{ComputeHead(i,b)} returns -1, otherwise
+ returns index in $E(root,b)$ of X.
+  \end{lemma}
+
+\begin{lemma}[Validity of \tt{head}]\label{head}
+No two blocks are written in the same index in \tt{n.blocks}.
+\end{lemma}
+%\begin{proof}
+%  \tt{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \tt{head} read.
+If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \tt{head}.
+If a process dies before line 51, another process will increment \tt{head} in line 54.
+%\end{proof}
+
+\begin{lemma}[Validity of \tt{super} and \tt{counter}]\label{superCounter}
+If \tt{super[i] $\neq$ null}, then \tt{super[i]} in node \tt{n} is the superblock of a block with \tt{time=i}.
+\end{lemma}
+%\begin{proof}
+%After a successful CAS in line 46 \tt{super} and \tt{counter} are modified in both children.
+\tt{super[i]} is supposed to be the superblock of a block with \tt{time}=i and \tt{counter} is the timer in each node.
+\tt{super[i]} and \tt{counter=i} are expected to update after a bunch of blocks with \tt{time=i} have been aggregated together into a block in the parent.
+If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \tt{time}.
+We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
+If a process dies right after line 48, then \tt{counter} will remain the same and \tt{super[i]} is correct.
+Furthermore we are sure that when \tt{super[i]} is read it will not be \tt{null}.
+%\end{proof}
+
+\begin{lemma}[Search Ranges]\label{search}
+  Preconditions of all invocation of \tt{BSearch} are satisfied.
+\end{lemma}
+%\begin{proof}
+%  
+%Line 83: \tt{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+%
+%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+%
+%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \tt{super[i]} is at most $p$, since at most $p$ processes could die.
+%
+%\end{proof}
+
 \end{document}

### 2022-05-04 article/psuedocode.tex
infinte loop problem test instance

--- 2022-04-26 article/psuedocode.tex
@@
 \renewcommand{\bf}[1]{\textbf{#1}}
+\newcommand{\nf}[1]{\normalfont{#1}}
 \newcommand{\cmt}[1]{\Comment{#1}}
 
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
+%\usepackage{amsmath,amssymb,amsthm}
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\theoremstyle{definition}
+%\newtheorem{definition}[theorem]{Definition}
 
 \begin{document}
+
+\section{Pseudocode}
 
@@
 
-%\Statex \bf{Structure}
+\Statex \bf{Structure}
 
@@
 
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
-\end{itemize}
+% \Statex $\diamondsuit$ \tt{\sl{Local}}
+% \begin{itemize}
+% \item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
+% \end{itemize}
 
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{Structures}}
+% \Statex
+% \Statex $\diamondsuit$ \tt{\sl{Structures}}
 
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
-\item \tt{\sl{BlockList} blocks}
-\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
-\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
+% \Statex $\blacktriangleright$ \tt{\sl{Node}}
+% \begin{itemize}
+% \item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
+% \item \tt{\sl{BlockList} blocks}
+%   \textsf{\tt{BlockList} is implemented with an array.}
+% \item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
+% \item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
 Since it is incremented after propagating, it may be behind by 1.}
-\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
-\end{itemize}
+% \item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
+% \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
-\begin{itemize}
+% %\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
+% %\begin{itemize}
+% %  
+% %\item \tt{\sl{PBRT} blocks}
+% %
+% %  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
+% %  
+% %\end{itemize}
 
-\item \tt{\sl{PBRT} blocks}
+% %\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
+% %\begin{itemize}
 
-  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
+% %\end{itemize}
 
-\end{itemize}
+% \Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
+% \begin{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
-\begin{itemize}
+%   \item \tt{\sl{int} \tt{last\sub{done}}}
 
-\item \tt{\sl{Block[]} blocks}
-
-  \textsf{\tt{BlockList} is implemented with an array.}
-
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
-\begin{itemize}
-
-  \item \tt{\sl{int} \tt{last\sub{done}}}
-
-  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
+%   \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
 A block is finished if all of its operations are finished.
@@
 
-\end{itemize}
-\pagebreak
+% \end{itemize}
+% \pagebreak
 
-\Statex $\blacktriangleright$ \tt{\sl{Block}} 
-%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
-\cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
+% \Statex $\blacktriangleright$ \tt{\sl{Block}} 
+% %\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
+% \cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
 \it{put the definitions before the pseudocode}}
 
-\begin{itemize}
-%  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
-%  \textsf{: \# enqueue, dequeue operations in the block}
-%  \item \tt{\sl{int} num, sum}
-%  \textsf{: total \# operations in block, prefix sum of \tt{num}}
+% \begin{itemize}
+% %  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
+% %  \textsf{: \# enqueue, dequeue operations in the block}
+% %  \item \tt{\sl{int} num, sum}
+% %  \textsf{: total \# operations in block, prefix sum of \tt{num}}
 
-  \item \tt{\sl{int} group}
-  \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
-\end{itemize}
+%   \item \tt{\sl{int} group}
+%   \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
+% \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}
+% \Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
+% \begin{itemize}
+%   \item \tt{\sl{Object} element}
+%   \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}
 
-    \item \tt{\sl{Object} response}
-  \textsf{: stores the response of the operation in the \tt{LeafBlock}.}
+%     \item \tt{\sl{Object} response}
+%   \textsf{: stores the response of the operation in the \tt{LeafBlock}.}
 
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
+%     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+%   \textsf{: \# enqueue, dequeue operations in the prefix for the block}
 
-\end{itemize}
+% \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~index of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}
-  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}
-  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}
-  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}
-  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
-\end{itemize}
+% \Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
+% \begin{itemize}
+%     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
+%   \textsf{:~~index of the last subblock of the block in the left and right child}
+%   \item \tt{\sl{int} sum\sub{enq-left}}
+%   \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+%   \item \tt{\sl{int} sum\sub{deq-left}}
+%   \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+%   \item \tt{\sl{int} sum\sub{enq-right}}
+%   \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
+%   \item \tt{\sl{int} sum\sub{deq-right}}
+%   \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
+% \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} length}
-  \textsf{: length of the queue after performing all operations in the prefix for this block}
-%  \item \tt{\sl{int} sum\sub{non-null deq}}
-%  \textsf{: count of non-null dequeus up to this block}
-  \item \tt{\sl{counter} num\sub{finished}}
-  \textsf{: number of finished operations in the block}
-    \item \tt{\sl{int} order}
-  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
-\end{itemize}
+% \Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
+% \begin{itemize}
+%   \item \tt{\sl{int} length}
+%   \textsf{: length of the queue after performing all operations in the prefix for this block}
+% %  \item \tt{\sl{int} sum\sub{non-null deq}}
+% %  \textsf{: count of non-null dequeus up to this block}
+%   \item \tt{\sl{counter} num\sub{finished}}
+%   \textsf{: number of finished operations in the block}
+%     \item \tt{\sl{int} order}
+%   \textsf{: the index of the block in the \tt{BlockList} containing the block.}
+% \end{itemize}
 
-%\Statex $\blacktriangleright$ \tt{Conventions}
-%\begin{itemize}
-%  \item \tt{i} : index of ith operation in the tree
-%  \item \tt{j} : index of jth operation in a node
-%  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
-%  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
-%\end{itemize}
+% %\Statex $\blacktriangleright$ \tt{Conventions}
+% %\begin{itemize}
+% %  \item \tt{i} : index of ith operation in the tree
+% %  \item \tt{j} : index of jth operation in a node
+% %  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
+% %  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
+% %\end{itemize}
 
@@
 
-\begin{footnotesize}
+% \begin{footnotesize}
 
-\it{Variable naming:}
-\begin{itemize}
-  \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
-  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
-\end{itemize}
+% \it{Variable naming:}
+% \begin{itemize}
+%   \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
+%   \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
+% \end{itemize}
 
-\it{Abbreviations:}
-\begin{itemize}
- \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
- \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
-\end{itemize}
-\end{footnotesize}
+% \it{Abbreviations:}
+% \begin{itemize}
+%  \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
+%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
+%   \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
+% \end{itemize}
+% \end{footnotesize}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{200}
-\begin{multicols}{2}
+% \begin{algorithm}
+% \caption{\tt{\sl{Queue}}}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{200}
+% \begin{multicols}{2}
 
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-%\State \tt{b.num\sub{enq}=1}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
+% \Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
+% \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
+% \State \tt{newBlock.element= e}
+% %\State \tt{b.num\sub{enq}=1}
+% \State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
+% \State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
+% \State \tt{leaf.}\Call{Append}{newBlock}
+% \EndFunction{Enqueue}
 
-\Statex
-\Function{Object}{Dequeue()}{}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
-\EndFunction{Dequeue}
+% \Statex
+% \Function{Object}{Dequeue()}{}
+% \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
+% \State \tt{newBlock.element= null}
+% \State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
+% \State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
+% \State \tt{leaf.}\Call{Append}{newBlock}
+% \State \Return\tt{leaf.}\Call{HelpDequeue()}{}
+% \EndFunction{Dequeue}
 
-\Statex
+% \Statex
+% \pagebreak
 
-\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
+% \Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
 Returns \tt{<-1,-->} if the queue is empty.}
-\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
-\State \Return \tt{<-1,-->}
-\Else
-\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}.
+% \If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
+% \State \Return \tt{<-1,-->}
+% \Else
+% \Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}.
 $r$th non-null dequeue returns the element of th $r$th enqueue.
@@
 Note that the $i$th dequeue in the given block is not a non-null dequeue.}
-\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
-\State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
-\EndIf
-\EndFunction{FindResponse}
+% \State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
+% \State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
+% \EndIf
+% \EndFunction{FindResponse}
 
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
+% \end{multicols}
+% \end{algorithmic}
+% \end{algorithm}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{300}
-\begin{multicols}{2}
+% \begin{algorithm}
+% \caption{\tt{\sl{Node}}}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{300}
+% \begin{multicols}{2}
 
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}}
-\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
+% \Function{void}{Propagate()}{}
+% \If{\bf{not} \Call{Refresh()}{}}
+% \State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
+% \EndIf
+% \If{\tt{this} \bf{is not} \tt{root}}
+% \State \tt{parent.}\Call{Propagate()}{}
+% \EndIf
+% \EndFunction{Propagate}
 
-\Statex
+% \Statex
 
-\Function{boolean}{Refresh()}{}
-\State \tt{s= size}
-%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
-\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{\tt{blocks.tryAppend(new, s)}}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
-\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
-\EndFor
-\State \tt{\Call{CAS}{size, h, h+1}}
-\State \Return{\tt{true}}
-\Else
-\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
-\State \Return{ \tt{false}}
-\EndIf
-\EndFunction{Refresh}
+% \Function{boolean}{Refresh()}{}
+% \State \tt{s= size}
+% %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
+% \State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
+% \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
+% \ElsIf{\tt{blocks.tryAppend(new, s)}}
+% \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
+% \State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
+% \State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
+% \EndFor
+% \State \tt{\Call{CAS}{size, h, h+1}}
+% \State \Return{\tt{true}}
+% \Else
+% \State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
+% \State \Return{ \tt{false}}
+% \EndIf
+% \EndFunction{Refresh}
 
-\Statex
+% \Statex
 
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
-\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
+% \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
+% \Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
 
-\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
+% \Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
 Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
-%\State \Return \tt{result block's index}
-\EndFunction{BSearch}
+% %\State \Return \tt{result block's index}
+% \EndFunction{BSearch}
 
-\pagebreak
+% \pagebreak
 
-\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}.
+% \Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
+% \Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}.
 Returns the created block as well as values read from each child's num\sub{propagated} field.
 These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
-\State \tt{block newBlock= \Call{new}{\sl{block}}}
-\State \tt{newBlock.group= num\sub{propagated}}
-\State \tt{newBlock.order= i}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.size} \label{lastLine}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{newBlock.end\sub{dir}= index\sub{last}}
-\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{this\sub{dir}= dir.num\sub{propagated}}
-\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
-\EndFor
-%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
-%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
-%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
-%\State \tt{b.sum= n.blocks[i-1].sum + b.num}
+% \State \tt{block newBlock= \Call{new}{\sl{block}}}
+% \State \tt{newBlock.group= num\sub{propagated}}
+% \State \tt{newBlock.order= i}
+% \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
+% \State \tt{index\sub{last}= dir.size} \label{lastLine}
+% \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
+% \State \tt{newBlock.end\sub{dir}= index\sub{last}}
+% \State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
+% \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
+% \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
+% \State \tt{this\sub{dir}= dir.num\sub{propagated}}
+% \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
+% \State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
+% \EndFor
+% %\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
+% %\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
+% %\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
+% %\State \tt{b.sum= n.blocks[i-1].sum + b.num}
 
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
-%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
-\EndIf
+% \If{\tt{this} \bf{is} \tt{root}}
+% \State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
+% %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
+% \EndIf
 
-\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
-\EndFunction{CreateBlock}
+% \State \Return \tt{<b, np\sub{left}, np\sub{right}>}
+% \EndFunction{CreateBlock}
 
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
+% \end{multicols}
+% \end{algorithmic}
+% \end{algorithm}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\caption{Node}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{400}
+% \begin{algorithm}
+% \caption{Node}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{400}
 
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
-\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element}
-\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
-\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
-\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
-\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
-\EndIf
-\EndFunction{GetEnq}
+% \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
+% \Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
+% \If{\tt{this} \bf{is} \tt{leaf}}
+% \State\Return \tt{blocks[b].element}
+% \ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
+% \State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
+% \State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
+% \Else
+% \State \tt{i= i-blocks[b].num\sub{enq-left}}
+% \State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
+% \State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
+% \EndIf
+% \EndFunction{GetEnq}
 
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
-\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>}
-\Else
-\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
-\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
-\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
-\If{\tt{dir {\keywordfont is} right}}
-%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
-\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
-\EndIf
-\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
-\EndIf
-\EndFunction{Index}
+% \Statex
+% \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
+% \Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
+% \If{\tt{this} \bf{is} \tt{root}}
+% \State\Return \tt{<b, i>}
+% \Else
+% \State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
+% \State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
+% \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
+% \If{\tt{dir {\keywordfont is} right}}
+% %\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
+% \State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
+% \EndIf
+% \State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
+% \EndIf
+% \EndFunction{Index}
 
-\end{algorithmic}
-\end{algorithm}
+% \end{algorithmic}
+% \end{algorithm}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\caption{Leaf}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{500}
+% \begin{algorithm}
+% \caption{Leaf}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{500}
 
-\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
-\State \tt{size+=1} \label{appendEnd} 
-\State \tt{blk.group= size} \label{appendStart} 
-\State \tt{blocks[size]= blk} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
+% \Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
+% \State \tt{size+=1} \label{appendEnd} 
+% \State \tt{blk.group= size} \label{appendStart} 
+% \State \tt{blocks[size]= blk} 
+% \State \tt{parent.}\Call{Propagate()}{} 
+% \EndFunction{Append}
 
-\Statex
+% \Statex
 
-\Function{Object}{HelpDequeue()}{}
-\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
-\State \tt{<r\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
- \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
+% \Function{Object}{HelpDequeue()}{}
+% \State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
+% \State \tt{<r\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
+%  \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
 If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
-\If{\tt{r\sub{enq}==-1}}
-\State \tt{output= null}
-\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
-\State \tt{last\sub{done}= b\sub{deq}}
-\EndIf
-\pagebreak
-\Else
-\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
-\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
-\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
-\State \tt{last\sub{done}= b\sub{deq}}
-\ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
-\State \tt{last\sub{done}= b\sub{enq}}
-\EndIf
-\EndIf
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
+% \If{\tt{r\sub{enq}==-1}}
+% \State \tt{output= null}
+% \State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
+% \If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
+% \State \tt{last\sub{done}= b\sub{deq}}
+% \EndIf
+% \pagebreak
+% \Else
+% \State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
+% \State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
+% \State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
+% \If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
+% \State \tt{last\sub{done}= b\sub{deq}}
+% \ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
+% \State \tt{last\sub{done}= b\sub{enq}}
+% \EndIf
+% \EndIf
+% \State \Return{\tt{output}}
+% \EndFunction{Dequeue}
 
-\end{algorithmic}
-\end{algorithm}
+% \end{algorithmic}
+% \end{algorithm}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\caption{BlockList}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{600}
+% \begin{algorithm}
+% \caption{BlockList}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{600}
 
-\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+% \Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
 Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
@@
 
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
-  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-\Statex \tt{root: }\sf{pointer to the root of the PBRT}
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
-\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-\State \tt{Help()}
-\State \tt{CollectGarbage()}
-\EndIf
-\State \tt{blk.num\sub{finished}= 0}
-\State \tt{*oldRoot= \&root.blocks.root}
-\State \tt{*newRoot= root.blocks.Append(blk).root}
-\State \Return \tt{CAS(root, oldRoot, newRoot)}
-\EndFunction{TryAppend}
+% \Statex
+% \Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
+%   \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+%   \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+% \Statex \tt{root: }\sf{pointer to the root of the PBRT}
+% \Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
+% \If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
+% \State \tt{Help()}
+% \State \tt{CollectGarbage()}
+% \EndIf
+% \State \tt{blk.num\sub{finished}= 0}
+% \State \tt{*oldRoot= \&root.blocks.root}
+% \State \tt{*newRoot= root.blocks.Append(blk).root}
+% \State \Return \tt{CAS(root, oldRoot, newRoot)}
+% \EndFunction{TryAppend}
 
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{blocks[]: }\sf{array of blocks}
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
+% \Statex
+% \Statex $\diamondsuit$ \tt{\sl{Array implementation}}
+% \Statex \tt{blocks[]: }\sf{array of blocks}
+% \Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
+% \State \Return{\tt{CAS(blocks[n], null, blk)}}
+% \EndFunction{TryAppend}
 
-\end{algorithmic}
-\end{algorithm}
+% \end{algorithmic}
+% \end{algorithm}
 
-%##########################################
+% %##########################################
 
-\begin{algorithm}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{800}
-\begin{multicols}{2}
+% \begin{algorithm}
+% \begin{algorithmic}[1]
+% \setcounter{ALG@line}{800}
+% \begin{multicols}{2}
 
-\Function{void}{Help}{}\cmt{Helps pending operations}
-\For{\tt{leaf l} \bf{in leaves}}
-\State{\tt{last= l.size-1}}
-\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
-\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
-\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
-\EndIf
-\EndFor
-\EndFunction{Help}
-\Statex
+% \Function{void}{Help}{}\cmt{Helps pending operations}
+% \For{\tt{leaf l} \bf{in leaves}}
+% \State{\tt{last= l.size-1}}
+% \cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
+% \If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
+% \State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
+% \EndIf
+% \EndFor
+% \EndFunction{Help}
+% \Statex
 
-\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
-\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
-\State \tt{t1,t2= RBT.split(order, s)}
-\State \tt{RBTRoot.CAS(t2.root)}
-\EndFunction{CollectGarbage}
+% \Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
+% \State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
+% \State \tt{t1,t2= RBT.split(order, s)}
+% \State \tt{RBTRoot.CAS(t2.root)}
+% \EndFunction{CollectGarbage}
 
-\Statex
+% \Statex
 
-\Function{Block}{FindMostRecentDone}{b}
-\For{\tt{leaf l} \bf{in leaves}}
-\State\tt{max= Max(l.maxOld, max)}
-\EndFor
-\State\Return\tt{max} \cmt{This snapshot suffies.}
-\EndFunction{FindMostRecentDone}
+% \Function{Block}{FindMostRecentDone}{b}
+% \For{\tt{leaf l} \bf{in leaves}}
+% \State\tt{max= Max(l.maxOld, max)}
+% \EndFor
+% \State\Return\tt{max} \cmt{This snapshot suffies.}
+% \EndFunction{FindMostRecentDone}
 
-\Statex
+% \Statex
 
-\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
+% \Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
 
-\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
+% \If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
 
-\State \Return \tt{this.leaf.response(block.order)}
-\EndIf
+% \State \Return \tt{this.leaf.response(block.order)}
+% \EndIf
 
-\EndFunction{FallBack}
+% \EndFunction{FallBack}
 
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
+% \end{multicols}
+% \end{algorithmic}
+% \end{algorithm}
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%                      %
+%                      %
+%                      %
+%                      %
+%                      %
+%                      %
+%                      %
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \clearpage
+% \section{Proof of Linearizability}
+
+% \begin{definition}
+% If \tt{n.blocks[i]==b}  we call \tt{i} the \emph{index} of block \tt{b} in node \tt{n}.
+Block \tt{b} is before block \tt{b$^\prime$} in node \tt{n} if and only if \tt{b}'s index is smaller than \tt{b$^\prime$}'s.
+Block \tt{b} is propagated to node \tt{n} or set \tt{S} if \tt{b} is in \tt{n.blocks} or \tt{S} or is a subblock of a block in \tt{n.blocks} or \tt{S}.
+% \end{definition}
+
+% %\begin{definition}
+% %  Block \tt{b} in node \tt{n} is new if \tt{b} is not a subblock of any block in \tt{n.parent.blocks[]}.
+% %\end{definition}
+
+% \begin{definition}
+%   Block \tt{b} in node \tt{n} is in the set \emph{Established(n, t)} if \tt{n.size} is greater than \tt{b}'s index at time $t$.
+% \end{definition}
+
+% \begin{lemma}[sizeProgress] \label{lem::sizeProgress}
+% {\nf{\tt{n.size}}} is non-decreasing over time.
+% \end{lemma}
+% %\begin{proof}
+% %  Simply because \tt{n.head} is only incremented.
+% %\end{proof}
+
+% \begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}({\nf{\tt{h=n.size}}}) might be 1 bit behind the first empty block in the node.
+% \end{lemma}
+% %\begin{proof}
+% % Because at the end of every \tt{Refresh()} with block size greater than 0 (Lines 53,56) \tt{n.head} is incremented.
+Maybe some process goes to sleep before incrementing the head, but after sleeping if \tt{h} does not increase then \tt{CAS} in Line 52 is going to be failed and nothing is going to be appended to \tt{n.blocks}.
+% %\end{proof}
+
+% %\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
+% %  If block \tt{b} in node \tt{n} is established, then all the blocks in \tt{n.blocks[]} before \tt{b} are established.
+If block \tt{b} in node \tt{n} is not new then all the blocks in \tt{n.blocks[]} before \tt{b} are not new.
+% %\end{lemma}
+% %\begin{proof}
+% %  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
+% %\end{proof}
+
+% \begin{lemma}[establishedOrder]\label{lem::establishedOrder}
+%   If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
+% \end{lemma}
+% %\begin{proof}
+% %Because blocks are only appended(not modified) with CAS to \tt{n.blocks[n.head]} and \tt{n.head} is non-decreasing.
+% %\end{proof}
+
+% \begin{lemma}[createBlock] \label{lem::createBlock}
+%   Suppose \tt{n.CreateBlock( h, x)} is invoked at time $t$.
+The blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ that are not propagated to $Established(n, t)$, are subblock of the block returned by \tt{CreateBlock(n, h, x)}.
+% \end{lemma}
+% %\begin{proof}
+% % We prove the claim for the left child.
+Blocks in \tt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
+Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
+See Figure~\ref{fig::createBlock}.
+The right child is the same.
+% %\end{proof}
+
+% \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
+%   Suppose \tt{Refresh(n)}'s \tt{CAS(n.blocks[h], null, new)} returns \tt{true}.
+Let $t$ be the time \tt{Refresh(n)} is invoked, blocks propgatated to $Established(n.left, t)$ and $Established(n.right, t)$ are propagated to in $Established(n, t)$ after \tt{CAS(n.blocks[h], null, new)}.
+% \end{lemma}
+% %\begin{proof}
+% %By Lemma~\ref{lem::createBlock} \tt{new} contains \tt{n}'s childrens' established blocks before Line 43 which is  appended to \tt{n.blocks} by \tt{CAS} in Line 48.
+% %\end{proof}
+
+% \begin{lemma}[falseRefresh] \label{lem:falseRefresh}
+%     If instance $r$ of \tt{Refresh(n)} returns \tt{false}, then there is another successful instance $r^\prime$ of \tt{Refresh(n)} that has performed a successful \tt{CAS(n.blocks[h], null, new)}(Line 49) after Line 43(\tt{h= n.head}) of $r$.
+% \end{lemma}
+% %\begin{proof}
+% %  If there is no other concurrent successful \tt{Refresh(n)} then \tt{Refresh(n)} would succeed in Line 48.
+So there is another \tt{Refresh(n)}, that has to \tt{CAS}ed successfully its block in \tt{n.blocks[h]} after Line 43 of \tt{Refresh(n)}.
+Otherwise the other \tt{Refresh(n)} should have read \tt{h$^\prime$>h} instead of \tt{h} for \tt{n.head}(Line 52).
+% %\end{proof}
+
+% \begin{lemma}[Double Refresh] \label{doubleRefresh}
+%   Consider two consecutive instances $r_1$,$r_2$ of \tt{Refresh(n)} by the same process (Lines 35,36).
+Let be the time before $r_1$ invoked.
+After $r_2$'s \tt{CAS} all the blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ are in $Established(n, t)$.
+% \end{lemma}
+% %\begin{proof}
+% %  
+% %
+% %If Line 35 (first Refresh which we call $R_1$) returns \tt{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
+If not, then there is another successful instance of \tt{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
+$R_1^\prime$ may or may not have propagated some subblocks of \tt{new} to \tt{n}.
+It is obvious that the \tt{new} constructed by the second Refresh in Line 36 contains the blocks in \tt{new} by $R_1$ which $R_1^\prime$ did not contain, since \tt{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
+If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
+If not, it is deduced that \tt{n.blocks[h]} was not \tt{null} before $R_2$'s \tt{CAS}.
+Furthermore, \tt{n.blocks[h]} was \tt{null} before reading \tt{h} by $R_1$.
+So there is a successful \tt{Refresh()} after the read of \tt{h} in $R_1$ and before the \tt{CAS} of $R_2$.
+This \tt{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
+See Figure~\ref{}.
+% %\end{proof}
+
+% \begin{figure}[hbt]
+%   \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
+%   \caption{$R_2^\prime$'s \tt{CAS} is executed after \tt{h1=n.head}.}
+% \end{figure}
+
+%   Block \tt{new} is created of new established subblocks of children of \tt{n}(Lemma \ref{lem::createBlock}, Line 46).
+If \tt{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \tt{n}.
+
+% \begin{lemma}[Double Refresh] \label{doublyRefresh}
+% All operations in \tt{n}'s children's blocks before line \tt{35} are guaranteed to be in \tt{n}'s blocks after Line~\tt{37}.
+% \end{lemma}
+% %\begin{proof}
+% %Suppose block \tt{b} with index \tt{i} is in the the left child of \tt{n} before the line 35.
+By Lemma~\ref{head} it follows that \tt{n.left.head} is greater than \tt{i}.
+\tt{Refresh()} calls \tt{CreateBlock()} and creates a block from blocks between \tt{n.blocks[n.head].end\textsubscript{left}} and \tt{n.left.head} in the left child, which contains \tt{b} as well.
+First it tries to append it in \tt{n.blocks.head} and if it was succsuful it continues recursivley.
+If not it tries again, and if the second call of \tt{Refresh()} in Line 36 fails.
+It means there is another \tt{Refresh} which has reah its \tt{i} after the Line 35, so it contains \tt{b} as well.
+% %\end{proof}
+% \tt{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
+If a \tt{Refresh()} procedure returns true it means it has appended the block created by \tt{CreateBlock()} into the parent node's sequence.
+So suppose two \tt{Refresh}es fail.
+Since the first \tt{Refresh()} was not successful, it means another CAS operation by a \tt{Refresh}, concurrent to the first \tt{Refresh()}, was successful before the second \tt{Refresh()}.
+So it means the second failed \tt{Refresh} is concurrent with a successful \tt{Refresh()} that assuredly has read block before the mentioned line \tt{35}.
+After all it means if any of the \tt{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
+
+% \begin{lemma}[Append]\label{append}
+%  When \text{\tt{Append(op)}} is finished, \tt{op} appears exactly once in a block of \tt{root.blocks}.
+% \end{lemma}
+% %\begin{proof}
+% %\tt{Append(op)} adds \tt{op} to \tt{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \tt{Propagate()} recursively propagates \tt{op} up to the root.
+% %By lemma \ref{doublyRefresh} we know that operation \tt{op} propagates from child to parent at each level.
+% %\end{proof}
+
+% \begin{lemma}[Block Size Upper Bound]\label{blockSize}
+% Each block in a node contains at most one operation from each processs.  
+% \end{lemma}
+% %\begin{proof}
+% %  Note that \tt{prevIndex} and \tt{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
+After a block has propagated to the parent the blocks between \tt{prevIndex} and \tt{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
+The number of new operations which have not propagated yet to the parent cannot be more than $p$.
+If so by the law of pigeonholes there is a process which has appended two concurrent operations.
+% %\end{proof}
+
+% \begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
+% Each block in a node  has at most $p$ subblocks.
+% \end{lemma}
+% %\begin{proof}
+% %  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
+% %\end{proof}
+
+% \begin{definition} [Ordering of operations inside a node] \label{ordering}
+% $\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
+
+% \begin{itemize}
+%   \item $E(n,i)$ is the sequence of enqueue operations that are member of \tt{n.blocks[i]} ordered by process id.
+%   \item 
+% $D(n,i)$ is the sequence of dequeue operations that are member of \tt{n.blocks[i]} ordered by process id.
+% \item $D(n)=D(n,1).D(n,2).D(n,3)...$
+% \item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+% \end{itemize}
+% \end{definition}
+
+% \begin{theorem}
+% The queue implementation is linearizable.
+% \end{theorem}
+% %\begin{proof}
+% %  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
+% %  \begin{enumerate}
+% %    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
+The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
+% %    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
+\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
+It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
+By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
+If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
+% %  \end{enumerate} 
+% %\end{proof}
+
+% \begin{lemma}[Get] \label{get}
+% \tt{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
+% \end{lemma}
+% %\begin{proof}
+% %It is obvious that \tt{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
+To find the \tt{i}th enqueue in block \tt{b} of an internal node \tt{n} in line 87 it is decided that it resides in the left child or the right child.
+This decision is made by Definition of $E(n,b)$.
+After that Lines 88, 92 search the proper subblocks of \tt{b}.
+From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \tt{prevBlock} and \tt{lastBlock} block of the \tt{CreateBlock()}.
+% %\end{proof}
+
+% \begin{lemma}[Index]
+%  Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
+% \end{lemma}
+% %\begin{proof}
+% %  \tt{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \tt{n} in \tt{n.parent} and then computes the order in $D(n.parent, superblock)$.
+Then calls \tt{index()} on \tt{n.parent} recursively.
+It is easy to see why the second is correct.
+Correctness of computing superblock comes from Lemma \ref{superBlock}.
+% %\end{proof}
+
+% \begin{lemma}[Computing SuperBlock]\label{superBlock}
+%   If \tt{Index(n,b,i)} performs line 101, then \tt{superblock} contains \tt{i}th Dequeue in \tt{b}th block of node \tt{n}.
+% \end{lemma}
+% %\begin{proof}
+% %\begin{enumerate}
+% % \item Value read for \tt{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
+% %  Values \tt{c\textsubscript{dir}} read in lines 23, \tt{super} are set before incrementing in lines 26,27.
+% %
+% % \item \tt{super[]} preserves order from child to parent; if in a child block \tt{b} is before \tt{c} then \tt{b.time} $\leq$ \tt{c.time} and \tt{super[b.time]} $\leq$ \tt{super[c.time]}\\$\blacktriangleright$
+% %  Follows from the order of lines 60, 48, 49.
+% %
+% %% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
+% %% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
+% % \item \tt{super[i+1]-super[i]}$\leq p$
+% %\\$\blacktriangleright$
+% % In a Refresh with successful CAS in line 46, \tt{super} and \tt{counter} are set for each child in lines 48,49.
+Assume the current value of the counter in node \tt{n} is \tt{i+1} and still \tt{super[i+1]} is not set.
+If an instance of successful \tt{Refresh(n)} finishes \tt{super[i+1]} is set a new value and a block is added after \tt{n.parent[sup[i]]}.
+There could be at most $p$ successful unfinished concurrent instances of \tt{Refresh()} that have not reached line 49.
+So the distance between \tt{super[i+1]} and \tt{super[i]} is less than $p$.
+% %
+% % \item Superblock of \tt{b} is within range $\pm 2p$ of the \tt{super[b.time]}.
+% %\\$\blacktriangleright$
+% %\tt{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
+It is trivial to see that \tt{n.super} and \tt{n.b.counter} are increasing.
+\tt{super(b)} is the real superblock of b.
+\tt{super(t]} is the index of the superblock of the last block with time \tt{t}.
+If \tt{b.time} is \tt{t} we have:
+% %$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+% %
+% %\end{enumerate}
+% %\end{proof}
+
+% \begin{lemma}[Computing Queue's Head] \label{computeHead}
+%   Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q.
+If Q is empty \tt{ComputeHead(i,b)} returns -1, otherwise
+%  returns index in $E(root,b)$ of X.
+%   \end{lemma}
+
+% \begin{lemma}[Validity of \tt{head}]\label{head}
+% No two blocks are written in the same index in \tt{n.blocks}.
+% \end{lemma}
+% %\begin{proof}
+% %  \tt{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \tt{head} read.
+If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \tt{head}.
+If a process dies before line 51, another process will increment \tt{head} in line 54.
+% %\end{proof}
+
+% \begin{lemma}[Validity of \tt{super} and \tt{counter}]\label{superCounter}
+% If \tt{super[i] $\neq$ null}, then \tt{super[i]} in node \tt{n} is the superblock of a block with \tt{time=i}.
+% \end{lemma}
+% %\begin{proof}
+% %After a successful CAS in line 46 \tt{super} and \tt{counter} are modified in both children.
+\tt{super[i]} is supposed to be the superblock of a block with \tt{time}=i and \tt{counter} is the timer in each node.
+\tt{super[i]} and \tt{counter=i} are expected to update after a bunch of blocks with \tt{time=i} have been aggregated together into a block in the parent.
+If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \tt{time}.
+We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
+If a process dies right after line 48, then \tt{counter} will remain the same and \tt{super[i]} is correct.
+Furthermore we are sure that when \tt{super[i]} is read it will not be \tt{null}.
+% %\end{proof}
+
+% \begin{lemma}[Search Ranges]\label{search}
+%   Preconditions of all invocation of \tt{BSearch} are satisfied.
+% \end{lemma}
+% %\begin{proof}
+% %  
+% %Line 83: \tt{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+% %
+% %Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+% %
+% %Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \tt{super[i]} is at most $p$, since at most $p$ processes could die.
+% %
+% %\end{proof}
 

### 2022-05-04 article/psuedocode.tex
fix the termination bug (defining \tt with \tt)
remove PBRT and fix lemmas till 9

--- 2022-04-26 article/psuedocode.tex
@@
 \renewcommand{\bf}[1]{\textbf{#1}}
+\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
 \newcommand{\cmt}[1]{\Comment{#1}}
@@
 \item \tt{\sl{BlockList} blocks}
+  \textsf{ implemented with an array.}
 \item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
@@
 
-\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
-\begin{itemize}
+%\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
+%\begin{itemize}
+%  
+%\item \tt{\sl{PBRT} blocks}
+%
+%  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
+%  
+%\end{itemize}
 
-\item \tt{\sl{PBRT} blocks}
+%\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
+%\begin{itemize}
 
-  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
+%\end{itemize}
 
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
-\begin{itemize}
-
-\item \tt{\sl{Block[]} blocks}
-
-  \textsf{\tt{BlockList} is implemented with an array.}
-
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{NonRootNode}}
+\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
 \begin{itemize}
@@
   \textsf{: number of finished operations in the block}
-    \item \tt{\sl{int} order}
-  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
+%    \item \tt{\sl{int} order}
+%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
 \end{itemize}
@@
 \Statex
+\pagebreak
 
-\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
+\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
 Returns \tt{<-1,-->} if the queue is empty.}
@@
 \State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
-\State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
+\State \Return \tt{<root.\Call{BSearch}{sum\sub{enq}, r\sub{enq}, root.FindMostRecentDone(), root.size}, r\sub{enq}>}
 \EndIf
@@
 \begin{algorithm}
+\caption{Root}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{500}
+
+\Function{Block}{FindMostRecentDone}{}
+\For{\tt{leaf l} \bf{in leaves}}
+\State\tt{max= Max(l.maxOld, max)}
+\EndFor
+\State\Return\tt{max} \cmt{This snapshot suffies.}
+\EndFunction{FindMostRecentDone}
+
+\end{algorithmic}
+\end{algorithm}
+%##########################################
+
+\begin{algorithm}
 \caption{Leaf}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{500}
+\setcounter{ALG@line}{600}
 
@@
 \State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
-\State \tt{<r\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
+\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
  \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
@@
 
+\Statex
+
+\Function{void}{Help}{}\cmt{Helps pending operations}
+
+\State{\tt{last= l.size-1}}
+\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
+\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
+\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
+\EndIf
+
+\EndFunction{Help}
+
 \end{algorithmic}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{600}
+\setcounter{ALG@line}{700}
 
@@
 \Statex
-\Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
-  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-\Statex \tt{root: }\sf{pointer to the root of the PBRT}
+\Statex $\diamondsuit$ \tt{\sl{root implementation}}
+%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
 \Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
 \If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-\State \tt{Help()}
-\State \tt{CollectGarbage()}
+\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
+\State \tt{l.Help()}
+\EndFor
 \EndIf
 \State \tt{blk.num\sub{finished}= 0}
-\State \tt{*oldRoot= \&root.blocks.root}
-\State \tt{*newRoot= root.blocks.Append(blk).root}
-\State \Return \tt{CAS(root, oldRoot, newRoot)}
+%\State \tt{*oldRoot= \&root.blocks.root}
+%\State \tt{*newRoot= root.blocks.Append(blk).root}
+%\State \Return \tt{CAS(root, oldRoot, newRoot)}
+\State \Return{\tt{CAS(blocks[n], null, blk)}}
 \EndFunction{TryAppend}
@@
 \begin{algorithm}
+\caption{Yet to decide how to handle.}
 \begin{algorithmic}[1]
 \setcounter{ALG@line}{800}
-\begin{multicols}{2}
 
-\Function{void}{Help}{}\cmt{Helps pending operations}
-\For{\tt{leaf l} \bf{in leaves}}
-\State{\tt{last= l.size-1}}
-\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
-\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
-\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
-\EndIf
-\EndFor
-\EndFunction{Help}
-\Statex
-
-\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
-\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
-\State \tt{t1,t2= RBT.split(order, s)}
-\State \tt{RBTRoot.CAS(t2.root)}
-\EndFunction{CollectGarbage}
-
-\Statex
-
-\Function{Block}{FindMostRecentDone}{b}
-\For{\tt{leaf l} \bf{in leaves}}
-\State\tt{max= Max(l.maxOld, max)}
-\EndFor
-\State\Return\tt{max} \cmt{This snapshot suffies.}
-\EndFunction{FindMostRecentDone}
-
-\Statex
+%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
+%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
+%\State \tt{t1,t2= RBT.split(order, s)}
+%\State \tt{RBTRoot.CAS(t2.root)}
+%\EndFunction{CollectGarbage}
 
@@
 
-\end{multicols}
 \end{algorithmic}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+                     %
+                     %
+                     %
+                     %
+                     %
+                     %
+                     %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+\section{Proof of Linearizability}
+
+\begin{definition}
+If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
+Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s.
+Block \nf{b} is propagated to node \nf{n} or set \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
+\end{definition}
+
+%\begin{definition}
+%  Block \nf{b} in node \nf{n} is new if \nf{b} is not a subblock of any block in \nf{n.parent.blocks[]}.
+%\end{definition}
+
+\begin{definition}
+  Block \nf{b} is is in the \emph{Established(n, t)} if \nf{b} is subblock of a \nf{b$^\prime$} such that \nf{n.size} at time $t$ is greater than index of \nf{b$^\prime$} .
+\end{definition}
+
+\begin{lemma}[sizeProgress] \label{lem::sizeProgress}
+\nf{n.size} is non-decreasing over time.
+\end{lemma}
+%\begin{proof}
+%  Simply because \nf{n.head} is only incremented.
+%\end{proof}
+
+\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}(\nf{h=n.size}) might be 1 bit behind the first empty block in the node.
+\end{lemma}
+%\begin{proof}
+% Because at the end of every \nf{Refresh()} with block size greater than 0 (Lines 53,56) \nf{n.head} is incremented.
+Maybe some process goes to sleep before incrementing the head, but after sleeping if \nf{h} does not increase then \nf{CAS} in Line 52 is going to be failed and nothing is going to be appended to \nf{n.blocks}.
+%\end{proof}
+
+%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
+%  If block \nf{b} in node \nf{n} is established, then all the blocks in \nf{n.blocks[]} before \nf{b} are established.
+If block \nf{b} in node \nf{n} is not new then all the blocks in \nf{n.blocks[]} before \nf{b} are not new.
+%\end{lemma}
+%\begin{proof}
+%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
+%\end{proof}
+
+\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
+  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
+\end{lemma}
+%\begin{proof}
+%Because blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing.
+%\end{proof}
+
+\begin{lemma}[createBlock] \label{lem::createBlock}
+  If \nf{n.CreateBlock(h, x)} is invoked at time $t$ and $b$ is the block returned, then $Established(n.end_{left}, t) \cup Established(n.end_{right}, t)/ Established(n, t) \subseteq b$ .
+\end{lemma}
+%\begin{proof}
+% We prove the claim for the left child.
+Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
+Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
+See Figure~\ref{fig::createBlock}.
+The right child is the same.
+%\end{proof}
+
+\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
+  Suppose \nf{Refresh(n)}'s \nf{TryAppend(new, s)} returns \nf{true}.
+Let $t$ be the time \nf{n.Refresh()} is invoked, blocks in $Established(n.end_{left}, t)$ and $Established(n.end_{right}, t)$ are in $Established(n, t)$ after \nf{TryAppend(new, s)}.
+\end{lemma}
+%\begin{proof}
+%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks before Line 43 which is  appended to \nf{n.blocks} by \nf{CAS} in Line 48.
+%\end{proof}
+
+\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
+    If instance $r$ of \nf{n.Refresh()} returns \nf{false}, then there is another successful instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)} with the same value \nf{s} read.
+\end{lemma}
+%\begin{proof}
+%  If there is no other concurrent successful \nf{Refresh(n)} then \nf{Refresh(n)} would succeed in Line 48.
+So there is another \nf{Refresh(n)}, that has to \nf{CAS}ed successfully its block in \nf{n.blocks[h]} after Line 43 of \nf{Refresh(n)}.
+Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
+%\end{proof}
+
+\begin{lemma}[Double Refresh] \label{doubleRefresh}
+  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process.
+Let $t_1$ be the time is $r_1$ invoked and $t_2$ be the time $r_2$ is invoked.
+After $r_2$'s \nf{TryAppend} we have $Established(n.end_{left}, t) \cup Established(n.end_{right}, t) \subseteq Established(n, t)$.
+\end{lemma}
+%\begin{proof}
+%  
+%
+%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
+If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
+$R_1^\prime$ may or may not have propagated some subblocks of \nf{new} to \nf{n}.
+It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
+If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
+If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}.
+Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$.
+So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$.
+This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
+See Figure~\ref{}.
+%\end{proof}
+
+%\begin{figure}[hbt]
+%  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
+%  \caption{$R_2^\prime$'s \nf{CAS} is executed after \nf{h1=n.head}.}
+%\end{figure}
+
+ Block \nf{new} is created of new established subblocks of children of \nf{n}(Lemma \ref{lem::createBlock}, Line 46).
+If \nf{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \nf{n}.
+
+\begin{lemma}[Double Refresh] \label{doublyRefresh}
+All operations in \nf{n}'s children's blocks before line \nf{35} are guaranteed to be in \nf{n}'s blocks after Line~\nf{37}.
+\end{lemma}
+%\begin{proof}
+%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
+By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}.
+\nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well.
+First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursivley.
+If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails.
+It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
+%\end{proof}
+\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
+If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
+So suppose two \nf{Refresh}es fail.
+Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}.
+So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
+After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
+
+\begin{lemma}[Append]\label{append}
+ When \text{\nf{Append(op)}} is finished, \nf{op} appears exactly once in a block of \nf{root.blocks}.
+\end{lemma}
+%\begin{proof}
+%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
+%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
+%\end{proof}
+
+\begin{lemma}[Block Size Upper Bound]\label{blockSize}
+Each block in a node contains at most one operation from each processs.  
+\end{lemma}
+%\begin{proof}
+%  Note that \nf{prevIndex} and \nf{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
+After a block has propagated to the parent the blocks between \nf{prevIndex} and \nf{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
+The number of new operations which have not propagated yet to the parent cannot be more than $p$.
+If so by the law of pigeonholes there is a process which has appended two concurrent operations.
+%\end{proof}
+
+\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
+Each block in a node  has at most $p$ subblocks.
+\end{lemma}
+%\begin{proof}
+%  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
+%\end{proof}
+
+\begin{definition} [Ordering of operations inside a node] \label{ordering}
+$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
+
+\begin{itemize}
+  \item $E(n,i)$ is the sequence of enqueue operations that are member of \nf{n.blocks[i]} ordered by process id.
+  \item 
+$D(n,i)$ is the sequence of dequeue operations that are member of \nf{n.blocks[i]} ordered by process id.
+\item $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+\end{itemize}
+\end{definition}
+
+\begin{theorem}
+The queue implementation is linearizable.
+\end{theorem}
+%\begin{proof}
+%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
+%  \begin{enumerate}
+%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
+The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
+%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
+\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
+It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
+By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
+If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
+%  \end{enumerate} 
+%\end{proof}
+
+\begin{lemma}[Get] \label{get}
+\nf{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
+\end{lemma}
+%\begin{proof}
+%It is obvious that \nf{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
+To find the \nf{i}th enqueue in block \nf{b} of an internal node \nf{n} in line 87 it is decided that it resides in the left child or the right child.
+This decision is made by Definition of $E(n,b)$.
+After that Lines 88, 92 search the proper subblocks of \nf{b}.
+From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \nf{prevBlock} and \nf{lastBlock} block of the \nf{CreateBlock()}.
+%\end{proof}
+
+\begin{lemma}[Index]
+ Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
+\end{lemma}
+%\begin{proof}
+%  \nf{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \nf{n} in \nf{n.parent} and then computes the order in $D(n.parent, superblock)$.
+Then calls \nf{index()} on \nf{n.parent} recursively.
+It is easy to see why the second is correct.
+Correctness of computing superblock comes from Lemma \ref{superBlock}.
+%\end{proof}
+
+\begin{lemma}[Computing SuperBlock]\label{superBlock}
+  If \nf{Index(n,b,i)} performs line 101, then \nf{superblock} contains \nf{i}th Dequeue in \nf{b}th block of node \nf{n}.
+\end{lemma}
+%\begin{proof}
+%\begin{enumerate}
+% \item Value read for \nf{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
+%  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.
+%
+% \item \nf{super[]} preserves order from child to parent; if in a child block \nf{b} is before \nf{c} then \nf{b.time} $\leq$ \nf{c.time} and \nf{super[b.time]} $\leq$ \nf{super[c.time]}\\$\blacktriangleright$
+%  Follows from the order of lines 60, 48, 49.
+%
+%% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
+%% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
+% \item \nf{super[i+1]-super[i]}$\leq p$
+%\\$\blacktriangleright$
+% In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
+Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set.
+If an instance of successful \nf{Refresh(n)} finishes \nf{super[i+1]} is set a new value and a block is added after \nf{n.parent[sup[i]]}.
+There could be at most $p$ successful unfinished concurrent instances of \nf{Refresh()} that have not reached line 49.
+So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
+%
+% \item Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
+%\\$\blacktriangleright$
+%\nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
+It is trivial to see that \nf{n.super} and \nf{n.b.counter} are increasing.
+\nf{super(b)} is the real superblock of b.
+\nf{super(t]} is the index of the superblock of the last block with time \nf{t}.
+If \nf{b.time} is \nf{t} we have:
+%$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+%
+%\end{enumerate}
+%\end{proof}
+
+\begin{lemma}[Computing Queue's Head] \label{computeHead}
+  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q. If Q is empty \nf{ComputeHead(i,b)} returns -1, otherwise
+ returns index in $E(root,b)$ of X.
+  \end{lemma}
+
+\begin{lemma}[Validity of \nf{head}]\label{head}
+No two blocks are written in the same index in \nf{n.blocks}.
+\end{lemma}
+%\begin{proof}
+%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read.
+If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}.
+If a process dies before line 51, another process will increment \nf{head} in line 54.
+%\end{proof}
+
+\begin{lemma}[Validity of \nf{super} and \nf{counter}]\label{superCounter}
+If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the superblock of a block with \nf{time=i}.
+\end{lemma}
+%\begin{proof}
+%After a successful CAS in line 46 \nf{super} and \nf{counter} are modified in both children.
+\nf{super[i]} is supposed to be the superblock of a block with \nf{time}=i and \nf{counter} is the timer in each node.
+\nf{super[i]} and \nf{counter=i} are expected to update after a bunch of blocks with \nf{time=i} have been aggregated together into a block in the parent.
+If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \nf{time}.
+We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
+If a process dies right after line 48, then \nf{counter} will remain the same and \nf{super[i]} is correct.
+Furthermore we are sure that when \nf{super[i]} is read it will not be \nf{null}.
+%\end{proof}
+
+\begin{lemma}[Search Ranges]\label{search}
+  Preconditions of all invocation of \nf{BSearch} are satisfied.
+\end{lemma}
+%\begin{proof}
+%  
+%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+%
+%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+%
+%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
+%
+%\end{proof}
+
 \end{document}

### 2022-05-13 article/psuedocode.tex
fix lemma claims

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
 \theoremstyle{definition}
@@
 \Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}}
-\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
+\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
+\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}\label{secondRefresh}
 \EndIf
@@
 \Function{boolean}{Refresh()}{}
-\State \tt{s= size}
+\State \tt{s= size} \label{readSize}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
@@
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
-\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
+\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
@@
 
-\begin{definition}
-If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
-Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s.
-Block \nf{b} is propagated to node \nf{n} or set \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
+\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.established here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
+Fix the logical order of definitions (cyclic refrences).
+
+\begin{definition}[Block]
+A block is an object that stores some statistics described in Algorithm Queue.
+It implicitly shows a set of operations.
+The set of operations of block $b$ are the operations in the leaf subblocks of $b$.
+We show the set of operations of block $b$, set of blocks $B$ by $ops(b)$, $ops(B)$.
+We also say $b$ contains $op$ if $op\in ops(b)$.
 \end{definition}
+
+\begin{definition}[Order]
+If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s. 
+\end{definition}
+
+\begin{definition}[Subblock]
+Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} or  is a subblock of a block in it.
+\end{definition}
+
+For simplicity we say block \nf{b} is propagated to node \nf{n} or set \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
 
@@
 \begin{definition}
-  Block \nf{b} is is in the \emph{Established(n, t)} if \nf{b} is subblock of a \nf{b$^\prime$} such that \nf{n.size} at time $t$ is greater than index of \nf{b$^\prime$} .
+  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$.
+Block \nf{b} in \nf{n.blocks} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
 \end{definition}
 
-\begin{lemma}[sizeProgress] \label{lem::sizeProgress}
-\nf{n.size} is non-decreasing over time.
+\begin{lemma}[headProgress] \label{lem::sizeProgress}
+\nf{n.established} is non-decreasing over time.
 \end{lemma}
@@
 
-\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}(\nf{h=n.size}) might be 1 bit behind the first empty block in the node.
+\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}(\nf{h=n.established}) might be 1 bit behind the first empty block in the node.
 \end{lemma}
-%\begin{proof}
-% Because at the end of every \nf{Refresh()} with block size greater than 0 (Lines 53,56) \nf{n.head} is incremented.
+\begin{proof}
+ Because at the end of every \nf{Refresh()} with block size greater than 0 (Lines 53,56) \nf{n.head} is incremented.
 Maybe some process goes to sleep before incrementing the head, but after sleeping if \nf{h} does not increase then \nf{CAS} in Line 52 is going to be failed and nothing is going to be appended to \nf{n.blocks}.
-%\end{proof}
+\end{proof}
 
@@
 \begin{lemma}[establishedOrder]\label{lem::establishedOrder}
-  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
+  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
 \end{lemma}
-%\begin{proof}
-%Because blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing.
-%\end{proof}
+\begin{proof}
+Because blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing.
+\end{proof}
 
 \begin{lemma}[createBlock] \label{lem::createBlock}
-  If \nf{n.CreateBlock(h, x)} is invoked at time $t$ and $b$ is the block returned, then $Established(n.end_{left}, t) \cup Established(n.end_{right}, t)/ Established(n, t) \subseteq b$ .
+  If \nf{b} is the block returned by \nf{n.CreateBlock(h, x)} invoked at time $t$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
 \end{lemma}
-%\begin{proof}
-% We prove the claim for the left child.
+\begin{proof}
+ We prove the claim for the left child.
 Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
@@
 The right child is the same.
-%\end{proof}
+\end{proof}
 
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Suppose \nf{Refresh(n)}'s \nf{TryAppend(new, s)} returns \nf{true}.
-Let $t$ be the time \nf{n.Refresh()} is invoked, blocks in $Established(n.end_{left}, t)$ and $Established(n.end_{right}, t)$ are in $Established(n, t)$ after \nf{TryAppend(new, s)}.
+  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated.
+Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
 \end{lemma}
-%\begin{proof}
-%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks before Line 43 which is  appended to \nf{n.blocks} by \nf{CAS} in Line 48.
-%\end{proof}
+\begin{proof}
+By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks before Line 43 which is  appended to \nf{n.blocks} by \nf{CAS} in Line 48.
+\end{proof}
 
 \begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-    If instance $r$ of \nf{n.Refresh()} returns \nf{false}, then there is another successful instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)} with the same value \nf{s} read.
+    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
+A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
 \end{lemma}
-%\begin{proof}
-%  If there is no other concurrent successful \nf{Refresh(n)} then \nf{Refresh(n)} would succeed in Line 48.
+\begin{proof}
+  If there is no other concurrent successful \nf{Refresh(n)} then \nf{Refresh(n)} would succeed in Line 48.
 So there is another \nf{Refresh(n)}, that has to \nf{CAS}ed successfully its block in \nf{n.blocks[h]} after Line 43 of \nf{Refresh(n)}.
 Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
-%\end{proof}
+\end{proof}
 
@@
   Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process.
-Let $t_1$ be the time is $r_1$ invoked and $t_2$ be the time $r_2$ is invoked.
-After $r_2$'s \nf{TryAppend} we have $Established(n.end_{left}, t) \cup Established(n.end_{right}, t) \subseteq Established(n, t)$.
+Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated.
+After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 \end{lemma}
-%\begin{proof}
-%  
-%
-%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
+\begin{proof}
+
+If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
 If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
@@
 See Figure~\ref{}.
-%\end{proof}
+\end{proof}
 
-%\begin{figure}[hbt]
-%  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
-%  \caption{$R_2^\prime$'s \nf{CAS} is executed after \nf{h1=n.head}.}
-%\end{figure}
+\begin{figure}[hbt]
+  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
+  \caption{$R_2^\prime$'s \nf{CAS} is executed after \nf{h1=n.head}.}
+\end{figure}
 
- Block \nf{new} is created of new established subblocks of children of \nf{n}(Lemma \ref{lem::createBlock}, Line 46).
+%  Block \nf{new} is created of new established subblocks of children of \nf{n}(Lemma \ref{lem::createBlock}, Line 46).
 If \nf{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \nf{n}.
 
-\begin{lemma}[Double Refresh] \label{doublyRefresh}
-All operations in \nf{n}'s children's blocks before line \nf{35} are guaranteed to be in \nf{n}'s blocks after Line~\nf{37}.
-\end{lemma}
-%\begin{proof}
-%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
+\begin{corollary}[Propagate Step] \label{doublyRefresh}
+All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s blocks after line~\ref{secondRefresh}.
+\end{corollary}
+\begin{proof}
+Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
 By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}.
 \nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well.
-First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursivley.
+First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively.
 If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails.
 It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
-%\end{proof}
+\end{proof}
 \nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
@@
 
-\begin{lemma}[Append]\label{append}
- When \text{\nf{Append(op)}} is finished, \nf{op} appears exactly once in a block of \nf{root.blocks}.
+\begin{lemma}[No Duplicates]\label{append}
+If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
 \end{lemma}
-%\begin{proof}
-%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
-%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
-%\end{proof}
+\begin{proof}
+\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
+By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
+\end{proof}
+\begin{corollary}
+  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
+\end{corollary}
 
 \begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block in a node contains at most one operation from each processs.  
+Each block contains at most one operation from each processs \nf{($\forall$ \nf{process p:}\#\nf{operations of p}$\in$\nf{ops(n.blocks[x]$\leq$1})}.
 \end{lemma}
@@
 \begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
-Each block in a node  has at most $p$ subblocks.
+Each block has at most $p$ subblocks.
 \end{lemma}
@@
 
-\begin{definition} [Ordering of operations inside a node] \label{ordering}
+\pagebreak
+\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
 $\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
@@
 \begin{itemize}
-  \item $E(n,i)$ is the sequence of enqueue operations that are member of \nf{n.blocks[i]} ordered by process id.
-  \item 
-$D(n,i)$ is the sequence of dequeue operations that are member of \nf{n.blocks[i]} ordered by process id.
-\item $D(n)=D(n,1).D(n,2).D(n,3)...$
-\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+\item We call operations before $op$ in the sequence of operations $S$, prefix of the $op$.
+  \item $E(n,i)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
+  \item $D(n,i)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
+\item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
+\item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
 \end{itemize}
 \end{definition}
+\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on deqeueus.}
 
@@
 
-\begin{lemma}[Get] \label{get}
-\nf{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
+\begin{lemma}[Get corretness] \label{get}
+\nf{n.GetENQ(b,i)} returns the $i$th Enqueue in $E(n,b)$.
 \end{lemma}
@@
 
-\begin{lemma}[Index]
- Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
+\begin{lemma}[Index correctness]
+ \nf{n.Index(b,i)} returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
 \end{lemma}
@@
 \begin{lemma}[Computing SuperBlock]\label{superBlock}
-  If \nf{Index(n,b,i)} performs line 101, then \nf{superblock} contains \nf{i}th Dequeue in \nf{b}th block of node \nf{n}.
+After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{superblock} contains the \nf{i}th dequeue in the \nf{b}th block of the node \nf{n}.
 \end{lemma}
@@
 \begin{lemma}[Computing Queue's Head] \label{computeHead}
-  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q. If Q is empty \nf{ComputeHead(i,b)} returns -1, otherwise
- returns index in $E(root,b)$ of X.
+  Let S be the state of an empty queue if the operations in prefix of $i$th dequeue in $D(root,b)$ are applied on it.
+\nf{FindResponse()} returns the index of the enqueue that is the head in S.
+If the  queue is empty in S it returns \nf{<-1,-->}.
   \end{lemma}
@@
 \begin{lemma}[Validity of \nf{head}]\label{head}
-No two blocks are written in the same index in \nf{n.blocks}.
+If block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} is going to remain b (not overwritten in future).
 \end{lemma}
@@
 \begin{lemma}[Validity of \nf{super} and \nf{counter}]\label{superCounter}
-If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the superblock of a block with \nf{time=i}.
+If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the index of the superblock of a block with \nf{time=i}$\pm$\nf{p}.
 \end{lemma}
@@
 
+\begin{lemma}[help]\label{help}
+After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone} are done.
+\end{lemma}
+
+\begin{lemma}[Root search range]\label{rootRange}
+  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the length of the queue.
+\end{lemma}
+
+\framebox[1.1\width]{TODO} Fallback safety lemmas.
+
 \end{document}

### 2022-05-13 article/psuedocode.tex
add proof for some of the first lemmas

@@
 \EndFor
-\State \tt{\Call{CAS}{size, h, h+1}}
+\State \tt{\Call{CAS}{size, h, h+1}} \label{incrementHead1}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
+\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
 \State \Return{ \tt{false}}
@@
 
+\framebox[1.1\width]{TODO} Fallback safety lemmas. Some parts are obsolete.
+
+\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
+Is lemma 7 necessary?
+Is lemma 13 induced trivially from lemma 8?
+
 \begin{definition}[Block]
@@
 \end{lemma}
-%\begin{proof}
-%  Simply because \nf{n.head} is only incremented.
-%\end{proof}
+\begin{proof}
+  Induced trivially from the pseudocode since \nf{n.established} is only incremented.
+\end{proof}
 
-\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line \ref{prevLine}(\nf{h=n.established}) might be 1 bit behind the first empty block in the node.
+\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.established}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
 \end{lemma}
 \begin{proof}
- Because at the end of every \nf{Refresh()} with block size greater than 0 (Lines 53,56) \nf{n.head} is incremented.
-Maybe some process goes to sleep before incrementing the head, but after sleeping if \nf{h} does not increase then \nf{CAS} in Line 52 is going to be failed and nothing is going to be appended to \nf{n.blocks}.
+At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
+If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
 \end{proof}
@@
 \begin{proof}
-Because blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing.
+Blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node grows.
 \end{proof}
@@
 \begin{proof}
- We prove the claim for the left child.
+ WLOG we prove the claim for the left child.
 Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
-Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
+Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$.
 See Figure~\ref{fig::createBlock}.
@@
 \begin{proof}
-By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks before Line 43 which is  appended to \nf{n.blocks} by \nf{CAS} in Line 48.
+By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
 \end{proof}
@@
 \begin{proof}
-  If there is no other concurrent successful \nf{Refresh(n)} then \nf{Refresh(n)} would succeed in Line 48.
-So there is another \nf{Refresh(n)}, that has to \nf{CAS}ed successfully its block in \nf{n.blocks[h]} after Line 43 of \nf{Refresh(n)}.
-Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
+  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313.
+So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}.
+%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
 \end{proof}
@@
 \end{lemma}
-\begin{proof}
-
-If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
+%\begin{proof}
+%  
+%
+%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
 If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
@@
 See Figure~\ref{}.
-\end{proof}
+%\end{proof}
 
@@
 \end{corollary}
-\begin{proof}
-Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
+%\begin{proof}
+%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
 By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}.
@@
 It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
-\end{proof}
+%\end{proof}
+
 \nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
@@
 \end{lemma}
-\begin{proof}
-\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
-By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
-\end{proof}
+%\begin{proof}
+%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
+%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
+%\end{proof}
+
 \begin{corollary}
@@
 
-\framebox[1.1\width]{TODO} Fallback safety lemmas.
-
 \end{document}

### 2022-05-18 article/psuedocode.tex
apply may 13 notes

@@
 \newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{observation}[theorem]{Observation}
 \theoremstyle{definition}
@@
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
+\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
@@
 \EndFor
-\State \tt{\Call{CAS}{size, h, h+1}} \label{incrementHead1}
+\State \tt{\Call{CAS}{size, s, s+1}} \label{incrementHead1}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
+\State \tt{\Call{CAS}{size, s, s+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
 \State \Return{ \tt{false}}
@@
 
+\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?
+
 \begin{definition}[Block]
@@
 \begin{definition}[Subblock]
-Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} or  is a subblock of a block in it.
+Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
 \end{definition}
 
-For simplicity we say block \nf{b} is propagated to node \nf{n} or set \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
+For simplicity we say block \nf{b} is propagated to node \nf{n} or to a set of blocks \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
 
@@
   Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$.
-Block \nf{b} in \nf{n.blocks} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
+Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
 \end{definition}
+
+\begin{observation}[Validity of \nf{head}]\label{head}
+Once block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
+\end{observation}
+
+\begin{lemma}
+  Each block is not duplicated.
+\end{lemma}
 
@@
 \begin{lemma}[createBlock] \label{lem::createBlock}
-  If \nf{b} is the block returned by \nf{n.CreateBlock(h, x)} invoked at time $t$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
+  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $b$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
 \end{lemma}
@@
  WLOG we prove the claim for the left child.
-Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
+Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the established operations at time $t$ by definition of subblock.
 Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$.
@@
 \begin{corollary}[Propagate Step] \label{doublyRefresh}
-All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s blocks after line~\ref{secondRefresh}.
+All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
 \end{corollary}
@@
 \begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
-Each block has at most $p$ subblocks.
+Each block has at most $p$ direct subblocks.
 \end{lemma}
@@
 \begin{itemize}
-\item We call operations before $op$ in the sequence of operations $S$, prefix of the $op$.
+\item We call operations strictly before $op$ in the sequence of operations $S$, prefix of the $op$.
   \item $E(n,i)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
@@
 \end{definition}
-\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on deqeueus.}
+\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on dequeues.}
 
-\begin{theorem}
-The queue implementation is linearizable.
-\end{theorem}
+\begin{lemma}[Search Ranges]\label{search}
+  Preconditions of all invocation of \nf{BSearch} are satisfied.
+\end{lemma}
 %\begin{proof}
-%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
-%  \begin{enumerate}
-%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
-The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
-%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
-\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
-It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
-By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
-If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
-%  \end{enumerate} 
+%  
+%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+%
+%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+%
+%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
+%
 %\end{proof}
 
-\begin{lemma}[Get corretness] \label{get}
+\begin{lemma}[Get correctness] \label{get}
 \nf{n.GetENQ(b,i)} returns the $i$th Enqueue in $E(n,b)$.
@@
 
+\begin{lemma}[help]\label{help}
+After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
+\end{lemma}
+
 \begin{lemma}[Index correctness]
- \nf{n.Index(b,i)} returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
+ \nf{n.Index(b,i)} returns the rank in $D(root)$ of the $i$th Dequeue in $D(n,b)$.
 \end{lemma}
@@
 \begin{lemma}[Computing Queue's Head] \label{computeHead}
-  Let S be the state of an empty queue if the operations in prefix of $i$th dequeue in $D(root,b)$ are applied on it.
-\nf{FindResponse()} returns the index of the enqueue that is the head in S.
+  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it.
+\nf{FindResponse()} returns the index in $E(root,b)$ of the enqueue that is the head in S.
 If the  queue is empty in S it returns \nf{<-1,-->}.
@@
 
-\begin{lemma}[Validity of \nf{head}]\label{head}
-If block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} is going to remain b (not overwritten in future).
-\end{lemma}
 %\begin{proof}
@@
 
-\begin{lemma}[Search Ranges]\label{search}
-  Preconditions of all invocation of \nf{BSearch} are satisfied.
-\end{lemma}
-%\begin{proof}
-%  
-%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-%
-%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-%
-%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
-%
-%\end{proof}
-
-\begin{lemma}[help]\label{help}
-After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone} are done.
-\end{lemma}
-
 \begin{lemma}[Root search range]\label{rootRange}
@@
 
+\begin{theorem}
+The queue implementation is linearizable.
+\end{theorem}
+%\begin{proof}
+%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
+%  \begin{enumerate}
+%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
+The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
+%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
+\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
+It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
+By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
+If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
+%  \end{enumerate} 
+%\end{proof}
+
 \end{document}

### 2022-05-20 article/psuedocode.tex
polish and add proof for some lemmas

@@
 
+\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
+block b $\in$ block c means b is subblock of c.
+block b $\in$ set B means b is in B.
+Merge these two to have shorter formulaes.
+
 \begin{definition}[Block]
-A block is an object that stores some statistics described in Algorithm Queue.
+A block is an object storing some statistics described in Algorithm Queue.
 It implicitly shows a set of operations.
@@
 \end{lemma}
+\begin{proof}
+  \nf{n.CreateBlock(i)} is successful for \nf{i=1,2,3,...} exactly one time and by definition of \nf{CreateBlock} the range of their blocks are disjoint so in propagation procedure on block can not be propagated twice (reference to propagation).
+\end{proof}
 
@@
 \begin{lemma}[createBlock] \label{lem::createBlock}
-  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $b$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(b)}.
+  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
 \end{lemma}
 \begin{proof}
- WLOG we prove the claim for the left child.
-Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the established operations at time $t$ by definition of subblock.
+From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
+We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}.
+It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$.
+If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see.
+Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it.
+Right child is the same.
+% WLOG we prove the claim for the left child.
+Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.
 Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$.
@@
 \begin{proof}
+Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$.
+By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}.
+Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10.
 By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
@@
 \end{lemma}
-%\begin{proof}
-%  
-%
-%If Line 35 (first Refresh which we call $R_1$) returns \nf{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \nf{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
-$R_1^\prime$ may or may not have propagated some subblocks of \nf{new} to \nf{n}.
-It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
+\begin{proof}
+
+If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
+If not, then there is another successful instance of \nf{n.Refresh()} $r_1^\prime$ by Lemma~\ref{lem:falseRefresh} for $r_2$.
+From $t_{i_{r_2}^\prime} > t_1$ and Lemma~\ref{lem::trueRefresh}, \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+%It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
 If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
@@
 See Figure~\ref{}.
-%\end{proof}
+\end{proof}
 
@@
 \end{corollary}
-%\begin{proof}
-%Suppose block \nf{b} with index \nf{i} is in the the left child of \nf{n} before the line 35.
-By Lemma~\ref{head} it follows that \nf{n.left.head} is greater than \nf{i}.
-\nf{Refresh()} calls \nf{CreateBlock()} and creates a block from blocks between \nf{n.blocks[n.head].end\textsubscript{left}} and \nf{n.left.head} in the left child, which contains \nf{b} as well.
-First it tries to append it in \nf{n.blocks.head} and if it was succsuful it continues recursively.
-If not it tries again, and if the second call of \nf{Refresh()} in Line 36 fails.
-It means there is another \nf{Refresh} which has reah its \nf{i} after the Line 35, so it contains \nf{b} as well.
-%\end{proof}
+\begin{proof}
+Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
+\end{proof}
 
@@
 
-\pagebreak
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
@@
 
-\begin{theorem}
+\begin{theorem}[Main]
 The queue implementation is linearizable.
@@
 
+\begin{lemma}[Time analysis]
+  \nf{n.GetEnq(b,i), n.Index(b,i)} take $O(\log^2 p)$ steps.
+Search in the root may take $O(\log Q+ p^2)$ steps.
+Helping is done every $p^2$ block appended to the root and takes $p\times \log^2p$ steps.
+Amortized time consumed for helping by each process is $O(\log^2 p)$.
+\end{lemma}
+
 \end{document}

### 2022-05-25 article/psuedocode.tex
apply may 20 feedback

@@
 
-\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.established here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
+\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
 Fix the logical order of definitions (cyclic refrences).
@@
 \framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
-block b $\in$ block c means b is subblock of c.
+block b $\sqsubseteq$ block c means b is subblock of c.
 block b $\in$ set B means b is in B.
@@
 Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
+Block b is direct superblock of block c if c is direct subblock of b.
 \end{definition}
@@
 \begin{definition}
-  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$.
+  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.head} is greater than index of \nf{b} at time $t$.
 Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
@@
 
-\begin{lemma}
-  Each block is not duplicated.
+\begin{lemma}[headProgress] \label{lem::sizeProgress}
+\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} is non-decreasing over \nf{i}.
 \end{lemma}
 \begin{proof}
-  \nf{n.CreateBlock(i)} is successful for \nf{i=1,2,3,...} exactly one time and by definition of \nf{CreateBlock} the range of their blocks are disjoint so in propagation procedure on block can not be propagated twice (reference to propagation).
+  Induced trivially from the pseudocode since \nf{n.head} is only incremented.
+\nf{n.blocks[i].
+Also end\sub{left, right}} are greater than or equal the previous values in the \nf{CreateBlock(i)} code (Lines~\ref{lastLine, prevLine}).
 \end{proof}
 
-\begin{lemma}[headProgress] \label{lem::sizeProgress}
-\nf{n.established} is non-decreasing over time.
+\begin{lemma}
+Every block has most one direct superblock.
 \end{lemma}
 \begin{proof}
-  Induced trivially from the pseudocode since \nf{n.established} is only incremented.
+To show this we are going to refer to the way n.blocks[] is partitioned while propagating up to n.parent.
+\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
+From we know that \nf{end\sub{left},end\sub{right}} are non decreasing, so the range of the subblocks of n.blocks are disjoint.
 \end{proof}
 
-\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.established}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
+\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.head}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
 \end{lemma}
@@
 
-\begin{lemma}[createBlock] \label{lem::createBlock}
-  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
-\end{lemma}
-\begin{proof}
-From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
-We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}.
+%\begin{lemma}[createBlock] \label{lem::createBlock}
+%  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
+%\end{lemma}
+%\begin{proof}
+%From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
+%We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}.
 It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$.
@@
 Right child is the same.
-% WLOG we prove the claim for the left child.
+%% WLOG we prove the claim for the left child.
 Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.
@@
 The right child is the same.
-\end{proof}
+%\end{proof}
 
@@
   Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated.
-Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
+Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
 \end{lemma}
 \begin{proof}
-Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$.
+From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
+So it remains to show \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})} which we call new operations $\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
+If \nf{TryAppend}returns true a block is appended to n.
+From the code of the \nf{CreatBlock} this block includes the establsihed blocks in n's children at $t_i$.
+Since the head in createblock is read after $t_i$.
+So the new operations are inthe block appended to $n$.
+
+%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$.
 By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}.
 Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10.
-By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
+%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
 \end{proof}
 
-\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
+%\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
+%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
 A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
-\end{lemma}
-\begin{proof}
-  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313.
+%\end{lemma}
+%\begin{proof}
+%  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313.
 So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}.
-%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
-\end{proof}
+%%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
+%\end{proof}
 
@@
 If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \nf{n.Refresh()} $r_1^\prime$ by Lemma~\ref{lem:falseRefresh} for $r_2$.
-From $t_{i_{r_2}^\prime} > t_1$ and Lemma~\ref{lem::trueRefresh}, \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+If not, then there is another successful instance of \nf{n.Refresh()} $r_2^\prime$ which has did TryAppend successfully into \nf{n.blocks[i+1]}.
+In figure 1 we see why the block it is appending contains established block in the n's children at $t_i$, since it create a block reading the head after $t_1$.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
-  \caption{$R_2^\prime$'s \nf{CAS} is executed after \nf{h1=n.head}.}
+  \center\includegraphics[width=6in]{pics/completedouble.png}
+  \caption{$t_t$ is after $R_2^\prime$'s reads n.head after n.head is incremented which is after the succesufil TryAppend(i) which is after $R_1$'s reading n.head that takes place at $t_i$.}
 \end{figure}

### 2022-05-30 article/psuedocode.tex
fix invariant lemma, simplify the double refresh pic

@@
 \begin{itemize}
-  \item \tt{b\sub{op}}: index of the block containing operation\tt{op}
+  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
   \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
@@
 
-\begin{observation}[Validity of \nf{head}]\label{head}
+\begin{observation}\label{head}
 Once block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
@@
 \begin{lemma}[headProgress] \label{lem::sizeProgress}
-\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} is non-decreasing over \nf{i}.
+\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} $\geq$ \nf{n.blocks[i-1].end\sub{left, right}}.
 \end{lemma}
@@
 
-\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.head}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
+\begin{lemma}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
 \end{lemma}
 \begin{proof}
-At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
+The value of \nf{n.head} is modified only in lines \ref{incrementHead1, incrementHead2} and a value is written into \nf{n.blocks[head]} is only in Line 313.
+It's obvious that writing into \nf{n.blocks[head]} does not change the state of the claim of the lemma and if it was right before 312 it is going to be right after 312.
+We show in both cases of the successful 313 or not the claim holds after the increment lines of \nf{n.head}.
+In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
+If the claim was held before it is going to be held after the \ref{incrementHead1}.
+Since wether\ref{incrementHead1} returns true or false it means \nf{n.head} has been incremented.
+The failiure case is also the same since it means \nf{n.blocks[head]} is written by some process.
+%At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
 If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
@@
 \begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/completedouble.png}
-  \caption{$t_t$ is after $R_2^\prime$'s reads n.head after n.head is incremented which is after the succesufil TryAppend(i) which is after $R_1$'s reading n.head that takes place at $t_i$.}
+  \center\includegraphics[width=6in]{pics/compactdouble.png}
+  \caption{$t1<$ $r_1$ reading head $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $R_2^\prime$ reading head $<$ \nf{TryAppend(i+1)} $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t2$}
 \end{figure}

### 2022-06-02 article/psuedocode.tex
complete some lemmas proofs

@@
 \newtheorem{definition}[theorem]{Definition}
+\newtheorem{invariant}[theorem]{Invariant}
 
@@
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{\tt{blocks.tryAppend(new, s)}}
+\ElsIf{\tt{blocks.tryAppend(new, s)}} \label{cas}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
@@
 
-\begin{lemma}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
-\end{lemma}
+\begin{corollary}[No Duplicates]\label{append}
+If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+\end{corollary}
+%\begin{proof}
+%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
+%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
+%\end{proof}
+
+\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
+\end{invariant}
 \begin{proof}
-The value of \nf{n.head} is modified only in lines \ref{incrementHead1, incrementHead2} and a value is written into \nf{n.blocks[head]} is only in Line 313.
-It's obvious that writing into \nf{n.blocks[head]} does not change the state of the claim of the lemma and if it was right before 312 it is going to be right after 312.
-We show in both cases of the successful 313 or not the claim holds after the increment lines of \nf{n.head}.
+Some value is written into \nf{n.blocks[head]} only in Line 313.
+It is obvious that writing into \nf{n.blocks[head]} does not change the state of the claim of the lemma.
+The value of \nf{n.head} is modified only in lines \ref{incrementHead1}, ][p\ref{incrementHead2}, we show in both cases of the successful \ref{cas} or not the claim holds after the increment lines of \nf{n.head}.
+If $h$ increments to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to claim the invariant still holds.
 In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
-If the claim was held before it is going to be held after the \ref{incrementHead1}.
-Since wether\ref{incrementHead1} returns true or false it means \nf{n.head} has been incremented.
-The failiure case is also the same since it means \nf{n.blocks[head]} is written by some process.
+Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readSize}.
+The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
 %At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
@@
   Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process.
-Let $t_1$ be the time $r_1$ is invoked and $t_2$ be the time $r_2$ terminated.
-After $r_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+Let $t_1$ be the time $R_1$ is invoked and $t_2$ be the time $R_2$ terminated.
+After $R_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 \end{lemma}
@@
 
-If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \nf{n.Refresh()} $r_2^\prime$ which has did TryAppend successfully into \nf{n.blocks[i+1]}.
-In figure 1 we see why the block it is appending contains established block in the n's children at $t_i$, since it create a block reading the head after $t_1$.
+If Line \ref{cas} of $R_1$ or $R_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
+If not, then there is another successful instance of \nf{n.Refresh()} $R_2^\prime$ which has did TryAppend successfully into \nf{n.blocks[i+1]}.
+Note that if $R_2$ reads some value greater than $i+1$ in Line \ref{readSize} it means there is a succesful \nf{Refresh()} instance started after Line \ref{readSize} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readSize} of $R_2$, form Lemma~\ref{lem::trueRefresh} by finish of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
+In the figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_i$, since it create a block reading the head after $t_1$.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
 
+\begin{corollary}
+  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
+\end{corollary}
+\begin{proof}
+  Follows from Lemma \ref{doubleRefresh}, \ref{append}.
+\end{proof}
+
 \nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
@@
 
-\begin{lemma}[No Duplicates]\label{append}
-If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+\begin{lemma}[Block Size Upper Bound]\label{blockSize}
+Each block contains at most one operation from each processs \nf{($\forall$ \nf{process p, block b:}\#\nf{operations of p}$\in$\nf{ops(b)$\leq$1}}.
 \end{lemma}
-%\begin{proof}
-%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
-%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
-%\end{proof}
-
-\begin{corollary}
-  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
-\end{corollary}
-
-\begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block contains at most one operation from each processs \nf{($\forall$ \nf{process p:}\#\nf{operations of p}$\in$\nf{ops(n.blocks[x]$\leq$1})}.
-\end{lemma}
-%\begin{proof}
-%  Note that \nf{prevIndex} and \nf{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
-After a block has propagated to the parent the blocks between \nf{prevIndex} and \nf{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
-The number of new operations which have not propagated yet to the parent cannot be more than $p$.
-If so by the law of pigeonholes there is a process which has appended two concurrent operations.
-%\end{proof}
+\begin{proof}
+Assume there are two operations $op_1, op_2$ from process $p$ in block $b$.
+A process cannot invoke two operations concurrently.
+WLOG assume \nf{leaf of p.Append(block containing op\sub{1})} has to be finished before \nf{leaf of p.Append(block containing op\sub{2})} starts.
+So before appending $op_2$ to the tree $op_1$ exists in every node.
+So there is a node that has two blocks containing $op_1$ since if not then $b$ cannot exist.
+This contradicts with \ref{append}.
+\end{proof}
+\textit{Is something wrong with this proof? It is more complicated than I thought.}
 
@@
 \end{lemma}
-%\begin{proof}
-%  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
-%\end{proof}
+\begin{proof}
+  It follows directly from Lemma \ref{blockSize} and the observation that each block contains at least one operation, induced from Line~\ref{addOP}. 
+\end{proof}
 
@@
 \item We call operations strictly before $op$ in the sequence of operations $S$, prefix of the $op$.
-  \item $E(n,i)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
-  \item $D(n,i)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[i])} ordered by their process id.
+  \item $E(n,b)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
+  \item $E(n,b,i)$ is the $i$th enqueue in $E(n,b)$.
+  \item $D(n,b)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
+  \item $D(n,b,i)$ is the $i$th enqueue in $D(n,b)$.
 \item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
@@
 \end{definition}
-\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only works on enqueues and \nf{IndexDEQ()} works on dequeues.}
-
-\begin{lemma}[Search Ranges]\label{search}
-  Preconditions of all invocation of \nf{BSearch} are satisfied.
-\end{lemma}
-%\begin{proof}
-%  
-%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-%
-%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-%
-%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
-%
-%\end{proof}
+\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}
 
@@
 \end{lemma}
-%\begin{proof}
-%It is obvious that \nf{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
-To find the \nf{i}th enqueue in block \nf{b} of an internal node \nf{n} in line 87 it is decided that it resides in the left child or the right child.
-This decision is made by Definition of $E(n,b)$.
+\begin{proof}
+It is easy to see that \nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
+In Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}.
+Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked.
+This recursive procedure ends in a leaf node.
 After that Lines 88, 92 search the proper subblocks of \nf{b}.
-From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \nf{prevBlock} and \nf{lastBlock} block of the \nf{CreateBlock()}.
-%\end{proof}
+The correctness of the invocation parameters is induced from the definition of ops(block), subblock.
+\end{proof}
+\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
+
+\begin{definition}
+An enqueue operation is \textit{finished} if its argument is returned by some process.
+A dequeue operation is \texttt{finished} if it returns \nf{null} or some value.
+Block \nf{b} is \textit{done} if all operations in \nf{ops(b)} are finished.
+\end{definition}
+\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
 
@@
 
-\begin{lemma}[Index correctness]
- \nf{n.Index(b,i)} returns the rank in $D(root)$ of the $i$th Dequeue in $D(n,b)$.
+\begin{lemma}[Computing SuperBlock]\label{superBlock}
+After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{n.parent.blocks[superblock]} contains $D(n,b,i)$.
 \end{lemma}
-%\begin{proof}
-%  \nf{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \nf{n} in \nf{n.parent} and then computes the order in $D(n.parent, superblock)$.
-Then calls \nf{index()} on \nf{n.parent} recursively.
-It is easy to see why the second is correct.
-Correctness of computing superblock comes from Lemma \ref{superBlock}.
-%\end{proof}
+\begin{proof}
+\begin{enumerate}
+ \item Value read for \nf{super[b.group]} in line 418 is not null.\\$\blacktriangleright$
+  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.
 
-\begin{lemma}[Computing SuperBlock]\label{superBlock}
-After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{superblock} contains the \nf{i}th dequeue in the \nf{b}th block of the node \nf{n}.
-\end{lemma}
-%\begin{proof}
-%\begin{enumerate}
-% \item Value read for \nf{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
-%  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.
-%
-% \item \nf{super[]} preserves order from child to parent; if in a child block \nf{b} is before \nf{c} then \nf{b.time} $\leq$ \nf{c.time} and \nf{super[b.time]} $\leq$ \nf{super[c.time]}\\$\blacktriangleright$
-%  Follows from the order of lines 60, 48, 49.
-%
-%% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-%% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-% \item \nf{super[i+1]-super[i]}$\leq p$
-%\\$\blacktriangleright$
-% In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
+ \item \nf{super[]} preserves order from child to parent; if in a child block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group} and \nf{super[b.group]} $\leq$ \nf{super[c.group]}\\$\blacktriangleright$
+  Follows from the order of lines 60, 48, 49.
+
+% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
+% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
+ \item \nf{super[i+1]-super[i]}$\leq p$
+\\$\blacktriangleright$
+ In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
 Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set.
@@
 So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
-%
-% \item Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
-%\\$\blacktriangleright$
-%\nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
+
+ \item Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
+\\$\blacktriangleright$
+\nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
 It is trivial to see that \nf{n.super} and \nf{n.b.counter} are increasing.
@@
 If \nf{b.time} is \nf{t} we have:
-%$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-%
-%\end{enumerate}
-%\end{proof}
+$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+
+\end{enumerate}
+\end{proof}
+
+\begin{lemma}[Index correctness]
+ \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D(n,b,i)$.
+\end{lemma}
+\begin{proof}
+We can see the base case \nf{root.IndexDEQ(b,i)} is trivial.
+  \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent}(Lemma \ref{superBlock}).
+Then the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
+It is easy to see why the second is correct.
+Correctness of computing superblock comes from Lemma \ref{superBlock}.
+\end{proof}
+\textit{Do I need to talk about the computation of the order in the parent which is based on the definition of ordering of dequeues in a block?}
+
+\begin{lemma}[Search Ranges]\label{search}
+  Preconditions of all invocation of \nf{BSearch} are satisfied.
+\end{lemma}
+\begin{proof}
+
+Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+
+Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+
+Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
+
+\end{proof}
 

### 2022-06-06 article/psuedocode.tex
apply jun 2 feedbacks

@@
 \begin{proof}
+Af first the invariant is true since initially 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}.
+The satisfiability of the invariant may be effected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
+
 Some value is written into \nf{n.blocks[head]} only in Line 313.
@@
 %\end{proof}
+
+\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
+If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
+So suppose two \nf{Refresh}es fail.
+Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}.
+So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
+After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
 
@@
 
-\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
-If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
-So suppose two \nf{Refresh}es fail.
-Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}.
-So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
-After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-
 \begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block contains at most one operation from each processs \nf{($\forall$ \nf{process p, block b:}\#\nf{operations of p}$\in$\nf{ops(b)$\leq$1}}.
+Each block contains at most one operation from each processs.
 \end{lemma}
 \begin{proof}
-Assume there are two operations $op_1, op_2$ from process $p$ in block $b$.
-A process cannot invoke two operations concurrently.
-WLOG assume \nf{leaf of p.Append(block containing op\sub{1})} has to be finished before \nf{leaf of p.Append(block containing op\sub{2})} starts.
-So before appending $op_2$ to the tree $op_1$ exists in every node.
-So there is a node that has two blocks containing $op_1$ since if not then $b$ cannot exist.
-This contradicts with \ref{append}.
+By proof of contradiction, assume there are more than one operation from process $p$ in block $b$ in node $n$.
+A process cannot invoke more than one operations concurrently.
+From $p$ 's operations in $b$, let $op_1$ be the first operation invoked and $op_2$ be the second one.
+Note that it is terminated before $op_2$ started.
+So before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
+So there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
+$op_1$ existing in $b$ an $b^\prime$ contradicts with \ref{append}.
 \end{proof}
-\textit{Is something wrong with this proof? It is more complicated than I thought.}
 
@@
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
+$\blacktriangleright$ Note that processes are numbered from 1 to $p$, left to right in the leaves of the tree and from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
 
@@
 \begin{lemma}[Get correctness] \label{get}
-\nf{n.GetENQ(b,i)} returns the $i$th Enqueue in $E(n,b)$.
+If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E(n,b,i)$.
 \end{lemma}
 \begin{proof}
-It is easy to see that \nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
-In Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}.
-Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked.
-This recursive procedure ends in a leaf node.
-After that Lines 88, 92 search the proper subblocks of \nf{b}.
-The correctness of the invocation parameters is induced from the definition of ops(block), subblock.
+We are going to prove this lemma by induction on the height of the tree.
+The base case for the leaves of the tree is pretty straight forward.
+Since leaf blocks contain exactly one operation then only \nf{GetENQ(b,1)} can be called on leaves.
+\nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
+For non leaf nodes in Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}.
+By definition of $E(n,b)$ operations from the left child come  before the operations of the right child.
+Having \nf{sum\sub{enq}}, the prefix sum of the number of enqueues we can compute the direct subblock containing the enqueue we are finding for with binary search.
+Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked which returns the correct operation by the hypothesis of the induction.
 \end{proof}

### 2022-06-16 article/psuedocode.tex
add lemmas, remove helping, apply comments

@@
 \begin{algorithm}
-\caption{Fields description}
+\caption{Tree Fields Description}
 \begin{algorithmic}[1]
@@
 
-\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
-\begin{itemize}
-
-  \item \tt{\sl{int} \tt{last\sub{done}}}
-
-  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
+%\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
+%\begin{itemize}
+%  
+%  \item \tt{\sl{int} \tt{last\sub{done}}}
+%
+%  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
 A block is finished if all of its operations are finished.
@@
 \it{put the definitions before the pseudocode}}
-
-\end{itemize}
-\pagebreak
+%  
+%\end{itemize}
 
@@
 \end{itemize}
+
+\pagebreak
 
@@
 %  \textsf{: count of non-null dequeus up to this block}
-  \item \tt{\sl{counter} num\sub{finished}}
-  \textsf{: number of finished operations in the block}
+%  \item \tt{\sl{counter} num\sub{finished}}
+%  \textsf{: number of finished operations in the block}
 %    \item \tt{\sl{int} order}
@@
 \end{multicols}
+\end{algorithmic}
+\end{algorithm}
+
+%##########################################
+
+\begin{algorithm}
+\caption{BlockList}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{700}
+
+\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
+
+%\Statex
+%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
+%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
+%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
+%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
+%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
+%\State \tt{l.Help()}
+%\EndFor
+%\EndIf
+%\State \tt{blk.num\sub{finished}= 0}
+%%\State \tt{*oldRoot= \&root.blocks.root}
+%%\State \tt{*newRoot= root.blocks.Append(blk).root}
+%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
+%\State \Return{\tt{CAS(blocks[n], null, blk)}}
+%\EndFunction{TryAppend}
+
+\Statex
+\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
+\Statex \tt{blocks[]: }\sf{array of blocks}
+\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
+\State \Return{\tt{CAS(blocks[n], null, blk)}}
+\EndFunction{TryAppend}
+
 \end{algorithmic}
@@
 \Statex
-\Function{Object}{Dequeue()}{}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
-\EndFunction{Dequeue}
-
-\Statex
-\pagebreak
 
@@
 \State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
-\State \Return \tt{<root.\Call{BSearch}{sum\sub{enq}, r\sub{enq}, root.FindMostRecentDone(), root.size}, r\sub{enq}>}
+\State \Return \tt{<root.\Call{DSearch}{r\sub{enq}, b}, r\sub{enq}>}
 \EndIf
 \EndFunction{FindResponse}
+
+\Statex
+
+\Function{Object}{Dequeue()}{}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
+\State \tt{newBlock.element= null}
+\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
+\State \tt{leaf.}\Call{Append}{newBlock}
+\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
+\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
+ \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
+If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
+\If{\tt{r\sub{enq}==-1}}
+\State \tt{output= null}
+\pagebreak
+\Else
+\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
+\EndIf
+\State \Return{\tt{output}}
+\EndFunction{Dequeue}
 
@@
 \end{algorithm}
+\pagebreak
 
@@
 
+\Statex
+\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
+\Function{int}{DSearch}{\sl{int} i, \sl{int} end}
+
+\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}.
+Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
+\State \tt{start= b-1}
+\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
+\State \tt{start= start-(b-start)}
+\EndWhile
+\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
+\EndFunction{DSearch}
+
 \pagebreak
@@
 \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{this\sub{dir}= dir.num\sub{propagated}}
+\State \tt{np\sub{dir}= dir.num\sub{propagated}}
 \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
+\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
 %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
@@
 
-\begin{algorithm}
-\caption{Root}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{500}
-
-\Function{Block}{FindMostRecentDone}{}
-\For{\tt{leaf l} \bf{in leaves}}
-\State\tt{max= Max(l.maxOld, max)}
-\EndFor
-\State\Return\tt{max} \cmt{This snapshot suffies.}
-\EndFunction{FindMostRecentDone}
-
-\end{algorithmic}
-\end{algorithm}
+%\begin{algorithm}
+%\caption{Root}
+%\begin{algorithmic}[1]
+%\setcounter{ALG@line}{500}
+%
+%\Function{Block}{FindMostRecentDone}{}
+%\For{\tt{leaf l} \bf{in leaves}}
+%\State\tt{max= Max(l.maxOld, max)}
+%\EndFor
+%\State\Return\tt{max} \cmt{This snapshot suffies.}
+%\EndFunction{FindMostRecentDone}
+%
+%\end{algorithmic}
+%\end{algorithm}
 %##########################################
@@
 
-\Statex
-
-\Function{Object}{HelpDequeue()}{}
-\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
-\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
- \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
-If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
-\If{\tt{r\sub{enq}==-1}}
-\State \tt{output= null}
-\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
-\State \tt{last\sub{done}= b\sub{deq}}
-\EndIf
-\pagebreak
-\Else
-\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
-\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
-\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
-\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
-\State \tt{last\sub{done}= b\sub{deq}}
-\ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
-\State \tt{last\sub{done}= b\sub{enq}}
-\EndIf
-\EndIf
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Function{void}{Help}{}\cmt{Helps pending operations}
-
-\State{\tt{last= l.size-1}}
-\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
-\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
-\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
-\EndIf
-
-\EndFunction{Help}
+%\Statex
+%
+%\Function{void}{Help}{}\cmt{Helps pending operations}
+%
+%\State{\tt{last= l.size-1}}
+%\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
+%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
+%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
+%\EndIf
+%
+%\EndFunction{Help}
 
@@
 
-\begin{algorithm}
-\caption{BlockList}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
-
-\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
-
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{root implementation}}
-%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
-\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
-\State \tt{l.Help()}
-\EndFor
-\EndIf
-\State \tt{blk.num\sub{finished}= 0}
-%\State \tt{*oldRoot= \&root.blocks.root}
-%\State \tt{*newRoot= root.blocks.Append(blk).root}
-%\State \Return \tt{CAS(root, oldRoot, newRoot)}
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
-
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{blocks[]: }\sf{array of blocks}
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
-
-\end{algorithmic}
-\end{algorithm}
-
-%##########################################
-
-\begin{algorithm}
-\caption{Yet to decide how to handle.}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{800}
-
-%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
-%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
-%\State \tt{t1,t2= RBT.split(order, s)}
-%\State \tt{RBTRoot.CAS(t2.root)}
-%\EndFunction{CollectGarbage}
-
-\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
-
-\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
-
-\State \Return \tt{this.leaf.response(block.order)}
-\EndIf
-
-\EndFunction{FallBack}
-
-\end{algorithmic}
-\end{algorithm}
+%\begin{algorithm}
+%\caption{Yet to decide how to handle.}
+%\begin{algorithmic}[1]
+%\setcounter{ALG@line}{800}
+%
+%%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
+%%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
+%%\State \tt{t1,t2= RBT.split(order, s)}
+%%\State \tt{RBTRoot.CAS(t2.root)}
+%%\EndFunction{CollectGarbage}
+%
+%\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
+%
+%\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
+%
+%\State \Return \tt{this.leaf.response(block.order)}
+%\EndIf
+%
+%\EndFunction{FallBack}
+%
+%\end{algorithmic}
+%\end{algorithm}
 
@@
 \clearpage
+
 \section{Proof of Linearizability}
@@
 \begin{definition}[Block]
-A block is an object storing some statistics described in Algorithm Queue.
-It implicitly shows a set of operations.
-The set of operations of block $b$ are the operations in the leaf subblocks of $b$.
-We show the set of operations of block $b$, set of blocks $B$ by $ops(b)$, $ops(B)$.
-We also say $b$ contains $op$ if $op\in ops(b)$.
-\end{definition}
-
-\begin{definition}[Order]
-If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s. 
+A block is an object storing some statistics, as described in Algorithm Queue.
+It implicitly represents a set of operations.
+If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
+Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s.
 \end{definition}
@@
 Block b is direct superblock of block c if c is direct subblock of b.
+The set of operations of block $b$ are the operations in the leaf sub blocks of $b$. We show the set of operations of block by \nf{ops(b)}. We also say $b$ contains $op$ if $op\in ops(b)$.
 \end{definition}
@@
 \begin{definition}
-  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.head} is greater than index of \nf{b} at time $t$.
+ A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if \nf{n.head} is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
 Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
@@
 \begin{observation}\label{head}
-Once block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
+Once a block \nf{b} is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
 \end{observation}
@@
 \begin{lemma}[headProgress] \label{lem::sizeProgress}
-\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} $\geq$ \nf{n.blocks[i-1].end\sub{left, right}}.
+\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
 \end{lemma}
 \begin{proof}
-  Induced trivially from the pseudocode since \nf{n.head} is only incremented.
-\nf{n.blocks[i].
-Also end\sub{left, right}} are greater than or equal the previous values in the \nf{CreateBlock(i)} code (Lines~\ref{lastLine, prevLine}).
+The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented.
+Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}.
+Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.size = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
 \end{proof}
@@
 \begin{proof}
-To show this we are going to refer to the way n.blocks[] is partitioned while propagating up to n.parent.
+To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}.
 \nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
-From we know that \nf{end\sub{left},end\sub{right}} are non decreasing, so the range of the subblocks of n.blocks are disjoint.
+Since \nf{end\sub{left},end\sub{right}} are non-decreasing, so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
 \end{proof}
@@
 \begin{corollary}[No Duplicates]\label{append}
-If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
 \end{corollary}
@@
 \begin{proof}
-Af first the invariant is true since initially 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}.
-The satisfiability of the invariant may be effected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
+The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}.
+The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
 
 Some value is written into \nf{n.blocks[head]} only in Line 313.
-It is obvious that writing into \nf{n.blocks[head]} does not change the state of the claim of the lemma.
-The value of \nf{n.head} is modified only in lines \ref{incrementHead1}, ][p\ref{incrementHead2}, we show in both cases of the successful \ref{cas} or not the claim holds after the increment lines of \nf{n.head}.
-If $h$ increments to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to claim the invariant still holds.
+It is obvious that writing into \nf{n.blocks[head]} presereves the invariant.
+The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}.
+Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case.
+If \nf{head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds.
 In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
@@
 \end{proof}
+\it{Explain More}
 
@@
 \begin{proof}
-Blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node grows.
+Blocks are only appended (not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node can only grows.
 \end{proof}
@@
 
-\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
+\nf{CreateBlock()} aggregates the blocks in the children that are not already established in the parent into one block.
 If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
@@
 Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}.
-So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
+So it means the second failed \nf{Refresh} is concurrent with another successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
 After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
@@
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated.
+  Let $t_i$ be the time a \nf{n.Refresh()} operation is invoked and $t_t$ be the time it terminates.
 Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
@@
 From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-So it remains to show \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})} which we call new operations $\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-If \nf{TryAppend}returns true a block is appended to n.
-From the code of the \nf{CreatBlock} this block includes the establsihed blocks in n's children at $t_i$.
-Since the head in createblock is read after $t_i$.
-So the new operations are inthe block appended to $n$.
+So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call new operations, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
+If \nf{TryAppend} returns true a block is appended to \nf{n}.
+From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$.
+Since the head in \nf{CreateBlock()} is read after $t_i$.
+So the new operations are in the block appended to $n$.
 
@@
 \end{proof}
+\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
 
@@
 \begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive failed instances $r_1$,$r_2$ of \nf{n.Refresh()} by some process.
+  Consider two consecutive failed instances $R_1$, $R_2$ of \nf{n.Refresh()} by some process.
 Let $t_1$ be the time $R_1$ is invoked and $t_2$ be the time $R_2$ terminated.
-After $R_2$'s \nf{TryAppend} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+W  e have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 \end{lemma}
@@
 If Line \ref{cas} of $R_1$ or $R_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \nf{n.Refresh()} $R_2^\prime$ which has did TryAppend successfully into \nf{n.blocks[i+1]}.
-Note that if $R_2$ reads some value greater than $i+1$ in Line \ref{readSize} it means there is a succesful \nf{Refresh()} instance started after Line \ref{readSize} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readSize} of $R_2$, form Lemma~\ref{lem::trueRefresh} by finish of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
-In the figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_i$, since it create a block reading the head after $t_1$.
+If not, then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]}.
+If $R_2$ reads some value greater than $i+1$ in Line \ref{readSize} it means a successful instance of \nf{Refresh()} started after Line \ref{readSize} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readSize} of $R_2$, from Lemma~\ref{lem::trueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
+In Figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_1$, since it create a block reading the head after $t_1$.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
 \end{proof}
+\it{last sentence need more detail and should be earlier. define i and tell why R2prime exists}
 
@@
 \end{figure}
+\it{this chain with more depth should be in the proof}
 
@@
 \item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
+\item $E(n,i)$ is the $i$th enqueue in $E(n)$.
 \item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item $D(n,i)$ is the $i$th dequeue in $D(n)$.
 \item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
@@
 
-\begin{definition}
-An enqueue operation is \textit{finished} if its argument is returned by some process.
+\begin{lemma}[DSearch correctness] \label{dsearch}
+If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E(root, i)$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E(root,b,i)$ in $\Theta(\log$(\nf{root.blocks[b].length+root.blocks[end].length}) steps.
+\end{lemma}
+\begin{proof}
+First we show \nf{end-b}$\leq$\nf{root.blocks[b].length+root.blocks[end].length}.
+We know each block size is greater than 0.
+So every block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or one \nf{Dequeue()}.
+There cannot be more than \nf{root.blocks[b].length} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the queue would become empty after bth block end before end and $E(n,i)$ could not be the response to to some \nf{DEQ} in \nf{end}.
+And since the lentgh of the queue would become \nf{root.blocks[end].length} in the end so there cannot be more than \nf{root.blocks[end].length} Dequeus in \nf{root.blocks[b..end]}.
+Cause if there was more then the end's lenght would become more than \nf{root.blocks[end].length}.
+
+Now that we know there are at most \nf{root.blocks[b].length+root.blocks[end].length} distance between end and b then with doubling search  in log\nf{root.blocks[b].length+root.blocks[end].length} steps we reach a block \nf{c} that the \nf{c.sum\sub{enq}} is less than i and the distance between c and end is not more than $2\times$\nf{root.blocks[b].length+root.blocks[end].length}.
+So the binary search takes $\Theta(\log$\nf{root.blocks[b].length+root.blocks[end].length)}$)$ steps.
+\end{proof}
+
+%\begin{definition}
+%An enqueue operation is \textit{finished} if its argument is returned by some process.
 A dequeue operation is \texttt{finished} if it returns \nf{null} or some value.
 Block \nf{b} is \textit{done} if all operations in \nf{ops(b)} are finished.
-\end{definition}
-\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
-
-\begin{lemma}[help]\label{help}
-After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
-\end{lemma}
+%\end{definition}
+%\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
+%
+%\begin{lemma}[help]\label{help}
+%After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
+%\end{lemma}
 
@@
 
-\begin{lemma}[Computing Queue's Head] \label{computeHead}
+\begin{lemma}[Computing Queue's Head block] \label{computeHead}
   Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it.
-\nf{FindResponse()} returns the index in $E(root,b)$ of the enqueue that is the head in S.
-If the  queue is empty in S it returns \nf{<-1,-->}.
+\nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$.
+If the queue is empty in $S$ then it returns \nf{<-1,-->}.
   \end{lemma}
-
+\begin{proof}
+  The length of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].length}.
+It is computed while creating the block in Line~\ref{computeLength}.
+If the length of a queue is greater than 0 then a \nf{Dequeue()} would decrease the length of the queue, otherwise the length of the queue remains 0.
+Having length of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the length of it.
+\end{proof}
+\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)} is the length of the queue after the block. I can only explain it here.
 %\begin{proof}

### 2022-06-22 article/psuedocode.tex
review pages 1-2

@@
 \newcommand{\cmt}[1]{\Comment{#1}}
+\newcommand{\head}{head}
+\newcommand{\size}{size}
 
@@
 \begin{itemize}
-\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is a pointer to the root node.}
+\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is the root node.}
 %such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
@@
 \begin{itemize}
-\item \tt{\sl{*Node} leaf:} \sf{ a pointer to the process's leaf in the tree.}
+\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
 \end{itemize}
@@
   \textsf{ implemented with an array.}
-\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
+\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}(-1). \tt{blocks[0]} is a block with all fields equal to zero.}
 \item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
@@
   \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. For \tt{enqueue} operations element is the input of the \tt{enqueue} and for \tt{dequeue} operations it is \tt{null}.}
-
-    \item \tt{\sl{Object} response}
-  \textsf{: stores the response of the operation in the \tt{LeafBlock}.}
+  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
 
@@
 \begin{itemize}
-  \item \tt{\sl{int} length}
-  \textsf{: length of the queue after performing all operations in the prefix for this block}
+  \item \tt{\sl{int} \size}
+  \textsf{: size of the queue after performing all operations in the prefix for this block}
 %  \item \tt{\sl{int} sum\sub{non-null deq}}
@@
 
-\begin{algorithm}
-\caption{BlockList}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
-
-\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
-
-%\Statex
-%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
-%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
-%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
-%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
-%\State \tt{l.Help()}
-%\EndFor
-%\EndIf
-%\State \tt{blk.num\sub{finished}= 0}
-%%\State \tt{*oldRoot= \&root.blocks.root}
-%%\State \tt{*newRoot= root.blocks.Append(blk).root}
-%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
-%\State \Return{\tt{CAS(blocks[n], null, blk)}}
-%\EndFunction{TryAppend}
-
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{blocks[]: }\sf{array of blocks}
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
-
-\end{algorithmic}
-\end{algorithm}
-
 \begin{footnotesize}
 
-\it{Variable naming:}
-\begin{itemize}
-  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
-  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
-\end{itemize}
+%\it{Variable naming:}
+%\begin{itemize}
+%  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
+%  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
+%\end{itemize}
 
@@
 %\State \tt{b.num\sub{enq}=1}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}}
+\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
+\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
+\State \tt{leaf.head+=1}
 \State \tt{leaf.}\Call{Append}{newBlock}
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block.
+\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{If $E(root,b_e,i_e)$ is the response to the $D(root,b_d,i_d)$ returns \tt{<b\sub{e},i\sub{e}>}.
 Returns \tt{<-1,-->} if the queue is empty.}
-\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
+\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - (i + root.blocks[b\sub{d}-1].sum\sub{deq}) $<$ 0}}
 \State \Return \tt{<-1,-->}
 \Else
-\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}.
-$r$th non-null dequeue returns the element of th $r$th enqueue.
-We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues.
-Note that the $i$th dequeue in the given block is not a non-null dequeue.}
-\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
-\State \Return \tt{<root.\Call{DSearch}{r\sub{enq}, b}, r\sub{enq}>}
+
+\State \tt{r\sub{enq}= i + root.blocks[b\sub{d}-1].sum\sub{deq} - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
+\State \cmt{size-enqs+deqs=null deqs}
+\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
 \EndIf
@@
 \State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
+\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
 \State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
+\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}\cmt{\tt{r\sub{deq}} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
 \State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
- \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist.
-If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
+ \cmt{\tt{$E(root,b_{enq},i_{enq})$ is response to the $D(root,b_{deq},i_{deq})$. If the response is \tt{null} then r\sub{enq} is -1.}} \label{deqRest}
 \If{\tt{r\sub{enq}==-1}}
@@
 \Function{boolean}{Refresh()}{}
-\State \tt{s= size} \label{readSize}
+\State \tt{s= \head} \label{readHead}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
@@
 \EndFor
-\State \tt{\Call{CAS}{size, s, s+1}} \label{incrementHead1}
+\State \tt{\Call{CAS}{\head, s, s+1}} \label{incrementHead1}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{size, s, s+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.} \label{incrementHead2}
+\State \tt{\Call{CAS}{\head, s, s+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
 \State \Return{ \tt{false}}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
-\Function{int}{DSearch}{\sl{int} i, \sl{int} end}
+\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}
 
@@
 \State \tt{start= start-(b-start)}
-\EndWhile
+\EndWhile \State
 \Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.size} \label{lastLine}
+\State \tt{index\sub{last}= dir.\head} \label{lastLine}
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
+\State \tt{newBlock.\size= max(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
 %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
@@
 \Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
-\State \tt{size+=1} \label{appendEnd} 
-\State \tt{blk.group= size} \label{appendStart} 
-\State \tt{blocks[size]= blk} 
+\State \tt{\head+=1} \label{appendEnd} 
+\State \tt{blk.group= \head} \label{appendStart} 
+\State \tt{blocks[\head]= blk} 
 \State \tt{parent.}\Call{Propagate()}{} 
@@
 %
-%\State{\tt{last= l.size-1}}
-%\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
+%\State{\tt{last= l.\head-1}}
+%\cmt{\tt{l.blocks[last]} can not be \tt{null} because \head increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
 %\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
@@
 %\EndFunction{Help}
+
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{BlockList}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{700}
+
+\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
+
+%\Statex
+%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
+%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
+%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
+%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
+%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
+%\State \tt{l.Help()}
+%\EndFor
+%\EndIf
+%\State \tt{blk.num\sub{finished}= 0}
+%%\State \tt{*oldRoot= \&root.blocks.root}
+%%\State \tt{*newRoot= root.blocks.Append(blk).root}
+%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
+%\State \Return{\tt{CAS(blocks[n], null, blk)}}
+%\EndFunction{TryAppend}
+
+\Statex
+%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
+\Statex \tt{blocks[]: }\sf{array of blocks}
+\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
+\State \Return{\tt{CAS(blocks[n], null, blk)}}
+\EndFunction{TryAppend}
 
@@
 
-\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
+\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.\head to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
 Fix the logical order of definitions (cyclic refrences).
@@
 
-\begin{lemma}[headProgress] \label{lem::sizeProgress}
+\begin{lemma}[headProgress] \label{lem::headProgress}
 \nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
@@
 Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}.
-Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.size = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
+Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.\head = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
 \end{proof}
@@
 In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
-Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readSize}.
+Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readHead}.
 The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
-%At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
+%At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
 If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
@@
 Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.
-Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$.
+Since the head is only increasing (Lemma~\ref{lem::headProgress}) the  claim holds after $t$.
 See Figure~\ref{fig::createBlock}.
@@
 Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10.
-%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
+%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readHead} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
 \end{proof}
@@
 %\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
+%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readHead} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
 A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
@@
 If not, then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]}.
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readSize} it means a successful instance of \nf{Refresh()} started after Line \ref{readSize} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readSize} of $R_2$, from Lemma~\ref{lem::trueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
+If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means a successful instance of \nf{Refresh()} started after Line \ref{readHead} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readHead} of $R_2$, from Lemma~\ref{lem::trueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
 In Figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_1$, since it create a block reading the head after $t_1$.
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E(root, i)$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E(root,b,i)$ in $\Theta(\log$(\nf{root.blocks[b].length+root.blocks[end].length}) steps.
+If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E(root, i)$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E(root,b,i)$ in $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) steps.
 \end{lemma}
 \begin{proof}
-First we show \nf{end-b}$\leq$\nf{root.blocks[b].length+root.blocks[end].length}.
+First we show \nf{end-b}$\leq$\nf{root.blocks[b].\size+root.blocks[end].\size}.
 We know each block size is greater than 0.
 So every block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or one \nf{Dequeue()}.
-There cannot be more than \nf{root.blocks[b].length} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the queue would become empty after bth block end before end and $E(n,i)$ could not be the response to to some \nf{DEQ} in \nf{end}.
-And since the lentgh of the queue would become \nf{root.blocks[end].length} in the end so there cannot be more than \nf{root.blocks[end].length} Dequeus in \nf{root.blocks[b..end]}.
-Cause if there was more then the end's lenght would become more than \nf{root.blocks[end].length}.
+There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the queue would become empty after bth block end before end and $E(n,i)$ could not be the response to to some \nf{DEQ} in \nf{end}.
+And since the lentgh of the queue would become \nf{root.blocks[end].\size} in the end so there cannot be more than \nf{root.blocks[end].\size} Dequeus in \nf{root.blocks[b..end]}.
+Cause if there was more then the end's lenght would become more than \nf{root.blocks[end].\size}.
 
-Now that we know there are at most \nf{root.blocks[b].length+root.blocks[end].length} distance between end and b then with doubling search  in log\nf{root.blocks[b].length+root.blocks[end].length} steps we reach a block \nf{c} that the \nf{c.sum\sub{enq}} is less than i and the distance between c and end is not more than $2\times$\nf{root.blocks[b].length+root.blocks[end].length}.
-So the binary search takes $\Theta(\log$\nf{root.blocks[b].length+root.blocks[end].length)}$)$ steps.
+Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} distance between end and b then with doubling search  in log\nf{root.blocks[b].\size+root.blocks[end].\size} steps we reach a block \nf{c} that the \nf{c.sum\sub{enq}} is less than i and the distance between c and end is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
+So the binary search takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps.
 \end{proof}
@@
 \end{proof}
+We call the dequeues that return some value \it{non-null dequeues}.
+$r$th non-null dequeue returns the element of th $r$th enqueue.
+We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= \size - \#enqueues.
+Note that the $i$th dequeue in the given block is not a non-null dequeue.
 
@@
 It is easy to see why the second is correct.
-Correctness of computing superblock comes from Lemma \ref{superBlock}.
+Correctness of computing superblock comes from Lemma \ref{superBlock}.\size
 \end{proof}
@@
 \begin{proof}
-  The length of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].length}.
+  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}.
 It is computed while creating the block in Line~\ref{computeLength}.
-If the length of a queue is greater than 0 then a \nf{Dequeue()} would decrease the length of the queue, otherwise the length of the queue remains 0.
-Having length of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the length of it.
+If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
+Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size of it.
 \end{proof}
-\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)} is the length of the queue after the block. I can only explain it here.
+\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
 %\begin{proof}
@@
 \begin{lemma}[Root search range]\label{rootRange}
-  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the length of the queue.
+  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
 \end{lemma}

### 2022-06-22 article/psuedocode.tex
review pagaes 3-5

@@
 \Function{boolean}{Refresh()}{}
-\State \tt{s= \head} \label{readHead}
+\State \tt{h= \head} \label{readHead}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
@@
 \If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
-\ElsIf{\tt{blocks.tryAppend(new, s)}} \label{cas}
+\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
@@
 \EndFor
-\State \tt{\Call{CAS}{\head, s, s+1}} \label{incrementHead1}
+\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
 \State \Return{\tt{true}}
 \Else
-\State \tt{\Call{CAS}{\head, s, s+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
+\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
 \State \Return{ \tt{false}}
@@
 
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
-\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}
-
-\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}.
-Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
-\State \tt{start= b-1}
-\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
-\State \tt{start= start-(b-start)}
-\EndWhile \State
-\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
-\EndFunction{DSearch}
-
 \pagebreak
@@
 \State \tt{newBlock.group= num\sub{propagated}}
-\State \tt{newBlock.order= i}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.\size= max(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
+\State \tt{newBlock.\size= max(root.blocks[i-1].\size + newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
 %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
@@
 \end{multicols}
+\end{algorithmic}
+\end{algorithm}
+\begin{algorithm}
+\caption{Root}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{800}
+\Statex
+\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
+\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}
+
+\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}.
+Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
+\State \tt{start= end-1}
+\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
+\State \tt{start= start-(end-start)}
+\EndWhile \State
+\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, end)}
+\EndFunction{DSearch}
 \end{algorithmic}
@@
 
-\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.\head to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before.
-Fix the logical order of definitions (cyclic refrences).
+\framebox[1.1\width]{TEST} Fix the logical order of definitions (cyclic refrences).
 
-\framebox[1.1\width]{TODO} Fallback safety lemmas. Some parts are obsolete.
-
-\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
+%\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
 Is lemma 7 necessary?
@@
 
-\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
+%\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
 block b $\sqsubseteq$ block c means b is subblock of c.
@@
 Merge these two to have shorter formulaes.
+
+\framebox[1.1\width]{Question} A good notation for \it{the index of the \nf{b}}?
 
@@
 If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
-Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s.
+Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if the index of the \nf{b} is smaller than the index of the \nf{b$^\prime$}'s.
 \end{definition}
@@
 Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
-Block b is direct superblock of block c if c is direct subblock of b.
-The set of operations of block $b$ are the operations in the leaf sub blocks of $b$. We show the set of operations of block by \nf{ops(b)}. We also say $b$ contains $op$ if $op\in ops(b)$.
+\end{definition}
+
+\begin{definition}[Superblock]
+  Block b is direct superblock of block c if c is a direct subblock of b.   Block b is superblock of block c if c is a subblock of b.
+\end{definition}
+
+\begin{definition}[Operations of a block]
+A block lb in a leaf represents one operation which if it is \tt{enqueue(x)} then \tt{lb.element=x}, otherwise \tt{element=null}.
+The set of operations of block $b$ are the operations in the subblocks of $b$.
+We show the set of operations of block b by \nf{ops(b)}.
 \end{definition}
@@
 For simplicity we say block \nf{b} is propagated to node \nf{n} or to a set of blocks \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
+We also say $b$ contains $op$ if $op\in ops(b)$.
 
@@
 \begin{definition}
- A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if \nf{n.head} is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
-Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
+ A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
+\emph{EST\sub{n, t}} is the set of established blocks at time $t$ of node $n$.
 \end{definition}

### 2022-06-22 article/psuedocode.tex
fix true refresh

@@
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{newBlock.end\sub{dir}= index\sub{last}}
+\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
 \State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
@@
 
+\framebox[1.1\width]{Question} How to remove the notion of time? To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t. Is it good?
+
 \begin{definition}[Block]
@@
 \begin{definition}[Subblock]
-Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (See line \ref{endDefLine} for the defined range).
 Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
@@
 
+\pagebreak
+
+\begin{lemma}[head Increment] \label{lem::headInc}
+If an n.Refresh  instance reaches Line \ref{cas} instance and reads head=h (\ref{readHead}) after it terminates head is greater than h.
+\end{lemma}
+\begin{proof}
+If Line \ref{incrementHead1} or \ref{incrementHead1} succeeded the claim holds, otherwise another process has incremented the head.
+\end{proof}
+
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Let $t_i$ be the time a \nf{n.Refresh()} operation is invoked and $t_t$ be the time it terminates.
-Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
+  Let $t_i$ be the time an instance of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates.
+Suppose the \nf{TryAppend(new, s)} of the \nf{n.Refresh()}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
 \end{lemma}
@@
 From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call new operations, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
-If \nf{TryAppend} returns true a block is appended to \nf{n}.
-From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$.
+So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
+If \nf{TryAppend} returns true a block \nf{new} is written into \nf{n.blocks[h]} (Line \ref{cas}).
+We show \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
+The proof for the right child's claim is the same.
+Let \nf{n.left.head} at \nf{t\sub{i}} be \nf{hli}.
+Let \nf{n.Refresh()} read \nf{head} equal to $h$(Line \ref{readHead}).
+By the lines \ref{prevLine},\ref{lastLine} the \nf{new} block in \nf{n.blocks[h]} contains \nf{n.left.blocks[ n.blocks[h-1].end\sub{left}+1..left.head]}.
+Since \nf{left.head} is read after \nf{t\sub{i}} then \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..left.head])}.
+By Lemma \ref{lem::establishedOrder} \nf{ops(n.left.blocks[0..n.blocks[h-1].end\sub{left}])}$\subseteq$\nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
+Since after line \ref{incrementHead2} we are sure that the \nf{head} is incremented (Lemma\ref{lem::headInc}) and \nf{n.head=h+1} at $t_t$ so the new block is established at $t_t$ and the new block contains the new operations which is what we wanted to show.
+% From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$.
 Since the head in \nf{CreateBlock()} is read after $t_i$.
@@
 \end{proof}
-\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
+\begin{figure}[hbt]
+  \center\includegraphics[width=5in]{pics/trueRefresh}
+  \caption{New established operations of the left child are in the new block.}
+\end{figure}
+%\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
 
@@
 %\end{proof}
+
+\pagebreak
 
@@
 
+\pagebreak
+
 \begin{corollary}[Propagate Step] \label{doublyRefresh}
@@
 \end{proof}
+
+\pagebreak
 
@@
 
+\pagebreak
+
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
@@
 
+\pagebreak
+
+Preconditions of all invocation of \nf{BSearch} are satisfied.
 \begin{lemma}[Get correctness] \label{get}
@@
 \textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
+
+\pagebreak
 
@@
 
+\pagebreak
+
 %\begin{definition}
@@
 %\end{lemma}
-
+Preconditions of all invocation of \nf{BSearch} are satisfied.
 \begin{lemma}[Computing SuperBlock]\label{superBlock}
@@
 
+\pagebreak
+
 \begin{lemma}[Index correctness]
@@
 
-\begin{lemma}[Search Ranges]\label{search}
-  Preconditions of all invocation of \nf{BSearch} are satisfied.
-\end{lemma}
-\begin{proof}
-
-Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-
-Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+%\begin{lemma}[Search Ranges]\label{search}
+%  Preconditions of all invocation of \nf{BSearch} are satisfied.
+%\end{lemma}
+%\begin{proof}
+%  
+%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+%
+%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
 Search is done over the left subblocks.
@@
 Line 92 is the same.
-
-Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
-
-\end{proof}
+%
+%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
+%
+%\end{proof}
 
@@
 
+\pagebreak
+
 \begin{lemma}[Validity of \nf{super} and \nf{counter}]\label{superCounter}
@@
 
-\begin{lemma}[Root search range]\label{rootRange}
-  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
-\end{lemma}
+\pagebreak
 
@@
 
+\pagebreak
+
 \begin{lemma}[Time analysis]
@@
 
+\begin{lemma}[Root search range]\label{rootRange}
+  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
+\end{lemma}
 \end{document}

### 2022-06-26 article/psuedocode.tex
review pages 13,15,18

@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
-\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
+\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.} \label{setSuper}
+\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
 \EndFor
@@
 \State \tt{block newBlock= \Call{new}{\sl{block}}}
-\State \tt{newBlock.group= num\sub{propagated}}
+\State \tt{newBlock.group= num\sub{propagated}}\label{setGroup}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{np\sub{dir}= dir.num\sub{propagated}}
+\State \tt{np\sub{dir}= dir.num\sub{propagated}} \label{setNP}
 \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
+\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.} \label{rightChildGet}
 \State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
@@
 
-\framebox[1.1\width]{Question} How to remove the notion of time? To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t. Is it good?
+\framebox[1.1\width]{Question} How to remove the notion of time?
+To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t.
+Is it good?
+Furthermore, can we remove the notion of established blocks?
 
@@
 
-\begin{definition}[Subblock]
+\begin{definition}[Subblock]\label{def::subblock}
 Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (See line \ref{endDefLine} for the defined range).
@@
 
-\begin{definition}[Operations of a block]
+\begin{definition}[Operations of a block]\label{def::ops}
 A block lb in a leaf represents one operation which if it is \tt{enqueue(x)} then \tt{lb.element=x}, otherwise \tt{element=null}.
@@
 \end{figure}
+\begin{lemma}[Precise True Refresh] \label{lem::prectrueRefresh}
+  Let $t_i$ be the time an instance of \nf{n.Refresh()} read the head (Line \ref{readHead}) and $t_t$ be the time its \nf{TryAppend(new, s)} terminates with  and returns \nf{true} (Line \ref{cas}).
+We have \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(n.blocks)}.
+\end{lemma}
 %\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
@@
 Let $t_1$ be the time $R_1$ is invoked and $t_2$ be the time $R_2$ terminated.
-W  e have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+We have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 \end{lemma}
@@
 If Line \ref{cas} of $R_1$ or $R_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]}.
+Let $R_1$ read $i$ and $R_2$ read $i+1$ from Line \ref{readHead}.
 If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means a successful instance of \nf{Refresh()} started after Line \ref{readHead} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readHead} of $R_2$, from Lemma~\ref{lem::trueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
+
+Since $R_2$'s \nf{TryAppend()} returns \nf{false} then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to append.
 In Figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_1$, since it create a block reading the head after $t_1$.
+By Lemma \ref{lem::prectrueRefresh} after $R_2^\prime$'s \nf{CAS} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(n.blocks)}.
+Also by Lemma \ref{lem::headInc} of $R_2$  head is more than $i+1$ after $R_2$\'s \ref{incrementHead2} line, so the block appended by $R_2^\prime$ to \nf{n} is established by then.
+To summarized $t_1$ is before $R_2^\prime$'s read head and $R_2^\prime$'s \nf{CAS} is before $R_2$'s termination.
+So \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
 \item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
-\item $E(n,i)$ is the $i$th enqueue in $E(n)$.
+\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
 \item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
-\item $D(n,i)$ is the $i$th dequeue in $D(n)$.
+\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
 \item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
@@
 For non leaf nodes in Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}.
-By definition of $E(n,b)$ operations from the left child come  before the operations of the right child.
+From Definition \ref{ordering} we know operations in a block are ordered by  their process id.
+Furthermore \nf{b.sum\sub{enq-lef}} stores the number of \nf{enqueue()} operations from the \nf{b}'s subblocks of the left child of \nf{n}.
+So if \nf{i} is greater than \nf{b.sum\sub{enq-lef}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue i the left child subblocks.
+By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of b which their range is \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+If the enqueue we're looking for was in the right child as there are \nf{b.sum\sub{enq-left}} enqueues before it we need to search for \nf{i-b.sum\sub{enq-left}} (Line \ref{rightChildGet}).
+of By definition of $E(n,b)$ operations from the left child come  before the operations of the right child.
 Having \nf{sum\sub{enq}}, the prefix sum of the number of enqueues we can compute the direct subblock containing the enqueue we are finding for with binary search.
@@
 \end{proof}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
+  \caption{The number of enqueues from the left and the right child}
+\end{figure}
+
 \textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
@@
  \item Value read for \nf{super[b.group]} in line 418 is not null.\\$\blacktriangleright$
-  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.
-
- \item \nf{super[]} preserves order from child to parent; if in a child block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group} and \nf{super[b.group]} $\leq$ \nf{super[c.group]}\\$\blacktriangleright$
-  Follows from the order of lines 60, 48, 49.
+  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
+So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
+
+ \item \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
+$\blacktriangleright$ Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.
+
+ \item Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\\$\blacktriangleright$ Line \ref{setSuper}.
 
@@
 \end{theorem}
-%\begin{proof}
-%  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
-%  \begin{enumerate}
-%    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
+\begin{proof}
+  We choose $L$ in Definition \ref{ordering} to be linearization ordering of operations and prove if we linearize operations as $L$ the queue works consistently.
+\end{proof}
+\begin{lemma}
+Operations in a block have a time point in common (There is a time $t$ all the operations are running).
+\end{lemma}
+
+\begin{lemma}[satisfiability]
+$L$ can be a linearization ordering.
+\end{lemma}
+\begin{proof}
+Once some operations are aggregated in one block they will be propagated together up to the root and we can linearize them in any order among themselves (previous lemma).
+Furthermore in L we arbitrary choose the order to be by process id, since it makes computations in  the blocks faster .
+\end{proof}
+
+\begin{lemma}[correctness]
+  If operations are applied as $L$ on a sequential queue, the sequence of the responses would be the same as our algorithm.
+\end{lemma}
+
+\begin{proof}
+\it{Old parts to review}
+  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
+  \begin{enumerate}
+    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
 The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
-%    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
+    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
 \\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
@@
 If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
-%  \end{enumerate} 
-%\end{proof}
+  \end{enumerate} 
+\end{proof}
 

### 2022-06-28 article/psuedocode.tex
wrap up the review

@@
 \State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
-\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
+\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
 \ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>}
+\State\Return \tt{<b, i>} \label{indexBaseCase}
 \Else
@@
   \item $E(n,b)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
-  \item $E(n,b,i)$ is the $i$th enqueue in $E(n,b)$.
+  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
   \item $D(n,b)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
-  \item $D(n,b,i)$ is the $i$th enqueue in $D(n,b)$.
+  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
 \item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
@@
 \begin{lemma}[Get correctness] \label{get}
-If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E(n,b,i)$.
+If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
 \end{lemma}
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E(root, i)$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E(root,b,i)$ in $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) steps.
+If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E_{root, i}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E_{root,b,i}$ in $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) steps.
 \end{lemma}
@@
 \begin{lemma}[Index correctness]
- \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D(n,b,i)$.
-\end{lemma}
-\begin{proof}
-We can see the base case \nf{root.IndexDEQ(b,i)} is trivial.
-  \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent}(Lemma \ref{superBlock}).
-Then the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
-It is easy to see why the second is correct.
-Correctness of computing superblock comes from Lemma \ref{superBlock}.\size
+ \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
+\end{lemma}
+\begin{proof}
+We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case \nf{root.IndexDEQ(b,i)} is trivial (Line \ref{indexBaseCase}).
+  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
+After that the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
+Then if the operation was propagated from the right child the number of dequeues from the left child are added to it, because the left child operations come before the right child operations (\ref{ordering}).
 \end{proof}
@@
 %\end{proof} 
-
-\pagebreak
-
-\begin{lemma}[Validity of \nf{super} and \nf{counter}]\label{superCounter}
+\begin{lemma}[super property]\label{superCounter}
 If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the index of the superblock of a block with \nf{time=i}$\pm$\nf{p}.
 \end{lemma}
+\pagebreak
+
 %\begin{proof}
@@
 Furthermore we are sure that when \nf{super[i]} is read it will not be \nf{null}.
+%\end{proof}
+
+\begin{lemma}[Computing Queue's Head block] \label{computeHead}
+  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it.
+\nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$.
+If the queue is empty in $S$ then it returns \nf{<-1,-->}.
+  \end{lemma}
+\begin{proof}
+  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}.
+It is computed while creating the block in Line~\ref{computeLength}.
+If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
+Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size of it.
+\end{proof}
+\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
+%\begin{proof}
+%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read.
+If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}.
+If a process dies before line 51, another process will increment \nf{head} in line 54.
 %\end{proof}
@@
 
-\begin{lemma}[Time analysis]
-  \nf{n.GetEnq(b,i), n.Index(b,i)} take $O(\log^2 p)$ steps.
+\begin{lemma}[Amortized time analysis]
+%  \nf{n.GetEnq(b,i), n.Index(b,i)} take $O(\log^2 p)$ steps.
 Search in the root may take $O(\log Q+ p^2)$ steps.
@@
 Amortized time consumed for helping by each process is $O(\log^2 p)$.
-\end{lemma}
-
-\begin{lemma}[Root search range]\label{rootRange}
-  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
-\end{lemma}
+\nf{Enqueue()} and \nf{Dequeue} take $O(\log62 p +q)$ steps (amortized anlysis), which $p$ is the number of processes and $q$ is the size of the queue at the time of invocation.
+\end{lemma}
+
+%\begin{lemma}[Root search range]\label{rootRange}
+%  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
+%\end{lemma}
 \end{document}

### 2022-07-04 article/psuedocode.tex
apply jun 30 feedback

@@
 \begin{itemize}
-\item \tt{\sl{Tree} tree} \textsf{: A binary tree of \tt{Node}s. \tt{root} is the root node.}
+\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
 %such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
@@
 \item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
-\item \tt{\sl{BlockList} blocks}
-  \textsf{ implemented with an array.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}(-1). \tt{blocks[0]} is a block with all fields equal to zero.}
+\item \tt{\sl{BlockList}}
+\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
 \item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
 Since it is incremented after propagating, it may be behind by 1.}
-\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
 \end{itemize}
@@
 %\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
-\cmt{For a \tt{block} in a \tt{blocklist} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
-\it{put the definitions before the pseudocode}}
 
@@
     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~index of the last subblock of the block in the left and right child}
+  \textsf{:~~indices of the last subblock of the block in the left and right child}
   \item \tt{\sl{int} sum\sub{enq-left}}
@@
  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\}, blocks[0].num\sub{x}=0})
+  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\})}
 \end{itemize}
 \end{footnotesize}
+
+\pagebreak
 
@@
 
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a block with element \tt{e} and appends it to the tree.}
+\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
@@
 \State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.head+=1}
 \State \tt{leaf.}\Call{Append}{newBlock}
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{If $E(root,b_e,i_e)$ is the response to the $D(root,b_d,i_d)$ returns \tt{<b\sub{e},i\sub{e}>}.
-Returns \tt{<-1,-->} if the queue is empty.}
-\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - (i + root.blocks[b\sub{d}-1].sum\sub{deq}) $<$ 0}}
-\State \Return \tt{<-1,-->}
-\Else
-
-\State \tt{r\sub{enq}= i + root.blocks[b\sub{d}-1].sum\sub{deq} - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
-\State \cmt{size-enqs+deqs=null deqs}
-\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
-\EndIf
-\EndFunction{FindResponse}
-
-\Statex
-
-\Function{Object}{Dequeue()}{}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
+\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
+\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
 \State \tt{newBlock.element= null}
@@
 \State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}\cmt{\tt{r\sub{deq}} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
-\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
- \cmt{\tt{$E(root,b_{enq},i_{enq})$ is response to the $D(root,b_{deq},i_{deq})$. If the response is \tt{null} then r\sub{enq} is -1.}} \label{deqRest}
-\If{\tt{r\sub{enq}==-1}}
+\State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
+\State \tt{<b\sub{enq}, i\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
+\label{deqRest}
+\If{\tt{i\sub{enq}==-1}}
 \State \tt{output= null}
@@
 \Else
-\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
+\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
 \EndIf
@@
 
+\Statex
+
+\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{Returns \tt{<b\sub{e},i\sub{e}>}, if $E_{root,b_e,i_e}$ is the response to the $D_{root,b_d,i_d}$.
+Returns \tt{<-1,-->} if the queue is empty.}
+\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
+\State \Return \tt{<-1,-->}
+\Else
+\State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
+\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
+\EndIf
+\EndFunction{FindResponse}
+
 \end{multicols}
@@
 \end{algorithm}
-\pagebreak
-
 %##########################################
@@
 \If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}\label{secondRefresh}
+\State \Call{Refresh()}{} \label{secondRefresh}
 \EndIf
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.} \label{setSuper}
+\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
 \State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
@@
 \Else
-\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \head. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
+\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
 \State \Return{ \tt{false}}
@@
 \Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}.
-Returns the created block as well as values read from each child's num\sub{propagated} field.
-These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
+\cmt{Creates a block to be inserted as $i$th \tt{block} in \tt{blocks}.
+Returns the created block as well as values read from each child's \tt{num\sub{propagated}} field.
+These values are used for incrementing the children's \tt{num\sub{propagated}} field if the block was appended to \tt{blocks} successfully.}
 \State \tt{block newBlock= \Call{new}{\sl{block}}}
@@
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.\size= max(root.blocks[i-1].\size + newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
+\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
 %\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
-\Function{<int, int>}{DSearch}{\sl{int} i, \sl{int} end}
-
-\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}.
-Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
+\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e} $\geq$ 1}}
+\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
+\cmt{Returns \tt{<b,i>} if $E_{root,e}=E_{root,b,i}$.}
 \State \tt{start= end-1}
-\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
-\State \tt{start= start-(end-start)}
-\EndWhile \State
-\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, end)}
+\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
+\State \tt{start= max(start-(end-start), 0)}
+\EndWhile
+\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
+\State \tt{i= e- root.blocks[b-1].sum\sub{enq}}
+\State\Return \tt{<b,i>}
 \EndFunction{DSearch}
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
-\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
+\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
 \If{\tt{this} \bf{is} \tt{leaf}}
 \State\Return \tt{blocks[b].element}
-\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
-\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
-\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
+\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{$E_{this,b,i}$ is in the left child of this node.}
+\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} 
+\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
 \Else
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.} \label{rightChildGet}
-\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
+\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[blocks[b].end\sub{right}-1].sum\sub{enq}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
+\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
 \EndIf
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
-\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
+\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 \Else
-\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
-\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
+\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
+\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
@@
 %\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
-\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
+\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child}
 \EndIf
@@
 \EndIf
-\EndFunction{Index}
+\EndFunction{IndexDeq}
 
@@
 %\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{blocks[]: }\sf{array of blocks}
+\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
+\Statex \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
+\Statex
 \Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
@@
 Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if the index of the \nf{b} is smaller than the index of the \nf{b$^\prime$}'s.
+For a \tt{block} in a \tt{BlockList} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
 \end{definition}
@@
 Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size  of it.
+
+  \tt{r\sub{enq}= (i\sub{d} + root.blocks[b\sub{d}-1].sum\sub{deq}) - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
 \end{proof}

### 2022-07-04 article/psuedocode.tex
review IndexDEQ

@@
 \newcommand{\head}{head}
-\newcommand{\size}{size}
+\newcommand{\size}{size }
 
@@
 %\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
-\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child}
+\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerRight}
 \EndIf
@@
 
+\begin{lemma}[Index correctness]
+ \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
+\end{lemma}
+\begin{proof}
+We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case \nf{root.IndexDEQ(b,i)} is trivial (Line \ref{indexBaseCase}).
+  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
+After that the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
+Then if the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
+\end{proof}
+\textit{Do I need to talk about the computation of the order in the parent which is based on the definition of ordering of dequeues in a block?
+Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}
+
 %\begin{definition}
@@
 %\end{lemma}
-Preconditions of all invocation of \nf{BSearch} are satisfied.
+
 \begin{lemma}[Computing SuperBlock]\label{superBlock}
@@
 \begin{proof}
-\begin{enumerate}
- \item Value read for \nf{super[b.group]} in line 418 is not null.\\$\blacktriangleright$
+Lemmas 28,29,30,31.
+\end{proof}
+\begin{lemma}
+Value read for \nf{super[b.group]} in line 418 is not null.\end{lemma}
+\begin{proof}
   Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
 So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
-
- \item \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
-$\blacktriangleright$ Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.
-
- \item Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\\$\blacktriangleright$ Line \ref{setSuper}.
+\end{proof}
+\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
+\end{lemma}
+\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}
+
+\begin{lemma} Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\end{lemma}
+\begin{proof} Line \ref{setSuper}.\end{proof}
+
+\begin{lemma}
+The number of the \nf{block}s with \nf{group=i} in a node is $\leq p$.  
+\end{lemma}
+\begin{proof}For the sake of simplicity we assumed all the blocks are propagated from the left child.
+\begin{figure}[hbt]
+  \center\includegraphics[width=5in]{pics/index1}
+\end{figure}  
+\end{proof}
 
@@
 % \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
- \item \nf{super[i+1]-super[i]}$\leq p$
-\\$\blacktriangleright$
+\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
+\begin{proof}
  In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
@@
 So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
-
- \item Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
-\\$\blacktriangleright$
-\nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
-It is trivial to see that \nf{n.super} and \nf{n.b.counter} are increasing.
+ \begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/index2}
+\end{figure}  
+\end{proof}
+\begin{lemma}[super property]\label{superCounter}
+If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
+\end{lemma}
+\begin{lemma} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
+\end{lemma}
+
+\begin{proof} \nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
 \nf{super(b)} is the real superblock of b.
@@
 $$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-
-\end{enumerate}
-\end{proof}
+\end{proof}
+
+\pagebreak
+
+%\begin{lemma}[Search Ranges]\label{search}
+%  Preconditions of all invocation of \nf{BSearch} are satisfied.
+%\end{lemma}
+%\begin{proof}
+%  
+%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
+%
+%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
+Search is done over the left subblocks.
+The start and end of the range are followed by definition.
+Line 92 is the same.
+%
+%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
+%
+%\end{proof}
 We call the dequeues that return some value \it{non-null dequeues}.
@@
 Note that the $i$th dequeue in the given block is not a non-null dequeue.
-
-\pagebreak
-
-\begin{lemma}[Index correctness]
- \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case \nf{root.IndexDEQ(b,i)} is trivial (Line \ref{indexBaseCase}).
-  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
-After that the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
-Then if the operation was propagated from the right child the number of dequeues from the left child are added to it, because the left child operations come before the right child operations (\ref{ordering}).
-\end{proof}
-\textit{Do I need to talk about the computation of the order in the parent which is based on the definition of ordering of dequeues in a block?}
-
-%\begin{lemma}[Search Ranges]\label{search}
-%  Preconditions of all invocation of \nf{BSearch} are satisfied.
-%\end{lemma}
-%\begin{proof}
-%  
-%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-%
-%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-%
-%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
-%
-%\end{proof}
-
 \begin{lemma}[Computing Queue's Head block] \label{computeHead}
@@
   \tt{r\sub{enq}= (i\sub{d} + root.blocks[b\sub{d}-1].sum\sub{deq}) - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
-\end{proof}
-\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
-%\begin{proof}
-%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read.
-If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}.
-If a process dies before line 51, another process will increment \nf{head} in line 54.
-%\end{proof} 
-\begin{lemma}[super property]\label{superCounter}
-If \nf{super[i] $\neq$ null}, then \nf{super[i]} in node \nf{n} is the index of the superblock of a block with \nf{time=i}$\pm$\nf{p}.
-\end{lemma}
-\pagebreak
-
-%\begin{proof}
-%After a successful CAS in line 46 \nf{super} and \nf{counter} are modified in both children.
-\nf{super[i]} is supposed to be the superblock of a block with \nf{time}=i and \nf{counter} is the timer in each node.
-\nf{super[i]} and \nf{counter=i} are expected to update after a bunch of blocks with \nf{time=i} have been aggregated together into a block in the parent.
-If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \nf{time}.
-We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
-If a process dies right after line 48, then \nf{counter} will remain the same and \nf{super[i]} is correct.
-Furthermore we are sure that when \nf{super[i]} is read it will not be \nf{null}.
-%\end{proof}
-
-\begin{lemma}[Computing Queue's Head block] \label{computeHead}
-  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it.
-\nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$.
-If the queue is empty in $S$ then it returns \nf{<-1,-->}.
-  \end{lemma}
-\begin{proof}
-  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}.
-It is computed while creating the block in Line~\ref{computeLength}.
-If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
-Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size of it.
 \end{proof}

### 2022-07-04 article/psuedocode.tex
add analysis

@@
 \end{lemma}
-\begin{lemma} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
+\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
 \end{lemma}
@@
 Amortized time consumed for helping by each process is $O(\log^2 p)$.
-\nf{Enqueue()} and \nf{Dequeue} take $O(\log62 p +q)$ steps (amortized anlysis), which $p$ is the number of processes and $q$ is the size of the queue at the time of invocation.
-\end{lemma}
+\nf{Enqueue()} and \nf{Dequeue} take $O(\log^2 p +q)$ steps (amortized anlysis), which $p$ is the number of processes and $q$ is the size of the queue at the time of invocation.
+\end{lemma}
+\begin{proof}
+\nf{Enqueue(x)} consists of creating a \nf{block(x)} and appending it to the tree.
+The first part takes constant time.
+To propagate \nf{x} to the root the algorithm tries two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
+Each \nf{Refresh} takes $O(1)$ steps since creating a block is done in constant time and does $O(1)$ \nf{CAS}es.
+Since the height of the tree is $O(\log p)$, \nf{Enqueue(x)} takes $O(\log p)$ steps.
+
+A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among operations, and returns the response.
+The first two part is similar to an \nf{Enqueue} operation.
+To compute the order there are some constant steps and \nf{IndexDeq} is called.
+\nf{IndexDeq} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree.
+In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$.
+Each search in \nf{GetEnq()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnq()} takes $O(\log^2 p)$ steps.
+
+If we split \nf{DSearch} time cost between the corresponding \nf{Enqueue}, \nf{Dequeue}, in amortized we have \nf{Enqueue} takes $O(\log p +q)$ and \nf{Dequeue} takes $O(\log^2 p +q)$ steps.
+\end{proof}
 

### 2022-07-06 article/psuedocode.tex
review main theorem

@@
 \State \tt{output= null}
-\pagebreak
 \Else
@@
 
-\Statex
+\pagebreak
 
@@
 
+\begin{lemma}
+Search in each level of \nf{IndexDeq()} takes $O(\log p)$ steps.  
+\end{lemma}
+\begin{proof}
+Show preconditions are satisfied and the range is $p$.  
+\end{proof}
+
 \pagebreak
@@
 \end{proof}
-\begin{lemma}
-Operations in a block have a time point in common (There is a time $t$ all the operations are running).
-\end{lemma}
+%\begin{lemma}
+%Operations in a block have a time point in common (There is a time $t$ all the operations are running).
+%\end{lemma}
 
@@
 \begin{proof}
-Once some operations are aggregated in one block they will be propagated together up to the root and we can linearize them in any order among themselves (previous lemma).
+To show this we need to say if in an execution, \nf{op\sub{1}} terminates before \nf{op\sub{2}} starts then \nf{op\sub{1}} is linearized before \nf{op\sub{2}}.
+If \nf{op\sub{1}} terminates before \nf{op\sub{2}} starts it means \nf{op\sub{1}.Append()} is terminated before \nf{op\sub{2}.Append()} starts.
+From Lemma \ref{append} \nf{op\sub{1}} is in \nf{root.blocks} before \nf{op\sub{2}} propagates so \nf{op\sub{1}} is linearized before \nf{op\sub{2}} by Definition \ref{ordering}.
+
+Once some operations are aggregated in one block they will be propagated together up to the root and we can linearize them in any order among themselves.
 Furthermore in L we arbitrary choose the order to be by process id, since it makes computations in the blocks faster .

### 2022-07-08 article/psuedocode.tex
apply jul 7 feedbacks

@@
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{s}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
+\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \label{invokeCreateBlock} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
 \If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
@@
 
-\begin{definition}[Subblock]\label{def::subblock}
-Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (See line \ref{endDefLine} for the defined range).
-Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
-\end{definition}
-
-\begin{definition}[Superblock]
-  Block b is direct superblock of block c if c is a direct subblock of b.   Block b is superblock of block c if c is a subblock of b.
-\end{definition}
-
-\begin{definition}[Operations of a block]\label{def::ops}
-A block lb in a leaf represents one operation which if it is \tt{enqueue(x)} then \tt{lb.element=x}, otherwise \tt{element=null}.
-The set of operations of block $b$ are the operations in the subblocks of $b$.
-We show the set of operations of block b by \nf{ops(b)}.
-\end{definition}
-
-For simplicity we say block \nf{b} is propagated to node \nf{n} or to a set of blocks \nf{S} if \nf{b} is in \nf{n.blocks} or \nf{S} or is a subblock of a block in \nf{n.blocks} or \nf{S}.
-We also say $b$ contains $op$ if $op\in ops(b)$.
-
-%\begin{definition}
-%  Block \nf{b} in node \nf{n} is new if \nf{b} is not a subblock of any block in \nf{n.parent.blocks[]}.
-%\end{definition}
-
-\begin{definition}
- A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
-\emph{EST\sub{n, t}} is the set of established blocks at time $t$ of node $n$.
-\end{definition}
-
-\begin{observation}\label{head}
-Once a block \nf{b} is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
-\end{observation}
-
-\begin{lemma}[headProgress] \label{lem::headProgress}
-\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
-\end{lemma}
-\begin{proof}
-The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented.
-Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}.
-Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.\head = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
-\end{proof}
-
-\begin{lemma}
-Every block has most one direct superblock.
-\end{lemma}
-\begin{proof}
-To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}.
-\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
-Since \nf{end\sub{left},end\sub{right}} are non-decreasing, so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
-\end{proof}
-
-\begin{corollary}[No Duplicates]\label{append}
-If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
-\end{corollary}
-%\begin{proof}
-%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
-%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
-%\end{proof}
-
 \begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
@@
 \it{Explain More}
+
+\begin{lemma}[headProgress] \label{lem::headProgress}
+\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}} and \nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
+\end{lemma}
+\begin{proof}
+The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented in the pseudocode in the lines \ref{incrementHead1} and \ref{incrementHead2} of the \nf{Refresh()} .
+
+Consider the block \nf{b} written into \nf{n.blocks[i]} by \nf{TryAppend()} at Line \ref{cas}.
+It is created by the \nf{CreateBlock()} called at Line \ref{invokeCreateBlock}.
+Prior to this call to \nf{CreateBlock()}, \nf{n.head=i} at Line \ref{readHead}, so \nf{n.blocks[i-1]} is already a non-null value \nf{b}$^\prime$ by invariant \ref{lem::headPosition}.
+Thus the \nf{CreateBlock(i-1)} that creates  \nf{b}$^\prime$ terminates before \nf{CreateBlock(i)} that creates \nf{b} is invoked.
+The value written into \nf{b.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock()}  was read from \nf{n.left.head} at Line \ref{lastLine} of \nf{CreateBlock}.
+Similarly, the value in \nf{n.blocks[i-1].end\sub{left}} was read from \nf{n.left.head}.
+Since \nf{n.left.head} is non-decreasing \nf{b}$^\prime$\nf{.end\sub{left}} $\leq$ \nf{b.end\sub{left}}.
+The proof for \nf{end\sub{right}} is similar.
+\end{proof}
+
+\begin{definition}[Subblock]\label{def::subblock}
+Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()}).
+Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if \nf{b} is a direct subblock of \texttt{n.blocks[i]} or a subblock of a direct subblock of \texttt{n.blocks[i]}.
+\end{definition}
+
+\begin{definition}[Superblock]
+  Block \nf{b} is \t{direct superblock} of block \nf{c} if c is a direct subblock of b.   Block \nf{b} is \it{superblock} of block \nf{c} if \nf{c} is a subblock of \nf{b}.
+\end{definition}
+
+\begin{definition}[Operations of a block]\label{def::ops}
+A leaf block \nf{b} in a leaf represents operation \nf{x} if \tt{b.element=x$\neq$null}.
+The set of operations of block \nf{b} are the operations in the subblocks of \nf{b}.
+We denote the set of operations of block \nf{b} by \nf{ops(b)}.
+
+We say block \nf{b} is \it{propagated to node} \nf{n} if \nf{b} is in \nf{n.blocks} or is a subblock of a block in \nf{n.blocks}. We also say \nf{b} contains \nf{op} if \nf{op}$\in$\nf{ops(b)}.
+\end{definition}
+
+%\begin{definition}
+%  Block \nf{b} in node \nf{n} is new if \nf{b} is not a subblock of any block in \nf{n.parent.blocks[]}.
+%\end{definition}
+
+\begin{definition}
+ A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
+\emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
+\end{definition}
+
+\begin{observation}\label{head}
+Once a block \nf{b} is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
+\end{observation}
+
+\begin{lemma}
+Every block has most one direct superblock.
+\end{lemma}
+\begin{proof}
+To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}.
+\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
+Since \nf{end\sub{left},end\sub{right}} are non-decreasing, so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
+\end{proof}
+
+\begin{corollary}[No Duplicates]\label{append}
+If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+\end{corollary}
+%\begin{proof}
+%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
+%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
+%\end{proof}
 

### 2022-07-12 article/psuedocode.tex
add detail to get()

@@
 \If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element}
-\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{$E_{this,b,i}$ is in the left child of this node.}
-\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} 
+\State\Return \tt{blocks[b].element} \label{getBaseCase}
+\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_{this,b,i}$ is in the left child of this node.}
+\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
 \State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
@@
 
-Preconditions of all invocation of \nf{BSearch} are satisfied.
+\begin{lemma}[Get BSearch correctness] \label{getSearch}
+Preconditions of invocation of \nf{BSearch} in \nf{GetENQ} are satisfied i.e. if the direct subblock containg $E_{n,b,i}$ is computed in both cases in Lines \ref{leftChildGet} and \ref{rightChildGet}.
+\end{lemma}
+\begin{proof}
+There are two cases that the subblock we're looking for is in the left child or the right child.
+If $i$th \nf{enqueue} of \nf{n.blocks[b]} was in the left child it would be in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by definition \ref{def::subblock}.
+\nf{sum\sub{enq}} is the sum of the number of enqueues in the prefix of a block.
+So $E_{n,b,i}$ is $E_{n.left, i+n.left.blocks[n.blocks[b].end_{left}-1]}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
+The parameters in \ref{leftChildGet} are correct so  this \nf{BSearch} returns the subblock containing $E_{n,b,i}$.
+
+If the enqueue we were looking for was in the right child as there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.right.blocks[n.blocks[b].end\sub{right}-1]} (Line \ref{rightChildGet}).
+Other parameters are the same as the left child.
+\end{proof}
+
 \begin{lemma}[Get correctness] \label{get}
@@
 We are going to prove this lemma by induction on the height of the tree.
-The base case for the leaves of the tree is pretty straight forward.
-Since leaf blocks contain exactly one operation then only \nf{GetENQ(b,1)} can be called on leaves.
-\nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
-For non leaf nodes in Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}.
-From Definition \ref{ordering} we know operations in a block are ordered by  their process id.
-Furthermore \nf{b.sum\sub{enq-lef}} stores the number of \nf{enqueue()} operations from the \nf{b}'s subblocks of the left child of \nf{n}.
-So if \nf{i} is greater than \nf{b.sum\sub{enq-lef}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue i the left child subblocks.
-By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of b which their range is \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-If the enqueue we're looking for was in the right child as there are \nf{b.sum\sub{enq-left}} enqueues before it we need to search for \nf{i-b.sum\sub{enq-left}} (Line \ref{rightChildGet}).
-of By definition of $E(n,b)$ operations from the left child come  before the operations of the right child.
-Having \nf{sum\sub{enq}}, the prefix sum of the number of enqueues we can compute the direct subblock containing the enqueue we are finding for with binary search.
-Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked which returns the correct operation by the hypothesis of the induction.
+The base case, the leaves of the tree, is pretty straight forward (Line \ref{getBaseCase}).
+Since leaf blocks contain exactly one operation thus only \nf{GetENQ(b,1)} can be called on a leaf block \nf{b}.
+\nf{l.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
+In the next paragraph we prove \nf{n.GetENQ()} returns $E_{n,b,i}$, if \nf{n.child.GetENQ()} works properly.
+
+For non leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}).
+From Definition \ref{ordering} we know operations in a block are ordered by their process id and since in leaves of the tree are ordered ascendingly by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
+Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}.
+So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child.
+By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} which their range is \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+We can compute the direct subblock containing the enqueue we are finding for with binary search by Lemma \ref{getSearch}.
+Finaly \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
 \end{proof}
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-  \caption{The number of enqueues from the left and the right child}
+  \center\includegraphics[width=7in]{pics/blockSumEnq.png}
+
+  \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}.
+\nf{n.blocks[b]} is  colored in the tope node by the color of the subblocks propagted to it.
+Left subblocks in red come before the right subblocks in blue and \nf{n.blocks[b].sum\sub{enq}=n.blocks[b].sum\sub{enq-left}+n.blocks[b].sum\sub{enq-right}}.
+  }\label{figGet}
 \end{figure}
 
-\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
+%\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
 

### 2022-07-13 article/psuedocode.tex
add detail dsearch

@@
 \renewcommand{\it}[1]{\textit{#1}}
+\renewcommand{\sc}[1]{\textsc{#1}}
 \renewcommand{\bf}[1]{\textbf{#1}}
@@
 \State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
-\State \tt{<b\sub{enq}, i\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
+\State \tt{output=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
 \label{deqRest}
-\If{\tt{i\sub{enq}==-1}}
-\State \tt{output= null}
-\Else
-\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
-\EndIf
+%\If{\tt{i\sub{enq}==-1}}
+%\State \tt{output= null}
+%\Else
+%\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
+%\EndIf
 \State \Return{\tt{output}}
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\cmt{Returns \tt{<b\sub{e},i\sub{e}>}, if $E_{root,b_e,i_e}$ is the response to the $D_{root,b_d,i_d}$.
-Returns \tt{<-1,-->} if the queue is empty.}
+\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\Statex\cmt{Returns the the response to the $D_{root,b_d,i_d}$. Returns \tt{null} if the queue is empty.}
 \If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
-\State \Return \tt{<-1,-->}
+\State \Return \tt{null}
 \Else
 \State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
-\State \Return \tt{root.\Call{DSearch}{r\sub{enq}, b\sub{d}}}
+\State \Return \tt{root.GetENQ(root.\Call{DSearch}{r\sub{enq}, b\sub{d}})}
 \EndIf
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e} $\geq$ 1}}
+\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
 \Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
@@
 \While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)}
+\State \tt{start= max(start-(end-start), 0)} \label{doubling}
 \EndWhile
 \State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}}
+\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
 \State\Return \tt{<b,i>}
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=7in]{pics/blockSumEnq.png}
+  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
 
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-If \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{i} and $E_{root, i}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]} then \nf{DSearch(i, end)} returns \nf{b} such that \nf{root.blocks[b]} contains $E_{root,b,i}$ in $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) steps.
-\end{lemma}
-\begin{proof}
-First we show \nf{end-b}$\leq$\nf{root.blocks[b].\size+root.blocks[end].\size}.
-We know each block size is greater than 0.
-So every block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or one \nf{Dequeue()}.
-There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the queue would become empty after bth block end before end and $E(n,i)$ could not be the response to to some \nf{DEQ} in \nf{end}.
-And since the lentgh of the queue would become \nf{root.blocks[end].\size} in the end so there cannot be more than \nf{root.blocks[end].\size} Dequeus in \nf{root.blocks[b..end]}.
-Cause if there was more then the end's lenght would become more than \nf{root.blocks[end].\size}.
-
-Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} distance between end and b then with doubling search  in log\nf{root.blocks[b].\size+root.blocks[end].\size} steps we reach a block \nf{c} that the \nf{c.sum\sub{enq}} is less than i and the distance between c and end is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
-So the binary search takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps.
-\end{proof}
+Assume \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{e} and $E_{root, e}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]}.
+\nf{DSearch(e, end)} returns \nf{<b, i>} such $E_{root,b,i}= E_{root,e}$ in $\Theta \big(\log($\nf{root.blocks[b].\size + root.blocks[end].\size} \hspace{-0.8em}$)\big)$ steps.
+\end{lemma}
+\begin{proof}
+First we show \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
+From line \ref{addOP}, we know that size of the every block in the tree is greater than 0.
+So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or at least one \nf{Dequeue()}.
+There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the  elements of the queue after \nf{blocks[b]} would become dequeued at some point, after \nf{root.blocks[b]} operations finish end before \nf{root.blocks[end]} operations start.
+And the response to to some \nf{dequeue()} in \nf{root.blocks[end]} could not be in $E(n,b)$.
+Furthermore since the size of the queue would become \nf{root.blocks[end].\size} in the end and there cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s so there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{enqueue()} operations in \nf{root.blocks[b..end]}.
+Cause if there were more, then the \nf{root.blocks[end]}'s size would become more than \nf{root.blocks[end].\size}.
+Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.
+So \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
+See Figure \ref{end-b}.
+
+Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
+Cause if otherwise, then \nf{(end-c)/2} satisfied the \nf{root.blocks[(end-c)/2].sum\sub{enq}<e}.
+Note that in the line \ref{doubling} the distance between \nf{end} and \nf{stat} is doubled.
+See figure \ref{fig::doubling}.
+
+After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
+So the routine non constant part is the binary search which takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps from the first paragraph.
+\end{proof}
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/doubling.png}
+
+  \caption{Distance relations between $b,c,end$}
+  \label{fig::doubling}
+\end{figure}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=3in]{pics/end-b.png}
+
+  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
+  \label{end-b}
+\end{figure}
 
@@
 %\end{proof}
+\it{Update the findresposne returning arguments.}
+
 We call the dequeues that return some value \it{non-null dequeues}.

### 2022-07-15 article/psuedocode.tex
add detail findResponse

@@
 \State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b\sub{deq}, i\sub{deq}>=} \Call{IndexDeq}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b\sub{deq}, i\sub{deq}} 
+\State \tt{<b, i>=} \Call{IndexDeq}{leaf.\head, 1}
+\State \tt{output=} \Call{FindResponse}{b, i} 
 \label{deqRest}
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b\sub{d}, \sl{int} i\sub{d}}\Statex\cmt{Returns the the response to the $D_{root,b_d,i_d}$. Returns \tt{null} if the queue is empty.}
-\If{\tt{ root.blocks[b\sub{d}-1].\size}\tt{ + root.blocks[b\sub{d}].num\sub{enq} - i\sub{d} $<$ 0}}
-\State \Return \tt{null}
+\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\Statex\cmt{Returns the the response to the $D_{root,b,i}$.}
+\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \State \Return \tt{null} \label{checkEmpty}\cmt{Check if the queue is empty.}
 \Else
-\State \tt{r\sub{enq}= i\sub{d} - root.blocks[b\sub{d}-1].size + root.blocks[b\sub{d}-1].sum\sub{enq}}
-\State \Return \tt{root.GetENQ(root.\Call{DSearch}{r\sub{enq}, b\sub{d}})}
+\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
+\Statex\cmt{$E_{root,e}$ is the response.}
+\State \Return \tt{root.GetENQ(root.\Call{DSearch}{e, b})}\label{findAnswer}
 \EndIf
@@
 %\end{proof}
-\it{Update the findresposne returning arguments.}
-
-We call the dequeues that return some value \it{non-null dequeues}.
-$r$th non-null dequeue returns the element of th $r$th enqueue.
-We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= \size - \#enqueues.
-Note that the $i$th dequeue in the given block is not a non-null dequeue.
-\begin{lemma}[Computing Queue's Head block] \label{computeHead}
-  Let $S$ be the state of an empty queue if the operations in prefix in $L$ of $i$th dequeue in $D(root,b)$ are applied on it.
-\nf{FindResponse()} returns \nf{(b, i)} which $E(root, b, i)$ is the the head of the queue in $S$.
-If the queue is empty in $S$ then it returns \nf{<-1,-->}.
-  \end{lemma}
-\begin{proof}
-  The \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$ is stored in the \nf{root.blocks[b].\size}.
-It is computed while creating the block in Line~\ref{computeLength}.
+
+\begin{definition}
+In an execution on a queue, the dequeue operations that return some value are called \it{non-null dequeues}.  
+\end{definition}
+\begin{observation} \label{responseToADeq}
+$k$th non-null dequeue in an execution returns the \nf{element} of $k$th enqueue.
+\end{observation}
+
+\begin{lemma}\label{sizeCorrectness}
+  \nf{root.blocks[b].\size}is the \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$.  
+\end{lemma}
+\begin{proof}
 If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
-Having \size of the queue after the previous block and number of enqueues and dequeues in the block, Line~\ref{computeLength} computes wether the queue becomes empty or the \size  of it.
-
-  \tt{r\sub{enq}= (i\sub{d} + root.blocks[b\sub{d}-1].sum\sub{deq}) - (root.blocks[b\sub{d}-1].size - root.blocks[b\sub{d}-1].sum\sub{enq} + root.blocks[b\sub{d}-1].sum\sub{deq})}
-\end{proof}
-\framebox[1.1\width]{HOW?} How to prove mathematicaly that \nf{ax(root.blocks[i-1].\size + b.num\sub{enq} - b.num\sub{deq}, 0)} is the \size of the queue after the block. I can only explain it here.
+By definition  of $L$ in definition \ref{ordering} enqueue operations come before dequeue operations in a block.
+If the queue won't become empty after $b$th block we can compute the size of the queue after $b$th block using with this equality \nf{root.blocks[b].size= root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} (Line \ref{computeLength}).
+Otherwise \nf{root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} would become negative.
+Since there are more dequeue operations than the size of the queue.
+See table \ref{qhistory}.
+\end{proof}
+
+\begin{lemma} \label{numberOfNND}
+If the operations are applied with the order of $L$, the number of non-null dequeues in the prefix for a block \nf{b} is \nf{b.sum\sub{enq}-b.\size}
+\end{lemma}
+\begin{proof}
+If there are $enqs$ enqueue operations in the prefix for \nf{b}, then the size of the queue after  prefix for \nf{b} is \nf{enqs - non-null deqs} by Observation 35.
+The correctness of the feild \size is showed in Lemma \ref{sizeCorrectness}.
+\end{proof}
+
+\begin{lemma}\label{nullReturn}
+$D_{root,b,i}$ returns \nf{null} iff \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0}.
+\end{lemma}
+
+\begin{lemma}[Computing Response] \label{computeHead}
+  Let $S$ be the state of an empty queue if the operations in $L$ are applied on it.
+\nf{FindResponse(b,i)} returns the element that is the the response to $D_{root,b,i}$ in $S$.
+If the queue is empty in $S$ then it returns \nf{null}.
+\end{lemma}
+\begin{proof}
+First note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_{root, b, i}$ we talk about what if the whole operation in the linearization are applied on a queue.
+
+$D_{root,b,i}$ is $D_{root,root.blocks[b-1].sum_{deq}+i}$ from the definition \ref{ordering} and \it{$sum_{enq}$}.
+$D_{root,b,i}$ returns \nf{null} if \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0} by Lemma \ref{nullReturn} (Line \ref{checkEmpty}).
+Otherwise if it is $d^\prime$th non-null dequeue in $L$ it returns $d^\prime$th enqueue by Observation \ref{responseToADeq} .
+By Lemma \ref{numberOfNND} there are \nf{root.blocks[b-1].sum\sub{enq} - root.blocks[b-1].\size} non-null dequeue operations before prefix for \nf{root.blocks[b-1]}.
+Note that the dequeues in \nf{root.blocks[b]} before the $i$th dequeue are non-null dequeues.
+So the response is $E_{root,i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}}$ (Line \ref{computeE}).
+See figure \ref{computeResponseDetail}.
+
+After computing \nf{e} we can find \nf{b,i} such that $E_{root,b,i}=E_{root,e}$ using \nf{DSearch} and then find its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
+\end{proof}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $E_{root,b,i}$.}
+\label{computeResponseDetail}
+\end{figure}
+
+\begin{table}[hbt]
+\centering
+  \begin{tabular}{c|c|c|c|c}
+    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
+    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
+        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
+            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
+                size & 0 & 2 & 2 & 0 \\ \hline
+  \end{tabular}
+  \caption{An example of root blocks fields. Blocks are from left to right and operations in the blocks are also from the left to right.}\label{qhistory}
+\end{table}
+
 %\begin{proof}

### 2022-07-27 article/psuedocode.tex
review get method

@@
 \ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_{this,b,i}$ is in the left child of this node.}
-\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+left.blocks[blocks[b].end\sub{left}-1].sum\sub{enq}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
+\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
 \State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[blocks[b].end\sub{right}-1].sum\sub{enq}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
+\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
 \State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
@@
 
-\begin{lemma}[Get BSearch correctness] \label{getSearch}
-Preconditions of invocation of \nf{BSearch} in \nf{GetENQ} are satisfied i.e. if the direct subblock containg $E_{n,b,i}$ is computed in both cases in Lines \ref{leftChildGet} and \ref{rightChildGet}.
-\end{lemma}
-\begin{proof}
-There are two cases that the subblock we're looking for is in the left child or the right child.
-If $i$th \nf{enqueue} of \nf{n.blocks[b]} was in the left child it would be in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by definition \ref{def::subblock}.
-\nf{sum\sub{enq}} is the sum of the number of enqueues in the prefix of a block.
-So $E_{n,b,i}$ is $E_{n.left, i+n.left.blocks[n.blocks[b].end_{left}-1]}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
-The parameters in \ref{leftChildGet} are correct so  this \nf{BSearch} returns the subblock containing $E_{n,b,i}$.
-
-If the enqueue we were looking for was in the right child as there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.right.blocks[n.blocks[b].end\sub{right}-1]} (Line \ref{rightChildGet}).
-Other parameters are the same as the left child.
-\end{proof}
-
 \begin{lemma}[Get correctness] \label{get}
@@
 \begin{proof}
-We are going to prove this lemma by induction on the height of the tree.
-The base case, the leaves of the tree, is pretty straight forward (Line \ref{getBaseCase}).
-Since leaf blocks contain exactly one operation thus only \nf{GetENQ(b,1)} can be called on a leaf block \nf{b}.
-\nf{l.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$.
-In the next paragraph we prove \nf{n.GetENQ()} returns $E_{n,b,i}$, if \nf{n.child.GetENQ()} works properly.
-
-For non leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}).
-From Definition \ref{ordering} we know operations in a block are ordered by their process id and since in leaves of the tree are ordered ascendingly by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
+We are going to prove this lemma by induction on the height of node \nf{n}.
+The base case, where \nf{n} os a leaf, is straightforward (Line \ref{getBaseCase}).
+Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf.
+\nf{n.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf \nf{n}.
+
+For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$.
+For non-leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}).
+From Definition \ref{ordering} we know enqueue operations in a block are ordered by their process id and since in leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
 Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}.
 So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child.
-By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} which their range is \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-We can compute the direct subblock containing the enqueue we are finding for with binary search by Lemma \ref{getSearch}.
-Finaly \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
+By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+
+If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
+Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$, which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
+We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in \ref{leftChildGet} are for searching $E_{n.left, i+eb}$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_{n,b,i}$.
+
+Else if the enqueue we are looking for is in the right child then there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
+Other parameters are assigned similar for the left child.
+So in both cases the direct subblock containing $E_{n,b,i}$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
+
+ Finally, \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
 \end{proof}
@@
   \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}.
-\nf{n.blocks[b]} is  colored in the tope node by the color of the subblocks propagted to it.
-Left subblocks in red come before the right subblocks in blue and \nf{n.blocks[b].sum\sub{enq}=n.blocks[b].sum\sub{enq-left}+n.blocks[b].sum\sub{enq-right}}.
+Enqueue operations from the left subblocks, colored red, are ordered before the enequeue operations from the right child, colored blue.
   }\label{figGet}

### 2022-07-27 article/psuedocode.tex
review dsearch

@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-Assume \nf{root.blocks[end].num\sub{enq}}$\geq$\nf{e} and $E_{root, e}$ is the response to some \nf{Dequeue()} in \nf{root.blocks[end]}.
-\nf{DSearch(e, end)} returns \nf{<b, i>} such $E_{root,b,i}= E_{root,e}$ in $\Theta \big(\log($\nf{root.blocks[b].\size + root.blocks[end].\size} \hspace{-0.8em}$)\big)$ steps.
-\end{lemma}
-\begin{proof}
+Assume \nf{root.blocks[end].sum\sub{enq}}$\geq$\nf{e} and $E_{root, e}$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
+\nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_{root,b,i}= E_{root,e}$.
+\end{lemma}
+\begin{proof}
+It is trivial to see that the doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} will find $E_{root,e}$ eventually.
+Because \nf{root.blocks[].sum\sub{enq}} is an increasing value from 0 to some value greater than \nf{e}.
+So there is a \nf{b} that \nf{root.blocks[b].sum\sub{enq}}$>e$ but \nf{root.blocks[b-1].sum\sub{enq}}$<e$.
+
 First we show \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
 From line \ref{addOP}, we know that size of the every block in the tree is greater than 0.
-So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue()} or at least one \nf{Dequeue()}.
-There cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s in \nf{root.blocks[b+1..end-1]}, since the  elements of the queue after \nf{blocks[b]} would become dequeued at some point, after \nf{root.blocks[b]} operations finish end before \nf{root.blocks[end]} operations start.
-And the response to to some \nf{dequeue()} in \nf{root.blocks[end]} could not be in $E(n,b)$.
-Furthermore since the size of the queue would become \nf{root.blocks[end].\size} in the end and there cannot be more than \nf{root.blocks[b].\size} \nf{Dequeue()}s so there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{enqueue()} operations in \nf{root.blocks[b..end]}.
-Cause if there were more, then the \nf{root.blocks[end]}'s size would become more than \nf{root.blocks[end].\size}.
+So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}.
+Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}.
+Then the  queue would become empty at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation.
+Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$.
+Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{Enqueue}s.
+Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size is \nf{root.blocks[end].\size}.
 Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.
@@
 Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
-Cause if otherwise, then \nf{(end-c)/2} satisfied the \nf{root.blocks[(end-c)/2].sum\sub{enq}<e}.
-Note that in the line \ref{doubling} the distance between \nf{end} and \nf{stat} is doubled.
-See figure \ref{fig::doubling}.
+Beause otherwise, then \nf{(end-c)/2} satisfied the \nf{root.blocks[(end-c)/2].sum\sub{enq}<e}.
+In line \ref{doubling} the differnece between \nf{end} and \nf{start} is doubled.
+See Figure \ref{fig::doubling}.
 
@@
 So the routine non constant part is the binary search which takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps from the first paragraph.
+
 \end{proof}

### 2022-08-02 article/psuedocode.tex
small  changes

@@
 
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
+\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
 \Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
@@
 A block is an object storing some statistics, as described in Algorithm Queue.
-It implicitly represents a set of operations.
+A block in a node's blocklist implicitly represents a set of operations.
 If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
@@
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-
-Some value is written into \nf{n.blocks[head]} only in Line 313.
-It is obvious that writing into \nf{n.blocks[head]} presereves the invariant.
+We show the invariant still holds after these two changes.
+
+In the algorithm, some value is appended to \nf{n.blocks[]} by writting into \nf{n.blocks[head]} only in Line 313.
+It is obvious that writing into \nf{n.blocks[head]} presereves the invariant, since the claims does not talk about \nf{n.blocks[head]}.
 The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}.
 Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case.
-If \nf{head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds.
+If \nf{n.head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds.
 In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
-Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readHead}.
+Note that wether \ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readHead}.
 The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
@@
 \begin{lemma}[head Increment] \label{lem::headInc}
-If an n.Refresh  instance reaches Line \ref{cas} instance and reads head=h (\ref{readHead}) after it terminates head is greater than h.
-\end{lemma}
-\begin{proof}
-If Line \ref{incrementHead1} or \ref{incrementHead1} succeeded the claim holds, otherwise another process has incremented the head.
+Let \nf{R=n.Refresh} be an instance of \nf{Refresh} that reaches Line \ref{cas}. After \nf{R} terminates \nf{n.head} is greater than h, the value read  in line \ref{readHead} of \nf{R}.
+\end{lemma}
+\begin{proof}
+If Line \ref{incrementHead1} or \ref{incrementHead1} are successful then the claim holds, otherwise another process has incremented the head from \nf{h} to \nf{h+1}.
 \end{proof}
@@
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Let $t_i$ be the time an instance of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates.
-Suppose the \nf{TryAppend(new, s)} of the \nf{n.Refresh()}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
+  Let $t_i$ be the time an instance \nf{R} of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates.
+Suppose the \nf{TryAppend(new, s)} of \nf{R}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
 \end{lemma}
@@
 \begin{corollary}[Propagate Step] \label{doublyRefresh}
-All operations in \nf{n}'s children's established blocks before line \ref{firstRefresh} are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
+All operations in \nf{n}'s children's established blocks before running line \ref{firstRefresh}  of a \nf{Propagate} routine are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
 \end{corollary}
 \begin{proof}
-Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
+If \ref{firstRefresh} or \ref{secondRefresh} succeed, the claim is true by \ref{lem::trueRefresh}.
+Otherwise Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
 \end{proof}
@@
 \begin{proof}
-  Follows from Lemma \ref{doubleRefresh}, \ref{append}.
+After \nf{Append(blk)}'s termination, \nf{blk} is in \nf{root.blocks} since \nf{blk} is established  in the leaf it has been added to.
+By induction on \ref{doublyRefresh} it is propagated up to the root.
+Finally \ref{append} shows  only one block in the root contains \nf{blk}.
 \end{proof}
@@
 By proof of contradiction, assume there are more than one operation from process $p$ in block $b$ in node $n$.
-A process cannot invoke more than one operations concurrently.
-From $p$ 's operations in $b$, let $op_1$ be the first operation invoked and $op_2$ be the second one.
-Note that it is terminated before $op_2$ started.
-So before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
+A process cannot invoke more than one operations concurrently, so one operation hasto be finished before the other starts.
+From $p$ 's operations in $b$, let $op_1$ be the operation invoked first, before $op_2$.
+Note that $op_1$ has terminated before $op_2$ started.
+So  by Corollary 17 before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
+Because $op_1$'s \nf{Append} is finished before $op_2$'s \nf{Append} starts.
 So there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
@@
 \begin{proof}
-  It follows directly from Lemma \ref{blockSize} and the observation that each block contains at least one operation, induced from Line~\ref{addOP}. 
+  It follows directly from Lemma \ref{blockSize} and the observation that each block appended to the tree contains at least one operation, induced from Line~\ref{addOP}. 
 \end{proof}

### 2022-08-03 article/psuedocode.tex
review findResponse()

@@
 %\end{proof}
+\begin{definition}
+   Assume the operations in $L$ are applied on an empty queue.
+If element of \nf{enqueue e} is the response to  \nf{dequeue d} then we say \nf{R(d)=e}.
+If \nf{d} 's response id \nf{null} (queue is empty) then \nf{R(d)=null}.
+\end{definition}
 
@@
 \begin{proof}
-If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
-By definition  of $L$ in definition \ref{ordering} enqueue operations come before dequeue operations in a block.
-If the queue won't become empty after $b$th block we can compute the size of the queue after $b$th block using with this equality \nf{root.blocks[b].size= root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} (Line \ref{computeLength}).
-Otherwise \nf{root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} would become negative.
-Since there are more dequeue operations than the size of the queue.
-See table \ref{qhistory}.
-\end{proof}
-
-\begin{lemma} \label{numberOfNND}
+ need to say?
+:: If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
+By definition \ref{ordering} enqueue operations come before dequeue operations in a block in $L$.
+
+We prove the claim by induction on \nf{b}.
+Base case \nf{b=0} is trivial since the queue is initialy empty and \nf{root.blocks[0].size=0}.
+For \nf{b=i} we are going to use the hypothesis for \nf{b=i-1}.
+If there are more than \nf{root.blocks[i-1].size+ root.blocks[i].sum\sub{enq}} dequeue operations in \nf{root.blocks[i]} then the queue would become empty after \nf{root.blocks[i]}.
+Otherwise we can compute the size of the queue after $b$th block using with this equality \nf{root.blocks[b].size= root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} (Line \ref{computeLength}).
+See Table \ref{qhistory} for an example of running some blocks of operations on an empty queue.
+\end{proof}
+
+\begin{lemma}[Duality of \#non-null dequeues and \nf{block.size}] \label{numberOfNND}
 If the operations are applied with the order of $L$, the number of non-null dequeues in the prefix for a block \nf{b} is \nf{b.sum\sub{enq}-b.\size}
@@
 \begin{proof}
-If there are $enqs$ enqueue operations in the prefix for \nf{b}, then the size of the queue after  prefix for \nf{b} is \nf{enqs - non-null deqs} by Observation 35.
-The correctness of the feild \size is showed in Lemma \ref{sizeCorrectness}.
+There are \nf{b.sum\sub{enq}} enqueue operations in the prefix for \nf{b}, then the size of the queue after the prefix for \nf{b} is \nf{\#enqs - \#non-null dequeues} in the prefix for \nf{b}, by Observation 35.
+So \nf{\#non-null dequeues} is \nf{b.sum\sub{enq}-b.\size}.
+The correctness of the \nf{block.\size}field is shown in Lemma \ref{sizeCorrectness}.
 \end{proof}
@@
 \begin{lemma}\label{nullReturn}
-$D_{root,b,i}$ returns \nf{null} iff \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0}.
+\nf{R(D\sub{root,b,i})} is \nf{null} iff \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0}.
 \end{lemma}
@@
 \begin{lemma}[Computing Response] \label{computeHead}
-  Let $S$ be the state of an empty queue if the operations in $L$ are applied on it.
-\nf{FindResponse(b,i)} returns the element that is the the response to $D_{root,b,i}$ in $S$.
-If the queue is empty in $S$ then it returns \nf{null}.
+\nf{FindResponse(b,i)} returns \nf{R(D\sub{root,b,i}).element}.
 \end{lemma}

### 2022-08-10 article/psuedocode.tex
add related, bug fix

@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.\head} \label{lastLine}
+\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
@@
 \Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
-\State \tt{\head+=1} \label{appendEnd} 
 \State \tt{blk.group= \head} \label{appendStart} 
 \State \tt{blocks[\head]= blk} 
+\State \tt{\head+=1} \label{appendEnd} 
 \State \tt{parent.}\Call{Propagate()}{} 
@@
 
+\begin{lemma}[head Increment] \label{lem::headInc}
+Let \nf{R} be an instance of \nf{Refresh} on node \nf{n} that reaches Line \ref{cas}. After \nf{R} terminates \nf{n.head} is greater than h, the value read  in line \ref{readHead} of \nf{R}.
+\end{lemma}
+\begin{proof}
+If Line \ref{incrementHead1} or \ref{incrementHead2} are successful then the claim holds, otherwise another process has incremented the head from \nf{h} to \nf{h+1}.
+\end{proof}
+
 \begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
@@
 
-In the algorithm, some value is appended to \nf{n.blocks[]} by writting into \nf{n.blocks[head]} only in Line 313.
-It is obvious that writing into \nf{n.blocks[head]} presereves the invariant, since the claims does not talk about \nf{n.blocks[head]}.
-The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}.
-Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case.
+In the algorithm, some value is appended to \nf{n.blocks[]} by writing into \nf{n.blocks[head]} only in Line 313.
+Writing into \nf{n.blocks[head]} preserves the invariant, since the claim does not talk about \nf{n.blocks[head]}.
+The value of \nf{n.head} is modified only in lines \ref{incrementHead1} and \ref{incrementHead2}.
+Depending on whether the \nf{TryAppend()} in Line~\ref{cas} succeeded or not, we show that the claim holds after the increment of \nf{n.head} in either case.
 If \nf{n.head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds.
 In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
-Note that wether \ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readHead}.
-The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
+Note that whether \ref{incrementHead1} or \ref{incrementHead1} return true or false,  after they finish we know  that \nf{n.head} has been incremented from the value read in Line \ref{readHead} (Lemma \ref{lem::headInc}).
+The failure case is also the same since it means some non-null value has been written into \nf{n.blocks[head]} by some process.
 %At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
@@
 \begin{lemma}[headProgress] \label{lem::headProgress}
-\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}} and \nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
-\end{lemma}
-\begin{proof}
-The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented in the pseudocode in the lines \ref{incrementHead1} and \ref{incrementHead2} of the \nf{Refresh()} .
+\nf{n.head} is non-decreasing over time.
+If \nf{n.blocks[i]$\neq$null} and \nf{i$.$0} then \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}} and \nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
+\end{lemma}
+\begin{proof}
+The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented in the pseudocode in lines \ref{incrementHead1} and \ref{incrementHead2} of \nf{Refresh()} .
 
 Consider the block \nf{b} written into \nf{n.blocks[i]} by \nf{TryAppend()} at Line \ref{cas}.
-It is created by the \nf{CreateBlock()} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock()}, \nf{n.head=i} at Line \ref{readHead}, so \nf{n.blocks[i-1]} is already a non-null value \nf{b}$^\prime$ by invariant \ref{lem::headPosition}.
+It is created by the \nf{CreateBlock(i)} called at Line \ref{invokeCreateBlock}.
+Prior to this call to \nf{CreateBlock(i)}, \nf{n.head=i} at Line \ref{readHead}, so \nf{n.blocks[i-1]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}.
 Thus the \nf{CreateBlock(i-1)} that creates  \nf{b}$^\prime$ terminates before \nf{CreateBlock(i)} that creates \nf{b} is invoked.
-The value written into \nf{b.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock()}  was read from \nf{n.left.head} at Line \ref{lastLine} of \nf{CreateBlock}.
-Similarly, the value in \nf{n.blocks[i-1].end\sub{left}} was read from \nf{n.left.head}.
+The value written into \nf{b.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock(i)}  was read from \nf{n.left.head-1} at Line \ref{lastLine} of \nf{CreateBlock(i)}.
+Similarly, the value in \nf{n.blocks[i-1].end\sub{left}} was read from \nf{n.left.head-1} during the call to \nf{CreateBlock(i-1)}.
 Since \nf{n.left.head} is non-decreasing \nf{b}$^\prime$\nf{.end\sub{left}} $\leq$ \nf{b.end\sub{left}}.
@@
 \begin{definition}[Subblock]\label{def::subblock}
-Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()}).
+Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} .
 Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if \nf{b} is a direct subblock of \texttt{n.blocks[i]} or a subblock of a direct subblock of \texttt{n.blocks[i]}.
@@
 
+%(The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()})
+
 \begin{definition}[Superblock]
-  Block \nf{b} is \t{direct superblock} of block \nf{c} if c is a direct subblock of b.   Block \nf{b} is \it{superblock} of block \nf{c} if \nf{c} is a subblock of \nf{b}.
+  Block \nf{b} is   \it{direct superblock} of block \nf{c} if c is a direct subblock of b.   Block \nf{b} is \it{superblock} of block \nf{c} if \nf{c} is a subblock of \nf{b}.
 \end{definition}
@@
 \begin{definition}[Operations of a block]\label{def::ops}
-A leaf block \nf{b} in a leaf represents operation \nf{x} if \tt{b.element=x$\neq$null}.
+A leaf block \nf{b} in a leaf represents \nf{enqueue(x)} if \tt{b.element=x$\neq$null}.
+Else if \tt{b.element=null} b represents a \nf{dequeue()}.
 The set of operations of block \nf{b} are the operations in the subblocks of \nf{b}.
@@
 \begin{definition}
- A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if the last value written into \nf{n.head} before $t$ is greater than the index of \nf{b} in \nf{n.blocks} at time $t$.
-\emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
+ A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if \nf{n.head}$>$ index of \nf{b} at time $t$. \emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
 \end{definition}
@@
 \begin{lemma}
-Every block has most one direct superblock.
+Every block has at most one direct superblock.
 \end{lemma}
@@
 \nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
-Since \nf{end\sub{left},end\sub{right}} are non-decreasing, so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
+Since \nf{end\sub{left},end\sub{right}} are non-decreasing (\nf{n.blocks[i].end\sub{left|right}>n.blocks[i-1].end\sub{left|right}}), so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
 \end{proof}
@@
 \begin{proof}
-Blocks are only appended (not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node can only grows.
-\end{proof}
+Blocks are only appended (not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node can only grow.
+\end{proof}
+\textit{useless?}
 
@@
 %\end{proof}
-
-\nf{CreateBlock()} aggregates the blocks in the children that are not already established in the parent into one block.
+%
+%\nf{CreateBlock()} aggregates the established blocks in the children that are not already established in the parent into one block.
 If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
@@
 After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-
-\pagebreak
-
-\begin{lemma}[head Increment] \label{lem::headInc}
-Let \nf{R=n.Refresh} be an instance of \nf{Refresh} that reaches Line \ref{cas}. After \nf{R} terminates \nf{n.head} is greater than h, the value read  in line \ref{readHead} of \nf{R}.
-\end{lemma}
-\begin{proof}
-If Line \ref{incrementHead1} or \ref{incrementHead1} are successful then the claim holds, otherwise another process has incremented the head from \nf{h} to \nf{h+1}.
-\end{proof}
+%\text
+\pagebreak
 
@@
   Let $t_i$ be the time an instance \nf{R} of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates.
-Suppose the \nf{TryAppend(new, s)} of \nf{R}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
-\end{lemma}
-\begin{proof}
-From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
-If \nf{TryAppend} returns true a block \nf{new} is written into \nf{n.blocks[h]} (Line \ref{cas}).
+If the \nf{TryAppend(new, s)} of \nf{R}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
+\end{lemma}
+\begin{proof}
+Since \nf{TryAppend} returns true a block \nf{new} is written into \nf{n.blocks[h]} in Line \ref{cas}.
+%So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
+
 We show \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-The proof for the right child's claim is the same.
-Let \nf{n.left.head} at \nf{t\sub{i}} be \nf{hli}.
-Let \nf{n.Refresh()} read \nf{head} equal to $h$(Line \ref{readHead}).
-By the lines \ref{prevLine},\ref{lastLine} the \nf{new} block in \nf{n.blocks[h]} contains \nf{n.left.blocks[ n.blocks[h-1].end\sub{left}+1..left.head]}.
-Since \nf{left.head} is read after \nf{t\sub{i}} then \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..left.head])}.
-By Lemma \ref{lem::establishedOrder} \nf{ops(n.left.blocks[0..n.blocks[h-1].end\sub{left}])}$\subseteq$\nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-Since after line \ref{incrementHead2} we are sure that the \nf{head} is incremented (Lemma\ref{lem::headInc}) and \nf{n.head=h+1} at $t_t$ so the new block is established at $t_t$ and the new block contains the new operations which is what we wanted to show.
+Let \nf{h} be the value \nf{n.Refresh()} reads from \nf{n.head} at line \ref{readHead}, \nf{h\sub{left,i}} be the value of \nf{n.left.head} at \nf{t\sub{i}} and \nf{h\sub{left,read}} be the value read from \nf{n.left.head-1} at line \ref{lastLine}.
+\nf{end\sub{left}} field of the block returned by \nf{CreateBlock(i)} is \nf{h\sub{left,read}}.
+By lines \ref{prevLine} and \ref{lastLine} the \nf{new} block in \nf{n.blocks[h]} contains \nf{n.left.blocks[n.blocks[h-1].end\sub{left}+1..h\sub{left,read}]}.
+Since \nf{left.head} is read after \nf{t\sub{i}} then \nf{h\sub{left,read}>h\sub{left,i}} which means \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..h\sub{left,read}])}.
+After the successful \nf{TryAppend} in line \ref{cas} we know all blocks in \nf{n.left.blocks[0..h\sub{left,read}-1} are subblocks of \nf{n.blocks[0..h]} by the definition of subblock.
+At \nf{t\sub{t}} we have \nf{n.head>h} by Lemma \ref{lem::headProgress}.
+So \nf{n.blocks[1..h]} are in \nf{EST\sub{n,t\sub{t}}} by definition of \nf{EST}.
+Note that after line \ref{incrementHead2} we are sure that the \nf{head} is incremented by Lemma\ref{lem::headInc}) which means \nf{n.head=h+1} at $t_t$ so the new block is established at $t_t$ and the new block contains the new operations which is what we wanted to show.
+The proof for \nf{ops(EST\sub{n.right, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}is the same.
+
 % From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$.
@@
 \end{figure}
-\begin{lemma}[Precise True Refresh] \label{lem::prectrueRefresh}
+\begin{lemma}[Stronger True Refresh] \label{lem::prectrueRefresh}
   Let $t_i$ be the time an instance of \nf{n.Refresh()} read the head (Line \ref{readHead}) and $t_t$ be the time its \nf{TryAppend(new, s)} terminates with  and returns \nf{true} (Line \ref{cas}).
@@
 \begin{proof}
-If \ref{firstRefresh} or \ref{secondRefresh} succeed, the claim is true by \ref{lem::trueRefresh}.
+If \ref{firstRefresh} or \ref{secondRefresh} succeed, the claim is true by Lemma \ref{lem::trueRefresh}.
 Otherwise Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
@@
 
-\begin{corollary}
-  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for some \nf{x} and only one \nf{x}.
+\begin{corollary}\label{lem::appendExactlyOnce}
+  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for exactly one \nf{x}.
 \end{corollary}
@@
 After \nf{Append(blk)}'s termination, \nf{blk} is in \nf{root.blocks} since \nf{blk} is established  in the leaf it has been added to.
-By induction on \ref{doublyRefresh} it is propagated up to the root.
-Finally \ref{append} shows  only one block in the root contains \nf{blk}.
+By applying Lemma \ref{doublyRefresh} inductively it is propagated up to the root.
+Finally Lemma \ref{append} shows  only one block in the root contains \nf{blk}.
 \end{proof}
@@
 \begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block contains at most one operation from each processs.
-\end{lemma}
-\begin{proof}
-By proof of contradiction, assume there are more than one operation from process $p$ in block $b$ in node $n$.
-A process cannot invoke more than one operations concurrently, so one operation hasto be finished before the other starts.
-From $p$ 's operations in $b$, let $op_1$ be the operation invoked first, before $op_2$.
-Note that $op_1$ has terminated before $op_2$ started.
-So  by Corollary 17 before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
+Each block contains at most one operation of each processs.
+\end{lemma}
+\begin{proof}
+To derive a contradiction, assume there are 2 operations $op_1$ and $op_2$ from process $p$ in block $b$ in node $n$.
+WLOG $op_1$ is invoked earlier than $op_2$.
+A process cannot invoke more than one operations concurrently, so one operation has to be finished before the other starts.
+So $op_1$ has terminated before $op_2$ started.
+By Corollary \ref{lem::appendExactlyOnce} before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
 Because $op_1$'s \nf{Append} is finished before $op_2$'s \nf{Append} starts.
 So there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
-$op_1$ existing in $b$ an $b^\prime$ contradicts with \ref{append}.
+$op_1$ existing in $b$ an $b^\prime$ contradicts Lemma \ref{append}.
 \end{proof}

### 2022-08-10 article/related.tex
add related, bug fix

--- 2022-03-23 proposal/proposal.tex
@@
 \documentclass[12pt]{article}
-\usepackage[a4paper]{geometry}
-\usepackage{hyperref}
+\renewcommand{\baselinestretch}{1.8}
+
+\usepackage[b4paper]{geometry}
+\usepackage{tikz-qtree}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\makeatletter
+\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
+\makeatother
 \usepackage{graphicx}
 \usepackage{subcaption}
-\usepackage{tikz-qtree}
+\usepackage{showkeys}
+\usepackage{amsmath}
 
-\renewcommand{\baselinestretch}{1.5}
+\usepackage{multicol}
+\setlength\columnsep{24pt}
+
+\algnewcommand\algorithmicforeach{\bf{for each}}
+\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
+
+\algdef{S}[FUNCTION]{Function}
+   [3]{{\tt{\sl{#1}}} \textproc{\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
+
+\algdef{E}[FUNCTION]{EndFunction}
+   [1]{\algorithmicend\ \tt{\textproc{#1}}}
+
+\algrenewcommand\Call[2]{\tt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+
+\newcommand\keywordfont{\sffamily\bfseries}
+\algrenewcommand\algorithmicend{{\keywordfont end}}
+\algrenewcommand\algorithmicfor{{\keywordfont for}}
+\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
+\algrenewcommand\algorithmicdo{{\keywordfont do}}
+\algrenewcommand\algorithmicuntil{{\keywordfont until}}
+\algrenewcommand\algorithmicfunction{{\keywordfont function}}
+\algrenewcommand\algorithmicif{{\keywordfont if}}
+\algrenewcommand\algorithmicthen{{\keywordfont then}}
+\algrenewcommand\algorithmicelse{{\keywordfont else}}
+\algrenewcommand\algorithmicreturn{{\keywordfont return}}
+
+\renewcommand\thealgorithm{}
+\newcommand{\setalglineno}[1]{%
+  \setcounter{ALC@line}{\numexpr#1-1}}
+
+\newcommand{\sub}[1]{\textsubscript{#1}}
+\renewcommand{\tt}[1]{\texttt{#1}}
+\renewcommand{\sl}[1]{\textsl{#1}}
+\renewcommand{\it}[1]{\textit{#1}}
+\renewcommand{\sc}[1]{\textsc{#1}}
+\renewcommand{\bf}[1]{\textbf{#1}}
+\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
+\newcommand{\cmt}[1]{\Comment{#1}}
+\newcommand{\head}{head}
+\newcommand{\size}{size }
+
+\usepackage{amsmath,amssymb,amsthm}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{observation}[theorem]{Observation}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{invariant}[theorem]{Invariant}
 
 \begin{document}
-
-\title{Thesis Proposal}
-\author{Hossein Naderibeni \\supervised by Eric Ruppert}
-\maketitle
-
-%\begin{enumerate}
-%  \item Importance of implementing shared data structures
-%\begin{itemize}
-%  \item how it helps developers
-%  \item why lock free data structures are better than using locks?
-%\end{itemize} 
-%  \item  What properties matter for a distributed data structure?
-%\begin{itemize}
-%  \item Correctness condition: Linearizability definition
-%  \item Wait-freeness
-%\end{itemize} 
-%  \item Previous works on distributed queues: more detailed
-%\begin{itemize}
-%  \item why we are studying queues
-%  \item $p$ factor in time complexity of all previous algorithms
-%  \item types of current queues in the literature
-%  \item previous impractical universal constructions that use long words
-%\end{itemize}
-%  \item Our contribution
-%\begin{itemize}
-%  \item polylogarithmic algorithm with respect to $p$ and $Q$ size of the queue
-%\end{itemize}
-%  \item Explaining ideas
-%\begin{itemize}
-%  \item using tournament tree to agree on the linearization order
-%  \item double refresh
-%  \item how to compute a dequeue response
-%  \item how to be sure the number of blocks in the root is related to the size of the queue
-%  \item memory management as a goal
-%\end{itemize}
-%\end{enumerate}
-
-\pagebreak
-
-\section{Introduction}
-
-\paragraph{}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-This is why distributed algorithms have become important.
-It is not hard to see why multiple processes cannot update sequential data structures designed for one process.
-For example, consider two processes trying to insert some values into a sequential linked list simultaneously.
-Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same.
-In this run, $p$'s update is overwritten.
-One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
-Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
-%We are going to resolve a common problem in previous shared queues which makes their worst case slow.
-
-The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[t]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[t]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption{\label{fig::badexample}An example of an execution that is not linearizable.
-Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every so often but never terminates.
-%From now on, we are going to focus on wait freedom.
-
-%Since executions on a shared data structure are different from sequential ones, the correctness condition also differs.
-The strongest correctness condition is linearizability, defined as "...".
-Lock-free is good, but a wait-free algorithm guarantees us every operation terminates after finite steps.
 
@@
 
-\begin{figure}[t]
-  \center\includegraphics[scale=0.5]{pics/msqueue}
-  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrows show new pointers and gray dashed arrows show the old pointers.}
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.4]{pics/msqueue}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
+In the first diagram the first element has been dequeued.
+Red arrows show new pointers and gray dashed arrows show the old pointers.}
 \end{figure}
@@
 
-\begin{figure}[t]
+\begin{figure}[hbt]
   \center\includegraphics[scale=0.3]{pics/baskets}
@@
 
-\begin{figure}[t]
+\begin{figure}[hbt]
   \center\includegraphics[scale=0.5]{pics/sundell}
@@
 
-\subsection{Universal constructions}
+\subsection{Universal Constructions}
 Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
@@
 Some impractical universal constructions are using big words as "...".
+\section{Adiitional ones}
+\subsection{Tsigas and Zhang}
+Tsigas and Zhang \cite{DBLP:conf/spaa/TsigasZ01} present a cyclic array queue that allows the head and tail to lag at most m nodes behind the actual head and tail of the queue, so the amortized number of CAS executions per operation is 1 + 1/m.
+Their algorithm is limited to bounded queues due to their static allocation.
 
-\section{Our Contribution}
-In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, avoiding the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays.
-We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
-Then we use the block tree as a stepping stone towards our queue algorithm.
-A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}).
-Each process has a leaf, and it appends its operations to its leaf.
-After that, the process tries to propagate its new operation up to the tree's root.
-An ordering of operations propagated up to a node is stored in that node.
-All processes agree on the sequence stored in the root and this is used as the linearization ordering.
-Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.
+\subsection{Kogan and Herlihy}
+Previous works present concurrent constructs that combine multiple operations into a single operation on the shared object.
+We chose to combine operations and apply them as batches, to increase scalability.
+The work of Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} is the closest to this work.
+They propose alternative definitions for linearizability of executions with batches, including MF-linearizability, which we use.
+They describe very simple implementations of stacks, queues, and linked lists that demonstrate the benefits of using futures.
 
-\begin{figure}[t]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root, which stores the final ordering of all operations.}
-\end{figure}
+\subsection{Yang and Mellor-Crummey}
+Because of the problem of failure of CASes one way is to use other primitive objects other than CAS objects.
+Yang and Mellor-Crummey \cite{DBLP:conf/ppopp/YangM16} utilize fetch-and-add as well, to form a wait-free queue.
+They present the first linearizable and wait-free FIFO queue based on fetch-and-add (FAA).
+While compare-and-swap (CAS) based non-blocking algorithms may perform poorly due to work wasted by CAS failures, algorithms that coordinate using FAA, which is guaranteed to succeed, can in principle perform better under high contention.
 
-The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
-Then, a dequeue operation uses the linearization ordering to compute its answer.
-
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation.
-The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
-%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
-Instead of storing operations explicitly in the nodes, we only keep track of the number of them.
-This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.
-
-\begin{figure}[t]
-\begin{center}
-\begin{subfigure}[b]{.49\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}}
-  \caption{Operations after $||$ are new.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.49\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}}
-  \caption{New operations are added to the parent node.}
-\end{subfigure}
-\caption{\label{fig::propagstep} Successful \texttt{Refresh}, operations in children after $||$ are new.}
-\end{center}
-\end{figure}
-
-We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
-\texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf.
-\texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root.
-The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
-After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
-Now we can implement Enqueue and Dequeue using our block tree.
-An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
-To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
-Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
-Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
-
-\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation.
-Each block stores a constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
-
-The algorithm works as a queue, but \texttt{Get(i)}  may take a long time since it has to find the block containing the $i$th operation at the root level.
-We came up with a garbage collection only for the root level to reduce the time to find this block to $O(\log q)$.
-The main idea is to remove a block after its operations are no longer needed.
-We will store the blocks in the root in a persistent red-black tree to be able only to maintain the remaining information of the queue.
-It is possible to search in persistent red-black trees with size $n$ in $O(\log n)$ steps and split the tree by a key in $O(\log n)$ steps.
-Finally, the algorithm works in $O(\log^2 p +\log q)$ steps.
-
-\section{Next steps}
-The current algorithm only works as a queue, but other data structures like stacks share similarities.
-It might be possible to come up with a generalization for all these types.
-We have not yet considered space usage, and we use infinite-sized arrays that are not practical.
-Our future work is to manage the memory of the tree and make it practical.
-
-\section{Planning}
-The plan is to finish the writing of the thesis by August 2021. The thesis will contain:
-\begin{itemize}
-  \item More detailed discussion of previous work
-  \item Description of the algorithm, layer by layer
-  \item Pseudocode of the algorithm
-  \item Proof of linearizability and time analysis
-\end{itemize}
-
-We expect to submit a paper of our work to \textit{36th International Symposium on Distributed Computing
-October 25-27, 2022, Augusta, Georgia, USA} by May 13.
+I could not compute its complexity but I guess it is hard or not important
 
 \bibliography{main}
-\bibliographystyle{ieeetr}
 

### 2022-08-14 article/psuedocode.tex
review double refresh, time analysis

@@
 \begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive failed instances $R_1$, $R_2$ of \nf{n.Refresh()} by some process.
+  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on \nf{n} by process $p$.
 Let $t_1$ be the time $R_1$ is invoked and $t_2$ be the time $R_2$ terminated.
-We have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+If $R_1$ and $R_2$ both fail and return false, then we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 \end{lemma}
@@
 Since $R_2$'s \nf{TryAppend()} returns \nf{false} then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to append.
-In Figure 1 we see why the block $R_2^\prime$ is appending contains established block in the n's children at $t_1$, since it create a block reading the head after $t_1$.
+Since $R_2^\prime$ creates the block after reading the value $i+1$ from \nf{n.head}(Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head} and the \nf{head}'s value is increaing by Lemma \ref{lem::headProgress} then $t_{R2^\prime \; \ref{readHead}} > t_{R1 \; \ref{readHead}} > t_1$ (See Figure \ref{fig::doubleRefresh}).
 By Lemma \ref{lem::prectrueRefresh} after $R_2^\prime$'s \nf{CAS} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(n.blocks)}.
-Also by Lemma \ref{lem::headInc} of $R_2$  head is more than $i+1$ after $R_2$\'s \ref{incrementHead2} line, so the block appended by $R_2^\prime$ to \nf{n} is established by then.
-To summarized $t_1$ is before $R_2^\prime$'s read head and $R_2^\prime$'s \nf{CAS} is before $R_2$'s termination.
-So \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2$\' terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by then (\nf{n.head>=i+2>i+1}).
+To summarize $t_1$ is before $R_2^\prime$'s \nf{read n.head} and $R_2^\prime$'s successful \nf{CAS} is before $R_2$'s termination.
+So by Lemma \ref{lem::prectrueRefresh} \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
 \end{proof}
-\it{last sentence need more detail and should be earlier. define i and tell why R2prime exists}
 
@@
   \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption{$t1<$ $r_1$ reading head $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $R_2^\prime$ reading head $<$ \nf{TryAppend(i+1)} $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t2$}
+  \caption{\label{fig::doubleRefresh}$t1<$ $r_1$ reading head $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $R_2^\prime$ reading head $<$ \nf{TryAppend(i+1)} $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t2$}
 \end{figure}
@@
 If \nf{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \nf{n}.
+
+\begin{definition}
+  $t_{before \; line}$ is the immediate time before running Line $line$.   $t_{before \; line}$ is the immediate time after running Line $line$.
+\end{definition} 
+\begin{corollary}
+  \nf{ops(EST\sub{n.left, t\sub{before \ref{firstRefresh}}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{before \ref{firstRefresh}}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{after \ref{secondRefresh}})}}
+\end{corollary}
+\begin{proof}
+  If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::trueRefresh} the claim holds.
+Also if first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::trueRefresh}.
+Finally, if both failed the claim is satisfied by Lemma \ref{doubleRefresh}.
+\end{proof}
 
@@
 Amortized time consumed for helping by each process is $O(\log^2 p)$.
-\nf{Enqueue()} and \nf{Dequeue} take $O(\log^2 p +q)$ steps (amortized anlysis), which $p$ is the number of processes and $q$ is the size of the queue at the time of invocation.
+\nf{Enqueue()} and \nf{Dequeue()}, each take $O(\log^2 p + \log q)$ steps in amortized analysis.
+Where $p$ is the number of processes and $q$ is the size of the queue at the time of invocation of operation.
 \end{lemma}
@@
 To propagate \nf{x} to the root the algorithm tries two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-Each \nf{Refresh} takes $O(1)$ steps since creating a block is done in constant time and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $O(\log p)$, \nf{Enqueue(x)} takes $O(\log p)$ steps.
-
-A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among operations, and returns the response.
+We can see from the code  that each \nf{Refresh} takes constant number of steps since creating a block is done in constant time and does $O(1)$ \nf{CAS}es.
+Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue(x)} takes $O(\log p)$ steps.
+
+A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response.
 The first two part is similar to an \nf{Enqueue} operation.
-To compute the order there are some constant steps and \nf{IndexDeq} is called.
+To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDeq()} is called.
 \nf{IndexDeq} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree.
@@
 
-%\begin{lemma}[Root search range]\label{rootRange}
-%  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
-%\end{lemma}
+\begin{lemma}[CASes invoked]
+An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
+\end{lemma}
+\begin{proof}
+  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} has 2 \nf{CAS}es, one in Line \ref{cas} and one in Lines \ref{incrementHead1} or \ref{incrementHead2}.
+\end{proof}
 \end{document}

### 2022-08-15 article/psuedocode.tex
review ordering,get

@@
 
+\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
+$\blacktriangleright$ Note that processes are numbered from 1 to $p$, left to right in the leaves of the tree and from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
+
+\begin{itemize}
+\item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
+  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} ordered by their process id.
+  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
+  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} ordered by their process id.
+  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
+\item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
+\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
+\item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
+\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+\end{itemize}
+\end{definition}
+\it{Note that in the non-root nodes we only need ordering of enqueues and dequeues among the operations of their own type.
+Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}
+
+\pagebreak
+
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
@@
 
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-$\blacktriangleright$ Note that processes are numbered from 1 to $p$, left to right in the leaves of the tree and from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
-
-\begin{itemize}
-\item We call operations strictly before $op$ in the sequence of operations $S$, prefix of the $op$.
-  \item $E(n,b)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
-  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
-  \item $D(n,b)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
-  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
-\item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
-\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
-\item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
-\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
-\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
-\end{itemize}
-\end{definition}
-\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}
-
-\pagebreak
-
 \begin{lemma}[Get correctness] \label{get}
@@
 We are going to prove this lemma by induction on the height of node \nf{n}.
-The base case, where \nf{n} os a leaf, is straightforward (Line \ref{getBaseCase}).
+The base case, where \nf{n} is a leaf, is straightforward (Line \ref{getBaseCase}).
 Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf.
-\nf{n.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf \nf{n}.
+\nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n}.
 
 For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$.
-For non-leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}).
-From Definition \ref{ordering} we know enqueue operations in a block are ordered by their process id and since in leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
-Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}.
-So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child.
+In the Line\ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks.
+From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered by their process id and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
+Furthermore the \nf{num\sub{enq-left}} field  in a block stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+So $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
+otherwise we should search for the $i$th enqueue in the left child.
 By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
@@
 If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
-Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$, which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
+Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
 We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
@@
 
-  \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}.
-Enqueue operations from the left subblocks, colored red, are ordered before the enequeue operations from the right child, colored blue.
+  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}.
+Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
   }\label{figGet}
@@
 
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3in]{pics/end-b.png}
-
-  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
-  \label{end-b}
-\end{figure}
+%\begin{figure}[hbt]  
+%  \center\includegraphics[width=3in]{pics/end-b.png}
+%
+%  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
+%  \label{end-b}
+%\end{figure}
 

### 2022-08-17 article/psuedocode.tex
review indexDeq proof

@@
 \item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.
-Since it is incremented after propagating, it may be behind by 1.}
+\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
 \end{itemize}
@@
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
+
+\If{\tt{dir {\keywordfont is} left}}
+\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-left}} \cmt{consider the enqueues in the previous blocks from the left child} \label{considerPreviousLeft}
+\EndIf
+
 \If{\tt{dir {\keywordfont is} right}}
-%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
-\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerRight}
+\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
+\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
 \EndIf
@@
 \item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
-  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} ordered by their process id.
+  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])}. $E(leaf,b)$ is the enqueue operation in \nf{ops(leaf.blocks[b])}.
+\begin{align*} 
+E(n,b) =&  E(n.left,n.blocks[b-1].end\sub{left}+1).E(n.left,n.blocks[b-1].end\sub{left}+2)...E(n.left,n.blocks[b].end\sub{left}). \\ 
+&E(n.right,n.blocks[b-1].end\sub{right}+1).E(n.right,n.blocks[b-1].end\sub{right}+2)...E(n.right,n.blocks[b].end\sub{right})
+\end{align*}
   \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} ordered by their process id.
-  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
 \item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
 \item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
+  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])}. $D(leaf,b)$ is the dequeue operation in \nf{ops(leaf.blocks[b])}.
+\begin{align*} 
+D(n,b) =&  D(n.left,n.blocks[b-1].end\sub{left}+1).D(n.left,n.blocks[b-1].end\sub{left}+2)...D(n.left,n.blocks[b].end\sub{left}). \\ 
+&D(n.right,n.blocks[b-1].end\sub{right}+1).D(n.right,n.blocks[b-1].end\sub{right}+2)...D(n.right,n.blocks[b].end\sub{right})
+\end{align*}
+    \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
 \item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
@@
   \center\includegraphics[width=6in]{pics/doubling.png}
-
   \caption{Distance relations between $b,c,end$}
@@
 
-\pagebreak
-
-\begin{lemma}[Index correctness]
- \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case \nf{root.IndexDEQ(b,i)} is trivial (Line \ref{indexBaseCase}).
-  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
-After that the order in $D(n.parent, superblock)$ is computed and \nf{index()} is called on \nf{n.parent} recursively.
-Then if the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
-\end{proof}
-\textit{Do I need to talk about the computation of the order in the parent which is based on the definition of ordering of dequeues in a block?
-Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}
+Hello Im hoseen master student in cs at York uni.
+I will defend my thesis by Nov this year and I was looking for ful time jobs starting from novomber.
+I have experince in both acdemia and indusrty.
+When I was a bachelors srudent I workd from 2 years as a Software dveloper.
+I have expereince with backend devlopment.
+I have implemented, deploeyd an maintained backend systems.
+
+\pagebreak
 
@@
 
-\begin{lemma}[Computing SuperBlock]\label{superBlock}
-After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{n.parent.blocks[superblock]} contains $D(n,b,i)$.
-\end{lemma}
-\begin{proof}
-Lemmas 28,29,30,31.
-\end{proof}
-\begin{lemma}
-Value read for \nf{super[b.group]} in line 418 is not null.\end{lemma}
-\begin{proof}
-  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
-So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
-\end{proof}
+\begin{lemma} Let $n.propagates$ be the number of groups of blocks that have been propagated from node \nf{n} to its parent (successful \nf{n.parent.Refresh()}).
+We have \nf{num\sub{propagated}}$\leq n.propagates\leq$\nf{num\sub{propagated}+p}. \nf{p} is the number of processes.
+\end{lemma}
+\begin{proof}
+    \tt{num\sub{propagated}} is incremented after propagating (Line \ref{incNP}). Since maybe some process falls sleep before incrementing \tt{num\sub{propagated}} it may be behind by \nf{p}.
+\end{proof}
+
 \begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
@@
 \end{lemma}
-\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.time]}.
+\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.group]}.
 \end{lemma}
@@
 \end{proof}
+
+\begin{lemma}[Computing SuperBlock]\label{superBlock}
+For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
+\end{lemma}
+\begin{proof}
+First we show the value read for \nf{super[b.group]} in line 418 is not null.
+  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
+So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
+Then by Lemma \ref{superRange}if we search in the range $p$, we can find the superblock.
+
+\end{proof}
+
+\begin{lemma}[Index correctness]
+ If \nf{n.blocks[b].num\sub{enq}}$\geq$\nf{i} then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
+\end{lemma}
+\begin{proof}
+We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case where \nf{n} is root is trivial (Line \ref{indexBaseCase}).
+  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
+After that the order in $D(n.parent, superblock)$ is computed.
+Note that by Lemma \ref{blockSize} in each block there is at most one operation from each process and operations of one type are ordered based on the order in the subblocks (See Figure \ref{fig::orderFromSubblocks}).
+Finally \nf{index()} is called on \nf{n.parent} recursively and it returns the correct response from induction hypothesis.
+If the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
+\end{proof}
+\textit{Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=4in]{pics/orderFromSubblocks.png}
+  \caption{Relation of ordering of operations of a block from its subblocks}
+  \label{fig::orderFromSubblocks}
+\end{figure}
 

### 2022-08-18 article/psuedocode.tex
apply aug 18 notes

@@
 
+\begin{corollary}[No Duplicates]\label{append}
+If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+\end{corollary}
+\begin{proof}
+  Operation \nf{op} is invoked only one time in an execution because every operations invoked is distinct.
+Since there is node \nf{n} which \nf{op} is in two different blocks of \nf{n}, there is node \nf{n}$^\prime$ that is the lowest height node in the tree that contains \nf{op} in two of its blocks \nf{b1,b2}.
+By Definition \ref{def::subblock}, \nf{b1} and \nf{b2} have  distinct subblocks(not only direct subblocks) and since \nf{op} is in only one leaf block, then it cannot be in both \nf{b1} and \nf{b2}.
+\end{proof}
+
 %(The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()})
@@
 To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}.
-\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}).
+\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine}, \ref{prevLine}).
 Since \nf{end\sub{left},end\sub{right}} are non-decreasing (\nf{n.blocks[i].end\sub{left|right}>n.blocks[i-1].end\sub{left|right}}), so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
@@
 
-\begin{corollary}[No Duplicates]\label{append}
-If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
-\end{corollary}
 %\begin{proof}
@@
 
+$\blacktriangleright$ Processes are numbered from 1 to $p$ and leaves of the tree are assigned from left to right.
+We will show in Lemma \ref{blockSize} that there is at most one operation from each process in a given block.
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-$\blacktriangleright$ Note that processes are numbered from 1 to $p$, left to right in the leaves of the tree and from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
 
 \begin{itemize}
-\item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
-  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])}. $E(leaf,b)$ is the enqueue operation in \nf{ops(leaf.blocks[b])}.
+\item The prefix of an operation $op$ in the sequence of operations $S$ is the sequence of operations strictly before $op$.
+  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} defined recursively as follows.
+$E(leaf,b)$ is the single enqueue operation in \nf{ops(leaf.blocks[b])} or an empty sequence if \nf{leaf.blocks[b].num\sub{enq}=0}.
+If \nf{n} is an internal node, then
 \begin{align*} 
-E(n,b) =&  E(n.left,n.blocks[b-1].end\sub{left}+1).E(n.left,n.blocks[b-1].end\sub{left}+2)...E(n.left,n.blocks[b].end\sub{left}). \\ 
-&E(n.right,n.blocks[b-1].end\sub{right}+1).E(n.right,n.blocks[b-1].end\sub{right}+2)...E(n.right,n.blocks[b].end\sub{right})
+E(n,b) =&  E(n.left,n.blocks[b-1].end\sub{left}+1)\cdot E(n.left,n.blocks[b-1].end\sub{left}+2)\cdots E(n.left,n.blocks[b].end\sub{left})\cdot \\ 
+&E(n.right,n.blocks[b-1].end\sub{right}+1)\cdot E(n.right,n.blocks[b-1].end\sub{right}+2)\cdots E(n.right,n.blocks[b].end\sub{right})
 \end{align*}
   \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
-\item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
+\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
 \item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])}. $D(leaf,b)$ is the dequeue operation in \nf{ops(leaf.blocks[b])}.
+  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} defined recursively as follows.
+$D(leaf,b)$ is the single dequeue operation in \nf{ops(leaf.blocks[b])} or an empty sequence if \nf{leaf.blocks[b].num\sub{deq}=0}.
+If \nf{n} is an internal node, then
 \begin{align*} 
-D(n,b) =&  D(n.left,n.blocks[b-1].end\sub{left}+1).D(n.left,n.blocks[b-1].end\sub{left}+2)...D(n.left,n.blocks[b].end\sub{left}). \\ 
-&D(n.right,n.blocks[b-1].end\sub{right}+1).D(n.right,n.blocks[b-1].end\sub{right}+2)...D(n.right,n.blocks[b].end\sub{right})
+D(n,b) =&  D(n.left,n.blocks[b-1].end\sub{left}+1)\cdot D(n.left,n.blocks[b-1].end\sub{left}+2)\cdots D(n.left,n.blocks[b].end\sub{left})\cdot \\ 
+&D(n.right,n.blocks[b-1].end\sub{right}+1)\cdot D(n.right,n.blocks[b-1].end\sub{right}+2)\cdots D(n.right,n.blocks[b].end\sub{right})
 \end{align*}
     \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
-\item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
+\item The order of the dequeue operations in the node $n$: $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
 \item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
-\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
 \end{itemize}
 \end{definition}
-\it{Note that in the non-root nodes we only need ordering of enqueues and dequeues among the operations of their own type.
-Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}
-
+
+\begin{definition}[Linearization] \label{def::lin}
+ $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+\end{definition}
+$\blacktriangleright$ In the non-root nodes, we only need ordering of enqueues and dequeues among the operations of their own type.
+Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works with dequeues.
 \pagebreak
@@
 \begin{proof}
-To derive a contradiction, assume there are 2 operations $op_1$ and $op_2$ from process $p$ in block $b$ in node $n$.
-WLOG $op_1$ is invoked earlier than $op_2$.
-A process cannot invoke more than one operations concurrently, so one operation has to be finished before the other starts.
-So $op_1$ has terminated before $op_2$ started.
-By Corollary \ref{lem::appendExactlyOnce} before appending $op_2$ to the tree $op_1$ exists in every node from the path of $p$'s leaf to the root.
-Because $op_1$'s \nf{Append} is finished before $op_2$'s \nf{Append} starts.
-So there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
-$op_1$ existing in $b$ an $b^\prime$ contradicts Lemma \ref{append}.
+To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$.
+Without loss of generality $op_1$ is invoked earlier than $op_2$.
+A process cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$.
+By Corollary \ref{lem::appendExactlyOnce}, before appending $op_2$ to the tree $op_1$ exists in every node on the path from $p$'s leaf to the root, because $op_1$'s \nf{Append} is finished before $op_2$'s \nf{Append} starts.
+So, there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
+Existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{append}.
 \end{proof}
@@
 \begin{proof}
-  It follows directly from Lemma \ref{blockSize} and the observation that each block appended to the tree contains at least one operation, induced from Line~\ref{addOP}. 
+The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to the tree contains at least one operation, due to the test on Line~\ref{addOP}.
+We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.
 \end{proof}
@@
 \begin{lemma}[Get correctness] \label{get}
-If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
+If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_{n,b,i}$.
 \end{lemma}
@@
 We are going to prove this lemma by induction on the height of node \nf{n}.
-The base case, where \nf{n} is a leaf, is straightforward (Line \ref{getBaseCase}).
-Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf.
-\nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n}.
+For the base case \nf{n} is a leaf.
+Leaf blocks each contain exactly one operation, so by the preconditions of \nf{GetENQ()}, only \nf{n.GetENQ(b,1)} can be called and \nf{n.blocks[b]} contains an enqueue.
+At Line \ref{getBaseCase} \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
 
 For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$.
-In the Line\ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks.
+In  Line \ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks.
 From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered by their process id and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
@@
 \begin{lemma}[Index correctness]
- If \nf{n.blocks[b].num\sub{enq}}$\geq$\nf{i} then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
+ If \nf{n.blocks[b].num\sub{deq}}$\geq$\nf{i} then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
 \end{lemma}
@@
 \begin{observation} \label{responseToADeq}
-$k$th non-null dequeue in an execution returns the \nf{element} of $k$th enqueue.
+In a sequential execution on a queue, $k$th non-null dequeue returns the \nf{element} of $k$th enqueue.
 \end{observation}

### 2022-08-18 article/psuedocode.tex
new notation for E(n,b,i)

@@
 \State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
-\Statex\cmt{$E_{root,e}$ is the response.}
+\Statex\cmt{$E_e(root)$ is the response.}
 \State \Return \tt{root.GetENQ(root.\Call{DSearch}{e, b})}\label{findAnswer}
@@
 \Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
-\cmt{Returns \tt{<b,i>} if $E_{root,e}=E_{root,b,i}$.}
+\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
 \State \tt{start= end-1}
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_{this,b,i}$.}
+\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
 \If{\tt{this} \bf{is} \tt{leaf}}
 \State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_{this,b,i}$ is in the left child of this node.}
+\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
 \State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
@@
 \end{align*}
-  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
+  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
 \item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
+\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
   \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} defined recursively as follows.
@@
 \end{align*}
-    \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
+    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
 \item The order of the dequeue operations in the node $n$: $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
+\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
 \end{itemize}
@@
 \begin{lemma}[Get correctness] \label{get}
-If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_{n,b,i}$.
+If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
 \end{lemma}
@@
 
-For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$.
+For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_i(n,b)$, if \nf{n.child.GetENQ(b,i)} returns $E_i(n.child,b)$.
 In  Line \ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks.
@@
 If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
-Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$.
+Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
 We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in \ref{leftChildGet} are for searching $E_{n.left, i+eb}$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_{n,b,i}$.
+The parameters in \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
 
@@
 Other parameters are assigned similar for the left child.
-So in both cases the direct subblock containing $E_{n,b,i}$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
-
- Finally, \nf{n.child.GetENQ()} is invoked on the subblock containing $E_{n,b,i}$ which returns $E_{n,b,i}$ by the hypothesis of the induction.
+So in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
+
+ Finally, \nf{n.child.GetENQ()} is invoked on the subblock containing $E_i(n,b)$ which returns $E_i(n,b)$ by the hypothesis of the induction.
 \end{proof}
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-Assume \nf{root.blocks[end].sum\sub{enq}}$\geq$\nf{e} and $E_{root, e}$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
-\nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_{root,b,i}= E_{root,e}$.
-\end{lemma}
-\begin{proof}
-It is trivial to see that the doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} will find $E_{root,e}$ eventually.
+Assume \nf{root.blocks[end].sum\sub{enq}}$\geq$\nf{e} and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
+\nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
+\end{lemma}
+\begin{proof}
+It is trivial to see that the doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} will find $E_e(root)$ eventually.
 Because \nf{root.blocks[].sum\sub{enq}} is an increasing value from 0 to some value greater than \nf{e}.
@@
 
-Hello Im hoseen master student in cs at York uni.
-I will defend my thesis by Nov this year and I was looking for ful time jobs starting from novomber.
-I have experince in both acdemia and indusrty.
-When I was a bachelors srudent I workd from 2 years as a Software dveloper.
-I have expereince with backend devlopment.
-I have implemented, deploeyd an maintained backend systems.
-
 \pagebreak
@@
 \begin{proof}
-First note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_{root, b, i}$ we talk about what if the whole operation in the linearization are applied on a queue.
+First note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_i(root, b)$ we talk about what if the whole operation in the linearization are applied on a queue.
 
@@
 Note that the dequeues in \nf{root.blocks[b]} before the $i$th dequeue are non-null dequeues.
-So the response is $E_{root,i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}}$ (Line \ref{computeE}).
+So the response is $E_{i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}}(root)$ (Line \ref{computeE}).
 See figure \ref{computeResponseDetail}.
 
-After computing \nf{e} we can find \nf{b,i} such that $E_{root,b,i}=E_{root,e}$ using \nf{DSearch} and then find its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
+After computing \nf{e} we can find \nf{b,i} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then find its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
 \end{proof}
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $E_{root,b,i}$.}
+  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $E_i(root,b)$.}
 \label{computeResponseDetail}

### 2022-08-23 thesis/thesis.tex
add intro+related+uc+blocktree

--- 2022-08-18 article/psuedocode.tex
@@
 
+\begin{abstract}
+In this work, we are going to introduce a novel lock-free queue implementation.
+Linearizability and lock-freedom are standard requirements for designing shared data structures.
+All existing linearizable, lock-free queues in the literature have a common problem in their worst case called CAS Retry Problem.
+Our contribution is solving this problem while outperforming the previous algorithms.
+\end{abstract}
+
+\section{Introduction}
+Shared data structures have become an essential field in distributed algorithms research.
+We are reaching the physical limits of how many transistors we can place on a CPU core.
+The industry solution to provide more computational power is to increase the number of cores of the CPU.
+This is why distributed algorithms have become important.
+It is not hard to see why multiple processes cannot update sequential data structures designed for one process.
+For example, consider two processes trying to insert some values into a sequential linked list simultaneously.
+Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same.
+In this run, $p$'s update is overwritten.
+One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released.
+Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
+Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
+
+The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
+A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
+
+The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
+Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty.
+The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
+Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
+The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
+
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.5]{pics/good}
+  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
+\end{figure}
+
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.5]{pics/bad}
+  \caption{\label{fig::badexample}An example of an execution that is not linearizable.
+Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
+\end{figure}
+
+There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
+An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
+We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
+A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm there might be an operation that takes infinitely many steps but never terminates.
+
+In section 2 we talk about previous queues and their common problems. We also talk about polylogarithmic construction of shared objects.
+
+Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
+He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
+Here, we first introduce a universal construction using $\textsc{O}(\log p)$ CAS operations~\cite{DBLP:conf/fsttcs/JayantiP05}.
+In section 3 we introduce a polylogarithmic step wait-free universal construction.
+Our main ideas in of the universal construction also appear in our Queue Algorithm (\ref{algQ}).
+The main short come of our universal construction is using big CAS objects.
+We use the universal construction as a stepping stone towards our queue algorithm, so we will not explain it in too much detail.
+
+In section 4 we introduce a concurrent wait-free datastructure, to agree on the order of the operations invoked on some processes.
+
+In section 5 we introduce our main work, the queue; prove its linearizability and wait-freeness.
+
+\section{Related Work}
+\subsection{List-based Queues}
+In the following paragraphs, we look at previous lock-free queues.
+Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
+A version of it is included in the standard Java Concurrency Package.
+Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
+Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue.
+To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
+If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
+This makes the amortized number of steps to be $\Omega(p)$ per enqueue.
+Similarly, dequeue can take $\Omega(p)$ steps.
+
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.4]{pics/msqueue}
+  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
+In the first diagram the first element has been dequeued.
+Red arrows show new pointers and gray dashed arrows show the old pointers.}
+\end{figure}
+
+Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination technique.
+The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
+An Elimination Queue consists of an MS-queue augmented with an elimination array.
+Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
+Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
+The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+
+Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
+Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
+Each basket has a time interval in which all its nodes' enqueue operations overlap.
+Since the operations in a basket are concurrent, we can order them in any way.
+Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
+However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
+
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.3]{pics/baskets}
+  \caption{Baskets queue idea.
+There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}.
+To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again.
+The successful process will be the next one in the basket and so on.}
+\end{figure}
+
+Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues.
+They use a doubly-linked list and do fewer \texttt{CAS} operations than MS-queue.
+But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one.
+The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
+
+Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
+Their queue is linearizable but not lock-free.
+Their main idea is that with knowledge of all the history of operations, it might be possible to answer queries faster than doing them one by one.
+In our work we also maintain the whole history.
+They present experiments that show their algorithm performs well in some situations.
+
+Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
+Global head and tail pointers point to arrays containing the first and last elements in the queue.
+Global pointers are up to date, but head and tail pointers may be behind in time.
+An enqueue or a dequeue searches in the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
+Their data structure is lock-free.
+Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.
+
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.5]{pics/sundell}
+    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
+\end{figure}
+
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique to achieve wait-freedom.
+Their step complexity is $\Omega(p)$ because of the helping mechanism.
+
+%todo: comparison
+%Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
+Their idea of batching allows a sequence of operations to be submitted as a batch for later execution.
+It supports a new notion introduced by the authors called Extended Medium Futures Linearizability.
+
+%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
+In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously.
+Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
+It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} .
+We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.
+
+\subsection{Universal Constructions}
+Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
+A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
+We can implement a concurrent queue using a universal construction.
+Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
+He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
+His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.
+
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.
+
+%FIFO queues have a wide range of use in OS and applications. The current state-of-the-art queues are implemented using linked lists; that's why they have factor p in their time complexity.
+
+%\href{https://docs.google.com/spreadsheets/d/1cL1tgXXdljkh462sMwkTVMHH_k0MBTmslIVM3xA5VS4/edit#gid=0}{Table of previous works}
+
+%There is a connection between queues and universal constructions.
+We can implement a universal construction using a queue.
+We can store operations in the shared queue and compute operations' responses using the queue's content.
+Some impractical universal constructions are using big words as "...".
+
+\pagebreak
+\section{ Universal Construction using Tournament Tree with Big CAS Objects }
+
+\paragraph{}
+A universal construction gets any sequential object and creates a concurrent version of the given datastructure.
+Our universal construction in Algorithm~\ref{alg1} relies on a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes.
+Nodes in the tree are CAS objects that store an ordering and support \tt{append()} and \tt{diff()} operations.
+Leaf $l_i$ stores the sequence of the operations invoked by $P_i$.
+Each internal node stores the sequence of operations propagated up to it.
+When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root.
+The history of operations stored at the root is the linearization ordering.
+The operation $op$ is linearized when it is appended to the root.
+
+\begin{center}
+\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
+          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
+\end{center}
+
+The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$.
+In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
+After a process adding a new operations to its leaf, it has to propagate new operations up to the root.
+\textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$).
+In each \tt{Propagate()} step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again.
+If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
+Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) terminates.
+
+\begin{figure}
+\begin{center}
+\begin{subfigure}[b]{.49\textwidth}
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}
+}
+  \caption{Operations after $||$ are new.}
+
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{.49\textwidth}
+
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}
+}
+  \caption{New operations are added to the parent node.}
+
+\end{subfigure}
+
+\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
+\end{center}
+\end{figure}
+
+\begin{algorithm}
+\caption{Universal Construction Idea}\label{alg1}
+\begin{algorithmic}[1]
+\begin{multicols}{2}
+
+\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
+\State \texttt{l\textsubscript{i}.\Call{append}{op}}
+\State \Call{Propagate}{parent of l\textsubscript{i}}
+\State \texttt{Run} \textsf{the sequence stored in root}
+\State \texttt{\Return op}\textsf{'s response from line 4}
+\EndFunction{Do}
+
+\Statex
+
+\Function{void}{Propagate}{\textsl{node} n}
+\If{\texttt{n==root}} \Return
+\ElsIf{!\Call{Refresh}{n}}
+\State \Call{Refresh}{$n$} \EndIf
+\State \Call{Propagate}{parent of n}
+\EndFunction{Propagate}
+
+\columnbreak
+
+\Function{boolean}{Refresh}{\textsl{node} n}
+\State \texttt{old=} \Call{Read}{n}
+\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
+\State \texttt{new= old$\cdot$new}
+\State \Return n.\Call{CAS}{old, new}
+\EndFunction{Refresh}
+
+\end{multicols}
+\end{algorithmic}
+\end{algorithm}
+
+$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
+The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}).
+However, to compute the result of an operation, we only use the total ordering that is stored at the root.
+Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable sized CAS objects and still achieve polygarithmic step complexity.
+
+\begin{figure}[h]
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]
+
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
+      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]
+
+\end{tikzpicture}
+\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
+\end{center}
+\end{figure}
+
+\pagebreak
+\section{Block Trees}
+Here we introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
+Then we use the block tree as a stepping stone towards our queue algorithm.
+A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}).
+Each process has a leaf, and it appends its operations to its leaf.
+After that, the process tries to propagate its new operation up to the tree's root.
+An ordering of operations propagated up to a node is stored in that node.
+All processes agree on the sequence stored in the root and this is used as the linearization ordering.
+Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.
+
+\begin{figure}[hbt]
+\begin{center}
+\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
+          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
+\end{center}
+\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
+Each node tries to propagate its operations up to the root, which stores the final ordering of all operations.}
+\end{figure}
+
+The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
+Then, a dequeue operation uses the linearization ordering to compute its answer.
+
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation.
+The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
+%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
+Instead of storing operations explicitly in the nodes, we only keep track of the number of them.
+This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
+To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.
+\pagebreak
+
+\begin{figure}[hbt]
+\begin{center}
+\begin{subfigure}[b]{.49\textwidth}
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}}
+  \caption{Operations after $||$ are new.}
+\end{subfigure}
+\hfill
+\begin{subfigure}[b]{.49\textwidth}
+  \centering
+  \resizebox{\columnwidth}{!}{
+  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\end{tikzpicture}}
+  \caption{New operations are added to the parent node.}
+\end{subfigure}
+\caption{\label{fig::propagstep} Successful \texttt{Refresh}, operations in children after $||$ are new.}
+\end{center}
+\end{figure}
+
+We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
+\texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf.
+\texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root.
+The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
+After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
+
+\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation.
+Each block stores a constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
+
+Block tree can be used to implement queue, but \texttt{Get(i)}  may take a long time since it has to find the block containing the $i$th operation at the root level.
+
+%\subsection{Idea in a nutshell}
+We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes.
+First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node.
+Second, each operation is linearized when its element is appended to the root.
+
+\paragraph{}
+ In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root.
+However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
+
+ \begin{figure}[h]
+\begin{center}
+\begin{tikzpicture}
+
+\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
+      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
+\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
+\\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
+\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
+\\ \hline\end{tabular}} ] ]
+
+\end{tikzpicture}
+\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves.
+Since we linearize operations when they are added to the root, ordering the blocks in the root is important.}
+\end{center}
+\end{figure}
+
+The definition of linearizability allows concurrent operations to be reordered arbitrarily.
+Thus, a group of concurrent operations can be appended to our root sequence as one block without specifying the order among the operations.
+
+%\begin{lemma}
+%Operations in the same block can be linearized in any order.
+%\end{lemma}
+
+\paragraph{}
+ We used unbounded CAS objects storing sequences as big words in the universal construction.
+One can represent sequences as arrays to overcome this implementation problem.
+Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.
+
+\paragraph{}
+Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied.
+This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations.
+Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set.
+Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)
+
+\pagebreak
+\begin{figure}
+
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
+	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
+	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]
+
+\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
+ [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
+ [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
+  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
+          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:block} Showing concurrent operation sets with blocks.
+Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
+Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
+Blocks between two lines $||$ are propagated together to the parent.
+For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent.
+The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.}
+\end{figure}
+
+Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
+For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
+Block (5,3) also points to elements from blocks (0,5) and (1,2).
+%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.
+
+\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
+\end{definition}
+
+\begin{definition}\{Subblock\}
+  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$.
+Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
+\end{definition}
+%{\TODO: describe how blocks work with a good example how exactly it is done}
+
+\paragraph{}
+We choose to linearize operations in a block from the left child before those from the right child as a convention.
+Operations within a block of the root can be ordered in any way that is convenient.
+In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
+So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
+All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
+%\begin{lemma}
+%\label{lem:block_size}
+%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
+%\end{lemma}
+%\begin{proof}
+%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
+%\end{proof}
+
+\paragraph{}
+In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes.
+So in a complete propagation, at most $2p$ blocks are merged into one block.
+\texttt{(maybe useful for analysis)}
+%\begin{lemma}
+%A block in the root cannot contain more than one operation from a process.
+%\end{lemma}
+%\begin{proof}
+%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
+%\end{proof}
+
+\pagebreak
+\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
+%\begin{definition}
+%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
+%\end{definition}
+\paragraph{}
+\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence.
+We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
+To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.
+
+Furthermore, in each block, we store the prefix sum of left and right elements.
+Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]
+
+\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:prefix} Using Prefix sums in blocks.
+When we want to find block b elements in its children, we can use binary search.
+The number below each block shows the count of elements in the previous blocks.}
+\end{figure}
+
+\begin{figure}[hbt]
+\centering
+  \includegraphics[width=5in]{pics/pointers}
+  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
+\end{figure}
+
+\paragraph{}
+Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
+From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
+So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
+
+\pagebreak
+\subsection{Block Tree Algorithm}
+\paragraph{}
+Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements.
+It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.
+
+\paragraph{Design of a block tree}
+
+Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
+Each internal node of the tree contains an array of blocks of elements.
+Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks.
+Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order.
+Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively.
+Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
+Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
+The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
+After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).
+
+In order to compute elements of a block faster we store prefix-sum blocks(block i has tuple(right-sum=$\#$right ops in previous block, left-sum=$\#$left ops in previous blocks)[See Figure \ref{fig:prefix}].
+Here is the algorithm to get elements of a block.
+
+\paragraph{Specification}
+A block tree is a shared data structure that stores a sequence of elements.
+It has two methods \texttt{Append(e)} and \texttt{Get(i)}.
+\texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence.
+\texttt{Get(i)} returns \texttt{i}th element stored in the sequence.
+
+\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.
+
+\paragraph{Membership}
+Element \texttt{e} is a member of block \texttt{b} in:
+\begin{itemize}
+  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
+  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
+\end{itemize}
+
+\pagebreak
+\paragraph{Order of elements inside node}
+Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
+\begin{itemize}
+  \item The block containing d is before the block containing e.
+  \item e and d are in the same block and d is in the left child and e is in the right child.
+  \item d is before e in the same child's order.
+\end{itemize}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=5.5in]{pics/tree}
+  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
+\end{figure}
+
+\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
+\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
+Other attributes are assigned values followed by definition.
+\begin{figure}[hbt]
+  \center\includegraphics[width=5.5in]{pics/createblock}
+  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
+\end{figure}
+
+%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}
+
+\pagebreak
+\paragraph{Double Refresh}
+Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
+\begin{proof}
+\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block.
+If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
+So suppose two \texttt{Refresh}es fail.
+Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}.
+So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}.
+After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
+\end{proof}
+\begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/doublerefresh.png}
+  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
+\end{figure}
+
+\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
+\begin{proof}
+Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks.
+So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end.
+Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
+\end{proof}
+
+\paragraph{Total Order}
+Sequence represented by the Block Tree is the sequence of the blocks stored in the root.
+
+\paragraph{Linearization Points}
+\texttt{Get(i)} is linearized when it terminates.
+\texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.
+
+\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
+\begin{proof}
+If there are more than $p$ subblocks, then there is more than one block from process pl.
+\texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9).
+So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
+\end{proof}
+
+\paragraph{Computing \texttt{Get(n, b, i)}}
+To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$.
+Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.
+
+\paragraph{How \texttt{Refresh(n)} works.}
+\begin{enumerate}
+  \item Read n's counter and head
+  \item Create block b
+  \item CAS b into n
+  \item If previous succeed:
+  \begin{enumerate}
+  \item Update sup of b's ending subblocks
+  \item Increment children's counters
+  \end{enumerate}
+\end{enumerate}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=6in]{pics/super}
+  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
+\end{figure}
+
+\paragraph{Computing superblock}
+\begin{enumerate}
+  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
+    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10.
+For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
+  \end{proof} 
+  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
+   Follows from the order of lines 37, 26, 27.
+ \end{proof}
+%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
+%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
+  \item \texttt{super[i+1]-super[i]}$\leq p$
+\begin{proof}
+ In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27.
+Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
+If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
+There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27.
+So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
+ \end{proof}
+  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
+\begin{proof}
+\texttt{super[i]} is the index of the superblock of a block containing block b.
+It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
+\texttt{super(b)} is the real superblock of b.
+\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
+If \texttt{b.time} is \texttt{t} we have:
+$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+  \end{proof} 
+\end{enumerate}
+
+\section{Implementing Queue using Block Tree}
+In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked.
+We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem).
+A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
+\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
+If the queue is empty it returns \texttt{null}.
+Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.
+So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
+%
+%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
+\paragraph{}
+Next, we describe how to use block tree to implement queues.
+The block tree, maintains the history of all operations, not only the current state of the queue.
+Now consider the following history of operations.
+What should each \texttt{Dequeue()} return?
+We can implement Enqueue and Dequeue using our block tree.
+An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
+To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
+Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
+Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
+
+\begin{table}[hbt]
+\centering
+  \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
+    \hline \texttt{ENQ(5)}& \texttt{ENQ(2)}& \texttt{DEQ()}& \texttt{ENQ(3)}&\texttt{DEQ()}& \texttt{DEQ()}& \texttt{DEQ()}& \texttt{ENQ(4)}& \texttt{ENQ(6)}& \texttt{DEQ()}\\ \hline
+  \end{tabular}
+  \caption{An example histoy of operations on the queue}
+\end{table}
+
+\begin{definition}
+A non-null dequeue is one that returns a non-null value.
+\end{definition}
+
+\paragraph{}
+In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order.
+Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null.
+If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}.
+So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.
+
+%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
+
+%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}
+
+\paragraph{}
+In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism.
+So now the problem is to find the result of each Dequeue.
+From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
+In the next example, operations in a cell are concurrent.
+\texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
+We will next describe how these values can be computed efficiently.
+
+\begin{table}[hbt]
+\centering
+  \begin{tabular}{c|c|c|c}
+    \hline \texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
+  \end{tabular}
+  \caption{An example history of operation blocks on the queue}
+\end{table}
+
+\pagebreak
+\paragraph{}
+Now, we claimed that by knowing the current size of the queue and the number of non-null dequeue operations before the current dequeue, we could compute the index of the resulting \texttt{Enqueue()}.
+We apply this approach to blocks; if we store the size of the queue after each block of operations happens and the number of non-null dequeues dequeues till a block, we can compute each dequeue's index of result in \textsc{O}$(1)$ steps.
+
+\begin{table}[hbt]
+\centering
+  \begin{tabular}{c|c|c|c|c}
+    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
+    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
+        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
+            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
+                size & 0 & 2 & 2 & 0 \\ \hline
+  \end{tabular}
+  \caption{Augmented history of operation blocks on the queue}
+\end{table}
+
+%\begin{definition}
+%  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
+%  
+%  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
+%  
+%\end{definition}
+
+Size and the number of non-null dequeues for $b$th block could be computed this way:\\
+\texttt{size[b]= max(size[b-1] +enqueues[b] -dequeues[b], 0)}\\
+\texttt{non-null dequeues[b]= non-null dequeues[b-1] +dequeues[b] -size[b-1] -enqueues[b]}
+
+Given \texttt{DEQ} is in block \texttt{b}, \texttt{response(DEQ)} would be:\\
+\texttt{(size[b-1]- index of DEQ in the block's dequeus >=0) ? ENQ[non-null dequeus[b-1]+ index of DEQ in the block's dequeus] : null;}
+
+\pagebreak
+\section{Main Algorithm}
+
+\paragraph{Specification}
+A Queue is a shared data structure that stores a sequence of elements.
+It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
+\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
+\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
+
+\begin{figure}[hbt]
+\centering
+  \includegraphics[width=5in]{pics/queue}
+  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
+\end{figure}
+
+\subsection{Pseudocode description}
+
+\paragraph{Tournament Tree}
+In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree.
+Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}.
+Each process adds $op$ to its leaf.
+In each internal node an ordering of operations in its subtree is stored.
+All processes agree on the total ordering of all operations stored in the root.
+This ordering will be the linearization of the operations.
+\paragraph{Implicit Storing Blocks}
+For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks.
+A Block is a constant size object that implicitly represents a sequence of operations.
+In each node there is an array of Blocks.
+
+\begin{definition}[Block]
+A block is an object that stores some statistics described in Algorithm Queue.   
+\end{definition}
+
+\begin{definition}[\label{def::subblock}Subblock]
+Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+\end{definition}
+Block $b$ contains subblocks in the left and right children.
+WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$.
+See Figure \ref{fig::createBlock} .
+
+\begin{definition}[Membership of an operation in a block]
+Operation \texttt{e} is a member of block \texttt{b} in:
+\begin{itemize}
+ \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
+ \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
+\end{itemize}
+
+\end{definition}
+
+We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s.
+In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}.
+Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
+There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step.
+It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases.
+Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}.
+This helps us to know which blocks have aggregated together to a block, not precisely though.
+We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.
+
+In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}.
+In order to make \texttt{BSearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
+
+To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}.
+\texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.
+
+\texttt{Enqueue(e)} just \texttt{append}s an \texttt{operation} with \texttt{element e} to the \texttt{root}.
+\texttt{Dequeue()} appends an \texttt{operation} to the root and computes its ordering and the \texttt{enqueue operation} containing the head before it calling \texttt{ComputeHead()} and then \texttt{gets} and returns the \texttt{operation}'s element.
+
+\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index.
+As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times.
+\texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.
+
+We only want to know the \texttt{element} of \texttt{enqueue} operations and compute ordering for \texttt{dequeue} operations.
+That's the reason here \texttt{Get()} searches between enqueues only and \texttt{Index()} returns ordering of a dequeue among dequeues.
+\texttt{Get(n, b ,i)} decides the requested element is in which child of n and continues to search recursively.
+\texttt{index(n, i, b)} calculates the ordering of the given operation in \texttt{n}'s parent each step and finally returns the result among total ordering.
+
+\pagebreak
 \section{Pseudocode}

### 2022-08-24 thesis/thesis.tex
apply aug 23 notes

@@
 \usepackage{subcaption}
-\usepackage{showkeys}
+%\usepackage{showkeys}
 \usepackage{amsmath}
@@
 \nf{n.head} is non-decreasing over time.
-If \nf{n.blocks[i]$\neq$null} and \nf{i$.$0} then \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}} and \nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
+If \nf{n.blocks[i]$\neq$null} and \nf{i$.$0} then $\nf{n.blocks[i].end\sub{left}} \geq \nf{n.blocks[i-1].end\sub{left}}$ and $\nf{n.blocks[i].end\sub{right}} \geq \nf{n.blocks[i-1].end\sub{right}}$.
 \end{lemma}
@@
 \end{lemma}
+\begin{definition}
+  An instance of \nf{Refresh()} is successful iff its \nf{TryAppend(new, s)} terminates with  and returns \nf{true}.
+\end{definition}
 %\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
@@
 
+\begin{definition}
+Let $_{R_1}t$ be the time $R_1$ is invoked and $t_{R_2}$ be the time $R_2$ terminates.
+  $_{line}t$ is the immediate time before running Line $line$.
+$t_{line}$ is the immediate time after running Line $line$.
+$_{line}t^{op}$ is the immediate time before running Line $line$ of operation $op$.
+$t_{line}^{op}$ is the immediate time after running Line $line$ of operation $op$.
+\end{definition}
+
 \begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on \nf{n} by process $p$.
-Let $t_1$ be the time $R_1$ is invoked and $t_2$ be the time $R_2$ terminated.
-If $R_1$ and $R_2$ both fail and return false, then we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
-\end{lemma}
-\begin{proof}
-
-If Line \ref{cas} of $R_1$ or $R_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}.
+  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on internal node \nf{n} by a process $p$.
+If $R_1$ and $R_2$ both fail and return false, then we have \nf{ops(EST\sub{n.left, \sub{R1}t})} $\cup$ \nf{ops(EST\sub{n.right, \sub{R1}t})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{R2}})}.
+\end{lemma}
+\begin{proof}
+
+If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means a successful instance of \nf{Refresh()} performed its Line \ref{readHead} after $t_{\ref{readHead}}^{R_1}$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before $t_{\ref{readHead}}^{R_2}$, from Lemma~\ref{lem::prectrueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
+
  Let $R_1$ read $i$ and $R_2$ read $i+1$ from Line \ref{readHead}.
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means a successful instance of \nf{Refresh()} started after Line \ref{readHead} of $R_1$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before \ref{readHead} of $R_2$, from Lemma~\ref{lem::trueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
-
-Since $R_2$'s \nf{TryAppend()} returns \nf{false} then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to append.
-Since $R_2^\prime$ creates the block after reading the value $i+1$ from \nf{n.head}(Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head} and the \nf{head}'s value is increaing by Lemma \ref{lem::headProgress} then $t_{R2^\prime \; \ref{readHead}} > t_{R1 \; \ref{readHead}} > t_1$ (See Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::prectrueRefresh} after $R_2^\prime$'s \nf{CAS} we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(n.blocks)}.
-Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2$\' terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by then (\nf{n.head>=i+2>i+1}).
-To summarize $t_1$ is before $R_2^\prime$'s \nf{read n.head} and $R_2^\prime$'s successful \nf{CAS} is before $R_2$'s termination.
-So by Lemma \ref{lem::prectrueRefresh} \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
+As $R_2$'s \nf{TryAppend()} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to append.
+Since $R_2^\prime$ creates the block after reading the value $i+1$ from \nf{n.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head} and the \nf{head}'s value is increasing by Lemma \ref{lem::headProgress} then $t_{R2^\prime \; \ref{readHead}} > t_{R1 \; \ref{readHead}} > _{R_1}t$ (see Figure \ref{fig::doubleRefresh}).
+By Lemma \ref{lem::prectrueRefresh} after $R_2^\prime$'s \nf{CAS} ($t_{\ref{cas}}^{R_2^\prime}$) we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(n.blocks)}.
+Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2^\prime$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by then ($\nf{n.head}\geq i+2 > i+1$).
+To summarize, $_{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} is before $R_2$'s termination.
+So, by Lemma \ref{lem::prectrueRefresh}, \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
 %It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
@@
   \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption{\label{fig::doubleRefresh}$t1<$ $r_1$ reading head $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $R_2^\prime$ reading head $<$ \nf{TryAppend(i+1)} $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t2$}
-\end{figure}
-\it{this chain with more depth should be in the proof}
+  \caption{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t_{R_2}$}
+\end{figure}
 
@@
 
-\begin{definition}
-  $t_{before \; line}$ is the immediate time before running Line $line$.   $t_{before \; line}$ is the immediate time after running Line $line$.
-\end{definition} 
 \begin{corollary}
-  \nf{ops(EST\sub{n.left, t\sub{before \ref{firstRefresh}}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{before \ref{firstRefresh}}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{after \ref{secondRefresh}})}}
+  \nf{ops(EST\sub{n.left, \sub{\ref{firstRefresh}}t})} $\cup$ \nf{ops(EST\sub{n.right, \sub{\ref{firstRefresh}}t})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{\ref{secondRefresh}})}}
 \end{corollary}
@@
 \end{proof}
-
-\pagebreak
 
@@
 \begin{lemma}[Get correctness] \label{get}
-If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
+If $\nf{n.blocks[b].num\sub{enq}} \geq \nf{i}$ then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
 \end{lemma}
@@
 We are going to prove this lemma by induction on the height of node \nf{n}.
-For the base case \nf{n} is a leaf.
-Leaf blocks each contain exactly one operation, so by the preconditions of \nf{GetENQ()}, only \nf{n.GetENQ(b,1)} can be called and \nf{n.blocks[b]} contains an enqueue.
-At Line \ref{getBaseCase} \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
-
-For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_i(n,b)$, if \nf{n.child.GetENQ(b,i)} returns $E_i(n.child,b)$.
-In  Line \ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks.
-From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered by their process id and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
-Furthermore the \nf{num\sub{enq-left}} field  in a block stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
-So $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
-otherwise we should search for the $i$th enqueue in the left child.
-By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+For the base case, \nf{n} is a leaf.
+Leaf blocks each contain exactly one operation, so by the hypothesis, only \nf{n.GetENQ(b,1)} can be called and only when \nf{n.blocks[b]} contains an enqueue.
+At Line \ref{getBaseCase}, \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
+
+For the induction step we prove \nf{n.GetENQ(b, i)} returns $E_i(n,b)$, assuming \nf{n.child.GetENQ(subblock, i)} returns $E_i(n.child,b)$.
+We argue that  Line \ref{leftOrRight} correctly decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
+From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered from left to right and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
+Furthermore the \nf{num\sub{enq-left}} field  in \nf{n.blocks[b]} stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+So the $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
+Otherwise we should search for the $i$th enqueue in the left child.
+By definition \ref{def::subblock} and \ref{def::ops} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
 
 If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
-Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Else if the enqueue we are looking for is in the right child then there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+Also, we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
+We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
+
+Otherwise the enqueue we are looking for is in the right child.
+Then, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
 So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
-Other parameters are assigned similar for the left child.
-So in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
-
- Finally, \nf{n.child.GetENQ()} is invoked on the subblock containing $E_i(n,b)$ which returns $E_i(n,b)$ by the hypothesis of the induction.
+Other parameters for the left child are chosen similarly to the way they were chosen for the right child.
+
+So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
+ Finally, \nf{n.child.GetENQ(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
 \end{proof}
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-Assume \nf{root.blocks[end].sum\sub{enq}}$\geq$\nf{e} and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
+Assume $\nf{root.blocks[end].sum\sub{enq}}\geq\nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
 \nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
@@
 Because \nf{root.blocks[].sum\sub{enq}} is an increasing value from 0 to some value greater than \nf{e}.
-So there is a \nf{b} that \nf{root.blocks[b].sum\sub{enq}}$>e$ but \nf{root.blocks[b-1].sum\sub{enq}}$<e$.
-
-First we show \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
+So there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}}>e$ but $\nf{root.blocks[b-1].sum\sub{enq}}<e$.
+
+First we show $\nf{end}-\nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size+1\big)$.
 From line \ref{addOP}, we know that size of the every block in the tree is greater than 0.
@@
 So \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
-See Figure \ref{end-b}.
+See Figure \ref{fig::doubling}.
 
@@
 \begin{lemma}[Index correctness]
- If \nf{n.blocks[b].num\sub{deq}}$\geq$\nf{i} then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
+ If $\nf{n.blocks[b].num\sub{deq}}\geq\nf{i}$ then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
 \end{lemma}

### 2022-08-26 thesis/thesis.tex
apply new content structure

@@
 \begin{document}
+\title{Wait-free Queues with Polylogarithmic Step Complexity}
+\author{Hossein Naderibeni \\supervised by Eric Ruppert}
+\maketitle
 
@@
 
+\tableofcontents
+\pagebreak
 \section{Introduction}
@@
 
+\pagebreak
 \section{Related Work}
@@
 
-\pagebreak
-\section{ Universal Construction using Tournament Tree with Big CAS Objects }
-
-\paragraph{}
-A universal construction gets any sequential object and creates a concurrent version of the given datastructure.
-Our universal construction in Algorithm~\ref{alg1} relies on a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes.
-Nodes in the tree are CAS objects that store an ordering and support \tt{append()} and \tt{diff()} operations.
-Leaf $l_i$ stores the sequence of the operations invoked by $P_i$.
-Each internal node stores the sequence of operations propagated up to it.
-When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root.
-The history of operations stored at the root is the linearization ordering.
-The operation $op$ is linearized when it is appended to the root.
-
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-
-The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$.
-In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
-After a process adding a new operations to its leaf, it has to propagate new operations up to the root.
-\textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$).
-In each \tt{Propagate()} step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again.
-If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
-Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) terminates.
-
-\begin{figure}
-\begin{center}
-\begin{subfigure}[b]{.49\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}
-}
-  \caption{Operations after $||$ are new.}
-
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.49\textwidth}
-
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
-\end{tikzpicture}
-}
-  \caption{New operations are added to the parent node.}
-
-\end{subfigure}
-
-\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
-\end{center}
-\end{figure}
-
-\begin{algorithm}
-\caption{Universal Construction Idea}\label{alg1}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
-\State \texttt{l\textsubscript{i}.\Call{append}{op}}
-\State \Call{Propagate}{parent of l\textsubscript{i}}
-\State \texttt{Run} \textsf{the sequence stored in root}
-\State \texttt{\Return op}\textsf{'s response from line 4}
-\EndFunction{Do}
-
-\Statex
-
-\Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{n==root}} \Return
-\ElsIf{!\Call{Refresh}{n}}
-\State \Call{Refresh}{$n$} \EndIf
-\State \Call{Propagate}{parent of n}
-\EndFunction{Propagate}
-
-\columnbreak
-
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{old=} \Call{Read}{n}
-\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
-\State \texttt{new= old$\cdot$new}
-\State \Return n.\Call{CAS}{old, new}
-\EndFunction{Refresh}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
-The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}).
-However, to compute the result of an operation, we only use the total ordering that is stored at the root.
-Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable sized CAS objects and still achieve polygarithmic step complexity.
-
-\begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]
-
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
-      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]
-
-\end{tikzpicture}
-\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
-\end{center}
-\end{figure}
-
-\pagebreak
-\section{Block Trees}
-Here we introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}.
-Then we use the block tree as a stepping stone towards our queue algorithm.
-A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}).
+\subsection{Attiya Fourier Lower Bound}
+\pagebreak
+
+\pagebreak
+\section{Our work}
+Jayanti introduced a wait-free polylogarithmic MESD queue, we are going to design a polylogarithmic MEMD queue based on some of their ideas.
+We apply two ideas from universal their work to create a new linearizable data structure agreeing on a sequence of elements among processes.
+First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node.
+Second, each operation is linearized when its element is appended to the root.
+We use the shared tournament tree among $p$ processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by procesess.
 Each process has a leaf, and it appends its operations to its leaf.
@@
 
-%\subsection{Idea in a nutshell}
-We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes.
-First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node.
-Second, each operation is linearized when its element is appended to the root.
-
 \paragraph{}
- In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root.
-However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
+We can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
 
@@
 
-%\begin{lemma}
-%Operations in the same block can be linearized in any order.
-%\end{lemma}
-
-\paragraph{}
- We used unbounded CAS objects storing sequences as big words in the universal construction.
-One can represent sequences as arrays to overcome this implementation problem.
-Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.
-
-\paragraph{}
-Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied.
-This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations.
-Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set.
-Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)
-
-\pagebreak
 \begin{figure}
@@
 
+In the original algorithm  we differ between the enqueues and the dequeues in a block. Later we discuss about the reason.
+
 Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
@@
 Block (5,3) also points to elements from blocks (0,5) and (1,2).
-%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.
-
-\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
-\end{definition}
-
-\begin{definition}\{Subblock\}
-  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$.
-Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
-\end{definition}
-%{\TODO: describe how blocks work with a good example how exactly it is done}
 
@@
 All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
-%\begin{lemma}
-%\label{lem:block_size}
-%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
-%\end{lemma}
-%\begin{proof}
-%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
-%\end{proof}
 
@@
 \texttt{(maybe useful for analysis)}
-%\begin{lemma}
-%A block in the root cannot contain more than one operation from a process.
-%\end{lemma}
-%\begin{proof}
-%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
-%\end{proof}
-
-\pagebreak
-\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
-%\begin{definition}
-%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
-%\end{definition}
+
 \paragraph{}
@@
 \pagebreak
-\subsection{Block Tree Algorithm}
-\paragraph{}
-Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements.
-It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.
-
-\paragraph{Design of a block tree}
+\subsection{Queue Decription}
+\paragraph{Design of a Queue}
 
@@
 \paragraph{Specification}
-A block tree is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Append(e)} and \texttt{Get(i)}.
-\texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence.
-\texttt{Get(i)} returns \texttt{i}th element stored in the sequence.
-
-\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.
-
-\paragraph{Membership}
-Element \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
-  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
-\end{itemize}
-
-\pagebreak
-\paragraph{Order of elements inside node}
-Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
-\begin{itemize}
-  \item The block containing d is before the block containing e.
-  \item e and d are in the same block and d is in the left child and e is in the right child.
-  \item d is before e in the same child's order.
-\end{itemize}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
-\end{figure}
+A Queue is a shared data structure that stores a sequence of elements.
+It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
+\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
+\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
 
@@
 
-\pagebreak
-\paragraph{Double Refresh}
-Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
-\begin{proof}
-\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block.
-If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
-So suppose two \texttt{Refresh}es fail.
-Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}.
-So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}.
-After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-\end{proof}
 \begin{figure}[hbt]
@@
 \end{figure}
-
-\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
-\begin{proof}
-Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks.
-So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end.
-Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
-\end{proof}
-
-\paragraph{Total Order}
-Sequence represented by the Block Tree is the sequence of the blocks stored in the root.
-
-\paragraph{Linearization Points}
-\texttt{Get(i)} is linearized when it terminates.
-\texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.
-
-\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
-\begin{proof}
-If there are more than $p$ subblocks, then there is more than one block from process pl.
-\texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9).
-So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
-\end{proof}
 
@@
 \paragraph{Computing superblock}
-\begin{enumerate}
-  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
-    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10.
-For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
-  \end{proof} 
-  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
-   Follows from the order of lines 37, 26, 27.
- \end{proof}
-%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-  \item \texttt{super[i+1]-super[i]}$\leq p$
-\begin{proof}
- In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27.
-Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
-If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27.
-So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
- \end{proof}
-  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
-\begin{proof}
-\texttt{super[i]} is the index of the superblock of a block containing block b.
-It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
-\texttt{super(b)} is the real superblock of b.
-\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
-If \texttt{b.time} is \texttt{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-  \end{proof} 
-\end{enumerate}
-
-\section{Implementing Queue using Block Tree}
+
+\paragraph{Implementing Queue using Block Tree}
 In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked.
@@
 
-\begin{definition}
+\paragraph{}
 A non-null dequeue is one that returns a non-null value.
-\end{definition}
-
-\paragraph{}
 In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order.
@@
 
-\pagebreak
-\section{Main Algorithm}
-
-\paragraph{Specification}
-A Queue is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
-\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
-\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
-
 \begin{figure}[hbt]
@@
 
-\paragraph{Tournament Tree}
+\paragraph{Tree}
 In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree.
@@
 
-\begin{definition}[Block]
-A block is an object that stores some statistics described in Algorithm Queue.   
-\end{definition}
-
-\begin{definition}[\label{def::subblock}Subblock]
-Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-\end{definition}
 Block $b$ contains subblocks in the left and right children.
@@
 See Figure \ref{fig::createBlock} .
-
-\begin{definition}[Membership of an operation in a block]
-Operation \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
- \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
- \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-\end{itemize}
-
-\end{definition}
 
@@
 \pagebreak
-\section{Pseudocode}
+\subsection{Pseudocode}
 
@@
 
-\section{Proof of Linearizability}
+\subsection{Proof of Correctness}
 
@@
 \pagebreak
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=5.5in]{pics/tree}
+  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
+\end{figure}
 
@@
 
-First we show $\nf{end}-\nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size+1\big)$.
+First we show $\nf{end}-\nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1\big)$.
 From line \ref{addOP}, we know that size of the every block in the tree is greater than 0.
@@
 Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.
-So \nf{end-b}$\leq 2 \times$\nf{root.blocks[b].\size+root.blocks[end].\size+1}.
+So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$.
 See Figure \ref{fig::doubling}.
@@
 Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
-Beause otherwise, then \nf{(end-c)/2} satisfied the \nf{root.blocks[(end-c)/2].sum\sub{enq}<e}.
+Beause otherwise, then \nf{(end-c)/2} satisfied the $\nf{root.blocks[(end-c)/2].sum\sub{enq}}<\nf{e}$.
 In line \ref{doubling} the differnece between \nf{end} and \nf{start} is doubled.
@@
 \end{proof}
+
+\pagebreak
+\subsection{Garbage Collection or Getting rid of the infinite Arrays}
+
+\pagebreak
+\section{Using Queues to Implement Vectors}
+Supporting Append, Read, Write in PolyLog time by modifying Get(Enq) Method. Create a Universal Construction Using our vector
+
+\pagebreak
+\section{Conclusion}
+possible directions for work
+
+Maybe Stacks
+
+Characterize what datastructure can be used for this approach, we already know: queue, fetch \& Inc, Vectors
+
 \end{document}

### 2022-08-31 thesis/thesis.tex
review some pages of desciption

@@
 \section{Our work}
-Jayanti introduced a wait-free polylogarithmic MESD queue, we are going to design a polylogarithmic MEMD queue based on some of their ideas.
-We apply two ideas from universal their work to create a new linearizable data structure agreeing on a sequence of elements among processes.
-First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node.
-Second, each operation is linearized when its element is appended to the root.
-We use the shared tournament tree among $p$ processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by procesess.
-Each process has a leaf, and it appends its operations to its leaf.
-After that, the process tries to propagate its new operation up to the tree's root.
+Jayanti and Petrovic introduced a wait-free polylogarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
+We are going to design a polylogarithmic multi-enqueuer multi-dequeuer queue using some of their ideas.
+But we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.
+We apply two ideas from their work to create a new shared data structure which enables processes to agree on the linearization ordering of the processes.
+We use the shared tournament tree among $p$ processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes.
+Each process has a leaf which the order of operations invoked by the process is stored in it.
+When the process wishes to do an operation it appends the operation to its leaf's sequence and after that, the process tries to propagate its new operation up to the tree's root.
 An ordering of operations propagated up to a node is stored in that node.
-All processes agree on the sequence stored in the root and this is used as the linearization ordering.
-Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.
-
+All processes agree on the sequence stored in the root that is used as the linearization ordering.
 \begin{figure}[hbt]
@@
 \end{center}
-\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root, which stores the final ordering of all operations.}
-\end{figure}
-
-The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}).
-Then, a dequeue operation uses the linearization ordering to compute its answer.
-
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation.
+\caption{ّ\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored.
+Each node tries to propagate its operations up to the root, which stores the totall ordering of all operations.}
+\end{figure}
+
+As we said in  each node the sequence of operations is stored.
+We implement the sequence using an array and appending to the sequence by doing \tt{CAS} operations on the first null element in the array.
+In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s both children (that have not already been propagated to $n$) and tries to append them into $n$.
 The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
-%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
-Instead of storing operations explicitly in the nodes, we only keep track of the number of them.
-This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.
-\pagebreak
+Because if both of the \nf{Refresh()}es fail there is another instance of \nf{Refresh()} in between which has succeeded to do \nf{CAS} and has already appended the operations the first \nf{Refresh} was trying to append.
+This mechanism makes us to do \nf{TryAppend}s twice instead of \nf{Append}ing to the array.
+From the next paragraph we explain our ideas which are different from Jayanti.
 
@@
 
-We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
-\texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf.
-\texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root.
-The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
-After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
-
-\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the sub-block or super-block containing the given operation.
-Each block stores a constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
-
-Block tree can be used to implement queue, but \texttt{Get(i)}  may take a long time since it has to find the block containing the $i$th operation at the root level.
-
-\paragraph{}
+Instead of storing operations explicitly in the nodes, we are going to keep track of some statistics of them.
+This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations.
+In each \nf{Refresh()} there is at most one operation from  each process trying to be propagated.
+Since one operation cannot invoke two operations concurrently.
+Also as the operations in a \textsc{Refresh()} step are concurrent we can linearize them  among themselves in any order we wish.
+Note that if two operations are in one \nf{Porpagate()} step in a node they are going to be propagated up to the root together.
+From now on instead of propagating sequence of operations in  \nf{Refresh} steps we propagate \nf{block}s of operations.
+The idea is that we can describe a blocks contents only using some numbers.
+Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
+In this way if we know the number of total operations in a block and the number of operations from the left child we can order the operations in a unique way which is a complete ordering.
 We can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
+In the next paragraphs we explain the reason behind our choice of ordering.
 
@@
 \end{figure}
+
+Previously we talked about storing the sequence of operations in the nodes of the tree.
+A process may wish to know information about the root ordering.
+Two functionalities are to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
+Since our algorithm is aimed for a queue, we make some assumptions here that we one wish only to know the order of a dequeue and one only tries to get an enqueue.
+We will explain in detail in the next but for now let's say \nf{enqueue}s and \nf{dequeue}s are appended to the tree and when one wants to find the response to a dequeu, it computes the order of the dequeue in linearization and using this information it computes which eneueue is the answer to the dequeue or if the answer is null.
+If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetENQ(n,b,i)}.
+\texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetENQ(n,b,i)} finds its sub-block recursively to reach a leaf.
+\texttt{Index()} is similar but more complicated, finding super-blocks from a leaf to the root.
+The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
+After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
+Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
+
+ \pagebreak
 
@@
 
+\pagebreak
+
 \paragraph{}
@@
 \texttt{(maybe useful for analysis)}
+
+\pagebreak
 
@@
 
-\pagebreak
-\subsection{Queue Decription}
 \paragraph{Design of a Queue}
@@
 
+\pagebreak
+
 \paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
@@
 \end{figure}
+
+\pagebreak
 
@@
 
+\pagebreak
+
 \paragraph{Computing superblock}

### 2022-08-31 thesis/thesis.tex
review DSearch

@@
 \cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
-\State \tt{start= end-1}
+\State \tt{start= end-1} \label{dsearchStart}
 \While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
@@
 \EndWhile
-\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)}
+\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchEnd}
 \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
@@
 \begin{lemma}[DSearch correctness] \label{dsearch}
-Assume $\nf{root.blocks[end].sum\sub{enq}}\geq\nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}.
-\nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
-It is trivial to see that the doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} will find $E_e(root)$ eventually.
-Because \nf{root.blocks[].sum\sub{enq}} is an increasing value from 0 to some value greater than \nf{e}.
-So there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}}>e$ but $\nf{root.blocks[b-1].sum\sub{enq}}<e$.
-
-First we show $\nf{end}-\nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1\big)$.
-From line \ref{addOP}, we know that size of the every block in the tree is greater than 0.
-So each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}.
+If $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$, \nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
+\end{lemma}
+\begin{proof}
+\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$.
+From Lemma we know \nf{sum\sub{enq}} fields of nf{root.blocks[]} are sorted in a non-decreasing order.
+Since \nf{root.blocks[0].sum\sub{enq}=0} and there is a \nf{block} in the root with \nf{sum\sub{enq}} value greater than \nf{e}, so there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}} \geq e$ but $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
+This block contains $E_i(root,b)$ and  the search on Line \ref{dsearchStart}-\ref{dsearchEnd} will eventually reach the \nf{b}.
+
+\end{proof}
+\begin{lemma}[DSearch Analysis]\label{dsearchTime}
+  Assume $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then  \nf{DSearch(e, end)} takes $\Theta(\log$\nf{root.blocks[b].\size+ root.blocks[end].\size}$)$ steps.
+\end{lemma}
+\begin{proof}
+First we show $\nf{end} - \nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1 \big)$.
+From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0.
+So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}.
 Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}.
-Then the  queue would become empty at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation.
+Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation.
 Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$.
-Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end].\size} \nf{Enqueue}s.
-Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size is \nf{root.blocks[end].\size}.
+Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}.
+Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}.
 Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.
@@
 
-Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2\times$\nf{root.blocks[b].\size+root.blocks[end].\size}.
+Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2 \times 2 \times \big(\nf{root.blocks[b].\size+root.blocks[end].\size}\big)$.
 Beause otherwise, then \nf{(end-c)/2} satisfied the $\nf{root.blocks[(end-c)/2].sum\sub{enq}}<\nf{e}$.
-In line \ref{doubling} the differnece between \nf{end} and \nf{start} is doubled.
+In line \ref{doubling} the difference between \nf{end} and \nf{start} is doubled.
 See Figure \ref{fig::doubling}.
@@
  After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-So the routine non constant part is the binary search which takes $\Theta(\log$\nf{root.blocks[b].\size+root.blocks[end].\size)}$)$ steps from the first paragraph.
+So the whole \nf{DSearch} routine takes $\Theta\big(\log(\nf{root.blocks[b].\size+root.blocks[end].\size})\big)$ steps.
 

### 2022-09-01 thesis/thesis.tex
new code for indexing super

@@
 
-\Statex
-
 \Statex $\diamondsuit$ \tt{\sl{Local}}
@@
 
-\Statex
-\Statex $\diamondsuit$ \tt{\sl{Structures}}
+%\Statex $\diamondsuit$ \tt{\sl{Structures}}
 
@@
 \item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
+%\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
 \end{itemize}
@@
 
-  \item \tt{\sl{int} group}
-  \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
+  \item \tt{\sl{int} super}
+  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
 \end{itemize}
-
-\pagebreak
 
@@
 \State \tt{h= \head} \label{readHead}
+\State \tt{hp= parent.\head} \label{readParentHead1}
+\State \tt{CAS(blocks[h-1].super, null, hp)}
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
+\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
+\State \tt{CAS(dir.blocks[h\sub{dir}], null, h)}
+\EndFor
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
-\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \label{invokeCreateBlock} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
+\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
+\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{Created block contains nothing.}
 \ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
-\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
+\State \tt{hp= parent.\head} \label{readParentHead2}
+\State \tt{\Call{CAS}{new.super, null, hp}} \label{setSuper}
+%\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
 \EndFor
@@
 
-\Statex
-
+\pagebreak
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
@@
 
-\pagebreak
+\Statex
 
 \Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\cmt{Creates a block to be inserted as $i$th \tt{block} in \tt{blocks}.
-Returns the created block as well as values read from each child's \tt{num\sub{propagated}} field.
-These values are used for incrementing the children's \tt{num\sub{propagated}} field if the block was appended to \tt{blocks} successfully.}
+\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
 \State \tt{block newBlock= \Call{new}{\sl{block}}}
-\State \tt{newBlock.group= num\sub{propagated}}\label{setGroup}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{np\sub{dir}= dir.num\sub{propagated}} \label{setNP}
 \State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
@@
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
-\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
+\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super-2, blocks[b].super+2}} \label{computeSuper}
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
@@
 Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$.
-Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}.
-Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}.
+Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}., because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}.
 Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.

### 2022-09-06 thesis/thesis.tex
add new super proof

@@
 \usepackage{subcaption}
-%\usepackage{showkeys}
+\usepackage{showkeys}
 \usepackage{amsmath}
@@
 
-\pagebreak
-
 \paragraph{}
@@
 \end{itemize}
-
-%\Statex $\blacktriangleright$ \tt{Conventions}
-%\begin{itemize}
-%  \item \tt{i} : index of ith operation in the tree
-%  \item \tt{j} : index of jth operation in a node
-%  \item \tt{b\sub{n}} : index of the block containing the operation n based on the scope
-%  \item Also we are not going to refer to blocks directly, only with their indices. Except while constructing a new block.
-%\end{itemize}
 
@@
 \end{algorithm}
+
+\begin{algorithm}
+\caption{BlockList}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{700}
+
+\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
+Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
+If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
+
+%\Statex
+%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
+%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
+%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
+%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
+%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
+%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
+%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
+%\State \tt{l.Help()}
+%\EndFor
+%\EndIf
+%\State \tt{blk.num\sub{finished}= 0}
+%%\State \tt{*oldRoot= \&root.blocks.root}
+%%\State \tt{*newRoot= root.blocks.Append(blk).root}
+%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
+%\State \Return{\tt{CAS(blocks[n], null, blk)}}
+%\EndFunction{TryAppend}
+
+\Statex
+%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
+\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
+\Statex
+\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
+\State \Return{\tt{CAS(blocks[n], null, blk)}}
+\EndFunction{TryAppend}
+
+\end{algorithmic}
+\end{algorithm}
 %##########################################
@@
 \State \tt{hp= parent.\head} \label{readParentHead1}
-\State \tt{CAS(blocks[h-1].super, null, hp)}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
+\If{\nf{blocks[h]!=null}}
+\State \tt{CAS(blocks[h].super, null, hp)} \label{helpSuper1}
+\EndIf
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
 \State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
 \State \tt{CAS(dir.blocks[h\sub{dir}], null, h)}
-\EndFor
+\EndFor \label{endHelpChild1}
 %\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
@@
 \Else
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
+\State \tt{hp= parent.\head} \label{readParentHead2}
+\State \tt{\Call{CAS}{new.super, null, hp}} \label{setSuper}
+%\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
+\EndFor
 \State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
@@
 %\EndFunction{Help}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{BlockList}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
-
-\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
-
-%\Statex
-%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
-%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
-%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
-%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
-%\State \tt{l.Help()}
-%\EndFor
-%\EndIf
-%\State \tt{blk.num\sub{finished}= 0}
-%%\State \tt{*oldRoot= \&root.blocks.root}
-%%\State \tt{*newRoot= root.blocks.Append(blk).root}
-%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
-%\State \Return{\tt{CAS(blocks[n], null, blk)}}
-%\EndFunction{TryAppend}
-
-\Statex
-%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
-\Statex \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
-\Statex
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
 
@@
 %\end{lemma}
-
-\begin{lemma} Let $n.propagates$ be the number of groups of blocks that have been propagated from node \nf{n} to its parent (successful \nf{n.parent.Refresh()}).
-We have \nf{num\sub{propagated}}$\leq n.propagates\leq$\nf{num\sub{propagated}+p}. \nf{p} is the number of processes.
-\end{lemma}
-\begin{proof}
-    \tt{num\sub{propagated}} is incremented after propagating (Line \ref{incNP}). Since maybe some process falls sleep before incrementing \tt{num\sub{propagated}} it may be behind by \nf{p}.
-\end{proof}
-
-\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
-\end{lemma}
-\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}
-
-\begin{lemma} Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\end{lemma}
-\begin{proof} Line \ref{setSuper}.\end{proof}
-
-\begin{lemma}
-The number of the \nf{block}s with \nf{group=i} in a node is $\leq p$.  
-\end{lemma}
-\begin{proof}For the sake of simplicity we assumed all the blocks are propagated from the left child.
-\begin{figure}[hbt]
-  \center\includegraphics[width=5in]{pics/index1}
-\end{figure}  
-\end{proof}
+%
+%\begin{lemma} Let $n.propagates$ be the number of groups of blocks that have been propagated from node \nf{n} to its parent (successful \nf{n.parent.Refresh()}).
+%We have \nf{num\sub{propagated}}$\leq n.propagates\leq$\nf{num\sub{propagated}+p}. \nf{p} is the number of processes.
+%\end{lemma}
+%\begin{proof}
+%    \tt{num\sub{propagated}} is incremented after propagating (Line \ref{incNP}). Since maybe some process falls sleep before incrementing \tt{num\sub{propagated}} it may be behind by \nf{p}.
+%\end{proof}
+%
+%
+%\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
+%\end{lemma}
+%\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}
+%
+%\begin{lemma} Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\end{lemma}
+%\begin{proof} Line \ref{setSuper}.\end{proof}
+
+%\begin{lemma}
+%The number of the \nf{block}s with \nf{super=i} in a node is $\leq p$.  
+%\end{lemma}
+%\begin{proof}For the sake of simplicity we assumed all the blocks are propagated from the left child.
+%\begin{figure}[hbt]
+%  \center\includegraphics[width=5in]{pics/index1}
+%\end{figure}  
+%\end{proof}
 
@@
 % \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
-\begin{proof}
- In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
+%\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
+%\begin{proof}
+% In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
 Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set.
@@
 So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
- \begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/index2}
-\end{figure}  
-\end{proof}
-\begin{lemma}[super property]\label{superCounter}
-If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
-\end{lemma}
-\begin{lemma}\label{superRange} Superblock of \nf{b} is within range $\pm 2p$ of the \nf{super[b.group]}.
-\end{lemma}
-
-\begin{proof} \nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
+% \begin{figure}[hbt]
+%  \center\includegraphics[width=4in]{pics/index2}
+%\end{figure}  
+%\end{proof}
+%\begin{lemma}[super property]\label{superCounter}
+%If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
+%\end{lemma}
+
+%\begin{proof} \nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
 \nf{super(b)} is the real superblock of b.
@@
 If \nf{b.time} is \nf{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-\end{proof}
+%$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
+%\end{proof}
+
+%\begin{lemma}
+%Search in each level of \nf{IndexDeq()} takes $O(\log p)$ steps.  
+%\end{lemma}
+%\begin{proof}
+%Show preconditions are satisfied and the range is $p$.  
+%\end{proof}
 
 \begin{lemma}
-Search in each level of \nf{IndexDeq()} takes $O(\log p)$ steps.  
-\end{lemma}
-\begin{proof}
-Show preconditions are satisfied and the range is $p$.  
-\end{proof}
-
+ Let the value When an instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{CAS} into \nf{n.blocks[h]} where \nf{h} is the value read from \nf{n.head}.
+\end{lemma}
+\begin{proof}
+  By Invariant \ref{lem::headPosition}, \nf{n.blocks[h]} may be \nf{null} or not \nf{null}.
+This also holds for \nf{n.parent.blocks[hp]}.
+Unless all other fields of a block that is set  while creating the block because of a technical problem we set the \nf{super} field of a block after appending it to the blocklist.
+\nf{super} field is supposed to show the approximate index of the superblock of a block, We set the value while appending the block or while a block is appended to the parent.
+These times the value of parent's \nf{head} is really close to the actual value of the superblock of the block.
+
+The \nf{Refresh} routine first helps to fill the \nf{super} field of \nf{n.blocks[h]} in case \nf{n.blocks[h]} is not null (Line \ref{helpSuper1}) and \nf{n.blocks[h].super} field is not set yet, and then the \nf{TryAppend} in Line \ref{cas} fails  and the \nf{Refresh} routine helps to increment \nf{n.head} value in Line \ref{incrementHead2}.
+
+After helping to fill the current \nf{block} at the \nf{head} index, \nf{Refresh} helps the left child and the right child to fill their \nf{super} field.
+A process may go to sleeps after Line \ref{cas} of its \nf{Refresh} and before Line \ref{setSuper}.
+So when a block is going to be appended to parent we help the children so the super field dos not change very much form its real value.
+After this step we are assured for all the blocks in the children that their index is less than the \nf{head}, the \nf{super} field is set.
+
+We claim after Line \ref{endHelpChild1} of \nf{n.Refresh()}, \nf{super} field of all the established blocks in \nf{n,n.left,n.right} is set and $\neq \nf{null}$.
+\end{proof}
+
+\begin{corollary}
+  Since Line \ref{incrementHead1} is executed after the Lines \ref{setSuper} and \ref{endHelpChild1}, then when \nf{n.head} is increased from \nf{h} to \nf{h+1}, we know that \nf{n.blocks[h]} is set.
+It might be two cases that the process which did \nf{TryAppend(new,h)} successfully, has incremented \nf{h} to \nf{h+1} or it might went to sleep and another process  which failed to do \nf{TryAppend(new,h)} comes and helps to set \nf{n.blocks[h].super} and increments \nf{n.head}.
+Or maybe a process that is trying to add a block to \nf{n.parent} after the \nf{TryAppend(new,h)} helps to set the \nf{super} field of \nf{n.blocks[h]}.
+\end{corollary}
+
+\begin{corollary}
+  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set by itself or the process that incremented the head from \nf{h} to \nf{h+1}, or the process that did the Lines 313 \ref{startHelpChild1} to 316 \ref{endHelpChild1}.
+\end{corollary}
+
+\it{How I can make this proof coming precise?}
+\begin{lemma}\label{superRange} If $\nf{n.blocks[b].super} \neq \nf{null}$ then, the superblock of \nf{n.blocks[b]} is in \nf{n.parent.blocks[b.super-1..b.super+1]}.
+\end{lemma}
+\begin{proof}
+  From the previous corollary, \nf{n.blocks[h].super} might be set while \nf{n.blocks[b]} is appended which is going to be one place behind the real super block, or it might set after appending which means one block behind the real value.
+Let us explain the possible cases on block \nf{b} in node \nf{n}.
+
+\nf{super} of the block field is set by the same process that appended the block.
+There are two possibilities that \nf{n.parent[hp]} is \nf{null} or not which \nf{hp} is the value of \nf{n.parent.head}/ In this case that \nf{super} field is the parent's \nf{head=hp} and when a block is going to be appended to the the \nf{hp+2} index of the parent it is going to contain the \nf{b}.
+One may wonder why, becuase it has read the parent's \nf{head} value \nf{hp+2} which is greater than \nf{hp+1} and it has to be incremented after \nf{n.bocks.TryAppend(new,h)}.
+
+The \nf{super} field is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b}.
+In this case also the \nf{super} value is going to be behind the index of the real superblock index in parent if the \nf{n.parent.blocks[hp]}  is already not \nf{null} and so the next of the next block appended to the parent is going to absolutely contain \nf{b} if the next block appending to \nf{n.parent} did not contain \nf{b} (Lemma \ref{trueRefresh}).
+One might wonder why the next block appended to the parent might not contain \nf{b} because it may read the blocks it was going to append before appending \nf{b} to \nf{n.blocks}.
+
+The \nf{super} field is set by the process doing the \nf{Refresh} on the parent.
+In this case if the \nf{Refresh} is successful then position of the superblock of \nf{b} is exactly equal to the \nf{super} value.
+But if it was failure it might be one bit before the real value.
+
+\end{proof}
+
+\pagebreak
 \begin{lemma}[Computing SuperBlock]\label{superBlock}

### 2022-09-08 thesis/thesis.tex
fix super helping

@@
 \State \tt{h= \head} \label{readHead}
-\State \tt{hp= parent.\head} \label{readParentHead1}
-\If{\nf{blocks[h]!=null}}
-\State \tt{CAS(blocks[h].super, null, hp)} \label{helpSuper1}
-\EndIf
+%\State \tt{hp= parent.\head} \label{readParentHead1}
+%\If{\nf{blocks[h]!=null}}
+%\State \tt{CAS(blocks[h].super, null, hp)} \label{helpSuper1}
+%\EndIf
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
 \State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\State \tt{CAS(dir.blocks[h\sub{dir}], null, h)}
+\If{\nf{dir.blocks[h\sub{dir}]!=null}}
+\State \tt{CAS(dir.blocks[h\sub{dir}].super, null, h)}
+\State \tt{CAS(dir.head, h\sub{dir}, h\sub{dir}+1)}
+\EndIf
 \EndFor \label{endHelpChild1}
-%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
 \State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{Created block contains nothing.}
-\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{hp= parent.\head} \label{readParentHead2}
-\State \tt{\Call{CAS}{new.super, null, hp}} \label{setSuper}
-%\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
-\EndFor
-\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
-\State \Return{\tt{true}}
-\Else
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
-\State \tt{hp= parent.\head} \label{readParentHead2}
-\State \tt{\Call{CAS}{new.super, null, hp}} \label{setSuper}
-%\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
-\EndFor
-\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
-\State \Return{ \tt{false}}
+\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
+%\cmt{Created block contains nothing.}
 \EndIf
+\State{\tt{result= blocks.tryAppend(new, h)}} \label{cas}
+\State \tt{hp= parent.\head} \label{readParentHead}
+\State \tt{\Call{CAS}{n.blocks[h].super, null, hp}} \label{setSuper1}
+\State \tt{\Call{CAS}{\head, h, h+1}} 
+%\cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
+\State \Return{ \tt{result}}
+
 \EndFunction{Refresh}
@@
 
-\begin{lemma}
- Let the value When an instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{CAS} into \nf{n.blocks[h]} where \nf{h} is the value read from \nf{n.head}.
-\end{lemma}
-\begin{proof}
-  By Invariant \ref{lem::headPosition}, \nf{n.blocks[h]} may be \nf{null} or not \nf{null}.
-This also holds for \nf{n.parent.blocks[hp]}.
-Unless all other fields of a block that is set  while creating the block because of a technical problem we set the \nf{super} field of a block after appending it to the blocklist.
-\nf{super} field is supposed to show the approximate index of the superblock of a block, We set the value while appending the block or while a block is appended to the parent.
-These times the value of parent's \nf{head} is really close to the actual value of the superblock of the block.
-
-The \nf{Refresh} routine first helps to fill the \nf{super} field of \nf{n.blocks[h]} in case \nf{n.blocks[h]} is not null (Line \ref{helpSuper1}) and \nf{n.blocks[h].super} field is not set yet, and then the \nf{TryAppend} in Line \ref{cas} fails  and the \nf{Refresh} routine helps to increment \nf{n.head} value in Line \ref{incrementHead2}.
-
-After helping to fill the current \nf{block} at the \nf{head} index, \nf{Refresh} helps the left child and the right child to fill their \nf{super} field.
-A process may go to sleeps after Line \ref{cas} of its \nf{Refresh} and before Line \ref{setSuper}.
-So when a block is going to be appended to parent we help the children so the super field dos not change very much form its real value.
-After this step we are assured for all the blocks in the children that their index is less than the \nf{head}, the \nf{super} field is set.
-
-We claim after Line \ref{endHelpChild1} of \nf{n.Refresh()}, \nf{super} field of all the established blocks in \nf{n,n.left,n.right} is set and $\neq \nf{null}$.
-\end{proof}
-
-\begin{corollary}
-  Since Line \ref{incrementHead1} is executed after the Lines \ref{setSuper} and \ref{endHelpChild1}, then when \nf{n.head} is increased from \nf{h} to \nf{h+1}, we know that \nf{n.blocks[h]} is set.
+ Let \nf{h} be the value read  from \nf{n.head} when a successful instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{TryAppend()} into \nf{n.blocks[h]}.
+We call the appended block \nf{b}.
+
+Despite other fields of a block that are set while creating the block, because of a technical problem we set the \nf{super} field of a block after appending it to the blocklist.
+\nf{super} field is supposed to show the approximate index of the superblock of a block, so we want \nf{super} to be as close to the index of the superblock of the block in the parent.
+We set the \nf{super} value to \nf{n.parent.head} after appending the block (Line \ref{}), since \nf{b}'s superblock comes after that in near future.
+Maybe the process goes to sleep at sometime after appending \nf{b} to \nf{n.blocks} and before setting the super.
+In this case we use help from the next \nf{Refresh}es on \nf{n} and \nf{n.parent}.
+
+    By Invariant \ref{lem::headPosition}, \nf{n.parent.blocks[hp]} may be \nf{null} or not \nf{null}, if \nf{hp} is the value read from \nf{n.parent.head} (Line \ref{}). 
+First Lines of the \nf{Refresh} routine(Lines \ref to \ref{}) helps the left child and the right child to fill their \nf{super} field.
+A process may go to sleep, after Line \ref{cas} of its \nf{Refresh} and before Line \ref{setSuper}.
+So when a block is going to be appended to \nf{n.parent.blocks} we help the children so the super field in the last block of the children does not differ very much form its real value.
+After this step we are assured for all the blocks in the children that their index is less than the child's \nf{head}, the \nf{super} field is set.
+Since we first help to fill the \nf{super} and then increment the \nf{head} (Lines \ref{}).
+From Lemma \ref{lem::trueRefresh} we knwo that blocks whose index are less than \nf{head} are propagated to the parent in a \nf{Refresh} instance.
+To summarize Line \ref{endHelpChild1} of \nf{n.Refresh()}, \nf{super} field of all the established blocks in \nf{n.left,n.right} is set and $\neq \nf{null}$.
+
+%
+%\begin{corollary}
+%  Since Line \ref{incrementHead1} is executed after the Lines \ref{setSuper} and \ref{endHelpChild1}, then when \nf{n.head} is increased from \nf{h} to \nf{h+1}, we know that \nf{n.blocks[h]} is set.
 It might be two cases that the process which did \nf{TryAppend(new,h)} successfully, has incremented \nf{h} to \nf{h+1} or it might went to sleep and another process  which failed to do \nf{TryAppend(new,h)} comes and helps to set \nf{n.blocks[h].super} and increments \nf{n.head}.
 Or maybe a process that is trying to add a block to \nf{n.parent} after the \nf{TryAppend(new,h)} helps to set the \nf{super} field of \nf{n.blocks[h]}.
+%\end{corollary}
+
+\begin{corollary}
+  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set before the next succseful \nf{n.Refresh} or the second next succesful \nf{n.parent.Refresh()}.
 \end{corollary}
-
-\begin{corollary}
-  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set by itself or the process that incremented the head from \nf{h} to \nf{h+1}, or the process that did the Lines 313 \ref{startHelpChild1} to 316 \ref{endHelpChild1}.
-\end{corollary}
-
-\it{How I can make this proof coming precise?}
-\begin{lemma}\label{superRange} If $\nf{n.blocks[b].super} \neq \nf{null}$ then, the superblock of \nf{n.blocks[b]} is in \nf{n.parent.blocks[b.super-1..b.super+1]}.
-\end{lemma}
-\begin{proof}
-  From the previous corollary, \nf{n.blocks[h].super} might be set while \nf{n.blocks[b]} is appended which is going to be one place behind the real super block, or it might set after appending which means one block behind the real value.
-Let us explain the possible cases on block \nf{b} in node \nf{n}.
-
-\nf{super} of the block field is set by the same process that appended the block.
-There are two possibilities that \nf{n.parent[hp]} is \nf{null} or not which \nf{hp} is the value of \nf{n.parent.head}/ In this case that \nf{super} field is the parent's \nf{head=hp} and when a block is going to be appended to the the \nf{hp+2} index of the parent it is going to contain the \nf{b}.
+\begin{proof}
+\nf{n.TryAppend(h+1)}cannot happen unless \nf{n.head} is incremented from \nf{h} to \nf{h+1}. The latter case is explained in the next page.  
+\end{proof}
+
+\pagebreak
+\begin{definition} We say \nf{R1} an instance of \nf{Refresh} is before \nf{R2} another instance of \nf{Refresh}.
+If \nf{R1}'s \nf{TryAppend} happens before \nf{R2}'s \nf{TryAppend}.
+We knew that \nf{CAS} objects are linearizable.
+\end{definition}
+
+\begin{lemma}\label{superRange} If $\nf{b.super} \neq \nf{null}$ then, the superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
+\end{lemma}
+\begin{proof}
+Let us explain the possible cases of setting \nf{super} of block \nf{b} in node \nf{n}.
+
+First case is that the \nf{super} field of \nf{b} is set by the same process that appended \nf{b}.
+There are two possibilities that \nf{n.parent[hp]} is already \nf{null} or not, when the value \nf{hp} is read from \nf{n.parent.head}.
+We argue how distant is the superblock of \nf{b} from \nf{n.parent.blocks[hp=b.super]}.
+If the successful \nf{Refresh} that does \nf{n.parent.TryAppend(new,hp+1)}, creates its \nf{new} block after \nf{n.head.CAS(h,h+1)} (Line \ref{}) then \nf{new} contains the not yet propagated established blocks from \nf{n}, including \nf{b} (blocks in \nf{n} with smaller index than \nf{h+1}).
+Otherwise, the next \nf{Refresh} on \nf{n.parent} which has read \nf{head=hp+2} contains \nf{b}.
+Because $\nf{t\sub{n.blocks.TryAppend(b,h)}} < \nf{t\sub{n.head.CAS(h,h+1)}} < \nf{t\sub{hp+2=read n.parent.head}} < \nf{t\sub{n.parent.head.CAS{hp+1,hp+2}}}$.
+See Lemma \ref{lem::trueRefresh}.
+In the case that \nf{n.parent.blocks[hp]!=null}, when a block is going to be appended to the the \nf{hp+1} index of the parent it is going to contain the \nf{b}.
 One may wonder why, becuase it has read the parent's \nf{head} value \nf{hp+2} which is greater than \nf{hp+1} and it has to be incremented after \nf{n.bocks.TryAppend(new,h)}.
-
-The \nf{super} field is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b}.
-In this case also the \nf{super} value is going to be behind the index of the real superblock index in parent if the \nf{n.parent.blocks[hp]}  is already not \nf{null} and so the next of the next block appended to the parent is going to absolutely contain \nf{b} if the next block appending to \nf{n.parent} did not contain \nf{b} (Lemma \ref{trueRefresh}).
-One might wonder why the next block appended to the parent might not contain \nf{b} because it may read the blocks it was going to append before appending \nf{b} to \nf{n.blocks}.
+The other case \nf{n.parent[hp]==null} is the same but one bit closer.
+
+The \nf{super} field of \nf{b} is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b} to \nf{n.blocks}.
+This \nf{Refresh} increments \nf{n.head} from \nf{h} to \nf{h+1}, if it was not already incremented (Lines \ref{}).
+The next successful \nf{Refresh} on \nf{n.parent} which reads \nf{n.head} with some value greater than \nf{h} propagates \nf{b} to \nf{n.parent}.
+If \nf{n.parent.blocks[hp]} was not null then \nf{n.parent.blocks[hp+1]} or \nf{n.parent.blocks[hp+2]} contains \nf{b}, otherwise the \nf{super} field in \nf{b} is the exact index of the superblock of \nf{b} or \nf{n.parent.blocks[hp+1]} contains \nf{b}.
+The second next block appended to the parent is going to absolutely contain \nf{b} if the next block appending to \nf{n.parent} did not contain \nf{b} (Lemma \ref{trueRefresh}).
+One might wonder why the next block appended to the parent might not contain \nf{b} because it might have read the blocks it was propagating before appending \nf{b} to \nf{n.blocks}.
 

### 2022-09-13 thesis/thesis.tex
minor changes

@@
 \If{\nf{dir.blocks[h\sub{dir}]!=null}}
-\State \tt{CAS(dir.blocks[h\sub{dir}].super, null, h)}
-\State \tt{CAS(dir.head, h\sub{dir}, h\sub{dir}+1)}
+\State \tt{dir.blocks[h\sub{dir}].super.CAS(null, h)}
+\State \tt{dir.head.CAS(h\sub{dir}, h\sub{dir}+1)}
 \EndIf
@@
 \EndIf
-\State{\tt{result= blocks.tryAppend(new, h)}} \label{cas}
+\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
 \State \tt{hp= parent.\head} \label{readParentHead}
-\State \tt{\Call{CAS}{n.blocks[h].super, null, hp}} \label{setSuper1}
-\State \tt{\Call{CAS}{\head, h, h+1}} 
+\State \tt{n.blocks[h].super.\Call{CAS}{null, hp}} \label{setSuper1}
+\State \tt{\head.\Call{CAS}{h, h+1}} 
 %\cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
@@
 
-\begin{corollary}
+\begin{lemma}
   After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set before the next succseful \nf{n.Refresh} or the second next succesful \nf{n.parent.Refresh()}.
-\end{corollary}
+Superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
+\end{lemma}
 \begin{proof}
@@
 
-\begin{lemma}\label{superRange} If $\nf{b.super} \neq \nf{null}$ then, the superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
+\begin{lemma}\label{superRange}  
 \end{lemma}
@@
 Let us explain the possible cases of setting \nf{super} of block \nf{b} in node \nf{n}.
+\nf{b.super} is set by an instance of \nf{n.Refresh} or \nf{n.parent.Refresh}.
+Since Lines \ref{} setting the \nf{super} is done before incrementing the \nf{head}, so if we know \nf{head} has advanced the super has to be set before.
 

### 2022-09-15 thesis/thesis.tex
type super new proof

--- 2022-09-08 thesis/thesis.tex
@@
 \If{\nf{dir.blocks[h\sub{dir}]!=null}}
-\State \tt{CAS(dir.blocks[h\sub{dir}].super, null, h)}
-\State \tt{CAS(dir.head, h\sub{dir}, h\sub{dir}+1)}
+\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
 \EndIf
@@
 \EndIf
-\State{\tt{result= blocks.tryAppend(new, h)}} \label{cas}
-\State \tt{hp= parent.\head} \label{readParentHead}
-\State \tt{\Call{CAS}{n.blocks[h].super, null, hp}} \label{setSuper1}
-\State \tt{\Call{CAS}{\head, h, h+1}} 
+\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
+
+\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
+\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
 %\cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
@@
 \EndFunction{Refresh}
+
+\Statex
+
+\Function{void}{Advance}{\sl{int} h, \sl{int} hp}
+
+\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
+\State \tt{head.CAS(h, h+1)} \label{incrementHead}
+\EndFunction{Advance}
 
@@
 
- Let \nf{h} be the value read  from \nf{n.head} when a successful instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{TryAppend()} into \nf{n.blocks[h]}.
+ Let \nf{h} be the value read  from \nf{n.head} when a successful instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{CAS(null, new)} into \nf{n.blocks[h]}.
 We call the appended block \nf{b}.
-
-Despite other fields of a block that are set while creating the block, because of a technical problem we set the \nf{super} field of a block after appending it to the blocklist.
+Despite other fields of a block that are set while creating the block, because we do not know the exact index of the superblock of the block we  can not set the \nf{super} field of a block while appending it to the blocklist.
+One approach is to set the super of a block when it is porpagated but this is not efficent because there might be $p$ blocks propagating together in a \nf{Refresh()} step.
+However we know that once a block is installed its stuperblock is going to be close to \nf{n.parent.head} at the time of installation.
 \nf{super} field is supposed to show the approximate index of the superblock of a block, so we want \nf{super} to be as close to the index of the superblock of the block in the parent.
-We set the \nf{super} value to \nf{n.parent.head} after appending the block (Line \ref{}), since \nf{b}'s superblock comes after that in near future.
+We set the \nf{super} value to \nf{n.parent.head} right after appending the block (Line \ref{}), since \nf{b}'s superblock comes after that in near future.
 Maybe the process goes to sleep at sometime after appending \nf{b} to \nf{n.blocks} and before setting the super.
@@
 
-    By Invariant \ref{lem::headPosition}, \nf{n.parent.blocks[hp]} may be \nf{null} or not \nf{null}, if \nf{hp} is the value read from \nf{n.parent.head} (Line \ref{}). 
-First Lines of the \nf{Refresh} routine(Lines \ref to \ref{}) helps the left child and the right child to fill their \nf{super} field.
+%    By Invariant \ref{lem::headPosition}, \nf{n.parent.blocks[hp]} may be \nf{null} or not \nf{null}, if \nf{hp} is the value read from \nf{n.parent.head} (Line \ref{}). 
+%First Lines of the \nf{Refresh} routine(Lines \ref to \ref{}) helps the left child and the right child to fill their \nf{super} field.
 A process may go to sleep, after Line \ref{cas} of its \nf{Refresh} and before Line \ref{setSuper}.
@@
 
-\begin{corollary}
-  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set before the next succseful \nf{n.Refresh} or the second next succesful \nf{n.parent.Refresh()}.
-\end{corollary}
-\begin{proof}
-\nf{n.TryAppend(h+1)}cannot happen unless \nf{n.head} is incremented from \nf{h} to \nf{h+1}. The latter case is explained in the next page.  
-\end{proof}
-
-\pagebreak
-\begin{definition} We say \nf{R1} an instance of \nf{Refresh} is before \nf{R2} another instance of \nf{Refresh}.
-If \nf{R1}'s \nf{TryAppend} happens before \nf{R2}'s \nf{TryAppend}.
+%\begin{lemma}
+%  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set before the next succseful \nf{n.Refresh} or the second next succesful \nf{n.parent.Refresh()}.
+Superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
+%\end{lemma}
+%\begin{proof}
+%\nf{n.TryAppend(h+1)}cannot happen unless \nf{n.head} is incremented from \nf{h} to \nf{h+1}. The latter case is explained in the next page.  
+%\end{proof}
+
+\begin{definition}  A Refresh is successful if it reaches Line \ref{cas} and returns true.
+If \nf{Refresh ()} instance \nf{R1} does its \nf{CAS} at Line \ref{cas} sooner than \nf{Refresh} instance \nf{R2} we say \nf{R1} has happened before \nf{R2}.
 We knew that \nf{CAS} objects are linearizable.
 \end{definition}
-
-\begin{lemma}\label{superRange} If $\nf{b.super} \neq \nf{null}$ then, the superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
-\end{lemma}
-\begin{proof}
-Let us explain the possible cases of setting \nf{super} of block \nf{b} in node \nf{n}.
-
-First case is that the \nf{super} field of \nf{b} is set by the same process that appended \nf{b}.
+%
+%\begin{lemma}\label{superRange}  
+%\end{lemma}
+%\begin{proof}
+%Let us explain the possible cases of setting \nf{super} of block \nf{b} in node \nf{n}.
+\nf{b.super} is set by an instance of \nf{n.Refresh} or \nf{n.parent.Refresh}.
+Since Lines \ref{} setting the \nf{super} is done before incrementing the \nf{head}, so if we know \nf{head} has advanced the super has to be set before.
+%
+%First case is that the \nf{super} field of \nf{b} is set by the same process that appended \nf{b}.
 There are two possibilities that \nf{n.parent[hp]} is already \nf{null} or not, when the value \nf{hp} is read from \nf{n.parent.head}.
@@
 The other case \nf{n.parent[hp]==null} is the same but one bit closer.
-
-The \nf{super} field of \nf{b} is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b} to \nf{n.blocks}.
+%
+%The \nf{super} field of \nf{b} is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b} to \nf{n.blocks}.
 This \nf{Refresh} increments \nf{n.head} from \nf{h} to \nf{h+1}, if it was not already incremented (Lines \ref{}).
@@
 One might wonder why the next block appended to the parent might not contain \nf{b} because it might have read the blocks it was propagating before appending \nf{b} to \nf{n.blocks}.
-
-The \nf{super} field is set by the process doing the \nf{Refresh} on the parent.
+%
+%The \nf{super} field is set by the process doing the \nf{Refresh} on the parent.
 In this case if the \nf{Refresh} is successful then position of the superblock of \nf{b} is exactly equal to the \nf{super} value.
 But if it was failure it might be one bit before the real value.
-
-\end{proof}
+%
+%\end{proof}
+
+Block \nf{b} is appended to \nf{n.blocks[h]} on Line \ref{cas}.
+After appending \nf{b}, \nf{super} field of \nf{b} is set on Line \ref{setSuper1} by the same process or another \nf{n.Refresh()} or  maybe an \nf{n.parent.Refresh()}.
+We want to argue how far \nf{b.super} is from the index of \nf{b}'s super block, which is created by a successful \nf{n.parent.Refresh()} that propagates \nf{b}.
+
+WLOG for the rest of this section assume \nf{n} is the left child of \nf{n.parent}. The first successful \nf{n.parent.Refresh()} that read some value
+greater than \nf{h} for \nf{left.head} is going to contain \nf{b} in its created block in Line \ref{invokeCreateBlock}.
+We call \nf{n.Refresh()} that added \nf{b} the \nf{R\sub{n}} and \nf{n.parent.Refresh()} that propagated \nf{b} to \nf{p}  the \nf{R\sub{p}} and we use \nf{p} instead of \nf{n.parent}.
+We also sometimes use \nf{n[h]} instead of \nf{n.blocks[h]}(maybe a good idea for the whole document?).
+
+\begin{lemma}
+After n blocks[h].CAS(null,b) succeeds an head cannot increase from h to h+1 unless be super is set.
+\end{lemma}
+\begin{proof}
+If an instance of n.Refresh() increments n.head on Line \ref{324}, then n.blocks[head].super was set at Line \ref{323}.
+\end{proof}
+
+\begin{corollary}
+  If n.blocks[i].super is null, then n.blocks[i+1] is null.
+\end{corollary}
+
+As per previous corollary, b.super has to be set before the next successful Refresh() on n after the Refresh() that put b into n.
+Because if b.super=null then the n head cannot advance and then the next n.Refresh() is going to fail.
+Now Let us talk about how the n.parent.Refresh()es that took place after the putting b into n, will help to set b.super.
+
+\begin{lemma}
+  If b$\in$n.parent.blocks[i] then b.super$\leq$i.
+\end{lemma}
+\begin{proof}
+  I did not quite understand the way Eric told me to prove this.
+\end{proof}
+
+\begin{lemma}
+If Rn=n.Refresh() puts b in n[h] at $t_{321}^{Rn}$, then the block created by the next successful P.Refresh() or the second next successful p.Refresh() contains b and b.super is set before the second next successful p.Refresh() after $t_{321}^{Rn}$.
+\end{lemma}
+
+\begin{proof}
+b.super could be set by helping of some n.Refresh() instance before the second next successful p.Refresh(), but even if no one helped it we prove the second next  successful p.Refresh() will help n to advance in Line \ref{setSuper} called by Line \ref{helpAdvance}.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=7in]{pics/timeLineSuper1.png}
+  \caption{b.super is set by some instance of Refresh() on n or n.parent showed by \nf{R'} and \nf{n.head} is incremented by some \nf{Refresh()} called \nf{R"}.}
+  \label{fig::timeLine1}
+\end{figure}
+
+\nf{Rp} is the first successful \nf{p.Refresh()} instance that has propagated \nf{b} into \nf{p.blocks[x]}.
+We want to know how great $\nf{x}-\nf{b.super}$ can be?
+\nf{p.head} is \nf{hp} when \nf{R'} reads it.
+From Lemma \ref{lem::headProgress} \nf{p.head} could only increase from \nf{hp} to \nf{hp+1} if $\nf{p[hp]}\neq\nf{null}$.
+In other words there should be a successful \nf{p.Refresh()} in order to \nf{p.head} advance.
+We claim there cannot be more than one \nf{p.Refresh()} after \nf{R'} reads \nf{p.head} and before $t_{\ref{lastLine}}^{Rp}$.
+
+Assume the first successful \nf{p.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{b.super}.
+It might happen maybe because the value read for $h_{left}$ in Line \ref{readChildHead} is less than \nf{h} or maybe $\nf{h}=\nf{h\sub{left}}$ and $\nf{left.blocks[h\sub{left}]}=\nf{null}$, which means \nf{n.head} is advanced but \nf{b} is still not installed in \nf{n[h]} which means \nf{R\sub{n}} has not reached to the Line \ref{cas}.
+
+Let the first successful \nf{p.Refresh()} be \nf{Rp1} and the second next successful \nf{p.Refresh()} be \nf{Rp2}.
+If \nf{Rp1} reads \nf{i} in Line \ref{readHead}, then \nf{Rp2} has to read \nf{i+1} in Line \ref{readHead}.
+See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances \nf{Rp1, Rp2} on \nf{p}.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/timeLineSuper2.png}
+  \caption{Rp2 reads p.head after $t_{321}^{Rp1}$, which is after $t_{321}^{Rn}$. Rp2 has to help increment n.head and set b.super.}
+  \label{fig::timeLine2}
+\end{figure}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=5in]{pics/timeLineSuper3.png}
+  \caption{ The second Refresh on p contains b and reads head>h.}
+  \label{fig::timeLine3}
+\end{figure}
+
+So \nf{b.super} has set by some process before the second next successful \nf{p.Refresh()} on Line \ref{setSuper1}.
+Since \nf{h} is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value fo \nf{left.head} greater than \nf{h} and propagates \nf{b} to \nf{p}.
+So if \nf{b} was not propagated already we are sure the second next successful \nf{p.Refresh()} propagates \nf{b}.
+\end{proof}
+
+\begin{corollary}
+If $\nf{b.super} \neq \nf{null}$, then it has at most 1 difference with the index of the superblock of \nf{b} in \nf{p}.
+\end{corollary}
 

### 2022-09-16 thesis/thesis.tex
may this commit be the last commit before the final commit on super lemma

@@
 \end{proof}
-\it{Explain More}
+
+\begin{observation} \label{nonDecreasingHead}
+  \nf{n.head} is non-decreasing over time.
+\end{observation}
+\begin{proof}
+  The claim follows trivially from the code since \nf{n.head} is only incremented in Line \ref{incrementHead} of \nf{Refresh()} .
+\end{proof}
 
 \begin{lemma}[headProgress] \label{lem::headProgress}
-\nf{n.head} is non-decreasing over time.
  If \nf{n.blocks[i]$\neq$null} and \nf{i$.$0} then $\nf{n.blocks[i].end\sub{left}} \geq \nf{n.blocks[i-1].end\sub{left}}$ and $\nf{n.blocks[i].end\sub{right}} \geq \nf{n.blocks[i-1].end\sub{right}}$.
@@
 \begin{proof}
-The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented in the pseudocode in lines \ref{incrementHead1} and \ref{incrementHead2} of \nf{Refresh()} .
 
@@
 
- Let \nf{h} be the value read  from \nf{n.head} when a successful instance of \nf{Refresh()} is called on the node \nf{n}, the instance is supposed to do a \nf{CAS(null, new)} into \nf{n.blocks[h]}.
-We call the appended block \nf{b}.
-Despite other fields of a block that are set while creating the block, because we do not know the exact index of the superblock of the block we  can not set the \nf{super} field of a block while appending it to the blocklist.
-One approach is to set the super of a block when it is porpagated but this is not efficent because there might be $p$ blocks propagating together in a \nf{Refresh()} step.
-However we know that once a block is installed its stuperblock is going to be close to \nf{n.parent.head} at the time of installation.
-\nf{super} field is supposed to show the approximate index of the superblock of a block, so we want \nf{super} to be as close to the index of the superblock of the block in the parent.
-We set the \nf{super} value to \nf{n.parent.head} right after appending the block (Line \ref{}), since \nf{b}'s superblock comes after that in near future.
-Maybe the process goes to sleep at sometime after appending \nf{b} to \nf{n.blocks} and before setting the super.
-In this case we use help from the next \nf{Refresh}es on \nf{n} and \nf{n.parent}.
+\begin{definition}\label{orderRefresh}  A Refresh is \it{successful} if it performs a successful \nf{CAS} on Line \ref{cas}.
+If \nf{Refresh} instance \nf{R\sub{1}} does its \nf{CAS} at Line \ref{cas} sooner than \nf{Refresh} instance \nf{R\sub{2}} we say \nf{R\sub{1}}  \it{happened} before \nf{R\sub{2}}.
+\end{definition}
+
+ Let \nf{i} be the value \nf{R\sub{n}}, a successful instance of \nf{Refresh()} on the node \nf{n}, reads from \nf{n.head}.
+\nf{R\sub{n}} does a successful \nf{CAS(null, new)} into \nf{n.blocks[h]}, where \nf{b} is the \nf{new}.
+Without loss of generality for the rest of this section assume \nf{n} is the left child of \nf{n.parent}.
+From now on we say \nf{p} as an abbreviation for \nf{n.parent}.
+Let \nf{R\sub{p}} be the first successful \nf{p.Refresh()} that reads some value greater than \nf{i} for \nf{left.head} and is contains \nf{b} in its created block \nf{s} in Line \ref{invokeCreateBlock}.
+From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating \nf{b}.
+\nf{R\sub{p}} does a successful \nf{CAS(null, new)} into \nf{p.blocks[j]}, where \nf{s} is the \nf{new}.
+
+Although other fields of \nf{b} are set while creating it, because the index of the superblock of \nf{b} is not known until it is propagated, \nf{R\sub{n}} cannot set the \nf{super} field of a \nf{b} while creating it.
+One approach is to set the \nf{super} field of \nf{b} after it is propagated by \nf{R\sub{b}} but this would not be efficient because there might be $p$ subblocks in \nf{s}.
+However, once \nf{b} is installed, its superblock is going to be close to \nf{n.parent.head} at the time of installation.
+One idea is that if we know the aproximate position of the superblock of \nf{b} then we can search for the real superblock when we wished to know the superblock of \nf{b} i.e.
+\nf{b.super} does not have to be the exact location of  the superblock of \nf{b}, but we want it to be close to \nf{j}.
+We can set \nf{b.super} to \nf{n.parent.head} while creating \nf{b}, but the problem is that there might be many \nf{p.Refresh}es could happen that contain blocks from the right child of \nf{p} and \nf{j} could be arbitrarily (right word?) greater than \nf{b.super}.
+We set \nf{b.super} to \nf{p.head} after appending \nf{b} to \nf{n.blocks} (Line \ref{setSuper1}).
+Maybe \nf{R\sub{n}} goes to sleep at some time after installing \nf{b} and before setting \nf{b.super}.
+In this case the next \nf{Refresh}es on \nf{n} and \nf{n.parent} help fill in the value of \nf{b.super}.
 
@@
 
-\begin{definition}  A Refresh is successful if it reaches Line \ref{cas} and returns true.
-If \nf{Refresh ()} instance \nf{R1} does its \nf{CAS} at Line \ref{cas} sooner than \nf{Refresh} instance \nf{R2} we say \nf{R1} has happened before \nf{R2}.
-We knew that \nf{CAS} objects are linearizable.
-\end{definition}
 %
@@
 Block \nf{b} is appended to \nf{n.blocks[h]} on Line \ref{cas}.
-After appending \nf{b}, \nf{super} field of \nf{b} is set on Line \ref{setSuper1} by the same process or another \nf{n.Refresh()} or  maybe an \nf{n.parent.Refresh()}.
-We want to argue how far \nf{b.super} is from the index of \nf{b}'s super block, which is created by a successful \nf{n.parent.Refresh()} that propagates \nf{b}.
-
-WLOG for the rest of this section assume \nf{n} is the left child of \nf{n.parent}. The first successful \nf{n.parent.Refresh()} that read some value
-greater than \nf{h} for \nf{left.head} is going to contain \nf{b} in its created block in Line \ref{invokeCreateBlock}.
-We call \nf{n.Refresh()} that added \nf{b} the \nf{R\sub{n}} and \nf{n.parent.Refresh()} that propagated \nf{b} to \nf{p}  the \nf{R\sub{p}} and we use \nf{p} instead of \nf{n.parent}.
-We also sometimes use \nf{n[h]} instead of \nf{n.blocks[h]}(maybe a good idea for the whole document?).
+After appending \nf{b}, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} by the same process or another process's \nf{n.Refresh()} or  maybe an \nf{n.parent.Refresh()}.
+We want to bound how far \nf{b.super} is from the index of \nf{b}'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates \nf{b}.
+
+\begin{observation} \label{setSupBeforeIncHead}
+After \nf{n.blocks[i].CAS(null, b)} succeeds, \nf{n.head} cannot increase from \nf{i} to \nf{i+1} unless \nf{b.super} is set.
+\end{observation}
+\begin{proof}
+From the Observation \ref{nonDecreasingHead} we know the only change to \nf{n.head} is on Line \ref{incrementHead} which is incrementing.
+Before an instance of \nf{Advance()} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
+\end{proof}
+
+\begin{corollary}
+  If \nf{n.blocks[i].super} is \nf{null}, then $\nf{n.head}<\nf{i}$ and \nf{n.blocks[i+1]} is \nf{null}.
+\end{corollary}
+\begin{proof}If \nf{b.super} is \nf{null} then \nf{n.head} cannot advance so the next \nf{n.Refresh()} will fail.
+By the previous corollary, \nf{b.super} has to be set before the next successful Refresh() on n after \nf{R\sub{n}}.
+\end{proof}
+Now let us talk about how the \nf{p.Refresh}es that took place after the putting \nf{b} into \nf{n}, will help to set \nf{b.super} and propagate \nf{b}.
 
 \begin{lemma}
-After n blocks[h].CAS(null,b) succeeds an head cannot increase from h to h+1 unless be super is set.
-\end{lemma}
-\begin{proof}
-If an instance of n.Refresh() increments n.head on Line \ref{324}, then n.blocks[head].super was set at Line \ref{323}.
-\end{proof}
-
-\begin{corollary}
-  If n.blocks[i].super is null, then n.blocks[i+1] is null.
-\end{corollary}
-
-As per previous corollary, b.super has to be set before the next successful Refresh() on n after the Refresh() that put b into n.
-Because if b.super=null then the n head cannot advance and then the next n.Refresh() is going to fail.
-Now Let us talk about how the n.parent.Refresh()es that took place after the putting b into n, will help to set b.super.
+  If $\nf{b}\in\nf{n.parent.blocks[i]}$ then $\nf{b.super}\leq\nf{i}$.
+\end{lemma}
+\begin{proof}
+For \nf{R\sub{p}} to contain block \nf{b}, it has to read \nf{n.head} greater than \nf{h} (see Line \ref{lastLine}).
+For \nf{n.head} to be greater than \nf{h} it means \nf{n.head} is incremented in Line \ref{incrementHead} which means \nf{b.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}).
+So if \nf{R\sub{p}} propagates \nf{b} it means \nf{b.super} was already set.
+Let \nf{j} be the value written in \nf{b.super}.
+\nf{j} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{b.super}.
+From Observation \ref{nonDecreasingHead} we know \nf{p.Head} is non-decreasing so $\nf{j}\leq\nf{i}$.
+The reader may wonder when the case $\nf{j}=\nf{i}$ happens, it happens when \nf{p.blocks[j]=null} while \nf{j} is read and \nf{R\sub{p}} puts its created block into \nf{n.blocks[j]}.
+\end{proof}
 
 \begin{lemma}
-  If b$\in$n.parent.blocks[i] then b.super$\leq$i.
-\end{lemma}
-\begin{proof}
-  I did not quite understand the way Eric told me to prove this.
-\end{proof}
-
-\begin{lemma}
-If Rn=n.Refresh() puts b in n[h] at $t_{321}^{Rn}$, then the block created by the next successful P.Refresh() or the second next successful p.Refresh() contains b and b.super is set before the second next successful p.Refresh() after $t_{321}^{Rn}$.
-\end{lemma}
-
-\begin{proof}
-b.super could be set by helping of some n.Refresh() instance before the second next successful p.Refresh(), but even if no one helped it we prove the second next  successful p.Refresh() will help n to advance in Line \ref{setSuper} called by Line \ref{helpAdvance}.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=7in]{pics/timeLineSuper1.png}
-  \caption{b.super is set by some instance of Refresh() on n or n.parent showed by \nf{R'} and \nf{n.head} is incremented by some \nf{Refresh()} called \nf{R"}.}
-  \label{fig::timeLine1}
-\end{figure}
-
-\nf{Rp} is the first successful \nf{p.Refresh()} instance that has propagated \nf{b} into \nf{p.blocks[x]}.
-We want to know how great $\nf{x}-\nf{b.super}$ can be?
+If \nf{R\sub=n.Refresh()} puts \nf{b} in \nf{n.blocks[h]} at Line \ref{cas}, then the block created by one of the next two successful \nf{p.Refresh}es according to the Definition \ref{orderRefresh} contains \nf{b} and \nf{b.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{p.Refresh()}.
+\end{lemma}
+
+\begin{proof}
+It is sufficient to prove one of the two successful \nf{p.Refresh()}es propagate \nf{b}.
+If the first successful \nf{p.Refresh()} propagated \nf{b} then the claim is true, so in the remaining part we assume the first \nf{p.Refresh()} did not propagate \nf{b} and prove the second \nf{p.Refresh()} propagates \nf{b}.
+
+\nf{b.super} is set by some instance of \nf{Refresh()} on \nf{n} or \nf{p} showed by \nf{R'} and \nf{n.head} is incremented by some \nf{Refresh()} called \nf{R"}.
+We want to know how great $\nf{j}-\nf{b.super}$ can be.
 \nf{p.head} is \nf{hp} when \nf{R'} reads it.
 From Lemma \ref{lem::headProgress} \nf{p.head} could only increase from \nf{hp} to \nf{hp+1} if $\nf{p[hp]}\neq\nf{null}$.
-In other words there should be a successful \nf{p.Refresh()} in order to \nf{p.head} advance.
-We claim there cannot be more than one \nf{p.Refresh()} after \nf{R'} reads \nf{p.head} and before $t_{\ref{lastLine}}^{Rp}$.
+In other words there should be a successful \nf{p.Refresh()} for \nf{p.head} to increase.
+We claim there cannot be another successful \nf{p.Refresh()} after \nf{R'} reads \nf{p.head} and before \nf{R\sub{p}} performs Line \ref{lastLine}.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=7in]{pics/timeLineSuper1.png}
+  \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
+  \label{fig::timeLine1}
+\end{figure}
 
 Assume the first successful \nf{p.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{b.super}.
-It might happen maybe because the value read for $h_{left}$ in Line \ref{readChildHead} is less than \nf{h} or maybe $\nf{h}=\nf{h\sub{left}}$ and $\nf{left.blocks[h\sub{left}]}=\nf{null}$, which means \nf{n.head} is advanced but \nf{b} is still not installed in \nf{n[h]} which means \nf{R\sub{n}} has not reached to the Line \ref{cas}.
+It might happen maybe because the value read for $h_{left}$ in Line \ref{readChildHead} is less than \nf{i} or maybe $\nf{i}=\nf{h\sub{left}}$ and $\nf{left.blocks[h\sub{left}]}=\nf{null}$, which means \nf{n.head} is advanced but \nf{b} is still not installed in \nf{n.blocks[i]} which means \nf{R\sub{n}} has not reached to the Line \ref{cas}.
 
 Let the first successful \nf{p.Refresh()} be \nf{Rp1} and the second next successful \nf{p.Refresh()} be \nf{Rp2}.
-If \nf{Rp1} reads \nf{i} in Line \ref{readHead}, then \nf{Rp2} has to read \nf{i+1} in Line \ref{readHead}.
+If \nf{Rp1} reads \nf{x} in Line \ref{readHead}, then \nf{Rp2} has to read \nf{x+1} in Line \ref{readHead} (iduced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
 See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances \nf{Rp1, Rp2} on \nf{p}.
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=5in]{pics/timeLineSuper3.png}
-  \caption{ The second Refresh on p contains b and reads head>h.}
+  \center\includegraphics[width=6in]{pics/timeLineSuper3.png}
+  \caption{ The second Refresh on \nf{p} contains \nf{b} and reads $\nf{n.head}>\nf{i}$.}
   \label{fig::timeLine3}
@@
 So \nf{b.super} has set by some process before the second next successful \nf{p.Refresh()} on Line \ref{setSuper1}.
-Since \nf{h} is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value fo \nf{left.head} greater than \nf{h} and propagates \nf{b} to \nf{p}.
+Since \nf{i} is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value fo \nf{left.head} greater than \nf{h} and propagates \nf{b} to \nf{p}.
 So if \nf{b} was not propagated already we are sure the second next successful \nf{p.Refresh()} propagates \nf{b}.
@@
 \begin{corollary}
-If $\nf{b.super} \neq \nf{null}$, then it has at most 1 difference with the index of the superblock of \nf{b} in \nf{p}.
+If \nf{b} has propagated to \nf{f}, then \nf{b.super} has at most 1 difference with the index of the superblock of \nf{b} in \nf{p}.
 \end{corollary}

### 2022-09-20 thesis/thesis.tex
review true Refresh

@@
 \pagebreak
-\section{Our work}
+\section{Our Queue}
 Jayanti and Petrovic introduced a wait-free polylogarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We are going to design a polylogarithmic multi-enqueuer multi-dequeuer queue using some of their ideas.
-But we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.
-We apply two ideas from their work to create a new shared data structure which enables processes to agree on the linearization ordering of the processes.
-We use the shared tournament tree among $p$ processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes.
+We benefit from some ideas of their work to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
+Our algorithm despite them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
+I our model there are $p$ procesess doing \nf{Enqueue(),Dequeue()} operations concurrently.
+We use a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes.
 Each process has a leaf which the order of operations invoked by the process is stored in it.
-When the process wishes to do an operation it appends the operation to its leaf's sequence and after that, the process tries to propagate its new operation up to the tree's root.
-An ordering of operations propagated up to a node is stored in that node.
-All processes agree on the sequence stored in the root that is used as the linearization ordering.
+When a process wishes to do an operation it appends the operation to its leaf and then tries to propagate its new operation up to the tree's root.
+In each node the ordering of operations propagated up to it is stored.
+All processes agree on the sequence stored in the root and it is defined to be the linearization ordering.
 \begin{figure}[hbt]
@@
 \caption{ّ\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root, which stores the totall ordering of all operations.}
-\end{figure}
-
-As we said in  each node the sequence of operations is stored.
-We implement the sequence using an array and appending to the sequence by doing \tt{CAS} operations on the first null element in the array.
-In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s both children (that have not already been propagated to $n$) and tries to append them into $n$.
-The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
-Because if both of the \nf{Refresh()}es fail there is another instance of \nf{Refresh()} in between which has succeeded to do \nf{CAS} and has already appended the operations the first \nf{Refresh} was trying to append.
-This mechanism makes us to do \nf{TryAppend}s twice instead of \nf{Append}ing to the array.
-From the next paragraph we explain our ideas which are different from Jayanti.
+Each node tries to propagate its operations up to the root, which stores the total ordering of all operations.}
+\end{figure}
+
+\it{Add sequence to nodes}
+
+We could implement the sequence stored in each node using an array of the queue operations and append some operations to the sequence by doing \tt{k-CAS} operation on the end of the array.
+To do a propagate step on node $n$ in the tree, we aggregate the operations from node $n$'s both children (that have not already been propagated to $n$) and try to append them into $n$.
+We call this procedure \sc{Refresh}$(n)$.
+The main idea is that if we call \sc{Refresh}$(n)$ twice, the operations in $n$'s children before the first \sc{Refresh}$(n)$ are guaranteed to be in $n$.
+Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh()} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append.
+This mechanism makes us overcome the \nf{CAS} Retry Problem.
 
@@
   \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $r_1$ & $l_1$ & $r_2$ & $l_2$ & $l_3$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c||c|c|}  \hline $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c||c|c|}  \hline $r_1$ & $r_2$ & $r_3$ & $r_4$\\ \hline \end{tabular}} ]
 \end{tikzpicture}}
@@
   \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
+\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $r_1$ & $l_1$ & $r_2$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $r_3$ & $r_4$  \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c|c||}  \hline $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ \\ \hline \end{tabular}}
+{\begin{tabular}{|c|c|c|c||}  \hline$r_1$ & $r_2$ & $r_3$ & $r_4$\\ \hline \end{tabular}} ]
 \end{tikzpicture}}
@@
 \end{subfigure}
-\caption{\label{fig::propagstep} Successful \texttt{Refresh}, operations in children after $||$ are new.}
+\caption{\label{fig::propagstep} Before and after of a \sc{Refresh}$(n)$ with successful \nf{CAS}.
+Operations propagating from the left child are numbered with $l$ and from the right child by $r$ and the operations in children after $||$ are new.}
 \end{center}
 \end{figure}
-
-Instead of storing operations explicitly in the nodes, we are going to keep track of some statistics of them.
-This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations.
-In each \nf{Refresh()} there is at most one operation from  each process trying to be propagated.
-Since one operation cannot invoke two operations concurrently.
-Also as the operations in a \textsc{Refresh()} step are concurrent we can linearize them  among themselves in any order we wish.
-Note that if two operations are in one \nf{Porpagate()} step in a node they are going to be propagated up to the root together.
-From now on instead of propagating sequence of operations in  \nf{Refresh} steps we propagate \nf{block}s of operations.
-The idea is that we can describe a blocks contents only using some numbers.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way if we know the number of total operations in a block and the number of operations from the left child we can order the operations in a unique way which is a complete ordering.
-We can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).
-In the next paragraphs we explain the reason behind our choice of ordering.
-
+\it{fix $||$}
 \begin{figure}[h]
@@
 
-Previously we talked about storing the sequence of operations in the nodes of the tree.
-A process may wish to know information about the root ordering.
-Two functionalities are to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization.
-Since our algorithm is aimed for a queue, we make some assumptions here that we one wish only to know the order of a dequeue and one only tries to get an enqueue.
-We will explain in detail in the next but for now let's say \nf{enqueue}s and \nf{dequeue}s are appended to the tree and when one wants to find the response to a dequeu, it computes the order of the dequeue in linearization and using this information it computes which eneueue is the answer to the dequeue or if the answer is null.
+The solution for  implementing the orderings in the tree told above is not efficient, because there are big \nf{CAS}es and operations information are copied all the way up to the root.
+Instead of storing operations explicitly in the nodes, we can keep track of some statistics of them.
+This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
+To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations.
+In each \nf{Refresh()} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Furthermore since the operations in a \textsc{Refresh()} step are concurrent we can linearize them  among themselves in any order we wish.
+Note that if two operations are in read one \textsc{Refresh()} step in a node they are going to be propagated up to the root together.
+Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
+In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.
+
+A process may wish to know the $i$th propagated operation or the rank of a propagated operation in the linearization.
+In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with its rank.
+\nf{enqueue}s and \nf{dequeue}s are appended to the tree and when we want to find the response to a \nf{dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null.
 If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetENQ(n,b,i)}.
@@
 Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
-
- \pagebreak
-
-The definition of linearizability allows concurrent operations to be reordered arbitrarily.
-Thus, a group of concurrent operations can be appended to our root sequence as one block without specifying the order among the operations.
-
-\begin{figure}
-
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
-	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]
-
-\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:block} Showing concurrent operation sets with blocks.
-Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
-Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
-Blocks between two lines $||$ are propagated together to the parent.
-For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent.
-The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.}
-\end{figure}
-
-In the original algorithm  we differ between the enqueues and the dequeues in a block. Later we discuss about the reason.
-
-Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
-For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
-Block (5,3) also points to elements from blocks (0,5) and (1,2).
-
-\paragraph{}
-We choose to linearize operations in a block from the left child before those from the right child as a convention.
-Operations within a block of the root can be ordered in any way that is convenient.
-In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
-So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
-All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
-
-\pagebreak
-
-\paragraph{}
-In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes.
-So in a complete propagation, at most $2p$ blocks are merged into one block.
-\texttt{(maybe useful for analysis)}
-
-\paragraph{}
-\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence.
-We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
-To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.
-
-Furthermore, in each block, we store the prefix sum of left and right elements.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]
-
-\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:prefix} Using Prefix sums in blocks.
-When we want to find block b elements in its children, we can use binary search.
-The number below each block shows the count of elements in the previous blocks.}
-\end{figure}
-
-\begin{figure}[hbt]
-\centering
-  \includegraphics[width=5in]{pics/pointers}
-  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
-\end{figure}
-
-\pagebreak
-
-\paragraph{}
-Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
-From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
-So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
-
-\paragraph{Design of a Queue}
-
-Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
-Each internal node of the tree contains an array of blocks of elements.
-Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks.
-Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order.
-Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively.
-Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
-Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
-The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
-After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).
-
-In order to compute elements of a block faster we store prefix-sum blocks(block i has tuple(right-sum=$\#$right ops in previous block, left-sum=$\#$left ops in previous blocks)[See Figure \ref{fig:prefix}].
-Here is the algorithm to get elements of a block.
-
-\paragraph{Specification}
-A Queue is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
-\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
-\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
-
-\pagebreak
-
-\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
-\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
-Other attributes are assigned values followed by definition.
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
-\end{figure}
-
-%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
-\end{figure}
-
-\pagebreak
-
-\paragraph{Computing \texttt{Get(n, b, i)}}
-To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$.
-Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.
-
-\paragraph{How \texttt{Refresh(n)} works.}
-\begin{enumerate}
-  \item Read n's counter and head
-  \item Create block b
-  \item CAS b into n
-  \item If previous succeed:
-  \begin{enumerate}
-  \item Update sup of b's ending subblocks
-  \item Increment children's counters
-  \end{enumerate}
-\end{enumerate}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/super}
-  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
-\end{figure}
-
-\pagebreak
-
-\paragraph{Computing superblock}
 
@@
 So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
-%
+
+\begin{figure}
+
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
+	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
+	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]
+
+\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
+ [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
+ [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
+  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
+          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:block} Showing concurrent operation sets with blocks.
+Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
+Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
+Blocks between two lines $||$ are propagated together to the parent.
+For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent.
+The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.
+Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
+For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
+Block (5,3) also points to elements from blocks (0,5) and (1,2).
+We choose to linearize operations in a block from the left child before those from the right child as a convention.
+Operations within a block of the root can be ordered in any way that is convenient.
+In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
+So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
+All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
+}
+\end{figure}
+
+\paragraph{}
+\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence.
+We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
+To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.
+
+Furthermore, in each block, we store the prefix sum of left and right elements.
+Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]
+
+\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:prefix} Using Prefix sums in blocks.
+When we want to find block b elements in its children, we can use binary search.
+The number below each block shows the count of elements in the previous blocks.}
+\end{figure}
+
+\begin{figure}[hbt]
+\centering
+  \includegraphics[width=5in]{pics/pointers}
+  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
+\end{figure}
+
+\paragraph{}
+Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
+From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
+So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
+
+\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
+\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
+Other attributes are assigned values followed by definition.
+\begin{figure}[hbt]
+  \center\includegraphics[width=5.5in]{pics/createblock}
+  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
+\end{figure}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/doublerefresh.png}
+  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
+\end{figure}
+
+\paragraph{Computing \texttt{Get(n, b, i)}}
+
+\paragraph{How \texttt{Refresh(n)} works.}
+
+\paragraph{Computing superblock}
+
 %\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
 \paragraph{}
-Next, we describe how to use block tree to implement queues.
-The block tree, maintains the history of all operations, not only the current state of the queue.
-Now consider the following history of operations.
-What should each \texttt{Dequeue()} return?
-We can implement Enqueue and Dequeue using our block tree.
-An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
-To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
+Now, we describe how to use the tree to implement a queue.
+Consider the following execution of operations.
+\texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
+What should a \texttt{Dequeue()} return?
+To compute the response of a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
 Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
@@
 
-\pagebreak
 \paragraph{}
@@
 
-\pagebreak
-
 \subsection{Pseudocode description}
+
+\paragraph{Specification}
+A Queue is a shared data structure that stores a sequence of elements.
+It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
+\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
+\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
 
@@
 
-\subsection{Proof of Correctness}
+\section{Proof of Correctness}
 
@@
 \end{proof}
-\textit{useless?}
 
@@
 
+\begin{definition}
+Let $^{op}t$ be the time $op$ is invoked and $t^{op}$ be the time $op$ terminates.
+  $_{l}t$ is the immediate time before running Line $l$.
+$t_{l}$ is the immediate time after running Line $l$.
+$^{op}t_{l}$ is the immediate time before running Line $l$ of operation $op$.
+$t_{l}^{op}$ is the immediate time after running Line $l$ of operation $op$.
+$v_l$ is the value of \nf{v} immediately before line $l$ for the process we are talking about.
+\end{definition}
+
+In the next Lemmas we show for every \nf{Refresh()} that has has succesuful \nf{CAS} in Line \ref{cas}, all the established operations in the children before the \nf{Refresh} are going to be established in the parent after the \nf{Refresh}.
+
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Let $t_i$ be the time an instance \nf{R} of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates.
-If the \nf{TryAppend(new, s)} of \nf{R}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
-\end{lemma}
-\begin{proof}
-Since \nf{TryAppend} returns true a block \nf{new} is written into \nf{n.blocks[h]} in Line \ref{cas}.
+If \nf{CAS} in Line \ref{cas} of instance $R$ of \nf{n.Refresh()} returns \nf{true}, we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
+ ops(n.blocks_{t_{\ref{cas}}^R})$.
+\end{lemma}
+\begin{proof}
 %So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
 
-We show \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
-Let \nf{h} be the value \nf{n.Refresh()} reads from \nf{n.head} at line \ref{readHead}, \nf{h\sub{left,i}} be the value of \nf{n.left.head} at \nf{t\sub{i}} and \nf{h\sub{left,read}} be the value read from \nf{n.left.head-1} at line \ref{lastLine}.
-\nf{end\sub{left}} field of the block returned by \nf{CreateBlock(i)} is \nf{h\sub{left,read}}.
-By lines \ref{prevLine} and \ref{lastLine} the \nf{new} block in \nf{n.blocks[h]} contains \nf{n.left.blocks[n.blocks[h-1].end\sub{left}+1..h\sub{left,read}]}.
-Since \nf{left.head} is read after \nf{t\sub{i}} then \nf{h\sub{left,read}>h\sub{left,i}} which means \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..h\sub{left,read}])}.
-After the successful \nf{TryAppend} in line \ref{cas} we know all blocks in \nf{n.left.blocks[0..h\sub{left,read}-1} are subblocks of \nf{n.blocks[0..h]} by the definition of subblock.
-At \nf{t\sub{t}} we have \nf{n.head>h} by Lemma \ref{lem::headProgress}.
-So \nf{n.blocks[1..h]} are in \nf{EST\sub{n,t\sub{t}}} by definition of \nf{EST}.
-Note that after line \ref{incrementHead2} we are sure that the \nf{head} is incremented by Lemma\ref{lem::headInc}) which means \nf{n.head=h+1} at $t_t$ so the new block is established at $t_t$ and the new block contains the new operations which is what we wanted to show.
-The proof for \nf{ops(EST\sub{n.right, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}is the same.
+We show $ops(EST_{\nf{n.left}, ^Rt})= ops(\nf{n.left.blocks[0..$n.left.head_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(\nf{n.blocks[0..}n.head_{\ref{readParentHead}}])$.
+%Let $h$ be the value \nf{n.Refresh()} reads from \nf{n.head} at Line \ref{readHead} and $h_{left}$ be the value read from \nf{n.left.head} at Line \ref{lastLine}.
+As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}.
+By the Definition \ref{def::subblock} for the Subblock  the block \nf{new} contains \nf{n.left.blocks[n.blocks[$n.head_{\ref{readHead}}-1$].end\sub{left}+1..$n.left.head_{\ref{lastLine}}-1$]} (see Lines \ref{prevLine} and \ref{lastLine}).
+% \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..h\sub{left,read}])}.
+After the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[0..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[0..$n.head_{\ref{readHead}}$]}.
+Because of the Lemma \ref{nonDecreasingHead} we have, $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{readParentHead}}$.
+From Lemma \ref{lem::establishedOrder} the inequality we proved is stronger than what we wanted to prove (TODO: make the lemma claim more general).
+The proof for the right child is the same.
 
@@
   \center\includegraphics[width=5in]{pics/trueRefresh}
-  \caption{New established operations of the left child are in the new block.}
-\end{figure}
-\begin{lemma}[Stronger True Refresh] \label{lem::prectrueRefresh}
-  Let $t_i$ be the time an instance of \nf{n.Refresh()} read the head (Line \ref{readHead}) and $t_t$ be the time its \nf{TryAppend(new, s)} terminates with  and returns \nf{true} (Line \ref{cas}).
-We have \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(n.blocks)}.
-\end{lemma}
+  \caption{New established operations of the left child are in the new block. (TO UPDATE)}
+\end{figure}
+
+\begin{corollary}[Weak True Refresh] \label{lem::prectrueRefresh}
+If \nf{CAS} in Line \ref{cas} of instance $R$ of \nf{n.Refresh()} returns \nf{true}, we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
+ ops(EST_{\nf{n}, t^R})$.
+\end{corollary}
+\begin{proof}
+The left side is the same as the Lemma \ref{lem::trueRefresh}, so it is sufficient to show when \nf{R} terminates established blocks in \nf{n} are super set of \nf{n.blocks} after the Line \ref{cas}.
+Because of Lemma \ref{lem::headInc}  we are sure that the \nf{n.head} is incremented after line \ref{incrementHead}.
+So the \nf{new} block appended to \nf{n} is established at $t^R$.
+\end{proof}
+
 \begin{definition}
-  An instance of \nf{Refresh()} is successful iff its \nf{TryAppend(new, s)} terminates with  and returns \nf{true}.
+  An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
 \end{definition}
@@
 \pagebreak
-
-\begin{definition}
-Let $_{R_1}t$ be the time $R_1$ is invoked and $t_{R_2}$ be the time $R_2$ terminates.
-  $_{line}t$ is the immediate time before running Line $line$.
-$t_{line}$ is the immediate time after running Line $line$.
-$_{line}t^{op}$ is the immediate time before running Line $line$ of operation $op$.
-$t_{line}^{op}$ is the immediate time after running Line $line$ of operation $op$.
-\end{definition}
 

### 2022-09-20 thesis/thesis.tex
double check page 20

@@
 
+\begin{definition}
+  An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
+\end{definition}
+
 In the next Lemmas we show for every \nf{Refresh()} that has has succesuful \nf{CAS} in Line \ref{cas}, all the established operations in the children before the \nf{Refresh} are going to be established in the parent after the \nf{Refresh}.
@@
 \begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-If \nf{CAS} in Line \ref{cas} of instance $R$ of \nf{n.Refresh()} returns \nf{true}, we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
+If $R$ is a succesful instance \nf{n.Refresh()}, then we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
  ops(n.blocks_{t_{\ref{cas}}^R})$.
@@
 Because of the Lemma \ref{nonDecreasingHead} we have, $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{readParentHead}}$.
-From Lemma \ref{lem::establishedOrder} the inequality we proved is stronger than what we wanted to prove (TODO: make the lemma claim more general).
+From Lemma \ref{lem::establishedOrder} the inequality we proved is stronger than what we wanted to prove (TODO: make the lemma 13 claim more general, if time passes n.blocks grows).
 The proof for the right child is the same.
@@
 \begin{corollary}[Weak True Refresh] \label{lem::prectrueRefresh}
-If \nf{CAS} in Line \ref{cas} of instance $R$ of \nf{n.Refresh()} returns \nf{true}, we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
+If $R$ is a complete instance \nf{n.Refresh()}, then we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
  ops(EST_{\nf{n}, t^R})$.
@@
 
-\begin{definition}
-  An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
-\end{definition}
 %\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}

### 2022-09-20 thesis/thesis.tex
review pages 17,18,19

@@
 
-\framebox[1.1\width]{TEST} Fix the logical order of definitions (cyclic refrences).
-
-%\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
+To prove a shared data structure works correctly it is sufficient to show it is linearizable.
+In our case which we create the linearization ordering in the root, we need to prove the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and if we do the linearization ordering created on a sequential queue we get the same results as our queue.
+The proof is structured like this.
+First we define and prove some facts about blocks and node's head properties.
+Then we introduce  the linearization ordering formally.
+Then we prove double \nf{Refresh} on a node is enough to propagate children's new operations up to the node.
+After this we prove some claims about the size of each block.
+Having propagates working  in a wait-free manner then we prove the correctness of \nf{DSearch()} and \nf{GetENQ()} and \nf{IndexDEQ}.
+Having these two main methods we prove the way we compute the response of a dequeue is correct and finally we prove the linearizability.
+
+%\framebox[1.1\width]{TEST} Fix the logical order of definitions (cyclic refrences).
+%
+%%\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
 Is lemma 7 necessary?
 Is lemma 13 induced trivially from lemma 8?
-
-\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?
-
-%\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
+%
+%\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?
+%
+%%\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
 block b $\sqsubseteq$ block c means b is subblock of c.
@@
 Merge these two to have shorter formulaes.
-
-\framebox[1.1\width]{Question} A good notation for \it{the index of the \nf{b}}?
-
-\framebox[1.1\width]{Question} How to remove the notion of time?
+%
+%\framebox[1.1\width]{Question} A good notation for \it{the index of the \nf{b}}?
+%
+%\framebox[1.1\width]{Question} How to remove the notion of time?
 To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t.
@@
 
-\begin{definition}[Block]
-A block is an object storing some statistics, as described in Algorithm Queue.
-A block in a node's blocklist implicitly represents a set of operations.
-If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}.
-Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if the index of the \nf{b} is smaller than the index of the \nf{b$^\prime$}'s.
-For a \tt{block} in a \tt{BlockList} we define \it{the prefix for the block} to be the blocks in the \tt{BlockList} up to and including the \tt{block}.
+A block is an object storing some statistics, as described in Algorithm Queue. A block in a node's blocklist implicitly represents a set of operations.
+\begin{definition}[Ordering of a block in a node]
+Let $b$ be \nf{n.blocks[$i$]} and $b^\prime$ be \nf{n.blocks[$j$]}.
+We call $i$ the \emph{index} of block $b$.
+Block $b$ is \it{before} block $b^\prime$ in node \nf{n} if and only if $i<j$.
+We define \it{the prefix for} $b$, to be the blocks in the \tt{n.blocks[0..$i$]}.
 \end{definition}
 
+Next we show that \nf{head} in a node can only be increased and when a \nf{Refresh} terminates \nf{head} is incremented by the process itself or maybe helped by another process.
+\begin{observation} \label{nonDecreasingHead}
+  \nf{head} is non-decreasing over time.
+\end{observation}
+\begin{proof}
+  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()} .
+\end{proof}
 \begin{lemma}[head Increment] \label{lem::headInc}
-Let \nf{R} be an instance of \nf{Refresh} on node \nf{n} that reaches Line \ref{cas}. After \nf{R} terminates \nf{n.head} is greater than h, the value read  in line \ref{readHead} of \nf{R}.
-\end{lemma}
-\begin{proof}
-If Line \ref{incrementHead1} or \ref{incrementHead2} are successful then the claim holds, otherwise another process has incremented the head from \nf{h} to \nf{h+1}.
-\end{proof}
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
+Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
+\end{lemma}
+\begin{proof}
+If the \nf{CAS} in Line \ref{incrementHead} is successful then the claim holds.
+Otherwise it has changed from the value that was read in Line \ref{readHead}, which by Observation \ref{nonDecreasingHead} it means another process has incremented the head.
+\end{proof}
+
+Now we show \nf{n.blocks[n.head]} is the last  block written into \nf{n} or the first empty block in \nf{n}.
+
+\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is $h$ then $\nf{n.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{n.blocks[i]}\neq\nf{null}$ for ${i<h}$.
 \end{invariant}
 \begin{proof}
-The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}.
+The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[}$x$\nf{]} is null for every $x>0$.
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
@@
 
-In the algorithm, some value is appended to \nf{n.blocks[]} by writing into \nf{n.blocks[head]} only in Line 313.
-Writing into \nf{n.blocks[head]} preserves the invariant, since the claim does not talk about \nf{n.blocks[head]}.
-The value of \nf{n.head} is modified only in lines \ref{incrementHead1} and \ref{incrementHead2}.
-Depending on whether the \nf{TryAppend()} in Line~\ref{cas} succeeded or not, we show that the claim holds after the increment of \nf{n.head} in either case.
-If \nf{n.head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds.
-In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore.
-Note that whether \ref{incrementHead1} or \ref{incrementHead1} return true or false,  after they finish we know  that \nf{n.head} has been incremented from the value read in Line \ref{readHead} (Lemma \ref{lem::headInc}).
-The failure case is also the same since it means some non-null value has been written into \nf{n.blocks[head]} by some process.
+In the algorithm, \nf{n.blocks} is modified only on Line \ref{cas} which updates \nf{n.blocks[$h$]} for $h$ read from \nf{n.head} in Line \ref{readHead}.Note that since the \nf{CAS} in Line \ref{cas} is successful it means \nf{n.head} has not changed from $h$ till doing the \nf{CAS}.
+Writing into \nf{n.blocks[n.head]} preserves the invariant, since the claim does not talk about the content of \nf{n.blocks[n.head]}.
+
+The value of \nf{n.head} is modified only in Line \ref{incrementHead}.
+If \nf{n.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
+When a process does Line \ref{incrementHead} it means Line \ref{cas} was finished which either case it is successful or not it means \nf{n.blocks[$h$]} is not \nf{null} anymore.
 %At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
@@
 
-\begin{observation} \label{nonDecreasingHead}
-  \nf{n.head} is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{n.head} is only incremented in Line \ref{incrementHead} of \nf{Refresh()} .
-\end{proof}
-
+The next lemma is used to prove the subblocks of two blocks in a node are disjoint. (I can't say its use without mentioning subblocks)
 \begin{lemma}[headProgress] \label{lem::headProgress}
- If \nf{n.blocks[i]$\neq$null} and \nf{i$.$0} then $\nf{n.blocks[i].end\sub{left}} \geq \nf{n.blocks[i-1].end\sub{left}}$ and $\nf{n.blocks[i].end\sub{right}} \geq \nf{n.blocks[i-1].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-
-Consider the block \nf{b} written into \nf{n.blocks[i]} by \nf{TryAppend()} at Line \ref{cas}.
-It is created by the \nf{CreateBlock(i)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock(i)}, \nf{n.head=i} at Line \ref{readHead}, so \nf{n.blocks[i-1]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}.
-Thus the \nf{CreateBlock(i-1)} that creates  \nf{b}$^\prime$ terminates before \nf{CreateBlock(i)} that creates \nf{b} is invoked.
-The value written into \nf{b.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock(i)}  was read from \nf{n.left.head-1} at Line \ref{lastLine} of \nf{CreateBlock(i)}.
-Similarly, the value in \nf{n.blocks[i-1].end\sub{left}} was read from \nf{n.left.head-1} during the call to \nf{CreateBlock(i-1)}.
-Since \nf{n.left.head} is non-decreasing \nf{b}$^\prime$\nf{.end\sub{left}} $\leq$ \nf{b.end\sub{left}}.
+ If $\nf{n.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
+\end{lemma}
+\begin{proof}
+
+Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}.
+$b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
+Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}.
+Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
+The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was minus one of the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
+Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was minus one of the value read from $\nf{n.left.head}-1$ during the call to \nf{CreateBlock($i-1$)}.
+As by Observation \ref{nonDecreasingHead} \nf{n.left.head} is non-decreasing, $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$.
 The proof for \nf{end\sub{right}} is similar.
@@
 
+We define the subblocks of a block recursively.
 \begin{definition}[Subblock]\label{def::subblock}
-Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} .
-Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if \nf{b} is a direct subblock of \texttt{n.blocks[i]} or a subblock of a direct subblock of \texttt{n.blocks[i]}.
+A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]} .
+Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
+We say block $b$ is \it{propagated to node} node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n.blocks$.
 \end{definition}
 
-\begin{corollary}[No Duplicates]\label{append}
-If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
+\begin{lemma} \label{lem::subblocksDistinct}
+Subblocks of any two blocks in node $n$ does not overlap.  
+\end{lemma}
+\begin{proof}
+We are going to prove by contradiction.
+Assume subblocks of $n.blocks[i]$ and $n.blocks[j]$ overlap which $i< j$.
+Without loss of generality assume left child subblocks of $n.blocks[i]$ overlap with the left child subblocks of $n.blocks[j]$.
+Direct left subblocks of $n.blocks[i]$ and $n.blocks[j]$ cannot overlap.
+Because of Lemma \ref{lem::headProgress} we have $\nf{n.blocks[$i-1$].end\sub{left}} \leq \nf{n.blocks[$i$].end\sub{left}} \leq \nf{n.blocks[$j-1$].end\sub{left}} \leq \nf{n.blocks[$j$].end\sub{left}}$.
+Since the tree height is finite we can go lower in the tree to reach a node that direct subblcoks of two different blocks in a node overlap.
+\end{proof}
+
+We define the operations of a block using the definition of subblocks.
+\begin{definition}[Operations of a block]\label{def::ops}
+A block $b$ in a leaf represents an \nf{Enqueue(e)} if $b.element=e\neq\nf{null}$ otherwise, if $b.element=\nf{null}$, $b$ represents a \nf{Dequeue()}.
+The set of operations of block \nf{b} is the union set of the operations in subblocks of \nf{b}.
+We denote the set of operations of block \nf{b} by \nf{ops(b)}.
+We also say \nf{b} contains \nf{op} if \nf{op}$\in$\nf{ops(b)}.
+\end{definition}
+
+\begin{definition}[Superblock]
+  Block \nf{b} is \it{superblock} of block \nf{c} if c is a direct subblock of b.
+\end{definition}
+\begin{corollary}
+Every block has at most one superblock.
 \end{corollary}
 \begin{proof}
-  Operation \nf{op} is invoked only one time in an execution because every operations invoked is distinct.
-Since there is node \nf{n} which \nf{op} is in two different blocks of \nf{n}, there is node \nf{n}$^\prime$ that is the lowest height node in the tree that contains \nf{op} in two of its blocks \nf{b1,b2}.
-By Definition \ref{def::subblock}, \nf{b1} and \nf{b2} have  distinct subblocks(not only direct subblocks) and since \nf{op} is in only one leaf block, then it cannot be in both \nf{b1} and \nf{b2}.
-\end{proof}
-
-%(The \nf{end\sub{left}, end\sub{right}} fields are written in \ref{endDefLine} of \nf{CreateBlock()})
-
-\begin{definition}[Superblock]
-  Block \nf{b} is   \it{direct superblock} of block \nf{c} if c is a direct subblock of b.   Block \nf{b} is \it{superblock} of block \nf{c} if \nf{c} is a subblock of \nf{b}.
+A block having more than one superblock is contradictory with Lemma \ref{lem::subblocksDistinct}.
+\end{proof}
+
+Operations are distinct \nf{Enqueue(e)} or \nf{Dequeue()}s invoked by processes. Next lemma proves that each operation appears once in the blocks of a node.
+\begin{lemma}[No Duplicates]\label{append}
+If $op$ is in $n.blocks[i]$ then there is no $j\neq i$ such that $op$ is in $ops(n.blocks[j])$.
+\end{lemma}
+\begin{proof}
+We prove this claim using Corollary \ref{lem::subblocksDistinct}.
+Assume $op$ is in the subblocks of both $n.blocks[i]$ and $n.blocks[j]$.
+From Corollary \ref{lem::subblocksDistinct} we know these blocks are different.
+So the leaf blocks containing $op$ are different, but since we assumed operations are distinctive it means one operation cannot be invoked two times.
+This leads us to contradictory with the hypothesis.
+\end{proof}
+
+\begin{definition}
+$n.blocks[i]$ is \emph{established} at time $t$ if $n.head>i$. \emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
 \end{definition}
 
-\begin{definition}[Operations of a block]\label{def::ops}
-A leaf block \nf{b} in a leaf represents \nf{enqueue(x)} if \tt{b.element=x$\neq$null}.
-Else if \tt{b.element=null} b represents a \nf{dequeue()}.
-The set of operations of block \nf{b} are the operations in the subblocks of \nf{b}.
-We denote the set of operations of block \nf{b} by \nf{ops(b)}.
-
-We say block \nf{b} is \it{propagated to node} \nf{n} if \nf{b} is in \nf{n.blocks} or is a subblock of a block in \nf{n.blocks}. We also say \nf{b} contains \nf{op} if \nf{op}$\in$\nf{ops(b)}.
-\end{definition}
-
-%\begin{definition}
-%  Block \nf{b} in node \nf{n} is new if \nf{b} is not a subblock of any block in \nf{n.parent.blocks[]}.
-%\end{definition}
-
-\begin{definition}
- A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if \nf{n.head}$>$ index of \nf{b} at time $t$. \emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
-\end{definition}
-
-\begin{observation}\label{head}
-Once a block \nf{b} is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
-\end{observation}
-
-\begin{lemma}
-Every block has at most one direct superblock.
-\end{lemma}
-\begin{proof}
-To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}.
-\nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine}, \ref{prevLine}).
-Since \nf{end\sub{left},end\sub{right}} are non-decreasing (\nf{n.blocks[i].end\sub{left|right}>n.blocks[i-1].end\sub{left|right}}), so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
-\end{proof}
-
-%\begin{proof}
-%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
-%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
-%\end{proof}
-
-%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
-%  If block \nf{b} in node \nf{n} is established, then all the blocks in \nf{n.blocks[]} before \nf{b} are established.
-If block \nf{b} in node \nf{n} is not new then all the blocks in \nf{n.blocks[]} before \nf{b} are not new.
-%\end{lemma}
-%\begin{proof}
-%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
-%\end{proof}
-
-\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
+Now we want to say block of a node grow over time.
+\begin{corollary}[Growing blocks]\label{lem::blocksOrder}
+  If  time $t<$ time $t^\prime$, then $ops(n.blocks)$ at time $t$ $\subseteq$ $ops(n.blocks)$ at time $t^\prime$.
+\end{corollary}
+\begin{proof}
+Blocks are only appended (not modified) with \nf{CAS} to $n.blocks[n.head]$ and $n.head$ is non-decreasing, so the set of blocks of a node contains the the set of blocks before the \nf{CAS}.
+\end{proof}
+
+\begin{corollary}[establishedOrder]\label{lem::establishedOrder}
   If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
-\end{lemma}
-\begin{proof}
-Blocks are only appended (not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node can only grow.
-\end{proof}
-
-%\begin{lemma}[createBlock] \label{lem::createBlock}
-%  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
-%\end{lemma}
-%\begin{proof}
-%From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
-%We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}.
-It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$.
-If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see.
-Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it.
-Right child is the same.
-%% WLOG we prove the claim for the left child.
-Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.
-Since the head is only increasing (Lemma~\ref{lem::headProgress}) the  claim holds after $t$.
-See Figure~\ref{fig::createBlock}.
-The right child is the same.
-%\end{proof}
-%
-%\nf{CreateBlock()} aggregates the established blocks in the children that are not already established in the parent into one block.
-If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence.
-So suppose two \nf{Refresh}es fail.
-Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}.
-So it means the second failed \nf{Refresh} is concurrent with another successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}.
-After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-%\text
+\end{corollary}
+
 \pagebreak
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
-\end{figure}
-
-$\blacktriangleright$ Processes are numbered from 1 to $p$ and leaves of the tree are assigned from left to right.
+  \caption{Order of operations in b: operations in leaves are ordered with numerical order in the drawing.}
+\end{figure}
+
+Now we define the ordering store in each node.
+In the non-root nodes we only need ordering of operations of a type among themselves.
+Processes are numbered from 1 to $p$ and leaves of the tree are assigned from left to right.
 We will show in Lemma \ref{blockSize} that there is at most one operation from each process in a given block.
@@
 \begin{itemize}
-\item The prefix of an operation $op$ in the sequence of operations $S$ is the sequence of operations strictly before $op$.
-  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in \nf{ops(leaf.blocks[b])} or an empty sequence if \nf{leaf.blocks[b].num\sub{enq}=0}.
-If \nf{n} is an internal node, then
+  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n.blocks[b])$ defined recursively as follows.
+$E(leaf,b)$ is the single enqueue operation in $ops(leaf.blocks[b])$ or an empty sequence if $leaf.blocks[b].num_{enq}=~0$.
+If $n$ is an internal node, then
 \begin{align*} 
-E(n,b) =&  E(n.left,n.blocks[b-1].end\sub{left}+1)\cdot E(n.left,n.blocks[b-1].end\sub{left}+2)\cdots E(n.left,n.blocks[b].end\sub{left})\cdot \\ 
-&E(n.right,n.blocks[b-1].end\sub{right}+1)\cdot E(n.right,n.blocks[b-1].end\sub{right}+2)\cdots E(n.right,n.blocks[b].end\sub{right})
+E(n,b) =&  E(n.left,n.blocks[b-1].end_{left}+1)\cdot E(n.left,n.blocks[b-1].end_{left}+2)\cdots E(n.left,n.blocks[b].end_{left})\cdot \\ 
+&E(n.right,n.blocks[b-1].end_{right}+1)\cdot E(n.right,n.blocks[b-1].end_{right}+2)\cdots E(n.right,n.blocks[b].end_{right})
 \end{align*}
@@
 \item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in \nf{ops(leaf.blocks[b])} or an empty sequence if \nf{leaf.blocks[b].num\sub{deq}=0}.
-If \nf{n} is an internal node, then
+  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n.blocks[b])$ defined recursively as follows.
+$D(leaf,b)$ is the single dequeue operation in $ops(leaf.blocks[b])$ or an empty sequence if $leaf.blocks[b].num\sub{deq}=~0$.
+If $n$ is an internal node, then
 \begin{align*} 
-D(n,b) =&  D(n.left,n.blocks[b-1].end\sub{left}+1)\cdot D(n.left,n.blocks[b-1].end\sub{left}+2)\cdots D(n.left,n.blocks[b].end\sub{left})\cdot \\ 
-&D(n.right,n.blocks[b-1].end\sub{right}+1)\cdot D(n.right,n.blocks[b-1].end\sub{right}+2)\cdots D(n.right,n.blocks[b].end\sub{right})
+D(n,b) =&  D(n.left,n.blocks[b-1].end_{left}+1)\cdot D(n.left,n.blocks[b-1].end_{left}+2)\cdots D(n.left,n.blocks[b].end_{left})\cdot \\ 
+&D(n.right,n.blocks[b-1].end_{right}+1)\cdot D(n.right,n.blocks[b-1].end_{right}+2)\cdots D(n.right,n.blocks[b].end_{right})
 \end{align*}
@@
 \end{definition}
-$\blacktriangleright$ In the non-root nodes, we only need ordering of enqueues and dequeues among the operations of their own type.
-Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works with dequeues.
 \pagebreak

### 2022-09-21 thesis/thesis.tex
review pages 21,22

@@
 
-%\Statex \bf{Structure}
-
 \Statex $\diamondsuit$ \tt{\sl{Shared}}
@@
 \item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-%such that \tt{root} is the root and the left child and the right child and the parent of \tt{Tree[i]} are \tt{Tree[2i+1]}, \tt{Tree[2i+2]} and \tt{Tree[i/2]}.
 \end{itemize}
@@
 
-%\Statex $\diamondsuit$ \tt{\sl{Structures}}
-
 \Statex $\blacktriangleright$ \tt{\sl{Node}}
 \begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
-\item \tt{\sl{BlockList}}
+\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
+\item \tt{\sl{blocks}} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
 \item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-%\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
 \end{itemize}
 
-%\Statex $\blacktriangleright$ \tt{\sl{Root} extends \sl{Node}}
-%\begin{itemize}
-%  
-%\item \tt{\sl{PBRT} blocks}
-%
-%  \textsf{\tt{BlockList} is implemented with a persistent red-black tree.}
-%  
-%\end{itemize}
-
-%\Statex $\blacktriangleright$ \tt{\sl{NonRootNode} extends \sl{Node}}
-%\begin{itemize}
-
-%\end{itemize}
-
-%\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
-%\begin{itemize}
-%  
-%  \item \tt{\sl{int} \tt{last\sub{done}}}
-%
-%  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the.
-A block is finished if all of its operations are finished.
-\tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response.
-\it{put the definitions before the pseudocode}}
-%  
-%\end{itemize}
-
 \Statex $\blacktriangleright$ \tt{\sl{Block}}
-%\cmt{If \tt{b} is \tt{blocks[i](i!=0)} then \tt{b[-1] is blocks[i-1].}}
 
 \begin{itemize}
-%  \item \tt{\sl{int} num\sub{enq}, num\sub{deq}}
-%  \textsf{: \# enqueue, dequeue operations in the block}
-%  \item \tt{\sl{int} num, sum}
-%  \textsf{: total \# operations in block, prefix sum of \tt{num}}
-
   \item \tt{\sl{int} super}
@@
   \textsf{: size of the queue after performing all operations in the prefix for this block}
-%  \item \tt{\sl{int} sum\sub{non-null deq}}
-%  \textsf{: count of non-null dequeus up to this block}
-%  \item \tt{\sl{counter} num\sub{finished}}
-%  \textsf{: number of finished operations in the block}
-%    \item \tt{\sl{int} order}
-%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
 \end{itemize}
@@
 
-%##########################################
-
 \begin{footnotesize}
-
-%\it{Variable naming:}
-%\begin{itemize}
-%  \item \tt{b\sub{op}}: index of the block containing operation \tt{op}
-%  \item \tt{r\sub{op}}: rank of operation \tt{op} i.e. the ordering among the operations of its type according to linearization ordering
-%\end{itemize}
 
@@
 
-%##########################################
-
 \begin{algorithm}
@@
 \State \tt{newBlock.element= e}
-%\State \tt{b.num\sub{enq}=1}
 \State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
@@
 \label{deqRest}
-%\If{\tt{i\sub{enq}==-1}}
-%\State \tt{output= null}
-%\Else
-%\State \tt{output= \Call{GetEnq}{b\sub{enq}, i\sub{enq}}}
-%\EndIf
 \State \Return{\tt{output}}
@@
 
-\begin{algorithm}
-\caption{BlockList}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{700}
-
-\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}.
-Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}.
-If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}
-
-%\Statex
-%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
-%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
-%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
-%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
-%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
-%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
-%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
-%\State \tt{l.Help()}
-%\EndFor
-%\EndIf
-%\State \tt{blk.num\sub{finished}= 0}
-%%\State \tt{*oldRoot= \&root.blocks.root}
-%%\State \tt{*newRoot= root.blocks.Append(blk).root}
-%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
-%\State \Return{\tt{CAS(blocks[n], null, blk)}}
-%\EndFunction{TryAppend}
-
-\Statex
-%\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
-\Statex \tt{\sl{block[]} blocks: }\sf{array of blocks}
-\Statex
-\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
-\State \Return{\tt{CAS(blocks[n], null, blk)}}
-\EndFunction{TryAppend}
-
-\end{algorithmic}
-\end{algorithm}
-%##########################################
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{300}
-\begin{multicols}{2}
-
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-%\State \tt{hp= parent.\head} \label{readParentHead1}
-%\If{\nf{blocks[h]!=null}}
-%\State \tt{CAS(blocks[h].super, null, hp)} \label{helpSuper1}
-%\EndIf
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-%\cmt{Created block contains nothing.}
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
-%\cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\Statex
-
-\Function{void}{Advance}{\sl{int} h, \sl{int} hp}
-
-\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\pagebreak
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
-\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
-
-\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
-Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
-%\State \Return \tt{result block's index}
-\EndFunction{BSearch}
-
-\Statex
-
-\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
-\State \tt{block newBlock= \Call{new}{\sl{block}}}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
-\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
-\EndFor
-%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
-%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
-%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
-%\State \tt{b.sum= n.blocks[i-1].sum + b.num}
-
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
-%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
-\EndIf
-
-\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
-\EndFunction{CreateBlock}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
 \begin{algorithm}
@@
 
+\begin{algorithm}
+\caption{Leaf}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{600}
+
+\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
+\State \tt{blk.group= \head} \label{appendStart} 
+\State \tt{blocks[\head]= blk} 
+\State \tt{\head+=1} \label{appendEnd} 
+\State \tt{parent.}\Call{Propagate()}{} 
+\EndFunction{Append}
+
+\end{algorithmic}
+\end{algorithm}
+
+\pagebreak
+
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{300}
+\begin{multicols}{2}
+
+\Function{void}{Propagate()}{}
+\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
+\State \Call{Refresh()}{} \label{secondRefresh}
+\EndIf
+\If{\tt{this} \bf{is not} \tt{root}}
+\State \tt{parent.}\Call{Propagate()}{}
+\EndIf
+\EndFunction{Propagate}
+
+\Statex
+
+\Function{boolean}{Refresh()}{}
+\State \tt{h= \head} \label{readHead}
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
+\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
+\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
+\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
+\EndIf
+\EndFor \label{endHelpChild1}
+\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
+\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
+\EndIf
+\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
+
+\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
+\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
+\State \Return{ \tt{result}}
+
+\EndFunction{Refresh}
+
+\Statex
+
+\Function{void}{Advance}{\sl{int} h, \sl{int} hp}
+
+\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
+\State \tt{head.CAS(h, h+1)} \label{incrementHead}
+\EndFunction{Advance}
+
+\pagebreak
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
+\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
+
+\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
+Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
+%\State \Return \tt{result block's index}
+\EndFunction{BSearch}
+
+\Statex
+
+\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
+\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
+\State \tt{block newBlock= \Call{new}{\sl{block}}}
+\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
+\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
+\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
+\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
+\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
+\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
+\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
+\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
+\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
+\EndFor
+\If{\tt{this} \bf{is} \tt{root}}
+\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
+\EndIf
+
+\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
+\EndFunction{CreateBlock}
+
+\end{multicols}
+\end{algorithmic}
+\end{algorithm}
+
 %##########################################
@@
 
-%##########################################
-
-%\begin{algorithm}
-%\caption{Root}
-%\begin{algorithmic}[1]
-%\setcounter{ALG@line}{500}
-%
-%\Function{Block}{FindMostRecentDone}{}
-%\For{\tt{leaf l} \bf{in leaves}}
-%\State\tt{max= Max(l.maxOld, max)}
-%\EndFor
-%\State\Return\tt{max} \cmt{This snapshot suffies.}
-%\EndFunction{FindMostRecentDone}
-%
-%\end{algorithmic}
-%\end{algorithm}
-%##########################################
-
-\begin{algorithm}
-\caption{Leaf}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{600}
-
-\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
-\State \tt{blk.group= \head} \label{appendStart} 
-\State \tt{blocks[\head]= blk} 
-\State \tt{\head+=1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-%\Statex
-%
-%\Function{void}{Help}{}\cmt{Helps pending operations}
-%
-%\State{\tt{last= l.\head-1}}
-%\cmt{\tt{l.blocks[last]} can not be \tt{null} because \head increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
-%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
-%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
-%\EndIf
-%
-%\EndFunction{Help}
-
-\end{algorithmic}
-\end{algorithm}
-
-%##########################################
-
-%\begin{algorithm}
-%\caption{Yet to decide how to handle.}
-%\begin{algorithmic}[1]
-%\setcounter{ALG@line}{800}
-%
-%%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
-%%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
-%%\State \tt{t1,t2= RBT.split(order, s)}
-%%\State \tt{RBTRoot.CAS(t2.root)}
-%%\EndFunction{CollectGarbage}
-%
-%\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
-%
-%\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
-%
-%\State \Return \tt{this.leaf.response(block.order)}
-%\EndIf
-%
-%\EndFunction{FallBack}
-%
-%\end{algorithmic}
-%\end{algorithm}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                     %
-                     %
-                     %
-                     %
-                     %
-                     %
-                     %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \clearpage
@@
 
-To prove a shared data structure works correctly it is sufficient to show it is linearizable.
-In our case which we create the linearization ordering in the root, we need to prove the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and if we do the linearization ordering created on a sequential queue we get the same results as our queue.
+To prove a shared data structure works correctly, it is sufficient to show it is linearizable.
+In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization operations get the same results as in our queue.
 The proof is structured like this.
-First we define and prove some facts about blocks and node's head properties.
-Then we introduce  the linearization ordering formally.
-Then we prove double \nf{Refresh} on a node is enough to propagate children's new operations up to the node.
-After this we prove some claims about the size of each block.
-Having propagates working  in a wait-free manner then we prove the correctness of \nf{DSearch()} and \nf{GetENQ()} and \nf{IndexDEQ}.
-Having these two main methods we prove the way we compute the response of a dequeue is correct and finally we prove the linearizability.
-
-%\framebox[1.1\width]{TEST} Fix the logical order of definitions (cyclic refrences).
-%
-%%\framebox[1.1\width]{Questions} When I write the lemmas since every claim in my mind is correct maybe I miss some fact that need proof or maybe I refer to some lemmas that are generally correct but not needed for the linearizability proof.
-Is lemma 7 necessary?
-Is lemma 13 induced trivially from lemma 8?
-%
-%\framebox[1.1\width]{TEST} Is it better to show \nf{ops(EST\sub{n, t})} with \nf{EST\sub{n, t}}?
-%
-%%\framebox[1.1\width]{TEST} How to merge notions of blocks and operations?
-block b $\sqsubseteq$ block c means b is subblock of c.
-block b $\in$ set B means b is in B.
-Merge these two to have shorter formulaes.
-%
-%\framebox[1.1\width]{Question} A good notation for \it{the index of the \nf{b}}?
-%
-%\framebox[1.1\width]{Question} How to remove the notion of time?
-To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t.
-Is it good?
-Furthermore, can we remove the notion of established blocks?
+First, we define and prove some facts about blocks and the node's \nf{head} field.
+Then, we introduce  the linearization ordering formally.
+Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
+After, this we prove some claims about the size and operation of each block which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}.
+Finally (2) is followed by proving the correctness of the way we compute the response of a dequeue.
 
@@
 \begin{definition}[Ordering of a block in a node]
-Let $b$ be \nf{n.blocks[$i$]} and $b^\prime$ be \nf{n.blocks[$j$]}.
+Let $b$ be \nf{$n$.blocks[$i$]} and $b^\prime$ be \nf{$n$.blocks[$j$]}.
 We call $i$ the \emph{index} of block $b$.
 Block $b$ is \it{before} block $b^\prime$ in node \nf{n} if and only if $i<j$.
-We define \it{the prefix for} $b$, to be the blocks in the \tt{n.blocks[0..$i$]}.
+We define \it{the prefix for} $b$ to be the blocks in \tt{n.blocks[0..$i$]}.
 \end{definition}
 
-Next we show that \nf{head} in a node can only be increased and when a \nf{Refresh} terminates \nf{head} is incremented by the process itself or maybe helped by another process.
+Next, we show that the value of \nf{head} in a node can be increased.
+After the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
 \begin{observation} \label{nonDecreasingHead}
-  \nf{head} is non-decreasing over time.
+For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
 \end{observation}
@@
 \end{proof}
-\begin{lemma}[head Increment] \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
+
+\begin{lemma} \label{lem::headInc}
+Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates, \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
 \end{lemma}
@@
 If the \nf{CAS} in Line \ref{incrementHead} is successful then the claim holds.
-Otherwise it has changed from the value that was read in Line \ref{readHead}, which by Observation \ref{nonDecreasingHead} it means another process has incremented the head.
+Otherwise $n.\nf{head}$ has changed from the value that was read in Line \ref{readHead}.
+By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
 \end{proof}
@@
 
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is $h$ then $\nf{n.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{n.blocks[i]}\neq\nf{null}$ for ${i<h}$.
+\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
 \end{invariant}
 \begin{proof}
-The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[}$x$\nf{]} is null for every $x>0$.
+Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}\neq\nf{null}$ for every $x>0$.
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
@@
 
-In the algorithm, \nf{n.blocks} is modified only on Line \ref{cas} which updates \nf{n.blocks[$h$]} for $h$ read from \nf{n.head} in Line \ref{readHead}.Note that since the \nf{CAS} in Line \ref{cas} is successful it means \nf{n.head} has not changed from $h$ till doing the \nf{CAS}.
-Writing into \nf{n.blocks[n.head]} preserves the invariant, since the claim does not talk about the content of \nf{n.blocks[n.head]}.
+In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
+Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}, because if so then $n.\nf{blocks[$h$].CAS}$ could not be successful.
+Writing into \nf{$n$.blocks[$n$.head$=h$]} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
 
@@
 If \nf{n.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-When a process does Line \ref{incrementHead} it means Line \ref{cas} was finished which either case it is successful or not it means \nf{n.blocks[$h$]} is not \nf{null} anymore.
-%At the end of every \nf{n.Refresh()} with a block  with \head greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}).
-If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
-\end{proof}
-
-The next lemma is used to prove the subblocks of two blocks in a node are disjoint. (I can't say its use without mentioning subblocks)
-\begin{lemma}[headProgress] \label{lem::headProgress}
- If $\nf{n.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-
+\nf{Advance()} is called  in Lines \ref{advance} and \ref{helpAdvance}.
+In case \nf{advance}, Line \ref{cas} was finished  before doing \ref{incrementHead} wether \ref{cas} is successful or not $n.\nf{blocks[h]}\neq\nf{null}$.
+In the other case also $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
+\end{proof}
+
+We define the subblocks of a block recursively.
+
+\begin{definition}[Subblock]\label{def::subblock}
+A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]}.
+Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
+We say block $b$ is \it{propagated} to node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n\nf{.blocks}$.
+\end{definition}
+
+The next lemma is used to prove the subblocks of two blocks in a node are disjoint.
+\begin{lemma} \label{lem::headProgress}
+ If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
+\end{lemma}
+\begin{proof}
 Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}.
-$b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
+Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
 Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}.
 Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
-The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was minus one of the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
-Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was minus one of the value read from $\nf{n.left.head}-1$ during the call to \nf{CreateBlock($i-1$)}.
-As by Observation \ref{nonDecreasingHead} \nf{n.left.head} is non-decreasing, $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$.
+The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
+Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}.
+By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$.
 The proof for \nf{end\sub{right}} is similar.
@@
 
-We define the subblocks of a block recursively.
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]} .
-Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
-We say block $b$ is \it{propagated to node} node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n.blocks$.
+\begin{lemma} \label{lem::subblocksDistinct}
+Subblocks of any two blocks in node $n$ do not overlap.  
+\end{lemma}
+\begin{proof}
+We are going to prove by contradiction.
+Consider the lowest \nf{node} $n$ in the tree that violates the claim, then subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
+Since $n$ is the lowest node in the tree violating the claim then direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
+Without loss of generality assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
+By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the ranges $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ and $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$j$].end\sub{left}}]$ cannot overlap.
+Therefore, direct subblocks of $n.blocks[i]$ and $n.blocks[j]$ cannot overlap.
+\end{proof}
+
+Now we can define the operations of a block using the definition of subblocks.
+\begin{definition}[Operations of a block]\label{def::ops}
+A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$ otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}.
+The set of operations of block \nf{b} is the union set of the operations in leaf subblocks of $b$.
+We denote the set of operations of block $b$ by ${ops(b)}$.
+We also say ${b}$ contains ${op}$ if $\nf{op}\in\nf{ops(b)}$.
 \end{definition}
 
-\begin{lemma} \label{lem::subblocksDistinct}
-Subblocks of any two blocks in node $n$ does not overlap.  
-\end{lemma}
-\begin{proof}
-We are going to prove by contradiction.
-Assume subblocks of $n.blocks[i]$ and $n.blocks[j]$ overlap which $i< j$.
-Without loss of generality assume left child subblocks of $n.blocks[i]$ overlap with the left child subblocks of $n.blocks[j]$.
-Direct left subblocks of $n.blocks[i]$ and $n.blocks[j]$ cannot overlap.
-Because of Lemma \ref{lem::headProgress} we have $\nf{n.blocks[$i-1$].end\sub{left}} \leq \nf{n.blocks[$i$].end\sub{left}} \leq \nf{n.blocks[$j-1$].end\sub{left}} \leq \nf{n.blocks[$j$].end\sub{left}}$.
-Since the tree height is finite we can go lower in the tree to reach a node that direct subblcoks of two different blocks in a node overlap.
-\end{proof}
-
-We define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $b$ in a leaf represents an \nf{Enqueue(e)} if $b.element=e\neq\nf{null}$ otherwise, if $b.element=\nf{null}$, $b$ represents a \nf{Dequeue()}.
-The set of operations of block \nf{b} is the union set of the operations in subblocks of \nf{b}.
-We denote the set of operations of block \nf{b} by \nf{ops(b)}.
-We also say \nf{b} contains \nf{op} if \nf{op}$\in$\nf{ops(b)}.
-\end{definition}
-
 \begin{definition}[Superblock]
-  Block \nf{b} is \it{superblock} of block \nf{c} if c is a direct subblock of b.
+  Block $b$ is \it{superblock} of block $c$ if $c$ is a direct subblock of $b$.
 \end{definition}
@@
 \begin{proof}
-A block having more than one superblock is contradictory with Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Operations are distinct \nf{Enqueue(e)} or \nf{Dequeue()}s invoked by processes. Next lemma proves that each operation appears once in the blocks of a node.
-\begin{lemma}[No Duplicates]\label{append}
-If $op$ is in $n.blocks[i]$ then there is no $j\neq i$ such that $op$ is in $ops(n.blocks[j])$.
-\end{lemma}
-\begin{proof}
-We prove this claim using Corollary \ref{lem::subblocksDistinct}.
+A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
+\end{proof}
+
+Operations are distinct \nf{Enqueue}s and \nf{Dequeue}s invoked by processes. Next lemma proves that each operation appears at most once in the blocks of a node.
+\begin{lemma}\label{lem::noDuplicates}
+If $op$ is in $n.blocks[i]$ then there is no $j\neq i$ such that $op$ is in $n.blocks[j]$.
+\end{lemma}
+\begin{proof}
+We prove this claim using Lemma \ref{lem::subblocksDistinct}.
 Assume $op$ is in the subblocks of both $n.blocks[i]$ and $n.blocks[j]$.
-From Corollary \ref{lem::subblocksDistinct} we know these blocks are different.
-So the leaf blocks containing $op$ are different, but since we assumed operations are distinctive it means one operation cannot be invoked two times.
+From Corollary \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
+Since each process puts each operation in only one block of its leaf then $op$ cannot be in two leaf blocks.
 This leads us to contradictory with the hypothesis.
@@
 \begin{definition}
-$n.blocks[i]$ is \emph{established} at time $t$ if $n.head>i$. \emph{EST\sub{n, t}} is the set of established blocks  of node \nf{n} at time $t$.
+$n\nf{.blocks[}i\nf{]}$ is \emph{established} at time $t$ if $n\nf{.head}>i$. \emph{EST\sub{n, t}} is the set of established blocks of node $n$ at time $t$.
 \end{definition}
@@
 Now we want to say block of a node grow over time.
-\begin{corollary}[Growing blocks]\label{lem::blocksOrder}
+\begin{observation}\label{lem::blocksOrder}
   If  time $t<$ time $t^\prime$, then $ops(n.blocks)$ at time $t$ $\subseteq$ $ops(n.blocks)$ at time $t^\prime$.
-\end{corollary}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n.blocks[n.head]$ and $n.head$ is non-decreasing, so the set of blocks of a node contains the the set of blocks before the \nf{CAS}.
-\end{proof}
-
-\begin{corollary}[establishedOrder]\label{lem::establishedOrder}
+\end{observation}
+\begin{proof}
+Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of blocks of a node after the \nf{CAS} contains the the set of blocks before the \nf{CAS}.
+\end{proof}
+
+\begin{corollary}\label{lem::establishedOrder}
   If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
 \end{corollary}
+\begin{proof}
+From Observations \ref{nonDecreasingHead},\ref{lem::blocksOrder}.  
+\end{proof}
 
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of operations in b: operations in leaves are ordered with numerical order in the drawing.}
-\end{figure}
-
-Now we define the ordering store in each node.
+  \caption{Order of operations in b. Operations in the leaves are ordered with numerical order shown in the drawing.}
+\end{figure}
+
+Now we define the ordering of operatiobs stored in each node.
 In the non-root nodes we only need ordering of operations of a type among themselves.
@@
 \begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n.blocks[b])$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf.blocks[b])$ or an empty sequence if $leaf.blocks[b].num_{enq}=~0$.
+  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
+$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{].num\sub{enq}}=0$.
 If $n$ is an internal node, then
 \begin{align*} 
-E(n,b) =&  E(n.left,n.blocks[b-1].end_{left}+1)\cdot E(n.left,n.blocks[b-1].end_{left}+2)\cdots E(n.left,n.blocks[b].end_{left})\cdot \\ 
-&E(n.right,n.blocks[b-1].end_{right}+1)\cdot E(n.right,n.blocks[b-1].end_{right}+2)\cdots E(n.right,n.blocks[b].end_{right})
+E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdot E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+2)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}})
 \end{align*}
@@
 \item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n.blocks[b])$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf.blocks[b])$ or an empty sequence if $leaf.blocks[b].num\sub{deq}=~0$.
+  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
+$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{].num\sub{deq}}=~0$.
 If $n$ is an internal node, then
 \begin{align*} 
-D(n,b) =&  D(n.left,n.blocks[b-1].end_{left}+1)\cdot D(n.left,n.blocks[b-1].end_{left}+2)\cdots D(n.left,n.blocks[b].end_{left})\cdot \\ 
-&D(n.right,n.blocks[b-1].end_{right}+1)\cdot D(n.right,n.blocks[b-1].end_{right}+2)\cdots D(n.right,n.blocks[b].end_{right})
+D(n,b) =&  D(n\nf{.left],n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot D(n\nf{.left}, n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdot D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+2)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}})
 \end{align*}
@@
 \begin{proof}
-%So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}.
 
 We show $ops(EST_{\nf{n.left}, ^Rt})= ops(\nf{n.left.blocks[0..$n.left.head_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(\nf{n.blocks[0..}n.head_{\ref{readParentHead}}])$.
-%Let $h$ be the value \nf{n.Refresh()} reads from \nf{n.head} at Line \ref{readHead} and $h_{left}$ be the value read from \nf{n.left.head} at Line \ref{lastLine}.
 As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}.
 By the Definition \ref{def::subblock} for the Subblock  the block \nf{new} contains \nf{n.left.blocks[n.blocks[$n.head_{\ref{readHead}}-1$].end\sub{left}+1..$n.left.head_{\ref{lastLine}}-1$]} (see Lines \ref{prevLine} and \ref{lastLine}).
-% \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..h\sub{left,read}])}.
+
 After the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[0..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[0..$n.head_{\ref{readHead}}$]}.
@@
 
-% From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$.
-Since the head in \nf{CreateBlock()} is read after $t_i$.
-So the new operations are in the block appended to $n$.
-
-%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$.
-By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}.
-Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10.
-%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readHead} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
 \end{proof}
@@
 
-%\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}
-
-%\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readHead} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}.
-A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
-%\end{lemma}
-%\begin{proof}
-%  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313.
-So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}.
-%%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
-%\end{proof}
-
-\pagebreak
-
+\pagebreak
+
+In the next lemma we show that if a process consecutively fails to do two instances of \nf{Refresh()} on node $n$, the established blocks in the children of $n$ before the first \nf{Refresh()} are going to be in $n$ after the second \nf{Refresh()}.
 \begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on internal node \nf{n} by a process $p$.
-If $R_1$ and $R_2$ both fail and return false, then we have \nf{ops(EST\sub{n.left, \sub{R1}t})} $\cup$ \nf{ops(EST\sub{n.right, \sub{R1}t})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{R2}})}.
-\end{lemma}
-\begin{proof}
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means a successful instance of \nf{Refresh()} performed its Line \ref{readHead} after $t_{\ref{readHead}}^{R_1}$ and finished its Line \ref{incrementHead1} or \ref{incrementHead2} before $t_{\ref{readHead}}^{R_2}$, from Lemma~\ref{lem::prectrueRefresh} by the end of this instance \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} has been propagated.
-
- Let $R_1$ read $i$ and $R_2$ read $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{TryAppend()} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{TryAppend()} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to append.
-Since $R_2^\prime$ creates the block after reading the value $i+1$ from \nf{n.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head} and the \nf{head}'s value is increasing by Lemma \ref{lem::headProgress} then $t_{R2^\prime \; \ref{readHead}} > t_{R1 \; \ref{readHead}} > _{R_1}t$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::prectrueRefresh} after $R_2^\prime$'s \nf{CAS} ($t_{\ref{cas}}^{R_2^\prime}$) we have \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(n.blocks)}.
-Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2^\prime$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by then ($\nf{n.head}\geq i+2 > i+1$).
-To summarize, $_{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} is before $R_2$'s termination.
-So, by Lemma \ref{lem::prectrueRefresh}, \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
-%It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
-If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
-If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}.
-Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$.
-So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$.
-This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
-See Figure~\ref{}.
+  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by a process $p$.
+If none of $R_1$ and $R_2$ is a succesful \nf{Refresh()}, then we have $ops(EST_{n.left, ^{R_1}t}) \cup ops(EST_{n.right, ^{R_1}t}) \subseteq ops(EST_{n, t^{R_2}})$.
+\end{lemma}
+\begin{proof}
+
+Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.   If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ is incremented  more than two times after $t_{\ref{readHead}}^{R_1}$.
+By Lemma \ref{lem::headPosition} when \nf{head} is incremented from $i$ a  successful \nf{CAS} in $n\nf{.blocks[i]}$ is done.
+Let $R_3$ be the \nf{Refresh()} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
+$R_3$ has read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block before $R_2$'s \nf{CAS} at Line \ref{cas}.
+So we have $^{R_1}t<t_{\ref{readHead}}^{R_m}<t_{\ref{cas}}^{R_m}<^{R_2}t_{\ref{readHead}}$.
+From \ref{lem::blocksOrder} and \ref{lem::trueRefresh} the claim holds.
+
+Now consider $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
+As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false} then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{CAS} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to \nf{CAS}.
+$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from \nf{n.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head}.
+By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
+By Lemma \ref{lem::prectrueRefresh} we have
+$ops(EST_{n.\nf{left}, t^{R_2^\prime}_{\ref{readHead}}}) \cup ops(EST_{n.\nf{right}, t^{R_2^\prime}_{\ref{readHead}}}) \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$
+.
+Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
+To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::blocksOrder} $ops(EST_{n.\nf{left}, t^{R_1}}) \cup ops(EST_{n.\nf{right}, t^{R_1}}) \subseteq ops(n.\nf{blocks}_{t^{R_2}})$
+
+Note that by Lemma \ref{lem::headInc} $R_1$ and $R_2$ both cannot read the same value $i$.
 \end{proof}
@@
 
-%  Block \nf{new} is created of new established subblocks of children of \nf{n}(Lemma \ref{lem::createBlock}, Line 46).
-If \nf{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \nf{n}.
-
-\begin{corollary}
+\begin{corollary} \label{lem::propagateStep}
   \nf{ops(EST\sub{n.left, \sub{\ref{firstRefresh}}t})} $\cup$ \nf{ops(EST\sub{n.right, \sub{\ref{firstRefresh}}t})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{\ref{secondRefresh}})}}
@@
 \begin{proof}
-  If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::trueRefresh} the claim holds.
-Also if first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::trueRefresh}.
-Finally, if both failed the claim is satisfied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-\pagebreak
-
-\begin{corollary}[Propagate Step] \label{doublyRefresh}
-All operations in \nf{n}'s children's established blocks before running line \ref{firstRefresh}  of a \nf{Propagate} routine are guaranteed to be in \nf{n}'s established blocks after line~\ref{secondRefresh}.
+  If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds.
+Else if first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
+Otherwise even if both failed, the claim is satisfied by Lemma \ref{doubleRefresh}.
+\end{proof}
+
+Now we say after \nf{Append($b$)} finishes, $b$ will be established in $root$.
+
+\begin{corollary}
+For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{root}, t^A})$.
 \end{corollary}
 \begin{proof}
-If \ref{firstRefresh} or \ref{secondRefresh} succeed, the claim is true by Lemma \ref{lem::trueRefresh}.
-Otherwise Lines \ref{firstRefresh} and \ref{secondRefresh} satisfy the preconditions of Lemma \ref{doubleRefresh}.
+$A$ adds $b$ to the assigned leaf of the process, establishes it and then calls \nf{Propagate()} on the parent of the leaf it appended $b$.
+For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} till $n$ is \nf{root}.
 \end{proof}
@@
 \begin{corollary}\label{lem::appendExactlyOnce}
-  After \nf{Append(blk)} finishes \nf{ops(blk)}$\subseteq$\nf{ops(root.blocks[x])} for exactly one \nf{x}.
+ After \nf{Append($b$)} finishes $b$ is in \nf{root.blocks[$x$]} for exactly one $x$.
 \end{corollary}
 \begin{proof}
-After \nf{Append(blk)}'s termination, \nf{blk} is in \nf{root.blocks} since \nf{blk} is established  in the leaf it has been added to.
-By applying Lemma \ref{doublyRefresh} inductively it is propagated up to the root.
-Finally Lemma \ref{append} shows  only one block in the root contains \nf{blk}.
-\end{proof}
+  Previous Corollary proves shows the claim for some $x$. By Lemma \ref{lem::appendExactlyOnce}
+\end{proof}
+
+\pagebreak
+Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnq()} and \nf{IndexDeq()}.
 
@@
 Without loss of generality $op_1$ is invoked earlier than $op_2$.
-A process cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$.
-By Corollary \ref{lem::appendExactlyOnce}, before appending $op_2$ to the tree $op_1$ exists in every node on the path from $p$'s leaf to the root, because $op_1$'s \nf{Append} is finished before $op_2$'s \nf{Append} starts.
-So, there is some block $b^\prime$  before $b$ in $n$ containing $op_1$.
-Existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{append}.
+Process $p$ cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$.
+By Corollary \ref{lem::appendExactlyOnce} before appending $op_2$, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$.
+Existence of $op_1$ in $b$ and $b^\prime$ contradicts with Lemma \ref{lem::noDuplicates}.
 \end{proof}
@@
 \end{proof}
-
-\pagebreak
-
-\begin{lemma}[Get correctness] \label{get}
-If $\nf{n.blocks[b].num\sub{enq}} \geq \nf{i}$ then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
-\end{lemma}
-\begin{proof}
-We are going to prove this lemma by induction on the height of node \nf{n}.
-For the base case, \nf{n} is a leaf.
-Leaf blocks each contain exactly one operation, so by the hypothesis, only \nf{n.GetENQ(b,1)} can be called and only when \nf{n.blocks[b]} contains an enqueue.
-At Line \ref{getBaseCase}, \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
-
-For the induction step we prove \nf{n.GetENQ(b, i)} returns $E_i(n,b)$, assuming \nf{n.child.GetENQ(subblock, i)} returns $E_i(n.child,b)$.
-We argue that  Line \ref{leftOrRight} correctly decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
-From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered from left to right and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
-Furthermore the \nf{num\sub{enq-left}} field  in \nf{n.blocks[b]} stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
-So the $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
-Otherwise we should search for the $i$th enqueue in the left child.
-By definition \ref{def::subblock} and \ref{def::ops} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-
-If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
-Also, we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise the enqueue we are looking for is in the right child.
-Then, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
-Other parameters for the left child are chosen similarly to the way they were chosen for the right child.
-
-So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
- Finally, \nf{n.child.GetENQ(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}.
-Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-%\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}
 
@@
 
-%\begin{figure}[hbt]  
-%  \center\includegraphics[width=3in]{pics/end-b.png}
-%
-%  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
-%  \label{end-b}
-%\end{figure}
-
-\pagebreak
-
-%\begin{definition}
-%An enqueue operation is \textit{finished} if its argument is returned by some process.
-A dequeue operation is \texttt{finished} if it returns \nf{null} or some value.
-Block \nf{b} is \textit{done} if all operations in \nf{ops(b)} are finished.
-%\end{definition}
-%\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
-%
-%\begin{lemma}[help]\label{help}
-%After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
-%\end{lemma}
-%
-%\begin{lemma} Let $n.propagates$ be the number of groups of blocks that have been propagated from node \nf{n} to its parent (successful \nf{n.parent.Refresh()}).
-%We have \nf{num\sub{propagated}}$\leq n.propagates\leq$\nf{num\sub{propagated}+p}. \nf{p} is the number of processes.
-%\end{lemma}
-%\begin{proof}
-%    \tt{num\sub{propagated}} is incremented after propagating (Line \ref{incNP}). Since maybe some process falls sleep before incrementing \tt{num\sub{propagated}} it may be behind by \nf{p}.
-%\end{proof}
-%
-%
-%\begin{lemma} \nf{super[]} preserves order from child to parent; i.e. if in node \nf{n} block \nf{b} is before \nf{c} then \nf{b.group} $\leq$ \nf{c.group}
-%\end{lemma}
-%\begin{proof} Line \ref{setGroup}. Since  \nf{num\sub{propagated}} is increasing.\end{proof}
-%
-%\begin{lemma} Let \nf{b, c} be in node \nf{n}, if \nf{b.group} $\leq$ \nf{c.group} then \nf{super[b.group]} $\leq$ \nf{super[c.group]}\end{lemma}
-%\begin{proof} Line \ref{setSuper}.\end{proof}
-
-%\begin{lemma}
-%The number of the \nf{block}s with \nf{super=i} in a node is $\leq p$.  
-%\end{lemma}
-%\begin{proof}For the sake of simplicity we assumed all the blocks are propagated from the left child.
-%\begin{figure}[hbt]
-%  \center\includegraphics[width=5in]{pics/index1}
-%\end{figure}  
-%\end{proof}
-
-% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-%\begin{lemma} \nf{super[i+1]-super[i]}$\leq p$\end{lemma}
-%\begin{proof}
-% In a Refresh with successful CAS in line 46, \nf{super} and \nf{counter} are set for each child in lines 48,49.
-Assume the current value of the counter in node \nf{n} is \nf{i+1} and still \nf{super[i+1]} is not set.
-If an instance of successful \nf{Refresh(n)} finishes \nf{super[i+1]} is set a new value and a block is added after \nf{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \nf{Refresh()} that have not reached line 49.
-So the distance between \nf{super[i+1]} and \nf{super[i]} is less than $p$.
-% \begin{figure}[hbt]
-%  \center\includegraphics[width=4in]{pics/index2}
-%\end{figure}  
-%\end{proof}
-%\begin{lemma}[super property]\label{superCounter}
-%If \nf{super[i] $\neq$ null} in node \nf{n}, then \nf{super[i]} is the index of the superblock of a block with \nf{time=i} in \nf{n.parent.blocks}.
-%\end{lemma}
-
-%\begin{proof} \nf{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
-\nf{super(b)} is the real superblock of b.
-\nf{super(t]} is the index of the superblock of the last block with time \nf{t}.
-If \nf{b.time} is \nf{t} we have:
-%$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-%\end{proof}
-
-%\begin{lemma}
-%Search in each level of \nf{IndexDeq()} takes $O(\log p)$ steps.  
-%\end{lemma}
-%\begin{proof}
-%Show preconditions are satisfied and the range is $p$.  
-%\end{proof}
+\pagebreak
+
+\begin{lemma}[Get correctness] \label{get}
+If $\nf{n.blocks[b].num\sub{enq}} \geq \nf{i}$ then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
+\end{lemma}
+\begin{proof}
+We are going to prove this lemma by induction on the height of node \nf{n}.
+For the base case, \nf{n} is a leaf.
+Leaf blocks each contain exactly one operation, so by the hypothesis, only \nf{n.GetENQ(b,1)} can be called and only when \nf{n.blocks[b]} contains an enqueue.
+At Line \ref{getBaseCase}, \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
+
+For the induction step we prove \nf{n.GetENQ(b, i)} returns $E_i(n,b)$, assuming \nf{n.child.GetENQ(subblock, i)} returns $E_i(n.child,b)$.
+We argue that  Line \ref{leftOrRight} correctly decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
+From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered from left to right and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
+Furthermore the \nf{num\sub{enq-left}} field  in \nf{n.blocks[b]} stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+So the $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
+Otherwise we should search for the $i$th enqueue in the left child.
+By definition \ref{def::subblock} and \ref{def::ops} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+
+If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
+Also, we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
+We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
+
+Otherwise the enqueue we are looking for is in the right child.
+Then, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
+Other parameters for the left child are chosen similarly to the way they were chosen for the right child.
+
+So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
+ Finally, \nf{n.child.GetENQ(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
+\end{proof}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
+
+  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}.
+Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
+  }\label{figGet}
+\end{figure}
+
+\pagebreak
 
@@
 In this case the next \nf{Refresh}es on \nf{n} and \nf{n.parent} help fill in the value of \nf{b.super}.
-
-%    By Invariant \ref{lem::headPosition}, \nf{n.parent.blocks[hp]} may be \nf{null} or not \nf{null}, if \nf{hp} is the value read from \nf{n.parent.head} (Line \ref{}). 
-%First Lines of the \nf{Refresh} routine(Lines \ref to \ref{}) helps the left child and the right child to fill their \nf{super} field.
-A process may go to sleep, after Line \ref{cas} of its \nf{Refresh} and before Line \ref{setSuper}.
-So when a block is going to be appended to \nf{n.parent.blocks} we help the children so the super field in the last block of the children does not differ very much form its real value.
-After this step we are assured for all the blocks in the children that their index is less than the child's \nf{head}, the \nf{super} field is set.
-Since we first help to fill the \nf{super} and then increment the \nf{head} (Lines \ref{}).
-From Lemma \ref{lem::trueRefresh} we knwo that blocks whose index are less than \nf{head} are propagated to the parent in a \nf{Refresh} instance.
-To summarize Line \ref{endHelpChild1} of \nf{n.Refresh()}, \nf{super} field of all the established blocks in \nf{n.left,n.right} is set and $\neq \nf{null}$.
-
-%
-%\begin{corollary}
-%  Since Line \ref{incrementHead1} is executed after the Lines \ref{setSuper} and \ref{endHelpChild1}, then when \nf{n.head} is increased from \nf{h} to \nf{h+1}, we know that \nf{n.blocks[h]} is set.
-It might be two cases that the process which did \nf{TryAppend(new,h)} successfully, has incremented \nf{h} to \nf{h+1} or it might went to sleep and another process  which failed to do \nf{TryAppend(new,h)} comes and helps to set \nf{n.blocks[h].super} and increments \nf{n.head}.
-Or maybe a process that is trying to add a block to \nf{n.parent} after the \nf{TryAppend(new,h)} helps to set the \nf{super} field of \nf{n.blocks[h]}.
-%\end{corollary}
-
-%\begin{lemma}
-%  After a successful \nf{TryAppend(new,h)}, \nf{n.blocks[h].super} is set before the next succseful \nf{n.Refresh} or the second next succesful \nf{n.parent.Refresh()}.
-Superblock of \nf{b} is in \nf{n.parent.blocks[b.super-1..b.super+2]}.
-%\end{lemma}
-%\begin{proof}
-%\nf{n.TryAppend(h+1)}cannot happen unless \nf{n.head} is incremented from \nf{h} to \nf{h+1}. The latter case is explained in the next page.  
-%\end{proof}
-
-%
-%\begin{lemma}\label{superRange}  
-%\end{lemma}
-%\begin{proof}
-%Let us explain the possible cases of setting \nf{super} of block \nf{b} in node \nf{n}.
-\nf{b.super} is set by an instance of \nf{n.Refresh} or \nf{n.parent.Refresh}.
-Since Lines \ref{} setting the \nf{super} is done before incrementing the \nf{head}, so if we know \nf{head} has advanced the super has to be set before.
-%
-%First case is that the \nf{super} field of \nf{b} is set by the same process that appended \nf{b}.
-There are two possibilities that \nf{n.parent[hp]} is already \nf{null} or not, when the value \nf{hp} is read from \nf{n.parent.head}.
-We argue how distant is the superblock of \nf{b} from \nf{n.parent.blocks[hp=b.super]}.
-If the successful \nf{Refresh} that does \nf{n.parent.TryAppend(new,hp+1)}, creates its \nf{new} block after \nf{n.head.CAS(h,h+1)} (Line \ref{}) then \nf{new} contains the not yet propagated established blocks from \nf{n}, including \nf{b} (blocks in \nf{n} with smaller index than \nf{h+1}).
-Otherwise, the next \nf{Refresh} on \nf{n.parent} which has read \nf{head=hp+2} contains \nf{b}.
-Because $\nf{t\sub{n.blocks.TryAppend(b,h)}} < \nf{t\sub{n.head.CAS(h,h+1)}} < \nf{t\sub{hp+2=read n.parent.head}} < \nf{t\sub{n.parent.head.CAS{hp+1,hp+2}}}$.
-See Lemma \ref{lem::trueRefresh}.
-In the case that \nf{n.parent.blocks[hp]!=null}, when a block is going to be appended to the the \nf{hp+1} index of the parent it is going to contain the \nf{b}.
-One may wonder why, becuase it has read the parent's \nf{head} value \nf{hp+2} which is greater than \nf{hp+1} and it has to be incremented after \nf{n.bocks.TryAppend(new,h)}.
-The other case \nf{n.parent[hp]==null} is the same but one bit closer.
-%
-%The \nf{super} field of \nf{b} is set by some process doing a \nf{Refresh} on \nf{n} after appending \nf{b} to \nf{n.blocks}.
-This \nf{Refresh} increments \nf{n.head} from \nf{h} to \nf{h+1}, if it was not already incremented (Lines \ref{}).
-The next successful \nf{Refresh} on \nf{n.parent} which reads \nf{n.head} with some value greater than \nf{h} propagates \nf{b} to \nf{n.parent}.
-If \nf{n.parent.blocks[hp]} was not null then \nf{n.parent.blocks[hp+1]} or \nf{n.parent.blocks[hp+2]} contains \nf{b}, otherwise the \nf{super} field in \nf{b} is the exact index of the superblock of \nf{b} or \nf{n.parent.blocks[hp+1]} contains \nf{b}.
-The second next block appended to the parent is going to absolutely contain \nf{b} if the next block appending to \nf{n.parent} did not contain \nf{b} (Lemma \ref{trueRefresh}).
-One might wonder why the next block appended to the parent might not contain \nf{b} because it might have read the blocks it was propagating before appending \nf{b} to \nf{n.blocks}.
-%
-%The \nf{super} field is set by the process doing the \nf{Refresh} on the parent.
-In this case if the \nf{Refresh} is successful then position of the superblock of \nf{b} is exactly equal to the \nf{super} value.
-But if it was failure it might be one bit before the real value.
-%
-%\end{proof}
 

### 2022-09-21 thesis/thesis.tex
minor fix

@@
 
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node's blocklist implicitly represents a set of operations.
+A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
 \begin{definition}[Ordering of a block in a node]
@@
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-
+\end{definition}
 \begin{itemize}
@@
 \begin{align*} 
-D(n,b) =&  D(n\nf{.left],n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot D(n\nf{.left}, n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot D(n\nf{.left}, n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
 &D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdot D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+2)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}})
@@
 \end{itemize}
-\end{definition}
 
@@
 
-%\begin{lemma}[Search Ranges]\label{search}
-%  Preconditions of all invocation of \nf{BSearch} are satisfied.
-%\end{lemma}
-%\begin{proof}
-%  
-%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-%
-%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-%
-%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
-%
-%\end{proof}
 \begin{definition}

### 2022-09-22 thesis/thesis.tex
review page 23

@@
 \State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile
-\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchEnd}
+\EndWhile \label{dsearchEnd}
+\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchBSearch}
 \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
@@
 \State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
+\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
+\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
 \EndFor
@@
 
-\begin{lemma}[DSearch correctness] \label{dsearch}
-If $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$, \nf{DSearch(e, end)} returns \nf{<b, i>} such that $E_i(root,b)= E_e(root)$.
+In the next two lemmas we analyze and prove the correctness of the doubling search we use to find the block containing $i$th \nf{Enqueue} in the root.
+Correctness is pretty staightforward from the hypothesis.
+For the time analysis we use the semantics of queue operations.
+\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
+If $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
 \end{lemma}
@@
 \nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$.
-From Lemma we know \nf{sum\sub{enq}} fields of nf{root.blocks[]} are sorted in a non-decreasing order.
-Since \nf{root.blocks[0].sum\sub{enq}=0} and there is a \nf{block} in the root with \nf{sum\sub{enq}} value greater than \nf{e}, so there is a \nf{b} that $\nf{root.blocks[b].sum\sub{enq}} \geq e$ but $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
-This block contains $E_i(root,b)$ and  the search on Line \ref{dsearchStart}-\ref{dsearchEnd} will eventually reach the \nf{b}.
-
-\end{proof}
-\begin{lemma}[DSearch Analysis]\label{dsearchTime}
-  Assume $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then  \nf{DSearch(e, end)} takes $\Theta(\log$\nf{root.blocks[b].\size+ root.blocks[end].\size}$)$ steps.
-\end{lemma}
-\begin{proof}
-First we show $\nf{end} - \nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1 \big)$.
+From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know \nf{sum\sub{enq-left}}, \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
+Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, $\nf{sum\sub{enq}}$ field of \nf{root.block[$0\cdot\cdot end$]} is  also non-decreasing.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a \nf{b} such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
+Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
+The doubling search on Lines \ref{dsearchStart}-\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[start+1].sum\sub{enq}}$.
+In Line \ref{dsearchBSearch} Binary Search finds $b$ in the range mentioned.
+Finally $i$ is computed from the definition of \nf{sum\sub{enq}}.
+\end{proof}
+
+\begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
+If $E_i(root,b)=E_e(root)$\nf{.element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $\Theta(\log$ \nf{root.blocks[$b$].size+ root.blocks[$end$].size}$)$ steps.
+\end{lemma}
+\begin{proof}
+First we show $end - b \leq 2 \times \big(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size} \big)$.
 From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0.
 So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}.
-Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}.
-Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation.
-Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$.
-Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}., because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}.
-Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}.
-So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$.
-See Figure \ref{fig::doubling}.
-
-Now that we know there are at most \nf{root.blocks[b].\size+root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]} then with doubling search in $\Theta \big(\log($\nf{root.blocks[b].\size+root.blocks[end].\size}$)\big)$ steps we reach \nf{start=c} that the \nf{root.blocks[c].sum\sub{enq}} is less than \nf{e} and \nf{end-c} is not more than $2 \times 2 \times \big(\nf{root.blocks[b].\size+root.blocks[end].\size}\big)$.
-Beause otherwise, then \nf{(end-c)/2} satisfied the $\nf{root.blocks[(end-c)/2].sum\sub{enq}}<\nf{e}$.
-In line \ref{doubling} the difference between \nf{end} and \nf{start} is doubled.
-See Figure \ref{fig::doubling}.
-
- After computing \nf{b}, the value \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-So the whole \nf{DSearch} routine takes $\Theta\big(\log(\nf{root.blocks[b].\size+root.blocks[end].\size})\big)$ steps.
-
+Suppose there were more than \nf{root.blocks[b].size} \nf{Dequeue}s in \nf{root.blocks[$b+1 \cdots end-1$]}.
+Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue()}.
+Furthermore in the execution of queue operations with the linearization ordering the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[end]}.
+There can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].size}.
+Therefore there cannot be more than \nf{root.blocks[$b$].size + root.blocks[$end-1$].size} \nf{Enqueue}s in \nf{root.blocks[$b+1\cdots end-1$]}.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size+ root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since each block has at least one operation thus there are at most $2 \times$\nf{root.blocks[$b$].size+ root.blocks[$end$].size} blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+So $end-b\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
+So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $\Theta \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
+See Figure \ref{fig::doubling}.After Line \ref{dsearchEnd}, Binary Search that finds \nf{b} also takes $\Theta\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
+Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
+So the claim is proved.
 \end{proof}
@@
   \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between $b,c,end$}
+  \caption{Distance relations between \nf{start}$,b,end$. \nf{start} is shown with $c$.}
   \label{fig::doubling}

### 2022-09-22 thesis/thesis.tex
apply 20,21,22

@@
 \item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{blocks}} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
+\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
 \item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
@@
 \Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
-\State \tt{blk.group= \head} \label{appendStart} 
 \State \tt{blocks[\head]= blk} 
@@
 The set of operations of block \nf{b} is the union set of the operations in leaf subblocks of $b$.
-We denote the set of operations of block $b$ by ${ops(b)}$.
+We denote the set of operations of block $b$ by ${ops(b)}$ and the union set of operations of set of blocks $B$ by $ops(B)$.
 We also say ${b}$ contains ${op}$ if $\nf{op}\in\nf{ops(b)}$.
@@
 \begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} at time $t$ if $n\nf{.head}>i$. \emph{EST\sub{n, t}} is the set of established blocks of node $n$ at time $t$.
+$n\nf{.blocks[}i\nf{]}$ is \emph{established} at time $t$ if $n\nf{.head}>i$.
+An operation is \it{established} if it is in an established block.
+$EST^n_t$ is the set of established operations in node $n$ at time $t$.
 \end{definition}
@@
 \begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$, then $ops(n.blocks)$ at time $t$ $\subseteq$ $ops(n.blocks)$ at time $t^\prime$.
+  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
 \end{observation}
@@
 \begin{corollary}\label{lem::establishedOrder}
-  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
+  If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
 \end{corollary}
 \begin{proof}
-From Observations \ref{nonDecreasingHead},\ref{lem::blocksOrder}.  
+From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
 \end{proof}
@@
 
-Now we define the ordering of operatiobs stored in each node.
+Now we define the ordering of operations stored in each node.
 In the non-root nodes we only need ordering of operations of a type among themselves.
@@
 \begin{definition}
-Let $^{op}t$ be the time $op$ is invoked and $t^{op}$ be the time $op$ terminates.
-  $_{l}t$ is the immediate time before running Line $l$.
-$t_{l}$ is the immediate time after running Line $l$.
-$^{op}t_{l}$ is the immediate time before running Line $l$ of operation $op$.
-$t_{l}^{op}$ is the immediate time after running Line $l$ of operation $op$.
-$v_l$ is the value of \nf{v} immediately before line $l$ for the process we are talking about.
+Let $t^{op}$ be the time $op$ is invoked and $^{op}t$ be the time $op$ terminates.
+  Let $t_{l}$ be the time immediately before executing Line $l$ and $_{l}t$ be the time immediately after executing Line $l$.
+Let $^{op}_{l}t$ is the time immediately before running Line $l$ of operation $op$ and   $t_{l}^{op}$ is the immediate time after running Line $l$ of operation $op$.
+In the text $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
 \end{definition}
 
-\begin{definition}
+\begin{definition} [Successful Refresh]
   An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
@@
 
-In the next Lemmas we show for every \nf{Refresh()} that has has succesuful \nf{CAS} in Line \ref{cas}, all the established operations in the children before the \nf{Refresh} are going to be established in the parent after the \nf{Refresh}.
-
-\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-If $R$ is a succesful instance \nf{n.Refresh()}, then we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
- ops(n.blocks_{t_{\ref{cas}}^R})$.
-\end{lemma}
-\begin{proof}
-
-We show $ops(EST_{\nf{n.left}, ^Rt})= ops(\nf{n.left.blocks[0..$n.left.head_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(\nf{n.blocks[0..}n.head_{\ref{readParentHead}}])$.
-As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}.
+In the next two results we show for every successful \nf{Refresh()}, all the operations established  in the children before the \nf{Refresh} are established in the parent after the \nf{Refresh}'s succesful \nf{CAS} at Line \ref{cas}.
+
+\begin{lemma} \label{lem::trueRefresh}
+If $R$ is a successful instance of \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n.blocks_{\ref{cas}})$.
+\end{lemma}
+\begin{proof}
+
+We show $EST_{n\nf{.left}}^{t^R}= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]})$.
+%As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}.
 By the Definition \ref{def::subblock} for the Subblock  the block \nf{new} contains \nf{n.left.blocks[n.blocks[$n.head_{\ref{readHead}}-1$].end\sub{left}+1..$n.left.head_{\ref{lastLine}}-1$]} (see Lines \ref{prevLine} and \ref{lastLine}).
-
-After the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[0..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[0..$n.head_{\ref{readHead}}$]}.
-Because of the Lemma \ref{nonDecreasingHead} we have, $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{readParentHead}}$.
-From Lemma \ref{lem::establishedOrder} the inequality we proved is stronger than what we wanted to prove (TODO: make the lemma 13 claim more general, if time passes n.blocks grows).
+Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$.
+Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[1..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[1..$n.head_{\ref{readHead}}$]}.
+Because of Lemma \ref{nonDecreasingHead} we have $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{cas}}$.
+From Observation \ref{lem::blocksOrder} the claim follows.
 The proof for the right child is the same.
-
 \end{proof}
@@
 
-\begin{corollary}[Weak True Refresh] \label{lem::prectrueRefresh}
-If $R$ is a complete instance \nf{n.Refresh()}, then we have $ops(EST_{\nf{n.left}, ^Rt}) \; \cup \; ops(EST_{\nf{n.right}, ^Rt}) \subseteq 
- ops(EST_{\nf{n}, t^R})$.
+\begin{corollary}\label{lem::prectrueRefresh}
+If $R$ is a complete instance \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
 \end{corollary}
 \begin{proof}
-The left side is the same as the Lemma \ref{lem::trueRefresh}, so it is sufficient to show when \nf{R} terminates established blocks in \nf{n} are super set of \nf{n.blocks} after the Line \ref{cas}.
-Because of Lemma \ref{lem::headInc}  we are sure that the \nf{n.head} is incremented after line \ref{incrementHead}.
-So the \nf{new} block appended to \nf{n} is established at $t^R$.
-\end{proof}
-
-\pagebreak
-
-In the next lemma we show that if a process consecutively fails to do two instances of \nf{Refresh()} on node $n$, the established blocks in the children of $n$ before the first \nf{Refresh()} are going to be in $n$ after the second \nf{Refresh()}.
-\begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by a process $p$.
-If none of $R_1$ and $R_2$ is a succesful \nf{Refresh()}, then we have $ops(EST_{n.left, ^{R_1}t}) \cup ops(EST_{n.right, ^{R_1}t}) \subseteq ops(EST_{n, t^{R_2}})$.
-\end{lemma}
-\begin{proof}
-
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.   If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ is incremented  more than two times after $t_{\ref{readHead}}^{R_1}$.
-By Lemma \ref{lem::headPosition} when \nf{head} is incremented from $i$ a  successful \nf{CAS} in $n\nf{.blocks[i]}$ is done.
-Let $R_3$ be the \nf{Refresh()} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ has read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block before $R_2$'s \nf{CAS} at Line \ref{cas}.
-So we have $^{R_1}t<t_{\ref{readHead}}^{R_m}<t_{\ref{cas}}^{R_m}<^{R_2}t_{\ref{readHead}}$.
-From \ref{lem::blocksOrder} and \ref{lem::trueRefresh} the claim holds.
-
-Now consider $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false} then there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{CAS} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from \nf{n.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from \nf{n.head}.
+The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when \nf{R} terminates the established blocks in $n$ are a superset of $n$\nf{.blocks} after Line \ref{cas}.
+Because of Lemma \ref{lem::headInc}  we are sure that $n$\nf{.head} is incremented after line \ref{incrementHead}.
+So the block \nf{new} appended to \nf{n} at Line \ref{cas} is established at $^Rt$.
+\end{proof}
+
+\pagebreak
+
+In the next lemma we show that if two consecutive instances of \nf{Refresh()} by the same process on node $n$ fail, then the established blocks in the children of $n$ before the first \nf{Refresh()} are guaranteed to be in $n$ after the second \nf{Refresh()}.
+
+\begin{lemma} \label{doubleRefresh}
+  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by process $p$.
+If neither $R_1$ nor $R_2$ is a successful \nf{Refresh()}, then we have $EST_{n\nf{.left}}^{t^R{_1}} \; \cup \; EST_{n\nf{.right}}^{t^R{_1}} \subseteq EST_{n}^{^R{_2}t}$.
+\end{lemma}
+\begin{proof}
+Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
+Note that by Lemma \ref{lem::headInc} $R_1$ and $R_2$ both cannot read the same value $i$.
+By Observation \ref{nonDecreasingHead} $R_2$ reads larger value of $n$\nf{.head} than $R_1$.
+
+Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
+As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{CAS} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to \nf{CAS}.
+$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
 By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
 By Lemma \ref{lem::prectrueRefresh} we have
-$ops(EST_{n.\nf{left}, t^{R_2^\prime}_{\ref{readHead}}}) \cup ops(EST_{n.\nf{right}, t^{R_2^\prime}_{\ref{readHead}}}) \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$
-.
-Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head}  head is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::blocksOrder} $ops(EST_{n.\nf{left}, t^{R_1}}) \cup ops(EST_{n.\nf{right}, t^{R_1}}) \subseteq ops(n.\nf{blocks}_{t^{R_2}})$
-
-Note that by Lemma \ref{lem::headInc} $R_1$ and $R_2$ both cannot read the same value $i$.
+$EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
+Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
+To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation \ref{lem::blocksOrder} $ops(EST_{n.\nf{left}, t^{R_1}}) \cup ops(EST_{n.\nf{right}, t^{R_1}}) \subseteq ops(n.\nf{blocks}_{t^{R_2}})$
+
+  If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented  more than two times since $_{\ref{readHead}}^{R_1}t$.
+By Lemma \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[i+1]}$ is non-null.
+Let $R_3$ be the \nf{Refresh()} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
+$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
+So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^R_{2}t$.
+From Observation \ref{lem::blocksOrder} on the operations before and after $R_3$'s \nf{CAS} and Lemma \ref{lem::trueRefresh} on $R_3$ the claim holds.
 \end{proof}
@@
   If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds.
-Else if first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
-Otherwise even if both failed, the claim is satisfied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we say after \nf{Append($b$)} finishes, $b$ will be established in $root$.
+If the first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
+Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
+\end{proof}
+
+Now we show that after \nf{Append($b$)} on a leaf finishes, $b$ will be established in \nf{root}.
 
 \begin{corollary}
-For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{root}, t^A})$.
+For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{n}, t^A})$ where $n\in\{\text{nodes in the path from the leaf to the root}\}$.
 \end{corollary}
 \begin{proof}
-$A$ adds $b$ to the assigned leaf of the process, establishes it and then calls \nf{Propagate()} on the parent of the leaf it appended $b$.
-For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} till $n$ is \nf{root}.
-\end{proof}
+$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$.
+For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} utill $n$ is \nf{root}.
+\end{proof}
+
+\pagebreak
 
 \begin{corollary}\label{lem::appendExactlyOnce}
- After \nf{Append($b$)} finishes $b$ is in \nf{root.blocks[$x$]} for exactly one $x$.
+ After \nf{Append($b$)} finishes, $b$ is in the nodes the path from the leaf to the root for exactly one time.
 \end{corollary}
 \begin{proof}
-  Previous Corollary proves shows the claim for some $x$. By Lemma \ref{lem::appendExactlyOnce}
-\end{proof}
-
-\pagebreak
+By the previous corollary and Lemma \ref{lem::appendExactlyOnce} there is exactly one block in each node containing $b$.
+\end{proof}
+
 Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnq()} and \nf{IndexDeq()}.
 
-\begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block contains at most one operation of each processs.
+\begin{lemma}\label{blockSize}
+Each block contains at most one operation of each process.
 \end{lemma}
@@
 Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $p$ cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$.
-By Corollary \ref{lem::appendExactlyOnce} before appending $op_2$, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+Process $p$ cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$ begins.
+By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append()}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
 This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$.
-Existence of $op_1$ in $b$ and $b^\prime$ contradicts with Lemma \ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
+The Existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
+\end{proof}
+
+\begin{lemma}\label{subBlocksBound}
 Each block has at most $p$ direct subblocks.
@@
 \begin{proof}
-The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to the tree contains at least one operation, due to the test on Line~\ref{addOP}.
+The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an initial node contains at least one operation, due to the test on Line~\ref{addOP}.
 We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.

### 2022-09-23 thesis/thesis.tex
review page 24

@@
 We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
-To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.
+To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BinarySearch faster.
 
@@
 \paragraph{}
-Starting from the root, \textsc{GetIndex}($i$) BSearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
+Starting from the root, \textsc{GetIndex}($i$) BinarySearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
 From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
-So BSearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
+So BinarySearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
 
@@
 In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}.
-In order to make \texttt{BSearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
+In order to make \texttt{BinarySearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
 
@@
 \EndWhile \label{dsearchEnd}
-\State \tt{b= root.BSearch(sum\sub{enq}, e, start, end)} \label{dsearchBSearch}
+\State \tt{b= root.BinarySearch(sum\sub{enq}, e, start, end)} \label{dsearchBinarySearch}
 \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
-\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
+\Function{int}{BinarySearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
 
@@
 %\State \Return \tt{result block's index}
-\EndFunction{BSearch}
+\EndFunction{BinarySearch}
 
@@
 \ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
-\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
+\State \tt{subBlock= left.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
 \State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
+\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
 \State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
@@
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
-\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super-2, blocks[b].super+2}} \label{computeSuper}
+\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super-2, blocks[b].super+2}} \label{computeSuper}
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
@@
 The doubling search on Lines \ref{dsearchStart}-\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[start+1].sum\sub{enq}}$.
-In Line \ref{dsearchBSearch} Binary Search finds $b$ in the range mentioned.
+In Line \ref{dsearchBinarySearch} Binary Search finds $b$ in the range mentioned.
 Finally $i$ is computed from the definition of \nf{sum\sub{enq}}.
@@
 \begin{lemma}[Get correctness] \label{get}
-If $\nf{n.blocks[b].num\sub{enq}} \geq \nf{i}$ then \nf{n.GetENQ(b,i)} returns the \nf{element} enqueued by $E_i(n,b)$.
+If $\nf{$n$.blocks[$b$].num\sub{enq}} \geq i$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
 \end{lemma}
@@
 We are going to prove this lemma by induction on the height of node \nf{n}.
-For the base case, \nf{n} is a leaf.
-Leaf blocks each contain exactly one operation, so by the hypothesis, only \nf{n.GetENQ(b,1)} can be called and only when \nf{n.blocks[b]} contains an enqueue.
-At Line \ref{getBaseCase}, \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
-
-For the induction step we prove \nf{n.GetENQ(b, i)} returns $E_i(n,b)$, assuming \nf{n.child.GetENQ(subblock, i)} returns $E_i(n.child,b)$.
-We argue that  Line \ref{leftOrRight} correctly decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
-From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered from left to right and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}).
-Furthermore the \nf{num\sub{enq-left}} field  in \nf{n.blocks[b]} stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
-So the $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}.
-Otherwise we should search for the $i$th enqueue in the left child.
-By definition \ref{def::subblock} and \ref{def::ops} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-
-If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}.
-Also, we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
+For the base case, $n$ is a leaf.
+Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called when $n$ is a leaf.
+Line \ref{getBaseCase} of \nf{$n$.GetENQ($b$, $1$)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
+
+For the induction step we prove if \nf{$n$.child.GetENQ($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$.
+From Definition  \ref{ordering} of $E(n,b)$ operations from the left subblocks come before the operations from the right subblocks in a block (See Figure \ref{figGet}).
+\nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} stores the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}.
+Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
+By definition \ref{def::ops} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range \texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]} $\cup$ \texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
+
+ There are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
 We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_i(n,b)$.
+The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
 
 Otherwise the enqueue we are looking for is in the right child.
-Then, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of it in \nf{n.blocks[b]} but not in \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
+Because left \nf{Enqueue}s are before the right \nf{Enqueues}, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead from the left child.
 So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
-Other parameters for the left child are chosen similarly to the way they were chosen for the right child.
+Other parameters for the left child are chosen similar to the right child.
 
@@
 \end{proof}
-\textit{Make sure to show preconditions of all invocation of \nf{BSearch} are satisfied.}
+\textit{Make sure to show preconditions of all invocation of \nf{BinarySearch} are satisfied.}
 

### 2022-09-26 thesis/thesis.tex
apply dsearch comments

@@
 
-\pagebreak
-
-In the next two lemmas we analyze and prove the correctness of the doubling search we use to find the block containing $i$th \nf{Enqueue} in the root.
-Correctness is pretty staightforward from the hypothesis.
-For the time analysis we use the semantics of queue operations.
+\nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} so that $i$th \nf{Enqueue} in $b$th block of root is $e$th \nf{Enqueue} in entire sequence stored in the root. 
 \begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
-If $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
+If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
 \end{lemma}
@@
 From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know \nf{sum\sub{enq-left}}, \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, $\nf{sum\sub{enq}}$ field of \nf{root.block[$0\cdot\cdot end$]} is  also non-decreasing.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a \nf{b} such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
+Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, $\nf{sum\sub{enq}}$ values of \nf{root.block[$0\cdot\cdot end$]} is  also non-decreasing.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-The doubling search on Lines \ref{dsearchStart}-\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[start+1].sum\sub{enq}}$.
+The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
 In Line \ref{dsearchBinarySearch} Binary Search finds $b$ in the range mentioned.
-Finally $i$ is computed from the definition of \nf{sum\sub{enq}}.
-\end{proof}
-
-\begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
-If $E_i(root,b)=E_e(root)$\nf{.element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $\Theta(\log$ \nf{root.blocks[$b$].size+ root.blocks[$end$].size}$)$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b \leq 2 \times \big(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size} \big)$.
-From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0.
-So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}.
-Suppose there were more than \nf{root.blocks[b].size} \nf{Dequeue}s in \nf{root.blocks[$b+1 \cdots end-1$]}.
-Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue()}.
-Furthermore in the execution of queue operations with the linearization ordering the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[end]}.
-There can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].size}.
-Therefore there cannot be more than \nf{root.blocks[$b$].size + root.blocks[$end-1$].size} \nf{Enqueue}s in \nf{root.blocks[$b+1\cdots end-1$]}.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size+ root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since each block has at least one operation thus there are at most $2 \times$\nf{root.blocks[$b$].size+ root.blocks[$end$].size} blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So $end-b\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $\Theta \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.After Line \ref{dsearchEnd}, Binary Search that finds \nf{b} also takes $\Theta\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-So the claim is proved.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$. \nf{start} is shown with $c$.}
-  \label{fig::doubling}
-\end{figure}
+Finally $i$, is computed from the definition of \nf{sum\sub{enq}}.
+\end{proof}
 
@@
 
+\begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
+If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
+\end{lemma}
+\begin{proof}
+First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
+Suppose there were more than \nf{root.blocks[b].size} \nf{Dequeue}s in \nf{root.blocks[$b+1 \cdots end-1$]}.
+Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue()}.
+Furthermore in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
+There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
+The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
+After an execution on a queue the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s in \nf{root.blocks[$b+1\cdots end-1$]}.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end$]} and since from line \ref{addOP} we know that \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+So $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
+
+So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
+See Figure \ref{fig::doubling}.
+After Line \ref{dsearchEnd}, the binary search that finds \nf{b} also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
+Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
+So the claim is proved.
+\end{proof}
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/doubling.png}
+  \caption{Distance relations between \nf{start}$,b,end$. \nf{start} is shown with $c$.}
+  \label{fig::doubling}
+\end{figure}
+
 \pagebreak

### 2022-09-26 thesis/thesis.tex
fix Index main lemma

@@
 \State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
-\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super-2, blocks[b].super+2}} \label{computeSuper}
+\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}
 \Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
 
-\If{\tt{dir {\keywordfont is} left}}
+\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
 \State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-left}} \cmt{consider the enqueues in the previous blocks from the left child} \label{considerPreviousLeft}
@@
 
-\If{\tt{dir {\keywordfont is} right}}
+\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
 \State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
 \State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
-\EndIf
+\EndIf \label{computeISuperEnd}
 \State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
@@
 
-\begin{corollary}
-If \nf{b} has propagated to \nf{f}, then \nf{b.super} has at most 1 difference with the index of the superblock of \nf{b} in \nf{p}.
+\begin{corollary} \label{superRelation}
+If \nf{$n$.blocks[$b$]} is propagated to $n.parent$, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
 \end{corollary}
 
-\pagebreak
-\begin{lemma}[Computing SuperBlock]\label{superBlock}
-For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
-\end{lemma}
-\begin{proof}
-First we show the value read for \nf{super[b.group]} in line 418 is not null.
-  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
+%\pagebreak
+%\begin{lemma}[Computing SuperBlock]\label{superBlock}
+%For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
+%\end{lemma}
+%\begin{proof}
+%First we show the value read for \nf{super[b.group]} in line 418 is not null.
+%  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
 So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
 Then by Lemma \ref{superRange}if we search in the range $p$, we can find the superblock.
-
-\end{proof}
-
-\begin{lemma}[Index correctness]
- If $\nf{n.blocks[b].num\sub{deq}}\geq\nf{i}$ then \nf{n.IndexDEQ(b,i)} returns the rank in $D(root)$ of $D_{n,b,i}$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of \nf{n} from the \nf{root}. We can see the base case where \nf{n} is root is trivial (Line \ref{indexBaseCase}).
-  In the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $i$th Dequeue in the $b$th block of \nf{n} in \nf{n.parent} by Lemma \ref{superBlock} (Line \ref{computeSuper}).
-After that the order in $D(n.parent, superblock)$ is computed.
-Note that by Lemma \ref{blockSize} in each block there is at most one operation from each process and operations of one type are ordered based on the order in the subblocks (See Figure \ref{fig::orderFromSubblocks}).
-Finally \nf{index()} is called on \nf{n.parent} recursively and it returns the correct response from induction hypothesis.
-If the operation was propagated from the right child the number of dequeues from the left child are added to it (Line \ref{considerRight}), because the left child operations come before the right child operations (Definition \ref{ordering}).
-\end{proof}
-\textit{Make sure to show preconditions of all invocation of \nf{BinarySearch} are satisfied.}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in]{pics/orderFromSubblocks.png}
-  \caption{Relation of ordering of operations of a block from its subblocks}
-  \label{fig::orderFromSubblocks}
-\end{figure}
+%
+%\end{proof}
+Now using Corollary \ref{superRelation} on each step of the \nf{IndexDEQ} we prove its correctness.
+
+\begin{lemma}[\nf{IndexDEQ} correctness]
+ If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDEQ($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
+\end{lemma}
+\begin{proof}
+We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
+For the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+So preconditions of the invocation of \nf{BinarySearch} is satisfied.
+After that the order of $D_i(n,b)$ in $D(n.parent, superblock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
+Note that by Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
+Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
+Finally \nf{IndexDEQ} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
+\end{proof}
 

### 2022-09-27 thesis/thesis.tex
fix main theoreom

@@
 We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
-For the non-root nodes \nf{n.IndexDEQ(b,i)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+For the non-root nodes \nf{$n$.IndexDEQ($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
 So preconditions of the invocation of \nf{BinarySearch} is satisfied.
@@
 \pagebreak
-
-\begin{definition}
-   Assume the operations in $L$ are applied on an empty queue.
-If element of \nf{enqueue e} is the response to  \nf{dequeue d} then we say \nf{R(d)=e}.
-If \nf{d} 's response id \nf{null} (queue is empty) then \nf{R(d)=null}.
+We have explained the picture of proof in the first paragraph of proof section. Should I repeat here?
+
+\begin{lemma}[satisfiability] \label{linearSat}
+$L$ can be a linearization ordering.
+\end{lemma}
+\begin{proof}
+A linearization for an execution is legal if, $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
+If $op_{1}$ terminates before $op_{2}$ starts,$op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started.
+From Lemma \ref{append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
+By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
+\end{proof}
+Once some operations are aggregated in one block they will get propagated up to the root together and they can linearized in any order among themselves.
+The way we have chosen to order the operations in a block makes \nf{GetENQ}, \nf{FindResponse} efficent.
+
+\begin{definition}\label{defNullDeq}
+If a \nf{Dequeue} operation returns null it is called \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
 \end{definition}
 
-\begin{definition}
-In an execution on a queue, the dequeue operations that return some value are called \it{non-null dequeues}.  
-\end{definition}
 \begin{observation} \label{responseToADeq}
@@
 
+\begin{definition}
+   Assume the operations in \nf{root.blocks} are applied on an empty queue in order of $L$.
+If element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$, we say ${Resp(d)=e}$, otherwise if ${d}$ is null dequeue then ${Resp(d)=\nf{null}}$.
+\end{definition}
+
+\begin{observation}\label{enqBeforeDeq}
+By Definition \ref{ordering} enqueue operations come before dequeue operations in a block.  
+\end{observation}
+
+In the next Lemma we talk about what the \nf{size} field in \nf{root} blocks store.
 \begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[b].\size}is the \size of the queue if the operations in the prefix for the $b$th block in the root are applied with the order of $L$.  
-\end{lemma}
-\begin{proof}
- need to say?
-:: If the \size of a queue is greater than 0 then a \nf{Dequeue()} would decrease the \size of the queue, otherwise the \size of the queue remains 0.
-By definition \ref{ordering} enqueue operations come before dequeue operations in a block in $L$.
-
-We prove the claim by induction on \nf{b}.
-Base case \nf{b=0} is trivial since the queue is initialy empty and \nf{root.blocks[0].size=0}.
-For \nf{b=i} we are going to use the hypothesis for \nf{b=i-1}.
-If there are more than \nf{root.blocks[i-1].size+ root.blocks[i].sum\sub{enq}} dequeue operations in \nf{root.blocks[i]} then the queue would become empty after \nf{root.blocks[i]}.
-Otherwise we can compute the size of the queue after $b$th block using with this equality \nf{root.blocks[b].size= root.blocks[b-1].size+ root.blocks[b].sum\sub{enq}- root.blocks[b].sum\sub{deq}} (Line \ref{computeLength}).
-See Table \ref{qhistory} for an example of running some blocks of operations on an empty queue.
-\end{proof}
-
-\begin{lemma}[Duality of \#non-null dequeues and \nf{block.size}] \label{numberOfNND}
-If the operations are applied with the order of $L$, the number of non-null dequeues in the prefix for a block \nf{b} is \nf{b.sum\sub{enq}-b.\size}
-\end{lemma}
-\begin{proof}
-There are \nf{b.sum\sub{enq}} enqueue operations in the prefix for \nf{b}, then the size of the queue after the prefix for \nf{b} is \nf{\#enqs - \#non-null dequeues} in the prefix for \nf{b}, by Observation 35.
-So \nf{\#non-null dequeues} is \nf{b.sum\sub{enq}-b.\size}.
-The correctness of the \nf{block.\size}field is shown in Lemma \ref{sizeCorrectness}.
-\end{proof}
+  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in order of~$L$.  
+\end{lemma}
+\begin{proof}
+
+We prove the claim by induction on $b$.
+Base case ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$.
+We are going to show the correctness of case $b=i$ assuming the correctness of $b=i-1$.
+By \ref{enqBeforeDeq} If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].sum\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}, otherwise size of the queue after $b$th block in root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].sum\sub{enq}}- \nf{root.blocks[}b\nf{].sum\sub{deq}}$  which is same as the assignment on Line \ref{computeLength}.
+\end{proof}
+
+Next Lemma is about the relation between \#enqueues, \#non-null dequeues and size.
+\begin{lemma} \label{numberOfNND}
+If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].\size}
+\end{lemma}
+\begin{proof}
+There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$.
+Size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is  ${\#\it{enqueues }- \#\it{non-null dequeues}}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
+By correctness of \nf{size} field from Lemma \ref{sizeCorrectness} $\it{\#non-null dequeues}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+\end{proof}
+
+\begin{corollary}\label{numNullDeqBlock}
+  If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].\size}
+\end{corollary}
 
@@
 \end{lemma}
+\begin{proof}
+From \ref{numNullDeqBlock}, \ref{defNullDeq} and \ref{enqBeforeDeq}.
+\end{proof}
 
 \begin{lemma}[Computing Response] \label{computeHead}
-\nf{FindResponse(b,i)} returns \nf{R(D\sub{root,b,i}).element}.
-\end{lemma}
-\begin{proof}
-First note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_i(root, b)$ we talk about what if the whole operation in the linearization are applied on a queue.
-
-$D_{root,b,i}$ is $D_{root,root.blocks[b-1].sum_{deq}+i}$ from the definition \ref{ordering} and \it{$sum_{enq}$}.
-$D_{root,b,i}$ returns \nf{null} if \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0} by Lemma \ref{nullReturn} (Line \ref{checkEmpty}).
-Otherwise if it is $d^\prime$th non-null dequeue in $L$ it returns $d^\prime$th enqueue by Observation \ref{responseToADeq} .
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[b-1].sum\sub{enq} - root.blocks[b-1].\size} non-null dequeue operations before prefix for \nf{root.blocks[b-1]}.
-Note that the dequeues in \nf{root.blocks[b]} before the $i$th dequeue are non-null dequeues.
+\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$\nf{.element}.
+\end{lemma}
+\begin{proof}
+Note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_i(root, b)$ we talk about what if the whole operation in the linearization are applied on a queue.
+
+$D_i(root,b)$ is $D_{root.blocks[b-1].sum_{deq}+i}(root)$ from the Definitions \ref{ordering} and \it{$sum_{deq}$}.
+$D_i(root,b)$ returns \nf{null} if $\nf{root.blocks[}$b-1$\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ by Lemma \ref{nullReturn} (Line \ref{checkEmpty}).
+Otherwise if it is $k$th non-null dequeue in $L$ it returns $k$th enqueue by Observation \ref{responseToADeq}.
+By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null dequeue operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
+Note that the dequeues in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
 So the response is $E_{i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}}(root)$ (Line \ref{computeE}).
@@
 
-After computing \nf{e} we can find \nf{b,i} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then find its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $E_i(root,b)$.}
+After computing \nf{e}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ by \nf{DSearch} and then finds its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
+\end{proof}
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
 \label{computeResponseDetail}
@@
 
-\begin{table}[hbt]
-\centering
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
-    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
-        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
-            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
-                size & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-  \caption{An example of root blocks fields. Blocks are from left to right and operations in the blocks are also from the left to right.}\label{qhistory}
-\end{table}
-
-%\begin{proof}
-%  \nf{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \nf{head} read.
-If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \nf{head}.
-If a process dies before line 51, another process will increment \nf{head} in line 54.
-%\end{proof}
-
-\pagebreak
+\begin{lemma}[correctness]\label{linearCorrect}
+  If operations are applied as $L$ on a sequential queue, the sequence of the responses would be the same as our algorithm.
+\end{lemma}
+\begin{proof}
+Enqueue operations do not return any value so they are fine. Response of a dequeue in sequential execution of $L$ is same as our algorithm from \ref{computeHead}.  
+\end{proof}
 
@@
 \begin{proof}
-  We choose $L$ in Definition \ref{ordering} to be linearization ordering of operations and prove if we linearize operations as $L$ the queue works consistently.
-\end{proof}
-%\begin{lemma}
-%Operations in a block have a time point in common (There is a time $t$ all the operations are running).
-%\end{lemma}
-
-\begin{lemma}[satisfiability]
-$L$ can be a linearization ordering.
-\end{lemma}
-\begin{proof}
-To show this we need to say if in an execution, \nf{op\sub{1}} terminates before \nf{op\sub{2}} starts then \nf{op\sub{1}} is linearized before \nf{op\sub{2}}.
-If \nf{op\sub{1}} terminates before \nf{op\sub{2}} starts it means \nf{op\sub{1}.Append()} is terminated before \nf{op\sub{2}.Append()} starts.
-From Lemma \ref{append} \nf{op\sub{1}} is in \nf{root.blocks} before \nf{op\sub{2}} propagates so \nf{op\sub{1}} is linearized before \nf{op\sub{2}} by Definition \ref{ordering}.
-
-Once some operations are aggregated in one block they will be propagated together up to the root and we can linearize them in any order among themselves.
-Furthermore in L we arbitrary choose the order to be by process id, since it makes computations in the blocks faster .
-\end{proof}
-
-\begin{lemma}[correctness]
-  If operations are applied as $L$ on a sequential queue, the sequence of the responses would be the same as our algorithm.
-\end{lemma}
-
-\begin{proof}
-\it{Old parts to review}
-  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
-  \begin{enumerate}
-    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
-The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
-    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
-\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
-It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
-By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
-If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
-  \end{enumerate} 
+From \ref{linearSat}, \ref{linearCorrect}.
 \end{proof}

### 2022-09-27 thesis/thesis.tex
finalize proof till L

@@
 
-\usepackage[b4paper,left=0.8in,right=0.8in,top=0.8in,bottom=0.8in]{geometry}
+\usepackage[letterpaper]{geometry}
 \usepackage{tikz-qtree}
@@
 \pagebreak
+\newpage
+\newgeometry{left=0.6in,right=0.6in,top=0.6in,bottom=0.8in}
+
 \subsection{Pseudocode}
-
 \begin{algorithm}
@@
 \setcounter{ALG@line}{200}
-\begin{multicols}{2}
 
@@
 
-\pagebreak
-
-\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\Statex\cmt{Returns the the response to the $D_{root,b,i}$.}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \State \Return \tt{null} \label{checkEmpty}\cmt{Check if the queue is empty.}
+\Statex
+
+\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
+\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \cmt{Check if the queue is empty.}
+\State \Return \tt{null} \label{checkEmpty}
 \Else
 \State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
-\Statex\cmt{$E_e(root)$ is the response.}
+\cmt{$E_e(root)$ is the response.}
 \State \Return \tt{root.GetENQ(root.\Call{DSearch}{e, b})}\label{findAnswer}
@@
 \State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}
-\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
 
@@
 
+\restoregeometry
+\newpage
 \clearpage
@@
 
-To prove a shared data structure works correctly, it is sufficient to show it is linearizable.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal i.e for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization operations get the same results as in our queue.
+We adopt linearizability as our definition of correctness.
+In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization order, operations get the same results as in our queue.
 The proof is structured like this.
@@
 Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After, this we prove some claims about the size and operation of each block which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}.
-Finally (2) is followed by proving the correctness of the way we compute the response of a dequeue.
+After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}.
+Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
 
@@
 We call $i$ the \emph{index} of block $b$.
-Block $b$ is \it{before} block $b^\prime$ in node \nf{n} if and only if $i<j$.
-We define \it{the prefix for} $b$ to be the blocks in \tt{n.blocks[0..$i$]}.
+Block $b$ is \it{before} block $b^\prime$ in node $n$ if and only if $i<j$.
+We define \it{the prefix for} block $b$ in node $n$ to be the blocks in \tt{$n$.blocks[0..$i$]}.
 \end{definition}
 
-Next, we show that the value of \nf{head} in a node can be increased.
-After the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
+Next, we show that the value of \nf{head} in a node can only be increased.
+By the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
 \begin{observation} \label{nonDecreasingHead}
@@
 \begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()} .
+  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()}.
 \end{proof}
@@
 \begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on node \nf{n}. After $R$ terminates, \nf{n.head} is greater than the value read in line \ref{readHead} of $R$.
+Let $R$ be an instance of \nf{Refresh} on a node $n$. After $R$ terminates, \nf{$n$.head} is greater than the value read in line \ref{readHead} of $R$.
 \end{lemma}
@@
 
-Now we show \nf{n.blocks[n.head]} is the last  block written into \nf{n} or the first empty block in \nf{n}.
+Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in $n$.
 
@@
 \begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}\neq\nf{null}$ for every $x>0$.
+Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show the invariant still holds after these two changes.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}, because if so then $n.\nf{blocks[$h$].CAS}$ could not be successful.
+We show the invariant still holds after such changes.
+
+In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
+Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
 Writing into \nf{$n$.blocks[$n$.head$=h$]} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
 
-The value of \nf{n.head} is modified only in Line \ref{incrementHead}.
-If \nf{n.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance()} is called  in Lines \ref{advance} and \ref{helpAdvance}.
-In case \nf{advance}, Line \ref{cas} was finished  before doing \ref{incrementHead} wether \ref{cas} is successful or not $n.\nf{blocks[h]}\neq\nf{null}$.
-In the other case also $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-\end{proof}
-
+The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance()}.
+If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
+\nf{Advance()} is called  in Lines \ref{helpAdvance} and \ref{advance}.
+For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
+For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance} whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
+\end{proof}
+
+\pagebreak
 We define the subblocks of a block recursively.
@@
 \begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of $i$th block in node $n$ if it is in \texttt{n.left.blocks[n.blocks[$i-1$].end\textsubscript{left}+1..n.blocks[$i$].end\textsubscript{left}]} or in \texttt{n.right.blocks[n.blocks[$i-1$].end\textsubscript{right}+1..n.blocks[$i$].end\textsubscript{right}]}.
-Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
+A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$ Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
 We say block $b$ is \it{propagated} to node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n\nf{.blocks}$.
@@
 \begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{n.blocks[$i$].end\sub{left}} \geq \nf{n.blocks[$i-1$].end\sub{left}}$ and $\nf{n.blocks[$i$].end\sub{right}} \geq \nf{n.blocks[$i-1$].end\sub{right}}$.
+ If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
 \end{lemma}
@@
 \begin{proof}
-We are going to prove by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim, then subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim then direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
+We are going to prove the lemma by contradiction.
+Consider the lowest \nf{node} $n$ in the tree that violates the claim.
+Then subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
+Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
 Without loss of generality assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
@@
 \end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$ otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}.
-The set of operations of block \nf{b} is the union set of the operations in leaf subblocks of $b$.
-We denote the set of operations of block $b$ by ${ops(b)}$ and the union set of operations of set of blocks $B$ by $ops(B)$.
-We also say ${b}$ contains ${op}$ if $\nf{op}\in\nf{ops(b)}$.
-\end{definition}
 
@@
 
-Operations are distinct \nf{Enqueue}s and \nf{Dequeue}s invoked by processes. Next lemma proves that each operation appears at most once in the blocks of a node.
+Now we can define the operations of a block using the definition of subblocks.
+\begin{definition}[Operations of a block]\label{def::ops}
+A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$.
+Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}.
+The set of operations of block \nf{b} is the union of the operations in leaf subblocks of $b$.
+We denote the set of operations of block $b$ by ${ops(b)}$ and the union  of operations of a set of blocks $B$ by $ops(B)$.
+We also say ${b}$ contains ${op}$ if $op\in{ops(b)}$.
+\end{definition}
+
+Operations are distinct \nf{Enqueue}s and \nf{Dequeue}s invoked by processes. The next lemma proves that each operation appears at most once in the blocks of a node.
 \begin{lemma}\label{lem::noDuplicates}
@@
 Since each process puts each operation in only one block of its leaf then $op$ cannot be in two leaf blocks.
-This leads us to contradictory with the hypothesis.
+This is a contradiction.
 \end{proof}
@@
 $n\nf{.blocks[}i\nf{]}$ is \emph{established} at time $t$ if $n\nf{.head}>i$.
-An operation is \it{established} if it is in an established block.
+An operation is \it{established} in node $n$ if it is in an established block of $n$.
 $EST^n_t$ is the set of established operations in node $n$ at time $t$.
@@
 
-Now we want to say block of a node grow over time.
+Now we want to say the blocks of a node grow over time.
 \begin{observation}\label{lem::blocksOrder}
@@
 Now we define the ordering of operations stored in each node.
-In the non-root nodes we only need ordering of operations of a type among themselves.
+In the non-root nodes we only need to order operations of a type among themselves.
 Processes are numbered from 1 to $p$ and leaves of the tree are assigned from left to right.
@@
   \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{].num\sub{enq}}=0$.
+$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
 If $n$ is an internal node, then
 \begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdot E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+2)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}})
+E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
 \end{align*}
@@
   \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{].num\sub{deq}}=~0$.
+$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
 If $n$ is an internal node, then
 \begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdot D(n\nf{.left}, n\nf{.blocks[}b-1\nf{].end\sub{left}}+2)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdot D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+2)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}})
+D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
 \end{align*}
     \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$: $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
+\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
 \item $D_i(n)$ is the $i$th dequeue in $D(n)$.
@@
 \begin{definition}[Linearization] \label{def::lin}
- $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
+ $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
 \end{definition}

### 2022-09-28 thesis/thesis.tex
review linearizability proof

@@
 
-\usepackage[letterpaper]{geometry}
+\usepackage[letterpaper, margin=1in]{geometry}
 \usepackage{tikz-qtree}
@@
 \algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} \textproc{\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
+   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
 
 \algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{\textproc{#1}}}
-
-\algrenewcommand\Call[2]{\tt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+   [1]{\algorithmicend\ \tt{{#1}}}
+
+\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
 
@@
 \nf{enqueue}s and \nf{dequeue}s are appended to the tree and when we want to find the response to a \nf{dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null.
-If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetENQ(n,b,i)}.
-\texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetENQ(n,b,i)} finds its sub-block recursively to reach a leaf.
+If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetEnqueue(n,b,i)}.
+\texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetEnqueue(n,b,i)} finds its sub-block recursively to reach a leaf.
 \texttt{Index()} is similar but more complicated, finding super-blocks from a leaf to the root.
@@
 \State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDeq}{leaf.\head, 1}
+\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
 \State \tt{output=} \Call{FindResponse}{b, i} 
@@
 \Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{checkEmpty}
+\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
+\State \Return \tt{null} \label{returnNull}
 \Else
@@
 \cmt{$E_e(root)$ is the response.}
-\State \Return \tt{root.GetENQ(root.\Call{DSearch}{e, b})}\label{findAnswer}
+\State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
 \EndIf
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
+\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
 \If{\tt{this} \bf{is} \tt{leaf}}
@@
 \State \tt{subBlock= left.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
-\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
+\State \Return\tt{left.}\Call{GetEnqueue}{subBlock, i} 
 \Else
@@
 \State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
-\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
+\State \Return\tt{right.}\Call{GetEnqueue}{subBlock, i} 
 \EndIf
-\EndFunction{GetEnq}
+\EndFunction{GetEnqueue}
 
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
-\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 \EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
+\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlock, i}
 \EndIf
-\EndFunction{IndexDeq}
+\EndFunction{IndexDequeue}
 
@@
 Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}.
+After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
 Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
+
+\subsection{Basic Properties}
+In this subsection we talk about some properties of blocks and fields of the tree nodes.
 
@@
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show the invariant still holds after such changes.
+We show that if the invariant holds before such a change  then it  still holds after the change.
 
 In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$n$.head$=h$]} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
+Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation \ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
+Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
 
@@
 
-\pagebreak
 We define the subblocks of a block recursively.
@@
 \pagebreak
+
+\subsection{Ordering Operations}
 
@@
 
+\subsection{Propagating Operations to the Root}
+In this section we explain why two \nf{Refresh}es are enough to propagate a nodes operations to its parent.
+
 \begin{definition}
@@
 We show $EST_{n\nf{.left}}^{t^R}= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]})$.
-%As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}.
-By the Definition \ref{def::subblock} for the Subblock  the block \nf{new} contains \nf{n.left.blocks[n.blocks[$n.head_{\ref{readHead}}-1$].end\sub{left}+1..$n.left.head_{\ref{lastLine}}-1$]} (see Lines \ref{prevLine} and \ref{lastLine}).
+
 Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$.
@@
 
-\begin{corollary}
+\begin{corollary}\label{lem::append}
 For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{n}, t^A})$ where $n\in\{\text{nodes in the path from the leaf to the root}\}$.
@@
 
-Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnq()} and \nf{IndexDeq()}.
+\pagebreak
+
+\subsection{Correctness of GetEnqueue and IndexDequeue}
+
+Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()} and \nf{IndexDequeue()}.
 
@@
 
-\pagebreak
-
-\begin{lemma}[Get correctness] \label{get}
-If $\nf{$n$.blocks[$b$].num\sub{enq}} \geq i$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
+\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
+If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
 \end{lemma}
@@
 We are going to prove this lemma by induction on the height of node \nf{n}.
-For the base case, $n$ is a leaf.
-Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called when $n$ is a leaf.
-Line \ref{getBaseCase} of \nf{$n$.GetENQ($b$, $1$)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .
-
-For the induction step we prove if \nf{$n$.child.GetENQ($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$.
-From Definition  \ref{ordering} of $E(n,b)$ operations from the left subblocks come before the operations from the right subblocks in a block (See Figure \ref{figGet}).
-\nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} stores the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+For the base case, suppose $n$ is a leaf.
+Leaf blocks each contain exactly one operation, so only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf.
+Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$.
+
+For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.
+From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
+\nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
 So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}.
 Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
-By definition \ref{def::ops} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range \texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]} $\cup$ \texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
-
- There are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise the enqueue we are looking for is in the right child.
-Because left \nf{Enqueue}s are before the right \nf{Enqueues}, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead from the left child.
-So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}).
-Other parameters for the left child are chosen similar to the right child.
-
-So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
- Finally, \nf{n.child.GetENQ(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
+By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
+\begin{align*}
+&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
+&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
+ \end{align*}
+First we consider the case where the \nf{Enqueue} we are looking is in the left child.
+There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
+We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
+
+Otherwise, the enqueue we are looking for is in the right child.
+Because \nf{Enqueue}s from the left subblocks are ordered before the \nf{Enqueue}s from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
+So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ (Line \ref{rightChildGet}).
+Other parameters for the right child are chosen similarly to the left child.
+
+So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
+ Finally, \nf{n.child.GetEnqueue(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
 \end{proof}
@@
 
-%\pagebreak
-%\begin{lemma}[Computing SuperBlock]\label{superBlock}
-%For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
-%\end{lemma}
-%\begin{proof}
-%First we show the value read for \nf{super[b.group]} in line 418 is not null.
-%  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}.
-So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading.
-Then by Lemma \ref{superRange}if we search in the range $p$, we can find the superblock.
-%
-%\end{proof}
-Now using Corollary \ref{superRelation} on each step of the \nf{IndexDEQ} we prove its correctness.
-
-\begin{lemma}[\nf{IndexDEQ} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDEQ($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
+Now using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue} we prove its correctness.
+
+\begin{lemma}[\nf{IndexDequeue} correctness]
+ If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
 \end{lemma}
@@
 We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDEQ($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
 So preconditions of the invocation of \nf{BinarySearch} is satisfied.
@@
 Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
-Finally \nf{IndexDEQ} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
-\end{proof}
-
-\pagebreak
-We have explained the picture of proof in the first paragraph of proof section. Should I repeat here?
-
-\begin{lemma}[satisfiability] \label{linearSat}
-$L$ can be a linearization ordering.
+Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
+\end{proof}
+
+\pagebreak
+\subsection{Linearizability}
+We now prove the two properties needed for linearizability.
+
+\begin{lemma} \label{linearSat}
+$L$ is a legal linearization ordering.
 \end{lemma}
@@
 A linearization for an execution is legal if, $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
-If $op_{1}$ terminates before $op_{2}$ starts,$op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started.
-From Lemma \ref{append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
+If $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started.
+From Lemma \ref{lem::append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
 By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
 \end{proof}
-Once some operations are aggregated in one block they will get propagated up to the root together and they can linearized in any order among themselves.
-The way we have chosen to order the operations in a block makes \nf{GetENQ}, \nf{FindResponse} efficent.
+Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves.
+The way we have chosen to order the operations in a block makes \nf{GetEnqueue} and \nf{FindResponse} efficient.
 
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns null it is called \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
+If a \nf{Dequeue} operation returns null it is called a \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
 \end{definition}
 
-\begin{observation} \label{responseToADeq}
-In a sequential execution on a queue, $k$th non-null dequeue returns the \nf{element} of $k$th enqueue.
-\end{observation}
-
 \begin{definition}
-   Assume the operations in \nf{root.blocks} are applied on an empty queue in order of $L$.
-If element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$, we say ${Resp(d)=e}$, otherwise if ${d}$ is null dequeue then ${Resp(d)=\nf{null}}$.
+   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
+${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
+Otherwise if ${d}$ is a null dequeue then ${Resp(d)=\nf{null}}$.
 \end{definition}
@@
 
-In the next Lemma we talk about what the \nf{size} field in \nf{root} blocks store.
+In the next lemma we show that the \nf{size} field in each \nf{root block} is computed correctly.
 \begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in order of~$L$.  
-\end{lemma}
-\begin{proof}
-
+  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
+\end{lemma}
+\begin{proof}
 We prove the claim by induction on $b$.
-Base case ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$.
-We are going to show the correctness of case $b=i$ assuming the correctness of $b=i-1$.
-By \ref{enqBeforeDeq} If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].sum\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}, otherwise size of the queue after $b$th block in root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].sum\sub{enq}}- \nf{root.blocks[}b\nf{].sum\sub{deq}}$  which is same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-Next Lemma is about the relation between \#enqueues, \#non-null dequeues and size.
+The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$.
+We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
+By Observation \ref{enqBeforeDeq} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].sum\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
+Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].sum\sub{enq}}- \nf{root.blocks[}b\nf{].sum\sub{deq}}$.
+In both cases, this is same as the assignment on Line \ref{computeLength}.
+\end{proof}
+
+The next lemma is useful to compute the number of non-null dequeues.
 \begin{lemma} \label{numberOfNND}
-If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].\size}
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$.
-Size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is  ${\#\it{enqueues }- \#\it{non-null dequeues}}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By correctness of \nf{size} field from Lemma \ref{sizeCorrectness} $\it{\#non-null dequeues}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
+\end{lemma}
+\begin{proof}
+There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
+$\#\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ $-$ $\#\it{non-null dequeues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
+By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness}, $\it{\#non-null dequeues}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
 \end{proof}
@@
 \begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].\size}
+  If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
 \end{corollary}
@@
 \begin{lemma}\label{nullReturn}
-\nf{R(D\sub{root,b,i})} is \nf{null} iff \nf{root.blocks[b-1].\size+ root.blocks[b].num\sub{enq}- i <0}.
-\end{lemma}
-\begin{proof}
-From \ref{numNullDeqBlock}, \ref{defNullDeq} and \ref{enqBeforeDeq}.
-\end{proof}
-
-\begin{lemma}[Computing Response] \label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$\nf{.element}.
-\end{lemma}
-\begin{proof}
-Note that by Definition \ref{ordering} the linearization ordering  of operations will not change as new operations come so instead of talking about the linearization of operations before the $E_i(root, b)$ we talk about what if the whole operation in the linearization are applied on a queue.
-
-$D_i(root,b)$ is $D_{root.blocks[b-1].sum_{deq}+i}(root)$ from the Definitions \ref{ordering} and \it{$sum_{deq}$}.
-$D_i(root,b)$ returns \nf{null} if $\nf{root.blocks[}$b-1$\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ by Lemma \ref{nullReturn} (Line \ref{checkEmpty}).
-Otherwise if it is $k$th non-null dequeue in $L$ it returns $k$th enqueue by Observation \ref{responseToADeq}.
+\nf{R(D\sub{root,b,i})} is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
+\end{lemma}
+\begin{proof}
+From Corollary \ref{numNullDeqBlock} and Observation \ref{enqBeforeDeq}.
+\end{proof}
+
+\begin{lemma}\label{computeHead}
+\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
+\end{lemma}
+\begin{proof}
+$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumDeq}.
+$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
+Otherwise if $D_i(root,b)$ is $e$th non-null dequeue in $L$ it should return $e$th enqueued value.
 By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null dequeue operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-Note that the dequeues in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
-So the response is $E_{i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}}(root)$ (Line \ref{computeE}).
-See figure \ref{computeResponseDetail}.
-
-After computing \nf{e}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ by \nf{DSearch} and then finds its \nf{element} using \nf{GetEnq} (Line \ref{findAnswer}).
+The dequeues in the \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
+So $Resp(D_i(root,b))$ is the \nf{element} field of $e$th non-null dequeue when $e=i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}$ (Line \ref{computeE}).
+See Figure \ref{computeResponseDetail}.
+
+After computing \nf{e} at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
 \end{proof}
@@
 
-\begin{lemma}[correctness]\label{linearCorrect}
-  If operations are applied as $L$ on a sequential queue, the sequence of the responses would be the same as our algorithm.
-\end{lemma}
-\begin{proof}
-Enqueue operations do not return any value so they are fine. Response of a dequeue in sequential execution of $L$ is same as our algorithm from \ref{computeHead}.  
+\begin{lemma}\label{linearCorrect}
+The responses to operations in our algorithm would be the same as in the sequential execution in the order given by $L$.
+\end{lemma}
+\begin{proof}
+Enqueue operations do not return any value. By \ref{computeHead} response of a dequeue in our algorithm is same as the response from the sequential execution of $L$.  
 \end{proof}
@@
 
-\pagebreak
-
+\paragraph{Remark} In fact our algorithm is strongly linearizable \cite{DBLP:conf/stoc/GolabHW11}.
+By Definition \ref{ordering} the linearization ordering of operations will not change as new operations come.
+\pagebreak
+
+\section{Analysis}
 \begin{lemma}[Amortized time analysis]
-%  \nf{n.GetEnq(b,i), n.Index(b,i)} take $O(\log^2 p)$ steps.
-Search in the root may take $O(\log Q+ p^2)$ steps.
-Helping is done every $p^2$ block appended to the root and takes $p\times \log^2p$ steps.
-Amortized time consumed for helping by each process is $O(\log^2 p)$.
 \nf{Enqueue()} and \nf{Dequeue()}, each take $O(\log^2 p + \log q)$ steps in amortized analysis.
@@
 The first two part is similar to an \nf{Enqueue} operation.
-To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDeq()} is called.
-\nf{IndexDeq} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree.
+To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDequeue()} is called.
+\nf{IndexDequeue} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree.
 In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$.
-Each search in \nf{GetEnq()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnq()} takes $O(\log^2 p)$ steps.
+Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
 
@@
 
-\begin{lemma}[CASes invoked]
+\begin{lemma}
 An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.

### 2022-09-29 thesis/thesis.tex
update figures

@@
 
-We show $EST_{n\nf{.left}}^{t^R}= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]})$.
+We show 
+\begin{align*}
+EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \\
+&\subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]}).
+\end{align*}
 
 Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$.
-Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[1..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[1..$n.head_{\ref{readHead}}$]}.
+Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[$1\cdots n.head_{\ref{readHead}}$]}.
 Because of Lemma \ref{nonDecreasingHead} we have $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{cas}}$.
@@
 \end{proof}
+
 \begin{figure}[hbt]
   \center\includegraphics[width=5in]{pics/trueRefresh}
-  \caption{New established operations of the left child are in the new block. (TO UPDATE)}
+  \caption{TODO :: Is this picture good or should I remove it? New established operations of the left child are in the new block.(TO UPDATE)}
 \end{figure}
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=7in]{pics/timeLineSuper1.png}
+  \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
   \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
@@
 
-Let the first successful \nf{p.Refresh()} be \nf{Rp1} and the second next successful \nf{p.Refresh()} be \nf{Rp2}.
-If \nf{Rp1} reads \nf{x} in Line \ref{readHead}, then \nf{Rp2} has to read \nf{x+1} in Line \ref{readHead} (iduced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
-See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances \nf{Rp1, Rp2} on \nf{p}.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/timeLineSuper2.png}
+Let the first successful \nf{p.Refresh()} be $R_{p1}$ and the second next successful \nf{p.Refresh()} be $R_{p2}$.
+If $R_{p1}$ reads \nf{x} in Line \ref{readHead}, then $R_{p2}$ has to read \nf{x+1} in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
+See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$  on ${p}$.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
   \caption{Rp2 reads p.head after $t_{321}^{Rp1}$, which is after $t_{321}^{Rn}$. Rp2 has to help increment n.head and set b.super.}
@@
 
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/timeLineSuper3.png}
-  \caption{ The second Refresh on \nf{p} contains \nf{b} and reads $\nf{n.head}>\nf{i}$.}
-  \label{fig::timeLine3}
-\end{figure}
+$$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
+$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh()$ after $R_n$.
+$t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh()} instances on a node.
+$t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code.
+The second Refresh on \nf{p} contains \nf{b} and reads $\nf{n.head}>\nf{i}$.
 
@@
 \begin{figure}[hbt]  
-  \center\includegraphics[width=2in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
+  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
 \label{computeResponseDetail}
@@
   \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$. \nf{start} is shown with $c$.}
+  \caption{Distance relations between \nf{start}$,b,end$.}
   \label{fig::doubling}

### 2022-09-29 backup/jul 14.tex
fix citations, review index deq

--- 2022-02-03 article/article.tex
@@
 
-\usepackage[b4paper,left=0.8in,right=0.8in,top=0.8in,bottom=0.8in]{geometry}
+\usepackage[a4paper,left=0.8in,right=0.8in,top=1.1in,bottom=1.1in]{geometry}
 \usepackage{tikz-qtree}
@@
 \usepackage{subcaption}
-\usepackage{showkeys}
-\usepackage{amsmath}
 
@@
 
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
 \title{Wait-free Queues with Polygarithmic Step Complexity}
@@
 \theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
+\newtheorem{definition}{Definition}
 
@@
 \paragraph{}
-We choose to linearize operations in a block from the left child before those from the right child as a convention.
-Operations within a block of the root can be ordered in any way that is convenient.
-In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
+We choose to linearize operations from the left child before those from the right child as a convention.
+In effect, this means that if there are concurrent operations in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
 So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
@@
 \paragraph{}
-Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements.
-It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.
+Our Block Tree is a linearizable implementation of a data structure that stores the elements appended by processes.
+It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{GetIndex}($i$) which returns the $i$th element in the sequence.
 
@@
 
-Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
+Each process is assigned with a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of operations by $p_i$ in the order they were invoked.
 Each internal node of the tree contains an array of blocks of elements.
@@
 Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order.
-Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively.
-Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
+Moreover, two numbers, left, right indicating the count of elements in the blocks from left and right child consecutively.
+Furthermore, two amounts, prefix left and prefix right, can be computed from the prefix sum of left and right values.
 Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
-The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
+Element $i$th in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
 After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).
@@
 Here is the algorithm to get elements of a block.
-
-\paragraph{Specification}
-A block tree is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Append(e)} and \texttt{Get(i)}.
-\texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence.
-\texttt{Get(i)} returns \texttt{i}th element stored in the sequence.
-
-\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.
-
-\paragraph{Membership}
-Element \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
-  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
-\end{itemize}
-
-\paragraph{Order of elements inside node}
-Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
-\begin{itemize}
-  \item The block containing d is before the block containing e.
-  \item e and d are in the same block and d is in the left child and e is in the right child.
-  \item d is before e in the same child's order.
-\end{itemize}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
-\end{figure}
-
-\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
-\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
-Other attributes are assigned values followed by definition.
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
-\end{figure}
-
-%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}
-
-\paragraph{Double Refresh}
-Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
-\begin{proof}
-\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block.
-If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
-So suppose two \texttt{Refresh}es fail.
-Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}.
-So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}.
-After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-\end{proof}
-\begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
-\end{figure}
-
-\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
-\begin{proof}
-Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks.
-So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end.
-Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
-\end{proof}
-
-\paragraph{Total Order}
-Sequence represented by the Block Tree is the sequence of the blocks stored in the root.
-
-\paragraph{Linearization Points}
-\texttt{Get(i)} is linearized when it terminates.
-\texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.
-
-\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
-\begin{proof}
-If there are more than $p$ subblocks, then there is more than one block from process pl.
-\texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9).
-So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
-\end{proof}
-
-\paragraph{Computing \texttt{Get(n, b, i)}}
-To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$.
-Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.
-
-\paragraph{How \texttt{Refresh(n)} works.}
-\begin{enumerate}
-  \item Read n's counter and head
-  \item Create block b
-  \item CAS b into n
-  \item If previous succeed:
-  \begin{enumerate}
-  \item Update sup of b's ending subblocks
-  \item Increment children's counters
-  \end{enumerate}
-\end{enumerate}
-
-%\paragraph{Time-Space Relation} There are at most $p^2$ blocks with same \texttt{time} attribute.
-%\begin{proof}
-%If a process dies at line 28, the children counter won't increase, and each process aggregates most block, and since there are at most p concurrent processes, then there will be at most $p^2$ blocks with same \texttt{time} attribute.
-%\end{proof}
-%
-%\paragraph{Time-Propagate() Relation} In a \texttt{CreateBlock()} procedure, the blocks that are seen have at most 2 different \texttt{time} attribute values.
-%
-%\paragraph{Time-Super Relation} sup[i+1]-sup[i] is $\leq p$
-%
-%\paragraph{\texttt{b}'s Superblock position} SuperBlock of \texttt{b} is within range $\pm p$ of the \texttt{\texttt{sup[b.time]}}.
-%\begin{proof}
-%
-%\end{proof}
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/super}
-  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
-\end{figure}
-
-\paragraph{Computing superblock}
-\begin{enumerate}
-  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
-    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10.
-For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
-  \end{proof} 
-  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
-   Follows from the order of lines 37, 26, 27.
- \end{proof}
-%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
-  \item \texttt{super[i+1]-super[i]}$\leq p$
-\begin{proof}
- In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27.
-Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
-If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27.
-So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
- \end{proof}
-  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
-\begin{proof}
-\texttt{super[i]} is the index of the superblock of a block containing block b.
-It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
-\texttt{super(b)} is the real superblock of b.
-\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
-If \texttt{b.time} is \texttt{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-  \end{proof} 
-\end{enumerate}
-
-%The block tree contains a tournament tree in which leaves are assigned to processes, and each process appends elements to its leaf and tries to propagate it up to the root.
-The final sequence is stored at the root.
-\texttt{Append} after appending to leaf calls \texttt{Propagate} which iterates over the path to root.
-In each \texttt{Propagate(n)} invocation on node n it tries ti \texttt{Refresh} the sequence stored at node n with the new elements in n's children.
-After adding the element to the root \texttt{Index(op)} finds the order of the element in the root.
-
-%\textsf{I finally thought maybe a sequence of elements is better for telling specification.}
 
@@
 
-\Statex $\blacktriangleright$ \texttt{\textsl{element} e}
+\Statex $\blacktriangleright$ \texttt{\textsl{operation} op}
 \begin{itemize}
@@
 \item \texttt{\textsl{int} pid}
-\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}.ops}}
+\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}'s array}}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \texttt{\textsl{leaf} l\textsubscript{i}}
+\begin{itemize}
+\item \texttt{\textsl{*node} parent}\textsf{: pointer to parent in Tree}
+\item \texttt{\textsl{operation[]} ops}\textsf{: ops invoked by process i}
+\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{ops}}
 \end{itemize}
@@
 \item \texttt{\textsl{*node} left, right, parent}
-\item \texttt{\textsl{block[]} blocks}\textsf{: blocks stored in \texttt{n}}
-\item \texttt{\textsl{int} head= 1}\textsf{: index of first empty cell of \texttt{blocks}}
-\item \texttt{\textsl{int} counter= 0}\textsf{}
-\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is \texttt{i}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{leaf node} l\textsubscript{i} extends \textsl{node}}
-\begin{itemize}
-\item \texttt{\textsl{operation[]} ops}\textsf{: elements that are invoked to \texttt{Append()} to the blok tree by process i}
+\item \texttt{\textsl{block[]} blocks}\textsf{: blocks propagated to \texttt{n}}
+\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{blocks}}
+\item \texttt{\textsl{int} counter= 0}\textsf{: \#groups of blocks propagated from \texttt{n}}
+\item \texttt{\textsl{int[]} super}\textsf{: approximate index of superblock of a block in the given group in \texttt{n}'s parent}
 \end{itemize}
@@
 \begin{itemize}
-  \item \texttt{\textsl{int} num\textsubscript{left}, sum\textsubscript{left}}
-
-  \textsf{~~\#operations from the left subblocks of b, prefix sum of num\textsubscript{left}}
-
-  \item \texttt{\textsl{int} num\textsubscript{right}, sum\textsubscript{right}}
-
-  \textsf{~\#operations from the right subblocks of b, prefix sum of num\textsubscript{right}}
-
-  \item \texttt{\textsl{int} sum}
-
-  \textsf{\# operation in b}
-
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
-
-  \textsf{~~index of \texttt{b}'s last subblock}
-  \item \texttt{\textsl{int} time}
-
+  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}}\textsf{: \#operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
+  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}\textsf{: index of \texttt{b}'s last subblock}
+  \item \texttt{\textsl{int} group}\textsf{: approximate group number}
 \end{itemize}
 
-%\Statex \textbf{Examples}
-%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
-%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
-%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
+\Statex \textbf{Examples}
+\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
+\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
+\Statex \hspace{0pt} \texttt{ n.blocks[n.last]}\textsf{: index of last block of node n}
+\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
+\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
 
@@
 \Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
-\State \texttt{op.loc= l\textsubscript{pid}.head}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.time= op.loc}
-\State \texttt{b.sum= 1}
-\State \texttt{l\textsubscript{pid}.ops[op.loc]= op}
-\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
-\State \texttt{op.head+= 1}
+\State \texttt{op.loc= l\textsubscript{pid}.last + 1}
+\State \texttt{l\textsubscript{pid}.ops[op.location]= op}
 \State \Call{Propagate}{l\textsubscript{pid}.parent}
-\State \Return \Call{Index}{l\textsubscript{pid}, op.loc, b} 
+\State \Return \Call{Index}{l\textsubscript{pid}, op.location, b} \Comment we don't know the block eachoperation resides in leaf
 \EndFunction{Append}
@@
 \Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{not} \Call{Refresh}{n}}
+\If{\texttt{n {\keywordfont is} root}} \Return
+\ElsIf{\texttt{not} \Call{Refresh}{n}}
 \State \Call{Refresh}{n}
-\EndIf
-\If{\texttt{n {\keywordfont is} root}} \Return
 \EndIf
@@
 \Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{i= n.head}
-\State \texttt{c= n.counter}
-\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, c}}
+\State \texttt{i= n.last}
+\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, n.counter}}
 \If{\Call{CAS}{n.blocks[i+1], null, new}}
+\State \texttt{\Call{CAS}{n.last, i, i+1}}
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
@@
 \State \Return{\texttt{true}}
-\Else 
-\State \Return{ \texttt{false}}
+\Else ~\Return{ \texttt{false}}
 \EndIf
-\State \texttt{\Call{CAS}{n.head, i, i+1}}
 \EndFunction{Index}
@@
 
-\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int c} 
+\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int g} 
 \Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
 \State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.time= c}
+\If{\texttt{n is leaf}}\Comment{TODO}
+\Else
+\State \texttt{b.group= g}
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{j= n.dir.head}
+\State \texttt{j= n.dir.last}
 \State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
 \State \texttt{end= n.dir.blocks[j]}
-\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].time}
+\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].group}
 \State \texttt{b.end\textsubscript{dir}= j}
-\State \texttt{b.num\textsubscript{dir}= end.sum - start.sum}
+\State \texttt{b.num\textsubscript{dir}= end.sum\textsubscript{\#dir} - start.sum\textsubscript{\#dir}}
 \State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
@@
 \State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
+\EndIf
 \State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{46}
 
 \Function{element}{Get}{\textsl{int} i} 
-\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.head}}
+\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.last}}
 \State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
@@
 
-\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
+\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with feild \texttt{f} $\geq$ \texttt{i}}
 \Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}
@@
 \Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}
-\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[i]}
+
+\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
+\If{\texttt{n.left {\keywordfont is} leaf}} \Return \texttt{n.left.ops[i]}
 \Else
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
 \State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
 \State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
+\EndIf
 \Else
 \State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
+\If{\texttt{n.right {\keywordfont is} leaf}} \Return \texttt{n.right.ops[i]}
+\Else 
 \State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
-\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} 
-\EndIf
+\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} \EndIf
 \EndIf
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition:~\texttt{i}th operation in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
-\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns rank in the  
-root of the \texttt{i}th operation in node \texttt{n}.}
+
+\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns order in  the root of \texttt{i}th operation in the node \texttt{n} which exists in block b}
 \If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
@@
 \State \texttt{dir= (n.parent.left==n)? left: right}
-\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].time]-p, sup[n.blocks[b].time]+p}}
+\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].group]-p, sup[n.blocks[b].group]+p}}
 \If{\texttt{dir {\keywordfont is} left}}
@@
 \subsection{Idea in a nutshell}
-With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log ^2p+\log n)$ steps.
-A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
-If the queue is empty it returns \texttt{null}.
-Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.
-So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
-%
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
-\paragraph{}
-Next, we describe how to use block tree to implement queues.
-The block tree, maintains the history of all operations, not only the current state of the queue.
-Now consider the following history of operations.
-What should each \texttt{Dequeue()} return?
+With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log p)$ steps.
+A queue consitsts of two operations, enqueue and dequeue.
+\textsc{Enqueue}$(n)$ appends element $n$ to the sequence stored, this is somehow similar to \textsc{Append}$(n)$.
+Since \textsc{Append}$(n)$  adds $n$ to the history of all operations.
+\textsc{Dequeue()} returns first not dequeued element among all elements.
+If index $i$ is the tail of the queue, we can return the dequeue response using \textsc{GetElement}$(i)$.
+So in the rest we modify block tree to compute $i$ for each \textsc{Dequeue}$()$ to achieve a FIFO queue.
+
+\subsection{What should a \textsc{Dequeue}$()$ return, regarding history of operations?}
+\paragraph{}
+In a basic look of we use block tree to implement queues the block tree, we maintain the sequence of all elements(history of all operations), not only the current state of the queue.
+Regarding the current state of the queue, if it is not empty, the \textsc{Dequeue}$()$ returns the first existing element.
+And if the queue is empty it returns \texttt{null}.
+Now look at all the history of operations, what should $i$th \textsc{Dequeue}$()$ return?
 
@@
 
+\paragraph{}
+In the example above, \textsc{Dequeue}$()$ operations return \texttt{5, 2, 3, null, 4} in order.
+If the queue size is greater than $0$ and $r$ Dequeue operations have returned a non-null response, then $i$th Dequeue returns $r+1$th Enqueue input.
+So in order to answer a Dequeue, there is only needed to know the size of the queue and $\#$ returning dequeues.
+
+\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
 \begin{definition}
-A non-null dequeue is one that returns a non-null value.
+After a sequence of operations, $\#$returning dequeues is the number of Dequeue operations that have returned non null value.
+\texttt{DEQ[i]} is the $i$th Dequeue.
+\texttt{ENQ[i]} is the argument of $i$th Enqueue.
 \end{definition}
 
-\paragraph{}
-In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order.
-Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null.
-If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}.
-So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.
-
-%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
-
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}
-
-\paragraph{}
-In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism.
-So now the problem is to find the result of each Dequeue.
-From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
-In the next example, operations in a cell are concurrent.
-\texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
-We will next describe how these values can be computed efficiently.
+\subsection{What should a \textsc{Dequeue}$()$ return, regarding history of blocks of operations?}
+
+\paragraph{}
+In the Block Tree idea, we did not store the sequence of operations themselves but instead stored blocks of concurrent operations to optimize \textsc{Propagate()} steps and increase parallelism.
+So now the problem is to find \texttt{ENQ[j]} for \texttt{DEQ[i]}.
+From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue.
+In the next example, operations in a cell are concurrent, and the order within a block is as said before.
+\texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4 , null} respectively.
 
@@
   \end{tabular}
-  \caption{An example history of operation blocks on the queue}
+  \caption{An example histoy of operation blocks on the queue}
 \end{table}
@@
 \paragraph{}
-Now, we claimed that by knowing the current size of the queue and the number of non-null dequeue operations before the current dequeue, we could compute the index of the resulting \texttt{Enqueue()}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations happens and the number of non-null dequeues dequeues till a block, we can compute each dequeue's index of result in \textsc{O}$(1)$ steps.
+In the last section, we claimed that by knowing the current size of the queue and $\#$returning dequeue operations before the current state, we could decide the index of result Enqueue.
+Now we apply this information to blocks; if we store the size of the queue after a block happens and the number returning dequeues in a block, we can compute every dequeue result in \textsc{O}$(1)$ steps.
 
@@
     \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
-    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
-        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
-            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
     size & 0 & 2 & 2 & 0 \\ \hline
+    \#enqs & 0 & 3 & 1 & 1 \\ \hline
+        \#deqs & 1 & 1 & 1 & 4 \\ \hline
+            \#redqs & 0 & 1 & 1 & 3 \\ \hline
   \end{tabular}
-  \caption{Augmented history of operation blocks on the queue}
+  \caption{Augmented histoy of operation blocks on the queue}
 \end{table}
 
-%\begin{definition}
-%  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
-%  
-%  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
-%  
-%\end{definition}
-
-Size and the number of non-null dequeues for $b$th block could be computed this way:\\
-\texttt{size[b]= max(size[b-1] +enqueues[b] -dequeues[b], 0)}\\
-\texttt{non-null dequeues[b]= non-null dequeues[b-1] +dequeues[b] -size[b-1] -enqueues[b]}
+\begin{definition}
+  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
+
+  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
+
+\end{definition}
 
 Given \texttt{DEQ} is in block \texttt{b}, \texttt{response(DEQ)} would be:\\
-\texttt{(size[b-1]- index of DEQ in the block's dequeus >=0) ? ENQ[non-null dequeus[b-1]+ index of DEQ in the block's dequeus] : null;}
+\texttt{(size[b-1]-index(DEQ)>=0) ? ENQ[INDEX(RDEQ)+index(DEQ)] : null;}
 
 \section{Main Algorithm}
-
-\paragraph{Specification}
-A Queue is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
-\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
-\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
-
-%Now that we know how to augment Block Tree root history to compute the response, we will present the complete algorithm.
-\texttt{Enqueue()} and \texttt{Dequeue()} are the main functions.
-\texttt{Enqueue(e)} adds \texttt{e} to the queue and \texttt{Dequeue()} returns the first element in the queue and removes it from the queue.
-Block Tree functionalities are very similar.
-\texttt{Append(e)} does what \texttt{Enqueue()} does and if we know the current index of the head of the queue we can call \texttt{Get()} to find its value.
-
-\begin{figure}[hbt]
-\centering
-  \includegraphics[width=5in]{pics/queue}
-  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
-\end{figure}
-
-\begin{algorithm}
-\caption{Queue \label{algQ}}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-%\Statex \textbf{Structure}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Operation}}
-\begin{itemize}
-%\item \texttt{\textsl{string} name}
-%\item \texttt{\textsl{string} type$\in$\{DEQ,ENQ\}}
-\item \texttt{\textsl{Object} element}\textsf{: argument of an enqueue, if null this is a dequeue}
-%\item \texttt{\textsl{int} pid}
-\item \texttt{\textsl{int} loc\textsf{: location in the leaf's \texttt{ops}}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Node}}
-\begin{itemize}
-\item \texttt{\textsl{*Node} left, right, parent}
-\item \texttt{\textsl{Block[]} blocks}\textsf{: index 0 contains an empty block with all fields equal to 0}
-\item \texttt{\textsl{int} head= 1}\textsf{: index of the first empty cell of \texttt{blocks}}
-\item \texttt{\textsl{int} counter= 0}\textsf{}
-\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is field \texttt{i}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Leaf} extends \textsl{Node}}
-\Statex \Comment{\textsf{\texttt{l\textsubscript{i}} is the Leaf for process \texttt{i} and \texttt{i} is in range \texttt{0} to \texttt{p-1}}}
-\begin{itemize}
-\item \texttt{\textsl{Operation[]} ops}\textsf{: invoked operations}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Root} extends \textsl{Node}}
-\begin{itemize}
-\item \texttt{override \textsl{Root Block[]} blocks}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Block}}
-
-\begin{itemize}
-  \item \texttt{\textsl{int} num\textsubscript{enq-left}, sum\textsubscript{enq-left}}
-
-  \textsf{~~\#enqueues from subblocks in left child, prefix sum of \texttt{num\textsubscript{enq-left}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{deq-left}, sum\textsubscript{deq-left}}
-
-  \textsf{~~\#dequeues from subblocks in left child, prefix sum of \texttt{num\textsubscript{deq-left}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{enq-right}, \texttt{sum\textsubscript{enq-right}}}
-
-  \textsf{~\#enqueues from subblocks in right child, prefix sum of \texttt{num\textsubscript{enq-right}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{deq-right}, \texttt{sum\textsubscript{deq-right}}}
-
-  \textsf{~\#dequeues from subblocks in right child, prefix sum of \texttt{num\textsubscript{deq-right}}}
-
-  \item \texttt{\textsl{int} num\textsubscript{enq}, num\textsubscript{deq}}
-
-  \textsf{\# enqueue, dequeue operations in the block}
-
-  \item \texttt{\textsl{int} sum\textsubscript{enq}, sum\textsubscript{deq}}
-
-  \textsf{sum of \# enqueue, dequeue operations in blocks up to this one}
-
-  \item \texttt{\textsl{int} num, sum}
-
-  \textsf{total \# operations in block, prefix sum of \texttt{num}}
-
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}
-
-  \textsf{~~index of this block's last subblock in the left and right child}
-  \item \texttt{\textsl{int} group}
-
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{Root Block} extends \textsl{Block}}
-\begin{itemize}
-  \item \texttt{\textsl{int} size}
-
-  \textsf{size of queue after this block's operations finish}
-
-  \item \texttt{\textsl{int} sum\textsubscript{non-null deq}}
-
-  \textsf{count of non-null dequeus up to this block}
-
-\end{itemize}
-
-%\Statex \textbf{Examples}
-%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
-%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
-%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
-%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
-
-\Statex
-
-\Function{void}{Enqueue}{\textsl{Object} e}
-\State \texttt{op= NEW(\textsl{operation})}
-%\State \texttt{op.type= ENQ}
-\State \texttt{op.element= e}
-%\State \texttt{op.pid= this.pid}
-\State \Call{Append}{op}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Function{Object}{Dequeue()}{}
-\State \texttt{op= NEW(\textsl{operation})}
-%\State \texttt{op.type= DEQ}
-%\State \texttt{op.pid= this.pid}
-\State  \Call{Append}{op} \Comment{$i$th DEQ in $b$th block}
-\State \texttt{<i, b>=} \Call{Index}{l\textsubscript{pid}, op.loc, 1}
-\State \texttt{res=} \Call{ComputeHead}{i, b} \Comment{Index of the enqueue whose argument should be returned}
-\If{\texttt{res }\textbf{is }\texttt{null}}
-\State \Return \texttt{null}
-\Else
-\State \Return \Call{Get}{res}
-\EndIf
-\EndFunction{Dequeue}
-
-\Statex
-
-\Function{int}{ComputeHead}{int i, int b}\Comment{Computes head of the queue when \texttt{i}th dequeue in \texttt{b}th block occurs. The dequeue should return the argument of the head enqueue.}
-\If{\texttt{root.blocks[b-1].size + root.blocks[b].num\textsubscript{enq} - i $<$ 0}}
-\State \Return \texttt{-1}
-\Else{ \Return \texttt{root.blocks[b-1].sum\textsubscript{non-null deq} + i}}
-\EndIf
-\EndFunction{ComputeHead}
-
-\Statex
-
-\Function{void}{Append}{\textsl{operation} op} 
-\State \texttt{pid= id of process performing \textsc{Append}}
-\State \texttt{op.loc= l\textsubscript{pid}.head}
-\State \texttt{op.head+= 1}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.group= op.loc}
-\If{\texttt{op.element==null}} 
-\texttt{b.sum\textsubscript{deq}=1} \Else \texttt{~b.sum\textsubscript{enq}=1} \EndIf
-%\State \texttt{b.sum\textsubscript{op.type}= 1}
-\State \texttt{l\textsubscript{pid}.ops[op.loc]= op} \label{addOP}
-\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
-\State \Call{Propagate}{l\textsubscript{pid}.parent} 
-\EndFunction{Append}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Queue Continued}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{33}
-\begin{multicols}{2}
-
-\Function{void}{Propagate}{\textsl{node} n}
-\If{\texttt{not} \Call{Refresh}{n}}
-\State \Call{Refresh}{n}
-\EndIf
-\If{\texttt{n.parent \textbf{is} null}} 
-\State \Call{Propagate}{n.parent}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{h= n.head}
-%\If{\texttt{n.blocks[h]!=null?}} \texttt{h+=1}\EndIf
-\State \texttt{t= n.counter}
-\State \texttt{<new, t\textsubscript{left}, t\textsubscript{right}>= \Call{CreateBlock}{n, h, t}}
-\If{\texttt{new.num==0}} \Return{\texttt{true}}
-\ElsIf{\Call{CAS}{n.blocks[h], null, new}}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{\Call{CAS}{n.dir.super[t\textsubscript{dir}], null, h+1}}
-\State \texttt{\Call{CAS}{n.dir.counter, t\textsubscript{dir}, c\textsubscript{dir}+1}}
-\EndFor
-\State \texttt{\Call{CAS}{n.head, h, h+1}}
-\State \Return{\texttt{true}}
-\Else
-\State \texttt{\Call{CAS}{n.head, h, h+1}}
-\State \Return{ \texttt{false}}
-\EndIf
-\EndFunction{Refresh}
-
-\Statex
-
-\Function{element}{Get}{\textsl{int} i}
-\Comment{Returns $i$th Enqueue.}
-\State \texttt{res= \Call{BSearch}{root, sum\textsubscript{enq}, i, 0, root.head}}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum\textsubscript{enq}}}
-\EndFunction{Get}
-
-\Statex
-
-\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
-\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}
-
-\Statex \Comment{\textmd{Does binary search for~the value \texttt{i} of the given prefix sum \texttt{feild}.
-\texttt{f} is one of \texttt{sum, sum\textsubscript{left}, sum\textsubscript{right}}.
-Returns the index of the leftmost block in \texttt{n.blocks[start..end]} whose \textsl{field} \texttt{f} is $\geq$ \texttt{i}}.}
-%\State \Return \texttt{result block's index}
-\EndFunction{BSearch}
-
-\pagebreak
-
-\Function{<Block, int, int>}{CreateBlock}{\textsl{node} n, \textsl{int} i, int t} 
-\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]} with time field=\texttt{t}. Returns the created block as well as values read from each child counter feild.}
-\State \texttt{block b= \Call{new}{\textsl{block}}}
-\State \texttt{b.group= t}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{lastIndex= n.dir.head} \label{lastLine}
-\State \texttt{prevIndex= n.blocks[i-1].end\textsubscript{dir}} \label{prevLine}
-\State \texttt{lastBlock= n.dir.blocks[lastIndex]}
-\State \texttt{prevBlock= n.dir.blocks[prevIndex]}
-\State \texttt{t\textsubscript{dir}= n.dir.counter}
-\State \texttt{b.end\textsubscript{dir}= lastIndex}
-\State \texttt{b.num\textsubscript{enq-dir}= lastBlock.sum\textsubscript{enq} - prevBlock.sum\textsubscript{enq}}
-\State \texttt{b.num\textsubscript{deq-dir}= lastBlock.sum\textsubscript{deq} - prevBlock.sum\textsubscript{deq}}
-\State \texttt{b.sum\textsubscript{enq-dir}= n.blocks[i-1].sum\textsubscript{enq-dir} + b.num\textsubscript{enq-dir}}
-\State \texttt{b.sum\textsubscript{deq-dir}= n.blocks[i-1].sum\textsubscript{deq-dir} + b.num\textsubscript{deq-dir}}
-\EndFor
-\State \texttt{b.num\textsubscript{enq}= b.num\textsubscript{enq-left} + b.num\textsubscript{enq-right}}
-\State \texttt{b.num\textsubscript{deq}= b.num\textsubscript{deq-left} + b.num\textsubscript{deq-right}}
-\State \texttt{b.num= b.num\textsubscript{enq} + b.num\textsubscript{deq}}
-\State \texttt{b.sum= n.blocks[i-1].sum + b.num}
-
-\If{\texttt{n.parent} \textbf{is} \texttt{null}}
-\State \texttt{b.size= max(root.blocks[i-1].size + b.num\textsubscript{enq} - b.num\textsubscript{deq}, 0)}
-\State \texttt{b.sum\textsubscript{non-null deq}= root.blocks[i-1].sum\textsubscript{non-null deq} + max( b.num\textsubscript{deq} - root.blocks[i-1].size - b.num\textsubscript{enq}, 0)}
-\EndIf
-
-\State \Return \texttt{b, t\textsubscript{left}, t\textsubscript{rightt}}
-\EndFunction{CreateBlock}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Queue Continued}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{83}
-
-\Statex $\leadsto$ \textsf{Precondition:~\texttt{n.blocks[b]} contains $\geq$\texttt{i} enqueues.}
-\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
-\Comment{\textmd{Returns the \texttt{i}th Enqueue in \texttt{b}th block of node \texttt{n}}}
-\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[b]}
-\Else
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{enq-left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
-\State \texttt{subBlock= \Call{BSearch}{n.left, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
-\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum\textsubscript{enq}} 
-\Else
-\State \texttt{i= i-n.blocks[b].num\textsubscript{enq-left}}
-\State\texttt{subBlock=\Call{BSearch}{n.right, sum\textsubscript{enq}, i, n.blocks[b-1].end\textsubscript{right}+1, n.blocks[b].end\textsubscript{right}}}
-\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum\textsubscript{enq}} 
-\EndIf
-\EndIf
-\EndFunction{Get}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \texttt{b}th block of node \texttt{n} has propagated up to the root and \texttt{i}th dequeue resides in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
-\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{root.blocks[b-1]+i, b}
-\Else
-\State \texttt{dir= (n.parent.left==n)? left: right}
-\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{deq-dir}, i, super[n.blocks[b].group]-p, super[n.blocks[b].group]+p}}
-\If{\texttt{dir {\keywordfont is} left}}
-\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq-right}}
-\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{deq} + n.blocks[superBlock].sum\textsubscript{deq-left}}
-\EndIf
-\State \Return\Call{Index}{n.parent, superBlock, i}
-\EndIf
-\EndFunction{Index}
-
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Pseudocode description}
-
-\paragraph{Tournament Tree}
-In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree.
-Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}.
-Each process adds $op$ to its leaf.
-In each internal node an ordering of operations in its subtree is stored.
-All processes agree on the total ordering of all operations stored in the root.
-This ordering will be the linearization of the operations.
-\paragraph{Implicit Storing Blocks}
-For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks.
-A Block is a constant size object that implicitly represents a sequence of operations.
-In each node there is an array of Blocks.
-
-\begin{definition}[Block]
-A block is an object that stores some statistics described in Algorithm Queue.   
-\end{definition}
-
-\begin{definition}[\label{def::subblock}Subblock]
-Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
-\end{definition}
-Block $b$ contains subblocks in the left and right children.
-WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$.
-See Figure \ref{fig::createBlock} .
-
-\begin{definition}[Membership of an operation in a block]
-Operation \texttt{e} is a member of block \texttt{b} in:
-\begin{itemize}
- \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
- \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
-\end{itemize}
-
-\end{definition}
-
-We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s.
-In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}.
-Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
-There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step.
-It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases.
-Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}.
-This helps us to know which blocks have aggregated together to a block, not precisely though.
-We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.
-
-In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}.
-In order to make \texttt{BSearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
-
-To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}.
-\texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.
-
-\texttt{Enqueue(e)} just \texttt{append}s an \texttt{operation} with \texttt{element e} to the \texttt{root}.
-\texttt{Dequeue()} appends an \texttt{operation} to the root and computes its ordering and the \texttt{enqueue operation} containing the head before it calling \texttt{ComputeHead()} and then \texttt{gets} and returns the \texttt{operation}'s element.
-
-\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index.
-As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times.
-\texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.
-
-We only want to know the \texttt{element} of \texttt{enqueue} operations and compute ordering for \texttt{dequeue} operations.
-That's the reason here \texttt{Get()} searches between enqueues only and \texttt{Index()} returns ordering of a dequeue among dequeues.
-\texttt{Get(n, b ,i)} decides the requested element is in which child of n and continues to search recursively.
-\texttt{index(n, i, b)} calculates the ordering of the given operation in \texttt{n}'s parent each step and finally returns the result among total ordering.
-
-\subsection{Complexity Analysis}
-
-\texttt{Enqueue()} operations do a constant number of steps and an \texttt{Append()} which calls \texttt{Propagate()}.
-Let $p$ be the number of processes and $m$ be the count of operations invoked.
-In a \texttt{Propagate()} step there are a constant number of steps taken at each level and the height of the tree is $\log(p)$.
-So \texttt{Enqueue()} takes $O(\log p)$ steps.
-
-\texttt{Get()} searches among all blocks at first and then iterates over a path from the root to a leafof the tree and in each step searches in a domain of size $p$ , so it takes $\log^2 p+\log m$.
-\texttt{Index()} is a path from a leaf to the root which each step calls a $\log p$ step search so it takes $\log^2p$.
-\texttt{Dequeue()} calls \texttt{Append()}, it also calls \texttt{Index()} and \texttt{Get()} that take $\log^2 p$, $\log^2p +\log n$.
-
-\subsection{Linearizability Proof}
-
-\begin{definition}
-If \texttt{n.blocks[i]==b}  we call \texttt{i} the \emph{index} of block \texttt{b} in node \texttt{n}.
-Block \texttt{b} is before block \texttt{b$^\prime$} in node \texttt{n} if and only if \texttt{b}'s index is smaller than \texttt{b$^\prime$}'s.
-Block \texttt{b} is propagated to node \texttt{n} or set \texttt{S} if \texttt{b} is in \texttt{n.blocks} or \texttt{S} or is a subblock of a block in \texttt{n.blocks} or \texttt{S}.
-\end{definition}
-
-%\begin{definition}
-%  Block \texttt{b} in node \texttt{n} is new if \texttt{b} is not a subblock of any block in \texttt{n.parent.blocks[]}.
-%\end{definition}
-
-\begin{definition}
-  Block \texttt{b} in node \texttt{n} is in \emph{Established(n, t)} if \texttt{n.head} is greater than \texttt{b}'s index at time $t$.
-\end{definition}
-
-\begin{lemma}[headProgress] \label{lem::headProgress}
-\texttt{n.head} is non-decreasing over time.
-\end{lemma}
-\begin{proof}
-  Simply because \texttt{n.head} is only incremented.
-\end{proof}
-
-\begin{lemma}[headPosition] \label{lem::headPosition} The value read in Line 52(\texttt{h=n.head}) might be 1 bit behind the first empty block.
-\end{lemma}
-\begin{proof}
- Because at the end of every \texttt{Refresh()} with block size greater than 0 (Lines 53,56) \texttt{n.head} is incremented.
-Maybe some process goes to sleep before incrementing the head, but after sleeping if \texttt{h} does not increase then \texttt{CAS} in Line 52 is going to be failed and nothing is going to be appended to \texttt{n.blocks}.
-\end{proof}
-
-%\begin{lemma}[oldnewOrder] \label{lem::oldnewOrder}
-%  If block \texttt{b} in node \texttt{n} is established, then all the blocks in \texttt{n.blocks[]} before \texttt{b} are established.
-If block \texttt{b} in node \texttt{n} is not new then all the blocks in \texttt{n.blocks[]} before \texttt{b} are not new.
-%\end{lemma}
-%\begin{proof}
-%  Note that blocks are only appended to a node in the algorithm. \textsf{CIRCLE REFERENCE PROBLEM. Shall we use induction to say older blocks are already appended to parent?}
-%\end{proof}
-
-\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
-  If  time $t<$ time $t^\prime$, then $Established(n, t)\subseteq Established(n, t^\prime)$.
-\end{lemma}
-\begin{proof}
-Because blocks are only appended(not modified) with CAS to \texttt{n.blocks[n.head]} and \texttt{n.head} is non-decreasing.
-\end{proof}
-
-\begin{lemma}[createBlock] \label{lem::createBlock}
-  Suppose \texttt{CreateBlock(n, h, x)} is invoked at time $t$.
-The blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ that are not propagated to $Established(n, t)$, are subblock of the block returned by \texttt{CreateBlock(n, h, x)}.
-\end{lemma}
-\begin{proof}
- We prove the claim for the left child.
-Blocks in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are all the new established operations at time $t$ by definition of Subblock.
-Line 70 is after $t$ and since the head is only increasing (Lemma~\ref{lem::headProgress}) the lemma holds.
-See Figure~\ref{fig::createBlock}.
-The right child is the same.
-\end{proof}
-
-\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
-  Suppose \texttt{Refresh(n)}'s \texttt{CAS(n.blocks[h], null, new)} returns \texttt{true}.
-Let $t$ be the time \texttt{Refresh(n)} is invoked, blocks propgatated to $Established(n.left, t)$ and $Established(n.right, t)$ are propagated to in $Established(n, t)$ after \texttt{CAS(n.blocks[h], null, new)}.
-\end{lemma}
-\begin{proof}
-By Lemma~\ref{lem::createBlock} \texttt{new} contains \texttt{n}'s childrens' established blocks before Line 43 which is  appended to \texttt{n.blocks} by \texttt{CAS} in Line 48.
-\end{proof}
-
-\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
-    If instance $r$ of \texttt{Refresh(n)} returns \texttt{false}, then there is another successful instance $r^\prime$ of \texttt{Refresh(n)} that has performed a successful \texttt{CAS(n.blocks[h], null, new)}(Line 49) after Line 43(\texttt{h= n.head}) of $r$.
-\end{lemma}
-\begin{proof}
-  If there is no other concurrent successful \texttt{Refresh(n)} then \texttt{Refresh(n)} would succeed in Line 48.
-So there is another \texttt{Refresh(n)}, that has to \texttt{CAS}ed successfully its block in \texttt{n.blocks[h]} after Line 43 of \texttt{Refresh(n)}.
-Otherwise the other \texttt{Refresh(n)} should have read \texttt{h$^\prime$>h} instead of \texttt{h} for \texttt{n.head}(Line 52).
-\end{proof}
-
-\begin{lemma}[Double Refresh] \label{doubleRefresh}
-  Consider two consecutive instances $r_1$,$r_2$ of \texttt{Refresh(n)} by the same process (Lines 35,36).
-Let be the time before $r_1$ invoked.
-After $r_2$'s \texttt{CAS} all the blocks propagated to $Established(n.left, t)$ and $Established(n.right, t)$ are in $Established(n, t)$.
-\end{lemma}
-\begin{proof}
-
-If Line 35 (first Refresh which we call $R_1$) returns \texttt{true}, the claim is held by Lemma~\ref{lem::trueRefresh}.
-If not, then there is another successful instance of \texttt{Refresh()} $R_1^\prime$ by Lemma~\ref{lem:falseRefresh}.
-$R_1^\prime$ may or may not have propagated some subblocks of \texttt{new} to \texttt{n}.
-It is obvious that the \texttt{new} constructed by the second Refresh in Line 36 contains the blocks in \texttt{new} by $R_1$ which $R_1^\prime$ did not contain, since \texttt{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}).
-If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds.
-If not, it is deduced that \texttt{n.blocks[h]} was not \texttt{null} before $R_2$'s \texttt{CAS}.
-Furthermore, \texttt{n.blocks[h]} was \texttt{null} before reading \texttt{h} by $R_1$.
-So there is a successful \texttt{Refresh()} after the read of \texttt{h} in $R_1$ and before the \texttt{CAS} of $R_2$.
-This \texttt{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds.
-See Figure~\ref{}.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
-  \caption{$R_2^\prime$'s \texttt{CAS} is executed after \texttt{h1=n.head}.}
-\end{figure}
-
- Block \texttt{new} is created of new established subblocks of children of \texttt{n}(Lemma \ref{lem::createBlock}, Line 46).
-If \texttt{CAS} in Line 48 succeeds then by Lemma~\ref{lem::trueRefresh} new established blocks will be in \texttt{n}.
-
-\begin{lemma}[Double Refresh] \label{doublyRefresh}
-All operations in \texttt{n}'s children's blocks before line \texttt{35} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{37}.
-\end{lemma}
-\begin{proof}
-Suppose block \texttt{b} with index \texttt{i} is in the the left child of \texttt{n} before the line 35.
-By Lemma~\ref{head} it follows that \texttt{n.left.head} is greater than \texttt{i}.
-\texttt{Refresh()} calls \texttt{CreateBlock()} and creates a block from blocks between \texttt{n.blocks[n.head].end\textsubscript{left}} and \texttt{n.left.head} in the left child, which contains \texttt{b} as well.
-First it tries to append it in \texttt{n.blocks.head} and if it was succsuful it continues recursivley.
-If not it tries again, and if the second call of \texttt{Refresh()} in Line 36 fails.
-It means there is another \texttt{Refresh} which has reah its \texttt{i} after the Line 35, so it contains \texttt{b} as well.
-\end{proof}
-\texttt{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block.
-If a \texttt{Refresh()} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence.
-So suppose two \texttt{Refresh}es fail.
-Since the first \texttt{Refresh()} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first \texttt{Refresh()}, was successful before the second \texttt{Refresh()}.
-So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh()} that assuredly has read block before the mentioned line \texttt{35}.
-After all it means if any of the \texttt{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
-
-\begin{lemma}[Append]\label{append}
- When \text{\texttt{Append(op)}} is finished, \texttt{op} appears exactly once in a block of \texttt{root.blocks}.
-\end{lemma}
-\begin{proof}
-\texttt{Append(op)} adds \texttt{op} to \texttt{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \texttt{Propagate()} recursively propagates \texttt{op} up to the root.
-By lemma \ref{doublyRefresh} we know that operation \texttt{op} propagates from child to parent at each level.
-\end{proof}
-
-\begin{lemma}[Block Size Upper Bound]\label{blockSize}
-Each block in a node contains at most one operation from each processs.  
-\end{lemma}
-\begin{proof}
-  Note that \texttt{prevIndex} and \texttt{lastIndex} defined in lines \ref{lastLine}, \ref{prevLine} are the indices defined in the Definition~\ref{def::subblock}.
-After a block has propagated to the parent the blocks between \texttt{prevIndex} and \texttt{lastIndex} make up to the parent (Lemma~\ref{doublyRefresh}).
-The number of new operations which have not propagated yet to the parent cannot be more than $p$.
-If so by the law of pigeonholes there is a process which has appended two concurrent operations.
-\end{proof}
-
-\begin{lemma}[Subblocks Upperbound]\label{subBlocksBound}
-Each block in a node  has at most $p$ subblocks.
-\end{lemma}
-\begin{proof}
-  From Line~\ref{addOP} it is induced directly that each block contains at least one operation. Now it follows directly by Lemma \ref{blockSize}.
-\end{proof}
-
-\begin{definition} [Ordering of operations inside a node] \label{ordering}
-$\blacktriangleright$ Note that from Lemma \ref{blockSize} we know there is at most one operation from each process in a given block.
-
-\begin{itemize}
-  \item $E(n,i)$ is the sequence of enqueue operations that are member of \texttt{n.blocks[i]} ordered by process id.
-  \item 
-$D(n,i)$ is the sequence of dequeue operations that are member of \texttt{n.blocks[i]} ordered by process id.
-\item $D(n)=D(n,1).D(n,2).D(n,3)...$
-\item $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
-\end{itemize}
-\end{definition}
-
-\begin{theorem}
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-  We show that the ordering $L$ stored in the root, satisfies the properties of a linearizable ordering.
-  \begin{enumerate}
-    \item If $op_1$ ends before $op_2$ begins in $E$, then $op_1$ comes before $op_2$ in $T$.\\$\blacktriangleright$ This is followed by Lemma \ref{append}.
-The time $op_1$ ends it is in root, before $op_2$, by Definition \ref{ordering} $op_1$ is before $op_2$.
-    \item Responses to operations in $E$ are same as they would be if done sequentially in order of $L$.
-\\$\blacktriangleright$ Enqueue operations do not have any response so it does no matter how they are ordered.
-It remains to prove  Dequeue $d$ returns the correct response according to the linearization order.
-By Lemma \ref{computeHead} it is deduced that the head of the queue at time of the linearization of $d$ is computed properly.
-If the Queue is not empty by Lemma \ref{get} we know that the returning response is the computed index element.
-  \end{enumerate} 
-\end{proof}
-
-\begin{lemma}[Get] \label{get}
-\texttt{Get(n,b,i)} returns $i$th Enqueue in $E(n,b)$.
-\end{lemma}
-\begin{proof}
-It is obvious that \texttt{Get(leaf l,b,1)} returns the operations stored in $b$th block of leaf $l$.
-To find the \texttt{i}th enqueue in block \texttt{b} of an internal node \texttt{n} in line 87 it is decided that it resides in the left child or the right child.
-This decision is made by Definition of $E(n,b)$.
-After that Lines 88, 92 search the proper subblocks of \texttt{b}.
-From Definition~\ref{def::subblock} weknow the subblocks of the $b$th block are within the \texttt{prevBlock} and \texttt{lastBlock} block of the \texttt{CreateBlock()}.
-\end{proof}
-
-\begin{lemma}[Index]
- Index(n,b,i) returns the rank in the $D(root)$ of ith Dequeue in $D(n,b)$.
-\end{lemma}
-\begin{proof}
-  \texttt{Index(n,b,i)} computes superblock of $i$th Dequeue in $b$th block of \texttt{n} in \texttt{n.parent} and then computes the order in $D(n.parent, superblock)$.
-Then calls \texttt{index()} on \texttt{n.parent} recursively.
-It is easy to see why the second is correct.
-Correctness of computing superblock comes from Lemma \ref{superBlock}.
-\end{proof}
-
-\begin{lemma}[Computing SuperBlock]\label{superBlock}
-  If \texttt{Index(n,b,i)} performs line 101, then \texttt{superblock} contains \texttt{i}th Dequeue in \texttt{b}th block of node \texttt{n}.
-\end{lemma}
-\begin{proof}
-\begin{enumerate}
- \item Value read for \texttt{super[b.time]} in line 101 is not null.\\$\blacktriangleright$
-  Values \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} are set before incrementing in lines 26,27.
-
- \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\\$\blacktriangleright$
-  Follows from the order of lines 60, 48, 49.
-
-% \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
-% \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
- \item \texttt{super[i+1]-super[i]}$\leq p$
-\\$\blacktriangleright$
- In a Refresh with successful CAS in line 46, \texttt{super} and \texttt{counter} are set for each child in lines 48,49.
-Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set.
-If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}.
-There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 49.
-So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
-
- \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
-\\$\blacktriangleright$
-\texttt{super[i]} is the index of the superblock of a block containing block b, followed by Lemma \ref{superCounter}.
-It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.
-\texttt{super(b)} is the real superblock of b.
-\texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}.
-If \texttt{b.time} is \texttt{t} we have:
-$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
-
-\end{enumerate}
-\end{proof}
-
-\begin{lemma}[Computing Queue's Head] \label{computeHead}
-  Let Q be state of the queue if the operations before ith Dequeue in $L(root)$ are applied on the Queue sequentially and X be the head of Q.
-If Q is empty \texttt{ComputeHead(i,b)} returns -1, otherwise
- returns index in $E(root,b)$ of X.
-  \end{lemma}
-
-\begin{lemma}[Validity of \texttt{head}]\label{head}
-No two blocks are written in the same index in \texttt{n.blocks}.
-\end{lemma}
-\begin{proof}
-  \texttt{head} is incremented in lines 51, 54 after trying to append a block to the index of the last \texttt{head} read.
-If it was successful, we have to do this, but if it was unsuccessful, it means it has appended to the index before, so we have to update the \texttt{head}.
-If a process dies before line 51, another process will increment \texttt{head} in line 54.
-\end{proof}
-
-\begin{lemma}[Validity of \texttt{super} and \texttt{counter}]\label{superCounter}
-If \texttt{super[i] $\neq$ null}, then \texttt{super[i]} in node \texttt{n} is the superblock of a block with \texttt{time=i}.
-\end{lemma}
-\begin{proof}
-After a successful CAS in line 46 \texttt{super} and \texttt{counter} are modified in both children.
-\texttt{super[i]} is supposed to be the superblock of a block with \texttt{time}=i and \texttt{counter} is the timer in each node.
-\texttt{super[i]} and \texttt{counter=i} are expected to update after a bunch of blocks with \texttt{time=i} have been aggregated together into a block in the parent.
-If the process dies before line 48 these values remain unchanged and incoming blocks will get the same \texttt{time}.
-We claim that our algorithm still works since at most $p$ processes die and it will not change our complexity.
-If a process dies right after line 48, then \texttt{counter} will remain the same and \texttt{super[i]} is correct.
-Furthermore we are sure that when \texttt{super[i]} is read it will not be \texttt{null}.
-\end{proof}
-
-\begin{lemma}[Search Ranges]\label{search}
-  Preconditions of all invocation of \texttt{BSearch} are satisfied.
-\end{lemma}
-\begin{proof}
-
-Line 83: \texttt{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
-
-Line 88: This search tries to find the ith enqueue, knowing that it is in the left child.
-Search is done over the left subblocks.
-The start and end of the range are followed by definition.
-Line 92 is the same.
-
-Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \texttt{super[i]} is at most $p$, since at most $p$ processes could die.
-
-\end{proof}
-
-\paragraph{Overall}
-
-The main difference between the main algorithm and the block tree is that we separate enqueues and dequeues, compute the number of non-null dequeues and the Queue's size in each block in the root.
-We emphasize that these changes work correctly; every other claim in Block trees applies here.
-
-\section{Make first level search fast}
-\begin{algorithm}
-\caption{Modified Queue \label{algQx}}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-%\Statex \textbf{Structure}
-
-%\Statex $\blacktriangleright$ \texttt{extends Queue Algorithm}
-
-%\Statex $\blacktriangleright$\texttt{\textsl{RBTNode} RBTRoot}  \Statex
-%  \textsf{pointer to an RBTNode, that is root of the RBTree}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{PRBTree[RBTNode]} rootBlocks}
-  \Statex \textsf{A persistant redblack tree supporting \texttt{append(n),get(i),split(j)}}. \texttt{append(n)} returns \texttt{true} in case successful.
-
-\Statex $\blacktriangleright$ \texttt{\textsl{RBTNode}}
-\begin{itemize}
-
-  \item \texttt{\textsl{RootBlock} block}
-
-  \item \texttt{\textsl{int} age}
-
-  \textsf{number of finished operations in the block}
-
-  \item \texttt{\textsl{int} step}
-
-  \textsf{If this node is the root of the RBTree, then step shows the number of appended nodes to the RBTree}
-
-%  \item \texttt{some RBTRee attributes like left,right children}
-
-\end{itemize}
-
-\Statex $\blacktriangleright$ \texttt{\textsl{leaf} extends Node}
-\begin{itemize}
-  \item \texttt{\textsl{int[]} response}
-
-  \textsf{leaf.response[i] stores response of leaf.ops[i]}
-
-  \item \texttt{\textsl{int} maxOld}
-
-  \textsf{Index of the youngest old block in the root that this process has seen yet.}
-
-\end{itemize}
-
-%\Statex $\blacktriangleright$ \texttt{\textsl{RBT(RBTnode n)}}
-%  \Statex \textsf{Gives us RBTree  persistentoperations if n is root of a RBTRee, for example RBT(n).append(x) appends x if to the redblack tree with n as root.
-A persistent append is an append which only contains the changed nodes.}
-
-\Statex
-
-\Function{boolean}{Refresh}{\textsl{node} n} \Comment if n is root
-\State \texttt{new=CreateBlock()} \Comment TODO
-\If{\texttt{new.num==0}} \Return{\texttt{true}}
-\ElsIf{\Call{RBTAppend}{new}}
-\Statex..\EndIf
-\EndFunction{Refresh}
-
-\Statex
-
-\Function{<int, int>}{Index}{\textsl{node} n, \textsl{int} b, \textsl{int} i} \Comment{Returns the order in the root among dequeus, of ith dequeue in bth block of node n}
-\If{\texttt{n {\keywordfont is} root}} \Return \texttt{search in the current RBTRee}
-\Else
-..
-\EndIf
-\EndFunction{Index}
-
-\Statex
-
-\Function{Object}{Dequeue()}{}
-\State ..
-\State \texttt{N\textsubscript{i}= RBTnode n that n.block contains the invocation of Dequeue}
-\State \texttt{N\textsubscript{r}= RBTnode n that n.block contains the response of Dequeue}
-\State \texttt{N\textsubscript{i}.age= N\textsubscript{i}.age+1} \Comment{this is a shared counter}
-\State \texttt{N\textsubscript{r}.age= N\textsubscript{r}.age+1}
-\If{\texttt{N\textsubscript{i} or N\textsubscript{r} become old}} \texttt{update maxOld}
-\EndIf
-\EndFunction{Dequeue}
-
-\pagebreak
-%\Function{void}{Update(Root Block B)}{}
-%\If{\texttt{B.todo==0}}
-%\If{\texttt{B.left.old-subtree} and \texttt{B.right.old-subtree}}
-%\State \texttt{B.old-subtree= True}
-%\If{\texttt{B.parent!=root}}
-%\State \texttt{Update(B.parent)}
-%\EndIf\EndIf
-%\EndIf
-%\EndFunction{Update}
-%\Statex
-
-\Function{void}{RBTAppend}{block b} \Comment{\textsf{adds block b to the \texttt{rootBlocks}}}
-\State \texttt{root= rootBlocks.root}
-\If{\texttt{root.step\%$p^2$==0}}
-\State \texttt{Help()}
-\State \texttt{CollectGarbage()}
-\EndIf
-\State \texttt{new= RBTnode(b,0,null)}
-%\State \texttt{r= pointer to root of the RBT(h).append(n)}
-%\State \texttt{*r.step= h.step+1}
-%\State \texttt{RBTRoot.CAS(h,r)}
-\State \Return \texttt{rootBlocks.append(new)}
-\EndFunction{RBTAppend}
-\Statex
-
-\Function{void}{Help}{}\Comment{Helps pending operations}
-\State{last= l.head-1}
-\For{\texttt{leaf l in leaves}}
-\If{\texttt{l.blocks[last] is not null}}
-\If{op \textbf{is} DEQ}
-\State\texttt{run \texttt{Dequeue()}  for \texttt{l.ops[last]} after Propagate()}
-\State \texttt{write the response to l.responses[last]}
-\EndIf
-\EndIf
-\EndFor
-\EndFunction{Help}
-\Statex
-
-\Function{void}{CollectGrabage}{}\Comment{Collects the old root blocks.}
-\State \texttt{l=FindYoungestOld(Root.Blocks.root)}
-\State \texttt{t1,t2= RBT.split(l)}
-\State \texttt{RBTRoot.CAS(t2.root)}
-\EndFunction{CollectGrabage}
-
-\Statex
-
-\Function{Block}{FindYoungestOld}{b}
-\State\Return  \texttt{read all maxOld values among leaves and decide the largest one} \Comment{There is no need to do a sasDK;Lnapgshot.}
-\EndFunction{findYoungestOld}
-
-\Statex
-
-\Function{response}{FallBack}{op i}
-
-\If{operation i in leaf l cannot find its desired RootBlock}
-
-\State \Return \texttt{l.response[i]}
-\EndIf
-
-\EndFunction{FallBack}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
+\paragraph{}
+Now that we know how to augment Block Tree root history to compute the response, we will present the complete algorithm.
+
+\section{Time Complexity}
 

### 2022-09-29 backup/jul12.tex
fix citations, review index deq

--- 2022-09-29 backup/jul 14.tex
@@
 \algdef{S}[FUNCTION]{Function}
-   [3]{{\texttt{\textsl{#1}}} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}
+   [3]{{\keywordfont #1} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}
 
@@
 \begin{itemize}
-%\item \texttt{\textsl{string} name}
+\item \texttt{\textsl{string} name}
 \item \texttt{\textsl{int} pid}
-\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}'s array}}
+\item \texttt{\textsl{int} loc \textsf{: location in leaf's array}}
 \end{itemize}
@@
 \begin{itemize}
-\item \texttt{\textsl{*node} parent}\textsf{: pointer to parent in Tree}
-\item \texttt{\textsl{operation[]} ops}\textsf{: ops invoked by process i}
-\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{ops}}
+\item \texttt{\textsl{*node} parent} \textsf{: pointer to parent in Tree}
+\item \texttt{\textsl{operation[]} ops} \textsf{: ops invoked by process i}
+\item \texttt{\textsl{int} last= 1} \textsf{: index of first empty cell of \texttt{ops}}
 \end{itemize}
@@
 \item \texttt{\textsl{*node} left, right, parent}
-\item \texttt{\textsl{block[]} blocks}\textsf{: blocks propagated to \texttt{n}}
-\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{blocks}}
-\item \texttt{\textsl{int} counter= 0}\textsf{: \#groups of blocks propagated from \texttt{n}}
-\item \texttt{\textsl{int[]} super}\textsf{: approximate index of superblock of a block in the given group in \texttt{n}'s parent}
+\item \texttt{\textsl{block[]} blocks} \textsf{: blocks propagated to \texttt{n}}
+\item \texttt{\textsl{int} last= 1} \textsf{: index of first empty cell of \texttt{blocks}}
 \end{itemize}
@@
 \begin{itemize}
-  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}}\textsf{: \#operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
-  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}\textsf{: index of \texttt{b}'s last subblock}
-  \item \texttt{\textsl{int} group}\textsf{: approximate group number}
+  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}} \textsf{:~\#~operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
+  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}} \textsf{: index of \texttt{b}'s last subblock}
 \end{itemize}
@@
 \Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
-\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}
+\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: ending right subblock of b position}
 
@@
 \State \Call{Propagate}{l\textsubscript{pid}.parent}
-\State \Return \Call{Index}{l\textsubscript{pid}, op.location, b} \Comment we don't know the block eachoperation resides in leaf
+\State \Return \Call{Index}{l\textsubscript{pid}, op.location}
 \EndFunction{Append}
@@
 
+\Function{element}{Get}{\textsl{int} i} 
+\State \texttt{res=\Call{BSearch}{root, sum, i, 0, root.last}}
+\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
+\EndFunction{Get}
+
+\Statex
+
+%\Function{index}{Index}{operation op, int pid}
+%\State \texttt{\textsl{int} i= op's order in l\textsubscript{pid}.ops}
+%\State \Return \Call{Index}{l\textsubscript{p}, i}
+%\EndFunction{Index}
+
+%\Statex
+
 \Function{void}{Propagate}{\textsl{node} n}
 \If{\texttt{n {\keywordfont is} root}} \Return
-\ElsIf{\texttt{not} \Call{Refresh}{n}}
-\State \Call{Refresh}{n}
+\Else
+\State \texttt{i= n.last}
+\State \texttt{\textsl{block} new=\Call{CreateBlock}{n, i+1}}
+\If{\Call{CAS}{n.blocks[i+1], null, new}}
+\State \texttt{\Call{CAS}{n.last, i, i+1}}
+\Else
+\State \texttt{i= n.last}
+\State \texttt{\textsl{block} new=\Call{CreateBlock}{n, i+1}}
+\If{\Call{CAS}{n.blocks[i+1], null, new}}
+\State \texttt{\Call{CAS}{n.last, i, i+1}}
+\EndIf
+\EndIf
 \EndIf
@@
 
-\Function{boolean}{Refresh}{\textsl{node} n}
-\State \texttt{i= n.last}
-\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, n.counter}}
-\If{\Call{CAS}{n.blocks[i+1], null, new}}
-\State \texttt{\Call{CAS}{n.last, i, i+1}}
-\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
-\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
-\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
-\EndFor
-\State \Return{\texttt{true}}
-\Else ~\Return{ \texttt{false}}
-\EndIf
-\EndFunction{Index}
-
-\Statex
-
-\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int g} 
+\Function{block}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
 \Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
 \State \texttt{block b= \Call{new}{\textsl{block}}}
-\If{\texttt{n is leaf}}\Comment{TODO}
-\Else
-\State \texttt{b.group= g}
 \ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
+\State \texttt{prev= n.blocks[i-1].ending\textsubscript{dir}}
 \State \texttt{j= n.dir.last}
-\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
 \State \texttt{end= n.dir.blocks[j]}
-\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].group}
 \State \texttt{b.end\textsubscript{dir}= j}
@@
 \State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
-\EndIf
-\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
+\State \Return \texttt{b}
 \EndFunction{CreateBlock}
@@
 \begin{algorithmic}[1]
-
-\Function{element}{Get}{\textsl{int} i} 
-\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.last}}
-\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
-\EndFunction{Get}
 
@@
 
-\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
+\Function{element}{Get}{node n, int b, int i} 
 \Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}
 
-\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
+\If{\texttt{i $\leq$ b.num\textsubscript{left}+1}} \Comment{\texttt{i} exists in left child of \texttt{n}}
 \If{\texttt{n.left {\keywordfont is} leaf}} \Return \texttt{n.left.ops[i]}
@@
 \Else
-\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
+\State \texttt{i= i-n.blocks[b].sum\textsubscript{left}}
 \If{\texttt{n.right {\keywordfont is} leaf}} \Return \texttt{n.right.ops[i]}
 \Else 
-\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
+\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].ending\textsubscript{right}, n.blocks[b].ending\textsubscript{right}}}
 \State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} \EndIf
@@
 
-\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns order in  the root of \texttt{i}th operation in the node \texttt{n} which exists in block b}
+\Function{index}{Index}{node n, int i} \Comment{Returns order in root of \texttt{i}th operation in node \texttt{n}}
 \If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
 \Else
-\State \texttt{dir= (n.parent.left==n)? left: right}
-\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].group]-p, sup[n.blocks[b].group]+p}}
+\State \texttt{nBlock= \Call{BSearch}{n, n.sum, i, 0, n.parent.last}}
+\Comment{\textmd{This search takes O(log n) steps}}
+\State \texttt{dir= (n.parent.left==n)?left:right}
+\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, 0, n.parent.last}}
+\Comment{\textmd{This search takes O(log n) steps}}
 \If{\texttt{dir {\keywordfont is} left}}
-\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
-\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
+\State \texttt{i+= n.parent.blocks[superBlock-1].sum + n.blocks[nBlock-1].sum}
+\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
 \EndIf
-\State \Return\Call{Index}{n.parent, i, superBlock}
+\State \Return\Call{Index}{n.parent,}
 \EndIf

### 2022-09-29 thesis/thesis.tex
fix citations, review index deq

@@
 \end{center}
-\caption{ّ\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored.
+\caption{\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored.
 Each node tries to propagate its operations up to the root, which stores the total ordering of all operations.}
@@
 
+ Let $i$ be the value $R_n$, a successful instance of \nf{Refresh()} on node \nf{n} reads from \nf{$n$.head}.
+$R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}.
+Let $p$ be $n$\nf{}.parent.
+Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$.
+From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating $b$.
+Let $R_p$ be the first successful \nf{p.Refresh()} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
+Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.
+
+Since the index of the superblock of ${b}$ is not known until $b$ is propagated, $R_n$ cannot set the \nf{super} field of ${b}$ while creating it.
+One approach is to set the \nf{super} field of ${b}$ by $R_n$ after propagating $b$ to $p$.
+This solution would not be efficient because there might be $p$ subblocks in the block $R_p$ propagated needing to update the \nf{super} field.
+However intuitively, once $b$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
+One idea is that if we know the approximate position of the superblock of $b$ then we can search for the real superblock when we wished to know the superblock of $b$ i.e.
+\nf{b.super} does not have to be the exact location of  the superblock of $b$, but we want it to be close to \nf{j}.
+We can set \nf{$b$.super} to \nf{$n$.parent.head} while creating $b$, but the problem is that there might be many \nf{$p$.Refresh}es that could happen after reading $p$\nf{.head} by $R_n$ and before propagating $b$ to $p$.
+If we set \nf{$b$.super} to \nf{$p$.head} after appending $b$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $b$ and before setting \nf{$b$.super}.
+In this case the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$b$.super}.
+
+Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}.
+After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh()} by the same process or another process or  maybe an \nf{$n$.parent.Refresh()}.
+We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates $b$.
+
 %\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
-\paragraph{}
+\paragraph{Queue from tree}
 Now, we describe how to use the tree to implement a queue.
@@
 Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}.
+Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value $b$$^\prime$ by Invariant \ref{lem::headPosition}.
 Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
@@
 Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}.
-The set of operations of block \nf{b} is the union of the operations in leaf subblocks of $b$.
+The set of operations of block $b$ is the union of the operations in leaf subblocks of $b$.
 We denote the set of operations of block $b$ by ${ops(b)}$ and the union  of operations of a set of blocks $B$ by $ops(B)$.
@@
 
-\subsection{Correctness of GetEnqueue and IndexDequeue}
-
-Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()} and \nf{IndexDequeue()}.
+\subsection{Correctness of GetEnqueue}
+
+Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()}.
 
@@
 So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
+Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
 By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
@@
 
-\pagebreak
-
-\begin{definition}\label{orderRefresh}  A Refresh is \it{successful} if it performs a successful \nf{CAS} on Line \ref{cas}.
-If \nf{Refresh} instance \nf{R\sub{1}} does its \nf{CAS} at Line \ref{cas} sooner than \nf{Refresh} instance \nf{R\sub{2}} we say \nf{R\sub{1}}  \it{happened} before \nf{R\sub{2}}.
+\clearpage
+
+\subsection{Correctness of IndexDequeue}
+\begin{definition}\label{orderRefresh} If \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
 \end{definition}
 
- Let \nf{i} be the value \nf{R\sub{n}}, a successful instance of \nf{Refresh()} on the node \nf{n}, reads from \nf{n.head}.
-\nf{R\sub{n}} does a successful \nf{CAS(null, new)} into \nf{n.blocks[h]}, where \nf{b} is the \nf{new}.
-Without loss of generality for the rest of this section assume \nf{n} is the left child of \nf{n.parent}.
-From now on we say \nf{p} as an abbreviation for \nf{n.parent}.
-Let \nf{R\sub{p}} be the first successful \nf{p.Refresh()} that reads some value greater than \nf{i} for \nf{left.head} and is contains \nf{b} in its created block \nf{s} in Line \ref{invokeCreateBlock}.
-From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating \nf{b}.
-\nf{R\sub{p}} does a successful \nf{CAS(null, new)} into \nf{p.blocks[j]}, where \nf{s} is the \nf{new}.
-
-Although other fields of \nf{b} are set while creating it, because the index of the superblock of \nf{b} is not known until it is propagated, \nf{R\sub{n}} cannot set the \nf{super} field of a \nf{b} while creating it.
-One approach is to set the \nf{super} field of \nf{b} after it is propagated by \nf{R\sub{b}} but this would not be efficient because there might be $p$ subblocks in \nf{s}.
-However, once \nf{b} is installed, its superblock is going to be close to \nf{n.parent.head} at the time of installation.
-One idea is that if we know the aproximate position of the superblock of \nf{b} then we can search for the real superblock when we wished to know the superblock of \nf{b} i.e.
-\nf{b.super} does not have to be the exact location of  the superblock of \nf{b}, but we want it to be close to \nf{j}.
-We can set \nf{b.super} to \nf{n.parent.head} while creating \nf{b}, but the problem is that there might be many \nf{p.Refresh}es could happen that contain blocks from the right child of \nf{p} and \nf{j} could be arbitrarily (right word?) greater than \nf{b.super}.
-We set \nf{b.super} to \nf{p.head} after appending \nf{b} to \nf{n.blocks} (Line \ref{setSuper1}).
-Maybe \nf{R\sub{n}} goes to sleep at some time after installing \nf{b} and before setting \nf{b.super}.
-In this case the next \nf{Refresh}es on \nf{n} and \nf{n.parent} help fill in the value of \nf{b.super}.
-
-Block \nf{b} is appended to \nf{n.blocks[h]} on Line \ref{cas}.
-After appending \nf{b}, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} by the same process or another process's \nf{n.Refresh()} or  maybe an \nf{n.parent.Refresh()}.
-We want to bound how far \nf{b.super} is from the index of \nf{b}'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates \nf{b}.
-
 \begin{observation} \label{setSupBeforeIncHead}
-After \nf{n.blocks[i].CAS(null, b)} succeeds, \nf{n.head} cannot increase from \nf{i} to \nf{i+1} unless \nf{b.super} is set.
+After \nf{$n$.blocks[$i$].CAS(null, $b$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ unless \nf{$b$.super} is set.
 \end{observation}
@@
 \end{corollary}
-\begin{proof}If \nf{b.super} is \nf{null} then \nf{n.head} cannot advance so the next \nf{n.Refresh()} will fail.
-By the previous corollary, \nf{b.super} has to be set before the next successful Refresh() on n after \nf{R\sub{n}}.
-\end{proof}
-Now let us talk about how the \nf{p.Refresh}es that took place after the putting \nf{b} into \nf{n}, will help to set \nf{b.super} and propagate \nf{b}.
+\begin{proof}
+By Lemma \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
+\end{proof}
+Now let us talk about how the \nf{$p$.Refresh}es that took place after the putting $b$ into \nf{n}, will help to set \nf{b.super} and propagate $b$.
 
 \begin{lemma}
-  If $\nf{b}\in\nf{n.parent.blocks[i]}$ then $\nf{b.super}\leq\nf{i}$.
-\end{lemma}
-\begin{proof}
-For \nf{R\sub{p}} to contain block \nf{b}, it has to read \nf{n.head} greater than \nf{h} (see Line \ref{lastLine}).
-For \nf{n.head} to be greater than \nf{h} it means \nf{n.head} is incremented in Line \ref{incrementHead} which means \nf{b.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}).
-So if \nf{R\sub{p}} propagates \nf{b} it means \nf{b.super} was already set.
-Let \nf{j} be the value written in \nf{b.super}.
-\nf{j} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{b.super}.
-From Observation \ref{nonDecreasingHead} we know \nf{p.Head} is non-decreasing so $\nf{j}\leq\nf{i}$.
-The reader may wonder when the case $\nf{j}=\nf{i}$ happens, it happens when \nf{p.blocks[j]=null} while \nf{j} is read and \nf{R\sub{p}} puts its created block into \nf{n.blocks[j]}.
+  If $b$ is a direct subblock of \nf{$n$.parent.blocks[$i$]} then $\nf{$b$.super}\leq i$.
+\end{lemma}
+\begin{proof}
+Instance $R_p$ of $n$\nf{.parent.Refresh()} contains block $b$, if it reads \nf{$n$.head} greater than ${h}$, the index of $b$ in $n$ (see Line \ref{lastLine}).
+For \nf{$n$.head} to be greater than ${h}$ it means \nf{$n$.head} is incremented in Line \ref{incrementHead} which means \nf{$b$.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}).
+So if $R_p$ propagates $b$ it means \nf{$b$.super} was already set.
+The value written in \nf{b.super} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{$b$.super}.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $b\nf{.super}\leq\nf{i}$.
+The reader may wonder when the case $b\nf{.super}=\nf{i}$ happens, it happens when $\nf{$n$.parent.blocks[$b$.super]}=\nf{null}$ while $b$\nf{.super} is read and $R_p$ puts its created block into \nf{n.blocks[$b$\nf{.super}]}.
 \end{proof}
@@
 \begin{lemma}
-If \nf{R\sub=n.Refresh()} puts \nf{b} in \nf{n.blocks[h]} at Line \ref{cas}, then the block created by one of the next two successful \nf{p.Refresh}es according to the Definition \ref{orderRefresh} contains \nf{b} and \nf{b.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{p.Refresh()}.
-\end{lemma}
-
-\begin{proof}
-It is sufficient to prove one of the two successful \nf{p.Refresh()}es propagate \nf{b}.
-If the first successful \nf{p.Refresh()} propagated \nf{b} then the claim is true, so in the remaining part we assume the first \nf{p.Refresh()} did not propagate \nf{b} and prove the second \nf{p.Refresh()} propagates \nf{b}.
-
-\nf{b.super} is set by some instance of \nf{Refresh()} on \nf{n} or \nf{p} showed by \nf{R'} and \nf{n.head} is incremented by some \nf{Refresh()} called \nf{R"}.
-We want to know how great $\nf{j}-\nf{b.super}$ can be.
-\nf{p.head} is \nf{hp} when \nf{R'} reads it.
-From Lemma \ref{lem::headProgress} \nf{p.head} could only increase from \nf{hp} to \nf{hp+1} if $\nf{p[hp]}\neq\nf{null}$.
-In other words there should be a successful \nf{p.Refresh()} for \nf{p.head} to increase.
-We claim there cannot be another successful \nf{p.Refresh()} after \nf{R'} reads \nf{p.head} and before \nf{R\sub{p}} performs Line \ref{lastLine}.
+If $R_n=$\nf{Refresh()} puts $b$ in \nf{$n$.blocks[$h$]} at Line \ref{cas}, then the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to the Definition \ref{orderRefresh} contains $b$ and \nf{$b$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh()}.
+\end{lemma}
+
+\begin{proof}
+It is sufficient to prove one of the two successful \nf{$n$.parent.Refresh()}es propagates $b$.
+If the first successful \nf{p.Refresh()} propagated $b$ then the claim is true.
+We prove if the first \nf{$n$.parent.Refresh()} did not propagate $b$, second \nf{$n$.parent.Refresh()} propagates $b$.
+
+\nf{$b$.super} is set by some instance of \nf{Refresh()} on $n$ or \nf{$n$.parent} showed by $R^\prime$ and \nf{$n$.head} is incremented by some \nf{Refresh()} called $R^\prime\prime$.
+Let \nf{$n$.parent.head} value be $hp$ when $R^\prime$ reads it.
+From Lemma \ref{lem::headProgress} \nf{$n$.parent.head} could only increase from ${hp}$ to ${hp+1}$ if $\nf{$n$.parent.blocks[$hp$]}\neq\nf{null}$.
+In other words there should be a successful \nf{$n$.parent.Refresh()} for \nf{$n$.parent.head} to be increased.
+Let $R_p$ be \nf{$n$.parent.Refresh} that propagated $b$ to \nf{$n$.parent}.
+We claim there cannot be another successful \nf{$n$.parent.Refresh()} after $R^\prime$ reads \nf{$n$.parent.head} and before $R_p$ performs Line \ref{lastLine}.
 
@@
 
-Assume the first successful \nf{p.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{b.super}.
-It might happen maybe because the value read for $h_{left}$ in Line \ref{readChildHead} is less than \nf{i} or maybe $\nf{i}=\nf{h\sub{left}}$ and $\nf{left.blocks[h\sub{left}]}=\nf{null}$, which means \nf{n.head} is advanced but \nf{b} is still not installed in \nf{n.blocks[i]} which means \nf{R\sub{n}} has not reached to the Line \ref{cas}.
-
-Let the first successful \nf{p.Refresh()} be $R_{p1}$ and the second next successful \nf{p.Refresh()} be $R_{p2}$.
-If $R_{p1}$ reads \nf{x} in Line \ref{readHead}, then $R_{p2}$ has to read \nf{x+1} in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
-See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$  on ${p}$.
+Without loss of generality  assume $n$ is the left child of \nf{$n$.parent}.
+As we mentioned the first successful \nf{$n$.parent.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{$b$.super}.
+It might happen maybe because the value $h_{left}$ read for $n$\nf{.head} in Line \ref{readChildHead} is less than or equal to $i$ and $\nf{left.blocks[$h_{left}$]}=\nf{null}$, which means \nf{$n$.head} is advanced but $b$ is still not installed in \nf{$n$.blocks[$i$]} which means $R_n$ has not reached  the Line \ref{cas}.
+
+Let the first successful \nf{$n$.parent.Refresh()} be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh()} be $R_{p2}$.
+If $R_{p1}$ reads $x$ in Line \ref{readHead}, then $R_{p2}$ has to read ${x+1}$ in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
+See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$ on \nf{$n$.parent}.
 
@@
   \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
-  \caption{Rp2 reads p.head after $t_{321}^{Rp1}$, which is after $t_{321}^{Rn}$. Rp2 has to help increment n.head and set b.super.}
+  \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
   \label{fig::timeLine2}
@@
 $t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code.
-The second Refresh on \nf{p} contains \nf{b} and reads $\nf{n.head}>\nf{i}$.
-
-So \nf{b.super} has set by some process before the second next successful \nf{p.Refresh()} on Line \ref{setSuper1}.
-Since \nf{i} is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value fo \nf{left.head} greater than \nf{h} and propagates \nf{b} to \nf{p}.
-So if \nf{b} was not propagated already we are sure the second next successful \nf{p.Refresh()} propagates \nf{b}.
+The second Refresh on \nf{p} contains $b$ and reads $\nf{n.head}>\nf{i}$.
+
+So \nf{$b$.super} has set by some process before the second next successful \nf{$n$.parent.Refresh()} on Line \ref{setSuper1}.
+Since $i$ is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value for \nf{left.head} greater than ${h}$ and propagates $b$ to \nf{$n$.parent}.
+So if $b$ was not propagated already we are sure the second next successful \nf{$n$.parent.Refresh()} propagates $b$.
 \end{proof}
@@
 \begin{corollary} \label{superRelation}
-If \nf{$n$.blocks[$b$]} is propagated to $n.parent$, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
+If \nf{$n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
 \end{corollary}
@@
 After that the order of $D_i(n,b)$ in $D(n.parent, superblock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-Note that by Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
+By Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
 Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
@@
 See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds \nf{b} also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
+After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
 Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
@@
 
+\section*{References}
+\bibliography{main.bib} 
+\bibliographystyle{plain}
+
 \end{document}

### 2022-09-30 thesis/thesis.tex
apply sept 29 notes

@@
 For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance} whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
+For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance}.
+Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
 \end{proof}
@@
 By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the ranges $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ and $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$j$].end\sub{left}}]$ cannot overlap.
-Therefore, direct subblocks of $n.blocks[i]$ and $n.blocks[j]$ cannot overlap.
+Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap.
 \end{proof}
@@
 \begin{definition}
-Let $t^{op}$ be the time $op$ is invoked and $^{op}t$ be the time $op$ terminates.
-  Let $t_{l}$ be the time immediately before executing Line $l$ and $_{l}t$ be the time immediately after executing Line $l$.
-Let $^{op}_{l}t$ is the time immediately before running Line $l$ of operation $op$ and   $t_{l}^{op}$ is the immediate time after running Line $l$ of operation $op$.
+Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
+We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear in the context.
 In the text $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
@@
 
-In the next two results we show for every successful \nf{Refresh()}, all the operations established  in the children before the \nf{Refresh} are established in the parent after the \nf{Refresh}'s succesful \nf{CAS} at Line \ref{cas}.
+In the next two results we show for every successful \nf{Refresh()}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
 
 \begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n.blocks_{\ref{cas}})$.
+If $R$ is a successful instance of \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
 \end{lemma}
@@
 We show 
-\begin{align*}
+\vspace{-2em}\begin{align*}
 EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \\
-&\subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]}).
+&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
 \end{align*}
 
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$.
-Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[$1\cdots n.head_{\ref{readHead}}$]}.
-Because of Lemma \ref{nonDecreasingHead} we have $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{cas}}$.
+Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
+Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
+Because of Lemma \ref{nonDecreasingHead} we have $n\nf{.left.head}_{309}-1<n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}<n\nf{.head}_{\ref{cas}}$.
 From Observation \ref{lem::blocksOrder} the claim follows.
@@
 
-\begin{figure}[hbt]
-  \center\includegraphics[width=5in]{pics/trueRefresh}
-  \caption{TODO :: Is this picture good or should I remove it? New established operations of the left child are in the new block.(TO UPDATE)}
-\end{figure}
+% \begin{figure}[hbt]
+%   \center\includegraphics[width=5in]{pics/trueRefresh}
+%   \caption{TODO :: Is this picture good or should I remove it? New established operations of the left child are in the new block.(TO UPDATE)}
+% \end{figure}
 
@@
 \begin{proof}
-The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when \nf{R} terminates the established blocks in $n$ are a superset of $n$\nf{.blocks} after Line \ref{cas}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n$\nf{.head} is incremented after line \ref{incrementHead}.
-So the block \nf{new} appended to \nf{n} at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-\pagebreak
-
-In the next lemma we show that if two consecutive instances of \nf{Refresh()} by the same process on node $n$ fail, then the established blocks in the children of $n$ before the first \nf{Refresh()} are guaranteed to be in $n$ after the second \nf{Refresh()}.
+The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
+Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
+Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
+So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
+\end{proof}
+
+\pagebreak
+
+In the next lemma we show that if two consecutive instances of \nf{Refresh()} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh()} are guaranteed to be in $n$ after the second \nf{Refresh()}.
 
@@
   Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by process $p$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh()}, then we have $EST_{n\nf{.left}}^{t^R{_1}} \; \cup \; EST_{n\nf{.right}}^{t^R{_1}} \subseteq EST_{n}^{^R{_2}t}$.
+If neither $R_1$ nor $R_2$ is a successful \nf{Refresh()}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
 \end{lemma}
@@
 Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-Note that by Lemma \ref{lem::headInc} $R_1$ and $R_2$ both cannot read the same value $i$.
-By Observation \ref{nonDecreasingHead} $R_2$ reads larger value of $n$\nf{.head} than $R_1$.
+By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ both cannot read the same value $i$.
+By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
 
 Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{n.Refresh()} that has done \nf{CAS} successfully into \nf{n.blocks[i+1]} before $R_2$ tries to \nf{CAS}.
+As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh()} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
 $R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
@@
 $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$ the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{n.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation \ref{lem::blocksOrder} $ops(EST_{n.\nf{left}, t^{R_1}}) \cup ops(EST_{n.\nf{right}, t^{R_1}}) \subseteq ops(n.\nf{blocks}_{t^{R_2}})$
+Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
+To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n.\nf{left}}^{^{R_2}t}$.
 
@@
 So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^R_{2}t$.
-From Observation \ref{lem::blocksOrder} on the operations before and after $R_3$'s \nf{CAS} and Lemma \ref{lem::trueRefresh} on $R_3$ the claim holds.
+From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh}, \ref{lem::headInc} on $R_3$ the claim holds.
 \end{proof}
@@
   \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{n.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{n.head} from $i+1$ to $i+2$ $<t_{R_2}$}
+  \caption{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
 \end{figure}
@@
 \begin{corollary} \label{lem::propagateStep}
-  \nf{ops(EST\sub{n.left, \sub{\ref{firstRefresh}}t})} $\cup$ \nf{ops(EST\sub{n.right, \sub{\ref{firstRefresh}}t})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{\ref{secondRefresh}})}}
+  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
 \end{corollary}
@@
 
-Now we show that after \nf{Append($b$)} on a leaf finishes, $b$ will be established in \nf{root}.
+Now we show that after \nf{Append($b$)} on a leaf finishes, the operation contained in $b$ will be established in \nf{root}.
 
 \begin{corollary}\label{lem::append}
-For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{n}, t^A})$ where $n\in\{\text{nodes in the path from the leaf to the root}\}$.
+For $A=l\nf{.Append(}b\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
 \end{corollary}
@@
 $A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$.
-For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} utill $n$ is \nf{root}.
+For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} untill $n$ is \nf{root}.
 \end{proof}
@@
 \begin{corollary}\label{lem::appendExactlyOnce}
- After \nf{Append($b$)} finishes, $b$ is in the nodes the path from the leaf to the root for exactly one time.
+ After $l$\nf{.Append($b$)} finishes, $b$ is subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
 \end{corollary}

### 2022-09-30 thesis/thesis.tex
add lemma sum enq

@@
 Starting from the root, \textsc{GetIndex}($i$) BinarySearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
-From lemma $\ref{lem:block_size}$ we know a block size is at most $p$.
+From lemma $\ref{blockSize}$ we know a block size is at most $p$.
 So BinarySearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
@@
 So now the problem is to find the result of each Dequeue.
-From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
+From lemma \ref{blockSize} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
 In the next example, operations in a cell are concurrent.
@@
 \end{definition}
+
+The next two observations are correct for both enqueues and deqeueues.
+
+\begin{observation}
+  $n\nf{.blocks[}b\nf{]}\nf{.num\sub{enq}}$ is the size of $E(n,b)$.
+\end{observation}
+
+\begin{observation}\label{lem::sumDeq}
+$\sum_{i=1}^{b} {E(n,i)}=n\nf{.blocks[}i\nf{].sum\sub{enq}}$.
+\end{observation}
+
+The next observation is correct for both directions.
+
+\begin{observation}\label{lem::numEnqLeft}
+Let $B$ be $n\nf{.blocks[}b\nf{]}$. $B\nf{.num\sub{enq-left}}$ is the number of \nf{Enqueue} operations in the direct subblocks of $B$ in $n$\nf{.left}.
+\end{observation}
+\begin{proof}
+Let $B^\prime$ be $n\nf{.blocks[}b-1\nf{]}$. We prove the claim by induction on height of node $n$. 
+\begin{align*}
+    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{x}}\\
+    =&B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock} \text{ code}\\
+    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}}
+\end{align*}
+\end{proof}
 \pagebreak
@@
 
-Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()}.
+First we prove some claims about the size and operations of a block. These lemmas will be used later for correctness and analysis of \nf{GetEnqueue()}.
 
@@
 From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-\nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+By Observation \ref{lem::numEnqLeft}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
 So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}.
@@
 \begin{proof}
-  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} has 2 \nf{CAS}es, one in Line \ref{cas} and one in Lines \ref{incrementHead1} or \ref{incrementHead2}.
+  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} invokes at most 3 \nf{CAS}es, one in Line \ref{cas} and two from \nf{Advance} in Line \ref{incrementHead}.
 \end{proof}

### 2022-10-04 thesis/thesis.tex
add hyperref
add current changes

@@
 %\usepackage{showkeys}
+\usepackage{hyperref}
 \usepackage{amsmath}
@@
 The main idea is that if we call \sc{Refresh}$(n)$ twice, the operations in $n$'s children before the first \sc{Refresh}$(n)$ are guaranteed to be in $n$.
-Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh()} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append.
+Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append.
 This mechanism makes us overcome the \nf{CAS} Retry Problem.
@@
 To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations.
-In each \nf{Refresh()} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
-Furthermore since the operations in a \textsc{Refresh()} step are concurrent we can linearize them  among themselves in any order we wish.
-Note that if two operations are in read one \textsc{Refresh()} step in a node they are going to be propagated up to the root together.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Furthermore since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish.
+Note that if two operations are in read one \textsc{Refresh} step in a node they are going to be propagated up to the root together.
 Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
@@
 
-\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
+\paragraph{\texttt{CreateBlock}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
 \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
@@
   \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
+  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
 \end{figure}
@@
   \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
+  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock}.}
 \end{figure}
@@
 
- Let $i$ be the value $R_n$, a successful instance of \nf{Refresh()} on node \nf{n} reads from \nf{$n$.head}.
+ Let $i$ be the value $R_n$, a successful instance of \nf{Refresh} on node \nf{n} reads from \nf{$n$.head}.
 $R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}.
@@
 Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$.
-From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating $b$.
-Let $R_p$ be the first successful \nf{p.Refresh()} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
+From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh} propagating $b$.
+Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
 Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.
@@
 Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}.
-After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh()} by the same process or another process or  maybe an \nf{$n$.parent.Refresh()}.
-We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates $b$.
+After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or  maybe an \nf{$n$.parent.Refresh}.
+We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh} that propagates $b$.
 
@@
 \paragraph{}
-In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism.
+In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate} steps and increase parallelism.
 So now the problem is to find the result of each Dequeue.
@@
 Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
-There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step.
+There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh} step.
 It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases.
@@
 \texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index.
-As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times.
+As we said later \texttt{Propagate} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh} two times.
 \texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.
@@
 \Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_i(this, b)=D_y(root,x)$.}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 \Else
-\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
-\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}
+\State \tt{dir= parent.left==n ? left: right} 
+\State \tt{sb= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1}\label{computeSuper}
 
 \If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-left}} \cmt{consider the enqueues in the previous blocks from the left child} \label{considerPreviousLeft}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
 \EndIf
@@
 \If{\tt{dir {\keywordfont is} right}} \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
-\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
+\State \tt{i+= blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
 \EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlock, i}
+\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
 \EndIf
@@
 Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
+By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
 \begin{observation} \label{nonDecreasingHead}
@@
 \begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()}.
+  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
 \end{proof}
@@
 
-The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance()}.
+The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
 If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance()} is called  in Lines \ref{helpAdvance} and \ref{advance}.
+\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
 For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
@@
 \begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
+  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
 \end{definition}
 
-In the next two results we show for every successful \nf{Refresh()}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
+In the next two results we show for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
 
 \begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
+If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
 \end{lemma}
@@
 \begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a complete instance \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
+If $R$ is a complete instance \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
 \end{corollary}
@@
 
-In the next lemma we show that if two consecutive instances of \nf{Refresh()} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh()} are guaranteed to be in $n$ after the second \nf{Refresh()}.
+In the next lemma we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
 
 \begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by process $p$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh()}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
+  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} on internal node $n$ by process $p$.
+If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
 \end{lemma}
@@
 Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh()} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
+As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
 $R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
@@
 By Lemma \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[i+1]}$ is non-null.
-Let $R_3$ be the \nf{Refresh()} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
+Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
 $R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
@@
 \begin{proof}
-  If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
+  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds.
+If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
 Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
@@
 \begin{proof}
-$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$.
-For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} untill $n$ is \nf{root}.
+$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $b$.
+For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} untill $n$ is \nf{root}.
 \end{proof}
@@
 
-First we prove some claims about the size and operations of a block. These lemmas will be used later for correctness and analysis of \nf{GetEnqueue()}.
+First we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
 
 \begin{lemma}\label{blockSize}
-Each block contains at most one operation of each process.
+Each block contains at most one operation of each process, and therefore at most $p$ operations in total.
 \end{lemma}
@@
 Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $p$ cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append()}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
+By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
 This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$.
-The Existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
+The existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
 \end{proof}
@@
 \begin{proof}
-The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an initial node contains at least one operation, due to the test on Line~\ref{addOP}.
+The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an internal node contains at least one operation, due to the test on Line~\ref{addOP}.
 We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.
@@
 
-\nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} so that $i$th \nf{Enqueue} in $b$th block of root is $e$th \nf{Enqueue} in entire sequence stored in the root. 
+\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} so that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
 \begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
@@
 \nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$.
-From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know \nf{sum\sub{enq-left}}, \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, $\nf{sum\sub{enq}}$ values of \nf{root.block[$0\cdot\cdot end$]} is  also non-decreasing.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[b-1].sum\sub{enq}} < e$.
+From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
+Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumDeq}.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
 The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-In Line \ref{dsearchBinarySearch} Binary Search finds $b$ in the range mentioned.
-Finally $i$, is computed from the definition of \nf{sum\sub{enq}}.
+In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned.
+Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumDeq}.
 \end{proof}
@@
 For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, so only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$.
-
-For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.
+Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf.
+Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
+
+For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.child, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
 From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
 By Observation \ref{lem::numEnqLeft}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
-So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}.
-By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
+So the $i$th \tt{enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
+Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[b]}.
+By Definitions \ref{def::ops} and \ref{def::subblock} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
 \begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
+&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \textrm{ or } \\
 &\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
  \end{align*}
-First we consider the case where the \nf{Enqueue} we are looking is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
+First we consider the case where the \nf{Enqueue} we are looking for is in the left child.
+There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \tt{enqueue}s in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
+We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
 
@@
 Because \nf{Enqueue}s from the left subblocks are ordered before the \nf{Enqueue}s from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ (Line \ref{rightChildGet}).
+So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
 Other parameters for the right child are chosen similarly to the left child.
@@
 So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
- Finally, \nf{n.child.GetEnqueue(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
+ Finally, \nf{n.child.GetEnqueue(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
 \end{proof}
@@
   \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}.
+\nf{n.blocks[b]} and its subblocks are shown by gray color.
 Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
@@
 \subsection{Correctness of IndexDequeue}
-\begin{definition}\label{orderRefresh} If \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
+Next few results show that the \nf{super} field of a block is accurate within one of actual index of the block's superblock in the parent node.
+Next we explain how it is used to compute the rank of a given \nf{dequeue} in the root.
+\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
 \end{definition}
@@
 \begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $b$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ unless \nf{$b$.super} is set.
+After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ untill \nf{$B$.super} is set.
 \end{observation}
 \begin{proof}
-From the Observation \ref{nonDecreasingHead} we know the only change to \nf{n.head} is on Line \ref{incrementHead} which is incrementing.
-Before an instance of \nf{Advance()} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
+From Observation \ref{nonDecreasingHead} we know the \nf{$n$.head} ichanges only by increment on Line \ref{incrementHead}.
+Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
 \end{proof}
@@
 \begin{corollary}
-  If \nf{n.blocks[i].super} is \nf{null}, then $\nf{n.head}<\nf{i}$ and \nf{n.blocks[i+1]} is \nf{null}.
+  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
 \end{corollary}
@@
 \end{proof}
-Now let us talk about how the \nf{$p$.Refresh}es that took place after the putting $b$ into \nf{n}, will help to set \nf{b.super} and propagate $b$.
+Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after the putting $B$ into $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
+
+\begin{observation}\label{lem::createBlockHead}
+If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads \nf{$n$.head} greater than $b$ in Line \ref{lastLine}. 
+\end{observation}
 
 \begin{lemma}
-  If $b$ is a direct subblock of \nf{$n$.parent.blocks[$i$]} then $\nf{$b$.super}\leq i$.
-\end{lemma}
-\begin{proof}
-Instance $R_p$ of $n$\nf{.parent.Refresh()} contains block $b$, if it reads \nf{$n$.head} greater than ${h}$, the index of $b$ in $n$ (see Line \ref{lastLine}).
-For \nf{$n$.head} to be greater than ${h}$ it means \nf{$n$.head} is incremented in Line \ref{incrementHead} which means \nf{$b$.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}).
-So if $R_p$ propagates $b$ it means \nf{$b$.super} was already set.
-The value written in \nf{b.super} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{$b$.super}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $b\nf{.super}\leq\nf{i}$.
-The reader may wonder when the case $b\nf{.super}=\nf{i}$ happens, it happens when $\nf{$n$.parent.blocks[$b$.super]}=\nf{null}$ while $b$\nf{.super} is read and $R_p$ puts its created block into \nf{n.blocks[$b$\nf{.super}]}.
+  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$sb$]} then $\nf{$B$.super}\leq sb$.
+\end{lemma}
+\begin{proof}
+By \ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ for \nf{$n$.head}, which means \nf{$n$.head} was incremented in Line \ref{incrementHead}.
+By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}, so if $R_p$ propagates $B$ it means \nf{$B$.super} was already set.
+The value written in \nf{$B$.super} was read in Line \ref{readHead} or Line \ref{readParentHead}, which are both before calling the \nf{Advance} that sets \nf{$B$.super}.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$.
+The reader may wonder when the case $b\nf{.super}=sb$ happens.
+This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{n.blocks[$b$\nf{.super}]} afterwards.
 \end{proof}
@@
 \begin{lemma}
-If $R_n=$\nf{Refresh()} puts $b$ in \nf{$n$.blocks[$h$]} at Line \ref{cas}, then the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to the Definition \ref{orderRefresh} contains $b$ and \nf{$b$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh()}.
-\end{lemma}
-
-\begin{proof}
-It is sufficient to prove one of the two successful \nf{$n$.parent.Refresh()}es propagates $b$.
-If the first successful \nf{p.Refresh()} propagated $b$ then the claim is true.
-We prove if the first \nf{$n$.parent.Refresh()} did not propagate $b$, second \nf{$n$.parent.Refresh()} propagates $b$.
-
-\nf{$b$.super} is set by some instance of \nf{Refresh()} on $n$ or \nf{$n$.parent} showed by $R^\prime$ and \nf{$n$.head} is incremented by some \nf{Refresh()} called $R^\prime\prime$.
+Let $R_n$ be the \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
+Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh}.
+\end{lemma}
+
+\begin{proof}
+Let the first successful \nf{$n$.parent.Refresh} after $R_n$ be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh} be $R_{p2}$.
+To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} neither by $R_{p1}$ nor by $R_{p2}$.
+
+Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is $\leq b$.
+From Observation \ref{nonDecreasingHead} the value  $R_{p2}$ reads in Line \ref{readChildHead} is also $\leq b$.
+
+$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so $n.$\nf{head} is $b$ at $_{\ref{cas}}^{R_n}t$.
+Since $R_{p2}$'s \nf{CAS} into $n$\nf{.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas}.
+$R_{p2}$'s \ref{readHead} is after $R_{p2}$'s \ref{incrementHead}.
+$R_p^\prime$'s \ref{incrementHead} is after $R_{p1}$'s \nf{CAS}.
+We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$.
+Since \ref{readChildHead} is after \ref{readHead} inc the code, $R_{p2}$ read a value $\geq b$ for $n.\nf{head}$.
+
+So $R_{p2}$ reads $n\nf{.head}=b$.
+After Line \ref{helpAdvance} of $R_{p2}$, $n$\nf{.head} is incremented from $b$.
+So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
+This is contradiction with our hypothesis.
+
+Furthermore, if $B$\nf{.super} is not setset earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
+
+% \nf{$b$.super} is set by some instance of \nf{Refresh} on $n$ or \nf{$n$.parent} showed by $R^\prime$ and \nf{$n$.head} is incremented by some \nf{Refresh} called $R^\prime\prime$.
 Let \nf{$n$.parent.head} value be $hp$ when $R^\prime$ reads it.
 From Lemma \ref{lem::headProgress} \nf{$n$.parent.head} could only increase from ${hp}$ to ${hp+1}$ if $\nf{$n$.parent.blocks[$hp$]}\neq\nf{null}$.
-In other words there should be a successful \nf{$n$.parent.Refresh()} for \nf{$n$.parent.head} to be increased.
+In other words there should be a successful \nf{$n$.parent.Refresh} for \nf{$n$.parent.head} to be increased.
 Let $R_p$ be \nf{$n$.parent.Refresh} that propagated $b$ to \nf{$n$.parent}.
-We claim there cannot be another successful \nf{$n$.parent.Refresh()} after $R^\prime$ reads \nf{$n$.parent.head} and before $R_p$ performs Line \ref{lastLine}.
+We claim there cannot be another successful \nf{$n$.parent.Refresh} after $R^\prime$ reads \nf{$n$.parent.head} and before $R_p$ performs Line \ref{lastLine}.
 
@@
 
-Without loss of generality  assume $n$ is the left child of \nf{$n$.parent}.
-As we mentioned the first successful \nf{$n$.parent.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{$b$.super}.
+% Without loss of generality  assume $n$ is the left child of \nf{$n$.parent}.
+As we mentioned the first successful \nf{$n$.parent.Refresh} after $t_{\ref{cas}}^{Rn}$ did not set \nf{$b$.super}.
 It might happen maybe because the value $h_{left}$ read for $n$\nf{.head} in Line \ref{readChildHead} is less than or equal to $i$ and $\nf{left.blocks[$h_{left}$]}=\nf{null}$, which means \nf{$n$.head} is advanced but $b$ is still not installed in \nf{$n$.blocks[$i$]} which means $R_n$ has not reached  the Line \ref{cas}.
 
-Let the first successful \nf{$n$.parent.Refresh()} be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh()} be $R_{p2}$.
+% Let the first successful \nf{$n$.parent.Refresh} be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh} be $R_{p2}$.
 If $R_{p1}$ reads $x$ in Line \ref{readHead}, then $R_{p2}$ has to read ${x+1}$ in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
-See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$ on \nf{$n$.parent}.
+See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh} instances $R_{p1}$,$R_{p2}$ on \nf{$n$.parent}.
 
@@
 \end{figure}
-
-$$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
-$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh()$ after $R_n$.
-$t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh()} instances on a node.
+TODO: update figures
+
+% $$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
+% $t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh$ after $R_n$.
+$t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh} instances on a node.
 $t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code.
@@
 
-So \nf{$b$.super} has set by some process before the second next successful \nf{$n$.parent.Refresh()} on Line \ref{setSuper1}.
-Since $i$ is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value for \nf{left.head} greater than ${h}$ and propagates $b$ to \nf{$n$.parent}.
-So if $b$ was not propagated already we are sure the second next successful \nf{$n$.parent.Refresh()} propagates $b$.
+% So \nf{$b$.super} has set by some process before the second next successful \nf{$n$.parent.Refresh} on Line \ref{setSuper1}.
+Since $i$ is read in the Line \ref{readHead} then the \nf{CreateBlock} in Line \ref{invokeCreateBlock} is going to read  some value for \nf{left.head} greater than ${h}$ and propagates $b$ to \nf{$n$.parent}.
+So if $b$ was not propagated already we are sure the second next successful \nf{$n$.parent.Refresh} propagates $b$.
 \end{proof}
@@
 \end{corollary}
+\begin{proof}
+\nf{super} field of a block is set after installing it  to the value of the parent node's \nf{head} by \nf{Advance}.
+From previous Lemma we know that the block is propagated by second next successful \nf{Refresh} on parent, so the value written in $b$.\nf{super} can  the \nf{super} field of a lemma i
+\end{proof}
+TODO: complete proof
 
@@
 \begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
+ If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
 \end{lemma}
@@
 We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-So preconditions of the invocation of \nf{BinarySearch} is satisfied.
-After that the order of $D_i(n,b)$ in $D(n.parent, superblock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
+For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
+By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
+If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{deqBeforeLeft}.
+Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
 Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
 \end{proof}
+TODO: add figure
 
@@
 \begin{proof}
-A linearization for an execution is legal if, $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
-If $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started.
+We must show that, every operation that terminates is in $L$ exactly once and if $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
+The first one is directly reasoned from Lemma \ref{lem::appendExactlyOnce}.
+For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
 From Lemma \ref{lem::append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
@@
 Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves.
-The way we have chosen to order the operations in a block makes \nf{GetEnqueue} and \nf{FindResponse} efficient.
+We have chosen to put \nf{enqueue}s in a block before \nf{dequeue}s (see Definition \ref{ordering}).
 
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns null it is called a \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
+If a \nf{Dequeue} operation returns null it is called a \it{null} \tt{dequeue}, otherwise it is called \it{non-null} \tt{dequeue}.
 \end{definition}
 
+Next we define  the responses that \nf{dequeue}s should return, according to the linearization.
 \begin{definition}
@@
 
-\begin{observation}\label{enqBeforeDeq}
-By Definition \ref{ordering} enqueue operations come before dequeue operations in a block.  
-\end{observation}
+% \begin{observation}\label{enqBeforeDeq}
+% By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block.  
+% \end{observation}
 
@@
 We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Observation \ref{enqBeforeDeq} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].sum\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].sum\sub{enq}}- \nf{root.blocks[}b\nf{].sum\sub{deq}}$.
+By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block.
+By Lemma \ref{lem::sumDeq} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
+Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
 In both cases, this is same as the assignment on Line \ref{computeLength}.
@@
 \begin{lemma} \label{numberOfNND}
-If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
+If operations in the root are applied with the order of $L$, the number of non-null \tt{dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
 \end{lemma}
@@
 There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
-$\#\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ $-$ $\#\it{non-null dequeues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness}, $\it{\#non-null dequeues}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \tt{dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
+By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::sumDeq}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
 \end{proof}
@@
 \begin{lemma}\label{nullReturn}
-\nf{R(D\sub{root,b,i})} is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-From Corollary \ref{numNullDeqBlock} and Observation \ref{enqBeforeDeq}.
+$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
+\end{lemma}
+\begin{proof}
+From Corollary \ref{numNullDeqBlock}.
 \end{proof}
@@
 $D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise if $D_i(root,b)$ is $e$th non-null dequeue in $L$ it should return $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null dequeue operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The dequeues in the \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
-So $Resp(D_i(root,b))$ is the \nf{element} field of $e$th non-null dequeue when $e=i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{deq}$ (Line \ref{computeE}).
+Otherwise, if $D_i(root,b)$ is the $e$th non-null dequeue in $L$ it should return the $e$th enqueued value.
+By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \tt{dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
+The \tt{dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
+So $D_i(root,b)$ is the $e$th non-null \tt{dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
 See Figure \ref{computeResponseDetail}.
@@
 \begin{proof}
-Enqueue operations do not return any value. By \ref{computeHead} response of a dequeue in our algorithm is same as the response from the sequential execution of $L$.  
+Enqueue operations do not return any value. By Lemma \ref{computeHead} response of a \tt{dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
 \end{proof}
@@
 \begin{proof}
-From \ref{linearSat}, \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as new operations come.
+The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
+\end{proof}
+
+\paragraph{Remark} In fact our algorithm is strongly linearizable defined in \cite{DBLP:conf/stoc/GolabHW11}.
+By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
 \pagebreak
@@
 \begin{proof}
-  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} invokes at most 3 \nf{CAS}es, one in Line \ref{cas} and two from \nf{Advance} in Line \ref{incrementHead}.
+  In each height of the tree at most 2 times \nf{Refresh} is invoked and  every \nf{Refresh} invokes at most 3 \nf{CAS}es, one in Line \ref{cas} and two from \nf{Advance} in Line \ref{incrementHead}.
 \end{proof}

### 2022-10-04 thesis/thesis.tex
finalize text

@@
 \usepackage{amsmath}
+\usepackage{relsize}
+\usepackage{enumitem}
 
@@
 \If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}, h}}} \label{helpAdvance}
+\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
 \EndIf
@@
 \State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-
+\State{\tt{this.\Call{Advance}{h}}} \label{advance}
+\State \Return{ \tt{result}}
+
+\EndFunction{Refresh}
+
+\Statex
+
+\Function{void}{Advance}{\sl{int} h}
 \State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State{\tt{this.\Call{Advance}{h, h\sub{p}}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\Statex
-
-\Function{void}{Advance}{\sl{int} h, \sl{int} hp}
-
 \State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
@@
 
-The next two observations are correct for both enqueues and deqeueues.
-
-\begin{observation}
-  $n\nf{.blocks[}b\nf{]}\nf{.num\sub{enq}}$ is the size of $E(n,b)$.
+\begin{observation}\label{sumToNum}
+$n.\nf{blocks[}i\nf{].sum\sub{x}}-n.\nf{blocks[}j\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$ ($i<j$)
 \end{observation}
 
-\begin{observation}\label{lem::sumDeq}
-$\sum_{i=1}^{b} {E(n,i)}=n\nf{.blocks[}i\nf{].sum\sub{enq}}$.
-\end{observation}
-
-The next observation is correct for both directions.
-
-\begin{observation}\label{lem::numEnqLeft}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$. $B\nf{.num\sub{enq-left}}$ is the number of \nf{Enqueue} operations in the direct subblocks of $B$ in $n$\nf{.left}.
-\end{observation}
-\begin{proof}
-Let $B^\prime$ be $n\nf{.blocks[}b-1\nf{]}$. We prove the claim by induction on height of node $n$. 
+\begin{lemma}\label{lem::numX}
+Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
+\begin{enumerate}[label=(\arabic*)]
+    \item If $n$ is an internal node $B\nf{.num\sub{x-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
+    \item If $n$ is an internal node $B\nf{.num\sub{x-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
+    \item $B\nf{.num\sub{x}}=\Big|E(n,b)\Big|$
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+ We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of claim for $n$.
 \begin{align*}
-    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{x}}\\
-    =&B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock} \text{ code}\\
-    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}}
+    B\nf{.num\sub{x-left}}=&B.\nf{sum\sub{x-left}}-B^\prime.\nf{sum\sub{x-left}}& \text{Definition of }\nf{num\sub{x}}\\
+    =&B^\prime.\nf{sum\sub{x-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}-B^\prime.\nf{sum\sub{x-left}} & \nf{CreateBlock}\\
+    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}\\
+    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{x}} &\text{Observation} \ref{sumToNum}
 \end{align*}
-\end{proof}
+The last line holds because of the induction hypothesis. (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
+\begin{align*} 
+E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
+&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
+\end{align*}
+By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{x-left}}+B\nf{.num\sub{x-right}}=B\nf{.num\sub{x}}$
+\end{proof}
+
+\begin{corollary}\label{lem::sumX}
+Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
+\begin{enumerate}[label=(\arabic*)]
+    \item If $n$ is an internal node $B\nf{.sum\sub{x-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
+    \item If $n$ is an internal node $B\nf{.sum\sub{x-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
+    \item $B\nf{.sum\sub{x}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
+\end{enumerate}
+\end{corollary}
+
 \pagebreak
@@
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumDeq}.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
@@
 In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned.
-Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumDeq}.
+Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumX}.
 \end{proof}
@@
 From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Observation \ref{lem::numEnqLeft}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
+By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
 So the $i$th \tt{enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
@@
 Let $R_n$ be the \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh}.
-\end{lemma}
-
-\begin{proof}
-Let the first successful \nf{$n$.parent.Refresh} after $R_n$ be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh} be $R_{p2}$.
+Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set after Line \ref{helpAdvance} of the the second successful \nf{$n$.parent.Refresh}.
+\end{lemma}
+
+\begin{proof}
+Let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
 To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} neither by $R_{p1}$ nor by $R_{p2}$.
@@
 
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so $n.$\nf{head} is $b$ at $_{\ref{cas}}^{R_n}t$.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas}.
-$R_{p2}$'s \ref{readHead} is after $R_{p2}$'s \ref{incrementHead}.
-$R_p^\prime$'s \ref{incrementHead} is after $R_{p1}$'s \nf{CAS}.
+$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}.
+Since $R_{p2}$'s \nf{CAS} into $n$\nf{parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}.
 We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$.
-Since \ref{readChildHead} is after \ref{readHead} inc the code, $R_{p2}$ read a value $\geq b$ for $n.\nf{head}$.
+Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$.
+
+$$_{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_p1}<_{\ref{incrementHead}}^{R_p^\prime}t<_{\ref{cas}}^{R_p2}t<_{\ref{readChildHead}}^{R_p2}$$
+TODO: draw figure
 
@@
 
-Furthermore, if $B$\nf{.super} is not setset earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
+Furthermore, if $B$\nf{.super} is not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
 
@@
 
-\begin{figure}[hbt]  
-  \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
-  \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
-  \label{fig::timeLine1}
-\end{figure}
+% \begin{figure}[hbt]  
+%   \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
+%   \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
+%   \label{fig::timeLine1}
+% \end{figure}
 
@@
 
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
-  \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
-  \label{fig::timeLine2}
-\end{figure}
-TODO: update figures
+% \begin{figure}[hbt]  
+%   \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
+%   \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
+%   \label{fig::timeLine2}
+% \end{figure}
 
@@
 \begin{corollary} \label{superRelation}
-If \nf{$n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
+If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
 \end{corollary}
 \begin{proof}
-\nf{super} field of a block is set after installing it  to the value of the parent node's \nf{head} by \nf{Advance}.
-From previous Lemma we know that the block is propagated by second next successful \nf{Refresh} on parent, so the value written in $b$.\nf{super} can  the \nf{super} field of a lemma i
-\end{proof}
-TODO: complete proof
+After that $B$ is installed $n$\nf{.parent.head} is read and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
+From previous Lemma we know that $B$ is propagated by second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
+To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_p2}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}+1$ and $n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$, so the value read for $n$\nf{.parent.head} is equal to or one less than the index of the superblock of $B$.
+\end{proof}
 
@@
 By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{deqBeforeLeft}.
+If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft}.
 Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
@@
 By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block.
-By Lemma \ref{lem::sumDeq} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
+By Lemma \ref{lem::numX} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
 Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
@@
 the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \tt{dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::sumDeq}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
 \end{proof}
@@
 \begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumDeq}.
+$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
 $D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
@@
 To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDequeue()} is called.
-\nf{IndexDequeue} does a search with range $p$ in each level (Lemma \ref{superRange}) which takes $O(log^2 p)$ in the tree.
+\nf{IndexDequeue} does a search with range $p$ in each level which takes $O(log^2 p)$ in the tree.
 In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$.

### 2022-10-05 thesis/thesis.tex
add figures, fix bugs, minor modifications

@@
 \State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
+\State \tt{i+= parent.blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
 \EndIf \label{computeISuperEnd}
@@
 $R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$.
-Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$.
-
-$$_{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_p1}<_{\ref{incrementHead}}^{R_p^\prime}t<_{\ref{cas}}^{R_p2}t<_{\ref{readChildHead}}^{R_p2}$$
-TODO: draw figure
+Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} (Line \ref{incrementHead}) after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}.
+We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
+Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
+So $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$ by Lemma \ref{nonDecreasingHead}.
+\begin{align*}
+\left.
+\begin{array}{r}
+ _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
+_{\ref{cas}}^{R_{p1}}<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
+_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
+\end{array}
+\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
+\end{align*}
 
@@
 
-Furthermore, if $B$\nf{.super} is not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
+Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
 
@@
 By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft}.
-Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}).
+If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
+Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
 Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
 \end{proof}
-TODO: add figure
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
+
+  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
+\end{figure}
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
+
+  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
+\end{figure}
 

### 2022-10-06 thesis/thesis.tex
Finalize proof, apply Oct 5 notes

@@
 
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
 \begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
-
+  \item \tt{\sl{int} \size}
+  \textsf{: size of the queue after performing all operations in the prefix for this block}
 \end{itemize}
@@
   \item \tt{\sl{int} sum\sub{enq-left}}
-  \textsf{: \# enqueue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+  \textsf{: \# enqueues in \tt{left.blocks[1..end\sub{left}]}}
   \item \tt{\sl{int} sum\sub{deq-left}}
-  \textsf{: \# dequeue operations in the prefix for \tt{left.blocks[end\sub{left}]}}
+  \textsf{: \# dequeues in \tt{left.blocks[1..end\sub{left}]}}
   \item \tt{\sl{int} sum\sub{enq-right}}
-  \textsf{: \# enqueue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
+  \textsf{: \# enqueues in \tt{right.blocks[1..end\sub{right}]}}
   \item \tt{\sl{int} sum\sub{deq-right}}
-  \textsf{: \# dequeue operations in the prefix for \tt{right.blocks[end\sub{right}]}}
+  \textsf{: \# dequeues in \tt{right.blocks[1..end\sub{right}]}}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
+\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
 \begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in the prefix for this block}
+  \item \tt{\sl{Object} element}
+  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+
+    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
 \end{itemize}
@@
 
-\it{Abbreviations:}
+\it{Abbreviations used in the code and the proof of correctness.}
 \begin{itemize}
  \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
- \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{$\emptyset$, enq, deq, enq-left, enq-right, deq-left, deq-right\})}
+%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
+  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{ enq, deq, enq-left, enq-right, deq-left, deq-right\})}
 \end{itemize}
@@
 
-\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
+\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree and returns its response.}
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
+\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the response to $D_{root,b,i}$.}
 \If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
@@
 \Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
-\cmt{Returns \tt{<b,i>} if $E_e(root)=E_i(root,b)$.}
+\cmt{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$.}
 \State \tt{start= end-1} \label{dsearchStart}
@@
 
-\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
+\Function{void}{Append}{\sl{block} blk} \cmt{Only called by the owner of the leaf.}
 \State \tt{blocks[\head]= blk} 
@@
 \State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, hp)} \label{setSuper1}
+\State \tt{blocks[h].super.CAS(null, h\sub{p}} \label{setSuper1}
 \State \tt{head.CAS(h, h+1)} \label{incrementHead}
@@
 \pagebreak
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} greater than or equal to \tt{i}}
+\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{i} of the given prefix sum \tt{field}.
+Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.
 \Function{int}{BinarySearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
-
-\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}.
-Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
-%\State \Return \tt{result block's index}
+\State \Return \tt{min\{j: blocks[j].f$\geq$i\}}
 \EndFunction{BinarySearch}
@@
 
-\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
-\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
-\State \tt{block newBlock= \Call{new}{\sl{block}}}
+\Statex\cmt{Creates and returns the block to be installed in \tt{blocks[i]}.
+Created block includes \tt{left.blocks[index\sub{prev}+1..index\sub{last}]} and \tt{right.blocks[index\sub{prev}+1..index\sub{last}]}.}
+\Function{Block}{CreateBlock}{\sl{int} i} 
+\State \tt{block new= \Call{new}{\sl{block}}}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
+\State \tt{new.end\sub{dir}= index\sub{last}} \label{endDefLine}
 \State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
 \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
+\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + }
+\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
+\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + }
+\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
 \EndFor
 \If{\tt{this} \bf{is} \tt{root}}
-\State \tt{newBlock.\size= max(root.blocks[i-1].\size { }+ newBlock.num\sub{enq} - newBlock.num\sub{deq}, 0)}\label{computeLength}
+\State \tt{new.\size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}}
+\Statex \hspace{8.5em} \tt{- new.num\sub{deq}, 0)}\label{computeLength}
 \EndIf
 
-\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
+\State \Return \tt{new}
 \EndFunction{CreateBlock}
@@
 \Else
-\State \tt{dir= parent.left==n ? left: right} 
-\State \tt{sb= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1}\label{computeSuper}
+\State \tt{dir= (parent.left==n ? left: right)} 
+\State \tt{sb= (parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1)}\label{computeSuper}
 
@@
 \State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\EndIf
-
-\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
+\Else \label{considerRight}
 \State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
@@
 
+The linearization ordering is given by the order that operations appear in the blocks in the root.
 \begin{definition}[Linearization] \label{def::lin}
@@
 \begin{observation}\label{sumToNum}
-$n.\nf{blocks[}i\nf{].sum\sub{x}}-n.\nf{blocks[}j\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$ ($i<j$)
+For any node $n$ and indices $i<j$ of \nf{blocks} in $in$, we have
+$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where \nf{x} in $\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
 \end{observation}
 
+Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
 \begin{lemma}\label{lem::numX}
-Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
+Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively.
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node $B\nf{.num\sub{x-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
-    \item If $n$ is an internal node $B\nf{.num\sub{x-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
-    \item $B\nf{.num\sub{x}}=\Big|E(n,b)\Big|$
+    \item If $n$ is an internal node $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
+    \item If $n$ is an internal node $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
+    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
 \end{enumerate}
@@
 \begin{proof}
- We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of claim for $n$.
+ We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
 \begin{align*}
-    B\nf{.num\sub{x-left}}=&B.\nf{sum\sub{x-left}}-B^\prime.\nf{sum\sub{x-left}}& \text{Definition of }\nf{num\sub{x}}\\
-    =&B^\prime.\nf{sum\sub{x-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}-B^\prime.\nf{sum\sub{x-left}} & \nf{CreateBlock}\\
-    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{x}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{x}}\\
-    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{x}} &\text{Observation} \ref{sumToNum}
+    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{enq}}\\
+    =& B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock}\\
+    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
+    =& \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} &\text{Observation } \ref{sumToNum}\\
+    =& \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big| & \textrm{Induction hypothesis (3)}
 \end{align*}
-The last line holds because of the induction hypothesis. (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
+The last line holds because of the induction hypothesis (3). (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
 \begin{align*} 
@@
 \end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{x-left}}+B\nf{.num\sub{x-right}}=B\nf{.num\sub{x}}$
-\end{proof}
-
+By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
+\end{proof}
+
+Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
 \begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{x} be in $\{\nf{enq},\nf{deq}\}$.
+Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{enq} be in $\{\nf{enq},\nf{deq}\}$.
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node $B\nf{.sum\sub{x-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
-    \item If $n$ is an internal node $B\nf{.sum\sub{x-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
-    \item $B\nf{.sum\sub{x}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
+    \item If $n$ is an internal node $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
+    \item If $n$ is an internal node $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
+    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
 \end{enumerate}
@@
 Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
 Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
@@
 
-\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} so that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
+\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
 \begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
@@
 \begin{proof}
-\nf{DSearch} performs a doubling search from \nf{root.blocks[end]} to \nf{root.blocks[0]} to find $E_e(root)$.
-From Lines \ref{setSumEnqLeft}, \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} is also non-decreasing.
+ From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
+Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
 Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-The doubling search on Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles its search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-In Line \ref{dsearchBinarySearch}, the binary search finds the $b$ required in the range mentioned.
-Finally $i$, is computed from the definition of \nf{sum\sub{enq}} and lemma \ref{lem::sumX}.
+Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
+Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blocks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
+By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
+Finally $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
 \end{proof}
@@
 \begin{proof}
-We are going to prove this lemma by induction on the height of node \nf{n}.
+We are going to prove this lemma by induction on the height of node $n$.
 For the base case, suppose $n$ is a leaf.
@@
 
-For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.child, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}.
-So the $i$th \tt{enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[b]}.
-By Definitions \ref{def::ops} and \ref{def::subblock} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
+For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{child}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
+From Definition \ref{ordering} of $E(n,b)$, so operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
+By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of $n$.
+So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
+Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
+By Definitions \ref{def::subblock} and \ref{def::ops} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
 \begin{align*}
@@
 First we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \tt{enqueue}s in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
+There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
 We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
@@
 So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
- Finally, \nf{n.child.GetEnqueue(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
+ Finally, \nf{$n$.child.GetEnqueue($subblock$, $i$)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
 \end{proof}
@@
 
-  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}.
-\nf{n.blocks[b]} and its subblocks are shown by gray color.
-Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
+  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
+Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
+\nf{Enqueue} operations from the left subblocks (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
   }\label{figGet}
@@
 \subsection{Correctness of IndexDequeue}
-Next few results show that the \nf{super} field of a block is accurate within one of actual index of the block's superblock in the parent node.
-Next we explain how it is used to compute the rank of a given \nf{dequeue} in the root.
+The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
+Then we explain how it is used to compute the rank of a given \nf{dequeue} in the root.
 \begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
@@
 \begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ untill \nf{$B$.super} is set.
+After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
 \end{observation}
 \begin{proof}
-From Observation \ref{nonDecreasingHead} we know the \nf{$n$.head} ichanges only by increment on Line \ref{incrementHead}.
+From Observation \ref{nonDecreasingHead} we know the \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
 Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
@@
 \end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after the putting $B$ into $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
+Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
 
 \begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads \nf{$n$.head} greater than $b$ in Line \ref{lastLine}. 
+If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
 \end{observation}
@@
 \begin{proof}
-By \ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ for \nf{$n$.head}, which means \nf{$n$.head} was incremented in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}, so if $R_p$ propagates $B$ it means \nf{$B$.super} was already set.
-The value written in \nf{$B$.super} was read in Line \ref{readHead} or Line \ref{readParentHead}, which are both before calling the \nf{Advance} that sets \nf{$B$.super}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$.
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that stores $n$\nf{.parent.blocks[$sb$]}.
+By \ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
+By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
+The value written in \nf{$B$.super} was read in Line \ref{readParentHead},s before the \nf{CAS} that sets \nf{$B$.super}.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$, since \nf{$n$.parent.head} is still equal to $sb$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Invariant \ref{lem::headProgress}.
 The reader may wonder when the case $b\nf{.super}=sb$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{n.blocks[$b$\nf{.super}]} afterwards.
+This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
 \end{proof}
@@
 \begin{lemma}
-Let $R_n$ be the \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set after Line \ref{helpAdvance} of the the second successful \nf{$n$.parent.Refresh}.
+Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
+Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
 \end{lemma}
@@
 Let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} neither by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is $\leq b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also $\leq b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads value $b$ for $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} (Line \ref{incrementHead}) after $R_{p1}$'s \ref{cas} and before $R_{p2}$'s \ref{readHead}.
+To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
+
+Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
+From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most $b$.
+
+$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
+Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
 We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, since \ref{readChildHead} is after \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-So $R_{p2}$ reads a value $\geq b$ for $n.\nf{head}$ by Lemma \ref{nonDecreasingHead}.
+Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
 \begin{align*}
@@
  _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
+_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
 _{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
@@
 
-So $R_{p2}$ reads $n\nf{.head}=b$.
-After Line \ref{helpAdvance} of $R_{p2}$, $n$\nf{.head} is incremented from $b$.
+So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Lemma \ref{nonDecreasingHead}.
+
+Therefore $R_{p2}$ reads $n\nf{.head}=b$.
+$R_{p2}$ calls $n$\nf{.Advance} at Line Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
 So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
@@
 
-Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$ in Line \ref{setSuper1} invoked from \ref{helpAdvance}.
+Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$ call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
 
@@
 \begin{proof}
-After that $B$ is installed $n$\nf{.parent.head} is read and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-From previous Lemma we know that $B$ is propagated by second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_p2}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}+1$ and $n\nf{.parent.head}_{_{\ref{cas}}^{R_p1}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$, so the value read for $n$\nf{.parent.head} is equal to or one less than the index of the superblock of $B$.
+% After that $B$ is installed $n$\nf{.parent.head} is read and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
+Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
+Before  $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
+From thr previous Lemma we know that $B$ is propagated by second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
+To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and by Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead} $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$.
+The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
+So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
 \end{proof}
@@
 \begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
+We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
+For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
+By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
 If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Else if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis.
+Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
+Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the correct response by induction hypothesis.
 \end{proof}
@@
 
-  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
+  \caption{The number of \nf{Dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
 \end{figure}
@@
 We must show that, every operation that terminates is in $L$ exactly once and if $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
-The first one is directly reasoned from Lemma \ref{lem::appendExactlyOnce}.
+The first claim is directly reasoned from Lemma \ref{lem::appendExactlyOnce}.
 For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Lemma \ref{lem::append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
+From Lemma \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
 By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
@@
 Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves.
-We have chosen to put \nf{enqueue}s in a block before \nf{dequeue}s (see Definition \ref{ordering}).
+We have chosen to put \nf{Enqueue}s in a block before \nf{Eequeue}s (see Definition \ref{ordering}).
 
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns null it is called a \it{null} \tt{dequeue}, otherwise it is called \it{non-null} \tt{dequeue}.
+If a \nf{Dequeue} operation returns null it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
 \end{definition}
 
-Next we define  the responses that \nf{dequeue}s should return, according to the linearization.
+Next we define  the responses that \nf{Dequeue}s should return, according to the linearization.
 \begin{definition}
@@
 % \begin{observation}\label{enqBeforeDeq}
-% By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block.  
+% By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block.  
 % \end{observation}
@@
 We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \tt{enqueue} operations come before \tt{dequeue} operations in a block.
-By Lemma \ref{lem::numX} if there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
+By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block.
+By Lemma \ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show ther number of \nf{Enqueue} and \nf{Dequeue} operations in it.
+If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
 Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
@@
 \begin{lemma} \label{numberOfNND}
-If operations in the root are applied with the order of $L$, the number of non-null \tt{dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
+If operations in the root are applied with the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
 \end{lemma}
@@
 There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \tt{dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \tt{dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
+By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
 \end{proof}
@@
 \begin{proof}
-From Corollary \ref{numNullDeqBlock}.
+From Corollary \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
 \end{proof}
@@
 Otherwise, if $D_i(root,b)$ is the $e$th non-null dequeue in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \tt{dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \tt{dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
-So $D_i(root,b)$ is the $e$th non-null \tt{dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
+By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
+The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
+So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
 See Figure \ref{computeResponseDetail}.
 
-After computing \nf{e} at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
+After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
+Correctness of \nf{DSearch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
 \end{proof}
@@
 \begin{proof}
-Enqueue operations do not return any value. By Lemma \ref{computeHead} response of a \tt{dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
+\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead} response of a \nf{Dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
 \end{proof}
@@
 
-\paragraph{Remark} In fact our algorithm is strongly linearizable defined in \cite{DBLP:conf/stoc/GolabHW11}.
+\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
 By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.

### 2022-10-07 thesis/thesis.tex
fill in general ideas to explain in description

@@
 \section{Our Queue}
-Jayanti and Petrovic introduced a wait-free polylogarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We benefit from some ideas of their work to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
-Our algorithm despite them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
-I our model there are $p$ procesess doing \nf{Enqueue(),Dequeue()} operations concurrently.
-We use a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes.
+Paragraph titles will be removed after polishing.
+
+\paragraph{What is a Queue} In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked.
+We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem).
+A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
+\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
+If the queue is empty it returns \texttt{null}.
+
+\paragraph{How we use Jayanti and why we are better} Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
+We use their idea of using a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
+Despite their work, our algorithm them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
+In our model there are $p$ processes doing \nf{Enqueue(),Dequeue()} operations concurrently.
+
+\paragraph{Introduce tree} There is a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes.
 Each process has a leaf which the order of operations invoked by the process is stored in it.
-When a process wishes to do an operation it appends the operation to its leaf and then tries to propagate its new operation up to the tree's root.
-In each node the ordering of operations propagated up to it is stored.
+When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
+In each node of the tree the ordering of operations propagated up to it is stored.
 All processes agree on the sequence stored in the root and it is defined to be the linearization ordering.
@@
 \end{figure}
-
-\it{Add sequence to nodes}
-
-We could implement the sequence stored in each node using an array of the queue operations and append some operations to the sequence by doing \tt{k-CAS} operation on the end of the array.
-To do a propagate step on node $n$ in the tree, we aggregate the operations from node $n$'s both children (that have not already been propagated to $n$) and try to append them into $n$.
-We call this procedure \sc{Refresh}$(n)$.
-The main idea is that if we call \sc{Refresh}$(n)$ twice, the operations in $n$'s children before the first \sc{Refresh}$(n)$ are guaranteed to be in $n$.
-Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append.
-This mechanism makes us overcome the \nf{CAS} Retry Problem.
+\it{TODO::Add sequence to nodes}
+
+\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, we can read the operations from node $n$'s both children that are not already in $n$, and then try to append them to the sequence stored in $n$.
+We call this procedure \tt{Refresh($n$)}.
+The main idea is that if a process invokes \tt{Refresh} on a node two times and in both times fail to append the new operations to $n$, the operations in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
+Because if both of the \tt{Refresh}$(n)$s fail to append then there is another instance of \nf{Refresh} in between which has succeeded to do append and has already appended the operations that the first \nf{Refresh} was trying to append.
 
@@
 \begin{center}
-\begin{subfigure}[b]{.49\textwidth}
+\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
   \centering
@@
   \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $r_1$ & $l_1$ & $r_2$ & $l_2$ & $l_3$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c||c|c|}  \hline $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c||c|c|}  \hline $r_1$ & $r_2$ & $r_3$ & $r_4$\\ \hline \end{tabular}} ]
+\Tree [.$r_1,l_1,l_2,r_2,l_3$
+$l_1,l_2,l_3,l_4,l_5$
+$r_1,r_2,r_3,r_4$ ]
 \end{tikzpicture}}
-  \caption{Operations after $||$ are new.}
+  \caption{Before the \nf{Refresh}.}
 \end{subfigure}
 \hfill
-\begin{subfigure}[b]{.49\textwidth}
+\begin{subfigure}[b]{.29\textwidth}
   \centering
@@
   \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
-\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $r_1$ & $l_1$ & $r_2$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $r_3$ & $r_4$  \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c|c||}  \hline $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ \\ \hline \end{tabular}}
-{\begin{tabular}{|c|c|c|c||}  \hline$r_1$ & $r_2$ & $r_3$ & $r_4$\\ \hline \end{tabular}} ]
+\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
+$l_1,l_2,l_3,l_4,l_5$
+$r_1,r_2,r_3,r_4$ ]
 \end{tikzpicture}}
-  \caption{New operations are added to the parent node.}
+  \caption{New operations are appended.}
 \end{subfigure}
-\caption{\label{fig::propagstep} Before and after of a \sc{Refresh}$(n)$ with successful \nf{CAS}.
-Operations propagating from the left child are numbered with $l$ and from the right child by $r$ and the operations in children after $||$ are new.}
+\hspace{6em}
+\caption{\label{fig::propagstep} Before and after of a \sc{Refresh}$(n)$ with successful \nf{CAS}. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
 \end{center}
 \end{figure}
-\it{fix $||$}
+
+\begin{figure}[hbt]
+  \center\includegraphics[width=4in]{pics/doublerefresh.png}
+  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock}.}
+\end{figure}
+\it{TODO::simplify the Figure, remove code invocations}
+
+\paragraph{Introduce CAS} We do not intend to get into our implementation but just for information we use \nf{CAS} (Compare \& Swap) operations to implement the Try to Append mechanism told in the previous paragraph.
+After a process appended its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node.
+So with poly-logarithmic \nf{CAS}es we can have a tree agreeing on the linearization.
+This solution makes us overcome the \nf{CAS} Retry Problem.
+
 \begin{figure}[h]
@@
 \end{tikzpicture}
-\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves.
-Since we linearize operations when they are added to the root, ordering the blocks in the root is important.}
+\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves.}
 \end{center}
@@
 
-The solution for  implementing the orderings in the tree told above is not efficient, because there are big \nf{CAS}es and operations information are copied all the way up to the root.
-Instead of storing operations explicitly in the nodes, we can keep track of some statistics of them.
-This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$).
-To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
-Furthermore since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish.
-Note that if two operations are in read one \textsc{Refresh} step in a node they are going to be propagated up to the root together.
+\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation's information is copied all the way up to the root.
+Instead we can keep track of some statistics of the operations propagated together.
+This allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
+To do that, we introduce \it{blocks} that contain the number of operations from the left and the right child in a \texttt{Refresh} procedure and only propagate the statistics block of the new operations, and we store arrays of blocks in each node of the tree In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Furthermore since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are in read one \textsc{Refresh} step in a node they are going to be propagated up to the root together.
 Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
@@
 
-A process may wish to know the $i$th propagated operation or the rank of a propagated operation in the linearization.
-In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with its rank.
-\nf{enqueue}s and \nf{dequeue}s are appended to the tree and when we want to find the response to a \nf{dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null.
-If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetEnqueue(n,b,i)}.
-\texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetEnqueue(n,b,i)} finds its sub-block recursively to reach a leaf.
-\texttt{Index()} is similar but more complicated, finding super-blocks from a leaf to the root.
-The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$.
-After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
-Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
-
-\paragraph{Implementing Queue using Block Tree}
-In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem).
-A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
-If the queue is empty it returns \texttt{null}.
-Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.
-So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
-
 \begin{figure}
-
 \begin{center}
@@
 
-\paragraph{}
-\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence.
-We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
-To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BinarySearch faster.
-
-Furthermore, in each block, we store the prefix sum of left and right elements.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see fig \ref{fig::pointer} and \ref{fig:prefix}).
+\paragraph{Introduce Get and Index} With what told till now we have a shared tree among procesess which they agree on the  implicit ordering stored in its root.
+Having an agreement by itself is not enough.
+A process may wish to know the $i$th propagated operation (1) or the rank of a propagated operation in the linearization (2).
+With these two queries we can design a universal construction.
+Let us do not focus on th universal construction and focus on how to use (1) and (2) to construct a queue.
+
+\paragraph{Introduce how to create Queue using Get and Index} In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with an specific rank.
+\nf{Enqueue}s and \nf{Dequeue}s are appended to the tree and when we want to find the response to a \nf{Dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null.
+
+\paragraph{Explain pieces of a block helping to do Get and Index}
+After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
+Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
+
+To get the $i$th enqueue in the root, we can find the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
+To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to determine the range of subblocks of a block to make binary search faster.
+In each block, we store the prefix sum of left and right elements.
+Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figures \ref{fig::pointer} and \ref{fig:prefix}).
+We know a block size is at most $p$, so binary search takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the search range which its size is \textsc{O}$(p)$.
 
@@
 
-\paragraph{}
-Starting from the root, \textsc{GetIndex}($i$) BinarySearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$.
-From lemma $\ref{blockSize}$ we know a block size is at most $p$.
-So BinarySearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.
-
-\paragraph{\texttt{CreateBlock}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
-\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
-Other attributes are assigned values followed by definition.
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock}.}
-\end{figure}
-
-\paragraph{Computing \texttt{Get(n, b, i)}}
-
-\paragraph{How \texttt{Refresh(n)} works.}
-
-\paragraph{Computing superblock}
-
- Let $i$ be the value $R_n$, a successful instance of \nf{Refresh} on node \nf{n} reads from \nf{$n$.head}.
-$R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}.
-Let $p$ be $n$\nf{}.parent.
-Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$.
-From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh} propagating $b$.
-Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
-Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.
-
-Since the index of the superblock of ${b}$ is not known until $b$ is propagated, $R_n$ cannot set the \nf{super} field of ${b}$ while creating it.
-One approach is to set the \nf{super} field of ${b}$ by $R_n$ after propagating $b$ to $p$.
-This solution would not be efficient because there might be $p$ subblocks in the block $R_p$ propagated needing to update the \nf{super} field.
-However intuitively, once $b$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-One idea is that if we know the approximate position of the superblock of $b$ then we can search for the real superblock when we wished to know the superblock of $b$ i.e.
-\nf{b.super} does not have to be the exact location of  the superblock of $b$, but we want it to be close to \nf{j}.
-We can set \nf{$b$.super} to \nf{$n$.parent.head} while creating $b$, but the problem is that there might be many \nf{$p$.Refresh}es that could happen after reading $p$\nf{.head} by $R_n$ and before propagating $b$ to $p$.
-If we set \nf{$b$.super} to \nf{$p$.head} after appending $b$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $b$ and before setting \nf{$b$.super}.
-In this case the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$b$.super}.
-
-Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}.
-After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or  maybe an \nf{$n$.parent.Refresh}.
-We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh} that propagates $b$.
-
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
 \paragraph{Queue from tree}
@@
 
-%\begin{definition}
-%  \texttt{index(op)}: index of the given Dequene among same type operation in conataing block.
-%  
-%  \texttt{INDEX(op)}: index of the given Dequene among same type operation in all operations.
-%  
-%\end{definition}
-
 Size and the number of non-null dequeues for $b$th block could be computed this way:\\
@@
 
-\subsection{Pseudocode description}
-
-\paragraph{Specification}
-A Queue is a shared data structure that stores a sequence of elements.
-It has two methods \texttt{Enqueue(e)} and \texttt{Dequeue()}.
-\texttt{Enqueue(e)} adds \texttt{e} to the end of the sequence.
-\texttt{Dequeue()} returns the first element stored in the sequence and removes it from the sequence.
-
-\paragraph{Tree}
-In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree.
-Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}.
-Each process adds $op$ to its leaf.
-In each internal node an ordering of operations in its subtree is stored.
-All processes agree on the total ordering of all operations stored in the root.
-This ordering will be the linearization of the operations.
-\paragraph{Implicit Storing Blocks}
-For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks.
-A Block is a constant size object that implicitly represents a sequence of operations.
-In each node there is an array of Blocks.
-
+\subsection{Details of Implementation}
+
+\paragraph{\tt{Block}}
 Block $b$ contains subblocks in the left and right children.
@@
 
-\texttt{Enqueue(e)} just \texttt{append}s an \texttt{operation} with \texttt{element e} to the \texttt{root}.
-\texttt{Dequeue()} appends an \texttt{operation} to the root and computes its ordering and the \texttt{enqueue operation} containing the head before it calling \texttt{ComputeHead()} and then \texttt{gets} and returns the \texttt{operation}'s element.
-
-\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index.
-As we said later \texttt{Propagate} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh} two times.
-\texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.
-
-We only want to know the \texttt{element} of \texttt{enqueue} operations and compute ordering for \texttt{dequeue} operations.
-That's the reason here \texttt{Get()} searches between enqueues only and \texttt{Index()} returns ordering of a dequeue among dequeues.
-\texttt{Get(n, b ,i)} decides the requested element is in which child of n and continues to search recursively.
-\texttt{index(n, i, b)} calculates the ordering of the given operation in \texttt{n}'s parent each step and finally returns the result among total ordering.
+\paragraph{\texttt{Enqueue(e)}}
+\paragraph{\texttt{Dequeue()}}
+\paragraph{\texttt{FindResponse(b,i)}}
+\paragraph{\tt{Append(blk)}}
+\paragraph{\texttt{Refresh(n)}}
+\paragraph{\tt{Propagate()}}
+\paragraph{\texttt{CreateBlock}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
+\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
+Other attributes are assigned values followed by definition.
+\begin{figure}[hbt]
+  \center\includegraphics[width=5.5in]{pics/createblock}
+  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
+\end{figure}
+
+\paragraph{\texttt{DSearch(e,end)}}
+\paragraph{\texttt{GetEnqueue(b,i)}}
+\paragraph{\tt{IndexDequeue(b,i)}}
+
+ Let $i$ be the value $R_n$, a successful instance of \nf{Refresh} on node \nf{n} reads from \nf{$n$.head}.
+$R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}.
+Let $p$ be $n$\nf{}.parent.
+Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$.
+From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh} propagating $b$.
+Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
+Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.
+
+Since the index of the superblock of ${b}$ is not known until $b$ is propagated, $R_n$ cannot set the \nf{super} field of ${b}$ while creating it.
+One approach is to set the \nf{super} field of ${b}$ by $R_n$ after propagating $b$ to $p$.
+This solution would not be efficient because there might be $p$ subblocks in the block $R_p$ propagated needing to update the \nf{super} field.
+However intuitively, once $b$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
+One idea is that if we know the approximate position of the superblock of $b$ then we can search for the real superblock when we wished to know the superblock of $b$ i.e.
+\nf{b.super} does not have to be the exact location of  the superblock of $b$, but we want it to be close to \nf{j}.
+We can set \nf{$b$.super} to \nf{$n$.parent.head} while creating $b$, but the problem is that there might be many \nf{$p$.Refresh}es that could happen after reading $p$\nf{.head} by $R_n$ and before propagating $b$ to $p$.
+If we set \nf{$b$.super} to \nf{$p$.head} after appending $b$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $b$ and before setting \nf{$b$.super}.
+In this case the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$b$.super}.
+
+Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}.
+After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or  maybe an \nf{$n$.parent.Refresh}.
+We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh} that propagates $b$.
 
@@
 
-\subsection{Pseudocode}
+\section{Pseudocode}
 \begin{algorithm}
@@
 \begin{definition}[Operations of a block]\label{def::ops}
-A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$.
-Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}.
+A block $b$ in a leaf represents an \nf{Enqueue} if $b\nf{.element}\neq\nf{null}$.
+Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue}.
 The set of operations of block $b$ is the union of the operations in leaf subblocks of $b$.
@@
 The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an internal node contains at least one operation, due to the test on Line~\ref{addOP}.
-We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.
+We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue} and \nf{Dequeue} routines.
 \end{proof}
@@
 From Definition \ref{ordering} of $E(n,b)$, so operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of $n$.
+By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks's subblocks in the left child of $n$.
 So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
@@
 The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{dequeue} in the root.
+Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
 \begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
@@
 
-  \caption{The number of \nf{dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
+  \caption{The number of \nf{Dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
 \end{figure}
@@
 \begin{lemma}[Amortized time analysis]
-\nf{Enqueue()} and \nf{Dequeue()}, each take $O(\log^2 p + \log q)$ steps in amortized analysis.
+\nf{Enqueue} and \nf{Dequeue}, each take $O(\log^2 p + \log q)$ steps in amortized analysis.
 Where $p$ is the number of processes and $q$ is the size of the queue at the time of invocation of operation.
@@
 
-A \nf{Dequeue()} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response.
+A \nf{Dequeue} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response.
 The first two part is similar to an \nf{Enqueue} operation.
@@
 \nf{IndexDequeue} does a search with range $p$ in each level which takes $O(log^2 p)$ in the tree.
-In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{enqueue} is invoked$)+\log$ size of the queue when \nf{dequeue} is invoked$)$.
+In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{Enqueue} is invoked$)+\log$ size of the queue when \nf{Dequeue} is invoked$)$.
 Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
@@
 \begin{lemma}
-An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
+An \nf{Enqueue} or \nf{Dequeue} operation, does at most $4\log p$ \nf{CAS} operations.
 \end{lemma}
@@
 \begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
+If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
 \end{lemma}
@@
 Suppose there were more than \nf{root.blocks[b].size} \nf{Dequeue}s in \nf{root.blocks[$b+1 \cdots end-1$]}.
-Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue()}.
+Then the element in the queue which is the response to the \nf{Dequeue} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue}.
 Furthermore in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.

### 2022-10-11 thesis/thesis.tex
complete method descriptions

@@
 
-\paragraph{\tt{Block}}
+\paragraph{Block}
 Block $b$ contains subblocks in the left and right children.
@@
 
-\paragraph{\texttt{Enqueue(e)}}
-\paragraph{\texttt{Dequeue()}}
-\paragraph{\texttt{FindResponse(b,i)}}
-\paragraph{\tt{Append(blk)}}
-\paragraph{\texttt{Refresh(n)}}
-\paragraph{\tt{Propagate()}}
-\paragraph{\texttt{CreateBlock}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children.
-\texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block.
-Other attributes are assigned values followed by definition.
+\paragraph{Enqueue(e)}
+An \nf{Enqueue} operation does not return a response, so to do an \nf{enqueue} operation in our algorithm it is sufficient to just append the \nf{Enqueue} operation to the root, then use its position in the linearization for future \nf{Dequeue} operations.
+\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
+
+\it{I believe the code is readable and we do not need references to code.}
+
+\paragraph{Dequeue()}
+\nf{Dequeue($e$)} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields, appends it to the tree.
+Then computes the position of the appended \nf{dequeue} operation in the root and after that finds the response of the \nf{dequeue} calling \nf{FindResponse}.
+
+\paragraph{FindResponse(b,i)}
+To compute the response of the $i$th \nf{dequeue} in $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
+If there are more dequeues than enqueues the queue would become empty before the requeusted dequeue.
+If the queue was not empty, Line \ref{ki} 222 computes the position of the eqneueue which is response to the dqeueue.
+Knowing the response is the $e$th enqueue in the root which is before the $b$th block we find the block and position contaiing the qnueue operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+
+\paragraph{Append(blk)}
+\nf{head} field is the index of the first empty index in \nf{blocks} array in a \nf{LeafBlock}.
+\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
+When \nf{Propagate} terminates it is guaranteed that the appended block is subblock of a block in the \nf{root}.
+There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
+
+\paragraph{n.Propagate()}
+\nf{Propagate($n$)} uses the double refresh idea and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
+Then invokes the \nf{Propagate} on \nf{$n$.parent} recursivly till it reaches the root.
+
+\paragraph{n.Refresh()}
+\nf{$n$.Refresh} goal is to  create a block of $n$'s children new blocks and append it to $n$\nf{.blocks}.
+From Lines 311 to 316 \ref{li} \nf{$n$.Refresh} helps to \nf{Advance} $n$'s children.
+\nf{Advance} increments the children's \nf{head} and set the \nf{super} field of the last block in $n$.
+The reason behind helping \nf{super} field is explained later in \nf{IndexDequeue} paragraph.
+After helping to \nf{Advance} children, the value $h$ from $n$\nf{.head} is read at Line \ref{readHead} and \nf{new} block is created in Line \ref{invokeCreateBlock}.
+\nf{new} is going to be inserted into $n\nf{.block[h]}$.
+The reason why reading \nf{head} and \nf{CreateBlock} are in two lines is that maybe the \nf{head} value changes from reading it till creating a block to be inserted to \nf{head}.
+Then if \nf{new} is empty \nf{Refresh} returns \nf{true} because there is no new operations to propagate  an it is unnecessary to add an empty block to the tree.
+Later we will assume all blocks contain at least one operation.
+Line \ref{cas} tries to install \nf{new}, if it was succsuful all is good, if not it means someone else has put a block earlier into \nf{head} or maybe the \nf{head} was not empty at first point.
+Anyway \nf{Refresh} tries to update the \nf{head} and \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
+
+\paragraph{CreateBlock} \texttt{$n$.CreateBlock(h)} returns a block containing new operations of \texttt{n}'s children.
+The \nf{new} block created in Line 333's fields are filled with symmtery for both left and right directions.
+\nf{index\sub{last}} is the index of the last subblock in the direction and \nf{index\sub{prev}} is the index of the prvious block of the first direct subblock to be aggregated into \nf{new}.
+With this said, \texttt{$n$.blocks[$i$\textsuperscript{$\prime$}].end\textsubscript{dir}} stores the index of the rightmost direct subblock of direction \nf{dir} child of \texttt{$n$.blocks[$i-1$]}.
+Then \nf{sum\sub{enq-dir}} is computed with the sum of the  from computing the number of enqueue operations in the \nf{new} block and the value stored in \texttt{$n$.blocks[$i-1$].sum\sub{enq-dir}}.
+\nf{sum\sub{deq-dir}} is also the same.
+Then if \nf{new} block is going to be installed in the \nf{root}, the \nf{size} field is computed.
+
+\it{should I get into details of computing size and sumenq?}
+
 \begin{figure}[hbt]
@@
 
-\paragraph{\texttt{DSearch(e,end)}}
-\paragraph{\texttt{GetEnqueue(b,i)}}
-\paragraph{\tt{IndexDequeue(b,i)}}
+\paragraph{DSearch(e,end)}
+We can show an operation in a node in two formats, the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block.
+If we know tuple of block and rank in the block of an operation ($E$) we can find the this tuple in the subblock containig the operation in poly log time.
+To find the response of a dequeue, the information we know about the response enquue is like the first way of showing the operation in the root ($E_i(\nf{root})$).
+So we have to find the block in the root containing the \nf{e}th enqueue in the root.
+We also know \nf{end} which the \nf{e}th enqueu is before or in the \nf{root.blocks[end]}.
+\nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line 806\ref{lo}).
+
+\paragraph{GetEnqueue(b,i)}
+\nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of \nf{i}th enqueue in $b$ th block of $n$.
+The range of subblocks of a block can be determined with \nf{end\sub{dir}} fields of the block and its previous block.
+Then the subblock can be found usig binary search on \nf{sum\sub{enq}} field.
+
+\paragraph{IndexDequeue(b,i)}
 
@@
 \Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
@@
 \EndFor \label{endHelpChild1}
+\State \tt{h= \head} \label{readHead}
 \State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}

### 2022-10-11 thesis/thesis.tex
apply oct 10 notes

@@
 
-In section 2 we talk about previous queues and their common problems. We also talk about polylogarithmic construction of shared objects.
+A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
+\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.
+\texttt{Dequeue()} removes and returns the first element among in the sequence.
+If the queue is empty it returns \texttt{null}.
+In section 2 we talk about previous queues and their common problems.
+We also talk about polylogarithmic construction of shared objects.
 
@@
 \pagebreak
-\section{Our Queue}
+\section{Queue Implementation}
 Paragraph titles will be removed after polishing.
 
-\paragraph{What is a Queue} In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem).
-A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence.
-If the queue is empty it returns \texttt{null}.
+\paragraph{What is a Queue} In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
+We design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of invocation.
+We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
 
 \paragraph{How we use Jayanti and why we are better} Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of using a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
-Despite their work, our algorithm them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
-In our model there are $p$ processes doing \nf{Enqueue(),Dequeue()} operations concurrently.
+We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
+Unlike their work, our algorithm does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
 
 \paragraph{Introduce tree} There is a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes.
-Each process has a leaf which the order of operations invoked by the process is stored in it.
+Each process has a leaf in which the operations invoked by the process are stored in order.
 When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-In each node of the tree the ordering of operations propagated up to it is stored.
-All processes agree on the sequence stored in the root and it is defined to be the linearization ordering.
+Each node of the tree keeps an ordering of operations propagated up to it.
+All processes agree on the sequence of operations in the root and this ordering is used as linearization ordering.
 \begin{figure}[hbt]
@@
 \end{center}
-\caption{\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored.
-Each node tries to propagate its operations up to the root, which stores the total ordering of all operations.}
+\caption{\label{fig::blocktree}Each of the processes $P_1,P_2...P_p$ has a leaf and in each node there is an ordering of operations stored.
+Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
 \end{figure}
@@
 
-\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, we can read the operations from node $n$'s both children that are not already in $n$, and then try to append them to the sequence stored in $n$.
-We call this procedure \tt{Refresh($n$)}.
-The main idea is that if a process invokes \tt{Refresh} on a node two times and in both times fail to append the new operations to $n$, the operations in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
-Because if both of the \tt{Refresh}$(n)$s fail to append then there is another instance of \nf{Refresh} in between which has succeeded to do append and has already appended the operations that the first \nf{Refresh} was trying to append.
+\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, and then tries to append them to the sequence stored in $n$.
+We call this procedure \tt{$n$.Refresh()}.
+A process doing a \nf{Refresh} on $n$ with successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent.
+The key idea is that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
+This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations after the first failed \nf{Refresh} and succeeded to do an append.
+This \nf{Refresh} appends the operations that the first failed \nf{Refresh} was trying to append.
 
@@
 \hspace{6em}
-\caption{\label{fig::propagstep} Before and after of a \sc{Refresh}$(n)$ with successful \nf{CAS}. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
+\caption{\label{fig::propagstep} Before and after of a \nf{$n$.Refresh} with successful append. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
 \end{center}
@@
   \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock}.}
+  \caption{The second failed Refresh is assuredly concurrent with a successful \texttt{Refresh}.}
 \end{figure}
@@
 
-\paragraph{Introduce CAS} We do not intend to get into our implementation but just for information we use \nf{CAS} (Compare \& Swap) operations to implement the Try to Append mechanism told in the previous paragraph.
-After a process appended its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node.
-So with poly-logarithmic \nf{CAS}es we can have a tree agreeing on the linearization.
-This solution makes us overcome the \nf{CAS} Retry Problem.
+\paragraph{Introduce CAS} We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s Try to Append mechanism in described in the previous paragraph.
+After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node.
+So with $O(\log p)$ \nf{CAS}es per operations we can have a tree agreeing on the linearization.
+This cooperative solution allows us to overcome the CAS Retry Problem.
 
@@
 \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
+\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
 \\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
 \\ \hline\end{tabular}} ] ]
-
 \end{tikzpicture}
-\caption{\label{fig::set} In each internal node, we store the set of all the operations propagated together, and one can arbitrarily linearize the sets of concurrent operations among themselves.}
+\caption{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
+In each internal node and one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh} among themselves.
+For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
+Since their execution time window have range in common they can be linearized in any order between themselves.}
 \end{center}
@@
 
-\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation's information is copied all the way up to the root.
-Instead we can keep track of some statistics of the operations propagated together.
-This allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{blocks} that contain the number of operations from the left and the right child in a \texttt{Refresh} procedure and only propagate the statistics block of the new operations, and we store arrays of blocks in each node of the tree In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
-Furthermore since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are in read one \textsc{Refresh} step in a node they are going to be propagated up to the root together.
+\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
+Instead we use implicit representation of the operations propagated together.
+Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
+We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}).
+Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
+To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step.
+A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure.
+A node stores an array of blocks of operations propagted up to it.
+A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
 Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
@@
 
-\paragraph{Introduce Get and Index} With what told till now we have a shared tree among procesess which they agree on the  implicit ordering stored in its root.
-Having an agreement by itself is not enough.
-A process may wish to know the $i$th propagated operation (1) or the rank of a propagated operation in the linearization (2).
-With these two queries we can design a universal construction.
-Let us do not focus on th universal construction and focus on how to use (1) and (2) to construct a queue.
-
-\paragraph{Introduce how to create Queue using Get and Index} In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with an specific rank.
-\nf{Enqueue}s and \nf{Dequeue}s are appended to the tree and when we want to find the response to a \nf{Dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null.
+\paragraph{Introduce Get and Index} So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
+With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ we can apply all the operations till $op$ in the root on a sequential instance of the object and then return the response.
+But Having this agreement by itself is not enough for an efficient queue.
+A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization.
+We will explain how to use (1) and (2) to construct a queue in the next paragraph.
 
 \paragraph{Explain pieces of a block helping to do Get and Index}
-After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}.
+After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2).
 Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
 
-To get the $i$th enqueue in the root, we can find the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element.
-To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to determine the range of subblocks of a block to make binary search faster.
-In each block, we store the prefix sum of left and right elements.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figures \ref{fig::pointer} and \ref{fig:prefix}).
-We know a block size is at most $p$, so binary search takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the search range which its size is \textsc{O}$(p)$.
+To get the $i$th enqueue in the root, we can find the block $B$ containing $i$th element in the root, and then recursively find the subblock of $B$ that contains $i$th element.
+To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search.
+We also store pointers to determine the range of subblocks of a block to make binary search faster.
+In each block, we store the prefix sum of operations from the left child and the right child.
+Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figures \ref{fig:prefix} and \ref{fig::pointer}).
+We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
 
@@
 
-\paragraph{Queue from tree}
-Now, we describe how to use the tree to implement a queue.
-Consider the following execution of operations.
-\texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree.
-What should a \texttt{Dequeue()} return?
-To compute the response of a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree.
-Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}.
-Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
+\paragraph{Introduce how to create Queue using Get and Index} In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank.
+\nf{Enqueue}s and \nf{Dequeue}s are appended to the tree and when we want to find the response to a \nf{Dequeue}, we compute the place of the \nf{dequeue} in the linearization.
+Then, using the rank of the \nf{Dequeue} among \nf{Deqeueue}s and the information about the size of the queue stored in the root we compute which \nf{Enqueue} is the answer to the \nf{Dequeue} or if the answer is null.
+
+\paragraph{}
+A non-null \nf{Dequeue} is one that returns a non-null value.
+If the queue is non-empty and $r$ \texttt{Dequeue} operations have returned a non-null response, then $i$th \texttt{Dequeue} returns the input of the $r+1$th \texttt{Enqueue}.
+So in order to answer a \nf{Dequeue}, it iss sufficient to know whether the queue is empty or not and the number of previous non-null \nf{dequeue}s.
+
+\paragraph{}
+We know we can linearize operations in a block in any order; here, we choose to decide to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
+In the next example, operations in a cell are concurrent.
+\texttt{Dequeue()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
+We will next describe how these values can be computed efficiently.
 
@@
 \centering
-  \begin{tabular}{c|c|c|c|c|c|c|c|c|c}
-    \hline \texttt{ENQ(5)}& \texttt{ENQ(2)}& \texttt{DEQ()}& \texttt{ENQ(3)}&\texttt{DEQ()}& \texttt{DEQ()}& \texttt{DEQ()}& \texttt{ENQ(4)}& \texttt{ENQ(6)}& \texttt{DEQ()}\\ \hline
+\begin{small}
+\begin{tabular}{c|c|c|c}
+    \hline \texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
   \end{tabular}
-  \caption{An example histoy of operations on the queue}
+\end{small}
+
+  \caption{An example history of operation blocks on the queue. \nf{Enqueue}s are shown with \nf{ENQ} and \nf{Dequeue}s are shown with \nf{DEQ}.}
 \end{table}
@@
 \paragraph{}
-A non-null dequeue is one that returns a non-null value.
-In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order.
-Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null.
-If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}.
-So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.
-
-%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
-
-%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}
-
-\paragraph{}
-In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate} steps and increase parallelism.
-So now the problem is to find the result of each Dequeue.
-From lemma \ref{blockSize} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations.
-In the next example, operations in a cell are concurrent.
-\texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
-We will next describe how these values can be computed efficiently.
+Now, we claimed that by knowing the size of the queue, we can compute the rank of the resulting \texttt{Enqueue()}.
+We apply this approach to blocks; if we store the size of the queue after each block of operations happens, we can compute each \nf{Dequeue}'s index of result in \textsc{O}$(1)$ steps.
 
@@
 \centering
-  \begin{tabular}{c|c|c|c}
-    \hline \texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
+\begin{small}
+  \begin{tabular}{c|c|c|c|c}
+    \hline &\texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
+    \tt{\#ENQs} & 0 & 3 & 1 & 1 \\ \hline
+    \tt{\#DEQs} & 1 & 1 & 1 & 4 \\ \hline
+    \tt{size} & 0 & 2 & 2 & 0 \\ \hline
   \end{tabular}
-  \caption{An example history of operation blocks on the queue}
-\end{table}
-
-\paragraph{}
-Now, we claimed that by knowing the current size of the queue and the number of non-null dequeue operations before the current dequeue, we could compute the index of the resulting \texttt{Enqueue()}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations happens and the number of non-null dequeues dequeues till a block, we can compute each dequeue's index of result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbt]
-\centering
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{DEQ()} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ()}& \texttt{ENQ(3)}, \texttt{DEQ()}&  \texttt{ENQ(4)}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}, \texttt{DEQ()}\\ \hline
-    \#enqueues & 0 & 3 & 1 & 1 \\ \hline
-        \#dequeues & 1 & 1 & 1 & 4 \\ \hline
-            \#non-null dequeues & 0 & 1 & 2 & 5 \\ \hline
-                size & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
+\end{small}
   \caption{Augmented history of operation blocks on the queue}
@@
 
-Size and the number of non-null dequeues for $b$th block could be computed this way:\\
-\texttt{size[b]= max(size[b-1] +enqueues[b] -dequeues[b], 0)}\\
-\texttt{non-null dequeues[b]= non-null dequeues[b-1] +dequeues[b] -size[b-1] -enqueues[b]}
-
-Given \texttt{DEQ} is in block \texttt{b}, \texttt{response(DEQ)} would be:\\
-\texttt{(size[b-1]- index of DEQ in the block's dequeus >=0) ? ENQ[non-null dequeus[b-1]+ index of DEQ in the block's dequeus] : null;}
+Size of the queue the $b$th block could be computed using these equations.
+$$\texttt{size[b] = max(size[b-1] + enqueues[b] - dequeues[b], 0)}$$
+$$\texttt{non-null dequeues[b] = enqueues[b] - size[b]}$$
+
+Given \texttt{DEQ} is in block $B$, response of \nf{DEQ}  is the \tt{Enqueue} with index \nf{non-null Dequeus[b-1]+ index of DEQ in the block's Dequeus} in the root, if \texttt{(size[b-1]- index of DEQ in the $B$'s Dequeus >=0)}.
+Otherwise it would be \nf{ null}
 
@@
 Block $b$ contains subblocks in the left and right children.
-WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$.
+The left subblocks of $b$ are some consecutive blocks in the left child starting from where the block prior to $b$ ended.
 See Figure \ref{fig::createBlock} .
@@
 We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s.
-In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}.
-Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
+In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
+Furthermore, in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
 There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh} step.

### 2022-10-11 thesis/thesis.tex
update double refresh fig
apply Oct 11 notes

@@
 All processes agree on the sequence of operations in the root and this ordering is used as linearization ordering.
-\begin{figure}[hbt]
+\begin{figure}[hbtp]
 \begin{center}
@@
 \end{figure}
-\it{TODO::Add sequence to nodes}
-
-\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, and then tries to append them to the sequence stored in $n$.
+
+\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
 We call this procedure \tt{$n$.Refresh()}.
@@
 
-\begin{figure}[hbt]
+\begin{figure}[hbtp]
 \begin{center}
@@
   \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
 \Tree [.$r_1,l_1,l_2,r_2,l_3$
@@
   \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
+  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
 \Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
@@
 
-\begin{figure}[hbt]
-  \center\includegraphics[width=4in]{pics/doublerefresh.png}
-  \caption{The second failed Refresh is assuredly concurrent with a successful \texttt{Refresh}.}
-\end{figure}
-\it{TODO::simplify the Figure, remove code invocations}
+\begin{figure}[hbpt]
+  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
+  \caption{The second failed Refreshof $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its operations after the first failed \nf{Refresh}.}
+\end{figure}
 
@@
 
-\begin{figure}[h]
+\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
+Instead we use implicit representation of the operations propagated together.
+Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
+We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}).
+Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
+To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step.
+A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
+A node stores an array of blocks of operations propagated up to it.
+A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
+Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
+In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.
+
+\begin{figure}[hbtp]
 \begin{center}
-\begin{tikzpicture}
+\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
 
@@
 
-\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
-Instead we use implicit representation of the operations propagated together.
-Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}).
-Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step.
-A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure.
-A node stores an array of blocks of operations propagted up to it.
-A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
-Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.
-
-\begin{figure}
+\begin{figure}[hbtp]
 \begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0.5cm},
-	level 3/.style={level distance=1.8cm,sibling distance=1.2cm}]
-
-\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} $\vdots$ $\vdots$ ] ] ]
-
+\begin{tikzpicture}
+[level 1/.style={level distance=3.3cm,sibling distance=0cm},
+	level 2/.style={level distance=2.5cm,sibling distance=0cm},
+	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
+\begin{small}
+
+\Tree [.{\begin{tabular}{||c|c|c||}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
+ [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
+ [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
+  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
+          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
+\end{small}
 \end{tikzpicture}
@@
 \caption{\label{fig:block} Showing concurrent operation sets with blocks.
+Blocks between two lines $||$ are propagated together to the parent.
 Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
-Block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
-Blocks between two lines $||$ are propagated together to the parent.
-For example, Blocks (2,3) and (4,1) from the leftmost leaf and (0,2) from its sibling are propagated together into the block (10,2) in their parent.
-The number underneath a group of blocks in a node indicates which block in the node's parent those blocks were propagated to.
-Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$.
-For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
-Block (5,3) also points to elements from blocks (0,5) and (1,2).
-We choose to linearize operations in a block from the left child before those from the right child as a convention.
-Operations within a block of the root can be ordered in any way that is convenient.
-In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids.
-So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2).
-All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
-}
+For example, block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
+The third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
+Block (5,3) is superblock of blocks (0,5) and (1,2) and blocks (5,1),(3,5) and (4,2) are subblocks of (14,6).}
 \end{figure}
@@
 After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2).
-Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.
-
-To get the $i$th enqueue in the root, we can find the block $B$ containing $i$th element in the root, and then recursively find the subblock of $B$ that contains $i$th element.
+Each block stores an extra constant amount of information (like prefix sums) to allow binary search to find the required block in a node quickly
+
+To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element.
 To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search.
@@
 In each block, we store the prefix sum of operations from the left child and the right child.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figures \ref{fig:prefix} and \ref{fig::pointer}).
+Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}).
 We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
 
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[level 1/.style={level distance=2.3cm,sibling distance=1cm}]
-
-\Tree [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline  0,0 & 3,8 & 9,8 & 23,14\end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline 0,0 & 0,3 & 4,5 & 9,6 \end{tabular}} ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline 0,0 & 2,1 & 7,1 & 11,3 & 14,7 \end{tabular}} ] ]
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:prefix} Using Prefix sums in blocks.
-When we want to find block b elements in its children, we can use binary search.
-The number below each block shows the count of elements in the previous blocks.}
-\end{figure}
-
-\begin{figure}[hbt]
+To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in.
+After a block is installed we set the approximate index of the superblock of the block in it.
+
+\begin{figure}[hbtp]
 \centering
-  \includegraphics[width=5in]{pics/pointers}
-  \caption{Block have pointers to the starting block of theirs for each child. \label{fig::pointer}}
-\end{figure}
-
-\paragraph{Introduce how to create Queue using Get and Index} In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank.
-\nf{Enqueue}s and \nf{Dequeue}s are appended to the tree and when we want to find the response to a \nf{Dequeue}, we compute the place of the \nf{dequeue} in the linearization.
-Then, using the rank of the \nf{Dequeue} among \nf{Deqeueue}s and the information about the size of the queue stored in the root we compute which \nf{Enqueue} is the answer to the \nf{Dequeue} or if the answer is null.
-
-\paragraph{}
-A non-null \nf{Dequeue} is one that returns a non-null value.
-If the queue is non-empty and $r$ \texttt{Dequeue} operations have returned a non-null response, then $i$th \texttt{Dequeue} returns the input of the $r+1$th \texttt{Enqueue}.
-So in order to answer a \nf{Dequeue}, it iss sufficient to know whether the queue is empty or not and the number of previous non-null \nf{dequeue}s.
-
-\paragraph{}
+  \includegraphics[width=4in, height=1.4in]{pics/pointers}
+  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
+\end{figure}
+
+\paragraph{Introduce how to create Queue using Get and Index}
+
+In an execution on a queue where no dequeue operation returns null, the $k$th dequeue returns the element of the $k$th enqueue.
+In general case a dequeue returns null if and only if the size of the queue after the previous operation is 0.
+Otherwise if the the dequeue is the $k$th non null returning dequeue it returns the element of $k$th enqueue.
+Having the size of the  queue after an operation we can compute the number of non null returning dequeues  from the number of enqueues up to block B and size of the queue after B.
+So if we store the size of the queue after each block operations in the root we can compute the index of the enqueue which is response to a dequeue in constant time.
+
+In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank.
 We know we can linearize operations in a block in any order; here, we choose to decide to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
@@
 \texttt{Dequeue()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
-We will next describe how these values can be computed efficiently.
-
-\begin{table}[hbt]
+
+\begin{table}[hbtp]
 \centering
@@
 
-\begin{table}[hbt]
+\begin{table}[hbtp]
 \centering
@@
 
+\it{Do you think the remainig part is overkill? I think it is to much detail which is also repeated in the proof.}
+
 Size of the queue the $b$th block could be computed using these equations.
@@
 Given \texttt{DEQ} is in block $B$, response of \nf{DEQ}  is the \tt{Enqueue} with index \nf{non-null Dequeus[b-1]+ index of DEQ in the block's Dequeus} in the root, if \texttt{(size[b-1]- index of DEQ in the $B$'s Dequeus >=0)}.
-Otherwise it would be \nf{ null}
-
-\begin{figure}[hbt]
-\centering
-  \includegraphics[width=5in]{pics/queue}
-  \caption{Fields stored in the Queue nodes. \label{fig::queue}}
-\end{figure}
+Otherwise it would be \nf{null}
 
@@
 
-\paragraph{Block}
+\paragraph{Node} (\it{Repetitive as the fields description?})
+ In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
+
+\paragraph{Block} (\it{Repetitive as the fields description?})
 Block $b$ contains subblocks in the left and right children.
 The left subblocks of $b$ are some consecutive blocks in the left child starting from where the block prior to $b$ ended.
-See Figure \ref{fig::createBlock} .
-
-We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s.
-In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-Furthermore, in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}.
-There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh} step.
-It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases.
-Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}.
-This helps us to know which blocks have aggregated together to a block, not precisely though.
-We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.
-
-In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{num\textsubscript{enq-left}}, \texttt{num\textsubscript{deq-left}}, \texttt{num\textsubscript{enq-right}}, \texttt{num\textsubscript{deq-right}}.
-In order to make \texttt{BinarySearch()}es faster we store prefix sums as well and there are some more general stats that help to make pseudocode more readable but not necessary.
-
-To compute the head of the \texttt{queue} before a \texttt{dequeue} two more fields are stored in the root \texttt{size} and \texttt{sum\textsubscript{non-null deq}}.
-\texttt{size} in a \texttt{block} shows the number of elements after the \texttt{block} has finished and \texttt{sum\textsubscript{non-null deq}} is the total number of non-null dequeues till the \texttt{block}.
+In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
+\nf{end\sub{left}} and \nf{end\sub{right}} field show the last subblock of a block in left and right child.
+\texttt{size} in a \texttt{block} in \nf{root} stores the size of the queue after the \texttt{block} has finished.
 
@@
 
-\it{I believe the code is readable and we do not need references to code.}
+\it{I believe the code is readable enough and we do not need references to code.}
 
@@
 If there are more dequeues than enqueues the queue would become empty before the requeusted dequeue.
-If the queue was not empty, Line \ref{ki} 222 computes the position of the eqneueue which is response to the dqeueue.
+If the queue was not empty, Line \ref{computeE} computes the position of the eqneueue which is response to the dqeueue.
 Knowing the response is the $e$th enqueue in the root which is before the $b$th block we find the block and position contaiing the qnueue operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
@@
 \nf{$n$.Refresh} goal is to  create a block of $n$'s children new blocks and append it to $n$\nf{.blocks}.
-From Lines 311 to 316 \ref{li} \nf{$n$.Refresh} helps to \nf{Advance} $n$'s children.
+Lines \ref{startHelpChild1}-\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
 \nf{Advance} increments the children's \nf{head} and set the \nf{super} field of the last block in $n$.
@@
 
-\it{should I get into details of computing size and sumenq?}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/createblock}
-  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
-\end{figure}
+\it{Should I get into details of computing size and sumenq?}
+
+%\begin{figure}[hbt]
+%  \center\includegraphics[width=5.5in]{pics/createblock}
+%  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
+%\end{figure}
 
@@
 We also know \nf{end} which the \nf{e}th enqueu is before or in the \nf{root.blocks[end]}.
-\nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line 806\ref{lo}).
+\nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line \ref{dsearchBinarySearch}).
 
@@
 \Function{boolean}{Refresh()}{}
+\State \tt{h= \head} \label{readHead}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
@@
 \EndFor \label{endHelpChild1}
-\State \tt{h= \head} \label{readHead}
 \State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}

### 2022-10-11 thesis/thesis.tex
update refrences

@@
 
-\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would no be poly-logarithmic time.
+\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be poly-logarithmic time.
 Instead we use implicit representation of the operations propagated together.
@@
 Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}, a piece of information about the operations in a \nf{Refresh} step.
-A block contain the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
+To do that, we introduce \it{block}s.
+A block is a piece of information about the operations in a \nf{Refresh} step.
+It contains the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
 A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in children into a new block and puts it in the parent.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
-Furthermore, since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
+A propagate step  aggregates the new blocks in children into a new block and puts it in the parents' blocks.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently which means there are less than or equal $p$ operations in a block.
+Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
 Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
@@
 \paragraph{Introduce Get and Index} So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ we can apply all the operations till $op$ in the root on a sequential instance of the object and then return the response.
-But Having this agreement by itself is not enough for an efficient queue.
+With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations till $op$ in the root on a sequential instance of the object and then return the response.
+Having this agreement by itself is not enough for an efficient queue.
 A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization.
-We will explain how to use (1) and (2) to construct a queue in the next paragraph.
+We will explain how to use (1) and (2) to construct a fast queue in the next paragraph.
 
@@
 After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-Each block stores an extra constant amount of information (like prefix sums) to allow binary search to find the required block in a node quickly
-
 To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element.
@@
 We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
 To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in.

### 2022-10-12 thesis/thesis.tex
add Future Directions

@@
 \pagebreak
-\subsection{Garbage Collection or Getting rid of the infinite Arrays}
-
-\pagebreak
-\section{Using Queues to Implement Vectors}
-Supporting Append, Read, Write in PolyLog time by modifying Get(Enq) Method. Create a Universal Construction Using our vector
-
-\pagebreak
-\section{Conclusion}
-possible directions for work
-
-Maybe Stacks
-
-Characterize what datastructure can be used for this approach, we already know: queue, fetch \& Inc, Vectors
-
-\section*{References}
+\section{Future directions}
+\paragraph{Infinite Sized Arrays}
+Our queue algorithm's usage of memory has two places to improve.
+First, we assumed arrays in the nodes are infinite sized.
+Second we do not handle garbage collection, dequeued operations are still remain in the nodes.
+We can use list to arrays to resolve the infinite sized arrays issue.
+To implement an infinite shared array we can create a small array of pointers called \nf{arr}.
+When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor]$]} points to an array or not.
+If not it creates a shared array with size $2^i$ and tries to \nf{CAS} it to the created array.
+After \nf{CAS} terminates whether it is successful or not \nf{arr[$\lfloor\log $ head$\rfloor]$]} points to an array.
+When a process wishes to access to $i$th element it looks upto \nf{arr[$\lfloor\log i\rfloor]$][$i-2^{\lfloor\log i\rfloor]}$]} Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ times \nf{CAS} has happened.
+And if the size of the all shared memory is $k$, \nf{arr} can be initialized to a $\log k$ sized array.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
+  \caption{Infinite sized Arrays}
+  \label{fig::doublingArray}
+\end{figure}
+
+\paragraph{Garbage Collection}
+To garbage collect we can keep track of the index of the most recent \nf{block} in the \nf{root} that has been term and then split the blocks from that \nf{block}.
+We can use Persistent Red Black Trees because they support split in an efficient method.
+
+\paragraph{Poly-logarithmic Wait-free Vectors}
+If we do not distinct between \nf{Enqueue}s and \nf{Dequeue}s and replace \nf{DSearch} with Binary Search on all the \nf{root} in our algorithm, we achieve a list with poly-logarithmic \nf{Append}, \nf{Get} and \nf{Index} methods with respect to the number of \nf{Append}s done.
+
+\paragraph{Stacks or maybe other objects that store sequences}
+The reason behind why our \nf{tree} worked well to implement a queue was two things.
+First, we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} fast.
+Two, the response of the \nf{Dequeue} is not very far from it (size of the queue).
+We think we can create wait-free poly-logarithmic version of other objects which their operations satisfy these two conditions.
+
 \bibliography{main.bib} 

### 2022-10-13 thesis/thesis.tex
apply  Oct 12 notes

@@
 \usepackage{enumitem}
+\usepackage{bold-extra}
 
@@
 \pagebreak
-\section{Queue Implementation}
-Paragraph titles will be removed after polishing.
-
-\paragraph{What is a Queue} In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
+\section{Queue Implementation} \label{DescriptQ}
+
+In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
 We design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of invocation.
@@
 
-\paragraph{How we use Jayanti and why we are better} Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
+Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
 We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations.
-
-\paragraph{Introduce tree} There is a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes.
+Unlike their work, our algorithm does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent dequeuers.
+
+There is a shared binary tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes.
 Each process has a leaf in which the operations invoked by the process are stored in order.
@@
 Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as linearization ordering.
+All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
 \begin{figure}[hbtp]
@@
 \end{center}
-\caption{\label{fig::blocktree}Each of the processes $P_1,P_2...P_p$ has a leaf and in each node there is an ordering of operations stored.
+\caption{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
 Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
@@
 
-\paragraph{Introduce Double Refresh} To do a propagate step on node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh()}.
-A process doing a \nf{Refresh} on $n$ with successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent.
-The key idea is that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in the node after the second failed \tt{Refresh}.
-This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations after the first failed \nf{Refresh} and succeeded to do an append.
-This \nf{Refresh} appends the operations that the first failed \nf{Refresh} was trying to append.
+To propagate operations to node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
+We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}).
+A \nf{Refresh} on $n$ with a successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent.
+We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
+This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations from $n$'s children after the first failed \nf{Refresh} and succeeded to do an append on $n$.
+This \nf{Refresh} helps to append the new operations in the $n$' children before the first failed \nf{Refresh}, in case they were not already appended.
 
@@
 \hspace{6em}
-\caption{\label{fig::propagstep} Before and after of a \nf{$n$.Refresh} with successful append. Operations propagating from the left child are numbered with $l$ and from the right child by $r$.}
+\caption{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
 \end{center}
@@
   \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption{The second failed Refreshof $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its operations after the first failed \nf{Refresh}.}
-\end{figure}
-
-\paragraph{Introduce CAS} We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s Try to Append mechanism in described in the previous paragraph.
+  \caption{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive \nf{Refresh}es.}
+\end{figure}
+
+The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh}.
+This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$' second failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
+We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
 After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node.
-So with $O(\log p)$ \nf{CAS}es per operations we can have a tree agreeing on the linearization.
+So with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
 This cooperative solution allows us to overcome the CAS Retry Problem.
 
-\paragraph{Introduce Blocks} It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be poly-logarithmic time.
-Instead we use implicit representation of the operations propagated together.
-Furthermore we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-We can only keep track of set of operations in each \nf{Refresh} and then define the linearization only in root (see Figure \ref{fig::set}).
+It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
+Instead we use an implicit representation of the operations propagated together.
+Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
+We can only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in root (see Figure \ref{fig::set}).
 Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
 To do that, we introduce \it{block}s.
-A block is a piece of information about the operations in a \nf{Refresh} step.
-It contains the number of operations from the left and the right child in a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
+A block stores information about the operations in a \nf{Refresh} step.
+It contains the number of operations from the left and the right child propagated to the node by a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
 A node stores an array of blocks of operations propagated up to it.
 A propagate step  aggregates the new blocks in children into a new block and puts it in the parents' blocks.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently which means there are less than or equal $p$ operations in a block.
+We call the aggregated blocks, subblocks of the new block and the new block, superblock of them.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+Thus, there are at most $p$ operations in a block.
 Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
@@
 \caption{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node and one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh} among themselves.
+In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
 For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time window have range in common they can be linearized in any order between themselves.}
+Since their execution time intervals overlap, they can be linearized in any order.}
 \end{center}
@@
 \end{center}
-\caption{\label{fig:block} Showing concurrent operation sets with blocks.
+\caption{\label{fig:block} Using blocks to represent operations.
 Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair(left, right) indicating the number of operations from the left and the right child, respectively.
-For example, block (12,6) in the root contains blocks (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child.
-Block (5,3) is superblock of blocks (0,5) and (1,2) and blocks (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-\paragraph{Introduce Get and Index} So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations till $op$ in the root on a sequential instance of the object and then return the response.
-Having this agreement by itself is not enough for an efficient queue.
+Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
+For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
+The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
+(5,3) is superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
+\end{figure}
+
+So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
+With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$.
+However, this approach is not enough for an efficient queue.
 A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization.
-We will explain how to use (1) and (2) to construct a fast queue in the next paragraph.
-
-\paragraph{Explain pieces of a block helping to do Get and Index}
+We explain how to implement (1) and (2) and use them to construct a fast queue below.
+
 After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing $i$th operation in the root, and then recursively find the subblock of $B$ that contains $i$th element.
-To make this recursive search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search.
-We also store pointers to determine the range of subblocks of a block to make binary search faster.
-In each block, we store the prefix sum of operations from the left child and the right child.
+To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in a child of the root that contains that $i$th operation.
+To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
+We also store pointers to determine the range of subblocks of a block to make the binary search faster.
+In each block, we store the prefix sum of operations from the left child and from the right child.
 Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}).
@@
 To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in.
-After a block is installed we set the approximate index of the superblock of the block in it.
+After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
 
@@
 
-\paragraph{Introduce how to create Queue using Get and Index}
-
-In an execution on a queue where no dequeue operation returns null, the $k$th dequeue returns the element of the $k$th enqueue.
-In general case a dequeue returns null if and only if the size of the queue after the previous operation is 0.
-Otherwise if the the dequeue is the $k$th non null returning dequeue it returns the element of $k$th enqueue.
-Having the size of the  queue after an operation we can compute the number of non null returning dequeues  from the number of enqueues up to block B and size of the queue after B.
-So if we store the size of the queue after each block operations in the root we can compute the index of the enqueue which is response to a dequeue in constant time.
-
-In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a \nf{Dequeue} and one tries to get an \nf{Enqueue} with an specific rank.
-We know we can linearize operations in a block in any order; here, we choose to decide to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-In the next example, operations in a cell are concurrent.
-\texttt{Dequeue()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
-
+In an execution on a queue where no dequeue operation returns null, the $k$th dequeue returns the argument of the $k$th enqueue.
+In the general case a dequeue returns null if and only if the size of the queue after the previous operation is 0.
+We refer to such a dequeue as a null dequeue.
+Otherwise, if the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
+Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B.
+So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue which is the response to a given dequeue in constant time.
+
+In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific rank.
+We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
+Consider the follwing operations, where operations in a cell are concurrent.
 \begin{table}[hbtp]
@@
 \begin{tabular}{c|c|c|c}
-    \hline \texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
+    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
   \end{tabular}
 \end{small}
-
-  \caption{An example history of operation blocks on the queue. \nf{Enqueue}s are shown with \nf{ENQ} and \nf{Dequeue}s are shown with \nf{DEQ}.}
 \end{table}
 
-\paragraph{}
-Now, we claimed that by knowing the size of the queue, we can compute the rank of the resulting \texttt{Enqueue()}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations happens, we can compute each \nf{Dequeue}'s index of result in \textsc{O}$(1)$ steps.
+\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
+Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
+We apply this approach to blocks; if we store the size of the queue after each block of operations happens, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
 
@@
   \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{DEQ} & \texttt{ENQ(5)}, \texttt{ENQ(2)}, \texttt{ENQ(1)}, \texttt{DEQ}& \texttt{ENQ(3)}, \texttt{DEQ}&  \texttt{ENQ(4)}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}, \texttt{DEQ}\\ \hline
-    \tt{\#ENQs} & 0 & 3 & 1 & 1 \\ \hline
-    \tt{\#DEQs} & 1 & 1 & 1 & 4 \\ \hline
-    \tt{size} & 0 & 2 & 2 & 0 \\ \hline
+    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
+    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
+    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
+    Size at end & 0 & 2 & 2 & 0 \\ \hline
   \end{tabular}
 \end{small}
-  \caption{Augmented history of operation blocks on the queue}
+  \caption{Augmented history of operation blocks on the queue.}
 \end{table}
 
-\it{Do you think the remainig part is overkill? I think it is to much detail which is also repeated in the proof.}
-
-Size of the queue the $b$th block could be computed using these equations.
-$$\texttt{size[b] = max(size[b-1] + enqueues[b] - dequeues[b], 0)}$$
-$$\texttt{non-null dequeues[b] = enqueues[b] - size[b]}$$
-
-Given \texttt{DEQ} is in block $B$, response of \nf{DEQ}  is the \tt{Enqueue} with index \nf{non-null Dequeus[b-1]+ index of DEQ in the block's Dequeus} in the root, if \texttt{(size[b-1]- index of DEQ in the $B$'s Dequeus >=0)}.
-Otherwise it would be \nf{null}
-
-\subsection{Details of Implementation}
-
-\paragraph{Node} (\it{Repetitive as the fields description?})
- In each \texttt{node} we store pointers to its parent and children, an array of \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{Block} (\it{Repetitive as the fields description?})
-Block $b$ contains subblocks in the left and right children.
-The left subblocks of $b$ are some consecutive blocks in the left child starting from where the block prior to $b$ ended.
-In each \texttt{block} we store 4 essential stats that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-\nf{end\sub{left}} and \nf{end\sub{right}} field show the last subblock of a block in left and right child.
-\texttt{size} in a \texttt{block} in \nf{root} stores the size of the queue after the \texttt{block} has finished.
-
-\paragraph{Enqueue(e)}
-An \nf{Enqueue} operation does not return a response, so to do an \nf{enqueue} operation in our algorithm it is sufficient to just append the \nf{Enqueue} operation to the root, then use its position in the linearization for future \nf{Dequeue} operations.
+\noindent The size of the queue after the $b$th block in the root could be computed as
+$$\textrm{max}\Big(\textrm{size after }b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
+\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
+$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
+
+Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
+Otherwise the response would be \nf{null}.
+
+\subsection{Details of the Implementation}
+Pseudocode for the queue implementation is given in Section \ref{algQ}.
+\paragraph{Node} 
+ In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
+
+\paragraph{Block}
+ The information stores in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
+If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation.
+If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
+The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the block prior to $B$ ended.
+In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
+\texttt{sum\textsubscript{enq-left}} is the total number of \nf{Enqueue} operations in theblocks before the last subblock of $B$ in the left child child, other fields semantics is also similar.
+The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively.
+The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
+
+\paragraph{\tt{Enqueue($e$)}}
+An \nf{Enqueue} operation does not return a response, so it is sufficient to just propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
 \nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
 
-\it{I believe the code is readable enough and we do not need references to code.}
-
-\paragraph{Dequeue()}
-\nf{Dequeue($e$)} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields, appends it to the tree.
-Then computes the position of the appended \nf{dequeue} operation in the root and after that finds the response of the \nf{dequeue} calling \nf{FindResponse}.
-
-\paragraph{FindResponse(b,i)}
-To compute the response of the $i$th \nf{dequeue} in $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more dequeues than enqueues the queue would become empty before the requeusted dequeue.
-If the queue was not empty, Line \ref{computeE} computes the position of the eqneueue which is response to the dqeueue.
-Knowing the response is the $e$th enqueue in the root which is before the $b$th block we find the block and position contaiing the qnueue operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{Append(blk)}
-\nf{head} field is the index of the first empty index in \nf{blocks} array in a \nf{LeafBlock}.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates it is guaranteed that the appended block is subblock of a block in the \nf{root}.
+\paragraph{\tt{Dequeue()}}
+\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields, appends it to the tree.
+Then it computes the position of the appended \nf{Dequeue} operation in the root and after that finds the response of the \nf{Dequeue} calling \nf{FindResponse}.
+
+\paragraph{\tt{FindResponse($b, i$)}}
+To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
+If there are more \nf{Dqeueue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dqeueue}.
+If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dqeueue}.
+Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block we find the block) and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+
+\paragraph{\tt{Append($B$)}}
+The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
+\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
+When \nf{Propagate} terminates it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
 There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
 
-\paragraph{n.Propagate()}
-\nf{Propagate($n$)} uses the double refresh idea and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then invokes the \nf{Propagate} on \nf{$n$.parent} recursivly till it reaches the root.
-
-\paragraph{n.Refresh()}
-\nf{$n$.Refresh} goal is to  create a block of $n$'s children new blocks and append it to $n$\nf{.blocks}.
-Lines \ref{startHelpChild1}-\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} and set the \nf{super} field of the last block in $n$.
-The reason behind helping \nf{super} field is explained later in \nf{IndexDequeue} paragraph.
-After helping to \nf{Advance} children, the value $h$ from $n$\nf{.head} is read at Line \ref{readHead} and \nf{new} block is created in Line \ref{invokeCreateBlock}.
-\nf{new} is going to be inserted into $n\nf{.block[h]}$.
+\paragraph{\tt{Propagate()}}
+\nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
+Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively untill it reaches the root.
+
+\paragraph{\tt{Refresh()}}
+The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
+The value $h$ from $n$\nf{.head} is read at Line \ref{readHead}.
+The new block created by \nf{Refresh} qill be inserted into $n\nf{.block[h]}$.
+Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
+\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks.
+The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
+After helping to \nf{Advance} children, \nf{new} block is created in Line \ref{invokeCreateBlock}.
 The reason why reading \nf{head} and \nf{CreateBlock} are in two lines is that maybe the \nf{head} value changes from reading it till creating a block to be inserted to \nf{head}.
@@
 
-\paragraph{CreateBlock} \texttt{$n$.CreateBlock(h)} returns a block containing new operations of \texttt{n}'s children.
+\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock(h)} returns a block containing new operations of \texttt{n}'s children.
 The \nf{new} block created in Line 333's fields are filled with symmtery for both left and right directions.
@@
 
-\paragraph{DSearch(e,end)}
+\paragraph{\tt{DSearch($e,end$)}}
 We can show an operation in a node in two formats, the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block.
@@
 
-\paragraph{GetEnqueue(b,i)}
+\paragraph{\tt{GetEnqueue($b,i$)}}
 \nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of \nf{i}th enqueue in $b$ th block of $n$.
@@
 
-\paragraph{IndexDequeue(b,i)}
+\paragraph{\tt{IndexDequeue($b,i$)}}
 
@@
 
-\section{Pseudocode}
+\subsection{Pseudocode} \label{algQ}
 \begin{algorithm}
@@
 \begin{itemize}
- \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for b$\geq$0 and x $\in$ \{enq, deq\}})
+ \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for internal blocks where b$\geq$0 and x $\in$ \{enq, deq\}})
 %  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for b>0 and x $\in$ \{ enq, deq, enq-left, enq-right, deq-left, deq-right\})}
+  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for all blocks where b>0 and x $\in$ \{ enq, deq, enq-left, enq-right, deq-left, deq-right\})}
 \end{itemize}
@@
 
-\Function{void}{Append}{\sl{block} blk} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= blk} 
+\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
+\State \tt{blocks[\head]= B} 
 \State \tt{\head+=1} \label{appendEnd} 

### 2022-10-15 thesis/thesis.tex
apply Oct 13 notes

@@
 The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively.
+The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
 The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
@@
 The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The value $h$ from $n$\nf{.head} is read at Line \ref{readHead}.
-The new block created by \nf{Refresh} qill be inserted into $n\nf{.block[h]}$.
+The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
+The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
 Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
@@
 After helping to \nf{Advance} children, \nf{new} block is created in Line \ref{invokeCreateBlock}.
-The reason why reading \nf{head} and \nf{CreateBlock} are in two lines is that maybe the \nf{head} value changes from reading it till creating a block to be inserted to \nf{head}.
-Then if \nf{new} is empty \nf{Refresh} returns \nf{true} because there is no new operations to propagate  an it is unnecessary to add an empty block to the tree.
-Later we will assume all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}, if it was succsuful all is good, if not it means someone else has put a block earlier into \nf{head} or maybe the \nf{head} was not empty at first point.
-Anyway \nf{Refresh} tries to update the \nf{head} and \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock(h)} returns a block containing new operations of \texttt{n}'s children.
-The \nf{new} block created in Line 333's fields are filled with symmtery for both left and right directions.
-\nf{index\sub{last}} is the index of the last subblock in the direction and \nf{index\sub{prev}} is the index of the prvious block of the first direct subblock to be aggregated into \nf{new}.
-With this said, \texttt{$n$.blocks[$i$\textsuperscript{$\prime$}].end\textsubscript{dir}} stores the index of the rightmost direct subblock of direction \nf{dir} child of \texttt{$n$.blocks[$i-1$]}.
-Then \nf{sum\sub{enq-dir}} is computed with the sum of the  from computing the number of enqueue operations in the \nf{new} block and the value stored in \texttt{$n$.blocks[$i-1$].sum\sub{enq-dir}}.
-\nf{sum\sub{deq-dir}} is also the same.
-Then if \nf{new} block is going to be installed in the \nf{root}, the \nf{size} field is computed.
-
-\it{Should I get into details of computing size and sumenq?}
-
-%\begin{figure}[hbt]
-%  \center\includegraphics[width=5.5in]{pics/createblock}
-%  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
-%\end{figure}
-
-\paragraph{\tt{DSearch($e,end$)}}
-We can show an operation in a node in two formats, the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block.
-If we know tuple of block and rank in the block of an operation ($E$) we can find the this tuple in the subblock containig the operation in poly log time.
-To find the response of a dequeue, the information we know about the response enquue is like the first way of showing the operation in the root ($E_i(\nf{root})$).
-So we have to find the block in the root containing the \nf{e}th enqueue in the root.
-We also know \nf{end} which the \nf{e}th enqueu is before or in the \nf{root.blocks[end]}.
-\nf{DSearch} uses the Doubling idea to find the range that the answer block is there (Lines \ref{dsearchStart}-\ref{dsearchEnd}) and then tries to find the response with binary search (Line \ref{dsearchBinarySearch}).
-
-\paragraph{\tt{GetEnqueue($b,i$)}}
-\nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of \nf{i}th enqueue in $b$ th block of $n$.
-The range of subblocks of a block can be determined with \nf{end\sub{dir}} fields of the block and its previous block.
-Then the subblock can be found usig binary search on \nf{sum\sub{enq}} field.
+Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there is no new operations to propagate and it is unnecessary to add an empty block to the tree.
+Later we will use the fact that all blocks contain at least one operation.
+Line \ref{cas} tries to install \nf{new}.
+If it was successful all is good.
+If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
+In this case, \nf{Refresh} helps update $n$\nf{.head} and the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
+
+\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
+The \nf{new} block is created in Line 333 and its fields are filled similarly for both left and right directions.
+The variable  \nf{index\sub{prev}} is the index of the block preceding the first direct subblock in direction \nf{dir} aggregated into \nf{new}.
+Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in \nf{dir} child.
+Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
+The field \nf{sum\sub{deq-dir}} is computed similarly.
+Then, if \nf{new} block is going to be installed in the \nf{root}, the \nf{size} field is also computed.
+
+\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DSearch($e, end$)}}
+We can show an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block.
+If we know the block and rank within the block of an operation we can find the subblock containig the operation and the operation's rank within that subblock in poly-log time.
+To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
+We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
+\nf{DSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
+A call to \nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of the \nf{i}th enqueue in \nf{b}th the block of $n$.
+The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
+Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
 
@@
 
- Let $i$ be the value $R_n$, a successful instance of \nf{Refresh} on node \nf{n} reads from \nf{$n$.head}.
-$R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}.
-Let $p$ be $n$\nf{}.parent.
-Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$.
-From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh} propagating $b$.
-Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}.
-Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.
-
-Since the index of the superblock of ${b}$ is not known until $b$ is propagated, $R_n$ cannot set the \nf{super} field of ${b}$ while creating it.
-One approach is to set the \nf{super} field of ${b}$ by $R_n$ after propagating $b$ to $p$.
-This solution would not be efficient because there might be $p$ subblocks in the block $R_p$ propagated needing to update the \nf{super} field.
-However intuitively, once $b$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-One idea is that if we know the approximate position of the superblock of $b$ then we can search for the real superblock when we wished to know the superblock of $b$ i.e.
-\nf{b.super} does not have to be the exact location of  the superblock of $b$, but we want it to be close to \nf{j}.
-We can set \nf{$b$.super} to \nf{$n$.parent.head} while creating $b$, but the problem is that there might be many \nf{$p$.Refresh}es that could happen after reading $p$\nf{.head} by $R_n$ and before propagating $b$ to $p$.
-If we set \nf{$b$.super} to \nf{$p$.head} after appending $b$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $b$ and before setting \nf{$b$.super}.
-In this case the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$b$.super}.
-
-Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}.
-After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or  maybe an \nf{$n$.parent.Refresh}.
-We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh} that propagates $b$.
+$n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block of the $i$th \nf{Dequeue} of $b$th block of $n$ in the root.
+Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
+Let $n$\nf{.parent} be $n$\nf{.parent}.
+Without loss of generality assume for the rest of this section a ${n}$ is the left child of $n$\nf{.parent}.
+Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${b}$ for \nf{left.head} therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
+Let the index of the block $R_p$ put in $n$\nf{.parent}\nf{.blocks} be $j$.
+
+Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating it.
+One approach for $R_n$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $n$\nf{.parent}.
+This solution would not be efficient because there might be $n$\nf{.parent} subblocks in the block $R_p$ propagated needing updates their \nf{super} field.
+However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
+If we know the approximate position of the superblock of $B$ then we can search for the real superblock when we it is needed.
+Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
+We can set \nf{$B$.super} to \nf{$n$.parent.head} while creating $B$, but the problem is that there might be many \nf{$n$\nf{.parent}.Refresh}es that could happen after $R_n$ reads $n$\nf{.parent}\nf{.head} and before propagating $B$ to $n$\nf{.parent}.
+If $R_n$ sets \nf{$B$.super} to \nf{$n$\nf{.parent}.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
+In this case, the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$B$.super}.
+
+Block $B$ is appended to \nf{$n$.blocks[i]} on Line \ref{cas}.
+After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
+We bound how far \nf{b.super} is from the index of $b$'s superblock by 1.
 
@@
 
-\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the response to $D_{root,b,i}$.}
+\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.}
 \If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
@@
 \State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
-\cmt{$E_e(root)$ is the response.}
+\cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue}in the root.}
 \State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
@@
 \end{algorithm}
+
+\pagebreak
 
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
+\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$,i.e.
+, the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}
+
 \Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
-\cmt{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$.}
 \State \tt{start= end-1} \label{dsearchStart}
@@
 
+\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
 \Function{void}{Propagate()}{}
@@
 
+\Statex $\triangleright$ \textmd{Creates a block containig new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
 \Function{boolean}{Refresh()}{}
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= index\sub{last}} \label{endDefLine}
-\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
+\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine}
 \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
+\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
 \State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + }
@@
 
-%##########################################
+\pagebreak
 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
+\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
 \State \Return\tt{right.}\Call{GetEnqueue}{subBlock, i} 
@@
 
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in $n$.
+Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
 
@@
 Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
-The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
+The value written into \nf{$b$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
 Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}.
@@
 \pagebreak
-\section{Future directions}
-\paragraph{Infinite Sized Arrays}
-Our queue algorithm's usage of memory has two places to improve.
-First, we assumed arrays in the nodes are infinite sized.
-Second we do not handle garbage collection, dequeued operations are still remain in the nodes.
-We can use list to arrays to resolve the infinite sized arrays issue.
-To implement an infinite shared array we can create a small array of pointers called \nf{arr}.
-When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor]$]} points to an array or not.
-If not it creates a shared array with size $2^i$ and tries to \nf{CAS} it to the created array.
-After \nf{CAS} terminates whether it is successful or not \nf{arr[$\lfloor\log $ head$\rfloor]$]} points to an array.
-When a process wishes to access to $i$th element it looks upto \nf{arr[$\lfloor\log i\rfloor]$][$i-2^{\lfloor\log i\rfloor]}$]} Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ times \nf{CAS} has happened.
-And if the size of the all shared memory is $k$, \nf{arr} can be initialized to a $\log k$ sized array.
-
+\section{Future Directions}
+We designed a tree to achieve agreement among processes in poly-log steps with respect to the total number of operations invoked and used it to implement a poly-log queue with respect to the size of the queue.
+We can use our solution for agreement to construct more poly-log wait-free linearizable objects.
+We call our solution \it{Block Tree}.
+In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.
+
+\paragraph{Reducing Space Usage}
+Our queue algorithm's usage of memory has places to improve.
+We assumed arrays in the nodes are unbounded.
+If there is an $N$, such that the $N$ is greater than the  maximum possible operations appended to the block tree, then we can use Array
+Segment~\cite{7073592} model to take $O(n+ \log N)$ space in each node, where $n$ is the total number of operations.
+To implement this we can create a static array of pointers called \nf{arr} of references to array segments.
+When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
+If not, it creates a shared array with size $2^i$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
+Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
+When a process wishes to access to $i$th element it looks up to \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]} which takes $O(1)$ steps.
+Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened.
+Furthermore at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]} which in average the time taken to allocate arrays in an execution is $O(1)$ per operation.
 \begin{figure}[hbt]  
   \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Infinite sized Arrays}
+  \caption{Array Segments}
   \label{fig::doublingArray}
@@
 \paragraph{Garbage Collection}
-To garbage collect we can keep track of the index of the most recent \nf{block} in the \nf{root} that has been term and then split the blocks from that \nf{block}.
-We can use Persistent Red Black Trees because they support split in an efficient method.
-
-\paragraph{Poly-logarithmic Wait-free Vectors}
-If we do not distinct between \nf{Enqueue}s and \nf{Dequeue}s and replace \nf{DSearch} with Binary Search on all the \nf{root} in our algorithm, we achieve a list with poly-logarithmic \nf{Append}, \nf{Get} and \nf{Index} methods with respect to the number of \nf{Append}s done.
-
-\paragraph{Stacks or maybe other objects that store sequences}
-The reason behind why our \nf{tree} worked well to implement a queue was two things.
-First, we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} fast.
-Two, the response of the \nf{Dequeue} is not very far from it (size of the queue).
-We think we can create wait-free poly-logarithmic version of other objects which their operations satisfy these two conditions.
+We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
+We can keep track of the \nf{block}s in the \nf{root} that all of their operations are terminated, i.e, all of its enqueues have been dequeued and the response of all of its dequeues is computed.
+We call these blocks \it{Finished Blocks}.
+If we help the old operations to compute their response, then we can say if block $B$ is finshed then all blcoks before $B$ are also finished.
+Knowing the most recent finished block we can release the memory taken by these blocks.
+To throw the garbage in the blocks away we cannot use arrays.
+We need a datastructure that supports \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$.
+\nf{split(i)} is to remove all the indices less than \nf{i}.
+We can use Persistent Red Black Trees~\cite{DBLP:conf/podc/BashariW21} for this usage.
+
+\begin{figure}[hbt]  
+  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
+  \caption{Finished blocks are shown with red color and unfinished blocks are shown with green color. All the subblocks of a finished block are also finished.}
+  \label{fig::doublingArray}
+\end{figure}
+
+\paragraph{Poly-logarithmic Wait-free Sequences}
+Assume a data structure storing a sequence that has supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
+\nf{append(e)} adds \nf{e} to the sequence, \nf{get(i)} gets the \nf{i}th element in the sequence and \nf{index(e)} computes the position of appended element \nf{e} in the sequence.
+We can modify our queue to design such data structure.
+If we do not distinct between \nf{Enqueue}s and \nf{Dequeue}s and replace \nf{DSearch} with Binary Search on all the \nf{root} in our algorithm, we achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
+
+\paragraph{Stacks}
+There are two reasons block tree worked well to implement a queue.
+Firstly, to respond a \nf{Dequeue} we do not need to look at all the history of operations, if a \nf{Dequeue} does not return null we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} in $O(\log n)$ if we keep the number of enqueues and the size.
+Secondly, the operations needed to look to respond a the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
+It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.
 

### 2022-10-16 thesis/thesis.tex
apply ananlysis feedback

@@
 
-In an execution on a queue where no dequeue operation returns null, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns null if and only if the size of the queue after the previous operation is 0.
-We refer to such a dequeue as a null dequeue.
+In an execution on a queue where no dequeue operation returns\nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
+In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0.
+We refer to such a dequeue as a\nf{null} dequeue.
 Otherwise, if the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
@@
 
-\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree and returns its response.}
+\Function{Object}{Dequeue()}{} \cmt{Creates a block with\nf{null} value element, appends it to the tree and returns its response.}
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
@@
 \EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(sum\sub{enq}, e, start, end)} \label{dsearchBinarySearch}
+\State \tt{b= root.BinarySearch( e, start, end)} \label{dsearchBinarySearch}
 \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
@@
 \pagebreak
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} greater than or equal to \tt{i}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{i} of the given prefix sum \tt{field}.
-Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.
-\Function{int}{BinarySearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}
-\State \Return \tt{min\{j: blocks[j].f$\geq$i\}}
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{i}}
+\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{i} of \tt{sum\sub{enq}} field.
+Returns the index of the leftmost block in \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{i}}.
+\Function{int}{BinarySearch}{ \sl{int} i, \sl{int} start, \sl{int} end}
+\State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$i\}}
 \EndFunction{BinarySearch}
@@
 \ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
-\State \tt{subBlock= left.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
+\State \tt{subBlock= left.\Call{BinarySearch}{ i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
 \State \Return\tt{left.}\Call{GetEnqueue}{subBlock, i} 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
+\State\tt{subBlock= right.\Call{BinarySearch}{ i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
 \State \Return\tt{right.}\Call{GetEnqueue}{subBlock, i} 
@@
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns null it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
+If a \nf{Dequeue} operation returns\nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
 \end{definition}
@@
 ${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null dequeue then ${Resp(d)=\nf{null}}$.
+Otherwise if ${d}$ is a\nf{null} dequeue then ${Resp(d)=\nf{null}}$.
 \end{definition}
@@
 \section{Analysis}
-\begin{lemma}[Amortized time analysis]
-\nf{Enqueue} and \nf{Dequeue}, each take $O(\log^2 p + \log q)$ steps in amortized analysis.
-Where $p$ is the number of processes and $q$ is the size of the queue at the time of invocation of operation.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue(x)} consists of creating a \nf{block(x)} and appending it to the tree.
-The first part takes constant time.
-To propagate \nf{x} to the root the algorithm tries two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes constant number of steps since creating a block is done in constant time and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue(x)} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block with null value element, appends it to the tree, computes its order among enqueue operations, and returns the response.
-The first two part is similar to an \nf{Enqueue} operation.
-To compute the order of a \nf{dqueue} in $D(n)$ there are some constant steps and \nf{IndexDequeue()} is called.
-\nf{IndexDequeue} does a search with range $p$ in each level which takes $O(log^2 p)$ in the tree.
-In the \nf{FindResponse()} routine \nf{DSearch()} in the root takes $\Theta(\log$(\nf{root.blocks[b].\size+root.blocks[end].\size}) by Lemma \ref{dsearch}, which is $O(\log$ size of the queue when \nf{Enqueue} is invoked$)+\log$ size of the queue when \nf{Dequeue} is invoked$)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-
-If we split \nf{DSearch} time cost between the corresponding \nf{Enqueue}, \nf{Dequeue}, in amortized we have \nf{Enqueue} takes $O(\log p +q)$ and \nf{Dequeue} takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
+
+Is this a Lemma or a Result or an Observation of Algorithm? Which is not used in main wait-free theorem.
 \begin{lemma}
-An \nf{Enqueue} or \nf{Dequeue} operation, does at most $4\log p$ \nf{CAS} operations.
-\end{lemma}
-\begin{proof}
-  In each height of the tree at most 2 times \nf{Refresh} is invoked and  every \nf{Refresh} invokes at most 3 \nf{CAS}es, one in Line \ref{cas} and two from \nf{Advance} in Line \ref{incrementHead}.
+An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
+\end{lemma}
+\begin{proof}
+  In each level of the tree  \nf{Refresh} is invoked at most two times and  every \nf{Refresh} invokes at most 7 \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
 \end{proof}
@@
 \begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[end]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
+If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
 \end{lemma}
@@
 First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-Suppose there were more than \nf{root.blocks[b].size} \nf{Dequeue}s in \nf{root.blocks[$b+1 \cdots end-1$]}.
-Then the element in the queue which is the response to the \nf{Dequeue} would become dequeued at some point before \nf{root.blocks[$end$]}'s first \nf{Dequeue}.
+There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
 Furthermore in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
 The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s in \nf{root.blocks[$b+1\cdots end-1$]}.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end$]} and since from line \ref{addOP} we know that \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+After an execution on a queue the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
+We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
 So $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
 
-So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
+So, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
 See Figure \ref{fig::doubling}.
@@
 Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-So the claim is proved.
 \end{proof}
@@
 \end{figure}
+
+\begin{lemma}[Worst Case Time analysis] \label{enqDeqTime}
+The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$,  where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_i$ is the size of the queue at time the response of the \nf{Dequeue} is linearized.
+\end{lemma}
+\begin{proof}
+\nf{Enqueue} consists of creating a block and appending it to the tree.
+The first part takes constant time.
+To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
+We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
+Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
+
+A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
+The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
+To compute the rank of a \nf{Dequeue} in $D(n)$ the dequeue calls \nf{IndexDequeue()}.
+\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(log p)$ steps.
+If the response to the dequeue is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
+Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log$ size of the queue when \nf{Enqueue} is linearized$)+\log$ size of the queue when \nf{Dequeue} is linearized$)$.
+Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
+\end{proof}
+
+\begin{lemma}[Amortized Time Analysis]
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
+\end{lemma}
+\begin{proof}
+If we split \nf{DSearch} time cost between the corresponding \nf{Enqueue}, \nf{Dequeue}, in each operation takes $O(\log^2 p +q)$ steps.
+\end{proof}
+
+\begin{theorem}
+The queue implementation is wait-free.
+\end{theorem}
+\begin{proof}
+To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
+This is directly concluded from Lemma \ref{enqDeqTime}.
+\end{proof}
 
@@
 There are two reasons block tree worked well to implement a queue.
-Firstly, to respond a \nf{Dequeue} we do not need to look at all the history of operations, if a \nf{Dequeue} does not return null we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} in $O(\log n)$ if we keep the number of enqueues and the size.
+Firstly, to respond a \nf{Dequeue} we do not need to look at all the history of operations, if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} in $O(\log n)$ if we keep the number of enqueues and the size.
 Secondly, the operations needed to look to respond a the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.

### 2022-10-17 thesis/thesis.tex
finalize description notes

@@
 In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of invocation.
+We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
 We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
@@
 We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent dequeuers.
+Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
 
@@
 
-To propagate operations to node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
+To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
 We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent.
+A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$.
 We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations from $n$'s children after the first failed \nf{Refresh} and succeeded to do an append on $n$.
-This \nf{Refresh} helps to append the new operations in the $n$' children before the first failed \nf{Refresh}, in case they were not already appended.
+We sketch the argument here.
 
@@
 
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh}.
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$' second failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
 We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node.
-So with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
+The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
+This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
+Process $K$'s \nf{Refresh} helps to append the new operations in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
+After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
+So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
 This cooperative solution allows us to overcome the CAS Retry Problem.
@@
 Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-We can only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in root (see Figure \ref{fig::set}).
-Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
+It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
+Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
 To do that, we introduce \it{block}s.
-A block stores information about the operations in a \nf{Refresh} step.
-It contains the number of operations from the left and the right child propagated to the node by a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example).
+A block stores information about the operations propagated by a \nf{Refresh}.
+It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
+See Figure \ref{fig:block} for an example.
 A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in children into a new block and puts it in the parents' blocks.
-We call the aggregated blocks, subblocks of the new block and the new block, superblock of them.
+A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks.
+We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
 In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
@@
 So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$.
+With this agreement on linearization ordering we can design a universal construction; for given object $O$ we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
 However, this approach is not enough for an efficient queue.
-A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) and use them to construct a fast queue below.
+We show that we can build an efficien queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
+We explain how to implement (1) and (2) in poly-logarithmic steps.
 
 After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in a child of the root that contains that $i$th operation.
+To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation.
+When we reach a block in a leaf, the operation is explicitly stored there.
 To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
@@
 We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
+
 To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in.
@@
 In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0.
-We refer to such a dequeue as a\nf{null} dequeue.
-Otherwise, if the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
+We refer to such a dequeue as a \it{null dequeue}.
+If the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
 Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue which is the response to a given dequeue in constant time.
+So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
 
@@
 We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the follwing operations, where operations in a cell are concurrent.
+Consider the following operations, where operations in a cell are concurrent.
 \begin{table}[hbtp]
@@
 
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
+Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
 Otherwise the response would be \nf{null}.
@@
 \subsection{Details of the Implementation}
-Pseudocode for the queue implementation is given in Section \ref{algQ}.
-\paragraph{Node} 
+Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
+\paragraph{\tt{Node}} 
  In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
 
-\paragraph{Block}
- The information stores in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
+\paragraph{\tt{Block}}
+ The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
 If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation.
@@
 In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-\texttt{sum\textsubscript{enq-left}} is the total number of \nf{Enqueue} operations in theblocks before the last subblock of $B$ in the left child child, other fields semantics is also similar.
+The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
+The other fields' semantics are similar.
 The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively.
@@
 \paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields, appends it to the tree.
-Then it computes the position of the appended \nf{Dequeue} operation in the root and after that finds the response of the \nf{Dequeue} calling \nf{FindResponse}.
+\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
+Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
 
@@
 To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dqeueue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dqeueue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dqeueue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block we find the block) and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
+If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dequeue}.
+Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
 
@@
 The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
+There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
 \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
 When \nf{Propagate} terminates it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
 
@@
 \nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively untill it reaches the root.
+Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
 
@@
 The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} children, \nf{new} block is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there is no new operations to propagate and it is unnecessary to add an empty block to the tree.
+After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
+Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate and it is unnecessary to add an empty block to the tree.
 Later we will use the fact that all blocks contain at least one operation.
@@
 If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps update $n$\nf{.head} and the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
+In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
 
 \paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The \nf{new} block is created in Line 333 and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first direct subblock in direction \nf{dir} aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in \nf{dir} child.
+The block \nf{new} is created in Line 333 and its fields are filled similarly for both left and right directions.
+The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
+Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
 Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
 The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} block is going to be installed in the \nf{root}, the \nf{size} field is also computed.
+Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
 
 \paragraph{\tt{GetEnqueue($b,i$)} and \tt{DSearch($e, end$)}}
-We can show an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block.
-If we know the block and rank within the block of an operation we can find the subblock containig the operation and the operation's rank within that subblock in poly-log time.
+We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
+If we know the block and rank within the block of an operation we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
 To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
@@
 \nf{DSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of the \nf{i}th enqueue in \nf{b}th the block of $n$.
+A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in $b$th the block of $n$.
 The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
@@
 
-$n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block of the $i$th \nf{Dequeue} of $b$th block of $n$ in the root.
+A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
 Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $n$\nf{.parent} be $n$\nf{.parent}.
-Without loss of generality assume for the rest of this section a ${n}$ is the left child of $n$\nf{.parent}.
-Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${b}$ for \nf{left.head} therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
-Let the index of the block $R_p$ put in $n$\nf{.parent}\nf{.blocks} be $j$.
+Let $par$ be $n$\nf{.parent}.
+Without loss of generality assume for the rest of this section a ${n}$ is the left child of $par$.
+Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
+Let $j$ be the index of the block that $R_{par}$ put in $par$\nf{.blocks}.
 
 Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating it.
-One approach for $R_n$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $n$\nf{.parent}.
-This solution would not be efficient because there might be $n$\nf{.parent} subblocks in the block $R_p$ propagated needing updates their \nf{super} field.
+One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
+This solution would not be efficient because there might be up to $p$ subblocks in the block $R_{par}$ propagated needing updates their \nf{super} field.
 However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when we it is needed.
+If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
 Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$n$.parent.head} while creating $B$, but the problem is that there might be many \nf{$n$\nf{.parent}.Refresh}es that could happen after $R_n$ reads $n$\nf{.parent}\nf{.head} and before propagating $B$ to $n$\nf{.parent}.
-If $R_n$ sets \nf{$B$.super} to \nf{$n$\nf{.parent}.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[i]} on Line \ref{cas}.
+We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
+If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
+In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
+
+Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
 After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We bound how far \nf{b.super} is from the index of $b$'s superblock by 1.
+We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most 1.
 

### 2022-10-17 thesis/thesis.tex
fix Future Directions

@@
 \begin{document}
-\title{Wait-free Queues with Polylogarithmic Step Complexity}
+\title{A Wait-free Queue with Polylogarithmic Step Complexity}
 \author{Hossein Naderibeni \\supervised by Eric Ruppert}
@@
 \section{Future Directions}
-We designed a tree to achieve agreement among processes in poly-log steps with respect to the total number of operations invoked and used it to implement a poly-log queue with respect to the size of the queue.
-We can use our solution for agreement to construct more poly-log wait-free linearizable objects.
-We call our solution \it{Block Tree}.
+We designed a tree to achieve agreement on an linearization of operations invoked by $p$ processes in an asynchronous model, which we will call \it{Block Tree}.
+We also implemented two queries to know information about the ordering agreed in the block tree.
+Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
+Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
 In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.
@@
 \paragraph{Reducing Space Usage}
-Our queue algorithm's usage of memory has places to improve.
-We assumed arrays in the nodes are unbounded.
-If there is an $N$, such that the $N$ is greater than the  maximum possible operations appended to the block tree, then we can use Array
-Segment~\cite{7073592} model to take $O(n+ \log N)$ space in each node, where $n$ is the total number of operations.
-To implement this we can create a static array of pointers called \nf{arr} of references to array segments.
+The \nf{blocks} arrays defined in our algorithm are unbounded.
+We could instead use the vector of Feldman, Valera-Leon and  Damian Segment~\cite{7073592} model to use $O(n)$ space in each node, where $n$ is the total number of operations.
+Their vector creates an array called \nf{arr} of pointers to array segments.
 When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array with size $2^i$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
+If not, it creates a shared array with size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
 Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access to $i$th element it looks up to \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]} which takes $O(1)$ steps.
-Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened.
-Furthermore at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]} which in average the time taken to allocate arrays in an execution is $O(1)$ per operation.
+When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
+Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
+Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}.
+Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
+The time taken to allocate arrays in an execution containing $n$ operations is $O(p\log n)$ which if $n>>p$ we can ignore the $p$ factor.
+The vector implementation has also a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
 \begin{figure}[hbt]  
@@
 We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} that all of their operations are terminated, i.e, all of its enqueues have been dequeued and the response of all of its dequeues is computed.
-We call these blocks \it{Finished Blocks}.
-If we help the old operations to compute their response, then we can say if block $B$ is finshed then all blcoks before $B$ are also finished.
-Knowing the most recent finished block we can release the memory taken by these blocks.
-To throw the garbage in the blocks away we cannot use arrays.
-We need a datastructure that supports \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$.
-\nf{split(i)} is to remove all the indices less than \nf{i}.
-We can use Persistent Red Black Trees~\cite{DBLP:conf/podc/BashariW21} for this usage.
+We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e, all of its enqueues have been dequeued and the responses of all of its dequeues have been computed.
+We call these blocks \it{finished blocks}.
+If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished.
+Knowing the most recent finished block in a node we can reclaim the memory taken by finished blocks.
+To throw the garbage in the blocks away we cannot use arrays (or vectors).
+We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$.
+Where \nf{split(i)} removes all the indices less than \nf{i}.
+We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/podc/BashariW21}.
 
@@
 \paragraph{Poly-logarithmic Wait-free Sequences}
-Assume a data structure storing a sequence that has supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-\nf{append(e)} adds \nf{e} to the sequence, \nf{get(i)} gets the \nf{i}th element in the sequence and \nf{index(e)} computes the position of appended element \nf{e} in the sequence.
+Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
+An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
 We can modify our queue to design such data structure.
-If we do not distinct between \nf{Enqueue}s and \nf{Dequeue}s and replace \nf{DSearch} with Binary Search on all the \nf{root} in our algorithm, we achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
+\nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} with replacing the \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
+ We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
 
@@
 There are two reasons block tree worked well to implement a queue.
-Firstly, to respond a \nf{Dequeue} we do not need to look at all the history of operations, if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is response of a \nf{Dequeue} in $O(\log n)$ if we keep the number of enqueues and the size.
-Secondly, the operations needed to look to respond a the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
+Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
+Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
 It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.
+
+\pagebreak
 

### 2022-10-17 thesis/thesis.tex
apply Oct 17 notes

@@
 \begin{abstract}
-In this work, we are going to introduce a novel lock-free queue implementation.
+In this work, we introduce a novel linearizable wait-free queue implementation.
 Linearizability and lock-freedom are standard requirements for designing shared data structures.
-All existing linearizable, lock-free queues in the literature have a common problem in their worst case called CAS Retry Problem.
-Our contribution is solving this problem while outperforming the previous algorithms.
+To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem.
+We show our algorithm resolves this problem and has worst case running time better than prior lock-free queues.
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
 \end{abstract}
@@
 Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
-
-The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+For the reasons told, we are interested in the design of the data structures that do not use locks.
+
+The question that may arise is ``What properties matter for a lock-free data structure?''.
+Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 
-The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
+The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
 Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty.
@@
   \caption{\label{fig::badexample}An example of an execution that is not linearizable.
-Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
+Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or \nf{null}.}
 \end{figure}
@@
 
+\it{is this place good for queue definition?}
 A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
@@
 If the queue is empty it returns \texttt{null}.
-In section 2 we talk about previous queues and their common problems.
-We also talk about polylogarithmic construction of shared objects.
-
-Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions.
-He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses.
-Here, we first introduce a universal construction using $\textsc{O}(\log p)$ CAS operations~\cite{DBLP:conf/fsttcs/JayantiP05}.
-In section 3 we introduce a polylogarithmic step wait-free universal construction.
-Our main ideas in of the universal construction also appear in our Queue Algorithm (\ref{algQ}).
-The main short come of our universal construction is using big CAS objects.
-We use the universal construction as a stepping stone towards our queue algorithm, so we will not explain it in too much detail.
-
-In section 4 we introduce a concurrent wait-free datastructure, to agree on the order of the operations invoked on some processes.
-
-In section 5 we introduce our main work, the queue; prove its linearizability and wait-freeness.
+
+\paragraph{Model} 
+A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e the scheduling of each process is independent from the scheduling of other processes.
+Our model is an asynchronous shared-memory distributed system of $p$ processes.
+Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement any lock-free data structures with multi-reader multi-writer registers alone.
+In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.
+
+\paragraph{Thesis outline} The rest of the thesis is organized as follows.
+Chapter 2 gives an outline of related word done in the area and the motivation of our implementation.
+Previous lock-free queues and their common problem is presented in Section 2.1.
+In Section 2.2 we mention some restricted lock-free queues.
+Section 2.3 talks about poly-logarithmic constructions of shared objects.
+Section 2.4 is devoted to the lower bound (COMPLETE).
+
+In Chapter 3 we introduce a poly-logarithmic step wait-free queue. Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.
+
+We prove the correctness of our queue in Chapter 4 by showing it is wait-free and linearizable.
+
+Chapter 5 presents performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
+
+Finally, we talk about some concluding remarks in Chapter 6.
+It contains how we can improve our algorithm memory usage and how to use its idea in designing other wait-free data structures in section 6.
 
@@
 \section{Related Work}
+In this Section, we look at previous lock-free queues.
 \subsection{List-based Queues}
-In the following paragraphs, we look at previous lock-free queues.
 Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
@@
 Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue.
+A shared variable Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue.
 To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
 If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps to be $\Omega(p)$ per enqueue.
+This makes the amortized number of steps $\Omega(p)$ per enqueue.
 Similarly, dequeue can take $\Omega(p)$ steps.
@@
 
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination technique.
+Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
 The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
@@
 Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
-However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
+However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
 
@@
 \end{figure}
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues.
-They use a doubly-linked list and do fewer \texttt{CAS} operations than MS-queue.
-But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one.
+\bf{TODO: make circles the same size}
+
+Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queues.
+MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node.
+They change the direction of storing the nodes and use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
+As previous algorithms, the worst case happens when contenion is high: $p$ concurrent enqueues have happened, which they have to be appended to the linked list one by one.
 The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
@@
 Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-Global head and tail pointers point to arrays containing the first and last elements in the queue.
+There is a global head which points to arrays containing the first elements in the queue.
+Also there is a head that points to the first element in the array that global head points.
+Global tail and tail pointers are the same.
 Global pointers are up to date, but head and tail pointers may be behind in time.
@@
 
-%todo: comparison
-%Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
+\bf{TODO} 
+Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
 Their idea of batching allows a sequence of operations to be submitted as a batch for later execution.
@@
 
-%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
-In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously.
+\bf{TODO}
+Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
+
+In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} .
-We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.
+It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09}.
+Our motivation is to overcome this problem and present a wait-free sublinear queue.
+
+\subsection{Restricted Queues}
+Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
+
+\bf{TODO}
+Which David?
 
@@
 He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.
-
+His universal construction can be used to create any data structure, but its implementation is not practical because of using \nf{CAS} instructions on unreasonably large-sized words of memory(\bf{TODO: how large?}).
+
+\bf{TODO: why introducing this result?}
 Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
@@
   \caption{Finished blocks are shown with red color and unfinished blocks are shown with green color. All the subblocks of a finished block are also finished.}
-  \label{fig::doublingArray}
+  \label{fig::finishedBlock}
 \end{figure}

### 2022-10-18 thesis/thesis.tex
add more related work

@@
 Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 is devoted to the lower bound (COMPLETE).
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue. Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.
+
+In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.
 
@@
 
-Chapter 5 presents performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
+Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
 
@@
 
-\bf{TODO} 
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed BQ: A Lock-Free Queue with Batching.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution.
-It supports a new notion introduced by the authors called Extended Medium Futures Linearizability.
-
-\bf{TODO}
-Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
+Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue with batching.
+Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
+They use some properties of the size of the queue before and after a batch which is similar to apart of our work.
+Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a new notion introduced by the authors called Extended Medium Futures Linearizability.
+
+Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses fast-path slow-path methodology.
+Their work is based on a circular queue using bounded memory.
+They design a slow phase which is wait-free bounded.
+If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase.
+Since there can be $p$ processes helping the worst-case complexity cannot be better than linear.
 
@@
 \subsection{Restricted Queues}
+
+David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
+Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinte memory.
+The author states that to reduce memory usage to be bounded the time per operation increases to linear.
+
 Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-
-\bf{TODO}
-Which David?
+Their algorithm time complexity is $O(\log n)$ where $n$ is the number of enqueuers.
+We will improve their algorithm to construct a multi-enqueuer multi-dequeuer queue.
+
+\bf{TODO: Where should be?}
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Their idea to achieve logarithmic complexity is to use a tree storing the list of elements added by processes and when a process wants to do Fetch\&Inc it appends its element to the tree and return the number of elements in the tree.
+Their work can be descibed a queue without dequeues.
 
@@
 Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-His universal construction can be used to create any data structure, but its implementation is not practical because of using \nf{CAS} instructions on unreasonably large-sized words of memory(\bf{TODO: how large?}).
-
-\bf{TODO: why introducing this result?}
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.
-
-%FIFO queues have a wide range of use in OS and applications. The current state-of-the-art queues are implemented using linked lists; that's why they have factor p in their time complexity.
-
-%\href{https://docs.google.com/spreadsheets/d/1cL1tgXXdljkh462sMwkTVMHH_k0MBTmslIVM3xA5VS4/edit#gid=0}{Table of previous works}
-
-%There is a connection between queues and universal constructions.
-We can implement a universal construction using a queue.
-We can store operations in the shared queue and compute operations' responses using the queue's content.
-Some impractical universal constructions are using big words as "...".
-
-\subsection{Attiya Fourier Lower Bound}
-\pagebreak
-
-\pagebreak
+Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
+Their universal construction cannot be used to create any data structure, but a set of objects called closed objects.
+We mention a non-practical poly-logarithmic universal construction in page 12.
+
+\pagebreak
+
 \section{Queue Implementation} \label{DescriptQ}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{100}
+\setcounter{ALG@line}{1}
 \begin{multicols}{2}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{200}
+\setcounter{ALG@line}{0}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{800}
+\setcounter{ALG@line}{25}
 \Statex
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{600}
+\setcounter{ALG@line}{34}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{300}
+\setcounter{ALG@line}{39}
 \begin{multicols}{2}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{400}
+\setcounter{ALG@line}{85}
 
@@
 
+\paragraph{Attiya--Fouren Lower Bound}
+Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes).
+Their result says any implementation of queues, using reads,
+writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$
+amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.
+
 \pagebreak

### 2022-10-19 thesis/thesis.tex
finalize introductio

@@
 \newtheorem{invariant}[theorem]{Invariant}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 The industry solution to provide more computational power is to increase the number of cores of the CPU.
-This is why distributed algorithms have become important.
-It is not hard to see why multiple processes cannot update sequential data structures designed for one process.
-For example, consider two processes trying to insert some values into a sequential linked list simultaneously.
+It is not hard to see why multiple processes cannot share a sequential data structure to work with.
+For example, consider two processes trying to append to a sequential linked list simultaneously.
 Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same.
 In this run, $p$'s update is overwritten.
-One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released.
+One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
-For the reasons told, we are interested in the design of the data structures that do not use locks.
-
-The question that may arise is ``What properties matter for a lock-free data structure?''.
-Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time data structure is locked.
+For these reasons, we are interested in the design of an efficient shared queue without using locks.
+
+A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
+\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.
+\texttt{Dequeue} removes and returns the first element in the queue.
+If the queue is empty it returns \texttt{null}.
+In a concurrent version of queue data structure, operations do not happen one at a time.
+The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 
+A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
+Our model is an asynchronous shared-memory distributed system of $p$ processes.
+Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement concurrent version of all data structures with multi-reader multi-writer registers alone, introducing a hieachy on how powerful are objects to solve consensus problem among processes.
+Objects like LL/SC, CAS can be used to reach consensus among any number of processors.
+CAS object implements an atomic Compare \& Swap(\nf{new, old}) operation, which if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns {true}, otherwise returns \nf{false}.
+In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.
+
+\bf{Concurrent execution: I didn't understed interleaving steps of processes}
+
 The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty.
+Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
 The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
@@
   \center\includegraphics[scale=0.5]{pics/bad}
-  \caption{\label{fig::badexample}An example of an execution that is not linearizable.
-Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or \nf{null}.}
+  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
+\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can be occurred before termination of \texttt{Enqueue(A)} and return \nf{null} or can be occurred after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
 \end{figure}
@@
 
-\it{is this place good for queue definition?}
-A queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.
-\texttt{Dequeue()} removes and returns the first element among in the sequence.
-If the queue is empty it returns \texttt{null}.
-
-\paragraph{Model} 
-A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement any lock-free data structures with multi-reader multi-writer registers alone.
-In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.
+We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free bounded poly-log time queue.
+We design a tournament tree in which processes try to propagate their operations up to the root to be linearized and they can do queries to get information about the linearization stored in the root.
+To solve the CAS Retry problem when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations.
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
+\nf{CAS} instructions cost more than other instructions for processor.
+Each operation in our queue does at most $14\log p$ \nf{CAS} operations.
+We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations.
+The garbage collection remains open which we have pointed to a way we think it can be handled.
 
@@
 Section 2.3 talks about poly-logarithmic constructions of shared objects.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue. In Section 3.1 we give a high-level description of our implementation. Section 3.2 contains the algorithm itself.
-
-We prove the correctness of our queue in Chapter 4 by showing it is wait-free and linearizable.
-
-Chapter 5 is devoted to performance analysis for our implementation. We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
-
-Finally, we talk about some concluding remarks in Chapter 6.
-It contains how we can improve our algorithm memory usage and how to use its idea in designing other wait-free data structures in section 6.
+Section 2.4 is summoned with a lower bound on the amortized time complexity of shared queues.
+
+In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
+In Section 3.1 we give a high-level description of our implementation.
+We also talk about the motivation and requirements in our design to achieve poly-log time.
+Section 3.2 contains the algorithm itself.
+
+We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
+
+Chapter 5 is devoted to performance analysis for our implementation.
+We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
+In the end we prove our queue is wait-free.
+
+Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
 
@@
 \section{Related Work}
-In this Section, we look at previous lock-free queues.
+In this section, we look at previous lock-free queues.
 \subsection{List-based Queues}
@@
 
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queues.
+Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
 MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node.
-They change the direction of storing the nodes and use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As previous algorithms, the worst case happens when contenion is high: $p$ concurrent enqueues have happened, which they have to be appended to the linked list one by one.
+They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
+Asin  previous algorithms, the worst case happens in the case where the contention is high when $p$ concurrent enqueues happene, their nodes have to be appended to the linked list one by one.
 The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
@@
 Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-Their queue is linearizable but not lock-free.
-Their main idea is that with knowledge of all the history of operations, it might be possible to answer queries faster than doing them one by one.
-In our work we also maintain the whole history.
-They present experiments that show their algorithm performs well in some situations.
+The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
+Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
 
 Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-There is a global head which points to arrays containing the first elements in the queue.
-Also there is a head that points to the first element in the array that global head points.
-Global tail and tail pointers are the same.
-Global pointers are up to date, but head and tail pointers may be behind in time.
-An enqueue or a dequeue searches in the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}).
-Their data structure is lock-free.
-Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.
+The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
+When the array is occupied an empty array is linked to the array and tail pointers are updated.
+Global head points to the array containing the first element in the queue, and head that points to the first element in that array.
+Global tail and tail pointers are similar (see~Figure~\ref{fig::sundell}).
+The pointers might be behind their real places, a process updates the position of the pointers when using them.
+When a process wants to write an element to current head it uses \nf{CAS}, so if $p$ processes try to enqueue simultaneously, the step (also \nf{CAS}) complexity remains $\Omega(p)$.
+Their design is lock-free but not wait-free.
 
@@
 
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue with batching.
+Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supprting futures.
+A future is an operation that can be applied on the queue when it is requested.
 Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the size of the queue before and after a batch which is similar to apart of our work.
-Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a new notion introduced by the authors called Extended Medium Futures Linearizability.
-
-Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses fast-path slow-path methodology.
+They use some properties of the size of the queue before and after a batch which is similar to a part of our work.
+Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a safety condition weaker than linearizabilty introduced by the authors, called Extended Medium Futures Linearizability.
+
+Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology.
 Their work is based on a circular queue using bounded memory.
@@
 If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase.
-Since there can be $p$ processes helping the worst-case complexity cannot be better than linear.
-
-In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously.
+Since there can be $p$ processes helping, the worst-case complexity cannot be better than $\Omega(p)$.
+
+In the worst-case, step complexity of all the list-based queues discussed above, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09}.
+It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
 Our motivation is to overcome this problem and present a wait-free sublinear queue.
@@
 Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-Their algorithm time complexity is $O(\log n)$ where $n$ is the number of enqueuers.
-We will improve their algorithm to construct a multi-enqueuer multi-dequeuer queue.
-
-\bf{TODO: Where should be?}
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the list of elements added by processes and when a process wants to do Fetch\&Inc it appends its element to the tree and return the number of elements in the tree.
-Their work can be descibed a queue without dequeues.
-
-\subsection{Universal Constructions}
+We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a poly-logarithmic multi-enqueuer multi-dequeuer queue.
+Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
+
+\subsection{Universal Constructions and other Poly-log Time Data Structures}
 Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
@@
 Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
+He also mentions that universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log n)$ worst case step complexity, using atomic access to $\Omega(n \log n)$ bit words.
 Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
 Their universal construction cannot be used to create any data structure, but a set of objects called closed objects.
-We mention a non-practical poly-logarithmic universal construction in page 12.
+We mention a non-practical universal construction with poly-log \nf{CAS} invocations in page 12.
+
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
+Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes and when a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree.
+If we have an array and two Fetch\&Inc objects head and tail we can implement a queue: to do an enqueue we can Fetch\&Inc head and write in the value fetched of the array and to do a dequeue we can Fetch\&Inc the tail and fetched the array's cell in the value returned.
+
+\subsection{Attiya--Fouren Lower Bound}
+Because of the \nf{CAS} retry problem in previous list based queues one might guess $p$ factor is inherent in time complexity of concurrent queues.
+Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes).
+Their result says any implementation of queues, using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
+The surprising point is that their result does not apply on our design and we manage to reach poly-log time complexity.
+It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.
 
@@
 We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
 
@@
 \section{Analysis}
-
-Is this a Lemma or a Result or an Observation of Algorithm? Which is not used in main wait-free theorem.
-\begin{lemma}
+In this section we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
+\begin{proposition}
 An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{lemma}
+\end{proposition}
 \begin{proof}
@@
 \section{Future Directions}
-We designed a tree to achieve agreement on an linearization of operations invoked by $p$ processes in an asynchronous model, which we will call \it{Block Tree}.
-We also implemented two queries to know information about the ordering agreed in the block tree.
+We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
+We  implemented two queries to know information about the ordering agreed in the block tree.
 Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
@@
 The \nf{blocks} arrays defined in our algorithm are unbounded.
-We could instead use the vector of Feldman, Valera-Leon and  Damian Segment~\cite{7073592} model to use $O(n)$ space in each node, where $n$ is the total number of operations.
-Their vector creates an array called \nf{arr} of pointers to array segments.
-When a process wishes to write into \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array with size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
+To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
+We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
+When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
+If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
+Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
 When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-Note that CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
+The CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
 Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}.
 Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(p\log n)$ which if $n>>p$ we can ignore the $p$ factor.
-The vector implementation has also a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
+The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
+The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
 \begin{figure}[hbt]  
@@
 If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node we can reclaim the memory taken by finished blocks.
+Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
 To throw the garbage in the blocks away we cannot use arrays (or vectors).
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$.
-Where \nf{split(i)} removes all the indices less than \nf{i}.
-We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/podc/BashariW21}.
+We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$, where \nf{split(i)} removes all the indices less than \nf{i}.
+If each process tries to do the garbage collection every $p^2$ operation on the queue then the amortized complexity remains the same.
+We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
+Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red black trees in a similar way.
 
@@
 We can modify our queue to design such data structure.
-\nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} with replacing the \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
+An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
  We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
@@
 \paragraph{Stacks}
-There are two reasons block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return\nf{null} we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
+There are two reasons the block tree worked well to implement a queue.
+Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
 Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
@@
 
-\paragraph{Attiya--Fouren Lower Bound}
-Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes).
-Their result says any implementation of queues, using reads,
-writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$
-amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}. It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.
-
 \pagebreak
@@
 \bibliography{main.bib} 
-\bibliographystyle{plain}
+\bibliographystyle{acm}
 

### 2022-10-19 thesis/thesis.tex
finalize draft

@@
 To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem.
-We show our algorithm resolves this problem and has worst case running time better than prior lock-free queues.
+We show our algorithm avoids this problem and has worst case running time better than prior lock-free queues.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
@@
 
-A system is called \it{asynchronous} when processes in the system run at arbitrary varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
+A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
 Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy showed \cite{10.1145/114005.102808} that we one cannot implement concurrent version of all data structures with multi-reader multi-writer registers alone, introducing a hieachy on how powerful are objects to solve consensus problem among processes.
+Herlihy showed \cite{10.1145/114005.102808} that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes.
 Objects like LL/SC, CAS can be used to reach consensus among any number of processors.
-CAS object implements an atomic Compare \& Swap(\nf{new, old}) operation, which if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns {true}, otherwise returns \nf{false}.
-In this work we use Compare \& Swap (\nf{CAS}) objects to reach universal synchronization.
+A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.
+In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.
 
@@
 
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free bounded poly-log time queue.
-We design a tournament tree in which processes try to propagate their operations up to the root to be linearized and they can do queries to get information about the linearization stored in the root.
-To solve the CAS Retry problem when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations.
+We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
+We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
+Processes can do queries to get information about the linearization stored in the root.
+To avoid the CAS Retry problem, when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
 \nf{CAS} instructions cost more than other instructions for processor.
-Each operation in our queue does at most $14\log p$ \nf{CAS} operations.
+Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
 We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations.
-The garbage collection remains open which we have pointed to a way we think it can be handled.
+We do not address garbage collection but we describe a way we think it can be handled.
 
@@
 \end{figure}
-\bf{TODO: make circles the same size}
+% \bf{TODO: make small circles the same size}
 
@@
 They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-Asin  previous algorithms, the worst case happens in the case where the contention is high when $p$ concurrent enqueues happene, their nodes have to be appended to the linked list one by one.
+As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
 The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
@@
 When the array is occupied an empty array is linked to the array and tail pointers are updated.
-Global head points to the array containing the first element in the queue, and head that points to the first element in that array.
-Global tail and tail pointers are similar (see~Figure~\ref{fig::sundell}).
-The pointers might be behind their real places, a process updates the position of the pointers when using them.
-When a process wants to write an element to current head it uses \nf{CAS}, so if $p$ processes try to enqueue simultaneously, the step (also \nf{CAS}) complexity remains $\Omega(p)$.
+A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
+Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
+A process updates the position of the pointers after it does an operation.
+One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
+They mention how to scan the arrays to update pointers while doing an operation.
+A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$.
 Their design is lock-free but not wait-free.
@@
 
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supprting futures.
-A future is an operation that can be applied on the queue when it is requested.
+Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
+In their queue, operations  return future objects instead of responses.
+Later when the response is needed it can be evaluated from the future object.
+They also define a weaker linearizabilty condition such that each operation can be linearized between its invocation and when its future is evaluated.
 Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
 They use some properties of the size of the queue before and after a batch which is similar to a part of our work.
-Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue and it supports a safety condition weaker than linearizabilty introduced by the authors, called Extended Medium Futures Linearizability.
+Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue.
 
@@
 Their work is based on a circular queue using bounded memory.
-They design a slow phase which is wait-free bounded.
-If a process has no progress other processes try to help it in its \nf{CAS} loop in its slow phase.
-Since there can be $p$ processes helping, the worst-case complexity cannot be better than $\Omega(p)$.
+When a process wishes to do an enqueue or a dequeue, it starts two paths.
+One is faster but its running time is not bounded the other one is slower but has a time bound.
+They show that these two paths do not effect each other and the queue remains consistent.
+If a process has no progress other processes help its slow path to finish.
+The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help.
+Because of this, the worst-case complexity cannot be better than $\Omega(p)$.
 
@@
 \subsection{Universal Constructions and other Poly-log Time Data Structures}
-Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}.
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
+A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}.
 We can implement a concurrent queue using a universal construction.
 Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log n)$ worst case step complexity, using atomic access to $\Omega(n \log n)$ bit words.
-Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-Their universal construction cannot be used to create any data structure, but a set of objects called closed objects.
-We mention a non-practical universal construction with poly-log \nf{CAS} invocations in page 12.
+He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst case step complexity, using atomic access to $\Omega(p \log p)$ bit words.
+Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}.
+Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object).
+We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.
+
+% TODO: page
 
 Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes and when a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree.
-If we have an array and two Fetch\&Inc objects head and tail we can implement a queue: to do an enqueue we can Fetch\&Inc head and write in the value fetched of the array and to do a dequeue we can Fetch\&Inc the tail and fetched the array's cell in the value returned.
+Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes.
+When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree.
+There are  some similarities between designing a queue and a Fetch\&Inc object.
+A Fetch\&Inc object can be constructed from a queue.
+The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation in a while.
 
 \subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list based queues one might guess $p$ factor is inherent in time complexity of concurrent queues.
-Attiya and Fouren proved a lower bound on amortized complexity of lock-free queues from the factor of contention (number of concurrent processes).
-Their result says any implementation of queues, using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that their result does not apply on our design and we manage to reach poly-log time complexity.
-It is an open problem to create the reduce the gap between our algorithm and Attiya--Fouren lower bound.
+Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in time complexity of concurrent queues.
+Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard ti $c$, the number of concurrent processes.
+Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
+The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.
 
@@
 
+\subsection{Attiya--Fouren Lower Bound}
+As we talked earlier in Section 2.4 Attiya--Fouren lower bound says a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
+Our amortized complexity is $O(log^2p+log q)$.
+It is an open problem to the reduce the gap between our algorithm and Attiya--Fouren lower bound.
+
 \pagebreak

### 2022-10-25 thesis/thesis.tex
apple franck comments

@@
 \begin{document}
-\title{A Wait-free Queue with Polylogarithmic Step Complexity}
+\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
 \author{Hossein Naderibeni \\supervised by Eric Ruppert}
@@
 To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem.
-We show our algorithm avoids this problem and has worst case running time better than prior lock-free queues.
+We show that our algorithm avoids this problem and has worst case running time better than prior lock-free queues.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
@@
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time data structure is locked.
+Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time the data structure is locked.
 For these reasons, we are interested in the design of an efficient shared queue without using locks.
@@
 If the queue is empty it returns \texttt{null}.
-In a concurrent version of queue data structure, operations do not happen one at a time.
+In a concurrent version of the queue data structure, operations do not happen one at a time.
 The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
@@
 Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy showed \cite{10.1145/114005.102808} that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes.
+Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes.
 Objects like LL/SC, CAS can be used to reach consensus among any number of processors.
@@
 
-\bf{Concurrent execution: I didn't understed interleaving steps of processes}
+% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
 
@@
   \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can be occurred before termination of \texttt{Enqueue(A)} and return \nf{null} or can be occurred after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
+\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occurr before termination of \texttt{Enqueue(A)} and return \nf{null} or can occurr after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
 \end{figure}
@@
 Processes can do queries to get information about the linearization stored in the root.
-To avoid the CAS Retry problem, when a process is propagating its operation to a node we create a helping mechanism that ensures operations are propagated after at most 2 \nf{CAS} invocations.
+We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
@@
 \paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of related word done in the area and the motivation of our implementation.
+Chapter 2 gives an outline of related work done in the area and the motivation of our implementation.
 Previous lock-free queues and their common problem is presented in Section 2.1.
@@
 Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 is summoned with a lower bound on the amortized time complexity of shared queues.
+Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
 
@@
 Chapter 5 is devoted to performance analysis for our implementation.
-We analyze the number of \nf{CAS} instructions out algorithm invokes and the worst-case and amortized time running of the queue.
+We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
 In the end we prove our queue is wait-free.
@@
 The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is occupied an empty array is linked to the array and tail pointers are updated.
+When the array is full, an empty array is linked to the array and tail pointers are updated.
 A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
@@
 When a process wishes to do an enqueue or a dequeue, it starts two paths.
-One is faster but its running time is not bounded the other one is slower but has a time bound.
+One is faster but its running time is not bounded and the other one is slower but has a time bound.
 They show that these two paths do not effect each other and the queue remains consistent.
-If a process has no progress other processes help its slow path to finish.
+If a process makes no progress other processes help its slow path to finish.
 The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help.
@@
 David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinte memory.
+Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
 The author states that to reduce memory usage to be bounded the time per operation increases to linear.
@@
 Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a poly-logarithmic multi-enqueuer multi-dequeuer queue.
+We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
 Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
@@
 Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Incs added by processes.
-When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and return the number of elements in the tree.
+Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
+When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and returns the number of elements in the tree.
 There are some similarities between designing a queue and a Fetch\&Inc object.
 A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation in a while.
+The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation after~\cite{DBLP:conf/podc/ChandraJT98}.
 
 \subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard ti $c$, the number of concurrent processes.
+Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
+Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
 Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
@@
 It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}.
+Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instruction on fixed-size objects in each \tt{Refresh}.
 To do that, we introduce \it{block}s.
@@
 It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure \ref{fig:block} for an example.
+See Figure~\ref{fig:block} for an example.
 A node stores an array of blocks of operations propagated up to it.
@@
 We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently.
+In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
 Thus, there are at most $p$ operations in a block.
@@
 So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering we can design a universal construction; for given object $O$ we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
+With this agreement on linearization ordering we can design a universal construction; for a given object we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
 However, this approach is not enough for an efficient queue.
-We show that we can build an efficien queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
+We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
 We explain how to implement (1) and (2) in poly-logarithmic steps.
 
-After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
+After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
 To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation.
@@
 
-In an execution on a queue where no dequeue operation returns\nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0.
+In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
+In the general case a dequeue returns \nf{null} if and only if the size of the queue after the previous operation is 0.
 We refer to such a dequeue as a \it{null dequeue}.
-If the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B.
+If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
+Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
 So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
@@
 
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively.
+\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
 Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations happens, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
+We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
 
@@
 \paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
+ In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
 
@@
 If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the block prior to $B$ ended.
+The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
+The right subblocks of $B$ are also defined similar.
 In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
@@
 The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively.
+The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
 The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
@@
 If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dequeue}.
+If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
 Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
@@
 \paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
+\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
 Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
 
-\paragraph{\tt{Refresh()}}
+\paragraph{\tt{Refresh()} and \tt{Advance()}}
 The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
@@
 \paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line 333 and its fields are filled similarly for both left and right directions.
+The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
 The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
 Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
+Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
 The field \nf{sum\sub{deq-dir}} is computed similarly.
@@
 Let $par$ be $n$\nf{.parent}.
-Without loss of generality assume for the rest of this section a ${n}$ is the left child of $par$.
+Without loss of generality assume for the rest of this section that ${n}$ is the left child of $par$.
 Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ put in $par$\nf{.blocks}.
+Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
 
@@
 One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks in the block $R_{par}$ propagated needing updates their \nf{super} field.
+This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
 However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
@@
   \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in the prefix for this block}
 \end{itemize}
@@
 
+\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
+\begin{itemize}
+  \item \tt{\sl{int} \size}
+  \textsf{: size of the queue after performing all operations in the prefix for this block}
+\end{itemize}
+
 \end{multicols}
@@
 
-\Function{Object}{Dequeue()}{} \cmt{Creates a block with\nf{null} value element, appends it to the tree and returns its response.}
+\Function{Object}{Dequeue()}{} \cmt{Creates a block with \nf{null} value element, appends it to the tree and returns its response.}
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
@@
 \Else
-\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
-\cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue}in the root.}
+\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
+\cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
 \State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$,i.e.
-, the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}
+\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}
 
@@
 
-\Statex $\triangleright$ \textmd{Creates a block containig new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
+\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
 \Function{boolean}{Refresh()}{}
@@
 \State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p}} \label{setSuper1}
+\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
 \State \tt{head.CAS(h, h+1)} \label{incrementHead}
@@
 \Function{Block}{CreateBlock}{\sl{int} i} 
-\State \tt{block new= \Call{new}{\sl{block}}}
+\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
@@
 \begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$. After $R$ terminates, \nf{$n$.head} is greater than the value read in line \ref{readHead} of $R$.
+Let $R$ be an instance of \nf{Refresh} on a node $n$ that its created block \nf{new} is not empty ($\nf{new.sum}\neq 0$).
+After $R$ terminates, \nf{$n$.head} is greater than the value read in line \ref{readHead} of $R$.
 \end{lemma}
@@
 In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation \ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
+Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
 Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
@@
 
-The next lemma is used to prove the subblocks of two blocks in a node are disjoint.
+The next lemma is used to prove that subblocks of two blocks in a node are disjoint.
 \begin{lemma} \label{lem::headProgress}
@@
 Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value $b$$^\prime$ by Invariant \ref{lem::headPosition}.
+Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value $b^\prime$ by Invariant \ref{lem::headPosition}.
 Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
-The value written into \nf{$b$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
+The value written into \nf{$b$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($i$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
 Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}.
@@
 
-Operations are distinct \nf{Enqueue}s and \nf{Dequeue}s invoked by processes. The next lemma proves that each operation appears at most once in the blocks of a node.
+The next lemma proves that each operation appears at most once in the blocks of a node.
 \begin{lemma}\label{lem::noDuplicates}
@@
 \begin{proof}
-We prove this claim using Lemma \ref{lem::subblocksDistinct}.
+We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
 Assume $op$ is in the subblocks of both $n.blocks[i]$ and $n.blocks[j]$.
-From Corollary \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
+From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
 Since each process puts each operation in only one block of its leaf then $op$ cannot be in two leaf blocks.
@@
 \begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} at time $t$ if $n\nf{.head}>i$.
+$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
 An operation is \it{established} in node $n$ if it is in an established block of $n$.
@@
 
-Now we want to say the blocks of a node grow over time.
+Now we want to say that blocks of a node grow over time.
 \begin{observation}\label{lem::blocksOrder}
@@
 
-The linearization ordering is given by the order that operations appear in the blocks in the root.
+The linearization ordering is given by the order in which operations appear in the blocks in the root.
 \begin{definition}[Linearization] \label{def::lin}
@@
 
-Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
+The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
 \begin{lemma}\label{lem::numX}
@@
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
+    \item If $n$ is an internal node then $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
+    \item If $n$ is an internal node then $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
     \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
@@
 \begin{proof}
- We prove the claim by induction on height of node $n$. Base case (3) for leaves is trivial. Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
+ We prove the claim by induction on the height of node $n$.
+Base case (3) for leaves is trivial and (1) and (2) are vacuously true for leaves.
+Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
 \begin{align*}
@@
 
-Next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
+The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
 \begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$ and \nf{enq} be in $\{\nf{enq},\nf{deq}\}$.
+Let $B$ be $n\nf{.blocks[}b\nf{]}$.
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
-    \item If $n$ is an internal node $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
+    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
+    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
     \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
@@
 \subsection{Propagating Operations to the Root}
-In this section we explain why two \nf{Refresh}es are enough to propagate a nodes operations to its parent.
+In this section we explain why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
 
@@
 Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear in the context.
+We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
 In the text $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
@@
 \vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \\
+EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
 &\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
@@
 
+In every node, \nf{blocks[$0$]} is an empty block without any operations.
 Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Lemma \ref{nonDecreasingHead} we have $n\nf{.left.head}_{309}-1<n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}<n\nf{.head}_{\ref{cas}}$.
+Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
+Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
 From Observation \ref{lem::blocksOrder} the claim follows.
@@
 \begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a complete instance \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
+If $R$ is a complete instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
 \end{corollary}
@@
 By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::prectrueRefresh} we have
-$EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n.\nf{left}}^{^{R_2}t}$.
-
-   If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented  more than two times since $_{\ref{readHead}}^{R_1}t$.
-By Lemma \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[i+1]}$ is non-null.
+By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
+Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
+To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation \ref{lem::blocksOrder} and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
+
+If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
+By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
 Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
 $R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^R_{2}t$.
+So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
 From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh}, \ref{lem::headInc} on $R_3$ the claim holds.
@@
 \begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}.
+  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true} then by Corollary \ref{lem::prectrueRefresh} the claim holds.
+If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded the claim still holds by Corollary \ref{lem::prectrueRefresh}.
 Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
@@
 $A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $b$.
-For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} untill $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary}\label{lem::appendExactlyOnce}
+For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
+\end{proof}
+
+\begin{corollary} \label{lem::appendExactlyOnce}
  After $l$\nf{.Append($b$)} finishes, $b$ is subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
@@
 \begin{proof}
-By the previous corollary and Lemma \ref{lem::appendExactlyOnce} there is exactly one block in each node containing $b$.
+By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing $b$.
 \end{proof}
@@
 
-\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the entire sequence stored in the root. 
+\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root. 
 \begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
@@
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Lemma \ref{lem::sumX}.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Corollary~\ref{lem::sumX}.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
@@
 For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf.
+Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and contains and \nf{Enqueue} operation.
 Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
 
-For the induction step we prove if \nf{$n$.child.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{child}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, so operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Observation \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks's subblocks in the left child of $n$.
+For the induction step we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
+From Definition \ref{ordering} of $E(n,b)$ we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
+By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks's subblocks in the left child of $n$.
 So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
@@
 There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
+We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
+The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
 
@@
 So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
- Finally, \nf{$n$.child.GetEnqueue($subblock$, $i$)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
+Let $subblock$ be the block in $n.\nf{dir}$ containing $E_i(n,b)$.
+Finally, \nf{$n$.child.GetEnqueue($subblock$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
 \end{proof}
@@
 \begin{proof}
-By Lemma \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
+By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
 \end{proof}
@@
 \begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that stores $n$\nf{.parent.blocks[$sb$]}.
-By \ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on $n$\nf{.parent.blocks[$sb$]}.
+By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
 By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super} was read in Line \ref{readParentHead},s before the \nf{CAS} that sets \nf{$B$.super}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$, since \nf{$n$.parent.head} is still equal to $sb$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Invariant \ref{lem::headProgress}.
+The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$, since \nf{$n$.parent.head} is still equal to $sb$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
 The reader may wonder when the case $b\nf{.super}=sb$ happens.
@@
 Let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-To derive a contradiction assume $B$ was not propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
+To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
 
@@
 $R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
+Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
 We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
+Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
 \begin{align*}
@@
 
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Lemma \ref{nonDecreasingHead}.
+So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
 
@@
 So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$ call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
-
-% \nf{$b$.super} is set by some instance of \nf{Refresh} on $n$ or \nf{$n$.parent} showed by $R^\prime$ and \nf{$n$.head} is incremented by some \nf{Refresh} called $R^\prime\prime$.
-Let \nf{$n$.parent.head} value be $hp$ when $R^\prime$ reads it.
-From Lemma \ref{lem::headProgress} \nf{$n$.parent.head} could only increase from ${hp}$ to ${hp+1}$ if $\nf{$n$.parent.blocks[$hp$]}\neq\nf{null}$.
-In other words there should be a successful \nf{$n$.parent.Refresh} for \nf{$n$.parent.head} to be increased.
-Let $R_p$ be \nf{$n$.parent.Refresh} that propagated $b$ to \nf{$n$.parent}.
-We claim there cannot be another successful \nf{$n$.parent.Refresh} after $R^\prime$ reads \nf{$n$.parent.head} and before $R_p$ performs Line \ref{lastLine}.
-
-% \begin{figure}[hbt]  
-%   \center\includegraphics[width=5.5in,height=0.65in]{pics/timeLineSuper1.png}
-%   \caption{Time relations between \nf{R\sub{n},R\sub{p},R',R"}}
-%   \label{fig::timeLine1}
-% \end{figure}
-
-% Without loss of generality  assume $n$ is the left child of \nf{$n$.parent}.
-As we mentioned the first successful \nf{$n$.parent.Refresh} after $t_{\ref{cas}}^{Rn}$ did not set \nf{$b$.super}.
-It might happen maybe because the value $h_{left}$ read for $n$\nf{.head} in Line \ref{readChildHead} is less than or equal to $i$ and $\nf{left.blocks[$h_{left}$]}=\nf{null}$, which means \nf{$n$.head} is advanced but $b$ is still not installed in \nf{$n$.blocks[$i$]} which means $R_n$ has not reached  the Line \ref{cas}.
-
-% Let the first successful \nf{$n$.parent.Refresh} be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh} be $R_{p2}$.
-If $R_{p1}$ reads $x$ in Line \ref{readHead}, then $R_{p2}$ has to read ${x+1}$ in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}).
-See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh} instances $R_{p1}$,$R_{p2}$ on \nf{$n$.parent}.
-
-% \begin{figure}[hbt]  
-%   \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
-%   \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
-%   \label{fig::timeLine2}
-% \end{figure}
-
-% $$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
-% $t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh$ after $R_n$.
-$t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh} instances on a node.
-$t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code.
-The second Refresh on \nf{p} contains $b$ and reads $\nf{n.head}>\nf{i}$.
-
-% So \nf{$b$.super} has set by some process before the second next successful \nf{$n$.parent.Refresh} on Line \ref{setSuper1}.
-Since $i$ is read in the Line \ref{readHead} then the \nf{CreateBlock} in Line \ref{invokeCreateBlock} is going to read  some value for \nf{left.head} greater than ${h}$ and propagates $b$ to \nf{$n$.parent}.
-So if $b$ was not propagated already we are sure the second next successful \nf{$n$.parent.Refresh} propagates $b$.
+This is in contradiction with our hypothesis.
+
+Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
 \end{proof}
@@
 Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From thr previous Lemma we know that $B$ is propagated by second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
+From the previous Lemma we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
 To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and by Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead} $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$.
@@
 Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the correct response by induction hypothesis.
+Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the correct response by the induction hypothesis.
 \end{proof}
@@
 \begin{proof}
-We must show that, every operation that terminates is in $L$ exactly once and if $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Lemma \ref{lem::appendExactlyOnce}.
+We must show that, every operation that terminates is in $L$ exactly once and if $op_{1}$ terminates before $op_{2}$ starts in an execution then $op_{1}$ is before $op_{2}$ in the linearization.
+The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
 For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Lemma \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
+From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
 By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
@@
 Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Eequeue}s (see Definition \ref{ordering}).
+We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
 
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns\nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
+If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
 \end{definition}
@@
 ${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a\nf{null} dequeue then ${Resp(d)=\nf{null}}$.
+Otherwise if ${d}$ is a \nf{null} dequeue then ${Resp(d)=\nf{null}}$.
 \end{definition}
@@
 We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{].size}=0$.
+The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an emprty block with \nf{size} field equal to $0$.
 We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
 By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block.
-By Lemma \ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show ther number of \nf{Enqueue} and \nf{Dequeue} operations in it.
+By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show ther number of \nf{Enqueue} and \nf{Dequeue} operations in it.
 If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
@@
 \begin{lemma} \label{numberOfNND}
-If operations in the root are applied with the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} enqueue operations in $\nf{root.blocks[}0\cdots b\nf{]}$. The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in order $L$ is 
+If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
+\end{lemma}
+\begin{proof}
+There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
+The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
 the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
@@
 \begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied with the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
+  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
 \end{corollary}
@@
 \begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm would be the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead} response of a \nf{Dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
+The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
+\end{lemma}
+\begin{proof}
+\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
 \end{proof}
@@
 We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
 So $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
@@
 
-\begin{lemma}[Worst Case Time analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$,  where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_i$ is the size of the queue at time the response of the \nf{Dequeue} is linearized.
+\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
+The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at time the response of the \nf{Dequeue} is linearized.
 \end{lemma}
@@
 To compute the rank of a \nf{Dequeue} in $D(n)$ the dequeue calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(log p)$ steps.
+\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
 If the response to the dequeue is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
@@
 \begin{proof}
-If we split \nf{DSearch} time cost between the corresponding \nf{Enqueue}, \nf{Dequeue}, in each operation takes $O(\log^2 p +q)$ steps.
+If we split the \nf{DSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
 \end{proof}
@@
 
-\paragraph{Stacks}
+\paragraph{Other Poly-log Wait-free Data structures}
 There are two reasons the block tree worked well to implement a queue.
@@
 \subsection{Attiya--Fouren Lower Bound}
-As we talked earlier in Section 2.4 Attiya--Fouren lower bound says a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
+As we talked earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
 Our amortized complexity is $O(log^2p+log q)$.
-It is an open problem to the reduce the gap between our algorithm and Attiya--Fouren lower bound.
+It is an open problem to the reduce the gap between our algorithm and the Attiya--Fouren lower bound.
 

### 2022-11-01 thesis/thesis.tex
add execution example

@@
 To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}).
+We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
 A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$.
@@
 
-\Tree [.{\begin{tabular}{||c|c|c||}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
+\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
  [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
@@
 \Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} 
+\State \tt{blocks[\head]= B} \label{appendLeaf}
 \State \tt{\head+=1} \label{appendEnd} 
@@
 \pagebreak
+\section{Appendix}
+\subsection{Example Execution of \nf{Propagate}}
+
+Here we want to demonstrate how four processes help each other to  propagate their operations up to the root. We show how the steps of each \nf{Propagate} works as the time goes.
+
+In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends block $B$ to the leaf.
+
+Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ and propagates both $P1$ and $P2$ operations.
+$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
+$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after the $P2$'s successful \nf{CAS} that put $ENQ_1,ENQ_2$.
+So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s \ref{readHead} is after $P2$'s \ref{cas}).
+
+Since the block $P1$ creates is empty it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
+$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it reaches $DEQ_4$ to the \nf{root} even before $P2$ puts its block to $n5$.
+
+Finally $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
+After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ to the \nf{root}.
+$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but for sure its operation is propagated after the second failed \nf{Refresh}.
+See Figure \ref{fig::exec4}
+
+All the fields of some blocks from the Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec45}.
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
+  \caption{$P1,P2,P4$ append operations. \label{fig::exec1}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
+  \caption{$P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
+  \caption{$P4$ puts its block to the \nf{root} before $P2$ appending its block to $n5$.\label{fig::exec3}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
+  \caption{$P3$ suceeds to the \nf{CAS} to the \nf{root}. $P1$'s work is done after two failed \nf{CAS}es. \label{fig::exec4}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
+  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
+\end{figure}
+
+\pagebreak
 

### 2022-11-07 thesis/thesis.tex
apply franck remained comments

@@
 In this section, we look at previous lock-free queues.
+In the worst-case, step complexity of all the lock-free queues, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously.
+Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
+
 \subsection{List-based Queues}
@@
 
-In the worst-case, step complexity of all the list-based queues discussed above, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+CAS Retry Problem is not limited to list-based queues; array-based queues also share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+In all implementations where there is a CAS object with contention $p$ ($p$ processes invoking CAS instructions concurrently), CAS Retry Problem exists.
 Our motivation is to overcome this problem and present a wait-free sublinear queue.
@@
 \end{corollary}
+\begin{proof} Result (1) can be proved using the previous Lemma.
+\begin{align*}
+    B\nf{.sum\sub{enq-left}}=&n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
+    =& \Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
+    &\vdots\\
+    &+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
+    =&\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
+\end{align*}
+
+We can prove (2) and (3) the same as (1).
+\end{proof}
 
@@
 Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation \ref{lem::blocksOrder} and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
+To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
+$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
+So we have
+$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
 
@@
 \begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$.
+To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $B$ in node $n$.
 Without loss of generality $op_1$ is invoked earlier than $op_2$.
@@
 Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$.
-The existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
+The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
+The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
 \end{proof}
@@
 \begin{proof}
-The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an internal node contains at least one operation, due to the test on Line~\ref{addOP}.
-We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue} and \nf{Dequeue} routines.
+From the definition of operations in a block in Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
+The claim follows from the equivalence of the number of operations in a block and the number of operations in the subblocks of the block and the observation that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
+Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
 \end{proof}
@@
 \nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
+
 \begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
@@
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$ by Corollary~\ref{lem::sumX}.
+By Corollary~\ref{lem::sumX} we know that \nf{sum\sub{enq}} field in a block shows the prefix sum of the number of \nf{Enqueue} operations in that block.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
@@
 \begin{proof}
-We must show that, every operation that terminates is in $L$ exactly once and if $op_{1}$ terminates before $op_{2}$ starts in an execution then $op_{1}$ is before $op_{2}$ in the linearization.
+We must show for any execution that every operation that terminates is in $L$ exactly once.
+Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
 The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
@@
 
-% \begin{observation}\label{enqBeforeDeq}
-% By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block.  
-% \end{observation}
-
 In the next lemma we show that the \nf{size} field in each \nf{root block} is computed correctly.
@@
 We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block.
+By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
 By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show ther number of \nf{Enqueue} and \nf{Dequeue} operations in it.
@@
 
-All the fields of some blocks from the Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec45}.
+All the fields of some blocks from the Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec5}.
 

### 2022-11-08 thesis/thesis.tex
fix typos, add bsearch, handle block plymorphism

@@
 Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case called the CAS Retry Problem.
-We show that our algorithm avoids this problem and has worst case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm, is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
+To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, the CAS Retry Problem.
+We show that our algorithm avoids this problem and has a worst-case running time better than prior lock-free queues.
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
 \end{abstract}
@@
 For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same.
+Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same.
 In this run, $p$'s update is overwritten.
-One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
+One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
 Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can make progress at the time the data structure is locked.
+Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
 For these reasons, we are interested in the design of an efficient shared queue without using locks.
@@
 A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.
+\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
 \texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty it returns \texttt{null}.
+If the queue is empty, it returns \texttt{null}.
 In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is ``What properties matter for implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+The question that may arise is ``What properties matter for the implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC, CAS can be used to reach consensus among any number of processors.
+Objects like LL/SC, CAS can be used to reach a consensus among any number of processors.
 A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.
-In this work we use Compare \& Swap (\nf{CAS}) objects to synchronize among  processes.
+In this work, we use Compare \& Swap (\nf{CAS}) objects to synchronize among processes.
 
@@
   \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occurr before termination of \texttt{Enqueue(A)} and return \nf{null} or can occurr after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
+\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before termination of \texttt{Enqueue(A)} and return \nf{null} or can occur after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
 \end{figure}
@@
 We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm there might be an operation that takes infinitely many steps but never terminates.
+A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
 
@@
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for processor.
+\nf{CAS} instructions cost more than other instructions for the processor.
 Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue, we present a way to reduce the memory size to the total number of operations.
-We do not address garbage collection but we describe a way we think it can be handled.
+We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
+Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
 
 \paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of related work done in the area and the motivation of our implementation.
+Chapter 2 gives an outline of related work done in the area and the motivation for our implementation.
 Previous lock-free queues and their common problem is presented in Section 2.1.
-In Section 2.2 we mention some restricted lock-free queues.
+In Section 2.2, we mention some restricted lock-free queues.
 Section 2.3 talks about poly-logarithmic constructions of shared objects.
@@
 In Section 3.1 we give a high-level description of our implementation.
-We also talk about the motivation and requirements in our design to achieve poly-log time.
+We also discuss the motivation and requirements in our design to achieve poly-log time.
 Section 3.2 contains the algorithm itself.
@@
 We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end we prove our queue is wait-free.
+In the end, we prove our queue is wait-free.
 
@@
 In this section, we look at previous lock-free queues.
-In the worst-case, step complexity of all the lock-free queues, includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously.
+In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
@@
   \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram the first element has been dequeued.
+In the first diagram, the first element has been dequeued.
 Red arrows show new pointers and gray dashed arrows show the old pointers.}
@@
 Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one by using \texttt{CAS} operations.
+Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
 However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
@@
   \caption{Baskets queue idea.
-There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}.
-To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again.
-The successful process will be the next one in the basket and so on.}
+At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
+To order the operations in a basket, the mechanism in the process algorithm is to \texttt{CAS} again.
+The successful process will be the next one in the basket, and so on.}
 \end{figure}
@@
 In their queue, operations  return future objects instead of responses.
-Later when the response is needed it can be evaluated from the future object.
-They also define a weaker linearizabilty condition such that each operation can be linearized between its invocation and when its future is evaluated.
+Later when the response is needed, it can be evaluated from the future object.
+They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
 Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the size of the queue before and after a batch which is similar to a part of our work.
-Their queue is not wait-free, in fact if the batch sizes are 1 then the queue is like MS-queue.
+They use some properties of the queue size before and after a batch, similar to a part of our work.
+Their queue is not wait-free, in fact if the batch sizes are 1, then the queue is like MS-queue.
 
@@
 When a process wishes to do an enqueue or a dequeue, it starts two paths.
-One is faster but its running time is not bounded and the other one is slower but has a time bound.
-They show that these two paths do not effect each other and the queue remains consistent.
-If a process makes no progress other processes help its slow path to finish.
-The helping phase suffers from CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help.
+One is faster but its running time is not bounded and the other one is slower but has a time-bound.
+They show that these two paths do not affect each other and the queue remains consistent.
+If a process makes no progress, other processes help its slow path to finish.
+The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help.
 Because of this, the worst-case complexity cannot be better than $\Omega(p)$.
@@
 Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded the time per operation increases to linear.
+The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
 
@@
 Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst case step complexity, using atomic access to $\Omega(p \log p)$ bit words.
+He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$ bit words.
 Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}.
@@
 
-It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
+It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
 Instead we use an implicit representation of the operations propagated together.
@@
 Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.
+In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
 
@@
 The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
+(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
 \end{figure}
@@
 So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering we can design a universal construction; for a given object we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
+With this agreement on linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$.
 However, this approach is not enough for an efficient queue.
@@
 After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation.
+To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
 When we reach a block in a leaf, the operation is explicitly stored there.
@@
 In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}).
+Moreover, for each block, we store two pointers to its last left and right subblock (see Figure \ref{fig::pointer}).
 We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
 
-To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in.
+To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that the operation is in.
 After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
@@
 In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the size of the queue after the previous operation is 0.
+In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
 We refer to such a dequeue as a \it{null dequeue}.
@@
 
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific rank.
+In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
 We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
@@
 \noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claimed that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
+Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
 We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
@@
 Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-Otherwise the response would be \nf{null}.
+Otherwise, the response would be \nf{null}.
 
 \subsection{Details of the Implementation}
-Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
+Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
 \paragraph{\tt{Node}} 
@@
  The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation.
+If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
 If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
 The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are also defined similar.
+The right subblocks of $B$ are also defined similarly.
 In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
@@
 \paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to just propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
+An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
 \nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
@@
 If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
 
@@
 \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
+When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
 
@@
 After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate and it is unnecessary to add an empty block to the tree.
+Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
 Later we will use the fact that all blocks contain at least one operation.
 Line \ref{cas} tries to install \nf{new}.
-If it was successful all is good.
+If it was successful, all is good.
 If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
@@
 We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
+If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
 To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
@@
 Let $par$ be $n$\nf{.parent}.
-Without loss of generality assume for the rest of this section that ${n}$ is the left child of $par$.
+Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
 Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
@@
 Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
+After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
 We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most 1.
@@
 
-\pagebreak
-
 \begin{algorithm}
@@
 
-\pagebreak
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{25}
+
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
+\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+\Function{int}{BinarySearch}{ \sl{int} x, \sl{int} start, \sl{int} end}
+% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
+\While{\nf{start<=end}}
+\State \sl{int} \tt{mid= (start+end)/2}
+\If{\nf{blocks[mid].sum\sub{enq}<x}}
+\State \nf{start= mid+1}
+\Else
+\State \nf{end= mid}
+\EndIf
+\EndWhile
+\State\Return \nf{start}
+\EndFunction{BinarySearch}
+
+\end{algorithmic}
+\end{algorithm}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
+\setcounter{ALG@line}{36}
 \Statex
@@
 \State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
+\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
 \State \tt{start= max(start-(end-start), 0)} \label{doubling}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{34}
+\setcounter{ALG@line}{45}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{39}
+\setcounter{ALG@line}{50}
 \begin{multicols}{2}
@@
 \EndFunction{Advance}
-
-\pagebreak
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{i}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{i} of \tt{sum\sub{enq}} field.
-Returns the index of the leftmost block in \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{i}}.
-\Function{int}{BinarySearch}{ \sl{int} i, \sl{int} start, \sl{int} end}
-\State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$i\}}
-\EndFunction{BinarySearch}
 
@@
 \If{\tt{this} \bf{is} \tt{root}}
+\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
 \State \tt{new.\size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{85}
+\setcounter{ALG@line}{94}
 
@@
 \State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
+\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
 \State \tt{subBlock= left.\Call{BinarySearch}{ i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
@@
 
+\subsection{Example Execution}
+
+Here we want to demonstrate how four processes help each other to  propagate their operations up to the root as time goes on and then compute their responses.
+
+In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends block $B$ to the leaf.
+
+Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ and propagates both $P1$ and $P2$ operations.
+$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
+$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after the $P2$'s successful \nf{CAS} that put $ENQ_1,ENQ_2$.
+So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s \ref{readHead} is after $P2$'s \ref{cas}).
+
+Since the block $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
+$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it reaches $DEQ_4$ to the \nf{root} even before $P2$ puts its block to $n5$.
+
+Finally, $P4$ goes to sleep, and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
+After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ to the \nf{root}.
+$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but for sure, its operation is propagated after the second failed \nf{Refresh}.
+See Figure \ref{fig::exec4}
+
+All the fields of some blocks from Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec5}.
+$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
+$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
+It computes the superblock of its leaf block, which is the second block in $n5$.
+Then computes the superblock of $ENQ_1,DEQ_2$.
+As we can see\nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
+Finally reaches the block $ENQ_1,ENQ_3,DEQ_2$.
+The queue size after the previous block is 0 so $DEQ_2$ 's response is $ENQ_1$'s element.
+Then it tries to get the $ENQ_1$'s element.
+$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$; because $1<ENQ_1,ENQ_3,DEQ_2$\nf{.sum\sub{enq-left}}.
+It reaches the leaf $n1$ and returns $ENQ_1$'s element.
+$DEQ_4$ returns its response the same.
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
+  \caption{$P1,P2,P4$ append operations. \label{fig::exec1}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
+  \caption{$P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
+  \caption{$P4$ puts its block to the \nf{root} before $P2$ appending its block to $n5$.\label{fig::exec3}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
+  \caption{$P3$ succeeds to the \nf{CAS} to the \nf{root}. $P1$'s work is done after two failed \nf{CAS}es. \label{fig::exec4}}
+\end{figure}
+
+\begin{figure}
+  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
+  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
+\end{figure}
+
+\clearpage
+
 \section{Proof of Correctness}
@@
 \subsection{Basic Properties}
-In this subsection we talk about some properties of blocks and fields of the tree nodes.
+In this subsection, we talk about some properties of blocks and fields of the tree nodes.
 
@@
 \begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful then the claim holds.
-Otherwise $n.\nf{head}$ has changed from the value that was read in Line \ref{readHead}.
+If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
+Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
 By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
@@
 The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change  then it  still holds after the change.
+We show that if the invariant holds before such a change, then it still holds after the change.
 
 In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
+Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
 Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
@@
 
-The next lemma is used to prove that subblocks of two blocks in a node are disjoint.
+The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
 \begin{lemma} \label{lem::headProgress}
@@
 Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
+Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
 By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the ranges $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ and $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$j$].end\sub{left}}]$ cannot overlap.
@@
 From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf then $op$ cannot be in two leaf blocks.
+Since each process puts each operation in only one block of its leaf, then $op$ cannot be in two leaf blocks.
 This is a contradiction.
@@
 \begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of blocks of a node after the \nf{CAS} contains the the set of blocks before the \nf{CAS}.
+Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
 \end{proof}
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of operations in b. Operations in the leaves are ordered with numerical order shown in the drawing.}
+  \caption{Order of operations in b. Operations in the leaves are ordered in the numerical order shown in the drawing.}
 \end{figure}
@@
 Now we define the ordering of operations stored in each node.
-In the non-root nodes we only need to order operations of a type among themselves.
-Processes are numbered from 1 to $p$ and leaves of the tree are assigned from left to right.
+In the non-root nodes, we only need to order operations of a type among themselves.
+Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
 We will show in Lemma \ref{blockSize} that there is at most one operation from each process in a given block.
@@
 
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
+The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
 \begin{lemma}\label{lem::numX}
@@
  We prove the claim by induction on the height of node $n$.
-Base case (3) for leaves is trivial and (1) and (2) are vacuously true for leaves.
+Base case (3) for leaves is trivial, and (1) and (2) are vacuously true for leaves.
 Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
@@
 \end{corollary}
-\begin{proof} Result (1) can be proved using the previous Lemma.
+\begin{proof} Result (1) can be proved using the previous lemma.
 \begin{align*}
@@
 \subsection{Propagating Operations to the Root}
-In this section we explain why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
+This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
 
@@
 We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
+In the text, $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
 \end{definition}
@@
 
-In the next two results we show for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
+In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
 
@@
 \begin{proof}
-The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
+The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
 Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
@@
 
-In the next lemma we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
+In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
 
@@
 \begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true} then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
+  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
+If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
+Otherwise, both failed, and the claim is satisfied by Lemma \ref{doubleRefresh}.
 \end{proof}
@@
 \begin{proof}
-$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $b$.
+$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the leaf's parent where it appended $b$.
 For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
@@
 \begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($b$)} finishes, $b$ is subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
+ After $l$\nf{.Append($b$)} finishes, $b$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
 \end{corollary}
@@
 
-First we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
+First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
 
 \begin{lemma}\label{blockSize}
-Each block contains at most one operation of each process, and therefore at most $p$ operations in total.
+Each block contains at most one operation of each process and, therefore, at most $p$ operations in total.
 \end{lemma}
@@
 \begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}} and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
+ From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
@@
 By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
+Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
 \end{proof}
@@
 \begin{proof}
-We are going to prove this lemma by induction on the height of node $n$.
+We will prove this lemma by induction on the node's height $n$.
 For the base case, suppose $n$ is a leaf.
@@
 
-For the induction step we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$ we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks's subblocks in the left child of $n$.
+For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
+From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
+By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
 So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
@@
  \end{align*}
-First we consider the case where the \nf{Enqueue} we are looking for is in the left child.
+First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
 There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
@@
 
-So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
+So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
 Let $subblock$ be the block in $n.\nf{dir}$ containing $E_i(n,b)$.
@@
 
-  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
+  \caption{The number and order of the enqueue operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
 Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
@@
 
-Furthermore, if $B$\nf{.super} was not set earlier it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
+Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
 \end{proof}
@@
 \begin{proof}
-% After that $B$ is installed $n$\nf{.parent.head} is read and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
+% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
 Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
 Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From the previous Lemma we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and by Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead} $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$.
+From the previous lemma we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
+To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and by Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead} $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$.
 The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
@@
 
-Now using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue} we prove its correctness.
+We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
 
@@
 By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeus in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
+If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
 Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the correct response by the induction hypothesis.
+Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
 \end{proof}
@@
 \end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves.
+Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
 We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
@@
 
-Next we define  the responses that \nf{Dequeue}s should return, according to the linearization.
+Next, we define  the responses that \nf{Dequeue}s should return, according to the linearization.
 \begin{definition}
@@
 
-In the next lemma we show that the \nf{size} field in each \nf{root block} is computed correctly.
+In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
 \begin{lemma}\label{sizeCorrectness}
@@
 We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an emprty block with \nf{size} field equal to $0$.
+The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
 We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
 By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show ther number of \nf{Enqueue} and \nf{Dequeue} operations in it.
+By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
 If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is same as the assignment on Line \ref{computeLength}.
+Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
+In both cases, this is the same as the assignment on Line \ref{computeLength}.
 \end{proof}
@@
 \section{Analysis}
-In this section we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
+In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
 \begin{proposition}
@@
 \begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times and  every \nf{Refresh} invokes at most 7 \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
+  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
 \end{proof}
@@
 There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
+Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
 The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
+After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
 We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of the every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
 So $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
@@
 \begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at time the response of the \nf{Dequeue} is linearized.
+The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
 \end{lemma}
@@
 The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$ the dequeue calls \nf{IndexDequeue()}.
+To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
 \nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the dequeue is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
+If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
 Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log$ size of the queue when \nf{Enqueue} is linearized$)+\log$ size of the queue when \nf{Dequeue} is linearized$)$.
@@
 Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs we talk about possible improvements on block trees and the data structures that we can implement with block trees.
+In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
 
@@
 The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
+To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
 We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
@@
 When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to to the \nf{CAS} on \nf{arr[$i$]}.
+The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
+Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
 Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
 The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely, since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
+The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
 \begin{figure}[hbt]  
@@
 We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e, all of its enqueues have been dequeued and the responses of all of its dequeues have been computed.
+We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all of its enqueues have been dequeued, and the responses of all of its dequeues have been computed.
 We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished then all blocks before $B$ are also finished.
+If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
 Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-To throw the garbage in the blocks away we cannot use arrays (or vectors).
+We cannot use arrays (or vectors) to throw the garbage blocks away.
 We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$, where \nf{split(i)} removes all the indices less than \nf{i}.
-If each process tries to do the garbage collection every $p^2$ operation on the queue then the amortized complexity remains the same.
-We can use a concurrent implementation of persistent red black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red black trees in a similar way.
+If each process tries to do the garbage collection every $p^2$ operation on the queue, then the amortized complexity remains the same.
+We can use a concurrent implementation of persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
+Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
 
@@
 An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such data structure.
+We can modify our queue to design such a data structure.
 An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
@@
 Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} is not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-It may be possible to create wait-free poly-logarithmic implementaion of other objects whose operations satisfy these two conditions.
+Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
+Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
 
 \subsection{Attiya--Fouren Lower Bound}
-As we talked earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
+As we talked about earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
 Our amortized complexity is $O(log^2p+log q)$.
-It is an open problem to the reduce the gap between our algorithm and the Attiya--Fouren lower bound.
+It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.
 
 \pagebreak
-\section{Appendix}
-\subsection{Example Execution of \nf{Propagate}}
-
-Here we want to demonstrate how four processes help each other to  propagate their operations up to the root. We show how the steps of each \nf{Propagate} works as the time goes.
-
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends block $B$ to the leaf.
-
-Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ and propagates both $P1$ and $P2$ operations.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after the $P2$'s successful \nf{CAS} that put $ENQ_1,ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s \ref{readHead} is after $P2$'s \ref{cas}).
-
-Since the block $P1$ creates is empty it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it reaches $DEQ_4$ to the \nf{root} even before $P2$ puts its block to $n5$.
-
-Finally $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ to the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but for sure its operation is propagated after the second failed \nf{Refresh}.
-See Figure \ref{fig::exec4}
-
-All the fields of some blocks from the Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec5}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption{$P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block to the \nf{root} before $P2$ appending its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption{$P3$ suceeds to the \nf{CAS} to the \nf{root}. $P1$'s work is done after two failed \nf{CAS}es. \label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
 

### 2022-11-11 thesis/thesis.tex
fix revisions, apply style guidelines

@@
-\documentclass[10pt]{article}
+\documentclass[11pt]{article}
 \renewcommand{\baselinestretch}{1.8}
@@
 \begin{document}
-\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
-\author{Hossein Naderibeni \\supervised by Eric Ruppert}
-\maketitle
-
-\begin{abstract}
+
+\begin{center}
+
+\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
+
+\vspace{2em}
+\Large{\bf{Hossein Naderibeni}}
+
+\vspace{3em}
+\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
+
+\vspace{2em}
+Graduate Program in Computer Science\\
+York University,\\
+Toronto, Ontario
+
+\vspace{2em}
+November, 2022
+
+\vspace{3em}
+\textcopyright Hossein Naderibeni, 2022
+
+\end{center}
+\thispagestyle{empty}
+\pagebreak
+
+\normalsize
+\section*{Abstract}
+
 In this work, we introduce a novel linearizable wait-free queue implementation.
 Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, the CAS Retry Problem.
-We show that our algorithm avoids this problem and has a worst-case running time better than prior lock-free queues.
+To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
+We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\end{abstract}
-
+\pagenumbering{roman}
+\pagebreak
+
+\section*{Acknowledgements}
+
+First, I want to thank my supervisor Eric Ruppert. ...
+
+I would also like to thank professor Franck van Breugel. ...
+
+I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
+
+I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
+
+\pagebreak
 \tableofcontents
+
 \pagebreak
+\listoffigures
+
+\pagebreak
+\listoftables
+\newpage
+
+\pagebreak
+
+\pagenumbering{arabic}
 \section{Introduction}
@@
 The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure to work with.
+It is not hard to see why multiple processes cannot share a sequential data structure.
 For example, consider two processes trying to append to a sequential linked list simultaneously.
@@
 In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is ``What properties matter for the implementation of a shared data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+The question that may arise is ``What properties matter for the implementation of a shared data structure?''.
+Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC, CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare \& Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}, otherwise it returns \nf{false}.
-In this work, we use Compare \& Swap (\nf{CAS}) objects to synchronize among processes.
+Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
+Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
+A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
+In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
 
@@
   \center\includegraphics[scale=0.5]{pics/good}
-  \caption{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
+  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
+Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
 \end{figure}
@@
   \center\includegraphics[scale=0.5]{pics/bad}
-  \caption{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before termination of \texttt{Enqueue(A)} and return \nf{null} or can occur after termination of \texttt{Enqueue(A)} and return \texttt{A}.}
+  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
+\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
 \end{figure}
@@
 Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated after at most 2 \nf{CAS} invocations.
+We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
@@
 \paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of related work done in the area and the motivation for our implementation.
+Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
 Previous lock-free queues and their common problem is presented in Section 2.1.
@@
 We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself.
+Section 3.2 contains the algorithm itself and there is an example of a concurrent exuection on the algorithm described in Section 3.3.
 
@@
 
-Chapter 5 is devoted to performance analysis for our implementation.
-We analyze the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
+Chapter 5 is devoted to complexity analysis for our implementation.
+We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
 In the end, we prove our queue is wait-free.
@@
 In this section, we look at previous lock-free queues.
-In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or dequeue simultaneously.
+In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
@@
   \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
+  \caption[MS-queue structure]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
 In the first diagram, the first element has been dequeued.
@@
 Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
+However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the worst-case step complexity remains $\Omega(p)$ per operation.
 
@@
   \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption{Baskets queue idea.
-At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, the mechanism in the process algorithm is to \texttt{CAS} again.
+  \caption[Baskets queue]{At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
+To order the operations in a basket, processes perform another \nf{CAS}.
 The successful process will be the next one in the basket, and so on.}
 \end{figure}
-% \bf{TODO: make small circles the same size}
 
 Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous added node to the new node.
+MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
 They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
 As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
+The worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
 
@@
   \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique to achieve wait-freedom.
+    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}]{\label{fig::sundell}Global pointers point to arrays.
+Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
+\end{figure}
+
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
 Their step complexity is $\Omega(p)$ because of the helping mechanism.
@@
 They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free, in fact if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology.
+Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
+
+Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
 Their work is based on a circular queue using bounded memory.
 When a process wishes to do an enqueue or a dequeue, it starts two paths.
-One is faster but its running time is not bounded and the other one is slower but has a time-bound.
+The fast path  ensures good performance while the slow path ensures termination.
 They show that these two paths do not affect each other and the queue remains consistent.
@@
 
-CAS Retry Problem is not limited to list-based queues; array-based queues also share the CAS retry problem as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-In all implementations where there is a CAS object with contention $p$ ($p$ processes invoking CAS instructions concurrently), CAS Retry Problem exists.
+The CAS Retry Problem is not limited to list-based queues; array-based queues also share it as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
 Our motivation is to overcome this problem and present a wait-free sublinear queue.
@@
 
-\subsection{Universal Constructions and other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object, introduced by Herlihy~\cite{10.1145/114005.102808}.
+\subsection{Universal Constructions and Other Poly-log Time Data Structures}
+A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
+The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
 We can implement a concurrent queue using a universal construction.
 Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$ bit words.
-Chandra, Jayanti and Tan  introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses using reasonably sized words~\cite{DBLP:conf/podc/ChandraJT98}.
-Their universal construction cannot be used to create any data structure, but a set of objects called closed objects (a queue is not a closed object).
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions on page 13.
+He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
+Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
+However their algorithm cannot be used to create a queue.
+We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the page 13.
 
@@
 Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do Fetch\&Inc it adds a Fetch\&Inc to the tree and returns the number of elements in the tree.
+When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
 There are some similarities between designing a queue and a Fetch\&Inc object.
 A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because it is the first wait-free data structure achieving poly-log implementation after~\cite{DBLP:conf/podc/ChandraJT98}.
+The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
+Also, it is one of the few wait-free data structures achieving poly-log complexity.
 
@@
 Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that their result does not contradict ours and we manage to reach poly-log time complexity.
+Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
+The surprising point is that since their result is about when contention is low, their lower bound does not contradict with our algorithm's complexity and we manage to reach poly-log time complexity.
 
@@
 
-There is a shared binary tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering of the operations invoked by processes.
-Each process has a leaf in which the operations invoked by the process are stored in order.
+There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
+The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree})Each process has a leaf in which the operations invoked by the process are stored in order.
 When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
@@
 \end{center}
-\caption{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
+\caption[Tournament tree]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
 Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
@@
 We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$.
+A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
 We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
@@
 \end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
+  \caption{Before the \nf{Refresh}}
 \end{subfigure}
@@
 \end{tikzpicture}}
-  \caption{New operations are appended.}
+  \caption{After the \nf{Refresh}}
 \end{subfigure}
 \hspace{6em}
-\caption{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append. Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
+\caption[A successful \nf{Refresh}]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
+Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
 \end{center}
@@
   \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
+  \caption[Two consecutive failed \nf{Refresh}es by a process]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
+\end{figure}
+
+We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
 The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
 This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
+Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
 After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
@@
 It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instruction on fixed-size objects in each \tt{Refresh}.
+Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
 To do that, we introduce \it{block}s.
@@
 A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks.
+A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
 We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
@@
 
-\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
+\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
       {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
@@
 \end{tikzpicture}
-\caption{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
+\caption[Showing the operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
 In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
@@
            [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c|c||}  \hline (2,1) & (5,0) & (4,2) & (3,4) & (6,3) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
+            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
 \end{small}
@@
 \end{center}
-\caption{\label{fig:block} Using blocks to represent operations.
+\caption[Showing the operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
 Blocks between two lines $||$ are propagated together to the parent.
@@
 So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$.
+With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning response for $op$.
 However, this approach is not enough for an efficient queue.
@@
 In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two pointers to its last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that the operation is in.
+Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
+We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
+
+To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
 After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
@@
 \end{small}
+\caption{An Execution on a queue separated by the operations in the root blocks}
 \end{table}
@@
 \noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
+$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
 \noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
@@
 
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
+Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
+\begin{center}
+    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
+\end{center}
+if \vspace{-3em}
+\begin{center}
+    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
+\end{center}
 Otherwise, the response would be \nf{null}.
@@
 The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are also defined similarly.
+The right subblocks of $B$ are defined similarly.
 In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
@@
 
+We now describe the routines used in the implementation.
+
 \paragraph{\tt{Enqueue($e$)}}
@@
 Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
 
@@
 Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks.
+\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
 The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
@@
 
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in $b$th the block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
 \paragraph{\tt{IndexDequeue($b,i$)}}
@@
 Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}.
+Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
 Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
 
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating it.
+Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
 One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
@@
 After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most 1.
+We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
+
+\paragraph{\tt{FindResponse($b, i$)}}
+To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
+If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
+If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
+Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+
+\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
+We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
+If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
+To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
+We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
+\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
+A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
+The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
+Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
 
@@
 \newpage
-\newgeometry{left=0.6in,right=0.6in,top=0.6in,bottom=0.8in}
 
 \subsection{Pseudocode} \label{algQ}
+We present our algorithm in pseudocode.
+Page 23 contains the description of the fields in the tree nodes and the blocks.
+Value of any uninitialized field is \nf{null}.
+Page 24 contains major routines and the rest is the auxiliary routines
+The abbreviations bellow are used in the pseudocode and the proof of correctness.
+\begin{itemize}
+ \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \{enq, deq\})
+%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
+  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
+\end{itemize}
+
 \begin{algorithm}
@@
 \setcounter{ALG@line}{1}
-\begin{multicols}{2}
 
@@
 
-\pagebreak
-
 \Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
@@
   \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}
-  \textsf{: \# enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}
-  \textsf{: \# dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}
-  \textsf{: \# enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}
-  \textsf{: \# dequeues in \tt{right.blocks[1..end\sub{right}]}}
+  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
+  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
 \end{itemize}
@@
 
-\end{multicols}
 \end{algorithmic}
 \end{algorithm}
-
-\begin{footnotesize}
-
-\it{Abbreviations used in the code and the proof of correctness.}
-\begin{itemize}
- \item \tt{blocks[b].sum\sub{x}=blocks[b].sum\sub{x-left}+blocks[b].sum\sub{x-right}}  \tt{ (for internal blocks where b$\geq$0 and x $\in$ \{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[b].num\sub{x}=blocks[b].sum\sub{x}-blocks[b-1].sum\sub{x}} \\ \tt{(for all blocks where b>0 and x $\in$ \{ enq, deq, enq-left, enq-right, deq-left, deq-right\})}
-\end{itemize}
-\end{footnotesize}
 
@@
 \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
+\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
 \EndIf
@@
 \Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{ \sl{int} x, \sl{int} start, \sl{int} end}
+\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
 % \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
@@
 \begin{algorithm}
-\caption{Root}
+\caption{\tt{\sl{Root}}}
 \begin{algorithmic}[1]
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_e(root)=E_i(root,b)$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within block \nf{b} of the \nf{root}.}
-
-\Function{<int, int>}{DSearch}{\sl{int} e, \sl{int} end}
+\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within the \nf{b}th block in the \nf{root}.}
+
+\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
 \State \tt{start= end-1} \label{dsearchStart}
@@
 \EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch( e, start, end)} \label{dsearchBinarySearch}
+\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
 \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
 \State\Return \tt{<b,i>}
-\EndFunction{DSearch}
+\EndFunction{DoublingSearch}
 \end{algorithmic}
@@
 \begin{algorithm}
-\caption{Leaf}
+\caption{\tt{\sl{Leaf}}}
 \begin{algorithmic}[1]
@@
 \State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head+=1} \label{appendEnd} 
+\State \tt{\head= \head+1} \label{appendEnd} 
 \State \tt{parent.}\Call{Propagate()}{} 
@@
 \setcounter{ALG@line}{50}
-\begin{multicols}{2}
 
@@
 
-\pagebreak
-
-\Function{void}{Advance}{\sl{int} h}
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{73}
+
+\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
 \State \tt{h\sub{p}= parent.\head} \label{readParentHead}
@@
 
-\Statex\cmt{Creates and returns the block to be installed in \tt{blocks[i]}.
-Created block includes \tt{left.blocks[index\sub{prev}+1..index\sub{last}]} and \tt{right.blocks[index\sub{prev}+1..index\sub{last}]}.}
-\Function{Block}{CreateBlock}{\sl{int} i} 
+\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
 \State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
@@
 \State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine}
+\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
 \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
 \State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + }
-\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + }
-\Statex \hspace{10.7em}\tt{block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
+\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
+\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
 \EndFor
@@
 \State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.\size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}}
-\Statex \hspace{8.5em} \tt{- new.num\sub{deq}, 0)}\label{computeLength}
+\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
 \EndIf
@@
 \begin{algorithm}
-\caption{Node}
+\caption{\tt{\sl{Node}}}
 \begin{algorithmic}[1]
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
+\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
 \If{\tt{this} \bf{is} \tt{leaf}}
 \State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
-\State \tt{subBlock= left.\Call{BinarySearch}{ i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
-\State \Return\tt{left.}\Call{GetEnqueue}{subBlock, i} 
+\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
+\State \tt{subBlockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
+\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
+\State \Return\tt{left.}\Call{GetEnqueue}{subBlockIndex, i} 
 \Else
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlock= right.\Call{BinarySearch}{ i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
-\State \Return\tt{right.}\Call{GetEnqueue}{subBlock, i} 
+\State\tt{subBlockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
+\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
+\State \Return\tt{right.}\Call{GetEnqueue}{subBlockIndex, i} 
 \EndIf
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_i(this, b)=D_y(root,x)$.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 \State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{sb= (parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1)}\label{computeSuper}
+\State \tt{superBlockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
+\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
 
 \If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superBlockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
 \Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superBlockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
+\State \tt{i+= parent.blocks[superBlockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
 \EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
+\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlockIndex, i}
 \EndIf
@@
 
-\restoregeometry
 \newpage
@@
 
-Here we want to demonstrate how four processes help each other to  propagate their operations up to the root as time goes on and then compute their responses.
-
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends block $B$ to the leaf.
-
-Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ and propagates both $P1$ and $P2$ operations.
+Here we want to demonstrate how four processes help each other to  propagate their operations up to the root and then compute their responses.
+
+Initially, each node in the tree contains one empty block.
+In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
+Line \ref{appendLeaf} appends a block to the process's leaf.
+
+Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
 $P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after the $P2$'s successful \nf{CAS} that put $ENQ_1,ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s \ref{readHead} is after $P2$'s \ref{cas}).
-
-Since the block $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it reaches $DEQ_4$ to the \nf{root} even before $P2$ puts its block to $n5$.
-
-Finally, $P4$ goes to sleep, and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ to the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but for sure, its operation is propagated after the second failed \nf{Refresh}.
-See Figure \ref{fig::exec4}
-
-All the fields of some blocks from Figure \ref{fig::exec4} are filled in Figure \ref{fig::exec5}.
+$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
+So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
+
+Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
+$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
+
+Then as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep, and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
+After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
+$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
+
+All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
 $P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
@@
 It computes the superblock of its leaf block, which is the second block in $n5$.
-Then computes the superblock of $ENQ_1,DEQ_2$.
-As we can see\nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-Finally reaches the block $ENQ_1,ENQ_3,DEQ_2$.
+Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
+As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
 The queue size after the previous block is 0 so $DEQ_2$ 's response is $ENQ_1$'s element.
-Then it tries to get the $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$; because $1<ENQ_1,ENQ_3,DEQ_2$\nf{.sum\sub{enq-left}}.
-It reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$DEQ_4$ returns its response the same.
+Then $P2$ tries to get $ENQ_1$'s element.
+$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{.sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
+Similarly , $P2$ determines that $ENQ_1$ came from the left child of $n5$.
+$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
+$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
+Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
 
@@
   \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations. \label{fig::exec1}}
+  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
 \end{figure}
@@
   \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption{$P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
+  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
 \end{figure}
@@
   \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block to the \nf{root} before $P2$ appending its block to $n5$.\label{fig::exec3}}
+  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
 \end{figure}
@@
   \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption{$P3$ succeeds to the \nf{CAS} to the \nf{root}. $P1$'s work is done after two failed \nf{CAS}es. \label{fig::exec4}}
+  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
+$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
+\label{fig::exec4}}
 \end{figure}
@@
 Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
+After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
 Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
@@
 Let $R$ be an instance of \nf{Refresh} on a node $n$ that its created block \nf{new} is not empty ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in line \ref{readHead} of $R$.
+After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
 \end{lemma}
@@
 
-The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
+The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
 If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
@@
 \begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}.
-Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value $b^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked.
-The value written into \nf{$b$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($i$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($i$)}.
-Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$.
+ If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
+\end{lemma}
+\begin{proof}
+Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
+Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
+Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
+Thus, the \nf{CreateBlock($b-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($b$)} that creates $B$ is invoked.
+The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
+Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
+By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
 The proof for \nf{end\sub{right}} is similar.
@@
 Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the ranges $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ and $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$j$].end\sub{left}}]$ cannot overlap.
+By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
 Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap.
@@
 \begin{definition}[Operations of a block]\label{def::ops}
-A block $b$ in a leaf represents an \nf{Enqueue} if $b\nf{.element}\neq\nf{null}$.
-Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue}.
-The set of operations of block $b$ is the union of the operations in leaf subblocks of $b$.
-We denote the set of operations of block $b$ by ${ops(b)}$ and the union  of operations of a set of blocks $B$ by $ops(B)$.
-We also say ${b}$ contains ${op}$ if $op\in{ops(b)}$.
+A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
+Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
+The set of operations of block $B$ is the union of the operations in leaf subblocks of $b$.
+We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks \bf{B} by $ops(\bf{B})$.
+We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
 \end{definition}
@@
 An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^n_t$ is the set of established operations in node $n$ at time $t$.
+$EST^t_n$ is the set of established operations in node $n$ at time $t$.
 \end{definition}
@@
 
-\pagebreak
-
 \subsection{Ordering Operations}
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption{Order of operations in b. Operations in the leaves are ordered in the numerical order shown in the drawing.}
+  \caption[Order of operations in a block]{Order of operations in the block b. Operations in the leaves are ordered in the numerical order shown in the drawing.}
 \end{figure}
@@
 Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively.
+
+If $n$ is an internal node, then
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
+    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
+    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
+\end{enumerate}
+And for every node n, we have
+\begin{enumerate}[label=(3)]
     \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
@@
 \begin{align*}
-    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{enq}}\\
-    =& B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock}\\
+    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
+    =& B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
     =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    =& \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} &\text{Observation } \ref{sumToNum}\\
-    =& \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big| & \textrm{Induction hypothesis (3)}
+    =& \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
+    =& \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
 \end{align*}
-The last line holds because of the induction hypothesis (3). (2) is similar to (1). Now we prove (3) starting from the Definition of $E(n,b)$.
+First line follows from the Definition of \nf{num\sub{enq}}.
+Second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
+\text{Observation } \ref{sumToNum} results the third line and the last line holds because of the induction hypothesis (3).
+(2) is similar to (1).
+Now we prove (3) starting from the Definition of $E(n,b)$.
 \begin{align*} 
@@
 
-\pagebreak
-
 \subsection{Propagating Operations to the Root}
@@
 \begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
+  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
+  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
 \end{definition}
@@
 
-% \begin{figure}[hbt]
-%   \center\includegraphics[width=5in]{pics/trueRefresh}
-%   \caption{TODO :: Is this picture good or should I remove it? New established operations of the left child are in the new block.(TO UPDATE)}
-% \end{figure}
-
 \begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a complete instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
+If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
 \end{corollary}
@@
   \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
+  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
 \end{figure}
@@
 The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
+The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
 \end{proof}
@@
 
-\nf{DSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DSearch} Correctness]\label{dsearch}
-If $\nf{root.blocks[$end$]}\neq \nf{null}$ and $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, \nf{DSearch($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
+\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
+
+\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
+If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
 \end{lemma}
@@
 By Corollary~\ref{lem::sumX} we know that \nf{sum\sub{enq}} field in a block shows the prefix sum of the number of \nf{Enqueue} operations in that block.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b$].sum\sub{enq}} \geq e$ and $\nf{root.blocks[$b-1$].sum\sub{enq}} < e$.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
@@
 
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetEnqueue($b$,~$i$)} returns $E_i(n,b)$.
+For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
 From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
@@
 Otherwise, the enqueue we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the \nf{Enqueue}s from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
+Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
 So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
@@
 
-  \caption{The number and order of the enqueue operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
+  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
 Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
@@
 \end{figure}
-
-\clearpage
 
@@
 \begin{proof}
-From Observation \ref{nonDecreasingHead} we know the \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
+From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
 Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
@@
 \begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$sb$]} then $\nf{$B$.super}\leq sb$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on $n$\nf{.parent.blocks[$sb$]}.
+  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superBlock$]} then $\nf{$B$.super}\leq superBlock$.
+\end{lemma}
+\begin{proof}
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on the superblock of B which is $n$\nf{.parent.blocks[$superBlock$]}.
 By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
@@
 The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq sb$, since \nf{$n$.parent.head} is still equal to $sb$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=sb$ happens.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superBlock$, since \nf{$n$.parent.head} is still equal to $superBlock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
+The reader may wonder when the case $b\nf{.super}=superBlock$ happens.
 This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
@@
 From the previous lemma we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and by Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead} $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$.
+To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
 The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
@@
 We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $sb$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, sb)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
+For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superBlock$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superBlock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
 By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$sb$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
+If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superBlock$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
 Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
@@
 
-  \caption{The number of \nf{Dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
 \end{figure}
@@
 
-  \caption{The number of \nf{Dequeue} operations before $E_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
-\end{figure}
-
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
+\end{figure}
+% \%TODO :Change sb to superblockIndex
+% \%TODO :Change subblock to subblockIndex in GetEnqueue proof
 \pagebreak
@@
 \begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null} \nf{Dequeue}.
+If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
 \end{definition}
@@
 
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DSearch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
+After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
+Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
 \end{proof}
@@
 \end{figure}
+\vspace{-1em}
 
@@
 
-\begin{lemma}[\nf{DSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
+\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
+If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
 \end{lemma}
@@
 If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log$ size of the queue when \nf{Enqueue} is linearized$)+\log$ size of the queue when \nf{Dequeue} is linearized$)$.
+Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log$ size of the queue when \nf{Enqueue} is linearized$)+\log$ size of the queue when \nf{Dequeue} is linearized$)$.
 Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
@@
 \begin{proof}
-If we split the \nf{DSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
+If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
 \end{proof}
@@
   \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption{Finished blocks are shown with red color and unfinished blocks are shown with green color. All the subblocks of a finished block are also finished.}
+  \caption[Blocks that can be safely garbage collected]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
+All the subblocks of a finished block are also finished.}
   \label{fig::finishedBlock}
@@
 We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
+An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
  We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
@@
 
-\subsection{Attiya--Fouren Lower Bound}
-As we talked about earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $O(\sqrt{\log \log p})$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when the number of concurrent processes is $O(\log\log p)$.
-Our amortized complexity is $O(log^2p+log q)$.
+\paragraph{Attiya--Fouren Lower Bound}
+As we talked about earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$.
+Our amortized complexity is $O(\log^2p+\log q)$.
 It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.

### 2022-11-12 thesis/thesis.tex
apply eric second half comments

@@
 \section*{Acknowledgements}
-
-First, I want to thank my supervisor Eric Ruppert. ...
-
-I would also like to thank professor Franck van Breugel. ...
-
-I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
+\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed.
+
+I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments bu also for his academic advice and motivations.
+
+Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
+
+Last but not the least, I would like to thank my parents, for providing me with unfailing support and continuous encouragement throughout my years of study.
+I wanted to thank my brother for being there for me whenever I needed any help.
 
 I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
+
+I dedicate this paper to my people, who are fighting for their freedom.
 
@@
 However their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the page 13.
+We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the Page 13.
 
@@
 We present our algorithm in pseudocode.
-Page 23 contains the description of the fields in the tree nodes and the blocks.
+Page~23 contains the description of the fields in the tree nodes and the blocks.
 Value of any uninitialized field is \nf{null}.
-Page 24 contains major routines and the rest is the auxiliary routines
+Page~24 contains major routines and the rest of this section consists of the auxiliary routines
 The abbreviations bellow are used in the pseudocode and the proof of correctness.
@@
     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in the prefix for the block}
+  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
 \end{itemize}
@@
   \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in the prefix for this block}
+  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
 \end{itemize}
@@
 \ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subBlockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
+\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
 \Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subBlockIndex, i} 
+\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
 \Else
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subBlockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
+\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
 \Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subBlockIndex, i} 
+\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
 \EndIf
@@
 \State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superBlockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
+\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
 \Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
@@
 \If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superBlockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
 \Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superBlockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superBlockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
+\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
 \EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlockIndex, i}
+\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
 \EndIf
@@
 We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization order, operations get the same results as in our queue.
+In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
 The proof is structured like this.
@@
 \begin{definition}[Ordering of a block in a node]
-Let $b$ be \nf{$n$.blocks[$i$]} and $b^\prime$ be \nf{$n$.blocks[$j$]}.
-We call $i$ the \emph{index} of block $b$.
-Block $b$ is \it{before} block $b^\prime$ in node $n$ if and only if $i<j$.
-We define \it{the prefix for} block $b$ in node $n$ to be the blocks in \tt{$n$.blocks[0..$i$]}.
+Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
 \end{definition}
@@
 \begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that its created block \nf{new} is not empty ($\nf{new.sum}\neq 0$).
+Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
 After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
@@
 For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance}.
+For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
 Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
@@
 \begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$ Block $b$ is a \emph{subblock} of block $c$ if $b$ is a direct subblock of $c$ or a subblock of a direct subblock of $c$.
-We say block $b$ is \it{propagated} to node $n$ if $b$ is in $n.blocks$ or is a subblock of a block in $n\nf{.blocks}$.
+A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
+Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
+We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
 \end{definition}
@@
 Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($b$)} that creates $B$ is invoked.
+Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
 The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
@@
 \begin{lemma} \label{lem::subblocksDistinct}
-Subblocks of any two blocks in node $n$ do not overlap.  
+Set of the subblocks of any two blocks in a node are distinct.  
 \end{lemma}
@@
 Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
+Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
 Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
@@
 By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap.
+Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
 \end{proof}
@@
 \begin{definition}[Superblock]
-  Block $b$ is \it{superblock} of block $c$ if $c$ is a direct subblock of $b$.
+  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
 \end{definition}
@@
 Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $b$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks \bf{B} by $ops(\bf{B})$.
+The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
+We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
 We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
@@
 \begin{lemma}\label{lem::noDuplicates}
-If $op$ is in $n.blocks[i]$ then there is no $j\neq i$ such that $op$ is in $n.blocks[j]$.
+For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
 \end{lemma}
@@
 We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n.blocks[i]$ and $n.blocks[j]$.
+Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
 From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, then $op$ cannot be in two leaf blocks.
+Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
 This is a contradiction.
@@
 
-\begin{corollary}\label{lem::establishedOrder}
-  If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-\end{corollary}
-\begin{proof}
-From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-\end{proof}
+% \begin{corollary}\label{lem::establishedOrder}
+%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
+% \end{corollary}
+% \begin{proof}
+% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
+% \end{proof}
 
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block]{Order of operations in the block b. Operations in the leaves are ordered in the numerical order shown in the drawing.}
+  \caption[Order of operations in a block]{Order of operations in the block B. Operations in the leaves are ordered in the numerical order shown in the drawing.}
 \end{figure}
@@
 Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves.
+In the non-root nodes, we only need to order operations of a type among themselves that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately.
 Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
@@
 
+The following observation follows from the Definition of \nf{num\sub{x}} on Page 22.
 \begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $in$, we have
+For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
 $$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where \nf{x} in $\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
@@
 \begin{lemma}\label{lem::numX}
-Let $B$, $B^\prime$ be $n\nf{.blocks[}b\nf{]}$, $n\nf{.blocks[}b-1\nf{]}$ respectively.
-
-If $n$ is an internal node, then
+Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
+
+\hspace{-\parindent}If $n$ is an internal node, then
 \begin{enumerate}[label=(\arabic*)]
@@
  We prove the claim by induction on the height of node $n$.
-Base case (3) for leaves is trivial, and (1) and (2) are vacuously true for leaves.
-Supposing the claim is true for $n$'s children, we prove the correctness of the claim for $n$.
+For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
+Supposing the claim is true for $n$'s children, we prove the claim for $n$.
 \begin{align*}
-    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    =& B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    =& n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    =& \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    =& \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
+    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
+    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
+    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
+    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
+    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
 \end{align*}
@@
 \begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$
+    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
+    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
+    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
 \end{enumerate}
@@
 \begin{align*}
-    B\nf{.sum\sub{enq-left}}=&n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    =& \Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\vdots\\
+    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
+    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
+    &\hspace{1em}\vdots\\
     &+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    =&\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
+    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
 \end{align*}
@@
 We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable \nf{v} immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable \nf{v} at time $t$.
+In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
 \end{definition}
@@
 Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ both cannot read the same value $i$.
+By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
 By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
@@
 So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh}, \ref{lem::headInc} on $R_3$ the claim holds.
+From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
 \end{proof}
@@
 If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is satisfied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($b$)} on a leaf finishes, the operation contained in $b$ will be established in \nf{root}.
+Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
+\end{proof}
+
+Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
 
 \begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}b\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
+For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
 \end{corollary}
 \begin{proof}
-$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the leaf's parent where it appended $b$.
-For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
+$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
+For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
 \end{proof}
@@
 \begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($b$)} finishes, $b$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
+ After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
 \end{corollary}
 \begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing $b$.
+By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
 \end{proof}
@@
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that \nf{sum\sub{enq}} field in a block shows the prefix sum of the number of \nf{Enqueue} operations in that block.
+By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the prefix blocks of that block in the node.
 Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
@@
 \begin{proof}
-We will prove this lemma by induction on the node's height $n$.
+We will prove this lemma by induction on the node $n$'s height.
 For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and contains and \nf{Enqueue} operation.
+Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
 Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
@@
 Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops} to find an operation in the subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
+By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
 \begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
+&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
+&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
  \end{align*}
@@
 
-Otherwise, the enqueue we are looking for is in the right child.
+Otherwise, the \nf{Enqueue} we are looking for is in the right child.
 Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
@@
 So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-Let $subblock$ be the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue($subblock$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
+$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
+Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
 \end{proof}
@@
 Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left subblocks (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
+\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
   }\label{figGet}
@@
 \begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superBlock$]} then $\nf{$B$.super}\leq superBlock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on the superblock of B which is $n$\nf{.parent.blocks[$superBlock$]}.
+  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
+\end{lemma}
+\begin{proof}
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on the superblock of B which is $n$\nf{.parent.blocks[$superblock$]}.
 By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
@@
 The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superBlock$, since \nf{$n$.parent.head} is still equal to $superBlock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superBlock$ happens.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
+The reader may wonder when the case $b\nf{.super}=superblock$ happens.
 This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem::secondRefreshSuper}
 Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
@@
 \begin{proof}
-Let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
+Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
 
 Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most $b$.
+From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
 
@@
 Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
+$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
 So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
@@
 
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line \ref{helpAdvance}.
+Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
 \end{proof}
@@
 Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From the previous lemma we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
+From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
 To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
@@
 We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superBlock$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superBlock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
+For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
+After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
 By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superBlock$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
+If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
 Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
@@
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.}\label{fig::blockSumDeqLeft}
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
+The index of the superblock is shown with $sb$.
+}\label{fig::blockSumDeqLeft}
 \end{figure}
@@
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.}\label{fig::blockSumDeqRight}
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
+The index of the superblock is shown with $sb$.
+}\label{fig::blockSumDeqRight}
 \end{figure}
@@
 % \%TODO :Change subblock to subblockIndex in GetEnqueue proof
-\pagebreak
+
 \subsection{Linearizability}
@@
 ${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a \nf{null} dequeue then ${Resp(d)=\nf{null}}$.
+Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
 \end{definition}
@@
 \begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue if the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
+  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
 \end{lemma}
@@
 \begin{proof}
-From Corollary \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
+Immediately from Corollary \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
 \end{proof}
@@
 $D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null dequeue in $L$ it should return the $e$th enqueued value.
+Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
 By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null dequeues.
+The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
 So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
@@
 \begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is same as the response from the sequential execution of $L$.  
+\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is same as its response in the sequential execution of $L$.  
 \end{proof}
@@
 After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-So, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
+We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
+Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
+So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
+
+Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
 See Figure \ref{fig::doubling}.
@@
 If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log$ size of the queue when \nf{Enqueue} is linearized$)+\log$ size of the queue when \nf{Dequeue} is linearized$)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ since there are $\leq p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
+Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
+Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
 \end{proof}
@@
 We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We  implemented two queries to know information about the ordering agreed in the block tree.
+We implemented two queries to compute information about the ordering agreed in the block tree.
 Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
@@
 We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all of its enqueues have been dequeued, and the responses of all of its dequeues have been computed.
+We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
 We call these blocks \it{finished blocks}.
@@
 We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$, where \nf{split(i)} removes all the indices less than \nf{i}.
-If each process tries to do the garbage collection every $p^2$ operation on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
+We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
+If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
+We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
 Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
@@
 \paragraph{Attiya--Fouren Lower Bound}
-As we talked about earlier in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$.
+As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$.
 Our amortized complexity is $O(\log^2p+\log q)$.

### 2022-11-13 thesis/thesis.tex
apply more comments

@@
 In this section, we look at previous lock-free queues.
-In the worst-case, step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously.
+The worst-case step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
@@
 Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their step complexity is $\Omega(p)$ because of the helping mechanism.
+Their worst-case step complexity is $\Omega(p)$ because of the helping mechanism.
 
@@
 
-\begin{lemma}[Amortized Time Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
+\begin{lemma}[Amortized Worst-case Analysis]
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in the worst-case, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
 \end{lemma}
@@
 \end{proof}
+
+\begin{corollary}
+    If the number of concurrent processes in an execution is less than $c$, then each \nf{GetEnqueue} takes $O(\log c)$ steps because each block has at most $c$ operations.
+So the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$.
+Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
+\end{corollary}
 
@@
 As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$.
-Our amortized complexity is $O(\log^2p+\log q)$.
+Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
 It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.

### 2022-11-14 thesis/thesis.tex
apply more comments 2

@@
 \vspace{3em}
-\textcopyright Hossein Naderibeni, 2022
+\textcopyright \; Hossein Naderibeni, 2022
 
@@
 \section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments bu also for his academic advice and motivations.
+\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
+Without his guidance and support, this thesis probably would not have been completed.
+He helped me in every step of this thesis, taught me the shared data structures, helped me find the problem, shape up the algorithm, and write it with its complications because of lots of new notions in it.
+Thank you again for always pushing to make it better.
+
+I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
 
@@
 
-Last but not the least, I would like to thank my parents, for providing me with unfailing support and continuous encouragement throughout my years of study.
-I wanted to thank my brother for being there for me whenever I needed any help.
+Last but not the least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
+I want to thank my brother for being there whenever I needed help.
 
@@
 
-I dedicate this paper to my people, who are fighting for their freedom.
+I dedicate this paper to all the people who are fighting for their freedom.
 
@@
 For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same.
-In this run, $p$'s update is overwritten.
+Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
+In this run, $P$'s update is overwritten, which makes queue inconsistent.
 One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
@@
 
-Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
+Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
 Their work is based on a circular queue using bounded memory.
@@
 
-\pagebreak
-\newpage
-
 \subsection{Pseudocode} \label{algQ}
 We present our algorithm in pseudocode.
-Page~23 contains the description of the fields in the tree nodes and the blocks.
+Page~22 contains the description of the fields in the tree nodes and the blocks.
 Value of any uninitialized field is \nf{null}.
-Page~24 contains major routines and the rest of this section consists of the auxiliary routines
+Page~23 contains major routines and the rest of this section consists of the auxiliary routines
 The abbreviations bellow are used in the pseudocode and the proof of correctness.
@@
 Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{blockSize} that there is at most one operation from each process in a given block.
+We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
 \begin{definition} [Ordering of operations inside the nodes] \label{ordering}
@@
 \begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} on internal node $n$ by process $p$.
+  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on internal node $n$.
 If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
@@
 
-\begin{lemma}\label{blockSize}
-Each block contains at most one operation of each process and, therefore, at most $p$ operations in total.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $B$ in node $n$.
+\begin{lemma}\label{oneOpinBlock}
+Each block contains at most one operation of each process.
+\end{lemma}
+\begin{proof}
+To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
 Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root.
+Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
+By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
 Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
@@
 
+\begin{corollary}\label{blockSize}
+    Each block contains at most $c$ operations in total, where $c$ is the number of concurrent processes ($c\leq p$).
+\end{corollary}
+
 \begin{lemma}\label{subBlocksBound}
-Each block has at most $p$ direct subblocks.
-\end{lemma}
-\begin{proof}
-From the definition of operations in a block in Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-The claim follows from the equivalence of the number of operations in a block and the number of operations in the subblocks of the block and the observation that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
+Each block has at most $c$ direct subblocks ($c\leq p$).
+\end{lemma}
+\begin{proof}
+From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
+We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
 Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
+By Corollary \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
 \end{proof}
@@
 By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the prefix blocks of that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
+Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
 Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
 Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blocks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
+Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
 By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
@@
 \end{figure}
-% \%TODO :Change sb to superblockIndex
-% \%TODO :Change subblock to subblockIndex in GetEnqueue proof
 
@@
 \begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
+  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
 \end{lemma}
@@
 \begin{proof}
-Immediately from Corollary \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
+The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
 \end{proof}
@@
 If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
+Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
 Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
@@
 
-\begin{corollary}
-    If the number of concurrent processes in an execution is less than $c$, then each \nf{GetEnqueue} takes $O(\log c)$ steps because each block has at most $c$ operations.
-So the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$.
+\begin{observation}
+    If the number of concurrent processes in an execution is $c$, then the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$.
 Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
-\end{corollary}
+\end{observation}
+\begin{proof}
+    The analysis is similar ro two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{GetEnqueue} takes $O(\log c)$ steps.
+\end{proof}
 

### 2022-11-15 thesis/thesis.tex
finalize thesis

@@
 Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me the shared data structures, helped me find the problem, shape up the algorithm, and write it with its complications because of lots of new notions in it.
+He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
 Thank you again for always pushing to make it better.
@@
 
-Last but not the least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there whenever I needed help.
+Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
+I want to thank my brother for being there for me whenever I needed help.
 
@@
 
-I dedicate this paper to all the people who are fighting for their freedom.
+I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
 
@@
 Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes queue inconsistent.
+In this run, $P$'s update is overwritten, which makes the queue inconsistent.
 One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
@@
 In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is ``What properties matter for the implementation of a shared data structure?''.
-Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
+The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
 To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
@@
 Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem is presented in Section 2.1.
+Previous lock-free queues and their common problem are presented in Section 2.1.
 In Section 2.2, we mention some restricted lock-free queues.
@@
 We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent exuection on the algorithm described in Section 3.3.
+Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
 
@@
 
-Chapter 5 is devoted to complexity analysis for our implementation.
+Chapter 5 is devoted to a complexity analysis for our implementation.
 We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
@@
 In this section, we look at previous lock-free queues.
-The worst-case step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the case where all $p$ processes try to do an enqueue or a dequeue simultaneously.
+The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
 Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
@@
 Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue.
+A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
 To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
@@
   \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
+  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
 In the first diagram, the first element has been dequeued.
@@
 Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the worst-case step complexity remains $\Omega(p)$ per operation.
+However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
 
@@
   \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue]{At a time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
+  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
 To order the operations in a basket, processes perform another \nf{CAS}.
@@
 As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The worst-case complexity is still $\Omega(p)$ \texttt{CAS}es.
+The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
 
@@
 They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the step (and \nf{CAS}) complexity remains $\Omega(p)$.
+A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
 Their design is lock-free but not wait-free.
@@
   \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}]{\label{fig::sundell}Global pointers point to arrays.
+    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
 Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
@@
 Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their worst-case step complexity is $\Omega(p)$ because of the helping mechanism.
+Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
 
@@
 If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the worst-case complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it as well~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
+Because of this, the amortized complexity cannot be better than $\Omega(p)$.
+
+The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
 Our motivation is to overcome this problem and present a wait-free sublinear queue.
@@
 Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of the Page 13.
+However, their algorithm cannot be used to create a queue.
+We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
 
@@
 The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-log complexity.
+Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
 
@@
 Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict with our algorithm's complexity and we manage to reach poly-log time complexity.
+The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
 
@@
 There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree})Each process has a leaf in which the operations invoked by the process are stored in order.
+The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
+Each process has a leaf in which the operations invoked by the process are stored in order.
 When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
@@
 \end{center}
-\caption[Tournament tree]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
+\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
 Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
@@
 \end{tikzpicture}}
-  \caption{Before the \nf{Refresh}}
+  \caption{Before the \nf{Refresh}.}
 \end{subfigure}
@@
 \end{tikzpicture}}
-  \caption{After the \nf{Refresh}}
+  \caption{After the \nf{Refresh}.}
 \end{subfigure}
 \hspace{6em}
-\caption[A successful \nf{Refresh}]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
+\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
 Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
@@
   \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
+  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
 \end{figure}
@@
 \end{tikzpicture}
-\caption[Showing the operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
+\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
 In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
@@
 \end{center}
-\caption[Showing the operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
+\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
 Blocks between two lines $||$ are propagated together to the parent.
@@
 So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning response for $op$.
+With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
 However, this approach is not enough for an efficient queue.
@@
 \end{small}
-\caption{An Execution on a queue separated by the operations in the root blocks}
+\caption{An execution on a queue separated by the operations in the root blocks.}
 \end{table}
@@
 We present our algorithm in pseudocode.
-Page~22 contains the description of the fields in the tree nodes and the blocks.
-Value of any uninitialized field is \nf{null}.
-Page~23 contains major routines and the rest of this section consists of the auxiliary routines
-The abbreviations bellow are used in the pseudocode and the proof of correctness.
+page~22 contains the description of the fields in the tree nodes and the blocks.
+The value of any uninitialized field is \nf{null}.
+page~23 contains major routines and the rest of this section consists of the auxiliary routines.
+The abbreviations below are used in the pseudocode and the proof of correctness.
 \begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \{enq, deq\})
+ \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
 %  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
@@
 
-\Function{Object}{Dequeue()}{} \cmt{Creates a block with \nf{null} value element, appends it to the tree and returns its response.}
+\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
+\Function{Object}{Dequeue()}{} 
 \State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
@@
 
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.}
+\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
+\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
 \If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
 \State \Return \tt{null} \label{returnNull}
-\Else
+\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
 \State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
 \State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
 \Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
 % \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<=end}}
-\State \sl{int} \tt{mid= (start+end)/2}
+\While{\nf{start<end}}
+\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
 \If{\nf{blocks[mid].sum\sub{enq}<x}}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within the \nf{b}th block in the \nf{root}.}
+\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
 
@@
 
-Here we want to demonstrate how four processes help each other to  propagate their operations up to the root and then compute their responses.
+Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
 
@@
 
-Then as in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
+Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
 $P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
@@
 
-Then as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep, and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
+Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
 After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
@@
 As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$ 's response is $ENQ_1$'s element.
+The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
 Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{.sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly , $P2$ determines that $ENQ_1$ came from the left child of $n5$.
+$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
+Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
 $P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
@@
 \begin{lemma} \label{lem::subblocksDistinct}
-Set of the subblocks of any two blocks in a node are distinct.  
+The sets of subblocks of any two blocks in a node are disjoint.  
 \end{lemma}
@@
   \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block]{Order of operations in the block B. Operations in the leaves are ordered in the numerical order shown in the drawing.}
+  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
 \end{figure}
@@
 Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately.
+In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
 Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
@@
 
-The following observation follows from the Definition of \nf{num\sub{x}} on Page 22.
+The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
 \begin{observation}\label{sumToNum}
 For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where \nf{x} in $\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
+$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
 \end{observation}
@@
     B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
+    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
+    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
     &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
@@
 \end{align*}
-First line follows from the Definition of \nf{num\sub{enq}}.
-Second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-\text{Observation } \ref{sumToNum} results the third line and the last line holds because of the induction hypothesis (3).
+The first line follows from the Definition of \nf{num\sub{enq}}.
+The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
+Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
 (2) is similar to (1).
-Now we prove (3) starting from the Definition of $E(n,b)$.
+Now we prove (3) starting from the definition of $E(n,b)$.
 \begin{align*} 
@@
     &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{1em}\vdots\\
-    &+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
+    &\hspace{2em}\vdots\\
+    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
     &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
@@
 \begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on internal node $n$.
+  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
 If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
@@
   \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
+  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
 \end{figure}
@@
 
-\begin{corollary}\label{blockSize}
-    Each block contains at most $c$ operations in total, where $c$ is the number of concurrent processes ($c\leq p$).
-\end{corollary}
+\begin{lemma}\label{blockSize}
+    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
+\end{lemma}
+\begin{proof}
+There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
+From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
+\end{proof}
 
 \begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks ($c\leq p$).
+Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
 \end{lemma}
@@
 Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Corollary \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
+By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
 \end{proof}
@@
 Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the prefix blocks of that block in the node.
+By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
 Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
@@
 
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
+  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
 Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
@@
 \begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS} on the superblock of B which is $n$\nf{.parent.blocks[$superblock$]}.
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
 By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
@@
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
 The index of the superblock is shown with $sb$.
@@
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
+  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
 The index of the superblock is shown with $sb$.
@@
 \begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is same as its response in the sequential execution of $L$.  
+\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
 \end{proof}
@@
 \begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in the worst-case, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
 \end{lemma}
@@
 \begin{observation}
-    If the number of concurrent processes in an execution is $c$, then the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$.
-Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
+    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
+Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
 \end{observation}
 \begin{proof}
-    The analysis is similar ro two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{GetEnqueue} takes $O(\log c)$ steps.
+    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
 \end{proof}
@@
   \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array Segments}
+  \caption{Array segments.}
   \label{fig::doublingArray}
@@
   \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
+  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
 All the subblocks of a finished block are also finished.}
@@
 \paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $O(\log\log p)$.
+As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
 Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower bound.
+It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
 

### 2022-11-15 thesis/thesis.tex
finalize

@@
-\documentclass[11pt]{article}
+\documentclass[11pt,letterpaper]{article}
 \renewcommand{\baselinestretch}{1.8}
@@
 \usepackage{bold-extra}
+\renewcommand*\contentsname{Table of Contents}
 
@@
 \begin{center}
-
+\pagenumbering{roman}
 \huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
@@
 \vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
+Graduate Program in Computer Science,\\
+York University\\
 Toronto, Ontario
@@
 \normalsize
+\addcontentsline{toc}{subsection}{Abstract}
 \section*{Abstract}
@@
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
+
 \pagebreak
 
+\addcontentsline{toc}{subsection}{Acknowledgements}
 \section*{Acknowledgements}
@@
 \pagebreak
+\addcontentsline{toc}{subsection}{Table of Contents}
 \tableofcontents
@@
 \pagebreak
+\addcontentsline{toc}{subsection}{List of Tables}
+\listoftables
+
+\pagebreak
+\addcontentsline{toc}{subsection}{List of Figures}
 \listoffigures
-
-\pagebreak
-\listoftables
 \newpage
@@
 \paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
+As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$ the number of concurrent processes, is $O(\log\log p)$.
 Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
@@
 
+\addcontentsline{toc}{section}{References}
 \bibliography{main.bib} 

### 2022-11-25 garbage_collect/main.tex
add garbage_collect+podc_papaer

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
+\documentclass{article}
+\usepackage[utf8]{inputenc}
 \renewcommand{\baselinestretch}{1.8}
-
+\usepackage[dvipsnames]{xcolor}
 \usepackage[letterpaper, margin=1in]{geometry}
@@
 \usepackage{subcaption}
-%\usepackage{showkeys}
 \usepackage{hyperref}
@@
 \renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
+\newcommand{\setalglineno}[1]{
   \setcounter{ALC@line}{\numexpr#1-1}}
@@
 
-\begin{center}
-
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\section*{Abstract}
-
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
-\pagebreak
-
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\tableofcontents
-
-\pagebreak
-\listoffigures
-
-\pagebreak
-\listoftables
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
-\section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
-
-\pagebreak
-\section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
-
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
-
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
-
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
-
-\subsection{Restricted Queues}
-
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
-
-% TODO: page
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
+\section{Feilds}
 \begin{algorithm}
@@
 \item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
+\item {\color{red}\tt{\sl{MaxbyProcess} lastDequeuedFrom} \textsf{Index of the most recent block in the root that has been deqeueued from.}}
 \end{itemize}
@@
 \item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
+\item {\color{red}\tt{\sl{int} garbageCollectRound}}
 \end{itemize}
@@
 \item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
+\item {\color{red}\tt{\sl{PBRT} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}}
 \item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
@@
   \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
+
+  \item {\color{red}\tt{\sl{object} response}}
 \end{itemize}
@@
 
+\section{Queue}
 \begin{algorithm}
@@
 \If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
+\State \tt{lastDequeudFrom.update(b)}
 \State \Return \tt{null} \label{returnNull}
@@
 \State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
+\State {\color{red}\tt{<x, y>= root.\Call{DoublingSearch}{e, b}}}
+\State {\color{red}\tt{lastDequeudFrom.update(x)}}
+\State \Return \tt{root.GetEnqueue(x,y)}\label{findAnswer}
 \EndIf
@@
 
+\section{Search+Append}
 \begin{algorithm}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+\Statex $\triangleright$ \textmd{\color{red} Update needed: search on RBT}.
 \Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
@@
 \Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
+{\color{red}\If{\tt{root.head\%p\textsuperscript{2}<p and garbageCollectRound<floor(root.head/p\textsuperscript{2})}}
+\State \Call{Help()}{}
+\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
+\State \tt{garbageCollectRound=floor(root.head/p\textsuperscript{2})}
+\EndIf}
 \State \tt{blocks[\head]= B} \label{appendLeaf}
@@
 
+\section{Propagate}
 \begin{algorithm}
@@
 
+\section{Index+Get}
+
 \begin{algorithm}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{\color{red} Update needed: return null when superblock in the root was not found.}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 
-\newpage
-\clearpage
+\color{red}
+\section{MaxByProcess}
 
-\subsection{Example Execution}
+\begin{algorithm}
+\caption{\tt{\sl{\color{red}MaxByProcess}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{121}
 
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
+\State \tt{\sl{int[p]} lastDequeuedbyProcess}
+\Statex
 
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
+\Function{int}{Get}{}
+\State \Return \tt{max(lastDequeuedbyProcess)}
+\EndFunction{Get}
 
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
+\Statex
+\Function{}{Update}{\sl{int} \tt{b}}
+\If{\tt{lastDequeuedbyProcess[pid]<b}}
+\State \tt{lastDequeuedbyProcess[pid]=b}
+\EndIf
+\EndFunction{Update}
 
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
+\end{algorithmic}
+\end{algorithm}
 
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
+\section{Help}
 
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
+\begin{algorithm}
+\caption{\tt{\sl{Tree}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{130}
 
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
+\Function{int}{Help}{}
+\State \bf{for each} \sl{process} \tt{P}
+\State \tt{h=P.leaf.head}
+\If{\tt{P.leaf.blocks[h].num\sub{deq}==1 and P.leaf.IndexDequeue(h,1)!=null}}
+\State \tt{<b, i>=} \Call{IndexDequeue}{h, 1}
+\State \tt{output=} \Call{FindResponse}{b, i} 
+\State \tt{P.leaf.blocks[h].response= output}
+\EndIf
+\State \bf{end for}
+\EndFunction{Help}
 
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
+\end{algorithmic}
+\end{algorithm}
 
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
+\section{FreeMemory}
 
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{140}
 
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
+\Function{}{FreeMemory}{\sl{int} \tt{b}}
+\If{not leaf}
+\State \tt{left.FreeMemory(blocks[i].end\sub{left})}
+\State \tt{right.FreeMemory(blocks[i].end\sub{right})}
+\EndIf
+\State \tt{\bf{while} !blocks.CAS(blocks.splitGreater(i))}
+\EndFunction{FreeMemory}
 
-\clearpage
+\end{algorithmic}
+\end{algorithm}
 
-\section{Proof of Correctness}
+\section{PBRT}
 
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
+\begin{algorithm}
+\caption{\tt{\sl{PBRT}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{140}
 
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
+\Statex \tt{\sl{PRBT} prbt}
+\Statex \sf{nodes store \tt{<key, sum\sub{enq}-> block}}
+\Statex \tt{[i] -> GetByBlock(i)}
 
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
+\Function{}{GetByBlock}{\sl{int} \tt{i}}
+\State \Return \tt{rbt.get(i)}
+\If{not found}
+\State \tt{return written response}
+\EndIf
+\EndFunction{FreeMemory}
 
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\pagebreak
-
-\pagebreak
-
-\bibliography{main.bib} 
-\bibliographystyle{acm}
+\end{algorithmic}
+\end{algorithm}
 

### 2022-11-25 old/highlevel/highlevel.tex
add garbage_collect+podc_papaer

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{fontspec}
\usepackage{xcolor}
\defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
\setmainfont{Open Sans}
\setmonofont{Monaco}[Color=gray]
\title{Intoducing the Best Shared Queue Ever}
%\author{Hossein Naderi}
\date{\today}
\begin{document}
\maketitle

\section{Description of the algorithm}

%In the following sections, we discuss how the abstract version algorithm works and explain its requirements for implementation. The general idea is to have processes agree on the linearization and then respond fast from the history.

\subsection{\texttt{Queue}}
\subsubsection{Definition}
\texttt{Queue} is an ordered set of elements.\subsubsection{Methods}
\paragraph{\texttt{Enqueue(element v)}}
Appends \texttt{v} after the last appended \texttt{element} to the queue.
\paragraph{\texttt{Dequeue()}}
Removes the first not dequeued \texttt{element} in the queue and returns it.
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/queue}
  \caption{\texttt{Queue}}
\end{figure}
\subsubsection{Implementation}
 We use a \texttt{Block tree} and a \texttt{Queue helper} implement a \texttt{Queue}. To do an enqueue, it is sufficient to append an enqueue operation to the block tree. For dequeues, we do the same but at last, compute the head using \texttt{Queue helper}, and if the queue is not empty, we return the corresponding enqueue argument.
\subsubsection{Complexity} \texttt{Enqueue(v)} and \texttt{Dequeue()} each take O(log Q+log$^{2}$p) steps.

\subsection{\texttt{‌Block tree}}
\subsubsection{Definition}
\texttt{Block tree} is a data structure that linearizes invocations of \texttt{operation}s by some procesess. \paragraph{Linearization} Operations are linearized when they are added to the root. The operations in a root block are linearized as the block's ordering.
\texttt{TODO}{problem: where should we define the linearization ordering?}
\subsubsection{Methods}
\paragraph{\texttt{Append(operation op)}}
Appends op to the linearization.
\paragraph{\texttt{Get(int i)}}
Returns ith linearized operation.
\paragraph{\texttt{Index(operation op)}}
If Append(op) is terminated, Index(op) returns the order of op in the linearization.
\paragraph{\texttt{Mark(int i)}} If operation i is marked Get(i) is not going to be invoked in future anymore.
\subsubsection{Implementation}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/tree}
  \caption{\texttt{Block tree}}
\end{figure}
\texttt{‌Block tree} is a binary tree of \texttt{node}s, such that a leaf is assigned to each process. Process \texttt{p} adds \texttt{op} to its leaf and propagates it to the \texttt{root node}, which stores the total ordering.
\paragraph{\texttt{Append(op)}}
Adds a block for op to the invoking process's leaf.
\paragraph{\texttt{Propagate(node n)}}
When \texttt{Propagate(node n)} is terminated, operations in n before calling \texttt{Propagate(node n)} are in the root.
\begin{enumerate}
  \item Refresh(n)
  \item If 1 was unsuccessful, do Refresh(n) one more time.
  \item Propagate(n.parent)
\end{enumerate}
\paragraph{\texttt{Get(i)}}
\begin{enumerate}
  \item B= find the root block of containing $i$th op
  \item Get(B, i-pre(B), root)
\end{enumerate}
\texttt{pre(B)} is the number of operations before block B in B's node.
\paragraph{\texttt{Index(op)}}
\begin{enumerate}
  \item B= the block in a leaf containing operation op
  \item Index(n,B,i+pre(B))
\end{enumerate}
\paragraph{\texttt{Mark(i)}} Mark i in the root node.
\subsubsection{Complexity}
\texttt{Append(op)}, \texttt{Get(i)} and \texttt{Index(op)} each take O(log$^{2}$p),
O(log (n-\#marked)+log$^{2}$p),
O(log$^{2}$p) steps.

\subsection{\texttt{Queue helper}}
\subsubsection{Definition}
Queue helper tells us the head of the queue at the time some dequeue operation has been linearized.
\subsubsection{Methods}
\paragraph{\texttt{ComputeHead(i)}} Computes head of the queue when ith dequeue occurs.
\paragraph{\texttt{Augment(i)}} Augments root block i.
\subsubsection{Implementation}
By adding the size and number of non-null dequeues in a block, we can compute the head in constant time.
\subsubsection{Complexity}
Constant time.

\subsection{\texttt{Node}}
\subsubsection{Definition}
A \texttt{node} contains the ordering of the blocks appended to it by time.
\subsubsection{Methods}
\paragraph{\texttt{Append(Block b)}}
Adds \texttt{block b} to the end of the \texttt{node}'s ordering.
\paragraph{\texttt{GetHead()}} Returns the index of the last block added.
\paragraph{\texttt{Refresh(Node n)}}
Creates a block from \texttt{n}'s children's blocks which are not already in \texttt{n}, and tries to append it to \texttt{n}.
\paragraph{\texttt{Get(B,i,n)}} Returns the $i$th operation in block B in node n.
\paragraph{\texttt{Index(n,B,i)}} Returns the superblock in n.parent containing block B in node n.
\subsubsection{Implementation}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/node}
  \caption{\texttt{Node}}
\end{figure}
\texttt{blocks[i]} stores the ith block. \texttt{time} shows the number of groups propagated to the parent. \texttt{super[i]} is the index of the superblock of the group i.
\paragraph{\texttt{Append(Block b)}} \texttt{CAS}es b to the first empty cell in A.
\paragraph{\texttt{GetHead()}} Returns the head field, it is updated every successful Refresh().
\paragraph{\texttt{Get(B,i,n)}}
\begin{enumerate}
  \item $SB$= find the subblock of B containing $i$th op
  \item Get(SB, i-pre(SB), B's node)
\end{enumerate}
\paragraph{\texttt{Index(n,B,i)}}
\begin{enumerate}
  \item $SB$= find the block in n's parent that contains ith op in block B of node n
  \item Index(n,B,i+pre(SB))
\end{enumerate}
\paragraph{\texttt{Refresh()}}
\begin{enumerate}
  \item Create a block from $n$'s children's blocks which are not already in $n$
  \item Append the block to $n$
  \item If 2 was successful, update the head field
\end{enumerate}
\subsubsection{Requirements} If Append() fails it means another successful instance of Append exists that has some time in common.
\subsubsection{Complexity} Each level of Get() takes $\log p$ steps, and there are $\log p$ levels. So Get may take O(log$^2$ p) steps. Complexity of finding the superblock of a block is O(log p).

\subsection{\texttt{Root node}}
A data structure that contains an ordering of not-marked blocks appended to it.
\paragraph{\texttt{Append(Block b)}}
Adds \texttt{block b} to the end of the \texttt{Root node}'s ordering.
\paragraph{\texttt{Mark(Block b)}}
Marks block b.
\paragraph{\texttt{Get(i)}}
Returns the block containing ith operation.

\subsection{\texttt{Block}}
\subsubsection{Definition}
A block is an ordering of some operations in a node. We can merge some blocks into a block, the initial blocks are called subblocks, and the latter is called superblock.

\texttt{TODO} {problem: where should we put end pointer of blocks that help get become faster?}
\subsubsection{Methods}
\paragraph{\texttt{CreateBlock(operation op)}}
Returns a \texttt{block} that contains \texttt{op}.
\paragraph{\texttt{CreateBlock(node n, int start, int last)}}
Returns a \texttt{block} that contains the blocks from n.start to n.last. The order of the input blocks remains the same.
\paragraph{\texttt{GetEnd()}}
Returns the last left and right subblock of the block.
\subsubsection{Implementation}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/block}
  \caption{\texttt{Block}}
\end{figure}
We can do the told before methods on blocks only by knowing
\begin{multicols}{2}
\begin{enumerate}
\item \#enqueue operation from left
\item \#enqueue operation from right
\item \#dequeue operation from left
\item \#dequeue operation from right
\item index of the last block from left
\item index of the last block from left
\item group number which tells the blocks which have propagated to the parent together
\end{enumerate}
\end{multicols}
\texttt{TODO} {problem: how to explain the separation of the enqueues from the dequeues? till now, we considered all operations from just one type}

\subsection{\texttt{Operation}}
\subsubsection{Implementation}
Each operation contains the arguments of the operation, the invoking process, and the position in the invoking's process operations.

\end{document}

### 2022-11-25 old/jayanti/maybe-final-edit.tex
add garbage_collect+podc_papaer

--- 2022-02-03 jayanti/describing-alogorithm.tex
@@
 \rhead{\rule[-1ex]{0pt}{3.5ex} Hossein Naderi }
-\newcommand\ANDREWID{Trying to explain the algorithm in simple language} 
+\newcommand\ANDREWID{Distributed Queue using $\Theta(\log n)$ Shared Memory Accesses} 
 \newcommand\HWNUM{1}
@@
 
-\question{Previous Work}
-
-\begin{table}[hbt]
-  \begin{center}
-  \begin{tabular}{l|c|c|c|c|c}
-    & Type & Progress Property & Conditions & Amortized Time Complexity & Space \\
-    \hline
-    & & & & & \\
-  \end{tabular}
-  \caption{}
-  \end{center}
-\end{table}
-
-\question{Goal}
-The model consists of $p$ processes. And the problem is to implement linearizable shared queue Q supporting $<ENQ, DEQ>$ operations between them.
-
-%
-%\begin{figure}[h]
-%\begin{center}
-%\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
-%
-%\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
-%%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
-%
-%%Curve Lines [id:da24404451870478872] 
-%\draw    (102,43) .. controls (143.58,47.95) and (138.12,104.85) .. (199.13,107.92) ;
-%\draw [shift={(201,108)}, rotate = 181.82] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%%Curve Lines [id:da9272135872377072] 
-%\draw    (102,121) .. controls (141.6,91.3) and (162.58,119.43) .. (224.12,117.08) ;
-%\draw [shift={(226,117)}, rotate = 537.27] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%%Curve Lines [id:da5453903537732101] 
-%\draw    (101,182) .. controls (158.42,196.85) and (149.19,136.23) .. (203.34,122.4) ;
-%\draw [shift={(205,122)}, rotate = 526.9300000000001] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%%Shape: Grid [id:dp5927814714859243] 
-%\draw  [draw opacity=0] (238,105) -- (350,105) -- (350,126) -- (238,126) -- cycle ; \draw   (238,105) -- (238,126)(258,105) -- (258,126)(278,105) -- (278,126)(298,105) -- (298,126)(318,105) -- (318,126)(338,105) -- (338,126) ; \draw   (238,105) -- (350,105)(238,125) -- (350,125) ; \draw    ;
-%%Curve Lines [id:da33961815970017184] 
-%\draw    (359,115) .. controls (398.6,85.3) and (399,117.35) .. (457.22,115.08) ;
-%\draw [shift={(459,115)}, rotate = 537.14] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%%Curve Lines [id:da9041409242658852] 
-%\draw    (359,115) .. controls (400.58,104.11) and (400.02,164.77) .. (458.22,160.16) ;
-%\draw [shift={(460,160)}, rotate = 534.29] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%%Curve Lines [id:da2855643108836211] 
-%\draw    (359,115) .. controls (379.79,51.64) and (412.34,55.91) .. (458.59,71.52) ;
-%\draw [shift={(460,72)}, rotate = 198.8] [color={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.75]    (10.93,-3.29) ..
-controls (6.95,-1.4) and (3.31,-0.3) ..
-(0,0) ..
-controls (3.31,0.3) and (6.95,1.4) ..
-(10.93,3.29)   ;
-%
-%\end{tikzpicture}
-%\caption{Multiple queries on the shared Queue}
-%\end{center}
-%\end{figure}
-
-%We show operation $o$ requested by process $p$ by $o_p$ and the shared queue by $Q$ and local queues by $q$.
-
-If $op$ finishes before $op^\prime$ starts, it has to take effect on the $Q$ before another.
-For concurrent operations from processes on $Q$, the implementation can decide which to put before the other.
-So we want to design an algorithm that gives processes responses and does their operation on the $Q$.
-Since the linearization is a complete ordering among given operations, our problem is to agree on a linearization of the operations.
-
-\begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}
-\Tree [.$n_0$ [.$n_1$ [.$n_3$ $p_1$ $p_2$ ] [.$n_4$ $p_3$ $p_4$ ] ]
-          [.$n_2$ [.$n_5$ $p_5$ $p_6$ ] [.$n_6$ $p_7$ $p_8$ ] ] ]
-\end{tikzpicture}
-\caption{The total ordering of all requests operation propagated up to the root is stored in the root.
-A request it done if it has reached to the root and has been added to the history.
-In each node we store which requests has been propagated up to it.}
-\end{center}
-\end{figure}
-
-A usual way \cite{JayantiP05} to agree on something is to use tournament trees.
-Consider a tournament tree where each process is assigned to one of its leaves.
-Each process adds an operation to its leaf and tries to propagate it up to the root.
-The first one that reaches the root is the first operation.
-Thus the tournament tree is like a competition between processes to determine the total ordering.
-Propagating an operation from a node means trying to move it one level higher than its parent.
-Sometimes, two operations from two nodes are propagated up to the parent concurrently; we have to choose one to be before the other arbitrarily.
-
-Now two problems arise:
-\begin{itemize}
-  \item Since the tournament tree is shared among all the processes, then at some point, there may be multiple processes trying to propagate their operations at the same node.
-So we need a lock-free procedure for propagating operations from children of a node to it.
-  \item One way to implement a shared object with the tournament tree is to store in each node the ordering of operations propagated up to that node.
-Merging of the two children of the root gives us the total ordering at some point.
-But keeping all these data is not memory-efficient.
-For example, all dequeues straight after a queue that gets empty can return null without knowing the whole history before.
-\end{itemize} 
-\pagebreak
-\question{First Try}
-
-In the first attempt, we store each subtree ordering of operation in its root.
-At each propagation step, we append children's new operations to the parent ordering.
-The merging step is heavy, and this way is not memory-efficient.
-Each ordering merge step is $O(p)$, and there will be $\log(p)$ propagate steps, and it will take $O(\#operations)$ to compute the operation response.
-
-\begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}
-
-\Tree [.{\begin{tabular}{l|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ & ...
-\\ \hline\end{tabular}}  [.{\begin{tabular}{l|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}}
-      {\begin{tabular}{l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline\end{tabular}} ] [.{\begin{tabular}{l|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-
-\end{tikzpicture}
-\caption{First Try: We show operation $i$ from process $j$ with $op_j^i$. In each node, we store the ordering of all the operations propagated up to it.}
-\end{center}
-\end{figure}
-
-\begin{algorithm}[h]
-\caption{First Try Algorithm}
-\begin{algorithmic}[1]
-
-\Procedure{Do}{node n, operation op}
-\State n.leaf.append(op)
-\State \Call{Propagate}{n}
-\State \Return \Call{Compute}{op}
-\EndProcedure
-\Statex
-\Procedure{Propagate}{node n}
-\If{r$\neq$root}
-\State \Call{MergeChilrenOrderingInto}{n.parent}
-\State \Call{Propagate}{r.parent}
-\EndIf
-\EndProcedure
-\Statex
-\Procedure{MergeChilrenOrderingInto}{node n}
-\State new=n.children.new-operations
-\State n.ordering.append(new)
-\EndProcedure
-
-\end{algorithmic}
-\end{algorithm}
-
-\question{Second Try}
-
-On the first try, we store the ordering of all the operations in the subtree in its root.
-But it's not necessary for concurrent operations.
-If two operations are propagated to a node simultaneously, they will be propagated up to the root together so we can store their ordering in the root.
-This improvement does not change the time complexity of merge steps.
-
-\begin{figure}[h]
-\begin{center}
-\begin{tikzpicture}
-
-\Tree [.{$\big\{op_2^1,op_2^2,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_1^1, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_2^2\big\},\big\{op_1^1\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline\end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-
-\end{tikzpicture}
-\caption{Second Try: In each internal node, we store the set of all the operations propagated together up to it, and one can arbitrarily linearize the total of concurrent operations.}
-\end{center}
-\end{figure}
-
-\pagebreak
-\question{Third Try}
-
-We call sets in the ordering from Algorithm 2 "blocks".
-Here we are proposing that if you store some constant statistics in each set, you can compute the set.
-If you know how many operations from the left child and right child are in block b, then with knowing the count of operations before block b in node n, we can find out which operation from each child of n has propagated to block b.
-This leads us to the main algorithm.
-Using this approach, we can make merge steps $O(\log p)$.
-As we said before, if we know some data like what is the size of the queue, it helps us to compute results faster.
-In this way, we propose our algorithm to compute the dequeue operation in $O(\log^2p)$.
-
-%\begin{figure}[h]
-%\begin{center}
-%\begin{tikzpicture}
-%  
-%\Tree [.{\begin{tabular}{l|c|c|c|c|c|c|c|c|c}  \hline $root$ & $\pi$ & $\chi$ & $\varphi$ & $\Delta$ & $\theta$ & $\Upsilon$ & $\xi$ & $\eta$ & ...
-\\ \hline\end{tabular}}  [.{\begin{tabular}{l|c|c|c|c}  \hline $n_1$ & $B_{11}$ & $B_{12}$ & $B_{13}$ & ...
-\\ \hline\end{tabular}}
-%      {\begin{tabular}{l|c|c|c}  \hline $p_1$ & $\varphi$ & $\Upsilon$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_2$ & $\Delta$ & $\theta$ & ...
-\\ \hline\end{tabular}} ] [.{\begin{tabular}{l|c|c|c}  \hline $n_2$ & $B_{21}$ & $B_{22}$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_3$ & $\chi$ & $\eta$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c}  \hline $p_4$ & $\pi$ & $\xi$ & ...
-\\ \hline\end{tabular}} ] ]
-%
-%\end{tikzpicture}
-%\caption{Third Try: In each node we store the summary of operation propagated together(some constant size statistics called blocks). Pros: Quick merges and low memory use.}
-%\end{center}
-%\end{figure}
-
-%If each propagation step is not heavy in our design, then there are at most $\log(p)$ steps.
-We are trying to implement steps in the nodes $\log(p)$ too.
-In the remainder, we focus on two parts of the algorithm; the first one deals with the lock-free property of the merge procedure on nodes of the tournament tree, and the second one is about why our design works fast and correctly as a shared queue.
-
-\question{Merge Step}
-In each merge step on node n, we read n.children new operations and try to append them to the n's ordering.
-Here we propose a wait-free approach using two CAS operations.
-First, we create a block of newly added operations to n's children(exist in n.children but not in n).
-After that, we try to append them to the last block; if it wasn't successful, we try again.
-If one of the $CAS$ operations was successful, then the block is propagated up to the root; otherwise, it means after the first unsuccessful CAS, another operation has come, and it reads current operations.
-
-\texttt{This new algorithm needs proof}
-
-\begin{algorithm}[h]
-\caption{Merge Step}
-\begin{algorithmic}[1]
-
-\Procedure{MergeChilrenOrderingInto}{node n}
-
-\State block=\Call{CreateBlock}{n.left,n.right}
-\State last=n.last \Comment{Index of the first empty cell of n's array of blocks.}
-\If{\Call{CAS}{n.last, last, last+1}}
-n[last]=block
-\ElsIf{\State\Call{CAS}{n.last, last+1, last+2}}
-n[last]=block
-\EndIf
-\EndProcedure
-
-\end{algorithmic}
-\end{algorithm}
-
-\question{Data structure details}
-Here we are talking about details of the information stored in blocks and the root.
-
-Tree Structure:
-\begin{itemize}
-  \item Leaf $l$ of the tournamnet tree is the list of the operations of process $p$.
-  \item Interval node $n$ stores an array of blocks($n.blocks$) and index of the first empty cell of the array($n.last$).
-  \item Root like interval nodes stores blocks with two additional data: size(size of the queue after the block), req($\#$returning deques in each block).
-\end{itemize}
-
-In each block we store:
-\begin{itemize}
-  \item Accumulative $\#$left enqs,$\#$right enqs,$\#$left deqs,$\#$right deqs
-  \item pointers to last block merged from left child and righ child
-\end{itemize}
-
-How to find the ith enq among all operations? Find the bock containing ith operation in the root using binary search. Decide the operation is in which child and continue recursively.
-
-\texttt{how to draw lists as nodes of the tree and draw edges from cells?}
+\question{Design}
+%We store an array of information blocks in each node of the tournament tree that will help us to do Get, Order queries.
+In merge steps of node n process p reads child left \& right the node n and create a block of info for the new operations in the children.
+After creating the block the process tries to CAS it to the end of the $n$'s array.
+Note that ordering is from left to right and from now on when we say $n$ we mean the value on node $n$.
 
 \begin{center}
-\begin{tikzpicture}
+\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
+	level 2/.style={level distance=2.5cm,sibling distance=0.5cm}]
 
-\Tree [.1 2 3 ]
+\Tree [.{\begin{tabular}{||c|c|c||}  \hline 0,11 & 12,6 & 8,36 \\ \hline\end{tabular}}
+ [.{\begin{tabular}{||c||c||}  \hline 10,2 & 5,3 \\  \hline \multicolumn{1}{c||}{2} & \multicolumn{1}{c}{3} \end{tabular}}
+ [.{\begin{tabular}{||c|c||c||}\hline 2,3 & 4,1 & 0,5 \\\hline\multicolumn{2}{c||}{1} & \multicolumn{1}{c}{2}\end{tabular}} ]
+  [.{\begin{tabular}{||c||c||}  \hline 0,2 & 1,2 \\  \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c}{2} \end{tabular}} ] ]
+          [.{\begin{tabular}{||c||c||c|c||}  \hline 3,8 & 6,0 & 14,6 & 0,16 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2} & \multicolumn{2}{c}{3}\end{tabular}}
+           [.{\begin{tabular}{||c||c||c|c||}  \hline 0,3 & 4,2 & 5,1 & 3,5 \\ \hline \multicolumn{1}{c||}{1} & \multicolumn{1}{c||}{2}& \multicolumn{2}{c}{3}\end{tabular}} ]
+            [.{\begin{tabular}{||c|c||c||c|c||}  \hline 2,1 & 5,0 & 4,2 & 3,4 & 6,3 \\ \hline \multicolumn{2}{c||}{1} & \multicolumn{1}{c||}{3} & \multicolumn{2}{c}{4}\end{tabular}} ] ] ]
 
@@
 
-And these coloumns in the root:\\
+\vspace{3em}
+\begin{center}
+\begin{tikzpicture}[level 1/.style={level distance=3.3cm,sibling distance=1cm},
+	level 2/.style={level distance=2.5cm,sibling distance=0.5cm}]
 
-{\begin{tabular}{c|c|c|c|c|c|c|c|c|}
+\Tree [.{\begin{tabular}{c|c|c|c|c|c|c|c|c|}
     \hline enq & 5,6 & 5,8 & 6,3 & 3,7 & 10,2 & 1,2 & 2,3 & ... \\
@@
     \hline size & 0 & 4 & 0 & 0 & 8 & 9 & 0 &... \\ \hline\end{tabular}}
+     [.{\begin{tabular}{c|c|c|c|c|c|c|}  \hline enq & 2,3 & 4,1 & 6,0 & 1,2 & ... \\ \hline deq & 1,2 & 3,0 & 1,4 & 2,5 & ... \\ \hline\end{tabular}}
+      {\begin{tabular}{c|c|c|c|c|}  \hline $p_1$ & &  & & ...
+\\ \hline\end{tabular}} {\begin{tabular}{c|c|c|c|c|}  \hline $p_2$ & & & & ...
+\\ \hline\end{tabular}} ] [.{\begin{tabular}{c|c|c|c|c|c|}  \hline enq & 2,4 & 6,2 & 2,1 & 1,6 & ...
+\\ \hline deq & 5,6 & 2,4 & 3,7 & 3,6 & ...
+\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c|c|}  \hline $p_3$ & & & & ...
+\\ \hline\end{tabular}} {\begin{tabular}{l|c|c|c|c|}  \hline $p_4$ & & & & ...
+\\ \hline\end{tabular}} ] ]
 
-\question{Complete Algorithm with the Logic related to the Queue}
+\end{tikzpicture}
+\end{center}
+
+\question{Description}
+
+In our model each process has an array which it appends its outgoing operation to it.
+Then there is a shared tournament tree that its leaves are operation lists of processes.
+Nodes of the tournament tree contain list of blocks.
+each block is a summarization of the operation propagated together.
+it has 4 numbers that tell us how many operation from which type has propagated from which child.
+In each block we order operation by this ordering(enqueus from left child, enqueues from right child, dequeues from left child, dequeues from right child) and by this rule we can find i-th operation type and the child that it has propagated up from.
+
+Propagate procedure: First we create a block of newly added operation to children of the given node, then try to apen it to the last of the block lists. If it fails we try again.
 
@@
 \Function{Create-Block}{n}
-\Statex \Comment constructs block of the new operation in children of n. if n is the root, it has extra fields: size, rdeqs.
+\Statex \Comment constructs block of the new operation in children of n. if n is root it has extra fields: size, rdeqs.
 \EndFunction
@@
 \end{algorithm}
-
-\begin{algorithm}
-\caption{Block Tree}
-\begin{algorithmic}[1]
-\begin{multicols}{2}
-
-\Statex $\blacktriangleright$ leaf $l_i$: list of operations
-\Statex $\blacktriangleright$ internal node $n$ of BT: list of blocks and index $last$
-\Statex $\blacktriangleright$ block $b$: 4 statistics of concurrent operations aggregated together to a block in a \textsc{Refresh} consisting [$left$:$ \#$left ops, $right$: $\#$right ops, $left-sum$: prefix sum left ops, $right-sum$: prefix sum right ops]
-\Statex $\blacktriangleright$ index $last$: index of last block of node $n$
-\Statex
-
-\Function{void}{Append}{operation op, pid i} 
-\State $l_i$.append($op$)
-\State \Call{Propagate}{parent of $l_i$}
-\EndFunction{Append}
-
-\Statex
-
-\Function{void}{Propagate}{node $n$}
-\If{$n$==root} \Return
-\Else
-\State $new$=\Call{CreateBlock}{n}
-\If{\Call{CAS}{$n$.last, now, now+1}}
-last block of $n$=now
-\ElsIf{\Call{CAS}{$n$.last, now, now+1}}
-last block of $n$=now
-\EndIf
-\EndIf
-\State \Call{Propagate}{parent of $n$}
-\EndFunction{Propagate}
-
-\Statex
-
-\Function{boolean}{CreateBlock}{node n}
-\State $current$= last block of $n$
-\State $start$= \Call{BSearch}{crrent.left-sum}
-\State $left$=0
-\ForEach{$block$}{ $start$: first null block of left child of $n$}
-\State $left$+=$block$.left
-\EndFor
-\State $left-sum$=current.left-sum+left
-\State do lines 14 to 19 for right
-\State \Return [left,right,left-sum,right-sum]
-\EndFunction{CreateBlock}
-
-\columnbreak
-
-\Function{element}{GetIndex}{block b, index i} 
-\State n:=b.node
-\If{i$\leq$b.left}
-\State sb=\Call{BSearch}{left child of $n$, i-b.right.sum}
-\State \Call{GetIndex}{sb, b.left-sum} 
-\Else ~sb=\Call{ BSearch}{right child of $n$, i-b.left.sum}
-\State \Call{GetIndex}{sb, b.right-sum} \EndIf
-\EndFunction{GetIndex}
-
-\Statex
-
-\Function{list}{GetElements}{block b} 
-\ForEach{block \textbf{in} \Call{GetSubBlocks}{b}}
-\State result.append(\Call{GetElements}{})
-\EndFor
-\State \Return result
-\EndFunction{GetElements}
-
-\Statex
-
-\Function{list}{GetSubBlocks}{b}
-\State n:=b.node
-\State b[-1]=b's previous block
-\ForEach{direction \{left, right\}}
-\State init=\Call{BSearch}{n.direction, b[-1].direction}
-\State end=\Call{BSearch}{n.direction, b.direction}
-\State result.append([init:end])
-\EndFor
-\State \Return result
-\EndFunction{GetSubBlocks}
-
-\end{multicols}
-\end{algorithmic}
-\end{algorithm}
-
-\bibliographystyle{plain}
-\bibliography{refrences}
+\pagebreak
 

### 2022-11-25 old/proposal/proposal.tex
add garbage_collect+podc_papaer

--- 2022-03-28 proposal/proposal.tex
@@
 
-In this work, we are going to introduce a novel lock-free queue.
-Linearizable lock-free data structures are the current standard requirements for designing distributed algorithms.
-All lock-free queues in the literature yet have a common problem in their worst case.
+In this work, we are going to introduce a novel lock-free queue implementation.
+Linearizability and lock-freedom are standard requirements for designing shared data structures.
+All existing linearizable, lock-free queues in the literature have a common problem in their worst case.
 Our algorithm solves this problem, performing better than them.
@@
 Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.
-Current approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
 
@@
 
-\begin{figure}[t]
+\begin{figure}[hbt]
   \center\includegraphics[scale=0.5]{pics/good}
@@
 
-\begin{figure}[t]
+\begin{figure}[hbt]
   \center\includegraphics[scale=0.5]{pics/bad}
@@
 
-\begin{figure}[t]
-  \center\includegraphics[scale=0.5]{pics/msqueue}
+\begin{figure}[hbt]
+  \center\includegraphics[scale=0.4]{pics/msqueue}
   \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
@@
 
-\begin{figure}[t]
+\begin{figure}[hbt]
 \begin{center}
@@
 
-\section{Next steps}
+\section{Next Steps}
 The current algorithm only works as a queue, but other data structures like stacks share similarities.

### 2022-11-25 podc_paper/podc2023sample.tex
add garbage_collect+podc_papaer

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt,letterpaper]{article}
-\renewcommand{\baselinestretch}{1.8}
-
-\usepackage[letterpaper, margin=1in]{geometry}
+%%
+%% This is file generated with the docstrip utility from the source file: 
+%%
+%% samples.dtx  (with options: `acmsmall-conf')
+%% With some modifications for PODC 2023.
+%% For the copyright see the source file.
+%%
+%%
+%% Commands for TeXCount
+%TC:macro \cite [option:text,text]
+%TC:macro \citep [option:text,text]
+%TC:macro \citet [option:text,text]
+%TC:envir table 0 1
+%TC:envir table* 0 1
+%TC:envir tabular [ignore] word
+%TC:envir displaymath 0 word
+%TC:envir math 0 word
+%TC:envir comment 0 0
+%%
+%%
+%% The first command in your LaTeX source must be the \documentclass
+%% command.
+%%
+%% For submission and review of your manuscript please change the
+%% command to \documentclass[manuscript, screen, review]{acmart}.
+%%
+%% When submitting camera ready or to TAPS, please change the command
+%% to \documentclass[sigconf]{acmart} or whichever template is required
+%% for your publication.
+%%
+%%
+\documentclass[acmsmall,nonacm,anonymous]{acmart}
+
 \usepackage{tikz-qtree}
@@
 
+%%
+%% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{acmcopyright}
+\copyrightyear{2023}
+\acmYear{2023}
+\acmDOI{XXXXXXX.XXXXXXX}
+
+%% These commands are for a PROCEEDINGS abstract or paper.
+\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
+  2023}{Orlando, FL}
+%%
+%%  Uncomment \acmBooktitle if the title of the proceedings is different
+%%  from ``Proceedings of ...''!
+%%
+%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
+%%  June 03--05, 2018, Woodstock, NY}
+\acmPrice{15.00}
+\acmISBN{978-1-4503-XXXX-X/18/06}
+
+%%
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+
+%%
+%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-\begin{center}
-\pagenumbering{roman}
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science,\\
-York University\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\addcontentsline{toc}{subsection}{Abstract}
-\section*{Abstract}
-
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
+\title{A Wait-free Queue with Poly-logarithmic Worst-caseStep Complexity}
+%\title{The Name of the Title Is Hope}
+
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+\author{Eric Ruppert}
+\affiliation{%
+  \institution{York University}
+  \streetaddress{P.O. Box 1212}
+  \city{Toronto}
+  \state{Ontario}
+  \country{Canada}
+  \postcode{43017-6221}
+}
+\authornote{Both authors contributed equally to this research.}
+\email{ruppert@cse.yorku.ca}
+\orcid{1234-5678-9012}
+
+\author{Hossein Naderibeni}
+\authornotemark[1]
+\email{hnaderi268@gmail.com}
+\affiliation{%
+  \institution{York University}
+  \streetaddress{P.O. Box 1212}
+  \city{Toronto}
+  \state{Ontario}
+  \country{Canada}
+  \postcode{43017-6221}
+}
+
+%%
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+\renewcommand{\shortauthors}{Trovato et al.}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
+\begin{abstract}
 In this work, we introduce a novel linearizable wait-free queue implementation.
@@
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-
-\pagebreak
-
-\addcontentsline{toc}{subsection}{Acknowledgements}
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\addcontentsline{toc}{subsection}{Table of Contents}
-\tableofcontents
-
-\pagebreak
-\addcontentsline{toc}{subsection}{List of Tables}
-\listoftables
-
-\pagebreak
-\addcontentsline{toc}{subsection}{List of Figures}
-\listoffigures
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
+\end{abstract}
+
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+\begin{CCSXML}
+<ccs2012>
+ <concept>
+  <concept_id>10010520.10010553.10010562</concept_id>
+  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+  <concept_significance>500</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010575.10010755</concept_id>
+  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+  <concept_significance>300</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010553.10010554</concept_id>
+  <concept_desc>Computer systems organization~Robotics</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10003033.10003083.10003095</concept_id>
+  <concept_desc>Networks~Network reliability</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Computer systems organization~Embedded systems}
+\ccsdesc[300]{Computer systems organization~Redundancy}
+\ccsdesc{Computer systems organization~Robotics}
+\ccsdesc[100]{Networks~Network reliability}
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{concurrent data structures, wait-free queues}
+
+\received{20 February 2007}
+\received[revised]{12 March 2009}
+\received[accepted]{5 June 2009}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
+
 \section{Introduction}
@@
 The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
 
@@
 
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
+The rest of the thesis is organized as follows.
 Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
@@
 
-\pagebreak
 \section{Related Work}
@@
 A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
+Their idea is to store the queue elements in a singly-linked list.
 A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
@@
 Similarly, dequeue can take $\Omega(p)$ steps.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
 
@@
 
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
-
 Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
@@
 A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
+Global tail and local tail pointers are similar.
 A process updates the position of the pointers after it does an operation.
@@
 Their design is lock-free but not wait-free.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
 
@@
 
-% TODO: page
-
 Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
@@
 The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
 
@@
 
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
-
 \section{Proof of Correctness}
@@
 
-\pagebreak
-
-\addcontentsline{toc}{section}{References}
-\bibliography{main.bib} 
-\bibliographystyle{acm}
+\section{Template Overview}
+As noted in the introduction, the ``\verb|acmart|'' document class can
+be used to prepare many different kinds of documentation --- a
+double-blind initial submission of a full-length technical paper and more --- all by
+selecting the appropriate {\itshape template style} and {\itshape
+  template parameters}.
+
+This document will explain the major features of the document
+class. For further information, the {\itshape \LaTeX\ User's Guide} is
+available from
+\url{https://www.acm.org/publications/proceedings-template}.
+
+\subsection{Template Styles}
+
+The primary parameter given to the ``\verb|acmart|'' document class is
+the {\itshape template style} which corresponds to the kind of publication
+or SIG publishing the work. This parameter is enclosed in square
+brackets and is a part of the {\verb|documentclass|} command:
+\begin{verbatim}
+  \documentclass[STYLE]{acmart}
+\end{verbatim}
+
+Journals use one of three template styles. All but three ACM journals
+use the {\verb|acmsmall|} template style:
+\begin{itemize}
+\item {\texttt{acmsmall}}: The default journal template style.
+\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
+\item {\texttt{acmtog}}: Used by TOG.
+\end{itemize}
+
+The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
+\begin{itemize}
+\item {\texttt{acmconf}}: The default proceedings template style.
+\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
+\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
+\end{itemize}
+
+The template style used for PODC 2023 submissions is {\verb|acmsmall|}.
+
+\subsection{Template Parameters}
+
+In addition to specifying the {\itshape template style} to be used in
+formatting your work, there are a number of {\itshape template parameters}
+which modify some part of the applied template style. A complete list
+of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
+
+Frequently-used parameters, or combinations of parameters, include:
+\begin{itemize}
+\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
+  conference submission. Anonymizes the work and includes line
+  numbers. Use with the \texttt{\acmSubmissionID} command to print the
+  submission's unique ID on each page of the work.
+\item{\texttt{authorversion}}: Produces a version of the work suitable
+  for posting by the author.
+\item{\texttt{screen}}: Produces colored hyperlinks.
+\end{itemize}
+
+This document uses the following string as the first command in the
+source file:
+\begin{verbatim}
+\documentclass[acmsmall,nonacm,anonymous]{acmart}
+\end{verbatim}
+
+\section{Modifications}
+
+Modifying the template --- including but not limited to: adjusting
+margins, typeface sizes, line spacing, paragraph and list definitions,
+and the use of the \verb|\vspace| command to manually adjust the
+vertical spacing between elements of your work --- is not allowed.
+
+{\bfseries Your document will be returned to you for revision if
+  modifications are discovered.}
+
+\section{Typefaces}
+
+The ``\verb|acmart|'' document class requires the use of the
+``Libertine'' typeface family. Your \TeX\ installation should include
+this set of packages. Please do not substitute other typefaces. The
+``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
+as they will override the built-in typeface families.
+
+\section{Title Information}
+
+The title of your work should use capital letters appropriately -
+\url{https://capitalizemytitle.com/} has useful rules for
+capitalization. Use the {\verb|title|} command to define the title of
+your work. If your work has a subtitle, define it with the
+{\verb|subtitle|} command.  Do not insert line breaks in your title.
+
+If your title is lengthy, you must define a short version to be used
+in the page headers, to prevent overlapping text. The \verb|title|
+command has a ``short title'' parameter:
+\begin{verbatim}
+  \title[short title]{full title}
+\end{verbatim}
+
+\section{CCS Concepts and User-Defined Keywords}
+
+Two elements of the ``acmart'' document class provide powerful
+taxonomic tools for you to help readers find your work in an online
+search.
+
+The ACM Computing Classification System ---
+\url{https://www.acm.org/publications/class-2012} --- is a set of
+classifiers and concepts that describe the computing
+discipline. Authors can select entries from this classification
+system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
+commands to be included in the \LaTeX\ source.
+
+User-defined keywords are a comma-separated list of words and phrases
+of the authors' choosing, providing a more flexible way of describing
+the research being presented.
+
+CCS concepts and user-defined keywords are required for for all
+articles over two pages in length, and are optional for one- and
+two-page articles (or abstracts).
+
+\section{Sectioning Commands}
+
+Your work should use standard \LaTeX\ sectioning commands:
+\verb|section|, \verb|subsection|, \verb|subsubsection|, and
+\verb|paragraph|. They should be numbered; do not remove the numbering
+from the commands.
+
+Simulating a sectioning command by setting the first word or words of
+a paragraph in boldface or italicized text is {\bfseries not allowed.}
+
+\section{Tables}
+
+The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
+package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
+high-quality tables.
+
+Table captions are placed {\itshape above} the table.
+
+Because tables cannot be split across pages, the best placement for
+them is typically the top of the page nearest their initial cite.  To
+ensure this proper ``floating'' placement of tables, use the
+environment \textbf{table} to enclose the table's contents and the
+table caption.  The contents of the table itself must go in the
+\textbf{tabular} environment, to be aligned properly in rows and
+columns, with the desired horizontal and vertical rules.  Again,
+detailed instructions on \textbf{tabular} material are found in the
+\textit{\LaTeX\ User's Guide}.
+
+Immediately following this sentence is the point at which
+Table~\ref{tab:freq} is included in the input file; compare the
+placement of the table here with the table in the printed output of
+this document.
+
+\begin{table}
+  \caption{Frequency of Special Characters}
+  \label{tab:freq}
+  \begin{tabular}{ccl}
+    \toprule
+    Non-English or Math&Frequency&Comments\\
+    \midrule
+    \O & 1 in 1,000& For Swedish names\\
+    $\pi$ & 1 in 5& Common in math\\
+    \$ & 4 in 5 & Used in business\\
+    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
+  \bottomrule
+\end{tabular}
+\end{table}
+
+To set a wider table, which takes up the whole width of the page's
+live area, use the environment \textbf{table*} to enclose the table's
+contents and the table caption.  As with a single-column table, this
+wide table will ``float'' to a location deemed more
+desirable. Immediately following this sentence is the point at which
+Table~\ref{tab:commands} is included in the input file; again, it is
+instructive to compare the placement of the table here with the table
+in the printed output of this document.
+
+\begin{table*}
+  \caption{Some Typical Commands}
+  \label{tab:commands}
+  \begin{tabular}{ccl}
+    \toprule
+    Command &A Number & Comments\\
+    \midrule
+    \texttt{{\char'134}author} & 100& Author \\
+    \texttt{{\char'134}table}& 300 & For tables\\
+    \texttt{{\char'134}table*}& 400& For wider tables\\
+    \bottomrule
+  \end{tabular}
+\end{table*}
+
+Always use midrule to separate table header rows from data rows, and
+use it only for this purpose. This enables assistive technologies to
+recognise table headers and support their users in navigating tables
+more easily.
+
+\section{Math Equations}
+You may want to display math equations in three distinct styles:
+inline, numbered or non-numbered display.  Each of the three are
+discussed in the next sections.
+
+\subsection{Inline (In-text) Equations}
+A formula that appears in the running text is called an inline or
+in-text formula.  It is produced by the \textbf{math} environment,
+which can be invoked with the usual
+\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
+the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
+and structures, from $\alpha$ to $\omega$, available in
+\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
+examples of in-text equations in context. Notice how this equation:
+\begin{math}
+  \lim_{n\rightarrow \infty}x=0
+\end{math},
+set here in in-line math style, looks slightly different when
+set in display style.  (See next section).
+
+\subsection{Display Equations}
+A numbered display equation---one set off by vertical space from the
+text and centered horizontally---is produced by the \textbf{equation}
+environment. An unnumbered display equation is produced by the
+\textbf{displaymath} environment.
+
+Again, in either environment, you can use any of the symbols and
+structures available in \LaTeX\@; this section will just give a couple
+of examples of display equations in context.  First, consider the
+equation, shown as an inline equation above:
+\begin{equation}
+  \lim_{n\rightarrow \infty}x=0
+\end{equation}
+Notice how it is formatted somewhat differently in
+the \textbf{displaymath}
+environment.  Now, we'll enter an unnumbered equation:
+\begin{displaymath}
+  \sum_{i=0}^{\infty} x + 1
+\end{displaymath}
+and follow it with another numbered equation:
+\begin{equation}
+  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
+\end{equation}
+just to demonstrate \LaTeX's able handling of numbering.
+
+\section{Figures}
+
+The ``\verb|figure|'' environment should be used for figures. One or
+more images can be placed within a figure. If your figure contains
+third-party material, you must clearly identify it as such, as shown
+in the example below.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{sample-franklin}
+  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
+    Ewing, Inc. [Public domain], via Wikimedia
+    Commons. (\url{https://goo.gl/VLCRBB}).}
+  \Description{A woman and a girl in white dresses sit in an open car.}
+\end{figure}
+
+Your figures should contain a caption which describes the figure to
+the reader.
+
+Figure captions are placed {\itshape below} the figure.
+
+Every figure should also have a figure description unless it is purely
+decorative. These descriptions convey what’s in the image to someone
+who cannot see it. They are also used by search engine crawlers for
+indexing images, and when images cannot be loaded.
+
+A figure description must be unformatted plain text less than 2000
+characters long (including spaces).  {\bfseries Figure descriptions
+  should not repeat the figure caption – their purpose is to capture
+  important information that is not already provided in the caption or
+  the main text of the paper.} For figures that convey important and
+complex new information, a short text description may not be
+adequate. More complex alternative descriptions can be placed in an
+appendix and referenced in a short figure description. For example,
+provide a data table capturing the information in a bar chart, or a
+structured list representing a graph.  For additional information
+regarding how best to write figure descriptions and why doing this is
+so important, please see
+\url{https://www.acm.org/publications/taps/describing-figures/}.
+
+\subsection{The ``Teaser Figure''}
+
+A ``teaser figure'' is an image, or set of images in one figure, that
+are placed after all author and affiliation information, and before
+the body of the article, spanning the page. If you wish to have such a
+figure in your article, place the command immediately before the
+\verb|\maketitle| command:
+\begin{verbatim}
+  \begin{teaserfigure}
+    \includegraphics[width=\textwidth]{sampleteaser}
+    \caption{figure caption}
+    \Description{figure description}
+  \end{teaserfigure}
+\end{verbatim}
+
+\section{Citations and Bibliographies}
+
+The use of \BibTeX\ for the preparation and formatting of one's
+references is strongly recommended. Authors' names should be complete
+--- use full first names (``Donald E. Knuth'') not initials
+(``D. E. Knuth'') --- and the salient identifying features of a
+reference should be included: title, year, volume, number, pages,
+article DOI, etc.
+
+The bibliography is included in your source document with these two
+commands, placed just before the \verb|\end{document}| command:
+\begin{verbatim}
+  \bibliographystyle{ACM-Reference-Format}
+  \bibliography{bibfile}
+\end{verbatim}
+where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
+suffix, of the \BibTeX\ file.
+
+Citations and references are numbered by default. A small number of
+ACM publications have citations and references formatted in the
+``author year'' style; for these exceptions, please include this
+command in the {\bfseries preamble} (before the command
+``\verb|\begin{document}|'') of your \LaTeX\ source:
+\begin{verbatim}
+  \citestyle{acmauthoryear}
+\end{verbatim}
+
+  Some examples.  A paginated journal article \cite{Abril07}, an
+  enumerated journal article \cite{Cohen07}, a reference to an entire
+  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
+  monograph/whole book in a series (see 2a in spec. document)
+  \cite{Harel79}, a divisible-book such as an anthology or compilation
+  \cite{Editor00} followed by the same example, however we only output
+  the series if the volume number is given \cite{Editor00a} (so
+  Editor00a's series should NOT be present since it has no vol. no.),
+  a chapter in a divisible book \cite{Spector90}, a chapter in a
+  divisible book in a series \cite{Douglass98}, a multi-volume work as
+  book \cite{Knuth97}, a couple of articles in a proceedings (of a
+  conference, symposium, workshop for example) (paginated proceedings
+  article) \cite{Andler79, Hagerup1993}, a proceedings article with
+  all possible elements \cite{Smith10}, an example of an enumerated
+  proceedings article \cite{VanGundy07}, an informally published work
+  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
+    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
+  master's thesis: \cite{anisi03}, an online document / world wide web
+  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
+  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
+  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
+  publication \cite{rous08}, 'YYYYb'-test for prolific author
+  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
+  contain 'duplicate' DOI and URLs (some SIAM articles)
+  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
+  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
+  couple of citations with DOIs:
+  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
+  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
+  Artifacts: \cite{R} and \cite{UMassCitations}.
+
+\section{Acknowledgments}
+
+Identification of funding sources and other support, and thanks to
+individuals and groups that assisted in the research and the
+preparation of the work should be included in an acknowledgment
+section, which is placed just before the reference section in your
+document.
+
+This section has a special environment:
+\begin{verbatim}
+  \begin{acks}
+  ...
+  \end{acks}
+\end{verbatim}
+so that the information contained therein can be more easily collected
+during the article metadata extraction phase, and to ensure
+consistency in the spelling of the section heading.
+
+Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
+
+\section{Appendices}
+
+If your work needs an appendix, add it before the
+``\verb|\end{document}|'' command at the conclusion of your source
+document.
+
+Start the appendix with the ``\verb|appendix|'' command:
+\begin{verbatim}
+  \appendix
+\end{verbatim}
+and note that in the appendix, sections are lettered, not
+numbered. This document has two appendices, demonstrating the section
+and subsection identification method.
+
+\section{Multi-language papers}
+
+Papers may be written in languages other than English or include
+titles, subtitles, keywords and abstracts in different languages (as a
+rule, a paper in a language other than English should include an
+English title and an English abstract).  Use \verb|language=...| for
+every language used in the paper.  The last language indicated is the
+main language of the paper.  For example, a French paper with
+additional titles and abstracts in English and German may start with
+the following command
+\begin{verbatim}
+\documentclass[sigconf, language=english, language=german,
+               language=french]{acmart}
+\end{verbatim}
+
+The title, subtitle, keywords and abstract will be typeset in the main
+language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
+begin title, subtitle and keywords, can be used to set these elements
+in the other languages.  The environment \verb|translatedabstract| is
+used to set the translation of the abstract.  These commands and
+environment have a mandatory first argument: the language of the
+second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
+of their usage.
+
+%%
+%% The acknowledgments section is defined using the "acks" environment
+%% (and NOT an unnumbered section). This ensures the proper
+%% identification of the section in the article metadata, and the
+%% consistent spelling of the heading.
+\begin{acks}
+To Robert, for the bagels and explaining CMYK and color spaces.
+\end{acks}
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{sample-base}
+
+%%
+%% If your work has an appendix, this is the place to put it.
+\appendix
+
+\section{Research Methods}
+
+\subsection{Part One}
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
+malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
+sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
+vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
+lacinia dolor. Integer ultricies commodo sem nec semper.
+
+\subsection{Part Two}
+
+Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
+ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
+ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
+eros. Vivamus non purus placerat, scelerisque diam eu, cursus
+ante. Etiam aliquam tortor auctor efficitur mattis.
+
+\section{Online Resources}
+
+Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
+pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
+enim maximus. Vestibulum gravida massa ut felis suscipit
+congue. Quisque mattis elit a risus ultrices commodo venenatis eget
+dui. Etiam sagittis eleifend elementum.
+
+Nam interdum magna at lectus dignissim, ac dignissim lorem
+rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
+massa et mattis lacinia.
 
 \end{document}
+\endinput
+%%
+%% End of file `sample-acmsmall-conf.tex'.

### 2022-11-28 garbage_collect/main.tex
add description

@@
 
+\pagebreak
+\section{Description}
+
+\color{black}
+In our original algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
+This makes the space used by the algorithm factor of the number of operations of invoked on the queue.
+In this section here we want to free the memory allocated by the operations that are no longer needed and make the space used polynomial of $p+q$.
+
+One way of handling garbage collection is to reallocate the space taken by each each operation right after it is not needed anymore, but it might cost too much to do that.
+So we garbage collect by batching.
+If we do the garbage collection every $p^2$ block appended to the root, its cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$.
+
+In our design a process attempts to do \nf{Garbage Collect} every $p^2$ block appended to the root.
+\nf{Garbage Collect} corresponded to the $kp^2$ root block is called the $k$th round.
+To know if it is $k$th round or nor we can use an if condition on the value of \nf{root.head}.
+But the problem is that a process might miss a round not because it is on idle but because it reads \nf{root.head} not to be $kp^2$.
+We handle that by checking with a window.
+
+\begin{lemma}
+The number of the blocks in the \nf{blocks} is $O(p^2+q)$.
+\end{lemma}
+
+\begin{lemma}
+When $\cfrac{\nf{root.head}}{p^2}$ incremnts from $r$ to $r+1$, at least one operation has executed Lines 48,49.
+\end{lemma}
+
+An \nf{Enqueue} operation cannot be remove after it is terminated since later it might be dequeues, however it can be remove from the tree after it has been computed to be the response of a \nf{Dequeue}.
+A \nf{Dequeue} operation can be removed from the tree safely after it has computed its response \nf{Enqueue}.
+Note that it is not needed for the \nf{Dequeue} to be terminated but only it is sufficient if it has computed its response.
+We say a block in the \nf{root} is \it{finished} if all of its operations can be removed.
+
+\begin{lemma}
+There are at most $p^2+q$ not finished blocks in the \nf{root.blocks} at a time.
+\end{lemma}
+
+The good property in a fifo queue is that if $i$th \nf{Enqueue} gets dequeued then it means $1$ to $i-1$th enqueue operation have also dequeued.
+This gives us the idea that if a block is finished all the blocks before it are also finished.
+However if an operation in a block goes to sleep for a long time then the block remains not finished.
+Since there are at most $p$ idle operations we can help them and then remove all the finished blocks safely.
+
+\begin{lemma}
+    If all current operations are helped, then there is a block in the root that all of its previous blocks are finished.
+\end{lemma}
+
+The idea above leads us to a poly-log data structure that supports throwing away all the values smaller than an index. Red-black trees do this for us.
+We can create a shared red-black tree just creaing a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
+\begin{observation}
+PBRT supports....
+\end{observation}
+
+\begin{lemma}
+    If we replace the arrays we used to implement \nf{blocks} with red-black trees the amortized complexity of the algorithm would be $PolyLog(p,q)$. And also the algorithm is correct.
+\end{lemma}
+
+We can help a \nf{Dequeue} by computing its response and writing it down. If the process in future failed to execute, it can read the helped value written down.
+
+\begin{lemma}
+The \nf{response} written is correct.
+\end{lemma}
+
+But how can we know which blocks in each node are finished or not? We can keep track of the last finished block.
+
+\begin{lemma}
+    If all current operations are helped, then the blocks before the newest block that some \nf{Enqueue} has been dequeued from is safe to remove.
+If the most current \nf{Deuque} returned \nf{null} then all the blocks before the block containing the \nf{Dequeue} can be removed.
+\end{lemma}
+
+There is a shared array among processes which they write the last block dequeud from in it.
+
+\begin{lemma}
+    \nf{lastDequeuedFrom}- index of the last finished block in the root is $O(p)$.
+\end{lemma}
+
+\begin{lemma}
+    After \nf{FreeMemory}, the space taken by each node is $O(p^2+q)$
+\end{lemma}
+
 \end{document}

### 2022-11-30 garbage_collect/main.tex
update psudeocode

@@
 \item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\item {\color{red}\tt{\sl{int} garbageCollectRound}}
+% \item {\color{red}\tt{\sl{int} garbageCollectRound}}
 \end{itemize}
@@
 \If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \tt{lastDequeudFrom.update(b)}
+\State \tt{ \color{red} lastDequeudFrom.update(b)}
 \State \Return \tt{null} \label{returnNull}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{\color{red} Update needed: search on RBT}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
+\Statex $\triangleright$ \textmd{\color{red} Update needed: search on RBT does not need start and end, we can search over whole the red-black tree.}.
+\Function{int}{BinarySearch}{\sl{int} x}
+\State \Return \tt{\color{red} min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
+% \While{\nf{start<end}}
+% \State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
+% \If{\nf{blocks[mid].sum\sub{enq}<x}}
+% \State \nf{start= mid+1}
+% \Else
+% \State \nf{end= mid}
+% \EndIf
+% \EndWhile
+% \State\Return \nf{start}
 \EndFunction{BinarySearch}
@@
 
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
+\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end} \cmt{{\color{red} I think with garbage collection, doubling search is not needed any more and a binary search on \nf{root.blocks} would be good enough.}}
+% \State \tt{start= end-1} \label{dsearchStart} 
+% \While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
+% \State \tt{start= max(start-(end-start), 0)} \label{doubling}
+% \EndWhile \label{dsearchEnd}
+% \State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
+% \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
+% \State\Return \tt{<b,i>}
 \EndFunction{DoublingSearch}
@@
 \Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-{\color{red}\If{\tt{root.head\%p\textsuperscript{2}<p and garbageCollectRound<floor(root.head/p\textsuperscript{2})}}
-\State \Call{Help()}{}
-\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
-\State \tt{garbageCollectRound=floor(root.head/p\textsuperscript{2})}
-\EndIf}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
+\State \tt{\color{red} blocks.TryAppend(B, head)} \label{appendLeaf}
 \State \tt{\head= \head+1} \label{appendEnd} 
@@
 \EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
+\State{\tt{\color{red} result= blocks.TryAppend(new, h)}} \label{cas}
 \State{\tt{this.\Call{Advance}{h}}} \label{advance}
@@
 
+\Statex
+\color{red}
+\Function{int}{GetLastDequeuedFrom}{}{\cmt{Returns the index that is safe to remove the blocks before that in the node.}}
+\State \tt{x= lastDequeuedFrom.Get()-1}
+\State \tt{n= root}
+\State \bf{while} \tt{n!=this} \bf{do}
+\State \tt{dir= left (if this is in left subtree of n): otherwise dir=right}
+\State \tt{    x=n.blocks[x].end\sub{dir}}
+\State\bf{end while}
+\EndFunction{GetLastDequeuedFrom}
+
 \end{algorithmic}
@@
 \State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,}  \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
+\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} \cmt{{\color{red} start and end values are not needed anymore?}}
 \State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
@@
 \State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
+\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} \cmt{{\color{red} start and end values are not needed anymore?}}
 \State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{\color{red} Update needed: return null when superblock in the root was not found.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{{\color{red} Update needed: return null when superblock in the root was not found.}}
 \If{\tt{this} \bf{is} \tt{root}}
@@
 \State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper} 
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
+\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1} \cmt{{\color{red} Preconditions might be not met.}}
 
@@
 \begin{algorithm}
+\color{red}
 \caption{\tt{\sl{\color{red}MaxByProcess}}}
@@
 \begin{algorithm}
+\color{red}
 \caption{\tt{\sl{Tree}}}
@@
 \begin{algorithm}
+\color{red}
 \caption{\tt{\sl{Node}}}
@@
 \If{not leaf}
-\State \tt{left.FreeMemory(blocks[i].end\sub{left})}
-\State \tt{right.FreeMemory(blocks[i].end\sub{right})}
+\State \tt{left.FreeMemory(blocks[i].end\sub{left}-1)}
+\State \tt{right.FreeMemory(blocks[i].end\sub{right}-1)}
 \EndIf
-\State \tt{\bf{while} !blocks.CAS(blocks.splitGreater(i))}
+\State \tt{blocks= blocks.splitGreater(i)} \cmt{I think CAS is not needed.}
 \EndFunction{FreeMemory}
@@
 \begin{algorithm}
+\color{red}
 \caption{\tt{\sl{PBRT}}}
@@
 
+\Statex
 \Function{}{GetByBlock}{\sl{int} \tt{i}}
@@
 \EndIf
+\EndFunction{GetByBlock}
+
+\Statex
+\Function{}{TryAppend}{\sl{block} \tt{B}, \sl{int} \tt{i}}
+\Statex {\color{red} Tries to append \nf{B} with key \nf{i} to the red-black tree.}
+{\color{red}\If{\tt{i\%p\textsuperscript{2}=0}}
+\State \Call{Help()}{}
+\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
+\State \tt{garbageCollectRound=floor(root.head/p\textsuperscript{2})}
+\EndIf}
 \EndFunction{FreeMemory}

### 2022-12-01 garbage_collect/main.tex
update lemmas

@@
 
-One way of handling garbage collection is to reallocate the space taken by each each operation right after it is not needed anymore, but it might cost too much to do that.
+One way of handling garbage collection is to reallocate the space taken by each operation right after it is not needed anymore, but it might cost too much to do that.
 So we garbage collect by batching.
-If we do the garbage collection every $p^2$ block appended to the root, its cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$.
+If we garbage collect the unnecessary blocks in a node every $p^2$ block appended to the node, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$.
 
-In our design a process attempts to do \nf{Garbage Collect} every $p^2$ block appended to the root.
-\nf{Garbage Collect} corresponded to the $kp^2$ root block is called the $k$th round.
-To know if it is $k$th round or nor we can use an if condition on the value of \nf{root.head}.
-But the problem is that a process might miss a round not because it is on idle but because it reads \nf{root.head} not to be $kp^2$.
-We handle that by checking with a window.
+In our design a process garbage collects a node when it attempts to append a block in the $kp^2$ position in the.
+\nf{Garbage Collect} corresponded to the $kp^2$th block in node $n$ is called the $k$th round of garbage collect on $n$.
+Every $p^2$  block appended to a node at least one garbage collect happens because the node's \nf{head} cannot advance until the garbage collect is done (see Lines ).
 
@@
 
+An \nf{Enqueue} operation can be removed from the tree after its \nf{element} has been computed to be the response of some \nf{Dequeue}.
+It is safe to remove a \nf{Dequeue} operation from the tree after the \nf{Dequeue} is terminated, but a \nf{Dequeue} might sleep for a long time and prevents to be garbage collected.
+In that situation other processes can help the \nf{Dequeue} to compute its response and wtite down its response somewhere.
+If we remove a \nf{Dequeue} from the tree after it is helped the \nf{Dequeue} can read the helped response writtern when it encountered a problem (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}).
+A \nf{Dequeue} operation can be removed from the tree after its response has been computed.
+We say a block  is \it{finished} if all of its operations can be removed.
+
 \begin{lemma}
-When $\cfrac{\nf{root.head}}{p^2}$ incremnts from $r$ to $r+1$, at least one operation has executed Lines 48,49.
+There are at most $O(p^2+q)$ unfinished blocks in a node's \nf{blocks} at a time.
 \end{lemma}
 
-An \nf{Enqueue} operation cannot be remove after it is terminated since later it might be dequeues, however it can be remove from the tree after it has been computed to be the response of a \nf{Dequeue}.
-A \nf{Dequeue} operation can be removed from the tree safely after it has computed its response \nf{Enqueue}.
-Note that it is not needed for the \nf{Dequeue} to be terminated but only it is sufficient if it has computed its response.
-We say a block in the \nf{root} is \it{finished} if all of its operations can be removed.
+If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means $1$ to $i-1$th \nf{Enqueue} operations are also dequeued.
+This gives us the idea that if a block is finished then all the blocks before it are also finished.
+If an operation in a block goes to sleep for a long time then other processes help the operation so the block get finished.
+Note that when an \nf{Enqueue} operation in a block is not finished the block cannot be finished until some \nf{Dequeue} dequeues that \nf{Enqueue}.
+Since there are at most $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely.
 
 \begin{lemma}
-There are at most $p^2+q$ not finished blocks in the \nf{root.blocks} at a time.
+    If all current operations are helped, then there is a block in the root that all of its previous blocks are finished. That block is the most recent block that has been dequeued from.
 \end{lemma}
 
-The good property in a fifo queue is that if $i$th \nf{Enqueue} gets dequeued then it means $1$ to $i-1$th enqueue operation have also dequeued.
-This gives us the idea that if a block is finished all the blocks before it are also finished.
-However if an operation in a block goes to sleep for a long time then the block remains not finished.
-Since there are at most $p$ idle operations we can help them and then remove all the finished blocks safely.
+The idea above leads us to a poly-log data structure that supports throwing away all the blocks with keys smaller than an index.
+Red-black trees do this for us.
+\nf{Get(i)}, \nf{Append()} and \nf{Split(i)} are logarithmic in block trees.
+We can create a shared red-black tree just creating a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
 
-\begin{lemma}
-    If all current operations are helped, then there is a block in the root that all of its previous blocks are finished.
-\end{lemma}
-
-The idea above leads us to a poly-log data structure that supports throwing away all the values smaller than an index. Red-black trees do this for us.
-We can create a shared red-black tree just creaing a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
 \begin{observation}
-PBRT supports....
+PBRT supports poly-log operations ....
 \end{observation}
@@
 
-But how can we know which blocks in each node are finished or not? We can keep track of the last finished block.
+But how can we know which blocks in each node are finished or not?
 
@@
     If all current operations are helped, then the blocks before the newest block that some \nf{Enqueue} has been dequeued from is safe to remove.
-If the most current \nf{Deuque} returned \nf{null} then all the blocks before the block containing the \nf{Dequeue} can be removed.
+If the most current \nf{Dequeue} returned \nf{null} then all the blocks before the block containing the \nf{Dequeue} can be removed.
 \end{lemma}
 
-There is a shared array among processes which they write the last block dequeud from in it.
+There is a shared array among processes which they write the last block dequeued from in it.
 
 \begin{lemma}
-    \nf{lastDequeuedFrom}- index of the last finished block in the root is $O(p)$.
+    $\nf{GetLastDequeuedFrom($n$} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
 \end{lemma}
@@
 \begin{lemma}
-    After \nf{FreeMemory}, the space taken by each node is $O(p^2+q)$
+    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$
 \end{lemma}

### 2022-12-13 podc_paper/podc-submission.tex
Simplified formatting; got rid of some bugs that caused Latex some trouble compiling

--- 2022-11-15 thesis/thesis.tex
@@
 \usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
+%\theoremstyle{definition}
+%\newtheorem{definition}[theorem]{Definition}
 \newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
+%\newtheorem{proposition}[theorem]{Proposition}
+
+%%
+%% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{none}  %FIX FOR CAMERA READY
+%\copyrightyear{2023}
+%\acmYear{2023}
+\acmDOI{XXXXXXX.XXXXXXX}
+\fancyfoot{} % REMOVE THIS LINE FOR CAMERA READY
+
+%% These commands are for a PROCEEDINGS abstract or paper.
+\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
+  2023}{Orlando, FL}
+%%
+%%  Uncomment \acmBooktitle if the title of the proceedings is different
+%%  from ``Proceedings of ...''!
+%%
+%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
+%%  June 03--05, 2018, Woodstock, NY}
+%\acmPrice{15.00}
+%\acmISBN{978-1-4503-XXXX-X/18/06}
+
+%%
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+
+%%
+%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-\begin{center}
-\pagenumbering{roman}
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science,\\
-York University\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\addcontentsline{toc}{subsection}{Abstract}
-\section*{Abstract}
-
+\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
+
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+\author{Eric Ruppert}
+\affiliation{%
+  \institution{York University}
+  \streetaddress{P.O. Box 1212}
+  \city{Toronto}
+  \state{Ontario}
+  \country{Canada}
+  \postcode{43017-6221}
+}
+\email{ruppert@cse.yorku.ca}
+\orcid{1234-5678-9012}
+
+\author{Hossein Naderibeni}
+\email{hnaderi268@gmail.com}
+\affiliation{%
+  \institution{York University}
+  \streetaddress{P.O. Box 1212}
+  \city{Toronto}
+  \state{Ontario}
+  \country{Canada}
+  \postcode{43017-6221}
+}
+
+\settopmatter{printacmref=false}  % REMOVE THIS LINE FOR CAMERA READY
+
+\begin{abstract}
 In this work, we introduce a novel linearizable wait-free queue implementation.
@@
 The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-
-\pagebreak
-
-\addcontentsline{toc}{subsection}{Acknowledgements}
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\addcontentsline{toc}{subsection}{Table of Contents}
-\tableofcontents
-
-\pagebreak
-\addcontentsline{toc}{subsection}{List of Tables}
-\listoftables
-
-\pagebreak
-\addcontentsline{toc}{subsection}{List of Figures}
-\listoffigures
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
+\end{abstract}
+
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{concurrent data structures, wait-free queues}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
+
 \section{Introduction}
@@
 \begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
+  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
   \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
@@
 \end{algorithm}
-
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
 
@@
 
-\pagebreak
-
-\pagebreak
-
-\addcontentsline{toc}{section}{References}
-\bibliography{main.bib} 
-\bibliographystyle{acm}
-
+\bibliographystyle{plain}
+\bibliography{sample-base.bib}
+
+\appendix
 \end{document}

### 2022-12-13 podc_paper/podc-submission.tex
Working on intro

@@
 \usepackage{bold-extra}
+\usepackage{showkeys}
 \renewcommand*\contentsname{Table of Contents}
@@
 \newcommand{\size}{size }
+
+\newcommand{\here}[1]{\bf{[[#1]]}}
 
@@
 \begin{abstract}
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
+We introduce a novel linearizable wait-free queue implementation using single-word 
+CAS instructions.
+Previous lock-free queues all take $\Omega(p)$ steps per operation in the worst case, 
+where $p$ is the number of processes that can access the queue.
+We achieve $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue.
+\here{mention space usage; update time bound if space-efficient version takes more time per op}
 \end{abstract}
@@
 \section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
+
+A great deal of research has been devoted to the design of lock-free queues,
+starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
+Their design uses a singly-linked list with pointers to the front and back nodes.
+To dequeue or enqueue an element, the front or back pointer is updated by a 
+compare-and-swap (CAS) instruction.
+If this CAS fails, the operation must retry.
+In the worst case, this means that each successful CAS may cause all other processes to
+fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
+Numerous papers have suggested modifications to the MS-queue \cite{list some}, but 
+all still have $\Omega(p)$ worst-case step complexity as a result of
+contention on the front and back of the queue.
+Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
+The same problem occurs in array-based implementations of queues \cite{list some}.
+
+This problem reoccurs with other data structures that keep track of a set of elements.
+For example, lock-free lists \cite{}, stacks \cite{} and search trees \cite{} 
+always have a term in their step complexity
+that depends linearly on the contention $c$, i.e., 
+the number of processes that access the data structure concurrently (which can be as high
+as $p$ in the worst case).
+It was conjectured that lock-free implementations of many types of data structures
+require $\Omega(c)$ steps per operation, even in an amortized sense \cite{Ruppert}.
+Attiya and Fouren \cite{AF} proved this conjecture for bag data structures, which allow processes
+to insert elements and delete an arbitrary element (chosen non-deterministically).
+They showed that any implementation of a bag from CAS instructions
+has $\Omega(c)$ amortized step complexity.
+Since a queue trivially implements a bag, the lower bound also holds for queues.
+However the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
+Thus, the lower bound could be stated more as $\Omega(\min(c,\log\log p))$ steps per operation.
+
+In this paper, we exploit this loophole to show that it is possible to implement a queue
+whose step complexity does not depend linearly on $p$.
+Our implementation's step complexity is polylogarithmic in $p$ and in $q$, the number of elements
+stored in the queue.
+
+Our queue uses a tournament tree where each process has its own leaf.
+The process tries to propagate its operations along the path from its leaf up to the root.
+Operations that reach the root are ordered, and this ordering is used to linearize all
+operations and compute their responses.
+\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
+We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
+The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
+\nf{CAS} instructions cost more than other instructions for the processor.
+Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
+We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
+Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
+------
 
@@
 We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
 

### 2022-12-15 podc_paper/introduction.tex
finished editing introduction

% !TEX root =  podc-submission.tex

\section{Introduction}

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
Numerous papers have suggested modifications to the MS-queue \cite{list some}, but 
all still have $\Omega(p)$ worst-case step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

This problem occurs more broadly in other data structures that keep track of a set of elements, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have a term in their step complexity
that depends linearly on the contention $c$, i.e., 
the number of processes that access the data structure concurrently (which can be as high
as $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which allow processes
to insert elements and remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole to show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure, but 
leave the problem of reclaiming that memory up to a garbage collector (such as the one Java provides).

Our queue uses a tournament tree where each process has its own leaf.
Each process propagates its operations along the path from its leaf up to the root.
The root of the tree provides an ordering for all the operations,
and this ordering is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
To ensure wait-freedom, we use a helping mechanism:  a process propagating its operation 
to a node $n$ helps propagate all operations from both of $n$'s children.
If we explicitly stored all operations in the nodes, this helping would become very costly.
Instead, we use an implicit representation of collections
of operations that allows for quickly merging two collections from the two children of a node,
and for quickly accessing any particular operation in the collection.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2022-12-15 podc_paper/podc-submission.tex
finished editing introduction

@@
 
-\usepackage{amsmath,amssymb,amsthm}
+\usepackage{amsmath,amsthm}
 %\newtheorem{theorem}{Theorem}
@@
 %\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
+\newtheorem{observation}{Observation}
 %\theoremstyle{definition}
 %\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
+\newtheorem{invariant}{Invariant}
 %\newtheorem{proposition}[theorem]{Proposition}
@@
 
-\section{Introduction}
-
-A great deal of research has been devoted to the design of lock-free queues,
-starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
-Their design uses a singly-linked list with pointers to the front and back nodes.
-To dequeue or enqueue an element, the front or back pointer is updated by a 
-compare-and-swap (CAS) instruction.
-If this CAS fails, the operation must retry.
-In the worst case, this means that each successful CAS may cause all other processes to
-fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
-Numerous papers have suggested modifications to the MS-queue \cite{list some}, but 
-all still have $\Omega(p)$ worst-case step complexity as a result of
-contention on the front and back of the queue.
-Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
-The same problem occurs in array-based implementations of queues \cite{list some}.
-
-This problem reoccurs with other data structures that keep track of a set of elements.
-For example, lock-free lists \cite{}, stacks \cite{} and search trees \cite{} 
-always have a term in their step complexity
-that depends linearly on the contention $c$, i.e., 
-the number of processes that access the data structure concurrently (which can be as high
-as $p$ in the worst case).
-It was conjectured that lock-free implementations of many types of data structures
-require $\Omega(c)$ steps per operation, even in an amortized sense \cite{Ruppert}.
-Attiya and Fouren \cite{AF} proved this conjecture for bag data structures, which allow processes
-to insert elements and delete an arbitrary element (chosen non-deterministically).
-They showed that any implementation of a bag from CAS instructions
-has $\Omega(c)$ amortized step complexity.
-Since a queue trivially implements a bag, the lower bound also holds for queues.
-However the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
-Thus, the lower bound could be stated more as $\Omega(\min(c,\log\log p))$ steps per operation.
-
-In this paper, we exploit this loophole to show that it is possible to implement a queue
-whose step complexity does not depend linearly on $p$.
-Our implementation's step complexity is polylogarithmic in $p$ and in $q$, the number of elements
-stored in the queue.
-
-Our queue uses a tournament tree where each process has its own leaf.
-The process tries to propagate its operations along the path from its leaf up to the root.
-Operations that reach the root are ordered, and this ordering is used to linearize all
-operations and compute their responses.
-\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-------
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-
-\section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list.
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
-
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
-
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
-
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar.
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
-
-\subsection{Restricted Queues}
-
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
+\input{introduction}
+\input{related}
 
@@
 
+\newpage
+
 \bibliographystyle{plain}

### 2022-12-15 podc_paper/related.tex
finished editing introduction

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
-\renewcommand{\baselinestretch}{1.8}
+% !TEX root =  podc-submission.tex
 
-\usepackage[letterpaper, margin=1in]{geometry}
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-%\usepackage{showkeys}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-
-\usepackage{multicol}
-\setlength\columnsep{24pt}
-
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\begin{document}
-
-\begin{center}
-
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\section*{Abstract}
-
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
-\pagebreak
-
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\tableofcontents
-
-\pagebreak
-\listoffigures
-
-\pagebreak
-\listoftables
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
-\section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
-
-\pagebreak
 \section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
+In this section, we look at previous lock-free queues. The amortized
+step complexity of all the lock-free queues includes an $\Omega(p)$
+term that comes from the worst case, where all $p$ processes try to do
+an enqueue or a dequeue simultaneously. Morrison and Afek call this
+the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
 
 \subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
+Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a
+lock-free queue which we refer to as the MS-queue. A version of it is
+included in the standard Java Concurrency Package. Their idea is to
+store the queue elements in a singly-linked list. A shared variable
+Head points to the first node in the linked list that has not been
+dequeued, and the Tail points to the last element in the queue. To
+insert a node into the linked list, they use atomic primitive
+operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try
+to enqueue simultaneously, only one can succeed, and the others have
+to retry. This makes the amortized number of steps $\Omega(p)$ per
+enqueue. Similarly, dequeue can take $\Omega(p)$ steps.
 
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
+Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a
+more sophisticated queue by using the \it{elimination} technique. The
+elimination mechanism has the dual purpose of allowing operations to
+complete in parallel and reducing contention for the queue. An
+Elimination Queue consists of an MS-queue augmented with an
+elimination array. Elimination works by allowing opposing pairs of
+concurrent operations such as an enqueue and a dequeue to exchange
+values when the queue is empty or when concurrent operations can be
+linearized to empty the queue. Their algorithm makes it possible for
+long-running operations to eliminate an opposing operation. The
+empirical evaluation showed the throughput of their work is better
+than the MS-queue, but the worst case is still the same; in case there
+are $p$ concurrent enqueues, their algorithm is not better than
+MS-queue.
 
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
+Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried
+to make the MS-queue more parallel by introducing the Baskets
+Queue. Their idea is to allow more parallelism by treating the
+simultaneous enqueue operations as a basket. Each basket has a time
+interval in which all its nodes' enqueue operations overlap. Since the
+operations in a basket are concurrent, we can order them in any
+way. Enqueues in a basket try to find their order in the basket one by
+one using \texttt{CAS} operations. However, like the previous
+algorithms, if there are $p$ concurrent enqueue operations in a
+basket, the amortized step complexity remains $\Omega(p)$ per
+operation.
 
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
+Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
+presented an optimistic approach to implement a queue. MS-queue uses
+two \nf{CAS}es to do an enqueue: one to change the tail to the new
+node and another one to change the next pointer of the previous node
+to the new node. They use a doubly-linked list to do fewer
+\texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in
+previous algorithms, the worst case happens in the case where the
+contention is high: when $p$ concurrent enqueues happen, their nodes
+have to be appended to the linked list one by one. The amortized
+complexity is still $\Omega(p)$ \texttt{CAS}es.
 
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
+Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new
+paradigm called flat combining. The key idea behind flat combining is
+to allow a combiner who has acquired the global lock on the data
+structure to learn about the requests of threads on the queue, combine
+them and apply the combined results on the data structure. Their queue
+is linearizable but not lock-free and they present experiments that
+show their algorithm performs well in some situations.
 
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
+Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10}
+introduced a new algorithm using a linked list of arrays. The queue is
+stored in a shared array where head and tail pointers point to the
+current elements in the queue. When the array is full, an empty array
+is linked to the array and tail pointers are updated. A global head
+points to the array containing the first element in the queue, and
+each process has a local head index that points to the first element
+in that array. Global tail and local tail pointers are similar. A
+process updates the position of the pointers after it does an
+operation. One process might go to sleep before setting the pointers,
+so the pointers might be behind their real places. They mention how to
+scan the arrays to update pointers while doing an operation. A process
+writes an element in the location head by a \nf{CAS} instruction, so
+if $p$ processes try to enqueue simultaneously, the amortized step
+(and \nf{CAS}) complexity remains $\Omega(p)$. Their design is
+lock-free but not wait-free.
 
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free
+queues based on the MS-queue and use Herlihy's helping
+technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their
+amortized step complexity is $\Omega(p)$ because of the helping
+mechanism.
 
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
+Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free
+queue supporting futures. In their queue, operations  return future
+objects instead of responses. Later when the response is needed, it
+can be evaluated from the future object. They also define a weaker
+linearizability condition such that each operation can be linearized
+between its invocation and when its future is evaluated. Their idea of
+batching allows a sequence of operations to be submitted as a batch
+for later execution on the MS-queue. They use some properties of the
+queue size before and after a batch, similar to a part of our
+work. Their queue is not wait-free: in fact, if the batch sizes are 1,
+then the queue is like MS-queue.
 
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
+Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a
+wait-free queue that uses the fast-path slow-path methodology
+introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their
+work is based on a circular queue using bounded memory. When a process
+wishes to do an enqueue or a dequeue, it starts two paths. The fast
+path  ensures good performance while the slow path ensures
+termination. They show that these two paths do not affect each other
+and the queue remains consistent. If a process makes no progress,
+other processes help its slow path to finish. The helping phase
+suffers from the CAS Retry Problem because processes compete in a
+\nf{CAS} loop to decide which succeeds to help. Because of this, the
+amortized complexity cannot be better than $\Omega(p)$.
 
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
+The CAS Retry Problem is not limited to list-based queues; array-based
+queues also share
+it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+Our motivation is to overcome this problem and present a wait-free
+sublinear queue.
 
@@
 
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
+David introduced the first sublinear concurrent queue
+\cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$
+steps for each operation, it is a single-enqueuer single-dequeuer
+queue and uses infinite memory. 
+The author states that to reduce memory usage to be bounded, the time
+per operation increases linearly.
 
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
+Jayanti and Petrovic introduced a wait-free poly-logarithmic
+multi-enqueuer single-dequeuer
+queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having
+a tournament tree among processes to agree on the linearization of
+operations to design a poly-logarithmic multi-enqueuer multi-dequeuer
+queue. Unlike their work, our algorithm does not put a limit on the
+number of concurrent dequeuers.
 
 \subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
+A \textit{universal construction} is an algorithm that can implement a
+shared version of any given sequential object. The first universal
+construction was introduced by
+Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent
+queue using a universal construction. Jayanti proved an $\Omega(\log
+p)$ lower bound on the worst-case shared-access time complexity of
+$p$-process universal
+constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that
+the universal construction by Afek, Dauber, and
+Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$
+worst-case step complexity, using atomic access to $\Omega(p \log
+p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal
+construction that achieves $\textsc{O}(\log^2 p)$ shared
+accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm
+cannot be used to create a queue. We mention a non-practical universal
+construction with a poly-log number of \nf{CAS} instructions in the
+last paragraph of page 13.
 
-% TODO: page
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
+Ellen and Woelfel introduced an implementation of a Fetch\&Inc object
+with step complexity of $O(\log p)$ using $O(\log n$)-bit
+\texttt{LL/SC} objects, where $n$ is the number of
+operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
+logarithmic complexity is to use a tree storing the Fetch\&Inc
+operations invoked by processes. When a process wants to do a
+Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number
+of elements in the tree. There are some similarities between designing
+a queue and a Fetch\&Inc object. A Fetch\&Inc object can be
+constructed from a queue. The algorithm by Ellen and Woelfel is
+interesting because of the similarities between Fetch\&Inc objects and
+queues. Also, it is one of the few wait-free data structures achieving
+poly-logarithmic complexity.
 
 \subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
-
-\section{Proof of Correctness}
-
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
-
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
-
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
-
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\pagebreak
-
-\pagebreak
-
-\bibliography{main.bib} 
-\bibliographystyle{acm}
-
-\end{document}
+Because of the \nf{CAS} retry problem in previous list-based queues
+one might guess the $\Omega(p)$ term is inherent in the time
+complexity of concurrent queues. Attiya and Fouren gave a lower bound
+on amortized complexity of lock-free queues with regard to $c$, the
+number of concurrent processes. Their result says if $c$ is
+$O(\log\log p)$, any implementation of queues using reads, writes and
+conditional operations like \nf{CAS} has $\Omega(c)$ amortized
+complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is
+that since their result is about when contention is low, their lower
+bound does not contradict our algorithm's complexity and we manage to
+reach poly-log time complexity.

### 2022-12-15 podc_paper/introduction.tex
working on related work section

@@
 fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
-Numerous papers have suggested modifications to the MS-queue \cite{list some}, but 
-all still have $\Omega(p)$ worst-case step complexity as a result of
+(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
+and divide the number of steps in the execution by the number of operations initated in the execution.)
+Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
+all still have $\Omega(p)$ amortized step complexity as a result of
 contention on the front and back of the queue.
@@
 Since a queue trivially implements a bag, this lower bound also holds for queues.
+At first glance, this would seem to settle the question of the step complexity of lock-free queues.
 However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
@@
 and $\Omega(p)$ CAS instructions in an amortized sense.
-For the space-bounded implementation, we unlink unneeded memory from our data structure, but 
-leave the problem of reclaiming that memory up to a garbage collector (such as the one Java provides).
+For the space-bounded implementation, we unlink unneeded memory from our data structure.
+We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
+garbage collector, such as the highly optimized one Java provides.
 

### 2022-12-15 podc_paper/related.tex
working on related work section

@@
 \section{Related Work}
-In this section, we look at previous lock-free queues. The amortized
-step complexity of all the lock-free queues includes an $\Omega(p)$
-term that comes from the worst case, where all $p$ processes try to do
-an enqueue or a dequeue simultaneously. Morrison and Afek call this
-the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
 
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a
-lock-free queue which we refer to as the MS-queue. A version of it is
-included in the standard Java Concurrency Package. Their idea is to
-store the queue elements in a singly-linked list. A shared variable
-Head points to the first node in the linked list that has not been
-dequeued, and the Tail points to the last element in the queue. To
-insert a node into the linked list, they use atomic primitive
-operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try
-to enqueue simultaneously, only one can succeed, and the others have
-to retry. This makes the amortized number of steps $\Omega(p)$ per
-enqueue. Similarly, dequeue can take $\Omega(p)$ steps.
+The MS-queue \cite{DBLP:conf/podc/MichaelS96} has stood the test of time.
+A version of it is
+included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
+However, as mentioned above, it suffers from the CAS retry problem
+which means it is lock-free but not wait-free and has an amortized step complexity
+of $\Theta(p)$ per operation.
 
@@
 
-\subsection{Restricted Queues}
+\paragraph{Other Primitives and Restricted Queues}
 
-David introduced the first sublinear concurrent queue
-\cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$
-steps for each operation, it is a single-enqueuer single-dequeuer
-queue and uses infinite memory. 
-The author states that to reduce memory usage to be bounded, the time
-per operation increases linearly.
+David introduced the first sublinear-time queue
+\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
+It uses fetch\&increment and swap primitive instructions and takes constant time per operation, but
+uses unbounded memory.  A modification to use bounded space
+is described, but it increases the time per operation to $\Omega(p)$.
 
 Jayanti and Petrovic introduced a wait-free poly-logarithmic
-multi-enqueuer single-dequeuer
-queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having
+queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it only works for a single dequeuer. 
+Our implementation uses their idea of having
 a tournament tree among processes to agree on the linearization of
-operations to design a poly-logarithmic multi-enqueuer multi-dequeuer
-queue. Unlike their work, our algorithm does not put a limit on the
-number of concurrent dequeuers.
+operations.
+
+\here{new result I found:}
+Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
+with $O(\sqrt{p})$ complexity using non-standard synchronization primitives
+called half-increment and half-max, which can be viewed as particular kinds of
+double-word read-modify-write operations.
+They use this as evidence that their primitives can be more efficient than CAS
+since previous CAS-based queues all required $\Omega(p)$ step complexity.
+Our new implementation counters this argument.
 

### 2022-12-15 podc_paper/introduction.tex
working on related work

@@
 
-A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
-starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
+A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
+As well as being a fundamental data structure, queues are needed in
+significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
+and sharing resources or tasks.
+The lock-free MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96} is a classic shared queue implementation.
 Their design uses a singly-linked list with pointers to the front and back nodes.

### 2022-12-15 podc_paper/related.tex
working on related work

@@
 
-The MS-queue \cite{DBLP:conf/podc/MichaelS96} has stood the test of time.
+Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
+in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
+is a lock-free queue that has stood the test of time.
 A version of it is
@@
 
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a
-more sophisticated queue by using the \it{elimination} technique. The
-elimination mechanism has the dual purpose of allowing operations to
-complete in parallel and reducing contention for the queue. An
-Elimination Queue consists of an MS-queue augmented with an
-elimination array. Elimination works by allowing opposing pairs of
-concurrent operations such as an enqueue and a dequeue to exchange
-values when the queue is empty or when concurrent operations can be
-linearized to empty the queue. Their algorithm makes it possible for
-long-running operations to eliminate an opposing operation. The
-empirical evaluation showed the throughput of their work is better
-than the MS-queue, but the worst case is still the same; in case there
-are $p$ concurrent enqueues, their algorithm is not better than
-MS-queue.
+A number of papers have described ways to cope with the high contention at the front
+and back of the MS-queue.
+Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
+added an elimination array that allows an enqueue to pass its value directly
+to a concurrent dequeue when the queue is empty or when concurrent operations 
+can be linearized to empty the queue.
+However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
+is still present.
+The baskets queue of
+Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
+is another approach to avoid contention:  concurrent enqueues are grouped into baskets
+and operations within a basket can be ordered arbitrarily.
+However, if there are $p$ concurrent enqueues in the same basket
+the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
+Both modifications thus still have $\Omega(p)$ amortized step complexity.
 
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried
-to make the MS-queue more parallel by introducing the Baskets
-Queue. Their idea is to allow more parallelism by treating the
-simultaneous enqueue operations as a basket. Each basket has a time
-interval in which all its nodes' enqueue operations overlap. Since the
-operations in a basket are concurrent, we can order them in any
-way. Enqueues in a basket try to find their order in the basket one by
-one using \texttt{CAS} operations. However, like the previous
-algorithms, if there are $p$ concurrent enqueue operations in a
-basket, the amortized step complexity remains $\Omega(p)$ per
-operation.
+-----
 
@@
 
+\here{Check:
+Turek Shasha Prakash locking without blocking (PODS92).
+Prakash Lee Johnson non blocking algorithms for concurrent data structures Tech rept 91-002 U of Florida 1991.
+}
+
 \paragraph{Other Primitives and Restricted Queues}
@@
 poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues
-one might guess the $\Omega(p)$ term is inherent in the time
-complexity of concurrent queues. Attiya and Fouren gave a lower bound
-on amortized complexity of lock-free queues with regard to $c$, the
-number of concurrent processes. Their result says if $c$ is
-$O(\log\log p)$, any implementation of queues using reads, writes and
-conditional operations like \nf{CAS} has $\Omega(c)$ amortized
-complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is
-that since their result is about when contention is low, their lower
-bound does not contradict our algorithm's complexity and we manage to
-reach poly-log time complexity.

### 2022-12-15 podc_paper/introduction.tex
related work; made bib entries consistent

--- 2022-12-15 podc_paper/introduction.tex
@@
 
-A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
-starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
+A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
+As well as being a fundamental data structure, queues are used in
+significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
+and sharing resources or tasks.
+The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
 Their design uses a singly-linked list with pointers to the front and back nodes.

### 2022-12-15 podc_paper/related.tex
related work; made bib entries consistent

@@
 
-Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
-in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
+Michael and Scott \cite{MS98} describe the early history of concurrent queue algorithms
+in the paper that described their MS-queue, which
 is a lock-free queue that has stood the test of time.
@@
 
-A number of papers have described ways to cope with the high contention at the front
+A number of papers have described ways to reduce contention at the front
 and back of the MS-queue.
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
-added an elimination array that allows an enqueue to pass its value directly
+Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
+added an elimination array that allows an enqueue to pass its enqueued value directly
 to a concurrent dequeue when the queue is empty or when concurrent operations 
@@
 Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
-is another approach to avoid contention:  concurrent enqueues are grouped into baskets
+is another approach to avoid contention:  concurrent enqueues are grouped into baskets.
+If an enqueue fails its CAS, it is put in the basket with the operation that succeeded,
 and operations within a basket can be ordered arbitrarily.
 However, if there are $p$ concurrent enqueues in the same basket
-the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
+the CAS retry problem occurs when they order themselves into a list using CAS instructions.
 Both modifications thus still have $\Omega(p)$ amortized step complexity.
+
+- MKLL18 batches several of a process's operations together and apply them as a group to the queue but this assumes processes can wait to do a batch all at once.
+In worst case batches of size 1 reduces to MS98
+- RC17 sketched how to add a helping mechanism to ensure wait-freedom (increases amortized time to Omega($P^2$) for enqueue)
+- Kogan Petrank PPoPP 2011 add helping to MS98 queue to make it wait-free; only makes complexity worse
+
+LS08 modified MS98; switched to doubly-linked list to reduce number of successful CASes needed for an op from 2 to 1 in optimistic approach, but could get inconsistent pointers that can be fixed when needed.
+Fixing the list, though rare in practice, is expensive, so Omega(nP) amortized time per op.
 
@@
 
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free
+Milman-Sela et al.~\cite{MKLLP22} designed a lock-free
 queue supporting futures. In their queue, operations  return future

### 2022-12-15 podc_paper/introduction.tex
working on related work section

@@
 significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
+packet processing \cite{DPDK},
 and sharing resources or tasks.
@@
 and divide the number of steps in the execution by the number of operations initated in the execution.)
-Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
+Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but
 all still have $\Omega(p)$ amortized step complexity as a result of

### 2022-12-15 podc_paper/related.tex
working on related work section

--- 2022-12-15 podc_paper/related.tex
@@
 
-Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
-in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
-is a lock-free queue that has stood the test of time.
+\paragraph{List-based Queues}
+The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
 A version of it is
 included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
-However, as mentioned above, it suffers from the CAS retry problem
-which means it is lock-free but not wait-free and has an amortized step complexity
+See the paper that introduced it for a survey of the early history of concurrent queues.
+As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
+Thus, it is lock-free but not wait-free and has an amortized step complexity
 of $\Theta(p)$ per operation.
 
-A number of papers have described ways to cope with the high contention at the front
-and back of the MS-queue.
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
-added an elimination array that allows an enqueue to pass its value directly
+A number of papers have described ways to reduce contention in the MS-queue.
+Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
+added an elimination array that allows an enqueue to pass its enqueued value directly
 to a concurrent dequeue when the queue is empty or when concurrent operations 
@@
 Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
-is another approach to avoid contention:  concurrent enqueues are grouped into baskets
-and operations within a basket can be ordered arbitrarily.
-However, if there are $p$ concurrent enqueues in the same basket
-the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
-Both modifications thus still have $\Omega(p)$ amortized step complexity.
+attempts to reduce contention by grouping concurrent enqueues into baskets.
+An enqueue that fails its CAS is put in the basket with the eqnueue that succeeded.
+Enqueues within a basket determine their order among themselves without having to access the back of the queue.
+However, if $p$ concurrent enqueues are in the same basket
+the CAS retry problem occurs when they order themselves using CAS instructions.
+Both modifications still have $\Omega(p)$ amortized step complexity.
+
+Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} described how to use futures to improve
+the performance of the MS-queue.
+Operations return future
+objects instead of responses. Later, when an operation's response is needed, it
+can be evaluated using the corresponding future object.
+This allows batches of enqueues or dequeues to be done at once on an MS-queue.
+However, the implementation satisfies a weaker correctness condition than linearizability.
+Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
+to mix enqueues and dequeues.
+\here{Is this worth saying:
+They use some properties of the queue size before and after a batch, similar to a part of our work.}
+In the worst case, where operations require their response right away,
+batches have size 1, and both of these implementations behave like a standard MS-queue.
+
+Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
+presented an optimistic  queue implementation. 
+%MS-queue uses
+%two \nf{CAS}es to do an enqueue: one to change the tail to the new
+%node and another one to change the next pointer of the previous node
+%to the new node. 
+In the MS-queue, an enqueue requires two CAS steps.
+The optimistic queue uses a doubly-linked list to reduce the number of
+\texttt{CAS} instructions to one in the best case. 
+Pointers in the doubly-linked list can be inconsistent, but are fixed when needed by traversing the list.
+Although this fixing is rare in practice, it yields an amortized complexity of $\Omega(qp)$ 
+steps per operation for worst-case executions.
+
+Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
+added Herlihy's helping
+technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
+a wait-free queue.
+Ramalhete and Correia \cite{RC17} added a different helping mechanism.
+In both cases, the helping mechanisms only add to the step complexity of the MS-queue.
 
@@
 
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
-presented an optimistic approach to implement a queue. MS-queue uses
-two \nf{CAS}es to do an enqueue: one to change the tail to the new
-node and another one to change the next pointer of the previous node
-to the new node. They use a doubly-linked list to do fewer
-\texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in
-previous algorithms, the worst case happens in the case where the
-contention is high: when $p$ concurrent enqueues happen, their nodes
-have to be appended to the linked list one by one. The amortized
-complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new
-paradigm called flat combining. The key idea behind flat combining is
-to allow a combiner who has acquired the global lock on the data
-structure to learn about the requests of threads on the queue, combine
-them and apply the combined results on the data structure. Their queue
-is linearizable but not lock-free and they present experiments that
-show their algorithm performs well in some situations.
+% OMIT, since it is not lock-free:
+%Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new
+%paradigm called flat combining. The key idea behind flat combining is
+%to allow a combiner who has acquired the global lock on the data
+%structure to learn about the requests of threads on the queue, combine
+%them and apply the combined results on the data structure. Their queue
+%is linearizable but not lock-free and they present experiments that
+%show their algorithm performs well in some situations.
 
@@
 lock-free but not wait-free.
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free
-queues based on the MS-queue and use Herlihy's helping
-technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their
-amortized step complexity is $\Omega(p)$ because of the helping
-mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free
-queue supporting futures. In their queue, operations  return future
-objects instead of responses. Later when the response is needed, it
-can be evaluated from the future object. They also define a weaker
-linearizability condition such that each operation can be linearized
-between its invocation and when its future is evaluated. Their idea of
-batching allows a sequence of operations to be submitted as a batch
-for later execution on the MS-queue. They use some properties of the
-queue size before and after a batch, similar to a part of our
-work. Their queue is not wait-free: in fact, if the batch sizes are 1,
-then the queue is like MS-queue.
 
@@
 \here{Check:
-Turek Shasha Prakash locking without blocking (PODS92).
-Prakash Lee Johnson non blocking algorithms for concurrent data structures Tech rept 91-002 U of Florida 1991.
+John Giacomoni et al. FastForward
+https://dl-acm-org.ezproxy.library.yorku.ca/doi/10.1145/1345206.1345215
 }
@@
 Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
-with $O(\sqrt{p})$ complexity using non-standard synchronization primitives
+with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
 called half-increment and half-max, which can be viewed as particular kinds of

### 2022-12-17 podc_paper/introduction.tex
working on related work

@@
 
-A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
-As well as being a fundamental data structure, queues are used in
+There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
+Besides being a fundamental data structure, queues are used in
 significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
@@
 The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
-Their design uses a singly-linked list with pointers to the front and back nodes.
+IT uses a singly-linked list with pointers to the front and back nodes.
 To dequeue or enqueue an element, the front or back pointer is updated by a 
@@
 contention on the front and back of the queue.
-Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
-The same problem occurs in array-based implementations of queues \cite{list-some}.
+Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
+The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
+Solutions that tried to sidestep this problem using fetch-and-increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
+rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.
 
-This problem occurs more broadly in other data structures that keep track of a set of elements, as observed by Ruppert \cite{Rup16}.
+Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert \cite{Rup16}.
 For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
-have a term in their step complexity
-that depends linearly on the contention $c$, i.e., 
-the number of processes that access the data structure concurrently (which can be as high
-as $p$ in the worst case).
+have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
+the number of processes that access the data structure concurrently (which can be $p$ in the worst case).
 Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
 that amortized $\Omega(c)$ steps per operation are indeed necessary
-for CAS-based implementations of lock-free bag data structures, which allow processes
-to insert elements and remove an arbitrary element (chosen non-deterministically).
+for CAS-based implementations of lock-free bag data structures, which provide operations
+to insert an element or remove an arbitrary element (chosen non-deterministically).
 Since a queue trivially implements a bag, this lower bound also holds for queues.
@@
 
-In this paper, we exploit this loophole to show that it is in fact possible to implement a linearizable, lock-free queue
+In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
 whose step complexity does not depend linearly on $p$.

### 2022-12-17 podc_paper/related.tex
working on related work

@@
 attempts to reduce contention by grouping concurrent enqueues into baskets.
-An enqueue that fails its CAS is put in the basket with the eqnueue that succeeded.
+An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
 Enqueues within a basket determine their order among themselves without having to access the back of the queue.
@@
 
------
+\paragraph{Array-Based Queues}
+Arrays can be used to implement queues with bounded capacity, with indices
+for the front and back elements updated using CAS instructions 
+\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
+the capacity constraint by using a linked list of arrays.
+These solutions also suffer from the CAS retry problem, with $\Omega(p)$ steps per operation.
+
+Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
+To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch-and-increment instructions.
+If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
+the enqueues fall back on using a CAS to add a new array to the linked list, 
+and the CAS retry problem reappears.
+% I looked at the CRQ algorithm to implement each of the circular arrays.
+% Here is an execution that causes livelock for 3 processes
+% enqueue(A) reads t=0 and increments tail to 1
+% enqueue(B) reads t=1 and increments tail to 2
+% (*) dequeue reads h=0 and increments head to 1
+% dequeue's CAS succeds changing array[0] to (1,R,bottom)
+% dequeue reads t=2 at line 53
+% enqueue(A) fails CAS on array[0]
+% dequeue reads h=1 and increments head to 2
+% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
+% enqueue(B) fails CAS on array[1]
+% enqueue(A) reads t=2 and increments tail to 3
+% enqueue(B) reads t=3 and increments tail to 4
+% dequeue reads t=4 at line 53
+% go back to (*) and repeat this sequence
+Similarly,
+other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
+The fast path has good performance and the slow path guarantees
+termination.
+In worst-case executions, where processes must execute the slow path,
+these implementations also take $\Omega(p)$ steps per operation, 
+due either to the CAS retry problem or helping mechanisms.
+
+%Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a
+%wait-free queue that uses the fast-path slow-path methodology. Their
+%work is based on a circular queue using bounded memory. When a process
+%wishes to do an enqueue or a dequeue, it starts two paths. 
+% They show that these two paths do not affect each other
+%and the queue remains consistent. If a process makes no progress,
+%other processes help its slow path to finish. The helping phase
+%suffers from the CAS Retry Problem because processes compete in a
+%\nf{CAS} loop to decide which succeeds to help. Because of this, the
+%amortized complexity cannot be better than $\Omega(p)$.
 
@@
 
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10}
-introduced a new algorithm using a linked list of arrays. The queue is
-stored in a shared array where head and tail pointers point to the
-current elements in the queue. When the array is full, an empty array
-is linked to the array and tail pointers are updated. A global head
-points to the array containing the first element in the queue, and
-each process has a local head index that points to the first element
-in that array. Global tail and local tail pointers are similar. A
-process updates the position of the pointers after it does an
-operation. One process might go to sleep before setting the pointers,
-so the pointers might be behind their real places. They mention how to
-scan the arrays to update pointers while doing an operation. A process
-writes an element in the location head by a \nf{CAS} instruction, so
-if $p$ processes try to enqueue simultaneously, the amortized step
-(and \nf{CAS}) complexity remains $\Omega(p)$. Their design is
-lock-free but not wait-free.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a
-wait-free queue that uses the fast-path slow-path methodology
-introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their
-work is based on a circular queue using bounded memory. When a process
-wishes to do an enqueue or a dequeue, it starts two paths. The fast
-path  ensures good performance while the slow path ensures
-termination. They show that these two paths do not affect each other
-and the queue remains consistent. If a process makes no progress,
-other processes help its slow path to finish. The helping phase
-suffers from the CAS Retry Problem because processes compete in a
-\nf{CAS} loop to decide which succeeds to help. Because of this, the
-amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based
-queues also share
-it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free
-sublinear queue.
-
-\here{Check:
-John Giacomoni et al. FastForward
-https://dl-acm-org.ezproxy.library.yorku.ca/doi/10.1145/1345206.1345215
-}
-
-\paragraph{Other Primitives and Restricted Queues}
-
-David introduced the first sublinear-time queue
-\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
-It uses fetch\&increment and swap primitive instructions and takes constant time per operation, but
-uses unbounded memory.  A modification to use bounded space
-is described, but it increases the time per operation to $\Omega(p)$.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic
-queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it only works for a single dequeuer. 
-Our implementation uses their idea of having
-a tournament tree among processes to agree on the linearization of
-operations.
-
-\here{new result I found:}
-Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
-with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
-called half-increment and half-max, which can be viewed as particular kinds of
-double-word read-modify-write operations.
-They use this as evidence that their primitives can be more efficient than CAS
-since previous CAS-based queues all required $\Omega(p)$ step complexity.
-Our new implementation counters this argument.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
+\paragraph{Universal Constructions and Other Poly-log Time Data Structures}
 A \textit{universal construction} is an algorithm that can implement a
@@
 
+Need to discuss Fatourou and Kallimanis TCS 55(3)
+Fatourou and Kallimanis \cite{FK14} gave a universal construction based on fetch-and-add and CAS
+and used it to implement a queue....
+
 Ellen and Woelfel introduced an implementation of a Fetch\&Inc object
@@
 poly-logarithmic complexity.
+
+\paragraph{Restricted Queues}
+David introduced the first sublinear-time queue
+\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
+It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
+uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
+
+Jayanti and Petrovic introduced a wait-free poly-logarithmic
+queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
+Our implementation uses their idea of having
+a tournament tree among processes to agree on the linearization of
+operations.
+
+\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?
+May not be worth mentioning because then $p$ is a constant, so complexity in terms of $p$ doesn't make sense.}
+
+\paragraph{Other Primitives}
+Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
+with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
+called half-increment and half-max, which can be viewed as particular kinds of
+double-word read-modify-write operations.
+They use this as evidence that their primitives can be more efficient than CAS
+since previous CAS-based queues all required $\Omega(p)$ step complexity.
+Our new implementation counters this argument.

### 2022-12-17 podc_paper/introduction.tex
working on related work

@@
 The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
-Solutions that tried to sidestep this problem using fetch-and-increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
+Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
 rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.
@@
 Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
-lock-free queues use an unbounded number of CAS instructions in the worst case
-and $\Omega(p)$ CAS instructions in an amortized sense.
+lock-free queues use 
+%an unbounded number of CAS instructions in the worst case and 
+$\Omega(p)$ CAS instructions in an amortized sense.
 For the space-bounded implementation, we unlink unneeded memory from our data structure.
 We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
-garbage collector, such as the highly optimized one Java provides.
+garbage collector, such as the highly optimized one that Java provides.
 

### 2022-12-17 podc_paper/related.tex
working on related work

@@
 Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
-To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch-and-increment instructions.
+To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
 If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
@@
 
-\paragraph{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a
-shared version of any given sequential object. The first universal
-construction was introduced by
-Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent
-queue using a universal construction. Jayanti proved an $\Omega(\log
-p)$ lower bound on the worst-case shared-access time complexity of
-$p$-process universal
-constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that
-the universal construction by Afek, Dauber, and
-Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$
-worst-case step complexity, using atomic access to $\Omega(p \log
-p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal
-construction that achieves $\textsc{O}(\log^2 p)$ shared
-accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm
-cannot be used to create a queue. We mention a non-practical universal
-construction with a poly-log number of \nf{CAS} instructions in the
-last paragraph of page 13.
-
-Need to discuss Fatourou and Kallimanis TCS 55(3)
-Fatourou and Kallimanis \cite{FK14} gave a universal construction based on fetch-and-add and CAS
-and used it to implement a queue....
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object
-with step complexity of $O(\log p)$ using $O(\log n$)-bit
-\texttt{LL/SC} objects, where $n$ is the number of
-operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
-logarithmic complexity is to use a tree storing the Fetch\&Inc
-operations invoked by processes. When a process wants to do a
-Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number
-of elements in the tree. There are some similarities between designing
-a queue and a Fetch\&Inc object. A Fetch\&Inc object can be
-constructed from a queue. The algorithm by Ellen and Woelfel is
-interesting because of the similarities between Fetch\&Inc objects and
-queues. Also, it is one of the few wait-free data structures achieving
-poly-logarithmic complexity.
+\paragraph{Universal Constructions}
+One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
+Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
+the universal construction of Afek, Dauber, and
+Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
+using atomic access to $\Omega(p \log p)$-bit words. 
+(Thus, in terms of operations on reasonably-sized words, their construction would still take $\Omega(p)$ steps per operation.)
+Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
+to implement a queue, but its step complexity is also $\Omega(P)$.
+\here{Double check this.}
 
@@
 Our new implementation counters this argument.
+
+\paragraph{Fetch\&Increment Objects}
+Ellen and Woelfel introduced an implementation of a Fetch\&Increment object
+with step complexity of $O(\log p)$ using $O(\log n$)-bit
+\texttt{LL/SC} objects, where $n$ is the number of
+operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
+logarithmic complexity is to use a tree storing the Fetch\&Increment
+operations invoked by processes. When a process wants to do a
+Fetch\&Increment it adds its Fetch\&Increment to the tree and returns the number
+of elements in the tree. There are some similarities between designing
+a queue and a Fetch\&Increment object. A Fetch\&Increment object can be
+constructed from a queue. The algorithm by Ellen and Woelfel is
+interesting because of the similarities between Fetch\&Increment objects and
+queues. Also, it is one of the few wait-free data structures achieving
+poly-logarithmic complexity.

### 2022-12-18 podc_paper/related.tex
Edited related work on fetch&inc

@@
 Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
-using atomic access to $\Omega(p \log p)$-bit words. 
-(Thus, in terms of operations on reasonably-sized words, their construction would still take $\Omega(p)$ steps per operation.)
+assuming that words can store $\Omega(p \log p)$ bits. 
+(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
 Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
@@
 \paragraph{Fetch\&Increment Objects}
-Ellen and Woelfel introduced an implementation of a Fetch\&Increment object
-with step complexity of $O(\log p)$ using $O(\log n$)-bit
-\texttt{LL/SC} objects, where $n$ is the number of
-operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea to achieve
-logarithmic complexity is to use a tree storing the Fetch\&Increment
-operations invoked by processes. When a process wants to do a
-Fetch\&Increment it adds its Fetch\&Increment to the tree and returns the number
-of elements in the tree. There are some similarities between designing
-a queue and a Fetch\&Increment object. A Fetch\&Increment object can be
-constructed from a queue. The algorithm by Ellen and Woelfel is
-interesting because of the similarities between Fetch\&Increment objects and
-queues. Also, it is one of the few wait-free data structures achieving
-poly-logarithmic complexity.
+Progress has been made on obtaining sublinear-time implementations of other objects 
+with consensus number 2.
+Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
+implemented fetch\&increment objects from LL/SC objects
+using a polylogarithmic number of steps per operation.
+Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
+which uses a tournament tree to keep track of the fetch\&increment operations
+that have been performed.  However, our construction requires much more intricate
+data structures, since the state of a queue cannot easily be represented as succinctly
+as the single-word state of a fetch\&increment object.

### 2022-12-18 podc_paper/algorithm.tex
moved algorithm description into separate file algorithm.tex

--- 2022-12-15 podc_paper/podc-submission.tex
@@
-%%
-%% This is file generated with the docstrip utility from the source file: 
-%%
-%% samples.dtx  (with options: `acmsmall-conf')
-%% With some modifications for PODC 2023.
-%% For the copyright see the source file.
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall,nonacm,anonymous]{acmart}
-
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-\usepackage{showkeys}
-\renewcommand*\contentsname{Table of Contents}
-
-\usepackage{multicol}
-\setlength\columnsep{24pt}
-
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\newcommand{\here}[1]{\bf{[[#1]]}}
-
-\usepackage{amsmath,amsthm}
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}{Observation}
-%\theoremstyle{definition}
-%\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}{Invariant}
-%\newtheorem{proposition}[theorem]{Proposition}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{none}  %FIX FOR CAMERA READY
-%\copyrightyear{2023}
-%\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-\fancyfoot{} % REMOVE THIS LINE FOR CAMERA READY
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
-  2023}{Orlando, FL}
-%%
-%%  Uncomment \acmBooktitle if the title of the proceedings is different
-%%  from ``Proceedings of ...''!
-%%
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-%\acmPrice{15.00}
-%\acmISBN{978-1-4503-XXXX-X/18/06}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Eric Ruppert}
-\affiliation{%
-  \institution{York University}
-  \streetaddress{P.O. Box 1212}
-  \city{Toronto}
-  \state{Ontario}
-  \country{Canada}
-  \postcode{43017-6221}
-}
-\email{ruppert@cse.yorku.ca}
-\orcid{1234-5678-9012}
-
-\author{Hossein Naderibeni}
-\email{hnaderi268@gmail.com}
-\affiliation{%
-  \institution{York University}
-  \streetaddress{P.O. Box 1212}
-  \city{Toronto}
-  \state{Ontario}
-  \country{Canada}
-  \postcode{43017-6221}
-}
-
-\settopmatter{printacmref=false}  % REMOVE THIS LINE FOR CAMERA READY
-
-\begin{abstract}
-We introduce a novel linearizable wait-free queue implementation using single-word 
-CAS instructions.
-Previous lock-free queues all take $\Omega(p)$ steps per operation in the worst case, 
-where $p$ is the number of processes that can access the queue.
-We achieve $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue.
-\here{mention space usage; update time bound if space-efficient version takes more time per op}
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{concurrent data structures, wait-free queues}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-\input{introduction}
-\input{related}
+% !TEX root =  podc-submission.tex
 
@@
 
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
+We use a static {\it tournament tree}, which is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
+assigned to each process.
 
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
+There is a shared binary tree among the processes to agree on one
+total ordering of the operations invoked by processes. The tree is
+called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each
+process has a leaf in which the operations invoked by the process are
+stored in order. When a process wishes to do an operation it appends
+the operation to its leaf and tries to propagate its new operation up
+to the tree's root. Each node of the tree keeps an ordering of
+operations propagated up to it. All processes agree on the sequence of
+operations in the root and this ordering is used as the linearization
+ordering.
 
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
+Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node
+there is an ordering of operations stored. Each process tries to
+propagate its operations up to the root, which stores a total ordering
+of all operations.
+
+To propagate operations to a node $n$ in the tree, a process observes
+the operations in both of $n$'s children that are not already in $n$,
+merges them to create an ordering and then tries to append the
+ordering to the sequence stored in $n$. We call this procedure
+\tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on
+$n$ with a successful append helps to propagate their operations up to
+$n$. We shall prove that if a process invokes \tt{Refresh} on the node
+$n$ two times and fails to append the new operations to $n$ both
+times, the operations that were in $n$'s children before the first
+\tt{Refresh} are guaranteed to be in $n$ after the second failed
+\tt{Refresh}. 
 We sketch the argument here.
@@
 \hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
+\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before
+  and after a \nf{$n$.Refresh} with a successful append. Operations
+  propagating from the left child are labelled with $l$ and from the
+  right child with $r$.} 
 \end{center}
@@
   \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
+  \caption[Two consecutive failed \nf{Refresh}es by a
+    process.]{\label{fig::simpleDoubleRefresh}Time relations between
+    the concurrent successful \nf{Refresh}es and the two consecutive
+    failed \nf{Refresh}es.} 
 \end{figure}
 
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
+We use \nf{CAS} (Compare\&Swap) instructions to implement the
+\nf{Refresh}'s attempt to append  described in the previous
+paragraph. 
+The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
+successful \texttt{Refresh} that has read its information after the
+invocation of the first failed \nf{Refresh} (see Figure
+\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
+a successful append during $P$'s first failed attempt, and some
+process $K$ performs a \nf{Refresh} that reads its information after
+$L$'s append and then performs a successful append during $P$'s second
+failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
+new operations that were in $n$'s children before $P$'s first failed
+\nf{Refresh}, in case they were not already appended. After a process
+appends its operation into its leaf it can call \nf{Refresh} on the
+path up to the root at most two times on each node. So, with $O(\log
+p)$ \nf{CAS}es an operation can ensure it appears in the
+linearization. This cooperative solution allows us to overcome the CAS
+Retry Problem.
 
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
+It is not efficient to explicitly store the sequence of operations in
+each node because each operation would have to be copied all the way
+up to the root; doing this would not be possible in poly-logarithmic
+time. 
+Instead we use an implicit representation of the operations propagated
+together. Furthermore, we do not need to maintain an ordering on
+operations propagated together in a node until they have reached the
+root. It is sufficient to only keep track of sets of operations
+propagated together in each \nf{Refresh} and then define the
+linearization ordering only in the root (see Figure
+\ref{fig::set}). Achieving a constant-sized implicit representation of
+operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions
+on fixed-size objects in each \tt{Refresh}. To do that, we introduce
+\it{block}s. A block stores information about the operations
+propagated by a \nf{Refresh}. It contains the number of operations
+from the left and the right child propagated to the node by the
+\texttt{Refresh} procedure. See Figure~\ref{fig:block} for an
+example. A node stores an array of blocks of operations propagated up
+to it. A propagate step  aggregates the new blocks in the children
+into a new block and stores the new block in the parent. We call the
+aggregated blocks \it{subblocks} of the new block and the new block
+the \it{superblock} of them.  In each \nf{Refresh} there is at most
+one operation from each process trying to be propagated, because one
+process cannot invoke two operations concurrently. Thus, there are at
+most $p$ operations in a block. Furthermore, since the operations in a
+\texttt{Refresh} step are concurrent we can linearize them  among
+themselves in any order we wish, because if two operations are read in
+one successful \tt{Refresh} step in a node they are going to be
+propagated up to the root together. Our choice is to put the
+operations propagated from the left child before the operations
+propagated from the right child. In this way, if we know the number of
+operations from the left child and the number of operations from the
+right child in a block, we have a complete ordering of the
+operations.
 
@@
 \end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
+\caption[The operations merged in a \nf{Refresh} step with
+  sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from
+  left to right. In each internal node one can arbitrarily linearize
+  the sets of concurrent operations  propagated together in a
+  \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated
+  together in one \nf{Propagate} step and they will be propagated up
+  to the root together. Since their execution time intervals overlap,
+  they can be linearized in any order.} 
 \end{center}
@@
 \end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
+\caption[The operations merged in a \nf{Refresh} step with (left,
+  right) blocks.]{\label{fig:block} Using blocks to represent
+  operations. Blocks between two lines $||$ are propagated together to
+  the parent. Each block consists of a pair (left, right) indicating
+  the number of operations from the left and the right child,
+  respectively. For example, (12,6) in the root contains  (10,2) from
+  the left child and (6,0) from the right child. The third block in
+  the root (8,36) is created by merging (5,3) from the left child and
+  (14,6) and (0,16) from the right child. (5,3) is the superblock of
+  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).} 
 \end{figure}
 
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
+So far, we have a shared tree that processes use to agree on the
+implicit ordering stored in its root. With this agreement on the
+linearization ordering, we can design a universal construction; for a
+given object, we can perform an operation $op$ by applying all the
+operations up until $op$ in the root on a local copy of the object and
+then returning the response for $op$. However, this approach is not
+enough for an efficient queue. 
+We show that we can build an efficient queue if we can compute two
+things about the ordering in the root: (1) the $i$th propagated
+operation and (2) the rank of a propagated operation in the
+linearization. We explain how to implement (1) and (2) in
+poly-logarithmic steps.
 
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
+After propagating an operation \textit{op} to the root, processes can
+find out information about the linearization ordering using (1) and
+(2).  
+To get the $i$th operation in the root, we find the block $B$
+containing the $i$th operation in the root, and then recursively find
+the subblock of $B$ in the descendent of the root that contains that
+$i$th operation. When we reach a block in a leaf, the operation is
+explicitly stored there. To make this search faster, instead of
+iterating over all blocks in the node, we store the prefix sum of the
+number of elements in the blocks sequence to permit a binary search
+for the required block. We also store pointers to determine the range
+of subblocks of a block to make the binary search faster. In each
+block, we store the prefix sum of operations from the left child and
+from the right child. Moreover, for each block, we store two
+attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of
+the last left and right subblock (see Figure \ref{fig::pointer}). We
+know a block size is at most $p$, so binary search takes at most
+\textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and
+\nf{end\sub{right}} indices of a block and its previous block reduce
+the search range size to \textsc{O}$(p)$.
 
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
+To compute the rank in the root of an operation in a leaf, we need to
+find the superblock of the block that the operation is in. After a
+block is installed in a node we store the approximate index of its
+superblock in it to make this faster.
 
@@
   \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
+  \caption{Each block stores the index of its last subblock in each
+    child. \label{fig::pointer}} 
 \end{figure}
 
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
+In an execution on a queue where no dequeue operation returns
+\nf{null}, the $k$th dequeue returns the argument of the $k$th
+enqueue. In the general case a dequeue returns \nf{null} if and only
+if the queue size after the previous operation is 0. We refer to such
+a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null
+dequeue, it returns the argument of the $k$th enqueue. Having the size
+of the queue after an operation we can compute the number of non-null
+dequeues from the number of enqueues before the operation. So, if we
+store the size of the queue after each block of operations in the
+root, we can compute the index of the enqueue whose argument is the
+response to a given dequeue in constant time.
 
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
+In our case of implementing a queue, a process only needs to compute
+the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific
+position.  We know we can linearize operations in a block in any
+order; here, we choose to put \nf{Enqueue} operations in a block
+before \nf{Dequeue} operations. Consider the following operations,
+where operations in a cell are concurrent. 
 \begin{table}[hbtp]
@@
 
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
+\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2,
+  1, 3, 4, null}, respectively. 
+Now, we claim that by knowing the size of the queue, we can compute
+the rank of the required \texttt{Enqueue} for any non-null
+\nf{Dequeue}. We apply this approach to blocks; if we store the size
+of the queue after each block of operations, we can compute the index
+of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
 
@@
 
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
+\noindent The size of the queue after the $b$th block in the root
+could be computed as 
+$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} +
+\#\textrm{\tt{Enqueue}s in }b\textrm{th block} -
+\#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$ 
+\noindent Moreover, the total number of non-null dequeues in blocks
+$1,2,...,b$ in the root is 
+$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} -
+\textrm{size after }b\textrm{th block}.$$
 
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
+Given a \texttt{Dequeue} is in block $B$, its response is the argument
+of the \tt{Enqueue} whose rank is 
 \begin{center}
@@
 \subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
+Section \ref{algQ} gives the pseudocode for the queue
+implementation. It uses the following two types of objects. 
 \paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
+ In each \texttt{Node} we store pointers to its parent and its left
+ and right child, an array of \nf{Block}s called \texttt{block}s and
+ the index \nf{head} of the first empty entry in \texttt{blocks}.
 
 \paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
+ The information stored in a \nf{Block} depends on whether  the
+ \nf{Block} is in an internal node or a leaf. If it is in a leaf, we
+ use a \nf{LeafBlock} which stores one operation. If a block $B$ is in
+ an internal node $n$, then it contains subblocks in the left and
+ right children of $n$. The left subblocks of $B$ are some consecutive
+ blocks in the left child of $n$ starting from where the left
+ subblocks of the block prior to $B$ ended. The right subblocks of $B$
+ are defined similarly. In each \texttt{block} we store four essential
+ fields that implicitly summarize which operations are in the block
+ \texttt{sum\textsubscript{enq-left}},
+ \texttt{sum\textsubscript{deq-left}},
+ \texttt{sum\textsubscript{enq-right}},
+ \texttt{sum\textsubscript{deq-right}}. The
+ \texttt{sum\textsubscript{enq-left}} field is the total number of
+ \nf{Enqueue} operations in the blocks before the last subblock of $B$
+ in the left child. The other fields' semantics are similar. The
+ \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of
+ the last subblock of a block in the left and the right child,
+ respectively. The approximate index of the superblock of non-root
+ blocks is stored in their \nf{super} field. The \texttt{size} field
+ in a block in the root node stores the size of the queue after the
+ operations in the block have been performed.
 
@@
 \paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
+An \nf{Enqueue} operation does not return a response, so it is
+sufficient to propagate the \nf{Enqueue} operation to the root and
+then use its position in the linearization for future \nf{Dequeue}
+operations. \nf{Enqueue($e$)} creates a \nf{LeafBlock} with
+$\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}}
+fields and then appends it to the tree.
 
 \paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
+\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and
+\nf{sum\sub{deq}} fields and appends it to the tree. Then, it computes
+the position of the appended \nf{Dequeue} operation in the root using
+\nf{IndexDequeue} and after that finds the response of the
+\nf{Dequeue} by calling \nf{FindResponse}.
 
 \paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
+The \nf{head} field is the index of the first empty slot in
+\nf{blocks} in a \nf{LeafBlock}. There are no multiple write accesses
+on \nf{head} and \nf{blocks} in a leaf because only the process that
+the leaf belongs to appends to it. \nf{Append($B$)} adds $B$ to the
+end of the \nf{blocks} field in the leaf, increments \nf{head} and
+then calls \nf{Propagate} on the leaf's \nf{parent}. When
+\nf{Propagate} terminates, it is guaranteed that the appended block is
+a subblock of a block in the \nf{root}.
 
 \paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
+\nf{Propagate} on node $n$ uses the double refresh idea described
+earlier and invokes two \nf{Refresh}es on $n$ in Lines
+\ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes
+\nf{Propagate} on \nf{$n$.parent} recursively until it reaches the
+root.
 
 \paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
+The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s
+children's new blocks and append it to $n$\nf{.blocks}. The variable
+\nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new
+block created by \nf{Refresh} will be inserted into
+$n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1}
+of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance}
+increments the children's \nf{head} if necessary and sets the
+\nf{super} field of their most recently appended blocks. The reason
+behind this helping is explained later when we discuss
+\nf{IndexDequeue}. After helping to \nf{Advance} the children, a new
+block called \nf{new} is created in Line
+\ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh}
+returns \nf{true} because there are no new operations to propagate,
+and it is unnecessary to add an empty block to the tree. Later we will
+use the fact that all blocks contain at least one operation. Line
+\ref{cas} tries to install \nf{new}. If it was successful, all is
+good. If not, it means someone else has already put a block in
+$n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance
+$n$\nf{.head} to \nf{h+1} and update the \nf{super} field of
+$n$\nf{.blocks[h]} at Line \ref{advance}.
 
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
+\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used
+by \nf{Refresh} to construct a block containing new operations of
+$n$'s children. 
+The block \nf{new} is created in Line \ref{initNewBlock} and its
+fields are filled similarly for both left and right directions. The
+variable  \nf{index\sub{prev}} is the index of the block preceding the
+first subblock in the child in direction \nf{dir} that is aggregated
+into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the
+index of the rightmost subblock of \nf{new} in the child. Then
+\nf{sum\sub{enq-dir}} is computed from the sum of the number of
+\nf{Enqueue} operations in the \nf{new} block from direction \nf{dir}
+and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The
+field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new}
+is going to be installed in the \nf{root}, the \nf{size} field is also
+computed.
 
@@
 
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
+A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank
+within the block in the root of the $i$th \nf{Dequeue} of the $b$th
+block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$
+that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let
+$par$ be $n$\nf{.parent}. Without loss of generality, assume for the
+rest of this section that ${n}$ is the left child of $par$. Let
+$R_{par}$ be the first successful \nf{$par$.Refresh} that reads some
+value greater than ${b}$ for \nf{left.head} and therefore contains
+${B}$ as a subblock of its created block in Line
+\ref{invokeCreateBlock}. Let $j$ be the index of the block that
+$R_{par}$ puts in $par$\nf{.blocks}.
 
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
+Since the index of the superblock of ${B}$ is not known until $B$ is
+propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while
+creating $B$. One approach for $R_{par}$ is to set the \nf{super}
+field of ${B}$ after propagating $B$ to $par$. This solution would not
+be efficient because there might be up to $p$ subblocks that $R_{par}$
+propagated, which need to update their \nf{super} field. However,
+intuitively, once $B$ is installed, its superblock is going to be
+close to \nf{$n$.parent.head} at the time of installation. If we know
+the approximate position of the superblock of $B$ then we can search
+for the real superblock when it is needed. Thus, \nf{$B$.super} does
+not have to be the exact location of  the superblock of $B$, but we
+want it to be close to $j$. We can set \nf{$B$.super} to
+\nf{$par$.head} while creating $B$, but the problem is that there
+might be many \nf{Refresh}es on $par$ that could happen after $R_n$
+reads $par$\nf{.head} and before propagating $B$ to $par$. If $R_n$
+sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to
+\nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at
+some time after installing $B$ and before setting \nf{$B$.super}. In
+this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the
+value of \nf{$B$.super}.
 
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
+Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}. After
+appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call
+to \nf{Advance} from \nf{$n$.Refresh} by the same or another process
+or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}. We shall
+show that this is sufficient to ensure that \nf{$B$.super} differs
+from the index of $B$'s superblock by at most~1.
 
 \paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
+To compute the response of the $i$th \nf{Dequeue} in the $b$th block
+of the root Line \ref{checkEmpty} computes whether the queue is empty
+or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue
+would become empty before the requested \nf{Dequeue}. If the queue is
+not empty, Line \ref{computeE} computes the rank $e$ of the
+\nf{Enqueue} whose argument is the response to the
+\nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the
+root (which is before the $b$th block), we find the block and position
+containing the \nf{Enqueue} operation using \nf{DoublingSearch} and
+after that \nf{GetEnqueue} finds its \nf{element}.
 
 \paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\section{Proof of Correctness}
-
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
-
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
-
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
-
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$ the number of concurrent processes, is $O(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\newpage
-
-\bibliographystyle{plain}
-\bibliography{sample-base.bib}
-
-\appendix
-\end{document}
+We can describe an operation in a node in two ways: the rank of the
+operation among all the operations in the node or the index of the
+block containing the operation in the node and the rank of the
+operation within that block. If we know the block and rank within the
+block of an operation, we can find the subblock containing the
+operation and the operation's rank within that subblock in poly-log
+time. To find the response of a \nf{Dequeue}, we know about the  rank
+of the response \nf{Enqueue} in the root (\nf{e} in Line
+\ref{computeE}). 
+We also know the \nf{e}th \nf{Enqueue} is in
+\nf{root.blocks[1..end]}. \nf{DoublingSearch} uses doubling to find
+the range that contains the answer block (Lines
+\ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the
+required indices with a binary search (Line
+\ref{dsearchBinarySearch}). 
+A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the
+$i$th enqueue in the $b$th block of $n$. The range of subblocks of a
+block is determined using the \nf{end\sub{left}} and
+\nf{end\sub{right}} fields of the block and its previous block. Then,
+the subblock is found using binary search on the \nf{sum\sub{enq}}
+field (Lines \ref{leftChildGet} and \ref{rightChildGet}).

### 2022-12-18 podc_paper/podc-submission.tex
working on related work
got rid of sample .bib entries
moved algorithm description into separate file algorithm.tex

@@
 \newcommand{\here}[1]{\bf{[[#1]]}}
+\newcommand{\ceil}[1]{\lceil #1 \rceil}
 
@@
 \input{related}
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
+\input{algorithm}
 
@@
 
+Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.
+
 \begin{figure}[hbt]  
@@
 \bibliographystyle{plain}
-\bibliography{sample-base.bib}
+\bibliography{queues.bib}
 

### 2022-12-18 podc_paper/related.tex
moved algorithm description into separate file algorithm.tex

@@
 which uses a tournament tree to keep track of the fetch\&increment operations
-that have been performed.  However, our construction requires much more intricate
-data structures, since the state of a queue cannot easily be represented as succinctly
+that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
+However, our construction requires more intricate
+data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
 as the single-word state of a fetch\&increment object.

### 2022-12-18 podc_paper/algorithm.tex
started working on description of the algorithm

@@
 
-We use a static {\it tournament tree}, which is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
+We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
+The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
 assigned to each process. 
+Each node of the tree stores an array of \emph{blocks} and each \emph{block} represents a 
+sequence of enqueues and a sequence of dequeues.
+To perform an operation on the queue, a process $P$ creates a block containing that single 
+operation and appends it to the array in $P$'s leaf.
+Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
+At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
+block and attempts to append this block to $v$'s array using a CAS instruction.
+This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
+then its operation has been propagated to $v$ by some other process.
+We define a total order on all operations that have been propagated to the root, which is used to
+linearize the operations.
+In this section, we assume for simplicity that each node has an infinite blocks array.
+We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.
 
-There is a shared binary tree among the processes to agree on one
-total ordering of the operations invoked by processes. The tree is
-called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each
-process has a leaf in which the operations invoked by the process are
-stored in order. When a process wishes to do an operation it appends
-the operation to its leaf and tries to propagate its new operation up
-to the tree's root. Each node of the tree keeps an ordering of
-operations propagated up to it. All processes agree on the sequence of
-operations in the root and this ordering is used as the linearization
-ordering.
+If $P$'s operation is an enqueue, it can terminate when the operation has been propagated to the root.
+If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
+dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
+(since the operation may have been propagated to the root by some other process).
+Then, $P$ determines whether the queue is empty when its dequeue is linearized. If so, it returns \nil.
+If not, $P$ computes the rank $r$ of its dequeue in the linearization ordering
+among all dequeues that do not return \nil,
+finds the $r$th enqueue in the linearization and returns that enqueue's value.
 
-Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node
-there is an ordering of operations stored. Each process tries to
-propagate its operations up to the root, which stores a total ordering
-of all operations.
+The primary challenge is thus figuring out what information to store in each block so that 
+the following tasks can be done efficiently (in a polylogarithmic number of steps).
+\begin{itemize}
+\item
+Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children.
+\item
+Given an operation in a leaf that has been propagated to the root, find that operation's location in the root's blocks array.
+\item
+Given the location of a dequeue operation in the root's blocks array, find the enqueue whose value it should return (or determine that the queue is empty when the dequeue is linearized).
+\end{itemize}
+
+---------
 

### 2022-12-19 podc_paper/algorithm.tex
Working on conclusion section

@@
 assigned to each process. 
-Each node of the tree stores an array of \emph{blocks} and each \emph{block} represents a 
+Each node of the tree stores an array of \emph{blocks}, where each block represents a 
 sequence of enqueues and a sequence of dequeues.
@@
 \end{itemize}
+
+\here{At what point would a figure illustrating of the tournament tree be useful?}
 

### 2022-12-19 podc_paper/conclusion.tex
Working on conclusion section

% !TEX root =  podc-submission.tex

\section{Future Directions}

Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
However, our approach does have a higher cost in the best case (for example, when an operation
runs in isolation).
It may be possible to make our queue adaptive by having an operation capture a starting node
in the tournament tree (as in \cite{??}) rather than having a statically assigned leaf as its starting point.
Another way to make our queue more practical might be to use it as the slow path in the
fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
get a queue that has good performance in practice while also having good worst-case step complexity.

There is a gap between our implementation, which takes $O(\log^2 p + \log q)$ steps per operation,
and Attiya and Fouren's $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17}.
It would be interesting to determine how the true step complexity of lock-free queues (or, more generally, bags)
depends on $p$.

We believe the approach used here to implement a lock-free queue 
could also be applied to obtain other lock-free
data structures with a polylogarithmic number of steps per operation.
For example, we could implement a kind of vector data structure that stores a sequence and
provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
\opa{get}{i} to read the \var{i}th element in the sequence, and
\opa{index}{e} to compute the position of element \var{e} in the sequence.
\here{Can omit rest of this paragraph to save space}
An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{get}{i} calls \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array, taking $O(\log n + \log^2p)$ when the vector has $n$ elements.  
An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and takes $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy}.
As future work we would like to investigate whether a similar approach could be used for stacks or deques.

### 2022-12-19 podc_paper/podc-submission.tex
started working on description of the algorithm
Working on conclusion section

@@
 \newcommand{\ceil}[1]{\lceil #1 \rceil}
+\newcommand{\nil}{nil}
+\newcommand{\opa}[2]{\nf{#1(#2)}}
+\newcommand{\op}[1]{\nf{#1}}
+\newcommand{\var}[1]{\nf{#1}}
 
@@
 
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
+\section{Reducing Space}
+\label{reducing}
 \paragraph{Reducing Space Usage}
@@
 
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$ the number of concurrent processes, is $O(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
+\input{conclusion}
 

### 2022-12-19 podc_paper/algorithm.tex
working on algorithm

@@
 
+We begin with an overview of the algorithm.
 We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
@@
 To perform an operation on the queue, a process $P$ creates a block containing that single 
-operation and appends it to the array in $P$'s leaf.
+operation and appends it to the blocks array in $P$'s leaf.
 Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
@@
 In this section, we assume for simplicity that each node has an infinite blocks array.
-We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.
+We describe in Section \ref{reducing} how to replace the infinite array by a representation that uses bounded space.
 
@@
 \end{itemize}
+
+An operation can only terminate after it a block containing it has been appended to the root's blocks array.
+So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
+$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
+Our linearization orders operations according to the block they belong to in the root's blocks array.
+Operations that appear in the same block are necessarily concurrent, so they can be ordered arbitrarily.
+We next describe how to order them so that computing results of dequeue operations is efficient.
+
+Each block $B$ in a node $v$ represents several consecutive blocks in each of $v$'s children.
+The blocks in the children are called the \emph{direct subblocks} of $B$.
+A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
+or a subblock of a direct subblock of $B$.
+Recall that blocks in the leaves contain a single operation.
+A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
+We order the operations within a block as follows.
 

### 2022-12-19 podc_paper/related.tex
working on algorithm

@@
 \paragraph{Array-Based Queues}
-Arrays can be used to implement queues with bounded capacity, with indices
-for the front and back elements updated using CAS instructions 
-\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
+%Similar to the front and back pointers of list-based queues, indices
+Dequeues and enqueues update
+indexes of the front and back elements using CAS instructions.
 Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
 the capacity constraint by using a linked list of arrays.
-These solutions also suffer from the CAS retry problem, with $\Omega(p)$ steps per operation.
+These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.
 
@@
 Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
-to implement a queue, but its step complexity is also $\Omega(P)$.
+to implement a queue, but its step complexity is also $\Omega(p)$.
 \here{Double check this.}
@@
 \paragraph{Fetch\&Increment Objects}
-Progress has been made on obtaining sublinear-time implementations of other objects 
+There are sublinear-time implementations of another object
 with consensus number 2.
@@
 Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
-which uses a tournament tree to keep track of the fetch\&increment operations
+which uses a tournament tree to keep track of the fetch\&increments
 that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.

### 2022-12-19 podc_paper/garbage-collection.tex
init garbage_collection, add pdf

--- 2022-12-01 garbage_collect/main.tex
@@
 
-\section{Feilds}
 \begin{algorithm}
@@
 
-\section{Queue}
 \begin{algorithm}
@@
 \State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State {\color{red}\tt{<x, y>= root.\Call{DoublingSearch}{e, b}}}
+\State {\color{red}\tt{<x, y>= root.\Call{BinarySearch}{e}}}
 \State {\color{red}\tt{lastDequeudFrom.update(x)}}
@@
 
-\section{Search+Append}
 \begin{algorithm}
@@
 \begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end} \cmt{{\color{red} I think with garbage collection, doubling search is not needed any more and a binary search on \nf{root.blocks} would be good enough.}}
-% \State \tt{start= end-1} \label{dsearchStart} 
-% \While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-% \State \tt{start= max(start-(end-start), 0)} \label{doubling}
-% \EndWhile \label{dsearchEnd}
-% \State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-% \State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-% \State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
 \caption{\tt{\sl{Leaf}}}
@@
 
-\section{Propagate}
 \begin{algorithm}
@@
 \State \tt{h= \head} \label{readHead}
+{\color{red}\If{\tt{h\%p\textsuperscript{2}=0}}
+\State \Call{Help()}{}
+\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
+\EndIf}
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
@@
 
-\section{Index+Get}
-
 \begin{algorithm}
@@
 \color{red}
-\section{MaxByProcess}
 
@@
 
-\section{Help}
-
 \begin{algorithm}
@@
 
-\section{FreeMemory}
-
 \begin{algorithm}
@@
 \end{algorithm}
-
-\section{PBRT}
 
@@
 
-\Statex
-\Function{}{TryAppend}{\sl{block} \tt{B}, \sl{int} \tt{i}}
-\Statex {\color{red} Tries to append \nf{B} with key \nf{i} to the red-black tree.}
-{\color{red}\If{\tt{i\%p\textsuperscript{2}=0}}
-\State \Call{Help()}{}
-\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
-\State \tt{garbageCollectRound=floor(root.head/p\textsuperscript{2})}
-\EndIf}
-\EndFunction{FreeMemory}
-
 \end{algorithmic}
@@
 \pagebreak
+\clearpage
 \section{Description}
@@
 \color{black}
-In our original algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
+In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
 This makes the space used by the algorithm factor of the number of operations of invoked on the queue.
-In this section here we want to free the memory allocated by the operations that are no longer needed and make the space used polynomial of $p+q$.
+Here in this section, we want to add a mechanism to free the memory allocated by the operations that are no longer needed and make the overall space used polynomial of $p+q$.
 
-One way of handling garbage collection is to reallocate the space taken by each operation right after it is not needed anymore, but it might cost too much to do that.
-So we garbage collect by batching.
-If we garbage collect the unnecessary blocks in a node every $p^2$ block appended to the node, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$.
+It is a common way to do garbage collection in batches.
+If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.
 
-In our design a process garbage collects a node when it attempts to append a block in the $kp^2$ position in the.
-\nf{Garbage Collect} corresponded to the $kp^2$th block in node $n$ is called the $k$th round of garbage collect on $n$.
-Every $p^2$  block appended to a node at least one garbage collect happens because the node's \nf{head} cannot advance until the garbage collect is done (see Lines ).
+In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}).
+Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).
 
 \begin{lemma}
-The number of the blocks in the \nf{blocks} is $O(p^2+q)$.
+Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
 \end{lemma}
 
-An \nf{Enqueue} operation can be removed from the tree after its \nf{element} has been computed to be the response of some \nf{Dequeue}.
-It is safe to remove a \nf{Dequeue} operation from the tree after the \nf{Dequeue} is terminated, but a \nf{Dequeue} might sleep for a long time and prevents to be garbage collected.
-In that situation other processes can help the \nf{Dequeue} to compute its response and wtite down its response somewhere.
-If we remove a \nf{Dequeue} from the tree after it is helped the \nf{Dequeue} can read the helped response writtern when it encountered a problem (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}).
-A \nf{Dequeue} operation can be removed from the tree after its response has been computed.
-We say a block  is \it{finished} if all of its operations can be removed.
+\begin{lemma}
+Total number of the blocks in the tree is $O(p^3+pq)$.
+\end{lemma}
+
+\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$ or $e.$\nf{element} has been computed to be $RESP(d)$.
+It is safe to remove \nf{Dequeue} $d$o from the tree after the $d$ is terminated or $Resp(d)$\nf{.element} has been computed.
+We can remove a block after the told conditions are satisfied for all ofits operations.
+A \nf{Dequeue} may go to sleep for a long time and prevent the \nf{block} to be removed.
+In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere.
+After writing down the response of th \nf{Dequeue} it is safe to remove.
+Because if the \nf{Dequeue} failed to compute its response it can read the response written (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}).
+
+\begin{definition}
+A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
+\end{definition}
+
+\begin{corollary}
+If a block is finished, then all of its subblocks are also finished.  
+\end{corollary}
 
 \begin{lemma}
-There are at most $O(p^2+q)$ unfinished blocks in a node's \nf{blocks} at a time.
+It is safe to remove a finished block from the tree.
 \end{lemma}
 
-If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means $1$ to $i-1$th \nf{Enqueue} operations are also dequeued.
+If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued.
 This gives us the idea that if a block is finished then all the blocks before it are also finished.
@@
 
+\begin{lemma}
+  The amortized step per process for \nf{GarbageCollect} is $O()$.
+\end{lemma}
+
+\begin{lemma}
+  Algorithm is wait-free and linearizable.
+\end{lemma}
+
 \end{document}

### 2022-12-19 podc_paper/algorithm.tex
working on algorithm description

--- 2022-12-18 podc_paper/algorithm.tex
@@
 
+\subsection{Overview of Implementation}
 We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
@@
 assigned to each process. 
-Each node of the tree stores an array of \emph{blocks} and each \emph{block} represents a 
+Each tree node  stores an array of \emph{blocks}, where each block represents a 
 sequence of enqueues and a sequence of dequeues.
+In this section, we assume for simplicity that each node has an infinite blocks array.
+Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.
+
 To perform an operation on the queue, a process $P$ creates a block containing that single 
-operation and appends it to the array in $P$'s leaf.
-Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
-At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
-block and attempts to append this block to $v$'s array using a CAS instruction.
-This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
-then its operation has been propagated to $v$ by some other process.
-We define a total order on all operations that have been propagated to the root, which is used to
+operation and appends it to the blocks array in $P$'s leaf.
+Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
+We shall define a total order on all operations that have been propagated to the root, which is used to
 linearize the operations.
-In this section, we assume for simplicity that each node has an infinite blocks array.
-We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.
 
-If $P$'s operation is an enqueue, it can terminate when the operation has been propagated to the root.
+To propagate an operation to a node $v$ in the tree, $P$ first observes
+the blocks in both of $v$'s children that are not already in $v$,
+creates a new block by combining information from those blocks, and attempts to append this 
+new block to $v$'s blocks array using a CAS instruction.
+Following \cite{}, we call this a 3-step sequence a
+\op{refresh} on $v$. %(see Figure \ref{fig::propagstep}).
+A successful \op{refresh} by $P$ may propagate pending operations by several other processes to $v$.
+This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
+new block into $v$'s array twice,
+then its operation has been propagated to $v$ by some other process, so $P$ can continue 
+onwards towards the root.
+
+Now suppose $P$'s operation has been propagated all the way to the root.
+If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
 If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
@@
 the following tasks can be done efficiently (in a polylogarithmic number of steps).
-\begin{itemize}
+\begin{enumerate}[label={(T\arabic*)}]
 \item
-Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children.
+\label{construct}
+Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{refresh}.
 \item
-Given an operation in a leaf that has been propagated to the root, find that operation's location in the root's blocks array.
+\label{findinroot}
+Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
 \item
-Given the location of a dequeue operation in the root's blocks array, find the enqueue whose value it should return (or determine that the queue is empty when the dequeue is linearized).
-\end{itemize}
+\label{findrank}
+Given a dequeue's position in the root, determine that the queue is empty when it occurs
+or determine the rank $r$ of the enqueue whose value it should return.
+\item
+\label{findenqueue}
+Find the $r$th enqueue in the linearization ordering.
+\end{enumerate}
+Since these tasks require knowing the linearization ordering, we next describe that ordering.
 
----------
+\subsection{Linearization Ordering}
 
-To propagate operations to a node $n$ in the tree, a process observes
-the operations in both of $n$'s children that are not already in $n$,
-merges them to create an ordering and then tries to append the
-ordering to the sequence stored in $n$. We call this procedure
-\tt{$n$.Refresh} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on
-$n$ with a successful append helps to propagate their operations up to
-$n$. We shall prove that if a process invokes \tt{Refresh} on the node
-$n$ two times and fails to append the new operations to $n$ both
-times, the operations that were in $n$'s children before the first
-\tt{Refresh} are guaranteed to be in $n$ after the second failed
-\tt{Refresh}. 
-We sketch the argument here.
+An operation can terminate only after a block containing it has been appended to the root's blocks array.
+So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
+$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
+Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
+Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.
 
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before
-  and after a \nf{$n$.Refresh} with a successful append. Operations
-  propagating from the left child are labelled with $l$ and from the
-  right child with $r$.} 
-\end{center}
-\end{figure}
+Each block in a leaf represents a single operation.
+Each block $B$ in an internal node $v$ results from merging
+several consecutive blocks from each of $v$'s children.
+The blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
+A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
+or a subblock of a direct subblock of $B$.
+A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
 
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a
-    process.]{\label{fig::simpleDoubleRefresh}Time relations between
-    the concurrent successful \nf{Refresh}es and the two consecutive
-    failed \nf{Refresh}es.} 
-\end{figure}
+The operations propagated by a \op{refresh} are all pending when the \op{refresh} occurs,
+so there is at most one operation per process.
+Hence, a block represents at most $p$ operations in total.  
+Moreover, we never append empty blocks, so 
+each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.
 
-We use \nf{CAS} (Compare\&Swap) instructions to implement the
-\nf{Refresh}'s attempt to append  described in the previous
-paragraph. 
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
-successful \texttt{Refresh} that has read its information after the
-invocation of the first failed \nf{Refresh} (see Figure
-\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
-a successful append during $P$'s first failed attempt, and some
-process $K$ performs a \nf{Refresh} that reads its information after
-$L$'s append and then performs a successful append during $P$'s second
-failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
-new operations that were in $n$'s children before $P$'s first failed
-\nf{Refresh}, in case they were not already appended. After a process
-appends its operation into its leaf it can call \nf{Refresh} on the
-path up to the root at most two times on each node. So, with $O(\log
-p)$ \nf{CAS}es an operation can ensure it appears in the
-linearization. This cooperative solution allows us to overcome the CAS
-Retry Problem.
+As mentioned above, we are free to order operations within a block however we like.
+For convenience, we order the enqueues and dequeues separately, and put the 
+operations that were propagated from the left child before the operations from the right child.
+More formally, we inductively define an sequence $E(B)$ of the enqueues represented by a block $B$.
+If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is either the single
+enqueue represented by the block (if $B$ stores an enqueue), or the empty sequence (if $B$ stores a dequeue).
+If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
+and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s enqueue sequence is the concatenation $E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)$.
+The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.
 
-It is not efficient to explicitly store the sequence of operations in
-each node because each operation would have to be copied all the way
-up to the root; doing this would not be possible in poly-logarithmic
-time. 
-Instead we use an implicit representation of the operations propagated
-together. Furthermore, we do not need to maintain an ordering on
-operations propagated together in a node until they have reached the
-root. It is sufficient to only keep track of sets of operations
-propagated together in each \nf{Refresh} and then define the
-linearization ordering only in the root (see Figure
-\ref{fig::set}). Achieving a constant-sized implicit representation of
-operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions
-on fixed-size objects in each \tt{Refresh}. To do that, we introduce
-\it{block}s. A block stores information about the operations
-propagated by a \nf{Refresh}. It contains the number of operations
-from the left and the right child propagated to the node by the
-\texttt{Refresh} procedure. See Figure~\ref{fig:block} for an
-example. A node stores an array of blocks of operations propagated up
-to it. A propagate step  aggregates the new blocks in the children
-into a new block and stores the new block in the parent. We call the
-aggregated blocks \it{subblocks} of the new block and the new block
-the \it{superblock} of them.  In each \nf{Refresh} there is at most
-one operation from each process trying to be propagated, because one
-process cannot invoke two operations concurrently. Thus, there are at
-most $p$ operations in a block. Furthermore, since the operations in a
-\texttt{Refresh} step are concurrent we can linearize them  among
-themselves in any order we wish, because if two operations are read in
-one successful \tt{Refresh} step in a node they are going to be
-propagated up to the root together. Our choice is to put the
-operations propagated from the left child before the operations
-propagated from the right child. In this way, if we know the number of
-operations from the left child and the number of operations from the
-right child in a block, we have a complete ordering of the
-operations.
+Finally, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
+linearization ordering is 
+$E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k)$.
+
+\here{Need some figures illustrating tournament tree, ordering of operations, representation as blocks.  Might be able to modify Fig \ref{fig::set} and \ref{fig:block}.}
 
@@
 
-So far, we have a shared tree that processes use to agree on the
-implicit ordering stored in its root. With this agreement on the
-linearization ordering, we can design a universal construction; for a
-given object, we can perform an operation $op$ by applying all the
-operations up until $op$ in the root on a local copy of the object and
-then returning the response for $op$. However, this approach is not
-enough for an efficient queue. 
-We show that we can build an efficient queue if we can compute two
-things about the ordering in the root: (1) the $i$th propagated
-operation and (2) the rank of a propagated operation in the
-linearization. We explain how to implement (1) and (2) in
-poly-logarithmic steps.
+\subsection{Representing Blocks}
+
+If a block contained an explicit representation of its sequences of enqueues and dequeues,
+it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
+Instead, the block stores an implicit representation of the sequences.
+
+---------
 
 After propagating an operation \textit{op} to the root, processes can
-find out information about the linearization ordering using (1) and
-(2).  
+find out information about the linearization ordering using \ref{findenqueue} and
+\ref{findinroot}.  
 To get the $i$th operation in the root, we find the block $B$

### 2022-12-19 podc_paper/correctness.tex
working on algorithm description

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
-\renewcommand{\baselinestretch}{1.8}
-
-\usepackage[letterpaper, margin=1in]{geometry}
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-%\usepackage{showkeys}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-
-\usepackage{multicol}
-\setlength\columnsep{24pt}
-
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\begin{document}
-
-\begin{center}
-
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\section*{Abstract}
-
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
-\pagebreak
-
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\tableofcontents
-
-\pagebreak
-\listoffigures
-
-\pagebreak
-\listoftables
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
-\section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
+% !TEX root =  podc-submission.tex
+
+\section{Proof of Correctness}
+
+\here{following material from algorithm description in thesis might be useful in sketch of correctness proof}
+\begin{figure}[hbpt]
+  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
+  \caption[Two consecutive failed \nf{Refresh}es by a
+    process.]{\label{fig::simpleDoubleRefresh}Time relations between
+    the concurrent successful \nf{Refresh}es and the two consecutive
+    failed \nf{Refresh}es.} 
 \end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
-
-\pagebreak
-\section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
-
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
-
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
-
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
-
-\subsection{Restricted Queues}
-
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
-
-% TODO: page
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
-
-\section{Proof of Correctness}
+We use \nf{CAS} (Compare\&Swap) instructions to implement the
+\nf{Refresh}'s attempt to append  described in the previous
+paragraph. 
+The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
+successful \texttt{Refresh} that has read its information after the
+invocation of the first failed \nf{Refresh} (see Figure
+\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
+a successful append during $P$'s first failed attempt, and some
+process $K$ performs a \nf{Refresh} that reads its information after
+$L$'s append and then performs a successful append during $P$'s second
+failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
+new operations that were in $n$'s children before $P$'s first failed
+\nf{Refresh}, in case they were not already appended. After a process
+appends its operation into its leaf it can call \nf{Refresh} on the
+path up to the root at most two times on each node. So, with $O(\log
+p)$ \nf{CAS}es an operation can ensure it appears in the
+linearization. This cooperative solution allows us to overcome the CAS
+Retry Problem.
 
@@
 By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\pagebreak
-
-\pagebreak
-
-\bibliography{main.bib} 
-\bibliographystyle{acm}
-
-\end{document}

### 2022-12-19 podc_paper/podc-submission.tex
working on algorithm description

@@
 \newcommand{\var}[1]{\nf{#1}}
+\newcommand{\com}{$\triangleright$}
 
@@
 \begin{algorithm}
-\caption{Tree Fields Description}
+\caption{Objects used in Tournament Tree Data Structure}
 \begin{algorithmic}[1]
@@
 \begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
+\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
+\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
+\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
+\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
+\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
 \end{itemize}
@@
   \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
+  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
 \begin{itemize}
     \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
+  \textsf{\com\ indices of the last subblock of the block in the left and right child}
+  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
+  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
+\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
 \begin{itemize}
   \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
 
     \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
+  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
+\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
 \begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
+  \item \tt{\sl{int} \size}%
+  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
 \end{itemize}
@@
 
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
+\Statex \com\ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
 \Function{Object}{Dequeue()}{} 
@@
 
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
+\Statex \com\ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
 \Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
+\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
 \Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
@@
 \Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
+\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
+\com\ the \nf{b}th block in the \nf{root}.}
 
@@
 
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
+\Statex \com\ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
 \Function{void}{Propagate()}{}
@@
 
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
+\Statex \com\ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
 \Function{boolean}{Refresh()}{}
@@
 
-\section{Proof of Correctness}
+\input{correctness}
 
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
+\input{analysis}
 
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
-
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
-
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\section{Reducing Space}
-\label{reducing}
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
+\input{boundedspace}
 

### 2022-12-19 podc_paper/analysis.tex
added .tex files for sections

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
-\renewcommand{\baselinestretch}{1.8}
-
-\usepackage[letterpaper, margin=1in]{geometry}
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-%\usepackage{showkeys}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-
-\usepackage{multicol}
-\setlength\columnsep{24pt}
-
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\begin{document}
-
-\begin{center}
-
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\section*{Abstract}
-
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
-\pagebreak
-
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\tableofcontents
-
-\pagebreak
-\listoffigures
-
-\pagebreak
-\listoftables
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
-\section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
-
-\pagebreak
-\section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
-
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
-
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
-
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
-
-\subsection{Restricted Queues}
-
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
-
-% TODO: page
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
-
-\section{Proof of Correctness}
-
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
-
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
-
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
-
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
+% !TEX root =  podc-submission.tex
 
@@
 \end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\pagebreak
-
-\pagebreak
-
-\bibliography{main.bib} 
-\bibliographystyle{acm}
-
-\end{document}

### 2022-12-19 podc_paper/boundedspace.tex
added .tex files for sections

--- 2022-11-15 thesis/thesis.tex
@@
-\documentclass[11pt]{article}
-\renewcommand{\baselinestretch}{1.8}
+% !TEX root =  podc-submission.tex
 
-\usepackage[letterpaper, margin=1in]{geometry}
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-%\usepackage{showkeys}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-
-\usepackage{multicol}
-\setlength\columnsep{24pt}
-
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{%
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\begin{document}
-
-\begin{center}
-
-\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}
-
-\vspace{2em}
-\Large{\bf{Hossein Naderibeni}}
-
-\vspace{3em}
-\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science
-
-\vspace{2em}
-Graduate Program in Computer Science\\
-York University,\\
-Toronto, Ontario
-
-\vspace{2em}
-November, 2022
-
-\vspace{3em}
-\textcopyright \; Hossein Naderibeni, 2022
-
-\end{center}
-\thispagestyle{empty}
-\pagebreak
-
-\normalsize
-\section*{Abstract}
-
-In this work, we introduce a novel linearizable wait-free queue implementation.
-Linearizability and lock-freedom are standard requirements for designing shared data structures.
-To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem.
-We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
-\pagenumbering{roman}
-\pagebreak
-
-\section*{Acknowledgements}
-\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert.
-Without his guidance and support, this thesis probably would not have been completed.
-He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it.
-Thank you again for always pushing to make it better.
-
-I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.
-
-Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.
-
-Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study.
-I want to thank my brother for being there for me whenever I needed help.
-
-I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.
-
-I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.
-
-\pagebreak
-\tableofcontents
-
-\pagebreak
-\listoffigures
-
-\pagebreak
-\listoftables
-\newpage
-
-\pagebreak
-
-\pagenumbering{arabic}
-\section{Introduction}
-Shared data structures have become an essential field in distributed algorithms research.
-We are reaching the physical limits of how many transistors we can place on a CPU core.
-The industry solution to provide more computational power is to increase the number of cores of the CPU.
-It is not hard to see why multiple processes cannot share a sequential data structure.
-For example, consider two processes trying to append to a sequential linked list simultaneously.
-Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same.
-In this run, $P$'s update is overwritten, which makes the queue inconsistent.
-One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure.
-Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing.
-Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked.
-For these reasons, we are interested in the design of an efficient shared queue without using locks.
-
-A sequential queue stores a sequence of elements and supports two operations, enqueue and dequeue.
-\texttt{Enqueue(e)} appends the element \texttt{e} to the sequence stored.
-\texttt{Dequeue} removes and returns the first element in the queue.
-If the queue is empty, it returns \texttt{null}.
-In a concurrent version of the queue data structure, operations do not happen one at a time.
-The question that may arise is, ``What properties matter for the implementation of a shared data structure?'' Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ.
-To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions.
-A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.
-
-A system is called \it{asynchronous} when processes in the system run at arbitrarily varying speeds, i.e., the scheduling of each process is independent from the scheduling of other processes.
-Our model is an asynchronous shared-memory distributed system of $p$ processes.
-Herlihy \cite{10.1145/114005.102808} showed that one cannot implement concurrent versions of all data structures with multi-reader multi-writer registers alone and introduced a hierarchy based on how powerful objects are to solve the consensus problem among processes.
-Objects like LL/SC or CAS can be used to reach a consensus among any number of processors.
-A CAS object provides an atomic Compare\&Swap(\nf{new, old}) operation: if the value stored in the object is \nf{old}, it updates the value to \nf{new} and returns \nf{true}; otherwise it returns \nf{false}.
-In this work, we use Compare\&Swap (\nf{CAS}) objects to synchronize among processes.
-
-% \bf{Concurrent execution: I didn't understed interleaving steps of processes}
-
-The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.
-Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty.
-The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation.
-Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}.
-The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/good}
-  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution.
-Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
-\end{figure}
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/bad}
-  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable.
-\texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
-\end{figure}
-
-There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom.
-An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps.
-We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates.
-A wait-free algorithm is also lock-free but not vice versa; in an infinite run of a lock-free algorithm, there might be an operation that takes infinitely many steps but never terminates.
-
-We will present a wait-free linearizable queue, which to the best of our knowledge  is the first wait-free queue whose operations run in a poly-logarithmic number of steps.
-We design a tournament tree in which each process tries to propagate its operations along a path from the process's leaf up to the root to be linearized.
-Processes can do queries to get information about the linearization stored in the root.
-We create a helping mechanism for when a process is propagating its operation to a node, such that new operations are ensured to be propagated one step further up the tree after at most 2 \nf{CAS} invocations.
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue}, is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\nf{CAS} instructions cost more than other instructions for the processor.
-Each operation in our queue does $\Theta(\log p)$ \nf{CAS} operations compared to $\Omega(p)$ \nf{CAS} steps for previous queues in the worst case.
-We use unbounded memory space in our queue and present a way to reduce the memory size to the total number of operations.
-Furthermore, we do not address garbage collection but we describe a way we think it can be handled.
-
-\paragraph{Thesis outline} The rest of the thesis is organized as follows.
-Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation.
-Previous lock-free queues and their common problem are presented in Section 2.1.
-In Section 2.2, we mention some restricted lock-free queues.
-Section 2.3 talks about poly-logarithmic constructions of shared objects.
-Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.
-
-In Chapter 3 we introduce a poly-logarithmic step wait-free queue.
-In Section 3.1 we give a high-level description of our implementation.
-We also discuss the motivation and requirements in our design to achieve poly-log time.
-Section 3.2 contains the algorithm itself and there is an example of a concurrent execution of the algorithm described in Section 3.3.
-
-We prove the correctness of our queue in Chapter 4 by showing it is linearizable.
-
-Chapter 5 is devoted to a complexity analysis for our implementation.
-We compute the number of \nf{CAS} instructions our algorithm invokes and the worst-case and amortized running time of the queue.
-In the end, we prove our queue is wait-free.
-
-Finally, we give some concluding remarks in Chapter 6. It contains how we can improve our algorithm's memory usage and how to use its idea in designing other wait-free data structures.
-
-\pagebreak
-\section{Related Work}
-In this section, we look at previous lock-free queues.
-The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously.
-Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.
-
-\subsection{List-based Queues}
-Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue.
-A version of it is included in the standard Java Concurrency Package.
-Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}).
-A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue.
-To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}.
-If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry.
-This makes the amortized number of steps $\Omega(p)$ per enqueue.
-Similarly, dequeue can take $\Omega(p)$ steps.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.4]{pics/msqueue}
-  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations.
-In the first diagram, the first element has been dequeued.
-Red arrows show new pointers and gray dashed arrows show the old pointers.}
-\end{figure}
-
-Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the \it{elimination} technique.
-The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue.
-An Elimination Queue consists of an MS-queue augmented with an elimination array.
-Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue.
-Their algorithm makes it possible for long-running operations to eliminate an opposing operation.
-The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.
-
-Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue.
-Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket.
-Each basket has a time interval in which all its nodes' enqueue operations overlap.
-Since the operations in a basket are concurrent, we can order them in any way.
-Enqueues in a basket try to find their order in the basket one by one using \texttt{CAS} operations.
-However, like the previous algorithms, if there are $p$ concurrent enqueue operations in a basket, the amortized step complexity remains $\Omega(p)$ per operation.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.3]{pics/baskets}
-  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}.
-To order the operations in a basket, processes perform another \nf{CAS}.
-The successful process will be the next one in the basket, and so on.}
-\end{figure}
-
-Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an optimistic approach to implement a queue.
-MS-queue uses two \nf{CAS}es to do an enqueue: one to change the tail to the new node and another one to change the next pointer of the previous node to the new node.
-They use a doubly-linked list to do fewer \texttt{CAS} operations in an \nf{Enqueue} than MS-queue.
-As in  previous algorithms, the worst case happens in the case where the contention is high: when $p$ concurrent enqueues happen, their nodes have to be appended to the linked list one by one.
-The amortized complexity is still $\Omega(p)$ \texttt{CAS}es.
-
-Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new paradigm called flat combining.
-The key idea behind flat combining is to allow a combiner who has acquired the global lock on the data structure to learn about the requests of threads on the queue, combine them and apply the combined results on the data structure.
-Their queue is linearizable but not lock-free and they present experiments that show their algorithm performs well in some situations.
-
-Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays.
-The queue is stored in a shared array where head and tail pointers point to the current elements in the queue.
-When the array is full, an empty array is linked to the array and tail pointers are updated.
-A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array.
-Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}).
-A process updates the position of the pointers after it does an operation.
-One process might go to sleep before setting the pointers, so the pointers might be behind their real places.
-They mention how to scan the arrays to update pointers while doing an operation.
-A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$.
-Their design is lock-free but not wait-free.
-
-\begin{figure}[hbt]
-  \center\includegraphics[scale=0.5]{pics/sundell}
-    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays.
-Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
-\end{figure}
-
-Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique~\cite{10.1145/114005.102808} to achieve wait-freedom.
-Their amortized step complexity is $\Omega(p)$ because of the helping mechanism.
-
-Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free queue supporting futures.
-In their queue, operations  return future objects instead of responses.
-Later when the response is needed, it can be evaluated from the future object.
-They also define a weaker linearizability condition such that each operation can be linearized between its invocation and when its future is evaluated.
-Their idea of batching allows a sequence of operations to be submitted as a batch for later execution on the MS-queue.
-They use some properties of the queue size before and after a batch, similar to a part of our work.
-Their queue is not wait-free: in fact, if the batch sizes are 1, then the queue is like MS-queue.
-
-Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-Their work is based on a circular queue using bounded memory.
-When a process wishes to do an enqueue or a dequeue, it starts two paths.
-The fast path  ensures good performance while the slow path ensures termination.
-They show that these two paths do not affect each other and the queue remains consistent.
-If a process makes no progress, other processes help its slow path to finish.
-The helping phase suffers from the CAS Retry Problem because processes compete in a \nf{CAS} loop to decide which succeeds to help.
-Because of this, the amortized complexity cannot be better than $\Omega(p)$.
-
-The CAS Retry Problem is not limited to list-based queues; array-based queues also share it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
-Our motivation is to overcome this problem and present a wait-free sublinear queue.
-
-\subsection{Restricted Queues}
-
-David introduced the first sublinear concurrent queue \cite{DBLP:conf/wdag/David04}.
-Even though his algorithm does $O(1)$ steps for each operation, it is a single-enqueuer single-dequeuer queue and uses infinite memory.
-The author states that to reduce memory usage to be bounded, the time per operation increases linearly.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}.
-We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a poly-logarithmic multi-enqueuer multi-dequeuer queue.
-Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.
-
-\subsection{Universal Constructions and Other Poly-log Time Data Structures}
-A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object.
-The first universal construction was introduced by Herlihy~\cite{10.1145/114005.102808}.
-We can implement a concurrent queue using a universal construction.
-Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}.
-He also mentions that the universal construction by Afek, Dauber, and Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$ worst-case step complexity, using atomic access to $\Omega(p \log p)$-bit words.
-Chandra, Jayanti and Tan introduced a semi-universal construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}.
-However, their algorithm cannot be used to create a queue.
-We mention a non-practical universal construction with a poly-log number of \nf{CAS} instructions in the last paragraph of page 13.
-
-% TODO: page
-
-Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}.
-Their idea to achieve logarithmic complexity is to use a tree storing the Fetch\&Inc operations invoked by processes.
-When a process wants to do a Fetch\&Inc it adds its Fetch\&Inc to the tree and returns the number of elements in the tree.
-There are some similarities between designing a queue and a Fetch\&Inc object.
-A Fetch\&Inc object can be constructed from a queue.
-The algorithm by Ellen and Woelfel is interesting because of the similarities between Fetch\&Inc objects and queues.
-Also, it is one of the few wait-free data structures achieving poly-logarithmic complexity.
-
-\subsection{Attiya--Fouren Lower Bound}
-Because of the \nf{CAS} retry problem in previous list-based queues one might guess the $\Omega(p)$ term is inherent in the time complexity of concurrent queues.
-Attiya and Fouren gave a lower bound on amortized complexity of lock-free queues with regard to $c$, the number of concurrent processes.
-Their result says if $c$ is $O(\log\log p)$, any implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17}.
-The surprising point is that since their result is about when contention is low, their lower bound does not contradict our algorithm's complexity and we manage to reach poly-log time complexity.
-
-\pagebreak
-
-\section{Queue Implementation} \label{DescriptQ}
-
-In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently.
-We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization.
-We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.
-
-There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes.
-The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}).
-Each process has a leaf in which the operations invoked by the process are stored in order.
-When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root.
-Each node of the tree keeps an ordering of operations propagated up to it.
-All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering.
-\begin{figure}[hbtp]
-\begin{center}
-\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
-          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
-\end{center}
-\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored.
-Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}
-\end{figure}
-
-To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$.
-We call this procedure \tt{$n$.Refresh} (see Figure \ref{fig::propagstep}).
-A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to $n$.
-We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
-We sketch the argument here.
-
-\begin{figure}[hbtp]
-\begin{center}
-\hspace{6em}\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{Before the \nf{Refresh}.}
-\end{subfigure}
-\hfill
-\begin{subfigure}[b]{.29\textwidth}
-  \centering
-  \resizebox{\columnwidth}{!}{
-  \begin{tikzpicture}[level 1/.style={level distance=1.2cm,sibling distance=0.5cm}]
-\Tree [.$r_1,l_1,l_2,r_2,l_3,l_4,l_5,r_3,r_4$
-$l_1,l_2,l_3,l_4,l_5$
-$r_1,r_2,r_3,r_4$ ]
-\end{tikzpicture}}
-  \caption{After the \nf{Refresh}.}
-\end{subfigure}
-\hspace{6em}
-\caption[A successful \nf{Refresh}.]{\label{fig::propagstep} Before and after a \nf{$n$.Refresh} with a successful append.
-Operations propagating from the left child are labelled with $l$ and from the right child with $r$.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh.drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a process.]{\label{fig::simpleDoubleRefresh}Time relations between the concurrent successful \nf{Refresh}es and the two consecutive failed \nf{Refresh}es.}
-\end{figure}
-
-We use \nf{CAS} (Compare\&Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}).
-This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}.
-Process $K$'s \nf{Refresh} helps to append the new operations that were in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended.
-After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node.
-So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization.
-This cooperative solution allows us to overcome the CAS Retry Problem.
-
-It is not efficient to explicitly store the sequence of operations in each node because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
-Instead we use an implicit representation of the operations propagated together.
-Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root.
-It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}).
-Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to do \texttt{CAS} instructions on fixed-size objects in each \tt{Refresh}.
-To do that, we introduce \it{block}s.
-A block stores information about the operations propagated by a \nf{Refresh}.
-It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure.
-See Figure~\ref{fig:block} for an example.
-A node stores an array of blocks of operations propagated up to it.
-A propagate step  aggregates the new blocks in the children into a new block and stores the new block in the parent.
-We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.
-In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one process cannot invoke two operations concurrently.
-Thus, there are at most $p$ operations in a block.
-Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together.
-Our choice is to put the operations propagated from the left child before the operations propagated from the right child.
-In this way, if we know the number of operations from the left child and the number of operations from the right child in a block, we have a complete ordering of the operations.
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from left to right.
-In each internal node one can arbitrarily linearize the sets of concurrent operations  propagated together in a \nf{Refresh}.
-For example $op_4^1$ and $op_3^2$ have propagated  together in one \nf{Propagate} step and they will be propagated up to the root together.
-Since their execution time intervals overlap, they can be linearized in any order.}
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left, right) blocks.]{\label{fig:block} Using blocks to represent operations.
-Blocks between two lines $||$ are propagated together to the parent.
-Each block consists of a pair (left, right) indicating the number of operations from the left and the right child, respectively.
-For example, (12,6) in the root contains  (10,2) from the left child and (6,0) from the right child.
-The third block in the root (8,36) is created by merging (5,3) from the left child and (14,6) and (0,16) from the right child.
-(5,3) is the superblock of  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).}
-\end{figure}
-
-So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root.
-With this agreement on the linearization ordering, we can design a universal construction; for a given object, we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$.
-However, this approach is not enough for an efficient queue.
-We show that we can build an efficient queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization.
-We explain how to implement (1) and (2) in poly-logarithmic steps.
-
-After propagating an operation \textit{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
-To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in the descendent of the root that contains that $i$th operation.
-When we reach a block in a leaf, the operation is explicitly stored there.
-To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block.
-We also store pointers to determine the range of subblocks of a block to make the binary search faster.
-In each block, we store the prefix sum of operations from the left child and from the right child.
-Moreover, for each block, we store two attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of the last left and right subblock (see Figure \ref{fig::pointer}).
-We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and \nf{end\sub{right}} indices of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to find the superblock of the block that the operation is in.
-After a block is installed in a node we store the approximate index of its superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each child. \label{fig::pointer}}
-\end{figure}
-
-In an execution on a queue where no dequeue operation returns \nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue.
-In the general case a dequeue returns \nf{null} if and only if the queue size after the previous operation is 0.
-We refer to such a dequeue as a \it{null dequeue}.
-If the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue.
-Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues before the operation.
-So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.
-
-In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific position.
-We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations.
-Consider the following operations, where operations in a cell are concurrent.
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2, 1, 3, 4, null}, respectively.
-Now, we claim that by knowing the size of the queue, we can compute the rank of the required \texttt{Enqueue} for any non-null \nf{Dequeue}.
-We apply this approach to blocks; if we store the size of the queue after each block of operations, we can compute the index of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root could be computed as
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$
-\noindent Moreover, the total number of non-null dequeues in blocks $1,2,...,b$ in the root is
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} - \textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
-
-\subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue implementation. It uses the following two types of objects.
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left and right child, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.
-
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf.
-If it is in a leaf, we use a \nf{LeafBlock} which stores one operation.
-If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$.
-The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the left subblocks of the block prior to $B$ ended.
-The right subblocks of $B$ are defined similarly.
-In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}.
-The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child.
-The other fields' semantics are similar.
-The \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of the last subblock of a block in the left and the right child, respectively.
-The approximate index of the superblock of non-root blocks is stored in their \nf{super} field.
-The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.
-
-We now describe the routines used in the implementation.
-
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is sufficient to propagate the \nf{Enqueue} operation to the root and then use its position in the linearization for future \nf{Dequeue} operations.
-\nf{Enqueue($e$)} creates a \nf{LeafBlock} with $\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and then appends it to the tree.
-
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree.
-Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.
-
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}.
-There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.
-\nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}.
-When \nf{Propagate} terminates, it is guaranteed that the appended block is a subblock of a block in the \nf{root}.
-
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described earlier and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}.
-Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.
-
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}.
-The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}.
-The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$.
-Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children.
-\nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recently appended blocks.
-The reason behind this helping is explained later when we discuss \nf{IndexDequeue}.
-After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}.
-Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate, and it is unnecessary to add an empty block to the tree.
-Later we will use the fact that all blocks contain at least one operation.
-Line \ref{cas} tries to install \nf{new}.
-If it was successful, all is good.
-If not, it means someone else has already put a block in $n$\nf{.blocks[h]}.
-In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.
-
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
-The block \nf{new} is created in Line \ref{initNewBlock} and its fields are filled similarly for both left and right directions.
-The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}.
-Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child.
-Then \nf{sum\sub{enq-dir}} is computed from the sum of the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}.
-The field \nf{sum\sub{deq-dir}} is computed similarly.
-Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.
-
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$.
-Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}.
-Let $par$ be $n$\nf{.parent}.
-Without loss of generality, assume for the rest of this section that ${n}$ is the left child of $par$.
-Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ as a subblock of its created block in Line \ref{invokeCreateBlock}.
-Let $j$ be the index of the block that $R_{par}$ puts in $par$\nf{.blocks}.
-
-Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating $B$.
-One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$.
-This solution would not be efficient because there might be up to $p$ subblocks that $R_{par}$ propagated, which need to update their \nf{super} field.
-However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation.
-If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed.
-Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$.
-We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$.
-If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}.
-In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.
-
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}.
-After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}.
-We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most~1.
-
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not.
-If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}.
-If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is the response to the \nf{Dequeue}.
-Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block), we find the block and position containing the \nf{Enqueue} operation using \nf{DoublingSearch} and after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block.
-If we know the block and rank within the block of an operation, we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time.
-To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
-We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}.
-\nf{DoublingSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in the $b$th block of $n$.
-The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block.
-Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\newpage
-\clearpage
-
-\subsection{Example Execution}
-
-Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.
-
-Initially, each node in the tree contains one empty block.
-In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$.
-Line \ref{appendLeaf} appends a block to the process's leaf.
-
-Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$.
-$P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$.
-$P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$.
-So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).
-
-Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}.
-$P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.
-
-Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root.
-After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}.
-$P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.
-
-All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}.
-$P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root.
-$P2$ needs to compute the index of the block that its \nf{Dequeue} is in.
-It computes the superblock of its leaf block, which is the second block in $n5$.
-Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$.
-As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2.
-The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element.
-Then $P2$ tries to get $ENQ_1$'s element.
-$ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$.
-Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$.
-$P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element.
-$P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$.
-Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
-  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
-  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
-  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
-  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.
-$P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}.
-\label{fig::exec4}}
-\end{figure}
-
-\begin{figure}
-  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
-  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
-\end{figure}
-
-\clearpage
-
-\section{Proof of Correctness}
-
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
-
-\subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
-
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
-
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
-\begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
-\end{proof}
-
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
-\end{lemma}
-\begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
-\end{proof}
-
-\begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
-\end{lemma}
-\begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
-\end{proof}
-
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
-
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
-
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
-\begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
-\end{proof}
-
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
-\subsection{Ordering Operations}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
-
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
-
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
-
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
-\end{proof}
-
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
-
-\subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
-\end{lemma}
-\begin{proof}
-
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
-
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
-\end{proof}
-
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
-\begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
-\end{proof}
-
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
-\begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
-\end{proof}
-
-\subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
-
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
-\end{lemma}
-\begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
-\end{proof}
-
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
-
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
-\end{lemma}
-\begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
-
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
-
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
-
-\begin{lemma} \label{linearSat}
-$L$ is a legal linearization ordering.
-\end{lemma}
-\begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
-\end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
-
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$.
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
-\end{proof}
-
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
-\end{lemma}
-\begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
-\end{proof}
-
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
-
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
-The queue implementation is linearizable.
-\end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-\pagebreak
-
-\section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
-
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
-
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
-\end{proof}
-
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
-
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
-The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
-
-\pagebreak
-\section{Future Directions}
-We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}.
-We implemented two queries to compute information about the ordering agreed in the block tree.
-Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes.
-Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects.
-In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.
-
+\section{Reducing Space}
+\label{reducing}
 \paragraph{Reducing Space Usage}
@@
 \end{figure}
-
-\paragraph{Poly-logarithmic Wait-free Sequences}
-Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}.
-An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence.
-We can modify our queue to design such a data structure.
-An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
- We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.
-
-\paragraph{Other Poly-log Wait-free Data structures}
-There are two reasons the block tree worked well to implement a queue.
-Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size.
-Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue.
-Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.
-
-\paragraph{Attiya--Fouren Lower Bound}
-As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$, the number of concurrent processes, is $\Omega(\log\log p)$.
-Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$.
-It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.
-
-\pagebreak
-
-\pagebreak
-
-\bibliography{main.bib} 
-\bibliographystyle{acm}
-
-\end{document}

### 2022-12-20 podc_paper/algorithm.tex
working on description of block fields

@@
 
+Each node has an infinite array called \var{blocks}.
+To simplify the code, each node has an empty block in \var{blocks[0]}.
+
 If a block contained an explicit representation of its sequences of enqueues and dequeues,
 it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
-Instead, the block stores an implicit representation of the sequences.
+Instead, the block stores an implicit representation of the sequences; see Figure \ref{object-fields}.
+
+\here{This parag may need to be rewritten if we use sum\sub{enq} field instead of sum\sub{enq-left} and sum\sub{enq-right}.
+Also, maybe it is too detailed at this point; perhaps instead just say that for task \ref{findenqueue} we want to be able to find the position of the $r$th enqueue in a block within a block in its child using binary search.}
+Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.
+This allows us to find the direct subblocks of any block easily.
+Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.
+Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
+First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
+so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
+This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
+Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
+to determine whether the enqueue came from the left or right child of the root.
+Suppose the enqueue came from the right child $v_r$.
+Then, we know that the index of the block in $v_r$ that contains the enqueue
+is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
+We can again do a binary search within this range.
+For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
+We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
+We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
+and the binary search within each other node along the path to a leaf can be done in $O(\log p)$ steps,
+for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
+All of the information for this search process can be derived from the 
+\var{sum\sub{enq-left}} and \var{sum\sub{enq-right}} fields.
+
+To facilitate \ref{findinroot}, each block has a field \var{super} that contains
+the (approximate) index of its superblock in the parent node's blocks array.
+We will ensure that this field's value differs from the true index of the superblock by at most 1.
+This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two blocks in the parent node.
+Thus, starting from an operation in a leaf's block, one can use these indices to track the 
+operation all the way up the path to the root, and determine the operation's location in a root block
+in $O(\log p)$ time.
+
+\begin{figure}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{1}
+
+\Statex $\diamondsuit$ \tt{\sl{Shared}}
+\begin{itemize}
+\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
+\end{itemize}
+
+\Statex $\diamondsuit$ \tt{\sl{Local}}
+\begin{itemize}
+\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{Node}}
+\begin{itemize}
+\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
+\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
+\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{Block}}
+
+\begin{itemize}
+  \item \tt{\sl{int} super}
+  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
+\begin{itemize}
+    \item \tt{int} \eleft, \eright
+  \textsf{\com\ indices of the last subblock of the block in the left and right child}
+  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
+  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
+  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
+\begin{itemize}
+  \item \tt{\sl{Object} element}
+  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+
+    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
+\begin{itemize}
+  \item \tt{\sl{int} \size}%
+  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
+\end{itemize}
+
+\end{algorithmic}
+\caption{Objects used in Tournament Tree Data Structure \label{object-fields}}
+\end{figure}
 

### 2022-12-20 podc_paper/podc-submission.tex
working on description of block fields

@@
 \algrenewcommand\algorithmicreturn{{\keywordfont return}}
+
+\newcommand{\eleft}{\var{end\sub{left}}}
+\newcommand{\eright}{\var{end\sub{right}}}
+\newcommand{\edir}{\var{end\sub{dir}}}
 
@@
 \begin{algorithm}
-\caption{Objects used in Tournament Tree Data Structure}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
-\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
-\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{\com\ indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
-\begin{itemize}
-  \item \tt{\sl{int} \size}%
-  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
 \caption{\tt{\sl{Queue}}}
@@
 \ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
+\State \tt{index\sub{prev}= blocks[i-1].\edir} \label{prevLine}
+\State \tt{new.\edir= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].\edir..dir.\head-1]}.}
 \State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
+\State \tt{block\sub{last}= dir.blocks[new.\edir]}
 \State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
@@
 \ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
+\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
+\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
 \State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
@@
 \State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
+\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
+\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
 \State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 

### 2022-12-21 podc_paper/algorithm.tex
working on description of blocks

@@
 Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
-We shall define a total order on all operations that have been propagated to the root, which is used to
-linearize the operations.
+We shall define a total order on all operations that have been propagated to the root, which 
+will serve as the linearization ordering of the operations.
 
@@
 Then, $P$ determines whether the queue is empty when its dequeue is linearized. If so, it returns \nil.
-If not, $P$ computes the rank $r$ of its dequeue in the linearization ordering
+If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
 among all dequeues that do not return \nil,
-finds the $r$th enqueue in the linearization and returns that enqueue's value.
+finds the $r$th enqueue in the linearization, and returns that enqueue's value.
 
@@
 
-\subsection{Representing Blocks}
+\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
 
@@
 To simplify the code, each node has an empty block in \var{blocks[0]}.
+The node also has a \var{head} index which is the position in the \var{blocks} array to be used
+for the next attempt to append a block to this node.
 
 If a block contained an explicit representation of its sequences of enqueues and dequeues,
-it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
-Instead, the block stores an implicit representation of the sequences; see Figure \ref{object-fields}.
+it would take $\Omega(p)$ time to construct a block, which would be much too slow for task \ref{construct}.
+Instead, the block stores an implicit representation of the sequences.
+We now explain how the fields of the blocks were chosen for this implicit representation. 
+See Figure \ref{object-fields}.
+
+A block in a leaf represents a single enqueue or dequeue.  The \var{element} field stores the value
+enqueued for an enqueue operation, or \var{null} if the operation is a dequeue.
 
@@
 Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.
-This allows us to find the direct subblocks of any block easily.
+This allows us to navigate to the direct subblocks of any block easily.
 Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.
@@
 \var{sum\sub{enq-left}} and \var{sum\sub{enq-right}} fields.
+\here{modify to \var{sum\sub{enq}} and \eleft/\eright.}
 
-To facilitate \ref{findinroot}, each block has a field \var{super} that contains
+To facilitate task \ref{findinroot}, each block has a field \var{super} that contains
 the (approximate) index of its superblock in the parent node's blocks array.
@@
 
+We now consider task \ref{findrank}.
+First, we must determine whether the queue is empty when a dequeue occurs.
+For this, each block in the root stores a \var{size} field that stores the number of elements
+in the queue after all operations in the linearization ordering up to that block (inclusive) 
+have been performed.
+We can easily determine which dequeues in a block $B_d$ in the root should return \nil using
+$B_{d-1}.\var{size}$ and the number of enqueues and dequeues in $B_d$.
+We can also compute the number of dequeues in blocks $B_1, \ldots, B_{d-1}$ that returned non-nil
+values as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
+Then, for any dequeue in $B_d$ that should return a non-nil value, we can also determine its
+rank among all non-nil dequeues in the linearization ordering, and hence the rank of the enqueue
+whose value it should return (among all enqueues).
+
+Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
+we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
+A \op{refresh} on node $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
+$h-1$ in the $B.\eleft$ and $B.\eright$.
+$B.\var{sum\sub{enq}}$ can be computed as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
+For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block,
+the number of enqueues in $B$ and the number of non-nil dequeues in $B$ (determined as described above).
+
+The only remaining field is $B.\var{super}$.  When the block 
+$B$ is created for a node $v$, we do not yet know where its
+superblock will eventually be installed in $B$'s parent.
+So, we leave this field blank.  After $B$ is installed 
+in $v.\var{blocks[i]}$, processes cooperate to fill it in 
+when they attempt to advance $v.\var{head}$ from $i$ to $i+1$.
+They use the value read from the \var{head} field of $v$'s parent at that time.
+As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
+shall prove that it is close.
+
 \begin{figure}
@@
 
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
+\Statex $\diamondsuit$ \tt{\sl{Shared Variable}}
 \begin{itemize}
@@
 
-\Statex $\diamondsuit$ \tt{\sl{Local}}
+\Statex $\diamondsuit$ \tt{\sl{Local Variable}}
 \begin{itemize}

### 2022-12-22 podc_paper/algorithm.tex
describing fields

@@
 (since the operation may have been propagated to the root by some other process).
-Then, $P$ determines whether the queue is empty when its dequeue is linearized. If so, it returns \nil.
+Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
+If so, it returns \nil\ and we call it a \emph{null dequeue}
 If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
-among all dequeues that do not return \nil,
+among all non-null dequeues,
 finds the $r$th enqueue in the linearization, and returns that enqueue's value.
@@
 \label{findrank}
-Given a dequeue's position in the root, determine that the queue is empty when it occurs
+Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
 or determine the rank $r$ of the enqueue whose value it should return.
@@
 \end{enumerate}
-Since these tasks require knowing the linearization ordering, we next describe that ordering.
+Since these tasks depends on the linearization ordering, we describe that ordering next.
 
@@
 
-\here{This parag may need to be rewritten if we use sum\sub{enq} field instead of sum\sub{enq-left} and sum\sub{enq-right}.
-Also, maybe it is too detailed at this point; perhaps instead just say that for task \ref{findenqueue} we want to be able to find the position of the $r$th enqueue in a block within a block in its child using binary search.}
 Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.
 This allows us to navigate to the direct subblocks of any block easily.
-Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.
-Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
-First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
-so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
-This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
-Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
-to determine whether the enqueue came from the left or right child of the root.
-Suppose the enqueue came from the right child $v_r$.
-Then, we know that the index of the block in $v_r$ that contains the enqueue
-is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
-We can again do a binary search within this range.
-For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
-We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
+Blocks also store prefix sums of the numbers of enqueues and dequeues:
+the block in $v.\var{blocks}[i]$ has two fields \var{sum\sub{enq}} and \var{sum\sub{deq}}
+that store the total number of enqueues and dequeues in $v.\var{blocks[1..i]}$.
+These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
+For example, consider finding the $r$th enqueue in the linearization ordering.
+We use the \var{sum\sub{enq}} fields of the root's blocks to do a binary search
+to locate which block in the root contains the operation.
+If we know which block $B$ in some node $v$ contains the enqueue,
+we can use the \var{sum\sub{enq}} field again determine which child of $v$ contains the enqueue
+and then to do a binary search
+among the direct subblocks of $B$ in that child.
+Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
+the enqueue.
+%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
+%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
+%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
+%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
+%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
+%to determine whether the enqueue came from the left or right child of the root.
+%Suppose the enqueue came from the right child $v_r$.
+%Then, we know that the index of the block in $v_r$ that contains the enqueue
+%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
+%We can again do a binary search within this range.
+%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
+%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
 We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
@@
 for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
-All of the information for this search process can be derived from the 
-\var{sum\sub{enq-left}} and \var{sum\sub{enq-right}} fields.
-\here{modify to \var{sum\sub{enq}} and \eleft/\eright.}
+All of the information needed for this search process can be derived from the 
+\eleft, \eright\ and \var{sum\sub{enq}} fields.
 
@@
 First, we must determine whether the queue is empty when a dequeue occurs.
-For this, each block in the root stores a \var{size} field that stores the number of elements
+To facilitate this, each block in the root stores a \var{size} field that stores the number of elements
 in the queue after all operations in the linearization ordering up to that block (inclusive) 
 have been performed.
-We can easily determine which dequeues in a block $B_d$ in the root should return \nil using
-$B_{d-1}.\var{size}$ and the number of enqueues and dequeues in $B_d$.
-We can also compute the number of dequeues in blocks $B_1, \ldots, B_{d-1}$ that returned non-nil
-values as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
-Then, for any dequeue in $B_d$ that should return a non-nil value, we can also determine its
-rank among all non-nil dequeues in the linearization ordering, and hence the rank of the enqueue
+We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
+$B_{d-1}.\var{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
+We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
+as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
+Then, for any non-null dequeue in $B_d$, we can also determine its
+rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
 whose value it should return (among all enqueues).
@@
 A \op{refresh} on node $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
-$h-1$ in the $B.\eleft$ and $B.\eright$.
-$B.\var{sum\sub{enq}}$ can be computed as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
+$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
+Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
 For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block,
-the number of enqueues in $B$ and the number of non-nil dequeues in $B$ (determined as described above).
+the number of enqueues in $B$ and the number of non-null dequeues in $B$ (determined as described above).
 
@@
 ---------
-
-After propagating an operation \textit{op} to the root, processes can
-find out information about the linearization ordering using \ref{findenqueue} and
-\ref{findinroot}.  
-To get the $i$th operation in the root, we find the block $B$
-containing the $i$th operation in the root, and then recursively find
-the subblock of $B$ in the descendent of the root that contains that
-$i$th operation. When we reach a block in a leaf, the operation is
-explicitly stored there. To make this search faster, instead of
-iterating over all blocks in the node, we store the prefix sum of the
-number of elements in the blocks sequence to permit a binary search
-for the required block. We also store pointers to determine the range
-of subblocks of a block to make the binary search faster. In each
-block, we store the prefix sum of operations from the left child and
-from the right child. Moreover, for each block, we store two
-attributes \nf{end\sub{left}} and \nf{end\sub{right}}, the indices of
-the last left and right subblock (see Figure \ref{fig::pointer}). We
-know a block size is at most $p$, so binary search takes at most
-\textsc{O}$(\log p)$ time, since the \nf{end\sub{left}} and
-\nf{end\sub{right}} indices of a block and its previous block reduce
-the search range size to \textsc{O}$(p)$.
-
-To compute the rank in the root of an operation in a leaf, we need to
-find the superblock of the block that the operation is in. After a
-block is installed in a node we store the approximate index of its
-superblock in it to make this faster.
-
-\begin{figure}[hbtp]
-\centering
-  \includegraphics[width=4in, height=1.4in]{pics/pointers}
-  \caption{Each block stores the index of its last subblock in each
-    child. \label{fig::pointer}} 
-\end{figure}
 

### 2022-12-22 podc_paper/related.tex

@@
 As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
-Thus, it is lock-free but not wait-free and has an amortized step complexity
+Thus, it is lock-free but not wait-free and has an amortized \here{Maybe we can say worst-case.
+I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.} step complexity
 of $\Theta(p)$ per operation.
@@
 In both cases, the helping mechanisms only add to the step complexity of the MS-queue.
+\here{I get what this sentence means, but from a high overview i think it is inaccurate.
+Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}
 
@@
 % go back to (*) and repeat this sequence
-Similarly,
+\here{Is it ok for a reader to talk about A and then start a sentence about similarity to B?
+My suggestion which is longer: Kogan and Petrank~\cite{10.1145/2370036.2145835} proposed the fast-path slow-path methodology which is similar to Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13}.
+Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} have used the methodology.} Similarly,
 other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.

### 2022-12-22 podc_paper/podc-submission.tex
some edits to related work

@@
 
+\def\draft{1}  % set to 1 for draft (include comments, color edits), 0 for clean final copy (omit comments, edits in black)
+\if\draft 1
+\newcommand{\mycomment}[3]{{\color{#2}{[\bf{#1: #3}]}}}
+\newcommand{\myedit}[2]{{\color{#1}{#2}}\normalcolor}
+\newcommand{\rebuttaledit}[2]{{\color{#1}{#2}}\normalcolor}
+\newcommand{\here}[1]{\bf{[[[#1]]]}}
+%\usepackage{showkeys}
+\else
+\newcommand{\mycomment}[3]{}
+\newcommand{\rebuttaledit}[2]{#2}
+\newcommand{\myedit}[2]{#2}
+\newcommand{\here}[1]{}  % \here's include comments to ourselves that readers should not see.
+\fi
+
+\newcommand{\Eric}[1]{\mycomment{Eric}{magenta}{#1}}
+\newcommand{\Hossein}[1]{\mycomment{Hossein}{blue}{#1}}
+\newcommand{\er}[1]{\myedit{magenta}{#1}}
+
 \algnewcommand\algorithmicforeach{\bf{for each}}
@@
 
-\newcommand{\here}[1]{\bf{[[#1]]}}
 \newcommand{\ceil}[1]{\lceil #1 \rceil}
@@
 
-\title{A Wait-free Queue with Poly-logarithmic Step Complexity}
+\title{A Wait-free Queue with Polylogarithmic Step Complexity}
 

### 2022-12-22 podc_paper/related.tex
some edits to related work

--- 2022-12-19 podc_paper/related.tex
@@
 As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
-Thus, it is lock-free but not wait-free and has an amortized step complexity
-of $\Theta(p)$ per operation.
+Thus, it is lock-free but not wait-free and has an amortized\Hossein{Maybe we can say worst-case.
+I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.}
+\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.
+Also, if we just said ``worst-case'' without ``amortized'' it would be confused with worst-case of an individual operation, which is infinite in this case.
+By the way, saying that previous implementations have $\Omega(p)$ step complexity, even in the amortized sense, makes the new contribution even more of a contrast.}
+step complexity of $\Theta(p)$ per operation.
 
@@
 a wait-free queue.
-Ramalhete and Correia \cite{RC17} added a different helping mechanism.
-In both cases, the helping mechanisms only add to the step complexity of the MS-queue.
+Then, Kogan and Petrank \cite{10.1145/2370036.2145835} introduced the 
+fast-path slow-path methodology for making data structures wait-free:
+The fast path has good performance and the slow path guarantees termination.
+They applied their methodology to combine the MS-queue (as the fast path)
+with their previous wait-free queue (as the slow path).
+Ramalhete and Correia \cite{RC17} added a different helping mechanism to the MS-queue.
+Although these approaches can improve practical performance,
+their amortized step complexity remains $\Omega(p)$. 
+\Hossein{I get what this sentence means, but from a high overview i think it is inaccurate.
+Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}
+\Eric{I modified the wording.
+I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussin array-based queues.}
 
@@
 % go back to (*) and repeat this sequence
-Similarly,
-other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
-The fast path has good performance and the slow path guarantees
-termination.
+This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
+Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} have also used this methodology
+on array-based queues.
 In worst-case executions, where processes must execute the slow path,

### 2022-12-23 podc_paper/algorithm.tex
Added some diagrams

@@
 sequence of enqueues and a sequence of dequeues.
+See Figure \ref{tournament} for an example.
 In this section, we assume for simplicity that each node has an infinite blocks array.
@@
 Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
-If so, it returns \nil\ and we call it a \emph{null dequeue}
+If so, it returns \var{null} and we call it a \emph{null dequeue}.
 If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
@@
 Since these tasks depends on the linearization ordering, we describe that ordering next.
+
+\begin{figure}[tb]
+\input{tournamentTree.pdf_t}
+\caption{An example tournament tree with four processes. 
+We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \var{blocks} arrays of the seven nodes.  The leftmost element of each \var{blocks} array is a dummy block.
+Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
+The fourth process's Deq\sub{6} is still in progress.
+The linearization order for this tree is
+Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
+\end{figure}
+
+\begin{figure}
+\input{implicit.pdf_t}
+\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
+The leaf blocks simply show the \var{element} field.
+Internal blocks show the \var{sum\sub{enq}} and \var{sum\sub{deq}} fields,
+and \eleft\ and \eright\ fields are shown using arrows as before.
+Root blocks also contain the additional \var{size} field.
+The \var{super} field is not shown.}
+\end{figure}
 
@@
 
-Finally, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
+Finally, when designing a total order for the operations propagated to the root, we choose
+to put each block's enqueues before its dequeues.
+Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
 linearization ordering is 
@@
 
-\here{Need some figures illustrating tournament tree, ordering of operations, representation as blocks.  Might be able to modify Fig \ref{fig::set} and \ref{fig:block}.}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture} [level 1/.style={level distance=1.8cm,sibling distance=0.6cm}, level 2/.style={level distance=1.4cm,sibling distance=0.4cm}]
-
-\Tree [.{$\big\{op_2^1,op_1^1,op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_2^2, op_4^2\big\},\big\{op_1^2\big\}...$ }  [.{ $\big\{op_2^1, op_1^1\big\},\big\{op_2^2\big\},\big\{op_1^2\big\}...$ }
-      {\begin{tabular}{|l|c|c|c}  \hline $op_1^1$ & $op_1^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c}  \hline $op_2^1$ & $op_2^2$ & ...
-\\ \hline \end{tabular}} ] [.{ $\big\{op_3^1\big\},\big\{op_4^1$, $op_3^2\big\},\big\{op_4^2\big\},...$ } {\begin{tabular}{|l|c|c|c}  \hline $op_3^1$ & $op_3^2$ & ...
-\\ \hline\end{tabular}} {\begin{tabular}{|l|c|c|c} \hline $op_4^1$ & $op_4^2$ & ...
-\\ \hline\end{tabular}} ] ]
-\end{tikzpicture}
-\caption[The operations merged in a \nf{Refresh} step with
-  sets.]{\label{fig::set} Leaves are for processes $P_1$ to $P_4$ from
-  left to right. In each internal node one can arbitrarily linearize
-  the sets of concurrent operations  propagated together in a
-  \nf{Refresh}. For example $op_4^1$ and $op_3^2$ have propagated
-  together in one \nf{Propagate} step and they will be propagated up
-  to the root together. Since their execution time intervals overlap,
-  they can be linearized in any order.} 
-\end{center}
-\end{figure}
-
-\begin{figure}[hbtp]
-\begin{center}
-\begin{tikzpicture}
-[level 1/.style={level distance=3.3cm,sibling distance=0cm},
-	level 2/.style={level distance=2.5cm,sibling distance=0cm},
-	level 3/.style={level distance=1.7cm,sibling distance=2cm}]
-\begin{small}
-
-\Tree [.{\begin{tabular}{|c|c|c|}  \hline (0,11) & (12,6) & (8,36) \\ \hline\end{tabular}}
- [.{\begin{tabular}{||c||c||}  \hline (10,2) & (5,3) \\\hline  \end{tabular}}
- [.{\begin{tabular}{||c|c||c||}\hline (2,3) & (4,1) & (0,5) \\\hline\end{tabular}} $\vdots$ $\vdots$ ]
-  [.{\begin{tabular}{||c||c||}  \hline (0,2) & (1,2)  \\ \hline  \end{tabular}} $\vdots$ $\vdots$ ] ]
-          [.{\begin{tabular}{||c||c||c|c||}  \hline (3,8) & (6,0) & (14,6) & (0,16) \\ \hline \end{tabular}}
-           [.{\begin{tabular}{||c||c||c|c||}  \hline (0,3) & (4,2) & (5,1) & (3,5) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ]
-            [.{\begin{tabular}{||c|c||c||c||}  \hline (2,1) & (5,0) & (4,2) & (9,7) \\ \hline \end{tabular}} $\vdots$ $\vdots$ ] ] ]
-\end{small}
-\end{tikzpicture}
-\end{center}
-\caption[The operations merged in a \nf{Refresh} step with (left,
-  right) blocks.]{\label{fig:block} Using blocks to represent
-  operations. Blocks between two lines $||$ are propagated together to
-  the parent. Each block consists of a pair (left, right) indicating
-  the number of operations from the left and the right child,
-  respectively. For example, (12,6) in the root contains  (10,2) from
-  the left child and (6,0) from the right child. The third block in
-  the root (8,36) is created by merging (5,3) from the left child and
-  (14,6) and (0,16) from the right child. (5,3) is the superblock of
-  (0,5) and (1,2) and (5,1),(3,5) and (4,2) are subblocks of (14,6).} 
-\end{figure}
-
 \subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
+\label{sec:fields}
 
@@
 We now explain how the fields of the blocks were chosen for this implicit representation. 
-See Figure \ref{object-fields}.
+Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} would actually be represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.
 
@@
 as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
-Then, for any non-null dequeue in $B_d$, we can also determine its
+Then, for any non-null dequeue in $B_d$, we can use this information to determine its
 rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
@@
 ---------
-
-In an execution on a queue where no dequeue operation returns
-\nf{null}, the $k$th dequeue returns the argument of the $k$th
-enqueue. In the general case a dequeue returns \nf{null} if and only
-if the queue size after the previous operation is 0. We refer to such
-a dequeue as a \it{null dequeue}. If the dequeue is the $k$th non-null
-dequeue, it returns the argument of the $k$th enqueue. Having the size
-of the queue after an operation we can compute the number of non-null
-dequeues from the number of enqueues before the operation. So, if we
-store the size of the queue after each block of operations in the
-root, we can compute the index of the enqueue whose argument is the
-response to a given dequeue in constant time.
 

### 2022-12-23 podc_paper/algorithm.tex
working on algorithm description

@@
 To simplify the code, each node has an empty block in \var{blocks[0]}.
-The node also has a \var{head} index which is the position in the \var{blocks} array to be used
+Each node also has a \var{head} index that stores the position in the \var{blocks} array to be used
 for the next attempt to append a block to this node.
@@
 If a block contained an explicit representation of its sequences of enqueues and dequeues,
-it would take $\Omega(p)$ time to construct a block, which would be much too slow for task \ref{construct}.
+it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
 Instead, the block stores an implicit representation of the sequences.
-We now explain how the fields of the blocks were chosen for this implicit representation. 
-Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} would actually be represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.
+We now explain how we designed the fields for this implicit representation. 
+Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.
 
@@
 enqueued for an enqueue operation, or \var{null} if the operation is a dequeue.
+
+\here{If space permits, we might want to add some examples in the following paragraphs that refer back to Figure \ref{implicit}.}
 
@@
 If we know which block $B$ in some node $v$ contains the enqueue,
-we can use the \var{sum\sub{enq}} field again determine which child of $v$ contains the enqueue
+we can use the \var{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
 and then to do a binary search
@@
 Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
-For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block,
-the number of enqueues in $B$ and the number of non-null dequeues in $B$ (determined as described above).
+For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block $B'$ and
+the number of enqueues and dequeues in $B$:
+$B.\var{size} = \max(0, B'.\var{size} + (B.\var{sum\sub{enq}}-B'.\var{sum\sub{enq}}) - (B.\var{sum\sub{deq}} - B'.\var{sum\sub{deq}}))$.
 
@@
 \begin{figure}
+\here{Does \var{leaf} have to be shared for helping when doing GC?}
 \begin{algorithmic}[1]
 \setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared Variable}}
+\Statex $\diamondsuit$ \tt{\sl{Shared variable}}
 \begin{itemize}
-\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
+\item \tt{\sl{Node} root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
 \end{itemize}
 
-\Statex $\diamondsuit$ \tt{\sl{Local Variable}}
+\Statex $\diamondsuit$ \tt{\sl{Thread-local variable}} 
 \begin{itemize}
-\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
+\item \tt{\sl{Node} leaf} \Comment{process's leaf in the tree}
 \end{itemize}
@@
 \begin{itemize}
-\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
-\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
-\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
+\item \tt{\sl{Node} left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
+\item \tt{\sl{Block[0..$\infty$]} blocks} \Comment{initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
+\item \tt{\sl{int} \head} \Comment{position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
 \end{itemize}
@@
 \begin{itemize}
+  	\item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+  		\Comment{number of enqueue, dequeue operations in \var{blocks} array up to this block (inclusive)}
   	\item \tt{\sl{int} super}
-  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
+  		\Comment{approximate index of the block's superblock in \var{parent.blocks}}
+	\item[\com] Blocks in internal nodes have the following additional fields
+	\item \tt{int} \eleft, \eright
+  		\Comment{index of last direct subblock in the left and right child}
+  	\item[\com] Blocks in leaf nodes have the following additional field
+  	\item \tt{\sl{Object} element}
+  		\Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+	\item[\com] Blocks in the root node have the following additional field
+	\item \tt{\sl{int} \size}%
+  		\Comment{size of the queue after performing all operations up to the end of this block}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
-\begin{itemize}
-    \item \tt{int} \eleft, \eright
-  \textsf{\com\ indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
+%\Statex {\com\  Blocks in internal nodes have the following additional fields}
+%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
+%\begin{itemize}
+%	\item \tt{int} \eleft, \eright
+%  		\Comment{index of last direct subblock in the left and right child}
+%\end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+%\Statex {\com\ Blocks in leaf nodes have the following additional field}
+%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
+%\begin{itemize}
+%  \item \tt{\sl{Object} element}
+%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+%\end{itemize}
 
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
-\begin{itemize}
-  \item \tt{\sl{int} \size}%
-  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
-\end{itemize}
+%\Statex {\com\ Blocks in the root node have the following additional field}
+%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
+%\begin{itemize}
+%  \item \tt{\sl{int} \size}%
+%  \Comment{size of the queue after performing all operations up to the end of this block}
+%\end{itemize}
 
@@
 \end{figure}
-
----------
-
-In our case of implementing a queue, a process only needs to compute
-the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific
-position.  We know we can linearize operations in a block in any
-order; here, we choose to put \nf{Enqueue} operations in a block
-before \nf{Dequeue} operations. Consider the following operations,
-where operations in a cell are concurrent. 
-\begin{table}[hbtp]
-\centering
-\begin{small}
-\begin{tabular}{c|c|c|c}
-    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-  \end{tabular}
-\end{small}
-\caption{An execution on a queue separated by the operations in the root blocks.}
-\end{table}
-
-\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2,
-  1, 3, 4, null}, respectively. 
-Now, we claim that by knowing the size of the queue, we can compute
-the rank of the required \texttt{Enqueue} for any non-null
-\nf{Dequeue}. We apply this approach to blocks; if we store the size
-of the queue after each block of operations, we can compute the index
-of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.
-
-\begin{table}[hbtp]
-\centering
-\begin{small}
-  \begin{tabular}{c|c|c|c|c}
-    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
-    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
-    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
-    Size at end & 0 & 2 & 2 & 0 \\ \hline
-  \end{tabular}
-\end{small}
-  \caption{Augmented history of operation blocks on the queue.}
-\end{table}
-
-\noindent The size of the queue after the $b$th block in the root
-could be computed as 
-$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} +
-\#\textrm{\tt{Enqueue}s in }b\textrm{th block} -
-\#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$ 
-\noindent Moreover, the total number of non-null dequeues in blocks
-$1,2,...,b$ in the root is 
-$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} -
-\textrm{size after }b\textrm{th block}.$$
-
-Given a \texttt{Dequeue} is in block $B$, its response is the argument
-of the \tt{Enqueue} whose rank is 
-\begin{center}
-    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
-\end{center}
-if \vspace{-3em}
-\begin{center}
-    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
-\end{center}
-Otherwise, the response would be \nf{null}.
 

### 2022-12-23 podc_paper/introduction.tex
Update introduction.tex
working on algorithm description

@@
 Besides being a fundamental data structure, queues are used in
-significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
+significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
 packet processing \cite{DPDK},
@@
 The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
-IT uses a singly-linked list with pointers to the front and back nodes.
+It uses a singly-linked list with pointers to the front and back nodes.
 To dequeue or enqueue an element, the front or back pointer is updated by a 

### 2022-12-28 podc_paper/algorithm.tex
turned comments back on
working on converting code to double column (and non-object-oriented) to save space

@@
 
-Each node has an infinite array called \var{blocks}.
-To simplify the code, each node has an empty block in \var{blocks[0]}.
-Each node also has a \var{head} index that stores the position in the \var{blocks} array to be used
-for the next attempt to append a block to this node.
+Each \node\ of the tournament tree has an infinite array called \var{blocks}.
+To simplify the code, each \node\ has an empty \block\ in \var{blocks[0]}.
+Each \node\ also has a \var{head} index that stores the position in the \var{blocks} array to be used
+for the next attempt to append a \block\ to this \node.
 
@@
 We now explain how we designed the fields for this implicit representation. 
-Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.
+Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.
 
+The information stored in a \block\ depends on whether it is in an internal node or a leaf.
 A block in a leaf represents a single enqueue or dequeue.  The \var{element} field stores the value
@@
 
-Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.
+Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.
 This allows us to navigate to the direct subblocks of any block easily.
@@
 to locate which block in the root contains the operation.
-If we know which block $B$ in some node $v$ contains the enqueue,
+If we know which block $B$ in some \node\ $v$ contains the enqueue,
 we can use the \var{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
@@
 We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
-and the binary search within each other node along the path to a leaf can be done in $O(\log p)$ steps,
+and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
 for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
@@
 To facilitate task \ref{findinroot}, each block has a field \var{super} that contains
-the (approximate) index of its superblock in the parent node's blocks array.
+the (approximate) index of its superblock in the parent \node's \var{blocks} array.
 We will ensure that this field's value differs from the true index of the superblock by at most 1.
-This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two blocks in the parent node.
+This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
 Thus, starting from an operation in a leaf's block, one can use these indices to track the 
@@
 we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
-A \op{refresh} on node $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
+A \op{refresh} on \node\ $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
 $h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
@@
 The only remaining field is $B.\var{super}$.  When the block 
-$B$ is created for a node $v$, we do not yet know where its
+$B$ is created for a \node\ $v$, we do not yet know where its
 superblock will eventually be installed in $B$'s parent.
@@
 
+\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
 \begin{figure}
@@
 \setcounter{ALG@line}{1}
-\Statex $\diamondsuit$ \tt{\sl{Shared variable}}
+\Statex \linecomment{Shared variable}
 \begin{itemize}
-\item \tt{\sl{Node} root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
+\item \var{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
 \end{itemize}
 
-\Statex $\diamondsuit$ \tt{\sl{Thread-local variable}} 
+\Statex \linecomment{Thread-local variable}
 \begin{itemize}
-\item \tt{\sl{Node} leaf} \Comment{process's leaf in the tree}
+\item \var{Node} \var{leaf} \Comment{process's leaf in the tree}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
+\Statex $\blacktriangleright$ \var{Node}
 \begin{itemize}
-\item \tt{\sl{Node} left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
-\item \tt{\sl{Block[0..$\infty$]} blocks} \Comment{initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
-\item \tt{\sl{int} \head} \Comment{position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
+\item \var{Node left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
+\item \var{Block[0..$\infty$] blocks} \Comment{initially \var{blocks[0]} contains an empty block with all integer fields equal to 0}
+\item \var{int \head} \Comment{position to attempt appending next \block\ to \var{blocks}, initially 1}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
+\Statex $\blacktriangleright$ \var{Block}
 
 \begin{itemize}
-  	\item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
+  	\item \var{int sum\sub{enq}, sum\sub{deq}}
   		\Comment{number of enqueue, dequeue operations in \var{blocks} array up to this block (inclusive)}
-  	\item \tt{\sl{int} super}
+  	\item \var{int super}
   		\Comment{approximate index of the block's superblock in \var{parent.blocks}}
-	\item[\com] Blocks in internal nodes have the following additional fields
-	\item \tt{int} \eleft, \eright
+	\item[\com] Blocks in internal \nodes\ have the following additional fields
+	\item \var{int} \eleft, \eright
   		\Comment{index of last direct subblock in the left and right child}
-  	\item[\com] Blocks in leaf nodes have the following additional field
-  	\item \tt{\sl{Object} element}
-  		\Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-	\item[\com] Blocks in the root node have the following additional field
-	\item \tt{\sl{int} \size}%
+  	\item[\com] Blocks in leaf \nodes\ have the following additional field
+  	\item \var{Object element}
+  		\Comment{if the block's operation is \op{Enqueue}(\var{x}) then \var{element=x}, otherwise \var{element=null}.}
+	\item[\com] Blocks in the root \node\ have the following additional field
+	\item \var{int \size}%
   		\Comment{size of the queue after performing all operations up to the end of this block}
@@
 \end{algorithmic}
-\caption{Objects used in Tournament Tree Data Structure \label{object-fields}}
+\caption{Objects used in tournament tree data structure \label{object-fields}}
 \end{figure}
@@
 \subsection{Details of the Implementation}
-Section \ref{algQ} gives the pseudocode for the queue
-implementation. It uses the following two types of objects. 
-\paragraph{\tt{Node}} 
- In each \texttt{Node} we store pointers to its parent and its left
- and right child, an array of \nf{Block}s called \texttt{block}s and
- the index \nf{head} of the first empty entry in \texttt{blocks}.
 
-\paragraph{\tt{Block}}
- The information stored in a \nf{Block} depends on whether  the
- \nf{Block} is in an internal node or a leaf. If it is in a leaf, we
- use a \nf{LeafBlock} which stores one operation. If a block $B$ is in
- an internal node $n$, then it contains subblocks in the left and
- right children of $n$. The left subblocks of $B$ are some consecutive
- blocks in the left child of $n$ starting from where the left
- subblocks of the block prior to $B$ ended. The right subblocks of $B$
- are defined similarly. In each \texttt{block} we store four essential
- fields that implicitly summarize which operations are in the block
- \texttt{sum\textsubscript{enq-left}},
- \texttt{sum\textsubscript{deq-left}},
- \texttt{sum\textsubscript{enq-right}},
- \texttt{sum\textsubscript{deq-right}}. The
- \texttt{sum\textsubscript{enq-left}} field is the total number of
- \nf{Enqueue} operations in the blocks before the last subblock of $B$
- in the left child. The other fields' semantics are similar. The
- \nf{end\sub{left}} and \nf{end\sub{right}} field store the index of
- the last subblock of a block in the left and the right child,
- respectively. The approximate index of the superblock of non-root
- blocks is stored in their \nf{super} field. The \texttt{size} field
- in a block in the root node stores the size of the queue after the
- operations in the block have been performed.
+We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{?}.
+We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
+$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
+the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
+We use a similar abbreviation \var{num\sub{deq}}.
 
-We now describe the routines used in the implementation.
+\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}
+\newcommand{\hangbox}[1]{\parbox[t]{\dimexpr\linewidth-\algorithmicindent\relax}{%
+    \setlength{\hangindent}{\algorithmicindent}%
+    #1}}
+\newcommand{\spac}{\medskip}
+\newcommand{\assign}{\mbox{:=}}
 
-\paragraph{\tt{Enqueue($e$)}}
-An \nf{Enqueue} operation does not return a response, so it is
-sufficient to propagate the \nf{Enqueue} operation to the root and
+\begin{figure}
+\begin{minipage}[t]{0.46\textwidth}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{0}
+
+\Function{void}{Enqueue}{\var{Object e}} 
+    \State \hangbox{let \var{new} be a new \var{Block} with \var{element} \assign\ \var{e},\\
+		\var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}+1},\\
+		\var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}}}
+    \State \Call{Append}{new}
+\EndFunction{Enqueue}
+
+\spac
+
+\Function{Object}{Dequeue()}{} 
+    \State \hangbox{let \var{n} be a new \var{Block} with \var{element} \assign\ \var{null},\\
+	    \var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}},\\
+	    \var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}+1}}
+    \State \Call{Append}{n}
+    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf.\head, 1}
+    \State \Return{ \Call{FindResponse}{b, i}}
+    \label{deqRest}
+\EndFunction{Dequeue}
+
+\spac
+
+\Function{void}{Append}{\var{Block B}} 
+    \State \linecomment{append block to leaf and propagate to root}
+    \State \var{leaf.blocks[leaf.\head] \assign\ B} \label{appendLeaf}
+    \State \var{leaf.\head\ \assign\ leaf.\head+1} \label{appendEnd} 
+    \State \Call{Propagate}{\var{leaf.parent}} 
+\EndFunction{Append}
+
+\spac
+
+\Function{void}{Propagate}{\var{Node v}}
+    \State \linecomment{propagates blocks from \var{v}'s children to root}
+    \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
+        \State \Call{Refresh}{v} \label{secondRefresh}
+    \EndIf
+    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
+        \State \Call{Propagate}{v.parent}
+    \EndIf
+\EndFunction{Propagate}
+
+\spac
+
+\Function{boolean}{Refresh}{\var{Node v}}
+    \State \linecomment{tries to append a new block to \var{v.blocks}}
+    \State \var{h \assign\ v.\head} \label{readHead}
+    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}} \label{startHelpChild1}
+        \State \var{h\sub{dir} \assign\ v.dir.\head} \label{readChildHead}
+        \If{\var{v.dir.blocks[h\sub{dir}] $\neq$ null}} \label{ifHeadnotNull}
+            \State{\Call{Advance}{v.dir, h\sub{dir}}} \label{helpAdvance}
+        \EndIf
+    \EndFor \label{endHelpChild1}
+    \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
+    \If{\var{new.num = 0}} \Return{\var{true}} \label{addOP} 
+    \EndIf
+    \State{\var{result} \assign\ \var{CAS(v.blocks[h], null, new)}} \label{cas}
+    \State{\var{\Call{Advance}{v, h}}} \label{advance}
+    \State \Return{ \tt{result}}
+\EndFunction{Refresh}
+
+\spac
+
+\Function{void}{Advance}{\var{Node v}, \var{int h}} 
+    \State \linecomment{set \nf{v.blocks[h].super} and }
+    \State \linecomment{increment \nf{v.head} from \nf{h} to \nf{h+1}}
+    \State \var{h\sub{p}} \assign\ \var{v.parent.\head} \label{readParentHead}
+    \State \var{CAS(v.blocks[h].super, null, h\sub{p})} \label{setSuper1}
+    \State \var{CAS(v.head, h, h+1)} \label{incrementHead}
+\EndFunction{Advance}
+
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.53\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{48}
+
+\Function{Block}{CreateBlock}{\var{Node v, int i}} 
+    \State\linecomment{create new block to be installed in \var{v.blocks[i]}}
+    \State let \var{new} be a new Block \label{initNewBlock}
+    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}}
+        \State \var{index\sub{prev} \assign\ v.blocks[i-1].\edir} \label{prevLine}
+        \State \var{new.\edir\ \assign\ v.dir.\head-1} \label{lastLine}
+        \State \linecomment{\var{new} contains \var{v.dir.blocks[index\sub{prev}+1..new.\edir]}}
+        \State \var{block\sub{prev} \assign\ v.dir.blocks[index\sub{prev}]}
+        \State \var{block\sub{last} \assign\ v.dir.blocks[new.\edir]}
+    \EndFor
+	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{enq} + \\
+			v.right.blocks[\eright].sum\sub{enq}}}
+	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{deq} + \\
+			v.right.blocks[\eright].sum\sub{deq}}}
+    \If{$\var{v} = \var{root}$}
+        \State \hangbox{\var{new.size \assign\ max(0, \\ v.blocks[i-1].\size\ + new.num\sub{enq} - new.num\sub{deq})}\label{computeLength}}
+    \EndIf
+    \State \Return \tt{new}
+\EndFunction{CreateBlock}
+
+\spac
+
+\Function{element}{FindResponse}{\var{int b, int i}}
+\State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
+\If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
+\State \linecomment{queue is empty when the dequeue is linearized}
+\State \Return \var{null} \label{returnNull}
+\Else \ \linecomment{response is the \var{e}th enqueue in the root}
+\State \var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - root.blocks[b-1].size}\label{computeE}
+\State \Return \var{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
+\EndIf
+\EndFunction{FindResponse}
+
+\end{algorithmic}
+\end{minipage}
+\caption{\tt{\sl{Queue}}}
+\end{figure}
+
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{25}
+
+\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
+\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
+\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
+% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
+\While{\nf{start<end}}
+\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
+\If{\nf{blocks[mid].sum\sub{enq}<x}}
+\State \nf{start \assign\ mid+1}
+\Else
+\State \nf{end \assign\ mid}
+\EndIf
+\EndWhile
+\State\Return \nf{start}
+\EndFunction{BinarySearch}
+
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\tt{\sl{Root}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{36}
+\Statex
+\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
+\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
+\com\ the \nf{b}th block in the \nf{root}.}
+
+\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
+\State \tt{start \assign\ end-1} \label{dsearchStart}
+\While{\tt{root.blocks[start].sum\sub{enq}}$> \assign\ $\tt{e}}
+\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
+\EndWhile \label{dsearchEnd}
+\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
+\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
+\State\Return \tt{<b,i>}
+\EndFunction{DoublingSearch}
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\tt{\sl{Node}}}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{94}
+
+\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
+\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
+\If{\tt{this} \bf{is} \tt{leaf}}
+\State\Return \tt{blocks[b].element} \label{getBaseCase}
+\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
+\State \tt{subblockIndex \assign left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
+\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
+\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
+\Else
+\State \tt{i \assign\ i-blocks[b].num\sub{enq-left}}
+\State\tt{subblockIndex \assign\ right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
+\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
+\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
+\EndIf
+\EndFunction{GetEnqueue}
+
+\Statex
+\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
+\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b}) \assign\ D_\tt{y}(\tt{root},\tt{x})$.}
+\If{\tt{this} \bf{is} \tt{root}}
+\State\Return \tt{<b, i>} \label{indexBaseCase}
+\Else
+\State \tt{dir\assign\ (parent.left=n ? left: right)} 
+\State \tt{superblockIndex \assign\ parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
+\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
+
+\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
+\Else \label{considerRight}
+\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
+\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
+\EndIf \label{computeISuperEnd}
+\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
+\EndIf
+\EndFunction{IndexDequeue}
+
+\end{algorithmic}
+\end{algorithm}
+
+An \op{Enqueue}($e$) operation appends a \block\ to the process's leaf.
+The block has $\var{element}=e$ and its \var{sum\sub{enq}} and \var{sum\sub{deq}} fields are set
+on line \ref{?}.
+It suffices to propagate the operation to the root and
 then use its position in the linearization for future \nf{Dequeue}
-operations. \nf{Enqueue($e$)} creates a \nf{LeafBlock} with
-$\nf{element}=e$, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}}
-fields and then appends it to the tree.
+operations.
 
-\paragraph{\tt{Dequeue()}}
-\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and
-\nf{sum\sub{deq}} fields and appends it to the tree. Then, it computes
-the position of the appended \nf{Dequeue} operation in the root using
-\nf{IndexDequeue} and after that finds the response of the
-\nf{Dequeue} by calling \nf{FindResponse}.
+A \op{Dequeue} also appends a \block\ to the process's leaf.
+The block has $\var{element}=null$ and its \var{sum\sub{enq}} and \var{sum\sub{deq}} fields are set
+on line \ref{?}.
+After propagating the operation to the root, it computes
+its position in the root using
+\nf{IndexDequeue} and then computes its response by calling \nf{FindResponse}.
 

### 2022-12-28 podc_paper/podc-submission.tex
working on algorithm description
turned comments back on
working on converting code to double column (and non-object-oriented) to save space

@@
 
+\newcommand{\node}{node}
+\newcommand{\nodes}{nodes}
+\newcommand{\block}{block}
+\newcommand{\blocks}{blocks}
+
 \algnewcommand\algorithmicforeach{\bf{for each}}
@@
 \algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
+   [3]{{\var{{#1}}} {\op{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
 
 \algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
+   [1]{\algorithmicend\ \op{{#1}}}
 
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+\algrenewcommand\Call[2]{\var{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+\usepackage{eqparbox}
 
@@
 \newcommand{\nil}{nil}
-\newcommand{\opa}[2]{\nf{#1(#2)}}
-\newcommand{\op}[1]{\nf{#1}}
-\newcommand{\var}[1]{\nf{#1}}
+\newcommand{\opa}[2]{\sf{#1(#2)}}
+\newcommand{\op}[1]{\sf{#1}}
+\newcommand{\var}[1]{\sf{#1}}
 \newcommand{\com}{$\triangleright$}
+\newcommand{\linecomment}[1]{\com\ #1}
 
@@
 
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex \com\ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex \com\ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
-\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start= mid+1}
-\Else
-\State \nf{end= mid}
-\EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
-\com\ the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start= end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
-\State \tt{start= max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{blocks[\head]= B} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex \com\ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex \com\ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].\edir} \label{prevLine}
-\State \tt{new.\edir= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].\edir..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.\edir]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
 \input{correctness}

### 2022-12-28 podc_paper/algorithm.tex
working on code

@@
 
-We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{?}.
+We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode}.
 We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
@@
 the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
-We use a similar abbreviation \var{num\sub{deq}}.
+We use \var{num\sub{deq}} similarly.
+\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}
 
@@
 \begin{figure}
-\begin{minipage}[t]{0.46\textwidth}
+\begin{minipage}[t]{0.465\textwidth}
 \begin{algorithmic}[1]
@@
     \State \Call{Append}{n}
-    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf.\head, 1}
+    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf, leaf.\head, 1}
     \State \Return{ \Call{FindResponse}{b, i}}
@@
 \Function{void}{Propagate}{\var{Node v}}
-    \State \linecomment{propagates blocks from \var{v}'s children to root}
+    \State \linecomment{propagate blocks from \var{v}'s children to root}
     \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
@@
 \Function{boolean}{Refresh}{\var{Node v}}
-    \State \linecomment{tries to append a new block to \var{v.blocks}}
+    \State \linecomment{try to append a new block to \var{v.blocks}}
     \State \var{h \assign\ v.\head} \label{readHead}
@@
     \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
-    \If{\var{new.num = 0}} \Return{\var{true}} \label{addOP} 
+    \If{\var{new = null}} \Return{\var{true}} \label{addOP} 
     \EndIf
@@
 
-\Function{void}{Advance}{\var{Node v}, \var{int h}} 
+\Function{void}{Advance}{\var{Node v, int h}} 
     \State \linecomment{set \nf{v.blocks[h].super} and }
@@
 \end{minipage}
-\begin{minipage}[t]{0.53\textwidth}
+\begin{minipage}[t]{0.525\textwidth}
 
@@
 \Function{Block}{CreateBlock}{\var{Node v, int i}} 
-    \State\linecomment{create new block to be installed in \var{v.blocks[i]}}
-    \State let \var{new} be a new Block \label{initNewBlock}
-    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}}
-        \State \var{index\sub{prev} \assign\ v.blocks[i-1].\edir} \label{prevLine}
-        \State \var{new.\edir\ \assign\ v.dir.\head-1} \label{lastLine}
-        \State \linecomment{\var{new} contains \var{v.dir.blocks[index\sub{prev}+1..new.\edir]}}
-        \State \var{block\sub{prev} \assign\ v.dir.blocks[index\sub{prev}]}
-        \State \var{block\sub{last} \assign\ v.dir.blocks[new.\edir]}
-    \EndFor
-	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{enq} + \\
-			v.right.blocks[\eright].sum\sub{enq}}}
-	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{deq} + \\
-			v.right.blocks[\eright].sum\sub{deq}}}
+    \State\linecomment{create new block for a \op{Refresh} to install in \var{v.blocks[i]}}
+    \State\linecomment{(return \var{null} if no blocks need to be propagated to \var{v})}
+    \State let \var{new} be a new \var{Block} \label{initNewBlock}
+    \State \var{new.\eleft} \assign\ $\var{v.left.head} - 1$
+    \State \var{new.\eright} \assign\ $\var{v.right.head} - 1$
+	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq} + \\
+			v.right.blocks[new.\eright].sum\sub{enq}}}
+	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq} + \\
+			v.right.blocks[new.\eright].sum\sub{deq}}}
+    \State \var{num\sub{enq}} \assign\ \var{new.sum\sub{enq} - v.blocks[i-1].sum\sub{enq}}
+    \State \var{num\sub{deq}} \assign\ \var{new.sum\sub{deq} - v.blocks[i-1].sum\sub{deq}}
     \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new.size \assign\ max(0, \\ v.blocks[i-1].\size\ + new.num\sub{enq} - new.num\sub{deq})}\label{computeLength}}
+        \State \hangbox{\var{new.size \assign\ max(0, v.blocks[i-1].\size\ + num\sub{enq} - num\sub{deq})}\label{computeLength}}
     \EndIf
-    \State \Return \tt{new}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
+        \State \Return \var{null}
+    \Else
+        \State \Return \var{new}
+    \EndIf
 \EndFunction{CreateBlock}
+
+\spac
+
+\Function{<int, int>}{IndexDequeue}{\var{Node v, int b, int i}}
+    \State \linecomment{return \var{<x, y>} such that \var{i}th dequeue in D(\var{v.blocks[b]})}
+    \State \linecomment{is \var{y}th dequeue of D(\var{root.blocks[x]})}
+    \State \linecomment{Precondition: \var{v.blocks[b]} has been propagated to root}
+    \State \linecomment{and $\var{v.blocks[b].num\sub{deq}}\geq \var{i}$}
+    \If{$\var{v} = \var{root}$}
+        \State\Return \var{<b, i>} \label{indexBaseCase}
+    \Else
+	    \State \var{dir \assign\ (v.parent.left = v ? left : right)} 
+    	\State \var{sup \assign\ v.blocks[b].super}
+	    \If{$\var{b} > \var{v.parent.blocks[sup].end\sub{dir}}$}
+	        \State{\var{sup} \assign\ $\var{sup}+1$}
+	    \EndIf
+	    \State \hangbox{\var{i += v.blocks[b-1].sum\sub{deq} - \\
+	    		v.blocks[v.parent.blocks[sup-1].end\sub{dir}].sum\sub{deq}}}
+        \If{$\var{dir} = \var{right}$} 
+        	\State \hangbox{\var{i += v.blocks[v.parent.blocks[sup].end\sub{left}].sum\sub{deq} -\\
+					v.blocks[v.parent.blocks[sup-1].end\sub{left}].sum\sub{deq}}\label{considerLeftBeforeRight}}
+        \EndIf \label{computeISuperEnd}
+        \State \Return\Call{IndexDequeue}{v.parent, sup, i}
+    \EndIf
+\EndFunction{IndexDequeue}
 
@@
     \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-\State \var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - root.blocks[b-1].size}\label{computeE}
+        \State \hangbox{\var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - \\
+	root.blocks[b-1].size}\label{computeE}}
         \State \Return \var{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
@@
 \end{minipage}
-\caption{\tt{\sl{Queue}}}
+\caption{Queue implementation.\label{pseudocode}}
 \end{figure}
@@
 \EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b}) \assign\ D_\tt{y}(\tt{root},\tt{x})$.}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir\assign\ (parent.left=n ? left: right)} 
-\State \tt{superblockIndex \assign\ parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
 

### 2022-12-29 podc_paper/algorithm.tex
working on code

@@
 \begin{figure}
-\begin{minipage}[t]{0.465\textwidth}
+\begin{minipage}[t]{0.461\textwidth}
 \begin{algorithmic}[1]
@@
 \end{minipage}
-\begin{minipage}[t]{0.525\textwidth}
+\begin{minipage}[t]{0.533\textwidth}
 
@@
     \State \var{new.\eright} \assign\ $\var{v.right.head} - 1$
-	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq} + \\
-			v.right.blocks[new.\eright].sum\sub{enq}}}
-	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq} + \\
-			v.right.blocks[new.\eright].sum\sub{deq}}}
+	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq}} + \\
+			\var{v.right.blocks[new.\eright].sum\sub{enq}}}
+	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq}} + \\
+			\var{v.right.blocks[new.\eright].sum\sub{deq}}}
     \State \var{num\sub{enq}} \assign\ \var{new.sum\sub{enq} - v.blocks[i-1].sum\sub{enq}}
@@
 	    \EndIf
-	    \State \hangbox{\var{i += v.blocks[b-1].sum\sub{deq} - \\
-	    		v.blocks[v.parent.blocks[sup-1].end\sub{dir}].sum\sub{deq}}}
+	    \State \linecomment{compute index \var{i} of dequeue in superblock}
+	    \State \hangbox{\var{i += v.blocks[b-1].sum\sub{deq}} - \\
+	    		\var{v.blocks[v.parent.blocks[sup-1].end\sub{dir}].sum\sub{deq}}}
         \If{$\var{dir} = \var{right}$} 
-        	\State \hangbox{\var{i += v.blocks[v.parent.blocks[sup].end\sub{left}].sum\sub{deq} -\\
-					v.blocks[v.parent.blocks[sup-1].end\sub{left}].sum\sub{deq}}\label{considerLeftBeforeRight}}
+        	\State \hangbox{\var{i += v.blocks[v.parent.blocks[sup].end\sub{left}].sum\sub{deq}} -\\
+					\var{v.blocks[v.parent.blocks[sup-1].end\sub{left}].sum\sub{deq}}\label{considerLeftBeforeRight}}
         \EndIf \label{computeISuperEnd}
@@
     \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-        \State \hangbox{\var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - \\
-	root.blocks[b-1].size}\label{computeE}}
-        \State \Return \var{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
+        \State $\var{e} \assign\ \var{i} + \var{root.blocks[b-1].sum\sub{enq}} - 
+	\var{root.blocks[b-1].size}$\label{computeE}
+		\State \linecomment{compute enqueue's block using binary search}
+		\State find min $\var{b}'\leq \var{b}$ with $\var{root.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{e}$
+		\State \linecomment{find rank of enqueue within its block}
+		\State $\var{i}' \assign\ \var{e} - \var{root.blocks}[\var{b}'-1].\var{sum\sub{enq}}$
+        \State \Return \Call{GetEnqueue}{root, $\var{b}'$, $\var{i}'$}\label{findAnswer}
     \EndIf
@@
 \State \tt{start \assign\ end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$> \assign\ $\tt{e}}
+\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
 \State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
@@
 \Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
+\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
 \If{\tt{this} \bf{is} \tt{leaf}}

### 2022-12-29 podc_paper/algorithm.tex
working on code

@@
 
-We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode}.
+We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
 We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
@@
     \If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
-        \State \linecomment{queue is empty when the dequeue is linearized}
-        \State \Return \var{null} \label{returnNull}
+        \State \Return \var{null} \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
     \Else \ \linecomment{response is the \var{e}th enqueue in the root}
@@
 
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
+%\begin{algorithm}
+%\caption{\tt{\sl{Node}}}
+%\begin{algorithmic}[1]
+%\setcounter{ALG@line}{25}
+%
+%\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
+%\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
+%\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
+%\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
+% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
+%\While{\nf{start<end}}
+%\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
+%\If{\nf{blocks[mid].sum\sub{enq}<x}}
+%\State \nf{start \assign\ mid+1}
+%\Else
+%\State \nf{end \assign\ mid}
+%\EndIf
+%\EndWhile
+%\State\Return \nf{start}
+%\EndFunction{BinarySearch}
+%
+%\end{algorithmic}
+%\end{algorithm}
+
+%\begin{algorithm}
+%\caption{\tt{\sl{Root}}}
+%\begin{algorithmic}[1]
+%\setcounter{ALG@line}{36}
+%\Statex
+%\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
+%\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
+%\com\ the \nf{b}th block in the \nf{root}.}
+%
+%\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
+%\State \tt{start \assign\ end-1} \label{dsearchStart}
+%\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
+%\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
+%\EndWhile \label{dsearchEnd}
+%\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
+%\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
+%\State\Return \tt{<b,i>}
+%\EndFunction{DoublingSearch}
+%\end{algorithmic}
+%\end{algorithm}
+
+\begin{figure}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
+\setcounter{ALG@line}{101}
 
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
-\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-\While{\nf{start<end}}
-\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
-\If{\nf{blocks[mid].sum\sub{enq}<x}}
-\State \nf{start \assign\ mid+1}
+\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \Comment{returns argument of ith enqueue in $E(\var{v.blocks[b]})$}
+    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v.blocks[b]})$ contains at least \var{i} enqueues}
+
+    \If{\var{v} is a leaf node}
+        \State\Return \var{blocks[b].element} \label{getBaseCase}
     \Else 
-\State \nf{end \assign\ mid}
+        \State \var{sum\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}$] from left child}
+        \State \var{prev\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b-1].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from left child}
+        \State \var{prev\sub{right}} \assign\ \var{v.right.blocks[v.blocks[b-1].\eright].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from right child}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v.left}}
+        	\State \linecomment{Find enqueue's block in \var{v.left} and its rank within block}
+            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eleft}+1..\var{v.blocks[b].\eleft}] s.t. $\var{v.left.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{left}}$
+            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.left.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{left}})$
+            \State \Return\Call{GetEnqueue}{\var{v.left}, $\var{b}'$, $\var{i}'$} 
+        \Else \Comment{required enqueue is in \var{v.right}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
+        	\State \linecomment{Find enqueue's block in \var{v.right} and its rank within block}
+            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eright}+1..\var{v.blocks[b].\eright}] s.t. $\var{v.right.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{right}}$
+            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.right.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{right}})$
+            \State \Return\Call{GetEnqueue}{\var{v.right}, $\var{b}'$, $\var{i}'$} 
         \EndIf
-\EndWhile
-\State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Root}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{36}
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
-\com\ the \nf{b}th block in the \nf{root}.}
-
-\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-\State \tt{start \assign\ end-1} \label{dsearchStart}
-\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
-\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
-\EndWhile \label{dsearchEnd}
-\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-\State\Return \tt{<b,i>}
-\EndFunction{DoublingSearch}
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex \assign left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i \assign\ i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex \assign\ right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
     \EndIf
 \EndFunction{GetEnqueue}
-
 \end{algorithmic}
-\end{algorithm}
+\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter be the rank within whole node rather than within the block?}}
+\end{figure}
 

### 2022-12-31 podc_paper/algorithm.tex
working on alg description, tidying code

@@
 
-\subsection{Overview of Implementation}
+\subsection{Overview}
 We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
@@
 Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
-If so, it returns \var{null} and we call it a \emph{null dequeue}.
+If so, it returns \nl\ and we call it a \emph{null dequeue}.
 If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
@@
 \caption{An example tournament tree with four processes. 
-We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \var{blocks} arrays of the seven nodes.  The leftmost element of each \var{blocks} array is a dummy block.
+We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
 Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
@@
 \caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
-The leaf blocks simply show the \var{element} field.
-Internal blocks show the \var{sum\sub{enq}} and \var{sum\sub{deq}} fields,
+The leaf blocks simply show the \fld{element} field.
+Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
 and \eleft\ and \eright\ fields are shown using arrows as before.
-Root blocks also contain the additional \var{size} field.
-The \var{super} field is not shown.}
+Root blocks also contain the additional \fld{size} field.
+The \fld{super} field is not shown.}
 \end{figure}
@@
 
-Each \node\ of the tournament tree has an infinite array called \var{blocks}.
-To simplify the code, each \node\ has an empty \block\ in \var{blocks[0]}.
-Each \node\ also has a \var{head} index that stores the position in the \var{blocks} array to be used
+Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
+To simplify the code, each \node\ has an empty \block\ in \fld{blocks[0]}.
+Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
 for the next attempt to append a \block\ to this \node.
@@
 The information stored in a \block\ depends on whether it is in an internal node or a leaf.
-A block in a leaf represents a single enqueue or dequeue.  The \var{element} field stores the value
-enqueued for an enqueue operation, or \var{null} if the operation is a dequeue.
+A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
+enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.
 
@@
 Blocks also store prefix sums of the numbers of enqueues and dequeues:
-the block in $v.\var{blocks}[i]$ has two fields \var{sum\sub{enq}} and \var{sum\sub{deq}}
-that store the total number of enqueues and dequeues in $v.\var{blocks[1..i]}$.
+the block in $v.\fld{blocks}[i]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
+that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{i}]$.
 These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
 For example, consider finding the $r$th enqueue in the linearization ordering.
-We use the \var{sum\sub{enq}} fields of the root's blocks to do a binary search
+We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
 to locate which block in the root contains the operation.
 If we know which block $B$ in some \node\ $v$ contains the enqueue,
-we can use the \var{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
+we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
 and then to do a binary search
@@
 All of the information needed for this search process can be derived from the 
-\eleft, \eright\ and \var{sum\sub{enq}} fields.
+\eleft, \eright\ and \fld{sum\sub{enq}} fields.
 
-To facilitate task \ref{findinroot}, each block has a field \var{super} that contains
-the (approximate) index of its superblock in the parent \node's \var{blocks} array.
+To facilitate task \ref{findinroot}, each block has a field \fld{super} that contains
+the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
 We will ensure that this field's value differs from the true index of the superblock by at most 1.
@@
 First, we must determine whether the queue is empty when a dequeue occurs.
-To facilitate this, each block in the root stores a \var{size} field that stores the number of elements
+To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
 in the queue after all operations in the linearization ordering up to that block (inclusive) 
@@
 We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
-$B_{d-1}.\var{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
+$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
 We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
-as $B_{d-1}.\var{sum\sub{enq}}-B_{d-1}.\var{size}$.
+as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
 Then, for any non-null dequeue in $B_d$, we can use this information to determine its
@@
 we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
-A \op{refresh} on \node\ $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
+A \op{refresh} on \node\ $v$ reads the value $h$ of the \fld{head} field of each of $v$'s children and stores 
 $h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
-Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
-For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block $B'$ and
+Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
+For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
 the number of enqueues and dequeues in $B$:
-$B.\var{size} = \max(0, B'.\var{size} + (B.\var{sum\sub{enq}}-B'.\var{sum\sub{enq}}) - (B.\var{sum\sub{deq}} - B'.\var{sum\sub{deq}}))$.
+$B.\fld{size} = \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}) - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}))$.
 
-The only remaining field is $B.\var{super}$.  When the block 
+The only remaining field is $B.\fld{super}$.  When the block 
 $B$ is created for a \node\ $v$, we do not yet know where its
@@
 So, we leave this field blank.  After $B$ is installed 
-in $v.\var{blocks[i]}$, processes cooperate to fill it in 
-when they attempt to advance $v.\var{head}$ from $i$ to $i+1$.
-They use the value read from the \var{head} field of $v$'s parent at that time.
+in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill it in 
+when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$.
+They use the value read from the \fld{head} field of $v$'s parent at that time.
 As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
@@
 \begin{itemize}
-\item \var{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
+\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
 \end{itemize}
@@
 \begin{itemize}
-\item \var{Node} \var{leaf} \Comment{process's leaf in the tree}
+\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \var{Node}
+\Statex $\blacktriangleright$ \typ{Node}
 \begin{itemize}
-\item \var{Node left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
-\item \var{Block[0..$\infty$] blocks} \Comment{initially \var{blocks[0]} contains an empty block with all integer fields equal to 0}
-\item \var{int \head} \Comment{position to attempt appending next \block\ to \var{blocks}, initially 1}
+\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
+\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
+\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
 \end{itemize}
 
-\Statex $\blacktriangleright$ \var{Block}
+\Statex $\blacktriangleright$ \typ{Block}
 
 \begin{itemize}
-  	\item \var{int sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueue, dequeue operations in \var{blocks} array up to this block (inclusive)}
-  	\item \var{int super}
-  		\Comment{approximate index of the block's superblock in \var{parent.blocks}}
+  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
+  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
+  	\item \typ{int} \fld{super}
+  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
 	\item[\com] Blocks in internal \nodes\ have the following additional fields
-	\item \var{int} \eleft, \eright
+	\item \typ{int} \eleft, \eright
   		\Comment{index of last direct subblock in the left and right child}
   	\item[\com] Blocks in leaf \nodes\ have the following additional field
-  	\item \var{Object element}
-  		\Comment{if the block's operation is \op{Enqueue}(\var{x}) then \var{element=x}, otherwise \var{element=null}.}
+  	\item \typ{Object} \fld{element}
+  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
 	\item[\com] Blocks in the root \node\ have the following additional field
-	\item \var{int \size}%
+	\item \typ{int} \size%
   		\Comment{size of the queue after performing all operations up to the end of this block}
@@
 We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
-We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
-$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
-the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
-We use \var{num\sub{deq}} similarly.
-\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}
+%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
+%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
+%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
+%We use \var{num\sub{deq}} similarly.
+%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}
 
@@
 
-\Function{void}{Enqueue}{\var{Object e}} 
-    \State \hangbox{let \var{new} be a new \var{Block} with \var{element} \assign\ \var{e},\\
-		\var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}+1},\\
-		\var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}}}
-    \State \Call{Append}{new}
+\Function{void}{Enqueue}{\typ{Object} \var{e}} 
+    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
+		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
+		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
+    \State \Call{Append}{\var{n}}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{} 
-    \State \hangbox{let \var{n} be a new \var{Block} with \var{element} \assign\ \var{null},\\
-	    \var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}},\\
-	    \var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}+1}}
-    \State \Call{Append}{n}
-    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf, leaf.\head, 1}
-    \State \Return{ \Call{FindResponse}{b, i}}
+    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
+	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
+	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
+    \State \Call{Append}{\var{n}}
+    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
+    \State \Return{ \Call{FindResponse}{\var{b, i}}}
     \label{deqRest}
@@
 
-\Function{void}{Append}{\var{Block B}} 
+\Function{void}{Append}{\typ{Block} \var{B}} 
     \State \linecomment{append block to leaf and propagate to root}
-    \State \var{leaf.blocks[leaf.\head] \assign\ B} \label{appendLeaf}
-    \State \var{leaf.\head\ \assign\ leaf.\head+1} \label{appendEnd} 
-    \State \Call{Propagate}{\var{leaf.parent}} 
+    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
+    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
+    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
 \EndFunction{Append}
@@
 
-\Function{void}{Propagate}{\var{Node v}}
+\Function{void}{Propagate}{\typ{Node} \var{v}}
     \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
-        \State \Call{Refresh}{v} \label{secondRefresh}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
     \EndIf
     \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
-        \State \Call{Propagate}{v.parent}
+        \State \Call{Propagate}{\var{v}.\fld{parent}}
     \EndIf
@@
 
-\Function{boolean}{Refresh}{\var{Node v}}
-    \State \linecomment{try to append a new block to \var{v.blocks}}
-    \State \var{h \assign\ v.\head} \label{readHead}
-    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}} \label{startHelpChild1}
-        \State \var{h\sub{dir} \assign\ v.dir.\head} \label{readChildHead}
-        \If{\var{v.dir.blocks[h\sub{dir}] $\neq$ null}} \label{ifHeadnotNull}
-            \State{\Call{Advance}{v.dir, h\sub{dir}}} \label{helpAdvance}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}}
+    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
+    \State \var{h} \assign\ \var{v}.\head \label{readHead}
+    \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
+        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
+        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
+            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
         \EndIf
     \EndFor \label{endHelpChild1}
-    \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
-    \If{\var{new = null}} \Return{\var{true}} \label{addOP} 
+    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
+    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
     \EndIf
-    \State{\var{result} \assign\ \var{CAS(v.blocks[h], null, new)}} \label{cas}
-    \State{\var{\Call{Advance}{v, h}}} \label{advance}
-    \State \Return{ \tt{result}}
+    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
+    \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
+    \State \Return{ \var{result}}
 \EndFunction{Refresh}
@@
 
-\Function{void}{Advance}{\var{Node v, int h}} 
-    \State \linecomment{set \nf{v.blocks[h].super} and }
-    \State \linecomment{increment \nf{v.head} from \nf{h} to \nf{h+1}}
-    \State \var{h\sub{p}} \assign\ \var{v.parent.\head} \label{readParentHead}
-    \State \var{CAS(v.blocks[h].super, null, h\sub{p})} \label{setSuper1}
-    \State \var{CAS(v.head, h, h+1)} \label{incrementHead}
+\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
+    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
+    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
+    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
+    \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
+    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
 \EndFunction{Advance}
@@
 
-\Function{Block}{CreateBlock}{\var{Node v, int i}} 
-    \State\linecomment{create new block for a \op{Refresh} to install in \var{v.blocks[i]}}
-    \State\linecomment{(return \var{null} if no blocks need to be propagated to \var{v})}
-    \State let \var{new} be a new \var{Block} \label{initNewBlock}
-    \State \var{new.\eleft} \assign\ $\var{v.left.head} - 1$
-    \State \var{new.\eright} \assign\ $\var{v.right.head} - 1$
-	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq}} + \\
-			\var{v.right.blocks[new.\eright].sum\sub{enq}}}
-	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq}} + \\
-			\var{v.right.blocks[new.\eright].sum\sub{deq}}}
-    \State \var{num\sub{enq}} \assign\ \var{new.sum\sub{enq} - v.blocks[i-1].sum\sub{enq}}
-    \State \var{num\sub{deq}} \assign\ \var{new.sum\sub{deq} - v.blocks[i-1].sum\sub{deq}}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
+    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State\linecomment{(return \nl\ if no blocks need to be propagated to \var{v})}
+    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
+    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$
+    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
+	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
+	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
+    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
+    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
     \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new.size \assign\ max(0, v.blocks[i-1].\size\ + num\sub{enq} - num\sub{deq})}\label{computeLength}}
+        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
     \EndIf
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
-        \State \Return \var{null}
+        \State \Return \nl
     \Else
@@
 
-\Function{<int, int>}{IndexDequeue}{\var{Node v, int b, int i}}
-    \State \linecomment{return \var{<x, y>} such that \var{i}th dequeue in D(\var{v.blocks[b]})}
-    \State \linecomment{is \var{y}th dequeue of D(\var{root.blocks[x]})}
-    \State \linecomment{Precondition: \var{v.blocks[b]} has been propagated to root}
-    \State \linecomment{and $\var{v.blocks[b].num\sub{deq}}\geq \var{i}$}
+\Function{<int, int>}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{return \var{<x, y>} such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
+    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
     \If{$\var{v} = \var{root}$}
@@
     \Else
-	    \State \var{dir \assign\ (v.parent.left = v ? left : right)} 
-    	\State \var{sup \assign\ v.blocks[b].super}
-	    \If{$\var{b} > \var{v.parent.blocks[sup].end\sub{dir}}$}
+	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$}
 	        \State{\var{sup} \assign\ $\var{sup}+1$}
@@
 	    \State \linecomment{compute index \var{i} of dequeue in superblock}
-	    \State \hangbox{\var{i += v.blocks[b-1].sum\sub{deq}} - \\
-	    		\var{v.blocks[v.parent.blocks[sup-1].end\sub{dir}].sum\sub{deq}}}
-        \If{$\var{dir} = \var{right}$} 
-        	\State \hangbox{\var{i += v.blocks[v.parent.blocks[sup].end\sub{left}].sum\sub{deq}} -\\
-					\var{v.blocks[v.parent.blocks[sup-1].end\sub{left}].sum\sub{deq}}\label{considerLeftBeforeRight}}
+	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
+	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
+        \If{$\fld{dir} = \fld{right}$} 
+        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} -$\\
+					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
         \EndIf \label{computeISuperEnd}
-        \State \Return\Call{IndexDequeue}{v.parent, sup, i}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
     \EndIf
@@
 
-\Function{element}{FindResponse}{\var{int b, int i}}
-    \State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
-    \If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
-        \State \Return \var{null} \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
+    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - $\\
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
+        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
     \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-        \State $\var{e} \assign\ \var{i} + \var{root.blocks[b-1].sum\sub{enq}} - 
-	\var{root.blocks[b-1].size}$\label{computeE}
+        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
 		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $\var{b}'\leq \var{b}$ with $\var{root.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{e}$
+		\State find min $\var{b}'\leq \var{b}$ with $\var{root}.\fld{blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{e}$
 		\State \linecomment{find rank of enqueue within its block}
-		\State $\var{i}' \assign\ \var{e} - \var{root.blocks}[\var{b}'-1].\var{sum\sub{enq}}$
-        \State \Return \Call{GetEnqueue}{root, $\var{b}'$, $\var{i}'$}\label{findAnswer}
+		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
+        \State \Return \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
     \EndIf
@@
 \end{minipage}
-\caption{Queue implementation.\label{pseudocode}}
+\caption{Queue implementation.\label{pseudocode1}}
 \end{figure}
@@
 
-\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \Comment{returns argument of ith enqueue in $E(\var{v.blocks[b]})$}
-    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v.blocks[b]})$ contains at least \var{i} enqueues}
+\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
 
     \If{\var{v} is a leaf node}
-        \State\Return \var{blocks[b].element} \label{getBaseCase}
+        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}$] from left child}
-        \State \var{prev\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b-1].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from left child}
-        \State \var{prev\sub{right}} \assign\ \var{v.right.blocks[v.blocks[b-1].\eright].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v.left}}
-        	\State \linecomment{Find enqueue's block in \var{v.left} and its rank within block}
-            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eleft}+1..\var{v.blocks[b].\eleft}] s.t. $\var{v.left.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{left}}$
-            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.left.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{left}})$
-            \State \Return\Call{GetEnqueue}{\var{v.left}, $\var{b}'$, $\var{i}'$} 
-        \Else \Comment{required enqueue is in \var{v.right}}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
+            \State \fld{dir} \assign\ \fld{left}
+        \Else
+            \State \fld{dir} \assign\ \fld{right}
             \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
-        	\State \linecomment{Find enqueue's block in \var{v.right} and its rank within block}
-            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eright}+1..\var{v.blocks[b].\eright}] s.t. $\var{v.right.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{right}}$
-            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.right.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{right}})$
-            \State \Return\Call{GetEnqueue}{\var{v.right}, $\var{b}'$, $\var{i}'$} 
         \EndIf
+        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
+        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
+$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$
+        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
+        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} 
     \EndIf
@@
 
-An \op{Enqueue}($e$) operation appends a \block\ to the process's leaf.
-The block has $\var{element}=e$ and its \var{sum\sub{enq}} and \var{sum\sub{deq}} fields are set
-on line \ref{?}.
+An \opemph{Enqueue}(\var{e}) appends a \block\ to the process's leaf.
+The block has $\fld{element}=\var{e}$ to indicate it represents an \op{Enqueue}(\var{e}) operation.
 It suffices to propagate the operation to the root and
@@
 
-A \op{Dequeue} also appends a \block\ to the process's leaf.
-The block has $\var{element}=null$ and its \var{sum\sub{enq}} and \var{sum\sub{deq}} fields are set
-on line \ref{?}.
+A \opemph{Dequeue} also appends a \block\ to the process's leaf.
+The block has $\fld{element}=\nl$ to indicate that it represents a \op{Dequeue} operation.
 After propagating the operation to the root, it computes
@@
 
-\paragraph{\tt{Append($B$)}}
-The \nf{head} field is the index of the first empty slot in
-\nf{blocks} in a \nf{LeafBlock}. There are no multiple write accesses
-on \nf{head} and \nf{blocks} in a leaf because only the process that
-the leaf belongs to appends to it. \nf{Append($B$)} adds $B$ to the
-end of the \nf{blocks} field in the leaf, increments \nf{head} and
-then calls \nf{Propagate} on the leaf's \nf{parent}. When
-\nf{Propagate} terminates, it is guaranteed that the appended block is
-a subblock of a block in the \nf{root}.
+\opemph{Append}(\var{B}) first adds the block \var{B} to the invoking process's leaf.
+The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
+so the \op{Append} writes \var{B} there and increments \fld{head}.
+Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
+\op{Append} then calls \op{Propagate} to ensure the operation represented by \var{B} is propagated to the root.
 
-\paragraph{\tt{Propagate()}}
-\nf{Propagate} on node $n$ uses the double refresh idea described
-earlier and invokes two \nf{Refresh}es on $n$ in Lines
-\ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes
-\nf{Propagate} on \nf{$n$.parent} recursively until it reaches the
-root.
+\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
+It uses the double refresh idea described
+above and invokes two \op{Refresh}es on \var{v} in Lines
+\ref{firstRefresh} and \ref{secondRefresh}. 
+If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
+that propagated blocks from \var{v}'s children to \var{v}.
+Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.
 
-\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s
-children's new blocks and append it to $n$\nf{.blocks}. The variable
-\nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new
-block created by \nf{Refresh} will be inserted into
-$n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1}
-of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance}
-increments the children's \nf{head} if necessary and sets the
-\nf{super} field of their most recently appended blocks. The reason
-behind this helping is explained later when we discuss
-\nf{IndexDequeue}. After helping to \nf{Advance} the children, a new
-block called \nf{new} is created in Line
-\ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh}
-returns \nf{true} because there are no new operations to propagate,
-and it is unnecessary to add an empty block to the tree. Later we will
-use the fact that all blocks contain at least one operation. Line
-\ref{cas} tries to install \nf{new}. If it was successful, all is
-good. If not, it means someone else has already put a block in
-$n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance
-$n$\nf{.head} to \nf{h+1} and update the \nf{super} field of
-$n$\nf{.blocks[h]} at Line \ref{advance}.
+%\paragraph{\tt{Refresh()} and \tt{Advance()}}
+The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
+in \var{v}'s
+children and append it to \var{v}.\fld{blocks}. 
+Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
+and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
+If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v},
+so \op{Refresh} terminates and returns true at line \ref{addOP}.
+Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
+Either this CAS succeeds or some other process has installed a new block in this location.
+Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
+from \var{h} to $\var{h}+1$
+and to fill in the \fld{super} field of the most recently appended block.
+The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
+Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
+other processes help keep \fld{head} up to date by also calling \op{Advance}, 
+either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.
+(The need for helping on line \ref{helpAdvance} is explained below, when we discuss the \op{IndexDequeue} routine.)
 
-\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used
-by \nf{Refresh} to construct a block containing new operations of
-$n$'s children. 
-The block \nf{new} is created in Line \ref{initNewBlock} and its
-fields are filled similarly for both left and right directions. The
-variable  \nf{index\sub{prev}} is the index of the block preceding the
-first subblock in the child in direction \nf{dir} that is aggregated
-into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the
-index of the rightmost subblock of \nf{new} in the child. Then
-\nf{sum\sub{enq-dir}} is computed from the sum of the number of
-\nf{Enqueue} operations in the \nf{new} block from direction \nf{dir}
-and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The
-field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new}
-is going to be installed in the \nf{root}, the \nf{size} field is also
-computed.
+\opemph{CreateBlock}(\var{v, i}) is used
+by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
+The block \nf{new} is created in Line \ref{initNewBlock}. 
+The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
+children, which are obtained by reading the \fld{head} index in \var{v}'s children.
+Recall that the \fld{sum\sub{enq}} field should store the number of enqueues in
+\var{v}.\fld{blocks}[1..\var{i}].
+Since these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright],
+the total number of these enqueues is $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
+Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
+subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues that are already in \var{v}.\fld{blocks}[$1..\var{i}-1$].
+The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
+Then, if \var{new}
+is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
+represents the number of elements in the queue after the operations in the block are performed.
+Finally, if the newly created block contains no operations, \op{CreateBlock} returns \nl\ to indicate
+that there is no need to append it to \var{v}.\fld{blocks}.
 
-\paragraph{\tt{IndexDequeue($b,i$)}}
-
-A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank
-within the block in the root of the $i$th \nf{Dequeue} of the $b$th
-block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$
-that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let
-$par$ be $n$\nf{.parent}. Without loss of generality, assume for the
+Once a \op{Dequeue} has been appended to a block of the process's leaf and propagated to the root,
+the \opemph{IndexDequeue} routine is used to find the \op{Dequeue}'s location in the root.
+More precisely, \opa{IndexDequeue}{v, b, i}
+computes the block in the root and the rank
+within that block  of the \var{i}th \op{Dequeue} of the block stored in \var{v}.\fld{blocks}[\var{b}].
+Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
+whose CAS wrote a block $B$ into \var{v}.\fld{blocks}[\var{b}].
+\here{---}
+Let $par$ be $n$\nf{.parent}. Without loss of generality, assume for the
 rest of this section that ${n}$ is the left child of $par$. Let

### 2022-12-31 podc_paper/related.tex
formatting

@@
 
-\paragraph{List-based Queues}
+\emph{List-based Queues.}
 The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
@@
 
-\paragraph{Array-Based Queues}
+\emph{Array-Based Queues.}
 Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
@@
 
-\paragraph{Universal Constructions}
+\emph{Universal Constructions.}
 One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
@@
 
-\paragraph{Restricted Queues}
+\emph{Restricted Queues.}
 David introduced the first sublinear-time queue
@@
 
-\paragraph{Other Primitives}
+\emph{Other Primitives.}
 Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
@@
 
-\paragraph{Fetch\&Increment Objects}
+\emph{Fetch\&Increment Objects.}
 There are sublinear-time implementations of another object

### 2022-12-31 podc_paper/algorithm.tex
fixed spacing for code
finished draft of algorithm description

@@
 \begin{figure}
-\begin{minipage}[t]{0.461\textwidth}
+\begin{minipage}[t]{0.465\textwidth}
 \begin{algorithmic}[1]
@@
 \end{minipage}
-\begin{minipage}[t]{0.533\textwidth}
+\begin{minipage}[t]{0.529\textwidth}
 
@@
     \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State\linecomment{(return \nl\ if no blocks need to be propagated to \var{v})}
     \State let \var{new} be a new \typ{Block} \label{initNewBlock}
-    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$
+    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
     \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
@@
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
-        \State \Return \nl
+        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
     \Else
@@
         \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
-$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$
+$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
         \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
@@
 either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.
-(The need for helping on line \ref{helpAdvance} is explained below, when we discuss the \op{IndexDequeue} routine.)
 
@@
 computes the block in the root and the rank
-within that block  of the \var{i}th \op{Dequeue} of the block stored in \var{v}.\fld{blocks}[\var{b}].
+within that block  of the \var{i}th \op{Dequeue} of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
 Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
-whose CAS wrote a block $B$ into \var{v}.\fld{blocks}[\var{b}].
-\here{---}
-Let $par$ be $n$\nf{.parent}. Without loss of generality, assume for the
-rest of this section that ${n}$ is the left child of $par$. Let
-$R_{par}$ be the first successful \nf{$par$.Refresh} that reads some
-value greater than ${b}$ for \nf{left.head} and therefore contains
-${B}$ as a subblock of its created block in Line
-\ref{invokeCreateBlock}. Let $j$ be the index of the block that
-$R_{par}$ puts in $par$\nf{.blocks}.
+whose CAS wrote a block \var{B} into \var{v}.\fld{blocks}[\var{b}].
+Let $par$ be \var{v}.\fld{parent}. Without loss of generality, assume 
+\var{v} is the left child of $par$. Let
+$R_{par}$ be the first successful \op{Refresh} on $par$ whose \op{CreateBlock} reads a
+value greater than \var{b} from $\var{par}.\fld{left.head}=\var{v}.\fld{head}$ at line \ref{createEndLeft} and therefore creates a block $B'$ that contains
+\var{B} as a subblock. 
+%Let $b'$ be the index of $B'$ in $par$.\fld{blocks}.
 
-Since the index of the superblock of ${B}$ is not known until $B$ is
-propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while
-creating $B$. One approach for $R_{par}$ is to set the \nf{super}
-field of ${B}$ after propagating $B$ to $par$. This solution would not
-be efficient because there might be up to $p$ subblocks that $R_{par}$
-propagated, which need to update their \nf{super} field. However,
-intuitively, once $B$ is installed, its superblock is going to be
-close to \nf{$n$.parent.head} at the time of installation. If we know
-the approximate position of the superblock of $B$ then we can search
-for the real superblock when it is needed. Thus, \nf{$B$.super} does
-not have to be the exact location of  the superblock of $B$, but we
-want it to be close to $j$. We can set \nf{$B$.super} to
-\nf{$par$.head} while creating $B$, but the problem is that there
-might be many \nf{Refresh}es on $par$ that could happen after $R_n$
-reads $par$\nf{.head} and before propagating $B$ to $par$. If $R_n$
-sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to
-\nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at
-some time after installing $B$ and before setting \nf{$B$.super}. In
-this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the
-value of \nf{$B$.super}.
+Since the index of \var{B}'s superblock is not known until $B$ is
+propagated to $par$, $R_v$ cannot set the \nf{super} field of ${B}$ when
+it creates \var{B}. 
+It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (because
+there could be up to $p$ direct subblocks).
+Intuitively, the index of \var{B}'s superblock will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
+Since we can search for the precise superblock when we need it, it suffices that \var{B}.\fld{super}
+is close to the actual index of \var{B}'s superblock.
+Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
+As mentioned earlier, other processes may help fill in this field by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that it is filled in soon after \var{B} is installed.
+We shall
+show that this ensures that \nf{$B$.super} differs
+from the index of $B'$ by at most~1.
 
-Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}. After
-appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call
-to \nf{Advance} from \nf{$n$.Refresh} by the same or another process
-or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}. We shall
-show that this is sufficient to ensure that \nf{$B$.super} differs
-from the index of $B$'s superblock by at most~1.
+To compute the response of the $i$th \nf{Dequeue} in the $b$th block
+of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
+If the queue is
+not empty, Line \ref{computeE} computes the rank \var{e} of the
+\nf{Enqueue} whose argument is the response to the
+\nf{Dequeue}. 
+A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b'$ of the block that contains 
+the \var{e}th enqueue.
+The index $b'$ is at most $b$ (since the enqueue is linearized before the dequeue).
+To find the left end of the range for the binary search for $b'$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
+Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.
 
-\paragraph{\tt{FindResponse($b, i$)}}
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block
-of the root Line \ref{checkEmpty} computes whether the queue is empty
-or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue
-would become empty before the requested \nf{Dequeue}. If the queue is
-not empty, Line \ref{computeE} computes the rank $e$ of the
-\nf{Enqueue} whose argument is the response to the
-\nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the
-root (which is before the $b$th block), we find the block and position
-containing the \nf{Enqueue} operation using \nf{DoublingSearch} and
-after that \nf{GetEnqueue} finds its \nf{element}.
-
-\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DoublingSearch($e, end$)}}
-We can describe an operation in a node in two ways: the rank of the
-operation among all the operations in the node or the index of the
-block containing the operation in the node and the rank of the
-operation within that block. If we know the block and rank within the
-block of an operation, we can find the subblock containing the
-operation and the operation's rank within that subblock in poly-log
-time. To find the response of a \nf{Dequeue}, we know about the  rank
-of the response \nf{Enqueue} in the root (\nf{e} in Line
-\ref{computeE}). 
-We also know the \nf{e}th \nf{Enqueue} is in
-\nf{root.blocks[1..end]}. \nf{DoublingSearch} uses doubling to find
-the range that contains the answer block (Lines
-\ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the
-required indices with a binary search (Line
-\ref{dsearchBinarySearch}). 
-A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the
-$i$th enqueue in the $b$th block of $n$. The range of subblocks of a
-block is determined using the \nf{end\sub{left}} and
-\nf{end\sub{right}} fields of the block and its previous block. Then,
-the subblock is found using binary search on the \nf{sum\sub{enq}}
-field (Lines \ref{leftChildGet} and \ref{rightChildGet}).
+\opemph{GetEnqueue}(\var{v,b,i}) returns the argument of the
+$i$th enqueue in the $b$th block of \typ{Node} $v$. 
+The range of subblocks of a block is determined using the \fld{end\sub{left}} and
+\fld{end\sub{right}} fields of the block and its previous block. Then,
+the exact subblock is found using binary search on the \fld{sum\sub{enq}}
+field (line \ref{getChild}).

### 2022-12-31 podc_paper/podc-submission.tex
working on code
working on alg description, tidying code
finished draft of algorithm description

@@
 \algdef{S}[FUNCTION]{Function}
-   [3]{{\var{{#1}}} {\op{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
+   [3]{{\typ{{#1}}} {\op{#2}}\ifthenelse{\equal{#3}{}}{}{(\var{#3})}}
 
@@
 
-\algrenewcommand\Call[2]{\var{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
+\algrenewcommand\Call[2]{\op{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}
 \usepackage{eqparbox}
@@
 
-\newcommand{\eleft}{\var{end\sub{left}}}
-\newcommand{\eright}{\var{end\sub{right}}}
-\newcommand{\edir}{\var{end\sub{dir}}}
-
 \renewcommand\thealgorithm{}
@@
 \newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
+\newcommand{\head}{\fld{head}}
+\newcommand{\size}{\fld{size}}
+\newcommand{\eleft}{\fld{end\sub{left}}}
+\newcommand{\eright}{\fld{end\sub{right}}}
+\newcommand{\edir}{\fld{end\sub{dir}}}
 
+\newcommand{\var}[1]{\mbox{\it{#1}}}
+\newcommand{\fld}[1]{\mbox{\it{#1}}}
+\newcommand{\typ}[1]{\mbox{\sf #1}}
 \newcommand{\ceil}[1]{\lceil #1 \rceil}
-\newcommand{\nil}{nil}
-\newcommand{\opa}[2]{\sf{#1(#2)}}
-\newcommand{\op}[1]{\sf{#1}}
-\newcommand{\var}[1]{\sf{#1}}
+\newcommand{\nl}{\mbox{\sf null}}
+\newcommand{\tr}{\mbox{\sf true}}
+\newcommand{\opa}[2]{\op{#1}(\var{#2})}
+\newcommand{\op}[1]{{\sf #1}}
+\newcommand{\opemph}[1]{{\sf \bf{#1}}}
 \newcommand{\com}{$\triangleright$}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{concurrent data structures, wait-free queues}
+%\keywords{concurrent data structures, wait-free queues}
 
@@
 \input{introduction}
+
 \input{related}
+
 \input{algorithm}
-
-\subsection{Pseudocode} \label{algQ}
-We present our algorithm in pseudocode.
-page~22 contains the description of the fields in the tree nodes and the blocks.
-The value of any uninitialized field is \nf{null}.
-page~23 contains major routines and the rest of this section consists of the auxiliary routines.
-The abbreviations below are used in the pseudocode and the proof of correctness.
-\begin{itemize}
- \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
-%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
-  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
-\end{itemize}
 

### 2022-12-31 podc_paper/algorithm.tex
minor edits

@@
     \State \Call{Append}{\var{n}}
-    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
@@
 
-\Function{<int, int>}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return \var{<x, y>} such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
     \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
@@
     \If{$\var{v} = \var{root}$}
-        \State\Return \var{<b, i>} \label{indexBaseCase}
+        \State\Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
     \Else
@@
         \If{$\fld{dir} = \fld{right}$} 
-        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} -$\\
+        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
 					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
@@
     \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
-    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - $\\
+    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
     		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
@@
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter be the rank within whole node rather than within the block?}}
+\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
 \end{figure}
@@
 field (line \ref{getChild}). 
+\here{explain a little more?}

### 2023-01-01 podc_paper/algorithm.tex
editing proof; fixed minor bug in algorithm (off-by-one error)

@@
 several consecutive blocks from each of $v$'s children.
-The blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
+The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
 A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
@@
 and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s enqueue sequence is the concatenation $E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)$.
-The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.
+The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.  We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
 
@@
 linearization ordering is 
-$E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k)$.
+\begin{equation}
+L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
+\end{equation}
+\label{linearization}
 
@@
 Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
-To simplify the code, each \node\ has an empty \block\ in \fld{blocks[0]}.
+To simplify the code, each \node\ has an empty \block\ in \fld{blocks}[0].
 Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
@@
 Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
+Thus, the direct subblocks of $v.\fld{blocks}[i]$ are
+\begin{eqnarray}\label{defsubblock}
+v.\fld{left.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{left}}+1..v.\fld{blocks}[i].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
+v.\fld{right.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{right}}+1..v.\fld{blocks}[i].\fld{end\sub{right}}].
+\end{eqnarray}
 This allows us to navigate to the direct subblocks of any block easily.
@@
 
-To facilitate task \ref{findinroot}, each block has a field \fld{super} that contains
+A block is called the \emph{superblock} of all of its direct subblocks.
+To facilitate task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
 the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
-We will ensure that this field's value differs from the true index of the superblock by at most 1.
+We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
 This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
@@
     \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
+		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
+		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}
     \State \Call{Append}{\var{n}}
@@
     \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
+	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
+	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}
     \State \Call{Append}{\var{n}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
-    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
-    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
+    \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
+    \State \var{leaf.}\head\ \assign\ \var{leaf.}\head+1 \label{appendEnd} 
+    \State \Call{Propagate}{\var{leaf.}\fld{parent}} 
 \EndFunction{Append}
@@
     \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
-    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
+    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
 	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
@@
     \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
-    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
-    \If{$\var{v} = \var{root}$}
-        \State\Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
+    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
+    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
+    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
     \Else
@@
     	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$}
-	        \State{\var{sup} \assign\ $\var{sup}+1$}
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
 	    \EndIf
@@
     \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
+    \State \linecomment{Preconditions:  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
+    \State \linecomment{contains at least \var{i} dequeues}
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{101}
+\setcounter{ALG@line}{99}
 
 \Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
 

### 2023-01-01 podc_paper/correctness.tex
editing proof; fixed minor bug in algorithm (off-by-one error)

@@
 
-\here{following material from algorithm description in thesis might be useful in sketch of correctness proof}
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a
-    process.]{\label{fig::simpleDoubleRefresh}Time relations between
-    the concurrent successful \nf{Refresh}es and the two consecutive
-    failed \nf{Refresh}es.} 
-\end{figure}
-We use \nf{CAS} (Compare\&Swap) instructions to implement the
-\nf{Refresh}'s attempt to append  described in the previous
-paragraph. 
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
-successful \texttt{Refresh} that has read its information after the
-invocation of the first failed \nf{Refresh} (see Figure
-\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
-a successful append during $P$'s first failed attempt, and some
-process $K$ performs a \nf{Refresh} that reads its information after
-$L$'s append and then performs a successful append during $P$'s second
-failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
-new operations that were in $n$'s children before $P$'s first failed
-\nf{Refresh}, in case they were not already appended. After a process
-appends its operation into its leaf it can call \nf{Refresh} on the
-path up to the root at most two times on each node. So, with $O(\log
-p)$ \nf{CAS}es an operation can ensure it appears in the
-linearization. This cooperative solution allows us to overcome the CAS
-Retry Problem.
+Consider any finite execution.  We must show that the linearization ordering $L$ defined in Section \ref{linearization} includes all operations that terminate and that 
+(1) if an operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ cannot appear
+before $op_1$ in $L$, and (2) each dequeue that terminates returns the same response as it would 
+in $L$.
 
-We adopt linearizability as our definition of correctness.
-In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue.
-The proof is structured like this.
+\here{need to update this outline later}
 First, we define and prove some facts about blocks and the node's \nf{head} field.
@@
 \subsection{Basic Properties}
-In this subsection, we talk about some properties of blocks and fields of the tree nodes.
 
-A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
-\begin{definition}[Ordering of a block in a node]
-Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
-\end{definition}
+Since \fld{head} fields are updated only on line \ref{incrementHead}, which increments it, we have the following observation.
+\begin{observation} \label{nonDecreasingHead}
+For each node \var{v},  \var{v}.\fld{head} is non-decreasing over time.
+\end{observation}
 
-Next, we show that the value of \nf{head} in a node can only be increased.
-By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
-\begin{observation} \label{nonDecreasingHead}
-For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
-\end{observation}
+\begin{lemma} \label{lem::headInc}
+Let $R$ be an instance of \opa{Refresh}{v} whose call to \op{CreateBlock} returns a non-\nl\ block.
+When $R$ terminates, \var{v}.\head\ is greater than the value $R$ reads from it at line \ref{readHead}.
+\end{lemma}
 \begin{proof}
-  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
+After $R$'s \op{CAS} at line \ref{incrementHead}, \var{v}.\head\ is no longer equal to the value \var{h}
+read at line \ref{readHead}.  The claim follows from \Cref{nonDecreasingHead}.
 \end{proof}
 
-\begin{lemma} \label{lem::headInc}
-Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$).
-After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
+Now we show $v.\fld{blocks}[v.\head]$ is either the last non-\nl\ block or the first \nl\ block in node $v$.
+
+\begin{invariant}\label{lem::headPosition} 
+For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
+\end{invariant}
+
+\begin{proof}
+Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claim~holds.
+
+Assume the claim holds before a change to $v.\fld{blocks}$, which can be made only
+by a successful \op{CAS} at line \ref{cas}.
+The \op{CAS} changes $v.\fld{blocks}[h]$ from \nl\ to a non-\nl\ value.
+Since $v.\fld{blocks}[h]$ is \nl\ before the CAS, $v.\head \leq h$ by the hypothesis.
+Since $h$ was read from $v.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
+$v.\head \geq h$ by \Cref{lem::headPosition}.
+So, $h=v.\head$ and a change to $v.\fld{blocks}[v.\head]$ preserves the invariant.
+
+Now, assume the claim holds before a change to $v.\head$, which can only be an increment from $h$ to $h+1$
+by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
+It suffices to show that $v.\fld{blocks}[head] \neq \nl$.
+\nf{Advance} is called either at line \ref{helpAdvance} 
+after testing that $v.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
+or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$. 
+\end{proof}
+
+It follows that blocks accessed by the \op{Enqueue}, \op{Dequeue} and \op{CreateBlock} routines are non-\nl.
+
+The following two lemmas show that no operation appears in more than one block of the root.
+\begin{lemma} \label{lem::headProgress}
+ If $b>0$ and $v.\fld{blocks}[b] \neq \nl$ then 
+ $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
+ $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
 \end{lemma}
 \begin{proof}
-If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds.
-Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}.
-By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
-\end{proof}
-
-Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.
-
-\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
-\end{invariant}
-\begin{proof}
-Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$.
-The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.
-We show that if the invariant holds before such a change, then it still holds after the change.
-
-In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}.
-Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail.
-Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.
-
-The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}.
-If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}.
-\nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}.
-For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}.
-For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}.
-Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}.
-\end{proof}
-
-We define the subblocks of a block recursively.
-
-\begin{definition}[Subblock]\label{def::subblock}
-A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
-Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.
-We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-\end{definition}
-
-The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
-\begin{lemma} \label{lem::headProgress}
- If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}.
-Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}.
-Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}.
-Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$.
-The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}.
-Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}.
-By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$.
-The proof for \nf{end\sub{right}} is similar.
+Let $B$ be the block in $v.\fld{blocks}[b]$.
+Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
+read $b$ from $v.\head$ at line \ref{readHead}.
+At that time, $v.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
+Thus, the \op{CreateBlock}($v,b-1$) that created $B'$ terminated before the \op{CreateBlock}($v,b$) that
+created $B$ started.
+It follows from \Cref{nonDecreasingHead} that the value that 
+line \ref{createEndLeft} of \op{CreateBlock}($v,b-1$) stores in $B'.\fld{end\sub{left}}$   
+is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($v,b$) 
+stores in $B.\fld{end\sub{left}}$.
+Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
+%of these calls to \op{CreateBlock} 
+satisfy the claim.
 \end{proof}
@@
 \begin{lemma} \label{lem::subblocksDistinct}
-The sets of subblocks of any two blocks in a node are disjoint.  
+If $B$ and $B'$ are two blocks in nodes at the same depth, their sets of subblocks are disjoint.
 \end{lemma}
 \begin{proof}
-We are going to prove the lemma by contradiction.
-Consider the lowest \nf{node} $n$ in the tree that violates the claim.
-Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$.
-Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap.
-Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$.
-By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$.
-Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
+We prove the lemma by reverse induction.
+If $B$ and $B'$ are in leaves, they have no subblocks, so the claim is true.
+Assume the claim is true for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
+Consider the direct subblocks of $B$ and $B'$ defined by (\ref{defsubblock}).
+If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
+If $B$ and $B'$ are in the same node, it follows from \Cref{lem::headProgress} that their direct subblocks are disjoint.
+Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
 \end{proof}
 
-\begin{definition}[Superblock]
-  Block $B$ is \it{superblock} of block $C$ if $C$ is a direct subblock of $B$.
-\end{definition}
-\begin{corollary}
-Every block has at most one superblock.
-\end{corollary}
-\begin{proof}
-A block having more than one superblock contradicts Lemma \ref{lem::subblocksDistinct}.
-\end{proof}
+It follows that each block has at most one superblock.
+Moreover, we can now prove each operation is contained in at most one block of each node,
+and hence appears at most once in the linearization~$L$.
 
-Now we can define the operations of a block using the definition of subblocks.
-\begin{definition}[Operations of a block]\label{def::ops}
-A block $B$ in a leaf represents an \nf{Enqueue} if $B\nf{.element}\neq\nf{null}$.
-Otherwise, if $B\nf{.element}=\nf{null}$, $B$ represents a \nf{Dequeue}.
-The set of operations of block $B$ is the union of the operations in leaf subblocks of $B$.
-We denote the set of operations of block $B$ by ${ops(B)}$ and the union  of operations of a set of blocks $\mathcal{B}$ by $ops(\mathcal{B})$.
-We also say $B$ contains ${op}$ if $op\in{ops(B)}$.
-\end{definition}
+\begin{lemma}\label{lem::noDuplicates}
+For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
+\end{lemma}
+-------
 
-The next lemma proves that each operation appears at most once in the blocks of a node.
-\begin{lemma}\label{lem::noDuplicates}
-For any node $n$, if $op$ is in $n.\nf{blocks[}i\nf{]}$ then there is no $j\neq i$ such that $op$ is in $n\nf{.blocks[}j\nf{]}$.
-\end{lemma}
 \begin{proof}
@@
 This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
+
+  We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
 
@@
 The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node. 
+\here{mention that precondition of IndexDequeue follows; also check that block of parent indexed by sup in that routine is non-null}
 Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
@@
 By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
+
+\here{following material from algorithm description in thesis might be useful in sketch of correctness proof}
+\begin{figure}[hbpt]
+  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
+  \caption[Two consecutive failed \nf{Refresh}es by a
+    process.]{\label{fig::simpleDoubleRefresh}Time relations between
+    the concurrent successful \nf{Refresh}es and the two consecutive
+    failed \nf{Refresh}es.} 
+\end{figure}
+We use \nf{CAS} (Compare\&Swap) instructions to implement the
+\nf{Refresh}'s attempt to append  described in the previous
+paragraph. 
+The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
+successful \texttt{Refresh} that has read its information after the
+invocation of the first failed \nf{Refresh} (see Figure
+\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
+a successful append during $P$'s first failed attempt, and some
+process $K$ performs a \nf{Refresh} that reads its information after
+$L$'s append and then performs a successful append during $P$'s second
+failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
+new operations that were in $n$'s children before $P$'s first failed
+\nf{Refresh}, in case they were not already appended. After a process
+appends its operation into its leaf it can call \nf{Refresh} on the
+path up to the root at most two times on each node. So, with $O(\log
+p)$ \nf{CAS}es an operation can ensure it appears in the
+linearization. This cooperative solution allows us to overcome the CAS
+Retry Problem.

### 2023-01-02 podc_paper/algorithm.tex
Working on proof

@@
 Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
-To simplify the code, each \node\ has an empty \block\ in \fld{blocks}[0].
+To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
+We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
 Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
@@
 		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}
+		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}\label{enqNew}
     \State \Call{Append}{\var{n}}
@@
 	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}
+	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
     \State \Call{Append}{\var{n}}
@@
 	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
 	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}

### 2023-01-02 podc_paper/correctness.tex
Working on proof

@@
 
-\begin{lemma}\label{lem::noDuplicates}
+\begin{corollary}\label{lem::noDuplicates}
 For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
-\end{lemma}
--------
-
+\end{corollary}
 \begin{proof}
-We prove this claim by contradiction using Lemma \ref{lem::subblocksDistinct}.
-Assume $op$ is in the subblocks of both $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$.
-From Lemma \ref{lem::subblocksDistinct} we know that the subblocks of these blocks are different, so there are two leaf blocks containing $op$.
-Since each process puts each operation in only one block of its leaf, $op$ cannot be in two leaf blocks.
-This is a contradiction.
+The operations contained in a block $B$ are those that appear in subblocks of $B$ in the leaves of the tree.
+Since each process puts each of its operations in only one block of its own leaf, an operation 
+cannot be in two different leaf blocks. 
+By \Cref{lem::subblocksDistinct}, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ have no subblocks in common, so the claim follows.
 \end{proof}
 
-\begin{definition}
-$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$.
-An operation is \it{established} in node $n$ if it is in an established block of $n$.
-$EST^t_n$ is the set of established operations in node $n$ at time $t$.
-\end{definition}
-
-Now we want to say that blocks of a node grow over time.
-\begin{observation}\label{lem::blocksOrder}
-  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of $ops(n.blocks)$ at time $t^\prime$.
-\end{observation}
-\begin{proof}
-Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-\end{proof}
+%\begin{definition}
+%$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \it{established} in node $n$ 
+%if it is in an established block of $n$. $EST^t_n$ is the set of established operations in node $n$ at time $t$.
+%\end{definition}
+%
+%Now we want to say that blocks of a node grow over time.
+%\begin{observation}\label{lem::blocksOrder}
+%  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of 
+%$ops(n.blocks)$ at time $t^\prime$.
+%\end{observation}
+%\begin{proof}
+%Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
+%\end{proof}
 
@@
 
-\subsection{Ordering Operations}
+The following shows that the values stored in \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields are accurate.
+\here{If space is tight, could easily move next proof to appendix and just say it follows easily from the definition of subblocks and the values stored in the fields on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq}, \ref{createSumDeq}.}
 
-\begin{figure}[hbt]
-  \center\includegraphics[width=5.5in]{pics/tree}
-  \caption[Order of operations in a block.]{Order of operations in the block $B$. Operations in the leaves are ordered in the numerical order shown in the drawing.}
-\end{figure}
+\begin{invariant}\label{lem::sum}
+If $B$ is a block stored in $v.\fld{blocks}[i]$,
+$B.\fld{sum\sub{enq}} = | E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i]) |$ and
+$B.\fld{sum\sub{deq}} = | D(v.\fld{blocks}[0])\cdots D(v.\fld{blocks}[i]) |$.
+\end{invariant}
+\begin{proof}
+Initially, each \fld{blocks} array only contains an empty block $B_0$ in location 0.
+By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
+Moreover, $B_0.\fld{sum\sub{enq}} = B_0.\fld{sum\sub{deq}} = 0$, so the claim is true.
 
-Now we define the ordering of operations stored in each node.
-In the non-root nodes, we only need to order operations of a type among themselves (that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately).
-Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right.
-We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
-\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
-\end{definition}
-\begin{itemize}
-  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
-\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
-\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
-  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows.
-$D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation.
-If $n$ is an internal node, then
-\begin{align*} 
-D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
-\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
-\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
-\end{itemize}
+We show that each installation of a block $B$ into some location $v.\fld{blocks}[i]$ preserves the claim,
+assuming the claim holds before this installation.  We consider two cases.
 
-The linearization ordering is given by the order in which operations appear in the blocks in the root.
-\begin{definition}[Linearization] \label{def::lin}
- $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
-\end{definition}
+If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
+For line \ref{enqNew}, $B$ represents a single enqueue operation, so $|E(B)|=1$ and $|D(B)|=0$.
+Moreover, $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
+$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, so the claim follows from the hypothesis.
+The proof for line \ref{deqNew}, where $B$ represents a single dequeue operation is similar.
 
-The following observation follows from the Definition of \nf{num\sub{x}} on page 22.
-\begin{observation}\label{sumToNum}
-For any node $n$ and indices $i<j$ of \nf{blocks} in $n$, we have
-$$n.\nf{blocks[}j\nf{].sum\sub{x}}-n.\nf{blocks[}i\nf{].sum\sub{x}}=\mathlarger{\sum}_{k=i+1}^{j}  n\nf{.blocks[}k\nf{].num\sub{x}}$$ where $\nf{x}\in\{\nf{enq}, \nf{deq}, \nf{enq-left}, \nf{enq-right}, \nf{deq-left}, \nf{deq-right}\}$.
-\end{observation}
-
-The next claim is also valid if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{lemma}\label{lem::numX}
-Let $B$ and $B^\prime$ be $n\nf{.blocks[}b\nf{]}$ and $n\nf{.blocks[}b-1\nf{]}$, respectively.
-
-\hspace{-\parindent}If $n$ is an internal node, then
-\begin{enumerate}[label=(\arabic*)]
-    \item  $B\nf{.num\sub{enq-left}}=\Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item $B\nf{.num\sub{enq-right}}=\Big|E(n.\nf{right},B^\prime.\nf{end\sub{right}}+1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-\end{enumerate}
-And for every node n, we have
-\begin{enumerate}[label=(3)]
-    \item $B\nf{.num\sub{enq}}=\Big|E(n,b)\Big|$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
- We prove the claim by induction on the height of node $n$.
-For the base case when $n$ is a leaf, statement (3) is trivial, and (1) and (2) are vacuously true.
-Supposing the claim is true for $n$'s children, we prove the claim for $n$.
-\begin{align*}
-    B\nf{.num\sub{enq-left}}&=B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\
-    &\quad-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}}\\
-    &= n.\nf{left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}- n.\nf{left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}\\
-    &= \sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}} \\
-    &= \Big|E(n.\nf{left},B^\prime.\nf{end\sub{left}}+1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-The first line follows from the Definition of \nf{num\sub{enq}}.
-The second line is similar to the way \nf{sum\sub{enq-left}} is computed in the \nf{CreateBlock} routine.
-Observation \ref{sumToNum} implies the third line and the last line holds because of the induction hypothesis (3).
-(2) is similar to (1).
-Now we prove (3) starting from the definition of $E(n,b)$.
-\begin{align*} 
-E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
-&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
-\end{align*}
-By (1) and (2) we have $\Big|E(n,b)\Big|=B\nf{.num\sub{enq-left}}+B\nf{.num\sub{enq-right}}=B\nf{.num\sub{enq}}$.
+Now suppose $v$ is an internal node. By the definition of subblocks in (\ref{defsubblock}), the
+subblocks of $v.\fld{blocks}[1..i]$ are $v.\fld{left.blocks}[1..B.\eleft]$ 
+and $v.\fld{right.blocks}[1..B.\eright]$.
+Thus, the enqueues in $E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i])$ are those in
+$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eleft])$ and
+$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eright])$.
+By the hypothesis, the total number of these enqueues is $v.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
+The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
 \end{proof}
 
-The next claim is also true if we replace \nf{enq} with \nf{deq} and $E$ with $D$.
-\begin{corollary}\label{lem::sumX}
-Let $B$ be $n\nf{.blocks[}b\nf{]}$.
-\begin{enumerate}[label=(\arabic*)]
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-left}}=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|$.
-    \item If $n$ is an internal node then $B\nf{.sum\sub{enq-right}}=\Big|E(n.\nf{right},1)\cdots E(n.\nf{right},B.\nf{end\sub{right}})\Big|$.
-    \item $B\nf{.sum\sub{enq}}=\Big|E(n,1)\cdot E(n,2)\cdots E(n,b)\Big|$.
-\end{enumerate}
-\end{corollary}
-\begin{proof} Result (1) can be proved using the previous lemma.
-\begin{align*}
-    B\nf{.sum\sub{enq-left}}&=n\nf{.blocks[}1\nf{].num\sub{enq-left}}+\dots+ n\nf{.blocks[}b\nf{].num\sub{enq-left}}\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},n\nf{.blocks[}1\nf{].end\sub{left}})\Big| +\\
-    &\hspace{2em}\vdots\\
-    &\quad+\Big|E(n.\nf{left},n\nf{.blocks[}b-1\nf{].end\sub{left}})\cdots E(n.\nf{left},n\nf{.blocks[}b\nf{].end\sub{left}})\Big|\\
-    &=\Big|E(n.\nf{left},1)\cdots E(n.\nf{left},B.\nf{end\sub{left}})\Big|
-\end{align*}
-
-We can prove (2) and (3) the same as (1).
-\end{proof}
+-------
 

### 2023-01-02 podc_paper/algorithm.tex
add garbage collection code to algorithm.tex

--- 2022-12-31 podc_paper/algorithm.tex
@@
 \begin{figure}
-\here{Does \var{leaf} have to be shared for helping when doing GC?}
+%\here{Does \var{leaf} have to be shared for helping when doing GC?}
 \begin{algorithmic}[1]
@@
 \begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
+\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process} % TODO: indent these lines in order to be more readable.
+I tries some times and i could not figure out doing that in algorithm.
+\item \gc{\typ{*Node[]} \var{leafOfProcess[1..p]} \Comment{Pointer to the leaf assigned to each process.}}
+\item \gc{\typ{int[]} \var{lastDequeuedByProcess[1..p]}}
 \end{itemize}
+
+\vspace{0.8em} %TODO: clean code
 
@@
 
+\vspace{0.4em} %TODO: clean code
+
 \Statex $\blacktriangleright$ \typ{Node}
@@
 \item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
+\gc{\item \typ{PBRT<Block, int>} \fld{blocks} \Comment{PBRT of blocks each node contains a block and an index key}}
 \item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
 \end{itemize}
+
+\vspace{0.8em} %TODO: clean code
 
@@
   		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
+  	\gc{\item \typ{Object} \fld{response}
+  		\Comment{if the block's operation is \opa{Dequeue}{} then \fld{response} is the response of the\\ operation if it has been helped.}}
 	\item[\com] Blocks in the root \node\ have the following additional field
@@
     \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
+    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryAppend}{\var{B}, \var{leaf}.\head}\label{appendLeafGC}}
     \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
@@
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
+
+\gc{\If{ \var{v} is root and \var{h}\%p\textsuperscript{2}==0}
+\State \Call{Help()}{}
+\State \Call{FreeMemory}{root, MAX(lastDequeuedByProcess)-1}
+\EndIf}
+
     \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
@@
     \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
+    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryAppend}{\var{new}, [\var{h}]} \label{GCcas}}
     \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
+    \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ \var{b}}
         \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
-    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
+            \EndIf
+\State \linecomment{response is the \var{e}th enqueue in the root}
         \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
@@
 		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
-        \State \Return \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
-    \EndIf
+        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
+        \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ $\var{b}'$}
+        \State \Return{\var{e}}
+
 \EndFunction{FindResponse}
@@
         \EndIf
-        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
+        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
@@
 \EndFunction{GetEnqueue}
+
+\gc{
+\Function{void}{Help}{}
+\State \bf{for} \var{i} \assign\ 1 to $p$
+\State h=leafOfProcess[\var{i}]
+\If{leafOfProcess[\var{i}].blocks[h].num\sub{deq}==1 and leafOfProcess[\var{i}].IndexDequeue(h,1)!=null}
+\State <b, i> \assign\ \Call{IndexDequeue}{leafOfProcess[\var{i}], h, 1}
+\State output \assign\ \Call{FindResponse}{b, i} 
+\State leafOfProcess[\var{i}].blocks[h].response \assign\ output
+\EndIf
+\State \bf{end for}
+\EndFunction{Help}
+
+\Function{}{FreeMemory}{\typ{Node} \var{n}, \typ{int} \var{i}}
+\State Split \var{n}.blocks.split over \var{i} and assign the grater tree to \var{n}.blocks with \tt{CAS}.
+\If{\var{n} is not a leaf}
+\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{left}}
+\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{right}}
+\EndIf
+\EndFunction{FreeMemory}
+
+}
 \end{algorithmic}

### 2023-01-03 podc_paper/boundedspace.tex
working on proof of propagation

@@
 Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.
+Mention that known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.
+But we do not need the full functionality:  in particular, we need an operation that tries to insert a node that is allowed to fail if a concurrent insertion occurs for our refresh operation.
 

### 2023-01-03 podc_paper/conclusion.tex
working on proof of propagation

@@
 An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and takes $O(\log p)$ steps.
-\here{doublecheck all preceding claims for accuracy}.
+\here{doublecheck all preceding claims for accuracy.}
 As future work we would like to investigate whether a similar approach could be used for stacks or deques.

### 2023-01-03 podc_paper/correctness.tex
working on proof of propagation

@@
 
-Since \fld{head} fields are updated only on line \ref{incrementHead}, which increments it, we have the following observation.
+A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
+when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
+Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
+(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
+Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
+(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
+Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
+which implies the following.
+
 \begin{observation} \label{nonDecreasingHead}
@@
 
--------
+This allows us to prove that every block a Refresh installs contains at least one operation.
+
+\begin{corollary}\label{blockNotEmpty}
+If a block $B$ is in $v.\fld{blocks}[i]$ where $i>0$, then $E(B)$ and $D(B)$ are not both empty.
+\end{corollary}
+\begin{proof}
+The \op{Refresh} that installed $B$ got $B$ as the response to its call to \op{CreateBlock} on line \ref{invokeCreateBlock}.
+Thus, at line \ref{testEmpty} $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
+By \Cref{lem::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
+so these sequences cannot both be empty.
+\end{proof}
 
 \subsection{Propagating Operations to the Root}
-This section explains why two \nf{Refresh}es are enough to propagate a node's operations to its parent.
+Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
+We say that node $v$ \emph{contains} an operation if some block in $v.\fld{blocks}$ contains the operation.
+\here{move this defn earlier and just recall it here?}
+Once a block is added to a node, it remains there forever.  Thus, if $v$ contains an operation at some time, it contains the operation at all later times too.
 
-  We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
-
-\begin{definition}
-Let $t^{op}$ be the time $op$ is invoked, $^{op}t$ be the time $op$ terminates, $t_{l}^{op}$ be the time immediately before running Line $l$ of operation $op$ and $^{op}_{l}t$ be the time immediately after running Line $l$ of operation $op$.
-We sometimes suppress $op$ and write $t_{l}$ or $_l t$ if $op$ is clear from the context.
-In the text, $v_l$ is the value of variable $v$ immediately after line $l$ for the process we are talking about and $v_t$ is the value of variable $v$ at time $t$.
-\end{definition}
-
-\begin{definition} [Successful Refresh]
-  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}.
-  % If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
-\end{definition}
-
-In the next two results, we show that for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.
-
-\begin{lemma} \label{lem::trueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
+\begin{lemma}\label{successfulRefresh}
+Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
+In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
+when $R$ executed line~\ref{readHead}.
 \end{lemma}
 \begin{proof}
+Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
+in the configuration $C$ immediately before $R$ executed line \ref{readHead}.
+Then some block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
+By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
+$v.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
+If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
+$v.\fld{left.head} > i$.
+Then, $R$ calls \op{CreateBlock}($v,h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
+\op{CreateBlock} reads a value greater than $i$ from $v.\fld{left.head}$ at line \ref{createEndLeft}.
+Thus, $new.\eleft \geq i$.  We consider two cases.
 
-We show 
-\vspace{-2em}\begin{align*}
-EST_{n\nf{.left}}^{t^R}&= ops(n\nf{.left.blocks[0..n.left.head$_{t^R}-1$]}) \\
-&\subseteq ops(n\nf{.blocks}_{\ref{cas}}) = ops(n\nf{.blocks[0..}n\nf{.head}_{\ref{cas}}\nf{]}).
-\end{align*}
+Suppose $R$'s call to \op{CreateBlock} returns the new block $B'$ and $R$'s \op{CAS} at line \ref{cas} 
+installs $B'$ in $v.\fld{blocks}$.
+Then, $B$ is a subblock of some block in $v$, since  $B'.\eleft$ is greater than or equal to $B$'s index
+in $v.\fld{left.blocks}$.
+Hence $B'$ contains $op$, as required.
 
-In every node, \nf{blocks[$0$]} is an empty block without any operations.
-Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n\nf{.left.head}_{\ref{lastLine}}-1$.
-Therefore, by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{$n$.left.blocks[$1\cdots n\nf{.left.head}_{\ref{lastLine}}-1$]} are subblocks of \nf{$n$.blocks[$1\cdots n\nf{.head}_{\ref{readHead}}$]}.
-Because of Observation \ref{nonDecreasingHead} we have $n\nf{.left.head}_{t^R}-1\leq n\nf{.left.head}_{\ref{lastLine}}-1$ and $n\nf{.head}_{\ref{readHead}}\leq n\nf{.head}_{\ref{cas}}$.
-From Observation \ref{lem::blocksOrder} the claim follows.
-The proof for the right child is the same.
+\Eric{I think this case was missing from the proof in the thesis}
+Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
+Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
+We formalize this intuition.
+The value computed at line \ref{createSumEnq} is
+\begin{eqnarray*}
+\var{num\sub{enq}} 
+&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
+&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
+&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
+\end{eqnarray*}
+It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
+$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
+$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
+Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
+Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
+these blocks contain no operations.
+By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq new.\eleft \geq i$.
+Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
 \end{proof}
 
-\begin{corollary}\label{lem::prectrueRefresh}
-If $R$ is a successful instance of \nf{$n$.Refresh} that terminates, then we have $$EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}.$$
-\end{corollary}
+We now show that if a process fails its \op{Refresh}($v$) twice, some other process must have succeeded.
+
+\begin{lemma}
+Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
+When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
+\end{lemma}
 \begin{proof}
-The left-hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$.
-Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.
-Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates.
-So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$.
+If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
+from Lemma \ref{successfulRefresh}.
+So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
+Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
+By \Cref{lem::headInc}, $h_2>h_1$.
+By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
+Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
+a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
+$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
+Thus, all operations contained in $v$'s children when $R_1$ begins
+are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
+By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
+which is before $R_2$ terminates.
 \end{proof}
 
-In the next lemma, we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.
-
-\begin{lemma} \label{doubleRefresh}
-  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on an internal node $n$.
-If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
-\end{lemma}
-\begin{proof}
-Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}.
-By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$.
-By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.
-
-Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}.
-As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}.
-$R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}.
-By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}).
-By Lemma \ref{lem::trueRefresh} we have $EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$.
-Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{$n$.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ into $n$\nf{.blocks[$i$]} is established by the time $R_2$ terminates.
-To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$), so we have $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}})$.
-$R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Lemma \ref{lem::headInc} $n$\nf{.head} has been incremented when $R_2$ terminates and the block $R_2^\prime$ put into $n$ is established by then.
-So we have
-$ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n}^{^{R_2}t}$.
-
-If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented at least two times since $_{\ref{readHead}}^{R_1}t$.
-By Invariant \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[$i+1$]}$ is non-null.
-Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$.
-$R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}.
-So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^{R_{2}}t$.
-From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh} and \ref{lem::headInc} on $R_3$ the claim holds.
-\end{proof}
-
-\begin{figure}[hbt]
-  \center\includegraphics[width=6in]{pics/compactdouble.png}
-  \caption[Time relations of the \nf{CAS}es when a process fails to \nf{Refresh} a node two times.]{\label{fig::doubleRefresh}$_{R_1}t<$ $t_{\ref{readHead}}^{R_1}$ $<$ incrementing \nf{$n$.head} from $i$ to $i+1$ $<$ $t_{\ref{readHead}}^{R_2^\prime}$ $<$ $t_{\ref{cas}}^{R_2^\prime}$ $<$ incrementing \nf{$n$.head} from $i+1$ to $i+2$ $<t_{R_2}$}
-\end{figure}
+-------
 

### 2023-01-03 podc_paper/algorithm.tex
working on proof of propagation
working on proof; some edits to intro to save space

--- 2023-01-02 podc_paper/algorithm.tex
@@
 See Figure \ref{tournament} for an example.
-In this section, we assume for simplicity that each node has an infinite blocks array.
-Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.
+In this section, we use an infinite array of blocks in each node.
+Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.
 
-To perform an operation on the queue, a process $P$ creates a block containing that single 
-operation and appends it to the blocks array in $P$'s leaf.
+To perform an operation on the queue, a process $P$ appends a new block containing that  
+operation to the blocks array in $P$'s leaf.
 Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
@@
 and \eleft\ and \eright\ fields are shown using arrows as before.
-Root blocks also contain the additional \fld{size} field.
+Root blocks also have the additional \fld{size} field.
 The \fld{super} field is not shown.}
@@
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmpty}
         \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}

### 2023-01-03 podc_paper/analysis.tex
working on proof; some edits to intro to save space

@@
 \section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the time complexity of our algorithm.
+In this section, we analyze the number of \nf{CAS} invocations and the step complexity of our algorithm.
+
+First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
+
+\begin{lemma}\label{blockSize}
+Each block $B$ in each node contains at most one operation of each process.
+If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
+\end{lemma}
+\begin{proof}
+Suppose $B$ contains an operation of process $p$.
+Let $op$ be the earliest operation by $p$ contained in $B$.
+When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
+Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
+those operations are invoked.
+
+Let $t$ be the earliest termination of any operation contained in $B$.
+By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
+are invoked before $t$.  Thus, they are all concurrently running at $t$, so there are at most
+$c$ operations contained in $B$.
+By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
+in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
+By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
+so there are at most $c$ direct subblocks.
+\end{proof}
+
 \begin{proposition}

### 2023-01-03 podc_paper/correctness.tex
working on proof; some edits to intro to save space

@@
 
-\begin{lemma}
+\begin{lemma}\label{lem::doubleRefresh}
 Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
@@
 
--------
-
-\begin{corollary} \label{lem::propagateStep}
-  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
-\end{corollary}
+\begin{lemma} \label{lem::appendExactlyOnce}
+When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
+\end{lemma}
 \begin{proof}
-  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true}, then by Corollary \ref{lem::prectrueRefresh} the claim holds.
-If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded, the claim still holds by Corollary \ref{lem::prectrueRefresh}.
-Otherwise, both failed, and the claim is implied by Lemma \ref{doubleRefresh}.
-\end{proof}
-
-Now we show that after \nf{Append($B$)} on a leaf finishes, the operation contained in $B$ will be established in \nf{root}.
-
-\begin{corollary}\label{lem::append}
-For $A=l\nf{.Append(}B\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
-\end{corollary}
-\begin{proof}
-$A$ adds $B$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $B$.
-For every node $n$, $n.$\nf{Propagate} appends $B$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} until $n$ is \nf{root}.
-\end{proof}
-
-\begin{corollary} \label{lem::appendExactlyOnce}
- After $l$\nf{.Append($B$)} finishes, $B$ is a subblock of exactly one block in each node along the path from $l$ to the \nf{root}.
-\end{corollary}
-\begin{proof}
-By the previous corollary and Lemma \ref{lem::noDuplicates} there is exactly one block in each node containing~$B$.
+The \op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
+performs a double \op{Refresh} on each node along the path from the leaf's parent to the root.
+By \Cref{lem::doubleRefresh}, this ensures $B$'s operation is contained in each node along this path.
+By \Cref{lem::noDuplicates}, it is contained in exactly one block in each node.
 \end{proof}
@@
 \subsection{Correctness of GetEnqueue}
-
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
-
-\begin{lemma}\label{oneOpinBlock}
-Each block contains at most one operation of each process.
-\end{lemma}
-\begin{proof}
-To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$.
-Without loss of generality $op_1$ is invoked earlier than $op_2$.
-Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins.
-By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root.
-Since $b$ contains $op_2$, it must be created after $op_2$ is invoked.
-The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$.
-The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
-\end{proof}
-
-\begin{lemma}\label{blockSize}
-    Each block contains at most $c$ operations, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-There is a time that all the operations in a block are concurrent, because otherwise if there is an operation in a block that has ended before another operation in that block starts, then by Corollary \ref{lem::appendExactlyOnce} these two operations couldn't be in the same block.
-From the definition of $c$ we know at any time in the execution there cannot be more than $c$ concurrent operations, and from the previous lemma we know a process has at most one operation in a block, so there cannot be a block with more than $c$ operations.
-\end{proof}
-
-\begin{lemma}\label{subBlocksBound}
-Each block has at most $c$ direct subblocks, where $c$ is the maximum number of concurrent operations at any time in the whole execution ($c\leq p$).
-\end{lemma}
-\begin{proof}
-From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block.
-We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}.
-Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
-By Lemma \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
-\end{proof}
+-------
 

### 2023-01-03 podc_paper/introduction.tex
working on proof; some edits to intro to save space

@@
 have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
-the number of processes that access the data structure concurrently (which can be $p$ in the worst case).
+the number of processes that access the data structure concurrently, which can be $p$ in the worst case.
 Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
 that amortized $\Omega(c)$ steps per operation are indeed necessary
-for CAS-based implementations of lock-free bag data structures, which provide operations
+for any CAS-based implementation of a lock-free bag data structure, which provide operations
 to insert an element or remove an arbitrary element (chosen non-deterministically).
@@
 
-In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
+We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
 whose step complexity does not depend linearly on $p$.
@@
 stored in the queue.
-For simplicity of presentation, we first give an unbounded-space construction that uses
+For ease of presentation, we first give an unbounded-space construction that uses
 $O(\log^2 p + \log q)$ steps per operation.
@@
 \here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
-Both implementations use single-word CAS instructions and reasonably-sized words.
+Both implementations use single-word CAS and reasonably-sized words.
 Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
@@
 %an unbounded number of CAS instructions in the worst case and 
-$\Omega(p)$ CAS instructions in an amortized sense.
+$\Omega(p)$ CAS instructions, even in an amortized sense.
 For the space-bounded implementation, we unlink unneeded memory from our data structure.
@@
 If we explicitly stored all operations in the nodes, this helping would become very costly.
-Instead, we use an implicit representation of collections
+Instead, we use a novel implicit representation of collections
 of operations that allows for quickly merging two collections from the two children of a node,

### 2023-01-03 podc_paper/related.tex
working on proof; some edits to intro to save space

@@
 Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
-added Herlihy's helping
-technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
-a wait-free queue.
-Then, Kogan and Petrank \cite{10.1145/2370036.2145835} introduced the 
-fast-path slow-path methodology for making data structures wait-free:
-The fast path has good performance and the slow path guarantees termination.
+used Herlihy's helping
+technique~\cite{10.1145/114005.102808} to make the MS-queue
+wait-free.
+Then, they introduced the 
+fast-path slow-path methodology \cite{10.1145/2370036.2145835} for making data structures wait-free:
+the fast path has good performance and the slow path guarantees termination.
 They applied their methodology to combine the MS-queue (as the fast path)
-with their previous wait-free queue (as the slow path).
+with their wait-free queue (as the slow path).
 Ramalhete and Correia \cite{RC17} added a different helping mechanism to the MS-queue.
-Although these approaches can improve practical performance,
-their amortized step complexity remains $\Omega(p)$. 
+Although these approaches can perform well in practice,
+the amortized step complexity remains~$\Omega(p)$. 
 \Hossein{I get what this sentence means, but from a high overview i think it is inaccurate.
@@
 \Eric{I modified the wording.
-I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussin array-based queues.}
+I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussing array-based queues.}
 
@@
 This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
-Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} have also used this methodology
+Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
 on array-based queues.
-In worst-case executions, where processes must execute the slow path,
-these implementations also take $\Omega(p)$ steps per operation, 
+In worst-case executions that use the slow path,
+they also take $\Omega(p)$ steps per operation, 
 due either to the CAS retry problem or helping mechanisms.
@@
 However, our construction requires more intricate
-data structures to represent sets of operations, since the state of a queue cannot easily be represented as succinctly
+data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
 as the single-word state of a fetch\&increment object.

### 2023-01-03 podc_paper/algorithm.tex
working on correctness of getEnqueue

@@
 operations that were propagated from the left child before the operations from the right child.
-More formally, we inductively define an sequence $E(B)$ of the enqueues represented by a block $B$.
-If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is either the single
-enqueue represented by the block (if $B$ stores an enqueue), or the empty sequence (if $B$ stores a dequeue).
+More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
+represented by a block $B$.
+If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
+and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
 If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
-and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s enqueue sequence is the concatenation $E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)$.
-The dequeue sequence $D(B)$ of a block $B$ is defined symmetrically.  We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
+and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
+\begin{eqnarray}
+E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
+D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
+\end{eqnarray}
+We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
 
@@
 Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
-Thus, the direct subblocks of $v.\fld{blocks}[i]$ are
+Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
 \begin{eqnarray}\label{defsubblock}
-v.\fld{left.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{left}}+1..v.\fld{blocks}[i].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
-v.\fld{right.blocks}[v.\fld{blocks}[i-1].\fld{end\sub{right}}+1..v.\fld{blocks}[i].\fld{end\sub{right}}].
+v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
+v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
 \end{eqnarray}
@@
 Blocks also store prefix sums of the numbers of enqueues and dequeues:
-the block in $v.\fld{blocks}[i]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{i}]$.
+the block in $v.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
+that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
 These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
@@
             \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
-        \EndIf
+        \EndIf \label{endChooseDir}
         \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
@@
 $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
-        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
-        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} 
+        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
+        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
     \EndIf
@@
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
+\caption{\label{pseudocode2}Queue implementation, continued.}
+%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
 \end{figure}

### 2023-01-03 podc_paper/correctness.tex
working on correctness of getEnqueue

@@
 \subsection{Correctness of GetEnqueue}
--------
 
-\nf{DoublingSearch($e$, $end$)} returns a pair \nf{<$b$, $i$>} such that the $i$th \nf{Enqueue} in the $b$th block of the root is the $e$th \nf{Enqueue} in the sequence stored in the root.
-
-\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
-If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
+\begin{lemma}\label{lem::get}
+If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
 \end{lemma}
 \begin{proof}
- From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order.
-Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing.
-By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the all blocks before that block in the node.
-Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$.
-Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.
-Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$.
-Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}.
-By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$.
-Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
+We prove the claim by induction on the height of node $v$.
+If $v$ is a leaf, the hypothesis implies that $i=1$ and the block $v.\fld{blocks}[b]$ represents 
+an enqueue whose argument is stored in $v.\fld{blocks}[b].\fld{element}$.
+\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.
+
+Assuming the claim holds for $v$'s children, we prove it for $v$.
+Let $B$ be $v.\fld{blocks}[b]$.
+By (\ref{defSeqs}),
+$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
+of $B$, which are listed in (\ref{defsubblock}).
+By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
+of enqueues in $E(B)$ that come from $B$'s subblocks in $v$'s left child.
+Thus, $dir$ is set to the direction for the child of $v$ that contains the required enqueue.
+Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
+Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
+By \Cref{lem::sum}, $v.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
+enqueues in $E'$ before the enqueues of block $B'$, so
+the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
+Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
+returns the required result, by the induction hypothesis.
 \end{proof}
 
-\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
-If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
-\end{lemma}
-\begin{proof}
-We will prove this lemma by induction on the node $n$'s height.
-For the base case, suppose $n$ is a leaf.
-Leaf blocks each contain exactly one operation, $n\nf{.blocks[$b$].sum\sub{enq}}\leq 1$, which means only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf and $n$\nf{.blocks[$b$]} must contain an \nf{Enqueue} operation.
-Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$, as required.
-
-For the induction step, we prove if \nf{$n$.dir.GetEnqueue($b^\prime$, $i$)} returns $E_i(n.\nf{dir}, b^\prime)$ then \nf{$n$.GetE-}\\\nf{nqueue($b$,~$i$)} returns $E_i(n,b)$.
-From Definition \ref{ordering} of $E(n,b)$, we know that operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}).
-By Lemma \ref{lem::numX}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue} operations from the blocks' subblocks in the left child of $n$.
-So the $i$th \nf{Enqueue} operation in $n\nf{.blocks[}b\nf{]}$ is propagated from the right child if and only if $i$ is greater than $n\nf{.blocks[}b\nf{]}$\nf{.num\sub{enq-left}}.
-Line \ref{leftOrRight} decides whether the $i$th enqueue in the $b$th block  of internal node $n$ is in the  left child or right child subblocks of \nf{$n$.blocks[$b$]}.
-By Definitions \ref{def::subblock} and \ref{def::ops}, to find an operation in the subblocks of \nf{$n$.blocks[$b$]} we need to search in the range
-\begin{align*}
-&\texttt{$n$.left.blocks[$n$.blocks[$b$-1].end\textsubscript{left}+1..$n$.blocks[$b$].end\textsubscript{left}]}  \textrm{ or } \\
-&\texttt{$n$.right.blocks[$n$.blocks[$b$-1].end\textsubscript{right}+1..$n$.blocks[$b$].end\textsubscript{right}]}.
- \end{align*}
-First, we consider the case where the \nf{Enqueue} we are looking for is in the left child.
-There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ \nf{Enqueue}s in the blocks of $n$\nf{.left} before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.\tt{left})$ which is $E_{i^\prime}(n.\tt{left},b^\prime)$ for some $b^\prime$ and $i^\prime$.
-We can compute $b^\prime$ and then search for the $i^\prime$th \nf{Enqueue} in \tt{$n$.left.blocks[$b^\prime$]}, where $i^\prime$ is $i+eb-n$\tt{.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}.
-The parameters in Line~\ref{leftChildGet} are for searching $E_{i+eb}(n\nf{.left})$ in \nf{$n$.left.blocks} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.
-
-Otherwise, the \nf{Enqueue} we are looking for is in the right child.
-Because \nf{Enqueue}s from the left subblocks are ordered before the ones from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child.
-So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ in the right child (Line \ref{rightChildGet}).
-Other parameters for the right child are chosen similarly to the left child.
-
-So, in both cases, the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
-$subblockIndex$ is the index of the block in $n.\nf{dir}$ containing $E_i(n,b)$.
-Finally, \nf{$n$.child.GetEnqueue( $subblockIndex$, $i$)} is invoked  and it returns $E_i(n,b)$\nf{.element} by the hypothesis of the induction.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumEnq.png}
-
-  \caption[\nf{Enqueue} operations propagated from the left and the right child to a node.]{The number and order of the \nf{Enqueue} operations propagated from the left and the right child to \nf{$n$.blocks[$b$]}.
-Both \nf{$n$.blocks[$b$]} and its subblocks are shown in grey.
-\nf{Enqueue} operations from the left child (colored red), are ordered before the \nf{Enqueue} operations from the right child (colored blue).
-  }\label{figGet}
-\end{figure}
+-------
 

### 2023-01-03 podc_paper/related.tex
working on correctness of getEnqueue

@@
 uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
-
-Jayanti and Petrovic introduced a wait-free poly-logarithmic
+Jayanti and Petrovic gave a wait-free poly-logarithmic
 queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
-Our implementation uses their idea of having
-a tournament tree among processes to agree on the linearization of
+We adapt their idea of processes using
+a tournament tree to agree on a linearization of
 operations.

### 2023-01-03 podc_paper/correctness.tex
working on proof

@@
 For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
+If $v\neq \var{root}$,  $v.\fld{blocks}[i].super \neq \nl$ for $0<i<v.head$.
 \end{invariant}
@@
 \begin{proof}
-Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claim~holds.
+Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.
 
@@
 by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
-It suffices to show that $v.\fld{blocks}[head] \neq \nl$.
+For the first two claims, it suffices to show that $v.\fld{blocks}[head] \neq \nl$.
 \nf{Advance} is called either at line \ref{helpAdvance} 
@@
 or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$.
+For the third claim, observe that prior to incrementing $v.\head$ at line \ref{incrementHead},
+the \op{CAS} at line \ref{setSuper1} ensures that $v.\fld{blocks}[i].super\neq \nl$.
 \end{proof}
@@
 
+\subsection{Correctness of IndexDequeue}
+
+The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
+
+\begin{lemma}
+  If $B=v.\var{blocks}[b]$ is a direct subblock of $v.\fld{parent.blocks}[s]$ then $B.\fld{super}\leq s$.
+\end{lemma}
 -------
 
-\subsection{Correctness of IndexDequeue}
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node. 
-\here{mention that precondition of IndexDequeue follows; also check that block of parent indexed by sup in that routine is non-null}
+\begin{proof}
+Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$s$]} into $n\nf{.parent}$.
+By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
+By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
+The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
+From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq s$, since \nf{$n$.parent.head} is still equal to $s$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
+The reader may wonder when the case $b\nf{.super}=s$ happens.
+This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
+\end{proof}
+
 Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
@@
 
-\begin{observation} \label{setSupBeforeIncHead}
-After \nf{$n$.blocks[$i$].CAS(null, $B$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ until \nf{$B$.super} is set.
-\end{observation}
-\begin{proof}
-From Observation \ref{nonDecreasingHead} we know that \nf{$n$.head} changes only by the increment on Line \ref{incrementHead}.
-Before an instance of \nf{Advance} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
-\end{proof}
-
-\begin{corollary}
-  If \nf{$n$.blocks[$i$].super} is \nf{null}, then $\nf{$n$.head}\leq i$ and \nf{$n$.blocks[$i+1$]} is \nf{null}.
-\end{corollary}
-\begin{proof}
-By Invariant \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
-\end{proof}
 Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
@@
 \end{observation}
-
-\begin{lemma}
-  If $B=n$\nf{.blocks[}$b$\nf{]} is a direct subblock of \nf{$n$.parent.blocks[$superblock$]} then $\nf{$B$.super}\leq superblock$.
-\end{lemma}
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$superblock$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq superblock$, since \nf{$n$.parent.head} is still equal to $superblock$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=superblock$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
 
@@
 We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
+\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}
 

### 2023-01-03 podc_paper/algorithm.tex
working on proof
minor edits

@@
 Following \cite{}, we call this a 3-step sequence a
-\op{refresh} on $v$. %(see Figure \ref{fig::propagstep}).
-A successful \op{refresh} by $P$ may propagate pending operations by several other processes to $v$.
+\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
+A successful \op{Refresh} by $P$ may propagate pending operations by several other processes to $v$.
 This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
@@
 \label{construct}
-Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{refresh}.
+Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{Refresh}.
 \item
@@
 
-The operations propagated by a \op{refresh} are all pending when the \op{refresh} occurs,
+The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
 so there is at most one operation per process.
@@
 
-Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
-we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
-A \op{refresh} on \node\ $v$ reads the value $h$ of the \fld{head} field of each of $v$'s children and stores 
-$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
+Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
+we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
+A \op{Refresh} on \node\ $v$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $v$'s children and stores 
+$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
 Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
@@
 superblock will eventually be installed in $B$'s parent.
-So, we leave this field blank.  After $B$ is installed 
-in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill it in 
-when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$.
-They use the value read from the \fld{head} field of $v$'s parent at that time.
+So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
+in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill in $B.\fld{super}$ 
+when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$,
+using a value they read from the \fld{head} field of $v$'s parent.
 As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
@@
     \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
+    \If{$\var{v}\neq \var{root}$}
 	    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
     	\State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
+	\EndIf
     \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{48}
+\setcounter{ALG@line}{50}
 
@@
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
-    \State \linecomment{Preconditions:  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
-    \State \linecomment{contains at least \var{i} dequeues}
+    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
+    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
+    %  \State \linecomment{contains at least \var{i} dequeues}
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{99}
+\setcounter{ALG@line}{102}
 

### 2023-01-04 podc_paper/algorithm.tex
working on proof (mainly indexdequeue)

@@
 
-An operation can terminate only after a block containing it has been appended to the root's blocks array.
+We shall show that an operation terminates only after a block containing it has been appended to the root's blocks array.
 So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
@@
 So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
-in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill in $B.\fld{super}$ 
-when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$,
+in $v.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
+when they attempt to advance $v.\fld{head}$ from $h$ to $h+1$,
 using a value they read from the \fld{head} field of $v$'s parent.
-As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
+As mentioned above, this might not be the exact index of $B$'s superblock, but we
 shall prove that it is close.
@@
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
     \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
@@
     	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}
 	    \EndIf
@@
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
-	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
+	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
         \If{$\fld{dir} = \fld{right}$} 
@@
 
-\opemph{GetEnqueue}(\var{v,b,i}) returns the argument of the
+\opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
 $i$th enqueue in the $b$th block of \typ{Node} $v$. 
@@
 the exact subblock is found using binary search on the \fld{sum\sub{enq}}
-field (line \ref{getChild}). 
+field (line~\ref{getChild}). 
 \here{explain a little more?}

### 2023-01-04 podc_paper/correctness.tex
working on proof (mainly indexdequeue)

@@
 
-The next few results show that the \nf{super} field of a block is accurate within one of the actual index of the block's superblock in the parent node.
+Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.
+\here{An alternative would be modify the code
+so that the correct index of the superblock is computed before storing it in $B.super$.
+In Advance, we could check if $h>v.parent.blocks[h_p].end_{left/right}$ and if so write $h_p+1$ instead of $h_p$ in $v.blocks[h].super$.
+Then, in IndexDequeue there would be no need to correct the value read from the super field.
+If we do this, the following lemma would be modified to prove that the value is perfectly accurate.}
+
+\begin{lemma}\label{superRelation}
+Let $B=v.\var{blocks}[b]$.
+  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
+\end{lemma}
+\begin{proof}
+We first show $B.\fld{super}\leq s$.
+Let $R_s$ be the instance of \op{Refresh}($v.\fld{parent}$) that installs the superblock of $B$ 
+in $v.\fld{parent.blocks}[s]$.
+By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $v.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
+By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
+that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $v.\fld{parent}.\head$ before $r$.
+When $r$ occurs, $v.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
+\ref{cas} succeeds.
+So, by \Cref{lem::headPosition}, $v.\fld{parent}.\head \leq s$ when $r$ occurs.
+Since the value stored $B.\fld{super}$ was read from $v.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
+
+Next, we show that $B.\fld{super}\geq s-1$.
+The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
+So, if $s\leq 2$, the claim is trivial.  Assume for the remainder that $s>2$.
+By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
+$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
+Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
+This read $r'$ must be before $B$ is installed in $v$;
+otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
+$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
+Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
+It is invoked either 
+at line \ref{helpAdvance} after seeing $v.\fld{blocks}[h]\neq \nl$ at line \ref{ifHeadnotNull}
+or at line \ref{advance} after ensuring $v.\fld{blocks}[h]\neq \nl$ at line~\ref{cas}.
+Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
+By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
+the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
+writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
+\end{proof}
+
+%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
+%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
+%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
+
+We prove \op{IndexDequeue}'s correctness using \Cref{superRelation} on each step of the \op{IndexDequeue}.
+\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}
 
 \begin{lemma}
-  If $B=v.\var{blocks}[b]$ is a direct subblock of $v.\fld{parent.blocks}[s]$ then $B.\fld{super}\leq s$.
-\end{lemma}
--------
-
-\begin{proof}
-Let $R_p$ be the instance of $n$\nf{.parent.Refresh} that does a successful \nf{CAS}(Line~\ref{cas}) and puts the superblock of $B$ which is $n$\nf{.parent.blocks[$s$]} into $n\nf{.parent}$.
-By Observation~\ref{lem::createBlockHead} if $R_p$ propagates $B$ it has to read a greater value than $b$ from \nf{$n$.head}, which means \nf{$n$.head} was incremented from $b$ to $b+1$ in Line \ref{incrementHead}.
-By Observation \ref{setSupBeforeIncHead} \nf{$B$.super} was already set in Line \ref{setSuper1}.
-The value written in \nf{$B$.super}, was read in Line \ref{readParentHead} before the \nf{CAS} that sets \nf{$B$.super} in Line \ref{setSuper1}.
-From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $B\nf{.super}\leq s$, since \nf{$n$.parent.head} is still equal to $s$ when $R_p$ executes its \nf{CAS} at Line \ref{cas} by Lemma \ref{lem::headProgress}.
-The reader may wonder when the case $b\nf{.super}=s$ happens.
-This can happen when $\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-\end{proof}
-
-Then we explain how it is used to compute the rank of a given \nf{Dequeue} in the root.
-\begin{definition}\label{orderRefresh} If a \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
-\end{definition}
-
-Now let us consider how the \nf{Refresh}es that took place on the parent of node $n$ after block $B$ was stored in $n$ will help to set \nf{$B$.super} and propagate $B$ to the parent.
-
-\begin{observation}\label{lem::createBlockHead}
-If the block created by an instance $R_p$ of $n$\nf{.parent.Refresh} contains block $B=n$\nf{.blocks[}$b$\nf{]} then $R_p$ reads a value greater than $b$ from $n$\nf{.head} in Line \ref{lastLine}. 
-\end{observation}
-
-\begin{lemma}\label{lem::secondRefreshSuper}
-Let $R_n$ be a \nf{Refresh} that puts $B$ in \nf{$n$.blocks[$b$]} at Line \ref{cas}.
-Then, the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to Definition \ref{orderRefresh} contains $B$ and \nf{$B$.super} is set when the second successful \nf{$n$.parent.Refresh} reaches Line \ref{invokeCreateBlock}.
-\end{lemma}
-
-\begin{proof}
-Let $R_{p1}$ and $R_{p2}$ be the next two successful \nf{$n$.parent.Refresh}es after $R_n$. To derive a contradiction assume $B$ was neither propagated to $n$\nf{.parent} by $R_{p1}$ nor by $R_{p2}$.
-
-Since $R_{p2}$'s created block does not contain $B$, by Observation \ref{lem::createBlockHead} the value $R_{p2}$ reads from $n$\nf{.head} in Line \ref{lastLine} is at most $b$.
-From Observation \ref{nonDecreasingHead} the value $R_{p2}$ reads in Line \ref{readChildHead} is also at most~$b$.
-
-$R_n$ puts $B$ into $n$\nf{.blocks[}$b$\nf{]} so ${R_n}$ reads the value $b$ from $n.$\nf{head}.
-Since $R_{p2}$'s \nf{CAS} into $n$\nf{.parent.blocks} is successful there should be a \nf{Refresh} instance $R_p^\prime$ on $n$\nf{.parent} that increments $n$\nf{.parent.head} (Line \ref{incrementHead}) after $R_{p1}$'s Line \ref{cas} and before $R_{p2}$'s Line \ref{readHead}.
-We assumed $t^{R_n}_{\ref{cas}}<t^{R_{p1}}_{\ref{cas}}<t^{R_{p2}}_{\ref{cas}}$ by Definition \ref{orderRefresh}.
-Finally, Line \ref{readChildHead} is after Line \ref{readHead} and $R_{p2}$'s Line \ref{readHead} is after $R_p^\prime$'s Line \ref{incrementHead}, which is after ${R_n}$'s $n$\nf{.blocks.CAS}.
-\begin{align*}
-\left.
-\begin{array}{r}
- _{\ref{cas}}^{R_n}t<_{\ref{cas}}^{R_{p1}}t\\
-_{\ref{cas}}^{R_{p1}}t<_{\ref{incrementHead}}^{R_{p^\prime}}t<_{\ref{readHead}}^{R_{p2}}t\\
-_{\ref{readHead}}^{R_{p2}}t<_{\ref{readChildHead}}^{R_{p2}}t\\
-\end{array}
-\right\} &\Longrightarrow  _{\ref{cas}}^{R_n}t<_{\ref{readChildHead}}^{R_{p2}}t
-\end{align*}
-
-So $R_{p2}$ reads a value greater than or equal to $b$ for $n.\nf{head}$ by Observation \ref{nonDecreasingHead}.
-
-Therefore $R_{p2}$ reads $n\nf{.head}=b$.
-$R_{p2}$ calls $n$\nf{.Advance} at Line \ref{helpAdvance}, which ensures $n$\nf{.head} is incremented from $b$.
-So the value $R_{p2}$ reads in Line \ref{lastLine} of \nf{CreateBlock} is greater than $b$ and $R_{p2}$'s created block contains $B$.
-This is in contradiction with our hypothesis.
-
-Furthermore, if $B$\nf{.super} was not set earlier, it is set by $R_{p2}$'s call to $n$\nf{.Advance} invoked from Line~\ref{helpAdvance}.
-\end{proof}
-
-\begin{corollary} \label{superRelation}
-If \nf{$B=n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$B$.super} is equal to or one less than the index of the superblock of $B$.
-\end{corollary}
-\begin{proof}
-% After that $B$ is installed, $n$\nf{.parent.head} is read, and $B.$\nf{super} field is set to the value read from the parent's \nf{head} (see Lines \ref{readParentHead} and \ref{setSuper1} of\nf{Advance}).
-Let $R_n$ be the \nf{$n$.Refresh} that put $B$ in $n$\nf{.blocks} and let $R_{p1}$ be the first successful \nf{$n$.parent.Refresh} after $R_n$ and $R_{p2}$ be the second next successful \nf{$n$.parent.Refresh}.
-Before $B$ can be propagated to $n$'s \nf{parent}, $n$\nf{.head}  must be greater than $b$, so by Observation \ref{setSupBeforeIncHead} $B$\nf{.super} is set.
-From Lemma \ref{lem::secondRefreshSuper} we know that $B$ is propagated by the second next successful \nf{Refresh}'s \nf{CAS} on $n$\nf{.parent.blocks}.
-To summarize, we have $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p2}}t}= n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}+1$ and  $n\nf{.parent.head}_{_{\ref{cas}}^{R_{p1}}t}\leq n\nf{.parent.head}_{_{\ref{cas}}^{R_n}t}$ from Definition \ref{orderRefresh} and Observation \ref{nonDecreasingHead}.
-The value that is set in $B$\nf{.super} is read from $n$\nf{.parent.head} after ${_{\ref{cas}}^{R_n}t}$.
-So $B$\nf{.super} is equal to or one less than the index of the superblock of $B$.
-\end{proof}
-
-We prove \nf{IndexDequeue}'s correctness using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue}.
-\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}
-
-\begin{lemma}[\nf{IndexDequeue} correctness]
- If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns $<x,y>$ such that $D_i(n,b)=D_y(\nf{root},x)$.
+If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
+ then \op{IndexDequeue}($v, b, i$) returns $\langle b',i'\rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
 \end{lemma}
 \begin{proof}
-We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (see Line \ref{indexBaseCase}).
-For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes $superblockIndex$, the index of the superblock of the $b$th block in $n$, in Line \ref{computeSuper} by Corollary \ref{superRelation}.
-After that, the position of $D_i(n,b)$ in $D(n.\nf{parent}, superblockIndex)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}.
-By Definition \ref{ordering}, \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right.
-If $D_i(n,b)$ was propagated from the left child, the number of dequeues in the left subblocks of \nf{$n$.parent.blocks[$superblockIndex$]} before \nf{$n$.blocks[$b$]} is considered in Line \ref{considerPreviousLeft} (see Figure \ref{fig::blockSumDeqLeft}).
-Otherwise, if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}) (see Figure \ref{fig::blockSumDeqRight}).
-Finally, \nf{IndexDequeue} is called on \nf{$n$.parent} recursively, and it returns the correct response by the induction hypothesis.
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqLeft.png}
+We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
+Assuming the claim holds for $v$'s parent, we prove it for $v$.
+Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
+\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
+By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
+The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a left child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a left child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqLeft}
-\end{figure}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/blockSumDeqRight.png}
+Next, the position of the required dequeue in $D(B')$ is computed in 
+lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
+We first add the number of dequeues in the subblocks of $B'$ in $v$ that preced $B$ on line \ref{computeISuperStart}.
+If $v$ is the right child of its parent, then all of the subblocks of $B'$ from $v$'s left sibling
+also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.
 
-  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child.]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child.
-The index of the superblock is shown with $sb$.
-}\label{fig::blockSumDeqRight}
-\end{figure}
+Finally, \op{IndexDequeue} is called recursively on $v$'s parent.
+Since $B$ has been propagated to the root, so has its superblock $B'$.
+Thus, all preconditions of the recursive call are met.
+By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}
+
+----------
 
@@
 
-\paragraph{Remark} In fact our algorithm is strongly linearizable as defined in \cite{DBLP:conf/stoc/GolabHW11}.
-By Definition \ref{ordering} the linearization ordering of operations will not change as blocks containing new operations are appended to the root.
-
 \here{following material from algorithm description in thesis might be useful in sketch of correctness proof}

### 2023-01-04 podc_paper/garbage-collection.tex
add garbage collection code to algorithm.tex
add garbage collection lemmas

@@
-\documentclass{article}
-\usepackage[utf8]{inputenc}
-\renewcommand{\baselinestretch}{1.8}
-\usepackage[dvipsnames]{xcolor}
-\usepackage[letterpaper, margin=1in]{geometry}
-\usepackage{tikz-qtree}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\makeatletter
-\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
-\makeatother
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{relsize}
-\usepackage{enumitem}
-\usepackage{bold-extra}
-\renewcommand*\contentsname{Table of Contents}
+% !TEX root =  podc-submission.tex
 
-\usepackage{multicol}
-\setlength\columnsep{24pt}
+\section{Garbage Collection}
 
-\algnewcommand\algorithmicforeach{\bf{for each}}
-\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
-
-\algdef{S}[FUNCTION]{Function}
-   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}
-
-\algdef{E}[FUNCTION]{EndFunction}
-   [1]{\algorithmicend\ \tt{{#1}}}
-
-\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}
-
-\newcommand\keywordfont{\sffamily\bfseries}
-\algrenewcommand\algorithmicend{{\keywordfont end}}
-\algrenewcommand\algorithmicfor{{\keywordfont for}}
-\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
-\algrenewcommand\algorithmicdo{{\keywordfont do}}
-\algrenewcommand\algorithmicuntil{{\keywordfont until}}
-\algrenewcommand\algorithmicfunction{{\keywordfont function}}
-\algrenewcommand\algorithmicif{{\keywordfont if}}
-\algrenewcommand\algorithmicthen{{\keywordfont then}}
-\algrenewcommand\algorithmicelse{{\keywordfont else}}
-\algrenewcommand\algorithmicreturn{{\keywordfont return}}
-
-\renewcommand\thealgorithm{}
-\newcommand{\setalglineno}[1]{
-  \setcounter{ALC@line}{\numexpr#1-1}}
-
-\newcommand{\sub}[1]{\textsubscript{#1}}
-\renewcommand{\tt}[1]{\texttt{#1}}
-\renewcommand{\sl}[1]{\textsl{#1}}
-\renewcommand{\it}[1]{\textit{#1}}
-\renewcommand{\sc}[1]{\textsc{#1}}
-\renewcommand{\bf}[1]{\textbf{#1}}
-\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
-\newcommand{\cmt}[1]{\Comment{#1}}
-\newcommand{\head}{head}
-\newcommand{\size}{size }
-
-\usepackage{amsmath,amssymb,amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}[theorem]{Observation}
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}[theorem]{Invariant}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\begin{document}
-
-\begin{algorithm}
-\caption{Tree Fields Description}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-
-\Statex $\diamondsuit$ \tt{\sl{Shared}}
-\begin{itemize}
-\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
-\item {\color{red}\tt{\sl{MaxbyProcess} lastDequeuedFrom} \textsf{Index of the most recent block in the root that has been deqeueued from.}}
-\end{itemize}
-
-\Statex $\diamondsuit$ \tt{\sl{Local}}
-\begin{itemize}
-\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
-% \item {\color{red}\tt{\sl{int} garbageCollectRound}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Node}}
-\begin{itemize}
-\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
-\item {\color{red}\tt{\sl{PBRT} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}}
-\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{Block}}
-
-\begin{itemize}
-  \item \tt{\sl{int} super}
-  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
-\begin{itemize}
-    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
-  \textsf{:~~indices of the last subblock of the block in the left and right child}
-  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
-  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
-  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
-\begin{itemize}
-  \item \tt{\sl{Object} element}
-  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-
-    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
-  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
-
-  \item {\color{red}\tt{\sl{object} response}}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
-\begin{itemize}
-  \item \tt{\sl{int} \size}
-  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
-\end{itemize}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Queue}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
-
-\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
-\State \tt{newBlock.element= e}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\EndFunction{Enqueue}
-
-\Statex
-
-\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
-\Function{Object}{Dequeue()}{} 
-\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
-\State \tt{newBlock.element= null}
-\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
-\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
-\State \tt{leaf.}\Call{Append}{newBlock}
-\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\label{deqRest}
-\State \Return{\tt{output}}
-\EndFunction{Dequeue}
-
-\Statex
-
-\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
-\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
-\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
-\State \tt{ \color{red} lastDequeudFrom.update(b)}
-\State \Return \tt{null} \label{returnNull}
-\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
-\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
-\State {\color{red}\tt{<x, y>= root.\Call{BinarySearch}{e}}}
-\State {\color{red}\tt{lastDequeudFrom.update(x)}}
-\State \Return \tt{root.GetEnqueue(x,y)}\label{findAnswer}
-\EndIf
-\EndFunction{FindResponse}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{25}
-
-\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-\Statex $\triangleright$ \textmd{\color{red} Update needed: search on RBT does not need start and end, we can search over whole the red-black tree.}.
-\Function{int}{BinarySearch}{\sl{int} x}
-\State \Return \tt{\color{red} min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-% \While{\nf{start<end}}
-% \State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
-% \If{\nf{blocks[mid].sum\sub{enq}<x}}
-% \State \nf{start= mid+1}
-% \Else
-% \State \nf{end= mid}
-% \EndIf
-% \EndWhile
-% \State\Return \nf{start}
-\EndFunction{BinarySearch}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Leaf}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{45}
-
-\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
-\State \tt{\color{red} blocks.TryAppend(B, head)} \label{appendLeaf}
-\State \tt{\head= \head+1} \label{appendEnd} 
-\State \tt{parent.}\Call{Propagate()}{} 
-\EndFunction{Append}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
-\Function{void}{Propagate()}{}
-\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
-\State \Call{Refresh()}{} \label{secondRefresh}
-\EndIf
-\If{\tt{this} \bf{is not} \tt{root}}
-\State \tt{parent.}\Call{Propagate()}{}
-\EndIf
-\EndFunction{Propagate}
-
-\Statex
-
-\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
-\Function{boolean}{Refresh()}{}
-\State \tt{h= \head} \label{readHead}
-{\color{red}\If{\tt{h\%p\textsuperscript{2}=0}}
-\State \Call{Help()}{}
-\State \tt{root.}\Call{FreeMemory()}{lastDequeuedFrom.Get()-1}
-\EndIf}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
-\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
-\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
-\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
-\EndIf
-\EndFor \label{endHelpChild1}
-\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
-\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
-\EndIf
-\State{\tt{\color{red} result= blocks.TryAppend(new, h)}} \label{cas}
-\State{\tt{this.\Call{Advance}{h}}} \label{advance}
-\State \Return{ \tt{result}}
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{73}
-
-\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
-\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
-\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
-\State \tt{head.CAS(h, h+1)} \label{incrementHead}
-\EndFunction{Advance}
-
-\Statex
-
-\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
-\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
-\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
-\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
-\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
-\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
-\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
-\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
-\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
-\EndFor
-\If{\tt{this} \bf{is} \tt{root}}
-\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
-\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
-\EndIf
-
-\State \Return \tt{new}
-\EndFunction{CreateBlock}
-
-\Statex
-\color{red}
-\Function{int}{GetLastDequeuedFrom}{}{\cmt{Returns the index that is safe to remove the blocks before that in the node.}}
-\State \tt{x= lastDequeuedFrom.Get()-1}
-\State \tt{n= root}
-\State \bf{while} \tt{n!=this} \bf{do}
-\State \tt{dir= left (if this is in left subtree of n): otherwise dir=right}
-\State \tt{    x=n.blocks[x].end\sub{dir}}
-\State\bf{end while}
-\EndFunction{GetLastDequeuedFrom}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{94}
-
-\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
-\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
-\If{\tt{this} \bf{is} \tt{leaf}}
-\State\Return \tt{blocks[b].element} \label{getBaseCase}
-\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
-\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,}  \label{leftChildGet}
-\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} \cmt{{\color{red} start and end values are not needed anymore?}}
-\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
-\Else
-\State \tt{i= i-blocks[b].num\sub{enq-left}}
-\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
-\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} \cmt{{\color{red} start and end values are not needed anymore?}}
-\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
-\EndIf
-\EndFunction{GetEnqueue}
-
-\Statex
-\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
-\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{{\color{red} Update needed: return null when superblock in the root was not found.}}
-\If{\tt{this} \bf{is} \tt{root}}
-\State\Return \tt{<b, i>} \label{indexBaseCase}
-\Else
-\State \tt{dir= (parent.left==n ? left: right)} 
-\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper} 
-\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1} \cmt{{\color{red} Preconditions might be not met.}}
-
-\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
-\Else \label{considerRight}
-\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
-\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
-\EndIf \label{computeISuperEnd}
-\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
-\EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{algorithm}
-
-\color{red}
-
-\begin{algorithm}
-\color{red}
-\caption{\tt{\sl{\color{red}MaxByProcess}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{121}
-
-\State \tt{\sl{int[p]} lastDequeuedbyProcess}
-\Statex
-
-\Function{int}{Get}{}
-\State \Return \tt{max(lastDequeuedbyProcess)}
-\EndFunction{Get}
-
-\Statex
-\Function{}{Update}{\sl{int} \tt{b}}
-\If{\tt{lastDequeuedbyProcess[pid]<b}}
-\State \tt{lastDequeuedbyProcess[pid]=b}
-\EndIf
-\EndFunction{Update}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\color{red}
-\caption{\tt{\sl{Tree}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{130}
-
-\Function{int}{Help}{}
-\State \bf{for each} \sl{process} \tt{P}
-\State \tt{h=P.leaf.head}
-\If{\tt{P.leaf.blocks[h].num\sub{deq}==1 and P.leaf.IndexDequeue(h,1)!=null}}
-\State \tt{<b, i>=} \Call{IndexDequeue}{h, 1}
-\State \tt{output=} \Call{FindResponse}{b, i} 
-\State \tt{P.leaf.blocks[h].response= output}
-\EndIf
-\State \bf{end for}
-\EndFunction{Help}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\color{red}
-\caption{\tt{\sl{Node}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{140}
-
-\Function{}{FreeMemory}{\sl{int} \tt{b}}
-\If{not leaf}
-\State \tt{left.FreeMemory(blocks[i].end\sub{left}-1)}
-\State \tt{right.FreeMemory(blocks[i].end\sub{right}-1)}
-\EndIf
-\State \tt{blocks= blocks.splitGreater(i)} \cmt{I think CAS is not needed.}
-\EndFunction{FreeMemory}
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}
-\color{red}
-\caption{\tt{\sl{PBRT}}}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{140}
-
-\Statex \tt{\sl{PRBT} prbt}
-\Statex \sf{nodes store \tt{<key, sum\sub{enq}-> block}}
-\Statex \tt{[i] -> GetByBlock(i)}
-
-\Statex
-\Function{}{GetByBlock}{\sl{int} \tt{i}}
-\State \Return \tt{rbt.get(i)}
-\If{not found}
-\State \tt{return written response}
-\EndIf
-\EndFunction{GetByBlock}
-
-\end{algorithmic}
-\end{algorithm}
-
-\pagebreak
-\clearpage
-\section{Description}
-
-\color{black}
 In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
@@
 
-\begin{lemma}
-Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
-\end{lemma}
+%\begin{lemma}
+%Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
+%\end{lemma}
+%
+%\begin{lemma}
+%Total number of the blocks in the tree is $O(p^3+pq)$.
+%\end{lemma}
 
-\begin{lemma}
-Total number of the blocks in the tree is $O(p^3+pq)$.
-\end{lemma}
-
-\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$ or $e.$\nf{element} has been computed to be $RESP(d)$.
-It is safe to remove \nf{Dequeue} $d$o from the tree after the $d$ is terminated or $Resp(d)$\nf{.element} has been computed.
-We can remove a block after the told conditions are satisfied for all ofits operations.
-A \nf{Dequeue} may go to sleep for a long time and prevent the \nf{block} to be removed.
+\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$.
+It is safe to remove \nf{Dequeue} $d$ from the tree after the $d$ terminates.
+We can remove a block after the conditions told are satisfied for all of its \nf{Enqueue}s and \nf{Dequeue}s.
+A \nf{Dequeue} may go to sleep for a long time and prevent a \nf{block} to be removed (the block it is in or the block its response is in).
 In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere.
-After writing down the response of th \nf{Dequeue} it is safe to remove.
-Because if the \nf{Dequeue} failed to compute its response it can read the response written (computing the \nf{Dequeue} index in the \nf{root} or getting the response \nf{Enqueue}).
+After writing down the response of th \nf{Dequeue} can be removed, since if the process wanted the response it can read from the helped response.
 
@@
 \begin{lemma}
-It is safe to remove a finished block from the tree.
+It is safe to remove a finished block from the non-leaf nodes.
 \end{lemma}
@@
 This gives us the idea that if a block is finished then all the blocks before it are also finished.
-If an operation in a block goes to sleep for a long time then other processes help the operation so the block get finished.
-Note that when an \nf{Enqueue} operation in a block is not finished the block cannot be finished until some \nf{Dequeue} dequeues that \nf{Enqueue}.
-Since there are at most $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely.
+If an operation in a block goes to sleep for a long time then other processes help the operation so the block gets finished.
+There are less than $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely in an amortized poly-log time.
 
 \begin{lemma}
-    If all current operations are helped, then there is a block in the root that all of its previous blocks are finished. That block is the most recent block that has been dequeued from.
+    If all \nf{Dequeue} operations in the root are helped, then all the blocks before a finished block in the root are also finished.
+\end{lemma}
+
+\begin{lemma}
+  Blocks before the most recent block that has been dequeued(computed to be dequeue) from are finished.
+If the most current \nf{Dequeue} returned(computed) \nf{null} then all the blocks before the block containing the \nf{Dequeue} are finished.
 \end{lemma}
@@
 
-\begin{lemma}
-    If all current operations are helped, then the blocks before the newest block that some \nf{Enqueue} has been dequeued from is safe to remove.
-If the most current \nf{Dequeue} returned \nf{null} then all the blocks before the block containing the \nf{Dequeue} can be removed.
-\end{lemma}
+\begin{observation}
+Every $p^2$ block appended to the root, \nf{FreeMemory} is invoked.  
+\end{observation}
 
-There is a shared array among processes which they write the last block dequeued from in it.
+To know the last block dequeied from we can implement a shared array among processes which they write the last root block they have dequeued from.
 
 \begin{lemma}
-    $\nf{GetLastDequeuedFrom($n$} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
+    $\nf{Max(LastDequeuedFrom)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
 \end{lemma}
@@
 \begin{lemma}
-    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$
+    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p+q)$.
 \end{lemma}
 
+\begin{corollary}
+The space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$.
+\end{corollary}
+
 \begin{lemma}
-  The amortized step per process for \nf{GarbageCollect} is $O()$.
+  The amortized step per process for the algorithm with garbage collection is $PolyLog(p+q)$.
 \end{lemma}
@@
 \end{lemma}
-
-\end{document}

### 2023-01-04 podc_paper/correctness.tex
finished initial revision of linearizability proof

@@
 \section{Proof of Correctness}
-
-Consider any finite execution.  We must show that the linearization ordering $L$ defined in Section \ref{linearization} includes all operations that terminate and that 
-(1) if an operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ cannot appear
-before $op_1$ in $L$, and (2) each dequeue that terminates returns the same response as it would 
-in $L$.
 
@@
 If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
-from Lemma \ref{successfulRefresh}.
+from \Cref{successfulRefresh}.
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
@@
 The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
-So, if $s\leq 2$, the claim is trivial.  Assume for the remainder that $s>2$.
+So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
 By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem::indexDequeue}
 If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
@@
 
-----------
+\subsection{Linearizability}
 
-\subsection{Linearizability}
-We now prove the two properties needed for linearizability.
+Consider any finite execution.  We must show that the linearization ordering $L$ defined in 
+(\ref{linearization}) is a legal permutation of a subset of the operations in 
+the execution, i.e., that it includes all operations that terminate and 
+if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does appear
+before $op_1$ in $L$.  Then we must show that each dequeue that terminates returns the 
+same response as it would in the sequential execution described by $L$.
 
@@
 \begin{proof}
-We must show for any execution that every operation that terminates is in $L$ exactly once.
-Also, if $op_{1}$ terminates before $op_{2}$ in starts in the execution, then $op_{1}$ is before $op_{2}$ in the linearization.
-The first claim is directly reasoned from Corollary \ref{lem::appendExactlyOnce}.
-For the latter, if $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append} has terminated before $op_{2}$\nf{.Append} started.
-From Corollary \ref{lem::append}, $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate.
-By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
+By \Cref{lem::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
+Since each operation creates a block $B$ containing the operation and then calls \op{Append}($B$), 
+the operation is propagated to the root before the operation terminates, by \Cref{lem::appendExactlyOnce},
+so it appears in $L$.
+Also, if $op_{1}$ terminates before $op_{2}$ begins, then $op_{1}$ 
+is propagated to the root before $op_2$ begins, so $op_2$ appears before $op_2$ in $L$.
 \end{proof}
-Once some operations are aggregated in one block, they will get propagated up to the root together, and they can be linearized in any order among themselves.
-We have chosen to put \nf{Enqueue}s in a block before \nf{Dequeue}s (see Definition \ref{ordering}).
 
-\begin{definition}\label{defNullDeq}
-If a \nf{Dequeue} operation returns \nf{null} it is called a \it{null} \nf{Dequeue}, otherwise it is called \it{non-null}~\nf{Dequeue}.
-\end{definition}
-
-Next, we define the responses that \nf{Dequeue}s should return, according to the linearization.
-\begin{definition}
-   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.
-${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$.
-Otherwise if ${d}$ is a null \nf{Dequeue} then ${Resp(d)=\nf{null}}$.
-\end{definition}
-
-In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
+In the next lemma, we show that the \fld{size} field in each block in the root is computed correctly.
 \begin{lemma}\label{sizeCorrectness}
-  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
+If the operations of $\var{root}.\fld{blocks}[0..b]$ are applied sequentially in the order of~$L$ on an initially empty queue, the resulting queue has $\var{root}.\fld{blocks}[b].\size$ elements.  
 \end{lemma}
@@
 We prove the claim by induction on $b$. 
-The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$.
-We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$.
-By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$.
-By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it.
-If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}.
-Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$.
-In both cases, this is the same as the assignment on Line \ref{computeLength}.
+The base case when ${b=0}$ is trivial, since the queue is initially empty and 
+$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
+Assuming the claim holds for $b-1$, we prove it for $b$.
+The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
+at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
+By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
+of block $B$ are performed.
+By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
+are the number of enqueues and dequeues contained in $B$.
+Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
+is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
 \end{proof}
 
-The next lemma is useful to compute the number of non-null dequeues.
-\begin{lemma} \label{numberOfNND}
-If operations in the root are applied in the order of $L$, the number of non-null \nf{Dequeue}s in $\nf{root.blocks[}0\cdots b\nf{]}$ is \nf{root.blocks[$b$].sum\sub{enq} $-$ root.blocks[$b$].size}.
+\begin{lemma}\label{linearCorrect}
+Each terminating dequeue returns the response it would in the sequential execution $L$.
 \end{lemma}
 \begin{proof}
-There are \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} \nf{Enqueue} operations in $\nf{root.blocks[}0\cdots b\nf{]}$ by Corollary~\ref{lem::sumX}.
-The size of the queue after doing $\nf{root.blocks[}0\cdots b\nf{]}$ in the order of $L$ is
-the number of $\it{enqueues}$ in $\nf{root.blocks[}0\cdots b\nf{]}$ minus the number of $\it{non-null \nf{Dequeue}s}$ in $\nf{root.blocks[}0\cdots b\nf{]}$.
-By the correctness of the \nf{size} field from Lemma \ref{sizeCorrectness} and \nf{sum\sub{enq}} field from Lemma \ref{lem::numX}, the number of $\it{non-null \nf{Dequeue}s}$ is \nf{root.blocks[$b$]}\nf{.sum\sub{enq}} $-$ \nf{root.blocks[$b$]}\nf{.size}.
+If a dequeue $D$ terminates, it is contained in some block in the root, by \Cref{lem::appendExactlyOnce}.
+By \Cref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
+returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
+$B=\var{root}.\fld{blocks}[b]$.
+$D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
+By \Cref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
+after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
+in the order given by $L$.
+By \Cref{lem::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
+is the number of enqueues in $B$.
+Since the enqueues in block $B$ precede the dequeues,
+the queue is empty when the $i$th dequeue of $B$ occurs if 
+$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
+So $D$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
+execution $L$.
+Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
+in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
+the number of non-\nl\ dequeues in that prefix of $L$.
+Hence, line \ref{computeE} sets $e$ to the rank of $D$ among all the non-\nl\ dequeues in $L$.
+Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
+By \Cref{lem::sum}, this enqueue is the ($i'$)the enqueue 
+in $E(\var{root}.\fld{blocks}[b'])$, where
+$b'$ and $i'$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
+By \Cref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
 \end{proof}
 
-\begin{corollary}\label{numNullDeqBlock}
-  If operations in the root are applied in the order of $L$, the number of non-null dequeues in $\nf{root.blocks[}b\nf{]}$ is \nf{root.blocks[$b$].num\sub{enq} $-$ root.blocks[$b$].\size $+$ root.blocks[$b-1$].size}.
-\end{corollary}
+Combining \Cref{linearSat} and \Cref{linearCorrect} gives us our main result.
 
-\begin{lemma}\label{nullReturn}
-$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
-\end{lemma}
-\begin{proof}
-The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
-\end{proof}
-
-\begin{lemma}\label{computeHead}
-\nf{FindResponse($b$, $i$)} returns $Resp(D_i(root,b))$.
-\end{lemma}
-\begin{proof}
-$D_i(root,b)$ is $D_{\nf{root.blocks[}b-1\nf{].sum\sub{deq}}+i}(root)$ by Definition \ref{ordering} and Lemma \ref{lem::sumX}.
-$D_i(root,b)$ returns \nf{null} at Line \ref{returnNull} if $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- i <0$ and $Resp(D_i(root,b))=\nf{null}$ in this case by Lemma \ref{nullReturn}.
-Otherwise, if $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} in $L$ it should return the $e$th enqueued value.
-By Lemma \ref{numberOfNND} there are \nf{root.blocks[$b-1$].sum\sub{enq} $-$ root.blocks[$b-1$].\size} non-null \nf{Dequeue} operations in $\nf{root.blocks[}0\cdots b-1\nf{]}$.
-The \nf{Dequeue}s in \nf{root.blocks[$b$]} before $D_i(root,b)$ are non-null \nf{Dequeue}s.
-So $D_i(root,b)$ is the $e$th non-null \nf{Dequeue} where $e= i + \nf{root.blocks[}b-1\nf{].sum\sub{deq}} - \nf{root.blocks[}b-1\nf{].size}$ (Line \ref{computeE}).
-See Figure \ref{computeResponseDetail}.
-
-After computing $e$ at Line \ref{computeE}, the code finds \nf{$b$,$i$} such that $E_i(root,b)=E_e(root)$ using \nf{DoublingSearch} and then finds its \nf{element} using \nf{GetEnqueue} (Line \ref{findAnswer}).
-Correctness of \nf{DoublingS-}\\\nf{earch} and \nf{GetEnqueue} routines are shown in Lemmas \ref{dsearch} and \ref{get}.
-\end{proof}
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=4in, height=2.5in]{pics/computeResponseDetail.png}\caption{The position of $D_i(root,b)$.}
-\label{computeResponseDetail}
-\end{figure}
-\vspace{-1em}
-
-\begin{lemma}\label{linearCorrect}
-The responses to operations in our algorithm are the same as in the sequential execution in the order given by $L$.
-\end{lemma}
-\begin{proof}
-\nf{Enqueue} operations do not return any value. By Lemma \ref{computeHead}, the response of a \nf{Dequeue} in our algorithm is the same as its response in the sequential execution of $L$.  
-\end{proof}
-
-\begin{theorem}[Main]
+\begin{theorem}
 The queue implementation is linearizable.
 \end{theorem}
-\begin{proof}
-The theorem follows from Lemmas \ref{linearSat} and \ref{linearCorrect}.
-\end{proof}
-
-\here{following material from algorithm description in thesis might be useful in sketch of correctness proof}
-\begin{figure}[hbpt]
-  \center\includegraphics[width=4in]{pics/doublyrefresh-drawio.png}
-  \caption[Two consecutive failed \nf{Refresh}es by a
-    process.]{\label{fig::simpleDoubleRefresh}Time relations between
-    the concurrent successful \nf{Refresh}es and the two consecutive
-    failed \nf{Refresh}es.} 
-\end{figure}
-We use \nf{CAS} (Compare\&Swap) instructions to implement the
-\nf{Refresh}'s attempt to append  described in the previous
-paragraph. 
-The second failed \nf{Refresh} of $P$ is assuredly concurrent with a
-successful \texttt{Refresh} that has read its information after the
-invocation of the first failed \nf{Refresh} (see Figure
-\ref{fig::simpleDoubleRefresh}). This is because some process $L$ does
-a successful append during $P$'s first failed attempt, and some
-process $K$ performs a \nf{Refresh} that reads its information after
-$L$'s append and then performs a successful append during $P$'s second
-failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the
-new operations that were in $n$'s children before $P$'s first failed
-\nf{Refresh}, in case they were not already appended. After a process
-appends its operation into its leaf it can call \nf{Refresh} on the
-path up to the root at most two times on each node. So, with $O(\log
-p)$ \nf{CAS}es an operation can ensure it appears in the
-linearization. This cooperative solution allows us to overcome the CAS
-Retry Problem.

### 2023-01-04 podc_paper/algorithm.tex
finished initial revision of linearizability proof
working on analysis section

@@
 L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
+\label{linearization}
 \end{equation}
-\label{linearization}
 
@@
     \State \Call{Append}{\var{n}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}
-    \State \Return{ \Call{FindResponse}{\var{b, i}}}
-    \label{deqRest}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}\label{invokeIndexDequeue}
+    \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
 \EndFunction{Dequeue}
@@
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
@@
 		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $\var{b}'\leq \var{b}$ with $\var{root}.\fld{blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{e}$
+		\State find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
 		\State \linecomment{find rank of enqueue within its block}
-		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
-        \State \Return \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\label{FRi}
+        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
     \EndIf
@@
 \nf{Dequeue}. 
-A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b'$ of the block that contains 
+A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b_e$ of the block that contains 
 the \var{e}th enqueue.
-The index $b'$ is at most $b$ (since the enqueue is linearized before the dequeue).
-To find the left end of the range for the binary search for $b'$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
+The index $b_e$ is at most $b$ (since the enqueue is linearized before the dequeue).
+To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
 Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

### 2023-01-04 podc_paper/analysis.tex
working on analysis section

@@
 \section{Analysis}
-In this section, we analyze the number of \nf{CAS} invocations and the step complexity of our algorithm.
+In this section, we analyze the number of steps and the number of \op{CAS} instructions performed by operations.
+First, we prove some claims about the size and operations of a block.
 
-First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.
+\begin{proposition}
+Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
+\end{proposition}
+\begin{proof}
+An operation invokes \nf{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
+A \nf{Refresh} does at most 7 \nf{CAS} steps: one in line \ref{cas} and two from each \nf{Advance} in line \ref{helpAdvance} or~\ref{advance}.
+\end{proof}
 
+\begin{lemma}\label{dsearchTime}
+The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
+\end{lemma}
+\begin{proof}
+Let the blocks in the root be $B_1, \ldots, B_\ell$.
+The doubling search used to compute $b_e$ takes $O(\log (b-b_e))$ steps,
+so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
+If $b \leq b_e+1$, then this is trivial, so assume for the rest of the proof that $b>b_e+1$.
+
+As shown in \Cref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
+Thus, there can be at most $B_{b_e}.\size$ dequeues in 
+$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
+all elements enqueued before the end of
+$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
+Furthermore, by \Cref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
+$B_1,\ldots,B_{b-1}$  is 
+$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
+Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
+So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
+$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
+Each of these $b-1-b_e$ blocks contains at least one operation, by \Cref{blockNotEmpty}.
+So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$
+\end{proof}
+
+The following lemma is useful in bounding the time to \op{GetEnqueue}
 \begin{lemma}\label{blockSize}
@@
 By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
-are invoked before $t$.  Thus, they are all concurrently running at $t$, so there are at most
-$c$ operations contained in $B$.
+are invoked before $t$.  Thus, all are  running concurrently at $t$, so $B$ contains at most $c$ operations.
 By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
@@
 By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
-so there are at most $c$ direct subblocks.
+$B$ has at most $c$ direct subblocks.
 \end{proof}
 
-\begin{proposition}
-An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
-\end{proposition}
-\begin{proof}
-  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
-\end{proof}
-
-\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
-If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
-\end{lemma}
-\begin{proof}
-First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$.
-There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}.
-Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}.
-The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}.
-After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution.
-We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s.
-Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}.
-So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.
-
-Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps.
-See Figure \ref{fig::doubling}.
-After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$.
-Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
-\end{proof}
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6in]{pics/doubling.png}
-  \caption{Distance relations between \nf{start}$,b,end$.}
-  \label{fig::doubling}
-\end{figure}
+--------
 

### 2023-01-04 podc_paper/correctness.tex
working on analysis section

@@
 
-\subsection{Correctness of GetEnqueue}
+\subsection{Correctness of Helping Routines}
+
+We first show that \op{getEnqueue} works correctly.
 
@@
 \end{proof}
-
-\subsection{Correctness of IndexDequeue}
 
@@
 If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
- then \op{IndexDequeue}($v, b, i$) returns $\langle b',i'\rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
+ then \op{IndexDequeue}($v, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
 \end{lemma}
@@
 
+Next, we show operations return the same response as they would in the sequential execution $L$.
+
 \begin{lemma}\label{linearCorrect}
@@
 Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
-By \Cref{lem::sum}, this enqueue is the ($i'$)the enqueue 
-in $E(\var{root}.\fld{blocks}[b'])$, where
-$b'$ and $i'$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
+By \Cref{lem::sum}, this enqueue is the $i_e$th enqueue 
+in $E(\var{root}.\fld{blocks}[b_e])$, where
+$b_e$ and $i_e$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
 By \Cref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
@@
 
-Combining \Cref{linearSat} and \Cref{linearCorrect} gives us our main result.
+Combining \Cref{linearSat} and \Cref{linearCorrect} provides our main result.
 

### 2023-01-05 podc_paper/algorithm-2.tex
add algorithm2

--- 2023-01-03 podc_paper/algorithm.tex
@@
 
-\section{Queue Implementation} \label{DescriptQ}
-
-\subsection{Overview}
-We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
-The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
-assigned to each process. 
-Each tree node  stores an array of \emph{blocks}, where each block represents a 
-sequence of enqueues and a sequence of dequeues.
-See Figure \ref{tournament} for an example.
-In this section, we use an infinite array of blocks in each node.
-Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.
-
-To perform an operation on the queue, a process $P$ appends a new block containing that  
-operation to the blocks array in $P$'s leaf.
-Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
-We shall define a total order on all operations that have been propagated to the root, which 
-will serve as the linearization ordering of the operations.
-
-To propagate an operation to a node $v$ in the tree, $P$ first observes
-the blocks in both of $v$'s children that are not already in $v$,
-creates a new block by combining information from those blocks, and attempts to append this 
-new block to $v$'s blocks array using a CAS instruction.
-Following \cite{}, we call this a 3-step sequence a
-\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
-A successful \op{Refresh} by $P$ may propagate pending operations by several other processes to $v$.
-This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
-new block into $v$'s array twice,
-then its operation has been propagated to $v$ by some other process, so $P$ can continue 
-onwards towards the root.
-
-Now suppose $P$'s operation has been propagated all the way to the root.
-If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
-If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
-dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
-(since the operation may have been propagated to the root by some other process).
-Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
-If so, it returns \nl\ and we call it a \emph{null dequeue}.
-If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
-among all non-null dequeues,
-finds the $r$th enqueue in the linearization, and returns that enqueue's value.
-
-The primary challenge is thus figuring out what information to store in each block so that 
-the following tasks can be done efficiently (in a polylogarithmic number of steps).
-\begin{enumerate}[label={(T\arabic*)}]
-\item
-\label{construct}
-Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{Refresh}.
-\item
-\label{findinroot}
-Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
-\item
-\label{findrank}
-Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
-or determine the rank $r$ of the enqueue whose value it should return.
-\item
-\label{findenqueue}
-Find the $r$th enqueue in the linearization ordering.
-\end{enumerate}
-Since these tasks depends on the linearization ordering, we describe that ordering next.
-
-\begin{figure}[tb]
-\input{tournamentTree.pdf_t}
-\caption{An example tournament tree with four processes. 
-We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
-Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
-The fourth process's Deq\sub{6} is still in progress.
-The linearization order for this tree is
-Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
-\end{figure}
-
-\begin{figure}
-\input{implicit.pdf_t}
-\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
-The leaf blocks simply show the \fld{element} field.
-Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
-and \eleft\ and \eright\ fields are shown using arrows as before.
-Root blocks also have the additional \fld{size} field.
-The \fld{super} field is not shown.}
-\end{figure}
-
-\subsection{Linearization Ordering}
-
-An operation can terminate only after a block containing it has been appended to the root's blocks array.
-So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
-$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
-Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
-Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.
-
-Each block in a leaf represents a single operation.
-Each block $B$ in an internal node $v$ results from merging
-several consecutive blocks from each of $v$'s children.
-The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
-A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
-or a subblock of a direct subblock of $B$.
-A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
-
-The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
-so there is at most one operation per process.
-Hence, a block represents at most $p$ operations in total.  
-Moreover, we never append empty blocks, so 
-each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.
-
-As mentioned above, we are free to order operations within a block however we like.
-For convenience, we order the enqueues and dequeues separately, and put the 
-operations that were propagated from the left child before the operations from the right child.
-More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
-represented by a block $B$.
-If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
-and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
-If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
-and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
-\begin{eqnarray}
-E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
-D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
-\end{eqnarray}
-We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
-
-Finally, when designing a total order for the operations propagated to the root, we choose
-to put each block's enqueues before its dequeues.
-Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
-linearization ordering is 
-\begin{equation}
-L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
-\end{equation}
-\label{linearization}
-
-\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
-\label{sec:fields}
-
-Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
-To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
-We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
-Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
-for the next attempt to append a \block\ to this \node.
-
-If a block contained an explicit representation of its sequences of enqueues and dequeues,
-it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
-Instead, the block stores an implicit representation of the sequences.
-We now explain how we designed the fields for this implicit representation. 
-Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.
-
-The information stored in a \block\ depends on whether it is in an internal node or a leaf.
-A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
-enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.
-
-\here{If space permits, we might want to add some examples in the following paragraphs that refer back to Figure \ref{implicit}.}
-
-Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
-Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
-\begin{eqnarray}\label{defsubblock}
-v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
-v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
-\end{eqnarray}
-This allows us to navigate to the direct subblocks of any block easily.
-Blocks also store prefix sums of the numbers of enqueues and dequeues:
-the block in $v.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
-These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
-For example, consider finding the $r$th enqueue in the linearization ordering.
-We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
-to locate which block in the root contains the operation.
-If we know which block $B$ in some \node\ $v$ contains the enqueue,
-we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
-and then to do a binary search
-among the direct subblocks of $B$ in that child.
-Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
-the enqueue.
-%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
-%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
-%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
-%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
-%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
-%to determine whether the enqueue came from the left or right child of the root.
-%Suppose the enqueue came from the right child $v_r$.
-%Then, we know that the index of the block in $v_r$ that contains the enqueue
-%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
-%We can again do a binary search within this range.
-%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
-%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
-We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
-and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
-for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
-All of the information needed for this search process can be derived from the 
-\eleft, \eright\ and \fld{sum\sub{enq}} fields.
-
-A block is called the \emph{superblock} of all of its direct subblocks.
-To facilitate task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
-the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
-We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
-This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
-Thus, starting from an operation in a leaf's block, one can use these indices to track the 
-operation all the way up the path to the root, and determine the operation's location in a root block
-in $O(\log p)$ time.
-
-We now consider task \ref{findrank}.
-First, we must determine whether the queue is empty when a dequeue occurs.
-To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
-in the queue after all operations in the linearization ordering up to that block (inclusive) 
-have been performed.
-We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
-$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
-We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
-as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
-Then, for any non-null dequeue in $B_d$, we can use this information to determine its
-rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
-whose value it should return (among all enqueues).
-
-Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
-we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
-A \op{Refresh} on \node\ $v$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $v$'s children and stores 
-$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
-Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
-For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
-the number of enqueues and dequeues in $B$:
-$B.\fld{size} = \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}) - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}))$.
-
-The only remaining field is $B.\fld{super}$.  When the block 
-$B$ is created for a \node\ $v$, we do not yet know where its
-superblock will eventually be installed in $B$'s parent.
-So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
-in $v.\fld{blocks}[\var{i}]$, processes cooperate to fill in $B.\fld{super}$ 
-when they attempt to advance $v.\fld{head}$ from $i$ to $i+1$,
-using a value they read from the \fld{head} field of $v$'s parent.
-As mentioned above, this might not be exactly the right index for $B$'s superblock, but we
-shall prove that it is close.
+\section{Garbage Collection}
 
@@
     \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}\label{enqNew}
+		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
+		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
     \State \Call{Append}{\var{n}}
@@
     \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
+	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
+	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
     \State \Call{Append}{\var{n}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
-    \State \var{leaf.}\head\ \assign\ \var{leaf.}\head+1 \label{appendEnd} 
-    \State \Call{Propagate}{\var{leaf.}\fld{parent}} 
+    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
+    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryAppend}{\var{B}, \var{leaf}.\head}\label{appendLeafGC}}
+    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
+    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
 \EndFunction{Append}
@@
     \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
-    \If{$\var{v}\neq \var{root}$}
     \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
     \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
-	\EndIf
     \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
+\setcounter{ALG@line}{48}
 
@@
     \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
-    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
+    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
 	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
 	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
@@
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmpty}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
         \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
@@
     \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
-    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
-    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
+    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \If{$\var{v} = \var{root}$}
+        \State\Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
     \Else
@@
     	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$}
+	        \State{\var{sup} \assign\ $\var{sup}+1$}
 	    \EndIf
@@
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
-    %  \State \linecomment{contains at least \var{i} dequeues}
+    \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{102}
+\setcounter{ALG@line}{101}
 
 \Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
+    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
 
@@
             \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
-        \EndIf \label{endChooseDir}
-        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
+        \EndIf
+        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
 $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
-        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
-        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
+        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
+        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} 
     \EndIf
 \EndFunction{GetEnqueue}
+
+\gc{
+\Function{void}{Help}{}
+\State \bf{for} \var{i} \assign\ 1 to $p$
+\State h=leafOfProcess[\var{i}]
+\If{leafOfProcess[\var{i}].blocks[h].num\sub{deq}==1 and leafOfProcess[\var{i}].IndexDequeue(h,1)!=null}
+\State <b, i> \assign\ \Call{IndexDequeue}{leafOfProcess[\var{i}], h, 1}
+\State output \assign\ \Call{FindResponse}{b, i} 
+\State leafOfProcess[\var{i}].blocks[h].response \assign\ output
+\EndIf
+\State \bf{end for}
+\EndFunction{Help}
+
+\Function{}{FreeMemory}{\typ{Node} \var{n}, \typ{int} \var{i}}
+\State Split \var{n}.blocks.split over \var{i} and assign the grater tree to \var{n}.blocks with \tt{CAS}.
+\If{\var{n} is not a leaf}
+\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{left}}
+\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{right}}
+\EndIf
+\EndFunction{FreeMemory}
+
+}
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued.}
-%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
+\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
 \end{figure}
 
-An \opemph{Enqueue}(\var{e}) appends a \block\ to the process's leaf.
-The block has $\fld{element}=\var{e}$ to indicate it represents an \op{Enqueue}(\var{e}) operation.
-It suffices to propagate the operation to the root and
-then use its position in the linearization for future \nf{Dequeue}
-operations.
+In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
+This makes the space used by the algorithm factor of the number of operations of invoked on the queue.
+Here in this section, we want to add a mechanism to free the memory allocated by the operations that are no longer needed and make the overall space used polynomial of $p+q$.
 
-A \opemph{Dequeue} also appends a \block\ to the process's leaf.
-The block has $\fld{element}=\nl$ to indicate that it represents a \op{Dequeue} operation.
-After propagating the operation to the root, it computes
-its position in the root using
-\nf{IndexDequeue} and then computes its response by calling \nf{FindResponse}.
+It is a common way to do garbage collection in batches.
+If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.
 
-\opemph{Append}(\var{B}) first adds the block \var{B} to the invoking process's leaf.
-The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
-so the \op{Append} writes \var{B} there and increments \fld{head}.
-Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
-\op{Append} then calls \op{Propagate} to ensure the operation represented by \var{B} is propagated to the root.
+In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}).
+Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).
 
-\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
-It uses the double refresh idea described
-above and invokes two \op{Refresh}es on \var{v} in Lines
-\ref{firstRefresh} and \ref{secondRefresh}. 
-If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
-that propagated blocks from \var{v}'s children to \var{v}.
-Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.
+%\begin{lemma}
+%Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
+%\end{lemma}
+%
+%\begin{lemma}
+%Total number of the blocks in the tree is $O(p^3+pq)$.
+%\end{lemma}
 
-%\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
-in \var{v}'s
-children and append it to \var{v}.\fld{blocks}. 
-Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
-and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
-If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v},
-so \op{Refresh} terminates and returns true at line \ref{addOP}.
-Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
-Either this CAS succeeds or some other process has installed a new block in this location.
-Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
-from \var{h} to $\var{h}+1$
-and to fill in the \fld{super} field of the most recently appended block.
-The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
-Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
-other processes help keep \fld{head} up to date by also calling \op{Advance}, 
-either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.
+\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$.
+It is safe to remove \nf{Dequeue} $d$ from the tree after the $d$ terminates.
+We can remove a block after the conditions told are satisfied for all of its \nf{Enqueue}s and \nf{Dequeue}s.
+A \nf{Dequeue} may go to sleep for a long time and prevent a \nf{block} to be removed (the block it is in or the block its response is in).
+In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere.
+After writing down the response of th \nf{Dequeue} can be removed, since if the process wanted the response it can read from the helped response.
 
-\opemph{CreateBlock}(\var{v, i}) is used
-by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
-The block \nf{new} is created in Line \ref{initNewBlock}. 
-The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
-children, which are obtained by reading the \fld{head} index in \var{v}'s children.
-Recall that the \fld{sum\sub{enq}} field should store the number of enqueues in
-\var{v}.\fld{blocks}[1..\var{i}].
-Since these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright],
-the total number of these enqueues is $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
-Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
-subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues that are already in \var{v}.\fld{blocks}[$1..\var{i}-1$].
-The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
-Then, if \var{new}
-is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
-represents the number of elements in the queue after the operations in the block are performed.
-Finally, if the newly created block contains no operations, \op{CreateBlock} returns \nl\ to indicate
-that there is no need to append it to \var{v}.\fld{blocks}.
+\begin{definition}
+A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
+\end{definition}
 
-Once a \op{Dequeue} has been appended to a block of the process's leaf and propagated to the root,
-the \opemph{IndexDequeue} routine is used to find the \op{Dequeue}'s location in the root.
-More precisely, \opa{IndexDequeue}{v, b, i}
-computes the block in the root and the rank
-within that block  of the \var{i}th \op{Dequeue} of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
-Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
-whose CAS wrote a block \var{B} into \var{v}.\fld{blocks}[\var{b}].
-Let $par$ be \var{v}.\fld{parent}. Without loss of generality, assume 
-\var{v} is the left child of $par$. Let
-$R_{par}$ be the first successful \op{Refresh} on $par$ whose \op{CreateBlock} reads a
-value greater than \var{b} from $\var{par}.\fld{left.head}=\var{v}.\fld{head}$ at line \ref{createEndLeft} and therefore creates a block $B'$ that contains
-\var{B} as a subblock. 
-%Let $b'$ be the index of $B'$ in $par$.\fld{blocks}.
+\begin{corollary}
+If a block is finished, then all of its subblocks are also finished.  
+\end{corollary}
 
-Since the index of \var{B}'s superblock is not known until $B$ is
-propagated to $par$, $R_v$ cannot set the \nf{super} field of ${B}$ when
-it creates \var{B}. 
-It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (because
-there could be up to $p$ direct subblocks).
-Intuitively, the index of \var{B}'s superblock will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
-Since we can search for the precise superblock when we need it, it suffices that \var{B}.\fld{super}
-is close to the actual index of \var{B}'s superblock.
-Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
-As mentioned earlier, other processes may help fill in this field by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that it is filled in soon after \var{B} is installed.
-We shall
-show that this ensures that \nf{$B$.super} differs
-from the index of $B'$ by at most~1.
+\begin{lemma}
+It is safe to remove a finished block from the non-leaf nodes.
+\end{lemma}
 
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block
-of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
-If the queue is
-not empty, Line \ref{computeE} computes the rank \var{e} of the
-\nf{Enqueue} whose argument is the response to the
-\nf{Dequeue}. 
-A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b'$ of the block that contains 
-the \var{e}th enqueue.
-The index $b'$ is at most $b$ (since the enqueue is linearized before the dequeue).
-To find the left end of the range for the binary search for $b'$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
-Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.
+If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued.
+This gives us the idea that if a block is finished then all the blocks before it are also finished.
+If an operation in a block goes to sleep for a long time then other processes help the operation so the block gets finished.
+There are less than $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely in an amortized poly-log time.
 
-\opemph{GetEnqueue}(\var{v,b,i}) returns the argument of the
-$i$th enqueue in the $b$th block of \typ{Node} $v$. 
-The range of subblocks of a block is determined using the \fld{end\sub{left}} and
-\fld{end\sub{right}} fields of the block and its previous block. Then,
-the exact subblock is found using binary search on the \fld{sum\sub{enq}}
-field (line \ref{getChild}). 
-\here{explain a little more?}
+\begin{lemma}
+    If all \nf{Dequeue} operations in the root are helped, then all the blocks before a finished block in the root are also finished.
+\end{lemma}
+
+\begin{lemma}
+  Blocks before the most recent block that has been dequeued(computed to be dequeue) from are finished.
+If the most current \nf{Dequeue} returned(computed) \nf{null} then all the blocks before the block containing the \nf{Dequeue} are finished.
+\end{lemma}
+
+The idea above leads us to a poly-log data structure that supports throwing away all the blocks with keys smaller than an index.
+Red-black trees do this for us.
+\nf{Get(i)}, \nf{Append()} and \nf{Split(i)} are logarithmic in block trees.
+We can create a shared red-black tree just creating a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
+
+\begin{observation}
+PBRT supports poly-log operations ....
+\end{observation}
+
+\begin{lemma}
+    If we replace the arrays we used to implement \nf{blocks} with red-black trees the amortized complexity of the algorithm would be $PolyLog(p,q)$. And also the algorithm is correct.
+\end{lemma}
+
+We can help a \nf{Dequeue} by computing its response and writing it down. If the process in future failed to execute, it can read the helped value written down.
+
+\begin{lemma}
+The \nf{response} written is correct.
+\end{lemma}
+
+But how can we know which blocks in each node are finished or not?
+
+\begin{observation}
+Every $p^2$ block appended to the root, \nf{FreeMemory} is invoked.  
+\end{observation}
+
+To know the last block dequeied from we can implement a shared array among processes which they write the last root block they have dequeued from.
+
+\begin{lemma}
+    $\nf{Max(LastDequeuedFrom)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
+\end{lemma}
+
+\begin{lemma}
+    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p+q)$.
+\end{lemma}
+
+\begin{corollary}
+The space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$.
+\end{corollary}
+
+\begin{lemma}
+  The amortized step per process for the algorithm with garbage collection is $PolyLog(p+q)$.
+\end{lemma}
+
+\begin{lemma}
+  Algorithm is wait-free and linearizable.
+\end{lemma}

### 2023-01-05 podc_paper/analysis.tex
typo
Finished draft of analysis section.  (For some reason, compiling on my laptop requires commenting out some \newtheorem commands in podc-submission.tex .)

@@
 
-\begin{lemma}\label{dsearchTime}
+\begin{lemma}\label{dSearchTime}
 The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
@@
 Each of these $b-1-b_e$ blocks contains at least one operation, by \Cref{blockNotEmpty}.
-So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$
+So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
 \end{proof}
@@
 
---------
+We now bound operations' running times in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.
 
-\begin{lemma}[Worst Case Time Analysis] \label{enqDeqTime}
-The worst case number of steps for an \nf{Enqueue} is $O(\log^2 p)$ and for a \nf{Dequeue}, is $O(\log^2 p + \log q_e+ \log q_d)$, where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and $q_e$ is the size of the queue at the time the response of the \nf{Dequeue} is linearized.
+\begin{lemma}\label{enqDeqTime}
+Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
+\Eric{In the thesis, Lemma 52 said enqueues take $O(\log^2 p)$ steps.  Was this a typo?}
+and each non-null \op{Dequeue} takes
+$O(\log p\log c + \log q_e+ \log q_d)$ steps,
+where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
+$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
 \end{lemma}
 \begin{proof}
-\nf{Enqueue} consists of creating a block and appending it to the tree.
-The first part takes constant time.
-To propagate the operation to the root the algorithm tries at most two \nf{Refresh}es in each node of the path from the leaf to the root (Lines \ref{firstRefresh}, \ref{secondRefresh}).
-We can see from the code  that each \nf{Refresh} takes a constant number of steps and does $O(1)$ \nf{CAS}es.
-Since the height of the tree is $\Theta(\log p)$, \nf{Enqueue} takes $O(\log p)$ steps.
+An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
+leaf and propagates it to the root.  The \op{Propagate} routine does $O(1)$ steps 
+at each node along the path from the process's leaf to the root.
+A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
+at each node along this path. 
+So, the total number of steps for either type of operation is $O(\log p)$.
 
-A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response.
-The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps.
-To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}.
-\nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps.
-If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps.
-Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$.
-Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
+A non-null \op{Dequeue} must, in addition, do a search at line \ref{FRb} and call \op{GetEnqueue}
+at line \ref{findAnswer}.
+\Eric{Modified:}
+By \Cref{dSearchTime}, the search takes $O(\log(q_e+q_d+p))$ steps,
+since the size of the queue can change by at most $p$ within one block of operations
+(by \Cref{blockSize}).
+\op{GetEnqueue} does a binary search within each node along a path from the root to a leaf.
+The search in each node $v$ is within the subblocks of one block in $v$'s parent.
+By \Cref{blockSize}, each search takes $O(\log c)$ time, for a total of $O(\log p\log c)$ steps.
 \end{proof}
 
-\begin{lemma}[Amortized Worst-case Analysis]
-The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} is $O(\log^2 p + \log q)$,  where $q$ is the size of the queue when the operation is linearized.
-\end{lemma}
-\begin{proof}
-If we split the \nf{DoublingSearch} time cost between the corresponding \nf{Enqueue} and \nf{Dequeue}, each operation takes $O(\log^2 p +q)$ steps.
-\end{proof}
+\Eric{I left out the amortized bound, since the worst-case bound seems stronger}
 
-\begin{observation}
-    If the maximum number of concurrent processes at any time in an execution is $c$, then the amortized worst-case step complexity is $O(\log p\log c + \log q)$ per operations.
-Furthermore, in a sequential, execution where $c=1$, the step complexity of our algorithm is $\Theta(\log p + \log q)$ per operation.
-\end{observation}
-\begin{proof}
-    The analysis is similar to the two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{BinarySearch} in each call of \nf{GetEnqueue} takes $O(\log c)$ steps.
-\end{proof}
-
-\begin{theorem}
+\begin{corollary}
 The queue implementation is wait-free.
-\end{theorem}
-\begin{proof}
-To prove the claim, it is sufficient to show that every \nf{Enqueue} and \nf{Dequeue} operation terminates after a finite number of its own steps.
-This is directly concluded from Lemma \ref{enqDeqTime}.
-\end{proof}
+\end{corollary}

### 2023-01-05 podc_paper/boundedspace.tex
working on description of GC

--- 2022-12-19 podc_paper/boundedspace.tex
@@
 
-\section{Reducing Space}
+\section{Reducing Space Usage}
 \label{reducing}
-\paragraph{Reducing Space Usage}
-The \nf{blocks} arrays defined in our algorithm are unbounded.
-To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}.
-We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}).
-When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not.
-If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}.
-Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array.
-When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps.
-The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}.
-Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}.
-Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps.
-The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$.
-The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
-\begin{figure}[hbt]  
-  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
-  \caption{Array segments.}
-  \label{fig::doublingArray}
-\end{figure}
+
+We now consider how to modify our implementation to bound the space it uses.
+We replace the \fld{blocks} array in each node by a red-black tree (RBT)
+that stores the blocks.
+Each block has an additional \fld{index} field
+that represents its position within the original \fld{blocks} array, and
+blocks in a RBT are sorted by \fld{index}.
+Accessing the block in entry $i$ in the \fld{blocks} array can then be 
+replaced by searching the RBT for the block with index $i$.
+The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
+is replaced by an attempt to insert a new block with index $i$ into the RBT.
+
+Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
+However, we do not require all the standard search tree functions.
+Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
+concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
+can fail if a concurrent \op{Refresh} does a successful CAS.
+Thus, we can use a particularly simple concurrent RBT implementation.
+A sequential RBT can be made persistent using the classic node-copying technique of 
+Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
+tree are performed by making a new copy of each node $v$ that must be modified, as well
+as each node along the path from the root to $v$.
+The tree reachable from the new copy of the root is the result of applying the tree operation.
+This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
+Once a process has performed an update to the RBT representing the blocks of a node 
+$v$ in the tournament tree, 
+it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
+A search in the RBT can simply read the pointer to the RBT root and perform a standard
+sequential search on it.
+Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.
+
+To prevent RBTs from growing without bound, we would like to discard
+blocks that are no longer needed.
+Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
+also keep the running time of our operations polylogarithmic.
+We msut keep a block $B$ if any element enqueued by the operations in $E(B)$ has
+not yet been dequeued, since a later dequeue may need to traverse the block to 
+find its response.
+Because of the FIFO nature of queues, this means that we can throw away 
+blocks whose indices are smaller than the block containing the oldest enqueue whose
+value is still in the queue.
+Fortunately, there is an efficient \op{Split} operation that can remove
+all nodes with indices less than a given threshold from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+
+To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
+After every $p^2$ successful insertions into the RBT representing $\var{root}.\fld{blocks}$,
+processes cooperate to perform \op{Split} operations on all nodes of the tournament tree  
+to discard unneeded blocks.
+We keep an array $\var{last}[1..p]$ where each process writes the index of the last block in
+\var{root} that it dequeued an element from.
+If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
+of $p^2$, it does GC by first helping all pending dequeue operations by other processes,
+then reading $\var{last}[1..p]$.
+
+\here{need to complete the description of GC}
 
@@
 
-Tarjan \cite[Sec.~4.2]{Tar83} described a split algorithm for red-black trees that runs in logarithmic time.
-
 \begin{figure}[hbt]  

### 2023-01-05 podc_paper/algorithm.tex
Finished draft of analysis section.  (For some reason, compiling on my laptop requires commenting out some \newtheorem commands in podc-submission.tex .)
working on description of GC
updates based on Hossein's comments of Jan 4

@@
 
-To propagate an operation to a node $v$ in the tree, $P$ first observes
+\Eric{Reworded following parag after Hossein's comments of Jan 4}
+To propagate operations from a node $v$'s children to $v$, $P$ first observes
 the blocks in both of $v$'s children that are not already in $v$,
 creates a new block by combining information from those blocks, and attempts to append this 
-new block to $v$'s blocks array using a CAS instruction.
+new block to $v$'s blocks array using a \op{CAS} instruction.
 Following \cite{}, we call this a 3-step sequence a
 \op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
-A successful \op{Refresh} by $P$ may propagate pending operations by several other processes to $v$.
-This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
-new block into $v$'s array twice,
-then its operation has been propagated to $v$ by some other process, so $P$ can continue 
-onwards towards the root.
+A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $v$.
+However, since a successful \op{Refresh} propagates multiple pending operations 
+from $v$'s children to $v$,
+we can prove that if two \op{Refresh}es by $P$ on $v$ fail,
+then $P$'s operation has been propagated to $v$ by some other process, so $P$ can continue 
+onwards towards the~root.
 
@@
 \label{construct}
-Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{Refresh}.
+Construct a block for node $v$ that represents the operations contained in consecutive blocks in $v$'s children, as required for a \op{Refresh}.
 \item
@@
 Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
-The fourth process's Deq\sub{6} is still in progress.
+The fourth process's Deq\sub{6} is still propagating towards the root.
 The linearization order for this tree is
@@
 
-We shall show that an operation terminates only after a block containing it has been appended to the root's blocks array.
+Performing a double refresh at each node along the path from the leaf to the root ensures 
+a block containing the operation is appended to the root before the operation termintes.
 So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
 $op_1$ will be in an earlier block than $op_2$ in root's blocks array.
-Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
-Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.
+Thus, we linearize operations according to the block they belong to in the root's blocks array.
+We can choose how to order operations in the same block, since they must be concurrent.
 
-Each block in a leaf represents a single operation.
+Each block in a leaf represents one operation.
 Each block $B$ in an internal node $v$ results from merging
@@
 The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
-A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
+A block $B'$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
 or a subblock of a direct subblock of $B$.
 A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
-
 The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
@@
 As mentioned above, we are free to order operations within a block however we like.
-For convenience, we order the enqueues and dequeues separately, and put the 
-operations that were propagated from the left child before the operations from the right child.
-More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
+We order the enqueues and dequeues separately, and put the 
+operations propagated from the left child before the operations from the right child.
+More formally, we inductively define sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
 represented by a block $B$.
@@
 and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
-If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
-and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
+If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
+$v$'s left child
+and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
 \begin{eqnarray}
@@
 \renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}
-\newcommand{\hangbox}[1]{\parbox[t]{\dimexpr\linewidth-\algorithmicindent\relax}{%
-    \setlength{\hangindent}{\algorithmicindent}%
-    #1}}
-\newcommand{\spac}{\medskip}
-\newcommand{\assign}{\mbox{:=}}
+
+\here{For consistency of notation, perhaps change $n$ to $B$ in code for Enqueue, Dequeue.  Maybe also new to B for CreateBlock, Refresh.}
 
@@
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
-    \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
+    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \label{startHelpChild1}
         \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
@@
     \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
-    \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
+    \State \Call{Advance}{\var{v, h}}\label{advance}
     \State \Return{ \var{result}}
@@
 If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
-that propagated blocks from \var{v}'s children to \var{v}.
+that propagated blocks that were in \var{v}'s children prior to line \ref{firstRefresh} to \var{v}.
 Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.
@@
 and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
-If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v},
-so \op{Refresh} terminates and returns true at line \ref{addOP}.
+If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v}.
+In this case, \op{Refresh} returns true at line \ref{addOP} and terminates.
 Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].

### 2023-01-05 podc_paper/correctness.tex
working on description of GC
updates based on Hossein's comments of Jan 4

@@
 
+\Eric{I'm using immutable to mean that they are written only when the object is created.  The \fld{super} field does change from null to some value after the block is created, so it 
+is written once, rather than 0 times.  In other words, we could have initialized
+ all the other fields when the Block constructor is called.}
 A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
@@
 \end{theorem}
+
+\here{Should there be a lemma somewhere in this section that explicitly shows that each
+object we dereference is non-null?}

### 2023-01-05 podc_paper/algorithm-2.tex
Finished draft of analysis section.  (For some reason, compiling on my laptop requires commenting out some \newtheorem commands in podc-submission.tex .)
minor edits

@@
 I tries some times and i could not figure out doing that in algorithm.
-\item \gc{\typ{*Node[]} \var{leafOfProcess[1..p]} \Comment{Pointer to the leaf assigned to each process.}}
-\item \gc{\typ{int[]} \var{lastDequeuedByProcess[1..p]}}
+\item \gc{\typ{*Node[]} \var{leafOfProcess}$[1..p]$ \Comment{Pointer to the leaf assigned to each process.}}
+\item \gc{\typ{int[]} \var{last}$[1..p]$ \Comment{index of last block each process has dequeued an element from}}
 \end{itemize}
@@
 \renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}
-\newcommand{\hangbox}[1]{\parbox[t]{\dimexpr\linewidth-\algorithmicindent\relax}{%
-    \setlength{\hangindent}{\algorithmicindent}%
-    #1}}
-\newcommand{\spac}{\medskip}
-\newcommand{\assign}{\mbox{:=}}
 
@@
 
-\gc{\If{ \var{v} is root and \var{h}\%p\textsuperscript{2}==0}
+\gc{    
+	\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
+	\State \var{max} \assign 0
+	\For {$k \assign 1..p$}
+		\State $\var{max} \assign \max(\var{max}, last[k])$
+	\EndFor 
 	\State \Call{Help()}{}
-\State \Call{FreeMemory}{root, MAX(lastDequeuedByProcess)-1}
-\EndIf}
+	\State \Call{FreeMemory}{$\var{root}, \var{max}-1$}
+	\EndIf
+}
 
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
-    \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ \var{b}}
+    \gc{\State \var{last[this process id]} \assign\ \var{b}}
         \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
@@
         \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
-        \gc{\State \var{lastDequeuedByProcess[this process id]} \assign\ $\var{b}'$}
+        \gc{\State \var{last[this process id]} \assign\ $\var{b}'$}
         \State \Return{\var{e}}
@@
 \begin{lemma}
-    $\nf{Max(LastDequeuedFrom)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
+    $\nf{Max(Last)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
 \end{lemma}

### 2023-01-05 podc_paper/correctness.tex
minor edits

@@
 
-\here{need to update this outline later}
-First, we define and prove some facts about blocks and the node's \nf{head} field.
-Then, we introduce  the linearization ordering formally.
-Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1).
-After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}.
-Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).
+After proving some basic invariants in Section \ref{sec::basicProperties},
+we show in Section \ref{sec::propagating} that performing a double refresh at each node
+is sufficient to propagate an operation to the root.
+In Section \ref{sec::helpingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
+routines correctly navigate through the tournament tree.
+Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.
 
 \subsection{Basic Properties}
+\label{sec::basicProperties}
 
@@
 \subsection{Propagating Operations to the Root}
+\label{sec::propagating}
+
 Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
@@
 \subsection{Correctness of Helping Routines}
+\label{sec::helpingCorrect}
 
@@
 \subsection{Linearizability}
+\label{sec::linearizability}
 

### 2023-01-06 podc_paper/algorithm-2.tex
Editing GC

@@
 
-\section{Garbage Collection}
+\section{Detailed Description of the Bounded-Space Implementation}
+\label{reducing-details}
 
-\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
-\begin{figure}
-%\here{Does \var{leaf} have to be shared for helping when doing GC?}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-\Statex \linecomment{Shared variable}
+Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
+We first describe the modifications to the implementation.
+It uses two additional shared arrays.
 \begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process} % TODO: indent these lines in order to be more readable.
-I tries some times and i could not figure out doing that in algorithm.
-\item \gc{\typ{*Node[]} \var{leafOfProcess}$[1..p]$ \Comment{Pointer to the leaf assigned to each process.}}
-\item \gc{\typ{int[]} \var{last}$[1..p]$ \Comment{index of last block each process has dequeued an element from}}
+\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
+\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
 \end{itemize}
-
-\vspace{0.8em} %TODO: clean code
-
-\Statex \linecomment{Thread-local variable}
+The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
+Each \typ{Block} has an additional field.
 \begin{itemize}
-\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
+\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
 \end{itemize}
-
-\vspace{0.4em} %TODO: clean code
-
-\Statex $\blacktriangleright$ \typ{Node}
+To facilitate helping, each \typ{Block} in a leaf has one more additional field.
 \begin{itemize}
-\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
-\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
-\gc{\item \typ{PBRT<Block, int>} \fld{blocks} \Comment{PBRT of blocks each node contains a block and an index key}}
-\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
+\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
 \end{itemize}
-
-\vspace{0.8em} %TODO: clean code
-
-\Statex $\blacktriangleright$ \typ{Block}
-
-\begin{itemize}
-  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
-  	\item \typ{int} \fld{super}
-  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
-	\item[\com] Blocks in internal \nodes\ have the following additional fields
-	\item \typ{int} \eleft, \eright
-  		\Comment{index of last direct subblock in the left and right child}
-  	\item[\com] Blocks in leaf \nodes\ have the following additional field
-  	\item \typ{Object} \fld{element}
-  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
-  	\gc{\item \typ{Object} \fld{response}
-  		\Comment{if the block's operation is \opa{Dequeue}{} then \fld{response} is the response of the\\ operation if it has been helped.}}
-	\item[\com] Blocks in the root \node\ have the following additional field
-	\item \typ{int} \size%
-  		\Comment{size of the queue after performing all operations up to the end of this block}
-\end{itemize}
-
-%\Statex {\com\  Blocks in internal nodes have the following additional fields}
-%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
-%\begin{itemize}
-%	\item \tt{int} \eleft, \eright
-%  		\Comment{index of last direct subblock in the left and right child}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in leaf nodes have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
-%\begin{itemize}
-%  \item \tt{\sl{Object} element}
-%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in the root node have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
-%\begin{itemize}
-%  \item \tt{\sl{int} \size}%
-%  \Comment{size of the queue after performing all operations up to the end of this block}
-%\end{itemize}
-
-\end{algorithmic}
-\caption{Objects used in tournament tree data structure \label{object-fields}}
-\end{figure}
-
-\subsection{Details of the Implementation}
-
-We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
-%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
-%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
-%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
-%We use \var{num\sub{deq}} similarly.
-%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}
 
@@
 
-In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate.
-This makes the space used by the algorithm factor of the number of operations of invoked on the queue.
-Here in this section, we want to add a mechanism to free the memory allocated by the operations that are no longer needed and make the overall space used polynomial of $p+q$.
-
-It is a common way to do garbage collection in batches.
 If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

### 2023-01-06 podc_paper/boundedspace.tex
Editing GC

@@
 
-We now consider how to modify our implementation to bound the space it uses.
+In our implementation, an operation remains in the \nf{blocks} arrays forever. 
+Thus, the space used is proportional to the number of operations that have been invoked.
+Now, we sketch a mechanism to remove blocks that are no longer needed to use space
+polynomial in $p$ and $q$, while maintaining polylogarithmic (amortized) step complexity.  For lack of space, details are in Appendix~\ref{reducing-details}.
+
 We replace the \fld{blocks} array in each node by a red-black tree (RBT)
@@
 
+Each process takes $\tilde{O}(p)$ to run GC.\footnote{We use $\tilde{O}$ notation
+to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
+Since all processes may run GC,
+a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
+Between two of these GC phases, there are at least $p^2$ operations, so
+GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
+
 \paragraph{Garbage Collection}

### 2023-01-06 podc_paper/boundedspace.tex
some editing of GC

@@
 also keep the running time of our operations polylogarithmic.
-We msut keep a block $B$ if any element enqueued by the operations in $E(B)$ has
-not yet been dequeued, since a later dequeue may need to traverse the block to 
-find its response.
-Because of the FIFO nature of queues, this means that we can throw away 
-blocks whose indices are smaller than the block containing the oldest enqueue whose
-value is still in the queue.
-Fortunately, there is an efficient \op{Split} operation that can remove
-all nodes with indices less than a given threshold from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+Blocks should be kept if they contain operations still in progress.
+Moreover, a block containing an \op{Enqueue}($x$) must be kept until $x$ is dequeued.
 
 To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
-After every $p^2$ successful insertions into the RBT representing $\var{root}.\fld{blocks}$,
-processes cooperate to perform \op{Split} operations on all nodes of the tournament tree  
-to discard unneeded blocks.
-We keep an array $\var{last}[1..p]$ where each process writes the index of the last block in
-\var{root} that it dequeued an element from.
 If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
-of $p^2$, it does GC by first helping all pending dequeue operations by other processes,
-then reading $\var{last}[1..p]$.
+of $p^2$, it does GC.
+To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
+each process writes the index of the last block in
+\var{root} containing an enqueue whose element it dequeued.
+To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
+Then, it helps complete every other process's pending dequeue 
+by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
+Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
+in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
+with indices smaller than $m$.
+Fortunately, there is an efficient \op{Split} operation that can remove
+all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
+of the discarded root blocks.
 
-\here{need to complete the description of GC}
+\here{Have to talk about repeating CAS to do GC at each node up to p times}
+
+Because of GC, an operation may not find a block $B$ that it requires when it searches a RBT.  If the operation
+is a dequeue, it must have been helped before $B$ was discarded, so it can simply read its response from
+its own leaf.
+\here{What about enqueues not finding required block?}
 
@@
 GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
-
-\paragraph{Garbage Collection}
-We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
-We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed.
-We call these blocks \it{finished blocks}.
-If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished.
-Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks.
-We cannot use arrays (or vectors) to throw the garbage blocks away.
-We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}.
-If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same.
-We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.
-
-\begin{figure}[hbt]  
-  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
-  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color.
-All the subblocks of a finished block are also finished.}
-  \label{fig::finishedBlock}
-\end{figure}

### 2023-01-06 podc_paper/algorithm.tex
minor edits
Rewording to save some space

@@
 new block to $v$'s blocks array using a \op{CAS} instruction.
-Following \cite{}, we call this a 3-step sequence a
+Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
 \op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
@@
 
-\begin{figure}[tb]
+\begin{figure}[t]
 \input{tournamentTree.pdf_t}
@@
 \end{figure}
+\here{Make sure two figures appear on same page of final version}
 
-\begin{figure}
+\begin{figure}[t]
 \input{implicit.pdf_t}
@@
 Performing a double refresh at each node along the path from the leaf to the root ensures 
-a block containing the operation is appended to the root before the operation termintes.
+a block containing the operation is appended to the root before the operation completes.
 So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
-$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
-Thus, we linearize operations according to the block they belong to in the root's blocks array.
+$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
+Thus, we linearize operations according to the block they belong to in the root's array.
 We can choose how to order operations in the same block, since they must be concurrent.
@@
 
-Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
+Each tournament tree \node\ has an infinite array called \fld{blocks}.
 To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
 We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
-Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
-for the next attempt to append a \block\ to this \node.
+Each \node's \fld{head} index  stores the position in the \fld{blocks} array to be used
+for the next attempt to append a \block.
 
@@
         \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
             \State \fld{dir} \assign\ \fld{left}
@@
 
-Once a \op{Dequeue} has been appended to a block of the process's leaf and propagated to the root,
-the \opemph{IndexDequeue} routine is used to find the \op{Dequeue}'s location in the root.
+Once a \op{Dequeue} is appended to a block of the process's leaf and propagated to the root,
+the \opemph{IndexDequeue} routine finds the \op{Dequeue}'s location in the root.
 More precisely, \opa{IndexDequeue}{v, b, i}
@@
 not empty, Line \ref{computeE} computes the rank \var{e} of the
-\nf{Enqueue} whose argument is the response to the
-\nf{Dequeue}. 
-A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b_e$ of the block that contains 
+\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
+A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
 the \var{e}th enqueue.
-The index $b_e$ is at most $b$ (since the enqueue is linearized before the dequeue).
+Since the enqueue is linearized before the dequeue, $b_e\leq b$.
 To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.

### 2023-01-06 podc_paper/analysis.tex
Rewording to save some space

@@
 An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
-leaf and propagates it to the root.  The \op{Propagate} routine does $O(1)$ steps 
-at each node along the path from the process's leaf to the root.
+leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
+at each node on the path from the process's leaf to the root.
 A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
-at each node along this path. 
+at each node on this path. 
 So, the total number of steps for either type of operation is $O(\log p)$.
 
-A non-null \op{Dequeue} must, in addition, do a search at line \ref{FRb} and call \op{GetEnqueue}
+A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
 at line \ref{findAnswer}.
 \Eric{Modified:}
-By \Cref{dSearchTime}, the search takes $O(\log(q_e+q_d+p))$ steps,
-since the size of the queue can change by at most $p$ within one block of operations
-(by \Cref{blockSize}).
-\op{GetEnqueue} does a binary search within each node along a path from the root to a leaf.
-The search in each node $v$ is within the subblocks of one block in $v$'s parent.
-By \Cref{blockSize}, each search takes $O(\log c)$ time, for a total of $O(\log p\log c)$ steps.
+By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
+since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
+\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
+Each node $v$'s search is within the subblocks of one block in $v$'s parent.
+By \Cref{blockSize}, each search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
 \end{proof}

### 2023-01-06 podc_paper/boundedspace.tex
Rewording to save some space

@@
 blocks in a RBT are sorted by \fld{index}.
+The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
+is replaced by an attempt to insert a new block with index $i$ into the RBT.
 Accessing the block in entry $i$ in the \fld{blocks} array can then be 
 replaced by searching the RBT for the block with index $i$.
-The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
-is replaced by an attempt to insert a new block with index $i$ into the RBT.
+The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
+using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by \Cref{lem::sum}.
 
@@
 To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
-If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
+If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
 of $p^2$, it does GC.
@@
 
-Because of GC, an operation may not find a block $B$ that it requires when it searches a RBT.  If the operation
-is a dequeue, it must have been helped before $B$ was discarded, so it can simply read its response from
+An operation $op$'s search of a RBT may fail to find the required block $B$ that has been removed 
+by another process's GC phase.  If $op$
+is a dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
 its own leaf.
@@
 
-Each process takes $\tilde{O}(p)$ to run GC.\footnote{We use $\tilde{O}$ notation
+Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
 to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
@@
 a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
-Between two of these GC phases, there are at least $p^2$ operations, so
+There are at least $p^2$ operations between two GC phases,  so
 GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.

### 2023-01-06 podc_paper/conclusion.tex
Rewording to save some space

--- 2022-12-19 podc_paper/conclusion.tex
@@
 It may be possible to make our queue adaptive by having an operation capture a starting node
-in the tournament tree (as in \cite{??}) rather than having a statically assigned leaf as its starting point.
+in the tournament tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
 Another way to make our queue more practical might be to use it as the slow path in the
@@
 We believe the approach used here to implement a lock-free queue 
-could also be applied to obtain other lock-free
+could be applied to obtain other lock-free
 data structures with a polylogarithmic number of steps per operation.
-For example, we could implement a kind of vector data structure that stores a sequence and
+For example, we could implement a  vector data structure that stores a sequence and
 provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
@@
 An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
-A \opa{get}{i} calls \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array, taking $O(\log n + \log^2p)$ when the vector has $n$ elements.  
-An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and takes $O(\log p)$ steps.
-\here{doublecheck all preceding claims for accuracy}.
-As future work we would like to investigate whether a similar approach could be used for stacks or deques.
+A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
+An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
+\here{doublecheck all preceding claims for accuracy.}
+We would like to investigate whether a similar approach could be used for stacks, deques or even priority queues.

### 2023-01-06 podc_paper/correctness.tex
Rewording to save some space

@@
 If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
-For line \ref{enqNew}, $B$ represents a single enqueue operation, so $|E(B)|=1$ and $|D(B)|=0$.
-Moreover, $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
-$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, so the claim follows from the hypothesis.
-The proof for line \ref{deqNew}, where $B$ represents a single dequeue operation is similar.
+For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
+Since $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
+$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
+The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.
 
@@
 \begin{proof}
-The \op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
-performs a double \op{Refresh} on each node along the path from the leaf's parent to the root.
-By \Cref{lem::doubleRefresh}, this ensures $B$'s operation is contained in each node along this path.
-By \Cref{lem::noDuplicates}, it is contained in exactly one block in each node.
+\op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
+does a double \op{Refresh}~on each internal node on the path $P$ from the leaf to the root.
+By \Cref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
+There is at most one such block in each node, by \Cref{lem::noDuplicates}.
 \end{proof}

### 2023-01-06 podc_paper/introduction.tex
Rewording to save some space

@@
 contention on the front and back of the queue.
-Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
+Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
 The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
@@
 whose step complexity does not depend linearly on $p$.
-Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
-stored in the queue.
+Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
 For ease of presentation, we first give an unbounded-space construction that uses
@@
 
-Our queue uses a tournament tree where each process has its own leaf.
+Our queue uses a tournament tree where each process has its own leaf.\here{Is "tournament tree" the right word here?  Maybe call it an ordering tree?}
 Each process propagates its operations along the path from its leaf up to the root.
-The root of the tree provides an ordering for all the operations,
-and this ordering is used to linearize the operations and compute their responses.
+Operations in the root are ordered, 
+and this order is used to linearize the operations and compute their responses.
 \here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
-To ensure wait-freedom, we use a helping mechanism:  a process propagating its operation 
-to a node $n$ helps propagate all operations from both of $n$'s children.
-If we explicitly stored all operations in the nodes, this helping would become very costly.
-Instead, we use a novel implicit representation of collections
-of operations that allows for quickly merging two collections from the two children of a node,
-and for quickly accessing any particular operation in the collection.
+Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
+to a node $v$ helps propagate all operations from both of $v$'s children.
+Explicitly storing all operations in the nodes would be too costly.
+Instead, we use a novel implicit representation of sets
+of operations that allows us to quickly merge two sets from the children of a node,
+and quickly access any  operation in a~set.
 \here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-01-06 podc_paper/related.tex
Rewording to save some space

@@
 
-A number of papers have described ways to reduce contention in the MS-queue.
+Many papers have described ways to reduce contention in the MS-queue.
 Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
 added an elimination array that allows an enqueue to pass its enqueued value directly
-to a concurrent dequeue when the queue is empty or when concurrent operations 
-can be linearized to empty the queue.
+to a concurrent dequeue when the queue is empty.
+% or when concurrent operations can be linearized to empty the queue.
 However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem

### 2023-01-06 podc_paper/algorithm-2.tex
Rewording to save some space
working on GC code

@@
 \label{reducing-details}
+
+\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLEAPP}{\com\ #1}}
 
@@
 \Function{void}{Enqueue}{\typ{Object} \var{e}} 
-    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
+    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
 		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}}
-    \State \Call{Append}{\var{n}}
+		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},\\
+		\gc{\fld{index} \assign\ \var{leaf}.\head}}
+    \State \Call{Append}{\var{B}}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{} 
-    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
+    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
 	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1}
-    \State \Call{Append}{\var{n}}
+	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,\\
+	    \gc{\fld{index}\assign \var{leaf}.\head}}
+    \State \Call{Append}{\var{B}}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
-    \label{deqRest}
+    \label{deqRestGC}
 \EndFunction{Dequeue}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \State \var{leaf}.\fld{blocks}[\var{leaf}.\head] \assign\ \var{B}\label{appendLeaf}
-    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryAppend}{\var{B}, \var{leaf}.\head}\label{appendLeafGC}}
-    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEnd} 
+    \State \gc{\var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
+    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
@@
     \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
     \EndIf
@@
     \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
-    \State \var{h} \assign\ \var{v}.\head \label{readHead}
+    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}
 
@@
 	\For {$k \assign 1..p$}
-		\State $\var{max} \assign \max(\var{max}, last[k])$
+		\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
 	\EndFor 
 	\State \Call{Help()}{}
-	\State \Call{FreeMemory}{$\var{root}, \var{max}-1$}
+	\State \Call{Discard}{$\var{root}, \var{max}-1$}
 	\EndIf
+	\State
 }
 
-    \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
-        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
-        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
-            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
+    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
+        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
+        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
+            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
         \EndIf
-    \EndFor \label{endHelpChild1}
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
+    \EndFor \label{endHelpChild1GC}
+    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
+    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
     \EndIf
-    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
-    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryAppend}{\var{new}, [\var{h}]} \label{GCcas}}
-    \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
+    \State \gc{\var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
+    \State \var{\Call{Advance}{\var{v, h}}}\label{advanceGC}
     \State \Return{ \var{result}}
@@
     \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
-    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
-    \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
-    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
+    \If{$\var{v}\neq \var{root}$}
+        \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHeadGC}
+        \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1GC}
+    \EndIf
+    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHeadGC}
 \EndFunction{Advance}
@@
     \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
-    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
+    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
+    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeftGC}
     \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
@@
 			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
+    \State \gc{\var{new}.\fld{index} \assign \var{h}}
+    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
     \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
     \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
+        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
     \EndIf
@@
     \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
-    \If{$\var{v} = \var{root}$}
-        \State\Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
+    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
     \Else
@@
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
-					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
-        \EndIf \label{computeISuperEnd}
+					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
+        \EndIf \label{computeISuperEndGC}
         \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
@@
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in \var{root}.\fld{blocks}[\var{b}]}
+    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
     		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
-    \gc{\State \var{last[this process id]} \assign\ \var{b}}
-        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
+        \State \gc{\var{last[this process id]} \assign\ \var{b}}
+        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
+    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
+        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
+		\State \linecomment{compute enqueue's block using binary search}
+		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
+		\State \linecomment{find rank of enqueue within its block}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
+        \State \gc{\var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}}\label{findAnswerGC}
+        \State \gc{\var{last[this process id]} \assign\ $b_e$}
+        \State \gc{\Return{\var{e}}}
     \EndIf
-\State \linecomment{response is the \var{e}th enqueue in the root}
-        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
-			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
-		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $\var{b}'\leq \var{b}$ with $\var{root}.\fld{blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{e}$
-		\State \linecomment{find rank of enqueue within its block}
-		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
-        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
-        \gc{\State \var{last[this process id]} \assign\ $\var{b}'$}
-        \State \Return{\var{e}}
-
 \EndFunction{FindResponse}
@@
 \end{minipage}
-\caption{Queue implementation.\label{pseudocode1}}
+\caption{Queue implementation.\label{pseudocode1GC}}
 \end{figure}
@@
 \Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v}.\fld{blocks}[\var{b}])$ contains at least \var{i} enqueues}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
 
     \If{\var{v} is a leaf node}
-        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
+        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
     \Else 
@@
         \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
             \State \fld{dir} \assign\ \fld{left}
@@
         \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
-$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
+$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
         \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
@@
 
-\Function{}{FreeMemory}{\typ{Node} \var{n}, \typ{int} \var{i}}
-\State Split \var{n}.blocks.split over \var{i} and assign the grater tree to \var{n}.blocks with \tt{CAS}.
-\If{\var{n} is not a leaf}
-\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{left}}
-\State \Call{FreeMemory}{\var{n}.left, n.blocks[i].end\sub{right}}
+\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{i}}
+    \State $T \assign \var{v}.\fld{blocks}$
+    \State $\langle T_1, T_2\rangle \assign \Call{Split}{T,i}$
+    \State \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T_2$}
+    \If{\var{v} is an internal \typ{Node}}
+        \State \Call{Discard}{\var{v}.\fld{left}, \var{v}.\fld{blocks}[\var{i}].\eleft}
+        \State \Call{Discard}{\var{v}.\fld{right}, \var{v}.\fld{blocks}[\var{i}].\eright}
     \EndIf
-\EndFunction{FreeMemory}
-
+\EndFunction{Discard}
 }
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
+\caption{\label{pseudocode2GC}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
 \end{figure}

### 2023-01-06 podc_paper/algorithm.tex
working on GC code

@@
 \Function{void}{Enqueue}{\typ{Object} \var{e}} 
-    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
+    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
 		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
 		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}\label{enqNew}
-    \State \Call{Append}{\var{n}}
+    \State \Call{Append}{\var{B}}
 \EndFunction{Enqueue}
@@
 \Function{Object}{Dequeue()}{} 
-    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
+    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
 	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
 	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
-    \State \Call{Append}{\var{n}}
+    \State \Call{Append}{\var{B}}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}\label{invokeIndexDequeue}

### 2023-01-06 podc_paper/algorithm-2.tex
finished draft of code for GC

@@
     \EndIf
-    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
+    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
         \State \Call{Propagate}{\var{v}.\fld{parent}}
@@
 	\EndFor 
-	\State \Call{Help()}{}
-	\State \Call{Discard}{$\var{root}, \var{max}-1$}
+	\State \Call{Help}{}
+	\State \Call{Discard}{$\var{root}, \var{max}-2$}
 	\EndIf
@@
 \end{minipage}
-\caption{Queue implementation.\label{pseudocode1GC}}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
 \end{figure}
@@
 
+\spac
+
 \gc{
+\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
+    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has been propagated to \var{root}}
+    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]\neq \nl$}
+    \If {$\var{v} = \var{root}$} \Return{true}
+    \Else
+    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
+	    \If{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} \var{sup} \assign\ $\var{sup}+1$
+	    \EndIf
+
+	    \State \Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}
+    \EndIf
+\EndFunction{Propagated}
+
+\spac
+
 \Function{void}{Help}{}
-\State \bf{for} \var{i} \assign\ 1 to $p$
-\State h=leafOfProcess[\var{i}]
-\If{leafOfProcess[\var{i}].blocks[h].num\sub{deq}==1 and leafOfProcess[\var{i}].IndexDequeue(h,1)!=null}
-\State <b, i> \assign\ \Call{IndexDequeue}{leafOfProcess[\var{i}], h, 1}
-\State output \assign\ \Call{FindResponse}{b, i} 
-\State leafOfProcess[\var{i}].blocks[h].response \assign\ output
+    \State \linecomment{help pending dequeues that have been propagated to root}
+    \For{\bf{each} $\ell$ in $\var{leaf}[1..k]$}
+        \State $\var{h} \assign \ell.\head$
+        \If{$\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$ and \Call{Propagated}{$\ell, h$}}
+            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
+            \State \var{output} \assign\ \Call{FindResponse}{$b, i$} 
+            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \var{output}$
         \EndIf
-\State \bf{end for}
+    \EndFor
 \EndFunction{Help}
 
-\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{i}}
+\spac
+
+\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
+    \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
+    \State \var{t\sub{left}} = \var{v}.\fld{blocks}[\var{t}].\eleft
+    \State \var{t\sub{right}} = \var{v}.\fld{blocks}[\var{t}].\eright
+	\For{\var{j} \assign 1 to $p$}
         \State $T \assign \var{v}.\fld{blocks}$
-    \State $\langle T_1, T_2\rangle \assign \Call{Split}{T,i}$
-    \State \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T_2$}
-    \If{\var{v} is an internal \typ{Node}}
-        \State \Call{Discard}{\var{v}.\fld{left}, \var{v}.\fld{blocks}[\var{i}].\eleft}
-        \State \Call{Discard}{\var{v}.\fld{right}, \var{v}.\fld{blocks}[\var{i}].\eright}
+        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} > t$}
+        \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
+    \EndFor
+    \If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}
+        \State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
+        \State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}
     \EndIf
@@
 \end{algorithmic}
-\caption{\label{pseudocode2GC}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
+\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
 \end{figure}

### 2023-01-06 podc_paper/algorithm-2.tex
typo
minor edits to GC code

@@
     \State \linecomment{append block to leaf and propagate to root}
-    \State \gc{\var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
+    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
     \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
@@
 
-\gc{    
-	\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
-	\State \var{max} \assign 0
-	\For {$k \assign 1..p$}
-		\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
-	\EndFor 
-	\State \Call{Help}{}
-	\State \Call{Discard}{$\var{root}, \var{max}-2$}
-	\EndIf
-	\State
-}
+	\gc{\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}}
+	\gc{\State \var{max} \assign 0}
+	\gc{\For {$k \assign 1..p$}}
+		\gc{\State $\var{max} \assign \max(\var{max}, \var{last}[k])$}
+	\gc{\EndFor}
+	\gc{\State \Call{Help}{}}
+	\gc{\State \Call{Discard}{$\var{root}, \var{max}-1$}}
+	\gc{\EndIf}
 
@@
     \EndIf
-    \State \gc{\var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
-    \State \var{\Call{Advance}{\var{v, h}}}\label{advanceGC}
+    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
+    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
     \State \Return{ \var{result}}
@@
 			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
-    \State \gc{\var{new}.\fld{index} \assign \var{h}}
+    \gc{\State \var{new}.\fld{index} \assign \var{h}}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
-        \State \gc{\var{last[this process id]} \assign\ \var{b}}
+        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}
         \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
@@
 		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
-        \State \gc{\var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}}\label{findAnswerGC}
-        \State \gc{\var{last[this process id]} \assign\ $b_e$}
-        \State \gc{\Return{\var{e}}}
+        \gc{
+        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+        \State \var{last}[\var{this process id}] \assign\ $b_e$
+        \State \Return{\var{e}}
+        }
     \EndIf
@@
     \State \linecomment{help pending dequeues that have been propagated to root}
-    \For{\bf{each} $\ell$ in $\var{leaf}[1..k]$}
+    \For{$\ell$ in $\var{leaf}[1..k]$}
         \State $\var{h} \assign \ell.\head$
@@
     \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
-    \State \var{t\sub{left}} = \var{v}.\fld{blocks}[\var{t}].\eleft
-    \State \var{t\sub{right}} = \var{v}.\fld{blocks}[\var{t}].\eright
+    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
+    \State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
 	\For{\var{j} \assign 1 to $p$}
         \State $T \assign \var{v}.\fld{blocks}$
-        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} > t$}
+        \State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} \geq t$}
         \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
@@
 
-If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.
+\begin{lemma}
+
+\end{lemma}
+
+-------
 

### 2023-01-07 podc_paper/algorithm.tex
finished draft of code for GC
minor edits; changed tournament tree to ordering tree

@@
 \subsection{Overview}
-We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
-The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
-assigned to each process. 
+Our \emph{\ordering\ tree} data structure is used to agree on a total ordering of the operations performed on the queue.
+It is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
+for each process. 
 Each tree node  stores an array of \emph{blocks}, where each block represents a 
 sequence of enqueues and a sequence of dequeues.
-See Figure \ref{tournament} for an example.
+See Figure \ref{orderingtree} for an example.
 In this section, we use an infinite array of blocks in each node.
@@
 \input{tournamentTree.pdf_t}
-\caption{An example tournament tree with four processes. 
+\caption{An example \ordering\ tree with four processes. 
 We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
@@
 The linearization order for this tree is
-Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
+Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
 \end{figure}
@@
 \input{implicit.pdf_t}
-\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
+\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
 The leaf blocks simply show the \fld{element} field.
@@
 
-Each tournament tree \node\ has an infinite array called \fld{blocks}.
-To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
-We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
+Each \node\ of the \ordering\ tree has an infinite array called \fld{blocks}.
+To simplify the code, \fld{blocks}[0] is initialized with an empty \block\ $B_0$, 
+where $E(B_0)$ and $D(B_0)$ are empty sequences.
 Each \node's \fld{head} index  stores the position in the \fld{blocks} array to be used
@@
 We now explain how we designed the fields for this implicit representation. 
-Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.
+Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.
 
@@
 \begin{figure}
-\here{Does \var{leaf} have to be shared for helping when doing GC?}
 \begin{algorithmic}[1]
@@
 \end{algorithmic}
-\caption{Objects used in tournament tree data structure \label{object-fields}}
+\caption{Objects used in the \ordering\ tree data structure \label{object-fields}}
 \end{figure}

### 2023-01-07 podc_paper/boundedspace.tex
finished draft of code for GC
minor edits; changed tournament tree to ordering tree

@@
 Once a process has performed an update to the RBT representing the blocks of a node 
-$v$ in the tournament tree, 
+$v$ in the \ordering\ tree, 
 it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
@@
 all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
-Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
+Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
 of the discarded root blocks.
@@
 GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
+Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-07 podc_paper/conclusion.tex
minor edits; changed tournament tree to ordering tree

@@
 It may be possible to make our queue adaptive by having an operation capture a starting node
-in the tournament tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
+in the \ordering\ tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
 Another way to make our queue more practical might be to use it as the slow path in the

### 2023-01-07 podc_paper/correctness.tex
minor edits; changed tournament tree to ordering tree

@@
 In Section \ref{sec::helpingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
-routines correctly navigate through the tournament tree.
+routines correctly navigate through the \ordering\ tree.
 Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.
@@
 
-\Eric{I'm using immutable to mean that they are written only when the object is created.  The \fld{super} field does change from null to some value after the block is created, so it 
-is written once, rather than 0 times.  In other words, we could have initialized
- all the other fields when the Block constructor is called.}
 A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
@@
 
+\here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
 \begin{corollary}\label{lem::noDuplicates}
@@
 Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.
-\here{An alternative would be modify the code
-so that the correct index of the superblock is computed before storing it in $B.super$.
-In Advance, we could check if $h>v.parent.blocks[h_p].end_{left/right}$ and if so write $h_p+1$ instead of $h_p$ in $v.blocks[h].super$.
-Then, in IndexDequeue there would be no need to correct the value read from the super field.
-If we do this, the following lemma would be modified to prove that the value is perfectly accurate.}
 

### 2023-01-07 podc_paper/introduction.tex
working on GC code
minor edits; changed tournament tree to ordering tree

@@
 
-Our queue uses a tournament tree where each process has its own leaf.\here{Is "tournament tree" the right word here?  Maybe call it an ordering tree?}
+Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
 Each process propagates its operations along the path from its leaf up to the root.
@@
 and this order is used to linearize the operations and compute their responses.
-\here{Either here or in related work section, talk about previous usage of tournament tree and how ours differs from it}
+\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
 Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 

### 2023-01-07 podc_paper/related.tex
minor edits; changed tournament tree to ordering tree

@@
 queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
-We adapt their idea of processes using
-a tournament tree to agree on a linearization of
-operations.
+Our \ordering\ tree is similar to the tree structure the use to agree on a linearization ordering.
 
@@
 Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
-which uses a tournament tree to keep track of the fetch\&increments
+which also uses a tree structure to keep track of the fetch\&increments
 that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.

### 2023-01-07 podc_paper/algorithm-3.tex
working on GC

--- 2023-01-06 podc_paper/algorithm-2.tex
@@
 \end{itemize}
+We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
+we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.
 
@@
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},\\
-		\gc{\fld{index} \assign\ \var{leaf}.\head}}
+		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,$\\
+		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},$\\
+		$\fld{index} \assign\ \var{leaf}.\head$}}
     \State \Call{Append}{\var{B}}
@@
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,\\
-	    \gc{\fld{index}\assign \var{leaf}.\head}}
+	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},$\\
+	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,$\\
+	    $\fld{index}\assign \var{leaf}.\head$}}
     \State \Call{Append}{\var{B}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head-1, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
+    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
     \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
+    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what index to use?}}
 \EndFunction{Append}
@@
     \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
+    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
     \EndIf
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}}
-    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
-    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}
-
-	\gc{\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}}
-	\gc{\State \var{max} \assign 0}
-	\gc{\For {$k \assign 1..p$}}
-		\gc{\State $\var{max} \assign \max(\var{max}, \var{last}[k])$}
-	\gc{\EndFor}
-	\gc{\State \Call{Help}{}}
-	\gc{\State \Call{Discard}{$\var{root}, \var{max}-1$}}
-	\gc{\EndIf}
-
+\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}, \typ{int} \var{h}}}
+    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
     \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
@@
     \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
-    \EndIf
-    \gc{\State \var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
-    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
-    \State \Return{ \var{result}}
-\EndFunction{Refresh}
+    \gc{\If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
+    \Else \ \Return \Call{Insert}{$T$, \var{new}}
+    \EndIf}
+\EndFunction{RefreshUpdate}
 
@@
 \setcounter{ALG@line}{48}
+\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
+    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
+    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}
+	\gc{
+	\State $\var{T} \assign\ \var{v}.\fld{blocks}$
+	\If{max \fld{index} in $T$ is not $h-1$} 
+		\State \linecomment \var{v}.\fld{blocks}[h] has already been added
+		\State\Return{false}
+	\ElsIf{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
+		\State \linecomment{garbage collection phase}
+		\State \var{max} \assign 0
+		\For {$k \assign 1..p$}
+			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
+		\EndFor
+		\State \Call{Help}{}
+		\State $\var{result} \assign \Call{Discard}{\var{root}, \var{max}-1}$
+    \Else\ \linecomment{do a standard refresh}
+        \State $T' \assign \Call{RefreshUpdate}{T, v, h}$
+        \If{$T' = \nl$} \Return{\tr}
+    	\Else\ $\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, T'}$
+		\EndIf
+	\EndIf
+	}
+    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
+    \State \Return{ \var{result}}
+\EndFunction{Refresh}
+
+\spac
 
@@
 \gc{
-\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
-    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has been propagated to \var{root}}
-    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]\neq \nl$}
-    \If {$\var{v} = \var{root}$} \Return{true}
-    \Else
-    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} \var{sup} \assign\ $\var{sup}+1$
-	    \EndIf
-
-	    \State \Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}
-    \EndIf
-\EndFunction{Propagated}
-
-\spac
+%\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
+%    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has been propagated to \var{root}}
+%    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]\neq \nl$}
+%    \If {$\var{v} = \var{root}$} \Return{true}
+%    \Else
+%    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+%	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
+%	    \If{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} \var{sup} \assign\ $\var{sup}+1$
+%	    \EndIf
+%	    \State \Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}
+%    \EndIf
+%\EndFunction{Propagated}
+%
+%\spac
 
 \Function{void}{Help}{}
-    \State \linecomment{help pending dequeues that have been propagated to root}
+    \State \linecomment{help pending operations}
+
     \For{$\ell$ in $\var{leaf}[1..k]$}
         \State $\var{h} \assign \ell.\head$
-        \If{$\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$ and \Call{Propagated}{$\ell, h$}}
+        \If {$\ell.\fld{blocks}[h] = \nl$} $h \assign h-1$ \Comment{index of last block written in $\ell$}
+        \EndIf
+    	\State \Call{Propagate}{$\ell.\fld{parent}$}
+        \If{$h>1$ and $\ell.\fld{blocks}[\var{h}].\fld{element} = \nl$} \Comment{operation is a  dequeue}
             \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, h, 1}$
-            \State \var{output} \assign\ \Call{FindResponse}{$b, i$} 
-            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \var{output}$
+            \State $\ell.\fld{blocks}[\var{h}].\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
         \EndIf
@@
 
-\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
+\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
     \State \linecomment{discard blocks with index less than or equal to threshold $t$ from RBT \var{v}.\fld{blocks} and all subblocks of those blocks}
+    \If{$v$ is an internal node}
 	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
@@
 		\For{\var{j} \assign 1 to $p$}
+			\State $h \assign \var{v}.\head$
         	\State $T \assign \var{v}.\fld{blocks}$
         	\State $T'  \assign \Call{Split}{T,t}$ \Comment{$T'$ is the root of new RBT containing $T$'s blocks with $\var{index} \geq t$}
-        \State \bf{exit when} \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{splitCASGC}
+			\State $T'' \assign \Call{RefreshUpdate}{T',v,h}$ \Comment{do a \op{Refresh} too}
+        	\State $\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, (T'' = \nl\ ?\ T' : T'')$}\label{splitCASGC}
+        	\State \bf{exit when} \var{result} \Comment{if \op{CAS} succeeds go to line \ref{Discardrecurse}}
     	\EndFor
-    \If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}
+    	\If{\var{v} is an internal \typ{Node}} \Comment{discard subblocks in descendants of $v$}\label{Discardrecurse}
         	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}
@@
     	\EndIf
+	\EndIf
+	\State \Return \var{result}
 \EndFunction{Discard}
@@
 
+\here{don't GC leaves of other processes}
+\here{insertion into rbt should only work if it is larger than any node already there}
+\here{line 39 checking if child head is not written should instead check if max node is = childHead}
 -------
@@
 \end{lemma}
+
+\here{Remark on an optimization:  instead of storing \head\ field separate from \fld{blocks}, we could store
+the maximum index of the RBT in the root of the RBT.  This would simplify things, but require modifying the proof more extensively.}

### 2023-01-07 podc_paper/algorithm-2.tex
minor edits; changed tournament tree to ordering tree
working on GC

--- 2023-01-06 podc_paper/algorithm-2.tex
@@
 \end{itemize}
-The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
 Each \typ{Block} has an additional field.
@@
 \end{itemize}
+The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
+Each RBT is initialized with an empty block with index 0.
+
+\here{Talk about what happens if a block is not found in an RBT}
+
+We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
+we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.
 
@@
 \Function{void}{Enqueue}{\typ{Object} \var{e}} 
+	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}},\\
-		\gc{\fld{index} \assign\ \var{leaf}.\head}}
+		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
+		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}},$\\
+		$\fld{index} \assign\ h+1$}}\label{enqNewGC}
     \State \Call{Append}{\var{B}}
@@
 \Function{Object}{Dequeue()}{} 
+	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    \fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[\var{leaf}.\head-1].\fld{sum\sub{deq}}+1,\\
-	    \gc{\fld{index}\assign \var{leaf}.\head}}
+	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
+	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1,$\\
+	    $\fld{index}\assign h+1$}}\label{deqNewGC}
     \State \Call{Append}{\var{B}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \var{leaf}.\head, 1}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{$h+1$}, 1}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \State \gc{\var{leaf}.\fld{blocks}.\Call{TryInsert}{\var{B}}\label{appendLeafGC}}
-    \State \var{leaf}.\head\ \assign\ \var{leaf}.\head+1 \label{appendEndGC} 
+    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
+    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what threshold to use for the split?}}
 \EndFunction{Append}
@@
     \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
+    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
     \EndIf
-    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
+    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
         \State \Call{Propagate}{\var{v}.\fld{parent}}
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}}
-    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
-    \State \var{h} \assign\ \var{v}.\head \label{readHeadGC}
-
 	\gc{
-	\If{ \var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
+\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
+    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
+	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
+    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
+	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
+		\State \linecomment{garbage collection phase}
 		\State \var{max} \assign 0
@@
 		\EndFor
-	\State \Call{Help()}{}
-	\State \Call{Discard}{$\var{root}, \var{max}-1$}
+		\State \Call{Help}{}
+		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
+    \Else\ \linecomment{do a standard refresh}
+        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
+        \If{$T' = \nl$} \Return{\tr}
+    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
 		\EndIf
-	\State
+	\EndIf
+%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
+\EndFunction{GCRefresh}
 }
-
-    \ForEach{\fld{dir} {\keywordfont{in}} \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
-        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHeadGC}
-        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNullGC}
-            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvanceGC}
-        \EndIf
-    \EndFor \label{endHelpChild1GC}
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC} 
-    \EndIf
-    \State \gc{\var{result} \assign\ \var{v}.\fld{blocks}.\Call{TryInsert}{\var{new}} \label{casGC}}
-    \State \var{\Call{Advance}{\var{v, h}}}\label{advanceGC}
-    \State \Return{ \var{result}}
-\EndFunction{Refresh}
-
-\spac
-
-\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
-    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
-    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
-    \If{$\var{v}\neq \var{root}$}
-        \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHeadGC}
-        \State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1GC}
-    \EndIf
-    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHeadGC}
-\EndFunction{Advance}
 
@@
 \setcounter{ALG@line}{48}
+\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}}
+    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
+    \State \Call{SetSuper}{$v$, $T$}
+    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
+    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
+    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
+    \Else \ \Return \Call{Insert}{$T$, \var{new}}\label{refreshInsertGC}
+    \EndIf
+\EndFunction{RefreshUpdate}
+}
+\spac
+
+\gc{
+\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
+	\State \linecomment{set \fld{super} field of last block in $T$}
+	\State \linecomment{and in $v$'s children's RBTs}
+	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
+	\If{$v\neq \var{root}$}	
+		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
+		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
+	\EndIf
+	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
+    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
+		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
+    \EndFor \label{endHelpChild1GC}
+\EndFunction{SetChildrenSuper}
+}
+
+\spac
 
@@
     \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
-    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeftGC}
-    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$
+    \gc{
+    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
+    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
+    \State \var{new}.\fld{index} \assign \var{i}
+    }
 	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
@@
 			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
-    \State \gc{\var{new}.\fld{index} \assign \var{h}}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
@@
 \EndFunction{CreateBlock}
+\end{algorithmic}
+\end{minipage}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}[t]{0.465\textwidth}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{48}
+\gc{
+\Function{void}{Help}{}
+    \State \linecomment{help pending operations}
+
+    \For{$\ell$ in $\var{leaf}[1..k]$}
+        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
+    	\State \Call{Propagate}{$\ell.\fld{parent}$}
+        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
+            \State \linecomment{operation is a  dequeue}
+            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
+            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
+        \EndIf
+    \EndFor
+\EndFunction{Help}
 
 \spac
+
+\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
+    \State \linecomment{discard blocks with index less than or equal}
+    \State \linecomment{to $t$ from $v$'s RBT and their subblocks}
+    \If{$v$ is an internal node}
+	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
+    	\State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
+		\For{\var{j} \assign 1 to $p$}
+        	\State $T \assign \var{v}.\fld{blocks}$\label{readT2GC}
+        	\State $T'  \assign \Call{Split}{T,t}$\label{split}
+			\State \linecomment{\op{Split} removes blocks with $\var{index} < t$}
+			\State $T'' \assign \Call{RefreshUpdate}{T',v}$ \label{RU2}
+        	\State \hangbox{$\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, \\
+					\hspace*{10mm}(T'' = \nl\ ?\ T' : T'')$}}\label{splitCASGC}
+        	\State \bf{exit when} \var{result} 
+	        \State \linecomment{if \op{CAS} succeeds go to line \ref{discardrecursecomment}}
+    	\EndFor
+    	\If{\var{v} is an internal \typ{Node}} 
+			\State \linecomment{discard subblocks in descendants of $v$}\label{discardrecursecomment}
+        	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}\label{discardrecurse1}
+        	\State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}\label{discardrecurse2}
+    	\EndIf
+	\EndIf
+	\State \Return \var{result}
+\EndFunction{Discard}
+}
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{48}
 
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
-        \State \gc{\var{last[this process id]} \assign\ \var{b}}
+        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}\label{setlast1}
         \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
@@
 		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
-        \State \gc{\var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}}\label{findAnswerGC}
-        \State \gc{\var{last[this process id]} \assign\ $b_e$}
-        \State \gc{\Return{\var{e}}}
+        \gc{
+        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+        \State \var{last}[\var{this process id}] \assign\ $b_e$\label{setlast2}
+        \State \Return{\var{e}}
+        }
     \EndIf
@@
 \end{minipage}
-\caption{Queue implementation.\label{pseudocode1GC}}
+
+\caption{Bounded-space queue implementation, continued.\label{pseudocode1GC}}
 \end{figure}
@@
 
-\gc{
-\Function{void}{Help}{}
-\State \bf{for} \var{i} \assign\ 1 to $p$
-\State h=leafOfProcess[\var{i}]
-\If{leafOfProcess[\var{i}].blocks[h].num\sub{deq}==1 and leafOfProcess[\var{i}].IndexDequeue(h,1)!=null}
-\State <b, i> \assign\ \Call{IndexDequeue}{leafOfProcess[\var{i}], h, 1}
-\State output \assign\ \Call{FindResponse}{b, i} 
-\State leafOfProcess[\var{i}].blocks[h].response \assign\ output
-\EndIf
-\State \bf{end for}
-\EndFunction{Help}
-
-\Function{void}{Discard}{\typ{Node} \var{v}, \typ{int} \var{i}}
-    \State $T \assign \var{v}.\fld{blocks}$
-    \State $\langle T_1, T_2\rangle \assign \Call{Split}{T,i}$
-    \State \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T_2$}
-    \If{\var{v} is an internal \typ{Node}}
-        \State \Call{Discard}{\var{v}.\fld{left}, \var{v}.\fld{blocks}[\var{i}].\eleft}
-        \State \Call{Discard}{\var{v}.\fld{right}, \var{v}.\fld{blocks}[\var{i}].\eright}
-    \EndIf
-\EndFunction{Discard}
-}
 \end{algorithmic}
-\caption{\label{pseudocode2GC}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?}}
+\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
 \end{figure}
 
-If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.
+\subsection{Correctness}
+Unfortunately, there are enough changes to the algorithm that a new proof of correctness
+is required.  Its structure mirrors the proof of the original algorithm, but requires additional
+reasoning to ensure GC does not interfere with other routines.
+
+\begin{observation}\label{RBTinsertions}
+When a block is inserted into an RBT $T$, its \fld{index} is one greater than the maximum \fld{index} in $T$.
+\end{observation}
+\begin{proof}
+If line \ref{appendLeafGC} inserts a block into a leaf's RBT, then it was created at 
+line \ref{enqNewGC} or \ref{deqNewGC} with \fld{index} $h+1$, where $h$ is the maximum \fld{index}
+in the leaf's RBT.  When the new block is inserted, the maximum \fld{index} in the RBT is still $h$,
+since only the owner of the leaf modifies the leaf's RBT.
+
+If line \ref{refreshInsertGC} inserts a block  into an internal node's RBT $T$,
+it is created by the call to \op{CreateBlock} at line \ref{invokeCreateBlockGC} and 
+has \fld{index} $h$, which is set at line \ref{refreshReadMax} to  the maximum \fld{index} in $T$ plus one.
+\end{proof}
+
+\begin{observation}\label{lastinroot}
+Each entry in \var{last} is the \var{index} of a block that has appeared in the root's RBT.
+\end{observation}
+\begin{proof}
+Only line \ref{setlast1} and \ref{setlast2} in \op{FindResponse} write values into \var{last}.
+The value written is either $b$ or $b_e \leq b$.
+By the precondition of \op{FindResponse}, $\var{root}.\fld{blocks}[b]$ exists, so the index $b$ has appeared
+in the root's RBT.
+By \Cref{RBTinsertions}, all values less than $b$ have appeared in the root's RBT too.
+\end{proof}
+
+\begin{observation}
+\label{edirGC} 
+If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
+$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
+\end{observation}
+\begin{proof}
+The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
+on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
+\end{proof}
+
+\begin{invariant}
+For each RBT $T$, 
+$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
+\end{invariant}
+\begin{proof}
+The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
+We show that each time a new RBT is constructed, it preserves the invariant.
+When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
+implies that the invariant is preserved.
+
+Suppose a new RBT is created by a \op{Split}($T,t$) operation $S$ on at line \ref{split} of a \op{Discard}($v,t$), 
+and that the invariant held prior to $S$.
+Clearly, $S$ preserves the property that indices are consecutive.  To show that the new tree
+is non-empty we first prove that the call to \op{Discard}($v,t$) has a parameter $t$ that is at least as big as a value previously
+read from the index field of some block in $v$'s RBT.
+If \op{Discard} was called from line \ref{discard}, then $t+1$ was read from an entry of \var{last}, so the claim follows from \Cref{lastinroot}.
+If \op{Discard} was called from line \ref{discardrecurse1} or \ref{discardrecurse2}, the claim follows
+from \Cref{edirGC}.
+Since the invariant held prior to $S$, no split before $S$ emptied $v$'s tree.  Therefore,
+some block with index at least $t$ is still in $T$ when $S$ occurs.  That block will be in the new tree created by $S$.
+\end{proof}
+
+Thus, calls to \op{MaxBlock} are well-defined.
+Also, this means that the sequence of blocks added to $v.\fld{blocks}$ have consecutive \fld{index} values, starting from 1, so the references to $v.\fld{blocks}[i]$ (meaning the block in $v$'s RBT with \fld{index} $i$) are well-defined.
+
+\begin{corollary}\label{maxNonDecreasing}
+The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
+\end{corollary}
+\begin{proof}
+The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
+a tree $T$ by a tree that is built from $T$ by inserting an element or removing the elements with 
+the smallest \fld{index} values.
+\end{proof}
+
+We now prove the analogue of Lemma \ref{lem::headProgress}.
+\begin{lemma}
+ If $b>0$ and a block with index $b$ has been inserted into $v.\fld{blocks}$ then 
+ $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
+ $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
+\end{lemma}
+\begin{proof}
+The block $B$ with index $b$ was installed into $v$'s RBT by the CAS at line \ref{casGC} or \ref{splitCASGC} 
+that changed the tree from $T$ to $T_{new}$.
+Before this CAS, the tree $T$ was read from $v.blocks$ at line \ref{readTGC} or \ref{readT2GC}, respectively,
+before the \op{RefreshUpdate} on line \ref{RU1} or \ref{RU2} that inserted $B$ to get $T_{new}$.
+Since the previous block $B'=v.\fld{blocks}[b-1]$ was already in $T$ (by \Cref{RBTinsertions}),
+the \op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the \op{CreateBlock}$(v,b)$ that created $B$ started.
+It follows from \Cref{maxNonDecreasing} that the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.
+Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
+\end{proof}
+
+\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} hold true for the modified algorithm.
+Their proofs are identical to those given in Section \ref{sec::basicProperties}.
+
+-------
 

### 2023-01-07 podc_paper/boundedspace.tex
working on GC

@@
 Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
-However, we do not require all the standard search tree functions.
+However, we do not require all the standard search tree operations.
 Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
@@
 can fail if a concurrent \op{Refresh} does a successful CAS.
+Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other node in the RBT.
 Thus, we can use a particularly simple concurrent RBT implementation.
@@
 in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
-with indices smaller than $m$.
+with indices smaller than $m-1$.
 Fortunately, there is an efficient \op{Split} operation that can remove
-all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
 Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks

### 2023-01-07 podc_paper/correctness.tex
working on GC

@@
 \begin{proof}
-We prove the lemma by reverse induction.
-If $B$ and $B'$ are in leaves, they have no subblocks, so the claim is true.
-Assume the claim is true for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
-Consider the direct subblocks of $B$ and $B'$ defined by (\ref{defsubblock}).
+We prove the lemma by reverse induction on the depth.
+If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
+Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
+Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
 If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.

### 2023-01-07 podc_paper/algorithm-2.tex
minor changes
working on GC; just realized there is a problem that splits might not get installed even after p tries at nodes other than the root

@@
 we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.
+
+\here{Walk reader through the code (and how it is different from basic versino)}
+
+\op{MaxBlock} is a query operation on a RBT that returns the block with the maximum \fld{index}.
+We could keep a pointer to the maximum block in the root of the RBT so that \op{MaxBlock}
+can be done in constant time.
 
@@
         \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
-        \If{$T' = \nl$} \Return{\tr}
+        \If{$T' = \nl$} \Return{\tr}\label{emptyblockGC}
     	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
@@
     \EndFor \label{endHelpChild1GC}
-\EndFunction{SetChildrenSuper}
+\EndFunction{SetSuper}
 }
@@
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
-        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
+        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
     \Else
@@
 
-\begin{invariant}
+\begin{invariant}\label{nonemptyConsec}
 For each RBT $T$, 
@@
 
-Thus, calls to \op{MaxBlock} are well-defined.
-Also, this means that the sequence of blocks added to $v.\fld{blocks}$ have consecutive \fld{index} values, starting from 1, so the references to $v.\fld{blocks}[i]$ (meaning the block in $v$'s RBT with \fld{index} $i$) are well-defined.
+Since RBTs are always non-empty, calls to \op{MaxBlock} are well-defined.  
+Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
+in $v$'s tree at some time during the execution.
+It follows from \Cref{RBTinsertions} and \Cref{nonemptyConsec} that this block is unique, if 
+it exists.
+We also use this notation in the code to indicate that a search of the RBT should be performed for the block
+with \fld{index} $b$.
+We describe later what should happen if the block is not found by the search because it has been removed
+by a GC phase.
+\here{remember to come back to this point}
 
@@
 The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
-a tree $T$ by a tree that is built from $T$ by inserting an element or removing the elements with 
-the smallest \fld{index} values.
+a tree $T$ by a tree that is built from $T$ by inserting an element and/or removing the elements with 
+the smallest \fld{index} values, without emptying the tree.  Thus, the maximum \fld{index} in the tree cannot decrease.
 \end{proof}
@@
 
-\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} hold true for the modified algorithm.
-Their proofs are identical to those given in Section \ref{sec::basicProperties}.
+\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
+hold true for the modified algorithm.  Their proofs are identical to those given in 
+Section \ref{sec::basicProperties} since they only depend on \Cref{lem::headProgress} 
+and the definition of subblocks given in (\ref{defsubblock}).
+The following pair of lemmas are analogous to \Cref{successfulRefresh} for the modified \op{Refresh} routine.
+We say that an operation has been \emph{propagated} to node $v$ in configuration $C$ if
+some block containing the operation has been in $v.blocks$ at or before $C$.
+
+\begin{lemma}\label{nullRefreshGC}
+Suppose $v.\fld{blocks}=T$ at some configuration $C$ and \op{RefreshUpdate}($T,v$) is invoked after $C$ and
+returns \nl.  Then, all operations that were propagated to $v$'s
+children before $C$ have been propagated to $v$ before $C$.
+\end{lemma}
+\begin{proof}
+Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $C$.
+Then, for some $i$, a block $B$ with \fld{index} $i$ has been in $v.\fld{left.blocks}$ before $C$.
+By Lemma \ref{maxNonDecreasing}, 
+the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ at all times during the \op{RefreshUpdate}.
+So, the value of $\var{new}.\eleft$ chosen at line \ref{createEndLeftGC} of the \op{RefreshUpdate}'s call
+to \op{CreateBlock} is at least $i$.
+
+Let $h$ be the value set at line \ref{refreshReadMax} of the \op{RefreshUpdate}.
+The \op{RefreshUpdate} returns \nl\ at line \ref{addOPGC}
+because the \op{CreateBlock}$(v,h)$  at line \ref{CBnullGC} returns \nl.
+By reasoning identical to the last paragraph of \Cref{successfulRefresh},
+it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ that
+the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
+$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
+contain no operations.
+Since each block contains at least one operation by \Cref{blockNotEmpty}
+these ranges must be empty, so $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
+This implies that the block $B$ containing $op$ is a subblock of some block that has appeared in $v.\fld{blocks}$, so $op$ has been propagated to $v$.
+\end{proof}
+
+\begin{lemma}\label{successfulRefreshGC}
+If a \op{CAS} at line \ref{casGC} or \ref{splitCASGC} modifies $v.\fld{blocks}$ from $T$ to $T'$,
+then all operations that were propagated to $v$'s children before $v.\fld{blocks}$ was set to $T$
+have been propagated to $v$ in the configuration after the CAS.
+\end{lemma}
+\begin{proof}
+Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $v.\fld{blocks}$ was set to $T$.
+Then, for some $i$, a block $B$ with \fld{index} $i$ that contains $op$
+has been in $v.\fld{left.blocks}$ before $v.\fld{blocks}$ was set to $T$.
+Let $c$ be the \op{CAS} instruction that changes $v.\fld{blocks}$ from $T$ to $T'$.
+If line \ref{casGC} performs $c$, then the \op{RefreshUpdate} $R$ at line \ref{RU1} that creates
+$T'$ began after $T$ was read from $v.\fld{blocks}$ at line \ref{readTGC}.
+Otherwise, line \ref{splitCASGC} performs $c$ and the \op{RefreshUpdate} $R$ at line \ref{RU2},
+which is invoked after $T$ was read from $v.\fld{blocks}$ at line \ref{readT2GC},
+either returned \nl\ or $T'$.
+
+If the \op{RefreshUpdate} $R$ returned \nl, the lemma follows from \Cref{nullRefreshGC}.
+So, assume \op{RefreshUpdate} $R$ returned $T'$.
+Then $T'$ contains a newly inserted block created by the call
+to \op{CreateBlock} on line \ref{invokeCreateBlockGC}.
+Since $R$ was invoked after $T$ was stored in $v.\fld{blocks}$, 
+the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ throughout $R$, by \Cref{maxNonDecreasing}.
+Thus. the \eleft\ field of the new block added to $T'$ by $R$'s call to \op{CreateBlock} is
+set to a value greater than or equal to $i$ at line \ref{createEndLeftGC}.
+By the definition of subblocks, it follows that the block $B$ containing $op$ is a subblock of a block
+that has been in $v.blocks$, as required.
+\end{proof}
+
+This allows us to show that a double \op{Refresh} propagates operations up the tree, 
+as in \Cref{lem::doubleRefresh}.
+
+\begin{lemma}\label{lem::doubleRefreshGC}
+Consider two consecutive terminating calls $R_1$, $R_2$ to \op{GCRefresh}($v$) by the same process.
+All operations propagated to $v$'s children when $R_1$ begins
+are propagated to $v$ when $R_2$ terminates.
+\end{lemma}
+\begin{proof}
+\end{proof}
 

### 2023-01-07 podc_paper/correctness.tex
working on GC; just realized there is a problem that splits might not get installed even after p tries at nodes other than the root

@@
 in the configuration $C$ immediately before $R$ executed line \ref{readHead}.
-Then some block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
+Then for some $i$, the block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
 By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
@@
 these blocks contain no operations.
-By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq new.\eleft \geq i$.
+By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
 Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
@@
 \begin{lemma}\label{lem::doubleRefresh}
-Consider two consecutive terminating calls $R_1$, $R_2$ to \nf{Refresh}($v$) by the same process.
+Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
 When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.

### 2023-01-07 podc_paper/algorithm-2.tex
Updated code to make GC run node-by-node instead

@@
 \label{reducing-details}
+
+\HERE{Idea:  replace last array with a bit in blocks that gets set when somebody dequeues an element in the block.
+then can use the max index of any marked block minus 1 as the last completed block.
+In order to ensure that blocks of dequeues are also cleared out,
+maybe dequeues that return null have to mark all blocks along the dequeue's path?
+}
 
@@
 
+\here{Emphasize that RBT nodes have \emph{pointers to} blocks.  Thus when RBT does node copying,
+it copies the pointer to the block; it does not make a copy of the block itself.
+This is important because if we mark a block that has been node-copied, we want the block to still be marked.  Similarly for super field.}
+
 \renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}
@@
 \Function{void}{Enqueue}{\typ{Object} \var{e}} 
-	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
+    \gc{
+    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
 	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
 		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
-		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}},$\\
-		$\fld{index} \assign\ h+1$}}\label{enqNewGC}
+		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}$,\\
+		$\fld{index}\assign h+1$}\label{enqNewGC}
+	}
     \State \Call{Append}{\var{B}}
@@
 \Function{Object}{Dequeue()}{} 
-	\gc{\State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
+    \gc{
+    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
 	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
-	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1,$\\
-	    $\fld{index}\assign h+1$}}\label{deqNewGC}
+	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1$,\\
+	    $\fld{index}\assign h+1$}\label{deqNewGC}
+	}
     \State \Call{Append}{\var{B}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{$h+1$}, 1}
-    \State \Return{ \Call{FindResponse}{\var{b, i}}}
+    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{h+1}, 1, \fa}$
+    \gc{
+    \State $\var{res} \assign \Call{FindResponse}{b, i}$
+	\If{$\var{res} = \nl$} \linecomment{mark blocks}
+		\State $\Call{IndexDequeue}{\var{leaf}, h+1, 1, \tr}$
+	\EndIf
+	\State \Return{\var{res}}
+    }
     \label{deqRestGC}
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{Insert}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
+    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
-    \gc{\State \here{Need to clean out my own leaf sometimes--how to determine what threshold to use for the split?}}
 \EndFunction{Append}
@@
 
+\gc{
+\Function{RBT}{AddBlock}{\typ{RBT} \var{T}, \typ{Block} \var{B}}
+    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
+    \If{\var{B}.\fld{index} is a multiple of $p^2$}
+    	\State \linecomment{Do garbage collection}
+        \State \Call{Help}{}
+        \State $m$ \assign\ maximum \fld{index} of marked block in $T$
+        \State $T' \assign\ \Call{Split}{T,m-1}$
+        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
+        \State \Return \Call{Insert}{$T',B$}
+	\Else\ \Return \Call{Insert}{$T, B$}
+    \EndIf
+\EndFunction{AddBlock}
+}
+\spac
+
 \Function{void}{Propagate}{\typ{Node} \var{v}}
     \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{\gc{GCRefresh}}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{\gc{GCRefresh}}{\var{v}} \label{secondRefreshGC}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
     \EndIf
@@
 
+\Function{boolean}{Refresh}{\typ{Node} \var{v}}
+	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
 	\gc{
-\Function{boolean}{GCRefresh}{\typ{Node} \var{v}}
-    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}, with GC}
-	\State $\var{T} \assign\ \var{v}.\fld{blocks}$\label{readTGC}
-    \State $\var{h} \assign\ \Call{MaxBlock}{T}.\fld{index}$ \label{readHeadGC}
-	\If{\var{v} is root and $\var{h} \mbox{ mod } p^2=0$}
-		\State \linecomment{garbage collection phase}
-		\State \var{max} \assign 0
-		\For {$k \assign 1..p$}
-			\State $\var{max} \assign \max(\var{max}, \var{last}[k])$
-		\EndFor
-		\State \Call{Help}{}
-		\State \Return $\Call{Discard}{\var{root}, \var{max}-1}$\label{discard}
-    \Else\ \linecomment{do a standard refresh}
-        \State $T' \assign \Call{RefreshUpdate}{T, v}$\label{RU1}
-        \If{$T' = \nl$} \Return{\tr}\label{emptyblockGC}
-    	\Else\ \Return $\Call{CAS}{\var{v}.\fld{blocks}, T, T'}$\label{casGC}
+	\State $T \assign v.\fld{blocks}$
+    \State \Call{SetSuper}{$v$, $T$}
+    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
+    }
+    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
+    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
+    \Else  
+    	\gc{
+	    \State $T' \assign \Call{AddBlock}{T, \var{new}}$
+		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
+		}
     \EndIf
-	\EndIf
-%    \State \Call{Advance}{\var{v, h}}\label{advanceGC}
-\EndFunction{GCRefresh}
-}
+
+\EndFunction{Refresh}
 
@@
 \setcounter{ALG@line}{48}
-\gc{\Function{RBT}{RefreshUpdate}{\typ{RBT} \var{T}, \typ{Node} \var{v}}
-    \State \linecomment{append new block to RBT $T$ read from $v.\fld{blocks}$}
-    \State \Call{SetSuper}{$v$, $T$}
-    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
-    \Else \ \Return \Call{Insert}{$T$, \var{new}}\label{refreshInsertGC}
-    \EndIf
-\EndFunction{RefreshUpdate}
-}
-\spac
 
@@
     \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
-    \State \var{new}.\fld{index} \assign \var{i}
+    \State $\var{new}.\fld{index} \assign i$
     }
@@
 \EndFunction{CreateBlock}
+
+\spac
+
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
+    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
+        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
+    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
+        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
+		\State \linecomment{compute enqueue's block using binary search}
+		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
+		\State \linecomment{find rank of enqueue within its block}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
+        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+    \EndIf
+\EndFunction{FindResponse}
+
 \end{algorithmic}
@@
 \EndFunction{Help}
-
-\spac
-
-\Function{boolean}{Discard}{\typ{Node} \var{v}, \typ{int} \var{t}}
-    \State \linecomment{discard blocks with index less than or equal}
-    \State \linecomment{to $t$ from $v$'s RBT and their subblocks}
-    \If{$v$ is an internal node}
-	    \State \var{t\sub{left}} \assign \var{v}.\fld{blocks}[\var{t}].\eleft
-    	\State \var{t\sub{right}} \assign \var{v}.\fld{blocks}[\var{t}].\eright
-		\For{\var{j} \assign 1 to $p$}
-        	\State $T \assign \var{v}.\fld{blocks}$\label{readT2GC}
-        	\State $T'  \assign \Call{Split}{T,t}$\label{split}
-			\State \linecomment{\op{Split} removes blocks with $\var{index} < t$}
-			\State $T'' \assign \Call{RefreshUpdate}{T',v}$ \label{RU2}
-        	\State \hangbox{$\var{result} \assign \Call{CAS}{\var{v}.\fld{blocks}, T, \\
-					\hspace*{10mm}(T'' = \nl\ ?\ T' : T'')$}}\label{splitCASGC}
-        	\State \bf{exit when} \var{result} 
-	        \State \linecomment{if \op{CAS} succeeds go to line \ref{discardrecursecomment}}
-    	\EndFor
-    	\If{\var{v} is an internal \typ{Node}} 
-			\State \linecomment{discard subblocks in descendants of $v$}\label{discardrecursecomment}
-        	\State \Call{Discard}{\var{v}.\fld{left}, \var{t\sub{left}}}\label{discardrecurse1}
-        	\State \Call{Discard}{\var{v}.\fld{right}, \var{t\sub{right}}}\label{discardrecurse2}
-    	\EndIf
-	\EndIf
-	\State \Return \var{result}
-\EndFunction{Discard}
 }
@@
 
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}, \gc{\typ{boolean} \var{mark}}}
     \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
@@
     \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \gc{
+    \If{\var{mark}} \var{v}.\fld{blocks}[\var{b}].\fld{mark} \assign \tr
+    \EndIf
+    }
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
@@
 \EndFunction{IndexDequeue}
-
-\spac
-
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
-        \gc{\State \var{last}[\var{this process id}] \assign\ \var{b}}\label{setlast1}
-        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
-    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
-			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
-		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
-		\State \linecomment{find rank of enqueue within its block}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
-        \gc{
-        \State \var{e} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
-        \State \var{last}[\var{this process id}] \assign\ $b_e$\label{setlast2}
-        \State \Return{\var{e}}
-        }
-    \EndIf
-\EndFunction{FindResponse}
 
@@
     \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
-
+	\gc{\State $\var{v}.\fld{blocks}[\var{b}].\fld{marked} := \tr$}
     \If{\var{v} is a leaf node}

### 2023-01-08 podc_paper/algorithm-2.tex
updated code to remove mark bits used for GC; went back to last array but one per node

@@
 \label{reducing-details}
-
-\HERE{Idea:  replace last array with a bit in blocks that gets set when somebody dequeues an element in the block.
-then can use the max index of any marked block minus 1 as the last completed block.
-In order to ensure that blocks of dequeues are also cleared out,
-maybe dequeues that return null have to mark all blocks along the dequeue's path?
-}
 
@@
 The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
+The nodes in the RBT each contain a pointer to a block, as well as the standard RBT node attributes: child pointers and a colour bit; this means that when an update on a RBT uses node-copying, the pointer to the block
+gets copied, rather than copying the block itself.)
 Each RBT is initialized with an empty block with index 0.
@@
     \State \linecomment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
+    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
@@
 \gc{
-\Function{RBT}{AddBlock}{\typ{RBT} \var{T}, \typ{Block} \var{B}}
+\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
     \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
@@
     	\State \linecomment{Do garbage collection}
+		\State $m \assign 0$
+		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
+		\EndFor
         \State \Call{Help}{}
-        \State $m$ \assign\ maximum \fld{index} of marked block in $T$
         \State $T' \assign\ \Call{Split}{T,m-1}$
@@
     	\gc{
-	    \State $T' \assign \Call{AddBlock}{T, \var{new}}$
+	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
 		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
@@
     \gc{
-    \If{\var{mark}} \var{v}.\fld{blocks}[\var{b}].\fld{mark} \assign \tr
+    \If{\var{mark} and $b>\var{v}.\fld{last}[\mbox{process id}]$}
+    	\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
     \EndIf
@@
     \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
-	\gc{\State $\var{v}.\fld{blocks}[\var{b}].\fld{marked} := \tr$}
+	\gc{
+	\If{$b > \var{v}.\fld{last}[\mbox{process id}]$}
+		\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
+	\EndIf
+	}
     \If{\var{v} is a leaf node}

### 2023-01-08 podc_paper/algorithm-2.tex
factored out subroutine CompleteDequeue, added some description of GC algorithm details

@@
 We first describe the modifications to the implementation.
-It uses two additional shared arrays.
+To avoid confusion, we use nodes to refer to the nodes of the ordering tree, and blocks to refer
+to the nodes of a RBT (since the RBT stores blocks).
+
+The space-bounded implementation uses an additional shared array so that processes can access one another's leaves to perform helping.
 \begin{itemize}
 \item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
-\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
 \end{itemize}
@@
 \end{itemize}
+Each \typ{Node} has an additional array field to help determine which of its blocks are safe to discard.
+\begin{itemize}
+\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the node that process $k$ }\\
+\mbox{ }\Comment{has accessed, either to find the enqueued item for its dequeue}\\
+\mbox{ }\Comment{or during a \nl\ dequeue}
+\end{itemize}
 The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
-The nodes in the RBT each contain a pointer to a block, as well as the standard RBT node attributes: child pointers and a colour bit; this means that when an update on a RBT uses node-copying, the pointer to the block
-gets copied, rather than copying the block itself.)
 Each RBT is initialized with an empty block with index 0.
+The \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
+longer needed; we can instead find the maximum \fld{index} of any block in the RBT.
+To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
+returns the block with the maximum \fld{index}.
+We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
+can be done in constant time, without affecting the time of other RBT operations.
 
 \here{Talk about what happens if a block is not found in an RBT}
+\here{Talk about \emph{marking} a block to indicate that any blocks with smaller indices will no longer
+be needed by operations that haven't started yet.}
 
-We also apply an optimization:  instead of storing a \head\ field separate from \fld{blocks},
-we can simply use the maximum \fld{index} of any block in the \fld{blocks} RBT.
+Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
+The \op{Propagate} and \op{FindResponse} routines are unchanged.
+Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
+visited by the routines to mark the blocks they visit in those nodes.
+Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
+to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
+of the \head\ field).
+\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
+separate routine called \op{CompleteDequeue} so that it can also be used by other processes
+helping to complete the operation.
 
-\here{Walk reader through the code (and how it is different from basic versino)}
+Once a dequeue has been propagated to the root, 
+\bf{\op CompleteDequeue} computes its answer.  This routine can also be called by helpers 
+who want to help complete the dequeue.
+In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
+it also updates \var{last} entries if the computed response is \nl.
+It does this by calling \op{IndexDequeue} again, this time with \tr\ as its final parameter.
+This extra parameter has been added to 
+\op{IndexDequeue} to indicate whether it should mark blocks that it traverses (which all contain the dequeue).
 
-\op{MaxBlock} is a query operation on a RBT that returns the block with the maximum \fld{index}.
-We could keep a pointer to the maximum block in the root of the RBT so that \op{MaxBlock}
-can be done in constant time.
+\here{Discuss new routines AddBlock, SetSuper, Help, and largely modified Refresh}
 
-\here{Emphasize that RBT nodes have \emph{pointers to} blocks.  Thus when RBT does node copying,
-it copies the pointer to the block; it does not make a copy of the block itself.
-This is important because if we mark a block that has been node-copied, we want the block to still be marked.  Similarly for super field.}
+\here{does RBT have to have pointer to block rather than block itself so that updates to super field work correctly?}
 
@@
     \gc{
-    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
+    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
 	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}}+1,$\\
-		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}$,\\
-		$\fld{index}\assign h+1$}\label{enqNewGC}
+		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
+		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,\\
+		$\fld{index}\assign h$}\label{enqNewGC}
 	}
@@
     \gc{
-    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}
+    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{enq}},$\\
-	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h].\fld{sum\sub{deq}}+1$,\\
-	    $\fld{index}\assign h+1$}\label{deqNewGC}
+	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
+	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,\\
+	    $\fld{index}\assign h$}\label{deqNewGC}
 	}
     \State \Call{Append}{\var{B}}
-    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, \gc{h+1}, 1, \fa}$
+    \gc{\State \Return \Call{CompleteDequeue}{\var{leaf}, $h$}}
+\EndFunction{Dequeue}
+
+\spac
+
 \gc{
+\Function{Object}{CompleteDequeue}{\var{leaf}, $h$}
+    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
+    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1, \fa}$
     \State $\var{res} \assign \Call{FindResponse}{b, i}$
-	\If{$\var{res} = \nl$} \linecomment{mark blocks}
-		\State $\Call{IndexDequeue}{\var{leaf}, h+1, 1, \tr}$
+	\If{$\var{res} = \nl$} \linecomment{update \var{last} values}
+		\State $\Call{IndexDequeue}{\var{leaf}, h, 1, \tr}$
 	\EndIf
 	\State \Return{\var{res}}
+\EndFunction{CompleteDequeue}
 }
-    \label{deqRestGC}
-\EndFunction{Dequeue}
 
@@
 
-\spac
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
 
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{48}
 \Function{boolean}{Refresh}{\typ{Node} \var{v}}
@@
 
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{48}
-
+\spac
 \gc{
@@
     \For{$\ell$ in $\var{leaf}[1..k]$}
-        \State $\var{last} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
+        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
     	\State \Call{Propagate}{$\ell.\fld{parent}$}
-        \If{$\var{last}.\fld{element}\hspace{-.1mm} =\hspace{-.1mm} \nl$\hspace{-.15mm} and\hspace{-.15mm} $\var{last}.\fld{index}\hspace{-.12mm}>\hspace{-.12mm}0$} 
+        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$} 
             \State \linecomment{operation is a  dequeue}
-            \State $\langle\var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\ell, \var{last}.\fld{index}, 1}$
-            \State $\var{last}.\fld{response} \assign\ \Call{FindResponse}{b, i}$ 
+            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDequeue}{\ell, \var{B}.\fld{index}}$ 
         \EndIf
@@
 
-\caption{Bounded-space queue implementation, continued.\label{pseudocode1GC}}
+\caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
 \end{figure}
@@
 \end{algorithmic}
-\caption{\label{pseudocode2GC}Bounded-space queue implementation, continued. }
+\caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }
 \end{figure}

### 2023-01-08 podc_paper/algorithm-2.tex
separated correctness argument for GC version into separate file.

@@
 	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
-		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}
+		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}\label{casGC}
 		}
@@
 \end{figure}
-
-\subsection{Correctness}
-Unfortunately, there are enough changes to the algorithm that a new proof of correctness
-is required.  Its structure mirrors the proof of the original algorithm, but requires additional
-reasoning to ensure GC does not interfere with other routines.
-
-\begin{observation}\label{RBTinsertions}
-When a block is inserted into an RBT $T$, its \fld{index} is one greater than the maximum \fld{index} in $T$.
-\end{observation}
-\begin{proof}
-If line \ref{appendLeafGC} inserts a block into a leaf's RBT, then it was created at 
-line \ref{enqNewGC} or \ref{deqNewGC} with \fld{index} $h+1$, where $h$ is the maximum \fld{index}
-in the leaf's RBT.  When the new block is inserted, the maximum \fld{index} in the RBT is still $h$,
-since only the owner of the leaf modifies the leaf's RBT.
-
-If line \ref{refreshInsertGC} inserts a block  into an internal node's RBT $T$,
-it is created by the call to \op{CreateBlock} at line \ref{invokeCreateBlockGC} and 
-has \fld{index} $h$, which is set at line \ref{refreshReadMax} to  the maximum \fld{index} in $T$ plus one.
-\end{proof}
-
-\begin{observation}\label{lastinroot}
-Each entry in \var{last} is the \var{index} of a block that has appeared in the root's RBT.
-\end{observation}
-\begin{proof}
-Only line \ref{setlast1} and \ref{setlast2} in \op{FindResponse} write values into \var{last}.
-The value written is either $b$ or $b_e \leq b$.
-By the precondition of \op{FindResponse}, $\var{root}.\fld{blocks}[b]$ exists, so the index $b$ has appeared
-in the root's RBT.
-By \Cref{RBTinsertions}, all values less than $b$ have appeared in the root's RBT too.
-\end{proof}
-
-\begin{observation}
-\label{edirGC} 
-If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
-$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
-\end{observation}
-\begin{proof}
-The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
-on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
-\end{proof}
-
-\begin{invariant}\label{nonemptyConsec}
-For each RBT $T$, 
-$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
-\end{invariant}
-\begin{proof}
-The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
-We show that each time a new RBT is constructed, it preserves the invariant.
-When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
-implies that the invariant is preserved.
-
-Suppose a new RBT is created by a \op{Split}($T,t$) operation $S$ on at line \ref{split} of a \op{Discard}($v,t$), 
-and that the invariant held prior to $S$.
-Clearly, $S$ preserves the property that indices are consecutive.  To show that the new tree
-is non-empty we first prove that the call to \op{Discard}($v,t$) has a parameter $t$ that is at least as big as a value previously
-read from the index field of some block in $v$'s RBT.
-If \op{Discard} was called from line \ref{discard}, then $t+1$ was read from an entry of \var{last}, so the claim follows from \Cref{lastinroot}.
-If \op{Discard} was called from line \ref{discardrecurse1} or \ref{discardrecurse2}, the claim follows
-from \Cref{edirGC}.
-Since the invariant held prior to $S$, no split before $S$ emptied $v$'s tree.  Therefore,
-some block with index at least $t$ is still in $T$ when $S$ occurs.  That block will be in the new tree created by $S$.
-\end{proof}
-
-Since RBTs are always non-empty, calls to \op{MaxBlock} are well-defined.  
-Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
-in $v$'s tree at some time during the execution.
-It follows from \Cref{RBTinsertions} and \Cref{nonemptyConsec} that this block is unique, if 
-it exists.
-We also use this notation in the code to indicate that a search of the RBT should be performed for the block
-with \fld{index} $b$.
-We describe later what should happen if the block is not found by the search because it has been removed
-by a GC phase.
-\here{remember to come back to this point}
-
-\begin{corollary}\label{maxNonDecreasing}
-The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
-\end{corollary}
-\begin{proof}
-The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
-a tree $T$ by a tree that is built from $T$ by inserting an element and/or removing the elements with 
-the smallest \fld{index} values, without emptying the tree.  Thus, the maximum \fld{index} in the tree cannot decrease.
-\end{proof}
-
-We now prove the analogue of Lemma \ref{lem::headProgress}.
-\begin{lemma}
- If $b>0$ and a block with index $b$ has been inserted into $v.\fld{blocks}$ then 
- $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
- $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-The block $B$ with index $b$ was installed into $v$'s RBT by the CAS at line \ref{casGC} or \ref{splitCASGC} 
-that changed the tree from $T$ to $T_{new}$.
-Before this CAS, the tree $T$ was read from $v.blocks$ at line \ref{readTGC} or \ref{readT2GC}, respectively,
-before the \op{RefreshUpdate} on line \ref{RU1} or \ref{RU2} that inserted $B$ to get $T_{new}$.
-Since the previous block $B'=v.\fld{blocks}[b-1]$ was already in $T$ (by \Cref{RBTinsertions}),
-the \op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the \op{CreateBlock}$(v,b)$ that created $B$ started.
-It follows from \Cref{maxNonDecreasing} that the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.
-Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
-\end{proof}
-
-\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
-hold true for the modified algorithm.  Their proofs are identical to those given in 
-Section \ref{sec::basicProperties} since they only depend on \Cref{lem::headProgress} 
-and the definition of subblocks given in (\ref{defsubblock}).
-The following pair of lemmas are analogous to \Cref{successfulRefresh} for the modified \op{Refresh} routine.
-We say that an operation has been \emph{propagated} to node $v$ in configuration $C$ if
-some block containing the operation has been in $v.blocks$ at or before $C$.
-
-\begin{lemma}\label{nullRefreshGC}
-Suppose $v.\fld{blocks}=T$ at some configuration $C$ and \op{RefreshUpdate}($T,v$) is invoked after $C$ and
-returns \nl.  Then, all operations that were propagated to $v$'s
-children before $C$ have been propagated to $v$ before $C$.
-\end{lemma}
-\begin{proof}
-Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $C$.
-Then, for some $i$, a block $B$ with \fld{index} $i$ has been in $v.\fld{left.blocks}$ before $C$.
-By Lemma \ref{maxNonDecreasing}, 
-the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ at all times during the \op{RefreshUpdate}.
-So, the value of $\var{new}.\eleft$ chosen at line \ref{createEndLeftGC} of the \op{RefreshUpdate}'s call
-to \op{CreateBlock} is at least $i$.
-
-Let $h$ be the value set at line \ref{refreshReadMax} of the \op{RefreshUpdate}.
-The \op{RefreshUpdate} returns \nl\ at line \ref{addOPGC}
-because the \op{CreateBlock}$(v,h)$  at line \ref{CBnullGC} returns \nl.
-By reasoning identical to the last paragraph of \Cref{successfulRefresh},
-it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ that
-the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
-$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
-contain no operations.
-Since each block contains at least one operation by \Cref{blockNotEmpty}
-these ranges must be empty, so $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
-This implies that the block $B$ containing $op$ is a subblock of some block that has appeared in $v.\fld{blocks}$, so $op$ has been propagated to $v$.
-\end{proof}
-
-\begin{lemma}\label{successfulRefreshGC}
-If a \op{CAS} at line \ref{casGC} or \ref{splitCASGC} modifies $v.\fld{blocks}$ from $T$ to $T'$,
-then all operations that were propagated to $v$'s children before $v.\fld{blocks}$ was set to $T$
-have been propagated to $v$ in the configuration after the CAS.
-\end{lemma}
-\begin{proof}
-Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $v.\fld{blocks}$ was set to $T$.
-Then, for some $i$, a block $B$ with \fld{index} $i$ that contains $op$
-has been in $v.\fld{left.blocks}$ before $v.\fld{blocks}$ was set to $T$.
-Let $c$ be the \op{CAS} instruction that changes $v.\fld{blocks}$ from $T$ to $T'$.
-If line \ref{casGC} performs $c$, then the \op{RefreshUpdate} $R$ at line \ref{RU1} that creates
-$T'$ began after $T$ was read from $v.\fld{blocks}$ at line \ref{readTGC}.
-Otherwise, line \ref{splitCASGC} performs $c$ and the \op{RefreshUpdate} $R$ at line \ref{RU2},
-which is invoked after $T$ was read from $v.\fld{blocks}$ at line \ref{readT2GC},
-either returned \nl\ or $T'$.
-
-If the \op{RefreshUpdate} $R$ returned \nl, the lemma follows from \Cref{nullRefreshGC}.
-So, assume \op{RefreshUpdate} $R$ returned $T'$.
-Then $T'$ contains a newly inserted block created by the call
-to \op{CreateBlock} on line \ref{invokeCreateBlockGC}.
-Since $R$ was invoked after $T$ was stored in $v.\fld{blocks}$, 
-the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ throughout $R$, by \Cref{maxNonDecreasing}.
-Thus. the \eleft\ field of the new block added to $T'$ by $R$'s call to \op{CreateBlock} is
-set to a value greater than or equal to $i$ at line \ref{createEndLeftGC}.
-By the definition of subblocks, it follows that the block $B$ containing $op$ is a subblock of a block
-that has been in $v.blocks$, as required.
-\end{proof}
-
-This allows us to show that a double \op{Refresh} propagates operations up the tree, 
-as in \Cref{lem::doubleRefresh}.
-
-\begin{lemma}\label{lem::doubleRefreshGC}
-Consider two consecutive terminating calls $R_1$, $R_2$ to \op{GCRefresh}($v$) by the same process.
-All operations propagated to $v$'s children when $R_1$ begins
-are propagated to $v$ when $R_2$ terminates.
-\end{lemma}
-\begin{proof}
-\end{proof}
-
--------
-
-In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}).
-Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).
-
-%\begin{lemma}
-%Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
-%\end{lemma}
-%
-%\begin{lemma}
-%Total number of the blocks in the tree is $O(p^3+pq)$.
-%\end{lemma}
-
-\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$.
-It is safe to remove \nf{Dequeue} $d$ from the tree after the $d$ terminates.
-We can remove a block after the conditions told are satisfied for all of its \nf{Enqueue}s and \nf{Dequeue}s.
-A \nf{Dequeue} may go to sleep for a long time and prevent a \nf{block} to be removed (the block it is in or the block its response is in).
-In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere.
-After writing down the response of th \nf{Dequeue} can be removed, since if the process wanted the response it can read from the helped response.
-
-\begin{definition}
-A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
-\end{definition}
-
-\begin{corollary}
-If a block is finished, then all of its subblocks are also finished.  
-\end{corollary}
-
-\begin{lemma}
-It is safe to remove a finished block from the non-leaf nodes.
-\end{lemma}
-
-If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued.
-This gives us the idea that if a block is finished then all the blocks before it are also finished.
-If an operation in a block goes to sleep for a long time then other processes help the operation so the block gets finished.
-There are less than $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely in an amortized poly-log time.
-
-\begin{lemma}
-    If all \nf{Dequeue} operations in the root are helped, then all the blocks before a finished block in the root are also finished.
-\end{lemma}
-
-\begin{lemma}
-  Blocks before the most recent block that has been dequeued(computed to be dequeue) from are finished.
-If the most current \nf{Dequeue} returned(computed) \nf{null} then all the blocks before the block containing the \nf{Dequeue} are finished.
-\end{lemma}
-
-The idea above leads us to a poly-log data structure that supports throwing away all the blocks with keys smaller than an index.
-Red-black trees do this for us.
-\nf{Get(i)}, \nf{Append()} and \nf{Split(i)} are logarithmic in block trees.
-We can create a shared red-black tree just creating a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
-
-\begin{observation}
-PBRT supports poly-log operations ....
-\end{observation}
-
-\begin{lemma}
-    If we replace the arrays we used to implement \nf{blocks} with red-black trees the amortized complexity of the algorithm would be $PolyLog(p,q)$. And also the algorithm is correct.
-\end{lemma}
-
-We can help a \nf{Dequeue} by computing its response and writing it down. If the process in future failed to execute, it can read the helped value written down.
-
-\begin{lemma}
-The \nf{response} written is correct.
-\end{lemma}
-
-But how can we know which blocks in each node are finished or not?
-
-\begin{observation}
-Every $p^2$ block appended to the root, \nf{FreeMemory} is invoked.  
-\end{observation}
-
-To know the last block dequeied from we can implement a shared array among processes which they write the last root block they have dequeued from.
-
-\begin{lemma}
-    $\nf{Max(Last)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
-\end{lemma}
-
-\begin{lemma}
-    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p+q)$.
-\end{lemma}
-
-\begin{corollary}
-The space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$.
-\end{corollary}
-
-\begin{lemma}
-  The amortized step per process for the algorithm with garbage collection is $PolyLog(p+q)$.
-\end{lemma}
-
-\begin{lemma}
-  Algorithm is wait-free and linearizable.
-\end{lemma}
-
-\here{Remark on an optimization:  instead of storing \head\ field separate from \fld{blocks}, we could store
-the maximum index of the RBT in the root of the RBT.  This would simplify things, but require modifying the proof more extensively.}

### 2023-01-08 podc_paper/correctness-2.tex
separated correctness argument for GC version into separate file.

% !TEX root =  podc-submission.tex

\subsection{Correctness}
Unfortunately, there are enough changes to the algorithm that a new proof of correctness
is required.  Its structure mirrors the proof of the original algorithm, but requires additional
reasoning to ensure GC does not interfere with other routines.

\begin{observation}\label{RBTinsertions}
When a block is inserted into an RBT $T$, its \fld{index} is one greater than the maximum \fld{index} in $T$.
\end{observation}
\begin{proof}
If line \ref{appendLeafGC} inserts a block into a leaf's RBT, then it was created at 
line \ref{enqNewGC} or \ref{deqNewGC} with \fld{index} $h+1$, where $h$ is the maximum \fld{index}
in the leaf's RBT.  When the new block is inserted, the maximum \fld{index} in the RBT is still $h$,
since only the owner of the leaf modifies the leaf's RBT.

If line \ref{refreshInsertGC} inserts a block  into an internal node's RBT $T$,
it is created by the call to \op{CreateBlock} at line \ref{invokeCreateBlockGC} and 
has \fld{index} $h$, which is set at line \ref{refreshReadMax} to  the maximum \fld{index} in $T$ plus one.
\end{proof}

\begin{observation}\label{lastinroot}
Each entry in \var{last} is the \var{index} of a block that has appeared in the root's RBT.
\end{observation}
\begin{proof}
Only line \ref{setlast1} and \ref{setlast2} in \op{FindResponse} write values into \var{last}.
The value written is either $b$ or $b_e \leq b$.
By the precondition of \op{FindResponse}, $\var{root}.\fld{blocks}[b]$ exists, so the index $b$ has appeared
in the root's RBT.
By \Cref{RBTinsertions}, all values less than $b$ have appeared in the root's RBT too.
\end{proof}

\begin{observation}
\label{edirGC} 
If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
\end{observation}
\begin{proof}
The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
\end{proof}

\begin{invariant}\label{nonemptyConsec}
For each RBT $T$, 
$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
\end{invariant}
\begin{proof}
The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
We show that each time a new RBT is constructed, it preserves the invariant.
When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
implies that the invariant is preserved.

Suppose a new RBT is created by a \op{Split}($T,t$) operation $S$ on at line \ref{split} of a \op{Discard}($v,t$), 
and that the invariant held prior to $S$.
Clearly, $S$ preserves the property that indices are consecutive.  To show that the new tree
is non-empty we first prove that the call to \op{Discard}($v,t$) has a parameter $t$ that is at least as big as a value previously
read from the index field of some block in $v$'s RBT.
If \op{Discard} was called from line \ref{discard}, then $t+1$ was read from an entry of \var{last}, so the claim follows from \Cref{lastinroot}.
If \op{Discard} was called from line \ref{discardrecurse1} or \ref{discardrecurse2}, the claim follows
from \Cref{edirGC}.
Since the invariant held prior to $S$, no split before $S$ emptied $v$'s tree.  Therefore,
some block with index at least $t$ is still in $T$ when $S$ occurs.  That block will be in the new tree created by $S$.
\end{proof}

Since RBTs are always non-empty, calls to \op{MaxBlock} are well-defined.  
Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in $v$'s tree at some time during the execution.
It follows from \Cref{RBTinsertions} and \Cref{nonemptyConsec} that this block is unique, if 
it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if the block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

\begin{corollary}\label{maxNonDecreasing}
The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
\end{corollary}
\begin{proof}
The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
a tree $T$ by a tree that is built from $T$ by inserting an element and/or removing the elements with 
the smallest \fld{index} values, without emptying the tree.  Thus, the maximum \fld{index} in the tree cannot decrease.
\end{proof}

We now prove the analogue of Lemma \ref{lem::headProgress}.
\begin{lemma}
 If $b>0$ and a block with index $b$ has been inserted into $v.\fld{blocks}$ then 
 $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
 $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
\end{lemma}
\begin{proof}
The block $B$ with index $b$ was installed into $v$'s RBT by the CAS at line \ref{casGC} or \ref{splitCASGC} 
that changed the tree from $T$ to $T_{new}$.
Before this CAS, the tree $T$ was read from $v.blocks$ at line \ref{readTGC} or \ref{readT2GC}, respectively,
before the \op{RefreshUpdate} on line \ref{RU1} or \ref{RU2} that inserted $B$ to get $T_{new}$.
Since the previous block $B'=v.\fld{blocks}[b-1]$ was already in $T$ (by \Cref{RBTinsertions}),
the \op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the \op{CreateBlock}$(v,b)$ that created $B$ started.
It follows from \Cref{maxNonDecreasing} that the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
hold true for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they only depend on \Cref{lem::headProgress} 
and the definition of subblocks given in (\ref{defsubblock}).
The following pair of lemmas are analogous to \Cref{successfulRefresh} for the modified \op{Refresh} routine.
We say that an operation has been \emph{propagated} to node $v$ in configuration $C$ if
some block containing the operation has been in $v.blocks$ at or before $C$.

\begin{lemma}\label{nullRefreshGC}
Suppose $v.\fld{blocks}=T$ at some configuration $C$ and \op{RefreshUpdate}($T,v$) is invoked after $C$ and
returns \nl.  Then, all operations that were propagated to $v$'s
children before $C$ have been propagated to $v$ before $C$.
\end{lemma}
\begin{proof}
Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $C$.
Then, for some $i$, a block $B$ with \fld{index} $i$ has been in $v.\fld{left.blocks}$ before $C$.
By Lemma \ref{maxNonDecreasing}, 
the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ at all times during the \op{RefreshUpdate}.
So, the value of $\var{new}.\eleft$ chosen at line \ref{createEndLeftGC} of the \op{RefreshUpdate}'s call
to \op{CreateBlock} is at least $i$.

Let $h$ be the value set at line \ref{refreshReadMax} of the \op{RefreshUpdate}.
The \op{RefreshUpdate} returns \nl\ at line \ref{addOPGC}
because the \op{CreateBlock}$(v,h)$  at line \ref{CBnullGC} returns \nl.
By reasoning identical to the last paragraph of \Cref{successfulRefresh},
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ that
the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
contain no operations.
Since each block contains at least one operation by \Cref{blockNotEmpty}
these ranges must be empty, so $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
This implies that the block $B$ containing $op$ is a subblock of some block that has appeared in $v.\fld{blocks}$, so $op$ has been propagated to $v$.
\end{proof}

\begin{lemma}\label{successfulRefreshGC}
If a \op{CAS} at line \ref{casGC} or \ref{splitCASGC} modifies $v.\fld{blocks}$ from $T$ to $T'$,
then all operations that were propagated to $v$'s children before $v.\fld{blocks}$ was set to $T$
have been propagated to $v$ in the configuration after the CAS.
\end{lemma}
\begin{proof}
Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $v.\fld{blocks}$ was set to $T$.
Then, for some $i$, a block $B$ with \fld{index} $i$ that contains $op$
has been in $v.\fld{left.blocks}$ before $v.\fld{blocks}$ was set to $T$.
Let $c$ be the \op{CAS} instruction that changes $v.\fld{blocks}$ from $T$ to $T'$.
If line \ref{casGC} performs $c$, then the \op{RefreshUpdate} $R$ at line \ref{RU1} that creates
$T'$ began after $T$ was read from $v.\fld{blocks}$ at line \ref{readTGC}.
Otherwise, line \ref{splitCASGC} performs $c$ and the \op{RefreshUpdate} $R$ at line \ref{RU2},
which is invoked after $T$ was read from $v.\fld{blocks}$ at line \ref{readT2GC},
either returned \nl\ or $T'$.

If the \op{RefreshUpdate} $R$ returned \nl, the lemma follows from \Cref{nullRefreshGC}.
So, assume \op{RefreshUpdate} $R$ returned $T'$.
Then $T'$ contains a newly inserted block created by the call
to \op{CreateBlock} on line \ref{invokeCreateBlockGC}.
Since $R$ was invoked after $T$ was stored in $v.\fld{blocks}$, 
the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ throughout $R$, by \Cref{maxNonDecreasing}.
Thus. the \eleft\ field of the new block added to $T'$ by $R$'s call to \op{CreateBlock} is
set to a value greater than or equal to $i$ at line \ref{createEndLeftGC}.
By the definition of subblocks, it follows that the block $B$ containing $op$ is a subblock of a block
that has been in $v.blocks$, as required.
\end{proof}

This allows us to show that a double \op{Refresh} propagates operations up the tree, 
as in \Cref{lem::doubleRefresh}.

\begin{lemma}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{GCRefresh}($v$) by the same process.
All operations propagated to $v$'s children when $R_1$ begins
are propagated to $v$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
\end{proof}

-------

In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}). Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).

%\begin{lemma}
%Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
%\end{lemma}
%
%\begin{lemma}
%Total number of the blocks in the tree is $O(p^3+pq)$.
%\end{lemma}

\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$. It is safe to remove \nf{Dequeue} $d$ from the tree after the $d$ terminates. We can remove a block after the conditions told are satisfied for all of its \nf{Enqueue}s and \nf{Dequeue}s. A \nf{Dequeue} may go to sleep for a long time and prevent a \nf{block} to be removed (the block it is in or the block its response is in). In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere. After writing down the response of th \nf{Dequeue} can be removed, since if the process wanted the response it can read from the helped response.

\begin{definition}
A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
\end{definition}

\begin{corollary}
If a block is finished, then all of its subblocks are also finished.  
\end{corollary}

\begin{lemma}
It is safe to remove a finished block from the non-leaf nodes.
\end{lemma}

If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued. This gives us the idea that if a block is finished then all the blocks before it are also finished. If an operation in a block goes to sleep for a long time then other processes help the operation so the block gets finished. There are less than $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely in an amortized poly-log time.

\begin{lemma}
    If all \nf{Dequeue} operations in the root are helped, then all the blocks before a finished block in the root are also finished.
\end{lemma}

\begin{lemma}
  Blocks before the most recent block that has been dequeued(computed to be dequeue) from are finished. If the most current \nf{Dequeue} returned(computed) \nf{null} then all the blocks before the block containing the \nf{Dequeue} are finished.
\end{lemma}

The idea above leads us to a poly-log data structure that supports throwing away all the blocks with keys smaller than an index. Red-black trees do this for us. \nf{Get(i)}, \nf{Append()} and \nf{Split(i)} are logarithmic in block trees.
We can create a shared red-black tree just creating a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.

\begin{observation}
PBRT supports poly-log operations ....
\end{observation}

\begin{lemma}
    If we replace the arrays we used to implement \nf{blocks} with red-black trees the amortized complexity of the algorithm would be $PolyLog(p,q)$. And also the algorithm is correct.
\end{lemma}

We can help a \nf{Dequeue} by computing its response and writing it down. If the process in future failed to execute, it can read the helped value written down.

\begin{lemma}
The \nf{response} written is correct.
\end{lemma}

But how can we know which blocks in each node are finished or not?

\begin{observation}
Every $p^2$ block appended to the root, \nf{FreeMemory} is invoked.  
\end{observation}

To know the last block dequeied from we can implement a shared array among processes which they write the last root block they have dequeued from.

\begin{lemma}
    $\nf{Max(Last)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
\end{lemma}

\begin{lemma}
    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p+q)$.
\end{lemma}

\begin{corollary}
The space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$.
\end{corollary}

\begin{lemma}
  The amortized step per process for the algorithm with garbage collection is $PolyLog(p+q)$.
\end{lemma}

\begin{lemma}
  Algorithm is wait-free and linearizable.
\end{lemma}

\here{Remark on an optimization:  instead of storing \head\ field separate from \fld{blocks}, we could store
the maximum index of the RBT in the root of the RBT.  This would simplify things, but require modifying the proof more extensively.}

### 2023-01-08 podc_paper/algorithm-2.tex
working on correctness of GC version

@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{0}
+\setcounter{ALG@line}{200}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{48}
+\setcounter{ALG@line}{248}
 \Function{boolean}{Refresh}{\typ{Node} \var{v}}
@@
 	\gc{
-	\State $T \assign v.\fld{blocks}$
+	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
     \State \Call{SetSuper}{$v$, $T$}
@@
     \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
-    \State $\var{new}.\fld{index} \assign i$
+    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
     }
@@
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
         \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{48}
+\setcounter{ALG@line}{307}
 \gc{
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{48}
+\setcounter{ALG@line}{318}
 
@@
     \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] has been propagated to root}
-    \State \linecomment{and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
+    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
     \gc{
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{101}
+\setcounter{ALG@line}{340}
 
 \Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
 	\gc{

### 2023-01-08 podc_paper/correctness-2.tex
working on correctness of GC version

@@
 reasoning to ensure GC does not interfere with other routines.
+The following observation describes
+how the set of blocks in the RBT can be modified.
 
-\begin{observation}\label{RBTinsertions}
-When a block is inserted into an RBT $T$, its \fld{index} is one greater than the maximum \fld{index} in $T$.
-\end{observation}
+\begin{lemma}\label{RBTupdates}
+Suppose a step of the algorithm changes $v.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
+If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
+$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some $m$.
+\end{lemma}
 \begin{proof}
-If line \ref{appendLeafGC} inserts a block into a leaf's RBT, then it was created at 
-line \ref{enqNewGC} or \ref{deqNewGC} with \fld{index} $h+1$, where $h$ is the maximum \fld{index}
-in the leaf's RBT.  When the new block is inserted, the maximum \fld{index} in the RBT is still $h$,
-since only the owner of the leaf modifies the leaf's RBT.
+The RBT of a node is updated only at line \ref{appendLeafGC} or \ref{casGC}.
 
-If line \ref{refreshInsertGC} inserts a block  into an internal node's RBT $T$,
-it is created by the call to \op{CreateBlock} at line \ref{invokeCreateBlockGC} and 
-has \fld{index} $h$, which is set at line \ref{refreshReadMax} to  the maximum \fld{index} in $T$ plus one.
+If line \ref{appendLeafGC} of an \op{Append} operation modifies $v.\fld{blocks}$,
+then $v$ is a leaf node, and no other process ever modifies $v.\fld{blocks}$.
+$T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
+$B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
+Either way, $B.\fld{index} = \max(I)+1$.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
+So the claim is satisfied.
+
+If line \ref{casGC} of a \op{Refresh} modifies $v.\fld{blocks}$, then $v$ is an internal node.
+After reading $T$ from $v.\fld{blocks}$ at line \ref{refreshReadTGC},
+then creates the  block \var{new},
+and then calls \op{AddBlock}($T,new$) to create $T'$.
+Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
+So the claim is satisfied.
 \end{proof}
 
-\begin{observation}\label{lastinroot}
-Each entry in \var{last} is the \var{index} of a block that has appeared in the root's RBT.
-\end{observation}
-\begin{proof}
-Only line \ref{setlast1} and \ref{setlast2} in \op{FindResponse} write values into \var{last}.
-The value written is either $b$ or $b_e \leq b$.
-By the precondition of \op{FindResponse}, $\var{root}.\fld{blocks}[b]$ exists, so the index $b$ has appeared
-in the root's RBT.
-By \Cref{RBTinsertions}, all values less than $b$ have appeared in the root's RBT too.
-\end{proof}
+Since each RBT starts with a single block with \fld{index} 0, the following is an easy consequence of \Cref{RBTupdates}.
+\begin{corollary}\label{nonEmptyIncreasing}
+The RBT stored in each node $v$ is never empty and always stores a set of blocks with consecutive indices.
+Moreover, its maximum \fld{index} can only increase over time.
+\end{corollary}
 
-\begin{observation}
-\label{edirGC} 
-If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
-$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
-\end{observation}
-\begin{proof}
-The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
-on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
-\end{proof}
-
-\begin{invariant}\label{nonemptyConsec}
-For each RBT $T$, 
-$T$ is non-empty and the blocks of $T$ have consecutive \fld{index} fields.
-\end{invariant}
-\begin{proof}
-The claim is true initially, since each RBT initially contains only the empty block with \fld{index} 0.
-We show that each time a new RBT is constructed, it preserves the invariant.
-When a new RBT is created by inserting an element into an existing RBT, \Cref{RBTinsertions}
-implies that the invariant is preserved.
-
-Suppose a new RBT is created by a \op{Split}($T,t$) operation $S$ on at line \ref{split} of a \op{Discard}($v,t$), 
-and that the invariant held prior to $S$.
-Clearly, $S$ preserves the property that indices are consecutive.  To show that the new tree
-is non-empty we first prove that the call to \op{Discard}($v,t$) has a parameter $t$ that is at least as big as a value previously
-read from the index field of some block in $v$'s RBT.
-If \op{Discard} was called from line \ref{discard}, then $t+1$ was read from an entry of \var{last}, so the claim follows from \Cref{lastinroot}.
-If \op{Discard} was called from line \ref{discardrecurse1} or \ref{discardrecurse2}, the claim follows
-from \Cref{edirGC}.
-Since the invariant held prior to $S$, no split before $S$ emptied $v$'s tree.  Therefore,
-some block with index at least $t$ is still in $T$ when $S$ occurs.  That block will be in the new tree created by $S$.
-\end{proof}
-
-Since RBTs are always non-empty, calls to \op{MaxBlock} are well-defined.  
+Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
 Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
 in $v$'s tree at some time during the execution.
-It follows from \Cref{RBTinsertions} and \Cref{nonemptyConsec} that this block is unique, if 
-it exists.
+It follows from \Cref{RBTupdates} and \Cref{nonemptyConsec} that 
+each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
+that has appeared in $v$'s RBT earlier.
+Thus, $v.\blocks[b]$ is unique, if it exists.
 We also use this notation in the code to indicate that a search of the RBT should be performed for the block
 with \fld{index} $b$.
-We describe later what should happen if the block is not found by the search because it has been removed
+We describe later what should happen if that block is not found by the search because it has been removed
 by a GC phase.
@@
 
-\begin{corollary}\label{maxNonDecreasing}
-The maximum \fld{index} in $v$'s RBT is non-decreasing over time.
-\end{corollary}
+We now establish that definition (\ref{defsubblock}) of a block's subblocks
+still makes sense.
+
+%\here{If next obs is only used once, can fold it into the proof that needs it}
+%\begin{observation}
+%\label{edirGC} 
+%If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
+%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
+%\end{observation}
+%\begin{proof}
+%The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
+%on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
+%\end{proof}
+
+We next prove the analogue of Lemma \ref{lem::headProgress}.
+\begin{customlemma}{$4'$}\label{lem::headProgressGC}
+ If $h>0$ and a block with index $h$ has been inserted into $v.\fld{blocks}$ then 
+ $v.\fld{blocks}[h-1].\fld{end\sub{left}} \leq v.\fld{blocks}[h].\fld{end\sub{left}}$ and 
+ $v.\fld{blocks}[h-1].\fld{end\sub{right}} \leq v.\fld{blocks}[h].\fld{end\sub{right}}$.
+\end{customlemma}
 \begin{proof}
-The CAS steps (on line \ref{casGC} and \ref{splitCASGC}) that update $v.\fld{blocks}$ always replace 
-a tree $T$ by a tree that is built from $T$ by inserting an element and/or removing the elements with 
-the smallest \fld{index} values, without emptying the tree.  Thus, the maximum \fld{index} in the tree cannot decrease.
-\end{proof}
-
-We now prove the analogue of Lemma \ref{lem::headProgress}.
-\begin{lemma}
- If $b>0$ and a block with index $b$ has been inserted into $v.\fld{blocks}$ then 
- $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
- $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
-\end{lemma}
-\begin{proof}
-The block $B$ with index $b$ was installed into $v$'s RBT by the CAS at line \ref{casGC} or \ref{splitCASGC} 
-that changed the tree from $T$ to $T_{new}$.
-Before this CAS, the tree $T$ was read from $v.blocks$ at line \ref{readTGC} or \ref{readT2GC}, respectively,
-before the \op{RefreshUpdate} on line \ref{RU1} or \ref{RU2} that inserted $B$ to get $T_{new}$.
-Since the previous block $B'=v.\fld{blocks}[b-1]$ was already in $T$ (by \Cref{RBTinsertions}),
-the \op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the \op{CreateBlock}$(v,b)$ that created $B$ started.
-It follows from \Cref{maxNonDecreasing} that the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.
+The block $B$ with index $h$ was installed into $v$'s RBT by the CAS at line \ref{casGC} 
+that changed the tree from $T$ to $T'$.
+Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $v.blocks$,
+line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
+and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
+Since $B'$ was already in $T$ before $B$ was created, the
+\op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the 
+\op{CreateBlock}$(v,b)$ that created $B$ started.
+By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ 
+stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
+of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  
 Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
@@
 \Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
-hold true for the modified algorithm.  Their proofs are identical to those given in 
-Section \ref{sec::basicProperties} since they only depend on \Cref{lem::headProgress} 
+hold for the modified algorithm.  Their proofs are identical to those given in 
+Section \ref{sec::basicProperties} since they  depend only on \Cref{lem::headProgress} (which can be replaced by \Cref{lem::headProgressGC})
 and the definition of subblocks given in (\ref{defsubblock}).
-The following pair of lemmas are analogous to \Cref{successfulRefresh} for the modified \op{Refresh} routine.
-We say that an operation has been \emph{propagated} to node $v$ in configuration $C$ if
-some block containing the operation has been in $v.blocks$ at or before $C$.
+Next, we prove an analogue of \Cref{successfulRefresh}.
+We say a node $v$ \emph{contains} an operation
+if some block containing the operation has previously appeared in $v$'s RBT (even if 
+the block has been removed from the RBT by a subsequent \op{Split}).
 
-\begin{lemma}\label{nullRefreshGC}
-Suppose $v.\fld{blocks}=T$ at some configuration $C$ and \op{RefreshUpdate}($T,v$) is invoked after $C$ and
-returns \nl.  Then, all operations that were propagated to $v$'s
-children before $C$ have been propagated to $v$ before $C$.
-\end{lemma}
+\begin{customlemma}{$9'$}
+\label{successfulRefreshGC}
+Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
+In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
+when $R$ executed line~\ref{refreshReadTGC}.
+\end{customlemma}
 \begin{proof}
-Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $C$.
-Then, for some $i$, a block $B$ with \fld{index} $i$ has been in $v.\fld{left.blocks}$ before $C$.
-By Lemma \ref{maxNonDecreasing}, 
-the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ at all times during the \op{RefreshUpdate}.
-So, the value of $\var{new}.\eleft$ chosen at line \ref{createEndLeftGC} of the \op{RefreshUpdate}'s call
-to \op{CreateBlock} is at least $i$.
+Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
+when $R$ executed line \ref{refreshReadTGC}.
+Let $i$ be the index of the block containing $op$ that was in $v$'s RBT before 
+$R$ executed line \ref{refreshReadTGC}.
+We consider two cases.
 
-Let $h$ be the value set at line \ref{refreshReadMax} of the \op{RefreshUpdate}.
-The \op{RefreshUpdate} returns \nl\ at line \ref{addOPGC}
-because the \op{CreateBlock}$(v,h)$  at line \ref{CBnullGC} returns \nl.
-By reasoning identical to the last paragraph of \Cref{successfulRefresh},
-it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ that
-the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
+Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
+installs $B'$ in $v.\fld{blocks}$.
+The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $v.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
+By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
+By the definition of subblocks, some block in $v$ contains $B$ as a subblock
+and therefore $v$ contains $op$.
+
+Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOPGC}.
+Let $h$ be the maximum \fld{index} in $T$ plus 1.
+By reasoning identical to the last paragraph of \Cref{successfulRefresh}'s proof,
+it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
+that the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
 $v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
 contain no operations.
-Since each block contains at least one operation by \Cref{blockNotEmpty}
-these ranges must be empty, so $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
-This implies that the block $B$ containing $op$ is a subblock of some block that has appeared in $v.\fld{blocks}$, so $op$ has been propagated to $v$.
-\end{proof}
-
-\begin{lemma}\label{successfulRefreshGC}
-If a \op{CAS} at line \ref{casGC} or \ref{splitCASGC} modifies $v.\fld{blocks}$ from $T$ to $T'$,
-then all operations that were propagated to $v$'s children before $v.\fld{blocks}$ was set to $T$
-have been propagated to $v$ in the configuration after the CAS.
-\end{lemma}
-\begin{proof}
-Suppose an operation $op$ has been propagated to $v$'s child (without loss of generality, $v.\fld{left}$) before $v.\fld{blocks}$ was set to $T$.
-Then, for some $i$, a block $B$ with \fld{index} $i$ that contains $op$
-has been in $v.\fld{left.blocks}$ before $v.\fld{blocks}$ was set to $T$.
-Let $c$ be the \op{CAS} instruction that changes $v.\fld{blocks}$ from $T$ to $T'$.
-If line \ref{casGC} performs $c$, then the \op{RefreshUpdate} $R$ at line \ref{RU1} that creates
-$T'$ began after $T$ was read from $v.\fld{blocks}$ at line \ref{readTGC}.
-Otherwise, line \ref{splitCASGC} performs $c$ and the \op{RefreshUpdate} $R$ at line \ref{RU2},
-which is invoked after $T$ was read from $v.\fld{blocks}$ at line \ref{readT2GC},
-either returned \nl\ or $T'$.
-
-If the \op{RefreshUpdate} $R$ returned \nl, the lemma follows from \Cref{nullRefreshGC}.
-So, assume \op{RefreshUpdate} $R$ returned $T'$.
-Then $T'$ contains a newly inserted block created by the call
-to \op{CreateBlock} on line \ref{invokeCreateBlockGC}.
-Since $R$ was invoked after $T$ was stored in $v.\fld{blocks}$, 
-the maximum \fld{index} in $v.\fld{left.blocks}$ is at least $i$ throughout $R$, by \Cref{maxNonDecreasing}.
-Thus. the \eleft\ field of the new block added to $T'$ by $R$'s call to \op{CreateBlock} is
-set to a value greater than or equal to $i$ at line \ref{createEndLeftGC}.
-By the definition of subblocks, it follows that the block $B$ containing $op$ is a subblock of a block
-that has been in $v.blocks$, as required.
+By \Cref{blockNotEmpty}, each block contains at least one operation, so
+these ranges must be empty, and $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
+This implies that the block $B$ containing $op$ is a subblock of some block that has appeared
+in $v$'s RBT, so $op$ is contained in $v$.
 \end{proof}

### 2023-01-08 podc_paper/correctness.tex
working on correctness of GC version

@@
 
-Now suppose $v$ is an internal node. By the definition of subblocks in (\ref{defsubblock}), the
+Now suppose $v$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
 subblocks of $v.\fld{blocks}[1..i]$ are $v.\fld{left.blocks}[1..B.\eleft]$ 
@@
 Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
-in the configuration $C$ immediately before $R$ executed line \ref{readHead}.
-Then for some $i$, the block $B$ in $v.\fld{left.blocks}[i]$ contains $op$.
+when $R$ executed line \ref{readHead}.
+Let $i$ be the index such that the block $B=v.\fld{left.blocks}[i]$ contains $op$.
 By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
@@
 in $v.\fld{left.blocks}$.
-Hence $B'$ contains $op$, as required.
+Hence $v$ contains $op$, as required.
 

### 2023-01-08 podc_paper/correctness-2.tex
working on GC section

@@
 We next prove the analogue of Lemma \ref{lem::headProgress}.
-\begin{customlemma}{$4'$}\label{lem::headProgressGC}
+\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
  If $h>0$ and a block with index $h$ has been inserted into $v.\fld{blocks}$ then 
@@
 
-\begin{customlemma}{$9'$}
+\begin{customlemma}{\ref{successfulRefresh}$'$}
 \label{successfulRefreshGC}
@@
 
-\begin{lemma}\label{lem::doubleRefreshGC}
-Consider two consecutive terminating calls $R_1$, $R_2$ to \op{GCRefresh}($v$) by the same process.
-All operations propagated to $v$'s children when $R_1$ begins
-are propagated to $v$ when $R_2$ terminates.
-\end{lemma}
+\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
+Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
+All operations contained $v$'s children when $R_1$ begins
+are contained in $v$ when $R_2$ terminates.
+\end{customlemma}
 \begin{proof}
+If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
+from \Cref{successfulRefreshGC}.
+So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
+\here{To be continued....}
+
 \end{proof}
 
+\here{next, should probably show that blocks are removed only after they are no longer needed; i.e., every dequeue has its answer (either returned, or written in its response field) and every enqueue has a matching dequeue that has either terminated or has a response written.}
+
 -------
+
+\subsection{Analysis}
+
+We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
+
+\begin{lemma}
+If the maximum \fld{index} in a node's RBT is a multiple of $p^2$, then
+it contains $O(q_{max}+p)$ blocks.
+\end{lemma}
+
+\begin{corollary}\label{RBTbound}
+At all times, the size of a node's RBT is $O(q_{max}+p^2)$. 
+\end{corollary}
+\begin{proof}
+
+\end{proof}
+
+\begin{proposition}
+The amortized step complexity is $O(\log p \log(p+q_{max}))$ per operation.
+\end{proposition}
+\begin{proof}
+It follows from \Cref{RBTbound}
+that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
+blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.
+
+First, we bound the number of steps taken \emph{excluding} the GC phase in line \ref{GCstart}--\ref{GCend}.
+An \op{Enqueue} or null \op{Dequeue} does $O(1)$ RBT operations and other work at each level of the tree during \op{Propagate},
+for a total of $O(\log p \log(p+q_{max}))$ steps.
+A non-null \op{Dequeue} must also search for a block in the root at line \ref{FRsearchGC}
+and call \op{GetEnqueue}, which does $O(1)$ RBT operations (including a search at line \ref{getChildGC}) and other work at each level of the tree.
+Thus, a \op{Dequeue} also takes $O(\log p \log(p+q_{max}))$ steps.
+
+Now we consider the additional steps a process takes to do GC in line \ref{GCstart}--\ref{GCend}.
+\here{To be continued}
+
+\end{proof}
 

### 2023-01-08 podc_paper/algorithm-2.tex
working on GC section
removed propagate from helping in GC code

@@
 \op{Dequeue} has been similarly modified, and the second half of the routine is now in a
-separate routine called \op{CompleteDequeue} so that it can also be used by other processes
+separate routine called \op{CompleteDeq} so that it can also be used by other processes
 helping to complete the operation.
@@
 Once a dequeue has been propagated to the root, 
-\bf{\op CompleteDequeue} computes its answer.  This routine can also be called by helpers 
+\bf{\op CompleteDeq} computes its answer.  This routine can also be called by helpers 
 who want to help complete the dequeue.
@@
     \State \Call{Append}{\var{B}}
-    \gc{\State \Return \Call{CompleteDequeue}{\var{leaf}, $h$}}
+    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}, $h$}}
 \EndFunction{Dequeue}
@@
 \gc{
-\Function{Object}{CompleteDequeue}{\var{leaf}, $h$}
+\Function{Object}{CompleteDeq}{\var{leaf}, $h$}
     \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
@@
 	\State \Return{\var{res}}
-\EndFunction{CompleteDequeue}
+\EndFunction{CompleteDeq}
 }
@@
     \If{\var{B}.\fld{index} is a multiple of $p^2$}
-    	\State \linecomment{Do garbage collection}
+    	\State \linecomment{Do garbage collection}\label{GCstart}
 		\State $m \assign 0$
@@
         \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
-        \State \Return \Call{Insert}{$T',B$}
+        \State \Return \Call{Insert}{$T',B$}\label{GCend}
 	\Else\ \Return \Call{Insert}{$T, B$}
@@
 
-\Function{void}{Propagate}{\typ{Node} \var{v}}
-    \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
-    \EndIf
-    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
-        \State \Call{Propagate}{\var{v}.\fld{parent}}
-    \EndIf
-\EndFunction{Propagate}
-
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{248}
 \Function{boolean}{Refresh}{\typ{Node} \var{v}}
@@
 \EndFunction{Refresh}
+
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{248}
+\Function{void}{Propagate}{\typ{Node} \var{v}}
+    \State \linecomment{propagate blocks from \var{v}'s children to root}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
+    \EndIf
+    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
+        \State \Call{Propagate}{\var{v}.\fld{parent}}
+    \EndIf
+\EndFunction{Propagate}
 
@@
 		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$
+		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
 		\State \linecomment{find rank of enqueue within its block}
@@
         \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
-    	\State \Call{Propagate}{$\ell.\fld{parent}$}
-        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$} 
-            \State \linecomment{operation is a  dequeue}
-            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDequeue}{\ell, \var{B}.\fld{index}}$ 
+    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
+        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
+        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
+            \State \linecomment{operation is a propagated dequeue}
+            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
         \EndIf
@@
 }
+
+\spac
+
+\gc{
+\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
+    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
+    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
+    \If {$\var{v} = \var{root}$} \Return{true}
+    \ElsIf {$\var{v}.\fld{super} = \nl$ or 
+    			\State \hspace*{4mm}$\var{v}.\fld{parent.blocks}[\var{v}.\fld{super}] = \nl$} 
+		\State\Return{\fa}
+    \Else
+    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} 
+				\State\var{sup} \assign\ $\var{sup}+1$
+	    \EndIf
+	    \State \hangbox{\Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and\\
+	    		 $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and\\
+			 	 \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}}
+    \EndIf
+\EndFunction{Propagated}
+}
+
 \end{algorithmic}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{318}
+\setcounter{ALG@line}{358}
 
@@
 \end{minipage}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{335}
+
+\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
+	\gc{
+	\If{$b > \var{v}.\fld{last}[\mbox{process id}]$}
+		\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
+	\EndIf
+	}
+    \If{\var{v} is a leaf node}
+        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
+    \Else 
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
+            \State \fld{dir} \assign\ \fld{left}
+        \Else
+            \State \fld{dir} \assign\ \fld{right}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
+        \EndIf
+        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
+        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
+$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
+        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
+        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
+    \EndIf
+\EndFunction{GetEnqueue}
+
+\end{algorithmic}
 
@@
 \begin{figure}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{340}
 
-\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
-	\gc{
-	\If{$b > \var{v}.\fld{last}[\mbox{process id}]$}
-		\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
-	\EndIf
-	}
-    \If{\var{v} is a leaf node}
-        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
-    \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
-            \State \fld{dir} \assign\ \fld{left}
-        \Else
-            \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
-        \EndIf
-        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
-        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
-$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
-        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
-        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$}
-    \EndIf
-\EndFunction{GetEnqueue}
-
-\end{algorithmic}
 \caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }

### 2023-01-08 podc_paper/correctness-2.tex
removed propagate from helping in GC code

@@
 Now we consider the additional steps a process takes to do GC in line \ref{GCstart}--\ref{GCend}.
-\here{To be continued}
-
+For each integer $r$, each process does GC on each node $v$ with a block whose \fld{index} is $rp^2$ at most
+once, since either its CAS succeeds, or another processes CAS advances the maximum 
+Each such helping phase takes $O(p)$ time to read the \var{last} array,
+$O(p \log p \log(p+q_{max}))$ time to \op{Help} (excluding any GC phases within \op{Help}, which will be accounted for at the other nodes),
+and \here{to be continued}
 \end{proof}

### 2023-01-08 podc_paper/correctness-2.tex
working on analysis of GC version

--- 2023-01-08 podc_paper/correctness-2.tex
@@
 We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
+Recall that GC is done on a node every $G$ times its RBT is updated, where we chose $G$ to be $p^2\ceil{\log p}$.
 
-\begin{lemma}
-If the maximum \fld{index} in a node's RBT is a multiple of $p^2$, then
+\begin{lemma}\label{boundAfterGC}
+If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
 it contains $O(q_{max}+p)$ blocks.
 \end{lemma}
+\begin{proof}
+\here{to be written}
+\end{proof}
 
 \begin{corollary}\label{RBTbound}
-At all times, the size of a node's RBT is $O(q_{max}+p^2)$. 
+At all times, the size of a node's RBT is $O(q_{max}+p+G)$. 
 \end{corollary}
 \begin{proof}
+Each update to a node's RBT adds at most one block to it, increasing its maximum \fld{index} by 1.
+Thus, there are at most $G$ updates since the last time its maximum \fld{index} was a multiple of $G$.
+The claim follows from \Cref{boundAfterGC}.
+\end{proof}
 
+\begin{theorem}\label{spaceBound}
+The size of our queue data structure is $O(pq_{max}+p^3\log p)$ words of memory.
+\end{theorem}
+\begin{proof}
+There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(p)$ memory for the
+\fld{last} array.  So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
+to be $p^2\ceil{\log p}$.
 \end{proof}
+
+Although individual operations may now become more expensive because they have to perform GC,
+we show that operations still have polylogarithmic amortized step complexity.
 
@@
 \begin{proof}
-It follows from \Cref{RBTbound}
+It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
 that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
@@
 A non-null \op{Dequeue} must also search for a block in the root at line \ref{FRsearchGC}
-and call \op{GetEnqueue}, which does $O(1)$ RBT operations (including a search at line \ref{getChildGC}) and other work at each level of the tree.
+and call \op{GetEnqueue}.  At each level of the tree, \op{GetEnqueue} does $O(1)$ RBT operations (including a search at line \ref{getChildGC}) and $O(1)$ other steps.
 Thus, a \op{Dequeue} also takes $O(\log p \log(p+q_{max}))$ steps.
 
-Now we consider the additional steps a process takes to do GC in line \ref{GCstart}--\ref{GCend}.
-\here{To be continued}
+Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}.
+If a process does GC in a call to \op{AddBlock}($v,T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on $v$}.
+We argue that each process $P$ can do an $r$th helping phase on $v$ at most once as follows.
+Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on $v$.
+Let $v,T,B$ be the arguments of $A$.
+Any call to \op{AddBlock} on internal node $v$ is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
+is the maximum \fld{index} in $T$ plus 1.
+The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
+or it fails because 
+some other \op{CAS} changes $v.blocks$ from $T$ to another tree.
+Either way, by \Cref{RBTupdates}, $v$'s RBT's maximum \fld{index} will be at least $rG$ at all times
+after this CAS.
+So if a subsequent \op{Refresh} by process $P$ 
+ever calls \op{AddBlock} on $v$ again, the block it passes as the third argument
+will have $\index>rG$, so $P$ will not perform an $r$th helping phase on $v$ again.
 
+Each helping phase takes $O(p)$ steps to read the \var{last} array,
+$O(p \log p \log(p+q_{max}))$ steps to \op{Help},
+followed by $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
+Thus, for each integer $r$ and each node $v$, a total of $O(p^2\log p\log(p+q_{max}))$ steps
+are performed by all processes during their $r$th helping phase on $v$.
+We can amortize these steps over the operations that appear in 
+$v.blocks[(r-1)p^2+1..p^2]$.
+By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
+so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
+is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
+Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$, which is
+within a constant factor of the steps it performs excluding GC.
 \end{proof}
+
+============
 

### 2023-01-09 podc_paper/boundedspace.tex
working on bounds for GC; discovered problem with using last array inside each node

@@
 To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
-If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
-of $p^2$, it does GC.
-To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
+If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
+To determine which blocks can be thrown away, the node has an array $\var{last}[1..p]$ where 
 each process writes the index of the last block in
-\var{root} containing an enqueue whose element it dequeued.
+containing an enqueue whose element it dequeued.
 To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
@@
 Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
-in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
+in $\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the RBT 
 with indices smaller than $m-1$.
@@
 all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
-Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
-of the discarded root blocks.
-
-\here{Have to talk about repeating CAS to do GC at each node up to p times}
 
@@
 Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
-to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
+to omit polylogarithmic factors; see Appendix \ref{sec::GCanalysis} for precise bounds.}
 Since all processes may run GC,
-a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
-There are at least $p^2$ operations between two GC phases,  so
+a total of $\tilde{O}(G)$ steps is spent on a GC phase.
+There are at least $G$ operations between two GC phases,  so
 GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.

### 2023-01-09 podc_paper/correctness-2.tex
working on bounds for GC; discovered problem with using last array inside each node

@@
 \subsection{Analysis}
+\label{sec::GCanalysis}
+
+\here{Danger:  what if I remove a block (by a split), which prevents another process from continuing on to mark ancestors/descendants of this node?
+Does existing helping take care of this?
+Or maybe this does not cause too many excess blocks to be left around?
+If not, maybe before splitting, we should mark descendants/ancesotrs of max marked block; in this case, all marking should go in the same direction up/down tree.}
 
@@
 Recall that GC is done on a node every $G$ times its RBT is updated, where we chose $G$ to be $p^2\ceil{\log p}$.
+
+We say the index $b$ \emph{is marked in node $v$} if some process has tested
+whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
+Intuitively, a block's index value is marked 
+when a dequeue passes through this block to find the enqueue
+whose argument it should return, or when a dequeue that returns null passes through the block.
+Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
+contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
+If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.
 
@@
 \begin{proof}
-\here{to be written}
+Consider the \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
+\fld{index} is a multiple of $G$.
+That \op{AddBlock} performs a GC phase.
+Let $C$ be the configuration after it executes line \ref{beforeLastScan}.
+Let $L'$ be the prefix of the linearization order $L$ consisting of blocks that have beed added to the
+root before $C$.
+$A$'s \op{Split} on line \ref{splitGC} removes from the RBT all blocks whose indices
+were marked in $v$ prior to line \ref{beforeLastScan}, except for $m$ and $m-1$.
+To bound the number of blocks left in the RBT after the split, 
+we get a bound on the number of operations in unmarked blocks in $v$.
+
+First consider an enqueue operation $E$ in an unmarked block in $v$.
+No dequeue that returns $E$'s value has completed (otherwise it would have marked the block in $v$).
+So, if the dequeue is in $L'$, then the dequeue is still pending at $C$.
+There can be at most $p$ dequeues pending at $C$.
+So, either $E$'s element is still in the
+
+An enqueue operation can be in an unmarked block in $v$
+only if the dequeue that returns its value has not yet completed.
+
+\here{to be continued}
 \end{proof}

### 2023-01-09 podc_paper/algorithm-2.tex
working on analysis of GC version
working on bounds for GC; discovered problem with using last array inside each node
modified code for new way of computing split point

@@
 
+\HERE{Update this parag after making changes}
 Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
@@
 In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
-it also updates \var{last} entries if the computed response is \nl.
-It does this by calling \op{IndexDequeue} again, this time with \tr\ as its final parameter.
-This extra parameter has been added to 
-\op{IndexDequeue} to indicate whether it should mark blocks that it traverses (which all contain the dequeue).
+it also updates \var{last} if the computed response is \nl.
 
@@
 
-\gc{
-\Function{Object}{CompleteDeq}{\var{leaf}, $h$}
-    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
-    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1, \fa}$
-    \State $\var{res} \assign \Call{FindResponse}{b, i}$
-	\If{$\var{res} = \nl$} \linecomment{update \var{last} values}
-		\State $\Call{IndexDequeue}{\var{leaf}, h, 1, \tr}$
-	\EndIf
-	\State \Return{\var{res}}
-\EndFunction{CompleteDeq}
-}
-
-\spac
-
 \Function{void}{Append}{\typ{Block} \var{B}} 
@@
 
-\gc{
-\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
-    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
-    \If{\var{B}.\fld{index} is a multiple of $p^2$}
-    	\State \linecomment{Do garbage collection}\label{GCstart}
-		\State $m \assign 0$
-		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
-		\EndFor
-        \State \Call{Help}{}
-        \State $T' \assign\ \Call{Split}{T,m-1}$
-        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
-        \State \Return \Call{Insert}{$T',B$}\label{GCend}
-	\Else\ \Return \Call{Insert}{$T, B$}
+\Function{void}{Propagate}{\typ{Node} \var{v}}
+    \State \linecomment{propagate blocks from \var{v}'s children to root}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
     \EndIf
-\EndFunction{AddBlock}
-}
+    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
+        \State \Call{Propagate}{\var{v}.\fld{parent}}
+    \EndIf
+\EndFunction{Propagate}
+
 \spac
@@
     	\gc{
-	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$
+	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
 		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}\label{casGC}
@@
 \EndFunction{Refresh}
+\spac
 
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{248}
-\Function{void}{Propagate}{\typ{Node} \var{v}}
-    \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
-    \EndIf
-    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
-        \State \Call{Propagate}{\var{v}.\fld{parent}}
-    \EndIf
-\EndFunction{Propagate}
-
-\spac
 \gc{
@@
 
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{250}
+\gc{
+\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
+    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
+    \If{\var{B}.\fld{index} is a multiple of $G$}
+    	\State \linecomment{Do garbage collection}\label{GCstart}
+        \State $s \assign \Call{SplitBlock}{v}.\var{index}$
+        \State \Call{Help}{}
+        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
+        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
+        \State \Return \Call{Insert}{$T',B$}\label{GCend}
+	\Else\ \Return \Call{Insert}{$T, B$}
+    \EndIf
+\EndFunction{AddBlock}
+}
 \spac
@@
 
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
-        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNullGC}
-    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
-			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
-		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
-		\State \linecomment{find rank of enqueue within its block}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
-        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
+    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
+    \Else
+	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
 	    \EndIf
-\EndFunction{FindResponse}
+	    \State \linecomment{compute index \var{i} of dequeue in superblock}
+	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
+	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
+        \If{$\fld{dir} = \fld{right}$} 
+        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
+					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
+        \EndIf \label{computeISuperEndGC}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
+    \EndIf
+\EndFunction{IndexDequeue}
 
@@
 \end{minipage}
-\caption{Bounded-space queue implementation.\label{pseudocode1GC}}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
 \end{figure}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{307}
+\setcounter{ALG@line}{300}
+\gc{
+\Function{Object}{CompleteDeq}{\var{leaf}, $h$}
+    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
+    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1}$
+    \State $\var{res} \assign \Call{FindResponse}{b, i}$
+	\State \Return{\var{res}}
+\EndFunction{CompleteDeq}
+}
+
+\spac
+
 \gc{
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{358}
+\setcounter{ALG@line}{353}
+\gc{
+\Function{Block}{SplitBlock}{\typ{Node} v}
+	\State \linecomment{figure out where to split $v$'s RBT}
+	\If{$v=\var{root}$}
+		\State $m \assign 0$
+		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
+		\EndFor
+		\State $B\assign \var{root}.\fld{blocks}[m-1]$
+	\Else
+		\State $B_p \assign \Call{SplitBlock}{v.parent}$
+		\State \var{dir} \assign ($v=v.\fld{parent.left}$ ? \fld{left} : \fld{right})
+		\State $B \assign v.\fld{blocks}[B_p.\edir]$
+	\EndIf
+	\State \Return ($B = \nl$ ? \Call{MinBlock}{v.\var{blocks}} : $B$)
+\EndFunction{SplitBlock}
+}
 
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}, \gc{\typ{boolean} \var{mark}}}
-    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
-    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+\spac
+
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
+    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
+    	\State \linecomment{ queue is empty when dequeue occurs}
         \gc{
-    \If{\var{mark} and $b>\var{v}.\fld{last}[\mbox{process id}]$}
-    	\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
+        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
         \EndIf
         }
-    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
-    \Else
-	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
+        \State \Return \nl \hfill \label{returnNullGC}
+    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
+        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
+		\State \linecomment{compute enqueue's block using binary search}
+		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
+		\State \linecomment{find rank of enqueue within its block}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
+		\gc{
+        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+        \If {$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
         \EndIf
-	    \State \linecomment{compute index \var{i} of dequeue in superblock}
-	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
-	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
-        \If{$\fld{dir} = \fld{right}$} 
-        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
-					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
-        \EndIf \label{computeISuperEndGC}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
+        \State \Return \var{res}
+		}
     \EndIf
-\EndFunction{IndexDequeue}
+\EndFunction{FindResponse}
 
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{335}
+\setcounter{ALG@line}{333}
 
@@
     \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
-	\gc{
-	\If{$b > \var{v}.\fld{last}[\mbox{process id}]$}
-		\State $\var{v}.\fld{last}[\mbox{process id}] \assign b$
-	\EndIf
-	}
     \If{\var{v} is a leaf node}
@@
 %\end{algorithm}
-
-\begin{figure}
-
-\caption{\label{pseudocode3GC}Bounded-space queue implementation, continued. }
-\end{figure}

### 2023-01-09 podc_paper/algorithm-2.tex
got the bound on the size of the RBTs

@@
 
-The space-bounded implementation uses an additional shared array so that processes can access one another's leaves to perform helping.
+The space-bounded implementation uses two additional shared arrays:
+the \var{leaf} array allows processes to access one another's leaves to perform helping, and
+the \var{last} array is used to determine which blocks are safe to discard.
 \begin{itemize}
 \item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
+\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the root  that process $k$}\\
+\mbox{ }\Comment{has observed to contain either an enqueue whose value has}\\
+\mbox{ }\Comment{been dequeued or a null dequeue }
 \end{itemize}
@@
 \end{itemize}
-Each \typ{Node} has an additional array field to help determine which of its blocks are safe to discard.
-\begin{itemize}
-\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the node that process $k$ }\\
-\mbox{ }\Comment{has accessed, either to find the enqueued item for its dequeue}\\
-\mbox{ }\Comment{or during a \nl\ dequeue}
-\end{itemize}
 The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
@@
 can be done in constant time, without affecting the time of other RBT operations.
+Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.
 
@@
 
-\HERE{Update this parag after making changes}
-Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
+\here{Update this parag after code is finalized--it has changed since I wrote this parag}
+Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode2GC}.
 The \op{Propagate} and \op{FindResponse} routines are unchanged.
@@
     	\State \linecomment{Do garbage collection}\label{GCstart}
-        \State $s \assign \Call{SplitBlock}{v}.\var{index}$
+        \State $s \assign \Call{SplitBlock}{v}.\var{index}$\label{callSplitBlock}
         \State \Call{Help}{}
@@
 	\EndIf
-	\State \Return ($B = \nl$ ? \Call{MinBlock}{v.\var{blocks}} : $B$)
+	\State \linecomment If $B$ was already discarded, use leftmost block instead
+	\State \Return ($B = \nl$ ? $\Call{MinBlock}{v.\var{blocks}}$ : $B$)
 \EndFunction{SplitBlock}

### 2023-01-09 podc_paper/correctness-2.tex
got the bound on the size of the RBTs

--- 2023-01-08 podc_paper/correctness-2.tex
@@
 in $v$'s tree at some time during the execution.
-It follows from \Cref{RBTupdates} and \Cref{nonemptyConsec} that 
+It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
 each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
@@
 \here{next, should probably show that blocks are removed only after they are no longer needed; i.e., every dequeue has its answer (either returned, or written in its response field) and every enqueue has a matching dequeue that has either terminated or has a response written.}
-
+one invariant will be that a null dequeue or a dequeued enqueue appears in the first block (or some later block) of each RBT.
+should imply that no operation will need any block before the first block in the RBT
 -------
@@
 \subsection{Analysis}
+\label{sec::GCanalysis}
+
+\here{need to prove a lemma analogous to \Cref{blockSize}}
 
 We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
-Recall that GC is done on a node every $G$ times its RBT is updated, where we chose $G$ to be $p^2\ceil{\log p}$.
+Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
+
+%We say the index $b$ \emph{is marked in node $v$} if some process has tested
+%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
+%Intuitively, a block's index value is marked 
+%when a dequeue passes through this block to find the enqueue
+%whose argument it should return, or when a dequeue that returns null passes through the block.
+%Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
+%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
+%If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.
 
@@
 If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
-it contains $O(q_{max}+p)$ blocks.
+it contains at most $2q_{max}+4p+1$ blocks.
 \end{lemma}
 \begin{proof}
-\here{to be written}
+Consider the invocation $A$ of \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
+\fld{index} is a multiple of $G$.
+$A$ performs a GC phase.
+Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
+That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
+array.
+Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
+Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
+$\ell$ is the last block added to the root's RBT before $C$.
+
+We first bound the number of operations in $L_2$.
+
+The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
+Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
+must still be in progress at $C$; otherwise the process that dequeued it would have
+set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
+contradicting the definition of $m$.
+So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
+Thus, there are at most $q_{max}+p$ enqueues in $L_2$.
+
+If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
+then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
+Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
+Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
+at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
+to a value greater than the index of the block of the root that contains the dequeue prior to $C$, 
+contradicting the definition of $m$.
+So, there are at most $p$ null dequeues in $L_2$.
+Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.
+
+$A$'s call to \op{SplitBlock} determines the index $s$ used to split $v$'s RBT by following 
+\eleft\ and \eright\ pointers from the root down to $v$.
+So, the block returned is a subblock of block $m$ of the root, unless
+at some point along the path of subblocks the subblock has already been removed by a split,
+in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.
+
+Next, we bound the number of operations in $v$'s blocks that are retained when $A$ does sets
+$T'\assign\op{Split}(T,s)$ operation.
+Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
+Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
+All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
+There are at most $p$ operations in block $m$ of the root by \Cref{blockSizeGC}
+and we showed above that there are at most $2q_{max} + 2p$ in blocks 
+$m+1..\ell$ of the root.
+Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
+Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
+block is inserted before $A$ sets $v$'s \fld{blocks} to the resulting RBT.
 \end{proof}
@@
 \begin{proof}
-There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(p)$ memory for the
-\fld{last} array.  So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
+There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(1)$ memory words.
+So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
 to be $p^2\ceil{\log p}$.
@@
 \begin{proposition}
-The amortized step complexity is $O(\log p \log(p+q_{max}))$ per operation.
+The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
 \end{proposition}
 \begin{proof}
+\here{this proof may need some modification since code was changed to change how split point is computed}
 It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$

### 2023-01-09 podc_paper/correctness-2.tex
cleaning up a few things

@@
 Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
-
+Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
 %We say the index $b$ \emph{is marked in node $v$} if some process has tested
@@
 
+The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
+\here{There may be room to improve this bound by a more careful analysis since the bounds so far
+didn't use the fact that operations only appear along one path in the tree; in other words
+doing a bound on total size of all RBTs at one level of the ordering tree at once, rather than
+on each node separately, might reduce it to qlogp + poly(p).  Might also be possible to be
+more precise about how it depends on "current" size of queue rather than max size.}
+
 \begin{theorem}\label{spaceBound}
-The size of our queue data structure is $O(pq_{max}+p^3\log p)$ words of memory.
+The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
 \end{theorem}
@@
 There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(1)$ memory words.
+Each process may hold pointers to $O(1)$ RBTs that are no longer current in local variables.
 So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
@@
 
+We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
+by the height of the tree and there are no loops.
+
 ============
@@
 
-The idea above leads us to a poly-log data structure that supports throwing away all the blocks with keys smaller than an index.
-Red-black trees do this for us.
-\nf{Get(i)}, \nf{Append()} and \nf{Split(i)} are logarithmic in block trees.
-We can create a shared red-black tree just creating a new path for the operation and then using \nf{CAS} to change the root of the tree. See [this] for more.
-
-\begin{observation}
-PBRT supports poly-log operations ....
-\end{observation}
-
 \begin{lemma}
@@
 
-\begin{observation}
-Every $p^2$ block appended to the root, \nf{FreeMemory} is invoked.  
-\end{observation}
-
-To know the last block dequeied from we can implement a shared array among processes which they write the last root block they have dequeued from.
-
-\begin{lemma}
-    $\nf{Max(Last)} - \text{index of the last finished block}$ in the node $n$ is $O(p)$.
-\end{lemma}
-
-\begin{lemma}
-    After \nf{FreeMemory}, the space taken by each node of the tree is $O(p+q)$.
-\end{lemma}
-
-\begin{corollary}
-The space taken by each node of the tree is $O(p^2+q)$. The total space in the tree is $PolyLog(p+q)$.
-\end{corollary}
-
-\begin{lemma}
-  The amortized step per process for the algorithm with garbage collection is $PolyLog(p+q)$.
-\end{lemma}
-
 \begin{lemma}
@@
 \end{lemma}
-
-\here{Remark on an optimization:  instead of storing \head\ field separate from \fld{blocks}, we could store
-the maximum index of the RBT in the root of the RBT.  This would simplify things, but require modifying the proof more extensively.}

### 2023-01-09 podc_paper/correctness-2.tex
minor edits

@@
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
-\here{To be continued....}
+Then some other \op{CAS} on $v.\fld{blocks}$ succeeds between the time each \op{Refresh}
+reads $v.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
+Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
+$R_3$ must have read $v.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
+The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
+\end{proof}
+
+\Cref{lem::appendExactlyOnce} can then be proved in the same way as in Section \ref{sec::propagating}.
+We modified the way \var{super} fields are computed, so we prove an analogue of \Cref{superRelation} to show that \fld{super} fields are (nearly) accurate.
+
+\begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
+Let $B=v.\var{blocks}[b]$.
+  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
+\end{customlemma}
+\begin{proof}
 
 \end{proof}
+
+The correctness of \op{GetEnqueue} and \op{indexDequeue}, which are identical to the original version of the implementation are dependent only on the fact that we do not discard blocks needed by those routines.
+The following lemmas show this.
+
+\here{prove some lemmas about not throwing away useful information, then restate correctness lemmas about getenq and indexdeq that say they work correctly OR the operation is already propagated/complete.}
 

### 2023-01-09 podc_paper/correctness.tex
working on GC section
minor edits

@@
 
-\Eric{I think this case was missing from the proof in the thesis}
+%\Eric{I think this case was missing from the proof in the thesis}
 Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
@@
 Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
-When $R_2$ terminates, $v$ contains all operations that $v$'s children contained when $R_1$ was invoked.
+All operations contained $v$'s children when $R_1$ begins
+are contained in $v$ when $R_2$ terminates.
 \end{lemma}
@@
 It is invoked either 
-at line \ref{helpAdvance} after seeing $v.\fld{blocks}[h]\neq \nl$ at line \ref{ifHeadnotNull}
-or at line \ref{advance} after ensuring $v.\fld{blocks}[h]\neq \nl$ at line~\ref{cas}.
+at line \ref{helpAdvance} after seeing $v.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
+or at line \ref{advance} after ensuring $v.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
 Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.

### 2023-01-09 podc_paper/algorithm-2.tex
cleaning up a few things
minor edits
got rid of super field in GC version

@@
 
-\Function{void}{Append}{\typ{Block} \var{B}} 
-    \State \linecomment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
-    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
-\EndFunction{Append}
-
-\spac
-
-\Function{void}{Propagate}{\typ{Node} \var{v}}
-    \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
-    \EndIf
-    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
-        \State \Call{Propagate}{\var{v}.\fld{parent}}
-    \EndIf
-\EndFunction{Propagate}
-
-\spac
-
-\Function{boolean}{Refresh}{\typ{Node} \var{v}}
-	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
-	\gc{
-	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
-    \State \Call{SetSuper}{$v$, $T$}
-    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
-    }
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\nl} \label{addOPGC}
-    \Else  
-    	\gc{
-	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
-		\State \Return{\Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}}\label{casGC}
-		}
-    \EndIf
-
-\EndFunction{Refresh}
-\spac
-
-\gc{
-\Function{void}{SetSuper}{\typ{Node} \var{v}, \typ{RBT} $T$} 
-	\State \linecomment{set \fld{super} field of last block in $T$}
-	\State \linecomment{and in $v$'s children's RBTs}
-	\State $\var{lastBlock} \assign \Call{MaxBlock}{T}$
-	\If{$v\neq \var{root}$}	
-		\State \var{sup} \assign\ \Call{MaxBlock}{\var{v}.\fld{parent.blocks}}.\fld{index}
-		\State \Call{CAS}{\var{lastBlock}.\fld{super}, \nl, \var{sup}}
-	\EndIf
-	\ForEach{\fld{dir} in \{\fld{left}, \fld{right}\}} \label{startHelpChild1GC}
-    	\State \var{lastChildBlock} \assign\ \Call{MaxBlock}{\var{v}.\fld{dir}.\fld{blocks}}
-		\State \Call{CAS}{\var{lastChildBlock}.\fld{super}, \nl, \var{lastBlock}.\fld{index}}
-    \EndFor \label{endHelpChild1GC}
-\EndFunction{SetSuper}
-}
-
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{250}
-\gc{
-\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
-    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
-    \If{\var{B}.\fld{index} is a multiple of $G$}
-    	\State \linecomment{Do garbage collection}\label{GCstart}
-        \State $s \assign \Call{SplitBlock}{v}.\var{index}$\label{callSplitBlock}
-        \State \Call{Help}{}
-        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
-        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
-        \State \Return \Call{Insert}{$T',B$}\label{GCend}
-	\Else\ \Return \Call{Insert}{$T, B$}
-    \EndIf
-\EndFunction{AddBlock}
-}
-\spac
-
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
-    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
-    \gc{
-    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
-    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
-    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
-    }
-	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
-	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
-    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
-    \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
-        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}\label{CBnullGC}
-    \Else
-        \State \Return \var{new}
-    \EndIf
-\EndFunction{CreateBlock}
-
-\spac
-
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
-    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
-    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
-    \Else
-	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$
-	    \EndIf
-	    \State \linecomment{compute index \var{i} of dequeue in superblock}
-	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
-	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}
-        \If{$\fld{dir} = \fld{right}$} 
-        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
-					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
-        \EndIf \label{computeISuperEndGC}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
-    \EndIf
-\EndFunction{IndexDequeue}
-
-\end{algorithmic}
-\end{minipage}
-\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
-\end{figure}
-
-\begin{figure}
-\begin{minipage}[t]{0.465\textwidth}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{300}
 \gc{
@@
 
+\Function{void}{Append}{\typ{Block} \var{B}} 
+    \State \linecomment{append block to leaf and propagate to root}
+    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
+    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
+\EndFunction{Append}
+
+\spac
+
 \gc{
-\Function{void}{Help}{}
-    \State \linecomment{help pending operations}
-
-    \For{$\ell$ in $\var{leaf}[1..k]$}
-        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
-    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
-        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
-        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
-            \State \linecomment{operation is a propagated dequeue}
-            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
+\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
+    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
+    \If{\var{B}.\fld{index} is a multiple of $G$}
+    	\State \linecomment{Do garbage collection}\label{GCstart}
+        \State $s \assign \Call{SplitBlock}{\var{v}}.\var{index}$\label{callSplitBlock}
+        \State \Call{Help}{}
+        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
+        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
+        \State \Return \Call{Insert}{$T',B$}\label{GCend}
+	\Else\ \Return \Call{Insert}{$T, B$}
     \EndIf
-    \EndFor
-\EndFunction{Help}
+\EndFunction{AddBlock}
 }
@@
 
-\gc{
-\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
-    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
-    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
-    \If {$\var{v} = \var{root}$} \Return{true}
-    \ElsIf {$\var{v}.\fld{super} = \nl$ or 
-    			\State \hspace*{4mm}$\var{v}.\fld{parent.blocks}[\var{v}.\fld{super}] = \nl$} 
-		\State\Return{\fa}
-    \Else
-    	\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-	    \State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\edir$} 
-				\State\var{sup} \assign\ $\var{sup}+1$
-	    \EndIf
-	    \State \hangbox{\Return{$\var{v}.\fld{parent.blocks}[\var{sup}] \neq \nl$ and\\
-	    		 $\var{b} \leq \var{v}.\fld{parent.blocks}[\var{sup}].\edir$ and\\
-			 	 \Call{Propagated}{\var{v}.\fld{parent}, \var{sup}}}}
-    \EndIf
-\EndFunction{Propagated}
-}
-
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{353}
 \gc{
@@
 
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{249}
+
+\Function{void}{Propagate}{\typ{Node} \var{v}}
+    \State \linecomment{propagate blocks from \var{v}'s children to root}
+    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
+    \EndIf
+    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
+        \State \Call{Propagate}{\var{v}.\fld{parent}}
+    \EndIf
+\EndFunction{Propagate}
+
 \spac
 
+\Function{boolean}{Refresh}{\typ{Node} \var{v}}
+	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
+	\gc{
+	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
+    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
+    }
+    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
+    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
+    \Else  
+    	\gc{
+	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
+		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
+		}
+    \EndIf
+
+\EndFunction{Refresh}
+
+\spac
+
+\gc{
+\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
+    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
+    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
+    \If {$\var{v} = \var{root}$} \Return{true}
+    \Else
+    	\State $T\assign \var{v}.\fld{parent.blocks}$
+		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+		\If{\Call{MaxBlock}{$T$}.\edir < b} \Return \fa
+		\Else
+			\State $B_p$ \assign min block in $T$ with $\edir \geq b$
+			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
+		\EndIf
+	\EndIf
+\EndFunction{Propagated}
+}
+
+\spac
+
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
+    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
+    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
+    \Else
+	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
+    	\gc{
+		\State $T\assign \var{v}.\fld{parent.blocks}$
+		\State $B_p$ \assign min block in $T$ with $\edir \geq b$
+		\State $B_p'$ \assign max block in $T$ with $\edir < b$
+	    }
+	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
+	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
+	    		$\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
+        \If{$\fld{dir} = \fld{right}$} 
+        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
+					$\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
+        \EndIf \label{computeISuperEndGC}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
+    \EndIf
+\EndFunction{IndexDequeue}
+
+\end{algorithmic}
+\end{minipage}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
+\end{figure}
+
+\begin{figure}
+\begin{minipage}[t]{0.465\textwidth}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{302}
+
+\gc{
+\Function{void}{Help}{}
+    \State \linecomment{help pending operations}
+
+    \For{$\ell$ in $\var{leaf}[1..k]$}
+        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
+    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
+        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
+        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
+            \State \linecomment{operation is a propagated dequeue}
+            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
+        \EndIf
+    \EndFor
+\EndFunction{Help}
+}
+\spac
+
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
+    \State\linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
+    \gc{
+    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
+    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
+    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
+    }
+	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
+	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
+    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
+    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \If{$\var{v} = \var{root}$}
+        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \\
+        	\var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
+    \EndIf
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
+        \State \Return \nl \hfill \com\ no blocks to be propagate to \var{v}\label{CBnullGC}
+    \Else
+        \State \Return \var{new}
+    \EndIf
+\EndFunction{CreateBlock}
+
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.529\textwidth}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{352}
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
@@
 
+\spac
+
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{333}
+\setcounter{ALG@line}{332}
 

### 2023-01-09 podc_paper/algorithm-2.tex
added text description of GC algorithms in appendix, now that code has (hopefully) stabilized

@@
 \end{itemize}
+
+The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
+Each RBT is initialized with an empty block with index~0.
+Any access to an entry of the \fld{blocks} array is replaced by a search in the RBT.
+The node's \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
+longer needed; we can instead simply find the maximum \fld{index} of any block in the RBT.
+To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
+returns the block with the maximum \fld{index}.
+We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
+can be done in constant time, without affecting the time of other RBT operations.
+Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.
+
+Blocks no longer require the \fld{super} field.  This was used to quickly find the blocks 
+superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
+by searching the parent's \fld{blocks} RBT instead.
 Each \typ{Block} has an additional field.
@@
 \end{itemize}
-The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
-Each RBT is initialized with an empty block with index 0.
-The \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
-longer needed; we can instead find the maximum \fld{index} of any block in the RBT.
-To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
-returns the block with the maximum \fld{index}.
-We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
-can be done in constant time, without affecting the time of other RBT operations.
-Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.
 
 \here{Talk about what happens if a block is not found in an RBT}
-\here{Talk about \emph{marking} a block to indicate that any blocks with smaller indices will no longer
-be needed by operations that haven't started yet.}
 
-\here{Update this parag after code is finalized--it has changed since I wrote this parag}
 Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode2GC}.
-The \op{Propagate} and \op{FindResponse} routines are unchanged.
-Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
-visited by the routines to mark the blocks they visit in those nodes.
-Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
-to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
+New or modified code appears in blue.
+The \op{Propagate} and \op{GetEnqueue} routines are unchanged.
+A few lines have been added to \op{FindResponse} to update the \var{last} array
+to ensure that it stores the value described above.
+Minor modifications have also
+been made to \op{Enqueue}, \op{Dequeue}, \op{CreateBlock}, \op{Refresh} and \op{Append}
+to accommodate  the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
 of the \head\ field).
-\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
+In addition, the second half of the \op{Dequeue} routine is now in a
 separate routine called \op{CompleteDeq} so that it can also be used by other processes
 helping to complete the operation.
+The \op{Refresh} routine no longer needs to set the \fld{super} field of blocks since
+that field has been removed.
+The \op{IndexDequeue} routine, which must trace the location of a dequeue along
+a path from its leaf to the root has a minor modification to search the \fld{blocks} RBT at 
+each level instead of using the \fld{super} field.
 
-Once a dequeue has been propagated to the root, 
-\bf{\op CompleteDeq} computes its answer.  This routine can also be called by helpers 
-who want to help complete the dequeue.
-In addition to calling \op{FindResponse} and \op{IndexDequeue} to compute the response,
-it also updates \var{last} if the computed response is \nl.
+The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
+used to implement GC.
+When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
+it calls the new \opemph{AddBlock} routine.
+Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
+the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p\ceil{\log p}$.
+This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
+The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
+calls \op{Help} to help all pending dequeues that have been propagated to the root
+(to ensure that all blocks before $s$ can safely be discarded),
+uses the standard RBT \op{Split} routine \cite{Tar83}
+to remove all blocks with \fld{index} less than $s$,
+inserts the new block
+and finally performs a \op{CAS} to swing the node's \fld{blocks} pointer to the new RBT.
 
-\here{Discuss new routines AddBlock, SetSuper, Help, and largely modified Refresh}
+To determine the oldest block in a node $v$ to keep, the \opemph{SplitBlock} routine  first
+finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
+or a null dequeue using the \var{last} array.  All enqueues before this block are either
+dequeued or will be dequeued by a dequeue that is in progress.  Once those pending dequeues have been
+helped to complete, it is safe to discard any blocks in the root older than $B_{root}$, as well
+as their subblocks.
+The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in $v$
+of $B_{root}$ (or any older block in the root).
+As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $v'$
+along the path from $v$ to the root is discarded by another GC phase.
+In this case, \op{SplitBlock} uses the last subblock in $v$ of the oldest block in $v'$ instead
+(since $v'$ determined that all blocks older than that are safe to discard).
 
-\here{does RBT have to have pointer to block rather than block itself so that updates to super field work correctly?}
+The \opemph{Help} routine is fairly straightforward:  it loops through all leaves and
+helps the dequeue that is in progress there if it has already been propagated to the root.
+The \opemph{Propagated} function is used to determine whether the dequeue has propagated to the root.
 
@@
         \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
-        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
+        \State \linecomment{\op{Split} removes blocks with $\fld{index} < s$}
         \State \Return \Call{Insert}{$T',B$}\label{GCend}
@@
 	\EndIf
-	\State \linecomment If $B$ was already discarded, use leftmost block instead
+	\State \linecomment If $B$ was discarded, use leftmost block instead
 	\State \Return ($B = \nl$ ? $\Call{MinBlock}{v.\var{blocks}}$ : $B$)
@@
 		\Else
-			\State $B_p$ \assign min block in $T$ with $\edir \geq b$
+			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
 			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
@@
 		\State $T\assign \var{v}.\fld{parent.blocks}$
-		\State $B_p$ \assign min block in $T$ with $\edir \geq b$
-		\State $B_p'$ \assign max block in $T$ with $\edir < b$
+		\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper1}
+		\State $B_p'$ \assign max block in $T$ with $\edir < b$\label{searchsuper2}
 	    }
@@
         \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
-        \If {$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
+        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
         \EndIf

### 2023-01-09 podc_paper/correctness-2.tex
added text description of GC algorithms in appendix, now that code has (hopefully) stabilized

@@
 
+\Cref{lem::headProgressGC} implies that 
+the nodes of an in-order traversal of any RBT have non-decreasing values of
+$\eleft$ (and of $\eright$).
+Thus, the searches for a block based on \eleft\ or \eright\ values 
+at lines \ref{searchsuper1}, \ref{searchsuper2} and \ref{searchsuper3},
+which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.
+
 \Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
@@
 and the definition of subblocks given in (\ref{defsubblock}).
+In particular, \Cref{lem::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
+values of \fld{sum\sub{enq}}  so the searches of a RBT based on
+\fld{sum\sub{enq}} values in lines
+\ref{getChildGC} and \ref{FRsearchGC} can be done in logarithmic time.
+
 Next, we prove an analogue of \Cref{successfulRefresh}.
@@
 \label{successfulRefreshGC}
-Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
+Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
 In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
@@
 We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
-by the height of the tree and there are no loops.
+by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
 

### 2023-01-09 podc_paper/algorithm-2.tex
minor edits

@@
 
-To determine the oldest block in a node $v$ to keep, the \opemph{SplitBlock} routine  first
+To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
 finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
@@
 as their subblocks.
-The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in $v$
+The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in \var{v}
 of $B_{root}$ (or any older block in the root).
-As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $v'$
-along the path from $v$ to the root is discarded by another GC phase.
-In this case, \op{SplitBlock} uses the last subblock in $v$ of the oldest block in $v'$ instead
-(since $v'$ determined that all blocks older than that are safe to discard).
+As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
+along the path from \var{v} to the root is discarded by another GC phase.
+In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
+(since $\var{v}'$ determined that all blocks older than that are safe to discard).
 
@@
 \Function{Block}{SplitBlock}{\typ{Node} v}
-	\State \linecomment{figure out where to split $v$'s RBT}
-	\If{$v=\var{root}$}
+	\State \linecomment{figure out where to split \var{v}'s RBT}
+	\If{$\var{v}=\var{root}$}
 		\State $m \assign 0$
-		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
+		\For{$k\assign 1..p$} $m\assign \max(m, \var{v}.\fld{last}[k])$
 		\EndFor
@@
 	\Else
-		\State $B_p \assign \Call{SplitBlock}{v.parent}$
-		\State \var{dir} \assign ($v=v.\fld{parent.left}$ ? \fld{left} : \fld{right})
-		\State $B \assign v.\fld{blocks}[B_p.\edir]$
+		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$
+		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
+		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$
 	\EndIf
 	\State \linecomment If $B$ was discarded, use leftmost block instead
-	\State \Return ($B = \nl$ ? $\Call{MinBlock}{v.\var{blocks}}$ : $B$)
+	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$)
 \EndFunction{SplitBlock}
@@
 \Function{boolean}{Refresh}{\typ{Node} \var{v}}
-	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
+	\State \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
 	\gc{
-	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
+	\State $T \assign \var{v}.\fld{blocks}$\label{refreshReadTGC}
     \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}

### 2023-01-09 podc_paper/analysis.tex
minor edits; changed tournament tree to ordering tree
minor edits

@@
 Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
-\Eric{In the thesis, Lemma 52 said enqueues take $O(\log^2 p)$ steps.  Was this a typo?}
 and each non-null \op{Dequeue} takes
@@
 at line \ref{findAnswer}.
-\Eric{Modified:}
 By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
@@
 
-\Eric{I left out the amortized bound, since the worst-case bound seems stronger}
-
 \begin{corollary}

### 2023-01-09 podc_paper/correctness-2.tex
minor edits

@@
 \begin{lemma}\label{RBTupdates}
-Suppose a step of the algorithm changes $v.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
+Suppose a step of the algorithm changes $\var{v}.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
 If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
@@
 
-If line \ref{appendLeafGC} of an \op{Append} operation modifies $v.\fld{blocks}$,
-then $v$ is a leaf node, and no other process ever modifies $v.\fld{blocks}$.
+If line \ref{appendLeafGC} of an \op{Append} operation modifies $\var{v}.\fld{blocks}$,
+then \var{v} is a leaf node, and no other process ever modifies $\var{v}.\fld{blocks}$.
 $T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
@@
 
-If line \ref{casGC} of a \op{Refresh} modifies $v.\fld{blocks}$, then $v$ is an internal node.
-After reading $T$ from $v.\fld{blocks}$ at line \ref{refreshReadTGC},
+If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
+After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
 then creates the  block \var{new},
@@
 \begin{corollary}\label{nonEmptyIncreasing}
-The RBT stored in each node $v$ is never empty and always stores a set of blocks with consecutive indices.
+The RBT stored in each node \var{v} is never empty and always stores a set of blocks with consecutive indices.
 Moreover, its maximum \fld{index} can only increase over time.
@@
 Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
-Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
-in $v$'s tree at some time during the execution.
+Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
+in \var{v}'s tree at some time during the execution.
 It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
-each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
-that has appeared in $v$'s RBT earlier.
-Thus, $v.\blocks[b]$ is unique, if it exists.
+each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
+that has appeared in \var{v}'s RBT earlier.
+Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
 We also use this notation in the code to indicate that a search of the RBT should be performed for the block
@@
 %\label{edirGC} 
-%If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
-%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
+%If $B$ is a block in the RBT of an internal node \var{v}, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
+%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $\var{v}.\var{dir}$.
 %\end{observation}
@@
 \begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
- If $h>0$ and a block with index $h$ has been inserted into $v.\fld{blocks}$ then 
- $v.\fld{blocks}[h-1].\fld{end\sub{left}} \leq v.\fld{blocks}[h].\fld{end\sub{left}}$ and 
- $v.\fld{blocks}[h-1].\fld{end\sub{right}} \leq v.\fld{blocks}[h].\fld{end\sub{right}}$.
+ If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
+ $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
+ $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
 \end{customlemma}
 \begin{proof}
-The block $B$ with index $h$ was installed into $v$'s RBT by the CAS at line \ref{casGC} 
+The block $B$ with index $h$ was installed into \var{v}'s RBT by the CAS at line \ref{casGC} 
 that changed the tree from $T$ to $T'$.
-Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $v.blocks$,
+Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
 line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
@@
 Next, we prove an analogue of \Cref{successfulRefresh}.
-We say a node $v$ \emph{contains} an operation
-if some block containing the operation has previously appeared in $v$'s RBT (even if 
+We say a node \var{v} \emph{contains} an operation
+if some block containing the operation has previously appeared in \var{v}'s RBT (even if 
 the block has been removed from the RBT by a subsequent \op{Split}).
@@
 Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
-In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
+In the configuration after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
 when $R$ executed line~\ref{refreshReadTGC}.
@@
 \begin{proof}
-Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
+Suppose \var{v}'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
 when $R$ executed line \ref{refreshReadTGC}.
-Let $i$ be the index of the block containing $op$ that was in $v$'s RBT before 
+Let $i$ be the index of the block containing $op$ that was in \var{v}'s RBT before 
 $R$ executed line \ref{refreshReadTGC}.
@@
 Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
-installs $B'$ in $v.\fld{blocks}$.
-The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $v.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
+installs $B'$ in $\var{v}.\fld{blocks}$.
+The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
 By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
-By the definition of subblocks, some block in $v$ contains $B$ as a subblock
-and therefore $v$ contains $op$.
+By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
+and therefore \var{v} contains $op$.
 
@@
 it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
-that the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
-$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
+that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
+$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$
 contain no operations.
 By \Cref{blockNotEmpty}, each block contains at least one operation, so
-these ranges must be empty, and $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
+these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
 This implies that the block $B$ containing $op$ is a subblock of some block that has appeared
-in $v$'s RBT, so $op$ is contained in $v$.
+in \var{v}'s RBT, so $op$ is contained in \var{v}.
 \end{proof}
@@
 \begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
-Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
-All operations contained $v$'s children when $R_1$ begins
-are contained in $v$ when $R_2$ terminates.
+Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
+All operations contained \var{v}'s children when $R_1$ begins
+are contained in \var{v} when $R_2$ terminates.
 \end{customlemma}
@@
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
-Then some other \op{CAS} on $v.\fld{blocks}$ succeeds between the time each \op{Refresh}
-reads $v.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
+Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
+reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
 Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
-$R_3$ must have read $v.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
+$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
 The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
@@
 \begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
-Let $B=v.\var{blocks}[b]$.
-  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
+Let $B=\var{v}.\fld{blocks}[b]$.
+  If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
 \end{customlemma}
@@
 Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
-%We say the index $b$ \emph{is marked in node $v$} if some process has tested
+%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
 %whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
@@
 %whose argument it should return, or when a dequeue that returns null passes through the block.
-%Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
+%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
 %contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
-%If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.
+%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.
 
@@
 \begin{proof}
-Consider the invocation $A$ of \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
+Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
 \fld{index} is a multiple of $G$.
@@
 
-$A$'s call to \op{SplitBlock} determines the index $s$ used to split $v$'s RBT by following 
-\eleft\ and \eright\ pointers from the root down to $v$.
+$A$'s call to \op{SplitBlock} determines the index $s$ used to split \var{v}'s RBT by following 
+\eleft\ and \eright\ pointers from the root down to \var{v}.
 So, the block returned is a subblock of block $m$ of the root, unless
@@
 
-Next, we bound the number of operations in $v$'s blocks that are retained when $A$ does sets
+Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$ does sets
 $T'\assign\op{Split}(T,s)$ operation.
@@
 Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
-block is inserted before $A$ sets $v$'s \fld{blocks} to the resulting RBT.
+block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
 \end{proof}
@@
 The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
-\here{There may be room to improve this bound by a more careful analysis since the bounds so far
+\here{For later version:  There may be room to improve this bound by a more careful analysis since the bounds so far
 didn't use the fact that operations only appear along one path in the tree; in other words
@@
 Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}.
-If a process does GC in a call to \op{AddBlock}($v,T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on $v$}.
-We argue that each process $P$ can do an $r$th helping phase on $v$ at most once as follows.
-Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on $v$.
-Let $v,T,B$ be the arguments of $A$.
-Any call to \op{AddBlock} on internal node $v$ is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
+If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
+We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once as follows.
+Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
+Let $\var{v},T,B$ be the arguments of $A$.
+Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
 is the maximum \fld{index} in $T$ plus 1.
@@
 or it fails because 
-some other \op{CAS} changes $v.blocks$ from $T$ to another tree.
-Either way, by \Cref{RBTupdates}, $v$'s RBT's maximum \fld{index} will be at least $rG$ at all times
+some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
+Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $rG$ at all times
 after this CAS.
 So if a subsequent \op{Refresh} by process $P$ 
-ever calls \op{AddBlock} on $v$ again, the block it passes as the third argument
-will have $\index>rG$, so $P$ will not perform an $r$th helping phase on $v$ again.
+ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
+will have $\index>rG$, so $P$ will not perform an $r$th helping phase on \var{v} again.
 
@@
 followed by $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
-Thus, for each integer $r$ and each node $v$, a total of $O(p^2\log p\log(p+q_{max}))$ steps
-are performed by all processes during their $r$th helping phase on $v$.
+Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
+are performed by all processes during their $r$th helping phase on \var{v}.
 We can amortize these steps over the operations that appear in 
-$v.blocks[(r-1)p^2+1..p^2]$.
+$\var{v}.\fld{blocks}[(r-1)p^2+1..p^2]$.
 By \Cref{blockNotEmpty}, there are at least $G$ such operations, 

### 2023-01-09 podc_paper/introduction.tex
minor edits

@@
 Besides being a fundamental data structure, queues are used in
-significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
+significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
 packet processing \cite{DPDK},
@@
 Then, we show how to modify the implementation to use 
-\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
+bounded space while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
 Both implementations use single-word CAS and reasonably-sized words.

### 2023-01-09 podc_paper/correctness-2.tex
working on gc invariant

@@
 
-The correctness of \op{GetEnqueue} and \op{indexDequeue}, which are identical to the original version of the implementation are dependent only on the fact that we do not discard blocks needed by those routines.
-The following lemmas show this.
+\subsubsection{GC Keeps Needed Blocks}
+
+The correctness of \op{GetEnqueue} and \op{indexDequeue}, which are very similar to the 
+original implementation, are dependent only on the fact that GC does not 
+discard blocks needed by those routines.
+The following lemmas are used to show this.
+
+We say a block is \emph{finished} if 
+\begin{itemize}
+\item
+it has been propagated to the root,
+\item
+the value of each enqueue contained in the block has either been returned by a dequeue
+or written in the \fld{result} field of a dequeue, and
+\item
+each dequeue contained in the block has either terminated or has a non-null \fld{result} field in the leaf block that represents it.
+\end{itemize}
+Intuitively, once a block is finished, operations no longer need to pass through the block
+to compute responses to operations.
+The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.
+
+\begin{observation}\label{subblocksFinished}
+A block is finished if and only all of its subblocks are finished.
+\end{observation}
+
+\begin{invariant}
+If the minimum \fld{index} of any block in \vv's RBT is $b_{min}$, then
+all blocks that were ever added to \vv\ with $\fld{index}$ at most $b_{min}$ are finished.
+\end{invariant}
+\begin{proof}
+The invariant is true initially, since the minimum \fld{index} block in \vv's RBT is the empty block, which is (vacuously) finished.
+
+We must show the invariant is preserved whenever $\vv.\fld{blocks}$ changes.
+The minimum \fld{index} of \vv's RBT can only change when \vv's RBT changes, either at line 
+\ref{appendLeafGC} (if \vv\ is a leaf) or at line \ref{casGC} (if \vv\ is an internal node).
+In either case, the step changes $\vv.\fld{blocks}$ is changed
+from $T$ to $T'$, where $T'$ is obtained by  a call to \op{AddBlock}(\vv, $T$, $B$).
+(In the case of a leaf \vv, this is true because only the process that owns the leaf ever
+writes to $\vv.\fld{blocks}$.)
+If the \op{AddBlock} does not do GC (lines \ref{GCstart}--\ref{GCend}), 
+then $T'$ is obtained by adding a new block to $T$, so by \Cref{} the minimum \fld{index}
+is the same.
+So consider the case where the \op{AddBlock} performs GC.
+We must show that any block of \vv\ whose index is less than or equal to the
+minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\vv.\fld{blocks}$.
+Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
+a block with a larger index), it suffices to show that all blocks that were ever added
+to \vv's RBT with \fld{index} at most $s$ are finished.
+
+\here{to be continued}
+might need assn that ops so far are consistent with L
+
+Claim:  After an \op{AddBlock} finishes its call to \op{Help} on line \ref{invokeHelp}
+all blocks of $v$ with \fld{index} less than or equal to $s$ are finished.
+
+\end{proof}
 

### 2023-01-10 podc_paper/correctness-2.tex
finished draft of proof for GC algorithm

--- 2023-01-09 podc_paper/correctness-2.tex
@@
 reasoning to ensure GC does not interfere with other routines.
+
+\subsubsection{Basic Properties}
+
 The following observation describes
@@
 
+\subsubsection{Propagating Operations to the Root}
+
 Next, we prove an analogue of \Cref{successfulRefresh}.
@@
 \Cref{lem::appendExactlyOnce} can then be proved in the same way as in Section \ref{sec::propagating}.
-We modified the way \var{super} fields are computed, so we prove an analogue of \Cref{superRelation} to show that \fld{super} fields are (nearly) accurate.
 
-\begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
-Let $B=\var{v}.\fld{blocks}[b]$.
-  If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
-\end{customlemma}
+\subsubsection{GC Keeps Needed Blocks}
+
+The correctness of \op{GetEnqueue} and \op{IndexDequeue}, which are very similar to the 
+original implementation, are dependent only on the fact that GC does not 
+discard blocks needed by those routines.
+The following results are used to show this.
+
+We say a block is \emph{finished} if 
+\begin{itemize}
+\item
+it has been propagated to the root,
+\item
+the value of each enqueue contained in the block has either been returned by a dequeue
+or written in the \fld{response} field of a dequeue, and
+\item
+each dequeue contained in the block has either terminated or has a non-null \fld{response} field in the leaf block that represents it.
+\end{itemize}
+Intuitively, once a block is finished, operations no longer need to pass through the block
+to compute responses to operations.
+The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.
+
+\begin{observation}\label{subblocksFinished}
+A block is finished if and only all of its subblocks are finished.
+\end{observation}
+
+\here{Technically, following proof might have to be wrapped up in a big induction with the argument
+that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
+is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
+Or alternatively, we could add an assumption that if every operation agrees with the linearization
+so far, then the invariant is preserved by each step.
+Then, when proving the linearization is correct, everything will work out.
+But just leave it as is for the submission, since this is a subtle point}
+
+\begin{invariant}\label{lem::finished}
+If the minimum \fld{index} of any block in \var{v}'s RBT is $b_{min}$, then
+each block with $\fld{index}$ at most $b_{min}$ that was ever added to \var{v} is finished.
+\end{invariant}
 \begin{proof}
+The invariant is true initially, since the minimum \fld{index} block in \var{v}'s RBT is the empty block, which is (vacuously) finished.
 
+We show that every step preserves the invariant.  
+We need only consider a step $st$ that modifies a node's RBT.
+The minimum \fld{index} of \var{v}'s RBT can only change when \var{v}'s RBT changes, either at line 
+\ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
+In either case, the step $st$ changes $\var{v}.\fld{blocks}$ is changed
+from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
+(In the case of a leaf \var{v}, this is true because only the process that owns the leaf ever
+writes to $\var{v}.\fld{blocks}$.)
+If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
+then $T'$ is obtained by adding a new block to $T$, so by \Cref{RBTupdates} the minimum \fld{index}
+is unchanged and the invariant is trivially preserved.
+So consider the case where $A$ performs GC.
+We must show that any block of \var{v} whose index is less than or equal to the
+minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
+Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
+a block with a larger index), it suffices to show that all blocks that were ever added
+to \var{v}'s RBT with \fld{index} at most $s$ are finished.
+
+We must examine how $A$'s call to the recursive algorithm \op{SplitBlock} computes the value of $s$.
+
+\begin{claim}\label{splitfinished}
+If one of the recursive calls to $\op{SplitBlock}(x)$ within $A$'s call to \op{SplitBlock},
+then that block and all earlier blocks in $x$ are finished when $st$ occurs.
+\end{claim}
+\begin{proof}[Proof of Claim]
+We prove this claim  by induction on the depth of $x$.
+
+For the base case, suppose $x$ is the root.
+\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
+that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
+By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
+are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
+Between the termination of $A$'s call to \op{Splitblock} at line \ref{callSplitBlock}
+and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
+Thus, after this helping (and before step $st$),
+$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.
+
+For the induction step, we assume the claim holds for $x$'s parent, and prove it for $x$.
+We consider two cases.
+
+If \op{SplitBlock}($x$) returns the minimum block of $x$'s RBT at line \ref{substitute},
+then the claim follows from the assumption that \Cref{lem::finished} holds at all times before $st$.
+
+Otherwise, \op{SplitBlock}($x$) returns the block $B$ at line \ref{substitute}.
+By the induction hypothesis, the block $B_p$ computed at line \ref{SBrecurse}
+(and all earlier blocks of $x$) are finished when $st$ occurs.
+By \Cref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
+is also finished when $st$ occurs.
+\renewcommand{\qedsymbol}{$\diamondsuit$}
 \end{proof}
 
-The correctness of \op{GetEnqueue} and \op{indexDequeue}, which are identical to the original version of the implementation are dependent only on the fact that we do not discard blocks needed by those routines.
-The following lemmas show this.
+If the \op{Split} at line \ref{splitGC} of $A$ modifies the RBT, then it discards
+all the blocks older than the one returned by \op{SplitBlock} at line \ref{callSplitBlock}.
+By \Cref{splitfinished}, the minimum block in the new tree will satisfy the invariant when $st$
+installs the new tree in $\var{v}.\fld{blocks}$.
+\end{proof}
 
-\here{prove some lemmas about not throwing away useful information, then restate correctness lemmas about getenq and indexdeq that say they work correctly OR the operation is already propagated/complete.}
+\begin{lemma}\label{blocknotfound}
+If a \op{Dequeue} operation fails to find a block in an RBT,
+%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
+then it has been propagated to the root and its result has been written in its \fld{response} field.
+If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
+\end{lemma}
+\begin{proof}
+Any block that GC removes from an RBT is finished, by \Cref{lem::finished}, so if an 
+\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
+the root (for example, during the \op{CreateBlock} routine), then a block containing
+the operation itself has been removed from a RBT, so the operation has propagated to the root,
+by the definition of finished.
+Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \op{response} field.
 
-\here{next, should probably show that blocks are removed only after they are no longer needed; i.e., every dequeue has its answer (either returned, or written in its response field) and every enqueue has a matching dequeue that has either terminated or has a response written.}
-one invariant will be that a null dequeue or a dequeued enqueue appears in the first block (or some later block) of each RBT.
-should imply that no operation will need any block before the first block in the RBT
--------
+After propagation to the root, a \op{Dequeue} must access blocks that contain the dequeue,
+as well as the enqueue whose value it will return (if it is not a null dequeue).
+If any of those blocks have been removed, it follows from \Cref{lem::finished} that the
+\op{Dequeue}'s result is written in its \fld{response} field.
+\end{proof}
+
+By \Cref{blocknotfound}, an operation that fails to find a block in an RBT can terminate.
+If it is a \op{Dequeue}, it can return the result written in its \fld{response} field.
+Since no other process updates the RBT in a process's leaf, the block containing the \fld{response}
+will be the last block in the leaf's RBT, and the last block of an RBT is never removed by GC.
+Thus, the \fld{response} field will still be there when a \op{Dequeue} needs it.
+
+\subsubsection{Linearizability}
+
+The correctness of the \op{IndexDequeue} and \op{GetEnqueue}  operations
+can be proved in the same way as in Section \ref{sec::helpingCorrect},
+since they are largely unchanged (except for the simplification that \op{IndexDequeue}
+can simply search for a block's superblock instead of using the block's \fld{super} 
+field to calculate the superblock's position).
+They will give the correct response, provided none of the blocks they need to access have been 
+removed by GC.  But as we have seen above, if that happens, the \op{Enqueue} or \op{Dequeue} can simply terminate.
+
+Similarly the results of Section \ref{sec::linearizability} 
+can be reproved in exactly the same way as for the original algorithm to establish
+that the space-bounded algorithm is linearizable.
 
@@
 
+It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log p(q_{max}+p^3)$.
+
 Although individual operations may now become more expensive because they have to perform GC,
@@
 some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
-Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $rG$ at all times
+Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
 after this CAS.
@@
 ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
-will have $\index>rG$, so $P$ will not perform an $r$th helping phase on \var{v} again.
+will have $\index>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.
 
@@
 by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
-
-============
-
-In our design a process attempts to collect the garbage, when it is going to to append a block in the $kp^2$ position in \nf{root.blocks} (see Line \ref{}).
-Every $p^2$  block appended to the root, one \nf{GarbageCollect} terminates because \nf{root.head} cannot advance until a \nf{GarbageCollect} garbage collect is done (see Lines ).
-
-%\begin{lemma}
-%Size of \nf{root.blocks} after \nf{GarbageCollect} is $O(p^2+q)$.
-%\end{lemma}
-%
-%\begin{lemma}
-%Total number of the blocks in the tree is $O(p^3+pq)$.
-%\end{lemma}
-
-\nf{Enqueue} operation $e$ can be removed from the tree after termination of \nf{Dequeue} $d$ where $Resp(d)=e$.
-It is safe to remove \nf{Dequeue} $d$ from the tree after the $d$ terminates.
-We can remove a block after the conditions told are satisfied for all of its \nf{Enqueue}s and \nf{Dequeue}s.
-A \nf{Dequeue} may go to sleep for a long time and prevent a \nf{block} to be removed (the block it is in or the block its response is in).
-In that situation other processes can help the \nf{Dequeue} by computing its response and writing it down somewhere.
-After writing down the response of th \nf{Dequeue} can be removed, since if the process wanted the response it can read from the helped response.
-
-\begin{definition}
-A block  is \it{finished} if all of its \nf{Enqueue}s have been computed to be the response of some \nf{Dequeue} and all of its \nf{Dequeues} responses have been computed.  
-\end{definition}
-
-\begin{corollary}
-If a block is finished, then all of its subblocks are also finished.  
-\end{corollary}
-
-\begin{lemma}
-It is safe to remove a finished block from the non-leaf nodes.
-\end{lemma}
-
-If the $i$th \nf{Enqueue} gets dequeued in a FIFO queue, it means the first $i-1$ \nf{Enqueue} operations have been already dequeued.
-This gives us the idea that if a block is finished then all the blocks before it are also finished.
-If an operation in a block goes to sleep for a long time then other processes help the operation so the block gets finished.
-There are less than $p$ idle operations, we can help them before garbage collection and then remove all the finished blocks safely in an amortized poly-log time.
-
-\begin{lemma}
-    If all \nf{Dequeue} operations in the root are helped, then all the blocks before a finished block in the root are also finished.
-\end{lemma}
-
-\begin{lemma}
-  Blocks before the most recent block that has been dequeued(computed to be dequeue) from are finished.
-If the most current \nf{Dequeue} returned(computed) \nf{null} then all the blocks before the block containing the \nf{Dequeue} are finished.
-\end{lemma}
-
-\begin{lemma}
-    If we replace the arrays we used to implement \nf{blocks} with red-black trees the amortized complexity of the algorithm would be $PolyLog(p,q)$. And also the algorithm is correct.
-\end{lemma}
-
-We can help a \nf{Dequeue} by computing its response and writing it down. If the process in future failed to execute, it can read the helped value written down.
-
-\begin{lemma}
-The \nf{response} written is correct.
-\end{lemma}
-
-But how can we know which blocks in each node are finished or not?
-
-\begin{lemma}
-  Algorithm is wait-free and linearizable.
-\end{lemma}

### 2023-01-10 podc_paper/algorithm-GC.tex
moved some proofs to appendix, some edits to cut space

--- 2023-01-09 podc_paper/algorithm-2.tex
@@
 finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
-or a null dequeue using the \var{last} array.  All enqueues before this block are either
-dequeued or will be dequeued by a dequeue that is in progress.  Once those pending dequeues have been
-helped to complete, it is safe to discard any blocks in the root older than $B_{root}$, as well
-as their subblocks.
+or a null dequeue using the \var{last} array.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
+dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
+helped to complete by line \ref{invokeHelp},
+it is safe to discard any blocks in the root older than $B_{root}$, as well
+as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
+are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
 The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in \var{v}
@@
         \State $s \assign \Call{SplitBlock}{\var{v}}.\var{index}$\label{callSplitBlock}
-        \State \Call{Help}{}
+        \State \Call{Help}{}\label{invokeHelp}
         \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
@@
 	\Else
-		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$
+		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
 		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
@@
 	\State \linecomment If $B$ was discarded, use leftmost block instead
-	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$)
+	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \label{substitute}
 \EndFunction{SplitBlock}
@@
 \end{minipage}
+\vspace*{-4mm}
 \caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}

### 2023-01-10 podc_paper/algorithm.tex
Updated code to make GC run node-by-node instead
finished draft of proof for GC algorithm
moved some proofs to appendix, some edits to cut space

@@
 
-The primary challenge is thus figuring out what information to store in each block so that 
-the following tasks can be done efficiently (in a polylogarithmic number of steps).
+We must choose what to store in each block so that the following tasks can be done efficiently.
 \begin{enumerate}[label={(T\arabic*)}]
@@
 Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
-\end{figure}
-\here{Make sure two figures appear on same page of final version}
-
-\begin{figure}[t]
 \input{implicit.pdf_t}
@@
     \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
-    \EndIf
+	\Else
 	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
@@
     	\State \Return{ \var{result}}
+	\EndIf
 \EndFunction{Refresh}
@@
 \end{minipage}
+\vspace*{-3mm}
 \caption{Queue implementation.\label{pseudocode1}}

### 2023-01-10 podc_paper/correctness-GC.tex
moved some proofs to appendix, some edits to cut space

--- 2023-01-10 podc_paper/correctness-2.tex
@@
 The correctness of the \op{IndexDequeue} and \op{GetEnqueue}  operations
-can be proved in the same way as in Section \ref{sec::helpingCorrect},
+can be proved in the same way as in Section \ref{sec::tracingCorrect},
 since they are largely unchanged (except for the simplification that \op{IndexDequeue}
@@
 
-\begin{theorem}\label{spaceBound}
+\begin{mytheorem}\label{spaceBound}
 The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
-\end{theorem}
+\end{mytheorem}
 \begin{proof}

### 2023-01-10 podc_paper/correctness.tex
minor edits
finished draft of proof for GC algorithm
moved some proofs to appendix, some edits to cut space

@@
 is sufficient to propagate an operation to the root.
-In Section \ref{sec::helpingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
+In Section \ref{sec::tracingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
 routines correctly navigate through the \ordering\ tree.
@@
 We say that node $v$ \emph{contains} an operation if some block in $v.\fld{blocks}$ contains the operation.
-\here{move this defn earlier and just recall it here?}
+%\here{move this defn earlier and just recall it here?}
 Once a block is added to a node, it remains there forever.  Thus, if $v$ contains an operation at some time, it contains the operation at all later times too.
@@
 
-\subsection{Correctness of Helping Routines}
-\label{sec::helpingCorrect}
+\subsection{Correctness of \opemph{GetEnqueue} and \opemph{IndexDequeue}}
+\label{sec::tracingCorrect}
 
-We first show that \op{getEnqueue} works correctly.
+See Appendix \ref{app::tracingDetails} for detailed proofs for this section.
 
-\begin{lemma}\label{lem::get}
-If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on the height of node $v$.
-If $v$ is a leaf, the hypothesis implies that $i=1$ and the block $v.\fld{blocks}[b]$ represents 
-an enqueue whose argument is stored in $v.\fld{blocks}[b].\fld{element}$.
-\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.
+We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
+This is proved by showing that the \fld{super} field of a block $B$ in node $v$ is read from the
+\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
+On one hand, $B.super$ is written before $B_s$ is installed:
+\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
+$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
+On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
+$B.\fld{super}$ is written after $B$ is installed, and \Cref{successfulRefresh} ensures that $B$ is propagated to the
+parent soon after.
 
-Assuming the claim holds for $v$'s children, we prove it for $v$.
-Let $B$ be $v.\fld{blocks}[b]$.
-By (\ref{defSeqs}),
-$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
-of $B$, which are listed in (\ref{defsubblock}).
-By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
-of enqueues in $E(B)$ that come from $B$'s subblocks in $v$'s left child.
-Thus, $dir$ is set to the direction for the child of $v$ that contains the required enqueue.
-Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
-Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
-By \Cref{lem::sum}, $v.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
-enqueues in $E'$ before the enqueues of block $B'$, so
-the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
-Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
-returns the required result, by the induction hypothesis.
-\end{proof}
-
-Next, we prove the \fld{super} field of a block is within one of the true index of the block's superblock.
-
-\begin{lemma}\label{superRelation}
+\begin{restatable}{lemma}{superRelationRes}
+\label{superRelation}
 Let $B=v.\var{blocks}[b]$.
   If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
-\end{lemma}
-\begin{proof}
-We first show $B.\fld{super}\leq s$.
-Let $R_s$ be the instance of \op{Refresh}($v.\fld{parent}$) that installs the superblock of $B$ 
-in $v.\fld{parent.blocks}[s]$.
-By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $v.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
-By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
-that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $v.\fld{parent}.\head$ before $r$.
-When $r$ occurs, $v.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
-\ref{cas} succeeds.
-So, by \Cref{lem::headPosition}, $v.\fld{parent}.\head \leq s$ when $r$ occurs.
-Since the value stored $B.\fld{super}$ was read from $v.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
-
-Next, we show that $B.\fld{super}\geq s-1$.
-The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
-So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
-By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
-$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
-Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
-This read $r'$ must be before $B$ is installed in $v$;
-otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
-$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
-Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
-It is invoked either 
-at line \ref{helpAdvance} after seeing $v.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
-or at line \ref{advance} after ensuring $v.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
-Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
-By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
-the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
-writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
-\end{proof}
+\end{restatable}
 
@@
 %into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
+Showing  \op{GetEnqueue} and \op{IndexDequeue} work correctly is mostly just checking that they correctly compute the index of the required block
+and the operation's rank within the block.  
+For \op{IndexDequeue}, we use \Cref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
+\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}
 
-We prove \op{IndexDequeue}'s correctness using \Cref{superRelation} on each step of the \op{IndexDequeue}.
-\here{mention somewhere why precondition of IndexDequeue is true; also check that block of parent indexed by sup in that routine is non-null}
-
-\begin{lemma}\label{lem::indexDequeue}
+\begin{restatable}{lemma}{indexDequeueRes}
+\label{lem::indexDequeue}
 If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
  then \op{IndexDequeue}($v, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
-Assuming the claim holds for $v$'s parent, we prove it for $v$.
-Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
-\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
-By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
-The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.
+\end{restatable}
 
-Next, the position of the required dequeue in $D(B')$ is computed in 
-lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
-We first add the number of dequeues in the subblocks of $B'$ in $v$ that preced $B$ on line \ref{computeISuperStart}.
-If $v$ is the right child of its parent, then all of the subblocks of $B'$ from $v$'s left sibling
-also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.
-
-Finally, \op{IndexDequeue} is called recursively on $v$'s parent.
-Since $B$ has been propagated to the root, so has its superblock $B'$.
-Thus, all preconditions of the recursive call are met.
-By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}
+\begin{restatable}{lemma}{getEnqRes}
+\label{lem::get}
+If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
+\end{restatable}
 
@@
 
-Consider any finite execution.  We must show that the linearization ordering $L$ defined in 
+We show that the linearization ordering $L$ defined in 
 (\ref{linearization}) is a legal permutation of a subset of the operations in 
 the execution, i.e., that it includes all operations that terminate and 
-if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does appear
-before $op_1$ in $L$.  Then we must show that each dequeue that terminates returns the 
-same response as it would in the sequential execution described by $L$.
+if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
+We also show result each dequeue returns is the same as in the sequential execution  $L$.
 
@@
 
-\begin{theorem}
+\begin{mytheorem}
 The queue implementation is linearizable.
-\end{theorem}
+\end{mytheorem}
 

### 2023-01-10 podc_paper/podc-submission.tex
editing proof; fixed minor bug in algorithm (off-by-one error)
add garbage collection code to algorithm.tex
working on proof; some edits to intro to save space
working on proof (mainly indexdequeue)
add garbage collection lemmas
add algorithm2
Finished draft of analysis section.  (For some reason, compiling on my laptop requires commenting out some \newtheorem commands in podc-submission.tex .)
minor edits
Editing GC
some editing of GC
Rewording to save some space
minor edits to GC code
minor edits; changed tournament tree to ordering tree
Updated code to make GC run node-by-node instead
separ

@@
 \usepackage{bold-extra}
-\usepackage{showkeys}
-\renewcommand*\contentsname{Table of Contents}
+\usepackage{cleveref}
+\usepackage{thmtools,thm-restate}
+
+%\renewcommand*\contentsname{Table of Contents}
 
@@
 \setlength\columnsep{24pt}
+%\renewcommand{\baselinestretch}{2}
 
-\def\draft{1}  % set to 1 for draft (include comments, color edits), 0 for clean final copy (omit comments, edits in black)
+\numberwithin{equation}{section}
+
+\def\draft{0}  % set to 1 for draft (include comments, color edits), 0 for clean final copy (omit comments, edits in black)
 \if\draft 1
+\usepackage{showkeys}
 \newcommand{\mycomment}[3]{{\color{#2}{[\bf{#1: #3}]}}}
@@
 \newcommand{\blocks}{blocks}
+\newcommand{\ordering}{ordering}
 
@@
 
-\newcommand{\var}[1]{\mbox{\it{#1}}}
+\newcommand{\hangbox}[1]{\parbox[t]{\dimexpr\linewidth-\algorithmicindent\relax}{%
+    \setlength{\hangindent}{\algorithmicindent}%
+    #1}}
+\newcommand{\spac}{\medskip}
+\newcommand{\assign}{\mbox{:=}}
+
+\newcommand{\var}[1]{\mbox{\textnormal{\it{#1}}}}
 \newcommand{\fld}[1]{\mbox{\it{#1}}}
@@
 \newcommand{\tr}{\mbox{\sf true}}
-\newcommand{\opa}[2]{\op{#1}(\var{#2})}
+\newcommand{\fa}{\mbox{\sf false}}
+\newcommand{\opa}[2]{\op{#1(\var{#2})}}
 \newcommand{\op}[1]{{\sf #1}}
@@
 \newcommand{\linecomment}[1]{\com\ #1}
+\newcommand{\gc}[1]{\color{blue}#1\color{black}}
 
 \usepackage{amsmath,amsthm}
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{observation}{Observation}
-%\theoremstyle{definition}
-%\newtheorem{definition}[theorem]{Definition}
-\newtheorem{invariant}{Invariant}
-%\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{mytheorem}[theorem]{Theorem}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{observation}[theorem]{Observation}
+\newtheorem{invariant}[theorem]{Invariant}
+\newtheorem{claim}{Claim}[theorem]
+
+\newtheorem{innercustomlemma}{Lemma}
+\newenvironment{customlemma}[1]
+  {\renewcommand\theinnercustomlemma{#1}\innercustomlemma}
+  {\endinnercustomlemma}
 
@@
 %% used to denote shared contribution to the research.
-\author{Eric Ruppert}
-\affiliation{%
-  \institution{York University}
-  \streetaddress{P.O. Box 1212}
-  \city{Toronto}
-  \state{Ontario}
-  \country{Canada}
-  \postcode{43017-6221}
-}
-\email{ruppert@cse.yorku.ca}
-\orcid{1234-5678-9012}
 
@@
 }
+\author{Eric Ruppert}
+\affiliation{%
+  \institution{York University}
+  \streetaddress{P.O. Box 1212}
+  \city{Toronto}
+  \state{Ontario}
+  \country{Canada}
+  \postcode{43017-6221}
+}
+\email{ruppert@cse.yorku.ca}
+\orcid{0000-0001-5613-8701}
 
@@
 CAS instructions.
-Previous lock-free queues all take $\Omega(p)$ steps per operation in the worst case, 
+Previous lock-free queues all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
 where $p$ is the number of processes that can access the queue.
-We achieve $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue.
-\here{mention space usage; update time bound if space-efficient version takes more time per op}
+We present a new wait-free queue that takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
+A bounded-space version of the implementation has amortized step complexity of $O(\log^2 p +\log q)$ per operation.
 \end{abstract}
@@
 
+%OMIT IN SUBMISSION FOR ANONYMITY
+%\paragraph{Acknowledgements}
+%We thank Franck van Breugel for his helpful comments.
+%Funding was provided by the Natural Sciences and Engineering Research Council of Canada.
+
 \newpage
@@
 
+\newpage
 \appendix
+
+\input{correctness-details}
+
+\input{algorithm-GC}
+
+\input{correctness-GC}
+
 \end{document}

### 2023-01-10 podc_paper/algorithm.tex
minor edits

@@
 and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
-\begin{eqnarray}
-E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
-D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
-\end{eqnarray}
+\begin{equation}
+E(B) = E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r) \mbox{ and }
+D(B) = D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
+\end{equation}
 We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
@@
 
-The information stored in a \block\ depends on whether it is in an internal node or a leaf.
-A block in a leaf represents a single enqueue or dequeue.  The \fld{element} field stores the value
+%The information stored in a \block\ depends on whether it is in an internal node or a leaf.
+A block in a leaf represents a single enqueue or dequeue.  The block's \fld{element} field stores the value
 enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.
@@
 	\item[\com] Blocks in internal \nodes\ have the following additional fields
-	\item \typ{int} \eleft, \eright
+	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{int} \eleft, \eright
   		\Comment{index of last direct subblock in the left and right child}
+	\end{itemize}
   	\item[\com] Blocks in leaf \nodes\ have the following additional field
-  	\item \typ{Object} \fld{element}
+  	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{Object} \fld{element}
   		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
+	\end{itemize}
 	\item[\com] Blocks in the root \node\ have the following additional field
-	\item \typ{int} \size%
+	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{int} \size%
   		\Comment{size of the queue after performing all operations up to the end of this block}
+	\end{itemize}
 \end{itemize}

### 2023-01-10 podc_paper/correctness-GC.tex
minor edits

@@
 
-\begin{proposition}
+\begin{mytheorem}
 The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
-\end{proposition}
+\end{mytheorem}
 \begin{proof}
@@
 by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
+Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
+
+\begin{customprop}{\bf{\ref{casbound}}$'$}
+Each operation performs $O(\log p)$ \op{CAS} instructions in the worst case.
+\end{customprop}

### 2023-01-10 podc_paper/introduction.tex
minor edits

@@
 In the worst case, this means that each successful CAS may cause all other processes to
-fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
+fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
 (To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
@@
 We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
-whose step complexity does not depend linearly on $p$.
-Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
+whose step complexity is sublinear in $p$.
+Our queue is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
 For ease of presentation, we first give an unbounded-space construction that uses
-$O(\log^2 p + \log q)$ steps per operation.
-Then, we show how to modify the implementation to use 
-bounded space while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
-Both implementations use single-word CAS and reasonably-sized words.
-Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
+$O(\log^2 p + \log q)$ steps per operation,
+and then modify it to bound the space
+while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
+Both versions use single-word CAS on reasonably-sized words.
+Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
 lock-free queues use 
@@
 $\Omega(p)$ CAS instructions, even in an amortized sense.
-For the space-bounded implementation, we unlink unneeded memory from our data structure.
+For the space-bounded version, we unlink unneeded memory from our data structure.
 We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
@@
 Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
-Each process propagates its operations along the path from its leaf up to the root.
+Each process adds its operations to its leaf.
+As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
+and avoids the CAS retry problem.
 Operations in the root are ordered, 
@@
 \here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
-Helping  ensures wait-freedom and avoids the CAS retry problem:  a process propagating its operation 
-to a node $v$ helps propagate all operations from both of $v$'s children.
 Explicitly storing all operations in the nodes would be too costly.

### 2023-01-10 podc_paper/podc-submission.tex
minor edits

@@
 
+\newtheorem{innercustomprop}{Proposition}
+\newenvironment{customprop}[1]
+  {\renewcommand\theinnercustomprop{#1}\innercustomprop}
+  {\endinnercustomprop}
+
 %%
@@
 \begin{abstract}
-We introduce a novel linearizable wait-free queue implementation using single-word 
+We present a novel linearizable wait-free queue implementation using single-word 
 CAS instructions.
-Previous lock-free queues all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
-where $p$ is the number of processes that can access the queue.
-We present a new wait-free queue that takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
+Previous lock-free queue implementations from \op{CAS} all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
+where $p$ is the number of processes that access the queue.
+Our new wait-free queue  takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
 A bounded-space version of the implementation has amortized step complexity of $O(\log^2 p +\log q)$ per operation.

### 2023-01-10 podc_paper/related.tex
minor edits

--- 2023-01-06 podc_paper/related.tex
@@
 queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
-We adapt their idea of processes using
-a tournament tree to agree on a linearization of
-operations.
+Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
 
@@
 \emph{Fetch\&Increment Objects.}
-There are sublinear-time implementations of another object
-with consensus number 2.
-Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
-implemented fetch\&increment objects from LL/SC objects
-using a polylogarithmic number of steps per operation.
+There are implementations of fetch\&increment objects that use a polylogarithmic number of steps per operation. \cite{ERW12,10.1007/978-3-642-41527-2_20}
 Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
-which uses a tournament tree to keep track of the fetch\&increments
+which also uses a tree structure to keep track of the fetch\&increments
 that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.

### 2023-01-10 podc_paper/algorithm.tex
edits to save space

@@
 
+\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
+\begin{figure}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{1}
+\Statex \linecomment{Shared variable}
+\begin{itemize}
+\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
+\end{itemize}
+
+\Statex \linecomment{Thread-local variable}
+\begin{itemize}
+\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \typ{Node}
+\begin{itemize}
+\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
+\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
+\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
+\end{itemize}
+
+\Statex $\blacktriangleright$ \typ{Block}
+
+\begin{itemize}
+  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
+  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
+  	\item \typ{int} \fld{super}
+  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
+	\item[\com] Blocks in internal \nodes\ have the following additional fields
+	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{int} \eleft, \eright
+  		\Comment{index of last direct subblock in the left and right child}
+	\end{itemize}
+  	\item[\com] Blocks in leaf \nodes\ have the following additional field
+  	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{Object} \fld{element}
+  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
+	\end{itemize}
+	\item[\com] Blocks in the root \node\ have the following additional field
+	\begin{itemize}[leftmargin=3mm]
+		\item[$\bullet$] \typ{int} \size%
+  		\Comment{size of the queue after performing all operations up to the end of this block}
+	\end{itemize}
+\end{itemize}
+
+%\Statex {\com\  Blocks in internal nodes have the following additional fields}
+%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
+%\begin{itemize}
+%	\item \tt{int} \eleft, \eright
+%  		\Comment{index of last direct subblock in the left and right child}
+%\end{itemize}
+
+%\Statex {\com\ Blocks in leaf nodes have the following additional field}
+%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
+%\begin{itemize}
+%  \item \tt{\sl{Object} element}
+%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+%\end{itemize}
+
+%\Statex {\com\ Blocks in the root node have the following additional field}
+%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
+%\begin{itemize}
+%  \item \tt{\sl{int} \size}%
+%  \Comment{size of the queue after performing all operations up to the end of this block}
+%\end{itemize}
+
+\end{algorithmic}
+\caption{Objects used in the \ordering\ tree data structure \label{object-fields}}
+\end{figure}
+
 Each \node\ of the \ordering\ tree has an infinite array called \fld{blocks}.
@@
 shall prove that it is close.
-
-\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
-\begin{figure}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{1}
-\Statex \linecomment{Shared variable}
-\begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
-\end{itemize}
-
-\Statex \linecomment{Thread-local variable}
-\begin{itemize}
-\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \typ{Node}
-\begin{itemize}
-\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
-\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
-\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
-\end{itemize}
-
-\Statex $\blacktriangleright$ \typ{Block}
-
-\begin{itemize}
-  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
-  	\item \typ{int} \fld{super}
-  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
-	\item[\com] Blocks in internal \nodes\ have the following additional fields
-	\begin{itemize}[leftmargin=3mm]
-		\item[$\bullet$] \typ{int} \eleft, \eright
-  		\Comment{index of last direct subblock in the left and right child}
-	\end{itemize}
-  	\item[\com] Blocks in leaf \nodes\ have the following additional field
-  	\begin{itemize}[leftmargin=3mm]
-		\item[$\bullet$] \typ{Object} \fld{element}
-  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
-	\end{itemize}
-	\item[\com] Blocks in the root \node\ have the following additional field
-	\begin{itemize}[leftmargin=3mm]
-		\item[$\bullet$] \typ{int} \size%
-  		\Comment{size of the queue after performing all operations up to the end of this block}
-	\end{itemize}
-\end{itemize}
-
-%\Statex {\com\  Blocks in internal nodes have the following additional fields}
-%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
-%\begin{itemize}
-%	\item \tt{int} \eleft, \eright
-%  		\Comment{index of last direct subblock in the left and right child}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in leaf nodes have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
-%\begin{itemize}
-%  \item \tt{\sl{Object} element}
-%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in the root node have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
-%\begin{itemize}
-%  \item \tt{\sl{int} \size}%
-%  \Comment{size of the queue after performing all operations up to the end of this block}
-%\end{itemize}
-
-\end{algorithmic}
-\caption{Objects used in the \ordering\ tree data structure \label{object-fields}}
-\end{figure}
 
@@
 
-    \If{\var{v} is a leaf node}
-        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
+    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
     \Else 
@@
 %\paragraph{\tt{Refresh()} and \tt{Advance()}}
-The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
+A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
 in \var{v}'s
-children and append it to \var{v}.\fld{blocks}. 
-Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
-and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
-If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v}.
-In this case, \op{Refresh} returns true at line \ref{addOP} and terminates.
-Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
-Either this CAS succeeds or some other process has installed a new block in this location.
-Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
+children and tries to append it to \var{v}.\fld{blocks}. 
+Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
+Line \ref{invokeCreateBlock} creates the new block to install in \var{v}.\fld{blocks}[\var{h}].
+If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
+Otherwise, the CAS at line \ref{cas} tries to install the new block into \var{v}.\fld{blocks}[\var{h}].
+Either this CAS succeeds or some other process has installed a  block in this location.
+Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
 from \var{h} to $\var{h}+1$
-and to fill in the \fld{super} field of the most recently appended block.
+and fill in the \fld{super} field of the most recently appended block.
 The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
-Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
-other processes help keep \fld{head} up to date by also calling \op{Advance}, 
-either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.
+A \op{Refresh} may pause after a successful CAS before calling \op{Advance} at line \ref{advance},
+so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
+either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.
 
@@
 by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
-The block \nf{new} is created in Line \ref{initNewBlock}. 
+The block \var{new} is created in Line \ref{initNewBlock}. 
 The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
 children, which are obtained by reading the \fld{head} index in \var{v}'s children.
-Recall that the \fld{sum\sub{enq}} field should store the number of enqueues in
-\var{v}.\fld{blocks}[1..\var{i}].
-Since these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright],
-the total number of these enqueues is $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
+Since the \fld{sum\sub{enq}} field should store the number of enqueues in
+\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright], line \ref{createSumEnq} sets
+\fld{sum\sub{enq}} to $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
 Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
@@
 represents the number of elements in the queue after the operations in the block are performed.
-Finally, if the newly created block contains no operations, \op{CreateBlock} returns \nl\ to indicate
-that there is no need to append it to \var{v}.\fld{blocks}.
+Finally, if the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
+ there is no need to install it.
 

### 2023-01-10 podc_paper/algorithm-GC.tex
edits to save space
dealt with comments from Hossein of 12:03 email on Jan 10

@@
 Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
-the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p\ceil{\log p}$.
+the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
 This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
@@
     \For{$\ell$ in $\var{leaf}[1..k]$}
-        \State $\var{B} \assign \Call{LastBlock}{\ell.\fld{blocks}}$
+        \State $\var{B} \assign \Call{MaxBlock}{\ell.\fld{blocks}}$
     	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{352}
+\setcounter{ALG@line}{351}
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
@@
     \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
-    \If{\var{v} is a leaf node}
-        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
+    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
     \Else 

### 2023-01-10 podc_paper/correctness-GC.tex
dealt with comments from Hossein of 12:03 email on Jan 10

@@
 \subsection{Correctness}
-Unfortunately, there are enough changes to the algorithm that a new proof of correctness
+There are enough changes to the algorithm that a new proof of correctness
 is required.  Its structure mirrors the proof of the original algorithm, but requires additional
@@
 We now establish that definition (\ref{defsubblock}) of a block's subblocks
-still makes sense.
+still makes sense by proving the analogue of Lemma \ref{lem::headProgress}.
 
@@
 
-We next prove the analogue of Lemma \ref{lem::headProgress}.
 \begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
@@
 \item
-each dequeue contained in the block has either terminated or has a non-null \fld{response} field in the leaf block that represents it.
+each dequeue contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
 \end{itemize}

### 2023-01-10 podc_paper/algorithm.tex
cutting to save space

@@
 If so, it returns \nl\ and we call it a \emph{null dequeue}.
-If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue in the linearization ordering
-among all non-null dequeues,
+If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue among all non-null dequeues in the linearization ordering,
 finds the $r$th enqueue in the linearization, and returns that enqueue's value.
@@
 \label{findrank}
-Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
+Given a dequeue's position in the root, determine if it is a null dequeue (i.e., if the queue is empty when it is linearized)
 or determine the rank $r$ of the enqueue whose value it should return.
@@
 \end{eqnarray}
-This allows us to navigate to the direct subblocks of any block easily.
+This allows us to navigate to a block's direct subblocks easily.
 Blocks also store prefix sums of the numbers of enqueues and dequeues:
-the block in $v.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
+the block in $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
 that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
-These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
-For example, consider finding the $r$th enqueue in the linearization ordering.
-We use the \fld{sum\sub{enq}} fields of the root's blocks to do a binary search
-to locate which block in the root contains the operation.
-If we know which block $B$ in some \node\ $v$ contains the enqueue,
-we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
+These fields allow us to pinpoint the  location of an operation among the subblocks of a given block.
+For example, consider finding the $r$th enqueue $E_r$ in the linearization.
+We do a binary search on \fld{sum\sub{enq}} fields of the root's blocks 
+to locate the block that contains $E_r$.
+If we know which block $B$ in some \node\ $v$ contains $E_r$,
+we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains $E_r$
 and then to do a binary search
@@
 Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
-the enqueue.
+$E_r$.
 %Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
@@
 and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
-for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
-All of the information needed for this search process can be derived from the 
+for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
+All information needed for this search process can be derived from the 
 \eleft, \eright\ and \fld{sum\sub{enq}} fields.
@@
 To facilitate task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
-the (approximate) index of its superblock in the parent \node's \fld{blocks} array.
-We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
+the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
+%We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
 This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
@@
 using a value they read from the \fld{head} field of $v$'s parent.
-As mentioned above, this might not be the exact index of $B$'s superblock, but we
-shall prove that it is close.
+%As mentioned above, this might not be the exact index of $B$'s superblock, but we
+%shall prove that it is close.
 
@@
             \State \fld{dir} \assign\ \fld{left}
-        \Else
+        \Else \Comment{required enqueue is in \var{v}.\fld{right}}
             \State \fld{dir} \assign\ \fld{right}
@@
 by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
-The block \var{new} is created in Line \ref{initNewBlock}. 
-The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
-children, which are obtained by reading the \fld{head} index in \var{v}'s children.
+%The block \var{new} is created in Line \ref{initNewBlock}. 
+The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
+children, obtained by reading the \fld{head} index in \var{v}'s children.
 Since the \fld{sum\sub{enq}} field should store the number of enqueues in
@@
 Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
-subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues that are already in \var{v}.\fld{blocks}[$1..\var{i}-1$].
+subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$].
 The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
@@
 
-Since the index of \var{B}'s superblock is not known until $B$ is
-propagated to $par$, $R_v$ cannot set the \nf{super} field of ${B}$ when
+Since the index of \var{B}'s superblock $B'$ is not known until $B$ is
+propagated to $par$, $R\sub{\var{v}}$ cannot set the \nf{super} field of ${B}$ when
 it creates \var{B}. 
-It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (because
+It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (since
 there could be up to $p$ direct subblocks).
-Intuitively, the index of \var{B}'s superblock will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
+Intuitively, the index of $B'$ will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
 Since we can search for the precise superblock when we need it, it suffices that \var{B}.\fld{super}
-is close to the actual index of \var{B}'s superblock.
+is close to the actual index of  $B'$.
 Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
-As mentioned earlier, other processes may help fill in this field by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that it is filled in soon after \var{B} is installed.
-We shall
-show that this ensures that \nf{$B$.super} differs
+As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
+We shall show \nf{$B$.\fld{super}}  differs
 from the index of $B'$ by at most~1.
@@
 of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
-If the queue is
-not empty, Line \ref{computeE} computes the rank \var{e} of the
+If not, line \ref{computeE} computes the rank \var{e} of the
 \op{Enqueue} whose argument is the \op{Dequeue}'s response. 

### 2023-01-10 podc_paper/correctness.tex
cutting to save space

@@
 \section{Proof of Correctness}
+\label{sec::correctness}
 
@@
 
-Now we show $v.\fld{blocks}[v.\head]$ is either the last non-\nl\ block or the first \nl\ block in node $v$.
+Now we show $\var{v}.\fld{blocks}[\var{v}.\head]$ is either the last non-\nl\ block or the first \nl\ block in node $\var{v}$.
 
 \begin{invariant}\label{lem::headPosition} 
-For $0 \leq i < v.\head$, $v.\fld{blocks}[i]\neq\nl$.  For $i>v.\head$, $v.\fld{blocks}[i]=\nl$.
-If $v\neq \var{root}$,  $v.\fld{blocks}[i].super \neq \nl$ for $0<i<v.head$.
+For $0 \leq i < \var{v}.\head$, $\var{v}.\fld{blocks}[i]\neq\nl$.  For $i>\var{v}.\head$, $\var{v}.\fld{blocks}[i]=\nl$.
+If $\var{v}\neq \var{root}$,  $\var{v}.\fld{blocks}[i].super \neq \nl$ for $0<i<\var{v}.\head$.
 \end{invariant}
@@
 \begin{proof}
-Initially, $v.\head=1$, $v.\fld{blocks}[0]\neq\nl$  and $v.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.
+Initially, $\var{v}.\head=1$, $\var{v}.\fld{blocks}[0]\neq\nl$  and $\var{v}.\fld{blocks}[i]=\nl$ for  $i>0$, so the claims~hold.
 
-Assume the claim holds before a change to $v.\fld{blocks}$, which can be made only
+Assume the claim holds before a change to $\var{v}.\fld{blocks}$, which can be made only
 by a successful \op{CAS} at line \ref{cas}.
-The \op{CAS} changes $v.\fld{blocks}[h]$ from \nl\ to a non-\nl\ value.
-Since $v.\fld{blocks}[h]$ is \nl\ before the CAS, $v.\head \leq h$ by the hypothesis.
-Since $h$ was read from $v.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
-$v.\head \geq h$ by \Cref{lem::headPosition}.
-So, $h=v.\head$ and a change to $v.\fld{blocks}[v.\head]$ preserves the invariant.
+The \op{CAS} changes $\var{v}.\fld{blocks}[h]$ from \nl\ to a non-\nl\ value.
+Since $\var{v}.\fld{blocks}[h]$ is \nl\ before the CAS, $\var{v}.\head \leq h$ by the hypothesis.
+Since $h$ was read from $\var{v}.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
+$\var{v}.\head \geq h$ by \Cref{lem::headPosition}.
+So, $h=\var{v}.\head$ and a change to $\var{v}.\fld{blocks}[\var{v}.\head]$ preserves the invariant.
 
-Now, assume the claim holds before a change to $v.\head$, which can only be an increment from $h$ to $h+1$
+Now, assume the claim holds before a change to $\var{v}.\head$, which can only be an increment from $h$ to $h+1$
 by a successful \op{CAS} at line \ref{incrementHead} of \op{Advance}.
-For the first two claims, it suffices to show that $v.\fld{blocks}[head] \neq \nl$.
+For the first two claims, it suffices to show that $\var{v}.\fld{blocks}[head] \neq \nl$.
 \nf{Advance} is called either at line \ref{helpAdvance} 
-after testing that $v.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
-or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $v.\fld{blocks}[h]\neq\nl$.
-For the third claim, observe that prior to incrementing $v.\head$ at line \ref{incrementHead},
-the \op{CAS} at line \ref{setSuper1} ensures that $v.\fld{blocks}[i].super\neq \nl$.
+after testing that $\var{v}.\fld{blocks}[h]\neq\nl$ at line \ref{ifHeadnotNull},
+or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $\var{v}.\fld{blocks}[h]\neq\nl$.
+For the third claim, observe that prior to incrementing $\var{v}.\head$ at line \ref{incrementHead},
+the \op{CAS} at line \ref{setSuper1} ensures that $\var{v}.\fld{blocks}[i].super\neq \nl$.
 \end{proof}
@@
 \begin{lemma} \label{lem::headProgress}
- If $b>0$ and $v.\fld{blocks}[b] \neq \nl$ then 
- $v.\fld{blocks}[b-1].\fld{end\sub{left}} \leq v.\fld{blocks}[b].\fld{end\sub{left}}$ and 
- $v.\fld{blocks}[b-1].\fld{end\sub{right}} \leq v.\fld{blocks}[b].\fld{end\sub{right}}$.
+ If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$ then 
+ $\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{left}}$ and 
+ $\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{right}}$.
 \end{lemma}
 \begin{proof}
-Let $B$ be the block in $v.\fld{blocks}[b]$.
+Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
 Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
-read $b$ from $v.\head$ at line \ref{readHead}.
-At that time, $v.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
-Thus, the \op{CreateBlock}($v,b-1$) that created $B'$ terminated before the \op{CreateBlock}($v,b$) that
+read $b$ from $\var{v}.\head$ at line \ref{readHead}.
+At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
+Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
 created $B$ started.
 It follows from \Cref{nonDecreasingHead} that the value that 
-line \ref{createEndLeft} of \op{CreateBlock}($v,b-1$) stores in $B'.\fld{end\sub{left}}$   
-is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($v,b$) 
+line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
+is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
 stores in $B.\fld{end\sub{left}}$.
@@
 \begin{corollary}\label{lem::noDuplicates}
-For  $i\neq j$, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ cannot both contain the same operation.
+For  $i\neq j$, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ cannot both contain the same operation.
 \end{corollary}
 \begin{proof}
-The operations contained in a block $B$ are those that appear in subblocks of $B$ in the leaves of the tree.
-Since each process puts each of its operations in only one block of its own leaf, an operation 
+A block $B$ contains the operations in $B$'s subblocks in leaves of the tree.
+An operation by process $P$ appears in one block of $P$'s leaf, so
+an operation 
 cannot be in two different leaf blocks. 
-By \Cref{lem::subblocksDistinct}, $v.\fld{blocks}[i]$ and $v.\fld{blocks}[j]$ have no subblocks in common, so the claim follows.
+By \Cref{lem::subblocksDistinct}, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ have no common subblocks, so the claim follows.
 \end{proof}
@@
 
-The following shows that the values stored in \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields are accurate.
-\here{If space is tight, could easily move next proof to appendix and just say it follows easily from the definition of subblocks and the values stored in the fields on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq}, \ref{createSumDeq}.}
+The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
+on line \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
+from the definition of subblocks.  See Appendix 
+\ref{app::tracingDetails} for a detailed proof.
 
-\begin{invariant}\label{lem::sum}
-If $B$ is a block stored in $v.\fld{blocks}[i]$,
-$B.\fld{sum\sub{enq}} = | E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i]) |$ and
-$B.\fld{sum\sub{deq}} = | D(v.\fld{blocks}[0])\cdots D(v.\fld{blocks}[i]) |$.
-\end{invariant}
-\begin{proof}
-Initially, each \fld{blocks} array only contains an empty block $B_0$ in location 0.
-By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
-Moreover, $B_0.\fld{sum\sub{enq}} = B_0.\fld{sum\sub{deq}} = 0$, so the claim is true.
-
-We show that each installation of a block $B$ into some location $v.\fld{blocks}[i]$ preserves the claim,
-assuming the claim holds before this installation.  We consider two cases.
-
-If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
-For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
-Since $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
-$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
-The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.
-
-Now suppose $v$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
-subblocks of $v.\fld{blocks}[1..i]$ are $v.\fld{left.blocks}[1..B.\eleft]$ 
-and $v.\fld{right.blocks}[1..B.\eright]$.
-Thus, the enqueues in $E(v.\fld{blocks}[0])\cdots E(v.\fld{blocks}[i])$ are those in
-$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eleft])$ and
-$E(v.\fld{left.blocks}[0]) \cdots E(v.\fld{left.blocks}[B.\eright])$.
-By the hypothesis, the total number of these enqueues is $v.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
-The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
-\end{proof}
+\begin{restatable}{invariant}{sumRes}
+\label{lem::sum}
+If $B$ is a block stored in $\var{v}.\fld{blocks}[i]$,
+$B.\fld{sum\sub{enq}} = | E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i]) |$ and
+$B.\fld{sum\sub{deq}} = | D(\var{v}.\fld{blocks}[0])\cdots D(\var{v}.\fld{blocks}[i]) |$.
+\end{restatable}
 
@@
 \begin{corollary}\label{blockNotEmpty}
-If a block $B$ is in $v.\fld{blocks}[i]$ where $i>0$, then $E(B)$ and $D(B)$ are not both empty.
+If a block $B$ is in $\var{v}.\fld{blocks}[i]$ where $i>0$, then $E(B)$ and $D(B)$ are not both empty.
 \end{corollary}
@@
 Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
-We say that node $v$ \emph{contains} an operation if some block in $v.\fld{blocks}$ contains the operation.
+We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
 %\here{move this defn earlier and just recall it here?}
-Once a block is added to a node, it remains there forever.  Thus, if $v$ contains an operation at some time, it contains the operation at all later times too.
+Since blocks are permanently added to nodes, if $\var{v}$ contains $op$ at some time, $v$ contains $op$ at all later times too.
 
 \begin{lemma}\label{successfulRefresh}
-Let $R$ be a call to \op{Refresh}($v$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
-In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
+Let $R$ be a call to \op{Refresh}($\var{v}$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
+In the configuration after that CAS (or termination, respectively), $\var{v}$ contains all operations that $\var{v}$'s children contained 
 when $R$ executed line~\ref{readHead}.
@@
 \begin{proof}
-Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
+Suppose $\var{v}$'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
 when $R$ executed line \ref{readHead}.
-Let $i$ be the index such that the block $B=v.\fld{left.blocks}[i]$ contains $op$.
+Let $i$ be the index such that the block $B=\var{v}.\fld{left.blocks}[i]$ contains $op$.
 By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
-$v.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
+$\var{v}.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
 If it is equal to $i$, $R$ calls \op{Advance} at line \ref{helpAdvance}, which ensures that 
-$v.\fld{left.head} > i$.
-Then, $R$ calls \op{CreateBlock}($v,h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
-\op{CreateBlock} reads a value greater than $i$ from $v.\fld{left.head}$ at line \ref{createEndLeft}.
+$\var{v}.\fld{left.head} > i$.
+Then, $R$ calls \op{CreateBlock}($\var{v},h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
+\op{CreateBlock} reads a value greater than $i$ from $\var{v}.\fld{left.head}$ at line \ref{createEndLeft}.
 Thus, $new.\eleft \geq i$.  We consider two cases.
@@
 Suppose $R$'s call to \op{CreateBlock} returns the new block $B'$ and $R$'s \op{CAS} at line \ref{cas} 
-installs $B'$ in $v.\fld{blocks}$.
-Then, $B$ is a subblock of some block in $v$, since  $B'.\eleft$ is greater than or equal to $B$'s index
-in $v.\fld{left.blocks}$.
-Hence $v$ contains $op$, as required.
+installs $B'$ in $\var{v}.\fld{blocks}$.
+Then, $B$ is a subblock of some block in $\var{v}$, since  $B'.\eleft$ is greater than or equal to $B$'s index
+in $\var{v}.\fld{left.blocks}$.
+Hence $\var{v}$ contains $op$, as required.
 
@@
 Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
-Intuitively, since there are no operations in $v$'s children to promote, $op$ is already in $v$.
+Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
 We formalize this intuition.
@@
 \var{num\sub{enq}} 
-&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - v.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
-&=& v.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + v.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
-&&\mbox{ }- v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
+&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
+&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
+&&\mbox{ }- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
 \end{eqnarray*}
 It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
-$v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
-$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$.
+$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
+$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$.
 Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
@@
 these blocks contain no operations.
-By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
-Hence, $B$ is already a subblock of some block in $v$, so $v$ contains $op$.
+By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
+Hence, $B$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
 \end{proof}
 
-We now show that if a process fails its \op{Refresh}($v$) twice, some other process must have succeeded.
+We now show that if a process fails its \op{Refresh}($\var{v}$) twice, some other process must have succeeded.
 
 \begin{lemma}\label{lem::doubleRefresh}
-Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
-All operations contained $v$'s children when $R_1$ begins
-are contained in $v$ when $R_2$ terminates.
+Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($\var{v}$) by the same process.
+All operations contained $\var{v}$'s children when $R_1$ begins
+are contained in $\var{v}$ when $R_2$ terminates.
 \end{lemma}
@@
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
-Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $v.\head$ at line \ref{readHead}.
+Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $\var{v}.\head$ at line \ref{readHead}.
 By \Cref{lem::headInc}, $h_2>h_1$.
-By \Cref{lem::headProgress}, $v.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
-Since $R_2$ fails its \op{CAS} on $v.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
-a successful \op{CAS} on $v.blocks[h_2]$ before $R_2$'s \op{CAS}.
-$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $v.\head$ and the value of $v.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
-Thus, all operations contained in $v$'s children when $R_1$ begins
-are also contained in $v$'s children when $R_3$ later executes line \ref{readHead}.
-By \Cref{successfulRefresh}, these operations are contained in $v$ when $R_3$ performs its successful \op{CAS},
+By \Cref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
+Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
+a successful \op{CAS} on $\var{v}.blocks[h_2]$ before $R_2$'s \op{CAS}.
+$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
+Thus, all operations contained in $\var{v}$'s children when $R_1$ begins
+are also contained in $\var{v}$'s children when $R_3$ later executes line \ref{readHead}.
+By \Cref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
 which is before $R_2$ terminates.
@@
 We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
-This is proved by showing that the \fld{super} field of a block $B$ in node $v$ is read from the
+This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from the
 \var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
@@
 \label{superRelation}
-Let $B=v.\var{blocks}[b]$.
-  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
+Let $B=\var{v}.\var{blocks}[b]$.
+  If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
 \end{restatable}
@@
 \label{lem::indexDequeue}
-If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
- then \op{IndexDequeue}($v, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
+If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
+ then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
 \end{restatable}
@@
 \label{lem::get}
-If $1\leq i\leq |E(v.\fld{blocks}[b])|$ then \op{getEnqueue}($v,b,i$) returns the argument of the $i$th enqueue in $E(v.\fld{blocks}[b])$.
+If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
 \end{restatable}

### 2023-01-10 podc_paper/algorithm.tex
shortening

@@
 For example, consider finding the $r$th enqueue $E_r$ in the linearization.
-We do a binary search on \fld{sum\sub{enq}} fields of the root's blocks 
-to locate the block that contains $E_r$.
-If we know which block $B$ in some \node\ $v$ contains $E_r$,
+A binary search for $r$ on \fld{sum\sub{enq}} fields of the root's blocks 
+finds the block  containing $E_r$.
+If we know a block $B$ in a \node\ $v$ contains $E_r$,
 we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains $E_r$
-and then to do a binary search
+and then do a binary search
 among the direct subblocks of $B$ in that child.
@@
 
-We now consider task \ref{findrank}.
-First, we must determine whether the queue is empty when a dequeue occurs.
-To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
-in the queue after all operations in the linearization ordering up to that block (inclusive) 
-have been performed.
-We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
-$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
-We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
-as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
-Then, for any non-null dequeue in $B_d$, we can use this information to determine its
-rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
-whose value it should return (among all enqueues).
+Now consider task \ref{findrank}.
+To determine whether the queue is empty when a dequeue occurs,
+each block in the root has a \fld{size} field storing the number of elements
+in the queue after all operations in the linearization up to that block (inclusive) 
+have been done.
+We can  determine which dequeues in a block $B_d$ in the root are null dequeues using
+$B_{d-1}.\fld{size}$, which is the size of the queue just before $B_d$'s operations, and the number of enqueues and dequeues in $B_d$.
+Moreover, the total number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
+is $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
+We can use this information to determine the
+rank of a non-null dequeue in $B_d$ among all non-null dequeues in the linearization, which is the rank  (among all enqueues) of the enqueue
+whose value the dequeue should return.
 
@@
 there could be up to $p$ direct subblocks).
-Intuitively, the index of $B'$ will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
-Since we can search for the precise superblock when we need it, it suffices that \var{B}.\fld{super}
-is close to the actual index of  $B'$.
-Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
-As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
-We shall show \nf{$B$.\fld{super}}  differs
-from the index of $B'$ by at most~1.
+Soon after $B$ is installed,
+a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
+set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
+We shall show that this happens soon enough that $B.\fld{super}$ differs from the true index of $B'$
+by at most 1.
+%Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
+%As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
+%We shall show \nf{$B$.\fld{super}}  differs
+%from the index of $B'$ by at most~1.
 

### 2023-01-10 podc_paper/analysis.tex
minor edits
shortening

@@
 
-\begin{proposition}
+\begin{proposition}\label{casbound}
 Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
@@
 
-We now bound operations' running times in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.
+We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.
 
-\begin{lemma}\label{enqDeqTime}
+\begin{mytheorem}\label{enqDeqTime}
 Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
@@
 $q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
-\end{lemma}
+\end{mytheorem}
 \begin{proof}

### 2023-01-10 podc_paper/boundedspace.tex
shortening

@@
 
-Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
-to omit polylogarithmic factors; see Appendix \ref{sec::GCanalysis} for precise bounds.}
-Since all processes may run GC,
-a total of $\tilde{O}(G)$ steps is spent on a GC phase.
-There are at least $G$ operations between two GC phases,  so
-GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
-Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.
+A GC phase on a node takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
+If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
+There are at least $G$ operations between two GC phases,  so for each node,
+GC  contributes only $O(\log(p+q))$ steps to each operation in an amortized sense.
+Adding up over all nodes an operation visits, 
+an operation takes $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
+%Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-10 podc_paper/conclusion.tex
shortening

@@
 \here{Can omit rest of this paragraph to save space}
-An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
-A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
-An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
+%An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
+%A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
+%An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
 \here{doublecheck all preceding claims for accuracy.}

### 2023-01-10 podc_paper/correctness.tex
shortening

@@
 
-After proving some basic invariants in Section \ref{sec::basicProperties},
-we show in Section \ref{sec::propagating} that performing a double refresh at each node
-is sufficient to propagate an operation to the root.
-In Section \ref{sec::tracingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
-routines correctly navigate through the \ordering\ tree.
-Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.
+After proving some basic properties in Section \ref{sec::basicProperties},
+we show in Section \ref{sec::propagating} that a double refresh at each node
+suffices to propagate an operation to the root.
+In Section \ref{sec::tracingCorrect} we show \op{GetEnqueue} and \op{IndexDequeue}
+correctly navigate through the  tree.
+Finally, we prove linearizability in Section \ref{sec::linearizability}.
 

### 2023-01-10 podc_paper/algorithm-GC.tex
cutting to save space
shortening and editing

@@
 Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
-the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
+the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $(p\ceil{\log p})^2$.
 This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
@@
 \vspace*{-4mm}
-\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $(p \ceil{\log p})^2$.}
 \end{figure}

### 2023-01-10 podc_paper/algorithm.tex
shortening and editing

@@
 
-Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
-Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
+Each block in an internal \node\ $\var{v}$ stores the indices of the last direct subblock in $\var{v}$'s left and right child in the fields \eleft\ and \eright.  
+Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
 \begin{eqnarray}\label{defsubblock}
-v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
-v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
+\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
+\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].
 \end{eqnarray}
-This allows us to navigate to a block's direct subblocks easily.
-Blocks also store prefix sums of the numbers of enqueues and dequeues:
-the block in $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
-These fields allow us to pinpoint the  location of an operation among the subblocks of a given block.
+The \eleft\ and \eright\ fields allow us to navigate to a block's direct subblocks.
+Blocks also store some prefix sums:
+%the block in 
+$\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
+that store the total numbers of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
+We use these to search for a particular operation.
+% pinpoint the  location of an operation among the subblocks of a given block.
 For example, consider finding the $r$th enqueue $E_r$ in the linearization.
@@
 among the direct subblocks of $B$ in that child.
-Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
-$E_r$.
+Thus, we work our way down the tree until we find the leaf block that  stores 
+$E_r$ explicitly.
 %Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
@@
 We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
-and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
+and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
 for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
-All information needed for this search process can be derived from the 
-\eleft, \eright\ and \fld{sum\sub{enq}} fields.
+%All information needed for this search process can be derived from the 
+%\eleft, \eright\ and \fld{sum\sub{enq}} fields.
 

### 2023-01-10 podc_paper/analysis.tex
shortening and editing

@@
 \begin{proposition}\label{casbound}
-Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
+Each \op{Enqueue} or \op{Dequeue} operation performs $O(\log p)$ \op{CAS} instructions.
 \end{proposition}
 \begin{proof}
-An operation invokes \nf{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
-A \nf{Refresh} does at most 7 \nf{CAS} steps: one in line \ref{cas} and two from each \nf{Advance} in line \ref{helpAdvance} or~\ref{advance}.
+An operation invokes \op{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
+A \op{Refresh} does at most 7 \op{CAS} steps: one in line \ref{cas} and two from each \op{Advance} in line \ref{helpAdvance} or~\ref{advance}.
 \end{proof}
@@
 $O(\log p\log c + \log q_e+ \log q_d)$ steps,
-where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
+where $q_d$ is the size of the queue when the \op{Dequeue} is linearized and 
 $q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.

### 2023-01-10 podc_paper/boundedspace.tex
shortening and editing

@@
 
-In our implementation, an operation remains in the \nf{blocks} arrays forever. 
-Thus, the space used is proportional to the number of operations that have been invoked.
-Now, we sketch a mechanism to remove blocks that are no longer needed to use space
+Operations remain in the \fld{blocks} arrays forever. 
+This uses space proportional to the number of enqueues that have been invoked.
+Now, we describe how to remove blocks that are no longer needed to use space
 polynomial in $p$ and $q$, while maintaining polylogarithmic (amortized) step complexity.  For lack of space, details are in Appendix~\ref{reducing-details}.
-
 We replace the \fld{blocks} array in each node by a red-black tree (RBT)
@@
 blocks in a RBT are sorted by \fld{index}.
-The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
+The attempt to install a new block in $\fld{blocks}[i]$  on line \ref{cas}
 is replaced by an attempt to insert a new block with index $i$ into the RBT.
-Accessing the block in entry $i$ in the \fld{blocks} array can then be 
-replaced by searching the RBT for the block with index $i$.
+Accessing the block in entry $i$ in the \fld{blocks} array is 
+replaced by searching the RBT for the  index $i$.
 The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
@@
 To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
-If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
+If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=(p\ceil{\log p})^2$, it does GC to remove obsolete blocks from the node's RBT.
 To determine which blocks can be thrown away, the node has an array $\var{last}[1..p]$ where 
@@
 is a dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
-its own leaf.
-\here{What about enqueues not finding required block?}
+its own leaf.  If $op$ is an enqueue, it can simply terminate.
 
-A GC phase on a node takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
-If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
+We show that GC ensures each RBT has size $O(pq+p^3\log p)$, so RBT operations take $O(\log(p+q))$ time.
+Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
+$O(\log p(\log p +\log q))$ steps.
+A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
+If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(\frac{G\log(p+q)}{\log p})$ steps.
 There are at least $G$ operations between two GC phases,  so for each node,
-GC  contributes only $O(\log(p+q))$ steps to each operation in an amortized sense.
+GC  contributes only $O(\frac{\log(p+q)}{\log p})$ steps to each operation in an amortized sense.
 Adding up over all nodes an operation visits, 
-an operation takes $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
+an operation spends $O(\log(p+q))$ steps doing GC in an amortized sense.
 %Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-10 podc_paper/correctness-GC.tex
shortening and editing

@@
 We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
-Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
+Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $(p\ceil{\log p})^2$.
 Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
@@
 \begin{mytheorem}\label{spaceBound}
-The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
+The queue data structure uses a maximum of $O(pq_{max}+p^3\log^2 p)$ words of memory at any time.
 \end{mytheorem}
@@
 So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
-to be $p^2\ceil{\log p}$.
+to be $(p\ceil{\log p})^2$.
 \end{proof}
 
-It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log p(q_{max}+p^3)$.
+\here{It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log^2 p(q_{max}+p^3)$.}
 
@@
 \here{this proof may need some modification since code was changed to change how split point is computed}
-It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
+It follows from \Cref{RBTbound} and the fact that $G=(p\ceil{\log p})^2$
 that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
@@
 so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
-is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
-Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$, which is
-within a constant factor of the steps it performs excluding GC.
+is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max})/\log p)$.
+Hence each operation's amortized number of GC steps is $O(\log(p+q_{max}))$, which is less than
+the bound on the number of non-GC steps it performs.
 \end{proof}

### 2023-01-10 podc_paper/correctness.tex
shortening and editing

@@
 A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
-when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
+when the block is created.
+% at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
 Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
-(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
+(from \nl\ to a non-\nl\ value), so it is written only once.
 Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
-(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
+(from \nl\ to a non-\nl\ value), so blocks are permanently added to nodes.
 Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
@@
 \begin{eqnarray*}
-\var{num\sub{enq}} 
-&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
-&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
-&&\mbox{ }- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
+\var{num\sub{enq}} \hspace*{-2mm}
+&=& \hspace*{-2mm} \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
+&=& \hspace*{-2mm}\var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
+&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
 \end{eqnarray*}
-It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total number of enqueues contained in the blocks
+By \Cref{lem::sum}, $num\sub{enq}$ is the  number of enqueues  in 
 $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
@@
 
-In the next lemma, we show that the \fld{size} field in each block in the root is computed correctly.
-\begin{lemma}\label{sizeCorrectness}
+A simple proof (in Appendix \ref{app::tracingDetails}) shows that \fld{size} fields are computed correctly.
+\begin{restatable}{lemma}{sizeCorrectRes}
+\label{sizeCorrectness}
 If the operations of $\var{root}.\fld{blocks}[0..b]$ are applied sequentially in the order of~$L$ on an initially empty queue, the resulting queue has $\var{root}.\fld{blocks}[b].\size$ elements.  
-\end{lemma}
-\begin{proof}
-We prove the claim by induction on $b$. 
-The base case when ${b=0}$ is trivial, since the queue is initially empty and 
-$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
-Assuming the claim holds for $b-1$, we prove it for $b$.
-The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
-at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
-By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
-of block $B$ are performed.
-By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
-are the number of enqueues and dequeues contained in $B$.
-Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
-is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
-\end{proof}
+\end{restatable}
 

### 2023-01-10 podc_paper/introduction.tex
shortening and editing

@@
 $\Omega(p)$ CAS instructions, even in an amortized sense.
-For the space-bounded version, we unlink unneeded memory from our data structure.
-We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
+For the space-bounded version, we unlink unneeded objects from our data structure.
+We do not address the orthogonal problem of reclaiming memory; we assume a safe
 garbage collector, such as the highly optimized one that Java provides.

### 2023-01-11 podc_paper/analysis.tex
Shortening--now it fits in 15 pages.

@@
 \section{Analysis}
-In this section, we analyze the number of steps and the number of \op{CAS} instructions performed by operations.
-First, we prove some claims about the size and operations of a block.
+We now analyze the number of steps and the number of \op{CAS} instructions performed by operations.
 
@@
 Let the blocks in the root be $B_1, \ldots, B_\ell$.
-The doubling search used to compute $b_e$ takes $O(\log (b-b_e))$ steps,
+The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
 so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
-If $b \leq b_e+1$, then this is trivial, so assume for the rest of the proof that $b>b_e+1$.
-
+If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
+%
 As shown in \Cref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
@@
 
-We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.
+We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.
 

### 2023-01-11 podc_paper/boundedspace.tex
Shortening--now it fits in 15 pages.

@@
 This uses space proportional to the number of enqueues that have been invoked.
-Now, we describe how to remove blocks that are no longer needed to use space
-polynomial in $p$ and $q$, while maintaining polylogarithmic (amortized) step complexity.  For lack of space, details are in Appendix~\ref{reducing-details}.
+Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
+polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic.  For lack of space, details are in Appendix~\ref{reducing-details}.
 We replace the \fld{blocks} array in each node by a red-black tree (RBT)
@@
 is replaced by an attempt to insert a new block with index $i$ into the RBT.
-Accessing the block in entry $i$ in the \fld{blocks} array is 
+Accessing the block in $\fld{blocks}[i]$ is 
 replaced by searching the RBT for the  index $i$.
@@
 
-To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
+To maintain good amortized time, we periodically do a garbage collection (GC) phase.
 If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=(p\ceil{\log p})^2$, it does GC to remove obsolete blocks from the node's RBT.
-To determine which blocks can be thrown away, the node has an array $\var{last}[1..p]$ where 
-each process writes the index of the last block in
-containing an enqueue whose element it dequeued.
+To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
+each process writes the index of the last block in the root
+containing a null dequeue or an enqueue whose element it dequeued.
 To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
@@
 Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
-in $\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the RBT 
-with indices smaller than $m-1$.
-Fortunately, there is an efficient \op{Split} operation that can remove
-all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+in $\fld{root.blocks}[1..m-1]$ have all been dequeued, so GC can discard all subblocks of those.
+Fortunately, there is an  RBT \op{Split} operation that can remove
+these blocks in logarithmic time \cite[Sec.~4.2]{Tar83}.
 
@@
 If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(\frac{G\log(p+q)}{\log p})$ steps.
-There are at least $G$ operations between two GC phases,  so for each node,
-GC  contributes only $O(\frac{\log(p+q)}{\log p})$ steps to each operation in an amortized sense.
+Thus, each node contributes $O(\frac{\log(p+q)}{\log p})$ steps to each operation in an amortized sense
+since there are at least $G$ operations between two GC phases.
 Adding up over all nodes an operation visits, 

### 2023-01-11 podc_paper/conclusion.tex
Shortening--now it fits in 15 pages.

@@
 
-Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
-However, our approach does have a higher cost in the best case (for example, when an operation
-runs in isolation).
-It may be possible to make our queue adaptive by having an operation capture a starting node
-in the \ordering\ tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
+Our focus was on optimizing step complexity for worst-case executions.
+However, our queue has a higher cost than the MS-queue in the best case (when an operation
+runs by itself).
+Perhaps our queue could be made adaptive by having an operation capture a starting node
+in the \ordering\ tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than starting at a statically assigned leaf.
 Another way to make our queue more practical might be to use it as the slow path in the
-fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
+fast-path slow-path methodology  \cite{10.1145/2370036.2145835} to
 get a queue that has good performance in practice while also having good worst-case step complexity.

### 2023-01-11 podc_paper/correctness.tex
Shortening--now it fits in 15 pages.

@@
 By \Cref{lem::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
-Since each operation creates a block $B$ containing the operation and then calls \op{Append}($B$), 
-the operation is propagated to the root before the operation terminates, by \Cref{lem::appendExactlyOnce},
+By \Cref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
 so it appears in $L$.

### 2023-01-11 podc_paper/correctness-details.tex
forgot to add the file correctness-details earlier

% !TEX root =  podc-submission.tex

\section{Detailed Proofs for Section \ref{sec::correctness}}
\label{app::tracingDetails}

Detailed proofs omitted from Section \ref{sec::correctness} for lack of space appear here.

\sumRes*
\begin{proof}
Initially, each \fld{blocks} array only contains an empty block $B_0$ in location 0.
By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
Moreover, $B_0.\fld{sum\sub{enq}} = B_0.\fld{sum\sub{deq}} = 0$, so the claim is true.

We show that each installation of a block $B$ into some location $\var{v}.\fld{blocks}[i]$ preserves the claim,
assuming the claim holds before this installation.  We consider two cases.

If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1]+1$ and
$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.

Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is $\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

\superRelationRes*
\begin{proof}
We first show $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($v.\fld{parent}$) that installs the superblock of $B$ 
in $v.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $v.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $v.\fld{parent}.\head$ before $r$.
When $r$ occurs, $v.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $v.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored $B.\fld{super}$ was read from $v.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $v$;
otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $v.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $v.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
\end{proof}

\indexDequeueRes*
\begin{proof}
We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $v$'s parent, we prove it for $v$.
Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.

Next, the position of the required dequeue in $D(B')$ is computed in 
lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
We first add the number of dequeues in the subblocks of $B'$ in $v$ that precede $B$ on line \ref{computeISuperStart}.
If $v$ is the right child of its parent, then all of the subblocks of $B'$ from $v$'s left sibling
also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.

Finally, \op{IndexDequeue} is called recursively on $v$'s parent.
Since $B$ has been propagated to the root, so has its superblock $B'$.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}

\getEnqRes*
\begin{proof}
We prove the claim by induction on the height of node $v$.
If $v$ is a leaf, the hypothesis implies that $i=1$ and the block $v.\fld{blocks}[b]$ represents 
an enqueue whose argument is stored in $v.\fld{blocks}[b].\fld{element}$.
\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.

Assuming the claim holds for $v$'s children, we prove it for $v$.
Let $B$ be $v.\fld{blocks}[b]$.
By (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of enqueues in $E(B)$ that come from $B$'s subblocks in $v$'s left child.
Thus, $dir$ is set to the direction for the child of $v$ that contains the required enqueue.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
By \Cref{lem::sum}, $v.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
enqueues in $E'$ before the enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

\sizeCorrectRes*
\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivial, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of enqueues and dequeues contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

### 2023-01-11 podc_paper/correctness-GC.tex
Shortening--now it fits in 15 pages.
added acmart.cls

@@
 
+\here{Should add some discussion about the fact that we keep the newest finished block, in case some operation contained in a block with index i also needs to look at fields of the block with index i-1}
+
 \begin{lemma}\label{blocknotfound}
@@
 All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
-There are at most $p$ operations in block $m$ of the root by \Cref{blockSizeGC}
+There are at most $p$ operations in block $m$ of the root %by \Cref{blockSizeGC}
 and we showed above that there are at most $2q_{max} + 2p$ in blocks 

### 2023-01-11 podc_paper/introduction.tex
added acmart.cls

@@
 (To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
-and divide the number of steps in the execution by the number of operations initated in the execution.)
+and divide the number of steps in the execution by the number of operations initiated in the execution.)
 Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but

### 2023-01-11 podc_paper/algorithm-GC.tex
Minor edits; fixed amortized complexity bound for GC

--- 2023-01-10 podc_paper/algorithm-GC.tex
@@
 
-Blocks no longer require the \fld{super} field.  This was used to quickly find the blocks 
+Blocks no longer require the \fld{super} field.  It was used to quickly find a block's 
 superblock in the parent node's \fld{blocks} array, but this can now be done efficiently 
@@
 The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
-used to implement GC.
+used to implement the garbage collection (GC) phase.
 When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
@@
 are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
-The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in \var{v}
-of $B_{root}$ (or any older block in the root).
+The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
+of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
 As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
@@
 In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
-(since $\var{v}'$ determined that all blocks older than that are safe to discard).
+(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).
 

### 2023-01-11 podc_paper/algorithm.tex
Minor edits; fixed amortized complexity bound for GC

@@
 To perform an operation on the queue, a process $P$ appends a new block containing that  
-operation to the blocks array in $P$'s leaf.
+operation to the \fld{blocks} array in $P$'s leaf.
 Then, $P$ attempts to propagate the operation to each node along the path from that leaf to the root of the tree.
@@
 \Eric{Reworded following parag after Hossein's comments of Jan 4}
-To propagate operations from a node $v$'s children to $v$, $P$ first observes
-the blocks in both of $v$'s children that are not already in $v$,
+To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ first observes
+the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
 creates a new block by combining information from those blocks, and attempts to append this 
-new block to $v$'s blocks array using a \op{CAS} instruction.
+new block to $\var{v}$'s \fld{blocks} array using a \op{CAS} instruction.
 Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
-\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
-A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $v$.
+\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
+A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $\var{v}$.
 However, since a successful \op{Refresh} propagates multiple pending operations 
-from $v$'s children to $v$,
-we can prove that if two \op{Refresh}es by $P$ on $v$ fail,
-then $P$'s operation has been propagated to $v$ by some other process, so $P$ can continue 
+from $\var{v}$'s children to $\var{v}$,
+we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
+then $P$'s operation has been propagated to $\var{v}$ by some other process, so $P$ can continue 
 onwards towards the~root.
@@
 \label{construct}
-Construct a block for node $v$ that represents the operations contained in consecutive blocks in $v$'s children, as required for a \op{Refresh}.
+Construct a block for node $\var{v}$ that represents the operations contained in consecutive blocks in $\var{v}$'s children, as required for a \op{Refresh}.
 \item
 \label{findinroot}
-Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
+Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's \fld{blocks} array.
 \item
@@
 Each block in a leaf represents one operation.
-Each block $B$ in an internal node $v$ results from merging
-several consecutive blocks from each of $v$'s children.
-The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
+Each block $B$ in an internal node $\var{v}$ results from merging
+several consecutive blocks from each of $\var{v}$'s children.
+The merged blocks in $\var{v}$'s children are called the \emph{direct subblocks} of $B$.
 A block $B'$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
@@
 and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
-If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
-$v$'s left child
-and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
+If $B$ is a block in an internal node $\var{v}$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
+$\var{v}$'s left child
+and $B^R_1,\ldots,B^R_r$ from $\var{v}$'s right child, then $B$'s operation sequences are defined by the concatenations 
 \begin{equation}
@@
 A block in a leaf represents a single enqueue or dequeue.  The block's \fld{element} field stores the value
-enqueued for an enqueue operation, or \nl\ if the operation is a dequeue.
+enqueued if the operation is an enqueue, or \nl\ if the operation is a dequeue.
 
@@
 
-Each block in an internal \node\ $\var{v}$ stores the indices of the last direct subblock in $\var{v}$'s left and right child in the fields \eleft\ and \eright.  
+Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child.  
 Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
@@
 finds the block  containing $E_r$.
-If we know a block $B$ in a \node\ $v$ contains $E_r$,
-we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains $E_r$
+If we know a block $B$ in a \node\ $\var{v}$ contains $E_r$,
+we can use the \fld{sum\sub{enq}} field again to determine which child of $\var{v}$ contains $E_r$
 and then do a binary search
@@
 we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
-A \op{Refresh} on \node\ $v$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $v$'s children and stores 
+A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
 $h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
-Then, we can compute $B.\fld{sum\sub{enq}}$ as $v.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + v.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
+Then, we can compute $B.\fld{sum\sub{enq}}$ as $\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
 For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
@@
 The only remaining field is $B.\fld{super}$.  When the block 
-$B$ is created for a \node\ $v$, we do not yet know where its
+$B$ is created for a \node\ $\var{v}$, we do not yet know where its
 superblock will eventually be installed in $B$'s parent.
-So, we leave $B.\fld{super}$ blank.  After $B$ is installed 
-in $v.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
-when they attempt to advance $v.\fld{head}$ from $h$ to $h+1$,
-using a value they read from the \fld{head} field of $v$'s parent.
+So, we leave $B.\fld{super}$ blank.  
+Soon after $B$ is installed,
+a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
+set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
+We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of $B'$
+by at most 1.
+%After $B$ is installed 
+%in $\var{v}.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
+%when they attempt to advance $\var{v}.\fld{head}$ from $h$ to $h+1$,
+%using a value they read from the \fld{head} field of $\var{v}$'s parent.
 %As mentioned above, this might not be the exact index of $B$'s superblock, but we
@@
 	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}
-	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}
-	    \EndIf
+    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\label{idsup1}
+	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
+	    \EndIf\label{idsup3}
 	    \State \linecomment{compute index \var{i} of dequeue in superblock}
@@
 
-%\begin{algorithm}
-%\caption{\tt{\sl{Node}}}
-%\begin{algorithmic}[1]
-%\setcounter{ALG@line}{25}
-%
-%\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-%\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
-%\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-%\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
-%\While{\nf{start<end}}
-%\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
-%\If{\nf{blocks[mid].sum\sub{enq}<x}}
-%\State \nf{start \assign\ mid+1}
-%\Else
-%\State \nf{end \assign\ mid}
-%\EndIf
-%\EndWhile
-%\State\Return \nf{start}
-%\EndFunction{BinarySearch}
-%
-%\end{algorithmic}
-%\end{algorithm}
-
-%\begin{algorithm}
-%\caption{\tt{\sl{Root}}}
-%\begin{algorithmic}[1]
-%\setcounter{ALG@line}{36}
-%\Statex
-%\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-%\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
-%\com\ the \nf{b}th block in the \nf{root}.}
-%
-%\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-%\State \tt{start \assign\ end-1} \label{dsearchStart}
-%\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
-%\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
-%\EndWhile \label{dsearchEnd}
-%\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-%\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-%\State\Return \tt{<b,i>}
-%\EndFunction{DoublingSearch}
-%\end{algorithmic}
-%\end{algorithm}
-
 \begin{figure}
@@
 It suffices to propagate the operation to the root and
-then use its position in the linearization for future \nf{Dequeue}
+then use its position in the linearization for future \op{Dequeue}
 operations.
@@
 its position in the root using
-\nf{IndexDequeue} and then computes its response by calling \nf{FindResponse}.
+\op{IndexDequeue} and then computes its response by calling \op{FindResponse}.
 
@@
 
-Once a \op{Dequeue} is appended to a block of the process's leaf and propagated to the root,
-the \opemph{IndexDequeue} routine finds the \op{Dequeue}'s location in the root.
+Once a dequeue is appended to a block of the process's leaf and propagated to the root,
+the \opemph{IndexDequeue} routine finds the dequeue's location in the root.
 More precisely, \opa{IndexDequeue}{v, b, i}
 computes the block in the root and the rank
-within that block  of the \var{i}th \op{Dequeue} of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
-Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
-whose CAS wrote a block \var{B} into \var{v}.\fld{blocks}[\var{b}].
-Let $par$ be \var{v}.\fld{parent}. Without loss of generality, assume 
-\var{v} is the left child of $par$. Let
-$R_{par}$ be the first successful \op{Refresh} on $par$ whose \op{CreateBlock} reads a
-value greater than \var{b} from $\var{par}.\fld{left.head}=\var{v}.\fld{head}$ at line \ref{createEndLeft} and therefore creates a block $B'$ that contains
-\var{B} as a subblock. 
-%Let $b'$ be the index of $B'$ in $par$.\fld{blocks}.
+within that block  of the \var{i}th dequeue of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
+Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may be off by one from the superblock's true index.
+The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the dequeue's 
+rank within the superblock's sequence of dequeues, using  (\ref{defSeqs}).
 
-Since the index of \var{B}'s superblock $B'$ is not known until $B$ is
-propagated to $par$, $R\sub{\var{v}}$ cannot set the \nf{super} field of ${B}$ when
-it creates \var{B}. 
-It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (since
-there could be up to $p$ direct subblocks).
-Soon after $B$ is installed,
-a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
-set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
-We shall show that this happens soon enough that $B.\fld{super}$ differs from the true index of $B'$
-by at most 1.
 %Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
 %As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
-%We shall show \nf{$B$.\fld{super}}  differs
+%We shall show {$B$.\fld{super}}  differs
 %from the index of $B'$ by at most~1.
 
-To compute the response of the $i$th \nf{Dequeue} in the $b$th block
+To compute the response of the $i$th \op{Dequeue} in the $b$th block
 of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
@@
 \opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
-$i$th enqueue in the $b$th block of \typ{Node} $v$. 
+$i$th enqueue in the $b$th block of \typ{Node} $\var{v}$. 
 The range of subblocks of a block is determined using the \fld{end\sub{left}} and

### 2023-01-11 podc_paper/analysis.tex
Minor edits; fixed amortized complexity bound for GC

@@
 
-The following lemma is useful in bounding the time to \op{GetEnqueue}
+The following lemma is useful in bounding the time for \op{GetEnqueue}
 \begin{lemma}\label{blockSize}
@@
 By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
-$B$ has at most $c$ direct subblocks.
+so $B$ has at most $c$ direct subblocks.
 \end{proof}
@@
 \op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
-Each node $v$'s search is within the subblocks of one block in $v$'s parent.
-By \Cref{blockSize}, each search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
+Each node $\var{v}$'s search is within the subblocks of one block in $\var{v}$'s parent.
+By \Cref{blockSize}, each such search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
 \end{proof}

### 2023-01-11 podc_paper/boundedspace.tex
Minor edits; fixed amortized complexity bound for GC

@@
 
-\section{Reducing Space Usage}
+\section{Bounding Space Usage}
 \label{reducing}
@@
 Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
-concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
-can fail if a concurrent \op{Refresh} does a successful CAS.
-Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other node in the RBT.
+concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
+can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
+Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
 Thus, we can use a particularly simple concurrent RBT implementation.
 A sequential RBT can be made persistent using the classic node-copying technique of 
-Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
-tree are performed by making a new copy of each node $v$ that must be modified, as well
-as each node along the path from the root to $v$.
-The tree reachable from the new copy of the root is the result of applying the tree operation.
+Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
+RBT are performed by making a new copy of each RBT node $x$ that must be modified, as well
+as each RBT node along the path from the RBT's root to~$x$.
+The RBT reachable from the new copy of the root is the result of applying the RBT operation.
 This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
 Once a process has performed an update to the RBT representing the blocks of a node 
-$v$ in the \ordering\ tree, 
-it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
+$\var{v}$ in the \ordering\ tree, 
+it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
 A search in the RBT can simply read the pointer to the RBT root and perform a standard
 sequential search on it.
-Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.
+Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.
 
@@
 To maintain good amortized time, we periodically do a garbage collection (GC) phase.
-If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=(p\ceil{\log p})^2$, it does GC to remove obsolete blocks from the node's RBT.
+If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
 To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
@@
 
-We show that GC ensures each RBT has size $O(pq+p^3\log p)$, so RBT operations take $O(\log(p+q))$ time.
+Our GC scheme ensures each RBT has  $O(q+p^2\log p)$ blocks, so RBT operations take $O(\log(p+q))$ time.
 Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
-$O(\log p(\log p +\log q))$ steps.
+$O(\log p\log(p + q))$ steps.
 A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
-If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(\frac{G\log(p+q)}{\log p})$ steps.
-Thus, each node contributes $O(\frac{\log(p+q)}{\log p})$ steps to each operation in an amortized sense
-since there are at least $G$ operations between two GC phases.
-Adding up over all nodes an operation visits, 
-an operation spends $O(\log(p+q))$ steps doing GC in an amortized sense.
+If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
+Since there are at least $G$ operations between two GC phases, each node contributes $O(\log(p+q))$ steps to each operation in an amortized sense.
+Adding up over all nodes an operation may have to do GC on, 
+an operation spends $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
+So the total amortized step complexity is  $O(\log p \log(p+q))$ per operation.
 %Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-11 podc_paper/conclusion.tex
Minor edits; fixed amortized complexity bound for GC

@@
 in the \ordering\ tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than starting at a statically assigned leaf.
-Another way to make our queue more practical might be to use it as the slow path in the
+A possible application of our queue  might be to use it as the slow path in the
 fast-path slow-path methodology  \cite{10.1145/2370036.2145835} to
@@
 depends on $p$.
+Since a queue is also a bag, it is the first lock-free bag we know of that has polylogarithmic step complexity.
 
@@
 could be applied to obtain other lock-free
-data structures with a polylogarithmic number of steps per operation.
-For example, we could implement a  vector data structure that stores a sequence and
-provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
-\opa{get}{i} to read the \var{i}th element in the sequence, and
-\opa{index}{e} to compute the position of element \var{e} in the sequence.
-\here{Can omit rest of this paragraph to save space}
+data structures with a polylogarithmic step complexity.
+For example, we can easily adapt our routines to implement a  vector data structure that stores a sequence and
+provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
+\opa{Get}{i} to read the \var{i}th element in the sequence, and
+\opa{Index}{e} to compute the position of element \var{e} in the sequence.
+%\here{Can omit rest of this paragraph to save space}
 %An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  

### 2023-01-11 podc_paper/correctness-GC.tex
Minor edits; fixed amortized complexity bound for GC

--- 2023-01-10 podc_paper/correctness-GC.tex
@@
 The following observation describes
-how the set of blocks in the RBT can be modified.
+how the set of blocks in a node's RBT can be modified.
 
@@
 
-Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$ does sets
-$T'\assign\op{Split}(T,s)$ operation.
+Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
+$T'\assign\op{Split}(T,s)$.
 Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
@@
 
-It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log p(q_{max}+p^3)$.
+\here{It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log^2 p(q_{max}+p^3)$.}
 
@@
 
-Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}.
+Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
+and show that the amortized number of GC steps each operation takes is also $O(\log p\log(p+q_{max}))$.
 If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
@@
 is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
-Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$, which is
-within a constant factor of the steps it performs excluding GC.
+Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$.
 \end{proof}

### 2023-01-11 podc_paper/correctness-details.tex
Minor edits; fixed amortized complexity bound for GC

@@
 \begin{proof}
-Initially, each \fld{blocks} array only contains an empty block $B_0$ in location 0.
+Initially, each \fld{blocks} array  contains only an empty block $B_0$ in location 0.
 By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
@@
 We first show $B.\fld{super}\leq s$.
-Let $R_s$ be the instance of \op{Refresh}($v.\fld{parent}$) that installs the superblock of $B$ 
-in $v.\fld{parent.blocks}[s]$.
-By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $v.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
+Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs the superblock of $B$ 
+in $\var{v}.\fld{parent.blocks}[s]$.
+By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
 By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
-that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $v.\fld{parent}.\head$ before $r$.
-When $r$ occurs, $v.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
+that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
+When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
 \ref{cas} succeeds.
-So, by \Cref{lem::headPosition}, $v.\fld{parent}.\head \leq s$ when $r$ occurs.
-Since the value stored $B.\fld{super}$ was read from $v.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
+So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
+Since the value stored $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
 
 Next, we show that $B.\fld{super}\geq s-1$.
-The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $v.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
+The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
 So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
-By \Cref{lem::headProgress}, $v.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
-$\op{Refresh}(v.\fld{parent})$ that installed the block in $v.\fld{parent.blocks}[s-1]$.
-Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $v.\fld{parent}.\head$ at line \ref{readHead}.
-This read $r'$ must be before $B$ is installed in $v$;
+By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
+$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
+Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
+This read $r'$ must be before $B$ is installed in $\var{v}$;
 otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
-$v.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
-Now, consider the call to \op{Advance}($v, b$) that writes $B.\fld{super}$.
+$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
+Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
 It is invoked either 
-at line \ref{helpAdvance} after seeing $v.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
-or at line \ref{advance} after ensuring $v.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
+at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
+or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
 Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
-By \Cref{nonDecreasingHead}, $v.\fld{parent}.\head$ is non-decreasing, so 
-the value this \op{Advance} reads in $v.\fld{parent}.\head$ and
-writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $v.\fld{parent}.\head$.
+By \Cref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
+the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
+writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
 \end{proof}
@@
 \begin{proof}
-We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
-Assuming the claim holds for $v$'s parent, we prove it for $v$.
-Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
-\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
+We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
+Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
+Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
+\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
 By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
-The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.
+The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]$.
 
@@
 lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
-We first add the number of dequeues in the subblocks of $B'$ in $v$ that precede $B$ on line \ref{computeISuperStart}.
-If $v$ is the right child of its parent, then all of the subblocks of $B'$ from $v$'s left sibling
+We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line \ref{computeISuperStart}.
+If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
 also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.
 
-Finally, \op{IndexDequeue} is called recursively on $v$'s parent.
+Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
 Since $B$ has been propagated to the root, so has its superblock $B'$.
@@
 \begin{proof}
-We prove the claim by induction on the height of node $v$.
-If $v$ is a leaf, the hypothesis implies that $i=1$ and the block $v.\fld{blocks}[b]$ represents 
-an enqueue whose argument is stored in $v.\fld{blocks}[b].\fld{element}$.
+We prove the claim by induction on the height of node $\var{v}$.
+If $\var{v}$ is a leaf, the hypothesis implies that $i=1$ and the block $\var{v}.\fld{blocks}[b]$ represents 
+an enqueue whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
 \op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.
 
-Assuming the claim holds for $v$'s children, we prove it for $v$.
-Let $B$ be $v.\fld{blocks}[b]$.
+Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
+Let $B$ be $\var{v}.\fld{blocks}[b]$.
 By (\ref{defSeqs}),
@@
 By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
-of enqueues in $E(B)$ that come from $B$'s subblocks in $v$'s left child.
-Thus, $dir$ is set to the direction for the child of $v$ that contains the required enqueue.
+of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
+Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue.
 Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
 Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
-By \Cref{lem::sum}, $v.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
+By \Cref{lem::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
 enqueues in $E'$ before the enqueues of block $B'$, so

### 2023-01-11 podc_paper/correctness.tex
Minor edits; fixed amortized complexity bound for GC

@@
 Let $R$ be an instance of \opa{Refresh}{v} whose call to \op{CreateBlock} returns a non-\nl\ block.
-When $R$ terminates, \var{v}.\head\ is greater than the value $R$ reads from it at line \ref{readHead}.
+When $R$ terminates, \var{v}.\head\ is strictly greater than the value $R$ reads from it at line \ref{readHead}.
 \end{lemma}
@@
 
-Assume the claim holds before a change to $\var{v}.\fld{blocks}$, which can be made only
+Assume the claims hold before a change to $\var{v}.\fld{blocks}$, which can be made only
 by a successful \op{CAS} at line \ref{cas}.
@@
 Since $\var{v}.\fld{blocks}[h]$ is \nl\ before the CAS, $\var{v}.\head \leq h$ by the hypothesis.
-Since $h$ was read from $\var{v}.\fld{blocks}[h]$ earlier at line \ref{readHead}, 
-$\var{v}.\head \geq h$ by \Cref{lem::headPosition}.
-So, $h=\var{v}.\head$ and a change to $\var{v}.\fld{blocks}[\var{v}.\head]$ preserves the invariant.
+Since $h$ was read from $\var{v}.\head$ earlier at line \ref{readHead}, the current value of 
+$\var{v}.\head$ is at least $h$ by \Cref{nonDecreasingHead}.
+So, $\var{v}.\head=h$ when the \op{CAS} occurs and a change to $\var{v}.\fld{blocks}[\var{v}.\head]$ preserves the invariant.
 
@@
 or at line \ref{advance} after the \op{CAS} at line \ref{cas} ensures $\var{v}.\fld{blocks}[h]\neq\nl$.
-For the third claim, observe that prior to incrementing $\var{v}.\head$ at line \ref{incrementHead},
+For the third claim, observe that prior to incrementing $\var{v}.\head$ to $i+1$ at line \ref{incrementHead},
 the \op{CAS} at line \ref{setSuper1} ensures that $\var{v}.\fld{blocks}[i].super\neq \nl$.
@@
 A block $B$ contains the operations in $B$'s subblocks in leaves of the tree.
-An operation by process $P$ appears in one block of $P$'s leaf, so
+An operation by process $P$ appears in just one block of $P$'s leaf, so
 an operation 
@@
 The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
-on line \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
+on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
 from the definition of subblocks.  See Appendix 
-\ref{app::tracingDetails} for a detailed proof.
+\ref{app::tracingDetails} for a detailed proof of \Cref{lem::sum}.
 
@@
 
-This allows us to prove that every block a Refresh installs contains at least one operation.
+This allows us to prove that every block a \op{Refresh} installs contains at least one operation.
 
@@
 The \op{Refresh} that installed $B$ got $B$ as the response to its call to \op{CreateBlock} on line \ref{invokeCreateBlock}.
-Thus, at line \ref{testEmpty} $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
+Thus, at line \ref{testEmpty}, $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
 By \Cref{lem::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
@@
 Then, $B$ is a subblock of some block in $\var{v}$, since  $B'.\eleft$ is greater than or equal to $B$'s index
-in $\var{v}.\fld{left.blocks}$.
+$i$ in $\var{v}.\fld{left.blocks}$.
 Hence $\var{v}$ contains $op$, as required.
@@
 &=& \hspace*{-2mm}\var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
-&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}
+&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
 \end{eqnarray*}
@@
 
-We now show that if a process fails its \op{Refresh}($\var{v}$) twice, some other process must have succeeded.
+We now show that a double \op{Refresh} propagates blocks as required.
 
@@
 By \Cref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
-which is before $R_2$ terminates.
+which is before $R_2$'s failed \op{CAS}.
 \end{proof}
@@
 if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
-We also show result each dequeue returns is the same as in the sequential execution  $L$.
+We also show the result each dequeue returns is the same as in the sequential execution  $L$.
 
@@
 Also, if $op_{1}$ terminates before $op_{2}$ begins, then $op_{1}$ 
-is propagated to the root before $op_2$ begins, so $op_2$ appears before $op_2$ in $L$.
+is propagated to the root before $op_2$ begins, so $op_1$ appears before $op_2$ in $L$.
 \end{proof}

### 2023-01-11 podc_paper/introduction.tex
Minor edits; fixed amortized complexity bound for GC

--- 2023-01-10 podc_paper/introduction.tex
@@
 (To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
-and divide the number of steps in the execution by the number of operations initated in the execution.)
+and divide the number of steps in the execution by the number of operations  in the execution.)
 Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but
@@
 that amortized $\Omega(c)$ steps per operation are indeed necessary
-for any CAS-based implementation of a lock-free bag data structure, which provide operations
+for any CAS-based implementation of a lock-free bag data structure, which provides operations
 to insert an element or remove an arbitrary element (chosen non-deterministically).
@@
 
-We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
-whose step complexity is sublinear in $p$.
-Our queue is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
-For ease of presentation, we first give an unbounded-space construction that uses
-$O(\log^2 p + \log q)$ steps per operation,
+We exploit this loophole.  We show  it is, in fact, possible to design a linearizable, wait-free queue
+with step complexity sublinear in $p$.
+Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
+For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
+dequeues take $O(\log^2 p + \log q)$ steps,
 and then modify it to bound the space
-while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
-Both versions use single-word CAS on reasonably-sized words.
+while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
+Both versions use single-word CAS on %reasonably-sized 
+small words.
 Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
@@
 \here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
-Explicitly storing all operations in the nodes would be too costly.
+Explicitly storing all operations in the tree nodes would be too costly.
 Instead, we use a novel implicit representation of sets

### 2023-01-11 podc_paper/related.tex
added acmart.cls
Minor edits; fixed amortized complexity bound for GC

--- 2023-01-07 podc_paper/related.tex
@@
 As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
-Thus, it is lock-free but not wait-free and has an amortized\Hossein{Maybe we can say worst-case.
+Thus, it is lock-free but not wait-free and has an amortized step complexity of $\Theta(p)$ per operation.
+
+%\Hossein{Maybe we can say worst-case.
 I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.}
-\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.
+%\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.
 Also, if we just said ``worst-case'' without ``amortized'' it would be confused with worst-case of an individual operation, which is infinite in this case.
 By the way, saying that previous implementations have $\Omega(p)$ step complexity, even in the amortized sense, makes the new contribution even more of a contrast.}
-step complexity of $\Theta(p)$ per operation.
 
@@
 \texttt{CAS} instructions to one in the best case. 
-Pointers in the doubly-linked list can be inconsistent, but are fixed when needed by traversing the list.
+Pointers in the doubly-linked list can be inconsistent, but are fixed when necessary by traversing the list.
 Although this fixing is rare in practice, it yields an amortized complexity of $\Omega(qp)$ 
@@
 (Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
-Fatourou and Kallimanis \cite{FK14} used their universal construction based on fetch\&add and LL/SC instructions
+Fatourou and Kallimanis \cite{FK14} used their own universal construction based on fetch\&add and LL/SC instructions
 to implement a queue, but its step complexity is also $\Omega(p)$.
@@
 uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
-Jayanti and Petrovic gave a wait-free poly-logarithmic
+Jayanti and Petrovic gave a wait-free polylogarithmic
 queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
-Our \ordering\ tree is similar to the tree structure the use to agree on a linearization ordering.
+Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
 
@@
 \emph{Fetch\&Increment Objects.}
-There are sublinear-time implementations of another object
-with consensus number 2.
-Two papers \cite{ERW12,10.1007/978-3-642-41527-2_20}
-implemented fetch\&increment objects from LL/SC objects
-using a polylogarithmic number of steps per operation.
+There are implementations of fetch\&increment objects that use a polylogarithmic number of steps per operation \cite{ERW12,10.1007/978-3-642-41527-2_20}.
 Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},

### 2023-01-11 podc_paper/algorithm-GC.tex
final edits

@@
 each level instead of using the \fld{super} field.
-
-The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
-used to implement the garbage collection (GC) phase.
-When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
-it calls the new \opemph{AddBlock} routine.
-Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
-the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
-This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
-The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
-calls \op{Help} to help all pending dequeues that have been propagated to the root
-(to ensure that all blocks before $s$ can safely be discarded),
-uses the standard RBT \op{Split} routine \cite{Tar83}
-to remove all blocks with \fld{index} less than $s$,
-inserts the new block
-and finally performs a \op{CAS} to swing the node's \fld{blocks} pointer to the new RBT.
-
-To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
-finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
-or a null dequeue using the \var{last} array.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
-dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
-helped to complete by line \ref{invokeHelp},
-it is safe to discard any blocks in the root older than $B_{root}$, as well
-as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
-are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
-The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
-of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
-As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
-along the path from \var{v} to the root is discarded by another GC phase.
-In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
-(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).
-
-The \opemph{Help} routine is fairly straightforward:  it loops through all leaves and
-helps the dequeue that is in progress there if it has already been propagated to the root.
-The \opemph{Propagated} function is used to determine whether the dequeue has propagated to the root.
 
@@
 		\EndFor
-		\State $B\assign \var{root}.\fld{blocks}[m-1]$
+		\State $B\assign \var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
 	\Else
@@
 		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
-		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$
+		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
 	\EndIf
@@
 		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-		\If{\Call{MaxBlock}{$T$}.\edir < b} \Return \fa
+		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa
 		\Else
@@
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRnumenqGC}
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
@@
 %\end{algorithm}
+
+The new routines, \op{AddBlock}, \op{SplitBlock}, \op{Help} and \op{Propagated} are
+used to implement the garbage collection (GC) phase.
+When a \op{Refresh} or \op{Append} wants to add a new block to a node's \fld{blocks} RBT,
+it calls the new \opemph{AddBlock} routine.
+Before attempting to add the block to a node's RBT, \op{AddBlock} triggers a GC phase on the RBT if 
+the new block's \fld{index} is a multiple of the constant $G$, which we choose to be $p^2\ceil{\log p}$.
+This ensures that obsolete blocks are removed from the RBT once every $G$ times a new block is added to it.
+The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
+calls \op{Help} to help all pending dequeues that have been propagated to the root
+(to ensure that all blocks before $s$ can safely be discarded),
+uses the standard RBT \op{Split} routine \cite{Tar83}
+to remove all blocks with \fld{index} less than $s$,
+inserts the new block
+and finally performs a \op{CAS} to swing the node's \fld{blocks} pointer to the new RBT.
+
+To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
+finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
+or a null dequeue using the \var{last} array.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
+dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
+helped to complete by line \ref{invokeHelp},
+it is safe to discard any blocks in the root older than $B_{root}$, as well
+as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
+are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
+The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
+of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
+As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
+along the path from \var{v} to the root is discarded by another GC phase.
+In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
+(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).
+
+The \opemph{Help} routine is fairly straightforward:  it loops through all leaves and
+helps the dequeue that is in progress there if it has already been propagated to the root.
+The \opemph{Propagated} function is used to determine whether the dequeue has propagated to the root.
+
+In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
+with index~$i$.
+A search for this block may sometimes not find it, if it has already been discarded
+by another process's GC.
+As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
+the enqueue can simply terminate because the fact that the block is gone means that another
+process has helped the enqueue reach the root of the ordering tree.
+Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
+written in the \fld{response} field of the 
+leaf block that represents the dequeue and terminate, since some other process
+will have written the \fld{response} there before discarding the needed block.
+We do not explicitly write this early termination every time we do a lookup in an RBT.
+There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
+on line \ref{SBlookup1} or \ref{SBlookup2} of
+\op{SplitBlock} because the required block has been discarded, 
+we continue doing GC, since we do not a GC phase on one node
+to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
+that were needed.
+Line \ref{substitute} says what to do in this case.

### 2023-01-11 podc_paper/correctness-GC.tex
final edits

@@
 with \fld{index} $b$.
-We describe later what should happen if that block is not found by the search because it has been removed
-by a GC phase.
-\here{remember to come back to this point}
 
-We now establish that definition (\ref{defsubblock}) of a block's subblocks
+We now establish that Definition (\ref{defsubblock}) of a block's subblocks
 still makes sense by proving the analogue of Lemma \ref{lem::headProgress}.
@@
 \begin{proof}
-The block $B$ with index $h$ was installed into \var{v}'s RBT by the CAS at line \ref{casGC} 
-that changed the tree from $T$ to $T'$.
-Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
+The block $B$ with index $h$ was installed into \var{v}'s RBT by the \op{CAS} at line \ref{casGC}.
+Suppose that \op{CAS} changed the tree from $T$ to $T'$.
+Before this \op{CAS}, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
 line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
@@
 Since $B'$ was already in $T$ before $B$ was created, the
-\op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the 
-\op{CreateBlock}$(v,b)$ that created $B$ started.
-By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ 
+\op{CreateBlock}$(\var{v},b-1)$ that created $B'$ terminated before the 
+\op{CreateBlock}$(\var{v},b)$ that created $B$ started.
+By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
 stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
-of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  
+of \op{CreateBlock}$(\var{v},b)$ stores in $B.\eleft$.  
 Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
@@
 Thus, the searches for a block based on \eleft\ or \eright\ values 
-at lines \ref{searchsuper1}, \ref{searchsuper2} and \ref{searchsuper3},
+at lines \ref{searchsuper3}, \ref{searchsuper1} and \ref{searchsuper2},
 which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.
@@
 are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
-Between the termination of $A$'s call to \op{Splitblock} at line \ref{callSplitBlock}
+Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
 and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
@@
 
-\here{Should add some discussion about the fact that we keep the newest finished block, in case some operation contained in a block with index i also needs to look at fields of the block with index i-1}
+We remark that \Cref{lem::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
+This is because the first unfinished block may still be traversed by an operation in the future,
+and when examining that block the operation may need information from the preceding block.  For example,
+when \op{FindResponse} is called on a block with index $b$, line \ref{FRnumenqGC} looks up the block with
+index $b-1$.
 
@@
 
-Similarly the results of Section \ref{sec::linearizability} 
+Similarly, the results of Section \ref{sec::linearizability} 
 can be reproved in exactly the same way as for the original algorithm to establish
@@
 
-We remark that the implementation remains wait-free:  the depth of recursion in each routine is  bounded
+The implementation remains wait-free:  the depth of recursion in each routine is  bounded
 by the height of the tree and the only loop is the counted loop in the \op{Help} routine.

### 2023-03-03 podc_paper/introduction.tex
Adding some recent references

@@
 significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
-packet processing \cite{DPDK},
+packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
 and sharing resources or tasks.

### 2023-03-03 podc_paper/related.tex
Adding some recent references

@@
 To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
+(It was shown recently that this implementation can be modified to use single-word CAS instructions rather than double-width CAS \cite{RK23}.)
 If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
@@
 Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
+Concurrently with our work, Johnen, Khattabi and Milani \cite{JKM23} build on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
+that achieves $O(\log p)$ time for enqueue operations and $O(k \log p)$ time for dequeue operations if there are only $k$ dequeuers.
 

### 2023-03-06 podc_paper/related.tex
added ref to Hossein's thesis

@@
 Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
-Concurrently with our work, Johnen, Khattabi and Milani \cite{JKM23} build on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
-that achieves $O(\log p)$ time for enqueue operations and $O(k \log p)$ time for dequeue operations if there are only $k$ dequeuers.
+Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
+that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.
 

### 2023-04-01 podc_paper/algorithm.tex
started reformatting for camera-ready

@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \input{tournamentTree.pdf_t}
@@
 Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
+\medskip
 \input{implicit.pdf_t}
@@
 The \fld{super} field is not shown.}
-\end{figure}
+\end{figure*}
 
@@
 and $B^R_1,\ldots,B^R_r$ from $\var{v}$'s right child, then $B$'s operation sequences are defined by the concatenations 
-\begin{equation}
-E(B) = E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r) \mbox{ and }
-D(B) = D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
-\end{equation}
+\begin{eqnarray}
+E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r) \mbox{ and }\nonumber\\
+D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
+\end{eqnarray}
 We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.
@@
 \begin{figure}
-\begin{algorithmic}[1]
+\begin{algorithmic}
 \setcounter{ALG@line}{1}
@@
 \begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
+\item \typ{Node} \var{root} \Comment{root of binary tree of \tt{Node}s with one leaf per process}
 \end{itemize}
@@
 \begin{itemize}
-\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
-\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
-\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
+\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{tree pointers initialized  when creating the tree}
+\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{blocks that have been propagated to this node;}\linebreak
+	\mbox{ }\Comment{\var{blocks}[0] is empty block whose integer fields are 0}
+\item \typ{int} \head \Comment{position to append next \block\ to \fld{blocks}, initially 1}
 \end{itemize}
@@
   	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
+  		\Comment{number of enqueues, dequeues in \fld{blocks} array}\linebreak
+		\mbox{ }\Comment{up to this block (inclusive)}
   	\item \typ{int} \fld{super}
-  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
+  		\Comment{approximate index of superblock in \fld{parent.blocks}}
 	\item[\com] Blocks in internal \nodes\ have the following additional fields
@@
 		\item[$\bullet$] \typ{Object} \fld{element}
-  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
+  		\Comment{\var{x} for \opa{Enqueue}{x} operation; otherwise \nl}
 	\end{itemize}
@@
 		\item[$\bullet$] \typ{int} \size%
-  		\Comment{size of the queue after performing all operations up to the end of this block}
+  		\Comment{size of queue after performing all operations up }\linebreak
+		\mbox{ }\Comment{to the end of this block}
 	\end{itemize}
@@
 \end{algorithmic}
-\caption{Objects used in the \ordering\ tree data structure \label{object-fields}}
+\caption{Objects used in the \ordering\ tree data structure.\label{object-fields}\Eric{Fix indentation}}
 \end{figure}
@@
 
-\begin{figure}
-\begin{minipage}[t]{0.465\textwidth}
+\begin{figure*}
+\begin{minipage}[t]{0.405\textwidth}
 \begin{algorithmic}[1]
@@
 
-\Function{void}{Append}{\typ{Block} \var{B}} 
-    \State \linecomment{append block to leaf and propagate to root}
+\Function{void}{Append}{\typ{Block} \var{B}} \com\ append block to leaf and propagate to root
     \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
@@
 
-\Function{void}{Propagate}{\typ{Node} \var{v}}
-    \State \linecomment{propagate blocks from \var{v}'s children to root}
+\Function{void}{Propagate}{\typ{Node} \var{v}} \com\ propagate blocks from \var{v}'s children to root
     \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}}
-    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \com\ try to append a new block to \var{v}.\fld{blocks}
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
@@
 
-\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} 
-    \State \linecomment{set \var{v}.\fld{blocks}[\var{h}].\fld{super} and }
-    \State \linecomment{increment \var{v}.\fld{head} from \var{h} to \var{h}+1}
+\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} \com\ set \var{v}.\fld{blocks}[\var{h}].\fld{super} and increment
+    \State \linecomment{\var{v}.\fld{head} from \var{h} to $\var{h}+1$}
     \If{$\var{v}\neq \var{root}$}
@@
 \end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
+\begin{minipage}[t]{0.585\textwidth}
 
@@
 
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
-    \State\linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \com\ create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]
     \State let \var{new} be a new \typ{Block} \label{initNewBlock}
@@
     \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
-	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
+	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
 			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
-	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
+	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + 
 			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
@@
 
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
-    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, has been }
-    \State \linecomment{propagated to root and contains at least \var{i} dequeues}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \com\ return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in
+    \State \linecomment{$D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
+    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least}
+    \State \linecomment{\var{i} dequeues}
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
@@
 	    \State \linecomment{compute index \var{i} of dequeue in superblock}
-	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
+	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
 	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
@@
 
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
@@
     %  \State \linecomment{contains at least \var{i} dequeues}
-    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
+    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
@@
 \caption{Queue implementation.\label{pseudocode1}}
-\end{figure}
+\end{figure*}
 
@@
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued.}
+\caption{\label{pseudocode2}Queue implementation, continued.\Eric{Fix code and combine with other figure}}
 %\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}

### 2023-04-01 podc_paper/podc-submission.tex
added abstract that was deleted by mistake
Minor edits; fixed amortized complexity bound for GC
started reformatting for camera-ready

@@
 %%
-\documentclass[acmsmall,nonacm,anonymous]{acmart}
+\documentclass[sigconf]{acmart}
+%\documentclass[acmsmall,nonacm]{acmart}
 
@@
 %% end of the preamble, start of the body of the document source.
-\begin{document}
 
@@
 \author{Hossein Naderibeni}
-\email{hnaderi268@gmail.com}
+%\email{hnaderi268@gmail.com}
 \affiliation{%
@@
 }
-\email{ruppert@cse.yorku.ca}
+%\email{ruppert@cse.yorku.ca}
 \orcid{0000-0001-5613-8701}
 
-\settopmatter{printacmref=false}  % REMOVE THIS LINE FOR CAMERA READY
-
-\begin{abstract}
-We present a novel linearizable wait-free queue implementation using single-word 
-CAS instructions.
-Previous lock-free queue implementations from \op{CAS} all have amortized step complexity of $\Omega(p)$ per operation in worst-case executions, 
-where $p$ is the number of processes that access the queue.
-Our new wait-free queue  takes $O(\log p)$ steps per enqueue and $O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.
-A bounded-space version of the implementation has amortized step complexity of $O(\log^2 p +\log q)$ per operation.
-\end{abstract}
+%\settopmatter{printacmref=false}  % REMOVE THIS LINE FOR CAMERA READY
+\copyrightyear{2023}
+\acmYear{2023}
+\setcopyright{acmlicensed}\acmConference[PODC '23]{xxx}{xxx}{xxx}
+\acmBooktitle{xxx}
+\acmPrice{xxx}
+\acmDOI{xxx}
+\acmISBN{xxx}
+\settopmatter{printacmref=true}
 
@@
 
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10003752.10003809.10010031</concept_id>
+<concept_desc>Theory of computation~Data structures design and analysis</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003809.10011778</concept_id>
+<concept_desc>Theory of computation~Concurrent algorithms</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Data structures design and analysis}
+\ccsdesc[500]{Theory of computation~Concurrent algorithms}
+
 %%
@@
 %% the work being presented. Separate the keywords with commas.
-%\keywords{concurrent data structures, wait-free queues}
+
+\keywords{concurrent data structures, wait-free queues}
 
@@
 %% information and builds the first part of the formatted document.
+
+\begin{document}
+
+\begin{abstract}
+We present a novel linearizable wait-free queue implementation using single-word CAS instructions.  
+Previous lock-free queue implementations from \op{CAS} all have amortized step complexity of 
+$\Omega(p)$ per operation in worst-case executions, where $p$ is the number of processes that 
+access the queue.  Our new wait-free queue  takes $O(\log p)$ steps per enqueue and 
+$O(\log^2 p +\log q)$ steps per dequeue, where $q$ is the size of the queue.  
+A bounded-space version of the implementation has $O(\log p \log(p+q))$ amortized step complexity  per operation.
+\end{abstract}
+
 \maketitle
+
+\Eric{Remove references to appendix and replace with tech report}
 
@@
 
-%OMIT IN SUBMISSION FOR ANONYMITY
-%\paragraph{Acknowledgements}
-%We thank Franck van Breugel for his helpful comments.
-%Funding was provided by the Natural Sciences and Engineering Research Council of Canada.
-
-\newpage
+\paragraph{Acknowledgements}
+We thank Franck van Breugel and the anonymous referees for their helpful comments.
+Funding was provided by the Natural Sciences and Engineering Research Council of Canada.
 
@@
 
-\newpage
-\appendix
-
-\input{correctness-details}
-
-\input{algorithm-GC}
-
-\input{correctness-GC}
+%\appendix
+%\input{correctness-details}
+%\input{algorithm-GC}
+%\input{correctness-GC}
 

### 2023-05-08 podc_paper/algorithm.tex
fixed code formatting

@@
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1,\\
-		\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}}\label{enqNew}
+		$\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
+		$\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
     \State \Call{Append}{\var{B}}
@@
     \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
-	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
+	    $\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
+	    $\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
     \State \Call{Append}{\var{B}}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}\label{invokeIndexDequeue}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$}\label{invokeIndexDequeue}
     \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
@@
     \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
-    \State \var{leaf.}\head\ \assign\ \var{leaf.}\head+1 \label{appendEnd} 
+    \State $\var{leaf.}\head\ \assign\ \var{leaf.}\head+1$ \label{appendEnd} 
     \State \Call{Propagate}{\var{leaf.}\fld{parent}} 
@@
 
-\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} \com\ set \var{v}.\fld{blocks}[\var{h}].\fld{super} and increment
-    \State \linecomment{\var{v}.\fld{head} from \var{h} to $\var{h}+1$}
-    \If{$\var{v}\neq \var{root}$}
-	    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
-    	\State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
-	\EndIf
-    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
-\EndFunction{Advance}
-
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.585\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
-
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \com\ create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
+    \State \linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
     \State let \var{new} be a new \typ{Block} \label{initNewBlock}
@@
     \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
-	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
-	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + 
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
+	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
+			\hspace*{11mm}\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
+	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
+			\hspace*{11mm}\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
@@
 \EndFunction{CreateBlock}
+
+\end{algorithmic}
+\end{minipage}
+\begin{minipage}[t]{0.585\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{50}
+\Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} \com\ set \var{v}.\fld{blocks}[\var{h}].\fld{super} and increment \var{v}.\fld{head} from \var{h} to $\var{h}+1$
+    \If{$\var{v}\neq \var{root}$}
+	    \State \var{h\sub{p}} \assign\ \var{v}.\fld{parent}.\head \label{readParentHead}
+    	\State \Call{CAS}{\var{v}.\fld{blocks}[\var{h}].\fld{super}, \nl, \var{h\sub{p}}} \label{setSuper1}
+	\EndIf
+    \State \Call{CAS}{\var{v}.\head, \var{h}, \var{h}+1} \label{incrementHead}
+\EndFunction{Advance}
 
@@
 
-\end{algorithmic}
-\end{minipage}
-\vspace*{-3mm}
-\caption{Queue implementation.\label{pseudocode1}}
-\end{figure*}
-
-\begin{figure}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{102}
-
 \Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
@@
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} 
+        \State \linecomment{\var{sum\sub{left}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}'s left child}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} 
+        \State \linecomment{\var{prev\sub{left}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}'s left child}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} 
+        \State \linecomment{\var{prev\sub{right}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}'s right child}
         \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
@@
         \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
-        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
-$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
+        \State \hangbox{find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such that\\
+        	 $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}}
         \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
@@
 \EndFunction{GetEnqueue}
+
 \end{algorithmic}
-\caption{\label{pseudocode2}Queue implementation, continued.\Eric{Fix code and combine with other figure}}
-%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
-\end{figure}
+\end{minipage}
+\vspace*{-3mm}
+\caption{Queue implementation.\label{pseudocode1}}
+\end{figure*}
 

### 2023-05-10 podc_paper/algorithm.tex
fixing bad line breaks

@@
 \begin{eqnarray}\label{defsubblock}
-\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
-\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].
+\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b\!-\!1].\fld{end\sub{left}}\!+\!1..\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
+\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b\!-\!1].\fld{end\sub{right}}\!+\!1..\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\!
 \end{eqnarray}
@@
 Thus, starting from an operation in a leaf's block, one can use these indices to track the 
-operation all the way up the path to the root, and determine the operation's location in a root block
+operation  up the path to the root, and determine the operation's location in a root block
 in $O(\log p)$ time.
@@
 we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
-A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and $h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
+A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
 $h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
-Then, we can compute $B.\fld{sum\sub{enq}}$ as $\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}$.
-For a block $B$ in the root, $B.\fld{size}$ can be computed using the \fld{size} field of the previous block $B'$ and
-the number of enqueues and dequeues in $B$:
-$B.\fld{size} = \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}) - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}))$.
+Then, we can compute 
+\begin{eqnarray*}
+B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
+&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
+\end{eqnarray*}
+For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
+the number of enqueues and dequeues in~$B$:
+\begin{eqnarray*}
+B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
+			&& \hspace*{19mm} - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).	
+\end{eqnarray*}
 
@@
 
-We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
+We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1}.
 %We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
@@
 \var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright], line \ref{createSumEnq} sets
-\fld{sum\sub{enq}} to $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
+\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
 Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
-subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$].
+subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{new}.\fld{sum\sub{enq}}.
 The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.

### 2023-05-10 podc_paper/conclusion.tex
fixing bad line breaks

@@
 
-There is a gap between our implementation, which takes $O(\log^2 p + \log q)$ steps per operation,
+There remains a gap between our implementation, which takes $O(\log^2 p + \log q)$ steps per operation,
 and Attiya and Fouren's $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17}.
@@
 depends on $p$.
-Since a queue is also a bag, it is the first lock-free bag we know of that has polylogarithmic step complexity.
+Since a queue is also a bag, our queue is the first lock-free bag we know of that has polylogarithmic step complexity.
 
@@
 \here{doublecheck all preceding claims for accuracy.}
-We would like to investigate whether a similar approach could be used for stacks, deques or even priority queues.
+We are investigating whether a similar approach can be used for stacks, deques or even priority queues.

### 2023-05-10 podc_paper/correctness.tex
fixing bad line breaks

@@
 \begin{lemma} \label{lem::headProgress}
- If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$ then 
- $\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{left}}$ and 
- $\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[b].\fld{end\sub{right}}$.
+ If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$, then
+ \begin{eqnarray*}
+ \var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{left}} \mbox{ and}\\
+ \var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{right}}.
+ \end{eqnarray*}
 \end{lemma}
@@
 \label{lem::sum}
-If $B$ is a block stored in $\var{v}.\fld{blocks}[i]$,
-$B.\fld{sum\sub{enq}} = | E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i]) |$ and
-$B.\fld{sum\sub{deq}} = | D(\var{v}.\fld{blocks}[0])\cdots D(\var{v}.\fld{blocks}[i]) |$.
+If $B$ is a block stored in $\var{v}.\fld{blocks}[i]$, then
+\begin{eqnarray*}
+B.\fld{sum\sub{enq}} &=& | E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i]) | \mbox{ and}\\
+B.\fld{sum\sub{deq}} &=& | D(\var{v}.\fld{blocks}[0])\cdots D(\var{v}.\fld{blocks}[i]) |.
+\end{eqnarray*}
 \end{restatable}
@@
 \var{num\sub{enq}} \hspace*{-2mm}
-&=& \hspace*{-2mm} \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
-&=& \hspace*{-2mm}\var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
-&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
+&=& \hspace*{-2mm} \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} \\
+&& + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}}\\
+&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
+&=& \hspace*{-2mm}\var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}}\\
+&& + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
+&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
+&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
 \end{eqnarray*}
-By \Cref{lem::sum}, $num\sub{enq}$ is the  number of enqueues  in 
+It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
 $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and

### 2023-05-10 podc_paper/podc-submission.tex
Small updates to bibliography
fixing bad line breaks

@@
 %\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-\fancyfoot{} % REMOVE THIS LINE FOR CAMERA READY
+%\fancyfoot{} % REMOVE THIS LINE FOR CAMERA READY
 
@@
 \acmYear{2023}
-\setcopyright{acmlicensed}\acmConference[PODC '23]{xxx}{xxx}{xxx}
-\acmBooktitle{xxx}
-\acmPrice{xxx}
-\acmDOI{xxx}
-\acmISBN{xxx}
+\setcopyright{acmlicensed}\acmConference[PODC '23]{ACM Symposium on Principles of Distributed Computing}{June 19--23, 2023}{Orlando, FL, USA}
+\acmBooktitle{ACM Symposium on Principles of Distributed Computing (PODC '23), June 19--23, 2023, Orlando, FL, USA}
+\acmPrice{15.00}
+\acmDOI{10.1145/3583668.3594565}
+\acmISBN{979-8-4007-0121-4/23/06}
+
 \settopmatter{printacmref=true}
@@
 
-\paragraph{Acknowledgements}
+\begin{acks}
+%\paragraph{Acknowledgements}
 We thank Franck van Breugel and the anonymous referees for their helpful comments.
 Funding was provided by the Natural Sciences and Engineering Research Council of Canada.
+\end{acks}
 
-\bibliographystyle{plain}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{queues.bib}

### 2023-05-10 podc_paper/related.tex
fixing bad line breaks

@@
 \emph{Fetch\&Increment Objects.}
-There are implementations of fetch\&increment objects that use a polylogarithmic number of steps per operation \cite{ERW12,10.1007/978-3-642-41527-2_20}.
-Our approach is similar to that of Ellen, Ramachandran and Woelfel \cite{ERW12},
-which also uses a tree structure to keep track of the fetch\&increments
-that have been performed, as in the universal construction of \cite{DBLP:conf/stoc/AfekDT95}.
+Ellen, Ramachandran and Woelfel \cite{ERW12} gave an implementation of 
+fetch\&increment objects that uses a polylogarithmic number of steps per operation.
+Like our queue, they also use a tree structure similar to the universal construction of 
+\cite{DBLP:conf/stoc/AfekDT95} to keep track of the operations
+that have been performed.
 However, our construction requires more intricate
@@
 as the single-word state of a fetch\&increment object.
+Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} gave an improved implementation of fetch\&increment with better step complexity.

### 2023-05-10 podc_paper/algorithm.tex
some formatting fixes, dealt with comments from reviewer A

@@
 If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
-dequeue must return.  To do this, $P$ first determines which block in the root contains its operation
-(since the operation may have been propagated to the root by some other process).
+dequeue must return.  To do this, $P$ first determines which block in the root contains its dequeue
+(since the dequeue may have been propagated to the root by some other process).
+$P$ does this by finding the dequeue's location in each node along the path from the leaf to the root.
 Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
@@
 \end{enumerate}
-Since these tasks depends on the linearization ordering, we describe that ordering next.
+Since these tasks depend on the linearization ordering, we describe that ordering next.
 
@@
 
-Finally, when designing a total order for the operations propagated to the root, we choose
-to put each block's enqueues before its dequeues.
+Finally, when linearizing the operations propagated to the root, we must
+choose how to order operations within a block.  Any ordering within a block would work, since all operations
+in a block are concurrent with one another.  We choose to put
+each block's enqueues before its dequeues.
 Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 

### 2023-05-10 podc_paper/introduction.tex
added ref to Hossein's thesis
started reformatting for camera-ready
some formatting fixes, dealt with comments from reviewer A

@@
 to insert an element or remove an arbitrary element (chosen non-deterministically).
-Since a queue trivially implements a bag, this lower bound also holds for queues.
-At first glance, this would seem to settle the question of the step complexity of lock-free queues.
-However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
-Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.
+Since a queue trivially implements a bag, this lower bound also applies to queues.
+Although this might seem to settle the step complexity of lock-free queues, the lower bound
+holds only if $c$ is $O(\log\log p)$ so it should be stated more precisely as
+%At first glance, this would seem to settle the question of the step complexity of lock-free queues.
+%However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
+%Thus, the lower bound could be stated more precisely as 
+an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.
 
@@
 and quickly access any  operation in a~set.
+A preliminary version of this work appeared in~\cite{Nad22}.
 \here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-05-10 podc_paper/related.tex
some formatting fixes, dealt with comments from reviewer A

@@
 
-\emph{List-based Queues.}
+\paragraph{List-based Queues.}
 The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
@@
 
-\emph{Array-Based Queues.}
+\paragraph{Array-Based Queues.}
 Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
@@
 
-\emph{Universal Constructions.}
+\paragraph{Universal Constructions.}
 One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
@@
 
-\emph{Restricted Queues.}
+\paragraph{Restricted Queues.}
 David introduced the first sublinear-time queue
@@
 
-\emph{Other Primitives.}
+\paragraph{Other Primitives.}
 Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
@@
 
-\emph{Fetch\&Increment Objects.}
+\paragraph{Fetch\&Increment Objects.}
 Ellen, Ramachandran and Woelfel \cite{ERW12} gave an implementation of 

### 2023-05-11 podc_paper/algorithm-GC.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

@@
 \item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
-\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the root  that process $k$}\\
-\mbox{ }\Comment{has observed to contain either an enqueue whose value has}\\
-\mbox{ }\Comment{been dequeued or a null dequeue }
+\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the}\\
+\mbox{ }\Comment{root that process $k$ has observed to contain either an en-}\\
+\mbox{ }\Comment{queue whose value has been dequeued or a null dequeue }
 \end{itemize}
@@
 \end{itemize}
-To facilitate helping, each \typ{Block} in a leaf has one more additional field.
+To facilitate helping, each \typ{Block} in a leaf has one more additional field that is used only for blocks that store a dequeue operation.
 \begin{itemize}
-\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
+\item \typ{Object} \var{response} \Comment{response of the dequeue in the block}
 \end{itemize}

### 2023-05-11 podc_paper/boundedspace.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

@@
 Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
-RBT are performed by making a new copy of each RBT node $x$ that must be modified, as well
+RBT make a new copy of each RBT node $x$ that must be modified, as well
 as each RBT node along the path from the RBT's root to~$x$.

### 2023-05-11 podc_paper/correctness-details.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

@@
 and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
-Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
-$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and
+Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in\linebreak
+$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in \linebreak
 $E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
-By the hypothesis, the total number of these enqueues is $\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
+By the hypothesis, the total number of these enqueues is \linebreak
+$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, \linebreak
+which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
 The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
@@
 \begin{proof}
-We first show $B.\fld{super}\leq s$.
-Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs the superblock of $B$ 
+We first show that $B.\fld{super}\leq s$.
+Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
 in $\var{v}.\fld{parent.blocks}[s]$.
 By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
-By \Cref{lem::headPosition}, $B.\fld{super} \neq \nl$ when $r$ occurs, which means
+By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
 that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
@@
 Next, we show that $B.\fld{super}\geq s-1$.
-The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ values are always at least 1, so $B.\fld{super} \geq 1$.
-So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the remainder of the proof.
-By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]\neq \nl$.  Let $R_{s-1}$ be the call to
+The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least 1, so $B.\fld{super} \geq 1$.
+So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
+By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
 $\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
@@
 We prove the claim by induction on $b$. 
-The base case when ${b=0}$ is trivial, since the queue is initially empty and 
+The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
 $\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 

### 2023-05-11 podc_paper/introduction.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

@@
 
-There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
+There has been a great deal of research in the past several decades on the design of shared queues.
 Besides being a fundamental data structure, queues are used in
-significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
+significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
 packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
 and sharing resources or tasks.
+We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
+appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
+is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.
+
 The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
@@
 
-We exploit this loophole.  We show  it is, in fact, possible to design a linearizable, wait-free queue
-with step complexity sublinear in $p$.
+We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
+to have step complexity sublinear in $p$.
 Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
+It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
 For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
@@
 while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
-Both versions use single-word CAS on %reasonably-sized 
-small words.
 Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
@@
 $\Omega(p)$ CAS instructions, even in an amortized sense.
+Both versions of our queue use single-word CAS on reasonably-sized 
+words.
+We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored in $O(1)$ words.
+This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
 For the space-bounded version, we unlink unneeded objects from our data structure.
@@
 Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
-Each process adds its operations to its leaf.
+A process adds its operations to its leaf.
 As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
@@
 \here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
-Explicitly storing all operations in the tree nodes would be too costly.
+Explicitly storing  operations in the tree nodes would be too costly.
 Instead, we use a novel implicit representation of sets

### 2023-05-11 podc_paper/related.tex
addressing reviewers' comments, squeezing space, fixing some formatting issues

@@
 The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
-A version of it is
-included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
-See the paper that introduced it for a survey of the early history of concurrent queues.
+The standard Java Concurrency Package includes a version of it.  %java.util.concurrent.ConcurrentLinkedQueue
+See \cite{MS98} for a survey of the early history of concurrent queues.
 As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
@@
 Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
-attempts to reduce contention by grouping concurrent enqueues into baskets.
+attempts to reduces contention by grouping concurrent enqueues into baskets.
 An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
-Enqueues within a basket determine their order among themselves without having to access the back of the queue.
+Enqueues within a basket order  themselves without having to access the back of the queue.
 However, if $p$ concurrent enqueues are in the same basket
@@
 
-Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} described how to use futures to improve
-the performance of the MS-queue.
+Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} improved the MS-queue's performance using \emph{futures}.
 Operations return future
 objects instead of responses. Later, when an operation's response is needed, it
-can be evaluated using the corresponding future object.
+is evaluated using the future object.
 This allows batches of enqueues or dequeues to be done at once on an MS-queue.
@@
 
+In the MS-queue, an enqueue requires two CAS steps.
 Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
-presented an optimistic  queue implementation. 
+presented an optimistic  queue implementation
 %MS-queue uses
@@
 %to the new node. 
-In the MS-queue, an enqueue requires two CAS steps.
-The optimistic queue uses a doubly-linked list to reduce the number of
+that uses a doubly-linked list to reduce the number of
 \texttt{CAS} instructions to one in the best case. 
 Pointers in the doubly-linked list can be inconsistent, but are fixed when necessary by traversing the list.
-Although this fixing is rare in practice, it yields an amortized complexity of $\Omega(qp)$ 
-steps per operation for worst-case executions.
+This fixing is rare in practice, but it yields an amortized complexity of $\Omega(qp)$ 
+steps per operation in the worst case.
 
@@
 \paragraph{Restricted Queues.}
-David introduced the first sublinear-time queue
+David gave the first sublinear-time queue
 \cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
@@
 Jayanti and Petrovic gave a wait-free polylogarithmic
-queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
+queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but only for a single dequeuer. 
 Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
-Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
+Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue  
 that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.
@@
 \paragraph{Other Primitives.}
-Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
+Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue 
 with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
-called half-increment and half-max, which can be viewed as particular kinds of
+called half-increment and half-max, which can be viewed as %particular kinds of
 double-word read-modify-write operations.

### 2023-05-11 podc_paper/algorithm-GC.tex
reformatted appendix code

@@
 
-\begin{figure}
-\begin{minipage}[t]{0.465\textwidth}
+\begin{figure*}
+\begin{minipage}[t]{0.41\textwidth}
 \begin{algorithmic}[1]
@@
 \Function{Object}{CompleteDeq}{\var{leaf}, $h$} 
-    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
+    \State \linecomment finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
     \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1}$
@@
 
-\Function{void}{Append}{\typ{Block} \var{B}} 
-    \State \linecomment{append block to leaf and propagate to root}
+\Function{void}{Append}{\typ{Block} \var{B}}  \linecomment{append block to leaf and propagate to root}
     \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
@@
 \Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} 
-    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
+    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
     \If{\var{B}.\fld{index} is a multiple of $G$}
@@
 
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{249}
+\spac
 
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}}
-	\State \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
 	\gc{
@@
 
-\spac
+\end{algorithmic}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{0.56\textwidth}
+
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{267}
 
 \gc{
-\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
-    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
+\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
     \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
@@
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
-    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
-    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
+    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
@@
 	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
-	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
-	    		$\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
+	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
+	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
         \If{$\fld{dir} = \fld{right}$} 
-        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
-					$\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
+        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
+					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
         \EndIf \label{computeISuperEndGC}
@@
 
-\end{algorithmic}
-\end{minipage}
-\vspace*{-4mm}
-\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
-\end{figure}
-
-\begin{figure}
-\begin{minipage}[t]{0.465\textwidth}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{302}
+\spac
 
 \gc{
-\Function{void}{Help}{}
-    \State \linecomment{help pending operations}
+\Function{void}{Help}{} \linecomment{help pending operations}
 
@@
     	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
-        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
-        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
+        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
             \State \linecomment{operation is a propagated dequeue}
@@
 
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
-    \State\linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
     \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
@@
     }
-	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
-	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
+	\State \var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}
+	\State \var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} +
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}
     \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
@@
     \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \\
-        	\var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
+        \State \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
+        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
     \EndIf
@@
 
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.529\textwidth}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{351}
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+\spac
+
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRnumenqGC}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
-    	\State \linecomment{ queue is empty when dequeue occurs}
+    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$
+    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \linecomment{queue is empty when dequeue occurs}
         \gc{
@@
 			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
-		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
-		\State \linecomment{find rank of enqueue within its block}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
+
+		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC} \linecomment{use binary search}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \linecomment{find rank of enqueue within its block}
 		\gc{
@@
 \end{minipage}
+\vspace*{-4mm}
+\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
+\end{figure*}
 
-\spac
-
+\begin{figure*}
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{332}
+\setcounter{ALG@line}{341}
 
@@
 \caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
-\end{figure}
+\end{figure*}
 

### 2023-05-11 podc_paper/algorithm.tex

@@
 creates a new block by combining information from those blocks, and attempts to append this 
-new block to $\var{v}$'s \fld{blocks} array using a \op{CAS} instruction.
+new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
 Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
@@
 If so, it returns \nl\ and we call it a \emph{null dequeue}.
-If not, $P$ computes the rank\footnote{We say that the $r$th element in a sequence has rank $r$ within that sequence.} $r$ of its dequeue among all non-null dequeues in the linearization ordering,
-finds the $r$th enqueue in the linearization, and returns that enqueue's value.
+If not, $P$ computes the rank $r$ of its dequeue among all non-null dequeues in the linearization ordering.  (We say that the $r$th element in a sequence has \emph{rank} $r$ within that sequence.)
+$P$ then returns the value of the $r$th enqueue in the linearization.
 
@@
 \label{construct}
-Construct a block for node $\var{v}$ that represents the operations contained in consecutive blocks in $\var{v}$'s children, as required for a \op{Refresh}.
+Construct a block for node $\var{v}$ that represents the operations in consecutive blocks in $\var{v}$'s children, as required for a \op{Refresh}.
 \item
@@
 \label{findrank}
-Given a dequeue's position in the root, determine if it is a null dequeue (i.e., if the queue is empty when it is linearized)
-or determine the rank $r$ of the enqueue whose value it should return.
+Given a dequeue's position in the root, decide if it is a null dequeue (i.e., if the queue is empty when it is linearized)
+or determine the rank $r$ of the enqueue whose value it returns.
 \item
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{50}
+\setcounter{ALG@line}{57}
 \Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} \com\ set \var{v}.\fld{blocks}[\var{h}].\fld{super} and increment \var{v}.\fld{head} from \var{h} to $\var{h}+1$
@@
 \EndFunction{FindResponse}
+
+\spac
 
@@
 \opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
-$i$th enqueue in the $b$th block of \typ{Node} $\var{v}$. 
-The range of subblocks of a block is determined using the \fld{end\sub{left}} and
-\fld{end\sub{right}} fields of the block and its previous block. Then,
-the exact subblock is found using binary search on the \fld{sum\sub{enq}}
-field (line~\ref{getChild}). 
-\here{explain a little more?}
+$i$th enqueue in the $b$th block $B$ of \typ{Node} $\var{v}$. 
+It recursively finds the location of the enqueue in each node along the path from $v$ to a leaf, which stores the argument explicitly.
+\op{GetEnqueue} first determines which child of \var{v} contains the enqueue, and
+then finds the range of blocks within that child that are subblocks of $B$ using information stored
+in $B$ and the block that precedes $B$ in $v$.
+\op{GetEnqueue} finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
+field (line~\ref{getChild}) and proceeds recursively down the tree.

### 2023-05-11 podc_paper/conclusion.tex

@@
 
-There remains a gap between our implementation, which takes $O(\log^2 p + \log q)$ steps per operation,
+A gap remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
 and Attiya and Fouren's $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17}.

### 2023-05-11 podc_paper/correctness-GC.tex

@@
 
-Each helping phase takes $O(p)$ steps to read the \var{last} array,
+Each helping phase takes $O(p)$ steps to read the \var{last} array, and
 $O(p \log p \log(p+q_{max}))$ steps to \op{Help},
-followed by $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
+and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
 Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps

### 2023-05-11 podc_paper/correctness.tex

@@
 % at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
-Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
-(from \nl\ to a non-\nl\ value), so it is written only once.
-Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
+Moreover, only a \op{CAS} at line \ref{setSuper1}  modifies  \fld{super}  
+(from \nl\ to a non-\nl\ value), so it is changed only once.
+Similarly, only a \op{CAS} at line \ref{cas} modifies an element of a node's \fld{blocks} array 
 (from \nl\ to a non-\nl\ value), so blocks are permanently added to nodes.
@@
 
-\begin{lemma} \label{lem::headInc}
+\begin{observation} \label{lem::headInc}
 Let $R$ be an instance of \opa{Refresh}{v} whose call to \op{CreateBlock} returns a non-\nl\ block.
 When $R$ terminates, \var{v}.\head\ is strictly greater than the value $R$ reads from it at line \ref{readHead}.
-\end{lemma}
+\end{observation}
 \begin{proof}
@@
 Let $R$ be a call to \op{Refresh}($\var{v}$) that performs a successful \op{CAS} on line \ref{cas} (or terminates at line \ref{addOP}).
-In the configuration after that CAS (or termination, respectively), $\var{v}$ contains all operations that $\var{v}$'s children contained 
+After that CAS (or termination, respectively), $\var{v}$ contains all operations that $\var{v}$'s children contained 
 when $R$ executed line~\ref{readHead}.
@@
 
-We now show that a double \op{Refresh} propagates blocks as required.
+We now show  a double \op{Refresh} propagates blocks as required.
 
@@
 We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
-This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from the
+This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from
 \var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
@@
 %into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
-Showing  \op{GetEnqueue} and \op{IndexDequeue} work correctly is mostly just checking that they correctly compute the index of the required block
+To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
 and the operation's rank within the block.  

### 2023-05-11 podc_paper/algorithm.tex
minor fixes

@@
 In this section, we use an infinite array of blocks in each node.
-Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.
+Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.
 
@@
 \op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
-A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $\var{v}$.
+A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
 However, since a successful \op{Refresh} propagates multiple pending operations 
@@
 Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
-and \eleft\ and \eright\ fields are shown using arrows as before.
+and \eleft\ and \eright\ fields are shown using arrows as in Figure \ref{orderingtree}.
 Root blocks also have the additional \fld{size} field.
@@
 
-Performing a double refresh at each node along the path from the leaf to the root ensures 
+Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
 a block containing the operation is appended to the root before the operation completes.
@@
 
-Finally, when linearizing the operations propagated to the root, we must
-choose how to order operations within a block.  Any ordering within a block would work, since all operations
-in a block are concurrent with one another.  We choose to put
+When linearizing the operations propagated to the root, we must
+choose how to order operations within a block.  
+We choose to put
 each block's enqueues before its dequeues.
 Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
-linearization ordering is 
+linearization ordering~is 
 \begin{equation}
@@
 $B$ is created for a \node\ $\var{v}$, we do not yet know where its
-superblock will eventually be installed in $B$'s parent.
+superblock will eventually be installed in \var{v}'s parent.
 So, we leave $B.\fld{super}$ blank.  
 Soon after $B$ is installed,
-a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
-set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
+%a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by 
+some process will
+set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
 We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of $B'$
@@
 \opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
-It uses the double refresh idea described
+It uses the double \op{Refresh} idea described
 above and invokes two \op{Refresh}es on \var{v} in Lines

### 2023-05-11 podc_paper/analysis.tex
minor fixes

@@
 An operation invokes \op{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
-A \op{Refresh} does at most 7 \op{CAS} steps: one in line \ref{cas} and two from each \op{Advance} in line \ref{helpAdvance} or~\ref{advance}.
+A \op{Refresh} does at most 5 \op{CAS} steps: one in line \ref{cas} and two during each \op{Advance} in line \ref{helpAdvance} or~\ref{advance}.
 \end{proof}
@@
 
-The following lemma is useful in bounding the time for \op{GetEnqueue}
+The following lemma helps bound the time for \op{GetEnqueue}.
 \begin{lemma}\label{blockSize}

### 2023-05-11 podc_paper/correctness-GC.tex
minor fixes

@@
 If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
-$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some $m$.
+$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some~$m$.
 \end{lemma}
@@
 After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
-then creates the  block \var{new},
+the \op{Refresh} then creates the  block \var{new},
 and then calls \op{AddBlock}($T,new$) to create $T'$.
@@
 \fld{sum\sub{enq}} values in lines
-\ref{getChildGC} and \ref{FRsearchGC} can be done in logarithmic time.
+\ref{FRsearchGC} and \ref{getChildGC} can be done in logarithmic time.
 
@@
 
-Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
+Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{casGC} 
 installs $B'$ in $\var{v}.\fld{blocks}$.

### 2023-05-11 podc_paper/introduction.tex
minor fixes

@@
 significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
-packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
+%packet processing \cite{DPDK}, 
+synchronization \cite{KAE23},\here{is this a good citation for this?}
 and sharing resources or tasks.
@@
 words.
-We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored in $O(1)$ words.
+We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).
+We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
 This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.

### 2023-05-11 podc_paper/related.tex
minor fixes

@@
 Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
-attempts to reduces contention by grouping concurrent enqueues into baskets.
+attempts to reduce contention by grouping concurrent enqueues into baskets.
 An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
@@
 Dequeues and enqueues update
-indexes of the front and back elements using CAS instructions.
+indices of the front and back elements using CAS instructions.
 Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
@@
 This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
-Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
-on array-based queues.
+Other array-based queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology.
 In worst-case executions that use the slow path,

### 2023-05-12 podc_paper/correctness.tex
minor fixes
Edited for camera-ready version:  changed references to appendix to arxiv version

@@
 on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
-from the definition of subblocks.  See Appendix 
-\ref{app::tracingDetails} for a detailed proof of \Cref{lem::sum}.
+from the definition of subblocks.  See 
+%Appendix \ref{app::tracingDetails} 
+the full version \cite{full} for a detailed proof of \Cref{lem::sum}.
 
@@
 
-Next, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
+In the next two lemmas, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
 We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
@@
 
-See Appendix \ref{app::tracingDetails} for detailed proofs for this section.
+See 
+%Appendix \ref{app::tracingDetails}
+the full version \cite{full} for detailed proofs for this section.
 
@@
 
-A simple proof (in Appendix \ref{app::tracingDetails}) shows that \fld{size} fields are computed correctly.
+A simple proof (in 
+%Appendix \ref{app::tracingDetails})
+the full version \cite{full})
+shows that \fld{size} fields are computed correctly.
 \begin{restatable}{lemma}{sizeCorrectRes}

### 2023-05-30 presentation/podc-presentation.tex
started working on presentation

\documentclass[compress]{beamer}

\usepackage{color}
\usepackage{amssymb}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage[normalem]{ulem} % for striking through text

\mode<presentation>
{
  \usetheme{Warsaw}
  \usecolortheme{beaver}
  \usefonttheme{structurebold}

  % or ...

  \setbeamercovered{transparent}
  % or whatever (possibly just delete it)
}
\logo{\includegraphics[width=15mm]{YorkLogo.pdf}}

\newcommand{\Red}[1]{{\color{red}#1}}
\newcommand{\Blue}[1]{{\color{blue}#1}}
\newcommand{\Green}[1]{{\color{green}#1}}

\newcommand{\tb}{\hspace*{10mm}}

\newcommand{\ignore}[1]{}

\title{A Wait-Free Queue with Polylogarithmic Step Complexity}

\author{Hossein Naderibeni \and Eric Ruppert}
\date{June 21, 2023}

\begin{document}

\begin{frame}
\maketitle
\end{frame}

\begin{frame}{Lock-Free Queue using CAS}
Michael and Scott Queue [PODC 1996]

\bigskip

% We store the elements of the queue in a singly-linked list.
% Enqueues add elements at the left end and dequeues remove elements from the right end.
% Each enqueue creates new node and performs a CAS on the tail node's next pointer to add it to the left end (and then advances the Tail pointer)
% Each dequeue uses a CAS to advance the Head pointer

\only<1>{\input{MSQueue-48to50.pdf_t}}
\only<2>{\input{MSQueue-48to51.pdf_t}}
\only<3>{\input{MSQueue-48to52.pdf_t}}
\only<4>{\input{MSQueue-48-50to53.pdf_t}}
\only<5>{\input{MSQueue-50to54.pdf_t}}
\end{frame}

\begin{frame}{CAS Retry Problem}

\end{frame}

\end{document}

### 2023-06-11 presentation/podc-presentation.tex
Added related work and final slide

@@
 
+\newcommand{\op}[1]{\mbox{\sc #1}}
 \newcommand{\tb}{\hspace*{10mm}}
+
+\renewcommand{\log}{\mbox{log }} % font looks funny for usual \log
 
@@
 \begin{frame}{CAS Retry Problem}
+Suppose $p$ processes want to enqueue simultaneously.
+\medskip
+
+\only<1>{\input{CASretry-49to50.pdf_t}}
+\only<2>{\input{CASretry-49to51.pdf_t}}
+\only<3->{\input{CASretry-49to52.pdf_t}}
+
+\medskip
+\visible<4->{
+\begin{itemize}
+\item one CAS of next pointer succeeds.
+\item other \op{Enqueues} must try again.
+\item<5> $\Rightarrow$ starvation and $\Omega(p)$ steps per operation (amortized)
+\end{itemize}
+}
+\end{frame}
+
+\begin{frame}{Other Lock-Free Queues}
+
+Other \Red{list-based} queues
+\begin{itemize}
+\item
+add elimination array \hfill {\small [Moir et al.\ 2005]}
+\item
+baskets queue \hfill {\small [Hoffman, Shalev, Shavit 2007]}
+\item
+doubly-linked list + optimism \hfill {\small [Ladan-Mozes, Shavit 2008]}
+\item
+fast path, slow path \hfill {\small [Kogan, Petrank 2012]}
+\item
+futures \hfill {\small [Kogan, Herlihy 2014]}
+\end{itemize}
+\visible<2->{All have CAS retry problem  \hfill {\small [Morrisson, Afek 2013]}}
+
+\smallskip
+
+\visible<3->{So do \Red{array-based} queues
+
+\smallskip
+
+All$^*$ previous queues take amortized $\Omega(p)$ steps per operation}
+
+\end{frame}
+
+\begin{frame}{$^*$Exceptions: Sublinear Time Queues}
+
+\begin{block}{Restricted queues}
+\begin{itemize}
+\item
+1 enqueuer, multiple dequeuers \hfill {\small [David 2004]}
+\item
+1 dequeuer, multiple enqueuers \hfill {\small [Jayanti, Petrovic 2005]}
+\end{itemize}
+\end{block}
+
+\pause
+
+\begin{block}{Other primitives}
+\begin{itemize}
+\item
+$O(\sqrt{p})$ using unusual double-word RMW instructions
+ \hfill {\small [Khanchandani, Wattenhofer 2018]}
+\end{itemize}
+\end{block}
+
+\pause
+
+\begin{block}{Universal constructions}
+\begin{itemize}
+\item
+$O(\log p)$ using huge words\\
+\mbox{ }\hfill{\small [Afek, Dauber, Touitou 1995; Jayanti 1998]}
+\item
+$\Omega(p)$ with reasonably-sized words
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{Lower Bound}
+
+All previous multi-enqueuer, multi-dequeuer queues take\\
+$\Omega(p)$ steps per operation.
+
+\medskip
+
+For many  data structures, fastest lock-free operations take
+$O(sequential\ complexity + contention)$ steps
+
+\pause
+
+\begin{block}{Lower Bound \hfill {\small [Attiya, Fouren 2017]}}
+\begin{itemize}
+\item
+Amortized step complexity for any bag is $\Omega(contention)$.
+\pause
+\item
+But lower bound holds only if $contention$ is $O(\log\log p)$
+\end{itemize}
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Our Result}
+
+\begin{block}{New Queue}
+\begin{itemize}
+\item $O(\log p)$ steps per \op{Enqueue}
+\item $O(\log^2 p + \log q)$ steps per \op{Dequeue}
+\item wait-free
+\item uses CAS on reasonable-size words
+\item bounded space version: $O(\log p\ \log(p+q))$ amortized steps per operation
+(relies on safe GC)
+\end{itemize}
+\end{block}
+
+$p = \#$ processes\\
+$q = \#$ elements in queue
+
+\medskip
+
+Previous lock-free queues take $\Omega(p)$ steps per operation.
+
+\end{frame}
+
+\begin{frame}{Future Directions}
+
+\begin{itemize}
+\item
+Practical implementation\\
+(perhaps slow path of fast path slow path method)
+\item
+Extend technique to other data structures\\
+ (stacks and deques are recently done)
+ \item
+Close gap between\\
+$\Omega(\log \log p)$ lower bound \hfill {\small [Attiya Fouren 2017]}\\
+$O(\log^2 p + \log q)$ upper bound \hfill {\small [this work]}
+
+\end{itemize}
 

### 2023-06-12 presentation/podc-presentation.tex
x
added slide about ordering tree

@@
 \newcommand{\Green}[1]{{\color{green}#1}}
+\newcommand{\Purple}[1]{{\color{violet}#1}}
 
@@
 
+\begin{frame}{Ordering Tree}
+\visible<5>{\hfill Use ordering in root as linearization\hfill}
+\medskip
+
+\only<1>{\input{OrderTree-50.pdf_t}}%
+\only<2>{\input{OrderTree-50to51.pdf_t}}%
+\only<3>{\input{OrderTree-50to52.pdf_t}}%
+\only<4->{\input{OrderTree-50to53.pdf_t}}
+
+\end{frame}
+
+\begin{frame}{Propagating Operations to the Root}
+
+\begin{enumerate}
+\item
+Append operation to your leaf
+\item
+At each node $v$ on path to root refresh {\it twice}:
+\begin{enumerate}[(a)]
+\item
+Read unpropagated operations in both of $v$'s children
+\item
+CAS them into $v$
+\end{enumerate}
+\end{enumerate}
+
+\pause
+\medskip
+
+Double Refresh:\\
+If your CAS on $v$ fails twice, then\\
+another process has propagated your operation to $v$.
+
+\pause
+\medskip
+Avoids CAS retry problem.
+\end{frame}
+
+\begin{frame}{Main Challenge}
+
+Refresh may have to propagate up to $p$ operations\\
+$\Rightarrow$ need an \Red{implicit representation}
+
+\end{frame}
+
 \begin{frame}{Future Directions}

### 2023-06-13 presentation/podc-presentation.tex
finished first draft of presentation

@@
 
-\only<1>{\input{MSQueue-48to50.pdf_t}}
-\only<2>{\input{MSQueue-48to51.pdf_t}}
-\only<3>{\input{MSQueue-48to52.pdf_t}}
-\only<4>{\input{MSQueue-48-50to53.pdf_t}}
-\only<5>{\input{MSQueue-50to54.pdf_t}}
+\only<1>{\input{MSQueue-1}}% 48-50
+\only<2>{\input{MSQueue-2}}% 48-51
+\only<3>{\input{MSQueue-3}}% 48-52
+\only<4>{\input{MSQueue-4}}% 48,50-53
+\only<5>{\input{MSQueue-5}}% 50-54
 \end{frame}
@@
 
-\only<1>{\input{CASretry-49to50.pdf_t}}
-\only<2>{\input{CASretry-49to51.pdf_t}}
-\only<3->{\input{CASretry-49to52.pdf_t}}
+\only<1>{\input{CASretry-1}}% 49-50
+\only<2>{\input{CASretry-2}}% 49-52
+\only<3->{\input{CASretry-3}}% 49-52
 
@@
 
-\begin{frame}{Our Result}
+\begin{frame}{Breaking Linear-Time Bottleneck}
 
-\begin{block}{New Queue}
+\begin{block}{Our New Queue}
 \begin{itemize}
@@
 \visible<5>{\hfill Use ordering in root as linearization\hfill}
-\medskip
 
-\only<1>{\input{OrderTree-50.pdf_t}}%
-\only<2>{\input{OrderTree-50to51.pdf_t}}%
-\only<3>{\input{OrderTree-50to52.pdf_t}}%
-\only<4->{\input{OrderTree-50to53.pdf_t}}
+\hspace*{-7mm}
+\only<1>{\input{OrderTree-1}}% 50
+\only<2>{\input{OrderTree-2}}% 50-51
+\only<3>{\input{OrderTree-3}}% 49-52
+\only<4->{\input{OrderTree-4}}% 50-53
 
@@
 
-Double Refresh:\\
+\Red{Double Refresh}:\\
 If your CAS on $v$ fails twice, then\\
@@
 
+\begin{frame}{Requirements for Implicit Representation}
+
+% 60% size; others are 80%
+\hspace*{7mm}
+\only<1>{\input{OrderTree-5}}% 50-53
+\only<2>{\input{OrderTree-6}}% 48,50-53
+\only<3>{\input{OrderTree-7}}% 50-54
+\only<4>{\input{OrderTree-8}}% 50-55
+\only<5>{\input{OrderTree-9}}% 50-56
+
+Must support following in polylog time
+\begin{itemize}
+\pause
+\item Refresh: promote batch of ops from children to parent
+\pause
+\item Find my \op{Dequeue} in root
+\pause
+\item Check if \op{Dequeue} returns null, or otherwise\\
+determine rank of \op{Dequeue} among non-null \op{Dequeues}
+\pause
+\item Find \op{Enqueue} of given rank (and its argument)
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Implicit Representation:  Blocks}
+
+\only<1>{\input{OrderTree-5}}% 50-53
+\only<2->{\input{OrderTree-10}}% 47,50-53
+
+\vfill
+
+\only<1-2>{\input{Blocks-0}}% none
+\only<3>{\input{Blocks-1}}% 50
+\only<4>{\input{Blocks-2}}% 49-50
+\only<5>{\input{Blocks-3}}% 47-51
+\only<6>{\input{Blocks-4}}% 46-51
+
+\end{frame}
+
+\begin{frame}{Adding a Block for a Refresh}
+\input{OrderTree-6}% 48,50-53
+
+\vfill
+
+\input{Blocks-5}% 46,48-50,52
+\end{frame}
+
+\begin{frame}{Tracing a \op{Dequeue} to the Root}
+\input{OrderTree-7}% 50-54
+
+\vfill
+
+\input{Blocks-6}% 46,48-51,53,54
+\end{frame}
+
+\begin{frame}{Check if \op{Dequeue} Returns Null}
+\input{OrderTree-7}% 50-54
+
+\vfill
+
+\input{Blocks-7}
+\end{frame}
+
+\begin{frame}{Rank of \op{Dequeue} Among Non-Null \op{Dequeues}}
+\input{OrderTree-8}%
+
+\vfill
+
+\input{Blocks-8}
+\end{frame}
+
+\begin{frame}{Find \op{Enqueue} of Given Rank}
+\input{OrderTree-9}
+
+\vfill
+
+\input{Blocks-9}
+\end{frame}
+
+\begin{frame}{Our Result}
+
+\begin{block}{New Wait-Free Queue}
+\begin{itemize}
+\item $O(\log p)$ steps per \op{Enqueue}
+\item $O(\mbox{log}^2 p + \log q)$ steps per \op{Dequeue}
+\item<2> Unbounded space
+\end{itemize}
+\end{block}
+
+$p = \#$ processes\\
+$q = \#$ elements in queue
+\end{frame}
+
+\begin{frame}{Bounding Space}
+
+\begin{itemize}
+\item Replace array of blocks with a red-black tree of blocks
+\item Periodically split RBT and discard obsolete blocks
+\item Processes  help one another to ensure blocks are obsolete
+\end{itemize}
+
+\begin{block}{Bounded-Space Queue}
+\begin{itemize}
+\item Amortized $O(\log p\ \log(p+q))$ steps per operation
+\item $O(pq+p^3\log p)$ space
+\item Still wait-free
+\end{itemize}
+\end{block}
+
+$p = \#$ processes\\
+$q = \#$ elements in queue
+
+\end{frame}
+
 \begin{frame}{Future Directions}
@@
 $\Omega(\log \log p)$ lower bound \hfill {\small [Attiya Fouren 2017]}\\
-$O(\log^2 p + \log q)$ upper bound \hfill {\small [this work]}
+$O(\mbox{log}^2 p + \log q)$ upper bound \hfill {\small [this work]}
 

### 2023-06-13 presentation/podc-presentation.tex
Some edits to slides

@@
 \maketitle
+\end{frame}
+
+\begin{frame}{Queues:  Breaking Linear-Time Bottleneck}
+
+Problem:  implement lock-free FIFO queue
+\begin{itemize}
+\item shared by $p$ processes
+\item use single-word CAS (reasonable-sized words)
+\item support multiple enqueuers, dequeuers
+\end{itemize}
+
+\medskip
+\pause
+
+Many previous solutions for this problem.
+
+\medskip
+
+All require $\Omega(p)$ steps per operation\\
+$\rightarrow$ Real obstacle to scalability
+
+\pause
+\begin{block}{Our New Queue}
+\begin{itemize}
+\item $O(\log p)$ steps per \op{Enqueue}
+\item $O(\mbox{log}^2 p + \log q)$ steps per \op{Dequeue} ($q=$ size of queue)
+\item wait-free
+\end{itemize}
+\end{block}
+
 \end{frame}
@@
 \item $O(\log p)$ steps per \op{Enqueue}
-\item $O(\log^2 p + \log q)$ steps per \op{Dequeue}
+\item $O(\mbox{log}^2 p + \log q)$ steps per \op{Dequeue}
 \item wait-free
@@
 % 60% size; others are 80%
-\hspace*{7mm}
+%\hspace*{7mm}
 \only<1>{\input{OrderTree-5}}% 50-53
@@
 \item $O(\mbox{log}^2 p + \log q)$ steps per \op{Dequeue}
+\item $O(\log p)$ CAS steps per \op{Dequeue}
 \item<2> Unbounded space
@@
 \begin{itemize}
-\item Replace array of blocks with a red-black tree of blocks
+\item Replace each array of blocks with a red-black tree of blocks
 \item Periodically split RBT and discard obsolete blocks

### 2023-06-19 presentation/podc-presentation.tex
minor changes

@@
 \renewcommand{\log}{\mbox{log }} % font looks funny for usual \log
+
+\newcommand{\itemnum}[1]{% reproduce beamer numbers as in enumerated lists
+\setcounter{enumi}{#1}\usebeamertemplate{enumerate item}% 
+}
 
@@
 
-\Red{Double Refresh}:\\
+\begin{block}{Double Refresh}
 If your CAS on $v$ fails twice, then\\
 another process has propagated your operation to $v$.
+\end{block}
 

### 2023-06-21 presentation/podc-presentation.tex
added photo
minor edits

@@
 
-\title{A Wait-Free Queue with Polylogarithmic Step Complexity}
+\title{A Wait-Free Queue with\\
+Polylogarithmic Step Complexity}
 
@@
 \maketitle
+\hspace*{29mm}\includegraphics[width=5cm]{Hossein.jpeg}
+
 \end{frame}
@@
 
-Problem:  implement lock-free FIFO queue
+Problem:  implement linearizable, lock-free FIFO queue
 \begin{itemize}
@@
 \medskip
-\visible<4->{
+\visible<3->{
 \begin{itemize}
-\item one CAS of next pointer succeeds.
-\item other \op{Enqueues} must try again.
+\item<3-> one CAS of next pointer succeeds.
+\item<4-> other \op{Enqueues} must try again.
 \item<5> $\Rightarrow$ starvation and $\Omega(p)$ steps per operation (amortized)
@@
 \end{itemize}
-\visible<2->{All have CAS retry problem  \hfill {\small [Morrisson, Afek 2013]}}
+\visible<2->{All have CAS retry problem  \hfill {\small [Morrison, Afek 2013]}}
 
@@
 
-\medskip
-
-Previous lock-free queues take $\Omega(p)$ steps per operation.
+%\medskip
+%
+%Previous lock-free queues take $\Omega(p)$ steps per operation.
 

### 2023-06-29 journal/algorithm-GC.tex
created directory for eventual journal version

--- 2023-05-11 podc_paper/algorithm-GC.tex
@@
 
-Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode2GC}.
+Pseudocode for the space-bounded implementation appears in Figures \ref{pseudocode1GC} and \ref{pseudocode2GC}.
 New or modified code appears in blue.
@@
 To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
-finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
-or a null dequeue using the \var{last} array.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
+uses the \var{last} array to
+find the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
+or a null dequeue.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
 dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
@@
 A search for this block may sometimes not find it, if it has already been discarded
-by another process's GC.
+by another process's GC phase.
 As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
@@
 will have written the \fld{response} there before discarding the needed block.
-We do not explicitly write this early termination every time we do a lookup in an RBT.
+We do not explicitly write this early termination in the pseudocode every time we do a lookup in an RBT.
 There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
@@
 \op{SplitBlock} because the required block has been discarded, 
-we continue doing GC, since we do not a GC phase on one node
+we continue doing GC, since we do not want a GC phase on one node
 to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks

### 2023-06-29 journal/algorithm.tex
created directory for eventual journal version

--- 2023-05-11 podc_paper/algorithm.tex
@@
 within that block  of the \var{i}th dequeue of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
-Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may be off by one from the superblock's true index.
+Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may differ from the superblock's true index by one.
 The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the dequeue's 
@@
 To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
-Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.
+Then, \op{GetEnqueue} traces down through the tree to find the required enqueue in a leaf.
 

### 2023-06-29 journal/boundedspace.tex
created directory for eventual journal version

--- 2023-05-11 podc_paper/boundedspace.tex
@@
 Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
-polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic.  For lack of space, details are in Appendix~\ref{reducing-details}.
+polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic.  For lack of space, details are in %Appendix~\ref{reducing-details}.
+\cite{full}.
 We replace the \fld{blocks} array in each node by a red-black tree (RBT)
@@
 Fortunately, there is an  RBT \op{Split} operation that can remove
-these blocks in logarithmic time \cite[Sec.~4.2]{Tar83}.
+these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
 

### 2023-06-29 journal/correctness-GC.tex
created directory for eventual journal version

--- 2023-05-11 podc_paper/correctness-GC.tex
@@
 
-Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{casGC} 
-installs $B'$ in $\var{v}.\fld{blocks}$.
+Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$
+that is installed in $\var{v}.\fld{blocks}$
+by $R$'s \op{CAS} at line \ref{casGC}.
 The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
@@
 \end{itemize}
-Intuitively, once a block is finished, operations no longer need to pass through the block
+Intuitively, once a block is finished, operations no longer need  the block
 to compute responses to operations.
@@
 \ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
-In either case, the step $st$ changes $\var{v}.\fld{blocks}$ is changed
+In either case, the step $st$ changes $\var{v}.\fld{blocks}$ 
 from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
@@
 By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
-are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
+are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
+the end of the \op{SplitBlock}.
 Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
-and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
+and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
 Thus, after this helping (and before step $st$),
@@
 By \Cref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
-is also finished when $st$ occurs.
+is also finished when $st$ occurs.  This completes the proof of \Cref{splitfinished}.
 \renewcommand{\qedsymbol}{$\diamondsuit$}
@@
 by the definition of finished.
-Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \op{response} field.
+Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.
 
@@
 \fld{index} is a multiple of $G$.
-$A$ performs a GC phase.
+Then, $A$ performs a GC phase.
 Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
@@
 Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
-must still be in progress at $C$; otherwise the process that dequeued it would have
+must still be in progress at~$C$; otherwise the process that dequeued it would have
 set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
@@
 \here{this proof may need some modification since code was changed to change how split point is computed}
-It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
+It follows from \Cref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
 that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
-blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.
+blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.
 
@@
 Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
-and show that the amortized number of GC steps each operation takes is also $O(\log p\log(p+q_{max}))$.
+and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
 If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
-We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once as follows.
+
+We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once.
 Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
@@
 ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
-will have $\index>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.
+will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.
 
 Each helping phase takes $O(p)$ steps to read the \var{last} array, and
-$O(p \log p \log(p+q_{max}))$ steps to \op{Help},
+$O(p \log p \log(p+q_{max}))$ steps in \op{Help},
 and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
@@
 We can amortize these steps over the operations that appear in 
-$\var{v}.\fld{blocks}[(r-1)p^2+1..p^2]$.
+$\var{v}.\fld{blocks}[(r-1)G+1..rG]$.
 By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
@@
 by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
-Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
+Since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
 

### 2023-06-29 journal/correctness-details.tex
created directory for eventual journal version

--- 2023-05-11 podc_paper/correctness-details.tex
@@
 For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
-Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1]+1$ and
-$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
-The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.
+Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
+$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
+The proof for line~\ref{deqNew}, where $B$ has a single dequeue, is similar.
 
@@
 So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
-Since the value stored $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
+Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
 
@@
 By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
-The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]$.
+The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.
 

### 2023-06-29 journal/conclusion.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound

--- 2023-05-11 podc_paper/conclusion.tex
@@
-% !TEX root =  podc-submission.tex
+% !TEX root =  queue.tex
 
@@
 
-A gap remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
-and Attiya and Fouren's $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17}.
-It would be interesting to determine how the true step complexity of lock-free queues (or, more generally, bags)
-depends on $p$.
-Since a queue is also a bag, our queue is the first lock-free bag we know of that has polylogarithmic step complexity.
+It would be interesting to close the gap that remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
+and Jayanti, Tarjan and Boix-Adser\`{a}'s $\Omega(\log p)$ lower bound \cite{JTB19}.
+For the more relaxed bag data structure the gap is larger between the  $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17} and our upper bound of $O(\log^2 p + \log q)$.
+Could the complexity for either queues or bags be made polylogarithmic in $p$ while being independent of the size $q$ of the data structure?
 
@@
 \here{doublecheck all preceding claims for accuracy.}
-We are investigating whether a similar approach can be used for stacks, deques or even priority queues.
+Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
+designed stacks and doubly-ended queues with polylogarithmic amortized step
+complexity.  This required a substantially different representation of the data stored in the ordering tree.
+\here{?  They also showed that a randomized implementation of the ordering tree technique can
+be used to obtain a (doubly-ended) queue with O(\here{?}) amortized expected step complexity,
+beating the  $\Omega(\min(c,\log\log p))$ lower bound on deterministic algorithms.}
+Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
+remains an open question.

### 2023-06-29 journal/related.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound

--- 2023-05-11 podc_paper/related.tex
@@
 Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} gave an improved implementation of fetch\&increment with better step complexity.
+
+\paragraph{Lower Bounds.}
+As mentioned in the introduction, it follows from Attiya and Fouren's lower bound on bag data structures
+\cite{DBLP:conf/opodis/AttiyaF17} that the
+amortized step complexity of operations on a queue is $\Omega(\min(c,\log\log p))$, where $c$ is contention.
+Subsequently, Jayanti, Tarjan and Boix-Adser\`{a} \cite{JTB19} proved a stronger $\Omega(\log p)$ lower bound on
+the amortized step complexity of queue operations.

### 2023-06-30 journal/related.tex
Added discussion of Dodds Haas and Kirsch's queue.

@@
 I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussing array-based queues.}
+
+Dodds, Haas and Kirsch described how a queue could be implemented similarly
+to their lock-free stack \cite{DHK15}.  To enqueue an item, a process adds the item, together with an associated timestamp,
+to the process's own single-enqueuer multi-dequeuer queue, which is implemented as a linked list.  
+Although enqueues are very efficient, dequeues take $\Omega(p)$ steps to look at all $p$ lists
+and return an item with the oldest timestamp.
+(Items added by concurrent enqueues may get the same timestamp.)
+Moreover, concurrent dequeues may compete to claim the same item using a CAS, resulting
+in a CAS retry problem.
 

### 2023-06-30 journal/related.tex
Updated information about Haas queue

--- 2023-06-29 journal/related.tex
@@
 I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussing array-based queues.}
+
+Haas \cite{Haa15} described a queue implementation based on timestamping.
+% This queue is also sketched in \cite{DHK15}.
+% A different? linearizability proof for this queue appears in Khyza et al.\ \cite{KDGP17}.
+To enqueue an item, a process adds the item, together with an associated timestamp,
+to the process's own single-enqueuer multi-dequeuer queue, which is implemented as a linked list.  
+Items added by concurrent enqueues may get the same timestamp.
+Although enqueues are very efficient, dequeues look at all $p$ lists
+to find and return an item with the oldest timestamp.
+This is compounded by the fact that dequeues may compete to claim the same item using a CAS, resulting
+in a CAS retry problem.  As a result, the amortized step complexity for dequeue operations
+is $\Omega(p^2)$.
 

### 2023-10-25 journal/related.tex
update to related work

@@
 \paragraph{Array-Based Queues.}
-Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
+Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,PMG16,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
 %Similar to the front and back pointers of list-based queues, indices
@@
 due either to the CAS retry problem or helping mechanisms.
+
+\here{Maybe mention experimental work of \cite{KLM+23} here or in intro}
 

### 2023-10-25 journal/algorithm.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
getting rid of some weird font macros
Moved Appendix A into main text

@@
 
-\Eric{Reworded following parag after Hossein's comments of Jan 4}
+%\Eric{Reworded following parag after Hossein's comments of Jan 4}
 To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ first observes
@@
 \begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of binary tree of \tt{Node}s with one leaf per process}
+\item \typ{Node} \var{root} \Comment{root of binary tree of \typ{Node}s with one leaf per process}
 \end{itemize}
@@
   	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueues, dequeues in \fld{blocks} array}\linebreak
-		\mbox{ }\Comment{up to this block (inclusive)}
+  		\Comment{number of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
   	\item \typ{int} \fld{super}
@@
 %\Statex {\com\  Blocks in internal nodes have the following additional fields}
-%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
+%\Statex $\blacktriangleright$ \typ{InternalBlock} extends \typ{Block} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
 %\begin{itemize}
-%	\item \tt{int} \eleft, \eright
+%	\item \typ{int} \eleft, \eright
 %  		\Comment{index of last direct subblock in the left and right child}
@@
 %\Statex {\com\ Blocks in leaf nodes have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
+%\Statex $\blacktriangleright$ \typ{LeafBlock} extends \typ{Block} \sf{\com\ the following additional field is used only for blocks in leaves}
 %\begin{itemize}
-%  \item \tt{\sl{Object} element}
-%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
+%  \item \typ{Object} \var{element}
+%  \Comment{if the block's operation is enqueue(x) then element=x, otherwise element=null.}
 %\end{itemize}
@@
 %\Statex {\com\ Blocks in the root node have the following additional field}
-%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
+%\Statex $\blacktriangleright$ \typ{RootBlock} extends \typ{InternalBlock} \sf{\com\ the following additional field is used only for blocks in the root}
 %\begin{itemize}
-%  \item \tt{\sl{int} \size}%
+%  \item \typ{int} \size%
 %  \Comment{size of the queue after performing all operations up to the end of this block}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \com\ propagate blocks from \var{v}'s children to root
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
         \State \Call{Refresh}{\var{v}} \label{secondRefresh}
@@
 
-%\paragraph{\tt{Refresh()} and \tt{Advance()}}
 A \opemph{Refresh} on node \var{v} creates a block representing the new blocks

### 2023-10-25 journal/correctness.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
getting rid of some weird font macros
Moved Appendix A into main text

--- 2023-05-12 podc_paper/correctness.tex
@@
 %\begin{definition}
-%$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \it{established} in node $n$ 
+%$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \emph{established} in node $n$ 
 %if it is in an established block of $n$. $EST^t_n$ is the set of established operations in node $n$ at time $t$.
@@
 on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
-from the definition of subblocks.  See 
-%Appendix \ref{app::tracingDetails} 
-the full version \cite{full} for a detailed proof of \Cref{lem::sum}.
+from the definition of subblocks.
 
-\begin{restatable}{invariant}{sumRes}
+\begin{invariant}
 \label{lem::sum}
@@
 \end{eqnarray*}
-\end{restatable}
+\end{invariant}
 
-This allows us to prove that every block a \op{Refresh} installs contains at least one operation.
+\begin{proof}
+Initially, each \fld{blocks} array  contains only an empty block $B_0$ in location 0.
+By definition, $E(B_0)$ and $D(B_0)$ are empty sequences.
+Moreover, $B_0.\fld{sum\sub{enq}} = B_0.\fld{sum\sub{deq}} = 0$, so the claim is true.
+
+We show that each installation of a block $B$ into some location $\var{v}.\fld{blocks}[i]$ preserves the claim,
+assuming the claim holds before this installation.  We consider two cases.
+
+If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
+For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
+Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
+$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
+The proof for line~\ref{deqNew}, where $B$ has a single dequeue, is similar.
+
+Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
+subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
+and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
+Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
+$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in 
+$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
+By the hypothesis, the total number of these enqueues is 
+$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
+which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
+The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
+\end{proof}
+
+\ref{lem::sum} allows us to prove that every block a \op{Refresh} installs contains at least one operation.
 
@@
 
-See 
-%Appendix \ref{app::tracingDetails}
-the full version \cite{full} for detailed proofs for this section.
-
 We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
@@
 
-\begin{restatable}{lemma}{superRelationRes}
+\begin{lemma}
 \label{superRelation}
@@
 If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
-\end{restatable}
+\end{lemma}
+\begin{proof}
+We first show that $B.\fld{super}\leq s$.
+Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
+in $\var{v}.\fld{parent.blocks}[s]$.
+By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
+By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
+that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
+When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
+\ref{cas} succeeds.
+So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
+Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
+
+Next, we show that $B.\fld{super}\geq s-1$.
+The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least 1, so $B.\fld{super} \geq 1$.
+So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
+By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
+$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
+Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
+This read $r'$ must be before $B$ is installed in $\var{v}$;
+otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
+$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
+Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
+It is invoked either 
+at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
+or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
+Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
+By \Cref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
+the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
+writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
+\end{proof}
 
@@
 
-\begin{restatable}{lemma}{indexDequeueRes}
+\begin{lemma}
 \label{lem::indexDequeue}
@@
  then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
-\end{restatable}
+\end{lemma}
 
-\begin{restatable}{lemma}{getEnqRes}
+\begin{proof}
+We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
+Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
+Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
+\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
+By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
+The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.
+
+Next, the position of the required dequeue in $D(B')$ is computed in 
+lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
+We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line \ref{computeISuperStart}.
+If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
+also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.
+
+Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
+Since $B$ has been propagated to the root, so has its superblock $B'$.
+Thus, all preconditions of the recursive call are met.
+By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}
+
+\begin{lemma}
 \label{lem::get}
 If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
-\end{restatable}
+\end{lemma}
+
+\begin{proof}
+We prove the claim by induction on the height of node $\var{v}$.
+If $\var{v}$ is a leaf, the hypothesis implies that $i=1$ and the block $\var{v}.\fld{blocks}[b]$ represents 
+an enqueue whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
+\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.
+
+Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
+Let $B$ be $\var{v}.\fld{blocks}[b]$.
+By (\ref{defSeqs}),
+$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
+of $B$, which are listed in (\ref{defsubblock}).
+By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
+of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
+Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue.
+Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
+Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
+By \Cref{lem::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
+enqueues in $E'$ before the enqueues of block $B'$, so
+the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
+Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
+returns the required result, by the induction hypothesis.
+\end{proof}
 
@@
 shows that \fld{size} fields are computed correctly.
-\begin{restatable}{lemma}{sizeCorrectRes}
+\begin{lemma}
 \label{sizeCorrectness}
 If the operations of $\var{root}.\fld{blocks}[0..b]$ are applied sequentially in the order of~$L$ on an initially empty queue, the resulting queue has $\var{root}.\fld{blocks}[b].\size$ elements.  
-\end{restatable}
+\end{lemma}
+
+\begin{proof}
+We prove the claim by induction on $b$. 
+The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
+$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
+Assuming the claim holds for $b-1$, we prove it for $b$.
+The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
+at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
+By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
+of block $B$ are performed.
+By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
+are the number of enqueues and dequeues contained in $B$.
+Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
+is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
+\end{proof}
 

### 2023-10-25 journal/related.tex
Moved Appendix A into main text

@@
 the amortized step complexity remains~$\Omega(p)$.
-\Hossein{I get what this sentence means, but from a high overview i think it is inaccurate.
-Since the step complexity of MS-queue is infinite, so nothing can be added to it and a wait-free queue is always better than a lock-free queue in the worst-case.}
-\Eric{I modified the wording.
-I also added the material about fast-path slow-path here (since I had forgotten that paper also applies it to a queue), and fixed the paragraph about FPSP method below when discussing array-based queues.}
 

### 2023-10-25 journal/algorithm.tex
redid formatting of main algorithm

@@
 \renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
-\begin{figure}
+\begin{figure*}
 \begin{algorithmic}
 \setcounter{ALG@line}{1}
-\Statex \linecomment{Shared variable}
-\begin{itemize}
-\item \typ{Node} \var{root} \Comment{root of binary tree of \typ{Node}s with one leaf per process}
-\end{itemize}
-
-\Statex \linecomment{Thread-local variable}
-\begin{itemize}
-\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
-\end{itemize}
 
@@
 	\item[\com] Blocks in internal \nodes\ have the following additional fields
-	\begin{itemize}[leftmargin=3mm]
+	\begin{itemize}[leftmargin=0mm]
 		\item[$\bullet$] \typ{int} \eleft, \eright
@@
   	\item[\com] Blocks in leaf \nodes\ have the following additional field
-  	\begin{itemize}[leftmargin=3mm]
+  	\begin{itemize}[leftmargin=0mm]
 		\item[$\bullet$] \typ{Object} \fld{element}
@@
 	\item[\com] Blocks in the root \node\ have the following additional field
-	\begin{itemize}[leftmargin=3mm]
+	\begin{itemize}[leftmargin=0mm]
 		\item[$\bullet$] \typ{int} \size%
@@
 \caption{Objects used in the \ordering\ tree data structure.\label{object-fields}\Eric{Fix indentation}}
-\end{figure}
+\end{figure*}
 
@@
 
-We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1}.
+We now discuss the queue implementation in more detail.  Pseudocode is provided in Figures \ref{code1}--\ref{code3}.
 %We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
@@
 \begin{figure*}
-\begin{minipage}[t]{0.405\textwidth}
 \begin{algorithmic}[1]
@@
 
+\Statex \linecomment{Shared variable}
+\begin{itemize}
+\item \typ{Node} \var{root} \Comment{root of ordering tree}
+\end{itemize}
+
+\Statex \linecomment{Thread-local variable}
+\begin{itemize}
+\item \typ{Node} \var{leaf} \Comment{process's leaf in the ordering tree}
+\end{itemize}
+
+\spac
+
 \Function{void}{Enqueue}{\typ{Object} \var{e}} 
-    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		$\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
-		$\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
+    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
+		\fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}+1$,\\
+		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
     \State \Call{Append}{\var{B}}
@@
 \Function{Object}{Dequeue()}{} 
-    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    $\fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
-	    $\fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
+    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
+	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
+	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
     \State \Call{Append}{\var{B}}
@@
 
-\Function{void}{Append}{\typ{Block} \var{B}} \com\ append block to leaf and propagate to root
+\Function{void}{Append}{\typ{Block} \var{B}} \Comment{append block to leaf and propagate to root}
     \State \var{leaf.}\fld{blocks}[\var{leaf.}\head] \assign\ \var{B}\label{appendLeaf}
@@
 
-\Function{void}{Propagate}{\typ{Node} \var{v}} \com\ propagate blocks from \var{v}'s children to root
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \com\ double refresh
+\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh}
         \State \Call{Refresh}{\var{v}} \label{secondRefresh}
     \EndIf
-    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
+    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
         \State \Call{Propagate}{\var{v}.\fld{parent}}
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}} \com\ try to append a new block to \var{v}.\fld{blocks}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block to \var{v}.\fld{blocks}}
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
@@
 \EndFunction{Refresh}
+\end{algorithmic}
+\caption{Queue implementation's main routines.\label{code1}}
+\end{figure*}
+
+\begin{figure*}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{39}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \Comment{create new block for \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State \hangbox{let \var{new} be a new \typ{Block} with fields\label{initNewBlock}
+    	\eleft\ \assign\ $\var{v}.\fld{left}.\head - 1$,\label{createEndLeft}
+    	\eright\ \assign\ $\var{v}.\fld{right}.\head - 1$,\label{createEndRight}\\
+		\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}\label{createSumEnq},\\
+		\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
+    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
+    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \If{$\var{v} = \var{root}$} \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
+    \EndIf
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \Comment{no blocks need to be propagated to \var{v}} \label{testEmpty} 
+    \Else\ \Return \var{new}
+    \EndIf
+\EndFunction{CreateBlock}
 
@@
 
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
-    \State \linecomment{create new block for a \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State let \var{new} be a new \typ{Block} \label{initNewBlock}
-    \State \var{new}.\eleft \assign\ $\var{v}.\fld{left}.\head - 1$\label{createEndLeft}
-    \State \var{new}.\eright \assign\ $\var{v}.\fld{right}.\head - 1$\label{createEndRight}
-	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
-			\hspace*{11mm}\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}\label{createSumEnq}
-	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
-			\hspace*{11mm}\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
-    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$}
-        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLength}
-    \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmpty}
-        \State \Return \nl \hfill \com\ no blocks need to be propagated to \var{v}
-    \Else
-        \State \Return \var{new}
-    \EndIf
-\EndFunction{CreateBlock}
-
-\end{algorithmic}
-\end{minipage}
-\begin{minipage}[t]{0.585\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{57}
 \Function{void}{Advance}{\typ{Node} \var{v}, \typ{int} \var{h}} \com\ set \var{v}.\fld{blocks}[\var{h}].\fld{super} and increment \var{v}.\fld{head} from \var{h} to $\var{h}+1$
@@
 
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \com\ return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in
-    \State \linecomment{$D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
-    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least}
-    \State \linecomment{\var{i} dequeues}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \com\ return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$
+    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
+    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} dequeues}
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
-        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
-    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
+        \State \Return \nl \hfill \Comment{queue is empty when dequeue occurs}\label{returnNull}
+    \Else \ \Comment{response is the \var{e}th enqueue in the root}
         \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
 			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
-		\State \linecomment{compute enqueue's block using binary search}
-		\State find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
-		\State \linecomment{find rank of enqueue within its block}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\label{FRi}
+		\State use binary search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
         \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
@@
 \EndFunction{FindResponse}
+\end{algorithmic}
+\caption{Queue implementation's helper routines. \label{code2}}
+\end{figure*}
 
-\spac
+\begin{figure*}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{85}
 
@@
         \EndIf \label{endChooseDir}
-        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
-        \State \hangbox{find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such that\\
-        	 $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}}
+        \State \linecomment{find enqueue's block in \var{v}.\fld{dir} and its rank within block}
+        \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\
+        	 that $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}}
         \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
@@
 \end{algorithmic}
-\end{minipage}
-\vspace*{-3mm}
-\caption{Queue implementation.\label{pseudocode1}}
+\caption{Queue implementation's \op{GetEnqueue} routine.\label{code3}}
 \end{figure*}

### 2023-10-25 journal/correctness.tex
redid formatting of main algorithm

@@
 \begin{eqnarray*}
-\var{num\sub{enq}} \hspace*{-2mm}
-&=& \hspace*{-2mm} \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} \\
+\var{num\sub{enq}}
+&=&  \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} \\
 && + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}}\\
 &&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
-&=& \hspace*{-2mm}\var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}}\\
+&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}}\\
 && + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\

### 2023-10-25 journal/algorithm-GC.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
getting rid of some weird font macros
reformatting GC code

@@
 
-Pseudocode for the space-bounded implementation appears in Figures \ref{pseudocode1GC} and \ref{pseudocode2GC}.
+Pseudocode for the space-bounded implementation appears in Figures \ref{code1GC} to \ref{code3GC}.
 New or modified code appears in blue.
@@
 \begin{figure*}
-\begin{minipage}[t]{0.41\textwidth}
 \begin{algorithmic}[1]
@@
     \gc{
-    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
-	\State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \var{e},\\
-		$\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
-		$\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,\\
-		$\fld{index}\assign h$}\label{enqNewGC}
-	}
+    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
+	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
+		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
+		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,
+		\fld{index}\ \assign $h$}\label{enqNewGC}}
     \State \Call{Append}{\var{B}}
@@
     \gc{
-    \State \var{h} \assign \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
-    \State \hangbox{let \var{B} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
-	    $\fld{sum\sub{enq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
-	    $\fld{sum\sub{deq}} \assign\ \var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,\\
-	    $\fld{index}\assign h$}\label{deqNewGC}
+    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
+    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
+	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
+	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,
+	    \fld{index} \assign $h$}\label{deqNewGC}
 	}
@@
 \gc{
-\Function{Object}{CompleteDeq}{\var{leaf}, $h$} 
-    \State \linecomment finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
-    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1}$
-    \State $\var{res} \assign \Call{FindResponse}{b, i}$
+\Function{Object}{CompleteDeq}{\var{leaf}, $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$
+    \State \var{res} \assign $\Call{FindResponse}{b, i}$
 	\State \Return{\var{res}}
@@
 
-\Function{void}{Append}{\typ{Block} \var{B}}  \linecomment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
+\Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}
+    \gc{\State \var{leaf}.\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
     \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
@@
 
+\Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh}
+        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
+    \EndIf
+    \If{$\var{v} \neq \var{root}$} \hfill \Comment{recurse up tree}
+        \State \Call{Propagate}{\var{v}.\fld{parent}}
+    \EndIf
+\EndFunction{Propagate}
+
+\spac
+
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block to $\var{v}.\fld{blocks}$}
 	\gc{
-\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} 
-    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
-    \If{\var{B}.\fld{index} is a multiple of $G$}
-    	\State \linecomment{Do garbage collection}\label{GCstart}
-        \State $s \assign \Call{SplitBlock}{\var{v}}.\var{index}$\label{callSplitBlock}
+	\State $T$ \assign\ $\var{v}.\fld{blocks}$\label{refreshReadTGC}
+    \State $\var{h}$ \assign\ $\Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
+    }
+    \State $\var{new}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$ \label{invokeCreateBlockGC}
+    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
+    \Else  
+    	\gc{
+	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
+		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
+		}
+    \EndIf
+\EndFunction{Refresh}
+
+\spac
+
+\gc{
+\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
+    \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
+        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\label{callSplitBlock}
         \State \Call{Help}{}\label{invokeHelp}
-        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
-        \State \linecomment{\op{Split} removes blocks with $\fld{index} < s$}
+        \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
         \State \Return \Call{Insert}{$T',B$}\label{GCend}
@@
 
-\spac
+\end{algorithmic}
+\caption{Bounded-space queue implementation's main routines.\label{code1GC}}
+\end{figure*}
 
+\begin{figure*}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{247}
 \gc{
-\Function{Block}{SplitBlock}{\typ{Node} v}
-	\State \linecomment{figure out where to split \var{v}'s RBT}
+\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{figure out where to split \var{v}'s RBT}
 	\If{$\var{v}=\var{root}$}
-		\State $m \assign 0$
-		\For{$k\assign 1..p$} $m\assign \max(m, \var{v}.\fld{last}[k])$
+		\State $m$ \assign\ 0
+		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
 		\EndFor
-		\State $B\assign \var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
+		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
 	\Else
-		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
-		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
-		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
+		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
+		\State \var{dir} \assign\ ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
+		\State $B$ \assign\ $\var{v}.\fld{blocks}[B_p.\edir]$\label{SBlookup2}
 	\EndIf
-	\State \linecomment If $B$ was discarded, use leftmost block instead
-	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \label{substitute}
+	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, use leftmost block instead}\label{substitute}
 \EndFunction{SplitBlock}
@@
 
-\Function{void}{Propagate}{\typ{Node} \var{v}}
-    \State \linecomment{propagate blocks from \var{v}'s children to root}
-    \If{\bf{not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \com\ double refresh
-        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
+\gc{
+\Function{void}{Help}{} \Comment{help pending operations}
+
+    \For{$\ell$ in $\var{leaf}[1..k]$}
+        \State $\var{B}$ \assign\ $\Call{MaxBlock}{\ell.\fld{blocks}}$
+    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
+        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
+            \State \linecomment{operation is a propagated dequeue}
+            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
         \EndIf
-    \If{$\var{v} \neq \var{root}$} \hfill \com\ recurse up tree
-        \State \Call{Propagate}{\var{v}.\fld{parent}}
-    \EndIf
-\EndFunction{Propagate}
-
+    \EndFor
+\EndFunction{Help}
+}
 \spac
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}} \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
 \gc{
-	\State $T \assign \var{v}.\fld{blocks}$\label{refreshReadTGC}
-    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
-    }
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
-    \Else  
-    	\gc{
-	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
-		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
-		}
-    \EndIf
-
-\EndFunction{Refresh}
-
-\end{algorithmic}
-\end{minipage}
-\hfill
-\begin{minipage}[t]{0.56\textwidth}
-
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{267}
-
-\gc{
-\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
+\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \Comment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
     \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
@@
     \Else
-    	\State $T\assign \var{v}.\fld{parent.blocks}$
+    	\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
 		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
@@
 		\Else
-			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
+			\State $B_p$ \assign\ minimum index block in $T$ with $\edir \geq b$\label{searchsuper3}
 			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
@@
 
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State \hangbox{let \var{new} be a new \typ{Block} with fields\label{initNewBlockGC}
+    \gc{
+    	\eleft\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC},\\
+    \eright\ \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$,\label{createEndRightGC}
+    \fld{index} \assign\ $i$\label{createIndexGC}
+    }\\
+	\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}\\
+	\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} +
+			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
+    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
+    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \If{$\var{v} = \var{root}$} \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
+        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
+    \EndIf
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \hfill \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
+    \Else\ \Return \var{new}
+    \EndIf
+\EndFunction{CreateBlock}
+
+\end{algorithmic}
+\caption{Bounded-space queue implementation's routines for GC and creating new \typ{Block}.\label{code2GC}}
+\end{figure*}
+
+\begin{figure*}
+\begin{algorithmic}[1]
+\setcounter{ALG@line}{290}
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
@@
     	\gc{
-		\State $T\assign \var{v}.\fld{parent.blocks}$
-		\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper1}
-		\State $B_p'$ \assign max block in $T$ with $\edir < b$\label{searchsuper2}
+		\State $T$ \assign\ $\var{v}.\fld{parent.blocks}$
+		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
+		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$\label{searchsuper2}
 	    }
@@
 
-\gc{
-\Function{void}{Help}{} \linecomment{help pending operations}
-
-    \For{$\ell$ in $\var{leaf}[1..k]$}
-        \State $\var{B} \assign \Call{MaxBlock}{\ell.\fld{blocks}}$
-    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
-        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
-            \State \linecomment{operation is a propagated dequeue}
-            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
-        \EndIf
-    \EndFor
-\EndFunction{Help}
-}
-\spac
-
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
-    \gc{
-    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
-    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
-    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
-    }
-	\State \var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}
-	\State \var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} +
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
-    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$}
-        \State \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
-        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
-    \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
-        \State \Return \nl \hfill \com\ no blocks to be propagate to \var{v}\label{CBnullGC}
-    \Else
-        \State \Return \var{new}
-    \EndIf
-\EndFunction{CreateBlock}
-
-\spac
-
 \Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$
-    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \linecomment{queue is empty when dequeue occurs}
+    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} -
+    		\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}
         \gc{
-        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
+        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
         \EndIf
@@
         \State \Return \nl \hfill \label{returnNullGC}
-    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
-        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
-			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
+    \Else \ \Comment{response is the \var{e}th enqueue in the root}
+        \State $\var{e}$\ \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size$\label{computeEGC}
 
-		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC} \linecomment{use binary search}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \linecomment{find rank of enqueue within its block}
+		\State use \gc{BST search in \var{root}.\fld{blocks} to find minimum index $b_e$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{e}$}\label{FRsearchGC} 
+		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of enqueue within its block}
 		\gc{
-        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
-        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
+        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
         \EndIf
@@
 
-\end{algorithmic}
-\end{minipage}
-\vspace*{-4mm}
-\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
-\end{figure*}
-
-\begin{figure*}
-\begin{algorithmic}[1]
-\setcounter{ALG@line}{341}
+\spac
 
@@
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRightGC} \cmt{required enqueue is in \var{v}.\fld{left}}
-            \State \fld{dir} \assign\ \fld{left}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\Comment{required enqueue is in \var{v}.\fld{left}}\label{leftOrRightGC} 
         \Else \Comment{required enqueue is in \var{v}.\fld{right}}
@@
         \EndIf
-        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
-        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t.
-$\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChildGC}
+        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
+        \State \gc{use BST search in \var{v}.\fld{dir.blocks} to find minimum index $\var{b}'$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$}\label{getChildGC}
         \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$
@@
 \end{algorithmic}
-
-\caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
+\caption{Bounded-space queue implementation's  routines to compute responses to operations.\label{code3GC}}
 \end{figure*}
@@
 %\begin{algorithm}
-%\caption{\tt{\sl{Node}}}
+%\caption{\typ{Node}}
 %\begin{algorithmic}[1]
@@
 %
-%\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
-%\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
-%\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
-%\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
-% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
+%\Statex $\leadsto$ \textsf{Precondition: \var{blocks[start..end]} contains a block with \fld{sum\sub{enq}} greater than or equal to \var{x}}
+%\Statex \com\ \textmd{Does a binary search for~the value \var{x} of \fld{sum\sub{enq}} field and returns the index of the leftmost block in\\
+%\com\ \var{blocks[start..end]} whose \fld{sum\sub{enq}} is $\geq$ \var{x}}.
+%\Function{int}{BinarySearch}{\typ{int} x, \typ{int} start, \typ{int} end}
+% \State \Return $\min\{j: \fld{blocks}[j].\fld{sum\sub{enq}} \geq x\}}
 %\While{\nf{start<end}}
-%\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
+%\State \typ{int} \var{mid} \assign\ floor((start+end)/2)
 %\If{\nf{blocks[mid].sum\sub{enq}<x}}
@@
 %\begin{algorithm}
-%\caption{\tt{\sl{Root}}}
+%\caption{\typ{Root}}
 %\begin{algorithmic}[1]
@@
 %\Statex
-%\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
-%\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
+%\Statex $\leadsto$ \textsf{Precondition: \var{root.blocks[end].sum\sub{enq} $\geq$ \var{e}}}
+%\Statex \com\ \textmd{Returns \var{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
 %\com\ the \nf{b}th block in the \nf{root}.}
 %
-%\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
-%\State \tt{start \assign\ end-1} \label{dsearchStart}
-%\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
-%\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
+%\Function{<int, int>}{DoublingSearch}{\typ{int} e, \typ{int} end}
+%\State \var{start \assign\ end-1} \label{dsearchStart}
+%\While{\var{root.blocks[start].sum\sub{enq}}$\geq$\var{e}}
+%\State \var{start \assign\ max(start-(end-start), 0)} \label{doubling}
 %\EndWhile \label{dsearchEnd}
-%\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
-%\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
-%\State\Return \tt{<b,i>}
+%\State \var{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
+%\State \var{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
+%\State\Return \var{<b,i>}
 %\EndFunction{DoublingSearch}

### 2023-10-25 journal/algorithm.tex
reformatting GC code

@@
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} 
-        \State \linecomment{\var{sum\sub{left}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}'s left child}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} 
-        \State \linecomment{\var{prev\sub{left}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}'s left child}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} 
-        \State \linecomment{\var{prev\sub{right}} is the number of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}'s right child}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
-            \State \fld{dir} \assign\ \fld{left}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
         \Else \Comment{required enqueue is in \var{v}.\fld{right}}
@@
         \EndIf \label{endChooseDir}
-        \State \linecomment{find enqueue's block in \var{v}.\fld{dir} and its rank within block}
+        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\

### 2023-10-25 journal/correctness.tex
reformatting GC code

@@
 
-\ref{lem::sum} allows us to prove that every block a \op{Refresh} installs contains at least one operation.
+Invariant \ref{lem::sum} allows us to prove that every block a \op{Refresh} installs contains at least one operation.
 
@@
 \label{sec::propagating}
+
+\here{cref doesn't seem to be working (it calls everything a theorem)}
 
@@
 
-A simple proof (in 
-%Appendix \ref{app::tracingDetails})
-the full version \cite{full})
-shows that \fld{size} fields are computed correctly.
+We next show that \fld{size} fields are computed correctly.
 \begin{lemma}
@@
 are the number of enqueues and dequeues contained in $B$.
-Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by $L$)
+Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by~$L$)
 is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
@@
 
-Next, we show operations return the same response as they would in the sequential execution $L$.
+Next, we show each operation returns the same response as it would in the sequential execution~$L$.
 
 \begin{lemma}\label{linearCorrect}
-Each terminating dequeue returns the response it would in the sequential execution $L$.
+Each terminating dequeue returns the response it would in the sequential execution~$L$.
 \end{lemma}

### 2023-10-25 journal/introduction.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
updated some related work

--- 2023-05-11 podc_paper/introduction.tex
@@
 $\Omega(p)$ CAS instructions, even in an amortized sense.
+Since a queue is also a bag, our queue is the first lock-free bag that we know of that has polylogarithmic step complexity.
 Both versions of our queue use single-word CAS on reasonably-sized 
@@
 and avoids the CAS retry problem.
-Operations in the root are ordered, 
-and this order is used to linearize the operations and compute their responses.
-\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
-Explicitly storing  operations in the tree nodes would be too costly.
-Instead, we use a novel implicit representation of sets
-of operations that allows us to quickly merge two sets from the children of a node,
-and quickly access any  operation in a~set.
+Operations reach the root in batches and are linearized in the order they reach the root.
+Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
+or applying them one-by-one to the queue would be too costly.
+Instead, we use a novel implicit representation of batches
+of queue operations that allows us to quickly merge two batches from the children of a node,
+and quickly access any  operation in a~batch.
 A preliminary version of this work appeared in~\cite{Nad22}.

### 2023-10-25 journal/related.tex
updated some related work

@@
 to mix enqueues and dequeues.
-\here{Is this worth saying:
-They use some properties of the queue size before and after a batch, similar to a part of our work.}
+%\here{Is this worth saying:
+%They use some properties of the queue size before and after a batch, similar to a part of our work.}
 In the worst case, where operations require their response right away,
@@
 \paragraph{Universal Constructions.}
+\Eric{Rewrote this paragraph on Oct 25}
 One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
+Afek, Dauber, and Touitou's universal construction~\cite{DBLP:conf/stoc/AfekDT95}
+introduced the technique where a process must climb a binary tree while helping any other processes it sees along the way to reach the root,
+and this technique is the basis of our queue implementation.
 Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
-the universal construction of Afek, Dauber, and
-Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
+their construction can be modified to use $O(\log p)$ steps per operation, 
 assuming that words can store $\Omega(p \log p)$ bits. 
-(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
+However, if more reasonably-sized $O(\log p)$-bit words words are used, the construction would take $\Omega(p\log p)$ steps per operation.
+There are two obstacles to making this construction more efficient:
+it is expensive to manipulate lists of up to $p$ operations that must be propagated along a path
+up to the root, and
+when a batch of up to $p$ operations reaches the root, it is expensive to perform all of them on the
+implemented data structure.
+In our work, we devise a novel way of doing this implicitly for batches of operations on a queue.
+
 Fatourou and Kallimanis \cite{FK14} used their own universal construction based on fetch\&add and LL/SC instructions
@@
 queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but only for a single dequeuer. 
-Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
+Like our queue, theirs uses the cooperative tree-climbing
+technique of Afek, Dauber and Touitou~\cite{DBLP:conf/stoc/AfekDT95}.
 Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue  
@@
 
-\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?
+%\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?
 May not be worth mentioning because then $p$ is a constant, so complexity in terms of $p$ doesn't make sense.}
@@
 double-word read-modify-write operations.
-They use this as evidence that their primitives can be more efficient than CAS
+They use this as evidence to argue that their primitives can be more efficient than CAS
 since previous CAS-based queues all required $\Omega(p)$ step complexity.
-Our new implementation counters this argument.
+Our new implementation undermines that argument.
 
@@
 as the single-word state of a fetch\&increment object.
-Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} gave an improved implementation of fetch\&increment with better step complexity.
+Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} then gave an improved implementation of fetch\&increment with better step complexity.
 
@@
 amortized step complexity of operations on a queue is $\Omega(\min(c,\log\log p))$, where $c$ is contention.
-Subsequently, Jayanti, Tarjan and Boix-Adser\`{a} \cite{JTB19} proved a stronger $\Omega(\log p)$ lower bound on
+Subsequently, Jayanti, Tarjan and Boix-Adser\`{a} \cite{JTB19} proved a  $\Omega(\log c)$ lower bound on
 the amortized step complexity of queue operations.
+The combination of these lower bounds is shown schematically in Figure \ref{graph-lb}.
+If we are interested in the complexity only as a function of $p$, the second result
+gives us an $\Omega(\log p)$ lower bound, since $c$ can be as high as $p$.
+
+\begin{figure}
+\input{graph-lb.pdf_t}
+\caption{Known lower bounds on amortized step complexity of queue operations when contention is~$c$ in a system of $p$ processes.\label{graph-lb}}
+\end{figure}

### 2023-10-26 journal/conclusion.tex
minor edits, mostly to Sec 1 and 2

@@
 Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
-designed stacks and doubly-ended queues with polylogarithmic amortized step
-complexity.  This required a substantially different representation of the data stored in the ordering tree.
+designed a doubly-ended queue with polylogarithmic amortized step
+complexity, which also yields a stack as a special case.  
+This required a substantially different representation of the data stored in the ordering tree.
 \here{?  They also showed that a randomized implementation of the ordering tree technique can

### 2023-10-26 journal/introduction.tex
minor edits, mostly to Sec 1 and 2

@@
 Besides being a fundamental data structure, queues are used in
-significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
+significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g.,\cite{BBFRSW21}),
+% ouroboros paper by Winter et al 2020 might be a source for queues used for mem management (in context of GPUs)
 %packet processing \cite{DPDK}, 
-synchronization \cite{KAE23},\here{is this a good citation for this?}
+synchronization \cite{KAE23},%\here{is this a good citation for this?}
 and sharing resources or tasks.
@@
 is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.
+We study the \emph{amortized step complexity} per operation, which is measured by taking the maximum, over all possible
+executions, of the number of steps in the execution divided by the total number of enqueue and dequeue operations in the execution.
 
-The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
+The lock-free \emph{MS-queue} of Michael and Scott \cite{MS98} is a classic shared queue implementation.
 It uses a singly-linked list with pointers to the front and back nodes.
@@
 fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
-(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
-and divide the number of steps in the execution by the number of operations  in the execution.)
 Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but
@@
 Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
-The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
-Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
+The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/opodis/GidenstamST10,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
+Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16}
 rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.
 
-Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert \cite{Rup16}.
+Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert~\cite{Rup16}.
 For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
@@
 Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
-It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
+It is also \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
 For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
@@
 $\Omega(p)$ CAS instructions, even in an amortized sense.
-Since a queue is also a bag, our queue is the first lock-free bag that we know of that has polylogarithmic step complexity.
+Since a queue is also a bag, our queue is the first lock-free bag  with polylogarithmic step complexity.
+
 Both versions of our queue use single-word CAS on reasonably-sized 
@@
 A preliminary version of this work appeared in~\cite{Nad22}.
-\here{Maybe say a little more about techniques used in the implementation, if space permits}
+%\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-10-26 journal/related.tex
minor edits, mostly to Sec 1 and 2

@@
 
-\here{Maybe mention experimental work of \cite{KLM+23} here or in intro}
+%\here{Maybe mention experimental work of \cite{KLM+23} here or in intro}
 
@@
 \paragraph{Other Primitives.}
+Li \cite{Li01} implemented a non-blocking linearizable queue using only the weak primitives
+test\&set, fetch\&add and swap.
+This implementation's amortized step complexity is not bounded
+as a function of the number of processes and the size of the queue.
+(It depends on the total number of enqueues in the execution.)
+
 Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue 
-with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
+with $O(\sqrt{p})$ step complexity using some strong non-standard synchronization primitives
 called half-increment and half-max, which can be viewed as %particular kinds of

### 2023-10-26 journal/algorithm.tex
started editing Sec 3

@@
 See Figure \ref{orderingtree} for an example.
-In this section, we use an infinite array of blocks in each node.
-Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.
+For ease of presentation,
+we use an infinite array of blocks in each node in this section.
+Then, in Section \ref{reducing}, we describe how to replace the infinite array by a representation that uses bounded space.
 
@@
 %\Eric{Reworded following parag after Hossein's comments of Jan 4}
-To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ first observes
+To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
+three steps.
+$P$ first observes
 the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
@@
 new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
-Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
+Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this three-step sequence a
 \op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
@@
 \caption{An example \ordering\ tree with four processes. 
+E(x) denotes an Enqueue(x) operation and D\sub{1} to D\sub{6} denote Dequeue operations.
 We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
 Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
-The fourth process's Deq\sub{6} is still propagating towards the root.
+The fourth process's D\sub{6} is still propagating towards the root.
 The linearization order for this tree is
-Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
+E(a) E(e) D\sub{2} $\mid$ E(b) D\sub{4} D\sub{5} $\mid$ E(d) E(f) E(h) D\sub{1} $\mid$ E(c) D\sub{3} $\mid$ E(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
 \medskip
@@
 Thus, we linearize operations according to the block they belong to in the root's array.
-We can choose how to order operations in the same block, since they must be concurrent.
+We can choose how to order operations within the same block, since they must be concurrent.
 
@@
 E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r) \mbox{ and }\nonumber\\
-D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
+D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r).
+\label{defSeqs}
 \end{eqnarray}
@@
 
-When linearizing the operations propagated to the root, we must
-choose how to order operations within a block.  
-We choose to put
-each block's enqueues before its dequeues.
+When linearizing operations within one of the root's blocks, we choose
+to put the block's enqueues before its dequeues.
 Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
@@
 \begin{equation}
-L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
-\label{linearization}
+L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).\label{linearization}
 \end{equation}
+
+\Eric{I've edited this far}
 

### 2023-10-30 journal/algorithm-GC.tex
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block to $\var{v}.\fld{blocks}$}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to $\var{v}.\fld{blocks}$}
 	\gc{
@@
     }
-    \State $\var{new}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$ \label{invokeCreateBlockGC}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
+    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$ \label{invokeCreateBlockGC}
+    \If{\var{B = \nl}} \Return{\tr} \label{addOPGC}
     \Else  
     	\gc{
-	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
+	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\label{refreshABGC}
 		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
@@
 
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State \hangbox{let \var{new} be a new \typ{Block} with fields\label{initNewBlockGC}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block $B$ to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlockGC}
     \gc{
@@
     }\\
-	\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}\\
-	\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} +
-			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
-    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$} \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
+	\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + 
+			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\\
+	\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} +
+			\var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}
+    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
+    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
         	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
@@
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \hfill \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
-    \Else\ \Return \var{new}
+    \Else\ \Return \var{B}
     \EndIf

### 2023-10-30 journal/algorithm.tex
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

--- 2023-10-25 journal/algorithm.tex
@@
 
-\subsection{Designing a Block Representation to Solve Tasks \ref{construct} to \ref{findenqueue}}
+\subsection{Representation of Blocks}
 \label{sec:fields}
@@
   	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{number of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
+  		\Comment{\# of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
   	\item \typ{int} \fld{super}
@@
 		\item[$\bullet$] \typ{int} \size%
-  		\Comment{size of queue after performing all operations up }\linebreak
-		\mbox{ }\Comment{to the end of this block}
-	\end{itemize}
+  		\Comment{size of queue after performing all operations up to end of this block}	\end{itemize}
 \end{itemize}
@@
 \end{algorithmic}
-\caption{Objects used in the \ordering\ tree data structure.\label{object-fields}\Eric{Fix indentation}}
+\caption{Objects used in the \ordering\ tree data structure.\label{object-fields}}
 \end{figure*}
 
+In this section, we describe how to represent blocks so that
+tasks \ref{construct} to \ref{findenqueue} can be done efficiently.
 Each \node\ of the \ordering\ tree has an infinite array called \fld{blocks}.
@@
 
-\here{If space permits, we might want to add some examples in the following paragraphs that refer back to Figure \ref{implicit}.}
-
 Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child.  
@@
 \begin{eqnarray}\label{defsubblock}
-\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b\!-\!1].\fld{end\sub{left}}\!+\!1..\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
-\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b\!-\!1].\fld{end\sub{right}}\!+\!1..\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\!
+\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{left}}+1..\nonumber\\
+		\var{v}.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
+\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[b-1].\fld{end\sub{right}}+1..\nonumber\\
+		\var{v}.\fld{blocks}[b].\fld{end\sub{right}}].\hspace*{3mm}
 \end{eqnarray}
@@
 For example, consider finding the $r$th enqueue $E_r$ in the linearization.
-A binary search for $r$ on \fld{sum\sub{enq}} fields of the root's blocks 
+A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
 finds the block  containing $E_r$.
@@
 and then do a binary search
-among the direct subblocks of $B$ in that child.
+of the \fld{sum\sub{enq}} fields of the direct subblocks of $B$ in that child.
 Thus, we work our way down the tree until we find the leaf block that  stores 
@@
 A block is called the \emph{superblock} of all of its direct subblocks.
-To facilitate task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
+To facilitate navigating upwards in the ordering tree for task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
 the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
@@
 
-\here{For consistency of notation, perhaps change $n$ to $B$ in code for Enqueue, Dequeue.  Maybe also new to B for CreateBlock, Refresh.}
-
 \begin{figure*}
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block to \var{v}.\fld{blocks}}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to \var{v}.\fld{blocks}}
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
@@
     \EndFor \label{endHelpChild1}
-    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
-    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
+    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
+    \If{$\var{B} = \nl$} \Return{\tr} \label{addOP} 
 	\Else
-	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
+	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}} \label{cas}
     	\State \Call{Advance}{\var{v, h}}\label{advance}
@@
 \setcounter{ALG@line}{39}
-\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \Comment{create new block for \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
-    \State \hangbox{let \var{new} be a new \typ{Block} with fields\label{initNewBlock}
+\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \Comment{create new block $B$ for \op{Refresh} to install in \var{v}.\fld{blocks}[\var{i}]}
+    \State \hangbox{let \var{B} be a new \typ{Block} with fields\label{initNewBlock}
     	\eleft\ \assign\ $\var{v}.\fld{left}.\head - 1$,\label{createEndLeft}
     	\eright\ \assign\ $\var{v}.\fld{right}.\head - 1$,\label{createEndRight}\\
-		\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}\label{createSumEnq},\\
-		\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
-    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
-    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$} \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
+		\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{enq}}\label{createSumEnq},\\
+		\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{deq}} + \var{v}.\fld{right.blocks}[\var{B}.\eright].\fld{sum\sub{deq}}}\label{createSumDeq}
+    \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
+    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
     \EndIf
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \Comment{no blocks need to be propagated to \var{v}} \label{testEmpty} 
-    \Else\ \Return \var{new}
+    \Else\ \Return \var{B}
     \EndIf
@@
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
-    \Else
+    \Else \Comment{First, find the superblock of $v.\fld{blocks}[b]$}
 	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
@@
     \Else \ \Comment{response is the \var{e}th enqueue in the root}
-        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
-			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
+        \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
+			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
 		\State use binary search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
@@
 An \opemph{Enqueue}(\var{e}) appends a \block\ to the process's leaf.
-The block has $\fld{element}=\var{e}$ to indicate it represents an \op{Enqueue}(\var{e}) operation.
+The block's \fld{element} field is \var{e} to indicate it represents an \op{Enqueue}(\var{e}) operation.
 It suffices to propagate the operation to the root and
@@
 A \opemph{Dequeue} also appends a \block\ to the process's leaf.
-The block has $\fld{element}=\nl$ to indicate that it represents a \op{Dequeue} operation.
-After propagating the operation to the root, it computes
-its position in the root using
-\op{IndexDequeue} and then computes its response by calling \op{FindResponse}.
+The block's \fld{element} field is \nl\ to indicate that it represents a \op{Dequeue} operation.
+After propagating the operation to the root, the \op{Dequeue} calls \op{IndexDequeue} to compute
+its position in the root 
+ and then calls \op{FindResponse} to compute its response.
 
@@
 It uses the double \op{Refresh} idea described
-above and invokes two \op{Refresh}es on \var{v} in Lines
+above and calls  \op{Refresh} on \var{v} twice in Lines
 \ref{firstRefresh} and \ref{secondRefresh}. 
-If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
+If both calls fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
 that propagated blocks that were in \var{v}'s children prior to line \ref{firstRefresh} to \var{v}.
@@
 Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
-Line \ref{invokeCreateBlock} creates the new block to install in \var{v}.\fld{blocks}[\var{h}].
+Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
 If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
-Otherwise, the CAS at line \ref{cas} tries to install the new block into \var{v}.\fld{blocks}[\var{h}].
+Otherwise, the CAS at line \ref{cas} tries to install $B$ into \var{v}.\fld{blocks}[\var{h}].
 Either this CAS succeeds or some other process has installed a  block in this location.
@@
 \opemph{CreateBlock}(\var{v, i}) is used
-by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
-%The block \var{new} is created in Line \ref{initNewBlock}. 
+by \op{Refresh} to construct a new block $B$ to be installed in \var{v}.\fld{blocks}[\var{i}].
 The \eleft\ and \eright\ fields store the indices of the last blocks appended to \var{v}'s
@@
 Since the \fld{sum\sub{enq}} field should store the number of enqueues in
-\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright], line \ref{createSumEnq} sets
-\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
+\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
+\fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
 Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
-subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{new}.\fld{sum\sub{enq}}.
-The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
-Then, if \var{new}
+subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
+The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
+Then, if \var{B}
 is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which

### 2023-10-30 journal/analysis.tex
created directory for eventual journal version
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

--- 2023-05-11 podc_paper/analysis.tex
@@
 %
-As shown in \Cref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
+As shown in Lemma \ref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
 Thus, there can be at most $B_{b_e}.\size$ dequeues in 
@@
 $E(B_{b_e})$ would be dequeued before $D(B_b)$. 
-Furthermore, by \Cref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
+Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
 $B_1,\ldots,B_{b-1}$  is 
@@
 $B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
-Each of these $b-1-b_e$ blocks contains at least one operation, by \Cref{blockNotEmpty}.
+Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
 So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
@@
 Let $op$ be the earliest operation by $p$ contained in $B$.
-When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
-Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
-those operations are invoked.
+When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
+Since $B$ is created before
+those operations are invoked, $B$ cannot contain any later operations by $p$.
 
 Let $t$ be the earliest termination of any operation contained in $B$.
-By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
+By Lemma \ref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
 are invoked before $t$.  Thus, all are  running concurrently at $t$, so $B$ contains at most $c$ operations.
@@
 in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
-By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
+By Corollary \ref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
 so $B$ has at most $c$ direct subblocks.
@@
 at line \ref{findAnswer}.
-By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
-since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
+By Lemma \ref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
+since the size of the queue can change by at most $p$ within one block (by Lemma \ref{blockSize}).
 \op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
 Each node $\var{v}$'s search is within the subblocks of one block in $\var{v}$'s parent.
-By \Cref{blockSize}, each such search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
+By Lemma \ref{blockSize}, each such search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
 \end{proof}

### 2023-10-30 journal/boundedspace.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
reformatting GC code
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

@@
 Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
-polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic.  For lack of space, details are in %Appendix~\ref{reducing-details}.
-\cite{full}.
-We replace the \fld{blocks} array in each node by a red-black tree (RBT)
+polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
 that stores the blocks.
@@
 The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
-using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by \Cref{lem::sum}.
+using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by Invariant \ref{inv::sum}.
 

### 2023-10-30 journal/conclusion.tex
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

@@
 This required a substantially different representation of the data stored in the ordering tree.
-\here{?  They also showed that a randomized implementation of the ordering tree technique can
-be used to obtain a (doubly-ended) queue with O(\here{?}) amortized expected step complexity,
-beating the  $\Omega(\min(c,\log\log p))$ lower bound on deterministic algorithms.}
 Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity

### 2023-10-30 journal/correctness-GC.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
getting rid of some weird font macros
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

@@
 After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
-the \op{Refresh} then creates the  block \var{new},
-and then calls \op{AddBlock}($T,new$) to create $T'$.
-Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
-The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
+the \op{Refresh} then creates the  block \var{B},
+and calls \op{AddBlock}($v,T,B$) to create $T'$.
+Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
 So the claim is satisfied.
@@
 
-Since each RBT starts with a single block with \fld{index} 0, the following is an easy consequence of \Cref{RBTupdates}.
+Since each RBT starts with a single block with \fld{index} 0, the following is an easy consequence of Lemma \ref{RBTupdates}.
 \begin{corollary}\label{nonEmptyIncreasing}
@@
 in \var{v}'s tree at some time during the execution.
-It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
+It follows from Lemma \ref{RBTupdates} and Corollary \ref{nonEmptyIncreasing} that 
 each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
@@
 \op{CreateBlock}$(\var{v},b)$ that created $B$ started.
-By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
+By Corollary \ref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(\var{v},b-1)$ 
 stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
@@
 
-\Cref{lem::headProgressGC} implies that 
+Lemma \ref{lem::headProgressGC} implies that 
 the nodes of an in-order traversal of any RBT have non-decreasing values of
@@
 
-\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
+Lemma \ref{lem::subblocksDistinct}, Corollary \ref{cor::noDuplicates}, Invariant \ref{inv::sum} and Corollary \ref{blockNotEmpty} all 
 hold for the modified algorithm.  Their proofs are identical to those given in 
-Section \ref{sec::basicProperties} since they  depend only on \Cref{lem::headProgress} (which can be replaced by \Cref{lem::headProgressGC})
+Section \ref{sec::basicProperties} since they  depend only on Lemma \ref{lem::headProgress} (which can be replaced by Lemma \ref{lem::headProgressGC})
 and the definition of subblocks given in (\ref{defsubblock}).
-In particular, \Cref{lem::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
+In particular, Invariant \ref{inv::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
 values of \fld{sum\sub{enq}}  so the searches of a RBT based on
@@
 
-Next, we prove an analogue of \Cref{successfulRefresh}.
+Next, we prove an analogue of Lemma \ref{successfulRefresh}.
 We say a node \var{v} \emph{contains} an operation
@@
 
-\begin{customlemma}{\ref{successfulRefresh}$'$}
-\label{successfulRefreshGC}
+\begin{customlemma}{\ref{successfulRefresh}$'$}\label{successfulRefreshGC}
 Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
@@
 when $R$ executed line \ref{refreshReadTGC}.
-Let $i$ be the index of the block containing $op$ that was in \var{v}'s RBT before 
+Let $i$ be the index of the block $B_\ell$ containing $op$ that was in \var{v}.\fld{left}'s RBT before 
 $R$ executed line \ref{refreshReadTGC}.
@@
 
-Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$
+Suppose $R$'s call to \op{CreateBlock} returns a new block $B$
 that is installed in $\var{v}.\fld{blocks}$
 by $R$'s \op{CAS} at line \ref{casGC}.
-The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
-By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
-By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
+The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
+By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
+By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
 and therefore \var{v} contains $op$.
@@
 Let $h$ be the maximum \fld{index} in $T$ plus 1.
-By reasoning identical to the last paragraph of \Cref{successfulRefresh}'s proof,
+By reasoning identical to the last paragraph of Lemma \ref{successfulRefresh}'s proof,
 it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
-that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
-$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$
+that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
+$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$
 contain no operations.
-By \Cref{blockNotEmpty}, each block contains at least one operation, so
-these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
-This implies that the block $B$ containing $op$ is a subblock of some block that has appeared
+By Corollary \ref{blockNotEmpty}, each block contains at least one operation, so
+these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft\geq i$.
+This implies that the block $B_\ell$ containing $op$ is a subblock of some block that has appeared
 in \var{v}'s RBT, so $op$ is contained in \var{v}.
@@
 This allows us to show that a double \op{Refresh} propagates operations up the tree, 
-as in \Cref{lem::doubleRefresh}.
+as in Lemma \ref{lem::doubleRefresh}.
 
@@
 If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
-from \Cref{successfulRefreshGC}.
+from Lemma \ref{successfulRefreshGC}.
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
@@
 $R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
-The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
+The claim follows from Lemma \ref{successfulRefreshGC} applied to $R_3$.
 \end{proof}
 
-\Cref{lem::appendExactlyOnce} can then be proved in the same way as in Section \ref{sec::propagating}.
+Lemma \ref{lem::appendExactlyOnce} can then be proved in the same way as in Section \ref{sec::propagating}.
 
@@
 
-\begin{invariant}\label{lem::finished}
+\begin{invariant}\label{inv::finished}
 If the minimum \fld{index} of any block in \var{v}'s RBT is $b_{min}$, then
@@
 If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
-then $T'$ is obtained by adding a new block to $T$, so by \Cref{RBTupdates} the minimum \fld{index}
+then $T'$ is obtained by adding a new block to $T$, so by Lemma \ref{RBTupdates} the minimum \fld{index}
 is unchanged and the invariant is trivially preserved.
@@
 If \op{SplitBlock}($x$) returns the minimum block of $x$'s RBT at line \ref{substitute},
-then the claim follows from the assumption that \Cref{lem::finished} holds at all times before $st$.
+then the claim follows from the assumption that Invariant \ref{inv::finished} holds at all times before $st$.
 
@@
 (and all earlier blocks of $x$) are finished when $st$ occurs.
-By \Cref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
-is also finished when $st$ occurs.  This completes the proof of \Cref{splitfinished}.
+By Observation \ref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
+is also finished when $st$ occurs.  This completes the proof of Claim \ref{splitfinished}.
 \renewcommand{\qedsymbol}{$\diamondsuit$}
@@
 all the blocks older than the one returned by \op{SplitBlock} at line \ref{callSplitBlock}.
-By \Cref{splitfinished}, the minimum block in the new tree will satisfy the invariant when $st$
+By Claim \ref{splitfinished}, the minimum block in the new tree will satisfy the invariant when $st$
 installs the new tree in $\var{v}.\fld{blocks}$.
@@
 
-We remark that \Cref{lem::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
+We remark that Invariant \ref{inv::finished} guarantees that GC keeps one block that is finished and discards all blocks with smaller indices.
 This is because the first unfinished block may still be traversed by an operation in the future,
@@
 \begin{proof}
-Any block that GC removes from an RBT is finished, by \Cref{lem::finished}, so if an 
+Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
 \op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
@@
 by the definition of finished.
-Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.
+Moreover, by Invariant \ref{inv::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.
 
@@
 as well as the enqueue whose value it will return (if it is not a null dequeue).
-If any of those blocks have been removed, it follows from \Cref{lem::finished} that the
+If any of those blocks have been removed, it follows from Invariant \ref{inv::finished} that the
 \op{Dequeue}'s result is written in its \fld{response} field.
@@
 
-By \Cref{blocknotfound}, an operation that fails to find a block in an RBT can terminate.
+By Lemma \ref{blocknotfound}, an operation that fails to find a block in an RBT can terminate.
 If it is a \op{Dequeue}, it can return the result written in its \fld{response} field.
@@
 
-\here{need to prove a lemma analogous to \Cref{blockSize}}
+\here{need to prove a lemma analogous to Lemma \ref{blockSize}}
 
@@
 Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
-Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
+Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.
 %We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
@@
 All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
-There are at most $p$ operations in block $m$ of the root %by \Cref{blockSizeGC}
+There are at most $p$ operations in block $m$ of the root 
 and we showed above that there are at most $2q_{max} + 2p$ in blocks 
@@
 Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
-Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
+Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
 block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
@@
 Thus, there are at most $G$ updates since the last time its maximum \fld{index} was a multiple of $G$.
-The claim follows from \Cref{boundAfterGC}.
+The claim follows from Lemma \ref{boundAfterGC}.
 \end{proof}
@@
 Each process may hold pointers to $O(1)$ RBTs that are no longer current in local variables.
-So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
+So the space bound follows from Corollary \ref{RBTbound} and the fact that $G$ is chosen
 to be $p^2\ceil{\log p}$.
@@
 \here{this proof may need some modification since code was changed to change how split point is computed}
-It follows from \Cref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
+It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
 that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
@@
 some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
-Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
+Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
 after this CAS.
@@
 $\var{v}.\fld{blocks}[(r-1)G+1..rG]$.
-By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
+By Corollary \ref{blockNotEmpty}, there are at least $G$ such operations, 
 so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
@@
 
-\begin{customprop}{\bf{\ref{casbound}}$'$}
+\begin{customprop}{\textbf{\ref{casbound}}$'$}
 Each operation performs $O(\log p)$ \op{CAS} instructions in the worst case.

### 2023-10-30 journal/correctness.tex
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

@@
 
-\begin{observation} \label{lem::headInc}
+\begin{observation} \label{obs::headInc}
 Let $R$ be an instance of \opa{Refresh}{v} whose call to \op{CreateBlock} returns a non-\nl\ block.
@@
 After $R$'s \op{CAS} at line \ref{incrementHead}, \var{v}.\head\ is no longer equal to the value \var{h}
-read at line \ref{readHead}.  The claim follows from \Cref{nonDecreasingHead}.
+read at line \ref{readHead}.  The claim follows from Observation \ref{nonDecreasingHead}.
 \end{proof}
@@
 
-\begin{invariant}\label{lem::headPosition} 
+\begin{invariant}\label{inv::headPosition} 
 For $0 \leq i < \var{v}.\head$, $\var{v}.\fld{blocks}[i]\neq\nl$.  For $i>\var{v}.\head$, $\var{v}.\fld{blocks}[i]=\nl$.
@@
 Since $h$ was read from $\var{v}.\head$ earlier at line \ref{readHead}, the current value of 
-$\var{v}.\head$ is at least $h$ by \Cref{nonDecreasingHead}.
+$\var{v}.\head$ is at least $h$ by Observation \ref{nonDecreasingHead}.
 So, $\var{v}.\head=h$ when the \op{CAS} occurs and a change to $\var{v}.\fld{blocks}[\var{v}.\head]$ preserves the invariant.
@@
 read $b$ from $\var{v}.\head$ at line \ref{readHead}.
-At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by \Cref{lem::headPosition}.
+At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
 Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
 created $B$ started.
-It follows from \Cref{nonDecreasingHead} that the value that 
+It follows from Observation \ref{nonDecreasingHead} that the value that 
 line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
@@
 If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
-If $B$ and $B'$ are in the same node, it follows from \Cref{lem::headProgress} that their direct subblocks are disjoint.
+If $B$ and $B'$ are in the same node, it follows from Lemma \ref{lem::headProgress} that their direct subblocks are disjoint.
 Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
@@
 \here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
-\begin{corollary}\label{lem::noDuplicates}
+\begin{corollary}\label{cor::noDuplicates}
 For  $i\neq j$, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ cannot both contain the same operation.
@@
 cannot be in two different leaf blocks. 
-By \Cref{lem::subblocksDistinct}, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ have no common subblocks, so the claim follows.
+By Lemma \ref{lem::subblocksDistinct}, $\var{v}.\fld{blocks}[i]$ and $\var{v}.\fld{blocks}[j]$ have no common subblocks, so the claim follows.
 \end{proof}
@@
 \begin{invariant}
-\label{lem::sum}
+\label{inv::sum}
 If $B$ is a block stored in $\var{v}.\fld{blocks}[i]$, then
@@
 
-Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and \Cref{lem::headProgress}, the
+Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the
 subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
@@
 
-Invariant \ref{lem::sum} allows us to prove that every block a \op{Refresh} installs contains at least one operation.
+Invariant \ref{inv::sum} allows us to prove that every block a \op{Refresh} installs contains at least one operation.
 
@@
 Thus, at line \ref{testEmpty}, $\var{num\sub{enq}}+\var{num\sub{deq}}\neq 0$.
-By \Cref{lem::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
+By Invariant \ref{inv::sum}, $\var{num\sub{enq}} = |E(B)|$ and $\var{num\sub{deq}} = |D(B)|$,
 so these sequences cannot both be empty.
@@
 when $R$ executed line \ref{readHead}.
-Let $i$ be the index such that the block $B=\var{v}.\fld{left.blocks}[i]$ contains $op$.
-By \Cref{nonDecreasingHead} and \Cref{lem::headProgress}, the value of $childHead$ that $R$ reads from
+Let $i$ be the index such that the block $B_\ell=\var{v}.\fld{left.blocks}[i]$ contains $op$.
+By Observation \ref{nonDecreasingHead} and Lemma \ref{lem::headProgress}, the value of $childHead$ that $R$ reads from
 $\var{v}.\fld{left.head}$ in line \ref{readChildHead} is at least $i$.
@@
 $\var{v}.\fld{left.head} > i$.
-Then, $R$ calls \op{CreateBlock}($\var{v},h$) in line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
+Then, $R$ creates a new block $B$ by calling \op{CreateBlock}($\var{v},h$) at line \ref{invokeCreateBlock}, where $h$ is the value $R$ reads at line \ref{readHead}.
 \op{CreateBlock} reads a value greater than $i$ from $\var{v}.\fld{left.head}$ at line \ref{createEndLeft}.
-Thus, $new.\eleft \geq i$.  We consider two cases.
+Thus, $B.\eleft \geq i$.  We consider two cases.
 
-Suppose $R$'s call to \op{CreateBlock} returns the new block $B'$ and $R$'s \op{CAS} at line \ref{cas} 
-installs $B'$ in $\var{v}.\fld{blocks}$.
-Then, $B$ is a subblock of some block in $\var{v}$, since  $B'.\eleft$ is greater than or equal to $B$'s index
+Suppose $R$'s \op{CAS} at line \ref{cas} 
+installs $B$ in $\var{v}.\fld{blocks}$.
+Then, $B_\ell$ is a subblock of some block in $\var{v}$, since  $B.\eleft$ is greater than or equal to $B_\ell$'s index
 $i$ in $\var{v}.\fld{left.blocks}$.
@@
 \var{num\sub{enq}}
-&=&  \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}} \\
-&& + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}}\\
+&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
+&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
 &&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
-&=& \var{v}.\fld{left.blocks}[new.\eleft].\fld{sum\sub{enq}}\\
-&& + \var{v}.\fld{right.blocks}[new.\eright].\fld{sum\sub{enq}} \\
+&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
+&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
 &&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
@@
 \end{eqnarray*}
-It follows from \Cref{lem::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
-$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
-$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$.
+It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
+$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
+$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
 Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
@@
 these blocks contain no operations.
-By \Cref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft \geq i$.
-Hence, $B$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
+By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
+Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
 \end{proof}
@@
 If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
-from \Cref{successfulRefresh}.
+from Lemma \ref{successfulRefresh}.
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
 Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $\var{v}.\head$ at line \ref{readHead}.
-By \Cref{lem::headInc}, $h_2>h_1$.
-By \Cref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
+By Observation \ref{obs::headInc}, $h_2>h_1$.
+By Lemma \ref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
 Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
 a successful \op{CAS} on $\var{v}.blocks[h_2]$ before $R_2$'s \op{CAS}.
-$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by \Cref{nonDecreasingHead}.
+$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by Observation \ref{nonDecreasingHead}.
 Thus, all operations contained in $\var{v}$'s children when $R_1$ begins
 are also contained in $\var{v}$'s children when $R_3$ later executes line \ref{readHead}.
-By \Cref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
+By Lemma \ref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
 which is before $R_2$'s failed \op{CAS}.
@@
 does a double \op{Refresh}~on each internal node on the path $P$ from the leaf to the root.
-By \Cref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
-There is at most one such block in each node, by \Cref{lem::noDuplicates}.
+By Lemma \ref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
+There is at most one such block in each node, by Corollary \ref{cor::noDuplicates}.
 \end{proof}
@@
 On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
-$B.\fld{super}$ is written after $B$ is installed, and \Cref{successfulRefresh} ensures that $B$ is propagated to the
+$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
 parent soon after.
@@
 By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
-By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
+By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
 that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
@@
 \ref{cas} succeeds.
-So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
-Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.
+So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
+Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.
 
@@
 So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
-By \Cref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
+By Lemma \ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
 $\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
@@
 This read $r'$ must be before $B$ is installed in $\var{v}$;
-otherwise, \Cref{successfulRefresh} would imply that $B$ is a subblock of one of 
+otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
 $\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
@@
 Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
-By \Cref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
+By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
 the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
@@
 and the operation's rank within the block.  
-For \op{IndexDequeue}, we use \Cref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
+For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
 \here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}
 
-\begin{lemma}
-\label{lem::indexDequeue}
+\begin{lemma}\label{lem::indexDequeue}
 If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
@@
 \op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
-By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
+By Lemma \ref{superRelation}, this index is either $B.super$ or $B.super+1$.
 The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.
@@
 
-\begin{lemma}
-\label{lem::get}
+\begin{lemma}\label{lem::get}
 If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
@@
 of $B$, which are listed in (\ref{defsubblock}).
-By \Cref{lem::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
+By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
 of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
@@
 Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
-By \Cref{lem::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
+By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
 enqueues in $E'$ before the enqueues of block $B'$, so
@@
 \begin{proof}
-By \Cref{lem::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
-By \Cref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
+By Corollary \ref{cor::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
+By Lemma \ref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
 so it appears in $L$.
@@
 of block $B$ are performed.
-By \Cref{lem::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
+By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
 are the number of enqueues and dequeues contained in $B$.
@@
 \begin{proof}
-If a dequeue $D$ terminates, it is contained in some block in the root, by \Cref{lem::appendExactlyOnce}.
-By \Cref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
+If a dequeue $D$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
+By Lemma \ref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
 returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
@@
 $D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
-By \Cref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
+By Lemma \ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
 after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
 in the order given by $L$.
-By \Cref{lem::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
+By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
 is the number of enqueues in $B$.
@@
 Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
-By \Cref{lem::sum}, this enqueue is the $i_e$th enqueue 
+By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
 in $E(\var{root}.\fld{blocks}[b_e])$, where
 $b_e$ and $i_e$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
-By \Cref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
+By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
 \end{proof}
 
-Combining \Cref{linearSat} and \Cref{linearCorrect} provides our main result.
+Combining Lemmas \ref{linearSat} and \ref{linearCorrect} provides our main result.
 

### 2023-10-30 journal/correctness.tex
minor edits

@@
 \begin{lemma} \label{lem::subblocksDistinct}
-If $B$ and $B'$ are two blocks in nodes at the same depth, their sets of subblocks are disjoint.
+If $B$ and $B'$ are two blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
 \end{lemma}
@@
 
-\here{Might be able to get rid of this corollary and just cite previous lemma instead to save space}
 \begin{corollary}\label{cor::noDuplicates}
@@
 The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
-on lines \ref{enqNew}, \ref{deqNew}, \ref{createSumEnq} and \ref{createSumDeq} follows easily
+on lines \ref{enqNew}, \ref{deqNew} %, \ref{createSumEnq} 
+and \ref{createSumDeq} follows easily
 from the definition of subblocks.
@@
 $B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
-The proof for line~\ref{deqNew}, where $B$ has a single dequeue, is similar.
+The proof for line~\ref{deqNew}, where $B$ represents a single dequeue, is similar.
 
@@
 \label{sec::propagating}
-
-\here{cref doesn't seem to be working (it calls everything a theorem)}
 
@@
 \label{sec::tracingCorrect}
+
+\er{edited this far}
 

### 2023-10-30 journal/algorithm-GC.tex
editing space bounded version

@@
 
-\section{Detailed Description of the Bounded-Space Implementation}
+\subsection{Detailed Description}
 \label{reducing-details}
@@
 
-Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
-We first describe the modifications to the implementation.
 To avoid confusion, we use nodes to refer to the nodes of the ordering tree, and blocks to refer
 to the nodes of a RBT (since the RBT stores blocks).
-
-The space-bounded implementation uses two additional shared arrays:
+The space-bounded implementation uses two  shared arrays:
 the \var{leaf} array allows processes to access one another's leaves to perform helping, and
 the \var{last} array is used to determine which blocks are safe to discard.
-\begin{itemize}
-\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
-\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the}\\
-\mbox{ }\Comment{root that process $k$ has observed to contain either an en-}\\
-\mbox{ }\Comment{queue whose value has been dequeued or a null dequeue }
-\end{itemize}
 
@@
 \begin{itemize}
-\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
+\item \typ{int} \var{index} 
 \end{itemize}
-To facilitate helping, each \typ{Block} in a leaf has one more additional field that is used only for blocks that store a dequeue operation.
+that represents the position this block would have in the \fld{blocks} array.
+To facilitate helping, each \typ{Block} in a leaf has one more additional field 
 \begin{itemize}
-\item \typ{Object} \var{response} \Comment{response of the dequeue in the block}
+\item \typ{Object} \var{response}
 \end{itemize}
-
-\here{Talk about what happens if a block is not found in an RBT}
+which is used only for blocks that store a dequeue operation and stores the response of the dequeue in the block.
 
@@
 \setcounter{ALG@line}{200}
+\State \linecomment Shared variables
+\State \typ{Node} root \Comment{root of ordering tree}
+\gc{
+\State \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
+\State \typ{int[]} \var{last}$[1..p]$\Comment{\var{last}$[k]$ is max \fld{index} of a root block that process $k$ saw}
+\State \mbox{ }\Comment{contains either a dequeued enqueue or a null dequeue}
+}
+
+\spac
 
 \Function{void}{Enqueue}{\typ{Object} \var{e}}
-    \gc{
-    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
+
+    \gc{\State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
 	\State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \var{e},
-		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}}+1,$\\
-		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}$,
-		\fld{index}\ \assign $h$}\label{enqNewGC}}
+		\fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}}+1,$\\
+		\fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}$,
+		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
     \State \Call{Append}{\var{B}}
@@
 \Function{Object}{Dequeue()}{} 
-    \gc{
-    \State \var{h} \assign\ \Call{MaxBlock}{\var{leaf}.\fld{blocks}}.\fld{index}+1
+    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
     \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
-	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{enq}},$\\
-	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}.\fld{blocks}[h-1].\fld{sum\sub{deq}}+1$,
-	    \fld{index} \assign $h$}\label{deqNewGC}
-	}
+	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
+	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
+	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
     \State \Call{Append}{\var{B}}
-    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}, $h$}}
+    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}
 \EndFunction{Dequeue}
@@
 \gc{
-\Function{Object}{CompleteDeq}{\var{leaf}, $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
+\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$
-    \State \var{res} \assign $\Call{FindResponse}{b, i}$
-	\State \Return{\var{res}}
+    \State \Return $\Call{FindResponse}{b, i}$
 \EndFunction{CompleteDeq}
@@
 \Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}
-    \gc{\State \var{leaf}.\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
-    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
+    \gc{\State \var{leaf}[\var{id}].\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}[\var{id}], \var{leaf}[\var{id}].\fld{blocks}, \var{B}}\label{appendLeafGC}}
+    \State \Call{Propagate}{\var{leaf}[\var{id}].\fld{parent}} 
 \EndFunction{Append}
@@
         \State \Return \Call{Insert}{$T',B$}\label{GCend}
-	\Else\ \Return \Call{Insert}{$T, B$}
+	\Else\ \Return \Call{Insert}{$T, B$}\label{noGC-insert}
     \EndIf
@@
 \end{algorithmic}
-\caption{Bounded-space queue implementation's main routines.\label{code1GC}}
+\caption{Bounded-space queue implementation's main routines for process number $id$.\label{code1GC}}
 \end{figure*}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{247}
+\setcounter{ALG@line}{251}
 \gc{
@@
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \hfill \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
     \Else\ \Return \var{B}
@@
 \begin{algorithmic}[1]
-\setcounter{ALG@line}{290}
+\setcounter{ALG@line}{296}
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
+    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th dequeue of $D(\var{root}.\fld{blocks}[b'])$}
     \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
-    \Else
+    \Else \Comment{First, find the superblock $B_p$ of $v.blocks[b]$}
 	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
@@
 		\State $B_p$ \assign\ min block in $T$ with $\edir \geq b$\label{searchsuper1}
-		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$\label{searchsuper2}
-	    }
+		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$ \Comment{predecessor of $B_p$}\label{searchsuper2}}
 	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
@@
     		\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}
-        \gc{
-        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}%
+        \gc{\If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
         \EndIf
@@
 calls \op{Help} to help all pending dequeues that have been propagated to the root
-(to ensure that all blocks before $s$ can safely be discarded),
+(to ensure that all blocks before $s$ can safely be discarded), and then
 uses the standard RBT \op{Split} routine \cite{Tar83}
-to remove all blocks with \fld{index} less than $s$,
-inserts the new block
-and finally performs a \op{CAS} to swing the node's \fld{blocks} pointer to the new RBT.
+to remove all blocks with \fld{index} less than $s$.
+Then \op{AddBlock}
+inserts the new block at line \ref{GCend} or \ref{noGC-insert}
+and returns the root of the resulting RBT.
+The \op{Append} or \op{Refresh} then stores this root into the node's \fld{blocks} field at line 
+ \ref{appendLeafGC} or \ref{casGC}.
 
@@
 of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
-As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
+While \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
 along the path from \var{v} to the root is discarded by another GC phase.
@@
 by another process's GC phase.
-As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
+As mentioned above, if this happens to an enqueue operation,
 the enqueue can simply terminate because the fact that the block is gone means that another
@@
 Line \ref{substitute} says what to do in this case.
+
+\er{Edited this far}

### 2023-10-30 journal/algorithm.tex
editing space bounded version

@@
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl \Comment{no blocks need to be propagated to \var{v}} \label{testEmpty} 
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
     \Else\ \Return \var{B}
@@
 
-\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \com\ return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$
-    \State \linecomment{is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
+\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
+    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
     \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} dequeues}

### 2023-10-31 journal/correctness-GC.tex
editing Sec 6

@@
 then \var{v} is a leaf node, and no other process ever modifies $\var{v}.\fld{blocks}$.
-$T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
+$T'$ was obtained from $T$ by calling \op{AddBlock}($v,T,B$).
 $B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
 Either way, $B.\fld{index} = \max(I)+1$.
-The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT at line \ref{splitGC} and then add $B$ to it.
 So the claim is satisfied.
@@
 Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
-We also use this notation in the code to indicate that a search of the RBT should be performed for the block
+We also use this notation in the code to indicate that a search of the RBT $v.\fld{blocks}$ should be performed for the block
 with \fld{index} $b$.
@@
 \begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
- If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
+ If $v$ is an internal node and a block with index $h>0$ has been inserted into $\var{v}.\fld{blocks}$ then 
  $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
@@
 at lines \ref{searchsuper3}, \ref{searchsuper1} and \ref{searchsuper2},
-which are used to look for the superblock of a node or its predecessor, can be done in logarithmic time.
+which are used to look for the superblock of a node or its predecessor, can be done 
+using an ordinary BST search.
 
@@
 and the definition of subblocks given in (\ref{defsubblock}).
-In particular, Invariant \ref{inv::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
-values of \fld{sum\sub{enq}}  so the searches of a RBT based on
+In particular, Invariant~\ref{inv::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
+values of \fld{sum\sub{enq}}  so the searches for a block based on
 \fld{sum\sub{enq}} values in lines
-\ref{FRsearchGC} and \ref{getChildGC} can be done in logarithmic time.
+\ref{FRsearchGC} and \ref{getChildGC} can be done using an ordinary BST search.
 
@@
 We say a node \var{v} \emph{contains} an operation
-if some block containing the operation has previously appeared in \var{v}'s RBT (even if 
-the block has been removed from the RBT by a subsequent \op{Split}).
+if some block containing the operation has previously appeared in the RBT \var{v}.\fld{block}
+(even if the block has been removed from the RBT by a subsequent \op{Split} during garbage collection).
 
@@
 By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
-and therefore \var{v} contains $op$.
+and therefore \var{v} contains~$op$.
 
@@
 $R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
-The claim follows from Lemma \ref{successfulRefreshGC} applied to $R_3$.
+The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
 \end{proof}
@@
 
-\here{Technically, following proof might have to be wrapped up in a big induction with the argument
-that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
-is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
-Or alternatively, we could add an assumption that if every operation agrees with the linearization
-so far, then the invariant is preserved by each step.
-Then, when proving the linearization is correct, everything will work out.
-But just leave it as is for the submission, since this is a subtle point}
+%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
+%that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
+%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
+%Or alternatively, we could add an assumption that if every operation agrees with the linearization
+%so far, then the invariant is preserved by each step.
+%Then, when proving the linearization is correct, everything will work out.
+%But just leave it as is for the submission, since this is a subtle point}
+% NO, I DON'T THINK THIS IS NECESSARY
 
@@
 minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
-Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
+Since $T'$ is obtained by discarding all blocks with \fld{index} values less than or equal to $s$ (and adding
 a block with a larger index), it suffices to show that all blocks that were ever added
@@
 
-We must examine how $A$'s call to the recursive algorithm \op{SplitBlock} computes the value of $s$.
+We must examine how $A$'s call at line \ref{callSplitBlock} to the recursive algorithm \op{SplitBlock} computes the value of $s$.
 
 \begin{claim}\label{splitfinished}
-If one of the recursive calls to $\op{SplitBlock}(x)$ within $A$'s call to \op{SplitBlock},
-then that block and all earlier blocks in $x$ are finished when $st$ occurs.
+If one of the recursive calls to $\op{SplitBlock}(x)$ within $A$'s call to \op{SplitBlock} returns a block $B$,
+then $B$ and all earlier blocks in $x$ are finished when $st$ occurs.
 \end{claim}
@@
 Thus, after this helping (and before step $st$),
-$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.
+$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks.
+Since \op{SplitBlock} returns \var{root}.\fld{blocks}[$m-1$], the claim is true.
 
@@
 index $b-1$.
+
+\er{edited this far}
 

### 2023-10-31 journal/correctness-GC.tex
editing Sec 6

@@
 %during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
-then it has been propagated to the root and its result has been written in its \fld{response} field.
+then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
 If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
@@
 
-\here{need to prove a lemma analogous to Lemma \ref{blockSize}}
+Lemma \ref{blockSize} shows that each block contains at most one operation of each process.
+Its proof depends only on Lemma \ref{lem::subblocksDistinct}, Corollary \ref{blockNotEmpty} and Lemma \ref{lem::appendExactlyOnce}, which are all still true for the space-bounded implementation.
+So the same proof of Lemma \ref{blockSize} still applies.
 
@@
 If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
-it contains at most $2q_{max}+4p+1$ blocks.
+it contains at most $3q_{max}+5p+1$ blocks.
 \end{lemma}
@@
 The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
-Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
-must still be in progress at~$C$; otherwise the process that dequeued it would have
-set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
-contradicting the definition of $m$.
+If the value enqueued by any enqueue in $L_2$  is not still in the queue
+at the end of $L_2$, then the dequeue in $L_2$ that dequeued that value must still be in progress
+at $C$; otherwise the process that performed that dequeue would have set its \var{last} entry
+to the index of the root block that contains the enqueue, which is greater than $m$, before $C$, contradicting
+the fact that all values in the $last$ array at $C$ are less than or equal to $m$.
 So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
@@
 at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
-to a value greater than the index of the block of the root that contains the dequeue prior to $C$, 
+to a value greater than the index of the root block that contains the dequeue prior to $C$, 
 contradicting the definition of $m$.
@@
 \eleft\ and \eright\ pointers from the root down to \var{v}.
-So, the block returned is a subblock of block $m$ of the root, unless
+So, the block returned is a subblock of the root block with index $m$, unless
 at some point along the path of subblocks the subblock has already been removed by a split,
-in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.
+in which case \op{SplitBlock} returns a subblock of a root block with index $m'>m$.
 
@@
 $T'\assign\op{Split}(T,s)$.
-Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
+Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$, by Lemma \ref{lem::appendExactlyOnce}.
 Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
 All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
-There are at most $p$ operations in block $m$ of the root 
-and we showed above that there are at most $2q_{max} + 2p$ in blocks 
+By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
+and we showed above that there are at most $3q_{max} + 3p$ in blocks 
 $m+1..\ell$ of the root.
-Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
-Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
+Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
+Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
 block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
@@
 The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
-\here{For later version:  There may be room to improve this bound by a more careful analysis since the bounds so far
-didn't use the fact that operations only appear along one path in the tree; in other words
-doing a bound on total size of all RBTs at one level of the ordering tree at once, rather than
-on each node separately, might reduce it to qlogp + poly(p).  Might also be possible to be
-more precise about how it depends on "current" size of queue rather than max size.}
+%\here{At first, I thought the following analysis could be tightened
+%by using the fact that operations only appear along one path in tree
+%so doing a bound on total size of all nodes on one level might work yield better bound.
+%But I don't think so, because we could have each process go in order to dequeue
+%everything and then enqueue q new items (where q>>p).  Then sets of enqueue operations
+%stored in each node at one level of tree would be disjoint (because of the way
+%GC is timed differently at each node).
 
@@
 
-\here{It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log^2 p(q_{max}+p^3)$.}
-
-Although individual operations may now become more expensive because they have to perform GC,
+Performing GC on a node takes $\Theta(p \log p\log(q_{max}+p))$ steps in the worst case (as explained in the following proof),
+so an individual operation can take up to $\Theta(p \log^2 p\log(q_{max}+p))$ steps
+if it helps at each node along the path from a leaf to the root.
+However,
 we show that operations still have polylogarithmic amortized step complexity.
@@
 \begin{proof}
-\here{this proof may need some modification since code was changed to change how split point is computed}
 It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
@@
 and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
-If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
+If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th GC phase on \var{v}}.
 
-We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once.
-Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
+We argue that each process $P$ can do an $r$th GC phase on \var{v} at most once.
+Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th GC phase on \var{v}.
 Let $\var{v},T,B$ be the arguments of $A$.
@@
 ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
-will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.
+will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th GC phase on \var{v} again.
 
-Each helping phase takes $O(p)$ steps to read the \var{last} array, and
+Each GC phase takes $O(p)$ steps for \op{SplitBlock} to read the \var{last} array and figure
+out where to split the $blocks$ RBT, 
 $O(p \log p \log(p+q_{max}))$ steps in \op{Help},
-and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
+and $O(\log(p+q_{max}))$ steps to split and insert a new node into the RBT.
 Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
@@
 by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
-Since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
+Moreover, since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
 

### 2023-10-31 journal/algorithm.tex
editing Sec 6
minor edits

@@
 B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
-			&& \hspace*{19mm} - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).	
+			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).	
 \end{eqnarray*}
@@
 Finally, if the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
- there is no need to install it.
+ there is no need to install~it.
 

### 2023-11-01 journal/algorithm.tex

@@
 \begin{equation}
-L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
-\label{linearization}
+L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).\label{linearization}
 \end{equation}

### 2023-11-01 journal/boundedspace.tex

@@
 
-\section{Bounding Space Usage}
+\section{Bounded-Space Implementation}
 \label{reducing}
 
-Operations remain in the \fld{blocks} arrays forever. 
-This uses space proportional to the number of enqueues that have been invoked.
+In the implementation described in Section \ref{DescriptQ}, 
+operations remain in the \fld{blocks} arrays forever. 
+Thus, the space used continues to grow as operations are invoked.
 Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
-polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
+polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
 that stores the blocks.
@@
 The RBT reachable from the new copy of the root is the result of applying the RBT operation.
-This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
+This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
 Once a process has performed an update to the RBT representing the blocks of a node 
@@
 
-To prevent RBTs from growing without bound, we would like to discard
+To prevent RBTs from growing without bound, we must discard
 blocks that are no longer needed.
@@
 Blocks should be kept if they contain operations still in progress.
-Moreover, a block containing an \op{Enqueue}($x$) must be kept until $x$ is dequeued.
+Moreover, a block containing an \op{Enqueue}($x$) operation must be kept until $x$ is dequeued.
 
@@
 
-Our GC scheme ensures each RBT has  $O(q+p^2\log p)$ blocks, so RBT operations take $O(\log(p+q))$ time.
-Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
-$O(\log p\log(p + q))$ steps.
-A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
-If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
-Since there are at least $G$ operations between two GC phases, each node contributes $O(\log(p+q))$ steps to each operation in an amortized sense.
-Adding up over all nodes an operation may have to do GC on, 
-an operation spends $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
-So the total amortized step complexity is  $O(\log p \log(p+q))$ per operation.
+%Our GC scheme ensures each RBT has  $O(q+p^2\log p)$ blocks, so RBT operations take $O(\log(p+q))$ time.
+%Excluding GC, an operation does $O(1)$ operations on RBTs at each level of the tree for a total of
+%$O(\log p\log(p + q))$ steps.
+%A GC phase takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
+%If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
+%Since there are at least $G$ operations between two GC phases, each node contributes $O(\log(p+q))$ steps to each operation in an amortized sense.
+%Adding up over all nodes an operation may have to do GC on, 
+%an operation spends $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
+%So the total amortized step complexity is  $O(\log p \log(p+q))$ per operation.
 %Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-11-01 journal/correctness.tex

@@
 
-\er{edited this far}
-
-We first show the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
-This is proved by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from
-\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in the parent node.
-On one hand, $B.super$ is written before $B_s$ is installed:
-\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
-$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
-On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
-$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
-parent soon after.
+In this section, we show that the routines used to compute responses to \op{Dequeue}
+operations work correctly.
+We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
+We do this by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from
+\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in $v$'s parent.
+In particular, $B.super$ is written after $B$ is installed, but before $B_s$ is installed.
+%On one hand, $B.super$ is written before $B_s$ is installed:
+%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
+%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
+%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
+%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
+%parent soon after.
 
@@
 in $\var{v}.\fld{parent.blocks}[s]$.
-By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
+By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
+obtains a value greater than $b$.
 By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
@@
 Next, we show that $B.\fld{super}\geq s-1$.
-The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least 1, so $B.\fld{super} \geq 1$.
+The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
 So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
-By Lemma \ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
+By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
 $\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
@@
 For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
-\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}
+%\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}
 
@@
 
-Next, the position of the required dequeue in $D(B')$ is computed in 
+Next, the position of the required dequeue in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
 lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
-We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line \ref{computeISuperStart}.
+We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
 If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
@@
 Let $B$ be $\var{v}.\fld{blocks}[b]$.
-By (\ref{defSeqs}),
+By Equation (\ref{defSeqs}),
 $E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
@@
 of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
-Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue.
+Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue operation.
 Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
@@
 
-We show that the linearization ordering $L$ defined in 
+We show that the linearization ordering $L$ defined in Equation
 (\ref{linearization}) is a legal permutation of a subset of the operations in 
@@
 if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
-We also show the result each dequeue returns is the same as in the sequential execution  $L$.
+We also show each completed dequeue returns the same result as it would in the sequential execution~$L$.
 
@@
 \begin{proof}
-If a dequeue $D$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
-By Lemma \ref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
-returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
+If a dequeue $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
+By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
+returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th dequeue in the block 
 $B=\var{root}.\fld{blocks}[b]$.
-$D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
-By Lemma \ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
+$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
+By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
 after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
@@
 $\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
-So $D$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
+So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
 execution $L$.
@@
 the number of non-\nl\ dequeues in that prefix of $L$.
-Hence, line \ref{computeE} sets $e$ to the rank of $D$ among all the non-\nl\ dequeues in $L$.
-Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
+Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ dequeues in $L$.
+Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th enqueue in $L$.
 By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
 in $E(\var{root}.\fld{blocks}[b_e])$, where
-$b_e$ and $i_e$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
+$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
 By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
@@
 
-\here{Should there be a lemma somewhere in this section that explicitly shows that each
-object we dereference is non-null?}
+%\here{Should there be a lemma somewhere in this section that explicitly shows that each
+%object we dereference is non-null?}

### 2023-11-01 journal/conclusion.tex
restored details about vector implementation in the concluding remarks

@@
 data structures with a polylogarithmic step complexity.
-For example, we can easily adapt our routines to implement a  vector data structure that stores a sequence and
+For example, we can easily adapt the routines of the implementation in Section \ref{DescriptQ} 
+to implement a  vector data structure that stores a sequence and
 provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
@@
 \opa{Index}{e} to compute the position of element \var{e} in the sequence.
-%\here{Can omit rest of this paragraph to save space}
-%An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
-%A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
-%An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
-\here{doublecheck all preceding claims for accuracy.}
+Only the \op{Append} operations need to be propagated to the root of the ordering tree
+since the other two operations do not affect the state of the object.
+An \opa{Append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
+A \opa{Get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
+An \op{Index} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps if the argument is a pointer to the leaf block that contains the element~$e$.
+
 Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have

### 2023-11-01 journal/related.tex
restored details about vector implementation in the concluding remarks

@@
 Fatourou and Kallimanis \cite{FK14} used their own universal construction based on fetch\&add and LL/SC instructions
-to implement a queue, but its step complexity is also $\Omega(p)$.
-\here{Double check this.}
+to implement a queue, but its step complexity is also $\Omega(p)$ when words are of reasonable size.
 

### 2023-11-01 journal/conclusion.tex
minor edit

@@
 
-We believe the approach used here to implement a lock-free queue 
+The approach used here to implement a lock-free queue 
 could be applied to obtain other lock-free
@@
 For example, we can easily adapt the routines of the implementation in Section \ref{DescriptQ} 
-to implement a  vector data structure that stores a sequence and
+to implement a restricted kind of vector data structure that stores a sequence and
 provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,

### 2024-05-16 journal/response.txt
started addressing reviewers' comments--see responses in response.txt

We thank the reviewers for their comments.  Below, we address them point-by-point.  Modifications to the manuscript are shown in magenta.

> Reviewer 1
> Minor corrections:
> 
> line 080, right column. Emphasize that the number of CAS operations
> (log p) is another measure of the step complexity, different from
> the amortized step complexity.

Done.

> line 465. The notion sum_enq and sum_deq is confusing. Probably use
> count_enq and count_deq instead ? Or cnt_enq and cnt_deq ?

We were concerned that count might sound like a count of the number of enqueue or dequeue operations in a single block.  We instead used sum to remind the reader that the fields of Block i store the prefix sum of the number of enqueue or dequeue operations in in Blocks 1..i (as mentioned when the fields are introduced at line 504 [check line number].

> lines 526 - 528, right column. It is not clear why the expression
> for B.size is correct if some of the dequeues are empty. Additional
> intuition here will be very helpful.

Some explanation has been added.

> line 543, right column. Is B' mentioned here is the same block
> mentioned in line 521, right column ? An additional explanation
> would be helpful.

No, this was a typo.

> lines 552 - 567, right column, and lines 686 - 711, right column.
> Adding explicit references to the lines of the code on Fig. 5 - 7
> would help to follow the explanation.

Done.

> line 558, right column. The sentence "the block's element field is
> $e$ to indicate it represents Enqueue(e) operation..." is
> confusing. It's not immediately clear why element $e$ indicates
> that the operation is Enqueue. Probably formulate it "the block's
> element field $e\neq null$ indicates..."
> 
> lines 594 - 608 (lines 24 - 37 in the code). Add a more detailed
> comments that explain the Refresh operation.
> 
> line 619 (line 44 in the code). It's not clear why the size of the
> queue is calculated correctly. Why if some of the dequeues that are
> counted by num_deq are empty and do not change the size of the
> queue ? Some intuition here would be helpful.
> 
> line 620 (line 46 in the code). Suggestion: consider to put the
> check whether the new block B contains no operations
> (num_deqs+num_enqs==0) immediately after the calculation of num_enq
> and num_deqs. First, complete the calculations that depend on the
> values of num_enq and num_deq, and then advance to the check
> whether v is root. If the change will be done, change the order of
> the explanation in lines 715 - 723, right column, accordingly.
> 
> line 634 (line 58 of the code). In the comment, probably add an
> additional remainder that D(v.blocks[b]) and D(root.blocks[b'])
> represent the sequence of the dequeues represented by the
> corresponding blocks.
> 
> line 640 (line 64 of the code). Add a comment that gives some
> intuition for this line.
> 
> lines 643 - 645 (lines 67 - 69 of the code). Add comments that help
> to understand the calculation of the dequeue's rank.
> 
> line 658 (line 81 in the code). Add additional comment that
> doubling search is used here to find the range of the binary
> search.
> 
> line 696, right column. "... and calls Refresh on $v$ twice in
> lines 226 and 227". The lines numbers should be 17 and 18.
> 
> line 701, right column. "... prior to line 226" should be "...
> prior to line 17"
> 
> line 706, right column. Add a description of the code in lines 26 -
> 29 of Refresh.
> 
> lines 771 - 790, left column. Add explicit references to the lines
> of the code in the description of GetEnqueue().
> 
> line 828, right column. Probably emphasize: if B and B' are two
> *different* blocks at the same depth...
> 
> line 870, right column, Probably should be "By the definition of
> *direct* subblocks (3.3)...".
> 
> lines 887, 892. left column. It look like there are two unnecessary
> empty lines.
> 
> line 919, right column. Should it be "the value computed in line
> *42*" (instead of 41) ?
> 
> line 983, right column. "... close to the time ..." is not clear in
> the context of asynchronous execution. Probably state it
> differently: "We do this by showing that B.super is written after B
> is installed, but before B_s is installed."
> 
> line 1003, right column, Probably should be "By the definition of
> *direct* subblocks (3.3)...".
> 
> line 1102, right column. A more detailed explanation would be
> helpful. Particulary, if op_1 precedes op_2, in which blocks op_1
> and op_2 can appear so that by (3.2) op_1 precedes op_2 in L ?
> 
> line 1255, left column. A short explanation why Corollary 23
> follows immediately from Theorem 22 will be helpful for easier
> reading.
> 
> line 1304, right column. "...will also keep the running time of our
> operations polylogarithmic *on $p$ and $q$*"
> 
> 
> Attachments:
> • https://reviewer-feedback.springernature.com/download/attachment/c514d3ef-d0cc-4718-9d05-877061314935
> 
> Reviewer 2
> A considerable amount of work in the literature of asynchronous concurrent data structures is devoted to the search of efficient wait-free or lock-free linearizable queues. The known queue implementations have unbounded step complexity and their amortized step complexity is at least linear on the number of processes. The current paper makes a significant progress in this matters, it proposes the first wait-free linearizable queue implementation whose worst-case step complexity is polylogarithmic on the total number of processes in the system and the number of elements in the queue. The implementation uses standard read, write and CAS instructions and memory locations of reasonable size.
> 
> The implementation is based in an ordering tree, which is a binary tree with each leaf associated to exactly one process. The main idea of the algorithm is quite simple: when a process wants to perform an operation, it first installs the operation in its leaf and then it propagates the operation in all the nodes in the path from the leaf to the root. During the process, the process might propagate other pending operations. The operation is linearized once it is installed at the root, and the response of the operation, in case it is a dequeue, can be determined by inspecting the previous operations installed in the root. The main idea of the algorithm is simple, what it is difficult is to implement the idea efficiently, i.e. in polylogarithimic number of steps.
> 
> Ordering trees has been used in the past in the universal construction of Afek, Dauber and Touitou, and in the polylogarithmic and linearizable single-enqueuer queue of Jayanti and Petrovic. Recently, the same approach was used by Johnen, Khattabi and Milani in their wait-free queue, but their implementation fails to have a polylogarithmic dequeue operation. 
> 
> After a complete related work section, the paper first shows an implementation that uses unbounded memory. The presentation of this algorithm is very nice, starting with a quite detailed overview of it, which made me follow the detailed implementation that comes after. The implementation is complicated to some extent, due to the arithmetic used to efficiently implement the simple idea the algorithm is based on. Furthermore, the linearizability proof and step complexity analysis are structured in simple pieces that makes the arguments convincing.
> 
> Then, the paper presents a modification of the first algorithm that uses bounded memory, more specifically polynomial on the number of processes and the size of the queue in the linearization, and with polylogarithmic amortized step complexity. As one would naturally expect, this version adds a sort of garbage collector mechanism that removes blocks of memory that are not needed anymore. I found this part of the paper harder to understand. I did not understand the need of using red-black trees, which I think complicates everything. I have no reason to think the implementation is wrong, all looks convincing to me, but I wonder if there is a way to simplify the presentation.
> 
> All in all, the paper is very good, delivering a solid contribution that is overall nicely presented. I recommend acceptance. I have only a few minor comments that appear below. 
> 
> Specific comments:
> 
> - ln 37, left. Missing blank space before ‘and’.
> 
> - I find the related work section complete. However, I wonder if it is worth citing a few works about memory reclamation, due to your bounded memory implementation.
> 
> - I do not understand why you do not cite the preliminary version in podc 2023. 
> 
> - ln 235, left. Repeated ‘words’.
> 
> - For completeness, I suggest to add a preliminaries section, with a reminder of the main definitions such as linearizability, step complexity, and so on. 
> 
> - ln 417, right. Are these direct blocks?
> 
> - ln 492-498. I would recall here what q stands for. Actually, I wonder if you could you use a letter that is more descriptive. 
> 
> - Sec 3.4, and later. At many places operations are written in different font styles. Please follow a uniform format. 
> 
> - ln 845-848, right. I got confused here. Lemma 4 is only about the indexes, isn't it? 
> 
> - ln 1084-1088, right. Here do you mean that L is a valid sequential execution of the queue.
> 
> - Lemma 15. After reading the proof I understood that legal linearization ordering only  means respecting real-time order. This is a concrete example why I think a preliminaries section might help.
> 
> - ln 1192-1993, right. Please formalize the notion of time in an execution,
> 
> - Theorem 22. What exactly is the size of the queue?
> 
> - Sec 6. Why do you need read-back trees? How is that they facilitate the task of keeping memory bounded? Why the first algorithm cannot be modified to remove bocks, let us say using some sort of epoch-based memory reclamation scheme?
> 
> - ln 1279-1283, left. Why? A whole path might need to be copied, isn’t it?
> 
> - ln 1442. Why not Help(), in accordance to Dequeue(), for example?
> 
> - Sec 7. Your algorithms are adaptive on the size of the queue but not on the number of processes. I wonder if an interesting future direction would be to make your algorithms fully adaptive.
>

### 2024-05-16 journal/algorithm.tex
started addressing reviewers' comments--see responses in response.txt

@@
 For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
-the number of enqueues and dequeues in~$B$:
+the number of enqueues and dequeues in~$B$
+\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
+The enqueues of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
+Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements
+(unless the size becomes 0 at some point during these dequeues).  Thus, we have}
 \begin{eqnarray*}
@@
 set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
-We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of $B'$
+We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of \rev{$B$'s superblock}
 by at most 1.
@@
 		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
-    \State \Call{Append}{\var{B}}
+    \State \Call{Append}{\var{B}}\label{enqAppend}
 \EndFunction{Enqueue}
@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}
+    \State \Call{Append}{\var{B}}\label{deqAppend}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$}\label{invokeIndexDequeue}
@@
     \State $\var{leaf.}\head\ \assign\ \var{leaf.}\head+1$ \label{appendEnd} 
-    \State \Call{Propagate}{\var{leaf.}\fld{parent}} 
+    \State \Call{Propagate}{\var{leaf.}\fld{parent}} \label{appendProp}
 \EndFunction{Append}
@@
 
-An \opemph{Enqueue}(\var{e}) appends a \block\ to the process's leaf.
-The block's \fld{element} field is \var{e} to indicate it represents an \op{Enqueue}(\var{e}) operation.
-It suffices to propagate the operation to the root and
-then use its position in the linearization for future \op{Dequeue}
-operations.
+An \opemph{Enqueue}(\var{e}) 
+\rev{creates a new Block at line \ref{enqNew} and sets its fields to indicate
+that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
+field.
+The call to \op{Append}, described below, on line \ref{enqAppend}
+appends the \block\ to the process's leaf and ensures it is propagated to the root.}
 
-A \opemph{Dequeue} also appends a \block\ to the process's leaf.
-The block's \fld{element} field is \nl\ to indicate that it represents a \op{Dequeue} operation.
-After propagating the operation to the root, the \op{Dequeue} calls \op{IndexDequeue} to compute
+A \opemph{Dequeue} also 
+\rev{creates a new Block at line \ref{deqNew} and sets its fields to indicate
+that the Block represents a \op{Dequeue}.
+It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
+it is propagated to the root.
+Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
 its position in the root 
@@
 The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
-so the \op{Append} writes \var{B} there and increments \fld{head}.
+so the \op{Append} writes \var{B} there \rev{at line \ref{appendLeaf}} and then increments \fld{head} \rev{at line \ref{appendEnd}}.
 Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
-\op{Append} then calls \op{Propagate} to ensure the operation represented by \var{B} is propagated to the root.
+\op{Append} then calls \op{Propagate} \rev{at line \ref{appendProp}} to ensure the operation represented by \var{B} is propagated to the root.
 

### 2024-05-16 journal/introduction.tex
started addressing reviewers' comments--see responses in response.txt

@@
 while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
-Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
+Moreover, \rev{if we count only CAS instructions,} each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
 lock-free queues use 
 %an unbounded number of CAS instructions in the worst case and 
-$\Omega(p)$ CAS instructions, even in an amortized sense.
+$\Omega(p)$, even in an amortized sense.
 Since a queue is also a bag, our queue is the first lock-free bag  with polylogarithmic step complexity.

### 2024-05-16 journal/response.txt
handling some more comments from reviewer 1

@@
+*** should add new comments to GC code similar to new comments in basic code
+
 We thank the reviewers for their comments.  Below, we address them point-by-point.  Modifications to the manuscript are shown in magenta.
@@
 > element field $e\neq null$ indicates..."
-> 
+
+This paragraph has been rewritten.
+
 > lines 594 - 608 (lines 24 - 37 in the code). Add a more detailed
 > comments that explain the Refresh operation.
-> 
+
+Done.
+
 > line 619 (line 44 in the code). It's not clear why the size of the
@@
 > queue ? Some intuition here would be helpful.
-> 
+
+Some explanation has been added in the text (line 526) [*** check line #].
+
 > line 620 (line 46 in the code). Suggestion: consider to put the
@@
 > the explanation in lines 715 - 723, right column, accordingly.
-> 
+
+Done.
+
 > line 634 (line 58 of the code). In the comment, probably add an

### 2024-05-16 journal/algorithm.tex
handling some more comments from reviewer 1

@@
 The enqueues of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
-Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements
-(unless the size becomes 0 at some point during these dequeues).  Thus, we have}
+Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
+(If subtracting this many elements would make the new size negative because there are more dequeues than there are elements, then the new size is 0.)  Thus, we have}
 \begin{eqnarray*}
@@
 		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
-    \State \Call{Append}{\var{B}}\label{enqAppend}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{enqAppend}
 \EndFunction{Enqueue}
@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}\label{deqAppend}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$}\label{invokeIndexDequeue}
-    \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{Root}.\fld{blocks}}}\label{invokeIndexDequeue}
+    \State \Return{ \Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \op{Dequeue} should return}}\label{deqRest}
 \EndFunction{Dequeue}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first fails, try again}}
         \State \Call{Refresh}{\var{v}} \label{secondRefresh}
@@
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
-    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \label{startHelpChild1}
+    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{\rev{help advance $v$'s children's \var{head} indices, if necessary}}\label{startHelpChild1}
         \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
@@
     \EndFor \label{endHelpChild1}
-    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
-    \If{$\var{B} = \nl$} \Return{\tr} \label{addOP} 
+    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlock}
+    \If{$\var{B} = \nl$} \Return{\tr}\Comment{\rev{if no new operations to propagate to $v$, do nothing}}\label{addOP} 
 	\Else
-	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}} \label{cas}
-    	\State \Call{Advance}{\var{v, h}}\label{advance}
-    	\State \Return{ \var{result}}
+	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{\rev{try to add $B$ to \var{v}.\fld{blocks}}}\label{cas}
+    	\State \Call{Advance}{\var{v, h}} \Comment{\rev{advance $v.head$ from $h$ to $h+1$}}\label{advance}
+    	\State \Return{ \var{result}}\Comment{\rev{return true iff \op{CAS} succeeded}}
 	\EndIf
@@
     \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
+    \ElsIf{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLength}
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
-    \Else\ \Return \var{B}
-    \EndIf
+    \State \Return \var{B}
 \EndFunction{CreateBlock}
@@
 An \opemph{Enqueue}(\var{e}) 
-\rev{creates a new Block at line \ref{enqNew} and sets its fields to indicate
+\rev{creates a new Block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
 that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
@@
 A \opemph{Dequeue} also 
-\rev{creates a new Block at line \ref{deqNew} and sets its fields to indicate
+\rev{creates a new Block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
 that the Block represents a \op{Dequeue}.
@@
 The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
-Then, if \var{B}
+\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
+ there is no need to install~it.
+Otherwise,} if \var{B}
 is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
 represents the number of elements in the queue after the operations in the block are performed.
-Finally, if the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
- there is no need to install~it.
 

### 2024-05-29 journal/response.txt
working on comments from reviewer 1

@@
 
-No, this was a typo.
+No, this was a typo, now corrected.
 
@@
 > corresponding blocks.
-> 
+
+Added reminder in caption of Figure.
+
 > line 640 (line 64 of the code). Add a comment that gives some
 > intuition for this line.
-> 
+
+Done.
+
 > lines 643 - 645 (lines 67 - 69 of the code). Add comments that help
 > to understand the calculation of the dequeue's rank.
-> 
+
+Done.
+
 > line 658 (line 81 in the code). Add additional comment that
@@
 > search.
-> 
+
+Done.  Some text has been added to the text description of FindResponse to explain that the doubling search first finds the left end of the range and then performs a binary search.
+
 > line 696, right column. "... and calls Refresh on $v$ twice in
@@
 > prior to line 17"
-> 
+
+These have been fixed.
+
 > line 706, right column. Add a description of the code in lines 26 -
 > 29 of Refresh.
-> 
+
+Done.
+
 > lines 771 - 790, left column. Add explicit references to the lines
 > of the code in the description of GetEnqueue().
-> 
+
+This has been done, and the description has been expanded.
+
 > line 828, right column. Probably emphasize: if B and B' are two
 > *different* blocks at the same depth...
-> 
+
+Done.
+
 > line 870, right column, Probably should be "By the definition of
 > *direct* subblocks (3.3)...".
-> 
+
+Yes, clarified.
+
 > lines 887, 892. left column. It look like there are two unnecessary
 > empty lines.
-> 
+
+This seems to be a strange effect of the double-spacing option for the draft.
+
 > line 919, right column. Should it be "the value computed in line
 > *42*" (instead of 41) ?
-> 
+
+Fixed.
+
 > line 983, right column. "... close to the time ..." is not clear in
@@
 > is installed, but before B_s is installed."
-> 
+
+This has been restated.
+
 > line 1003, right column, Probably should be "By the definition of
 > *direct* subblocks (3.3)...".
-> 
+
+Fixed.
+
 > line 1102, right column. A more detailed explanation would be
@@
 > and op_2 can appear so that by (3.2) op_1 precedes op_2 in L ?
-> 
+
+This has been expanded.
+
 > line 1255, left column. A short explanation why Corollary 23

### 2024-05-29 journal/algorithm.tex
working on comments from reviewer 1

@@
 So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
-$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
+$op_1$ will be in an earlier block than $op_2$ in the root's \fld{blocks} array.
 Thus, we linearize operations according to the block they belong to in the root's array.
@@
 \end{itemize}
-
-%\Statex {\com\  Blocks in internal nodes have the following additional fields}
-%\Statex $\blacktriangleright$ \typ{InternalBlock} extends \typ{Block} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
-%\begin{itemize}
-%	\item \typ{int} \eleft, \eright
-%  		\Comment{index of last direct subblock in the left and right child}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in leaf nodes have the following additional field}
-%\Statex $\blacktriangleright$ \typ{LeafBlock} extends \typ{Block} \sf{\com\ the following additional field is used only for blocks in leaves}
-%\begin{itemize}
-%  \item \typ{Object} \var{element}
-%  \Comment{if the block's operation is enqueue(x) then element=x, otherwise element=null.}
-%\end{itemize}
-
-%\Statex {\com\ Blocks in the root node have the following additional field}
-%\Statex $\blacktriangleright$ \typ{RootBlock} extends \typ{InternalBlock} \sf{\com\ the following additional field is used only for blocks in the root}
-%\begin{itemize}
-%  \item \typ{int} \size%
-%  \Comment{size of the queue after performing all operations up to the end of this block}
-%\end{itemize}
-
 \end{algorithmic}
@@
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
-    \Else \Comment{First, find the superblock of $v.\fld{blocks}[b]$}
+    \Else \Comment{first, find the superblock of $v.\fld{blocks}[b]$}
 	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\label{idsup1}
+    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{\rev{actual index of superblock is either \var{sup} or $\var{sup}+1$}}\label{idsup1}
 	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
@@
 	    \State \linecomment{compute index \var{i} of dequeue in superblock}
+	    \State \linecomment{\rev{add \# dequeues in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
 	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
-        \If{$\fld{dir} = \fld{right}$} 
+        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# dequeues from left sibling}}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
@@
 			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
-		\State use binary search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
+		\State use \rev{doubling} search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
 		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
@@
 \end{algorithmic}
-\caption{Queue implementation's helper routines. \label{code2}}
+\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of dequeues represented by block $B$.}\label{code2}}
 \end{figure*}
@@
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ from $v.\fld{left}$}}
         \EndIf \label{endChooseDir}
@@
 An \opemph{Enqueue}(\var{e}) 
-\rev{creates a new Block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
-that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
+\rev{creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
+that the block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
 field.
@@
 A \opemph{Dequeue} also 
-\rev{creates a new Block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
-that the Block represents a \op{Dequeue}.
+\rev{creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
+that the block represents a \op{Dequeue}.
 It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
@@
 Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
+\rev{Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
+for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.}
 Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
@@
 The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
-A \op{Refresh} may pause after a successful CAS before calling \op{Advance} at line \ref{advance},
+A \op{Refresh} may pause between a successful CAS at line \ref{cas} and calling \op{Advance} at line \ref{advance},
 so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
@@
 \op{Enqueue} whose argument is the \op{Dequeue}'s response. 
-A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
+\rev{The routine next searches for}
+the index $b_e$ of the block in \var{root}.\fld{blocks} that contains 
 the \var{e}th enqueue.
 Since the enqueue is linearized before the dequeue, $b_e\leq b$.
-To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
-Then, \op{GetEnqueue} traces down through the tree to find the required enqueue in a leaf.
+To find the left end of the range \rev{to search for $b_e$, \op{FindResponse} does} a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
+\rev{Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.}
+Finally, the call to \op{GetEnqueue} \rev{at line \ref{findAnswer}} traces down through the tree to find the required enqueue in a leaf.
 
@@
 It recursively finds the location of the enqueue in each node along the path from $v$ to a leaf, which stores the argument explicitly.
-\op{GetEnqueue} first determines which child of \var{v} contains the enqueue, and
-then finds the range of blocks within that child that are subblocks of $B$ using information stored
-in $B$ and the block that precedes $B$ in $v$.
-\op{GetEnqueue} finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
-field (line~\ref{getChild}) and proceeds recursively down the tree.
+\rev{The test at line \ref{leftOrRight} determines which child of \var{v} contains
+the enqueue by checking whether $i$ is less than or equal to the number of  
+enqueues in $B$ that came from \var{v}'s left child.
+Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$} 
+using information stored
+in $B$ and the block that precedes $B$ in $v$,
+\rev{and then}
+finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
+field \rev{of the child's \var{blocks} array.
+Line \ref{getChildIndex} computes the index of the enqueue within the block
+in the child, and line \ref{getRecurse}
+proceeds recursively down the tree.}

### 2024-05-29 journal/correctness.tex
working on comments from reviewer 1

@@
 \begin{lemma} \label{lem::subblocksDistinct}
-If $B$ and $B'$ are two blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
+If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
 \end{lemma}
@@
 
-Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the
+Now suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
 subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
@@
 We formalize this intuition.
-The value computed at line \ref{createSumEnq} is
+The value computed at line \ref{computeNumEnq} is
 \begin{eqnarray*}
@@
 We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
-We do this by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from
-\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in $v$'s parent.
-In particular, $B.super$ is written after $B$ is installed, but before $B_s$ is installed.
+We do this by showing that \rev{the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
+\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.}
 %On one hand, $B.super$ is written before $B_s$ is installed:
@@
 in $\var{v}.\fld{parent.blocks}[s]$.
-By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
+By the definition of \rev{direct} subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
 obtains a value greater than $b$.
@@
 So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
-By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
+By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ \rev{is assigned a non-\nl\ value before $B$'s superblock is installed}.  Let $R_{s-1}$ be the call to
 $\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
@@
 so it appears in $L$.
-Also, if $op_{1}$ terminates before $op_{2}$ begins, then $op_{1}$ 
-is propagated to the root before $op_2$ begins, so $op_1$ appears before $op_2$ in $L$.
+\rev{Suppose an operation} $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
+is propagated to the root before $op_2$ begins, \rev{and therefore before $op_2$ is propagated to the root.
+Thus, $op_1$ appears in an earlier block of \var{root}.\fld{blocks} than $op_2$.
+So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.}
 \end{proof}

### 2024-05-29 journal/response.txt
finished handling comments by reviewer 1

@@
 > reading.
-> 
+
+We found it easier to combine Theorem 22 and Corollary 23 and their
+proofs (now stated together as Theorem 22).
+
 > line 1304, right column. "...will also keep the running time of our
 > operations polylogarithmic *on $p$ and $q$*"
-> 
-> 
+
+Done.
+
 > Attachments:

### 2024-05-29 journal/analysis.tex
finished handling comments by reviewer 1

@@
 When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
-Since $B$ is created before
-those operations are invoked, $B$ cannot contain any later operations by $p$.
+$B$ cannot contain any later operations by $p$, since $B$ is created before
+those operations are invoked.
 
@@
 \begin{mytheorem}\label{enqDeqTime}
-Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
+\rev{The queue implementation is wait-free and}
+each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
 and each non-null \op{Dequeue} takes
@@
 at line \ref{findAnswer}.
-By Lemma \ref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
+\rev{Since the doubling search must search a finite portion of the root's
+\var{blocks} array, it terminates within a finite number of steps.}
+Moreover, by Lemma \ref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
 since the size of the queue can change by at most $p$ within one block (by Lemma \ref{blockSize}).
@@
 \end{proof}
-
-\begin{corollary}
-The queue implementation is wait-free.
-\end{corollary}

### 2024-05-29 journal/boundedspace.tex
finished handling comments by reviewer 1

@@
 Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
-polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
+polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic \rev{in $p$ and $q$}. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
 that stores the blocks.
@@
 Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
-also keep the running time of our operations polylogarithmic.
+also keep the running time of our operations polylogarithmic \rev{in $p$ and $q$}.
 Blocks should be kept if they contain operations still in progress.

### 2024-05-30 journal/algorithm-GC.tex
handling some more comments from reviewer 1
working on comments from reviewer 1
added more comments to pseudocode (esp. GC version)

@@
 \State \linecomment Shared variables
-\State \typ{Node} root \Comment{root of ordering tree}
+\State \typ{Node} root \Comment{root of ordering tree}%
 \gc{
-\State \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
+\State \typ{Node[]} \var{leaf}$[1..p]$\Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
 \State \typ{int[]} \var{last}$[1..p]$\Comment{\var{last}$[k]$ is max \fld{index} of a root block that process $k$ saw}
-\State \mbox{ }\Comment{contains either a dequeued enqueue or a null dequeue}
+\State \mbox{ }\Comment{contains either a dequeued enqueue or a null dequeue}%
 }
@@
 		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
-    \State \Call{Append}{\var{B}}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}
 \EndFunction{Enqueue}
@@
 	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
-    \State \Call{Append}{\var{B}}
-    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}%
+    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{\rev{compute response \op{Dequeue} should return}}
 \EndFunction{Dequeue}
@@
 \Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$
-    \State \Return $\Call{FindResponse}{b, i}$
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{\rev{find where \op{Dequeue} is in \var{root}.\fld{blocks}}}
+    \State \Return $\Call{FindResponse}{b, i}$\Comment{\rev{find response \op{Dequeue} should return}}
 \EndFunction{CompleteDeq}
@@
 
-\Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}
+\Function{void}{Append}{\typ{Block} \var{B}}  \Comment{append block to leaf and propagate to root}%
     \gc{\State \var{leaf}[\var{id}].\fld{blocks} \assign\ \Call{AddBlock}{\var{leaf}[\var{id}], \var{leaf}[\var{id}].\fld{blocks}, \var{B}}\label{appendLeafGC}}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh}
-        \State \Call{Refresh}{\var{v}} \label{secondRefresh}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \Comment{double refresh: \rev{if first fails, try again}}
+        \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
     \EndIf
@@
 
-\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to $\var{v}.\fld{blocks}$}
+\Function{boolean}{Refresh}{\typ{Node} \var{v}} \Comment{try to append a new block $B$ to $\var{v}.\fld{blocks}$}%
 	\gc{
@@
     }
-    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$ \label{invokeCreateBlockGC}
-    \If{\var{B = \nl}} \Return{\tr} \label{addOPGC}
+    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlockGC}
+    \If{\var{B = \nl}} \Return{\tr}\Comment{\rev{if no new operations to propagate to \var{v}, do nothing}}\label{addOPGC}
     \Else  
     	\gc{
-	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\label{refreshABGC}
-		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
+	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\Comment{\rev{$T'$ is a copy of $T$ with $B$ added (and GC if needed)}}\label{refreshABGC}
+		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\Comment{\rev{try installing $T'$ as \var{v}'s \fld{blocks} tree}}\label{casGC}%
 		}
@@
 \gc{
-\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
+\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} \Comment{do GC if necessary; add block $\var{B}\neq\nl$ to \var{T}}
     \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
-        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\label{callSplitBlock}
-        \State \Call{Help}{}\label{invokeHelp}
+        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\Comment{\rev{find lowest index of block to keep}}\label{callSplitBlock}
+        \State \Call{Help}{}\Comment{\rev{help complete pending \op{Dequeues}}}\label{invokeHelp}
         \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
-        \State \Return \Call{Insert}{$T',B$}\label{GCend}
-	\Else\ \Return \Call{Insert}{$T, B$}\label{noGC-insert}
+        \State \Return \Call{Insert}{$T',B$}\Comment{\rev{add $B$ to the tree}}\label{GCend}
+	\Else\ \Return \Call{Insert}{$T, B$}\Comment{\rev{if GC not needed, just add $B$ to the tree}}\label{noGC-insert}
     \EndIf
@@
 \gc{
-\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{figure out where to split \var{v}'s RBT}
-	\If{$\var{v}=\var{root}$}
-		\State $m$ \assign\ 0
+\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{\rev{find lowest index of block GC should keep in \var{v}}}
+	\If{$\var{v}=\var{root}$} \Comment{return block before the last block containing a}
+		\State $m$ \assign\ 0\Comment{dequeued enqueue or null dequeue}
 		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
@@
 		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
-	\Else
+	\Else \Comment{\rev{return last sublock of first block kept in parent}}
 		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
@@
 	\EndIf
-	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, use leftmost block instead}\label{substitute}
+	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$) \Comment{If $B$ was discarded, keep all blocks}\label{substitute}
 \EndFunction{SplitBlock}
@@
 \gc{
-\Function{void}{Help}{} \Comment{help pending operations}
+\Function{void}{Help}{} \Comment{help pending \op{Dequeues}}
 
@@
             \State \linecomment{operation is a propagated dequeue}
-            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
+            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{\rev{compute response and store it in leaf block}}
         \EndIf
@@
 		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa
-		\Else
+		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa\ \Comment{\rev{\var{v}.\fld{blocks}[\var{b}] has not propagated to \var{v}.\fld{parent}}}
+		\Else \Comment{\rev{check if superblock of \var{v}.\fld{blocks}[\var{b}] propagated to root}}
 			\State $B_p$ \assign\ minimum index block in $T$ with $\edir \geq b$\label{searchsuper3}
@@
     \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
-    \If{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
+    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
+    \ElsIf{$\var{v} = \var{root}$} \var{B}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
         	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
     \EndIf
-    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmptyGC}\label{CBnullGC}
-    \Else\ \Return \var{B}
-    \EndIf
+    \State\Return \var{B}
 \EndFunction{CreateBlock}
@@
 	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
+	    \State \linecomment{\rev{add \# dequeues in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
 	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
-        \If{$\fld{dir} = \fld{right}$} 
+        \If{$\fld{dir} = \fld{right}$}\Comment{\rev{add \# dequeues in $B_p$ from left sibling}}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
@@
         \EndIf \label{computeISuperEndGC}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}} \Comment{\rev{recurse up the tree}}
     \EndIf
@@
 		\State use \gc{BST search in \var{root}.\fld{blocks} to find minimum index $b_e$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{e}$}\label{FRsearchGC} 
-		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of enqueue within its block}
+		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of enqueue within its block}%
 		\gc{
-        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
+        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that enqueue}}\label{findAnswerGC}
         \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
@@
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
         \EndIf
@@
 The \op{Append} or \op{Refresh} then stores this root into the node's \fld{blocks} field at line 
- \ref{appendLeafGC} or \ref{casGC}.
+ \ref{appendLeafGC} or~\ref{casGC}.
 
@@
 Line \ref{substitute} says what to do in this case.
-
-\er{Edited this far}

### 2024-05-30 journal/algorithm.tex
added more comments to pseudocode (esp. GC version)

@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{Root}.\fld{blocks}}}\label{invokeIndexDequeue}
-    \State \Return{ \Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \op{Dequeue} should return}}\label{deqRest}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root}.\fld{blocks}}}\label{invokeIndexDequeue}
+    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \op{Dequeue} should return}}\label{deqRest}
 \EndFunction{Dequeue}
@@
 	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
-        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# dequeues from left sibling}}
+        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# dequeues in superblock from left sibling}}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
@@
         \EndIf \label{computeISuperEnd}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}} \Comment{\rev{recurse up the tree}}
     \EndIf
@@
 		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
-        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
+        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that enqueue}}\label{findAnswer}
     \EndIf
@@
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ from $v.\fld{left}$}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
         \EndIf \label{endChooseDir}
@@
 \end{algorithmic}
-\caption{Queue implementation's \op{GetEnqueue} routine.\label{code3}}
+\caption{Queue implementation's \op{GetEnqueue} routine.  \rev{Recall that $E(B)$ is the sequence of enqueues represented by block $B$}.\label{code3}}
 \end{figure*}

### 2024-05-30 journal/response.txt
added more comments to pseudocode (esp. GC version)
working on reviewer 2's comments

@@
-*** should add new comments to GC code similar to new comments in basic code
-
 We thank the reviewers for their comments.  Below, we address them point-by-point.  Modifications to the manuscript are shown in magenta.
@@
 
-Done.
+Done.  Quite a few new comments have been added throughout the pseudocode.
 
@@
 
-> Attachments:
-> • https://reviewer-feedback.springernature.com/download/attachment/c514d3ef-d0cc-4718-9d05-877061314935
-> 
 > Reviewer 2
-> A considerable amount of work in the literature of asynchronous concurrent data structures is devoted to the search of efficient wait-free or lock-free linearizable queues.
-The known queue implementations have unbounded step complexity and their amortized step complexity is at least linear on the number of processes.
-The current paper makes a significant progress in this matters, it proposes the first wait-free linearizable queue implementation whose worst-case step complexity is polylogarithmic on the total number of processes in the system and the number of elements in the queue.
-The implementation uses standard read, write and CAS instructions and memory locations of reasonable size.
-> 
-> The implementation is based in an ordering tree, which is a binary tree with each leaf associated to exactly one process.
-The main idea of the algorithm is quite simple: when a process wants to perform an operation, it first installs the operation in its leaf and then it propagates the operation in all the nodes in the path from the leaf to the root.
-During the process, the process might propagate other pending operations.
-The operation is linearized once it is installed at the root, and the response of the operation, in case it is a dequeue, can be determined by inspecting the previous operations installed in the root.
-The main idea of the algorithm is simple, what it is difficult is to implement the idea efficiently, i.e.
-in polylogarithimic number of steps.
-> 
-> Ordering trees has been used in the past in the universal construction of Afek, Dauber and Touitou, and in the polylogarithmic and linearizable single-enqueuer queue of Jayanti and Petrovic.
-Recently, the same approach was used by Johnen, Khattabi and Milani in their wait-free queue, but their implementation fails to have a polylogarithmic dequeue operation.
-> 
-> After a complete related work section, the paper first shows an implementation that uses unbounded memory.
-The presentation of this algorithm is very nice, starting with a quite detailed overview of it, which made me follow the detailed implementation that comes after.
-The implementation is complicated to some extent, due to the arithmetic used to efficiently implement the simple idea the algorithm is based on.
-Furthermore, the linearizability proof and step complexity analysis are structured in simple pieces that makes the arguments convincing.
 >
@@
 I have no reason to think the implementation is wrong, all looks convincing to me, but I wonder if there is a way to simplify the presentation.
-> 
-> All in all, the paper is very good, delivering a solid contribution that is overall nicely presented. I recommend acceptance. I have only a few minor comments that appear below. 
-> 
+
+For the space-bounded version, we replace an infinite array by a red-black tree (using the array index as the key).
+This allows us to append elements, access elements, and split off the obsolete elements, all in logarithmic time.
+We realize that this additional data structure complicates the algorithm, but we were unfortunately unable to come up with a simpler way to do it.
+We hoped that presenting the simpler, space-unbounded version first would make it easier to understand the space-bounded version.
+We did our best to explain the transformation from the arrays to the RBTs in the first page of Section 6.
+
 > Specific comments:
@@
 > - ln 37, left. Missing blank space before ‘and’.
-> 
+
+Fixed.
+
 > - I find the related work section complete. However, I wonder if it is worth citing a few works about memory reclamation, due to your bounded memory implementation.
-> 
+
+Although our space-bounded algorithm unlinks objects that are no longer needed from our data structure, we consider the problem of reclaiming that memory to be an orthogonal issue.
+There is a rather large body of literature on memory reclamation for concurrent data structures, so we have added (at line 110 of the introduction) a reference to a very recent paper that includes a large bibliography of work in this area, to help the reader find relevant literature on it.
+
 > - I do not understand why you do not cite the preliminary version in podc 2023.
-> 
+
+This omission has been corrected.
+
 > - ln 235, left. Repeated ‘words’.

### 2024-05-30 journal/introduction.tex
fix typo: add misssing space
working on reviewer 2's comments

@@
 %packet processing \cite{DPDK}, 
-synchronization \cite{KAE23},%\here{is this a good citation for this?}
+synchronization \cite{KAE23}, %\here{is this a good citation for this?}
  and sharing resources or tasks.
@@
 
-Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert~\cite{Rup16}.
+Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed in~\cite{Rup16}.
 For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
@@
 garbage collector, such as the highly optimized one that Java provides.
+\rev{See \cite{SBM24} and the references therein for a variety of approaches to
+safe memory reclamation.}
 
@@
 and quickly access any  operation in a~batch.
-A preliminary version of this work appeared in~\cite{Nad22}.
+Preliminary versions of this work appeared in~\cite{Nad22,NR23}.
 %\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2024-05-31 journal/response.txt
working on reviewer 2's comments

@@
 > - ln 235, left. Repeated ‘words’.
-> 
+
+Fixed.
+
 > - For completeness, I suggest to add a preliminaries section, with a reminder of the main definitions such as linearizability, step complexity, and so on.
-> 
+
+We tried to avoid the need for too many definitions and preferred to introduce them as needed instead of in a preliminaries section, since we are using fairly standard terminology.
+The terms mentioned above were explained in the first paragraph of the paper.
+A more formal explanation of what must be proved to establish linearizability is given in the first paragraph of Section 4.4.
+Similarly, wait-free is defined when it is first used.
+For completeness, we added a definition of CAS in a footnote on page 1.
+
 > - ln 417, right. Are these direct blocks?
-> 
+
+These are the blocks actually stored in root.blocks.  (The adjective direct is only applied to subblocks.)
+
 > - ln 492-498. I would recall here what q stands for. Actually, I wonder if you could you use a letter that is more descriptive.
-> 
+
+We added the reminder.  Since q stands for queue, we thought q would remind readers of the queue's size.
+
 > - Sec 3.4, and later. At many places operations are written in different font styles. Please follow a uniform format.
-> 
+
+Done.
+
 > - ln 845-848, right. I got confused here. Lemma 4 is only about the indexes, isn't it? 

### 2024-05-31 journal/algorithm-GC.tex
working on reviewer 2's comments

@@
 \end{itemize}
-which is used only for blocks that store a dequeue operation and stores the response of the dequeue in the block.
+which is used only for blocks that store a \dequeue\ operation and stores the response of the \dequeue\ in the block.
 
@@
 that field has been removed.
-The \op{IndexDequeue} routine, which must trace the location of a dequeue along
+The \op{IndexDequeue} routine, which must trace the location of a \dequeue\ along
 a path from its leaf to the root has a minor modification to search the \fld{blocks} RBT at 
@@
 \State \typ{int[]} \var{last}$[1..p]$\Comment{\var{last}$[k]$ is max \fld{index} of a root block that process $k$ saw}
-\State \mbox{ }\Comment{contains either a dequeued enqueue or a null dequeue}%
+\State \mbox{ }\Comment{contains either a dequeued \enqueue\ or a null \dequeue}%
 }
@@
 \gc{
-\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$}
+\Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated \dequeue\ in $\var{leaf}.\fld{blocks}[h]$}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{\rev{find where \op{Dequeue} is in \var{root}.\fld{blocks}}}
@@
 	\If{$\var{v}=\var{root}$} \Comment{return block before the last block containing a}
-		\State $m$ \assign\ 0\Comment{dequeued enqueue or null dequeue}
+		\State $m$ \assign\ 0\Comment{dequeued \enqueue\ or null \dequeue}
 		\For{$k$ \assign\ $1..p$} $m$ \assign\ $\max(m, \var{v}.\fld{last}[k])$
@@
         \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
-            \State \linecomment{operation is a propagated dequeue}
+            \State \linecomment{operation is a propagated \dequeue}
             \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{\rev{compute response and store it in leaf block}}
@@
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
-    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th dequeue of $D(\var{root}.\fld{blocks}[b'])$}
+    \State \linecomment{return $\langle b', i' \rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is ($i'$)th \dequeue\ of $D(\var{root}.\fld{blocks}[b'])$}
     \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
@@
 		\State $B_p'$ \assign\ max block in $T$ with $\edir < b$ \Comment{predecessor of $B_p$}\label{searchsuper2}}
-	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
-	    \State \linecomment{\rev{add \# dequeues in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}}
+	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock $B_p$}
+	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
 	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
-        \If{$\fld{dir} = \fld{right}$}\Comment{\rev{add \# dequeues in $B_p$ from left sibling}}
+        \If{$\fld{dir} = \fld{right}$}\Comment{\rev{add \# \dequeues\ in $B_p$ from left sibling}}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
@@
 
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th \dequeue\ in $D(\var{root}.\fld{blocks}[\var{b}])$}
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
@@
     		\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
-    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when dequeue occurs}%
+    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \Comment{queue is empty when \dequeue\ occurs}%
         \gc{\If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]$\ \assign\ $b$
@@
         \State \Return \nl \hfill \label{returnNullGC}
-    \Else \ \Comment{response is the \var{e}th enqueue in the root}
+    \Else \ \Comment{response is the \var{e}th \enqueue\ in the root}
         \State $\var{e}$\ \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
@@
 		\State use \gc{BST search in \var{root}.\fld{blocks} to find minimum index $b_e$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{e}$}\label{FRsearchGC} 
-		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of enqueue within its block}%
+		\State $i_e$ \assign\ $\var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \Comment{find rank of \enqueue\ within its block}%
 		\gc{
-        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that enqueue}}\label{findAnswerGC}
+        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswerGC}
         \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
@@
 
-\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} enqueues}
+\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th \enqueue\ in $E(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] exists and contains at least \var{i} \enqueues}
     \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCaseGC}
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\Comment{required enqueue is in \var{v}.\fld{left}}\label{leftOrRightGC} 
-        \Else \Comment{required enqueue is in \var{v}.\fld{right}}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\Comment{required \enqueue\ is in \var{v}.\fld{left}}\label{leftOrRightGC} 
+        \Else \Comment{required \enqueue\ is in \var{v}.\fld{right}}
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{\rev{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
         \EndIf
-        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
+        \State \linecomment{find \enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State \gc{use BST search in \var{v}.\fld{dir.blocks} to find minimum index $\var{b}'$ of a \typ{Block} with $\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$}\label{getChildGC}
@@
 The GC phase uses \op{SplitBlock} to determine the index $s$ of the oldest block to keep,
-calls \op{Help} to help all pending dequeues that have been propagated to the root
+calls \op{Help} to help all pending \dequeues\ that have been propagated to the root
 (to ensure that all blocks before $s$ can safely be discarded), and then
@@
 uses the \var{last} array to
-find the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
-or a null dequeue.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
-dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
+find the most recent block $B_{root}$ in the root that contains either an \enqueue\ that has been dequeued
+or a null \dequeue.  By the FIFO property of queues, all \enqueues\ in blocks before $B_{root}$ are either
+dequeued or will be dequeued by a \dequeue\ that is currently in progress.  Once those pending \dequeues\ have been
 helped to complete by line \ref{invokeHelp},
@@
 The \opemph{Help} routine is fairly straightforward:  it loops through all leaves and
-helps the dequeue that is in progress there if it has already been propagated to the root.
-The \opemph{Propagated} function is used to determine whether the dequeue has propagated to the root.
+helps the \dequeue\ that is in progress there if it has already been propagated to the root.
+The \opemph{Propagated} function is used to determine whether the \dequeue\ has propagated to the root.
 
@@
 by another process's GC phase.
-As mentioned above, if this happens to an enqueue operation,
-the enqueue can simply terminate because the fact that the block is gone means that another
-process has helped the enqueue reach the root of the ordering tree.
-Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
+As mentioned above, if this happens to an \enqueue\ operation,
+the \enqueue\ can simply terminate because the fact that the block is gone means that another
+process has helped the \enqueue\ reach the root of the ordering tree.
+Similarly, if  a \dequeue\ operation performs a failed search on a RBT, the \dequeue\ can return the value 
 written in the \fld{response} field of the 
-leaf block that represents the dequeue and terminate, since some other process
+leaf block that represents the \dequeue\ and terminate, since some other process
 will have written the \fld{response} there before discarding the needed block.

### 2024-05-31 journal/algorithm.tex
working on reviewer 2's comments

@@
 Each tree node  stores an array of \emph{blocks}, where each block represents a 
-sequence of enqueues and a sequence of dequeues.
+sequence of \enqueue\ operations and a sequence of \dequeue\ operations.
 See Figure \ref{orderingtree} for an example.
@@
 new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
-Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this three-step sequence a
+\rev{Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does} this three-step sequence a
 \op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
@@
 Now suppose $P$'s operation has been propagated all the way to the root.
-If $P$'s operation is an enqueue, it has obtained a place in the linearization ordering and can terminate.
-If $P$'s operation is a dequeue, $P$ must use information in the tree to compute the value that the
-dequeue must return.  To do this, $P$ first determines which block in the root contains its dequeue
-(since the dequeue may have been propagated to the root by some other process).
-$P$ does this by finding the dequeue's location in each node along the path from the leaf to the root.
-Then, $P$ determines whether the queue is empty when its dequeue is linearized. 
-If so, it returns \nl\ and we call it a \emph{null dequeue}.
-If not, $P$ computes the rank $r$ of its dequeue among all non-null dequeues in the linearization ordering.  (We say that the $r$th element in a sequence has \emph{rank} $r$ within that sequence.)
-$P$ then returns the value of the $r$th enqueue in the linearization.
+If $P$'s operation is an \enqueue, it has obtained a place in the linearization ordering and can terminate.
+If $P$'s operation is a \dequeue, $P$ must use information in the tree to compute the value that the
+\dequeue\ must return.  To do this, $P$ first determines which block in the root contains its 
+\dequeue\ (since the \dequeue\ may have been propagated to the root by some other process).
+$P$ does this by finding the \dequeue's location in each node along the path from the leaf to the root.
+Then, $P$ determines whether the queue is empty when its \dequeue\ is linearized. 
+If so, it returns \nl\ and we call it a \emph{null \dequeue}.
+If not, $P$ computes the rank $r$ of its \dequeue\ among all non-null \dequeues\ in the linearization ordering.  (We say that the $r$th element in a sequence has \emph{rank} $r$ within that sequence.)
+$P$ then returns the value of the $r$th \enqueue\ in the linearization.
 
@@
 \label{findinroot}
-Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's \fld{blocks} array.
+Given a \dequeue\ in a leaf that has been propagated to the root, find that operation's position in the root's \fld{blocks} array.
 \item
 \label{findrank}
-Given a dequeue's position in the root, decide if it is a null dequeue (i.e., if the queue is empty when it is linearized)
-or determine the rank $r$ of the enqueue whose value it returns.
+Given a \dequeue's position in the root, decide if it is a null \dequeue\ (i.e., if the queue is empty when it is linearized)
+or determine the rank $r$ of the \enqueue\ whose value it returns.
 \item
 \label{findenqueue}
-Find the $r$th enqueue in the linearization ordering.
+Find the $r$th \enqueue\ in the linearization ordering.
 \end{enumerate}
@@
 \caption{An example \ordering\ tree with four processes. 
-E(x) denotes an Enqueue(x) operation and D\sub{1} to D\sub{6} denote Dequeue operations.
-We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
+E(x) denotes an Enqueue(x) operation and D\sub{1} to D\sub{6} denote \dequeue\ operations.
+We show explicitly the \enqueue\ sequence and \dequeue\ sequence represented by each block in the \fld{blocks} arrays of the seven nodes.
+The leftmost element of each \fld{blocks} array is a dummy block.
 Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
@@
 As mentioned above, we are free to order operations within a block however we like.
-We order the enqueues and dequeues separately, and put the 
+We order the \enqueue\ and \dequeue\ operations separately, and put the 
 operations propagated from the left child before the operations from the right child.
-More formally, we inductively define sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
-represented by a block $B$.
-If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
-and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
+More formally, we inductively define \enqueue\ and \dequeue\ sequences $E(B)$ and $D(B)$
+for block $B$.
+If $B$ is a block in a leaf representing an \enqueue\ operation, its \enqueue\ sequence $E(B)$ is that operation
+and its \dequeue\ sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a \dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
 If $B$ is a block in an internal node $\var{v}$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
@@
 When linearizing operations within one of the root's blocks, we choose
-to put the block's enqueues before its dequeues.
+to put the block's \enqueues\ before its \dequeues.
 Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
@@
   	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
-  		\Comment{\# of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
+  		\Comment{\# of \enqueues, \dequeues\ in \fld{blocks} array up to this block (inclusive)}
   	\item \typ{int} \fld{super}
@@
 
-If a block contained an explicit representation of its sequences of enqueues and dequeues,
+If a block contained an explicit representation of its sequences of \enqueues\ and \dequeues,
 it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
@@
 %The information stored in a \block\ depends on whether it is in an internal node or a leaf.
-A block in a leaf represents a single enqueue or dequeue.  The block's \fld{element} field stores the value
-enqueued if the operation is an enqueue, or \nl\ if the operation is a dequeue.
+A block in a leaf represents a single \enqueue\ or \dequeue.  The block's \fld{element} field stores the value
+enqueued if the operation is an \enqueue, or \nl\ if the operation is a \dequeue.
 
@@
 $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total numbers of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
+that store the total numbers of \enqueues\ and \dequeues\ in $\var{v}.\fld{blocks}[1..\var{b}]$.
 We use these to search for a particular operation.
 % pinpoint the  location of an operation among the subblocks of a given block.
-For example, consider finding the $r$th enqueue $E_r$ in the linearization.
+For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
 A binary search for $r$ on the \fld{sum\sub{enq}} fields of the root's blocks 
@@
 $E_r$ explicitly.
-%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
-%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
-%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
-%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
-%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
-%to determine whether the enqueue came from the left or right child of the root.
-%Suppose the enqueue came from the right child $v_r$.
-%Then, we know that the index of the block in $v_r$ that contains the enqueue
-%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
-%We can again do a binary search within this range.
-%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
-%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
+
+\rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
 We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
@@
 Now consider task \ref{findrank}.
-To determine whether the queue is empty when a dequeue occurs,
+To determine whether the queue is empty when a \dequeue\ occurs,
 each block in the root has a \fld{size} field storing the number of elements
@@
 have been done.
-We can  determine which dequeues in a block $B_d$ in the root are null dequeues using
-$B_{d-1}.\fld{size}$, which is the size of the queue just before $B_d$'s operations, and the number of enqueues and dequeues in $B_d$.
-Moreover, the total number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
+We can  determine which \dequeues\ in a block $B_d$ in the root are null \dequeues\ using
+$B_{d-1}.\fld{size}$, which is the size of the queue just before $B_d$'s operations, and the number of \enqueues\ and \dequeues\ in $B_d$.
+Moreover, the total number of non-null \dequeues\ in blocks $B_1, \ldots, B_{d-1}$ 
 is $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
 We can use this information to determine the
-rank of a non-null dequeue in $B_d$ among all non-null dequeues in the linearization, which is the rank  (among all enqueues) of the enqueue
-whose value the dequeue should return.
+rank of a non-null \dequeue\ in $B_d$ among all non-null \dequeues\ in the linearization, which is the rank  (among all \enqueues) of the \enqueue\ whose value the \dequeue\ should return.
 
@@
 For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
-the number of enqueues and dequeues in~$B$
+the number of \enqueues\ and \dequeues\ in~$B$
 \rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
-The enqueues of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
-Then the dequeues of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
-(If subtracting this many elements would make the new size negative because there are more dequeues than there are elements, then the new size is 0.)  Thus, we have}
+The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
+Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
+(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
 \begin{eqnarray*}
@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
     \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root}.\fld{blocks}}}\label{invokeIndexDequeue}
-    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \op{Dequeue} should return}}\label{deqRest}
+    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
 \EndFunction{Dequeue}
@@
 \Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
-    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
-    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} dequeues}
+    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th \dequeue\ of $D(\var{root}.\fld{blocks}[\var{b}'])$}
+    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} \dequeues}
     \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
@@
 	    \EndIf\label{idsup3}
-	    \State \linecomment{compute index \var{i} of dequeue in superblock}
-	    \State \linecomment{\rev{add \# dequeues in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
+	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock}
+	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
 	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
-        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# dequeues in superblock from left sibling}}
+        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# \dequeues\ in superblock from left sibling}}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
@@
 
-\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$
+\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th \dequeue\ in $D(\var{root}.\fld{blocks}[\var{b}])$
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
@@
     \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
-        \State \Return \nl \hfill \Comment{queue is empty when dequeue occurs}\label{returnNull}
-    \Else \ \Comment{response is the \var{e}th enqueue in the root}
+        \State \Return \nl \hfill \Comment{queue is empty when \dequeue\ occurs}\label{returnNull}
+    \Else \ \Comment{response is the \var{e}th \enqueue\ in the root}
         \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
@@
 		\State use \rev{doubling} search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
-		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
-        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that enqueue}}\label{findAnswer}
+		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of \enqueue\ within its block}\label{FRi}
+        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswer}
     \EndIf
@@
 \end{algorithmic}
-\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of dequeues represented by block $B$.}\label{code2}}
+\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of \dequeues\ represented by block $B$.}\label{code2}}
 \end{figure*}
@@
 
-\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th enqueue in $E(\var{v}.\fld{blocks}[\var{b}])$}
-    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} enqueues}
+\Function{element}{GetEnqueue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} \Comment{returns argument of \var{i}th \enqueue\ in $E(\var{v}.\fld{blocks}[\var{b}])$}
+    \State \linecomment{Preconditions: $\var{i}\geq 1$ and \var{v}.\fld{blocks}[\var{b}] is non-\nl\ and contains at least \var{i} \enqueues}
 
@@
     \Else 
-        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
-        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
-        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
-        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
-        \Else \Comment{required enqueue is in \var{v}.\fld{right}}
+        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
+        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ \enqueues\ in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
+        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required \enqueue\ is in \var{v}.\fld{left}}
+        \Else \Comment{required \enqueue is in \var{v}.\fld{right}}
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
         \EndIf \label{endChooseDir}
-        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
+        \State \linecomment{find \enqueue\'s block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\
@@
 \end{algorithmic}
-\caption{Queue implementation's \op{GetEnqueue} routine.  \rev{Recall that $E(B)$ is the sequence of enqueues represented by block $B$}.\label{code3}}
+\caption{Queue implementation's \op{GetEnqueue} routine.  \rev{Recall that $E(B)$ is the sequence of \enqueues\ represented by block $B$}.\label{code3}}
 \end{figure*}
@@
 \rev{creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
-that the block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
+that the block represents an \op{Enqueue}.  The argument \var{e} of the \enqueue\ is stored in the block's \fld{element}
 field.
@@
 \rev{creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
-that the block represents a \op{Dequeue}.
+that the block represents a \dequeue.
 It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
 it is propagated to the root.
-Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
+Then,} the \dequeue\ calls \op{IndexDequeue} to compute
 its position in the root 
@@
 children, obtained by reading the \fld{head} index in \var{v}'s children.
-Since the \fld{sum\sub{enq}} field should store the number of enqueues in
-\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
+Since the \fld{sum\sub{enq}} field should store the number of \enqueues\ in
+\var{v}.\fld{blocks}[1..\var{i}] and these \enqueues\ come from \var{v}.\fld{left.blocks}[1..\var{B}.\eleft] and \var{v}.\fld{blocks}[1..\var{B}.\eright], line \ref{createSumEnq} sets
 \fld{sum\sub{enq}} to the sum of $\var{v}.\fld{left.blocks}[\var{B}.\eleft].\fld{sum\sub{enq}}$ and $\var{v}.\fld{right}.\fld{blocks}[\var{B}.\eright].\fld{sum\sub{enq}}$.
-Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
-subtracting  the number of enqueues  in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
+Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of \enqueues\ in the new block by
+subtracting  the number of \enqueues\ in \var{v}.\fld{blocks}[$1..\var{i}-1$] from \var{B}.\fld{sum\sub{enq}}.
 The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
@@
 
-Once a dequeue is appended to a block of the process's leaf and propagated to the root,
-the \opemph{IndexDequeue} routine finds the dequeue's location in the root.
+Once a \dequeue\ is appended to a block of the process's leaf and propagated to the root,
+the \opemph{IndexDequeue} routine finds the \dequeue's location in the root.
 More precisely, \opa{IndexDequeue}{v, b, i}
 computes the block in the root and the rank
-within that block  of the \var{i}th dequeue of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
+within that block  of the \var{i}th \dequeue\ of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
 Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may differ from the superblock's true index by one.
-The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the dequeue's 
-rank within the superblock's sequence of dequeues, using  (\ref{defSeqs}).
+The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the \dequeue's 
+rank within the superblock's sequence of \dequeues, using  (\ref{defSeqs}).
 
@@
 
-To compute the response of the $i$th \op{Dequeue} in the $b$th block
+To compute the response of the $i$th \dequeue\ in the $b$th block
 of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
 If not, line \ref{computeE} computes the rank \var{e} of the
-\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
+\op{Enqueue} whose argument is the \dequeue's response. 
 \rev{The routine next searches for}
 the index $b_e$ of the block in \var{root}.\fld{blocks} that contains 
-the \var{e}th enqueue.
-Since the enqueue is linearized before the dequeue, $b_e\leq b$.
+the \var{e}th \enqueue.
+Since the \enqueue\ is linearized before the \dequeue, $b_e\leq b$.
 To find the left end of the range \rev{to search for $b_e$, \op{FindResponse} does} a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
 \rev{Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.}
-Finally, the call to \op{GetEnqueue} \rev{at line \ref{findAnswer}} traces down through the tree to find the required enqueue in a leaf.
+Finally, the call to \op{GetEnqueue} \rev{at line \ref{findAnswer}} traces down through the tree to find the required \enqueue\ in a leaf.
 
 \opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
-$i$th enqueue in the $b$th block $B$ of \typ{Node} $\var{v}$. 
-It recursively finds the location of the enqueue in each node along the path from $v$ to a leaf, which stores the argument explicitly.
+$i$th \enqueue\ in the $b$th block $B$ of \typ{Node} $\var{v}$. 
+It recursively finds the location of the \enqueue\ in each node along the path from $v$ to a leaf, which stores the argument explicitly.
 \rev{The test at line \ref{leftOrRight} determines which child of \var{v} contains
-the enqueue by checking whether $i$ is less than or equal to the number of  
-enqueues in $B$ that came from \var{v}'s left child.
+the \enqueue\ by checking whether $i$ is less than or equal to the number of  
+\enqueues\ in $B$ that came from \var{v}'s left child.
 Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$} 
@@
 \rev{and then}
-finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
+finds the exact subblock containing the \enqueue\ using a binary search on the \fld{sum\sub{enq}}
 field \rev{of the child's \var{blocks} array.
-Line \ref{getChildIndex} computes the index of the enqueue within the block
+Line \ref{getChildIndex} computes the index of the \enqueue\ within the block
 in the child, and line \ref{getRecurse}

### 2024-05-31 journal/analysis.tex
working on reviewer 2's comments

@@
 \begin{lemma}\label{dSearchTime}
-The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
+The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th \enqueue\ takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
 \end{lemma}
@@
 %
-As shown in Lemma \ref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
-Thus, there can be at most $B_{b_e}.\size$ dequeues in 
+As shown in Lemma \ref{linearCorrect}, the \dequeue\ that calls \op{FindResponse} is in $B_b$ and is supposed to return an \enqueue\ in $B_{b_e}$.
+Thus, there can be at most $B_{b_e}.\size$ \dequeues\ in 
 $D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,

### 2024-05-31 journal/boundedspace.tex
working on reviewer 2's comments

@@
 each process writes the index of the last block in the root
-containing a null dequeue or an enqueue whose element it dequeued.
+containing a null \dequeue\ or an \enqueue\ whose element it dequeued.
 To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
-Then, it helps complete every other process's pending dequeue 
-by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
+Then, it helps complete every other process's pending \dequeue\ by 
+computing the \dequeue's response and writing it in the block in the leaf that represents the \dequeue.
 Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
@@
 by another process's GC phase.  If $op$
-is a dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
-its own leaf.  If $op$ is an enqueue, it can simply terminate.
+is a \dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
+its own leaf.  If $op$ is an \enqueue, it can simply terminate.
 

### 2024-05-31 journal/conclusion.tex
working on reviewer 2's comments

@@
 A \opa{Get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
-An \op{Index} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps if the argument is a pointer to the leaf block that contains the element~$e$.
+An \op{Index} is similar to \op{IndexDequeue} (except operating on \enqueues\ instead of \dequeues) and would take $O(\log p)$ steps if the argument is a pointer to the leaf block that contains the element~$e$.
 

### 2024-05-31 journal/correctness-GC.tex
working on reviewer 2's comments

@@
 Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
-In the configuration after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
+Immediately after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
 when $R$ executed line~\ref{refreshReadTGC}.
@@
 \item
-the value of each enqueue contained in the block has either been returned by a dequeue
-or written in the \fld{response} field of a dequeue, and
+the value of each \enqueue\ contained in the block has either been returned by a \dequeue\ or 
+written in the \fld{response} field of a \dequeue, and
 \item
-each dequeue contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
+each \dequeue\ contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
 \end{itemize}
@@
 %\here{Technically, following proof might have to be wrapped up in a big induction with the argument
-%that outputs are linearized correctly, to argue that the matchup between enqueues and dequeues
+%that outputs are linearized correctly, to argue that the matchup between \enqueues\ and \dequeues\ 
 %is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
@@
 \op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
-that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
-By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
+that contains an operation that is either a null \dequeue\ or an \enqueue\ whose value is the response for a \dequeue\ that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
+By the FIFO property of queues, the values enqueued by \enqueues\ in $\var{root}.\fld{blocks}[1..m-1]$
 are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
@@
 Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
-and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
+and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending \dequeues\ at line \ref{invokeHelp}.
 Thus, after this helping (and before step $st$),
@@
 
-\er{edited this far}
-
 \begin{lemma}\label{blocknotfound}
@@
 by the definition of finished.
-Moreover, by Invariant \ref{inv::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.
+Moreover, by Invariant \ref{inv::finished}, if the operation is a \dequeue, then its result is in its \fld{response} field.
 
-After propagation to the root, a \op{Dequeue} must access blocks that contain the dequeue,
-as well as the enqueue whose value it will return (if it is not a null dequeue).
+After propagation to the root, a \op{Dequeue} must access blocks that contain the \dequeue,
+as well as the \enqueue\ whose value it will return (if it is not a null \dequeue).
 If any of those blocks have been removed, it follows from Invariant \ref{inv::finished} that the
@@
 %Intuitively, a block's index value is marked 
-%when a dequeue passes through this block to find the enqueue
-%whose argument it should return, or when a dequeue that returns null passes through the block.
+%when a \dequeue\ passes through this block to find the \enqueue
+%whose argument it should return, or when a \dequeue\ that returns null passes through the block.
 %Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
-%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
+%contains either an \enqueue whose value is being dequeued or a \nl\ \dequeue.
 %If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.
@@
 Then, $A$ performs a GC phase.
-Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
+Let \rev{$t$ be the time} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
 That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
@@
 Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
-$\ell$ is the last block added to the root's RBT before $C$.
+$\ell$ is the last block added to the root's RBT before $t$.
 
@@
 
-The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
-If the value enqueued by any enqueue in $L_2$  is not still in the queue
-at the end of $L_2$, then the dequeue in $L_2$ that dequeued that value must still be in progress
-at $C$; otherwise the process that performed that dequeue would have set its \var{last} entry
-to the index of the root block that contains the enqueue, which is greater than $m$, before $C$, contradicting
-the fact that all values in the $last$ array at $C$ are less than or equal to $m$.
-So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
-Thus, there are at most $q_{max}+p$ enqueues in $L_2$.
+The number of \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
+If the value enqueued by any \enqueue\ in $L_2$  is not still in the queue
+at the end of $L_2$, then the \dequeue\ in $L_2$ that dequeued that value must still be in progress
+at time $t$; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
+to the index of the root block that contains the \enqueue, which is greater than $m$, before $t$, contradicting
+the fact that all values in the $last$ array at $t$ are less than or equal to $m$.
+So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$.
+Thus, there are at most $q_{max}+p$ \enqueues\ in $L_2$.
 
-If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
+If a \dequeue\ in $L_2$ returns a non-null value in the sequential execution $L$,
 then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
-Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
-Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
-at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
-to a value greater than the index of the root block that contains the dequeue prior to $C$, 
+Thus, there are at most $q_{max}+(q_{max}+p)$ \dequeues\ in $L_2$ that return non-null values.
+Any \dequeue\ in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
+at $t$; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
+to a value greater than the index of the root block that contains the \dequeue\ prior to $t$, 
 contradicting the definition of $m$.
-So, there are at most $p$ null dequeues in $L_2$.
-Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.
+So, there are at most $p$ null \dequeues\ in $L_2$.
+Thus, there are at most $2q_{max} + 2p$ \dequeues\ in $L_2$.
 
@@
 $T'\assign\op{Split}(T,s)$.
-Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$, by Lemma \ref{lem::appendExactlyOnce}.
+Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma \ref{lem::appendExactlyOnce}.
 Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.

### 2024-05-31 journal/correctness.tex
working on reviewer 2's comments

@@
 If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
-For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
+For line \ref{enqNew}, $B$ represents a single \enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
 Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
 $B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
-The proof for line~\ref{deqNew}, where $B$ represents a single dequeue, is similar.
+The proof for line~\ref{deqNew}, where $B$ represents a single \dequeue, is similar.
 
-Now suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
+Now, suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
 subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
 and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
-Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
-$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in 
+Thus, the \enqueues\ in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
+$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and %those in 
 $E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
-By the hypothesis, the total number of these enqueues is 
+By the hypothesis, the total number of these \enqueues\ is 
 $\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
@@
 \begin{eqnarray*}
-\var{num\sub{enq}}
+&&\var{num\sub{enq}}\\
 &=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
@@
 \end{eqnarray*}
-It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
+It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
 $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
 $\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
-Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
+Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
 Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
@@
 By Lemma \ref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
-There is at most one such block in each node, by Corollary \ref{cor::noDuplicates}.
+There is at most one such block in each node, by Corollary~\ref{cor::noDuplicates}.
 \end{proof}
@@
 If $\var{v}.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(\var{v}.\fld{blocks}[b])|$, 
- then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
+ then \op{IndexDequeue}($\var{v}, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th \dequeue\ in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th \dequeue\ of $D(\var{root}.\fld{blocks}[b'])$.
 \end{lemma}
@@
 
-Next, the position of the required dequeue in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
+Next, the position of the required \dequeue\ in $D(B')$ (as defined by Equation (\ref{defSeqs})) is computed in 
 lines \ref{computeISuperStart}--\ref{computeISuperEnd}. 
-We first add the number of dequeues in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
+We first add the number of \dequeues\ in the subblocks of $B'$ in $\var{v}$ that precede $B$ on line~\ref{computeISuperStart}.
 If $\var{v}$ is the right child of its parent, then all of the subblocks of $B'$ from $\var{v}$'s left sibling
-also precede the required dequeue, so we add the number of dequeues in those subblocks in line \ref{considerLeftBeforeRight}.
+also precede the required \dequeue, so we add the number of \dequeues\ in those subblocks in line \ref{considerLeftBeforeRight}.
 
@@
 Thus, all preconditions of the recursive call are met.
-By the induction hypothesis, the recursive call returns the location of the required dequeue in the root.\end{proof}
+By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}
 
 \begin{lemma}\label{lem::get}
-If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th enqueue in $E(\var{v}.\fld{blocks}[b])$.
+If $1\leq i\leq |E(\var{v}.\fld{blocks}[b])|$ then \op{getEnqueue}($\var{v},b,i$) returns the argument of the $i$th \enqueue\ in $E(\var{v}.\fld{blocks}[b])$.
 \end{lemma}
@@
 If $\var{v}$ is a leaf, the hypothesis implies that $i=1$ and the block $\var{v}.\fld{blocks}[b]$ represents 
-an enqueue whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
-\op{GetEnqueue} returns the argument of this enqueue at line \ref{getBaseCase}.
+an \enqueue\ whose argument is stored in $\var{v}.\fld{blocks}[b].\fld{element}$.
+\op{GetEnqueue} returns the argument of this \enqueue\ at line \ref{getBaseCase}.
 
@@
 By Equation (\ref{defSeqs}),
-$E(B)$ is obtained by concatenating the enqueue sequences of the direct subblocks
+$E(B)$ is obtained by concatenating the \enqueue\ sequences of the direct subblocks
 of $B$, which are listed in (\ref{defsubblock}).
 By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
-of enqueues in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
-Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required enqueue operation.
-Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required enqueue within the portion $E'$ of $E(B)$ that comes from that child.
-Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required enqueue.
+of \enqueues\ in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
+Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required \enqueue\ operation.
+Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required \enqueue\ within the portion $E'$ of $E(B)$ that comes from that child.
+Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required \enqueue.
 By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
-enqueues in $E'$ before the enqueues of block $B'$, so
-the value $i'$ computed on line \ref{getChildIndex} is the position of the required enqueue within $E(B')$.
+\enqueues\ in $E'$ before the \enqueues of block $B'$, so
+the value $i'$ computed on line \ref{getChildIndex} is the position of the required \enqueue\ within $E(B')$.
 Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
@@
 (\ref{linearization}) is a legal permutation of a subset of the operations in 
-the execution, i.e., that it includes all operations that terminate and 
+the concurrent execution, i.e., that it includes all operations that terminate and 
 if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
-We also show each completed dequeue returns the same result as it would in the sequential execution~$L$.
+We also show each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would in the sequential execution~$L$ \rev{(i.e., the sequential execution of the operations in the order given by $L$)}.
 
@@
 By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
-are the number of enqueues and dequeues contained in $B$.
-Hence, the size of the queue after the operations of $B$ are performed (with enqueues before dequeues as specified by~$L$)
+are the number of \enqueues\ and \dequeues\ contained in $B$.
+Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
 is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
@@
 \begin{lemma}\label{linearCorrect}
-Each terminating dequeue returns the response it would in the sequential execution~$L$.
+Each terminating \dequeue\ returns the response it would in the sequential execution~$L$.
 \end{lemma}
 \begin{proof}
-If a dequeue $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
+If a \dequeue\ $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
 By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
-returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th dequeue in the block 
+returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th \dequeue\ in the block 
 $B=\var{root}.\fld{blocks}[b]$.
@@
 By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
-is the number of enqueues in $B$.
-Since the enqueues in block $B$ precede the dequeues,
-the queue is empty when the $i$th dequeue of $B$ occurs if 
+is the number of \enqueues\ in $B$.
+Since the \enqueues\ in block $B$ precede the \dequeues,
+the queue is empty when the $i$th \dequeue\ of $B$ occurs if 
 $\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
@@
 in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
-the number of non-\nl\ dequeues in that prefix of $L$.
-Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ dequeues in $L$.
-Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th enqueue in $L$.
-By Invariant \ref{inv::sum}, this enqueue is the $i_e$th enqueue 
-in $E(\var{root}.\fld{blocks}[b_e])$, where
+the number of non-\nl\ \dequeues\ in that prefix of $L$.
+Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
+Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
+By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
 $b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
-By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
+By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required \enqueue.
 \end{proof}

### 2024-05-31 journal/introduction.tex
working on reviewer 2's comments

@@
 To dequeue or enqueue an element, the front or back pointer is updated by a 
-compare-and-swap (CAS) instruction.
+compare-and-swap (CAS) instruction\footnote{\rev{A CAS(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the CAS leaves \var{X} unchanged and returns \fa.}}.
 If this CAS fails, the operation must retry.
@@
 and quickly access any  operation in a~batch.
-Preliminary versions of this work appeared in~\cite{Nad22,NR23}.
+Preliminary \rev{versions of this work appeared in~\cite{Nad22,NR23}}.
 %\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2024-05-31 journal/response.txt
working on reviewer 2's comments

@@
 > - ln 845-848, right. I got confused here. Lemma 4 is only about the indexes, isn't it?
-> 
+
+Since expression (3.3) defines subblocks in terms of indexes, we can apply Lemma 4.  Some more explanation has been added to the proof.
+
 > - ln 1084-1088, right. Here do you mean that L is a valid sequential execution of the queue.
@@
 Why the first algorithm cannot be modified to remove bocks, let us say using some sort of epoch-based memory reclamation scheme?
-> 
+
+As mentioned above, we could not figure out a simpler way to do this.
+
 > - ln 1279-1283, left. Why? A whole path might need to be copied, isn’t it?
-> 
+
+Since all accesses are through the root, one cannot reach a node without traversing all nodes along the path from the root to that node.
+Copying those nodes then only increases the running time by a constant factor.
+(The sentence has been reworded to avoid confusion.)
+
 > - ln 1442. Why not Help(), in accordance to Dequeue(), for example?
-> 
+
+We removed the parentheses from Dequeue for consistency.
+
 > - Sec 7. Your algorithms are adaptive on the size of the queue but not on the number of processes. I wonder if an interesting future direction would be to make your algorithms fully adaptive.
->
+
+Yes, this was mentioned in sentence 3 of Sec 7.

### 2024-05-31 journal/boundedspace.tex
working on reviewer 2's comments

@@
 The RBT reachable from the new copy of the root is the result of applying the RBT operation.
-This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
+This \rev{increases the running time of any routine designed for a (sequential) RBT only by a constant factor, assuming that nodes are accessed only by following pointers from the root (as is the case for all RBT operations we require)}.
 Once a process has performed an update to the RBT representing the blocks of a node 

### 2024-05-31 journal/correctness.tex
working on reviewer 2's comments

@@
 If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
-If $B$ and $B'$ are in the same node, it follows from Lemma \ref{lem::headProgress} that their direct subblocks are disjoint.
-Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
+\rev{Suppose $B$ and $B'$ are in the same node $v$.
+By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
+so the intervals of direct subblocks of different blocks in $v$ in $v$'s left child defined by (\ref{defsubblock}) are non-overlapping.
+Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
+In all cases, the direct subblocks of $B$ and $B'$} (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
 \end{proof}

### 2024-05-31 journal/analysis.tex
minor edit

@@
 
-We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.
+We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue \rev{(i.e., the number of elements in the queue)}.
 

### 2024-06-03 journal/response.txt
minor edit
finished dealing with reviewer 2's comments

@@
 
-This seems to be a strange effect of the double-spacing option for the draft.
+This seems to be a strange effect of the double-spacing option for the draft.  It should look okay in the final version.
 
@@
 >
-> Then, the paper presents a modification of the first algorithm that uses bounded memory, more specifically polynomial on the number of processes and the size of the queue in the linearization, and with polylogarithmic amortized step complexity.
-As one would naturally expect, this version adds a sort of garbage collector mechanism that removes blocks of memory that are not needed anymore.
-I found this part of the paper harder to understand.
-I did not understand the need of using red-black trees, which I think complicates everything.
-I have no reason to think the implementation is wrong, all looks convincing to me, but I wonder if there is a way to simplify the presentation.
+> Then, the paper presents a modification of the first algorithm that
+> uses bounded memory, more specifically polynomial on the number of
+> processes and the size of the queue in the linearization, and with
+> polylogarithmic amortized step complexity. As one would naturally
+> expect, this version adds a sort of garbage collector mechanism that
+> removes blocks of memory that are not needed anymore. I found this
+> part of the paper harder to understand. I did not understand the need
+> of using red-black trees, which I think complicates everything. I
+> have no reason to think the implementation is wrong, all looks
+> convincing to me, but I wonder if there is a way to simplify the
+> presentation.
 
@@
 
-We tried to avoid the need for too many definitions and preferred to introduce them as needed instead of in a preliminaries section, since we are using fairly standard terminology.
-The terms mentioned above were explained in the first paragraph of the paper.
-A more formal explanation of what must be proved to establish linearizability is given in the first paragraph of Section 4.4.
-Similarly, wait-free is defined when it is first used.
-For completeness, we added a definition of CAS in a footnote on page 1.
+A new section on the model of computation has been added to include
+these definitions.
 
@@
 > - ln 1084-1088, right. Here do you mean that L is a valid sequential execution of the queue.
-> 
+
+Technically, L is just a sequence of operations (without the
+corresponding responses), rather than a sequential execution.
+The wording has been clarified.
+
 > - Lemma 15. After reading the proof I understood that legal linearization ordering only  means respecting real-time order. This is a concrete example why I think a preliminaries section might help.
-> 
+
+The changes to the paragraph before Lemma 15 should help to clarify this.
+
 > - ln 1192-1993, right. Please formalize the notion of time in an execution,
-> 
+
+When revising, we were more careful about the use of "time".  (For
+example, we now avoid using it when we mean step complexity.)  Since
+the model section now defines an execution as a sequence of steps
+(where the steps of processes are interleaved), some references to
+time are now replaced by "step".  We have left some references to time
+in informal descriptions, hoping that they will reinforce the intution
+behind the arguments.
+
 > - Theorem 22. What exactly is the size of the queue?
-> 
+
+We added a clarification of what we mean just before the statement
+of this theorem.
+
 > - Sec 6.

### 2024-06-03 journal/algorithm.tex
working on reviewer 2's comments
finished dealing with reviewer 2's comments

@@
 
-If a block contained an explicit representation of its sequences of \enqueues\ and \dequeues,
-it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
+If a block contained an explicit \rev{listing} of its sequences of \enqueues\ and \dequeues,
+it would take $\Omega(p)$ \rev{steps} to construct a block, which would be too slow for task \ref{construct}.
 Instead, the block stores an implicit representation of the sequences.
@@
 operation  up the path to the root, and determine the operation's location in a root block
-in $O(\log p)$ time.
+in $O(\log p)$ \rev{steps}.
 
@@
 Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
-we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
+we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ \rev{steps}.
 A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
@@
 
-\Function{Object}{Dequeue()}{} 
+\Function{Object}{Dequeue}{} 
     \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,

### 2024-06-03 journal/analysis.tex
finished dealing with reviewer 2's comments

--- 2024-05-31 journal/analysis.tex
@@
 
-The following lemma helps bound the time for \op{GetEnqueue}.
+The following lemma helps bound the \rev{number of steps \op{GetEnqueue} takes}.
 \begin{lemma}\label{blockSize}
@@
 We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.
+\rev{The bounds refer to the size of the queue when certain operations are linearized,
+meaning the number of elements that would be in a queue after the operations in the prefix
+of the linearization order up to that operation were performed sequentially.}.
 

### 2024-06-03 journal/correctness-GC.tex
finished dealing with reviewer 2's comments

@@
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
-Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
-reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
+Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between \rev{each \op{Refresh}'s 
+read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS}} at line \ref{casGC}.
 Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
@@
 If \op{SplitBlock}($x$) returns the minimum block of $x$'s RBT at line \ref{substitute},
-then the claim follows from the assumption that Invariant \ref{inv::finished} holds at all times before $st$.
+then the claim follows from the assumption that Invariant \ref{inv::finished} holds before $st$.
 
@@
 Then, $A$ performs a GC phase.
-Let \rev{$t$ be the time} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
+Let \rev{$s$ be the step} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
 That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
@@
 Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
-$\ell$ is the last block added to the root's RBT before $t$.
+$\ell$ is the last block added to the root's RBT before $s$.
 
@@
 at the end of $L_2$, then the \dequeue\ in $L_2$ that dequeued that value must still be in progress
-at time $t$; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
-to the index of the root block that contains the \enqueue, which is greater than $m$, before $t$, contradicting
-the fact that all values in the $last$ array at $t$ are less than or equal to $m$.
-So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$.
-Thus, there are at most $q_{max}+p$ \enqueues\ in $L_2$.
+when $s$ occurs; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
+to the index of the root block that contains the \enqueue, which is greater than $m$, before $s$, contradicting
+the fact that all values in the $last$ array are less than or equal to $m$ when $s$ occurs.
+So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of~$L_2$.
+Thus, there are at most $q_{max}+p$ \enqueues\ in~$L_2$.
 
@@
 Any \dequeue\ in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
-at $t$; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
-to a value greater than the index of the root block that contains the \dequeue\ prior to $t$, 
+when $s$ occurs; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
+to a value greater than the index of the root block that contains the \dequeue\ prior to $s$, 
 contradicting the definition of $m$.
@@
 some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
-Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
+Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will always be at least $r\cdot G$ 
 after this CAS.

### 2024-06-05 journal/response.txt

@@
 We were concerned that count might sound like a count of the number of enqueue or dequeue operations in a single block.
-We instead used sum to remind the reader that the fields of Block i store the prefix sum of the number of enqueue or dequeue operations in in Blocks 1..i (as mentioned when the fields are introduced at line 504 [check line number].
+We instead used sum to remind the reader that the fields of Block i store the prefix sum of the number of enqueue or dequeue operations in in Blocks 1..i (as mentioned when the fields are introduced on page 11).
 
@@
 
-Done.  Quite a few new comments have been added throughout the pseudocode.
+Done.  We have added many comments throughout the pseudocode.
 
@@
 
-Some explanation has been added in the text (line 526) [*** check line #].
+Some explanation has been added in the text (see magenta text on p.12)
 
@@
 
-Done.  Some text has been added to the text description of FindResponse to explain that the doubling search first finds the left end of the range and then performs a binary search.
+Done.  We added to the text description of FindResponse (p.17) to explain that the doubling search first finds the left end of the range and then performs a binary search.
 
@@
 We hoped that presenting the simpler, space-unbounded version first would make it easier to understand the space-bounded version.
-We did our best to explain the transformation from the arrays to the RBTs in the first page of Section 6.
+We did our best to explain the transformation from the arrays to the RBTs in the first page of Section 7.
 
@@
 Although our space-bounded algorithm unlinks objects that are no longer needed from our data structure, we consider the problem of reclaiming that memory to be an orthogonal issue.
-There is a rather large body of literature on memory reclamation for concurrent data structures, so we have added (at line 110 of the introduction) a reference to a very recent paper that includes a large bibliography of work in this area, to help the reader find relevant literature on it.
+There is a rather large body of literature on memory reclamation for concurrent data structures, so we have added (at line 110, left column, of the introduction) a reference to a very recent paper that includes a large bibliography of work in this area, to help the reader find relevant literature on it.
 
@@
 
-A new section on the model of computation has been added to include
+A new section 3 on the model of computation has been added to include
 these definitions.
@@
 
-Yes, this was mentioned in sentence 3 of Sec 7.
+Yes, this was mentioned in sentence 3 of the conclusion (now Section 8).

### 2024-06-05 journal/algorithm-GC.tex
working on reviewer 2's comments

@@
 We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
-can be done in constant time, without affecting the time of other RBT operations.
+can be done in \rev{$O(1)$ steps}, without affecting the step complexity of other RBT operations.
 Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.
@@
 
-\Function{Object}{Dequeue()}{} 
+\Function{Object}{Dequeue}{} 
     \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}

### 2024-06-05 journal/boundedspace.tex

--- 2024-05-31 journal/boundedspace.tex
@@
 The RBT reachable from the new copy of the root is the result of applying the RBT operation.
-This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
+This \rev{increases the number of steps taken by any routine designed for a (sequential) RBT only by a constant factor, assuming that nodes are accessed only by following pointers from the root (as is the case for all RBT operations we require)}.
 Once a process has performed an update to the RBT representing the blocks of a node 
@@
 Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
-also keep the running time of our operations polylogarithmic \rev{in $p$ and $q$}.
+also keep the \rev{step complexity} of our operations polylogarithmic \rev{in $p$ and $q$}.
 Blocks should be kept if they contain operations still in progress.
@@
 
-To maintain good amortized time, we periodically do a garbage collection (GC) phase.
+To maintain good amortized \rev{step complexity}, we periodically do a garbage collection (GC) phase.
 If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
@@
 Fortunately, there is an  RBT \op{Split} operation that can remove
-these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+these obsolete blocks from an RBT using \rev{a logarithmic number of steps} \cite[Sec.~4.2]{Tar83}.
 

### 2024-06-05 journal/correctness.tex

@@
 read $b$ from $\var{v}.\head$ at line \ref{readHead}.
-At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
+At that \rev{read}, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
 Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
@@
 
-We show that the linearization ordering $L$ defined in Equation
+We \rev{next show in Lemma \ref{linearSat}} that the linearization ordering $L$ defined in Equation
 (\ref{linearization}) is a legal permutation of a subset of the operations in 
@@
 if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
-We also show each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would in the sequential execution~$L$ \rev{(i.e., the sequential execution of the operations in the order given by $L$)}.
+\rev{Then, we show in Lemma \ref{linearCorrect}} that each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would \rev{if the operations were performed sequentially in the order given by $L$}.
 
@@
 
-Next, we show each operation returns the same response as it would in the sequential execution~$L$.
+Next, we show each operation returns the same response as it would \rev{if the operations were performed sequentially in the order $L$}.
 
 \begin{lemma}\label{linearCorrect}
-Each terminating \dequeue\ returns the response it would in the sequential execution~$L$.
+Each terminating \dequeue\ returns the response it would \rev{if the operations
+were performed sequentially in the order given by the linearization $L$}.
 \end{lemma}
@@
 So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
-execution $L$.
+execution \rev{defined by} $L$.
 Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
-in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
+in the sequential execution \rev{defined by} $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
 the number of non-\nl\ \dequeues\ in that prefix of $L$.
 Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
-Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
+Thus, in the sequential execution \rev{defined by} $L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
 By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
@@
 
+\rev{We remark that, if execution $\alpha$ is a prefix of another execution $\alpha'$,
+then the linearization associated with $\alpha$ is a prefix of the linearization
+assocated with $\alpha'$, so our implementation is also strongly linearizable,
+according to the definition in \cite{GHW11}.}
+
 %\here{Should there be a lemma somewhere in this section that explicitly shows that each

### 2024-06-05 journal/preliminaries.tex

% !TEX root =  queue.tex

\section{Model}

\rev{
We use a  standard model of an asynchronous, shared-memory system.
A collection of $p$ processes can communicate by accessing
shared memory locations using read, write and compare-and-swap (CAS) instructions.
A \op{CAS}(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the \op{CAS} leaves \var{X} unchanged and returns \fa.
We also assume processes can allocate a block of shared memory to represent an object.

We are concerned with implementing a first-in, first-out queue in such a system,
which requires an algorithm that a process can execute to perform
an enqueue or dequeue operation.
We treat individual memory instructions as atomic, so an execution
of such an implementation can be thought of as a sequence of steps by processes
where in each step a process performs at most one shared-memory instruction 
(and receives the response of that instruction).
The subsequence of steps performed by a process should follow that process's
algorithm for performing a sequence of enqueue and dequeue operations.
The order in which the steps of processes are interleaved is arbitrary, and the implementation must be correct for all
possible interleavings.

We use the standard correctness requirement of linearizability \cite{HW90} for 
our queue.  This means that for every execution of our queue implementation,
there should be a sequential ordering (called the linearization) of 
a subset of the operations, including all completed operations,
satisfying two properties.  First, if one operation terminates before another
begins, they must appear in the same order in the linearization.
Second, each completed operation must return the same response that it would
if the operations were done sequentially in their linearization order.
(In fact, our implementation also satisfies the stronger requirement of being
strongly linearizable \cite{GHW11}.)
The termination condition that our queue satisfies is wait-freedom:
no operation by a process can take infinitely many steps of the process to complete.

The step complexity of an operation is the number of steps it performs.
The amortized step complexity is the total number of steps by all processes
divided by the number of operations they perform.
More formally, to show that the amortized step complexity is $O(T_e)$ per enqueue
and $O(T_d)$ per dequeue, we must show that in every execution, 
if $n_e$ enqueues and $n_d$ dequeues are invoked, the total number
of steps taken by all processes is $O(n_e\cdot T_e + n_d\cdot T_d)$.
}

### 2024-06-05 journal/algorithm.tex
some edits from final proofreading

@@
 we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
-then $P$'s operation has been propagated to $\var{v}$ by some other process, so $P$ can continue 
+then $P$'s operation has been propagated to $\var{v}$ by some other process.
+\rev{Thus, after performing a double \op{Refresh},} $P$ can continue 
 onwards towards the~root.
@@
 \caption{An example \ordering\ tree with four processes. 
-E(x) denotes an Enqueue(x) operation and D\sub{1} to D\sub{6} denote \dequeue\ operations.
+E(x) denotes an \enqueue(x) operation and D\sub{1} to D\sub{6} denote \dequeue\ operations.
 We show explicitly the \enqueue\ sequence and \dequeue\ sequence represented by each block in the \fld{blocks} arrays of the seven nodes.
@@
 Thus, we linearize operations according to the block they belong to in the root's array.
-We can choose how to order operations within the same block, since they must be concurrent.
+We can freely choose how to order operations within the same block, since they must be concurrent.
 
@@
 
-Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child.  
+Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, \rev{respectively}.  
 Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
@@
 $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total numbers of \enqueues\ and \dequeues\ in $\var{v}.\fld{blocks}[1..\var{b}]$.
+that store the total numbers of \enqueues\ and \dequeues, \rev{respectively}, in $\var{v}.\fld{blocks}[1..\var{b}]$.
 We use these to search for a particular operation.
@@
 \rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
-We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
+We shall show \rev{in Section \ref{sec-analysis}} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
 and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,

### 2024-06-05 journal/preliminaries.tex
some edits from final proofreading

@@
 A collection of $p$ processes can communicate by accessing
-shared memory locations using read, write and compare-and-swap (CAS) instructions.
+shared memory locations using read, write and compare-and-swap (\op{CAS}) instructions.
 A \op{CAS}(\var{X}, \var{old}, \var{new}) instruction updates a variable \var{X} to the value \var{new} and returns \tr\ if \var{X}'s value is \var{old}; otherwise the \op{CAS} leaves \var{X} unchanged and returns \fa.
@@
 
-We are concerned with implementing a first-in, first-out queue in such a system,
-which requires an algorithm that a process can execute to perform
+We are concerned with implementing a first-in, first-out queue in such a system.
+An implementation provides algorithms that a process can execute to perform
 an enqueue or dequeue operation.
@@
 The subsequence of steps performed by a process should follow that process's
-algorithm for performing a sequence of enqueue and dequeue operations.
+algorithms for performing the process's desired sequence of enqueue and dequeue operations.
 The order in which the steps of processes are interleaved is arbitrary, and the implementation must be correct for all
@@
 divided by the number of operations they perform.
-More formally, to show that the amortized step complexity is $O(T_e)$ per enqueue
-and $O(T_d)$ per dequeue, we must show that in every execution, 
-if $n_e$ enqueues and $n_d$ dequeues are invoked, the total number
-of steps taken by all processes is $O(n_e\cdot T_e + n_d\cdot T_d)$.
+More formally, to show that the amortized step complexity is $O(T)$ per operation, we must show that in every execution, 
+if $m$ enqueue and dequeue operations are invoked, the total number
+of steps taken by all processes is $O(m\cdot T)$.
 }

### 2024-06-06 journal/algorithm.tex
some more edits from final proofreading

@@
 (If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
-\begin{eqnarray*}
-B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
-			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).
-\end{eqnarray*}
+\begin{eqnarray}
+B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\nonumber\\ 
+			&& \hspace*{10mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).\label{size-eqn}
+\end{eqnarray}
 
@@
 set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
-We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of \rev{$B$'s superblock}
+We shall show that this happens soon enough \rev{to ensure} that $B.\fld{super}$ can differ from the true index of \rev{$B$'s superblock}
 by at most 1.
@@
 		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}.\fld{blocks} and propagate to root}}\label{enqAppend}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf.}\fld{blocks} and propagate to root}}\label{enqAppend}
 \EndFunction{Enqueue}
@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root}.\fld{blocks}}}\label{invokeIndexDequeue}
+    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf.}\fld{blocks} and propagate to root}}\label{deqAppend}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root.}\fld{blocks}}}\label{invokeIndexDequeue}
     \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first fails, try again}}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first \op{Refresh} fails, try again}}
         \State \Call{Refresh}{\var{v}} \label{secondRefresh}
@@
     \State \var{num\sub{enq}} \assign\ $\var{B}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnq}
-    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
+    \State \var{num\sub{deq}} \assign\ $\var{B}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$\label{computeNumDeq}
     \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$} \Return \nl\ \Comment{no blocks need to be propagated to \var{v}}\label{testEmpty} 
@@
         \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required \enqueue\ is in \var{v}.\fld{left}}
-        \Else \Comment{required \enqueue is in \var{v}.\fld{right}}
+        \Else \Comment{required \enqueue\ is in \var{v}.\fld{right}}
             \State \fld{dir} \assign\ \fld{right}
@@
         \EndIf \label{endChooseDir}
-        \State \linecomment{find \enqueue\'s block in \var{v}.\fld{dir.blocks} and its rank within block}
+        \State \linecomment{find \enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
         \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\
@@
 The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
-\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
+\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ at line \ref{testEmpty} to indicate
  there is no need to install~it.
 Otherwise,} if \var{B}
-is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
+is going to be installed in the root, line \ref{computeLength} \rev{uses Equation (\ref{size-eqn}) to compute $B$'s} \fld{size} field, which
 represents the number of elements in the queue after the operations in the block are performed.
@@
 the \var{e}th \enqueue.
-Since the \enqueue\ is linearized before the \dequeue, $b_e\leq b$.
+Since the \enqueue\ is linearized before the \dequeue, \linebreak $b_e\leq b$.
 To find the left end of the range \rev{to search for $b_e$, \op{FindResponse} does} a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.

### 2024-06-06 journal/correctness.tex
some more edits from final proofreading

@@
 After proving some basic properties in Section \ref{sec::basicProperties},
-we show in Section \ref{sec::propagating} that a double refresh at each node
+we show in Section \ref{sec::propagating} that a double \op{Refresh} at each node
 suffices to propagate an operation to the root.
@@
 If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
-Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
+Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two \rev{different} blocks in nodes at depth $d$.
 Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
@@
 $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
-$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
+$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..$ $B.\eright]$.
 Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
@@
 By Observation \ref{obs::headInc}, $h_2>h_1$.
-By Lemma \ref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
+By Lemma~\ref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line~\ref{readHead}.
 Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
@@
 \begin{proof}
-We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
+We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see line \ref{indexBaseCase}).
 Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
@@
 Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
-Since $B$ has been propagated to the root, so has its superblock $B'$.
+Since $B$ has been propagated to the root, its superblock $B'$ \rev{has also been propagated to the root}.
 Thus, all preconditions of the recursive call are met.
@@
 By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
-\enqueues\ in $E'$ before the \enqueues of block $B'$, so
+\enqueues\ in $E'$ before the \enqueues\ of block $B'$, so
 the value $i'$ computed on line \ref{getChildIndex} is the position of the required \enqueue\ within $E(B')$.
@@
 By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
-are the number of \enqueues\ and \dequeues\ contained in $B$.
+\rev{computed on line \ref{computeNumEnq} and \ref{computeNumDeq}}
+are the number of \enqueues\ and \dequeues\ contained in $B$, respectively.
 Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
-is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
+is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, \rev{as described in Equation (\ref{size-eqn})}.
 \end{proof}

### 2024-06-08 journal/analysis.tex
some edits from final proofreading
some final edits from proofreading

@@
 \section{Analysis}
+\label{sec-analysis}
+
 We now analyze the number of steps and the number of \op{CAS} instructions performed by operations.
@@
 Thus, there can be at most $B_{b_e}.\size$ \dequeues\ in 
-$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
+$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution defined by $L$,
 all elements enqueued before the end of
@@
 Let $op$ be the earliest operation by $p$ contained in $B$.
-When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
+When $op$ terminates, $op$ is contained in $B$, by Lemma \ref{lem::appendExactlyOnce}.
 $B$ cannot contain any later operations by $p$, since $B$ is created before

### 2024-06-08 journal/conclusion.tex
some final edits from proofreading

@@
 It would be interesting to close the gap that remains between our queue, which takes $O(\log^2 p + \log q)$ steps per operation,
-and Jayanti, Tarjan and Boix-Adser\`{a}'s $\Omega(\log p)$ lower bound \cite{JTB19}.
+and Jayanti, Tarjan and Boix-Adser\`{a}'s $\Omega(\log p)$ lower bound~\cite{JTB19}.
 For the more relaxed bag data structure the gap is larger between the  $\Omega(\min(c,\log\log p))$ lower bound \cite{DBLP:conf/opodis/AttiyaF17} and our upper bound of $O(\log^2 p + \log q)$.
@@
 remains an open question.
+\rev{The technique of cooperatively propagating information up a tree with a double refresh
+at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.}

### 2024-06-08 journal/correctness-GC.tex
some final edits from proofreading

@@
 Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
-The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT \rev{at line \ref{splitGC}}
+and then add $B$ to it \rev{at line \ref{GCend} or \ref{noGC-insert}}.
 So the claim is satisfied.
@@
 We say a node \var{v} \emph{contains} an operation
-if some block containing the operation has previously appeared in the RBT \var{v}.\fld{block}
+if some block containing the operation has previously appeared in the RBT \var{v}.\fld{blocks}
 (even if the block has been removed from the RBT by a subsequent \op{Split} during garbage collection).
@@
 The \op{CreateBlock} set $B.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
-By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
+By Corollary \ref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than~$i$.
 By the definition of subblocks, some block in \var{v} contains $B_\ell$ as a subblock
@@
 contain no operations.
-By Corollary \ref{blockNotEmpty}, each block contains at least one operation, so
+By Corollary~\ref{blockNotEmpty}, each block contains at least one operation, so
 these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft\geq i$.
@@
 This allows us to show that a double \op{Refresh} propagates operations up the tree, 
-as in Lemma \ref{lem::doubleRefresh}.
+as in Lemma~\ref{lem::doubleRefresh}.
 
@@
 This is because the first unfinished block may still be traversed by an operation in the future,
-and when examining that block the operation may need information from the preceding block.  For example,
+and when examining that block the operation may need information from the \emph{preceding block}.  For example,
 when \op{FindResponse} is called on a block with index $b$, line \ref{FRnumenqGC} looks up the block with
@@
 %during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
-then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
+then \rev{the \dequeue} has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
 If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
@@
 $T'\assign\op{Split}(T,s)$.
-Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma \ref{lem::appendExactlyOnce}.
+Since $T$ was read before $t$, any operation in $T$ is either in progress at $t$ or has been propagated to the root before $t$, by Lemma~\ref{lem::appendExactlyOnce}.
 Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
@@
 \begin{proof}
-Each update to a node's RBT adds at most one block to it, increasing its maximum \fld{index} by 1.
+Each update to a node's RBT adds at most one block to it, increasing its maximum \fld{index} by~1.
 Thus, there are at most $G$ updates since the last time its maximum \fld{index} was a multiple of $G$.
@@
 
-Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
+Now, we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
 and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.

### 2024-06-08 journal/preliminaries.tex
fixed defn of wait-free

@@
 The termination condition that our queue satisfies is wait-freedom:
-no operation by a process can take infinitely many steps of the process to complete.
+each operation by a process is guaranteed to terminate within a finite number of the process's own steps.
 

### 2024-07-10 journal/related.tex
Removed repeated words
updating a bit of related work

@@
 Thus, it is lock-free but not wait-free and has an amortized step complexity of $\Theta(p)$ per operation.
-
-%\Hossein{Maybe we can say worst-case.
-I see that you have used amortized step complexity consistently everywhere but maybe it is fair to say that these papers were not designed to improve the worst-case amortized step complexity.}
-%\Eric{Amortized step complexity is defined in first parag of paper to be worst case over all executions of the average number of steps per operation, so I think the worst-case aspect of it is clear.
-Also, if we just said ``worst-case'' without ``amortized'' it would be confused with worst-case of an individual operation, which is infinite in this case.
-By the way, saying that previous implementations have $\Omega(p)$ step complexity, even in the amortized sense, makes the new contribution even more of a contrast.}
 
@@
 to mix enqueues and dequeues.
-%\here{Is this worth saying:
-%They use some properties of the queue size before and after a batch, similar to a part of our work.}
 In the worst case, where operations require their response right away,
@@
 assuming that words can store $\Omega(p \log p)$ bits. 
-However, if more reasonably-sized $O(\log p)$-bit words words are used, the construction would take $\Omega(p\log p)$ steps per operation.
+However, if more reasonably-sized $O(\log p)$-bit words are used, the construction would take $\Omega(p\log p)$ steps per operation.
 There are two obstacles to making this construction more efficient:
@@
 (It depends on the total number of enqueues in the execution.)
+
+\HERE{discuss \cite{ACH18} who seem to have showed that strongly linearizable (define) queue is impossible using weak primitives; can also mention that HW22 observed that the MS queue is not strongly linearizable but can easily be made so}
 

### 2024-07-16 journal/algorithm.tex
cleaning up some .tex files (removing comments, \rev{})

@@
 
-%\Eric{Reworded following parag after Hossein's comments of Jan 4}
 To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ performs the following
@@
 new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
-\rev{Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does} this three-step sequence a
+Using terminology from \cite{DBLP:conf/fsttcs/JayantiP05}, we call the procedure that does this three-step sequence a
 \op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
@@
 then $P$'s operation has been propagated to $\var{v}$ by some other process.
-\rev{Thus, after performing a double \op{Refresh},} $P$ can continue 
+Thus, after performing a double \op{Refresh}, $P$ can continue 
 onwards towards the~root.
@@
 
-If a block contained an explicit \rev{listing} of its sequences of \enqueues\ and \dequeues,
-it would take $\Omega(p)$ \rev{steps} to construct a block, which would be too slow for task \ref{construct}.
+If a block contained an explicit listing of its sequences of \enqueues\ and \dequeues,
+it would take $\Omega(p)$ steps to construct a block, which would be too slow for task \ref{construct}.
 Instead, the block stores an implicit representation of the sequences.
@@
 
-%The information stored in a \block\ depends on whether it is in an internal node or a leaf.
 A block in a leaf represents a single \enqueue\ or \dequeue.  The block's \fld{element} field stores the value
@@
 
-Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, \rev{respectively}.  
+Each block in an internal \node\ $\var{v}$ has fields \eleft\ and \eright\ that store the indices of the block's last direct subblock in $\var{v}$'s left and right child, respectively.  
 Thus, the direct subblocks of $\var{v}.\fld{blocks}[b]$ are
@@
 Blocks also store some prefix sums:
-%the block in 
 $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
-that store the total numbers of \enqueues\ and \dequeues, \rev{respectively}, in $\var{v}.\fld{blocks}[1..\var{b}]$.
+that store the total numbers of \enqueues\ and \dequeues, respectively, in $\var{v}.\fld{blocks}[1..\var{b}]$.
 We use these to search for a particular operation.
-% pinpoint the  location of an operation among the subblocks of a given block.
 For example, consider finding the $r$th \enqueue\ $E_r$ in the linearization.
@@
 
-\rev{Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.}
-We shall show \rev{in Section \ref{sec-analysis}} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
+Recall that $p$ is the number of processes and $q$ is the number of elements in the queue.
+We shall show in Section \ref{sec-analysis} that the binary search in the root can be done in $O(\log p + \log q)$ steps,
 and the binary search within each other \node\ along the path to a leaf takes $O(\log p)$ steps,
 for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
-%All information needed for this search process can be derived from the 
-%\eleft, \eright\ and \fld{sum\sub{enq}} fields.
 
@@
 the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
-%We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
 This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
@@
 operation  up the path to the root, and determine the operation's location in a root block
-in $O(\log p)$ \rev{steps}.
+in $O(\log p)$ steps.
 
@@
 Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
-we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ \rev{steps}.
+we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ steps.
 A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
@@
 the number of \enqueues\ and \dequeues\ in~$B$
-\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
+as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
 The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
 Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
-(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
+(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have
 \begin{eqnarray}
@@
 Soon after $B$ is installed,
-%a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by 
 some process will
 set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
-We shall show that this happens soon enough \rev{to ensure} that $B.\fld{super}$ can differ from the true index of \rev{$B$'s superblock}
-by at most 1.
+We shall show that this happens soon enough to ensure that $B.\fld{super}$ can differ from the true index of 
+$B$'s superblock by at most 1.
 %After $B$ is installed 
@@
 We now discuss the queue implementation in more detail.  Pseudocode is provided in Figures \ref{code1}--\ref{code3}.
-%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
-%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
-%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
-%We use \var{num\sub{deq}} similarly.
-%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}
 
@@
 		\fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}$}\label{enqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf.}\fld{blocks} and propagate to root}}\label{enqAppend}
+    \State \Call{Append}{\var{B}}\Comment{append \op{Enqueue} to \var{leaf.}\fld{blocks} and propagate to root}\label{enqAppend}
 \EndFunction{Enqueue}
@@
 	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf.}\fld{blocks} and propagate to root}}\label{deqAppend}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root.}\fld{blocks}}}\label{invokeIndexDequeue}
-    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
+    \State \Call{Append}{\var{B}}\Comment{append \dequeue\ to \var{leaf.}\fld{blocks} and propagate to root}\label{deqAppend}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{find where \op{Deqeueue} is in \var{root.}\fld{blocks}}\label{invokeIndexDequeue}
+    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{find response \dequeue\ should return}\label{deqRest}
 \EndFunction{Dequeue}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  \rev{if first \op{Refresh} fails, try again}}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefresh}  \hfill \Comment{double refresh:  if first \op{Refresh} fails, try again}
         \State \Call{Refresh}{\var{v}} \label{secondRefresh}
@@
     \State \var{h} \assign\ \var{v}.\head \label{readHead}
-    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{\rev{help advance $v$'s children's \var{head} indices, if necessary}}\label{startHelpChild1}
+    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \Comment{help advance $v$'s children's \var{head} indices, if necessary}\label{startHelpChild1}
         \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
@@
     \EndFor \label{endHelpChild1}
-    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlock}
-    \If{$\var{B} = \nl$} \Return{\tr}\Comment{\rev{if no new operations to propagate to $v$, do nothing}}\label{addOP} 
+    \State \var{B} \assign\ \Call{CreateBlock}{\var{v, h}}\Comment{create new block to append to \var{v}.\fld{blocks}}\label{invokeCreateBlock}
+    \If{$\var{B} = \nl$} \Return{\tr}\Comment{if no new operations to propagate to $v$, do nothing}\label{addOP} 
 	\Else
-	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{\rev{try to add $B$ to \var{v}.\fld{blocks}}}\label{cas}
-    	\State \Call{Advance}{\var{v, h}} \Comment{\rev{advance $v.head$ from $h$ to $h+1$}}\label{advance}
-    	\State \Return{ \var{result}}\Comment{\rev{return true iff \op{CAS} succeeded}}
+	    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{B}}\Comment{try to add $B$ to \var{v}.\fld{blocks}}\label{cas}
+    	\State \Call{Advance}{\var{v, h}} \Comment{advance $v.head$ from $h$ to $h+1$}\label{advance}
+    	\State \Return{ \var{result}}\Comment{return true iff \op{CAS} succeeded}
 	\EndIf
@@
 	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{\rev{actual index of superblock is either \var{sup} or $\var{sup}+1$}}\label{idsup1}
+    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{actual index of superblock is either \var{sup} or $\var{sup}+1$}\label{idsup1}
 	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
@@
 	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock}
-	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
+	    \State \linecomment{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
 	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
-        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# \dequeues\ in superblock from left sibling}}
+        \If{$\fld{dir} = \fld{right}$} \Comment{add \# \dequeues\ in superblock from left sibling}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
@@
         \EndIf \label{computeISuperEnd}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}} \Comment{\rev{recurse up the tree}}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}} \Comment{recurse up the tree}
     \EndIf
@@
     \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
-    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
-    %  \State \linecomment{contains at least \var{i} dequeues}
     \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
@@
 			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
-		\State use \rev{doubling} search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
+		\State use doubling search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
 		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of \enqueue\ within its block}\label{FRi}
-        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswer}
+        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{find argument of that \enqueue}\label{findAnswer}
     \EndIf
@@
 \end{algorithmic}
-\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of \dequeues\ represented by block $B$.}\label{code2}}
+\caption{Queue implementation's helper routines. Recall that $D(B)$ is the sequence of \dequeues\ represented by block $B$.\label{code2}}
 \end{figure*}
@@
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}
         \EndIf \label{endChooseDir}
@@
 \end{algorithmic}
-\caption{Queue implementation's \op{GetEnqueue} routine.  \rev{Recall that $E(B)$ is the sequence of \enqueues\ represented by block $B$}.\label{code3}}
+\caption{Queue implementation's \op{GetEnqueue} routine.  Recall that $E(B)$ is the sequence of \enqueues\ represented by block $B$.\label{code3}}
 \end{figure*}
@@
 
-%Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
-%As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
-%We shall show {$B$.\fld{super}}  differs
-%from the index of $B'$ by at most~1.
-
 To compute the response of the $i$th \dequeue\ in the $b$th block
@@
 \op{Enqueue} whose argument is the \dequeue's response. 
-\rev{The routine next searches for}
+The routine next searches for
 the index $b_e$ of the block in \var{root}.\fld{blocks} that contains 
@@
 Since the \enqueue\ is linearized before the \dequeue, \linebreak $b_e\leq b$.
-To find the left end of the range \rev{to search for $b_e$, \op{FindResponse} does} a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
-\rev{Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.}
-Finally, the call to \op{GetEnqueue} \rev{at line \ref{findAnswer}} traces down through the tree to find the required \enqueue\ in a leaf.
+To find the left end of the range to search for $b_e$, \op{FindResponse} does a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
+Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.
+Finally, the call to \op{GetEnqueue} at line \ref{findAnswer} traces down through the tree to find the required \enqueue\ in a leaf.
 
@@
 It recursively finds the location of the \enqueue\ in each node along the path from $v$ to a leaf, which stores the argument explicitly.
-\rev{The test at line \ref{leftOrRight} determines which child of \var{v} contains
+The test at line \ref{leftOrRight} determines which child of \var{v} contains
 the \enqueue\ by checking whether $i$ is less than or equal to the number of  
 \enqueues\ in $B$ that came from \var{v}'s left child.
-Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$} 
+Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$
 using information stored
 in $B$ and the block that precedes $B$ in $v$,
-\rev{and then}
+and then
 finds the exact subblock containing the \enqueue\ using a binary search on the \fld{sum\sub{enq}}
-field \rev{of the child's \var{blocks} array.
+field of the child's \var{blocks} array.
 Line \ref{getChildIndex} computes the index of the \enqueue\ within the block
 in the child, and line \ref{getRecurse}
-proceeds recursively down the tree.}
+proceeds recursively down the tree.

### 2024-07-16 journal/analysis.tex
cleaning up some .tex files (removing comments, \rev{})

@@
 If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
-%
+
 As shown in Lemma \ref{linearCorrect}, the \dequeue\ that calls \op{FindResponse} is in $B_b$ and is supposed to return an \enqueue\ in $B_{b_e}$.
@@
 
-The following lemma helps bound the \rev{number of steps \op{GetEnqueue} takes}.
+The following lemma helps bound the number of steps \op{GetEnqueue} takes.
 \begin{lemma}\label{blockSize}
@@
 We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.
-\rev{The bounds refer to the size of the queue when certain operations are linearized,
+The bounds refer to the size of the queue when certain operations are linearized,
 meaning the number of elements that would be in a queue after the operations in the prefix
-of the linearization order up to that operation were performed sequentially.}.
+of the linearization order up to that operation were performed sequentially.
 
@@
 at line \ref{findAnswer}.
-\rev{Since the doubling search must search a finite portion of the root's
-\var{blocks} array, it terminates within a finite number of steps.}
+Since the doubling search must search a finite portion of the root's
+\var{blocks} array, it terminates within a finite number of steps.
 Moreover, by Lemma \ref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,

### 2024-07-16 journal/boundedspace.tex
cleaning up some .tex files (removing comments, \rev{})

--- 2024-05-31 journal/boundedspace.tex
@@
 Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
-polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic \rev{in $p$ and $q$}. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
+polynomial in $p$ and $q$, while ensuring the (amortized) step complexity is still polylogarithmic 
+in $p$ and $q$. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
 that stores the blocks.
@@
 The RBT reachable from the new copy of the root is the result of applying the RBT operation.
-This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
+This increases the number of steps taken by any routine designed for a (sequential) RBT only by a constant factor, assuming that nodes are accessed only by following pointers from the root (as is the case for all RBT operations we require).
 Once a process has performed an update to the RBT representing the blocks of a node 
@@
 Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
-also keep the running time of our operations polylogarithmic \rev{in $p$ and $q$}.
+also keep the step complexity of our operations polylogarithmic in $p$ and $q$.
 Blocks should be kept if they contain operations still in progress.
@@
 
-To maintain good amortized time, we periodically do a garbage collection (GC) phase.
+To maintain good amortized step complexity, we periodically do a garbage collection (GC) phase.
 If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
@@
 Fortunately, there is an  RBT \op{Split} operation that can remove
-these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
+these obsolete blocks from an RBT using a logarithmic number of steps \cite[Sec.~4.2]{Tar83}.
 

### 2024-07-16 journal/correctness.tex
cleaning up some .tex files (removing comments, \rev{})

@@
 A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
-when the block is created.
-% at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
+when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
 Moreover, only a \op{CAS} at line \ref{setSuper1}  modifies  \fld{super}  
@@
 read $b$ from $\var{v}.\head$ at line \ref{readHead}.
-At that \rev{read}, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
+At that read, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
 Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
@@
 Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
-%of these calls to \op{CreateBlock} 
+of these calls to \op{CreateBlock} 
 satisfy the claim.
@@
 \begin{lemma} \label{lem::subblocksDistinct}
-If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
+If $B$ and $B'$ are two different blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
 \end{lemma}
@@
 If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
-Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two \rev{different} blocks in nodes at depth $d$.
+Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two different blocks in nodes at depth $d$.
 Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
 If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
-\rev{Suppose $B$ and $B'$ are in the same node $v$.
+Suppose $B$ and $B'$ are in the same node $v$.
 By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
@@
 Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
-In all cases, the direct subblocks of $B$ and $B'$} (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
+In all cases, the direct subblocks of $B$ and $B'$ (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
 \end{proof}
@@
 
-%\begin{definition}
-%$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \emph{established} in node $n$ 
-%if it is in an established block of $n$. $EST^t_n$ is the set of established operations in node $n$ at time $t$.
-%\end{definition}
-%
-%Now we want to say that blocks of a node grow over time.
-%\begin{observation}\label{lem::blocksOrder}
-%  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of 
-%$ops(n.blocks)$ at time $t^\prime$.
-%\end{observation}
-%\begin{proof}
-%Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
-%\end{proof}
-
-% \begin{corollary}\label{lem::establishedOrder}
-%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
-% \end{corollary}
-% \begin{proof}
-% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
-% \end{proof}
-
 The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
-on lines \ref{enqNew}, \ref{deqNew} %, \ref{createSumEnq} 
+on lines \ref{enqNew}, \ref{deqNew}
 and \ref{createSumDeq} follows easily
@@
 
-Now, suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
+Now, suppose $\var{v}$ is an internal node. By the definition of direct subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the direct
 subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
@@
 We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
-%\here{move this defn earlier and just recall it here?}
 Since blocks are permanently added to nodes, if $\var{v}$ contains $op$ at some time, $v$ contains $op$ at all later times too.
@@
 
-%\Eric{I think this case was missing from the proof in the thesis}
 Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
@@
 We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.
-We do this by showing that \rev{the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
-\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.}
-%On one hand, $B.super$ is written before $B_s$ is installed:
-%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
-%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
-%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
-%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
-%parent soon after.
+We do this by showing that the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
+\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.
 
@@
 in $\var{v}.\fld{parent.blocks}[s]$.
-By the definition of \rev{direct} subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
+By the definition of direct subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft}
 obtains a value greater than $b$.
@@
 So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
-By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ \rev{is assigned a non-\nl\ value before $B$'s superblock is installed}.  Let $R_{s-1}$ be the call to
+By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ 
+is assigned a non-\nl\ value before $B$'s superblock is installed.  Let $R_{s-1}$ be the call to
 $\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
@@
 
-%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
-%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
-%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
 To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
@@
 For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
-%\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}
 
@@
 Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
-Since $B$ has been propagated to the root, its superblock $B'$ \rev{has also been propagated to the root}.
+Since $B$ has been propagated to the root, its superblock $B'$ has also been propagated to the root.
 Thus, all preconditions of the recursive call are met.
@@
 
-We \rev{next show in Lemma \ref{linearSat}} that the linearization ordering $L$ defined in Equation
+We next show in Lemma \ref{linearSat} that the linearization ordering $L$ defined in Equation
 (\ref{linearization}) is a legal permutation of a subset of the operations in 
@@
 if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.
-\rev{Then, we show in Lemma \ref{linearCorrect}} that each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would \rev{if the operations were performed sequentially in the order given by $L$}.
+Then, we show in Lemma \ref{linearCorrect} that each completed \dequeue\ returns the same result in the concurrent execution as it would if the operations were performed sequentially in the order given by $L$.
 
@@
 so it appears in $L$.
-\rev{Suppose an operation} $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
-is propagated to the root before $op_2$ begins, \rev{and therefore before $op_2$ is propagated to the root.
+Suppose an operation $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
+is propagated to the root before $op_2$ begins, and therefore before $op_2$ is propagated to the root.
 Thus, $op_1$ appears in an earlier block of \var{root}.\fld{blocks} than $op_2$.
-So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.}
+So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.
 \end{proof}
@@
 By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
-\rev{computed on line \ref{computeNumEnq} and \ref{computeNumDeq}}
+computed on line \ref{computeNumEnq} and \ref{computeNumDeq}
 are the number of \enqueues\ and \dequeues\ contained in $B$, respectively.
 Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
-is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, \rev{as described in Equation (\ref{size-eqn})}.
+is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, as described in Equation (\ref{size-eqn}).
 \end{proof}
 
-Next, we show each operation returns the same response as it would \rev{if the operations were performed sequentially in the order $L$}.
+Next, we show each operation returns the same response as it would if the operations were performed sequentially in the order $L$.
 
 \begin{lemma}\label{linearCorrect}
-Each terminating \dequeue\ returns the response it would \rev{if the operations
-were performed sequentially in the order given by the linearization $L$}.
+Each terminating \dequeue\ returns the response it would if the operations
+were performed sequentially in the order given by the linearization $L$.
 \end{lemma}
@@
 So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
-execution \rev{defined by} $L$.
+execution defined by $L$.
 Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
-in the sequential execution \rev{defined by} $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
+in the sequential execution defined by $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
 the number of non-\nl\ \dequeues\ in that prefix of $L$.
 Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
-Thus, in the sequential execution \rev{defined by} $L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
+Thus, in the sequential execution defined by $L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
 By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
@@
 
-\rev{We remark that, if execution $\alpha$ is a prefix of another execution $\alpha'$,
+We remark that, if execution $\alpha$ is a prefix of another execution $\alpha'$,
 then the linearization associated with $\alpha$ is a prefix of the linearization
 assocated with $\alpha'$, so our implementation is also strongly linearizable,
-according to the definition in \cite{GHW11}.}
-
-%\here{Should there be a lemma somewhere in this section that explicitly shows that each
-%object we dereference is non-null?}
+according to the definition in \cite{GHW11}.

### 2024-07-16 journal/introduction.tex
finished dealing with reviewer 2's comments
some edits from final proofreading
cleaning up some .tex files (removing comments, \rev{})

--- 2024-05-30 journal/introduction.tex
@@
 %packet processing \cite{DPDK}, 
-synchronization \cite{KAE23}, %\here{is this a good citation for this?}
+synchronization \cite{KAE23}, 
  and sharing resources or tasks.
@@
 holds only if $c$ is $O(\log\log p)$ so it should be stated more precisely as
-%At first glance, this would seem to settle the question of the step complexity of lock-free queues.
-%However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
-%Thus, the lower bound could be stated more precisely as 
 an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.
@@
 while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
-Moreover, \rev{if we count only CAS instructions,} each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
+Moreover, if we count only CAS instructions, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
 lock-free queues use 
-%an unbounded number of CAS instructions in the worst case and 
 $\Omega(p)$, even in an amortized sense.
@@
 garbage collector, such as the highly optimized one that Java provides.
-\rev{See \cite{SBM24} and the references therein for a variety of approaches to
-safe memory reclamation.}
+See \cite{SBM24} and the references therein for a variety of approaches to
+safe memory reclamation.
 
@@
 and quickly access any  operation in a~batch.
+
 Preliminary versions of this work appeared in~\cite{Nad22,NR23}.

### 2024-07-16 journal/related.tex
cleaning up some .tex files (removing comments, \rev{})

@@
 to a concurrent dequeue when the queue is empty.
-% or when concurrent operations can be linearized to empty the queue.
 However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
@@
 presented an optimistic  queue implementation
-%MS-queue uses
-%two \nf{CAS}es to do an enqueue: one to change the tail to the new
-%node and another one to change the next pointer of the previous node
-%to the new node. 
 that uses a doubly-linked list to reduce the number of
@@
 Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,PMG16,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
-%Similar to the front and back pointers of list-based queues, indices
 Dequeues and enqueues update
@@
 with $O(\sqrt{p})$ step complexity using some strong non-standard synchronization primitives
-called half-increment and half-max, which can be viewed as %particular kinds of
+called half-increment and half-max, which can be viewed as 
 double-word read-modify-write operations.

### 2024-07-16 journal/algorithm-GC.tex
some final edits from proofreading
cleaning tex files (removing comments etc)

@@
 We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
-can be done in \rev{$O(1)$ steps}, without affecting the step complexity of other RBT operations.
+can be done in $O(1)$ steps, without affecting the step complexity of other RBT operations.
 Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.
@@
 		\gc{\fld{index}\ \assign $h$}}\label{enqNewGC}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Enqueue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}
+    \State \Call{Append}{\var{B}}\Comment{append \op{Enqueue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}
 \EndFunction{Enqueue}
@@
 	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
-    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}%
-    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{\rev{compute response \op{Dequeue} should return}}
+    \State \Call{Append}{\var{B}}\Comment{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}%
+    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{compute response \op{Dequeue} should return}
 \EndFunction{Dequeue}
@@
 \Function{Object}{CompleteDeq}{\typ{Node} \var{leaf}, \typ{int} $h$} \Comment{finish propagated \dequeue\ in $\var{leaf}.\fld{blocks}[h]$}
-    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{\rev{find where \op{Dequeue} is in \var{root}.\fld{blocks}}}
-    \State \Return $\Call{FindResponse}{b, i}$\Comment{\rev{find response \op{Dequeue} should return}}
+    \State $\langle \var{b}, \var{i}\rangle$ \assign\ $\Call{IndexDequeue}{\var{leaf}, h, 1}$\Comment{find where \op{Dequeue} is in \var{root}.\fld{blocks}}
+    \State \Return $\Call{FindResponse}{b, i}$\Comment{find response \op{Dequeue} should return}
 \EndFunction{CompleteDeq}
@@
 \Function{void}{Propagate}{\typ{Node} \var{v}} \Comment{propagate blocks from \var{v}'s children to root}
-    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \Comment{double refresh: \rev{if first fails, try again}}
+    \If{{\keywordfont not} \Call{Refresh}{\var{v}}} \label{firstRefreshGC}  \hfill \Comment{double refresh: if first \op{Refresh} fails, try again}
         \State \Call{Refresh}{\var{v}} \label{secondRefreshGC}
@@
     }
-    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$\Comment{\rev{create new block to append to \var{v}.\fld{blocks}}}\label{invokeCreateBlockGC}
-    \If{\var{B = \nl}} \Return{\tr}\Comment{\rev{if no new operations to propagate to \var{v}, do nothing}}\label{addOPGC}
+    \State $\var{B}$ \assign\ $\Call{CreateBlock}{\var{v, h}}$\Comment{create new block to append to \var{v}.\fld{blocks}}\label{invokeCreateBlockGC}
+    \If{\var{B = \nl}} \Return{\tr}\Comment{if no new operations to propagate to \var{v}, do nothing}\label{addOPGC}
     \Else  
     	\gc{
-	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\Comment{\rev{$T'$ is a copy of $T$ with $B$ added (and GC if needed)}}\label{refreshABGC}
-		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\Comment{\rev{try installing $T'$ as \var{v}'s \fld{blocks} tree}}\label{casGC}%
+	    \State $T'$ \assign\ $\Call{AddBlock}{v, T, \var{B}}$\Comment{$T'$ is a copy of $T$ with $B$ added (and GC if needed)}\label{refreshABGC}
+		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\Comment{try installing $T'$ as \var{v}'s \fld{blocks} tree}\label{casGC}%
 		}
@@
     \If{\var{B}.\fld{index} is a multiple of $G$} \Comment{do garbage collection}\label{GCstart}
-        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\Comment{\rev{find lowest index of block to keep}}\label{callSplitBlock}
-        \State \Call{Help}{}\Comment{\rev{help complete pending \op{Dequeues}}}\label{invokeHelp}
+        \State $s$ \assign\ \Call{SplitBlock}{\var{v}}.\var{index}\Comment{find lowest index of block to keep}\label{callSplitBlock}
+        \State \Call{Help}{}\Comment{help complete pending \op{Dequeues}}\label{invokeHelp}
         \State $T'$ \assign\ \Call{Split}{$T,s$} \Comment{\op{Split} removes blocks with $\fld{index} < s$}\label{splitGC}
-        \State \Return \Call{Insert}{$T',B$}\Comment{\rev{add $B$ to the tree}}\label{GCend}
-	\Else\ \Return \Call{Insert}{$T, B$}\Comment{\rev{if GC not needed, just add $B$ to the tree}}\label{noGC-insert}
+        \State \Return \Call{Insert}{$T',B$}\Comment{add $B$ to the tree}\label{GCend}
+	\Else\ \Return \Call{Insert}{$T, B$}\Comment{if GC not needed, just add $B$ to the tree}\label{noGC-insert}
     \EndIf
@@
 \gc{
-\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{\rev{find lowest index of block GC should keep in \var{v}}}
+\Function{Block}{SplitBlock}{\typ{Node} v} \Comment{find lowest index of block GC should keep in \var{v}}
 	\If{$\var{v}=\var{root}$} \Comment{return block before the last block containing a}
@@
 		\State $B$ \assign\ $\var{root}.\fld{blocks}[m-1]$\label{SBlookup1}
-	\Else \Comment{\rev{return last sublock of first block kept in parent}}
+	\Else \Comment{return last sublock of first block kept in parent}
 		\State $B_p$ \assign\ $\Call{SplitBlock}{\var{v}.\fld{parent}}$\label{SBrecurse}
@@
             \State \linecomment{operation is a propagated \dequeue}
-            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{\rev{compute response and store it in leaf block}}
+            \State $\var{B}.\fld{response}$ \assign\ $\Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ \Comment{compute response and store it in leaf block}
         \EndIf
@@
 		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
-		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa\ \Comment{\rev{\var{v}.\fld{blocks}[\var{b}] has not propagated to \var{v}.\fld{parent}}}
-		\Else \Comment{\rev{check if superblock of \var{v}.\fld{blocks}[\var{b}] propagated to root}}
+		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa\ \Comment{\var{v}.\fld{blocks}[\var{b}] has not propagated to \var{v}.\fld{parent}}
+		\Else \Comment{check if superblock of \var{v}.\fld{blocks}[\var{b}] propagated to root}
 			\State $B_p$ \assign\ minimum index block in $T$ with $\edir \geq b$\label{searchsuper3}
@@
 	    \State \linecomment{compute index \var{i} of \dequeue\ in superblock $B_p$}
-	    \State \linecomment{\rev{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}}
+	    \State \linecomment{add \# \dequeues\ in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $B_p$}
 	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
 	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
-        \If{$\fld{dir} = \fld{right}$}\Comment{\rev{add \# \dequeues\ in $B_p$ from left sibling}}
+        \If{$\fld{dir} = \fld{right}$}\Comment{add \# \dequeues\ in $B_p$ from left sibling}
         	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
@@
         \EndIf \label{computeISuperEndGC}
-        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}} \Comment{\rev{recurse up the tree}}
+        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}} \Comment{recurse up the tree}
     \EndIf
@@
 		\gc{
-        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{\rev{find argument of that \enqueue}}\label{findAnswerGC}
+        \State \var{res} \assign\ \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\Comment{find argument of that \enqueue}\label{findAnswerGC}
         \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]$ \assign\ $b_e$
@@
             \State \fld{dir} \assign\ \fld{right}
-            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{\rev{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}}
+            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$\Comment{deduct \# \enqueues\ in $v.\fld{blocks}[b]$ that came from $v.\fld{left}$}
         \EndIf

### 2024-07-16 journal/algorithm.tex
cleaning tex files (removing comments etc)

@@
 An \opemph{Enqueue}(\var{e}) 
-\rev{creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
+creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
 that the block represents an \op{Enqueue}.  The argument \var{e} of the \enqueue\ is stored in the block's \fld{element}
@@
 The call to \op{Append}, described below, on line \ref{enqAppend}
-appends the \block\ to the process's leaf and ensures it is propagated to the root.}
+appends the \block\ to the process's leaf and ensures it is propagated to the root.
 
 A \opemph{Dequeue} also 
-\rev{creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
+creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
 that the block represents a \dequeue.
@@
 it is propagated to the root.
-Then,} the \dequeue\ calls \op{IndexDequeue} to compute
+Then, the \dequeue\ calls \op{IndexDequeue} to compute
 its position in the root 
@@
 The leaf's \fld{head} field stores the first empty slot in the leaf's \fld{blocks} array,
-so the \op{Append} writes \var{B} there \rev{at line \ref{appendLeaf}} and then increments \fld{head} \rev{at line \ref{appendEnd}}.
+so the \op{Append} writes \var{B} there at line \ref{appendLeaf} and then increments \fld{head} at line \ref{appendEnd}.
 Since \op{Append} writes only to the process's own leaf, there cannot be concurrent updates to a leaf.
-\op{Append} then calls \op{Propagate} \rev{at line \ref{appendProp}} to ensure the operation represented by \var{B} is propagated to the root.
+\op{Append} then calls \op{Propagate} at line \ref{appendProp} to ensure the operation represented by \var{B} is propagated to the root.
 
@@
 Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
-\rev{Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
-for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.}
+Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
+for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.
 Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
@@
 The values of \var{B}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
-\rev{If the new block contains no operations, \op{CreateBlock} returns \nl\ at line \ref{testEmpty} to indicate
+If the new block contains no operations, \op{CreateBlock} returns \nl\ at line \ref{testEmpty} to indicate
  there is no need to install~it.
-Otherwise,} if \var{B}
-is going to be installed in the root, line \ref{computeLength} \rev{uses Equation (\ref{size-eqn}) to compute $B$'s} \fld{size} field, which
+Otherwise, if \var{B}
+is going to be installed in the root, line \ref{computeLength} uses Equation (\ref{size-eqn}) to compute $B$'s \fld{size} field, which
 represents the number of elements in the queue after the operations in the block are performed.

### 2024-07-16 journal/analysis.tex
cleaning tex files (removing comments etc)

@@
 
-\begin{mytheorem}\label{enqDeqTime}
-\rev{The queue implementation is wait-free and}
+\begin{theorem}\label{enqDeqTime}
+The queue implementation is wait-free and
 each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
@@
 $q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
-\end{mytheorem}
+\end{theorem}
 \begin{proof}

### 2024-07-16 journal/conclusion.tex
cleaning tex files (removing comments etc)

@@
 remains an open question.
-\rev{The technique of cooperatively propagating information up a tree with a double refresh
-at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.}
+The technique of cooperatively propagating information up a tree with a double refresh
+at each node was also used recently to obtain lock-free augmented search trees \cite{FR24}.

### 2024-07-16 journal/correctness-GC.tex
cleaning tex files (removing comments etc)

@@
 Line \ref{createIndexGC} sets $\var{B}.\var{index} = \max(I)+1$.
-The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT \rev{at line \ref{splitGC}}
-and then add $B$ to it \rev{at line \ref{GCend} or \ref{noGC-insert}}.
+The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT at line \ref{splitGC}
+and then add $B$ to it at line \ref{GCend} or \ref{noGC-insert}.
 So the claim is satisfied.
@@
 still makes sense by proving the analogue of Lemma \ref{lem::headProgress}.
-
-%\here{If next obs is only used once, can fold it into the proof that needs it}
-%\begin{observation}
-%\label{edirGC} 
-%If $B$ is a block in the RBT of an internal node \var{v}, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
-%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $\var{v}.\var{dir}$.
-%\end{observation}
-%\begin{proof}
-%The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
-%on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
-%\end{proof}
 
@@
 So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
-Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between \rev{each \op{Refresh}'s 
-read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS}} at line \ref{casGC}.
+Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between each \op{Refresh}'s 
+read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS} at line \ref{casGC}.
 Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
@@
 \end{observation}
-
-%\here{Technically, following proof might have to be wrapped up in a big induction with the argument
-%that outputs are linearized correctly, to argue that the matchup between \enqueues\ and \dequeues\ 
-%is the same as it would be in the linearization ordering.  (Check whether this is really needed.)
-%Or alternatively, we could add an assumption that if every operation agrees with the linearization
-%so far, then the invariant is preserved by each step.
-%Then, when proving the linearization is correct, everything will work out.
-%But just leave it as is for the submission, since this is a subtle point}
-% NO, I DON'T THINK THIS IS NECESSARY
 
@@
 %during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
-then \rev{the \dequeue} has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
+then the \dequeue\ has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
 If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
@@
 Part of the proof of the following lemma is similar to the proof of Lemma \ref{dSearchTime}.
-%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
-%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
-%Intuitively, a block's index value is marked 
-%when a \dequeue\ passes through this block to find the \enqueue
-%whose argument it should return, or when a \dequeue\ that returns null passes through the block.
-%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
-%contains either an \enqueue whose value is being dequeued or a \nl\ \dequeue.
-%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.
 
@@
 Then, $A$ performs a GC phase.
-Let \rev{$s$ be the step} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
+Let $s$ be the step when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
 That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
@@
 The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
-%\here{At first, I thought the following analysis could be tightened
-%by using the fact that operations only appear along one path in tree
-%so doing a bound on total size of all nodes on one level might work yield better bound.
-%But I don't think so, because we could have each process go in order to dequeue
-%everything and then enqueue q new items (where q>>p).  Then sets of enqueue operations
-%stored in each node at one level of tree would be disjoint (because of the way
-%GC is timed differently at each node).
 
-\begin{mytheorem}\label{spaceBound}
+\begin{theorem}\label{spaceBound}
 The queue data structure uses a maximum of $O(pq_{max}+p^3\log p)$ words of memory at any time.
-\end{mytheorem}
+\end{theorem}
 \begin{proof}
@@
 
-\begin{mytheorem}
+\begin{theorem}
 The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
-\end{mytheorem}
+\end{theorem}
 \begin{proof}

### 2024-07-16 journal/related.tex
cleaning tex files (removing comments etc)

@@
 \paragraph{Universal Constructions.}
-\Eric{Rewrote this paragraph on Oct 25}
 One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
@@
 
-\paragraph{Other Primitives.}
+\paragraph{Queues Implemented from Other Primitives.}
 Li \cite{Li01} implemented a non-blocking linearizable queue using only the weak primitives
@@
 
-\HERE{discuss \cite{ACH18} who seem to have showed that strongly linearizable (define) queue is impossible using weak primitives; can also mention that HW22 observed that the MS queue is not strongly linearizable but can easily be made so}
+%Could discuss \cite{ACH18} who seem to have showed that strongly linearizable 
+%(define) queue is impossible using weak primitives; can also mention that HW22 
+%observed that the MS queue is not strongly linearizable but can easily be made so}
 
